### 2015-05-29 LSP/lsp-seriesinfo/cmle-info.tex
initial commit paper

%% -*- coding:utf-8 -*-
{\large Computational Models of Language Evolution}

\bigskip

Editors: Luc Steels, Remi van Trijp

\bigskip

In this series:

\begin{enumerate}
\item Steels, Luc. The Talking Heads experiment: origins of words and meanings.
\item Vogt, Paul. How mobile robots can self-organize a vocabulary.
\item Bleys, Joris. Language strategies for the domain of colour.
\item van Trijp, Remi. The evolution of case grammar.
\item Spranger, Michael. The evolution of grounded spatial language.
\end{enumerate}

% 
% \vfill
% 
% \gdef\lsISSN{2363-5568}%
% \hfill ISSN: \lsISSN
%

### 2015-05-29 LSP/lsp-seriesinfo/silp-info.tex
initial commit paper

%% -*- coding:utf-8 -*-

{\raggedleft\IfFileExists{./LSP/lsp-seriesinfo/labphon-logo.pdf}{\includegraphics[width=5cm]{./LSP/lsp-seriesinfo/labphon-logo.pdf}}{LabPhon-Logo}}

\bigskip

{\large Studies in Laboratory Phonology}

\bigskip

Chief Editor:  Martine Grice% (Chief Editor, University of Cologne, Germany),
\\
Editors:    Doris Mücke, % (University of Cologne, Germany),
    Taehong Cho % (Hanyang University, Seoul, South Korea)

\bigskip

In this series:

\begin{enumerate}
\item Cangemi, Francesco. Prosodic detail in Neapolitan Italian
\end{enumerate}

\vfill

\gdef\lsISSN{2363-5576}%
\hfill ISSN: \lsISSN

### 2015-05-29 chapters/encoding.tex
initial commit paper

\section{Encoding}
\label{encoding}

There are many in-depth histories of the origin and development of writing systems (e.g.~Robinson 1997, Powell 2012), a story that we will not repeat here. However, the history of turning writing into computer readable code is not so often told, so we decided to offer a short survey of the major developments here. This history turns out to be intimately related to the history of telegraphic communication.\footnote{Because of the recent history as summarised in this section, we have used mostly rather ephemeral internet sources. When not references by traditional literature in the bibliography, we have specifically used \url{http://www.unicode.org/history/} and various Wikipedia pages for the information presented here. A useful survey of the historical development of the physical hardware of telegraphy and telecommunication is Huurdeman (2003). Most books that discuss the development of encoding of telegraphic communication focus of cryptography (e.g.~Singh 1999) and forego the rather interesting story of `open' encoding that is related here.}

Writing systems have existed for roughly 6000 years, allowing people to exchange messages through time and space. Additionally, to bridge large geographic distances, \textsc{telegraphic systems} of communication (from Greek \emph{τῆλε γράφειν} `distant writing') have a long and widespread history since ancient times. The most widespread telegraphic systems worldwide are so-called whistled languages (Meyer 2015), but also drumming languages (Meyer et al. 2012) and smoke, fire, hydraulic or flag signals are forms of telegraphy.

Telegraphy was reinvigorated in the end of the eighteenth century through the introduction of semaphoric systems by Claude Chapelle (since then using flags, flashing lights, or various specially designed contraptions) to convey messages over large distances. The ``innovation'' of those systems was that all characters of the written language were replaced one-to-one by visual signals. Since then, all telegraphic systems have taken this principle, namely that any language to be transmitted first has to be turned into some orthographic system, which subsequently is encoded for transmission by the sender, and then turned back into orthographic representation at the receiver side.\footnote{Sound and video-based telecommunication of course takes a different approach by ignoring the written version of language and directly encode sound waves or light patterns.} This of course implies that the usefulness of any such telegraphic encoding completely depends on the sometimes rather haphazard structure of orthographic systems.

In the nineteenth century, \textsc{electric telegraphy} lead to a new approach in which written language characters were encoded by signals sent through a copper wire. Originally, ``bisignal'' codes were used, consisting of two different signals. For example, Carl Friedrich Gauss in 1833 used positive and negative current (Mania 2008:282). More famous and influential, Samuel Morse in 1836 used long and short pulses.

In those \textsc{bisignal codes} each character from the written language was encoded with a different number of signals (between one and five), so two different separators are needed: one between signals and one between characters. For example, in Morse-code there is a short pause between signals and a long pause between characters. Actually, Morse-code also includes an extra long pause between words. Interestingly, it took a long time to consider the written word boundary---using white-space---as a bona-fide character that should simply be encoded with its own code point. This happened only with the revision of the Baudot-code (see below) by Donald Murray in 1901, in which he introduced a specific white-space code. This principle has been followed ever since.

From those bisignal encodings, true \textsc{binary codes} developed with a fixed length of signals per character. In such systems only a single separator between signals is needed, because the separation between characters can be established by counting until a fixed number of signals.\footnote{Of course, no explicit separator is needed when the timing of the signals is known, which is the principle used in all modern telecommunication systems. An important modern consideration is also how to know where to start counting when you did not catch the start of a message, something that is known in Unicode as ``self synchronization''.}

In the context of electric telegraphy, such a binary code system was first established by Émile Baudot in 1870, using a fixed combination of five signals for each written character.\footnote{True binary codes have a longer history, going at least back to the Baconian cipher devised by Francis Bacon in 1605. However, the proposal by Baudot was the quintessential proposal leading to all modern systems.} There are $2^5 = 32$ possible combination when using five binary signals; an encoding today designated as ``5-bit''. These codes are sufficient for all Latin letters, but of course they do not suffice for all written symbols, including punctuation and digits. As a solution, the Baudot code uses a so-called ``shift'' character, which signifies that from that point onwards---until shifted back---a different encoding is used, allowing for yet another set of 32 codes. In effect, this means that the Baudot code, and the \textsc{International Telegraph Alphabet} (ITA) derived from it, had an extra `bit' of information, so the encoding is actually 6-bit (with $2^6 = 64$ different possible characters). For decades, this encoding was the standard for all telegraphy and it is still in limited use today.

To also allow for different uppercase and lowercase letters and a large variety of control characters to be used in the newly developing technology of computers, the American Standards Association decided to propose a new 7-bit encoding in 1963 (with $2^7 = 128$ different possible characters), known as the \textsc{American Standard Code for Information Interchange} (ASCII), geared towards the encoding of English orthography.

With the ascent of other orthographies in computer usage, the wish to encode further variation of Latin letters (like German <ß> or various letters with diacritics like <è>) led the Digital Equipment Corporation to introduce an 8-bit \textsc{Multinational Character Set} (MCS, with $2^8 = 256$ different possible characters), first used with the introduction of the {\small VT}220 Terminal in 1983.

Because 256 characters were clearly not enough for the many different characters needed in the world's writing systems, the ISO/IEC~8859 standard in 1987 extended the MCS to include 16 different 8-bit code pages. For example, part 5 was used for Cyrillic characters, part 6 for Arabic, and part 7 for Greek.\footnote{In effect, because $16 = 2^4$, this means that ISO/IEC~8859 was actually a $8+4=12$-bit encoding, though with very many duplicates by design (e.g.~all ASCII codes were repeated in each 8-bit code page). To be precise, ISO/IEC~8859 used the 7-bit ASCII as the basis for each code page, and defined 16 different 7-bit extensions, leading to $(1+16)\cdot{2^7} = 2.176$ possible characters.}

This system almost immediately was understood to be insufficient and impractical, so various initiatives to extend and reorganize the encoding started in the 1980s. This led, for example, to various proprietary encodings from Microsoft (e.g.~Windows Latin 1) and Apple (e.g.~Mac OS Roman), which one still sometimes encounters today.

More wide-ranging, various people in the 1980s started to develop true international code sets. In the United States, a group of computer scientists formed the \textsc{Unicode Consortium}, proposing a 16-bit encoding in 1991 (with $2^{16} = 65.536$ different possible characters). At the same time in Europe, the \textsc{International Organization for Standardization} (ISO) was working on ISO~10646 to supplant the ISO/IEC~8859 standard. Their first draft of the \textsc{Universal Character Set} (UCS) in 1990 was 31-bit (with theoretically $2^{31} = 2.147.483.648$ possible characters, but because of some technical restrictions only 679.477.248 were allowed).

Since 1991, the Unicode Consortium and the ISO jointly develop the \textsc{unicode standard}, or ISO/IEC~10646, leading to the current system including the original 16-bit Unicode proposal as the \textsc{basic multilingual plane}, and 16 additional planes of 16-bit for further extensions (with in total $(1+16) \cdot 2^{16} = 1.114.112$ possible characters). The most recent version of the Unicode Standard (7.0) was published in June 2014 and it defines 112.956 different characters (Unicode Consortium 2014).

### 2015-05-29 chapters/introduction.tex
initial commit paper

\chapter{Writing Systems}
\label{writing_systems}

\section{Introduction}
\label{introduction}

Writing systems arise and develop in a complex mixture of cultural, technological and practical pressures. They tend to be highly conservative, in that people who have learned to read and write in a specific way (however impractical or tedious) are mostly unwilling to change their habits, e.g.~they tend to resist spelling reforms. In all literate societies there exists a strong socio-political mainstream that tries to force unification of writing (for example by strongly enforcing `right' from `wrong' writing in schools). However, there is also a large community of users who take as many liberties in their writing as they can get away with.

For example, the writing of tone diacritics in Yoruba is often proclaimed to be the `right' way to write, although many users of Yoruba writing seem to be perfectly fine with leaving them out. As pointed out by the proponents of the `official' rules, there are some homographs when leaving out the tone diacritics (Olúmúyìw 2013: 44). However, writing systems (and the languages they represent) are normally full of homophones, which is normally not a problem at all for speakers of the language. More importantly, writing is not just a purely functional tool, but just as importantly it is a mechanism to signal social affiliation. By showing that you `know the rules' of expressing yourself in writing, others will more easily accept you as a worthy participant in their group. And that just as well holds for obeying to the `official' rules when writing a job application, as for obeying to the `informal' rules when writing an SMS to classmates in school. The case of Yoruba writing is an exemplary case, as even after more than a century of efforts to standardize the writing systems, there is still a wide range of variation in daily use (Olúmúyìw 2013).

The sometimes cumbersome and sometimes illogical structure, and the enormous variability of existing writing systems is a fact of life scholars have to accept and should try to adapt to as good as possible. Our plea here is a proposal for a formalization to do exactly that.

When considering the worldwide linguistic diversity, including all lesser-studied and endangered languages, there exist numerous different orthographies using symbols from the same scripts. For example, there are hundreds of orthographies using Latin-based alphabetic scripts. All of these orthographies use the same symbols, but these symbols differ in meaning and usage throughout the various orthographies. To be able to computationally use and compare different orthographies, we need a way to specify all orthographic idiosyncrasies in a computer-readable format (a process called `tailoring' in Unicode parlance). We call such specifications \textsc{orthography profiles}. Ideally, these specifications have to be integrated into so-called Unicode locale descriptions, though we will argue that in practice this is often not the most useful solution for the kind of problems arising in the daily practice of linguistics. Consequently, a central goal of this paper is to flesh out the linguistic challenges for locale descriptions, and work out suggestions to improve their structure for usage in a linguistic context. Conversely, we also aim to improve linguists' understanding and appreciation for the accomplishments of the Unicode Consortium in the development of the Unicode Standard.

The necessity to computationally use and compare different orthographies most forcefully arises in the context of language comparison. Concretely, in our current research our goal is to develop quantitative methods for language comparison and historical analysis in order to investigate worldwide linguistic variation and to model the historical and areal processes that underlie linguistic diversity (cf.~Steiner et al. 2010, List 2012a, 2012b, List \& Moran 2013, Moran \& Prokic 2013). In this work, it is crucial to be able to flexibly process across numerous resources with different orthographies. In many cases even different resources on the `same' language use different orthographic conventions. Another orthographic challenge that we encounter regularly in our linguistic practice is electronic resources on a particular language that claim to follow a specific orthographic convention (often a resource-specific convention), but on closer inspection such resources are almost always not consistently encoded. Thus a second goal of our orthography profiles is to allow for an easy specification of orthographic conventions, and use such profiles to check consistency and to report errors to be corrected.

A central step in our proposed solution to this problem is the tailored grapheme separation of strings of symbols, a process we call \textsc{grapheme tokenization}. Basically, given some strings of symbols (e.g.~morphemes, words, sentences) in a specific source, our first processing step is to specify how these strings have to be separated into graphemes, considering the specific orthographic conventions used in a particular source document. Our experience is that such a graphemic tokenization can be performed without extensive in-depth knowledge about the phonetic and phonological details of the language in question. For example, the specification that <ou> is a grapheme of English is a much easier task than to specify what exactly the phonetic values of this grapheme are in any specific occurrence in English words. Grapheme separation is a task that can be performed relatively reliably and with limited availability of time and resources (compare, for example, the task of creating a complete phonetic or phonological normalization).

Although grapheme tokenization is only one part of the solution, it is an important and highly fruitful processing step. Given a grapheme tokenization, various subsequent tasks become easier, like (a) temporarily reducing the orthography in a processing pipeline, e.g.~only distinguishing high versus low vowels; (b) normalizing orthographies across sources (often including temporary reduction of oppositions), e.g.~specifying an (approximate) mapping to the International Phonetic Alphabet; (c) using co-occurrence statistics across different languages (or different sources in the same language) to estimate the probability of grapheme matches, e.g.~with the goal to find regular sound changes between related languages or transliterations between different sources in the same language.

Before we deal with these proposals, in the first part of this paper (Sections 2 through 5) we give an extended introduction to the notion of encoding (Section 2) and writing systems, both from a linguistic perspective and from the perspective of the Unicode Consortium (Section 3). We consider the Unicode Standard to be a breakthrough (and ongoing) development that fundamentally changed the way we look at writing systems, and we aim to provide here a slightly more in-depth survey of the many techniques that are available in the standard. A good appreciation for the solutions that the Unicode Standard also allows for a thorough understanding of possible pitfalls that one might encounter when using it (Section 4). As an example of the current state-of-the-art, we discuss the rather problematic marriage of the International Phonetic Alphabet (IPA) with Unicode (Section 5).

The second part of the paper (Sections 6 and 7) describes our proposals for how to deal with the Unicode Standard in the daily practice of (comparative) linguists. First, we discuss the challenges of characterizing a writing system. To solve these problems, we propose the notions of orthography profiles, closely related to Unicode locale descriptions (Section 6). Finally, we discuss practical issues with actual examples (Section 7). We provide reference implementation of our proposals in R and in Python, available as open-source libraries.

The following conventions are followed in this paper. All phonemic and phonetic representations are given in the International Phonetic Alphabet (IPA), unless noted otherwise (International Phonetic Association 2005). Standard conventions are used for distinguishing between graphemic < >, phonemic / / and phonetic [ ] representations. For character descriptions, we follow the notational conventions of the Unicode Standard (Unicode Consortium 2014). Character names are represented in small capital letters (e.g.~\textsc{latin small letter schwa}) and code points are expressed as U\emph{+n} where \emph{n} is a four to six digit hexadecimal number, e.g.~U+0256, which can be rendered as the glyph <ə>.

### 2015-05-29 chapters/ipa_meets_unicode.tex
initial commit paper

\chapter{IPA meets Unicode}
\label{ipa-meets-unicode}

\section{Introducing the International Phonetic Alphabet (IPA)}
\label{introducing-the-international-phonetic-alphabet-ipa}

The International Phonetic Alphabet (IPA) is a common standard in linguistics to transcribe sounds of spoken language into some Latin-based characters (International Phonetic Association 1999). Although IPA is reasonably easily adhered to with pen and paper, it is not trivial to encode IPA characters electronically. Early work addressing the need for a universal computing environment for writing systems and their computational complexity is discussed in Simons 1989. For a long time, linguists (like all other computer users) were limited to ASCII-encoded 7-bit characters, which only includes Latin characters, numbers and some punctuation and symbols. Restricted to these standard character sets that lacked IPA support or other language-specific graphemes that they needed, linguists devised their own solutions (cf.~Bird and Simons 2003). For example, some chose to represent unavailable graphemes with substitutes, e.g.~the combination of to represent . Tech-savvy linguists redefined selected characters from a character encoding by mapping custom made fonts to those code points. However, one linguist's electronic text would not render properly on another linguist's computer without access to the same font. Further, if two character encodings defined two character sets differently, then data could not be reliably and correctly displayed. This is a common example of the non-interoperability of data and data formats.

To alleviate this problem, during the late 1980s, SAMPA (Speech Assessment Methods Phonetic Alphabet) was created to represent IPA symbols with 7-bit printable ASCII character sequences, e.g. for [ɸ]. Two problems with SAMPA are that (i) it is only a partial encoding of the IPA and (ii) it encodes different languages in separate data tables, instead of a universal alphabet, like IPA. SAMPA tables are derived from phonemes appearing in several European languages that were developed as part of a European Commission-funded project to address technical problems like electronic mail exchange (what is now simply called email). SAMPA is essentially a hack to work around displaying IPA characters, but it provided speech technology and other fields a basis that has been widely adopted and used in code. So, SAMPA was a collection of tables to be compared, instead of a large universal table representing all languages. An extended version of SAMPA, called X-SAMPA, set out to include every symbol in the IPA chart including all diacritics (Wells nd.). X-SAMPA was considered more universally applicable because it consisted of one table that encoded the set of characters that represented phones/segments in IPA across languages. SAMPA and X-SAMPA have been widely used for speech technology and as an encoding system in computational linguistics. Eventually, ASCII-encoding of the IPA became depreciated through the advent of the Unicode Standard. Note however that many popular software packages used for linguistic analyses still require ASCII input, e.g.~RuG/L04\footnote{http://www.let.rug.nl/kleiweg/L04/} and SplitsTree4.\footnote{http://www.splitstree.org/}

Still, there are a few pitfalls to be aware of when using the Unicode Standard to encode IPA. As we have said before, from a linguistic perspective it might look like the Unicode Consortium is making incomprehensible decisions, but it is important to realize that the consortium has tried and is continuing to try to be as consistent as possible across a wide range of use cases, and it does place linguistic traditions above other orthographic possibilities. In general, we strongly suggest linguists not to complain about any decisions in the Unicode Standard, but to try and understand the rationale of the Unicode Consortium (which in our experience is almost always well-conceived) and devise ways to work with any unexpected behaviour. Many of the current problems derive from the fact that the IPA is clearly historically based on the Latin script, but different enough from most other Latin-based writing systems to warrant special attention. This ambivalent status of the IPA glyphs (partly Latin, partly special) is unfortunately also attested in the treatment of IPA in the Unicode Standard. In retrospect, it might have been better to consider the IPA (and other transcription systems) to be a special or new kind of script within the Unicode Standard, and treat the obvious similarity to Latin glyphs as a historical relic. All IPA glyphs would then have their own code points, instead of the current situation in which some IPA glyphs have special code points, while others are treated as being identical to the `regular' Latin characters. Yet, the current situation, however unfortunate, is unlikely to change, so as linguists we will have to learn to deal with the specific pitfalls of IPA within the Unicode Standard. In this section, we will describe these pitfalls in some detail.

\section{Pitfall: There is no single complete Unicode code block for IPA}
\label{pitfall-there-is-no-single-complete-unicode-code-block-for-ipa}

The ambivalent nature of IPA glyphs arises because, on the one hand, the IPA uses Latin-based glyphs like , or . From this perspective, the IPA seems to be just another orthographic tradition using Latin characters, all of which do not get a special treatment within the Unicode Standard (just like e.g.~the French, German, or Danish orthographic traditions do not have a special status). On the other hand, the IPA uses many special symbols (like turned , mirrored and/or extended Latin glyphs ) not found in any other Latin-based writing system. For this reason, and already in the first version of the Unicode Standard (Version 1.0 from 1991), a special block with code points, called ``IPA Extensions'' was included. As noted in Section 4, Pitfall 1, the Unicode Standard code space is subdivided into character blocks, which generally encode characters from a single script. However, as is illustrated by the IPA, characters that form a single writing system may be dispersed across several different character blocks. With its diverse collection of symbols from various scripts and diacritics, the IPA is spread across over 13 blocks in the Unicode Standard:\footnote{This number of blocks depends on whether only IPA-sanctioned symbols are counted or if the phonetic symbols commonly found in the literature are also included, see Moran 2012, Appendix C.}
\begin{itemize}
	\item Basic Latin (30 characters), e.g. <a, b, c, d, e> 
	\item Latin-1 Supplement (4 characters): <æ, ç, ð, ø\textgreater{} 
	\item Latin Extended-A (3 characters): <ħ, ŋ, œ> 
	\item Latin Extended-B (5 characters): <ǀ, ǁ, ǂ, ǃ, ȵ> 
	\item IPA Extensions (70 characters), e.g.: <ɐ, ɑ, ɔ> 
	\item Spacing Modifier Letters (20 characters), e.g.: <ʰ ʷ ˥> 
	\item Combining Diacritical Marks (33 characters), e.g.: <̝ ̥ ̪> 
	\item Greek and Coptic (3 characters): <β, θ, χ> 
	\item Phonetic Extensions (2 characters): <ᴅ, ᴴ> 
	\item Phonetic Extensions Supplement (3 characters): <ᶑ , ᶾ, ᶣ> 
	\item Superscripts and Subscripts (1 character): <ⁿ> 
	\item Arrows (2 characters): \textless{}↑, ↓\textgreater{} 
	\item Latin Extended-C (1 character): <ⱱ> 
\end{itemize}

\section{Pitfall: There are many IPA homoglyphs in Unicode}
\label{pitfall-there-are-many-ipa-homoglyphs-in-unicode}

Another problem is the large number of homoglyphs, i.e.~different characters that have highly similar glyphs (or even completely identical, depending on the font rendering). For example, a speaker of Russian should ideally not use the CYRILLIC SMALL LETTER A at code point U+0430 for IPA transcriptions, but instead the LATIN SMALL LETTER A at code point U+0061, although visually they are mostly indistinguishable, and the Cyrillic character is more easily typed on a Cyrillic keyboard. Another example we commonly encounter is the use of LATIN SMALL LETTER G at U+0067, instead of the sanctioned Unicode Standard IPA character for the voiced velar stop LATIN SMALL LETTER SCRIPT G at U+0261. One begins to question whether this issue is at all apparent to the working linguist, or if they simply use the U+0067 because it is easily keyboarded and thus saves time, whereas the latter must be cumbersomely inserted as a special symbol in most software.\footnote{This issue was recently addressed by the International Phonetic Association, which has taken the stance that both the keyboard LATIN SMALL LETTER G and the LATIN SMALL LETTER SCRIPT G are valid input characters for the voiced velar plosive. Unfortunately, this decision further introduces ambiguity for linguists trying to adhere to a strict Unicode Standard IPA encoding.}

Furthermore, on the one hand even linguists are unlikely to distinguish between the LATIN SMALL LETTER SCHWA at code point U+0259 and LATIN SMALL LETTER TURNED E at U+01DD. On the other hand, non-linguists are unlikely to distinguish any semantic difference between an open back unrounded vowel , the LATIN SMALL LETTER ALPHA at U+0251, and the open front unrounded vowel , LATIN SMALL LETTER A at U+0061. But even among linguists this distinction leads to problems. For example, as pointed out by Mielke (2009), there is a problem stemming from the fact that about 75\% of languages are reported to have a five-vowel system (Maddieson 1984). Historically, linguistic descriptions tend not to include precise audio recording and measurements of formants, so this may lead one to ask if the many characters that are used in phonological description reflects a ``transcriptional bias''. The common use of in transcriptions could be in part due to the ease of typing the letter on an English keyboard (or for older descriptions, the typewriter). We found it to be exceedingly rare that a linguist uses for a low back unrounded vowel.\footnote{One example is Vidal 2001a:75, in which the author states: ``The definition of Pilagá /a/ as {[}+back{]} results from its behavior in certain phonological contexts. For instance, uvular and pharyngeal consonants only occur around /a/ and /o/. Hence, the characterization of /a/ and /o/ as a natural class of (i.e., {[}+back{]} vowels), as opposed to /i/ and /e/.''} They simply use $<$a$>$ as long as there is no opposition to $<$ɑ$>$.\footnote{See Thomason's Language Log post, ``Why I don't love the International Phonetic Alphabet'', at: http://itre.cis.upenn.edu/\textasciitilde{}myl/languagelog/archives/005287.html.} Making things even more problematic, there is an old typographic tradition that the double-story uses a single-story in italics. This leads to the unfortunate effect that in most well-designed fonts the italics of and use the same glyph. If this distinction has to be kept upright in italics, the only solution we can currently offer is to use `slanted' glyphs (i.e.~artificially italicised glyphs) instead of real italics (i.e.~special italics glyphs designed by a typographer).\footnote{For example, the widely used IPA font Doulos SIL (http://scripts.sil.org/cms/scripts/page.php?item\_id=DoulosSIL) does not have real italics. This leads some word-processing software, like Microsoft Word, to produce slanted glyphs instead. That particular combination of font and software application will thus lead to the desired effect. However, note that when the text is transferred to another font (i.e.~one that includes real italics) and/or to another software application (like Apple Pages, which does not perform slanting), then this visual appearance will be lost. In this case we are thus still in the pre-Unicode situation in which the choice of font and rendering software actually matters. The ideal solution from a linguistic point of view would be the introduction of a new IPA code point for a different kind of which explicitly specifies that it should still be rendered as a double-story character when italicized. After informal discussion with various Unicode players, our impression is that this highly restricted problem is not sufficiently urgent to introduce even more -like characters in Unicode (which already lead to much confusion, see Section 4, Pitfall 4). This is a clear situation in which the Unicode Consortium is not just thinking about linguists, but has a more wide-ranging practical view to consider.}

Some other homoglyphs related to encoding IPA in the Unicode Standard are:
\begin{itemize}
	\item The uses of the apostrophe has led to long discussions on the Unicode Standard email list. An English keyboard inputs \textless{}`\textgreater{} APOSTROPHE at U+0027, although the `preferred' Unicode apostrophe is the \textless{}'\textgreater{} RIGHT SINGLE QUOTATION MARK at U+2019. Yet the glottal stop/glottalization/ejective marker is another completely different character, the MODIFIER LETTER APOSTROPHE at U+02BC, but unfortunately looks mostly highly similar to U+2019. 
	\item There is ambiguous encoding of IPA segments within the Unicode Standard. An example is the U+02C1 MODIFIER LETTER REVERSED GLOTTAL STOP vs the U+02E4 MODIFIER LETTER SMALL REVERSED GLOTTAL STOP . Both are denoted in the Unicode Standard as the `pharyngealized diacritic' and both appear in various resources representing phonetic data online. This is thus an example for which the Unicode Standard does not solve the linguistic standardization problem. 
	\item There is at least one case in which the character name assigned by the Unicode Consortium does not match the IPA's description: in the Unicode Standard at U+01C3 is labeled LATIN LETTER RETROFLEX CLICK, but in IPA is an alveolar or postalveolar click (not retroflex). This naming is probably best seen as a simple error in the Unicode Standard. Note that most linguists simply seem to use or <ou>) while ligatures are single characters (e.g. <ʧ> LATIN SMALL LETTER TESH DIGRAPH at U+02A7). Ligatures arose in the context of printing easier-to-read texts, and are included in the Unicode Standard for reasons of legacy encoding. However, their usage is discouraged by the Unicode core specification. Specifically related to IPA, various phonetic combinations of characters (typically affricates) are available as single code-points in the Unicode Standard, but are designated as ``ligatures'' or ``digraphs'' (confusingly both names appear interchangeably). Such glyphs might be used by software to produce a pleasing display, but they should not be hard-coded into the text itself. In the context of IPA, characters like the following ligatures should thus \emph{not} be used. Instead a combination of two characters is preferred: 
	\item <ʣ> LATIN SMALL LETTER DZ DIGRAPH at U+02A3 (use <d> + <z> instead) 
	\item <ʧ> LATIN SMALL LETTER TESH DIGRAPH at U+02A7 (use <t> + <ʃ> instead) 
	\item <ʩ> LATIN SMALL LETTER FENG DIGRAPH at U+02A9 (use <f> + <ŋ> instead) 
\end{itemize}

However, there are a few Unicode characters that are historically ligatures, but which are today considered as simple characters in the Unicode Standard and thus should be used when writing IPA, namely:
\begin{itemize}
	\item <ɮ> LATIN SMALL LETTER LEZH at U+026E 
	\item <œ> LATIN SMALL LIGATURE OE at U+0153 
	\item <ɶ> LATIN LETTER SMALL CAPITAL OE at U+0276 
	\item <æ> LATIN SMALL LETTER AE at U+00E6 
\end{itemize}

\section{Pitfall: The notion of diacritic differs between IPA and Unicode}
\label{pitfall-the-ipa-notion-of-diacritics-is-not-the-same-as-the-unicode-standards-notion-of-diacritics}

Another pitfall is diacritics. The problem is that the meaning of the term `diacritics' as used by the IPA is not the same as it used in the Unicode Standard. Specifically, diacritics in the IPA-sense are either so-called \textsc{Combining Diacritical Marks} or \textsc{Spacing Modifier Letters} in the Unicode Standard. Crucially, Combining Diacritical Marks are by definition combined with the character before them (to form so-called default grapheme clusters, see Section 3). In contrast, Spacing Modifier Letters are by definition \emph{not} combined into grapheme clusters with the preceding character, but simply treated as separate letters. In the context of the IPA, the following IPA-diacritics are actually Spacing Modifier Letters in the Unicode Standard:
\begin{itemize}
	\item Length marks, namely <ː> MODIFIER LETTER TRIANGULAR COLON at U+02D0 and <ˑ> MODIFIER LETTER HALF TRIANGULAR COLON at U+02D1. 
	\item Tone letters, like <˥> MODIFIER LETTER EXTRA-HIGH TONE BAR at U+02E5, and others like this. 
	\item Superscript letters, like <ʰ> MODIFIER LETTER SMALL H at U+02B0 or <ˤ> MODIFIER LETTER SMALL REVERSED GLOTTAL STOP at U+02E4, and others like this. 
	\item <˞> MODIFIER LETTER RHOTIC HOOK at U+02DE. 
\end{itemize}

Although linguists might expect these characters to belong together with the character in front of them, at least for <ʰ> MODIFIER LETTER SMALL H at U+02B0 the Unicode Consortium's decision to treat it as a separate character is also linguistically correct, because according to the IPA it can be used both for aspiration (more precisely postaspiration following the base character) and preaspiration (preceding the base character). Note that there is a mechanism in Unicode to force separate characters to be combined (namely by using the ZERO WIDTH JOINER at U+200D), but this seems to be a rather impractical, and probably not enforceable solution to us.

\section{Pitfall: There is no unique diacritic ordering in IPA, nor in Unicode}
\label{pitfall-neither-the-ipa-nor-the-unicode-standard-enforce-a-unique-diacritic-ordering}

Also related to diacritics is the question of ordering. To our knowledge, the International Phonetic Association does not specify a specific ordering for diacritics that combine with phonetic base symbols; this exercise is left to the reasoning of the transcriber. However, such marks have to be explicitly ordered if sequences of them are to be interoperable and compatible. An example is a labialized aspirated alveolar plosive: <tʷʰ>. There is nothing holding linguists back from using <tʰʷ> instead (with exactly the same intended meaning). However, from a technical standpoint, these two sequences are different, e.g.~if both sequences are used in a document, searching for <tʷʰ> will not find any instances of <tʰʷ>, and vice versa. Likewise, a creaky voiced syllabic dental nasal can be encoded in various orders, e.g. <n̪̰̩>, <n̩̰̪> or <n̩̪̰>.

In accordance with the absence of any specification of ordering in the IPA, the Unicode Standard likewise does not propose any standard orderings. Both leave it to the user to be consistent. However, there is one aspect of ordering for which the Unicode Standard does present a canonical solution, which is uncontroversial from a linguistic perspective. Diacritics in the Unicode Standard (i.e.~Combining Diacritical Marks, see above) are classified in Canonical Combining Classes. In practice, the diacritics are distinguished by their position relative to the base character.\footnote{See http://unicode.org/reports/tr44/\#Canonical\_Combining\_Class\_Values for a detailed description.} When applying a Unicode normalization (NFC or NFD, see previous section), the diacritics in different positions are put in a specified order. This process therefore harmonizes the difference between different encodings, for example, of a midtone creaky voice vowel <ḛ̄>. This grapheme cluster can be encoded either as <e> + <̄> + <̰> or as <e> + <̰> + <̄> . To prevent this twofold encoding, the Unicode Standard specifies the second ordering as canonical (diacritics below before diacritics above).

When encoding a string according to the Unicode Standard, it is possible to do this either using the NFC (``composition'') or NFD (``decomposition'') normalization. Decomposition implies that precomposed characters (like <á> LATIN SMALL LETTER A WITH ACUTE at U+00E1) will be split into its parts. This might sound preferable for a linguistic analysis, as the different diacritics are separated from the base characters. However, note that most attached elements like strokes (e.g.~in the <ɨ>), retroflex hooks (e.g.~in <ʐ>) or rhotic hooks (e.g.~in <ɝ>) will not be decomposed, but strangely enough a cedilla (like in <ç>) will be decomposed. In general, Unicode decomposition does not behave like a feature decomposition as expected from a linguistic perspective. It is thus important to consider Unicode decomposition only as a technical procedure, and not assume that it is linguistically sensible.

Facing the problem of specifying a consistent ordering of diacritics while developing a large database of phonological inventories from the world's languages, Moran (2012: 540) defines a set of diacritic ordering conventions.\footnote{The most recent version of these conventions is online: http://phoible.github.io/conventions/} The conventions are influenced by the linguistic literature, though some ad-hoc decisions had to be taken. The goal was to explicitly define all character sequences so that the vast variety of phonemes found in descriptions of the world's language were normalized into consistent character sequences, e.g.~if one language description uses and another , when both are intended to be phonetically equivalent, then a decision to normalize to one form was taken. For example, when a character sequence contains more than one character in Spacing Modifier Letters, the order that is proposed is the following (where <\textbar{}> indicates ``or'' and <→> indicates ``precedes''):
\begin{itemize}
	\itemsep1pt\parskip0pt\parsep0pt 
	\item \textsc{Spacing Modifier Letters ordering:} ( unreleased <̚> \textbar{} lateral release <ˡ> \textbar{} nasal release <ⁿ>) → ( palatalized <ʲ>) → ( labialized <ʷ>) → ( velarized <ˠ>) → ( pharyngealized <ˤ>) → ( aspirated <ʰ> \textbar{} ejective <ʼ>) → ( long <ː> \textbar{} half-long <ˑ>) 
\end{itemize}

If a character sequence contains more than one diacritic below the base character, then the place feature is applied first (dental, laminal, apical, fronted, backed, lowered, raised), then the laryngeal setting (voiced, voiceless, creaky voice, breathy voice) and finally the syllabic or non-syllabic marker (for vowels, ATR gets put on between the place and laryngeal setting). So:
\begin{itemize}
	\itemsep1pt\parskip0pt\parsep0pt 
	\item \textsc{Combining Diacritical Marks (below) ordering:} ( dental <t̪> \textbar{} laminal <t̻> \textbar{} apical <t̺>) → ( fronted <u̟> \textbar{} backed <e̠>) →( lowered <e̞> \textbar{} raised <e̝>) → ( ATR <e̘ e̙>) → ( voiced <s̬> \textbar{} voiceless <n̥> \textbar{} creaky voice <b̰> \textbar{} breathy voice <b̤>) → ( syllabic <n̩> \textbar{} non-syllabic <e̯>) 
\end{itemize}

Character sequences with diacritics above the base character were not problematic in Moran 2012 because they include only the centralized, mid-centralized and nasalized combining characters. Moran (2012) marks tones as singletons with Space Modifier Letters, e.g. \textless{}˦\textgreater{} for a phonemic high tone, instead of accent diacritics, alleviating potential conflicts. Building on the work of Moran (2012), if a character sequence contains more than one diacritic above the base character, we propose:
\begin{itemize}
	\itemsep1pt\parskip0pt\parsep0pt 
	\item \textsc{Combining Diacritical Marks (above) ordering:} (centralized <ë> \textbar{} mid-centralized <e̽>) → (extra short <ĕ>) →( tone accents, e.g. <è> ) → ( Spacing Modifier Letters ) → ( tone letters, e.g. <e˦>) 
\end{itemize}

### 2015-05-29 chapters/originaltext.tex
initial commit paper

The Unicode Cookbook for Linguists Managing orthographies in multilingual datasets using orthography profiles

Steven Moran \& Michael Cysouw

\textsc{Abstract}

In this paper, we define the notion of an orthography profile, which is a language-specific (or even resource-specific) description of the units and rules that are needed to adequately model a writing system. An orthography profile describes the Unicode code points, characters, graphemes and orthographic rules as used in a writing system. Our orthography profiles are used to describe writing systems and possibly even to transliterate them into another (for example IPA-like) transcriptions with the goal to harmonize orthographies across languages and resources on an ad-hoc basis. The central aspect of our approach is that we introduce an intermediate step of graphemic tokenization, which is closer to linguistic practice and easier to manage than transducer-based transliteration. We leverage orthography profiles to enable comparative analysis of languages with different writing systems and to perform quality control on datasets with highly specific orthographic systems. We also describe how our proposals relate to the current Unicode specification, and introduce some of the more delicate aspects of the Unicode Standard for a linguistic audience.

\textsc{Contents}

{[}{[}TOC{]}{]}

\section{1. Introduction}\label{introduction}

Writing systems arise and develop in a complex mixture of cultural, technological and practical pressures. They tend to be highly conservative, in that people who have learned to read and write in a specific way (however impractical or tedious) are mostly unwilling to change their habits, e.g.~they tend to resist spelling reforms. In all literate societies there exists a strong socio-political mainstream that tries to force unification of writing (for example by strongly enforcing `right' from `wrong' writing in schools). However, there is also a large community of users who take as many liberties in their writing as they can get away with.

For example, the writing of tone diacritics in Yoruba is often proclaimed to be the `right' way to write, although many users of Yoruba writing seem to be perfectly fine with leaving them out. As pointed out by the proponents of the `official' rules, there are some homographs when leaving out the tone diacritics (Olúmúyìw 2013: 44). However, writing systems (and the languages they represent) are normally full of homophones, which is normally not a problem at all for speakers of the language. More importantly, writing is not just a purely functional tool, but just as importantly it is a mechanism to signal social affiliation. By showing that you `know the rules' of expressing yourself in writing, others will more easily accept you as a worthy participant in their group. And that just as well holds for obeying to the `official' rules when writing a job application, as for obeying to the `informal' rules when writing an SMS to classmates in school. The case of Yoruba writing is an exemplary case, as even after more than a century of efforts to standardize the writing systems, there is still a wide range of variation in daily use (Olúmúyìw 2013).

The sometimes cumbersome and sometimes illogical structure, and the enormous variability of existing writing systems is a fact of life scholars have to accept and should try to adapt to as good as possible. Our plea here is a proposal for a formalization to do exactly that.

When considering the worldwide linguistic diversity, including all lesser-studied and endangered languages, there exist numerous different orthographies using symbols from the same scripts. For example, there are hundreds of orthographies using Latin-based alphabetic scripts. All of these orthographies use the same symbols, but these symbols differ in meaning and usage throughout the various orthographies. To be able to computationally use and compare different orthographies, we need a way to specify all orthographic idiosyncrasies in a computer-readable format (a process called `tailoring' in Unicode parlance). We call such specifications \textsc{orthography profiles}. Ideally, these specifications have to be integrated into so-called Unicode locale descriptions, though we will argue that in practice this is often not the most useful solution for the kind of problems arising in the daily practice of linguistics. Consequently, a central goal of this paper is to flesh out the linguistic challenges for locale descriptions, and work out suggestions to improve their structure for usage in a linguistic context. Conversely, we also aim to improve linguists' understanding and appreciation for the accomplishments of the Unicode Consortium in the development of the Unicode Standard.

The necessity to computationally use and compare different orthographies most forcefully arises in the context of language comparison. Concretely, in our current research our goal is to develop quantitative methods for language comparison and historical analysis in order to investigate worldwide linguistic variation and to model the historical and areal processes that underlie linguistic diversity (cf.~Steiner et al. 2010, List 2012a, 2012b, List \& Moran 2013, Moran \& Prokic 2013). In this work, it is crucial to be able to flexibly process across numerous resources with different orthographies. In many cases even different resources on the `same' language use different orthographic conventions. Another orthographic challenge that we encounter regularly in our linguistic practice is electronic resources on a particular language that claim to follow a specific orthographic convention (often a resource-specific convention), but on closer inspection such resources are almost always not consistently encoded. Thus a second goal of our orthography profiles is to allow for an easy specification of orthographic conventions, and use such profiles to check consistency and to report errors to be corrected.

A central step in our proposed solution to this problem is the tailored grapheme separation of strings of symbols, a process we call \textsc{grapheme tokenization}. Basically, given some strings of symbols (e.g.~morphemes, words, sentences) in a specific source, our first processing step is to specify how these strings have to be separated into graphemes, considering the specific orthographic conventions used in a particular source document. Our experience is that such a graphemic tokenization can be performed without extensive in-depth knowledge about the phonetic and phonological details of the language in question. For example, the specification that <ou> is a grapheme of English is a much easier task than to specify what exactly the phonetic values of this grapheme are in any specific occurrence in English words. Grapheme separation is a task that can be performed relatively reliably and with limited availability of time and resources (compare, for example, the task of creating a complete phonetic or phonological normalization).

Although grapheme tokenization is only one part of the solution, it is an important and highly fruitful processing step. Given a grapheme tokenization, various subsequent tasks become easier, like (a) temporarily reducing the orthography in a processing pipeline, e.g.~only distinguishing high versus low vowels; (b) normalizing orthographies across sources (often including temporary reduction of oppositions), e.g.~specifying an (approximate) mapping to the International Phonetic Alphabet; (c) using co-occurrence statistics across different languages (or different sources in the same language) to estimate the probability of grapheme matches, e.g.~with the goal to find regular sound changes between related languages or transliterations between different sources in the same language.

Before we deal with these proposals, in the first part of this paper (Sections 2 through 5) we give an extended introduction to the notion of encoding (Section 2) and writing systems, both from a linguistic perspective and from the perspective of the Unicode Consortium (Section 3). We consider the Unicode Standard to be a breakthrough (and ongoing) development that fundamentally changed the way we look at writing systems, and we aim to provide here a slightly more in-depth survey of the many techniques that are available in the standard. A good appreciation for the solutions that the Unicode Standard also allows for a thorough understanding of possible pitfalls that one might encounter when using it (Section 4). As an example of the current state-of-the-art, we discuss the rather problematic marriage of the International Phonetic Alphabet (IPA) with Unicode (Section 5).

The second part of the paper (Sections 6 and 7) describes our proposals for how to deal with the Unicode Standard in the daily practice of (comparative) linguists. First, we discuss the challenges of characterizing a writing system. To solve these problems, we propose the notions of orthography profiles, closely related to Unicode locale descriptions (Section 6). Finally, we discuss practical issues with actual examples (Section 7). We provide reference implementation of our proposals in R and in Python, available as open-source libraries.

The following conventions are followed in this paper. All phonemic and phonetic representations are given in the International Phonetic Alphabet (IPA), unless noted otherwise (International Phonetic Association, 2005). Standard conventions are used for distinguishing between graphemic <\textgreater{}, phonemic / / and phonetic {[} {]} representations. For character descriptions, we follow the notational conventions of the Unicode Standard (Unicode Consortium 2014). Character names are represented in small capital letters (e.g.~LATIN SMALL LETTER SCHWA) and code points are expressed as U\emph{+n} where \emph{n} is a four to six digit hexadecimal number, e.g.~U+0256, which can be rendered as the glyph <ə>.

\section{2. Encoding}\label{encoding}

There are many in-depth histories of the origin and development of writing systems (e.g.~Robinson 1997, Powell 2012), a story that we will not repeat here. However, the history of turning writing into computer readable code is not so often told, so we decided to offer a short survey of the major developments here. This history turns out to be intimately related to the history of telegraphic communication.\footnote{Because of the recent history as summarised in this section, we have used mostly rather ephemeral internet sources, specifically http://www.unicode.org/history/ and various Wikipedia pages for the information presented here (when not references by traditional literature in the bibliography). A useful survey of the historical development of the physical hardware of telegraphy and telecommunication is Huurdeman (2003). Most books that discuss the development of encoding of telegraphic communication focus of cryptography (e.g.~Singh 1999) and forego the rather interesting story of `open' encoding that is related here.)

Writing systems have existed for roughly 6000 years, allowing people to exchange messages through time and space. Additionally, to bridge large geographic distances, \textsc{telegraphic systems} of communication (from Greek \emph{τῆλε γράφειν} ``distant writing'') have a long and widespread history since ancient times. The most widespread telegraphic systems worldwide are so-called whistled languages (Meyer 2015), but also drumming languages (Meyer et al. 2012) and smoke, fire, hydraulic or flag signals are forms of telegraphy. Telegraphy was reinvigorated in the end of the eighteenth century through the introduction of semaphoric systems by Claude Chapelle (since then using flags, flashing lights, or various specially designed contraptions) to convey messages over large distances. The `innovation' of those systems was that all characters of the written language were replaced one-to-one by visual signals. Since then, all telegraphic systems have taken this principle, namely that any language to be transmitted first has to be turned into some orthographic system, which subsequently is encoded for transmission by the sender, and then turned back into orthographic representation at the receiver side.\footnote{Sound and video-based telecommunication of course takes a different approach by ignoring the written version of language and directly encode sound waves or light patterns.) This of course implies that the usefulness of any such telegraphic encoding completely depends on the often rather haphazard structure of orthographic systems.

In the nineteenth century, \textsc{electric telegraphy} lead to a new approach in which written language characters were encoded by signals sent through a copper wire. Originally, `bisignal' codes were used, consisting of two different signals. For example, Carl Friedrich Gauss in 1833 used positive and negative current (Mania 2008: 282). More famous and influential, Samuel Morse in 1836 used long and short pulses. In those bisignal systems each character from the written language was encoded with a different number of signals (between one and five), so two different separators are needed: one between signals and one between characters (in Morse-code there is a short pause between signals and a long pause between characters).\footnote{Actually, Morse-code also includes an extra long pause between words. Interestingly, it took a long time to consider the written white-space as a bona-fide character that should simply be encoded with its own code point instead of with a different kind of pause. This happened only with the revision of the Baudot-code (see below) by Donald Murray in 1901, in which he finally introduced a white-space code. This principle has been followed ever since, and for example was included in the second version of the International Telegraph Alphabet (ITA2) in 1930.)

From those encodings, true \textsc{binary codes} developed with a fixed length of signals per character. In such systems only a single separator between signals is needed, because the separation between characters can be established by counting until a fixed number of signals.\footnote{Of course, no explicit separator is needed when the timing of the signals is known, which is the principle used in all modern telecommunication systems. An important modern consideration is also how to know where to start counting when you did not catch the start of a message, something that is known in Unicode as `self synchronizing'.) In the context of electric telegraphy, such a system was first established by Émile Baudot in 1870, using a fixed combination of five signals for each written character.\footnote{True binary codes have a longer history, going at least back to the Baconian cipher devised by Francis Bacon in 1605.) There are 25 = 32 possible combination when using five binary signals; an encoding today designated as `5-bit'. These codes are sufficient for all Latin letters, but of course they do not suffice for all written symbols, including punctuation and digits. As a solution, the Baudot code uses a so-called `shift' character, which signifies that from that point onwards (until shifted back) a different encoding is used, allowing for yet another set of 32 codes. In effect, this means that the Baudot code, and the \textsc{International Telegraph Alphabet (ITA)} derived from it, had an extra `bit' of information, so the encoding is actually 6-bit (with 26 = 64 different possible characters). For decades, this encoding was the standard for all telegraphy and it is still in limited use today.

To also allow for different uppercase and lowercase letters and a large variety of control characters, the American Standards Association decided to propose a new 7-bit encoding in 1963 (with 27 = 128 different possible characters), known as the \textsc{American Standard Code for Information Interchange(ASCII)}, geared towards the encoding of English orthography. With the ascent of other orthographies in computer usage, the wish to encode further variation of Latin letters (like German <ß> or various letters with diacritics like <è>) led the Digital Equipment Corporation to introduce an 8-bit \textsc{Multinational Character Set (MCS)} (with 28 = 256 different possible characters), first used with the introduction of the VT220 Terminal in 1983. Because 256 characters were clearly not enough for the many different characters needed in the world's writing systems, the ISO/IEC 8859 standard in 1987 extended the MCS to include 16 different 8-bit code pages. For example, part 5 was used for Cyrillic characters, part 6 for Arabic, and part 7 for Greek.\footnote{In effect, because 16 = 24, this means that ISO/IEC 8859 was actually a 8+4 = 12-bit encoding, though with very many duplicates by design (e.g.~all ASCII codes were repeated in each 8-bit code page). To be precise, ISO/IEC 8859 used the 7-bit ASCII as a basis for each code page, and defined 16 different 7-bit extensions, leading to (1+16) times 27 = 2,176 possible characters.)

This system almost immediately was understood to be insufficient and impractical, so various initiatives to extend and reorganize the encoding started in the 1980s. This led, for example, to various proprietary encodings from Microsoft (e.g.~Windows Latin 1) and Apple (e.g.~Mac OS Roman), which one still sometimes encounters today. More wide-ranging, various people in the 1980s started to develop true international code sets. In the United States, a group of computer scientists formed the Unicode Consortium, proposing a 16-bit encoding in 1991 (with 216 = 65,536 different possible characters). At the same time in Europe, the International Organization for Standardization (ISO) was working on ISO 10646 to supplant the ISO 8859 standard. Their first draft of the \textsc{Universal Character Set (UCS)} in 1990 was 31-bit (with theoretically 231 = 2,147,483,648 possible characters, but because of some technical restrictions only 679,477,248 were allowed). Since 1991, the Unicode Consortium and the International Organization for Standardization have jointly developed the \textsc{Unicode Standard}, or ISO/IEC 10646, leading to the current system including the original 16-bit Unicode proposal as the \textsc{Basic Multilingual Plane}, and 16 additional planes of 16-bit for further extensions (with in total 1+16 times 216 = 1,114,112 possible characters). The most recent version of the Unicode Standard (7.0) was published in June 2014 and it defines 112,956 different characters (Unicode Consortium 2014).

\section{3. Terminology}\label{terminology}

\subsection{The linguistic tradition}\label{the-linguistic-tradition}

In this section we provide a very brief overview of the linguistic terminology concerning writing systems before turning to the slightly different computational terminology in the next section on the Unicode Standard. Linguistically speaking, a \textsc{writing system} is a symbolic system that uses visible or tactile signs to represent language in a systematic way. The term `writing system' has two mutually exclusive meanings. First, it may refer to the way a particular language is written. In this sense the term refers to the writing system of a particular language, as, for example, in \emph{``the Serbian writing system uses two scripts: Latin and Cyrillic''}. Second, the term `writing system' may refer to a type of symbolic system as used among the world's languages to represent languages, as, for example, in \emph{``alphabetic writing system''}. In this latter sense the term refers to how scripts have been classified according to the way that they encode language, as in, for example, \emph{``the Latin and Cyrillic scripts are both alphabetic writing systems''}. To avoid confusion, this second notion of `writing system' would have been more aptly called \textsc{script system}. The term \textsc{script} refers to a collection of distinct symbols as employed by one or more orthographies. For example, both Serbian and Russian are written with subsets of the Cyrillic script. A single language, like Serbian or Japanese, can also be written using orthographies based on different scripts. Over the years linguists have typologized script systems in a variety of ways, with the tripartite classification of logographic, syllabic, and alphabetic remaining the most popular, even though there are at least half a dozen different types of script systems that can be distinguished (Daniels, 1990, 1996).

Only focussing here on the first sense, we distinguish two different kinds of writing systems as used to write a particular language, namely transcriptions and orthographies. First, \textsc{transcription} is a scientific procedure (and also the result of that procedure) for graphically representing the sounds of human speech at the phonetic level. It incorporates a set of unambiguous symbols to represent speech sounds, including conventions that specify how these symbols should be combined. A \textsc{transcription system} is a specific system of symbols and rules used for transcription of the sounds of a spoken language variety. In principle, a transcription system should be language-independent, in that it should be applicable to all spoken human languages. The International Phonetic Alphabet (IPA) is a commonly used transcription system that provides a medium for transcribing languages at the phonetic level. However, there is a long history of alternative kinds of transcription systems (see Kemp 1994) and today various alternatives are in widespread use (e.g.~X-SAMPA and Cyrillic-based phonetic transcription systems).\footnote{Most users of IPA do not follow the standard explicitly and many dialects based on it have emerged, e.g.~the Africanist and Americanist transcription systems. ) Note that IPA symbols are also often used to represent language on a phonemic level. However, it is important to realize that in this usage the IPA symbols are not a transcription system, but rather an orthography (though with strong links to the pronunciation). Further, a transcription system does not need to be highly detailed, but it can also be a system of broad sound classes. Although such an approximative transcription is not normally used in linguistics, it is widespread in technological approaches (e.g.~Soundex, Beider \& Morse 2010)\footnote{http://stevemorse.org/phonetics/bmpm.htm , http://stevemorse.org/phonetics/bmpm2.htm) and it is sometimes fruitfully used in automatic approaches to historical linguistics (Dolgopolsky 1986, ASJP\footnote{http://asjp.clld.org/)).

Second, an \textsc{orthography} specifies the symbols, punctuations, and the rules in which a specific language is written in a standardized way. Orthographies are often based on a phonemic analysis, but they almost always include idiosyncrasies because of historical developments (like sound changes or loans) and because of the widely-followed principle of lexical integrity (i.e.~the attempt to write the same lexical root in a consistent way, also when synchronic phonemic rules change the pronunciation, for example with final devoicing in many Germanic languages). All orthographies are language-specific (and often even resource-specific), although individual symbols or rules might be shared between languages. A \textsc{practical orthography} is a strongly phoneme-based writing system designed for practical use by speakers. The mapping relation between phonemes and graphemes in practical orthographies is purposely shallow, i.e.~there is mostly a systematic and faithful mapping from a phoneme to a grapheme. Practical orthographies are intended to jumpstart written materials development by correlating a writing system with the sound units of a language (cf.~Meinhof \& Jones 1928). Symbols from the IPA are often used by linguists in the development of practical orthographies for languages without writing systems, though this usage of IPA symbols should not be confused with transcription (as defined above). Further, a \textsc{transliteration} is a mapping between two different orthographies. It is the process of ``recording the graphic symbols of one writing system in terms of the corresponding graphic symbols of a second writing system'' (Kemp 2006:396). In straightforward cases, such a transliteration is simply a matter of replacing one symbol with another. However, there are widespread complications, like one-to-many or many-to-many mappings, which are not always easy, or even possible, to solve without listing all cases individually (cf.~Moran 2012, chp 2).

Breaking it down further, a script consists of \textsc{graphemes}, and graphemes consist of \textsc{characters}. In the linguistic terminology of writing systems, a \textsc{character} is a general term for any self-contained element in a writing system.\footnote{There is a second interpretation of the term `character', i.e.~a conventional term for a unit in the Chinese writing system (Daniels 1996). This interpretation will not be further explored in this paper.) Although in literate societies most people have a strong intuition about what the characters are in their particular orthography or orthographies, it turns out that the separation of an orthography into separate characters is far from trivial. The widespread intuitive notion of a character is strongly biased towards educational traditions, like the alphabet taught at schools, and technological possibilities, like the available type pieces in a printer's job case, the keys on a typewriter, or the symbols displayed in Microsoft Word's symbol browser. In practice, characters often consist of multiple building blocks, each of which could be considered a character in its own right. For example, although a Chinese character may be considered to be a single basic unanalyzable unit, at a more fine-grained level of analysis the internal structure of Chinese characters is often comprised of smaller semantic and phonetic units that should be considered characters (Sproat 2000). In alphabetic scripts, this problem is most forcefully exemplified by diacritics. A \textsc{diacritic} is a mark, or series of marks, that may be above, below, or through other characters (Gaultney 2002). Diacritics are sometimes used to distinguish homophonous words, but they are more often used to indicate a modified pronunciation (Daniels \& Bright 1996, xli). The central question is whether, for example, , , and should be considered four characters, or different combinations of three characters. In general, multiple characters together can form another character, and it is not always possible to decide on principled grounds what should be the basic building blocks of an orthography.

For that reason, it is better to analyse an orthography as a collection of graphemes. A \textsc{grapheme} is the basic, minimally distinctive symbol of a particular writing system, alike to the phoneme is an abstract representation of a distinct sound in a specific language. The term `grapheme' was modeled after the term `phoneme' and represents a contrastive graphical unit in a writing system (see Kohrt 1986 for a historical overview of the term grapheme). Most importantly, a single grapheme regularly consists of multiple characters, like <th>, <ou> and <gh> in English (note that each character in these graphemes is also a separate grapheme in English). Such complex graphemes are often used to represent single phonemes. So, a combination of characters is used to represent a single phoneme. Note that the opposite is also found in writing systems, in cases in which a single character represents a combination of two or more phonemes. For example, <x> in English orthography represents a combination of the phonemes /k/ and /s/. Further, conditioned or free variants of a grapheme are called \textsc{allographs}. For example, the distinctive forms of Greek sigma are conditioned, with <σ> being used word-internally and <ς> being used at the end of a word. In sum, there are many-to-many relationships between phonemes and graphemes as they are expressed in the myriad of language- and resource-specific orthographies.

This exposition of the linguistic terminology involved in describing writing systems has been purposely brief. We have highlighted some of the linguistic notions that are pertinent, yet sometimes confused with, the technological definitions developed for the computational processing of the world's writing systems, which we describe in the next section.

\subsection{The Unicode approach}\label{the-unicode-approach}

The conceptualisation and terminology of writing systems was rejuvenated through the development of the Unicode Standard, with major input from Mark Davis, co-founder and long-term president of the Unicode Consortium. For many years, `exotic' writing systems and phonetic transcription systems on personal computers were constrained by the American Standard Code for Information Interchange* *(ASCII) character encoding scheme (based on the Latin script), which only allowed for a strongly limited number of different symbols to be encoded. This implied that users could either use and adopt the (extended) Latin alphabet or they could assign new symbols to the small number of code points in the ASCII encoding scheme to be rendered by a specifically designed font (Bird \& Simons 2003). In this situation, it was necessary to specify the font together with each document to ensure the rightful display of its content. To alleviate this problem of assigning different symbols to the same code points, in the late 80's and early 90's the Unicode Consortium set itself the ambitious goal of developing a single universal character encoding to provide a unique number, a code point, for every character in the world's writing systems. Nowadays, the Unicode Standard is the default encoding of the technologies that support the World Wide Web and for all modern operating systems, software and programming languages.

The Unicode Standard represents a massive step forward because it aims to eradicate the distinction between `universal' (ASCII) versus `language/resource-particular' (Font) by adding as much as possible language-specific information into the universal standard. However, there are still language/resource-specific specifications necessary for the proper usage of Unicode, as will be discussed below. Within the Unicode structure many of these specifications can be captured by so-called `Locales' or `Locale Descriptions', so we are moving to a new distinction of `universal' (Unicode Standard) versus `language-particular' (Locale Description). The major gain is a much larger compatibility on the universal level (because Unicode standardizes a much larger portion of writing system diversity), and much better possibilities for automatic processing on the language-particular level (because Locale Descriptions are computer readable specifications).

Each version of the Unicode Standard (as of writing at version 7) consists of a set of specifications and guidelines that include (i) a core specification, (ii) code charts, (iii) standard annexes and (iv) a character database.\footnote{All documents of the Unicode Standard are available at http://www.unicode.org/versions/latest/. For a quick survey of the use of terminology inside the Unicode Standard, their glossary is particularly useful, available at http://www.unicode.org/glossary/. For a general introduction to the principles of Unicode, Chapter 2 of the core specification, called ``General Structure'', is particularly insightful. Different from many other documents of the Unicode Standard, this general introduction is relatively easy to read and illustrated with many interesting examples from various orthography traditions all over the world. ) The \textsc{core specification} is a book directed toward human readers that describes the formal standard for encoding multilingual text. The \textsc{code charts} provide a humanly readable online reference to the character contents of the Unicode Standard in the form of PDF files. The \textsc{Unicode Standard Annexes (UAX)} are a set of technical standards that describe the implementation of the Unicode Standard for software development, Web standards, programming languages, etc. The \textsc{Unicode Character Database (UCD)} is a set of computer-readable text files that describe the character properties, including a set of rich character and writing system semantics, for each character in the Unicode Standard. In this section, we introduce the basic Unicode concepts, but we will leave out many details. Please consult the above mentioned full documentation for a more detailed discussion. Further note that the Unicode Standard is exactly that, namely a standard. It normatively describes notions and rules to be followed. In the actual practice of applying this standard in a computational setting, a specific implementation is necessary. The most widely used implementation of the Unicode Standard is the \textsc{International Components for Unicode (ICU)}, which offers C/C++ and Java libraries trying to implement to Unicode Standard.\footnote{http://icu-project.org )

The Unicode Standard is a \textsc{character encoding system} which goal it is to support the interchange and processing of written characters and text in a computational setting. Underlyingly, the character encoding is represented by a range of numerical values called a \textsc{code space}, which is used to encode a set of characters. A \textsc{code point} is a unique non-negative integer within a code space (i.e.~within a certain numerical range). In the Unicode Standard character encoding system, an \textsc{abstract character}, for example a ``LATIN SMALL LETTER P'', is mapped to a particular code point, such as the decimal value 112, normally represented in hexadecimal and in Unicode parlance as (U+)0070.\footnote{Hexadecimal (`base-16') 0070 is equivalent to decimal (`base-10') 112, which can be calculated by considering 0·163 + 0·162 + 7·161 + 0·160 = 7·16 = 112. Underlyingly, computers will of course treat this code point binary (`base-2') as 11100000, i.e.~112 = 1·27 + 1·26 + 1·25 + 0·24 + 0·23 + 0·22 + 0·21 + 0·20 = 64 + 32 + 16.) That encoded abstract character is rendered on a computer screen (or printed page) as a \textsc{glyph}, e.g. <p>, depending on the \textsc{font} and the context in which that character appears.

In Unicode Standard terminology, an (abstract) \textsc{character} is the basic encoding unit. The term `character' can be quite confusing due to its alternative definitions across different scientific disciplines and because in general the word `character' means many different things to different people. It is therefore often preferable to refer to Unicode characters simply as \textsc{code points}, because there is a one-to-one mapping between Unicode characters and their numeric representation. The Unicode Consortium (2012) defines a character as ``the smallest component of written language that has semantic value. It refers to the abstract meaning and/or general shape, rather than a specific shape, though in code tables some form of visual representation is essential for the reader's understanding.'' Unicode defines characters as abstractions of orthographic symbols, and it does not define visualizations for these characters (although it does presents examples). In contrast, a \textsc{glyph} is a concrete graphical representation of a character as it appears when rendered (or rasterized) and displayed on an electronic device or on printed paper. For example, <reinstert Gs here> are different glyphs of the same character, i.e.~they may be rendered differently depending on the typography being used, but they all share the same code point. From the perspective of Unicode they are `the same thing'. In this approach, a \textsc{font} is then simply a collection of glyphs linked to code points. Allography is not specified in Unicode (expect for a few exceptional cases, due to legacy encoding issues), but can be specified in a font as a \textsc{contextual variant} (aka presentation form).

Each code point in the Unicode Standard is associated with a set of \textsc{character properties} as defined by the Unicode character property model.\footnote{The character property model is described in http://www.unicode.org/reports/tr23/, but the actual properties are described in http://www.unicode.org/reports/tr44/. A simplified overview of the properties is available at http://userguide.icu-project.org/strings/properties. The actual code tables listing all properties for all Unicode code points are available at http://www.unicode.org/Public/UCD/latest/ucd/. ) Basically, those properties are just a long list of values for each character. For example, code point U+0047 has the following properties (among many others):
\begin{itemize}
	\item Name: LATIN CAPITAL LETTER G 
	\item Alphabetic: YES 
	\item Uppercase: YES 
	\item Script: LATIN 
	\item Extender: NO 
	\item Simple\_Lowercase\_Mapping: 0067 
\end{itemize}

These properties contain the basic information of the Unicode Standard and they are necessary to define the correct behavior and conformance required for interoperability in and across different software implementations (as defined in the Unicode Standard Annexes). The character properties assigned to each code point is based on each character's behavior in the real-world writing traditions. For example, the corresponding lowercase character to U+0047 is U+0067 (though note that the relation between uppercase and lowercase is in many situations much more complex than this, and Unicode has further specifications for those cases). Another use of properties is to define the script of a character. The Unicode Standard defines the term \textsc{script} as, ``A collection of letters and other written signs used to represent textual information in one or more writing systems. For example, Russian is written with a subset of the Cyrillic script; Ukrainian is written with a different subset. The Japanese writing system uses several scripts.'' In practice, script is simply defined for each character as the explicit ``Script'' property in the Unicode Character Database.

One frequently references property is the \textsc{block} property, which is often used in software applications to impose some structure to the large number of Unicode characters. Each character in Unicode belongs to a specific block. These blocks are basically an organisational structure to alleviate the administrative burden of keeping Unicode up-to-date. Blocks consist of characters that in some way belong together, so that characters are easier to find. Some blocks are connected with a specific script, like the Hebrew block or the Gujarati block. However, blocks are predefined ranges of code points, and often there will come a point after which the range is completely filled. Any extra characters will have to be assigned somewhere else. There is, for example, a block `Arabic', which contains most Arabic symbols. However, there is also a block `Arabic Supplement', `Arabic Presentation Forms-A' and `Arabic Presentation Form B'. The situation with Latin symbols is even more extra. In general, the names for block should be taken as a definitional statement. For example, many IPA symbols are not located in the aptly-names block ``IPA extensions'', but in other blocks (see Section 5).

There are many cases in which a sequence of characters (i.e.~a sequence of more than one code point) represents what a user perceives as an individual unit in a particular orthographic writing system. For this reason the Unicode Standard differentiates between \textsc{abstract character} and \textsc{user-perceived character}. Sequences of multiple code points that correspond to a single user-perceived characters are called** grapheme clusters** in Unicode parlance. Grapheme clusters come in two flavors: (default) grapheme clusters and tailored grapheme clusters.

The (default) \textsc{grapheme clusters} are locale-independent graphemes, i.e.~they always apply when a particular combination of characters occurs independent of the writing system in which they are used. These character combinations are defined in the Unicode Standard as functioning as one \textsc{text element}.\footnote{The Unicode Standard defines text element as: ``A minimum unit of text in relation to a particular text process, in the context of a given writing system. In general, the mapping between text elements and code points is many-to-many.'') The simplest example of a grapheme cluster is a base character followed by a letter modifier character. For example, the sequence + <̃\textgreater{} (LATIN SMALL LETTER N at U+006E followed by COMBINING TILDE at U+0303) combines visually into , a user-percieved character in writing systems like that of Spanish. So, what the user perceives as a single character actually involves a multi-code-point sequence. Note that this specific sequence can also be represented with a single \textsc{precomposed} code point, LATIN SMALL LETTER N WITH TILDE at U+00F1, but this is not the case for all multi-code point character sequences. The problem that there multiple encodings possible for the same text element has been acknowledged early on in the Unicode Standard (e.g.~for , the sequence U+006E U+0303 should in all situations be treated identically to the precomposed U+00F1), and a system of \textsc{canonical equivalence} is available for such situations. Basically, the Unicode Standard offers different kind of normalizations to either decompose all precomposed characters (called \textsc{NFD}, \textsc{Normalization Form Canonical Decomposition}), or compose as much as possible combinations (called \textsc{NFC}, \textsc{Normalization Form Canonical Composition}). In current practice of software development, NFC seems to be preferred in most situations and is widely proposed as the preferred canonical form.

More difficult for text processing, because less standardized, is what the Unicode Standard terms \textsc{tailored grapheme clusters}. Tailored grapheme clusters are locale-dependent graphemes, i.e.~such combination of characters do not function as text elements in all situations. For example, the sequence + for the Slovak digraph or the sequence in the Sisaala practical orthography (pronounced as IPA /tʃ/; Moran 2006). These grapheme clusters are ``tailored'' in the sense that they must be specified on a language-by-language or a writing system-by-writing system basis. The Unicode Standard provides technological specifications for creating locale specific data in so-called \textsc{Unicode Locale Descriptions}, i.e.~a set of specification that defines a set of language-specific elements (e.g.~tailored grapheme clusters, collation order, capitalization-equivalence), as well as other special information, like how to format numbers, dates, or currencies. Locale descriptions are saved in the \textsc{Common Locale Data Repository (CLDR)},\footnote{http://cldr.unicode.org/ ) a repository of language-specific definitions of writing system properties, each of which describes specific usages of characters. Each locale can be encoded in a document using the \textsc{Locale Data Markup Language (LDML)}. LDML is an XML format and vocabulary for the exchange of structured locale data.

Unicode Locale Descriptions allow users to define language- or resource-specific writing systems or orthographies. The Unicode Consortium defines \textsc{writing system} only very loosely, as it is not a central concept in the Unicode Standard. A writing system is, ``A set of rules for using one or more scripts to write a particular language. Examples include the American English writing system, the British English writing system, the French writing system, and the Japanese writing system.''. However, there are various drawbacks of locale descriptions for the daily practice of linguistic work in a multilingual setting (see Section 7).

\section{4. Unicode Pitfalls}\label{unicode-pitfalls}

\subsection{Wrong it is not}\label{wrong-it-is-not}

In this section, we describe some of the most common pitfalls that we have encountered when using the Unicode Standard in our own work, or in discussion with other linguists. This section is not meant as a criticism of the decisions made by the Unicode Consortium; on the contrary, we aim to highlight where the technological aspects of the Unicode Standard diverge from many users' intuitions. What have sometimes been referred to as problems or inconsistencies in the Unicode Standard are mostly due to legacy compatibility issues, which can lead to unexpected behavior by linguists using the standard. However, there are also some cases in which the Unicode Standard has made decisions that theoretically could have been taken differently, but for some reason or another (mostly very good reasons) were accepted as they are now. We call behavior that executes without error but does something different than the user expected---often unknowingly---a ``pitfall''.

In this context, it is important to realize that the Unicode Standard was not developed to solve linguistic problems per se, but to offer a consistent computational environment for written language. In those cases in which the Unicode Standard behaves differently as expected, we think it is important not to dismiss Unicode as `wrong' or `deficient', because our experience is that in almost all cases the behaviour of the Unicode Standard has been particularly well thought through. The Unicode Consortium considers important practical use-cases that from a linguistic point of view are normally not considered. Our general guideline for dealing with the Unicode Standard is to accept it as it is, and not to battle windmills. Alternatively, of course, it is possible to actively engage in the development of the standard itself, an effort that is highly appreciated by the Unicode Consortium.

\subsection{Pitfall: Blocks}\label{pitfall-blocks}

The Unicode code space is subdivided into blocks of contiguous code points. For example, the block called ``Cyrillic'' runs from U+0400 till U+04FF. These blocks arose as an attempt at ordering the enormous amount of characters in Unicode, but the ideas of blocks very quickly ran into problems. First, the size of a block is fixed, so when a block is full, a new block will have to be instantiated somewhere further in the code space. For example, this led to the blocks ``Cyrillic Supplement'', ``Cyrillic Extended-A'' (which are also already full) and ``Cyrillic Extended-B''. Second, when a specific character already exists, then it is not duplicated in another block, although the name of the block might indicate that a specific symbol should be available there. In general, names of blocks are just an approximate indication of the kind of characters that will be in the block.

The problem with blocks arises because finding the right character among the thousands of Unicode characters is not easy. Many software applications present blocks as a primary search mechanism, because the block names suggest where to look for a particular character. However, when a user searches for an IPA character in the block ``IPA Extensions'' then many IPA characters will not be found there. For example, the velar nasal <ŋ> is not part of the block ``IPA Extensions'' because it was already included as LATIN SMALL LETTER ENG at U+014B in the block ``Latin Extensions-A''.

In general, finding a specific character in the Unicode Standard is often not trivial. The names of the blocks can help, but it is not (and never was supposed to be) a foolproof structure. It is not the goal nor aim of the Unicode Consortium to provide a user interface to the Unicode Standard. If one often encounters the problem of needing to find a suitable character, there are various other useful services for end users available.\footnote{The Unicode website offers a basic interface to the code charts at http://www.unicode.org/charts/index.html. As a more flexible interface, we particularly like PopChar from Macility, available for both Macintosh and Windows. However, there are various free websites that offer search interfaces to the Unicode code tables, like http://unicode-search.net or http://unicode-search.net. A further useful approach for searching characters using shape matching is http://shapecatcher.com. )

\subsection{Pitfall: Missing glyphs}\label{pitfall-missing-glyphs}

The Unicode Standard is often praised (and deservedly so) for solving many of the perennial problems with the interchange and display of the world's writing systems. However, a common complaint from users is that, while the praise may be true, they mostly just see some boxes on their screen instead of those promised symbols. The problem of course is that users' computers do not have any glyphs installed matching the Unicode code points in the file they are trying to inspect. It is important to realize that internally in the computer everything still works as expected: any handling of Unicode code points works independently of how they are displayed on the screen. So, although a user might only see boxes being displayed, this user should be assured that everything is still in order.

The central problem behind the missing glyphs is that designing actual glyphs includes a lot of different considerations and it is a time-consuming process. Many traditional expectations of how specific characters should look like have to be taken into account when designing glyphs. Those expectations are often not well documented, and it is mostly up to the knowledge and experience of the font designer to try and conform to them as good as possible. Therefore, most designers produce fonts only including glyphs for certains parts of the Unicode Standard, namely for those characters they feel comfortable with. At the same time, the number of characters defined by the Unicode Standard is growing with each new version, so it is neigh impossible for any designer to produce glyphs for all characters. The result of this is that, almost necessarily, each font only includes glyphs for a subset of the characters in the Unicode Standard.

The simple solution to missing glyphs is thus to install additional fonts providing additional glyphs. For the more exotic characters there is often not much choice. There are a few particularly large fonts that might be considered. First, there is the \textsc{Everson Mono} font made by Michael Everson, which currently includes 9,756 different glyphs (not including Chinese) updated up to Unicode 7.0.\footnote{Everson Mono is available as shareware at http://www.evertype.com/emono/. ) Already a bit older is the \textsc{Titus Cyberbit Basic} font made by Jost Gippert and Carl-Martin Bunz, which includes 10,044 different glyphs (not including Chinese), but not including newer characters added after Unicode 4.0.\footnote{Titus Cyberbit Basic is available at http://titus.fkidg1.uni-frankfurt.de/unicode/tituut.asp. )

Further, we suggest to always install at least one so-called \textsc{fall-back font,} which provides glyphs that at least show the user some information about the underlyingly encoded character. Apple Macintoshes have such a font (which is invisible to the user), which is designed by Michael Everson and made available for other systems through the Unicode Consortium.\footnote{The Apple/Everson fallback font is available at http://www.unicode.org/policies/lastresortfont\_eula.html. ) Further, the \textsc{GNU Unifont} is a clever way to produce bitmaps approximating the intended glyph of each available character, updated to Unicode 7.0.\footnote{The GNU Unifont is available at http://unifoundry.com/unifont.html. ) Finally, the Summer Institute of Linguistics provides a \textsc{SIL Unicode BMP Fallback Font}, currently available up to Unicode version 6.1. This font does not even attempt to show a real glyph, but only shows the hexadecimal code inside a box for each character, so a user can at least see the Unicode codepoint of the character to be displayed.\footnote{The SIL Unicode BMP Fallback Font is available at http://scripts.sil.org/UnicodeBMPFallbackFont. )

\subsection{Pitfall: Faulty rendering}\label{pitfall-faulty-rendering}

A similar complaint to missing glyphs, discussed previously, is that while there might be a glyph being displayed, it does not `look right'. There are two reasons for unexpected visual display, namely automatic font substitution and faulty rendering. Like missing glyphs, any such problems are independent from the Unicode Standard. The Unicode Standard only includes very general information about characters and leaves the specific visual display to others to decide on. Any faulty display is thus not to be blamed on the Unicode Consortium, but on a complex interplay of different mechanisms happening in a computer to turn Unicode codepoints into visual symbols. We will only sketch a few aspects of this complex interplay here.

Most modern software applications (like Microsoft Word) offer some approach to \textsc{automatic font substitution}. This means that when a text is written in a specific font (e.g.~Times New Roman) and a inserted Unicode character does not have a glyph within this font, then the software application will automatically search for another font to display the glyph. The result will be that some specific glyph will look slightly different from the others. This mechanism works differently depending on the software application, and mostly only limited user influence is expected and little feedback is given, which might be rather frustrating to font-aware users.\footnote{For example, Apple Pages does not give any feedback that a font is being replaced, and the user does not seem to have any influence on the choice of replacement (except by manually marking all occurrences). In contrast, Microsoft Word does indicate the font replacement by showing the name of the font replacement, but it simply changes the font, so any text written after the replacement is written in a different font as before. Both behaviors leave much to be desired.)

The other problem with visual display is related to the so-called `font rendering'. \textsc{Font rendering }refers to the process of the actual positioning of Unicode characters on a page of written text. This positioning is actually a highly complex problem, and many things can go wrong in the process. Well-known rendering problems, like proportional glyph size or ligatures are reasonably well understood. However, even within the Latin script, the positioning of diacritics relative to a base character is still a widespread problem. Especially when more than one diacritic is supposed to be placed above (or below) each other, this often leads to unexpected effects in many modern software applications. The problems arising in Arabic and in many southeast Asian scripts (like Devanagari or Burmese) are even much more complex. There are basically three different approaches to font rendering. The most widespread is Adobe's and Microsoft's \emph{OpenType }system. This approach makes it relatively easy for font developers, as the font itself does not include all details about the precise placement of individual characters. For those details, additional script-descriptions are necessary. All of those systems can lead to unexpected behaviour.\footnote{For more details about OpenType, see http://www.adobe.com/products/type/opentype.html and http://www.microsoft.com/typography/otspec/. Additional systems for complex text layout are, among others, Microsoft's DirectWrite https://msdn.microsoft.com/library/dd368038.aspx and the open-source project HarfBuzz http://www.freedesktop.org/wiki/Software/HarfBuzz/. ) Alternative systems are Apple's \emph{Apple Advanced Typography} (AAT) and the open-source \emph{Graphite} system from the Summer Institute of Linguistics (SIL).\footnote{More information about AAT can be found at https://developer.apple.com/fonts/. SIL's Graphite is described in detail at http://scripts.sil.org/cms/scripts/page.php?site\_id=projects\&item\_id=graphite\_home. ) In both of these systems, a larger burden is placed on the description inside the font.

\subsection{Pitfall: Homoglyphs}\label{pitfall-homoglyphs}

Homoglyphs are visually indistinguishable glyphs (or highly similar glyphs) that have different code points in the Unicode Standard and thus different character semantics. As a principle, the Unicode Standard does not specify how a character appears visually on the page or the screen. So in most cases, a different appearance is caused by the specific design of a font, or by user-settings like size or boldface. Taking an example already discussed in the previous section, the following symbols <reinsert Gs here> are different glyphs of the same character, i.e.~they may be rendered differently depending on the typography being used, but they all share the same code point (viz. LATIN SMALL LETTER G at U+0067). In contrast, the symbols <A А Α Ꭺ ᗅ ᴀ ꓮＡ𐊠 𐌀 𝖠 𝙰> are all different code points, although they look highly similar -- in some cases even sharing exactly the same glyph in some fonts. All these different A-like characters include the following code points in the Unicode Standard:
\begin{itemize}
	\item <A\textgreater{} LATIN CAPITAL LETTER A, at U+0041 
	\item <А> CYRILLIC CAPITAL LETTER A, at U+0410 
	\item <Α> GREEK CAPITAL LETTER ALPHA, at U+0391 
	\item <Ꭺ> CHEROKEE LETTER GO, at U+13AA 
	\item <ᗅ> CANADIAN SYLLABICS CARRIER GHO, at U+15C5 
	\item <ᴀ> LATIN LETTER SMALL CAPITAL A, at U+1D00 
	\item <ꓮ> LISU LETTER A, at U+A4EE 
	\item <Ａ> FULLWIDTH LATIN CAPITAL LETTER A, at U+FF21 
	\item <𐊠> CARIAN LETTER A, at U+102A0 
	\item <𐌀> OLD ITALIC LETTER A, at U+10300 
	\item <𝖠> MATHEMATICAL SANS-SERIF CAPITAL A, U+1D5A0 
	\item <𝙰> MATHEMATICAL MONOSPACE CAPITAL A, at U+1D670 
\end{itemize}

The existence of such homoglyphs is partly due to legacy compatibility, but for the most part these characters are simply different characters that happen to look similar.\footnote{A particularly nice interface to look for homoglyphs is http://shapecatcher.com, based on the principle of recognizing shapes (Belongie et al. 2002).) Yet, they are supposed to behave different from the perspective of a font designer. For example, when designing a Cyrillic font, the will have different aesthetics and different traditional expectation compared to a Latin .

Such homoglyphs are a widespread problem for consistent encoding. Although for most users it looks like the words and are identical, in actual fact they do not even share a single code point.\footnote{The first words consists completely of Latin characters, namely U+0076, U+006F, U+0063, U+0065, U+0073, while the second is a mix of Cyrillic and Greek characters, namely U+03BD, U+03BF, U+0041, U+0435, U+0455.) For computers these two words are completely different entities. Commonly, when users with Cyrillic or Greek keyboards have to type some Latin-based orthography, they mix similar looking Cyrillic or Greek characters into their text, because those characters are so much easier to type. Similarly, when users want to enter an unusual symbol, they normally search by visual impression in their favourite software application, and just pick something that looks reasonably alike to what they expect the glyph to look like.

\subsection{Pitfall: Canonical equivalence}\label{pitfall-canonical-equivalence}

For some characters, there is more than one possible encoding in the Unicode Standard. This is a possible pitfall, as this would mean that for the computer there exist multiple different entities that for a user are the same. This would, for example, lead to problems with searching, as the computer would search for specific encodings, and not find all expected characters. As a solution, the Unicode Standard includes a notion of \textsc{canonical equivalence}. Different encodings are explicitly declared as equivalent in the Unicode Standard code tables. Further, to harmonize all encodings in a specific piece of text, the Unicode Standard proposes a mechanism of \textsc{normalization}.

Consider for example the characters and following Unicode code points:
\begin{enumerate}
	\def\labelenumi{\arabic{enumi}.} 
	\item LATIN CAPITAL LETTER A WITH RING ABOVE (U+00C5) 
	\item ANGSTROM SIGN (U+212B) 
	\item LATIN CAPITAL LETTER A (U+0041) + COMBINING RING ABOVE (U+030A) 
\end{enumerate}

The character, represented here by glyph , is encoded in the Unicode Standard in the first two examples by a single-character sequence; each is assigned a different code point. In the third example, the glyph is encoded in a multiple-character sequence that is composed of two character code points. All three sequences are \textsc{canonically equivalent}, i.e.~they are strings that represent the same abstract character and because they are not distinguishable by the user, the Unicode Standard requires them to be treated the same in regards to their behavior and appearance. Nevertheless, they are encoded differently. For example, if one were to search an electronic text (with software that does not apply Unicode Standard normalization) for ANGSTROM SIGN (U+212B), instances of LATIN CAPITAL LETTER A WITH RING ABOVE (U+00C5) would not be found.

In other words, there are equivalent sequences of Unicode characters that should be normalized, i.e.~transformed into a unique Unicode-sanctioned representation of a character sequence called a \textsc{normalization form}. Unicode provides a \emph{Unicode Normalization Algorithm}, which essentially puts combining marks into a specific logical order and it defines decomposition and composition transformation rules to convert each string into one of four normalization forms. We will discuss here the two most relevant normalization forms: NFC and NFD.

The first of the three characters above is considered the \textsc{Normalization Form C (NFC)}, where ``C'' stands for composition. When the process of NFC normalization is applied to the character sequences in 2 and 3, both sequences are normalized into the `pre-composed' character sequence in 1. Thus all three canonical character sequences are standardized into one composition form in NFC. The other central Unicode normalization form is the \textsc{Normalization Form D (NFD)}, where ``D'' stands for decomposition. When NFD is applied to the three examples above, all three, including importantly the single-character sequences in 1 and 2, are normalized into the `decomposed' multiple-sequence of characters in 3. Again, all three are then logically equivalent and therefore comparable and syntactically interoperable.

As illustrated, some characters in the Unicode Standard have alternative representations (in fact, many do), but the Unicode Normalization Algorithm can be used to transform certain sequences of characters into canonical (or compatible) forms to test for equivalency. To determine equivalence, each character in the Unicode Standard is associated with a combining class, which is formally defined as a character property called ``Canonical Combining Class'' which is specified in the Unicode Character Database. The combining class assigned to each code point is a numeric value between 0 and 254 and is used by the Canonical Ordering Algorithm to determine which sequences of combining marks are canonically equivalent. Normalization forms, as very briefly described above, can be used to ensure character equivalence by ordering character sequences so that they can be faithfully compared.

It is very important to note that any software applications that is Unicode Standard compliant is free to change the character stream from one representation to another. This means that a software application may compose, decompose or reorder characters as its developers desire; as long as the resultant strings are canonically equivalent to the original. This might lead to unexpected behaviour for users. Various players, like the Unicode Consortium, the W3C, or the TEI recommend NFC in most user-directed situations, and some software applications that we tested indeed seem to automatically convert strings into NFC.\footnote{See the summary of various recommendation here: http://www.win.tue.nl/\textasciitilde{}aeb/linux/uc/nfc\_vs\_nfd.html. ) This means in practice that if a user, for example, enters <a> and <`>, i.e.~LATIN SMALL LETTER A at U+0061 and COMBINING GRAVE ACCENT at U+0300, this might be converted into <à>, i.e.~LATIN SMALL LETTER A WITH GRAVE at U+00E0.\footnote{The behaviour of software applications can be quite erratic in this respect. For example, Apple's TextEdit does not do any conversion on text entry. However, when you copy and paste some text inside the same document in ``rich text mode'' (i.e.~RTF-format), it will be transformed into NFC on paste. Saving a document does not do any conversion to the glyphs on screen, but it will save the characters in NFC.)

\subsection{Pitfall: Characters are not glyphs}\label{pitfall-characters-are-not-glyphs}

A central principle of Unicode is the distinction between character and glyph. A character is the abstract notion of a symbol in a writing system, while a glyph is the concrete drawing of such a symbol. In practice, there is a complex interaction between characters and glyphs. A single Unicode character may of course be rendered as a single glyph. However, a character may also be a `piece' of a glyph, or vice-versa. Actually, all possible relations between glyphs and characters are attested.

First, \textsc{a single character may have different contextually determined glyphs.} For example, characters in writing systems like Hebrew and Arabic have different glyphs depending on where they appear in a word. Some letters in Hebrew change their form at the end of the word, and in Arabic, primary letters have four contextually-sensitive variants (isolated, word initial, medial and final). Second, \textsc{a single character may be rendered as a sequence of multiple glyphs.} For example, in Tamil one Unicode character may result in a combination of a consonant and vowel, which are rendered as two adjacent glyphs by fonts that supports Tamil. Third, \textsc{a single glyph may be a combination of multiple characters. }For example, the ligature <ﬁ>, a single glyph, is the result of two characters, <f> and <i>, that have undergone glyph substitution by font rendering. Like contextually-determined glyphs, ligatures are (intended) artifacts of text processing instructions. Finally, \textsc{a single glyph may be a part of a character} (e.g.~diacritics).

Further, the rendering of a glyph is also always dependent on the font being used. For example, the Unicode character LATIN SMALL LETTER G appears as and in the Computer Modern and Courier fonts because their typefaces are designed differently. Furthermore, font face may change the visual appearance of a character, for example Times New Roman two-story <a> changes to a single-story a in italics <\emph{a}>. This becomes a real problem for some phonetic typesetting (see Section 5, Pitfall 2).

In sum, character-to-glyph mappings are complex technological issues that the Unicode Consortium has had to address in the development of the Unicode Standard, but for the lay user they can be utterly confusing because visual rendering does not indicate logical encoding, or, in other words, what you see isn't (necessarily) what you get.

\subsection{Pitfall: File formats}\label{pitfall-file-formats}

Unicode is a character encoding standard, but these characters of course actually appear inside some kind of file. The most basic Unicode-based file format is pure line-based text, i.e.~strings of Unicode-encoded characters separated by line breaks (note that these line breaks are what for most people intuitively corresponds to paragraph breaks). Unfortunately, even within this apparently basic setting there exist a multitude of variants. In general, these different possibilities are well-understood in the software industry, and nowadays they normally do not lead to any problems for the end user. However, there are some situations in which a user is suddenly confronted with cryptic questions in the user interface involving abbreviations like LF, CR, BE, LE or BOM. Most prominently this occurs with exporting or importing data in several software applications from Microsoft. Basically, there are two different issues involved. First, the encoding of** line breaks** and, second, the encoding of the Unicode characters into \textsc{code units} and the related issue of \textsc{endianness}.

The issue with \textsc{line breaks} originated with the instructions necessary to direct a printing head to a new line. This involves two movements, known as \textsc{carriage return (CR)} (returning the printing head to the start of the line on the page) and \textsc{line feed (LF)} (moving the printing head to the next line on the page). Physically, these are two different events, but conceptually together they form one action. In the history of computing, various encodings of line breaks have been used (e.g.~CR+LF, LF+CR, only LF, or only CR). Currently, all Unix and Unix-derived systems use only LF as code for a line break, while software from Microsoft still uses a combination of CR and LF. Today, most software applications recognize both options, and are able to deal with either encoding of line breaks (until rather recently this was not the case, and using the wrong line breaks could lead to unexpected errors). Our impression is that there is a strong tendency in software development to standardize on the simpler ``only LF'' encoding for line breaks, and we suggest that everybody use this encoding whenever possible.

A code unit is a sequence of bits used to encode character in a encoding. The issue with code units stems from the question how to separate a stream of binary ones and zero, i.e.~bits, into chunks representing Unicode characters. Depending on different use cases, the Unicode Standard offers three different approaches, called UTF-32, UTF-16 and UTF-8.\footnote{The letters UTF stand for either ``Unicode Transformation Format'' or ``Universal Coded Character Set Transformation Format'', but the notion of `transformation' is a legacy notion that does not have meaning anymore. Nevertheless, the designation ``UTF'' (in capitals) has become an official standard designation, but should probably best be read as simply ``Unicode Format''.) The details of this issue is extensively explained in section 2.5 of the Unicode core specification. Basically, \textsc{UTF-32} encodes each character in 32 bits (32 `binary units', i.e.~32 zeros or ones) and is the most disk-space-consuming variant of the three. However, it is the most efficient encoding processing-wise, because the computer simply has to separate each character after 32 bits. In contrast, \textsc{UTF-16} uses only 16 bits per character, which is sufficient for the large majority of Unicode characters, but not for all of them. A special system of `surrogates' is defined within the Unicode Standard to deal with these additional characters. The effect is a more disk-space efficient encoding (approximately half the size), while adding a limited computational overhead. Finally, \textsc{UTF-8} is a more complex system that dynamically encodes each character with the minimally necessary number of bits, choosing either 8, 16 or 32 bits depending on the character. This represents again a strong reduction in space (particularly due to the high frequency of data using ASCII characters that need only 8 bits) at the expense of even more computation necessary to process such strings. However, because of the ever growing computational power of modern machines, the processing overhead is in most practical situations a non-issue, while saving on space is still useful, particularly for sending texts over the Internet. As an effect, UTF-8 has become the dominant encoding on the World Wide Web. We suggest that everybody uses UTF-8 as their default encoding.

A related problem is a general issue about how to store information in computer memory, which is known as \textsc{endianness}. The details of this issue go beyond the scope of this paper. It suffices to realize that there is a difference between \textsc{big-endian} (BE) storage and \textsc{little-endian} (LE) storage. The Unicode Standard offers a possibility to explicitly indicate what kind of storage is used by starting a file with a so-called `Byte Order Mark' (BOM). However, the Unicode Standard does not require the usage of BOM, preferring other non-Unicode methods to signal to computers which kind of endianness is used. This issue only arises with UTF-32 and UTF-16 encodings, and when using the preferred UTF-8, using a BOM is theoretically possible, but strongly dispreferred according to the Unicode Standard. We suggest that everyone tries to prevent the inclusion of BOM in your data.

Together with the topic of normalisation (NFC vs.~NFD, see Pitfall 5 above), this leads to the following nicely cryptic file-encoding recommendation, namely to preferably use \textsc{UTF-8 in NFC without BOM with LF line breaks}, whenever possible.

\section{5. IPA meets Unicode}\label{ipa-meets-unicode}

\subsection{Introducing the International Phonetic Alphabet (IPA)}\label{introducing-the-international-phonetic-alphabet-ipa}

The International Phonetic Alphabet (IPA) is a common standard in linguistics to transcribe sounds of spoken language into some Latin-based characters (International Phonetic Association 1999). Although IPA is reasonably easily adhered to with pen and paper, it is not trivial to encode IPA characters electronically. Early work addressing the need for a universal computing environment for writing systems and their computational complexity is discussed in Simons 1989. For a long time, linguists (like all other computer users) were limited to ASCII-encoded 7-bit characters, which only includes Latin characters, numbers and some punctuation and symbols. Restricted to these standard character sets that lacked IPA support or other language-specific graphemes that they needed, linguists devised their own solutions (cf.~Bird and Simons 2003). For example, some chose to represent unavailable graphemes with substitutes, e.g.~the combination of to represent . Tech-savvy linguists redefined selected characters from a character encoding by mapping custom made fonts to those code points. However, one linguist's electronic text would not render properly on another linguist's computer without access to the same font. Further, if two character encodings defined two character sets differently, then data could not be reliably and correctly displayed. This is a common example of the non-interoperability of data and data formats.

To alleviate this problem, during the late 1980s, SAMPA (Speech Assessment Methods Phonetic Alphabet) was created to represent IPA symbols with 7-bit printable ASCII character sequences, e.g. for {[}ɸ{]}. Two problems with SAMPA are that (i) it is only a partial encoding of the IPA and (ii) it encodes different languages in separate data tables, instead of a universal alphabet, like IPA. SAMPA tables are derived from phonemes appearing in several European languages that were developed as part of a European Commission-funded project to address technical problems like electronic mail exchange (what is now simply called email). SAMPA is essentially a hack to work around displaying IPA characters, but it provided speech technology and other fields a basis that has been widely adopted and used in code. So, SAMPA was a collection of tables to be compared, instead of a large universal table representing all languages. An extended version of SAMPA, called X-SAMPA, set out to include every symbol in the IPA chart including all diacritics (Wells nd.). X-SAMPA was considered more universally applicable because it consisted of one table that encoded the set of characters that represented phones/segments in IPA across languages. SAMPA and X-SAMPA have been widely used for speech technology and as an encoding system in computational linguistics. Eventually, ASCII-encoding of the IPA became depreciated through the advent of the Unicode Standard. Note however that many popular software packages used for linguistic analyses still require ASCII input, e.g.~RuG/L04\footnote{http://www.let.rug.nl/kleiweg/L04/) and SplitsTree4.\footnote{http://www.splitstree.org/)

Still, there are a few pitfalls to be aware of when using the Unicode Standard to encode IPA. As we have said before, from a linguistic perspective it might look like the Unicode Consortium is making incomprehensible decisions, but it is important to realize that the consortium has tried and is continuing to try to be as consistent as possible across a wide range of use cases, and it does place linguistic traditions above other orthographic possibilities. In general, we strongly suggest linguists not to complain about any decisions in the Unicode Standard, but to try and understand the rationale of the Unicode Consortium (which in our experience is almost always well-conceived) and devise ways to work with any unexpected behaviour. Many of the current problems derive from the fact that the IPA is clearly historically based on the Latin script, but different enough from most other Latin-based writing systems to warrant special attention. This ambivalent status of the IPA glyphs (partly Latin, partly special) is unfortunately also attested in the treatment of IPA in the Unicode Standard. In retrospect, it might have been better to consider the IPA (and other transcription systems) to be a special or new kind of script within the Unicode Standard, and treat the obvious similarity to Latin glyphs as a historical relic. All IPA glyphs would then have their own code points, instead of the current situation in which some IPA glyphs have special code points, while others are treated as being identical to the `regular' Latin characters. Yet, the current situation, however unfortunate, is unlikely to change, so as linguists we will have to learn to deal with the specific pitfalls of IPA within the Unicode Standard. In this section, we will describe these pitfalls in some detail.

\subsection{Pitfall: There is no single complete Unicode code block for IPA}\label{pitfall-there-is-no-single-complete-unicode-code-block-for-ipa}

The ambivalent nature of IPA glyphs arises because, on the one hand, the IPA uses Latin-based glyphs like , or . From this perspective, the IPA seems to be just another orthographic tradition using Latin characters, all of which do not get a special treatment within the Unicode Standard (just like e.g.~the French, German, or Danish orthographic traditions do not have a special status). On the other hand, the IPA uses many special symbols (like turned , mirrored and/or extended Latin glyphs ) not found in any other Latin-based writing system. For this reason, and already in the first version of the Unicode Standard (Version 1.0 from 1991), a special block with code points, called ``IPA Extensions'' was included. As noted in Section 4, Pitfall 1, the Unicode Standard code space is subdivided into character blocks, which generally encode characters from a single script. However, as is illustrated by the IPA, characters that form a single writing system may be dispersed across several different character blocks. With its diverse collection of symbols from various scripts and diacritics, the IPA is spread across over 13 blocks in the Unicode Standard:\footnote{This number of blocks depends on whether only IPA-sanctioned symbols are counted or if the phonetic symbols commonly found in the literature are also included, see Moran 2012, Appendix C.)
\begin{itemize}
	\item Basic Latin (30 characters), e.g. <a, b, c, d, e> 
	\item Latin-1 Supplement (4 characters): <æ, ç, ð, ø\textgreater{} 
	\item Latin Extended-A (3 characters): <ħ, ŋ, œ> 
	\item Latin Extended-B (5 characters): <ǀ, ǁ, ǂ, ǃ, ȵ> 
	\item IPA Extensions (70 characters), e.g.: <ɐ, ɑ, ɔ> 
	\item Spacing Modifier Letters (20 characters), e.g.: <ʰ ʷ ˥> 
	\item Combining Diacritical Marks (33 characters), e.g.: <̝ ̥ ̪> 
	\item Greek and Coptic (3 characters): <β, θ, χ> 
	\item Phonetic Extensions (2 characters): <ᴅ, ᴴ> 
	\item Phonetic Extensions Supplement (3 characters): <ᶑ , ᶾ, ᶣ> 
	\item Superscripts and Subscripts (1 character): <ⁿ> 
	\item Arrows (2 characters): \textless{}↑, ↓\textgreater{} 
	\item Latin Extended-C (1 character): <ⱱ> 
\end{itemize}

\subsection{Pitfall: There are many IPA homoglyphs in Unicode}\label{pitfall-there-are-many-ipa-homoglyphs-in-unicode}

Another problem is the large number of homoglyphs, i.e.~different characters that have highly similar glyphs (or even completely identical, depending on the font rendering). For example, a speaker of Russian should ideally not use the CYRILLIC SMALL LETTER A at code point U+0430 for IPA transcriptions, but instead the LATIN SMALL LETTER A at code point U+0061, although visually they are mostly indistinguishable, and the Cyrillic character is more easily typed on a Cyrillic keyboard. Another example we commonly encounter is the use of LATIN SMALL LETTER G at U+0067, instead of the sanctioned Unicode Standard IPA character for the voiced velar stop LATIN SMALL LETTER SCRIPT G at U+0261. One begins to question whether this issue is at all apparent to the working linguist, or if they simply use the U+0067 because it is easily keyboarded and thus saves time, whereas the latter must be cumbersomely inserted as a special symbol in most software.\footnote{This issue was recently addressed by the International Phonetic Association, which has taken the stance that both the keyboard LATIN SMALL LETTER G and the LATIN SMALL LETTER SCRIPT G are valid input characters for the voiced velar plosive. Unfortunately, this decision further introduces ambiguity for linguists trying to adhere to a strict Unicode Standard IPA encoding.)

Furthermore, on the one hand even linguists are unlikely to distinguish between the LATIN SMALL LETTER SCHWA at code point U+0259 and LATIN SMALL LETTER TURNED E at U+01DD. On the other hand, non-linguists are unlikely to distinguish any semantic difference between an open back unrounded vowel , the LATIN SMALL LETTER ALPHA at U+0251, and the open front unrounded vowel , LATIN SMALL LETTER A at U+0061. But even among linguists this distinction leads to problems. For example, as pointed out by Mielke (2009), there is a problem stemming from the fact that about 75\% of languages are reported to have a five-vowel system (Maddieson 1984). Historically, linguistic descriptions tend not to include precise audio recording and measurements of formants, so this may lead one to ask if the many characters that are used in phonological description reflects a ``transcriptional bias''. The common use of in transcriptions could be in part due to the ease of typing the letter on an English keyboard (or for older descriptions, the typewriter). We found it to be exceedingly rare that a linguist uses for a low back unrounded vowel.\footnote{One example is Vidal 2001a:75, in which the author states: ``The definition of Pilagá /a/ as {[}+back{]} results from its behavior in certain phonological contexts. For instance, uvular and pharyngeal consonants only occur around /a/ and /o/. Hence, the characterization of /a/ and /o/ as a natural class of (i.e., {[}+back{]} vowels), as opposed to /i/ and /e/.'') They simply use as long as there is no opposition to .\footnote{See Thomason's Language Log post, ``Why I don't love the International Phonetic Alphabet'', at: http://itre.cis.upenn.edu/\textasciitilde{}myl/languagelog/archives/005287.html.) Making things even more problematic, there is an old typographic tradition that the double-story uses a single-story in italics. This leads to the unfortunate effect that in most well-designed fonts the italics of and use the same glyph. If this distinction has to be kept upright in italics, the only solution we can currently offer is to use `slanted' glyphs (i.e.~artificially italicised glyphs) instead of real italics (i.e.~special italics glyphs designed by a typographer).\footnote{For example, the widely used IPA font Doulos SIL (http://scripts.sil.org/cms/scripts/page.php?item\_id=DoulosSIL) does not have real italics. This leads some word-processing software, like Microsoft Word, to produce slanted glyphs instead. That particular combination of font and software application will thus lead to the desired effect. However, note that when the text is transferred to another font (i.e.~one that includes real italics) and/or to another software application (like Apple Pages, which does not perform slanting), then this visual appearance will be lost. In this case we are thus still in the pre-Unicode situation in which the choice of font and rendering software actually matters. The ideal solution from a linguistic point of view would be the introduction of a new IPA code point for a different kind of which explicitly specifies that it should still be rendered as a double-story character when italicized. After informal discussion with various Unicode players, our impression is that this highly restricted problem is not sufficiently urgent to introduce even more -like characters in Unicode (which already lead to much confusion, see Section 4, Pitfall 4). This is a clear situation in which the Unicode Consortium is not just thinking about linguists, but has a more wide-ranging practical view to consider.)

Some other homoglyphs related to encoding IPA in the Unicode Standard are:
\begin{itemize}
	\item The uses of the apostrophe has led to long discussions on the Unicode Standard email list. An English keyboard inputs \textless{}`\textgreater{} APOSTROPHE at U+0027, although the `preferred' Unicode apostrophe is the \textless{}'\textgreater{} RIGHT SINGLE QUOTATION MARK at U+2019. Yet the glottal stop/glottalization/ejective marker is another completely different character, the MODIFIER LETTER APOSTROPHE at U+02BC, but unfortunately looks mostly highly similar to U+2019. 
	\item There is ambiguous encoding of IPA segments within the Unicode Standard. An example is the U+02C1 MODIFIER LETTER REVERSED GLOTTAL STOP vs the U+02E4 MODIFIER LETTER SMALL REVERSED GLOTTAL STOP . Both are denoted in the Unicode Standard as the `pharyngealized diacritic' and both appear in various resources representing phonetic data online. This is thus an example for which the Unicode Standard does not solve the linguistic standardization problem. 
	\item There is at least one case in which the character name assigned by the Unicode Consortium does not match the IPA's description: in the Unicode Standard at U+01C3 is labeled LATIN LETTER RETROFLEX CLICK, but in IPA is an alveolar or postalveolar click (not retroflex). This naming is probably best seen as a simple error in the Unicode Standard. Note that most linguists simply seem to use or <ou>) while ligatures are single characters (e.g. <ʧ> LATIN SMALL LETTER TESH DIGRAPH at U+02A7). Ligatures arose in the context of printing easier-to-read texts, and are included in the Unicode Standard for reasons of legacy encoding. However, their usage is discouraged by the Unicode core specification. Specifically related to IPA, various phonetic combinations of characters (typically affricates) are available as single code-points in the Unicode Standard, but are designated as ``ligatures'' or ``digraphs'' (confusingly both names appear interchangeably). Such glyphs might be used by software to produce a pleasing display, but they should not be hard-coded into the text itself. In the context of IPA, characters like the following ligatures should thus \emph{not} be used. Instead a combination of two characters is preferred: 
	\item <ʣ> LATIN SMALL LETTER DZ DIGRAPH at U+02A3 (use <d> + <z> instead) 
	\item <ʧ> LATIN SMALL LETTER TESH DIGRAPH at U+02A7 (use <t> + <ʃ> instead) 
	\item <ʩ> LATIN SMALL LETTER FENG DIGRAPH at U+02A9 (use <f> + <ŋ> instead) 
\end{itemize}

However, there are a few Unicode characters that are historically ligatures, but which are today considered as simple characters in the Unicode Standard and thus should be used when writing IPA, namely:
\begin{itemize}
	\item <ɮ> LATIN SMALL LETTER LEZH at U+026E 
	\item <œ> LATIN SMALL LIGATURE OE at U+0153 
	\item <ɶ> LATIN LETTER SMALL CAPITAL OE at U+0276 
	\item <æ> LATIN SMALL LETTER AE at U+00E6 
\end{itemize}

\subsection{Pitfall: The IPA notion of diacritics is not the same as the Unicode Standard's notion of diacritics}\label{pitfall-the-ipa-notion-of-diacritics-is-not-the-same-as-the-unicode-standards-notion-of-diacritics}

Another pitfall is diacritics. The problem is that the meaning of the term `diacritics' as used by the IPA is not the same as it used in the Unicode Standard. Specifically, diacritics in the IPA-sense are either so-called \textsc{Combining Diacritical Marks} or \textsc{Spacing Modifier Letters} in the Unicode Standard. Crucially, Combining Diacritical Marks are by definition combined with the character before them (to form so-called default grapheme clusters, see Section 3). In contrast, Spacing Modifier Letters are by definition \emph{not} combined into grapheme clusters with the preceding character, but simply treated as separate letters. In the context of the IPA, the following IPA-diacritics are actually Spacing Modifier Letters in the Unicode Standard:
\begin{itemize}
	\item Length marks, namely <ː> MODIFIER LETTER TRIANGULAR COLON at U+02D0 and <ˑ> MODIFIER LETTER HALF TRIANGULAR COLON at U+02D1. 
	\item Tone letters, like <˥> MODIFIER LETTER EXTRA-HIGH TONE BAR at U+02E5, and others like this. 
	\item Superscript letters, like <ʰ> MODIFIER LETTER SMALL H at U+02B0 or <ˤ> MODIFIER LETTER SMALL REVERSED GLOTTAL STOP at U+02E4, and others like this. 
	\item <˞> MODIFIER LETTER RHOTIC HOOK at U+02DE. 
\end{itemize}

Although linguists might expect these characters to belong together with the character in front of them, at least for <ʰ> MODIFIER LETTER SMALL H at U+02B0 the Unicode Consortium's decision to treat it as a separate character is also linguistically correct, because according to the IPA it can be used both for aspiration (more precisely postaspiration following the base character) and preaspiration (preceding the base character). Note that there is a mechanism in Unicode to force separate characters to be combined (namely by using the ZERO WIDTH JOINER at U+200D), but this seems to be a rather impractical, and probably not enforceable solution to us.

\subsection{Pitfall: Neither the IPA nor the Unicode Standard enforce a unique diacritic ordering}\label{pitfall-neither-the-ipa-nor-the-unicode-standard-enforce-a-unique-diacritic-ordering}

Also related to diacritics is the question of ordering. To our knowledge, the International Phonetic Association does not specify a specific ordering for diacritics that combine with phonetic base symbols; this exercise is left to the reasoning of the transcriber. However, such marks have to be explicitly ordered if sequences of them are to be interoperable and compatible. An example is a labialized aspirated alveolar plosive: <tʷʰ>. There is nothing holding linguists back from using <tʰʷ> instead (with exactly the same intended meaning). However, from a technical standpoint, these two sequences are different, e.g.~if both sequences are used in a document, searching for <tʷʰ> will not find any instances of <tʰʷ>, and vice versa. Likewise, a creaky voiced syllabic dental nasal can be encoded in various orders, e.g. <n̪̰̩>, <n̩̰̪> or <n̩̪̰>.

In accordance with the absence of any specification of ordering in the IPA, the Unicode Standard likewise does not propose any standard orderings. Both leave it to the user to be consistent. However, there is one aspect of ordering for which the Unicode Standard does present a canonical solution, which is uncontroversial from a linguistic perspective. Diacritics in the Unicode Standard (i.e.~Combining Diacritical Marks, see above) are classified in Canonical Combining Classes. In practice, the diacritics are distinguished by their position relative to the base character.\footnote{See http://unicode.org/reports/tr44/\#Canonical\_Combining\_Class\_Values for a detailed description.) When applying a Unicode normalization (NFC or NFD, see previous section), the diacritics in different positions are put in a specified order. This process therefore harmonizes the difference between different encodings, for example, of a midtone creaky voice vowel <ḛ̄>. This grapheme cluster can be encoded either as <e> + <̄> + <̰> or as <e> + <̰> + <̄> . To prevent this twofold encoding, the Unicode Standard specifies the second ordering as canonical (diacritics below before diacritics above).

When encoding a string according to the Unicode Standard, it is possible to do this either using the NFC (``composition'') or NFD (``decomposition'') normalization. Decomposition implies that precomposed characters (like <á> LATIN SMALL LETTER A WITH ACUTE at U+00E1) will be split into its parts. This might sound preferable for a linguistic analysis, as the different diacritics are separated from the base characters. However, note that most attached elements like strokes (e.g.~in the <ɨ>), retroflex hooks (e.g.~in <ʐ>) or rhotic hooks (e.g.~in <ɝ>) will not be decomposed, but strangely enough a cedilla (like in <ç>) will be decomposed. In general, Unicode decomposition does not behave like a feature decomposition as expected from a linguistic perspective. It is thus important to consider Unicode decomposition only as a technical procedure, and not assume that it is linguistically sensible.

Facing the problem of specifying a consistent ordering of diacritics while developing a large database of phonological inventories from the world's languages, Moran (2012: 540) defines a set of diacritic ordering conventions.\footnote{The most recent version of these conventions is online: http://phoible.github.io/conventions/) The conventions are influenced by the linguistic literature, though some ad-hoc decisions had to be taken. The goal was to explicitly define all character sequences so that the vast variety of phonemes found in descriptions of the world's language were normalized into consistent character sequences, e.g.~if one language description uses and another , when both are intended to be phonetically equivalent, then a decision to normalize to one form was taken. For example, when a character sequence contains more than one character in Spacing Modifier Letters, the order that is proposed is the following (where <\textbar{}> indicates ``or'' and <→> indicates ``precedes''):
\begin{itemize}
	\itemsep1pt\parskip0pt\parsep0pt 
	\item \textsc{Spacing Modifier Letters ordering:} ( unreleased <̚> \textbar{} lateral release <ˡ> \textbar{} nasal release <ⁿ>) → ( palatalized <ʲ>) → ( labialized <ʷ>) → ( velarized <ˠ>) → ( pharyngealized <ˤ>) → ( aspirated <ʰ> \textbar{} ejective <ʼ>) → ( long <ː> \textbar{} half-long <ˑ>) 
\end{itemize}

If a character sequence contains more than one diacritic below the base character, then the place feature is applied first (dental, laminal, apical, fronted, backed, lowered, raised), then the laryngeal setting (voiced, voiceless, creaky voice, breathy voice) and finally the syllabic or non-syllabic marker (for vowels, ATR gets put on between the place and laryngeal setting). So:
\begin{itemize}
	\itemsep1pt\parskip0pt\parsep0pt 
	\item \textsc{Combining Diacritical Marks (below) ordering:} ( dental <t̪> \textbar{} laminal <t̻> \textbar{} apical <t̺>) → ( fronted <u̟> \textbar{} backed <e̠>) →( lowered <e̞> \textbar{} raised <e̝>) → ( ATR <e̘ e̙>) → ( voiced <s̬> \textbar{} voiceless <n̥> \textbar{} creaky voice <b̰> \textbar{} breathy voice <b̤>) → ( syllabic <n̩> \textbar{} non-syllabic <e̯>) 
\end{itemize}

Character sequences with diacritics above the base character were not problematic in Moran 2012 because they include only the centralized, mid-centralized and nasalized combining characters. Moran (2012) marks tones as singletons with Space Modifier Letters, e.g. \textless{}˦\textgreater{} for a phonemic high tone, instead of accent diacritics, alleviating potential conflicts. Building on the work of Moran (2012), if a character sequence contains more than one diacritic above the base character, we propose:
\begin{itemize}
	\itemsep1pt\parskip0pt\parsep0pt 
	\item \textsc{Combining Diacritical Marks (above) ordering:} (centralized <ë> \textbar{} mid-centralized <e̽>) → (extra short <ĕ>) →( tone accents, e.g. <è> ) → ( Spacing Modifier Letters ) → ( tone letters, e.g. <e˦>) 
\end{itemize}

\section{6. Orthography profiles}\label{orthography-profiles}

\subsection{Characterizing writing systems}\label{characterizing-writing-systems}

At this point in the course of rapid ongoing developments, we are left with a situation in which the Unicode Standard offers a highly detailed and flexible approach to deal computationally with writing systems, but it has unfortunately not influenced the linguistic practice very much. In many practical situations, the Unicode Standard is far too complex for the day-to-day practice in linguistics because it does not offer practical solutions for the down-to-earth problems of many linguists. In this section, we propose some simple practical guidelines and methods to improve on this situation.

Our central aims for linguistics, to be approached with a Unicode-based solution, are: (i) to improve the consistency of the encoding of sources, (ii) to transparently document knowledge about the writing system (including transliteration), and (iii) to do all of that in a way that is easy and quick to manage for many different sources with many different writing systems. The central concept in our proposal is the \textsc{orthography profile}, a simple tab-separated CSV text file, that characterizes and documents a writing system. We also offer basic implementations in Python and R to assist with the production of such files, and to apply orthography profiles for consistency testing, grapheme tokenization and transliteration. Not only can orthography profiles be helpful in the daily practice of linguistics, they also succinctly document the orthographic details of a specific source, and, as such, might fruitfully be published alongside sources (e.g.~in digital archives). Also, in high-level linguistic analyses in which the graphemic detail is of central importance (e.g.~phonotactic or comparative-historical studies), orthography profiles can transparently document the decisions that have been taken in the interpretation of the orthography in the sources used.

Given these goals, Unicode locale descriptions (see Section 3) might seem like the ideal orthography profiles. However, there are various practical obstacles preventing the use of such locale descriptions in the daily linguistic practice, namely: (i) the XML-structure is too verbose to easily and quickly produce or correct manually, (ii) locale descriptions are designed for a wide scope on information (like date formats or names of weekdays) most of which is not applicable for documenting writing systems, and (iii) most crucially, even if someone made the effort to produce a technically correct locale description for a specific source at hand, then it is nigh impossible to deploy the description. This is because a locale description has to be submitted to and accepted by the Unicode Common Locale Data Repository. The repository is (rightly so) not interested in descriptions that only apply to a limited set of sources (e.g.~only a single specific dictionary).

The major challenge then is developing an infrastructure to identify the elements that are individual graphemes in a source, specifically for the enormous variety of sources using some kind of alphabetic writing system. Authors of source documents (e.g.~dictionaries, wordlists, corpora) use a variety of writing systems that range from their own idiosyncratic transcriptions to already well-established practical or longstanding orthographies. Although the IPA is one practical choice as a sound-based normalization for writing systems (which can act as an interlingual pivot to attain interoperability across writing systems), graphemes in each writing system must also be identified and standardized if interoperability across different sources is to be achieved. In most cases, this amounts to more than simply mapping a grapheme to an IPA segment because graphemes must first be identified in context (e.g.~is the sequence one sound or two sounds or both?) and strings must be tokenized, which may include taking orthographic rules into account (e.g. between vowels is /n/ and after a vowel but before a consonant is a nasalized vowel /ṽ/). In our experience, data from each source must be individually tokenized into graphemes so that its orthographic structure is identified and its contents can be extracted. To extract data for analysis, a source-by-source approach is required before an orthography profile can be created. For example, almost each available lexicon on the world's languages is idiosyncratic in its orthography and thus requires lexicon-specific approaches to identify graphemes in the writing system and to map graphemes to phonemes, if desired.

Thus, our key proposal for the characterization of a writing system is to use a grapheme tokenization as an inter-orthographic pivot. Basically, any source document is tokenized by graphemes, and only then a mapping to IPA (or any other orthographic conversion) is performed. An orthography profile then is a description of the units and rules that are needed to adequately model a graphemic tokenization for a language variety as described in a particular source document. An orthography profile summarizes the Unicode (tailored) graphemes and orthographic rules used to write a language (the details of the structure and assumptions of such a profile will be presented in the next section).

As an example of graphemic tokenization, note the three different levels of technological and linguistic elements that interact in the hypothetical lexical form <tsʰṍ̰shi> :
\begin{enumerate}
	\def\labelenumi{\arabic{enumi}.} 
	\item code points (10 text elements): t s ʰ o ̃ ̰ ´ s h i 
	\item grapheme clusters (7 text elements): t s ʰ ṍ̰ s h i 
	\item tailored grapheme clusters (4 text elements): tsʰ ṍ̰ sh i 
\end{enumerate}

In (1), the string <tsʰṍ̰shi> has been tokenized into ten Unicode code points (using NFD normalization), delimited here by space. Unicode tokenization is required because sequences of code points can differ in their visual and logical orders. For example, <õ̰> is ambiguous to whether it is the sequence of + <̰> + <̃> or + <̃> + <̰>. Although these two variants are visually homoglyphs, computationally they are different. Unicode normalization should be applied to this string to reorder the code points into a canonical order, allowing the data to be treated canonically equivalently for search and comparison. In (2), the Unicode code points have been logically normalized and visually organized into grapheme clusters, as specified by the Unicode Standard. The combining character sequence <õ̰> is normalized and visually grouped together. Note that , the MODIFIER LETTER SMALL H at U+02B0, is not grouped with . This is because it belongs to Spacing Modifier Letters category in the Unicode Standard. These characters are underspecified for the direction in which they modify a host character. For example, can indicate either pre- or post-aspiration (whereas the nasalization or creaky diacritic is defined in the Unicode Standard to apply to a specified base character). Finally, to arrive at the graphemic tokenization in (3), tailored grapheme clusters are needed (as for example specified in an orthography profile). For example, this orthography profile would specify that the sequence of characters , and form a single grapheme , and that and form a grapheme . The orthography profile could also specify orthographic rules, e.g.~when tokenization graphemes, in say English, the in the forms and should be treated as distinct sequences depending on their contexts.

\subsection{Informal description of orthography profiles}\label{informal-description-of-orthography-profiles}

An orthography profile describes the Unicode code points, characters, graphemes and orthographic rules in a writing system. An orthography profile is a language-specific (and often even resource-specific) description of the units and rules that are needed to adequately model a writing system. An important assumption of our work is that we assume a resource is encoded in Unicode (or has been converted to Unicode). Any data source that the Unicode Standard is unable to capture, will also not be captured by an orthography profile.

Informally, an orthography profile specifies the graphemes (or, in Unicode parlance, ``tailored grapheme clusters'') that are expected to occur in any data to be analysed or checked for consistency. These graphemes are first identified throughout the whole data (a step which we call `tokenization'), and possibly simply returned as such, possibly including error messages about any parts of the data that are not specified by the orthography profile. Once the graphemes are identified, they might also be changed into other graphemes (a step which we call `transliteration'). When a grapheme has different possible transliterations, then these differences should be separated by contextual specification, possibly down to listing individual exceptional cases.

In practice, we foresee a workflow in which orthography profiles are iteratively refined, while at the same time inconsistencies and errors in the data to be tokenized are corrected. In some more complex use-cases there might even be a need for multiple different orthography profiles to be applied in sequence (see Section 8 on various exemplary use-cases). The result of any such workflow will normally be a cleaned dataset and an explicit description of the orthographic structure in the form of an orthography profile. Subsequently, the orthography profiles can be easily distributed in scholarly channels alongside the cleaned data, for example in supplementary material added to journal papers or in electronic archives.

\subsection{Formal specification of orthography profiles}\label{formal-specification-of-orthography-profiles}

The formal specifications of an orthography profile (or simply `profile' for short) are the following:
\begin{enumerate}
	\def\labelenumi{\arabic{enumi}.} 
	\item \textsc{A profile is a} \textsc{Unicode UTF-8 encoded text file} (ideally using NFC, no-BOM, and LF; see Section 4, Pitfall ``File Formats'') that includes the information pertinent to the orthography. 
	\item \textsc{A profile is a} \textsc{tab-separated CSV file with an obligatory header line}. A minimal profile can have just a single column, in which case there will of course no tabs, but the first line will still be the header. For all columns we assume the name in the header of the CSV file to be crucial. The actual ordering of the columns is unimportant. 
	\item \textsc{Lines starting with a hash \textless{}\#\textgreater{} are ignored.} Comments and metadata can be included inside the file, but only as complete lines in the profile, to be marked by lines starting with hash ``\#'' (NUMBER SIGN at U+0023). Hashes somewhere else in the file are to be treated literally, i.e.~hashes are only to be ignored when they occur at the start of a line.\footnote{Comments that belong to specific lines will have to be put in a separate column of the CSV file, e.g.~add a column called `comments'. Further, if the content of a profile contains a hash at the start of a line, either reorder the columns so the hash does not occur at the start of the line, or add a dummy column in front of the data to not have the data start with a hash.) 
	\item \textsc{Metadata are given in commented lines at the beginning of the text file in a basic ``tag: value'' format. }Metadata about the orthographic description given in the orthography profile includes, minimally, (i) author, (ii) date, (iii) title, (iv) a stable language identifier encoded in BCP 47/ISO 639-3, and (v) bibliographic data for resource(s) that illustrate the orthography described in the profile. 
\end{enumerate}

The content of a profile consists of lines, each describing a grapheme of the orthography, using the following columns:
\begin{enumerate}
	\def\labelenumi{\arabic{enumi}.} 
	\item \textsc{A minimal profile consists of a single column with header `graphemes'}, listing each of the different graphemes in a separate line. 
	\item \textsc{Optional columns called `left' and `right' can be used to specify the left and right context of the grapheme, respectively.} The same grapheme can occur multiple times with different contextual specifications, for example to distinguish different pronunciations depending on the context. 
	\item \textsc{The columns `grapheme', `left' and `right' can use regular expression metacharacters.} If regular expressions are used, then all literal usage of the special symbols, like full stops ``.'' or dollar signs ``\$'' (so-called ``metacharacters'') have to be explicitly escaped by adding a backslash before them (i.e.~use ``.'' or ``\$''). Note that any specification of context automatically expects regular expressions, so it is probably better to always escape all regular expression metacharacters when used literally in the orthography, i.e.~the following symbols will need to be preceded by a backslash: {[} {]} ( ) \{ \} ~+ * . - ! ? \^{} \$ . 
	\item \textsc{An optional column called `class' can be used to specify classes of graphemes}, for example to define a class of vowels. Users can simply add ad-hoc identifiers in this column to indicate a group of graphemes, which can then be used in the description of the graphemes or the context. The identifiers should of course be chosen such that they do not conflate with any symbols used in the orthography themselves. Note that such classes only refer to the graphemes, not to the context. 
	\item \textsc{Columns describing transliterations for each graphemes can be added and named at will}. Often more than a single possible transliteration will be of interest. Any software application using these profiles should use the names of these columns to select a specific transliteration column. 
	\item \textsc{Any other columns can be added freely, but will mostly be ignored by any software application using the profiles}. As orthography profiles are also intended to be read and interpreted by humans, it is often highly useful to add extra information on the graphemes in further columns, like for example Unicode codepoints, Unicode names, frequency of occurrence, examples of occurrence, explanation of the contextual restrictions, or comments. 
\end{enumerate}

For the automatic processing of the profiles, the following technical standards will be expected:
\begin{enumerate}
	\def\labelenumi{\arabic{enumi}.} 
	\item \textsc{Each line of a profile will be interpreted as a regular expression. }Software applications using profiles can also offer to interpret a profile in the literal sense to avoid the necessity for the user to escape regular expressions metacharacters in the profile. However, this only is possible when no contexts or classes are described, so this seems only useful in the most basic orthographies. 
	\item \textsc{The `class' column will be used to produce explicit ``or'' chains of regular expressions}, which will then be inserted in the `graphemes', `left' and `right' columns at the position indicated by the class-identifiers. For example, a class ``V'' as a context specification might be replaced by a regular expression like ``(a\textbar{}e\textbar{}i\textbar{}o\textbar{}u\textbar{}ei\textbar{}au)''. Only the graphemes themselves are included here, not any contexts specified for the elements of the class. 
	\item \textsc{The `left' and `right' contexts will be included into the regular expressions by using lookbehind and lookahead}. Basically, the actual regular expression syntax of lookbehind and lookahead is simply hidden to the users by allowing them to only specify the contexts themselves. Internally, the contexts in the columns ``left'' and ``right'' are combined with the column ``graphemes'' to form a complex regular expression like ``(?\textless{}=left)graphemes(?=right)''. 
	\item \textsc{The regular expressions will be applied in the order as specified in the profile, from top to bottom.} A software implementation can offer help in figuring out the optimal ordering of the regular expressions, but should then explicitly report on the order used. 
\end{enumerate}

The actual implementation of the profile on some text-string will function as follows:
\begin{enumerate}
	\def\labelenumi{\arabic{enumi}.} 
	\item \textsc{All graphemes are matched in the text before they are tokenized or transliterated}. In this way, there is no necessity for the user to consider `feeding' and `bleeding' situations, in which the application of a rule either changes the text so another rule suddenly applies (``feeding'') or prevents another rule to apply (`bleeding'). 
	\item \textsc{The matching of the graphemes can occur either globally or linearly. }From a computer science perspective, the most natural way to match graphemes from a profile in some text is by walking linearly through the text-string from left to right, and at each position go through all graphemes in the profile to see which one matches, then go to the position at the end of the matched grapheme and start over. This is basically how a finite state transducer works, which is a well-established technique in computer science. However, from a linguistic point of view, our experience is that most linguists find it more natural to think from a global perspective. In this approach, the first grapheme in the profile is matched everywhere in the text-string first, before moving to the next grapheme in the profile. Theoretically, these approaches will lead to different results, though in practice of actual natural language orthographies they almost always lead to the same result. Still, we suggest that any software application using orthography profiles should offer both approaches (i.e. ``global'' or ``linear'') to the user. The approach used should be documented in the metadata as ``tokenization method''. 
	\item \textsc{The matching of the graphemes can occur either in NFC or NFD. }By default, both the profile and the text-string to be tokenized should be treated as NFC (see section 4, Pitfall ``Canonical equivalence'' above). However, in some use-cases it turns out to be practical to treat both text and profile as NFD. This typically happens when very many different combinations of diacritics occur in the data. An NFD-profile can then be used to first check which individual diacritics are used, before turning to the more cumbersome inspection of all combinations. We suggest that any software application using orthography profiles should offer both approaches (i.e. ``NFC'' or ``NFD'') to the user. The approach used should be documented in the metadata as ``unicode normalization''. 
	\item \textsc{The text-string is always returned in tokenized form} by separating the matched graphemes by a user-specified symbols-string. Any transliteration will be returned on top of the tokenization. 
	\item \textsc{Leftover characters (i.e.~characters that are not matched by the profile) should be reported to the user as errors.} Typically, the unmatched character are replaced in the tokenization by a user-specified symbol-string. 
\end{enumerate}

Any software application offering to use orthography profile:
\begin{enumerate}
	\def\labelenumi{\arabic{enumi}.} 
	\item \textsc{should offer user-options} to specify:
	\begin{enumerate}
		\def\labelenumii{\arabic{enumii}.} 
		\item \textsc{the name of the column to be used for transliteration} (if any). 
		\item \textsc{the symbol-string to be inserted between graphemes.} Optionally, a warning might be given if the chosen string includes characters from the orthography itself. 
		\item \textsc{the symbol-string to be inserted for unmatched strings} in the tokenized and transliterated output. 
		\item \textsc{the tokenization method}, i.e.~whether the tokenization should proceed ``global'' or ``linear''. 
		\item \textsc{unicode normalization}, i.e.~whether the text-string and profile should use ``NFC'' or ``NFD''. 
	\end{enumerate}
	\item \textsc{might offer user-options }to:
	\begin{enumerate}
		\def\labelenumii{\arabic{enumii}.} \setcounter{enumii}{5} 
		\item \textsc{assist in the ordering of the graphemes.} In our experience, it makes sense to apply larger graphemes before shorter graphemes, and to apply graphemes with context before graphemes without context. Further, frequently relevant rules might be applied after rarely relevant rules (though frequency is difficult to establish in practice, as it depends on the available data). Also, if this all fails to give any decisive ordering between rules, it seems useful to offer linguists the option to reverse the ordering from any manual specified ordering, because linguists tend to write the more general rule first, before turning to exceptions or special cases. 
		\item \textsc{assist in dealing with upper and lower case characters.} It seems practical to offer some basic case matching, so characters like <a> and <A> are treated equally. This will be useful in many concrete cases, although the user should be warned that case matching does not function universally in the same way across orthographies. Ideally, users should prepare orthography profiles with all lowercase and uppercase variants explicitly mentioned, so by default no case matching should be performed. 
		\item \textsc{treat the profile literal}, i.e.~to not interpret regular expression metacharacters. Matching graphemes literally often leads to strong speed increase, and would allow users to not needing to worry about escaping metacharacters. However, in our experience all actually interesting use-cases of orthography profiles include some contexts, which automatically prevents any literal interpretation, so by default the matching should not be literal. 
	\end{enumerate}
	\item \textsc{should return the following information} to the user:
	\begin{enumerate}
		\def\labelenumii{\arabic{enumii}.} \setcounter{enumii}{8} 
		\item \textsc{the original text-strings to be processed in the used Unicode normalization}, i.e.~in either NFC or NFD as specified by the user. 
		\item \textsc{the tokenized strings}, with additionally any transliterated strings, if transliteration is requested. 
		\item \textsc{a survey of all errors encountered}, ideally both in which text-strings any errors occurred and which characters in the text-strings lead to errors. 
		\item \textsc{a reordered profile}, when any automatic reordering is offered 
	\end{enumerate}
\end{enumerate}

\section{7. Use cases}\label{use-cases}

We now present several use cases that have motivated the development of orthography profiles. These include:
\begin{itemize}
	\item grapheme tokenization 
	\item normalization of orthographic systems to attain interoperability across different source documents, 
	\item cognate identification for detecting the similarity between words from different languages 
\end{itemize}

An important assumption of our work is that input sources are encoded in Unicode (UTF-8 to be precise). Anything that the Unicode Standard is unable to capture, cannot be captured by an orthography profile. We also use Unicode normalization form NFD to decompose all incoming text input into normalized and (Unicode) logically ordered strings. This process is of fundamental importance when working with text data in the Unicode Standard.

\emph{Grapheme tokenization}

Once text has been normalized, an additional and straightforward process is Unicode grapheme tokenization. More recently this regular expression, often available as short cut ``\X'', identifies all sequences of ``base'' characters followed by 0 or more combining diacritics. For example, the sequence, <ŋ͡mṍraː> would tokenize as a space-delimited sequence as: <ŋ͡ m ṍ r a ː>. For a first pass at identifying grapheme clusters, this is a straightforward tokenization. However, the linguist will note that both the tie bar (COMBINING DOUBLE INVERTED BREVE, a Unicode Combining Modifier) and the length marker (MODIFIER LETTER TRIANGULAR COLON. a Unicode Letter Modifier) do not appear ``correctly'', i.e.~the tie bar is grouped with the first character in the sequence and the length marker appears singly. This is necessary, as defined by the Unicode Standard, because of these character's semantic properties -- with Unicode we cannot know if, for example, the MODIFIER LETTER TRIANGULAR COLON should appear before or after the base character that it modifies, cf.~the IPA aspiration marker, , the MODIFIER LETTER SMALL H, which is also a Unicode Letter Modifier and for linguists a diacritic that be be used for pre- or post-aspiration. These ambiguities of position in tokenization are not uncommon in IPA, thus orthographic (or source) normalization and tokenization is needed.

*Cross-orthography **normalization*

Orthographic tokenization requires a specification of the tailored grapheme clusters (linguists: graphemes) to separate orthographic units within an input source. In the simplest case, the Unicode graphemic tokenization with regular expression ``\X'' captures the grapheme clusters, i.e.~the input data is graphemically tokenized and there are no additional string sequences required to capture each grapheme, e.g.~no specification. This scenario is the most unlikely.

The creation of an orthography typically requires that someone describes knowledge about the input data source's writing system. For example, when orthographically tokenizing a dictionary as input, the sequences of grapheme clusters need to be explicitly stated or a regular expression rule can be specified to transform the input. We illustrate these methods with examples from Thiesen 1998, a bilingual dictionary of Diccionario Bora-Castellano, and with Huber \& Huber 1992, a comparative vocabulary of sixty-nine indigenous languages of Columbia.

Huber \& Huber 1992

({[}a\textbar{}á\textbar{}e\textbar{}é\textbar{}i\textbar{}í\textbar{}o\textbar{}ó\textbar{}u\textbar{}ú{]})(n)(\s)({[}a\textbar{}á\textbar{}e\textbar{}é\textbar{}i\textbar{}í\textbar{}o\textbar{}ó\textbar{}u\textbar{}ú{]}), \1 \2 \4

A preliminary orthography profile can be generated with code available online.\footnote{https://github.com/bambooforest/tokenizer) The Python script creates a unigram model for individual Unicode characters or Unicode grapheme clusters and their frequency in the source input. We find that this gives a broad overview of the input at the character and grapheme levels and it oftens finds typographic errors in the source input by identifying rarely occurring or one-off characters in the input. The models also provides the basis for the ``quick'' development of an initial orthography profile by saving the user keyboarding time.

Note that these symbols are still phonetically unspecified, i.e.~symbol may or may not represent, say, a front open unrounded vowel.

The orthography profile is designed so that specified strings of characters are captured for tokenization.

\% how the tokenization works

The graphemic cluster specifications in the orthography profile are read into a trie data structure, which is used to tokenize the incoming input source by identifying sentinels (tokenization boundaries) in the input stream, which we then separate grapheme clusters by spaces.

An illustration is given in Figure N.

Comparative wordlists often start as the compilation of different sources with different writing systems. The Dogon comparative wordlist is an example of a compilation of slightly divergent transcription practices, which we will use to illustrate our point and to orthographically normalize an input source by transforming different transcriptions into a single IPA-like transcription.

*Orthographic normalization / cross-source **normalization*

Once we have normalized and tokenized our text elements, we can apply various sorts of source normalization across resources to create interoperable and comparable data.

Sound-based normalization is practical because \ldots{}

We use orthography profiles in our work to describe cross-linguistic difference between writing systems.

\emph{Cognate detection}

The identification of cognates is a non-trivial task, which involves identifying the semantic and phonetic similarity of words.

use grapheme correspondences to approximate cognacy and sound correspondences

to identify cognates and sound changes

Examples

Orthographic similarity != Phonetic similarity != Regular sound change

Cognate identification is further complicated by the fact that languages are written differently, and even more so for the low resource languages that we work with, linguists' phonetic transcriptions are heterogeneous and typically document-specific, i.e.~two linguists working on the same language variety most often develop different practical orthographies by using different symbols and different orthographic rules. Thus the essence of language documentation and description is diverse and ephemeral in nature. The encoding of linguistic data and linguistic data models is diverse and abstracting away from individual analytical preferences requires a document-by-document approach and a mechanism for standardization of writing to form an interlingual pivot so that different descriptions can be normalized into one encoding system to undertake analysis.

\% Orthography profile implementation in Lingpy / generic tokenizer

\% describe Tokenizer's function

\% describe algorithm for tokenizing IPA (Unicode grapheme cluster tokenization and then alignment of Letter Modifiers and Tie Bars)

\% insert an example of orthographic tokenization from PAD

\section{move Letter Modifiers left}\label{move-letter-modifiers-left}

(\s)(\p{Lm}), \2

which catches:

\section{Diacritics: Unicode letter modifiers (Lm)}\label{diacritics-unicode-letter-modifiers-lm}

\section{({[}ː\textbar{}ʰ\textbar{}ʼ\textbar{}ʿ\textbar{}ʾ{]})}\label{ux2d0ux2b0ux2bcux2bfux2be}

from the set of this:

\section{Diacritics: all}\label{diacritics-all}

\section{({[}ː\textbar{}ʰ\textbar{}ʼ\textbar{}ʿ\textbar{}ʾ\textbar{}̜\textbar{}̠\textbar{}̟\textbar{}͡\textbar{}̪\textbar{}̰\textbar{}̣\textbar{}̥\textbar{}̩\textbar{}‿\textbar{}.\textbar{}→\textbar{}˻\textbar{}+\textbar{}̃\textbar{}̫\textbar{}̤{]})}\label{ux2d0ux2b0ux2bcux2bfux2be.}

and allows us to with one regular expression do this:

PAD examples

A side effect of character tokenization is identification of outliers, e.g.

Some examples in Python

Some examples in R

{[}UNICODE USE-CASES{]}

The most basic overall Unicode character property is the \textsc{General Category}, which categorizes Unicode characters into: \emph{Letters, Punctuation, Symbols, Marks, Numbers, Separators, }and* Other*. Each Unicode character property also has a character property value. So each General Category is denoted by a property value by a single letter abbreviation, e.g. ``L'' for Letter. Each General Category property also has subcategories that are also properties with property values. For example, ``Letters'' is broken down into Uppercase Letter (Lu), Lowercase Letter (Ll), Titlecase Letter (Lt), Modifier Letter (Lm) and Other Letter (Lo).\footnote{The subcategory ``Other Letter'' includes characters such as uncased letters as in the Chinese writing system.) By defining character properties for each code point, algorithmic implementations such as text processing applications or regular expression engines that are conformant to the Unicode Standard can be used to recognize whole (sub)categories of characters or ranges of characters, e.g.~to match proper names in English in title case or more generally to match character types, such as scripts, or symbols and integers in email addresses, phone numbers, etc.

Script information is useful for algorithmic processes, such as collation, searching and for multilingual text processing. For example, if a regular expression engine includes functionality to match characters at the level of script, then it provides the ability to identify characters, e.g. \match{Greek}, which tests whether letters in a text belong to the Greek script or not. This functionality can be used to determine the boundaries between different writing systems that appear in the same document. In Section 6 (Use cases of grapheme tokenization) we provide an example of a regular expression match at the level of character property.

Unicode-aware regular expression engines may contain a meta-character, ``\X'', that matches this level of grapheme cluster, i.e.~a base character followed by some number of accents or diacritical marks. Grapheme clusters are important to determine word or line boundaries, for collation (`ordering') and for user interface interaction (e.g.~mouse selection, backspacing, cursor movement).

\section{9. Conclusion}\label{conclusion}

In our work we focus on the writing systems used in lesser-described and endangered languages and our orthography profiles are used to describe writing systems and to transpose them into phonetic transcription. Of course the International Phonetic Alphabet and other transcription systems are essentially just orthographies that are an approximation of sound. Nevertheless, sound-based normalization is practical in automatically identifying cognates and sound changes with quantitative methods. We leverage orthography profiles to enable comparative analysis of languages with different writing systems.

An orthography profile lists the graphemes in a particular description of language data, e.g.~a wordlist, dictionary or corpus. Building on the knowledge that can be extracted from that description by tokenizing words by the graphemes made explicit in the orthography profile, it is straightforward to undertake other analyses of the data. For example, various ngram models of the data can be extracted with a few lines of code. A unigram model with counts, frequencies and positive log probability (``plogs''; ``entropy under this model'') (Goldsmith \& Riggle, 2010) provides a fair amount of information about a given data source. Essentially, the orthography profile provides the description that allows this information to be calculated based on the mapping of sequences of characters into graphemic units. Bigram models are also straightforwardly extracted. In the case of bigrams, mutual information can be captured and used in various other statistical analyses, such as quantitative language comparison, inferring phylogenetic trees, etc.

\section{10. Nice text blocks to be included somewhere}\label{nice-text-blocks-to-be-included-somewhere}

Looking at the variety of writing systems used for the world's languages, there are various challenges for the computational processing of resources in which such writing systems are used.

A widespread mechanism to derive new symbols for a writing system is to use a combination of multiple symbols already available. That can lead to completely new symbols, like <æ> or <w>, which are still rather transparently derived from <a e> and <v v>, respectively. More typically, many orthographies use sequences of symbols as special characters, like <sch> or <gh>. To identify such `grapheme clusters' is one of the central problems to be tackled with orthography profiles. A further frequently used mechanism is to add extra material to an already available symbol, either to make it better visible (this is probably the reason for the addition of a dot on top of the <i>) or to derive new symbols. The widespread use of diacritics is the obvious result of such processes (Gaultney 2002). However, also the derivation of new symbols by added extra strokes is widespread (and not only in Far-Eastern character-based writing systems), like for example <ŋ> or <ç>.

The use of punctuation (Daniëls 1994, Humez \& Humez 2008) and spaces (Saenger 1997) to improve readability. {[}e.g.~space after punctuation, but not before{]}

Currently, the Unicode Standard is the standard character encoding for linguistic data (Anderson 2003). However, there are still hundreds of different encoding systems that were invented independently to capture orthographic diversity as different nations adopted and developed computer systems. These different encoding systems are problematic and in conflict with one another because different standards were formalized differently and for different purposes by different standards committees in different countries. No unified encoding scheme contained enough code points to encode all characters, so two different encoding schemes possibly used the same code point for different characters, or used different code points to represent the same character. Because computers support multiple character encoding schemes, data risked being corrupted when handled by different applications and encodings. The Unicode Standard was devised to alleviate these problems.

For orthographies, identifying graphemes can be even more challenging than identifying phones and phonemes in phonetic transcription because although transcriptions may not adhere strictly to IPA, they tend to have straightforward mappings between sounds and symbols. On the other hand, orthographies can introduce orthographic rules, which add an additional challenges in identifying graphemes in words, as mentioned above. Thus for resources not in IPA or IPA-like transcriptions, graphemes often must be manually identified, whether they are encoded as singletons or multi-character sequences. The identification of graphemes and the formulation of orthographic rules are used to create an orthography profile.

Character-to-glyph mappings are particularly not clear-cut for linguists using the IPA.\footnote{We give more linguistic examples in Section 5.) Just look in the IPA Extensions block.\footnote{Character ``blocks'' are described in Section 3 (The Unicode approach).) There are several characters that are glyphs (variants) of the same grapheme in the Latin block, e.g. and \& and .\footnote{LATIN SMALL LETTER ALPHA U+0251 vs LATIN SMALL LETTER A U+0061 , and LATIN SMALL LETTER SCRIPT G U+0261 vs LATIN SMALL LETTER G U+0067 .) Other characters like

, , do not appear at all in the IPA Extensions block, but as separate phonetic symbols; they are already encoded in the Basic Latin, Latin Extended-A and Greek \& Coptic blocks. When a linguist transcribes an IPA

on a QWERTY keyboard, if it is from the Basic Latin group, it is valid Unicode IPA (it is not valid IPA if it is from the Cyrillic Block). However, keyboard and and because they have been given specific code points in the IPA Extensions block, where they have been given special status as IPA characters, and which most often go unnoticed by linguists.\footnote{In Section 5 we discuss in detail the Unicode IPA alphabet and its characters dispersal across different Unicode Standards scripts and blocks.)

The Unicode Standard divides text elements roughly into two classes of graphemes. The cover term \textsc{grapheme cluster}\footnote{Formerly referred to as ``locale-independent graphemes'' to emphasize that the term grapheme is used differently in linguistics.) is used to emphasize the difference between the term \emph{grapheme }as used by linguists and what the users perceive as a character (a basic unit of a writing system for a language), which may not be a single Unicode character.\footnote{http://www.unicode.org/reports/tr29/) The Unicode Consortium makes a further distinction between grapheme clusters and tailored grapheme clusters.

The simplest example of a grapheme cluster is essentially a base character followed by a Letter Modifier character. For example, the sequence + \textless{}\textasciitilde{}\textgreater{} (COMBINING TILDE at U+0303) combines visually into , a user-percieved character in writing systems like that of Spanish.\footnote{Although the code point sequence can also be represented with a single code point, LATIN SMALL LETTER N WITH TILDE at U+00F1, it is not the case that all multi-code point character sequences have a single code point equivalent, which can be deduced from normalization. Often for commonly-used letters in European languages, a single code point form exists.)

What the user perceives as a single character may actually involved a multi-code point sequence.\footnote{Unicode-aware regular expression engines may contain a meta-character, ``\X'', that matches this level of grapheme cluster, i.e.~a base character followed by some number of accents or diacritical marks. As mentioned, grapheme clusters are important for collation algorithms, but they are also used to determine other processes such as word or line boundaries and for user interface interaction (e.g.~mouse selection, backspacing, cursor movement).)

More challenging for text processing is what the Unicode Consortium terms tailored grapheme clusters. For example, the sequence + for the Slovak digraph or the sequence . These grapheme clusters are ``tailored'' in the sense that they must be specified on a language-by-language basis. Algorithms can be adapted to use as input these tailored language-specific orthographies for transliteration, localization, etc. In fact, the Unicode Standard provides a method for defining language-specific tailored writing systems called the Unicode Locale Description. Unicode Locale Descriptions allow users to define a set of extended grapheme clusters. Unicode Locale Descriptions are saved in the Common Locale Data Repository (CLDR),\footnote{http://cldr.unicode.org/) a repository of language-specific definitions of writing system properties, each of which describes specific usages of characters. Each can be encoded in a Unicode Locale Data Markup Language (LDML) document. LDML is an XML format and vocabulary for the exchange of structured locale data. It is a format used not only for CLDR, but also for general interchange of locale data, such as in Microsoft's .NET. We return to LDML in Section 5 (Orthography profiles and locale descriptions).

We have noted that the rendering of electronic text requires that characters are encoded and stored in a computer's memory. These character sequences are then mapped from character code points to glyphs by software that takes into account a character's context. Transparent to most users is the fact that different logical orders of characters may result in the same visual output. For example, phonetic characters with certain combinations of diacritics may be homoglyphs (see below), even while the logical order of their character sequences are not the same.\footnote{One example is a vowel that is both nasalized and creaky voice, e.g. {[}ḛ̃{]}. See discussion below.) Thus some type of standardization, what the Unicode Consortium calls ``normalization'' of the ordering of (some classes of) characters, is required to make sure that all data are logically consistent and therefore easily comparable and equally searchable. The aim of standardization is to attain syntactic \textsc{interoperability} of data.\footnote{By interoperability we mean the ability to ubiquitously exchange and merge disparate (textual) data sets, and data encoding formats, to facilitate data sharing and to ``effortlessly'' undertake comparison and analysis. For a discussion of syntactic versus semantic interoperability, see Ide and Pustejovsky, 2010.)

The IPA is often used by linguists in undertaking initial phonetic and phonemic analyses of languages. However, once the distinctive sounds of a given language have been identified (and a phonological description of the target language has been described), a linguist may continue his or her documentation and description of higher-order structures like words and sentences with a broad phoneme-based alphabet. In fact, literary materials development often leapfrogs from a phonemic alphabet to locale-specific practical orthographies, in which numerous factors affect a language's orthography development, e.g.~politics, purpose, speaker communities familiarity with other writing systems, technological limitations.

One simple example is illustrated with two genealogically related languages spoken in Northern Ghana: Western Sisaala\footnote{Western Sisaala lacks /kp/.) and Sisaala Tumulung.\footnote{Sisaala Tumulung (ISO 639-3: sil) is spoken around the city Tumu, which is about 50 kilometers geographically to the east of Western Sisaala (ssl) speaking areas.) Both languages contain the same basic set of contrastive sounds, but their orthographies encode affricates (and vowels) differently graphemically; the influence stem from orthographies of government-sponsored literacy languages, which are taught to children in schools in these areas. In Western Sisaala the sound /tʃ/ is written and /dʒ/ as due to familiarity of these graphemes by students learning Dagaare (Moran 2006). Spoken further to the east, Sisaala Tumuling encodes the sound /tʃ/ as and /dʒ/ as (Blass 1975).

If encoded in the Unicode Standard, the design of both orthographies has to specify that certain sequences of characters, such as + , are single graphemes (compare English and which is ambiguous). Design is important, for example, for collation ordering (sorting), i.e.~the sequence of characters that is treated as a unit for ordering purposes, such as in dictionaries. Indeed, the characters , and are three unique graphemes consisting of two Unicode characters from the Latin script, but developers of a software program that recognizes the Western Sisaala alphabet might want to specify the sort order , , \ldots{} . Thus there is a need for orthography-specific specifications that define sequences of characters as graphemes, whether technologically, linguistically, or both, for multilingual text processing.

A major challenge that the developers of the Unicode Standard faced is that there exists no universal set of fundamental units of text because the divisions of text into elements, such as graphemes, is dependent on language-specific writing systems and orthographies, as just shown. Moreover, the division of electronic text into elements is process dependent, e.g.~sorting and spell checking may or may not need to take capitalization into account, depending on the orthography in question. Thus the encoding and processing of text requires well defined \textsc{text elements} and rules for interpreting them; these elements and rules are language-specific when we talk about major languages or well established orthographies, but more often in our research we face input from lesser-studied languages documented in author-specific transcription or practical orthography.\footnote{And this is usually down to the document-specific level, the description of which has been coined ``doculect'' for purposes of discussing differences of opinion in language descriptions (Cysouw and Good, 2013).)

In Section XXX we noted that the Unicode Consortium makes a crucial distinction between characters and glyphs. The relations between characters and glyphs is a complex one and it is typically hidden from the user. Technologically, we must distinguish between characters and glyphs because:
\begin{enumerate}
	\def\labelenumi{\arabic{enumi}.} 
	\item The logical order of a sequence of characters (in computer memory) may not be the same as the visual order of their (rendered) glyphs. 
	\item There is not always a one-to-one mapping between characters and glyphs. 
\end{enumerate}

Regarding the first point, a user cannot assume that contiguously rendered text on-screen actually reflects the computer's internal storage order of individual and sequential characters in memory, i.e.~the \textsc{logical order}. In other words, in the Unicode Standard the visual order of glyphs may not be the same as the logical order of their digitally encoded characters; contiguous display is not indicative of contiguous text.* *A nice example is Hebrew text.\footnote{http://www.macchiato.com/unicode/globalization-gotchas) Another example is the glyph <ṍ>, which may consist of one, two or three characters in encoded in various physical orders. The possible mismatch in sequences becomes clearer when we consider the second point.

When a base character, i.e.~any graphic character not in the General Category of Combining Mark, is followed by one or more characters having the property General Category of Combining Mark, they can be combined visually into one or more glyphs using the Canonical Ordering Algorithm to create Unicode \textsc{grapheme clusters}. As we have shown, this input may be ambiguous from a technological perspective if normalization is not taken into account. For example, a linguist may insert a single character or a character sequence for to denote a high back unrounded vowel with high tone:
\begin{itemize}
	\item LATIN SMALL LETTER U WITH ACUTE (U+00FA) 
	\item LATIN SMALL LETTER U (U+0075) + COMBINING ACUTE ACCENT (U+0301) 
\end{itemize}

The normalization algorithms in the Unicode Standard are used to make these character sequences canonically equivalent for purposes such as ubiquitous search and comparison.

However, the standard does not specify many (or most) language-specific graphemes, or in Unicode parlance, \textsc{tailored grapheme clusters}, such as English .\footnote{Some (unfortunate) exceptions included for legacy purposes, and used incorrectly by linguists, include digraphs like LATIN SMALL LETTER TS DIGRAPH at U+02A6 or LATIN SMALL LETTER DEZH DIGRAPH at U+02A4. These and other digraphs are encoded in the Unicode Standard's IPA Extensions block.) For technological reasons, these language-specific tailored grapheme clusters must be defined outside of the Unicode Standard, in what are called (language-specific) Unicode Locale Descriptions.\footnote{http://www.unicode.org/reports/tr35/) The encoding of these multiple character sequences in Locale Descriptions is beyond the need (or technical ability) of many linguists; thus we have developed a simpler approach which we used in our own research and describe in detailed in Section XXX. Basically, the description of writing systems for lower resource languages, and in particular for document-specific descriptions (e.g.~different descriptions of the same language with different orthographies), requires an orthographic description at the (tailored) grapheme (cluster) level.

An orthography profile is defined as a simple delimited flat text file encoded in the UTF-8, with character sequences in Unicode Normalization Form D. Metadata aside, the column delimited contents of an orthography profile are essential a state-table, i.e.~an implementation of a state transition table, but for ease of use encoded in a simple flat text file.

In an orthography profile, all character strings are converted into NFD because each character in a grapheme may have phonetic value, and by using NFD, all graphemes are decomposed into a standardized order. For example, a vowel that is both nasalized and creaky looks like in IPA, where is some vowel. Although visually the same, a nasalized and creaky vowel can be composed of several different character sequences, as illustrated with in 1-3:
\begin{enumerate}
	\def\labelenumi{\arabic{enumi}.} 
	\item <õ̰> LATIN SMALL LETTER O + COMBINING TILDE + COMBINING TILDE BELOW: U+006F + U+0303 + U+0330 
	\item <õ̰> LATIN SMALL LETTER O + COMBINING TILDE BELOW + COMBINING TILDE: U+006F + U+0330 + U+0303 
	\item <õ̰> LATIN SMALL LETTER O WITH TILDE + COMBINING TILDE BELOW: U+00F5 + U+0330 
\end{enumerate}

Applying NFD to these three character sequences results in one standard sequence; in this example the character sequence is that shown 1. NFD makes different sequences of input interoperable and it retains all of the phonetic information captured by the separate characters that combine to form a vowel with nasalization and creaky voice phonation. This is an important step to consider in linguistic research because each character has phonetic value, i.e.~the combining tilde denotes that the base character is produce by a lowering of the velum, which allows air to escape through the nose when the sound is produced. The tilde below also has meaning: it represents a phonation type in which the vocal cords are vibrating, but they are also being held slightly further apart, allowing more air to escape, which produces a murmured sound. To enable comparative analyses of languages at a deep phonetic level, retaining a code point for each character of phonetic value is deciduous. For example, if the sound systems of two languages are to be compared, one would not want to lose a possibly contrastive phonological feature like nasalization, which may have been collapsed into a single character in its logical representation.

Our approach is similar in nature to the LDML, but we use a simpler state-table method, instead of generating an XML format for the ``exchange of structured locale data''. An orthography profile is a specification of the tailored grapheme clusters in a particular source document and any orthographic re-write rules needed after the application of Unicode normalization and orthographic profile tokenization. Orthography profiles can of course be used for transliteration between scripts, e.g.~from practical orthography to IPA and vice versa, by specifying mapping (replacements) from the list of tailored grapheme clusters in the source document and a set of replacement characters. This will be made clear in the next section in which we provide some use cases.

To summarize, LDML is basically overkill for our intentions of identifying and delimiting grapheme boundaries (or in Unicode parlance ``tailored grapheme clusters''). LDML is appropriate, even necessary, for complex localization complexes for software applications -- localization of, say, an operating system or software application across cultures may require the instantiation of different date stamps an other local standards. For linguistic purposes of source document data tokenization, we therefore develop a simple-to-implement process of ``orthographic'' tokenization, using standard Unicode Standard-specified processes (e.g.~normalization and grapheme cluster tokenization) with an additional level of linguistic ``tailored grapheme cluster'' specification.

\textsc{REFERENCES}

Nancy Ide and James Pustejovsky. 2010. What does inter- operability mean, anyway? toward an operational def- inition of interoperability for language technology. In Proceedings of the Second International Conference on Global Interoperability for Language Resources. Hong Kong, China.

Huber, Randal Q. and Robert B. Reed (compilers).
\begin{enumerate}
	\def\labelenumi{\arabic{enumi}.} \setcounter{enumi}{1991} 
	\itemsep1pt\parskip0pt\parsep0pt 
	\item Vocabulario comparativo: Palabras selectas de lenguas indígenas de Colombia (Comparative vocabulary: Selected words in indigenous languages of Colombia). Bogota: Instituto Lingüísti- co de Verano. 
\end{enumerate}

Thiesen, W. (1998). Diccionario bora-castellano, castellano-bora (1st ed.). Lima Perú: Instituto Linguístico de Verano.Retrieved from: http://www.worldcat.org/title/diccionario-bora-castellano-castellano-bora/oclc/40505215\&referer=brief\_results

\textsc{APPENDIX - ORTHOGRAPHY PROFILE EXAMPLE}

\textsc{APPENDIX - UNICODE IPA}

Provide a clear and full Unicode IPA chart with additional information as an appendix
\begin{verbatim}
	- create this chart

	- perhaps we should spell out conformance rules with regard to this chart? i.e. if a source claims conformance to our Unicode IPA chart, it should:

	- identify the sections (or range) of the table that the data conforms to

	- the source is in accordance with not only the Unicode IPA characters, but their intrinsic meanings (e.g. an Africanist might use <y> for /j/, but this level of conformance requires that <j> symbolizes /j/, etc.)

	- someone who wants to use an Africanist transcription system can declare that they use the Unicode IPA chart, except for a range of characters (but they provide the mapping to Unicode IPA chart) 
\end{verbatim}

\textsc{Notes}

\url{http://www.unicode.org/reports/tr29/}

The default boundary determination mechanism specified in this annex provides a straightforward and efficient way to determine some of the most significant boundaries in text: user-perceived characters, words, and sentences. Boundaries used in line breaking (also called \emph{word wrapping}) are defined in {[}\href{http://www.unicode.org/reports/tr41/tr41-11.html\#UAX14}{UAX14}{]}.

\textsc{Nice quotes}

vanZundert: ``How does one design and plan a general unified infrastructure for a target that is all over the place? A `one size fits all' approach would be a disastrous underestimation of the specific needs of humanities research. The essence of humanities research is in its diverse, heterogeneous and ephemeral nature. But the need of big infrastructures to be designed well before implementation and use, means that every single design decision closes the resulting system to some category of users. Anything specific can potentially break the data model and the defined workflow. Of course big infrastructure architects also know this; to compensate, they look for possibilities to make more abstract models and malleable workflows. But abstract away far enough and the model becomes rather meaningless. If the design recommendation of big infrastructure is that we will use XML, what was the use of all that time and effort? Big infrastructure design seems to routinely underestimate this problem. To give but one simple example: the token seems such a logical, granular, easy to automate conceptual catch-all for text models. Can there possibly be a more generalizable model for text than as a series of tokens? You may think so, until someone asks how you handle languages that do not use spaces, or how you account for prefixes and suffixes, let alone ambiguous functions. Regarding text as a series of tokens denies in any case several other essential humanistic aspects of text: its typography, its materiality, its referentiality to name but a few. At second glance, from the humanities research perspective, there is pretty little actually generalizable about tokenization.''

\% Ladefoged, Peter. 1990. Some Reflections on the IPA. In UCLA Working Papers in Phonetics, 74. ``The most interesting ambiguities in the Association's theoretical concepts concern what the symbols are symbolizing.''

There are several ambiguous aspects of the IPA, but the most interesting concerns the theoretical concepts that its symbols are symbolizing \cite[64]{Ladefoged1990a}. Although the IPA is \texttt{intended to be a set of symbols for representing all the possible sounds of the world\textquotesingle{}s languages\textquotesingle{}\textquotesingle{} \textbackslash{}citep\{IPA2005\}, Ladefoged points out that the IPA also states that IPA symbols are}\textit{not} symbols for phones they are simply shorthand for what a phonologist would regard as a bundle of features'' \cite{Ladefoged1990a}. From the IPA:
\begin{quote}

	``The representation of these sounds uses a set of phonetic categories which describe how each sound is made. These categories define classes of sounds that operate in phonological rules and historical sounds changes. The symbols of the IPA are shorthand ways of indicating certain intersections of these categories. Thus [p] is a shorthand way of designating the intersection of the categories voiceless, bilabial, and plosive; [m] is the intersection of the categories voiced, bilabial, and nasal; and so on.'' 
\end{quote}

\noindent Thus, the ``association is deliberately not explicit about what is meant by a phonological contrast'' \cite[65]{Ladefoged1990a}.

### 2015-05-29 chapters/orthography_profiles.tex
initial commit paper

--- 2015-05-29 chapters/originaltext.tex
@@
-The Unicode Cookbook for Linguists Managing orthographies in multilingual datasets using orthography profiles
+\chapter{Orthography profiles}
+\label{orthography-profiles}
 
-Steven Moran \& Michael Cysouw
-
-\textsc{Abstract}
-
-In this paper, we define the notion of an orthography profile, which is a language-specific (or even resource-specific) description of the units and rules that are needed to adequately model a writing system.
-An orthography profile describes the Unicode code points, characters, graphemes and orthographic rules as used in a writing system.
-Our orthography profiles are used to describe writing systems and possibly even to transliterate them into another (for example IPA-like) transcriptions with the goal to harmonize orthographies across languages and resources on an ad-hoc basis.
-The central aspect of our approach is that we introduce an intermediate step of graphemic tokenization, which is closer to linguistic practice and easier to manage than transducer-based transliteration.
-We leverage orthography profiles to enable comparative analysis of languages with different writing systems and to perform quality control on datasets with highly specific orthographic systems.
-We also describe how our proposals relate to the current Unicode specification, and introduce some of the more delicate aspects of the Unicode Standard for a linguistic audience.
-
-\textsc{Contents}
-
-{[}{[}TOC{]}{]}
-
-\section{1. Introduction}\label{introduction}
-
-Writing systems arise and develop in a complex mixture of cultural, technological and practical pressures.
-They tend to be highly conservative, in that people who have learned to read and write in a specific way (however impractical or tedious) are mostly unwilling to change their habits, e.g.~they tend to resist spelling reforms.
-In all literate societies there exists a strong socio-political mainstream that tries to force unification of writing (for example by strongly enforcing `right' from `wrong' writing in schools).
-However, there is also a large community of users who take as many liberties in their writing as they can get away with.
-
-For example, the writing of tone diacritics in Yoruba is often proclaimed to be the `right' way to write, although many users of Yoruba writing seem to be perfectly fine with leaving them out.
-As pointed out by the proponents of the `official' rules, there are some homographs when leaving out the tone diacritics (Olúmúyìw 2013: 44).
-However, writing systems (and the languages they represent) are normally full of homophones, which is normally not a problem at all for speakers of the language.
-More importantly, writing is not just a purely functional tool, but just as importantly it is a mechanism to signal social affiliation.
-By showing that you `know the rules' of expressing yourself in writing, others will more easily accept you as a worthy participant in their group.
-And that just as well holds for obeying to the `official' rules when writing a job application, as for obeying to the `informal' rules when writing an SMS to classmates in school.
-The case of Yoruba writing is an exemplary case, as even after more than a century of efforts to standardize the writing systems, there is still a wide range of variation in daily use (Olúmúyìw 2013).
-
-The sometimes cumbersome and sometimes illogical structure, and the enormous variability of existing writing systems is a fact of life scholars have to accept and should try to adapt to as good as possible.
-Our plea here is a proposal for a formalization to do exactly that.
-
-When considering the worldwide linguistic diversity, including all lesser-studied and endangered languages, there exist numerous different orthographies using symbols from the same scripts.
-For example, there are hundreds of orthographies using Latin-based alphabetic scripts.
-All of these orthographies use the same symbols, but these symbols differ in meaning and usage throughout the various orthographies.
-To be able to computationally use and compare different orthographies, we need a way to specify all orthographic idiosyncrasies in a computer-readable format (a process called `tailoring' in Unicode parlance).
-We call such specifications \textsc{orthography profiles}.
-Ideally, these specifications have to be integrated into so-called Unicode locale descriptions, though we will argue that in practice this is often not the most useful solution for the kind of problems arising in the daily practice of linguistics.
-Consequently, a central goal of this paper is to flesh out the linguistic challenges for locale descriptions, and work out suggestions to improve their structure for usage in a linguistic context.
-Conversely, we also aim to improve linguists' understanding and appreciation for the accomplishments of the Unicode Consortium in the development of the Unicode Standard.
-
-The necessity to computationally use and compare different orthographies most forcefully arises in the context of language comparison.
-Concretely, in our current research our goal is to develop quantitative methods for language comparison and historical analysis in order to investigate worldwide linguistic variation and to model the historical and areal processes that underlie linguistic diversity (cf.~Steiner et al.
-2010, List 2012a, 2012b, List \& Moran 2013, Moran \& Prokic 2013).
-In this work, it is crucial to be able to flexibly process across numerous resources with different orthographies.
-In many cases even different resources on the `same' language use different orthographic conventions.
-Another orthographic challenge that we encounter regularly in our linguistic practice is electronic resources on a particular language that claim to follow a specific orthographic convention (often a resource-specific convention), but on closer inspection such resources are almost always not consistently encoded.
-Thus a second goal of our orthography profiles is to allow for an easy specification of orthographic conventions, and use such profiles to check consistency and to report errors to be corrected.
-
-A central step in our proposed solution to this problem is the tailored grapheme separation of strings of symbols, a process we call \textsc{grapheme tokenization}.
-Basically, given some strings of symbols (e.g.~morphemes, words, sentences) in a specific source, our first processing step is to specify how these strings have to be separated into graphemes, considering the specific orthographic conventions used in a particular source document.
-Our experience is that such a graphemic tokenization can be performed without extensive in-depth knowledge about the phonetic and phonological details of the language in question.
-For example, the specification that <ou> is a grapheme of English is a much easier task than to specify what exactly the phonetic values of this grapheme are in any specific occurrence in English words.
-Grapheme separation is a task that can be performed relatively reliably and with limited availability of time and resources (compare, for example, the task of creating a complete phonetic or phonological normalization).
-
-Although grapheme tokenization is only one part of the solution, it is an important and highly fruitful processing step.
-Given a grapheme tokenization, various subsequent tasks become easier, like (a) temporarily reducing the orthography in a processing pipeline, e.g.~only distinguishing high versus low vowels; (b) normalizing orthographies across sources (often including temporary reduction of oppositions), e.g.~specifying an (approximate) mapping to the International Phonetic Alphabet; (c) using co-occurrence statistics across different languages (or different sources in the same language) to estimate the probability of grapheme matches, e.g.~with the goal to find regular sound changes between related languages or transliterations between different sources in the same language.
-
-Before we deal with these proposals, in the first part of this paper (Sections 2 through 5) we give an extended introduction to the notion of encoding (Section 2) and writing systems, both from a linguistic perspective and from the perspective of the Unicode Consortium (Section 3).
-We consider the Unicode Standard to be a breakthrough (and ongoing) development that fundamentally changed the way we look at writing systems, and we aim to provide here a slightly more in-depth survey of the many techniques that are available in the standard.
-A good appreciation for the solutions that the Unicode Standard also allows for a thorough understanding of possible pitfalls that one might encounter when using it (Section 4).
-As an example of the current state-of-the-art, we discuss the rather problematic marriage of the International Phonetic Alphabet (IPA) with Unicode (Section 5).
-
-The second part of the paper (Sections 6 and 7) describes our proposals for how to deal with the Unicode Standard in the daily practice of (comparative) linguists.
-First, we discuss the challenges of characterizing a writing system.
-To solve these problems, we propose the notions of orthography profiles, closely related to Unicode locale descriptions (Section 6).
-Finally, we discuss practical issues with actual examples (Section 7).
-We provide reference implementation of our proposals in R and in Python, available as open-source libraries.
-
-The following conventions are followed in this paper.
-All phonemic and phonetic representations are given in the International Phonetic Alphabet (IPA), unless noted otherwise (International Phonetic Association, 2005).
-Standard conventions are used for distinguishing between graphemic <\textgreater{}, phonemic / / and phonetic {[} {]} representations.
-For character descriptions, we follow the notational conventions of the Unicode Standard (Unicode Consortium 2014).
-Character names are represented in small capital letters (e.g.~LATIN SMALL LETTER SCHWA) and code points are expressed as U\emph{+n} where \emph{n} is a four to six digit hexadecimal number, e.g.~U+0256, which can be rendered as the glyph <ə>.
-
-\section{2. Encoding}\label{encoding}
-
-There are many in-depth histories of the origin and development of writing systems (e.g.~Robinson 1997, Powell 2012), a story that we will not repeat here.
-However, the history of turning writing into computer readable code is not so often told, so we decided to offer a short survey of the major developments here.
-This history turns out to be intimately related to the history of telegraphic communication.\footnote{Because of the recent history as summarised in this section, we have used mostly rather ephemeral internet sources, specifically http://www.unicode.org/history/ and various Wikipedia pages for the information presented here (when not references by traditional literature in the bibliography).
-A useful survey of the historical development of the physical hardware of telegraphy and telecommunication is Huurdeman (2003).
-Most books that discuss the development of encoding of telegraphic communication focus of cryptography (e.g.~Singh 1999) and forego the rather interesting story of `open' encoding that is related here.)
-
-Writing systems have existed for roughly 6000 years, allowing people to exchange messages through time and space.
-Additionally, to bridge large geographic distances, \textsc{telegraphic systems} of communication (from Greek \emph{τῆλε γράφειν} ``distant writing'') have a long and widespread history since ancient times.
-The most widespread telegraphic systems worldwide are so-called whistled languages (Meyer 2015), but also drumming languages (Meyer et al.
-2012) and smoke, fire, hydraulic or flag signals are forms of telegraphy.
-Telegraphy was reinvigorated in the end of the eighteenth century through the introduction of semaphoric systems by Claude Chapelle (since then using flags, flashing lights, or various specially designed contraptions) to convey messages over large distances.
-The `innovation' of those systems was that all characters of the written language were replaced one-to-one by visual signals.
-Since then, all telegraphic systems have taken this principle, namely that any language to be transmitted first has to be turned into some orthographic system, which subsequently is encoded for transmission by the sender, and then turned back into orthographic representation at the receiver side.\footnote{Sound and video-based telecommunication of course takes a different approach by ignoring the written version of language and directly encode sound waves or light patterns.) This of course implies that the usefulness of any such telegraphic encoding completely depends on the often rather haphazard structure of orthographic systems.
-
-In the nineteenth century, \textsc{electric telegraphy} lead to a new approach in which written language characters were encoded by signals sent through a copper wire.
-Originally, `bisignal' codes were used, consisting of two different signals.
-For example, Carl Friedrich Gauss in 1833 used positive and negative current (Mania 2008: 282).
-More famous and influential, Samuel Morse in 1836 used long and short pulses.
-In those bisignal systems each character from the written language was encoded with a different number of signals (between one and five), so two different separators are needed: one between signals and one between characters (in Morse-code there is a short pause between signals and a long pause between characters).\footnote{Actually, Morse-code also includes an extra long pause between words.
-Interestingly, it took a long time to consider the written white-space as a bona-fide character that should simply be encoded with its own code point instead of with a different kind of pause.
-This happened only with the revision of the Baudot-code (see below) by Donald Murray in 1901, in which he finally introduced a white-space code.
-This principle has been followed ever since, and for example was included in the second version of the International Telegraph Alphabet (ITA2) in 1930.)
-
-From those encodings, true \textsc{binary codes} developed with a fixed length of signals per character.
-In such systems only a single separator between signals is needed, because the separation between characters can be established by counting until a fixed number of signals.\footnote{Of course, no explicit separator is needed when the timing of the signals is known, which is the principle used in all modern telecommunication systems.
-An important modern consideration is also how to know where to start counting when you did not catch the start of a message, something that is known in Unicode as `self synchronizing'.) In the context of electric telegraphy, such a system was first established by Émile Baudot in 1870, using a fixed combination of five signals for each written character.\footnote{True binary codes have a longer history, going at least back to the Baconian cipher devised by Francis Bacon in 1605.) There are 25 = 32 possible combination when using five binary signals; an encoding today designated as `5-bit'.
-These codes are sufficient for all Latin letters, but of course they do not suffice for all written symbols, including punctuation and digits.
-As a solution, the Baudot code uses a so-called `shift' character, which signifies that from that point onwards (until shifted back) a different encoding is used, allowing for yet another set of 32 codes.
-In effect, this means that the Baudot code, and the \textsc{International Telegraph Alphabet (ITA)} derived from it, had an extra `bit' of information, so the encoding is actually 6-bit (with 26 = 64 different possible characters).
-For decades, this encoding was the standard for all telegraphy and it is still in limited use today.
-
-To also allow for different uppercase and lowercase letters and a large variety of control characters, the American Standards Association decided to propose a new 7-bit encoding in 1963 (with 27 = 128 different possible characters), known as the \textsc{American Standard Code for Information Interchange(ASCII)}, geared towards the encoding of English orthography.
-With the ascent of other orthographies in computer usage, the wish to encode further variation of Latin letters (like German <ß> or various letters with diacritics like <è>) led the Digital Equipment Corporation to introduce an 8-bit \textsc{Multinational Character Set (MCS)} (with 28 = 256 different possible characters), first used with the introduction of the VT220 Terminal in 1983.
-Because 256 characters were clearly not enough for the many different characters needed in the world's writing systems, the ISO/IEC 8859 standard in 1987 extended the MCS to include 16 different 8-bit code pages.
-For example, part 5 was used for Cyrillic characters, part 6 for Arabic, and part 7 for Greek.\footnote{In effect, because 16 = 24, this means that ISO/IEC 8859 was actually a 8+4 = 12-bit encoding, though with very many duplicates by design (e.g.~all ASCII codes were repeated in each 8-bit code page).
-To be precise, ISO/IEC 8859 used the 7-bit ASCII as a basis for each code page, and defined 16 different 7-bit extensions, leading to (1+16) times 27 = 2,176 possible characters.)
-
-This system almost immediately was understood to be insufficient and impractical, so various initiatives to extend and reorganize the encoding started in the 1980s.
-This led, for example, to various proprietary encodings from Microsoft (e.g.~Windows Latin 1) and Apple (e.g.~Mac OS Roman), which one still sometimes encounters today.
-More wide-ranging, various people in the 1980s started to develop true international code sets.
-In the United States, a group of computer scientists formed the Unicode Consortium, proposing a 16-bit encoding in 1991 (with 216 = 65,536 different possible characters).
-At the same time in Europe, the International Organization for Standardization (ISO) was working on ISO 10646 to supplant the ISO 8859 standard.
-Their first draft of the \textsc{Universal Character Set (UCS)} in 1990 was 31-bit (with theoretically 231 = 2,147,483,648 possible characters, but because of some technical restrictions only 679,477,248 were allowed).
-Since 1991, the Unicode Consortium and the International Organization for Standardization have jointly developed the \textsc{Unicode Standard}, or ISO/IEC 10646, leading to the current system including the original 16-bit Unicode proposal as the \textsc{Basic Multilingual Plane}, and 16 additional planes of 16-bit for further extensions (with in total 1+16 times 216 = 1,114,112 possible characters).
-The most recent version of the Unicode Standard (7.0) was published in June 2014 and it defines 112,956 different characters (Unicode Consortium 2014).
-
-\section{3. Terminology}\label{terminology}
-
-\subsection{The linguistic tradition}\label{the-linguistic-tradition}
-
-In this section we provide a very brief overview of the linguistic terminology concerning writing systems before turning to the slightly different computational terminology in the next section on the Unicode Standard.
-Linguistically speaking, a \textsc{writing system} is a symbolic system that uses visible or tactile signs to represent language in a systematic way.
-The term `writing system' has two mutually exclusive meanings.
-First, it may refer to the way a particular language is written.
-In this sense the term refers to the writing system of a particular language, as, for example, in \emph{``the Serbian writing system uses two scripts: Latin and Cyrillic''}.
-Second, the term `writing system' may refer to a type of symbolic system as used among the world's languages to represent languages, as, for example, in \emph{``alphabetic writing system''}.
-In this latter sense the term refers to how scripts have been classified according to the way that they encode language, as in, for example, \emph{``the Latin and Cyrillic scripts are both alphabetic writing systems''}.
-To avoid confusion, this second notion of `writing system' would have been more aptly called \textsc{script system}.
-The term \textsc{script} refers to a collection of distinct symbols as employed by one or more orthographies.
-For example, both Serbian and Russian are written with subsets of the Cyrillic script.
-A single language, like Serbian or Japanese, can also be written using orthographies based on different scripts.
-Over the years linguists have typologized script systems in a variety of ways, with the tripartite classification of logographic, syllabic, and alphabetic remaining the most popular, even though there are at least half a dozen different types of script systems that can be distinguished (Daniels, 1990, 1996).
-
-Only focussing here on the first sense, we distinguish two different kinds of writing systems as used to write a particular language, namely transcriptions and orthographies.
-First, \textsc{transcription} is a scientific procedure (and also the result of that procedure) for graphically representing the sounds of human speech at the phonetic level.
-It incorporates a set of unambiguous symbols to represent speech sounds, including conventions that specify how these symbols should be combined.
-A \textsc{transcription system} is a specific system of symbols and rules used for transcription of the sounds of a spoken language variety.
-In principle, a transcription system should be language-independent, in that it should be applicable to all spoken human languages.
-The International Phonetic Alphabet (IPA) is a commonly used transcription system that provides a medium for transcribing languages at the phonetic level.
-However, there is a long history of alternative kinds of transcription systems (see Kemp 1994) and today various alternatives are in widespread use (e.g.~X-SAMPA and Cyrillic-based phonetic transcription systems).\footnote{Most users of IPA do not follow the standard explicitly and many dialects based on it have emerged, e.g.~the Africanist and Americanist transcription systems.
-) Note that IPA symbols are also often used to represent language on a phonemic level.
-However, it is important to realize that in this usage the IPA symbols are not a transcription system, but rather an orthography (though with strong links to the pronunciation).
-Further, a transcription system does not need to be highly detailed, but it can also be a system of broad sound classes.
-Although such an approximative transcription is not normally used in linguistics, it is widespread in technological approaches (e.g.~Soundex, Beider \& Morse 2010)\footnote{http://stevemorse.org/phonetics/bmpm.htm , http://stevemorse.org/phonetics/bmpm2.htm) and it is sometimes fruitfully used in automatic approaches to historical linguistics (Dolgopolsky 1986, ASJP\footnote{http://asjp.clld.org/)).
-
-Second, an \textsc{orthography} specifies the symbols, punctuations, and the rules in which a specific language is written in a standardized way.
-Orthographies are often based on a phonemic analysis, but they almost always include idiosyncrasies because of historical developments (like sound changes or loans) and because of the widely-followed principle of lexical integrity (i.e.~the attempt to write the same lexical root in a consistent way, also when synchronic phonemic rules change the pronunciation, for example with final devoicing in many Germanic languages).
-All orthographies are language-specific (and often even resource-specific), although individual symbols or rules might be shared between languages.
-A \textsc{practical orthography} is a strongly phoneme-based writing system designed for practical use by speakers.
-The mapping relation between phonemes and graphemes in practical orthographies is purposely shallow, i.e.~there is mostly a systematic and faithful mapping from a phoneme to a grapheme.
-Practical orthographies are intended to jumpstart written materials development by correlating a writing system with the sound units of a language (cf.~Meinhof \& Jones 1928).
-Symbols from the IPA are often used by linguists in the development of practical orthographies for languages without writing systems, though this usage of IPA symbols should not be confused with transcription (as defined above).
-Further, a \textsc{transliteration} is a mapping between two different orthographies.
-It is the process of ``recording the graphic symbols of one writing system in terms of the corresponding graphic symbols of a second writing system'' (Kemp 2006:396).
-In straightforward cases, such a transliteration is simply a matter of replacing one symbol with another.
-However, there are widespread complications, like one-to-many or many-to-many mappings, which are not always easy, or even possible, to solve without listing all cases individually (cf.~Moran 2012, chp 2).
-
-Breaking it down further, a script consists of \textsc{graphemes}, and graphemes consist of \textsc{characters}.
-In the linguistic terminology of writing systems, a \textsc{character} is a general term for any self-contained element in a writing system.\footnote{There is a second interpretation of the term `character', i.e.~a conventional term for a unit in the Chinese writing system (Daniels 1996).
-This interpretation will not be further explored in this paper.) Although in literate societies most people have a strong intuition about what the characters are in their particular orthography or orthographies, it turns out that the separation of an orthography into separate characters is far from trivial.
-The widespread intuitive notion of a character is strongly biased towards educational traditions, like the alphabet taught at schools, and technological possibilities, like the available type pieces in a printer's job case, the keys on a typewriter, or the symbols displayed in Microsoft Word's symbol browser.
-In practice, characters often consist of multiple building blocks, each of which could be considered a character in its own right.
-For example, although a Chinese character may be considered to be a single basic unanalyzable unit, at a more fine-grained level of analysis the internal structure of Chinese characters is often comprised of smaller semantic and phonetic units that should be considered characters (Sproat 2000).
-In alphabetic scripts, this problem is most forcefully exemplified by diacritics.
-A \textsc{diacritic} is a mark, or series of marks, that may be above, below, or through other characters (Gaultney 2002).
-Diacritics are sometimes used to distinguish homophonous words, but they are more often used to indicate a modified pronunciation (Daniels \& Bright 1996, xli).
-The central question is whether, for example, , , and should be considered four characters, or different combinations of three characters.
-In general, multiple characters together can form another character, and it is not always possible to decide on principled grounds what should be the basic building blocks of an orthography.
-
-For that reason, it is better to analyse an orthography as a collection of graphemes.
-A \textsc{grapheme} is the basic, minimally distinctive symbol of a particular writing system, alike to the phoneme is an abstract representation of a distinct sound in a specific language.
-The term `grapheme' was modeled after the term `phoneme' and represents a contrastive graphical unit in a writing system (see Kohrt 1986 for a historical overview of the term grapheme).
-Most importantly, a single grapheme regularly consists of multiple characters, like <th>, <ou> and <gh> in English (note that each character in these graphemes is also a separate grapheme in English).
-Such complex graphemes are often used to represent single phonemes.
-So, a combination of characters is used to represent a single phoneme.
-Note that the opposite is also found in writing systems, in cases in which a single character represents a combination of two or more phonemes.
-For example, <x> in English orthography represents a combination of the phonemes /k/ and /s/.
-Further, conditioned or free variants of a grapheme are called \textsc{allographs}.
-For example, the distinctive forms of Greek sigma are conditioned, with <σ> being used word-internally and <ς> being used at the end of a word.
-In sum, there are many-to-many relationships between phonemes and graphemes as they are expressed in the myriad of language- and resource-specific orthographies.
-
-This exposition of the linguistic terminology involved in describing writing systems has been purposely brief.
-We have highlighted some of the linguistic notions that are pertinent, yet sometimes confused with, the technological definitions developed for the computational processing of the world's writing systems, which we describe in the next section.
-
-\subsection{The Unicode approach}\label{the-unicode-approach}
-
-The conceptualisation and terminology of writing systems was rejuvenated through the development of the Unicode Standard, with major input from Mark Davis, co-founder and long-term president of the Unicode Consortium.
-For many years, `exotic' writing systems and phonetic transcription systems on personal computers were constrained by the American Standard Code for Information Interchange* *(ASCII) character encoding scheme (based on the Latin script), which only allowed for a strongly limited number of different symbols to be encoded.
-This implied that users could either use and adopt the (extended) Latin alphabet or they could assign new symbols to the small number of code points in the ASCII encoding scheme to be rendered by a specifically designed font (Bird \& Simons 2003).
-In this situation, it was necessary to specify the font together with each document to ensure the rightful display of its content.
-To alleviate this problem of assigning different symbols to the same code points, in the late 80's and early 90's the Unicode Consortium set itself the ambitious goal of developing a single universal character encoding to provide a unique number, a code point, for every character in the world's writing systems.
-Nowadays, the Unicode Standard is the default encoding of the technologies that support the World Wide Web and for all modern operating systems, software and programming languages.
-
-The Unicode Standard represents a massive step forward because it aims to eradicate the distinction between `universal' (ASCII) versus `language/resource-particular' (Font) by adding as much as possible language-specific information into the universal standard.
-However, there are still language/resource-specific specifications necessary for the proper usage of Unicode, as will be discussed below.
-Within the Unicode structure many of these specifications can be captured by so-called `Locales' or `Locale Descriptions', so we are moving to a new distinction of `universal' (Unicode Standard) versus `language-particular' (Locale Description).
-The major gain is a much larger compatibility on the universal level (because Unicode standardizes a much larger portion of writing system diversity), and much better possibilities for automatic processing on the language-particular level (because Locale Descriptions are computer readable specifications).
-
-Each version of the Unicode Standard (as of writing at version 7) consists of a set of specifications and guidelines that include (i) a core specification, (ii) code charts, (iii) standard annexes and (iv) a character database.\footnote{All documents of the Unicode Standard are available at http://www.unicode.org/versions/latest/.
-For a quick survey of the use of terminology inside the Unicode Standard, their glossary is particularly useful, available at http://www.unicode.org/glossary/.
-For a general introduction to the principles of Unicode, Chapter 2 of the core specification, called ``General Structure'', is particularly insightful.
-Different from many other documents of the Unicode Standard, this general introduction is relatively easy to read and illustrated with many interesting examples from various orthography traditions all over the world.
-) The \textsc{core specification} is a book directed toward human readers that describes the formal standard for encoding multilingual text.
-The \textsc{code charts} provide a humanly readable online reference to the character contents of the Unicode Standard in the form of PDF files.
-The \textsc{Unicode Standard Annexes (UAX)} are a set of technical standards that describe the implementation of the Unicode Standard for software development, Web standards, programming languages, etc.
-The \textsc{Unicode Character Database (UCD)} is a set of computer-readable text files that describe the character properties, including a set of rich character and writing system semantics, for each character in the Unicode Standard.
-In this section, we introduce the basic Unicode concepts, but we will leave out many details.
-Please consult the above mentioned full documentation for a more detailed discussion.
-Further note that the Unicode Standard is exactly that, namely a standard.
-It normatively describes notions and rules to be followed.
-In the actual practice of applying this standard in a computational setting, a specific implementation is necessary.
-The most widely used implementation of the Unicode Standard is the \textsc{International Components for Unicode (ICU)}, which offers C/C++ and Java libraries trying to implement to Unicode Standard.\footnote{http://icu-project.org )
-
-The Unicode Standard is a \textsc{character encoding system} which goal it is to support the interchange and processing of written characters and text in a computational setting.
-Underlyingly, the character encoding is represented by a range of numerical values called a \textsc{code space}, which is used to encode a set of characters.
-A \textsc{code point} is a unique non-negative integer within a code space (i.e.~within a certain numerical range).
-In the Unicode Standard character encoding system, an \textsc{abstract character}, for example a ``LATIN SMALL LETTER P'', is mapped to a particular code point, such as the decimal value 112, normally represented in hexadecimal and in Unicode parlance as (U+)0070.\footnote{Hexadecimal (`base-16') 0070 is equivalent to decimal (`base-10') 112, which can be calculated by considering 0·163 + 0·162 + 7·161 + 0·160 = 7·16 = 112.
-Underlyingly, computers will of course treat this code point binary (`base-2') as 11100000, i.e.~112 = 1·27 + 1·26 + 1·25 + 0·24 + 0·23 + 0·22 + 0·21 + 0·20 = 64 + 32 + 16.) That encoded abstract character is rendered on a computer screen (or printed page) as a \textsc{glyph}, e.g.
-<p>, depending on the \textsc{font} and the context in which that character appears.
-
-In Unicode Standard terminology, an (abstract) \textsc{character} is the basic encoding unit.
-The term `character' can be quite confusing due to its alternative definitions across different scientific disciplines and because in general the word `character' means many different things to different people.
-It is therefore often preferable to refer to Unicode characters simply as \textsc{code points}, because there is a one-to-one mapping between Unicode characters and their numeric representation.
-The Unicode Consortium (2012) defines a character as ``the smallest component of written language that has semantic value.
-It refers to the abstract meaning and/or general shape, rather than a specific shape, though in code tables some form of visual representation is essential for the reader's understanding.'' Unicode defines characters as abstractions of orthographic symbols, and it does not define visualizations for these characters (although it does presents examples).
-In contrast, a \textsc{glyph} is a concrete graphical representation of a character as it appears when rendered (or rasterized) and displayed on an electronic device or on printed paper.
-For example, <reinstert Gs here> are different glyphs of the same character, i.e.~they may be rendered differently depending on the typography being used, but they all share the same code point.
-From the perspective of Unicode they are `the same thing'.
-In this approach, a \textsc{font} is then simply a collection of glyphs linked to code points.
-Allography is not specified in Unicode (expect for a few exceptional cases, due to legacy encoding issues), but can be specified in a font as a \textsc{contextual variant} (aka presentation form).
-
-Each code point in the Unicode Standard is associated with a set of \textsc{character properties} as defined by the Unicode character property model.\footnote{The character property model is described in http://www.unicode.org/reports/tr23/, but the actual properties are described in http://www.unicode.org/reports/tr44/.
-A simplified overview of the properties is available at http://userguide.icu-project.org/strings/properties.
-The actual code tables listing all properties for all Unicode code points are available at http://www.unicode.org/Public/UCD/latest/ucd/.
-) Basically, those properties are just a long list of values for each character.
-For example, code point U+0047 has the following properties (among many others):
-\begin{itemize}
-	\item Name: LATIN CAPITAL LETTER G 
-	\item Alphabetic: YES 
-	\item Uppercase: YES 
-	\item Script: LATIN 
-	\item Extender: NO 
-	\item Simple\_Lowercase\_Mapping: 0067 
-\end{itemize}
-
-These properties contain the basic information of the Unicode Standard and they are necessary to define the correct behavior and conformance required for interoperability in and across different software implementations (as defined in the Unicode Standard Annexes).
-The character properties assigned to each code point is based on each character's behavior in the real-world writing traditions.
-For example, the corresponding lowercase character to U+0047 is U+0067 (though note that the relation between uppercase and lowercase is in many situations much more complex than this, and Unicode has further specifications for those cases).
-Another use of properties is to define the script of a character.
-The Unicode Standard defines the term \textsc{script} as, ``A collection of letters and other written signs used to represent textual information in one or more writing systems.
-For example, Russian is written with a subset of the Cyrillic script; Ukrainian is written with a different subset.
-The Japanese writing system uses several scripts.'' In practice, script is simply defined for each character as the explicit ``Script'' property in the Unicode Character Database.
-
-One frequently references property is the \textsc{block} property, which is often used in software applications to impose some structure to the large number of Unicode characters.
-Each character in Unicode belongs to a specific block.
-These blocks are basically an organisational structure to alleviate the administrative burden of keeping Unicode up-to-date.
-Blocks consist of characters that in some way belong together, so that characters are easier to find.
-Some blocks are connected with a specific script, like the Hebrew block or the Gujarati block.
-However, blocks are predefined ranges of code points, and often there will come a point after which the range is completely filled.
-Any extra characters will have to be assigned somewhere else.
-There is, for example, a block `Arabic', which contains most Arabic symbols.
-However, there is also a block `Arabic Supplement', `Arabic Presentation Forms-A' and `Arabic Presentation Form B'.
-The situation with Latin symbols is even more extra.
-In general, the names for block should be taken as a definitional statement.
-For example, many IPA symbols are not located in the aptly-names block ``IPA extensions'', but in other blocks (see Section 5).
-
-There are many cases in which a sequence of characters (i.e.~a sequence of more than one code point) represents what a user perceives as an individual unit in a particular orthographic writing system.
-For this reason the Unicode Standard differentiates between \textsc{abstract character} and \textsc{user-perceived character}.
-Sequences of multiple code points that correspond to a single user-perceived characters are called** grapheme clusters** in Unicode parlance.
-Grapheme clusters come in two flavors: (default) grapheme clusters and tailored grapheme clusters.
-
-The (default) \textsc{grapheme clusters} are locale-independent graphemes, i.e.~they always apply when a particular combination of characters occurs independent of the writing system in which they are used.
-These character combinations are defined in the Unicode Standard as functioning as one \textsc{text element}.\footnote{The Unicode Standard defines text element as: ``A minimum unit of text in relation to a particular text process, in the context of a given writing system.
-In general, the mapping between text elements and code points is many-to-many.'') The simplest example of a grapheme cluster is a base character followed by a letter modifier character.
-For example, the sequence + <̃\textgreater{} (LATIN SMALL LETTER N at U+006E followed by COMBINING TILDE at U+0303) combines visually into , a user-percieved character in writing systems like that of Spanish.
-So, what the user perceives as a single character actually involves a multi-code-point sequence.
-Note that this specific sequence can also be represented with a single \textsc{precomposed} code point, LATIN SMALL LETTER N WITH TILDE at U+00F1, but this is not the case for all multi-code point character sequences.
-The problem that there multiple encodings possible for the same text element has been acknowledged early on in the Unicode Standard (e.g.~for , the sequence U+006E U+0303 should in all situations be treated identically to the precomposed U+00F1), and a system of \textsc{canonical equivalence} is available for such situations.
-Basically, the Unicode Standard offers different kind of normalizations to either decompose all precomposed characters (called \textsc{NFD}, \textsc{Normalization Form Canonical Decomposition}), or compose as much as possible combinations (called \textsc{NFC}, \textsc{Normalization Form Canonical Composition}).
-In current practice of software development, NFC seems to be preferred in most situations and is widely proposed as the preferred canonical form.
-
-More difficult for text processing, because less standardized, is what the Unicode Standard terms \textsc{tailored grapheme clusters}.
-Tailored grapheme clusters are locale-dependent graphemes, i.e.~such combination of characters do not function as text elements in all situations.
-For example, the sequence + for the Slovak digraph or the sequence in the Sisaala practical orthography (pronounced as IPA /tʃ/; Moran 2006).
-These grapheme clusters are ``tailored'' in the sense that they must be specified on a language-by-language or a writing system-by-writing system basis.
-The Unicode Standard provides technological specifications for creating locale specific data in so-called \textsc{Unicode Locale Descriptions}, i.e.~a set of specification that defines a set of language-specific elements (e.g.~tailored grapheme clusters, collation order, capitalization-equivalence), as well as other special information, like how to format numbers, dates, or currencies.
-Locale descriptions are saved in the \textsc{Common Locale Data Repository (CLDR)},\footnote{http://cldr.unicode.org/ ) a repository of language-specific definitions of writing system properties, each of which describes specific usages of characters.
-Each locale can be encoded in a document using the \textsc{Locale Data Markup Language (LDML)}.
-LDML is an XML format and vocabulary for the exchange of structured locale data.
-
-Unicode Locale Descriptions allow users to define language- or resource-specific writing systems or orthographies.
-The Unicode Consortium defines \textsc{writing system} only very loosely, as it is not a central concept in the Unicode Standard.
-A writing system is, ``A set of rules for using one or more scripts to write a particular language.
-Examples include the American English writing system, the British English writing system, the French writing system, and the Japanese writing system.''.
-However, there are various drawbacks of locale descriptions for the daily practice of linguistic work in a multilingual setting (see Section 7).
-
-\section{4. Unicode Pitfalls}\label{unicode-pitfalls}
-
-\subsection{Wrong it is not}\label{wrong-it-is-not}
-
-In this section, we describe some of the most common pitfalls that we have encountered when using the Unicode Standard in our own work, or in discussion with other linguists.
-This section is not meant as a criticism of the decisions made by the Unicode Consortium; on the contrary, we aim to highlight where the technological aspects of the Unicode Standard diverge from many users' intuitions.
-What have sometimes been referred to as problems or inconsistencies in the Unicode Standard are mostly due to legacy compatibility issues, which can lead to unexpected behavior by linguists using the standard.
-However, there are also some cases in which the Unicode Standard has made decisions that theoretically could have been taken differently, but for some reason or another (mostly very good reasons) were accepted as they are now.
-We call behavior that executes without error but does something different than the user expected---often unknowingly---a ``pitfall''.
-
-In this context, it is important to realize that the Unicode Standard was not developed to solve linguistic problems per se, but to offer a consistent computational environment for written language.
-In those cases in which the Unicode Standard behaves differently as expected, we think it is important not to dismiss Unicode as `wrong' or `deficient', because our experience is that in almost all cases the behaviour of the Unicode Standard has been particularly well thought through.
-The Unicode Consortium considers important practical use-cases that from a linguistic point of view are normally not considered.
-Our general guideline for dealing with the Unicode Standard is to accept it as it is, and not to battle windmills.
-Alternatively, of course, it is possible to actively engage in the development of the standard itself, an effort that is highly appreciated by the Unicode Consortium.
-
-\subsection{Pitfall: Blocks}\label{pitfall-blocks}
-
-The Unicode code space is subdivided into blocks of contiguous code points.
-For example, the block called ``Cyrillic'' runs from U+0400 till U+04FF.
-These blocks arose as an attempt at ordering the enormous amount of characters in Unicode, but the ideas of blocks very quickly ran into problems.
-First, the size of a block is fixed, so when a block is full, a new block will have to be instantiated somewhere further in the code space.
-For example, this led to the blocks ``Cyrillic Supplement'', ``Cyrillic Extended-A'' (which are also already full) and ``Cyrillic Extended-B''.
-Second, when a specific character already exists, then it is not duplicated in another block, although the name of the block might indicate that a specific symbol should be available there.
-In general, names of blocks are just an approximate indication of the kind of characters that will be in the block.
-
-The problem with blocks arises because finding the right character among the thousands of Unicode characters is not easy.
-Many software applications present blocks as a primary search mechanism, because the block names suggest where to look for a particular character.
-However, when a user searches for an IPA character in the block ``IPA Extensions'' then many IPA characters will not be found there.
-For example, the velar nasal <ŋ> is not part of the block ``IPA Extensions'' because it was already included as LATIN SMALL LETTER ENG at U+014B in the block ``Latin Extensions-A''.
-
-In general, finding a specific character in the Unicode Standard is often not trivial.
-The names of the blocks can help, but it is not (and never was supposed to be) a foolproof structure.
-It is not the goal nor aim of the Unicode Consortium to provide a user interface to the Unicode Standard.
-If one often encounters the problem of needing to find a suitable character, there are various other useful services for end users available.\footnote{The Unicode website offers a basic interface to the code charts at http://www.unicode.org/charts/index.html.
-As a more flexible interface, we particularly like PopChar from Macility, available for both Macintosh and Windows.
-However, there are various free websites that offer search interfaces to the Unicode code tables, like http://unicode-search.net or http://unicode-search.net.
-A further useful approach for searching characters using shape matching is http://shapecatcher.com.
-)
-
-\subsection{Pitfall: Missing glyphs}\label{pitfall-missing-glyphs}
-
-The Unicode Standard is often praised (and deservedly so) for solving many of the perennial problems with the interchange and display of the world's writing systems.
-However, a common complaint from users is that, while the praise may be true, they mostly just see some boxes on their screen instead of those promised symbols.
-The problem of course is that users' computers do not have any glyphs installed matching the Unicode code points in the file they are trying to inspect.
-It is important to realize that internally in the computer everything still works as expected: any handling of Unicode code points works independently of how they are displayed on the screen.
-So, although a user might only see boxes being displayed, this user should be assured that everything is still in order.
-
-The central problem behind the missing glyphs is that designing actual glyphs includes a lot of different considerations and it is a time-consuming process.
-Many traditional expectations of how specific characters should look like have to be taken into account when designing glyphs.
-Those expectations are often not well documented, and it is mostly up to the knowledge and experience of the font designer to try and conform to them as good as possible.
-Therefore, most designers produce fonts only including glyphs for certains parts of the Unicode Standard, namely for those characters they feel comfortable with.
-At the same time, the number of characters defined by the Unicode Standard is growing with each new version, so it is neigh impossible for any designer to produce glyphs for all characters.
-The result of this is that, almost necessarily, each font only includes glyphs for a subset of the characters in the Unicode Standard.
-
-The simple solution to missing glyphs is thus to install additional fonts providing additional glyphs.
-For the more exotic characters there is often not much choice.
-There are a few particularly large fonts that might be considered.
-First, there is the \textsc{Everson Mono} font made by Michael Everson, which currently includes 9,756 different glyphs (not including Chinese) updated up to Unicode 7.0.\footnote{Everson Mono is available as shareware at http://www.evertype.com/emono/.
-) Already a bit older is the \textsc{Titus Cyberbit Basic} font made by Jost Gippert and Carl-Martin Bunz, which includes 10,044 different glyphs (not including Chinese), but not including newer characters added after Unicode 4.0.\footnote{Titus Cyberbit Basic is available at http://titus.fkidg1.uni-frankfurt.de/unicode/tituut.asp.
-)
-
-Further, we suggest to always install at least one so-called \textsc{fall-back font,} which provides glyphs that at least show the user some information about the underlyingly encoded character.
-Apple Macintoshes have such a font (which is invisible to the user), which is designed by Michael Everson and made available for other systems through the Unicode Consortium.\footnote{The Apple/Everson fallback font is available at http://www.unicode.org/policies/lastresortfont\_eula.html.
-) Further, the \textsc{GNU Unifont} is a clever way to produce bitmaps approximating the intended glyph of each available character, updated to Unicode 7.0.\footnote{The GNU Unifont is available at http://unifoundry.com/unifont.html.
-) Finally, the Summer Institute of Linguistics provides a \textsc{SIL Unicode BMP Fallback Font}, currently available up to Unicode version 6.1.
-This font does not even attempt to show a real glyph, but only shows the hexadecimal code inside a box for each character, so a user can at least see the Unicode codepoint of the character to be displayed.\footnote{The SIL Unicode BMP Fallback Font is available at http://scripts.sil.org/UnicodeBMPFallbackFont.
-)
-
-\subsection{Pitfall: Faulty rendering}\label{pitfall-faulty-rendering}
-
-A similar complaint to missing glyphs, discussed previously, is that while there might be a glyph being displayed, it does not `look right'.
-There are two reasons for unexpected visual display, namely automatic font substitution and faulty rendering.
-Like missing glyphs, any such problems are independent from the Unicode Standard.
-The Unicode Standard only includes very general information about characters and leaves the specific visual display to others to decide on.
-Any faulty display is thus not to be blamed on the Unicode Consortium, but on a complex interplay of different mechanisms happening in a computer to turn Unicode codepoints into visual symbols.
-We will only sketch a few aspects of this complex interplay here.
-
-Most modern software applications (like Microsoft Word) offer some approach to \textsc{automatic font substitution}.
-This means that when a text is written in a specific font (e.g.~Times New Roman) and a inserted Unicode character does not have a glyph within this font, then the software application will automatically search for another font to display the glyph.
-The result will be that some specific glyph will look slightly different from the others.
-This mechanism works differently depending on the software application, and mostly only limited user influence is expected and little feedback is given, which might be rather frustrating to font-aware users.\footnote{For example, Apple Pages does not give any feedback that a font is being replaced, and the user does not seem to have any influence on the choice of replacement (except by manually marking all occurrences).
-In contrast, Microsoft Word does indicate the font replacement by showing the name of the font replacement, but it simply changes the font, so any text written after the replacement is written in a different font as before.
-Both behaviors leave much to be desired.)
-
-The other problem with visual display is related to the so-called `font rendering'.
-\textsc{Font rendering }refers to the process of the actual positioning of Unicode characters on a page of written text.
-This positioning is actually a highly complex problem, and many things can go wrong in the process.
-Well-known rendering problems, like proportional glyph size or ligatures are reasonably well understood.
-However, even within the Latin script, the positioning of diacritics relative to a base character is still a widespread problem.
-Especially when more than one diacritic is supposed to be placed above (or below) each other, this often leads to unexpected effects in many modern software applications.
-The problems arising in Arabic and in many southeast Asian scripts (like Devanagari or Burmese) are even much more complex.
-There are basically three different approaches to font rendering.
-The most widespread is Adobe's and Microsoft's \emph{OpenType }system.
-This approach makes it relatively easy for font developers, as the font itself does not include all details about the precise placement of individual characters.
-For those details, additional script-descriptions are necessary.
-All of those systems can lead to unexpected behaviour.\footnote{For more details about OpenType, see http://www.adobe.com/products/type/opentype.html and http://www.microsoft.com/typography/otspec/.
-Additional systems for complex text layout are, among others, Microsoft's DirectWrite https://msdn.microsoft.com/library/dd368038.aspx and the open-source project HarfBuzz http://www.freedesktop.org/wiki/Software/HarfBuzz/.
-) Alternative systems are Apple's \emph{Apple Advanced Typography} (AAT) and the open-source \emph{Graphite} system from the Summer Institute of Linguistics (SIL).\footnote{More information about AAT can be found at https://developer.apple.com/fonts/.
-SIL's Graphite is described in detail at http://scripts.sil.org/cms/scripts/page.php?site\_id=projects\&item\_id=graphite\_home.
-) In both of these systems, a larger burden is placed on the description inside the font.
-
-\subsection{Pitfall: Homoglyphs}\label{pitfall-homoglyphs}
-
-Homoglyphs are visually indistinguishable glyphs (or highly similar glyphs) that have different code points in the Unicode Standard and thus different character semantics.
-As a principle, the Unicode Standard does not specify how a character appears visually on the page or the screen.
-So in most cases, a different appearance is caused by the specific design of a font, or by user-settings like size or boldface.
-Taking an example already discussed in the previous section, the following symbols <reinsert Gs here> are different glyphs of the same character, i.e.~they may be rendered differently depending on the typography being used, but they all share the same code point (viz.
-LATIN SMALL LETTER G at U+0067).
-In contrast, the symbols <A А Α Ꭺ ᗅ ᴀ ꓮＡ𐊠 𐌀 𝖠 𝙰> are all different code points, although they look highly similar -- in some cases even sharing exactly the same glyph in some fonts.
-All these different A-like characters include the following code points in the Unicode Standard:
-\begin{itemize}
-	\item <A\textgreater{} LATIN CAPITAL LETTER A, at U+0041 
-	\item <А> CYRILLIC CAPITAL LETTER A, at U+0410 
-	\item <Α> GREEK CAPITAL LETTER ALPHA, at U+0391 
-	\item <Ꭺ> CHEROKEE LETTER GO, at U+13AA 
-	\item <ᗅ> CANADIAN SYLLABICS CARRIER GHO, at U+15C5 
-	\item <ᴀ> LATIN LETTER SMALL CAPITAL A, at U+1D00 
-	\item <ꓮ> LISU LETTER A, at U+A4EE 
-	\item <Ａ> FULLWIDTH LATIN CAPITAL LETTER A, at U+FF21 
-	\item <𐊠> CARIAN LETTER A, at U+102A0 
-	\item <𐌀> OLD ITALIC LETTER A, at U+10300 
-	\item <𝖠> MATHEMATICAL SANS-SERIF CAPITAL A, U+1D5A0 
-	\item <𝙰> MATHEMATICAL MONOSPACE CAPITAL A, at U+1D670 
-\end{itemize}
-
-The existence of such homoglyphs is partly due to legacy compatibility, but for the most part these characters are simply different characters that happen to look similar.\footnote{A particularly nice interface to look for homoglyphs is http://shapecatcher.com, based on the principle of recognizing shapes (Belongie et al.
-2002).) Yet, they are supposed to behave different from the perspective of a font designer.
-For example, when designing a Cyrillic font, the will have different aesthetics and different traditional expectation compared to a Latin .
-
-Such homoglyphs are a widespread problem for consistent encoding.
-Although for most users it looks like the words and are identical, in actual fact they do not even share a single code point.\footnote{The first words consists completely of Latin characters, namely U+0076, U+006F, U+0063, U+0065, U+0073, while the second is a mix of Cyrillic and Greek characters, namely U+03BD, U+03BF, U+0041, U+0435, U+0455.) For computers these two words are completely different entities.
-Commonly, when users with Cyrillic or Greek keyboards have to type some Latin-based orthography, they mix similar looking Cyrillic or Greek characters into their text, because those characters are so much easier to type.
-Similarly, when users want to enter an unusual symbol, they normally search by visual impression in their favourite software application, and just pick something that looks reasonably alike to what they expect the glyph to look like.
-
-\subsection{Pitfall: Canonical equivalence}\label{pitfall-canonical-equivalence}
-
-For some characters, there is more than one possible encoding in the Unicode Standard.
-This is a possible pitfall, as this would mean that for the computer there exist multiple different entities that for a user are the same.
-This would, for example, lead to problems with searching, as the computer would search for specific encodings, and not find all expected characters.
-As a solution, the Unicode Standard includes a notion of \textsc{canonical equivalence}.
-Different encodings are explicitly declared as equivalent in the Unicode Standard code tables.
-Further, to harmonize all encodings in a specific piece of text, the Unicode Standard proposes a mechanism of \textsc{normalization}.
-
-Consider for example the characters and following Unicode code points:
-\begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item LATIN CAPITAL LETTER A WITH RING ABOVE (U+00C5) 
-	\item ANGSTROM SIGN (U+212B) 
-	\item LATIN CAPITAL LETTER A (U+0041) + COMBINING RING ABOVE (U+030A) 
-\end{enumerate}
-
-The character, represented here by glyph , is encoded in the Unicode Standard in the first two examples by a single-character sequence; each is assigned a different code point.
-In the third example, the glyph is encoded in a multiple-character sequence that is composed of two character code points.
-All three sequences are \textsc{canonically equivalent}, i.e.~they are strings that represent the same abstract character and because they are not distinguishable by the user, the Unicode Standard requires them to be treated the same in regards to their behavior and appearance.
-Nevertheless, they are encoded differently.
-For example, if one were to search an electronic text (with software that does not apply Unicode Standard normalization) for ANGSTROM SIGN (U+212B), instances of LATIN CAPITAL LETTER A WITH RING ABOVE (U+00C5) would not be found.
-
-In other words, there are equivalent sequences of Unicode characters that should be normalized, i.e.~transformed into a unique Unicode-sanctioned representation of a character sequence called a \textsc{normalization form}.
-Unicode provides a \emph{Unicode Normalization Algorithm}, which essentially puts combining marks into a specific logical order and it defines decomposition and composition transformation rules to convert each string into one of four normalization forms.
-We will discuss here the two most relevant normalization forms: NFC and NFD.
-
-The first of the three characters above is considered the \textsc{Normalization Form C (NFC)}, where ``C'' stands for composition.
-When the process of NFC normalization is applied to the character sequences in 2 and 3, both sequences are normalized into the `pre-composed' character sequence in 1.
-Thus all three canonical character sequences are standardized into one composition form in NFC.
-The other central Unicode normalization form is the \textsc{Normalization Form D (NFD)}, where ``D'' stands for decomposition.
-When NFD is applied to the three examples above, all three, including importantly the single-character sequences in 1 and 2, are normalized into the `decomposed' multiple-sequence of characters in 3.
-Again, all three are then logically equivalent and therefore comparable and syntactically interoperable.
-
-As illustrated, some characters in the Unicode Standard have alternative representations (in fact, many do), but the Unicode Normalization Algorithm can be used to transform certain sequences of characters into canonical (or compatible) forms to test for equivalency.
-To determine equivalence, each character in the Unicode Standard is associated with a combining class, which is formally defined as a character property called ``Canonical Combining Class'' which is specified in the Unicode Character Database.
-The combining class assigned to each code point is a numeric value between 0 and 254 and is used by the Canonical Ordering Algorithm to determine which sequences of combining marks are canonically equivalent.
-Normalization forms, as very briefly described above, can be used to ensure character equivalence by ordering character sequences so that they can be faithfully compared.
-
-It is very important to note that any software applications that is Unicode Standard compliant is free to change the character stream from one representation to another.
-This means that a software application may compose, decompose or reorder characters as its developers desire; as long as the resultant strings are canonically equivalent to the original.
-This might lead to unexpected behaviour for users.
-Various players, like the Unicode Consortium, the W3C, or the TEI recommend NFC in most user-directed situations, and some software applications that we tested indeed seem to automatically convert strings into NFC.\footnote{See the summary of various recommendation here: http://www.win.tue.nl/\textasciitilde{}aeb/linux/uc/nfc\_vs\_nfd.html.
-) This means in practice that if a user, for example, enters <a> and <`>, i.e.~LATIN SMALL LETTER A at U+0061 and COMBINING GRAVE ACCENT at U+0300, this might be converted into <à>, i.e.~LATIN SMALL LETTER A WITH GRAVE at U+00E0.\footnote{The behaviour of software applications can be quite erratic in this respect.
-For example, Apple's TextEdit does not do any conversion on text entry.
-However, when you copy and paste some text inside the same document in ``rich text mode'' (i.e.~RTF-format), it will be transformed into NFC on paste.
-Saving a document does not do any conversion to the glyphs on screen, but it will save the characters in NFC.)
-
-\subsection{Pitfall: Characters are not glyphs}\label{pitfall-characters-are-not-glyphs}
-
-A central principle of Unicode is the distinction between character and glyph.
-A character is the abstract notion of a symbol in a writing system, while a glyph is the concrete drawing of such a symbol.
-In practice, there is a complex interaction between characters and glyphs.
-A single Unicode character may of course be rendered as a single glyph.
-However, a character may also be a `piece' of a glyph, or vice-versa.
-Actually, all possible relations between glyphs and characters are attested.
-
-First, \textsc{a single character may have different contextually determined glyphs.} For example, characters in writing systems like Hebrew and Arabic have different glyphs depending on where they appear in a word.
-Some letters in Hebrew change their form at the end of the word, and in Arabic, primary letters have four contextually-sensitive variants (isolated, word initial, medial and final).
-Second, \textsc{a single character may be rendered as a sequence of multiple glyphs.} For example, in Tamil one Unicode character may result in a combination of a consonant and vowel, which are rendered as two adjacent glyphs by fonts that supports Tamil.
-Third, \textsc{a single glyph may be a combination of multiple characters.
-}For example, the ligature <ﬁ>, a single glyph, is the result of two characters, <f> and <i>, that have undergone glyph substitution by font rendering.
-Like contextually-determined glyphs, ligatures are (intended) artifacts of text processing instructions.
-Finally, \textsc{a single glyph may be a part of a character} (e.g.~diacritics).
-
-Further, the rendering of a glyph is also always dependent on the font being used.
-For example, the Unicode character LATIN SMALL LETTER G appears as and in the Computer Modern and Courier fonts because their typefaces are designed differently.
-Furthermore, font face may change the visual appearance of a character, for example Times New Roman two-story <a> changes to a single-story a in italics <\emph{a}>.
-This becomes a real problem for some phonetic typesetting (see Section 5, Pitfall 2).
-
-In sum, character-to-glyph mappings are complex technological issues that the Unicode Consortium has had to address in the development of the Unicode Standard, but for the lay user they can be utterly confusing because visual rendering does not indicate logical encoding, or, in other words, what you see isn't (necessarily) what you get.
-
-\subsection{Pitfall: File formats}\label{pitfall-file-formats}
-
-Unicode is a character encoding standard, but these characters of course actually appear inside some kind of file.
-The most basic Unicode-based file format is pure line-based text, i.e.~strings of Unicode-encoded characters separated by line breaks (note that these line breaks are what for most people intuitively corresponds to paragraph breaks).
-Unfortunately, even within this apparently basic setting there exist a multitude of variants.
-In general, these different possibilities are well-understood in the software industry, and nowadays they normally do not lead to any problems for the end user.
-However, there are some situations in which a user is suddenly confronted with cryptic questions in the user interface involving abbreviations like LF, CR, BE, LE or BOM.
-Most prominently this occurs with exporting or importing data in several software applications from Microsoft.
-Basically, there are two different issues involved.
-First, the encoding of** line breaks** and, second, the encoding of the Unicode characters into \textsc{code units} and the related issue of \textsc{endianness}.
-
-The issue with \textsc{line breaks} originated with the instructions necessary to direct a printing head to a new line.
-This involves two movements, known as \textsc{carriage return (CR)} (returning the printing head to the start of the line on the page) and \textsc{line feed (LF)} (moving the printing head to the next line on the page).
-Physically, these are two different events, but conceptually together they form one action.
-In the history of computing, various encodings of line breaks have been used (e.g.~CR+LF, LF+CR, only LF, or only CR).
-Currently, all Unix and Unix-derived systems use only LF as code for a line break, while software from Microsoft still uses a combination of CR and LF.
-Today, most software applications recognize both options, and are able to deal with either encoding of line breaks (until rather recently this was not the case, and using the wrong line breaks could lead to unexpected errors).
-Our impression is that there is a strong tendency in software development to standardize on the simpler ``only LF'' encoding for line breaks, and we suggest that everybody use this encoding whenever possible.
-
-A code unit is a sequence of bits used to encode character in a encoding.
-The issue with code units stems from the question how to separate a stream of binary ones and zero, i.e.~bits, into chunks representing Unicode characters.
-Depending on different use cases, the Unicode Standard offers three different approaches, called UTF-32, UTF-16 and UTF-8.\footnote{The letters UTF stand for either ``Unicode Transformation Format'' or ``Universal Coded Character Set Transformation Format'', but the notion of `transformation' is a legacy notion that does not have meaning anymore.
-Nevertheless, the designation ``UTF'' (in capitals) has become an official standard designation, but should probably best be read as simply ``Unicode Format''.) The details of this issue is extensively explained in section 2.5 of the Unicode core specification.
-Basically, \textsc{UTF-32} encodes each character in 32 bits (32 `binary units', i.e.~32 zeros or ones) and is the most disk-space-consuming variant of the three.
-However, it is the most efficient encoding processing-wise, because the computer simply has to separate each character after 32 bits.
-In contrast, \textsc{UTF-16} uses only 16 bits per character, which is sufficient for the large majority of Unicode characters, but not for all of them.
-A special system of `surrogates' is defined within the Unicode Standard to deal with these additional characters.
-The effect is a more disk-space efficient encoding (approximately half the size), while adding a limited computational overhead.
-Finally, \textsc{UTF-8} is a more complex system that dynamically encodes each character with the minimally necessary number of bits, choosing either 8, 16 or 32 bits depending on the character.
-This represents again a strong reduction in space (particularly due to the high frequency of data using ASCII characters that need only 8 bits) at the expense of even more computation necessary to process such strings.
-However, because of the ever growing computational power of modern machines, the processing overhead is in most practical situations a non-issue, while saving on space is still useful, particularly for sending texts over the Internet.
-As an effect, UTF-8 has become the dominant encoding on the World Wide Web.
-We suggest that everybody uses UTF-8 as their default encoding.
-
-A related problem is a general issue about how to store information in computer memory, which is known as \textsc{endianness}.
-The details of this issue go beyond the scope of this paper.
-It suffices to realize that there is a difference between \textsc{big-endian} (BE) storage and \textsc{little-endian} (LE) storage.
-The Unicode Standard offers a possibility to explicitly indicate what kind of storage is used by starting a file with a so-called `Byte Order Mark' (BOM).
-However, the Unicode Standard does not require the usage of BOM, preferring other non-Unicode methods to signal to computers which kind of endianness is used.
-This issue only arises with UTF-32 and UTF-16 encodings, and when using the preferred UTF-8, using a BOM is theoretically possible, but strongly dispreferred according to the Unicode Standard.
-We suggest that everyone tries to prevent the inclusion of BOM in your data.
-
-Together with the topic of normalisation (NFC vs.~NFD, see Pitfall 5 above), this leads to the following nicely cryptic file-encoding recommendation, namely to preferably use \textsc{UTF-8 in NFC without BOM with LF line breaks}, whenever possible.
-
-\section{5. IPA meets Unicode}\label{ipa-meets-unicode}
-
-\subsection{Introducing the International Phonetic Alphabet (IPA)}\label{introducing-the-international-phonetic-alphabet-ipa}
-
-The International Phonetic Alphabet (IPA) is a common standard in linguistics to transcribe sounds of spoken language into some Latin-based characters (International Phonetic Association 1999).
-Although IPA is reasonably easily adhered to with pen and paper, it is not trivial to encode IPA characters electronically.
-Early work addressing the need for a universal computing environment for writing systems and their computational complexity is discussed in Simons 1989.
-For a long time, linguists (like all other computer users) were limited to ASCII-encoded 7-bit characters, which only includes Latin characters, numbers and some punctuation and symbols.
-Restricted to these standard character sets that lacked IPA support or other language-specific graphemes that they needed, linguists devised their own solutions (cf.~Bird and Simons 2003).
-For example, some chose to represent unavailable graphemes with substitutes, e.g.~the combination of to represent .
-Tech-savvy linguists redefined selected characters from a character encoding by mapping custom made fonts to those code points.
-However, one linguist's electronic text would not render properly on another linguist's computer without access to the same font.
-Further, if two character encodings defined two character sets differently, then data could not be reliably and correctly displayed.
-This is a common example of the non-interoperability of data and data formats.
-
-To alleviate this problem, during the late 1980s, SAMPA (Speech Assessment Methods Phonetic Alphabet) was created to represent IPA symbols with 7-bit printable ASCII character sequences, e.g.
-for {[}ɸ{]}.
-Two problems with SAMPA are that (i) it is only a partial encoding of the IPA and (ii) it encodes different languages in separate data tables, instead of a universal alphabet, like IPA.
-SAMPA tables are derived from phonemes appearing in several European languages that were developed as part of a European Commission-funded project to address technical problems like electronic mail exchange (what is now simply called email).
-SAMPA is essentially a hack to work around displaying IPA characters, but it provided speech technology and other fields a basis that has been widely adopted and used in code.
-So, SAMPA was a collection of tables to be compared, instead of a large universal table representing all languages.
-An extended version of SAMPA, called X-SAMPA, set out to include every symbol in the IPA chart including all diacritics (Wells nd.).
-X-SAMPA was considered more universally applicable because it consisted of one table that encoded the set of characters that represented phones/segments in IPA across languages.
-SAMPA and X-SAMPA have been widely used for speech technology and as an encoding system in computational linguistics.
-Eventually, ASCII-encoding of the IPA became depreciated through the advent of the Unicode Standard.
-Note however that many popular software packages used for linguistic analyses still require ASCII input, e.g.~RuG/L04\footnote{http://www.let.rug.nl/kleiweg/L04/) and SplitsTree4.\footnote{http://www.splitstree.org/)
-
-Still, there are a few pitfalls to be aware of when using the Unicode Standard to encode IPA.
-As we have said before, from a linguistic perspective it might look like the Unicode Consortium is making incomprehensible decisions, but it is important to realize that the consortium has tried and is continuing to try to be as consistent as possible across a wide range of use cases, and it does place linguistic traditions above other orthographic possibilities.
-In general, we strongly suggest linguists not to complain about any decisions in the Unicode Standard, but to try and understand the rationale of the Unicode Consortium (which in our experience is almost always well-conceived) and devise ways to work with any unexpected behaviour.
-Many of the current problems derive from the fact that the IPA is clearly historically based on the Latin script, but different enough from most other Latin-based writing systems to warrant special attention.
-This ambivalent status of the IPA glyphs (partly Latin, partly special) is unfortunately also attested in the treatment of IPA in the Unicode Standard.
-In retrospect, it might have been better to consider the IPA (and other transcription systems) to be a special or new kind of script within the Unicode Standard, and treat the obvious similarity to Latin glyphs as a historical relic.
-All IPA glyphs would then have their own code points, instead of the current situation in which some IPA glyphs have special code points, while others are treated as being identical to the `regular' Latin characters.
-Yet, the current situation, however unfortunate, is unlikely to change, so as linguists we will have to learn to deal with the specific pitfalls of IPA within the Unicode Standard.
-In this section, we will describe these pitfalls in some detail.
-
-\subsection{Pitfall: There is no single complete Unicode code block for IPA}\label{pitfall-there-is-no-single-complete-unicode-code-block-for-ipa}
-
-The ambivalent nature of IPA glyphs arises because, on the one hand, the IPA uses Latin-based glyphs like , or .
-From this perspective, the IPA seems to be just another orthographic tradition using Latin characters, all of which do not get a special treatment within the Unicode Standard (just like e.g.~the French, German, or Danish orthographic traditions do not have a special status).
-On the other hand, the IPA uses many special symbols (like turned , mirrored and/or extended Latin glyphs ) not found in any other Latin-based writing system.
-For this reason, and already in the first version of the Unicode Standard (Version 1.0 from 1991), a special block with code points, called ``IPA Extensions'' was included.
-As noted in Section 4, Pitfall 1, the Unicode Standard code space is subdivided into character blocks, which generally encode characters from a single script.
-However, as is illustrated by the IPA, characters that form a single writing system may be dispersed across several different character blocks.
-With its diverse collection of symbols from various scripts and diacritics, the IPA is spread across over 13 blocks in the Unicode Standard:\footnote{This number of blocks depends on whether only IPA-sanctioned symbols are counted or if the phonetic symbols commonly found in the literature are also included, see Moran 2012, Appendix C.)
-\begin{itemize}
-	\item Basic Latin (30 characters), e.g. <a, b, c, d, e> 
-	\item Latin-1 Supplement (4 characters): <æ, ç, ð, ø\textgreater{} 
-	\item Latin Extended-A (3 characters): <ħ, ŋ, œ> 
-	\item Latin Extended-B (5 characters): <ǀ, ǁ, ǂ, ǃ, ȵ> 
-	\item IPA Extensions (70 characters), e.g.: <ɐ, ɑ, ɔ> 
-	\item Spacing Modifier Letters (20 characters), e.g.: <ʰ ʷ ˥> 
-	\item Combining Diacritical Marks (33 characters), e.g.: <̝ ̥ ̪> 
-	\item Greek and Coptic (3 characters): <β, θ, χ> 
-	\item Phonetic Extensions (2 characters): <ᴅ, ᴴ> 
-	\item Phonetic Extensions Supplement (3 characters): <ᶑ , ᶾ, ᶣ> 
-	\item Superscripts and Subscripts (1 character): <ⁿ> 
-	\item Arrows (2 characters): \textless{}↑, ↓\textgreater{} 
-	\item Latin Extended-C (1 character): <ⱱ> 
-\end{itemize}
-
-\subsection{Pitfall: There are many IPA homoglyphs in Unicode}\label{pitfall-there-are-many-ipa-homoglyphs-in-unicode}
-
-Another problem is the large number of homoglyphs, i.e.~different characters that have highly similar glyphs (or even completely identical, depending on the font rendering).
-For example, a speaker of Russian should ideally not use the CYRILLIC SMALL LETTER A at code point U+0430 for IPA transcriptions, but instead the LATIN SMALL LETTER A at code point U+0061, although visually they are mostly indistinguishable, and the Cyrillic character is more easily typed on a Cyrillic keyboard.
-Another example we commonly encounter is the use of LATIN SMALL LETTER G at U+0067, instead of the sanctioned Unicode Standard IPA character for the voiced velar stop LATIN SMALL LETTER SCRIPT G at U+0261.
-One begins to question whether this issue is at all apparent to the working linguist, or if they simply use the U+0067 because it is easily keyboarded and thus saves time, whereas the latter must be cumbersomely inserted as a special symbol in most software.\footnote{This issue was recently addressed by the International Phonetic Association, which has taken the stance that both the keyboard LATIN SMALL LETTER G and the LATIN SMALL LETTER SCRIPT G are valid input characters for the voiced velar plosive.
-Unfortunately, this decision further introduces ambiguity for linguists trying to adhere to a strict Unicode Standard IPA encoding.)
-
-Furthermore, on the one hand even linguists are unlikely to distinguish between the LATIN SMALL LETTER SCHWA at code point U+0259 and LATIN SMALL LETTER TURNED E at U+01DD.
-On the other hand, non-linguists are unlikely to distinguish any semantic difference between an open back unrounded vowel , the LATIN SMALL LETTER ALPHA at U+0251, and the open front unrounded vowel , LATIN SMALL LETTER A at U+0061.
-But even among linguists this distinction leads to problems.
-For example, as pointed out by Mielke (2009), there is a problem stemming from the fact that about 75\% of languages are reported to have a five-vowel system (Maddieson 1984).
-Historically, linguistic descriptions tend not to include precise audio recording and measurements of formants, so this may lead one to ask if the many characters that are used in phonological description reflects a ``transcriptional bias''.
-The common use of in transcriptions could be in part due to the ease of typing the letter on an English keyboard (or for older descriptions, the typewriter).
-We found it to be exceedingly rare that a linguist uses for a low back unrounded vowel.\footnote{One example is Vidal 2001a:75, in which the author states: ``The definition of Pilagá /a/ as {[}+back{]} results from its behavior in certain phonological contexts.
-For instance, uvular and pharyngeal consonants only occur around /a/ and /o/.
-Hence, the characterization of /a/ and /o/ as a natural class of (i.e., {[}+back{]} vowels), as opposed to /i/ and /e/.'') They simply use as long as there is no opposition to .\footnote{See Thomason's Language Log post, ``Why I don't love the International Phonetic Alphabet'', at: http://itre.cis.upenn.edu/\textasciitilde{}myl/languagelog/archives/005287.html.) Making things even more problematic, there is an old typographic tradition that the double-story uses a single-story in italics.
-This leads to the unfortunate effect that in most well-designed fonts the italics of and use the same glyph.
-If this distinction has to be kept upright in italics, the only solution we can currently offer is to use `slanted' glyphs (i.e.~artificially italicised glyphs) instead of real italics (i.e.~special italics glyphs designed by a typographer).\footnote{For example, the widely used IPA font Doulos SIL (http://scripts.sil.org/cms/scripts/page.php?item\_id=DoulosSIL) does not have real italics.
-This leads some word-processing software, like Microsoft Word, to produce slanted glyphs instead.
-That particular combination of font and software application will thus lead to the desired effect.
-However, note that when the text is transferred to another font (i.e.~one that includes real italics) and/or to another software application (like Apple Pages, which does not perform slanting), then this visual appearance will be lost.
-In this case we are thus still in the pre-Unicode situation in which the choice of font and rendering software actually matters.
-The ideal solution from a linguistic point of view would be the introduction of a new IPA code point for a different kind of which explicitly specifies that it should still be rendered as a double-story character when italicized.
-After informal discussion with various Unicode players, our impression is that this highly restricted problem is not sufficiently urgent to introduce even more -like characters in Unicode (which already lead to much confusion, see Section 4, Pitfall 4).
-This is a clear situation in which the Unicode Consortium is not just thinking about linguists, but has a more wide-ranging practical view to consider.)
-
-Some other homoglyphs related to encoding IPA in the Unicode Standard are:
-\begin{itemize}
-	\item The uses of the apostrophe has led to long discussions on the Unicode Standard email list.
-An English keyboard inputs \textless{}`\textgreater{} APOSTROPHE at U+0027, although the `preferred' Unicode apostrophe is the \textless{}'\textgreater{} RIGHT SINGLE QUOTATION MARK at U+2019.
-Yet the glottal stop/glottalization/ejective marker is another completely different character, the MODIFIER LETTER APOSTROPHE at U+02BC, but unfortunately looks mostly highly similar to U+2019.
-	\item There is ambiguous encoding of IPA segments within the Unicode Standard.
-An example is the U+02C1 MODIFIER LETTER REVERSED GLOTTAL STOP vs the U+02E4 MODIFIER LETTER SMALL REVERSED GLOTTAL STOP .
-Both are denoted in the Unicode Standard as the `pharyngealized diacritic' and both appear in various resources representing phonetic data online.
-This is thus an example for which the Unicode Standard does not solve the linguistic standardization problem.
-	\item There is at least one case in which the character name assigned by the Unicode Consortium does not match the IPA's description: in the Unicode Standard at U+01C3 is labeled LATIN LETTER RETROFLEX CLICK, but in IPA is an alveolar or postalveolar click (not retroflex).
-This naming is probably best seen as a simple error in the Unicode Standard.
-Note that most linguists simply seem to use or <ou>) while ligatures are single characters (e.g.
-<ʧ> LATIN SMALL LETTER TESH DIGRAPH at U+02A7).
-Ligatures arose in the context of printing easier-to-read texts, and are included in the Unicode Standard for reasons of legacy encoding.
-However, their usage is discouraged by the Unicode core specification.
-Specifically related to IPA, various phonetic combinations of characters (typically affricates) are available as single code-points in the Unicode Standard, but are designated as ``ligatures'' or ``digraphs'' (confusingly both names appear interchangeably).
-Such glyphs might be used by software to produce a pleasing display, but they should not be hard-coded into the text itself.
-In the context of IPA, characters like the following ligatures should thus \emph{not} be used.
-Instead a combination of two characters is preferred:
-	\item <ʣ> LATIN SMALL LETTER DZ DIGRAPH at U+02A3 (use <d> + <z> instead) 
-	\item <ʧ> LATIN SMALL LETTER TESH DIGRAPH at U+02A7 (use <t> + <ʃ> instead) 
-	\item <ʩ> LATIN SMALL LETTER FENG DIGRAPH at U+02A9 (use <f> + <ŋ> instead) 
-\end{itemize}
-
-However, there are a few Unicode characters that are historically ligatures, but which are today considered as simple characters in the Unicode Standard and thus should be used when writing IPA, namely:
-\begin{itemize}
-	\item <ɮ> LATIN SMALL LETTER LEZH at U+026E 
-	\item <œ> LATIN SMALL LIGATURE OE at U+0153 
-	\item <ɶ> LATIN LETTER SMALL CAPITAL OE at U+0276 
-	\item <æ> LATIN SMALL LETTER AE at U+00E6 
-\end{itemize}
-
-\subsection{Pitfall: The IPA notion of diacritics is not the same as the Unicode Standard's notion of diacritics}\label{pitfall-the-ipa-notion-of-diacritics-is-not-the-same-as-the-unicode-standards-notion-of-diacritics}
-
-Another pitfall is diacritics.
-The problem is that the meaning of the term `diacritics' as used by the IPA is not the same as it used in the Unicode Standard.
-Specifically, diacritics in the IPA-sense are either so-called \textsc{Combining Diacritical Marks} or \textsc{Spacing Modifier Letters} in the Unicode Standard.
-Crucially, Combining Diacritical Marks are by definition combined with the character before them (to form so-called default grapheme clusters, see Section 3).
-In contrast, Spacing Modifier Letters are by definition \emph{not} combined into grapheme clusters with the preceding character, but simply treated as separate letters.
-In the context of the IPA, the following IPA-diacritics are actually Spacing Modifier Letters in the Unicode Standard:
-\begin{itemize}
-	\item Length marks, namely <ː> MODIFIER LETTER TRIANGULAR COLON at U+02D0 and <ˑ> MODIFIER LETTER HALF TRIANGULAR COLON at U+02D1. 
-	\item Tone letters, like <˥> MODIFIER LETTER EXTRA-HIGH TONE BAR at U+02E5, and others like this. 
-	\item Superscript letters, like <ʰ> MODIFIER LETTER SMALL H at U+02B0 or <ˤ> MODIFIER LETTER SMALL REVERSED GLOTTAL STOP at U+02E4, and others like this. 
-	\item <˞> MODIFIER LETTER RHOTIC HOOK at U+02DE. 
-\end{itemize}
-
-Although linguists might expect these characters to belong together with the character in front of them, at least for <ʰ> MODIFIER LETTER SMALL H at U+02B0 the Unicode Consortium's decision to treat it as a separate character is also linguistically correct, because according to the IPA it can be used both for aspiration (more precisely postaspiration following the base character) and preaspiration (preceding the base character).
-Note that there is a mechanism in Unicode to force separate characters to be combined (namely by using the ZERO WIDTH JOINER at U+200D), but this seems to be a rather impractical, and probably not enforceable solution to us.
-
-\subsection{Pitfall: Neither the IPA nor the Unicode Standard enforce a unique diacritic ordering}\label{pitfall-neither-the-ipa-nor-the-unicode-standard-enforce-a-unique-diacritic-ordering}
-
-Also related to diacritics is the question of ordering.
-To our knowledge, the International Phonetic Association does not specify a specific ordering for diacritics that combine with phonetic base symbols; this exercise is left to the reasoning of the transcriber.
-However, such marks have to be explicitly ordered if sequences of them are to be interoperable and compatible.
-An example is a labialized aspirated alveolar plosive: <tʷʰ>.
-There is nothing holding linguists back from using <tʰʷ> instead (with exactly the same intended meaning).
-However, from a technical standpoint, these two sequences are different, e.g.~if both sequences are used in a document, searching for <tʷʰ> will not find any instances of <tʰʷ>, and vice versa.
-Likewise, a creaky voiced syllabic dental nasal can be encoded in various orders, e.g.
-<n̪̰̩>, <n̩̰̪> or <n̩̪̰>.
-
-In accordance with the absence of any specification of ordering in the IPA, the Unicode Standard likewise does not propose any standard orderings.
-Both leave it to the user to be consistent.
-However, there is one aspect of ordering for which the Unicode Standard does present a canonical solution, which is uncontroversial from a linguistic perspective.
-Diacritics in the Unicode Standard (i.e.~Combining Diacritical Marks, see above) are classified in Canonical Combining Classes.
-In practice, the diacritics are distinguished by their position relative to the base character.\footnote{See http://unicode.org/reports/tr44/\#Canonical\_Combining\_Class\_Values for a detailed description.) When applying a Unicode normalization (NFC or NFD, see previous section), the diacritics in different positions are put in a specified order.
-This process therefore harmonizes the difference between different encodings, for example, of a midtone creaky voice vowel <ḛ̄>.
-This grapheme cluster can be encoded either as <e> + <̄> + <̰> or as <e> + <̰> + <̄> .
-To prevent this twofold encoding, the Unicode Standard specifies the second ordering as canonical (diacritics below before diacritics above).
-
-When encoding a string according to the Unicode Standard, it is possible to do this either using the NFC (``composition'') or NFD (``decomposition'') normalization.
-Decomposition implies that precomposed characters (like <á> LATIN SMALL LETTER A WITH ACUTE at U+00E1) will be split into its parts.
-This might sound preferable for a linguistic analysis, as the different diacritics are separated from the base characters.
-However, note that most attached elements like strokes (e.g.~in the <ɨ>), retroflex hooks (e.g.~in <ʐ>) or rhotic hooks (e.g.~in <ɝ>) will not be decomposed, but strangely enough a cedilla (like in <ç>) will be decomposed.
-In general, Unicode decomposition does not behave like a feature decomposition as expected from a linguistic perspective.
-It is thus important to consider Unicode decomposition only as a technical procedure, and not assume that it is linguistically sensible.
-
-Facing the problem of specifying a consistent ordering of diacritics while developing a large database of phonological inventories from the world's languages, Moran (2012: 540) defines a set of diacritic ordering conventions.\footnote{The most recent version of these conventions is online: http://phoible.github.io/conventions/) The conventions are influenced by the linguistic literature, though some ad-hoc decisions had to be taken.
-The goal was to explicitly define all character sequences so that the vast variety of phonemes found in descriptions of the world's language were normalized into consistent character sequences, e.g.~if one language description uses and another , when both are intended to be phonetically equivalent, then a decision to normalize to one form was taken.
-For example, when a character sequence contains more than one character in Spacing Modifier Letters, the order that is proposed is the following (where <\textbar{}> indicates ``or'' and <→> indicates ``precedes''):
-\begin{itemize}
-	\itemsep1pt\parskip0pt\parsep0pt 
-	\item \textsc{Spacing Modifier Letters ordering:} ( unreleased <̚> \textbar{} lateral release <ˡ> \textbar{} nasal release <ⁿ>) → ( palatalized <ʲ>) → ( labialized <ʷ>) → ( velarized <ˠ>) → ( pharyngealized <ˤ>) → ( aspirated <ʰ> \textbar{} ejective <ʼ>) → ( long <ː> \textbar{} half-long <ˑ>)
-\end{itemize}
-
-If a character sequence contains more than one diacritic below the base character, then the place feature is applied first (dental, laminal, apical, fronted, backed, lowered, raised), then the laryngeal setting (voiced, voiceless, creaky voice, breathy voice) and finally the syllabic or non-syllabic marker (for vowels, ATR gets put on between the place and laryngeal setting).
-So:
-\begin{itemize}
-	\itemsep1pt\parskip0pt\parsep0pt 
-	\item \textsc{Combining Diacritical Marks (below) ordering:} ( dental <t̪> \textbar{} laminal <t̻> \textbar{} apical <t̺>) → ( fronted <u̟> \textbar{} backed <e̠>) →( lowered <e̞> \textbar{} raised <e̝>) → ( ATR <e̘ e̙>) → ( voiced <s̬> \textbar{} voiceless <n̥> \textbar{} creaky voice <b̰> \textbar{} breathy voice <b̤>) → ( syllabic <n̩> \textbar{} non-syllabic <e̯>)
-\end{itemize}
-
-Character sequences with diacritics above the base character were not problematic in Moran 2012 because they include only the centralized, mid-centralized and nasalized combining characters.
-Moran (2012) marks tones as singletons with Space Modifier Letters, e.g.
-\textless{}˦\textgreater{} for a phonemic high tone, instead of accent diacritics, alleviating potential conflicts.
-Building on the work of Moran (2012), if a character sequence contains more than one diacritic above the base character, we propose:
-\begin{itemize}
-	\itemsep1pt\parskip0pt\parsep0pt 
-	\item \textsc{Combining Diacritical Marks (above) ordering:} (centralized <ë> \textbar{} mid-centralized <e̽>) → (extra short <ĕ>) →( tone accents, e.g.
-<è> ) → ( Spacing Modifier Letters ) → ( tone letters, e.g.
-<e˦>)
-\end{itemize}
-
-\section{6. Orthography profiles}\label{orthography-profiles}
-
-\subsection{Characterizing writing systems}\label{characterizing-writing-systems}
+\section{Characterizing writing systems}
+\label{characterizing-writing-systems}
 
@@
 In most cases, this amounts to more than simply mapping a grapheme to an IPA segment because graphemes must first be identified in context (e.g.~is the sequence one sound or two sounds or both?) and strings must be tokenized, which may include taking orthographic rules into account (e.g.
-between vowels is /n/ and after a vowel but before a consonant is a nasalized vowel /ṽ/).
+between vowels is /n/ and after a vowel but before a consonant is a nasalized vowel /ṽ/).
 In our experience, data from each source must be individually tokenized into graphemes so that its orthographic structure is identified and its contents can be extracted.
@@
 	\item code points (10 text elements): t s ʰ o ̃ ̰ ´ s h i 
-	\item grapheme clusters (7 text elements): t s ʰ ṍ̰ s h i 
-	\item tailored grapheme clusters (4 text elements): tsʰ ṍ̰ sh i 
+	\item grapheme clusters (7 text elements): t s ʰ ṍ̰ s h i 
+	\item tailored grapheme clusters (4 text elements): tsʰ ṍ̰ sh i 
 \end{enumerate}
 
-In (1), the string <tsʰṍ̰shi> has been tokenized into ten Unicode code points (using NFD normalization), delimited here by space.
+In (1), the string <tsʰṍ̰shi> has been tokenized into ten Unicode code points (using NFD normalization), delimited here by space.
 Unicode tokenization is required because sequences of code points can differ in their visual and logical orders.
-For example, <õ̰> is ambiguous to whether it is the sequence of + <̰> + <̃> or + <̃> + <̰>.
+For example, <õ̰> is ambiguous to whether it is the sequence of + <̰> + <̃> or + <̃> + <̰>.
 Although these two variants are visually homoglyphs, computationally they are different.
@@
 In (2), the Unicode code points have been logically normalized and visually organized into grapheme clusters, as specified by the Unicode Standard.
-The combining character sequence <õ̰> is normalized and visually grouped together.
+The combining character sequence <õ̰> is normalized and visually grouped together.
 Note that , the MODIFIER LETTER SMALL H at U+02B0, is not grouped with .
@@
 
-\subsection{Informal description of orthography profiles}\label{informal-description-of-orthography-profiles}
+\section{Informal description of orthography profiles}
+\label{informal-description-of-orthography-profiles}
 
@@
 
-\subsection{Formal specification of orthography profiles}\label{formal-specification-of-orthography-profiles}
+\section{Formal specification of orthography profiles}
+\label{formal-specification-of-orthography-profiles}
 
@@
 Hashes somewhere else in the file are to be treated literally, i.e.~hashes are only to be ignored when they occur at the start of a line.\footnote{Comments that belong to specific lines will have to be put in a separate column of the CSV file, e.g.~add a column called `comments'.
-Further, if the content of a profile contains a hash at the start of a line, either reorder the columns so the hash does not occur at the start of the line, or add a dummy column in front of the data to not have the data start with a hash.)
+Further, if the content of a profile contains a hash at the start of a line, either reorder the columns so the hash does not occur at the start of the line, or add a dummy column in front of the data to not have the data start with a hash.}
 	\item \textsc{Metadata are given in commented lines at the beginning of the text file in a basic ``tag: value'' format.
@@
 \end{enumerate}
-
-\section{7. Use cases}\label{use-cases}
-
-We now present several use cases that have motivated the development of orthography profiles. These include:
-\begin{itemize}
-	\item grapheme tokenization 
-	\item normalization of orthographic systems to attain interoperability across different source documents, 
-	\item cognate identification for detecting the similarity between words from different languages 
-\end{itemize}
-
-An important assumption of our work is that input sources are encoded in Unicode (UTF-8 to be precise).
-Anything that the Unicode Standard is unable to capture, cannot be captured by an orthography profile.
-We also use Unicode normalization form NFD to decompose all incoming text input into normalized and (Unicode) logically ordered strings.
-This process is of fundamental importance when working with text data in the Unicode Standard.
-
-\emph{Grapheme tokenization}
-
-Once text has been normalized, an additional and straightforward process is Unicode grapheme tokenization.
-More recently this regular expression, often available as short cut ``\X'', identifies all sequences of ``base'' characters followed by 0 or more combining diacritics.
-For example, the sequence, <ŋ͡mṍraː> would tokenize as a space-delimited sequence as: <ŋ͡ m ṍ r a ː>.
-For a first pass at identifying grapheme clusters, this is a straightforward tokenization.
-However, the linguist will note that both the tie bar (COMBINING DOUBLE INVERTED BREVE, a Unicode Combining Modifier) and the length marker (MODIFIER LETTER TRIANGULAR COLON.
-a Unicode Letter Modifier) do not appear ``correctly'', i.e.~the tie bar is grouped with the first character in the sequence and the length marker appears singly.
-This is necessary, as defined by the Unicode Standard, because of these character's semantic properties -- with Unicode we cannot know if, for example, the MODIFIER LETTER TRIANGULAR COLON should appear before or after the base character that it modifies, cf.~the IPA aspiration marker, , the MODIFIER LETTER SMALL H, which is also a Unicode Letter Modifier and for linguists a diacritic that be be used for pre- or post-aspiration.
-These ambiguities of position in tokenization are not uncommon in IPA, thus orthographic (or source) normalization and tokenization is needed.
-
-*Cross-orthography **normalization*
-
-Orthographic tokenization requires a specification of the tailored grapheme clusters (linguists: graphemes) to separate orthographic units within an input source.
-In the simplest case, the Unicode graphemic tokenization with regular expression ``\X'' captures the grapheme clusters, i.e.~the input data is graphemically tokenized and there are no additional string sequences required to capture each grapheme, e.g.~no specification.
-This scenario is the most unlikely.
-
-The creation of an orthography typically requires that someone describes knowledge about the input data source's writing system.
-For example, when orthographically tokenizing a dictionary as input, the sequences of grapheme clusters need to be explicitly stated or a regular expression rule can be specified to transform the input.
-We illustrate these methods with examples from Thiesen 1998, a bilingual dictionary of Diccionario Bora-Castellano, and with Huber \& Huber 1992, a comparative vocabulary of sixty-nine indigenous languages of Columbia.
-
-Huber \& Huber 1992
-
-({[}a\textbar{}á\textbar{}e\textbar{}é\textbar{}i\textbar{}í\textbar{}o\textbar{}ó\textbar{}u\textbar{}ú{]})(n)(\s)({[}a\textbar{}á\textbar{}e\textbar{}é\textbar{}i\textbar{}í\textbar{}o\textbar{}ó\textbar{}u\textbar{}ú{]}), \1 \2 \4
-
-A preliminary orthography profile can be generated with code available online.\footnote{https://github.com/bambooforest/tokenizer) The Python script creates a unigram model for individual Unicode characters or Unicode grapheme clusters and their frequency in the source input.
-We find that this gives a broad overview of the input at the character and grapheme levels and it oftens finds typographic errors in the source input by identifying rarely occurring or one-off characters in the input.
-The models also provides the basis for the ``quick'' development of an initial orthography profile by saving the user keyboarding time.
-
-Note that these symbols are still phonetically unspecified, i.e.~symbol may or may not represent, say, a front open unrounded vowel.
-
-The orthography profile is designed so that specified strings of characters are captured for tokenization.
-
-\% how the tokenization works
-
-The graphemic cluster specifications in the orthography profile are read into a trie data structure, which is used to tokenize the incoming input source by identifying sentinels (tokenization boundaries) in the input stream, which we then separate grapheme clusters by spaces.
-
-An illustration is given in Figure N.
-
-Comparative wordlists often start as the compilation of different sources with different writing systems.
-The Dogon comparative wordlist is an example of a compilation of slightly divergent transcription practices, which we will use to illustrate our point and to orthographically normalize an input source by transforming different transcriptions into a single IPA-like transcription.
-
-*Orthographic normalization / cross-source **normalization*
-
-Once we have normalized and tokenized our text elements, we can apply various sorts of source normalization across resources to create interoperable and comparable data.
-
-Sound-based normalization is practical because \ldots{}
-
-We use orthography profiles in our work to describe cross-linguistic difference between writing systems.
-
-\emph{Cognate detection}
-
-The identification of cognates is a non-trivial task, which involves identifying the semantic and phonetic similarity of words.
-
-use grapheme correspondences to approximate cognacy and sound correspondences
-
-to identify cognates and sound changes
-
-Examples
-
-Orthographic similarity != Phonetic similarity != Regular sound change
-
-Cognate identification is further complicated by the fact that languages are written differently, and even more so for the low resource languages that we work with, linguists' phonetic transcriptions are heterogeneous and typically document-specific, i.e.~two linguists working on the same language variety most often develop different practical orthographies by using different symbols and different orthographic rules.
-Thus the essence of language documentation and description is diverse and ephemeral in nature.
-The encoding of linguistic data and linguistic data models is diverse and abstracting away from individual analytical preferences requires a document-by-document approach and a mechanism for standardization of writing to form an interlingual pivot so that different descriptions can be normalized into one encoding system to undertake analysis.
-
-\% Orthography profile implementation in Lingpy / generic tokenizer
-
-\% describe Tokenizer's function
-
-\% describe algorithm for tokenizing IPA (Unicode grapheme cluster tokenization and then alignment of Letter Modifiers and Tie Bars)
-
-\% insert an example of orthographic tokenization from PAD
-
-\section{move Letter Modifiers left}\label{move-letter-modifiers-left}
-
-(\s)(\p{Lm}), \2
-
-which catches:
-
-\section{Diacritics: Unicode letter modifiers (Lm)}\label{diacritics-unicode-letter-modifiers-lm}
-
-\section{({[}ː\textbar{}ʰ\textbar{}ʼ\textbar{}ʿ\textbar{}ʾ{]})}\label{ux2d0ux2b0ux2bcux2bfux2be}
-
-from the set of this:
-
-\section{Diacritics: all}\label{diacritics-all}
-
-\section{({[}ː\textbar{}ʰ\textbar{}ʼ\textbar{}ʿ\textbar{}ʾ\textbar{}̜\textbar{}̠\textbar{}̟\textbar{}͡\textbar{}̪\textbar{}̰\textbar{}̣\textbar{}̥\textbar{}̩\textbar{}‿\textbar{}.\textbar{}→\textbar{}˻\textbar{}+\textbar{}̃\textbar{}̫\textbar{}̤{]})}\label{ux2d0ux2b0ux2bcux2bfux2be.}
-
-and allows us to with one regular expression do this:
-
-PAD examples
-
-A side effect of character tokenization is identification of outliers, e.g.
-
-Some examples in Python
-
-Some examples in R
-
-{[}UNICODE USE-CASES{]}
-
-The most basic overall Unicode character property is the \textsc{General Category}, which categorizes Unicode characters into: \emph{Letters, Punctuation, Symbols, Marks, Numbers, Separators, }and* Other*.
-Each Unicode character property also has a character property value.
-So each General Category is denoted by a property value by a single letter abbreviation, e.g.
-``L'' for Letter.
-Each General Category property also has subcategories that are also properties with property values.
-For example, ``Letters'' is broken down into Uppercase Letter (Lu), Lowercase Letter (Ll), Titlecase Letter (Lt), Modifier Letter (Lm) and Other Letter (Lo).\footnote{The subcategory ``Other Letter'' includes characters such as uncased letters as in the Chinese writing system.) By defining character properties for each code point, algorithmic implementations such as text processing applications or regular expression engines that are conformant to the Unicode Standard can be used to recognize whole (sub)categories of characters or ranges of characters, e.g.~to match proper names in English in title case or more generally to match character types, such as scripts, or symbols and integers in email addresses, phone numbers, etc.
-
-Script information is useful for algorithmic processes, such as collation, searching and for multilingual text processing.
-For example, if a regular expression engine includes functionality to match characters at the level of script, then it provides the ability to identify characters, e.g.
-\match{Greek}, which tests whether letters in a text belong to the Greek script or not.
-This functionality can be used to determine the boundaries between different writing systems that appear in the same document.
-In Section 6 (Use cases of grapheme tokenization) we provide an example of a regular expression match at the level of character property.
-
-Unicode-aware regular expression engines may contain a meta-character, ``\X'', that matches this level of grapheme cluster, i.e.~a base character followed by some number of accents or diacritical marks.
-Grapheme clusters are important to determine word or line boundaries, for collation (`ordering') and for user interface interaction (e.g.~mouse selection, backspacing, cursor movement).
-
-\section{9. Conclusion}\label{conclusion}
-
-In our work we focus on the writing systems used in lesser-described and endangered languages and our orthography profiles are used to describe writing systems and to transpose them into phonetic transcription.
-Of course the International Phonetic Alphabet and other transcription systems are essentially just orthographies that are an approximation of sound.
-Nevertheless, sound-based normalization is practical in automatically identifying cognates and sound changes with quantitative methods.
-We leverage orthography profiles to enable comparative analysis of languages with different writing systems.
-
-An orthography profile lists the graphemes in a particular description of language data, e.g.~a wordlist, dictionary or corpus.
-Building on the knowledge that can be extracted from that description by tokenizing words by the graphemes made explicit in the orthography profile, it is straightforward to undertake other analyses of the data.
-For example, various ngram models of the data can be extracted with a few lines of code.
-A unigram model with counts, frequencies and positive log probability (``plogs''; ``entropy under this model'') (Goldsmith \& Riggle, 2010) provides a fair amount of information about a given data source.
-Essentially, the orthography profile provides the description that allows this information to be calculated based on the mapping of sequences of characters into graphemic units.
-Bigram models are also straightforwardly extracted.
-In the case of bigrams, mutual information can be captured and used in various other statistical analyses, such as quantitative language comparison, inferring phylogenetic trees, etc.
-
-\section{10. Nice text blocks to be included somewhere}\label{nice-text-blocks-to-be-included-somewhere}
-
-Looking at the variety of writing systems used for the world's languages, there are various challenges for the computational processing of resources in which such writing systems are used.
-
-A widespread mechanism to derive new symbols for a writing system is to use a combination of multiple symbols already available.
-That can lead to completely new symbols, like <æ> or <w>, which are still rather transparently derived from <a e> and <v v>, respectively.
-More typically, many orthographies use sequences of symbols as special characters, like <sch> or <gh>.
-To identify such `grapheme clusters' is one of the central problems to be tackled with orthography profiles.
-A further frequently used mechanism is to add extra material to an already available symbol, either to make it better visible (this is probably the reason for the addition of a dot on top of the <i>) or to derive new symbols.
-The widespread use of diacritics is the obvious result of such processes (Gaultney 2002).
-However, also the derivation of new symbols by added extra strokes is widespread (and not only in Far-Eastern character-based writing systems), like for example <ŋ> or <ç>.
-
-The use of punctuation (Daniëls 1994, Humez \& Humez 2008) and spaces (Saenger 1997) to improve readability. {[}e.g.~space after punctuation, but not before{]}
-
-Currently, the Unicode Standard is the standard character encoding for linguistic data (Anderson 2003).
-However, there are still hundreds of different encoding systems that were invented independently to capture orthographic diversity as different nations adopted and developed computer systems.
-These different encoding systems are problematic and in conflict with one another because different standards were formalized differently and for different purposes by different standards committees in different countries.
-No unified encoding scheme contained enough code points to encode all characters, so two different encoding schemes possibly used the same code point for different characters, or used different code points to represent the same character.
-Because computers support multiple character encoding schemes, data risked being corrupted when handled by different applications and encodings.
-The Unicode Standard was devised to alleviate these problems.
-
-For orthographies, identifying graphemes can be even more challenging than identifying phones and phonemes in phonetic transcription because although transcriptions may not adhere strictly to IPA, they tend to have straightforward mappings between sounds and symbols.
-On the other hand, orthographies can introduce orthographic rules, which add an additional challenges in identifying graphemes in words, as mentioned above.
-Thus for resources not in IPA or IPA-like transcriptions, graphemes often must be manually identified, whether they are encoded as singletons or multi-character sequences.
-The identification of graphemes and the formulation of orthographic rules are used to create an orthography profile.
-
-Character-to-glyph mappings are particularly not clear-cut for linguists using the IPA.\footnote{We give more linguistic examples in Section 5.) Just look in the IPA Extensions block.\footnote{Character ``blocks'' are described in Section 3 (The Unicode approach).) There are several characters that are glyphs (variants) of the same grapheme in the Latin block, e.g.
-and \& and .\footnote{LATIN SMALL LETTER ALPHA U+0251 vs LATIN SMALL LETTER A U+0061 , and LATIN SMALL LETTER SCRIPT G U+0261 vs LATIN SMALL LETTER G U+0067 .) Other characters like
-
-, , do not appear at all in the IPA Extensions block, but as separate phonetic symbols; they are already encoded in the Basic Latin, Latin Extended-A and Greek \& Coptic blocks.
-When a linguist transcribes an IPA
-
-on a QWERTY keyboard, if it is from the Basic Latin group, it is valid Unicode IPA (it is not valid IPA if it is from the Cyrillic Block).
-However, keyboard and and because they have been given specific code points in the IPA Extensions block, where they have been given special status as IPA characters, and which most often go unnoticed by linguists.\footnote{In Section 5 we discuss in detail the Unicode IPA alphabet and its characters dispersal across different Unicode Standards scripts and blocks.)
-
-The Unicode Standard divides text elements roughly into two classes of graphemes.
-The cover term \textsc{grapheme cluster}\footnote{Formerly referred to as ``locale-independent graphemes'' to emphasize that the term grapheme is used differently in linguistics.) is used to emphasize the difference between the term \emph{grapheme }as used by linguists and what the users perceive as a character (a basic unit of a writing system for a language), which may not be a single Unicode character.\footnote{http://www.unicode.org/reports/tr29/) The Unicode Consortium makes a further distinction between grapheme clusters and tailored grapheme clusters.
-
-The simplest example of a grapheme cluster is essentially a base character followed by a Letter Modifier character.
-For example, the sequence + \textless{}\textasciitilde{}\textgreater{} (COMBINING TILDE at U+0303) combines visually into , a user-percieved character in writing systems like that of Spanish.\footnote{Although the code point sequence can also be represented with a single code point, LATIN SMALL LETTER N WITH TILDE at U+00F1, it is not the case that all multi-code point character sequences have a single code point equivalent, which can be deduced from normalization.
-Often for commonly-used letters in European languages, a single code point form exists.)
-
-What the user perceives as a single character may actually involved a multi-code point sequence.\footnote{Unicode-aware regular expression engines may contain a meta-character, ``\X'', that matches this level of grapheme cluster, i.e.~a base character followed by some number of accents or diacritical marks.
-As mentioned, grapheme clusters are important for collation algorithms, but they are also used to determine other processes such as word or line boundaries and for user interface interaction (e.g.~mouse selection, backspacing, cursor movement).)
-
-More challenging for text processing is what the Unicode Consortium terms tailored grapheme clusters.
-For example, the sequence + for the Slovak digraph or the sequence .
-These grapheme clusters are ``tailored'' in the sense that they must be specified on a language-by-language basis.
-Algorithms can be adapted to use as input these tailored language-specific orthographies for transliteration, localization, etc.
-In fact, the Unicode Standard provides a method for defining language-specific tailored writing systems called the Unicode Locale Description.
-Unicode Locale Descriptions allow users to define a set of extended grapheme clusters.
-Unicode Locale Descriptions are saved in the Common Locale Data Repository (CLDR),\footnote{http://cldr.unicode.org/) a repository of language-specific definitions of writing system properties, each of which describes specific usages of characters.
-Each can be encoded in a Unicode Locale Data Markup Language (LDML) document.
-LDML is an XML format and vocabulary for the exchange of structured locale data.
-It is a format used not only for CLDR, but also for general interchange of locale data, such as in Microsoft's .NET.
-We return to LDML in Section 5 (Orthography profiles and locale descriptions).
-
-We have noted that the rendering of electronic text requires that characters are encoded and stored in a computer's memory.
-These character sequences are then mapped from character code points to glyphs by software that takes into account a character's context.
-Transparent to most users is the fact that different logical orders of characters may result in the same visual output.
-For example, phonetic characters with certain combinations of diacritics may be homoglyphs (see below), even while the logical order of their character sequences are not the same.\footnote{One example is a vowel that is both nasalized and creaky voice, e.g.
-{[}ḛ̃{]}.
-See discussion below.) Thus some type of standardization, what the Unicode Consortium calls ``normalization'' of the ordering of (some classes of) characters, is required to make sure that all data are logically consistent and therefore easily comparable and equally searchable.
-The aim of standardization is to attain syntactic \textsc{interoperability} of data.\footnote{By interoperability we mean the ability to ubiquitously exchange and merge disparate (textual) data sets, and data encoding formats, to facilitate data sharing and to ``effortlessly'' undertake comparison and analysis.
-For a discussion of syntactic versus semantic interoperability, see Ide and Pustejovsky, 2010.)
-
-The IPA is often used by linguists in undertaking initial phonetic and phonemic analyses of languages.
-However, once the distinctive sounds of a given language have been identified (and a phonological description of the target language has been described), a linguist may continue his or her documentation and description of higher-order structures like words and sentences with a broad phoneme-based alphabet.
-In fact, literary materials development often leapfrogs from a phonemic alphabet to locale-specific practical orthographies, in which numerous factors affect a language's orthography development, e.g.~politics, purpose, speaker communities familiarity with other writing systems, technological limitations.
-
-One simple example is illustrated with two genealogically related languages spoken in Northern Ghana: Western Sisaala\footnote{Western Sisaala lacks /kp/.) and Sisaala Tumulung.\footnote{Sisaala Tumulung (ISO 639-3: sil) is spoken around the city Tumu, which is about 50 kilometers geographically to the east of Western Sisaala (ssl) speaking areas.) Both languages contain the same basic set of contrastive sounds, but their orthographies encode affricates (and vowels) differently graphemically; the influence stem from orthographies of government-sponsored literacy languages, which are taught to children in schools in these areas.
-In Western Sisaala the sound /tʃ/ is written and /dʒ/ as due to familiarity of these graphemes by students learning Dagaare (Moran 2006).
-Spoken further to the east, Sisaala Tumuling encodes the sound /tʃ/ as and /dʒ/ as (Blass 1975).
-
-If encoded in the Unicode Standard, the design of both orthographies has to specify that certain sequences of characters, such as + , are single graphemes (compare English and which is ambiguous).
-Design is important, for example, for collation ordering (sorting), i.e.~the sequence of characters that is treated as a unit for ordering purposes, such as in dictionaries.
-Indeed, the characters , and are three unique graphemes consisting of two Unicode characters from the Latin script, but developers of a software program that recognizes the Western Sisaala alphabet might want to specify the sort order , , \ldots{} .
-Thus there is a need for orthography-specific specifications that define sequences of characters as graphemes, whether technologically, linguistically, or both, for multilingual text processing.
-
-A major challenge that the developers of the Unicode Standard faced is that there exists no universal set of fundamental units of text because the divisions of text into elements, such as graphemes, is dependent on language-specific writing systems and orthographies, as just shown.
-Moreover, the division of electronic text into elements is process dependent, e.g.~sorting and spell checking may or may not need to take capitalization into account, depending on the orthography in question.
-Thus the encoding and processing of text requires well defined \textsc{text elements} and rules for interpreting them; these elements and rules are language-specific when we talk about major languages or well established orthographies, but more often in our research we face input from lesser-studied languages documented in author-specific transcription or practical orthography.\footnote{And this is usually down to the document-specific level, the description of which has been coined ``doculect'' for purposes of discussing differences of opinion in language descriptions (Cysouw and Good, 2013).)
-
-In Section XXX we noted that the Unicode Consortium makes a crucial distinction between characters and glyphs.
-The relations between characters and glyphs is a complex one and it is typically hidden from the user.
-Technologically, we must distinguish between characters and glyphs because:
-\begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item The logical order of a sequence of characters (in computer memory) may not be the same as the visual order of their (rendered) glyphs. 
-	\item There is not always a one-to-one mapping between characters and glyphs. 
-\end{enumerate}
-
-Regarding the first point, a user cannot assume that contiguously rendered text on-screen actually reflects the computer's internal storage order of individual and sequential characters in memory, i.e.~the \textsc{logical order}.
-In other words, in the Unicode Standard the visual order of glyphs may not be the same as the logical order of their digitally encoded characters; contiguous display is not indicative of contiguous text.* *A nice example is Hebrew text.\footnote{http://www.macchiato.com/unicode/globalization-gotchas) Another example is the glyph <ṍ>, which may consist of one, two or three characters in encoded in various physical orders.
-The possible mismatch in sequences becomes clearer when we consider the second point.
-
-When a base character, i.e.~any graphic character not in the General Category of Combining Mark, is followed by one or more characters having the property General Category of Combining Mark, they can be combined visually into one or more glyphs using the Canonical Ordering Algorithm to create Unicode \textsc{grapheme clusters}.
-As we have shown, this input may be ambiguous from a technological perspective if normalization is not taken into account.
-For example, a linguist may insert a single character or a character sequence for to denote a high back unrounded vowel with high tone:
-\begin{itemize}
-	\item LATIN SMALL LETTER U WITH ACUTE (U+00FA) 
-	\item LATIN SMALL LETTER U (U+0075) + COMBINING ACUTE ACCENT (U+0301) 
-\end{itemize}
-
-The normalization algorithms in the Unicode Standard are used to make these character sequences canonically equivalent for purposes such as ubiquitous search and comparison.
-
-However, the standard does not specify many (or most) language-specific graphemes, or in Unicode parlance, \textsc{tailored grapheme clusters}, such as English .\footnote{Some (unfortunate) exceptions included for legacy purposes, and used incorrectly by linguists, include digraphs like LATIN SMALL LETTER TS DIGRAPH at U+02A6 or LATIN SMALL LETTER DEZH DIGRAPH at U+02A4.
-These and other digraphs are encoded in the Unicode Standard's IPA Extensions block.) For technological reasons, these language-specific tailored grapheme clusters must be defined outside of the Unicode Standard, in what are called (language-specific) Unicode Locale Descriptions.\footnote{http://www.unicode.org/reports/tr35/) The encoding of these multiple character sequences in Locale Descriptions is beyond the need (or technical ability) of many linguists; thus we have developed a simpler approach which we used in our own research and describe in detailed in Section XXX.
-Basically, the description of writing systems for lower resource languages, and in particular for document-specific descriptions (e.g.~different descriptions of the same language with different orthographies), requires an orthographic description at the (tailored) grapheme (cluster) level.
-
-An orthography profile is defined as a simple delimited flat text file encoded in the UTF-8, with character sequences in Unicode Normalization Form D.
-Metadata aside, the column delimited contents of an orthography profile are essential a state-table, i.e.~an implementation of a state transition table, but for ease of use encoded in a simple flat text file.
-
-In an orthography profile, all character strings are converted into NFD because each character in a grapheme may have phonetic value, and by using NFD, all graphemes are decomposed into a standardized order.
-For example, a vowel that is both nasalized and creaky looks like in IPA, where is some vowel.
-Although visually the same, a nasalized and creaky vowel can be composed of several different character sequences, as illustrated with in 1-3:
-\begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item <õ̰> LATIN SMALL LETTER O + COMBINING TILDE + COMBINING TILDE BELOW: U+006F + U+0303 + U+0330 
-	\item <õ̰> LATIN SMALL LETTER O + COMBINING TILDE BELOW + COMBINING TILDE: U+006F + U+0330 + U+0303 
-	\item <õ̰> LATIN SMALL LETTER O WITH TILDE + COMBINING TILDE BELOW: U+00F5 + U+0330 
-\end{enumerate}
-
-Applying NFD to these three character sequences results in one standard sequence; in this example the character sequence is that shown 1.
-NFD makes different sequences of input interoperable and it retains all of the phonetic information captured by the separate characters that combine to form a vowel with nasalization and creaky voice phonation.
-This is an important step to consider in linguistic research because each character has phonetic value, i.e.~the combining tilde denotes that the base character is produce by a lowering of the velum, which allows air to escape through the nose when the sound is produced.
-The tilde below also has meaning: it represents a phonation type in which the vocal cords are vibrating, but they are also being held slightly further apart, allowing more air to escape, which produces a murmured sound.
-To enable comparative analyses of languages at a deep phonetic level, retaining a code point for each character of phonetic value is deciduous.
-For example, if the sound systems of two languages are to be compared, one would not want to lose a possibly contrastive phonological feature like nasalization, which may have been collapsed into a single character in its logical representation.
-
-Our approach is similar in nature to the LDML, but we use a simpler state-table method, instead of generating an XML format for the ``exchange of structured locale data''.
-An orthography profile is a specification of the tailored grapheme clusters in a particular source document and any orthographic re-write rules needed after the application of Unicode normalization and orthographic profile tokenization.
-Orthography profiles can of course be used for transliteration between scripts, e.g.~from practical orthography to IPA and vice versa, by specifying mapping (replacements) from the list of tailored grapheme clusters in the source document and a set of replacement characters.
-This will be made clear in the next section in which we provide some use cases.
-
-To summarize, LDML is basically overkill for our intentions of identifying and delimiting grapheme boundaries (or in Unicode parlance ``tailored grapheme clusters'').
-LDML is appropriate, even necessary, for complex localization complexes for software applications -- localization of, say, an operating system or software application across cultures may require the instantiation of different date stamps an other local standards.
-For linguistic purposes of source document data tokenization, we therefore develop a simple-to-implement process of ``orthographic'' tokenization, using standard Unicode Standard-specified processes (e.g.~normalization and grapheme cluster tokenization) with an additional level of linguistic ``tailored grapheme cluster'' specification.
-
-\textsc{REFERENCES}
-
-Nancy Ide and James Pustejovsky.
-2010.
-What does inter- operability mean, anyway?
-toward an operational def- inition of interoperability for language technology.
-In Proceedings of the Second International Conference on Global Interoperability for Language Resources.
-Hong Kong, China.
-
-Huber, Randal Q. and Robert B. Reed (compilers).
-\begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} \setcounter{enumi}{1991} 
-	\itemsep1pt\parskip0pt\parsep0pt 
-	\item Vocabulario comparativo: Palabras selectas de lenguas indígenas de Colombia (Comparative vocabulary: Selected words in indigenous languages of Colombia).
-Bogota: Instituto Lingüísti- co de Verano.
-\end{enumerate}
-
-Thiesen, W.
-(1998).
-Diccionario bora-castellano, castellano-bora (1st ed.).
-Lima Perú: Instituto Linguístico de Verano.Retrieved from: http://www.worldcat.org/title/diccionario-bora-castellano-castellano-bora/oclc/40505215\&referer=brief\_results
-
-\textsc{APPENDIX - ORTHOGRAPHY PROFILE EXAMPLE}
-
-\textsc{APPENDIX - UNICODE IPA}
-
-Provide a clear and full Unicode IPA chart with additional information as an appendix
-\begin{verbatim}
-	- create this chart
-
-	- perhaps we should spell out conformance rules with regard to this chart? i.e. if a source claims conformance to our Unicode IPA chart, it should:
-
-	- identify the sections (or range) of the table that the data conforms to
-
-	- the source is in accordance with not only the Unicode IPA characters, but their intrinsic meanings (e.g.
-an Africanist might use <y> for /j/, but this level of conformance requires that <j> symbolizes /j/, etc.)
-
-	- someone who wants to use an Africanist transcription system can declare that they use the Unicode IPA chart, except for a range of characters (but they provide the mapping to Unicode IPA chart) 
-\end{verbatim}
-
-\textsc{Notes}
-
-\url{http://www.unicode.org/reports/tr29/}
-
-The default boundary determination mechanism specified in this annex provides a straightforward and efficient way to determine some of the most significant boundaries in text: user-perceived characters, words, and sentences.
-Boundaries used in line breaking (also called \emph{word wrapping}) are defined in {[}\href{http://www.unicode.org/reports/tr41/tr41-11.html\#UAX14}{UAX14}{]}.
-
-\textsc{Nice quotes}
-
-vanZundert: ``How does one design and plan a general unified infrastructure for a target that is all over the place?
-A `one size fits all' approach would be a disastrous underestimation of the specific needs of humanities research.
-The essence of humanities research is in its diverse, heterogeneous and ephemeral nature.
-But the need of big infrastructures to be designed well before implementation and use, means that every single design decision closes the resulting system to some category of users.
-Anything specific can potentially break the data model and the defined workflow.
-Of course big infrastructure architects also know this; to compensate, they look for possibilities to make more abstract models and malleable workflows.
-But abstract away far enough and the model becomes rather meaningless.
-If the design recommendation of big infrastructure is that we will use XML, what was the use of all that time and effort?
-Big infrastructure design seems to routinely underestimate this problem.
-To give but one simple example: the token seems such a logical, granular, easy to automate conceptual catch-all for text models.
-Can there possibly be a more generalizable model for text than as a series of tokens?
-You may think so, until someone asks how you handle languages that do not use spaces, or how you account for prefixes and suffixes, let alone ambiguous functions.
-Regarding text as a series of tokens denies in any case several other essential humanistic aspects of text: its typography, its materiality, its referentiality to name but a few.
-At second glance, from the humanities research perspective, there is pretty little actually generalizable about tokenization.''
-
-\% Ladefoged, Peter.
-1990.
-Some Reflections on the IPA.
-In UCLA Working Papers in Phonetics, 74.
-``The most interesting ambiguities in the Association's theoretical concepts concern what the symbols are symbolizing.''
-
-There are several ambiguous aspects of the IPA, but the most interesting concerns the theoretical concepts that its symbols are symbolizing \cite[64]{Ladefoged1990a}.
-Although the IPA is \texttt{intended to be a set of symbols for representing all the possible sounds of the world\textquotesingle{}s languages\textquotesingle{}\textquotesingle{} \textbackslash{}citep\{IPA2005\}, Ladefoged points out that the IPA also states that IPA symbols are}\textit{not} symbols for phones they are simply shorthand for what a phonologist would regard as a bundle of features'' \cite{Ladefoged1990a}.
-From the IPA:
-\begin{quote}
-
-	``The representation of these sounds uses a set of phonetic categories which describe how each sound is made.
-These categories define classes of sounds that operate in phonological rules and historical sounds changes.
-The symbols of the IPA are shorthand ways of indicating certain intersections of these categories.
-Thus [p] is a shorthand way of designating the intersection of the categories voiceless, bilabial, and plosive; [m] is the intersection of the categories voiced, bilabial, and nasal; and so on.''
-\end{quote}
-
-\noindent Thus, the ``association is deliberately not explicit about what is meant by a phonological contrast'' \cite[65]{Ladefoged1990a}.

### 2015-05-29 chapters/pitfalls.tex
initial commit paper

--- 2015-05-29 chapters/originaltext.tex
@@
-The Unicode Cookbook for Linguists Managing orthographies in multilingual datasets using orthography profiles
+\chapter{Unicode Pitfalls}
+\label{unicode-pitfalls}
 
-Steven Moran \& Michael Cysouw
-
-\textsc{Abstract}
-
-In this paper, we define the notion of an orthography profile, which is a language-specific (or even resource-specific) description of the units and rules that are needed to adequately model a writing system.
-An orthography profile describes the Unicode code points, characters, graphemes and orthographic rules as used in a writing system.
-Our orthography profiles are used to describe writing systems and possibly even to transliterate them into another (for example IPA-like) transcriptions with the goal to harmonize orthographies across languages and resources on an ad-hoc basis.
-The central aspect of our approach is that we introduce an intermediate step of graphemic tokenization, which is closer to linguistic practice and easier to manage than transducer-based transliteration.
-We leverage orthography profiles to enable comparative analysis of languages with different writing systems and to perform quality control on datasets with highly specific orthographic systems.
-We also describe how our proposals relate to the current Unicode specification, and introduce some of the more delicate aspects of the Unicode Standard for a linguistic audience.
-
-\textsc{Contents}
-
-{[}{[}TOC{]}{]}
-
-\section{1. Introduction}\label{introduction}
-
-Writing systems arise and develop in a complex mixture of cultural, technological and practical pressures.
-They tend to be highly conservative, in that people who have learned to read and write in a specific way (however impractical or tedious) are mostly unwilling to change their habits, e.g.~they tend to resist spelling reforms.
-In all literate societies there exists a strong socio-political mainstream that tries to force unification of writing (for example by strongly enforcing `right' from `wrong' writing in schools).
-However, there is also a large community of users who take as many liberties in their writing as they can get away with.
-
-For example, the writing of tone diacritics in Yoruba is often proclaimed to be the `right' way to write, although many users of Yoruba writing seem to be perfectly fine with leaving them out.
-As pointed out by the proponents of the `official' rules, there are some homographs when leaving out the tone diacritics (Olúmúyìw 2013: 44).
-However, writing systems (and the languages they represent) are normally full of homophones, which is normally not a problem at all for speakers of the language.
-More importantly, writing is not just a purely functional tool, but just as importantly it is a mechanism to signal social affiliation.
-By showing that you `know the rules' of expressing yourself in writing, others will more easily accept you as a worthy participant in their group.
-And that just as well holds for obeying to the `official' rules when writing a job application, as for obeying to the `informal' rules when writing an SMS to classmates in school.
-The case of Yoruba writing is an exemplary case, as even after more than a century of efforts to standardize the writing systems, there is still a wide range of variation in daily use (Olúmúyìw 2013).
-
-The sometimes cumbersome and sometimes illogical structure, and the enormous variability of existing writing systems is a fact of life scholars have to accept and should try to adapt to as good as possible.
-Our plea here is a proposal for a formalization to do exactly that.
-
-When considering the worldwide linguistic diversity, including all lesser-studied and endangered languages, there exist numerous different orthographies using symbols from the same scripts.
-For example, there are hundreds of orthographies using Latin-based alphabetic scripts.
-All of these orthographies use the same symbols, but these symbols differ in meaning and usage throughout the various orthographies.
-To be able to computationally use and compare different orthographies, we need a way to specify all orthographic idiosyncrasies in a computer-readable format (a process called `tailoring' in Unicode parlance).
-We call such specifications \textsc{orthography profiles}.
-Ideally, these specifications have to be integrated into so-called Unicode locale descriptions, though we will argue that in practice this is often not the most useful solution for the kind of problems arising in the daily practice of linguistics.
-Consequently, a central goal of this paper is to flesh out the linguistic challenges for locale descriptions, and work out suggestions to improve their structure for usage in a linguistic context.
-Conversely, we also aim to improve linguists' understanding and appreciation for the accomplishments of the Unicode Consortium in the development of the Unicode Standard.
-
-The necessity to computationally use and compare different orthographies most forcefully arises in the context of language comparison.
-Concretely, in our current research our goal is to develop quantitative methods for language comparison and historical analysis in order to investigate worldwide linguistic variation and to model the historical and areal processes that underlie linguistic diversity (cf.~Steiner et al.
-2010, List 2012a, 2012b, List \& Moran 2013, Moran \& Prokic 2013).
-In this work, it is crucial to be able to flexibly process across numerous resources with different orthographies.
-In many cases even different resources on the `same' language use different orthographic conventions.
-Another orthographic challenge that we encounter regularly in our linguistic practice is electronic resources on a particular language that claim to follow a specific orthographic convention (often a resource-specific convention), but on closer inspection such resources are almost always not consistently encoded.
-Thus a second goal of our orthography profiles is to allow for an easy specification of orthographic conventions, and use such profiles to check consistency and to report errors to be corrected.
-
-A central step in our proposed solution to this problem is the tailored grapheme separation of strings of symbols, a process we call \textsc{grapheme tokenization}.
-Basically, given some strings of symbols (e.g.~morphemes, words, sentences) in a specific source, our first processing step is to specify how these strings have to be separated into graphemes, considering the specific orthographic conventions used in a particular source document.
-Our experience is that such a graphemic tokenization can be performed without extensive in-depth knowledge about the phonetic and phonological details of the language in question.
-For example, the specification that <ou> is a grapheme of English is a much easier task than to specify what exactly the phonetic values of this grapheme are in any specific occurrence in English words.
-Grapheme separation is a task that can be performed relatively reliably and with limited availability of time and resources (compare, for example, the task of creating a complete phonetic or phonological normalization).
-
-Although grapheme tokenization is only one part of the solution, it is an important and highly fruitful processing step.
-Given a grapheme tokenization, various subsequent tasks become easier, like (a) temporarily reducing the orthography in a processing pipeline, e.g.~only distinguishing high versus low vowels; (b) normalizing orthographies across sources (often including temporary reduction of oppositions), e.g.~specifying an (approximate) mapping to the International Phonetic Alphabet; (c) using co-occurrence statistics across different languages (or different sources in the same language) to estimate the probability of grapheme matches, e.g.~with the goal to find regular sound changes between related languages or transliterations between different sources in the same language.
-
-Before we deal with these proposals, in the first part of this paper (Sections 2 through 5) we give an extended introduction to the notion of encoding (Section 2) and writing systems, both from a linguistic perspective and from the perspective of the Unicode Consortium (Section 3).
-We consider the Unicode Standard to be a breakthrough (and ongoing) development that fundamentally changed the way we look at writing systems, and we aim to provide here a slightly more in-depth survey of the many techniques that are available in the standard.
-A good appreciation for the solutions that the Unicode Standard also allows for a thorough understanding of possible pitfalls that one might encounter when using it (Section 4).
-As an example of the current state-of-the-art, we discuss the rather problematic marriage of the International Phonetic Alphabet (IPA) with Unicode (Section 5).
-
-The second part of the paper (Sections 6 and 7) describes our proposals for how to deal with the Unicode Standard in the daily practice of (comparative) linguists.
-First, we discuss the challenges of characterizing a writing system.
-To solve these problems, we propose the notions of orthography profiles, closely related to Unicode locale descriptions (Section 6).
-Finally, we discuss practical issues with actual examples (Section 7).
-We provide reference implementation of our proposals in R and in Python, available as open-source libraries.
-
-The following conventions are followed in this paper.
-All phonemic and phonetic representations are given in the International Phonetic Alphabet (IPA), unless noted otherwise (International Phonetic Association, 2005).
-Standard conventions are used for distinguishing between graphemic <\textgreater{}, phonemic / / and phonetic {[} {]} representations.
-For character descriptions, we follow the notational conventions of the Unicode Standard (Unicode Consortium 2014).
-Character names are represented in small capital letters (e.g.~LATIN SMALL LETTER SCHWA) and code points are expressed as U\emph{+n} where \emph{n} is a four to six digit hexadecimal number, e.g.~U+0256, which can be rendered as the glyph <ə>.
-
-\section{2. Encoding}\label{encoding}
-
-There are many in-depth histories of the origin and development of writing systems (e.g.~Robinson 1997, Powell 2012), a story that we will not repeat here.
-However, the history of turning writing into computer readable code is not so often told, so we decided to offer a short survey of the major developments here.
-This history turns out to be intimately related to the history of telegraphic communication.\footnote{Because of the recent history as summarised in this section, we have used mostly rather ephemeral internet sources, specifically http://www.unicode.org/history/ and various Wikipedia pages for the information presented here (when not references by traditional literature in the bibliography).
-A useful survey of the historical development of the physical hardware of telegraphy and telecommunication is Huurdeman (2003).
-Most books that discuss the development of encoding of telegraphic communication focus of cryptography (e.g.~Singh 1999) and forego the rather interesting story of `open' encoding that is related here.)
-
-Writing systems have existed for roughly 6000 years, allowing people to exchange messages through time and space.
-Additionally, to bridge large geographic distances, \textsc{telegraphic systems} of communication (from Greek \emph{τῆλε γράφειν} ``distant writing'') have a long and widespread history since ancient times.
-The most widespread telegraphic systems worldwide are so-called whistled languages (Meyer 2015), but also drumming languages (Meyer et al.
-2012) and smoke, fire, hydraulic or flag signals are forms of telegraphy.
-Telegraphy was reinvigorated in the end of the eighteenth century through the introduction of semaphoric systems by Claude Chapelle (since then using flags, flashing lights, or various specially designed contraptions) to convey messages over large distances.
-The `innovation' of those systems was that all characters of the written language were replaced one-to-one by visual signals.
-Since then, all telegraphic systems have taken this principle, namely that any language to be transmitted first has to be turned into some orthographic system, which subsequently is encoded for transmission by the sender, and then turned back into orthographic representation at the receiver side.\footnote{Sound and video-based telecommunication of course takes a different approach by ignoring the written version of language and directly encode sound waves or light patterns.) This of course implies that the usefulness of any such telegraphic encoding completely depends on the often rather haphazard structure of orthographic systems.
-
-In the nineteenth century, \textsc{electric telegraphy} lead to a new approach in which written language characters were encoded by signals sent through a copper wire.
-Originally, `bisignal' codes were used, consisting of two different signals.
-For example, Carl Friedrich Gauss in 1833 used positive and negative current (Mania 2008: 282).
-More famous and influential, Samuel Morse in 1836 used long and short pulses.
-In those bisignal systems each character from the written language was encoded with a different number of signals (between one and five), so two different separators are needed: one between signals and one between characters (in Morse-code there is a short pause between signals and a long pause between characters).\footnote{Actually, Morse-code also includes an extra long pause between words.
-Interestingly, it took a long time to consider the written white-space as a bona-fide character that should simply be encoded with its own code point instead of with a different kind of pause.
-This happened only with the revision of the Baudot-code (see below) by Donald Murray in 1901, in which he finally introduced a white-space code.
-This principle has been followed ever since, and for example was included in the second version of the International Telegraph Alphabet (ITA2) in 1930.)
-
-From those encodings, true \textsc{binary codes} developed with a fixed length of signals per character.
-In such systems only a single separator between signals is needed, because the separation between characters can be established by counting until a fixed number of signals.\footnote{Of course, no explicit separator is needed when the timing of the signals is known, which is the principle used in all modern telecommunication systems.
-An important modern consideration is also how to know where to start counting when you did not catch the start of a message, something that is known in Unicode as `self synchronizing'.) In the context of electric telegraphy, such a system was first established by Émile Baudot in 1870, using a fixed combination of five signals for each written character.\footnote{True binary codes have a longer history, going at least back to the Baconian cipher devised by Francis Bacon in 1605.) There are 25 = 32 possible combination when using five binary signals; an encoding today designated as `5-bit'.
-These codes are sufficient for all Latin letters, but of course they do not suffice for all written symbols, including punctuation and digits.
-As a solution, the Baudot code uses a so-called `shift' character, which signifies that from that point onwards (until shifted back) a different encoding is used, allowing for yet another set of 32 codes.
-In effect, this means that the Baudot code, and the \textsc{International Telegraph Alphabet (ITA)} derived from it, had an extra `bit' of information, so the encoding is actually 6-bit (with 26 = 64 different possible characters).
-For decades, this encoding was the standard for all telegraphy and it is still in limited use today.
-
-To also allow for different uppercase and lowercase letters and a large variety of control characters, the American Standards Association decided to propose a new 7-bit encoding in 1963 (with 27 = 128 different possible characters), known as the \textsc{American Standard Code for Information Interchange(ASCII)}, geared towards the encoding of English orthography.
-With the ascent of other orthographies in computer usage, the wish to encode further variation of Latin letters (like German <ß> or various letters with diacritics like <è>) led the Digital Equipment Corporation to introduce an 8-bit \textsc{Multinational Character Set (MCS)} (with 28 = 256 different possible characters), first used with the introduction of the VT220 Terminal in 1983.
-Because 256 characters were clearly not enough for the many different characters needed in the world's writing systems, the ISO/IEC 8859 standard in 1987 extended the MCS to include 16 different 8-bit code pages.
-For example, part 5 was used for Cyrillic characters, part 6 for Arabic, and part 7 for Greek.\footnote{In effect, because 16 = 24, this means that ISO/IEC 8859 was actually a 8+4 = 12-bit encoding, though with very many duplicates by design (e.g.~all ASCII codes were repeated in each 8-bit code page).
-To be precise, ISO/IEC 8859 used the 7-bit ASCII as a basis for each code page, and defined 16 different 7-bit extensions, leading to (1+16) times 27 = 2,176 possible characters.)
-
-This system almost immediately was understood to be insufficient and impractical, so various initiatives to extend and reorganize the encoding started in the 1980s.
-This led, for example, to various proprietary encodings from Microsoft (e.g.~Windows Latin 1) and Apple (e.g.~Mac OS Roman), which one still sometimes encounters today.
-More wide-ranging, various people in the 1980s started to develop true international code sets.
-In the United States, a group of computer scientists formed the Unicode Consortium, proposing a 16-bit encoding in 1991 (with 216 = 65,536 different possible characters).
-At the same time in Europe, the International Organization for Standardization (ISO) was working on ISO 10646 to supplant the ISO 8859 standard.
-Their first draft of the \textsc{Universal Character Set (UCS)} in 1990 was 31-bit (with theoretically 231 = 2,147,483,648 possible characters, but because of some technical restrictions only 679,477,248 were allowed).
-Since 1991, the Unicode Consortium and the International Organization for Standardization have jointly developed the \textsc{Unicode Standard}, or ISO/IEC 10646, leading to the current system including the original 16-bit Unicode proposal as the \textsc{Basic Multilingual Plane}, and 16 additional planes of 16-bit for further extensions (with in total 1+16 times 216 = 1,114,112 possible characters).
-The most recent version of the Unicode Standard (7.0) was published in June 2014 and it defines 112,956 different characters (Unicode Consortium 2014).
-
-\section{3. Terminology}\label{terminology}
-
-\subsection{The linguistic tradition}\label{the-linguistic-tradition}
-
-In this section we provide a very brief overview of the linguistic terminology concerning writing systems before turning to the slightly different computational terminology in the next section on the Unicode Standard.
-Linguistically speaking, a \textsc{writing system} is a symbolic system that uses visible or tactile signs to represent language in a systematic way.
-The term `writing system' has two mutually exclusive meanings.
-First, it may refer to the way a particular language is written.
-In this sense the term refers to the writing system of a particular language, as, for example, in \emph{``the Serbian writing system uses two scripts: Latin and Cyrillic''}.
-Second, the term `writing system' may refer to a type of symbolic system as used among the world's languages to represent languages, as, for example, in \emph{``alphabetic writing system''}.
-In this latter sense the term refers to how scripts have been classified according to the way that they encode language, as in, for example, \emph{``the Latin and Cyrillic scripts are both alphabetic writing systems''}.
-To avoid confusion, this second notion of `writing system' would have been more aptly called \textsc{script system}.
-The term \textsc{script} refers to a collection of distinct symbols as employed by one or more orthographies.
-For example, both Serbian and Russian are written with subsets of the Cyrillic script.
-A single language, like Serbian or Japanese, can also be written using orthographies based on different scripts.
-Over the years linguists have typologized script systems in a variety of ways, with the tripartite classification of logographic, syllabic, and alphabetic remaining the most popular, even though there are at least half a dozen different types of script systems that can be distinguished (Daniels, 1990, 1996).
-
-Only focussing here on the first sense, we distinguish two different kinds of writing systems as used to write a particular language, namely transcriptions and orthographies.
-First, \textsc{transcription} is a scientific procedure (and also the result of that procedure) for graphically representing the sounds of human speech at the phonetic level.
-It incorporates a set of unambiguous symbols to represent speech sounds, including conventions that specify how these symbols should be combined.
-A \textsc{transcription system} is a specific system of symbols and rules used for transcription of the sounds of a spoken language variety.
-In principle, a transcription system should be language-independent, in that it should be applicable to all spoken human languages.
-The International Phonetic Alphabet (IPA) is a commonly used transcription system that provides a medium for transcribing languages at the phonetic level.
-However, there is a long history of alternative kinds of transcription systems (see Kemp 1994) and today various alternatives are in widespread use (e.g.~X-SAMPA and Cyrillic-based phonetic transcription systems).\footnote{Most users of IPA do not follow the standard explicitly and many dialects based on it have emerged, e.g.~the Africanist and Americanist transcription systems.
-) Note that IPA symbols are also often used to represent language on a phonemic level.
-However, it is important to realize that in this usage the IPA symbols are not a transcription system, but rather an orthography (though with strong links to the pronunciation).
-Further, a transcription system does not need to be highly detailed, but it can also be a system of broad sound classes.
-Although such an approximative transcription is not normally used in linguistics, it is widespread in technological approaches (e.g.~Soundex, Beider \& Morse 2010)\footnote{http://stevemorse.org/phonetics/bmpm.htm , http://stevemorse.org/phonetics/bmpm2.htm) and it is sometimes fruitfully used in automatic approaches to historical linguistics (Dolgopolsky 1986, ASJP\footnote{http://asjp.clld.org/)).
-
-Second, an \textsc{orthography} specifies the symbols, punctuations, and the rules in which a specific language is written in a standardized way.
-Orthographies are often based on a phonemic analysis, but they almost always include idiosyncrasies because of historical developments (like sound changes or loans) and because of the widely-followed principle of lexical integrity (i.e.~the attempt to write the same lexical root in a consistent way, also when synchronic phonemic rules change the pronunciation, for example with final devoicing in many Germanic languages).
-All orthographies are language-specific (and often even resource-specific), although individual symbols or rules might be shared between languages.
-A \textsc{practical orthography} is a strongly phoneme-based writing system designed for practical use by speakers.
-The mapping relation between phonemes and graphemes in practical orthographies is purposely shallow, i.e.~there is mostly a systematic and faithful mapping from a phoneme to a grapheme.
-Practical orthographies are intended to jumpstart written materials development by correlating a writing system with the sound units of a language (cf.~Meinhof \& Jones 1928).
-Symbols from the IPA are often used by linguists in the development of practical orthographies for languages without writing systems, though this usage of IPA symbols should not be confused with transcription (as defined above).
-Further, a \textsc{transliteration} is a mapping between two different orthographies.
-It is the process of ``recording the graphic symbols of one writing system in terms of the corresponding graphic symbols of a second writing system'' (Kemp 2006:396).
-In straightforward cases, such a transliteration is simply a matter of replacing one symbol with another.
-However, there are widespread complications, like one-to-many or many-to-many mappings, which are not always easy, or even possible, to solve without listing all cases individually (cf.~Moran 2012, chp 2).
-
-Breaking it down further, a script consists of \textsc{graphemes}, and graphemes consist of \textsc{characters}.
-In the linguistic terminology of writing systems, a \textsc{character} is a general term for any self-contained element in a writing system.\footnote{There is a second interpretation of the term `character', i.e.~a conventional term for a unit in the Chinese writing system (Daniels 1996).
-This interpretation will not be further explored in this paper.) Although in literate societies most people have a strong intuition about what the characters are in their particular orthography or orthographies, it turns out that the separation of an orthography into separate characters is far from trivial.
-The widespread intuitive notion of a character is strongly biased towards educational traditions, like the alphabet taught at schools, and technological possibilities, like the available type pieces in a printer's job case, the keys on a typewriter, or the symbols displayed in Microsoft Word's symbol browser.
-In practice, characters often consist of multiple building blocks, each of which could be considered a character in its own right.
-For example, although a Chinese character may be considered to be a single basic unanalyzable unit, at a more fine-grained level of analysis the internal structure of Chinese characters is often comprised of smaller semantic and phonetic units that should be considered characters (Sproat 2000).
-In alphabetic scripts, this problem is most forcefully exemplified by diacritics.
-A \textsc{diacritic} is a mark, or series of marks, that may be above, below, or through other characters (Gaultney 2002).
-Diacritics are sometimes used to distinguish homophonous words, but they are more often used to indicate a modified pronunciation (Daniels \& Bright 1996, xli).
-The central question is whether, for example, , , and should be considered four characters, or different combinations of three characters.
-In general, multiple characters together can form another character, and it is not always possible to decide on principled grounds what should be the basic building blocks of an orthography.
-
-For that reason, it is better to analyse an orthography as a collection of graphemes.
-A \textsc{grapheme} is the basic, minimally distinctive symbol of a particular writing system, alike to the phoneme is an abstract representation of a distinct sound in a specific language.
-The term `grapheme' was modeled after the term `phoneme' and represents a contrastive graphical unit in a writing system (see Kohrt 1986 for a historical overview of the term grapheme).
-Most importantly, a single grapheme regularly consists of multiple characters, like <th>, <ou> and <gh> in English (note that each character in these graphemes is also a separate grapheme in English).
-Such complex graphemes are often used to represent single phonemes.
-So, a combination of characters is used to represent a single phoneme.
-Note that the opposite is also found in writing systems, in cases in which a single character represents a combination of two or more phonemes.
-For example, <x> in English orthography represents a combination of the phonemes /k/ and /s/.
-Further, conditioned or free variants of a grapheme are called \textsc{allographs}.
-For example, the distinctive forms of Greek sigma are conditioned, with <σ> being used word-internally and <ς> being used at the end of a word.
-In sum, there are many-to-many relationships between phonemes and graphemes as they are expressed in the myriad of language- and resource-specific orthographies.
-
-This exposition of the linguistic terminology involved in describing writing systems has been purposely brief.
-We have highlighted some of the linguistic notions that are pertinent, yet sometimes confused with, the technological definitions developed for the computational processing of the world's writing systems, which we describe in the next section.
-
-\subsection{The Unicode approach}\label{the-unicode-approach}
-
-The conceptualisation and terminology of writing systems was rejuvenated through the development of the Unicode Standard, with major input from Mark Davis, co-founder and long-term president of the Unicode Consortium.
-For many years, `exotic' writing systems and phonetic transcription systems on personal computers were constrained by the American Standard Code for Information Interchange* *(ASCII) character encoding scheme (based on the Latin script), which only allowed for a strongly limited number of different symbols to be encoded.
-This implied that users could either use and adopt the (extended) Latin alphabet or they could assign new symbols to the small number of code points in the ASCII encoding scheme to be rendered by a specifically designed font (Bird \& Simons 2003).
-In this situation, it was necessary to specify the font together with each document to ensure the rightful display of its content.
-To alleviate this problem of assigning different symbols to the same code points, in the late 80's and early 90's the Unicode Consortium set itself the ambitious goal of developing a single universal character encoding to provide a unique number, a code point, for every character in the world's writing systems.
-Nowadays, the Unicode Standard is the default encoding of the technologies that support the World Wide Web and for all modern operating systems, software and programming languages.
-
-The Unicode Standard represents a massive step forward because it aims to eradicate the distinction between `universal' (ASCII) versus `language/resource-particular' (Font) by adding as much as possible language-specific information into the universal standard.
-However, there are still language/resource-specific specifications necessary for the proper usage of Unicode, as will be discussed below.
-Within the Unicode structure many of these specifications can be captured by so-called `Locales' or `Locale Descriptions', so we are moving to a new distinction of `universal' (Unicode Standard) versus `language-particular' (Locale Description).
-The major gain is a much larger compatibility on the universal level (because Unicode standardizes a much larger portion of writing system diversity), and much better possibilities for automatic processing on the language-particular level (because Locale Descriptions are computer readable specifications).
-
-Each version of the Unicode Standard (as of writing at version 7) consists of a set of specifications and guidelines that include (i) a core specification, (ii) code charts, (iii) standard annexes and (iv) a character database.\footnote{All documents of the Unicode Standard are available at http://www.unicode.org/versions/latest/.
-For a quick survey of the use of terminology inside the Unicode Standard, their glossary is particularly useful, available at http://www.unicode.org/glossary/.
-For a general introduction to the principles of Unicode, Chapter 2 of the core specification, called ``General Structure'', is particularly insightful.
-Different from many other documents of the Unicode Standard, this general introduction is relatively easy to read and illustrated with many interesting examples from various orthography traditions all over the world.
-) The \textsc{core specification} is a book directed toward human readers that describes the formal standard for encoding multilingual text.
-The \textsc{code charts} provide a humanly readable online reference to the character contents of the Unicode Standard in the form of PDF files.
-The \textsc{Unicode Standard Annexes (UAX)} are a set of technical standards that describe the implementation of the Unicode Standard for software development, Web standards, programming languages, etc.
-The \textsc{Unicode Character Database (UCD)} is a set of computer-readable text files that describe the character properties, including a set of rich character and writing system semantics, for each character in the Unicode Standard.
-In this section, we introduce the basic Unicode concepts, but we will leave out many details.
-Please consult the above mentioned full documentation for a more detailed discussion.
-Further note that the Unicode Standard is exactly that, namely a standard.
-It normatively describes notions and rules to be followed.
-In the actual practice of applying this standard in a computational setting, a specific implementation is necessary.
-The most widely used implementation of the Unicode Standard is the \textsc{International Components for Unicode (ICU)}, which offers C/C++ and Java libraries trying to implement to Unicode Standard.\footnote{http://icu-project.org )
-
-The Unicode Standard is a \textsc{character encoding system} which goal it is to support the interchange and processing of written characters and text in a computational setting.
-Underlyingly, the character encoding is represented by a range of numerical values called a \textsc{code space}, which is used to encode a set of characters.
-A \textsc{code point} is a unique non-negative integer within a code space (i.e.~within a certain numerical range).
-In the Unicode Standard character encoding system, an \textsc{abstract character}, for example a ``LATIN SMALL LETTER P'', is mapped to a particular code point, such as the decimal value 112, normally represented in hexadecimal and in Unicode parlance as (U+)0070.\footnote{Hexadecimal (`base-16') 0070 is equivalent to decimal (`base-10') 112, which can be calculated by considering 0·163 + 0·162 + 7·161 + 0·160 = 7·16 = 112.
-Underlyingly, computers will of course treat this code point binary (`base-2') as 11100000, i.e.~112 = 1·27 + 1·26 + 1·25 + 0·24 + 0·23 + 0·22 + 0·21 + 0·20 = 64 + 32 + 16.) That encoded abstract character is rendered on a computer screen (or printed page) as a \textsc{glyph}, e.g.
-<p>, depending on the \textsc{font} and the context in which that character appears.
-
-In Unicode Standard terminology, an (abstract) \textsc{character} is the basic encoding unit.
-The term `character' can be quite confusing due to its alternative definitions across different scientific disciplines and because in general the word `character' means many different things to different people.
-It is therefore often preferable to refer to Unicode characters simply as \textsc{code points}, because there is a one-to-one mapping between Unicode characters and their numeric representation.
-The Unicode Consortium (2012) defines a character as ``the smallest component of written language that has semantic value.
-It refers to the abstract meaning and/or general shape, rather than a specific shape, though in code tables some form of visual representation is essential for the reader's understanding.'' Unicode defines characters as abstractions of orthographic symbols, and it does not define visualizations for these characters (although it does presents examples).
-In contrast, a \textsc{glyph} is a concrete graphical representation of a character as it appears when rendered (or rasterized) and displayed on an electronic device or on printed paper.
-For example, <reinstert Gs here> are different glyphs of the same character, i.e.~they may be rendered differently depending on the typography being used, but they all share the same code point.
-From the perspective of Unicode they are `the same thing'.
-In this approach, a \textsc{font} is then simply a collection of glyphs linked to code points.
-Allography is not specified in Unicode (expect for a few exceptional cases, due to legacy encoding issues), but can be specified in a font as a \textsc{contextual variant} (aka presentation form).
-
-Each code point in the Unicode Standard is associated with a set of \textsc{character properties} as defined by the Unicode character property model.\footnote{The character property model is described in http://www.unicode.org/reports/tr23/, but the actual properties are described in http://www.unicode.org/reports/tr44/.
-A simplified overview of the properties is available at http://userguide.icu-project.org/strings/properties.
-The actual code tables listing all properties for all Unicode code points are available at http://www.unicode.org/Public/UCD/latest/ucd/.
-) Basically, those properties are just a long list of values for each character.
-For example, code point U+0047 has the following properties (among many others):
-\begin{itemize}
-	\item Name: LATIN CAPITAL LETTER G 
-	\item Alphabetic: YES 
-	\item Uppercase: YES 
-	\item Script: LATIN 
-	\item Extender: NO 
-	\item Simple\_Lowercase\_Mapping: 0067 
-\end{itemize}
-
-These properties contain the basic information of the Unicode Standard and they are necessary to define the correct behavior and conformance required for interoperability in and across different software implementations (as defined in the Unicode Standard Annexes).
-The character properties assigned to each code point is based on each character's behavior in the real-world writing traditions.
-For example, the corresponding lowercase character to U+0047 is U+0067 (though note that the relation between uppercase and lowercase is in many situations much more complex than this, and Unicode has further specifications for those cases).
-Another use of properties is to define the script of a character.
-The Unicode Standard defines the term \textsc{script} as, ``A collection of letters and other written signs used to represent textual information in one or more writing systems.
-For example, Russian is written with a subset of the Cyrillic script; Ukrainian is written with a different subset.
-The Japanese writing system uses several scripts.'' In practice, script is simply defined for each character as the explicit ``Script'' property in the Unicode Character Database.
-
-One frequently references property is the \textsc{block} property, which is often used in software applications to impose some structure to the large number of Unicode characters.
-Each character in Unicode belongs to a specific block.
-These blocks are basically an organisational structure to alleviate the administrative burden of keeping Unicode up-to-date.
-Blocks consist of characters that in some way belong together, so that characters are easier to find.
-Some blocks are connected with a specific script, like the Hebrew block or the Gujarati block.
-However, blocks are predefined ranges of code points, and often there will come a point after which the range is completely filled.
-Any extra characters will have to be assigned somewhere else.
-There is, for example, a block `Arabic', which contains most Arabic symbols.
-However, there is also a block `Arabic Supplement', `Arabic Presentation Forms-A' and `Arabic Presentation Form B'.
-The situation with Latin symbols is even more extra.
-In general, the names for block should be taken as a definitional statement.
-For example, many IPA symbols are not located in the aptly-names block ``IPA extensions'', but in other blocks (see Section 5).
-
-There are many cases in which a sequence of characters (i.e.~a sequence of more than one code point) represents what a user perceives as an individual unit in a particular orthographic writing system.
-For this reason the Unicode Standard differentiates between \textsc{abstract character} and \textsc{user-perceived character}.
-Sequences of multiple code points that correspond to a single user-perceived characters are called** grapheme clusters** in Unicode parlance.
-Grapheme clusters come in two flavors: (default) grapheme clusters and tailored grapheme clusters.
-
-The (default) \textsc{grapheme clusters} are locale-independent graphemes, i.e.~they always apply when a particular combination of characters occurs independent of the writing system in which they are used.
-These character combinations are defined in the Unicode Standard as functioning as one \textsc{text element}.\footnote{The Unicode Standard defines text element as: ``A minimum unit of text in relation to a particular text process, in the context of a given writing system.
-In general, the mapping between text elements and code points is many-to-many.'') The simplest example of a grapheme cluster is a base character followed by a letter modifier character.
-For example, the sequence + <̃\textgreater{} (LATIN SMALL LETTER N at U+006E followed by COMBINING TILDE at U+0303) combines visually into , a user-percieved character in writing systems like that of Spanish.
-So, what the user perceives as a single character actually involves a multi-code-point sequence.
-Note that this specific sequence can also be represented with a single \textsc{precomposed} code point, LATIN SMALL LETTER N WITH TILDE at U+00F1, but this is not the case for all multi-code point character sequences.
-The problem that there multiple encodings possible for the same text element has been acknowledged early on in the Unicode Standard (e.g.~for , the sequence U+006E U+0303 should in all situations be treated identically to the precomposed U+00F1), and a system of \textsc{canonical equivalence} is available for such situations.
-Basically, the Unicode Standard offers different kind of normalizations to either decompose all precomposed characters (called \textsc{NFD}, \textsc{Normalization Form Canonical Decomposition}), or compose as much as possible combinations (called \textsc{NFC}, \textsc{Normalization Form Canonical Composition}).
-In current practice of software development, NFC seems to be preferred in most situations and is widely proposed as the preferred canonical form.
-
-More difficult for text processing, because less standardized, is what the Unicode Standard terms \textsc{tailored grapheme clusters}.
-Tailored grapheme clusters are locale-dependent graphemes, i.e.~such combination of characters do not function as text elements in all situations.
-For example, the sequence + for the Slovak digraph or the sequence in the Sisaala practical orthography (pronounced as IPA /tʃ/; Moran 2006).
-These grapheme clusters are ``tailored'' in the sense that they must be specified on a language-by-language or a writing system-by-writing system basis.
-The Unicode Standard provides technological specifications for creating locale specific data in so-called \textsc{Unicode Locale Descriptions}, i.e.~a set of specification that defines a set of language-specific elements (e.g.~tailored grapheme clusters, collation order, capitalization-equivalence), as well as other special information, like how to format numbers, dates, or currencies.
-Locale descriptions are saved in the \textsc{Common Locale Data Repository (CLDR)},\footnote{http://cldr.unicode.org/ ) a repository of language-specific definitions of writing system properties, each of which describes specific usages of characters.
-Each locale can be encoded in a document using the \textsc{Locale Data Markup Language (LDML)}.
-LDML is an XML format and vocabulary for the exchange of structured locale data.
-
-Unicode Locale Descriptions allow users to define language- or resource-specific writing systems or orthographies.
-The Unicode Consortium defines \textsc{writing system} only very loosely, as it is not a central concept in the Unicode Standard.
-A writing system is, ``A set of rules for using one or more scripts to write a particular language.
-Examples include the American English writing system, the British English writing system, the French writing system, and the Japanese writing system.''.
-However, there are various drawbacks of locale descriptions for the daily practice of linguistic work in a multilingual setting (see Section 7).
-
-\section{4. Unicode Pitfalls}\label{unicode-pitfalls}
-
-\subsection{Wrong it is not}\label{wrong-it-is-not}
+\section{Wrong it is not}
+\label{wrong-it-is-not}
 
@@
 
-\subsection{Pitfall: Blocks}\label{pitfall-blocks}
+\section{Pitfall: Blocks}
+\label{pitfall-blocks}
 
@@
 However, there are various free websites that offer search interfaces to the Unicode code tables, like http://unicode-search.net or http://unicode-search.net.
-A further useful approach for searching characters using shape matching is http://shapecatcher.com.
-)
+A further useful approach for searching characters using shape matching is http://shapecatcher.com.}
 
-\subsection{Pitfall: Missing glyphs}\label{pitfall-missing-glyphs}
+\section{Pitfall: Missing glyphs}
+\label{pitfall-missing-glyphs}
 
@@
 There are a few particularly large fonts that might be considered.
-First, there is the \textsc{Everson Mono} font made by Michael Everson, which currently includes 9,756 different glyphs (not including Chinese) updated up to Unicode 7.0.\footnote{Everson Mono is available as shareware at http://www.evertype.com/emono/.
-) Already a bit older is the \textsc{Titus Cyberbit Basic} font made by Jost Gippert and Carl-Martin Bunz, which includes 10,044 different glyphs (not including Chinese), but not including newer characters added after Unicode 4.0.\footnote{Titus Cyberbit Basic is available at http://titus.fkidg1.uni-frankfurt.de/unicode/tituut.asp.
-)
+First, there is the \textsc{Everson Mono} font made by Michael Everson, which currently includes 9,756 different glyphs (not including Chinese) updated up to Unicode 7.0.\footnote{Everson Mono is available as shareware at http://www.evertype.com/emono/.} Already a bit older is the \textsc{Titus Cyberbit Basic} font made by Jost Gippert and Carl-Martin Bunz, which includes 10,044 different glyphs (not including Chinese), but not including newer characters added after Unicode 4.0.\footnote{Titus Cyberbit Basic is available at http://titus.fkidg1.uni-frankfurt.de/unicode/tituut.asp.}
 
 Further, we suggest to always install at least one so-called \textsc{fall-back font,} which provides glyphs that at least show the user some information about the underlyingly encoded character.
-Apple Macintoshes have such a font (which is invisible to the user), which is designed by Michael Everson and made available for other systems through the Unicode Consortium.\footnote{The Apple/Everson fallback font is available at http://www.unicode.org/policies/lastresortfont\_eula.html.
-) Further, the \textsc{GNU Unifont} is a clever way to produce bitmaps approximating the intended glyph of each available character, updated to Unicode 7.0.\footnote{The GNU Unifont is available at http://unifoundry.com/unifont.html.
-) Finally, the Summer Institute of Linguistics provides a \textsc{SIL Unicode BMP Fallback Font}, currently available up to Unicode version 6.1.
-This font does not even attempt to show a real glyph, but only shows the hexadecimal code inside a box for each character, so a user can at least see the Unicode codepoint of the character to be displayed.\footnote{The SIL Unicode BMP Fallback Font is available at http://scripts.sil.org/UnicodeBMPFallbackFont.
-)
+Apple Macintoshes have such a font (which is invisible to the user), which is designed by Michael Everson and made available for other systems through the Unicode Consortium.\footnote{The Apple/Everson fallback font is available at http://www.unicode.org/policies/lastresortfont\_eula.html.} Further, the \textsc{GNU Unifont} is a clever way to produce bitmaps approximating the intended glyph of each available character, updated to Unicode 7.0.\footnote{The GNU Unifont is available at http://unifoundry.com/unifont.html.} Finally, the Summer Institute of Linguistics provides a \textsc{SIL Unicode BMP Fallback Font}, currently available up to Unicode version 6.1.
+This font does not even attempt to show a real glyph, but only shows the hexadecimal code inside a box for each character, so a user can at least see the Unicode codepoint of the character to be displayed.\footnote{The SIL Unicode BMP Fallback Font is available at http://scripts.sil.org/UnicodeBMPFallbackFont.}
 
-\subsection{Pitfall: Faulty rendering}\label{pitfall-faulty-rendering}
+\section{Pitfall: Faulty rendering}
+\label{pitfall-faulty-rendering}
 
@@
 In contrast, Microsoft Word does indicate the font replacement by showing the name of the font replacement, but it simply changes the font, so any text written after the replacement is written in a different font as before.
-Both behaviors leave much to be desired.)
+Both behaviors leave much to be desired.}
 
@@
 All of those systems can lead to unexpected behaviour.\footnote{For more details about OpenType, see http://www.adobe.com/products/type/opentype.html and http://www.microsoft.com/typography/otspec/.
-Additional systems for complex text layout are, among others, Microsoft's DirectWrite https://msdn.microsoft.com/library/dd368038.aspx and the open-source project HarfBuzz http://www.freedesktop.org/wiki/Software/HarfBuzz/.
-) Alternative systems are Apple's \emph{Apple Advanced Typography} (AAT) and the open-source \emph{Graphite} system from the Summer Institute of Linguistics (SIL).\footnote{More information about AAT can be found at https://developer.apple.com/fonts/.
-SIL's Graphite is described in detail at http://scripts.sil.org/cms/scripts/page.php?site\_id=projects\&item\_id=graphite\_home.
-) In both of these systems, a larger burden is placed on the description inside the font.
+Additional systems for complex text layout are, among others, Microsoft's DirectWrite https://msdn.microsoft.com/library/dd368038.aspx and the open-source project HarfBuzz http://www.freedesktop.org/wiki/Software/HarfBuzz/.} Alternative systems are Apple's \emph{Apple Advanced Typography} (AAT) and the open-source \emph{Graphite} system from the Summer Institute of Linguistics (SIL).\footnote{More information about AAT can be found at https://developer.apple.com/fonts/.
+SIL's Graphite is described in detail at http://scripts.sil.org/cms/scripts/page.php?site\_id=projects\&item\_id=graphite\_home.} In both of these systems, a larger burden is placed on the description inside the font.
 
-\subsection{Pitfall: Homoglyphs}\label{pitfall-homoglyphs}
+\section{Pitfall: Homoglyphs}
+\label{pitfall-homoglyphs}
 
@@
 LATIN SMALL LETTER G at U+0067).
-In contrast, the symbols <A А Α Ꭺ ᗅ ᴀ ꓮＡ𐊠 𐌀 𝖠 𝙰> are all different code points, although they look highly similar -- in some cases even sharing exactly the same glyph in some fonts.
+In contrast, the symbols <> are all different code points, although they look highly similar -- in some cases even sharing exactly the same glyph in some fonts.
 All these different A-like characters include the following code points in the Unicode Standard:
 \begin{itemize}
-	\item <A\textgreater{} LATIN CAPITAL LETTER A, at U+0041 
-	\item <А> CYRILLIC CAPITAL LETTER A, at U+0410 
-	\item <Α> GREEK CAPITAL LETTER ALPHA, at U+0391 
-	\item <Ꭺ> CHEROKEE LETTER GO, at U+13AA 
-	\item <ᗅ> CANADIAN SYLLABICS CARRIER GHO, at U+15C5 
-	\item <ᴀ> LATIN LETTER SMALL CAPITAL A, at U+1D00 
-	\item <ꓮ> LISU LETTER A, at U+A4EE 
-	\item <Ａ> FULLWIDTH LATIN CAPITAL LETTER A, at U+FF21 
-	\item <𐊠> CARIAN LETTER A, at U+102A0 
-	\item <𐌀> OLD ITALIC LETTER A, at U+10300 
-	\item <𝖠> MATHEMATICAL SANS-SERIF CAPITAL A, U+1D5A0 
-	\item <𝙰> MATHEMATICAL MONOSPACE CAPITAL A, at U+1D670 
+	\item <A> LATIN CAPITAL LETTER A, at U+0041 
+	\item <> CYRILLIC CAPITAL LETTER A, at U+0410 
+	\item <> GREEK CAPITAL LETTER ALPHA, at U+0391 
+	\item <> CHEROKEE LETTER GO, at U+13AA 
+	\item <> CANADIAN SYLLABICS CARRIER GHO, at U+15C5 
+	\item <> LATIN LETTER SMALL CAPITAL A, at U+1D00 
+	\item <> LISU LETTER A, at U+A4EE 
+	\item <> FULLWIDTH LATIN CAPITAL LETTER A, at U+FF21 
+	\item <> CARIAN LETTER A, at U+102A0 
+	\item <> OLD ITALIC LETTER A, at U+10300 
+	\item <> MATHEMATICAL SANS-SERIF CAPITAL A, U+1D5A0 
+	\item <> MATHEMATICAL MONOSPACE CAPITAL A, at U+1D670 
 \end{itemize}
@@
 The existence of such homoglyphs is partly due to legacy compatibility, but for the most part these characters are simply different characters that happen to look similar.\footnote{A particularly nice interface to look for homoglyphs is http://shapecatcher.com, based on the principle of recognizing shapes (Belongie et al.
-2002).) Yet, they are supposed to behave different from the perspective of a font designer.
+2002).} Yet, they are supposed to behave different from the perspective of a font designer.
 For example, when designing a Cyrillic font, the will have different aesthetics and different traditional expectation compared to a Latin .
@@
 Such homoglyphs are a widespread problem for consistent encoding.
-Although for most users it looks like the words and are identical, in actual fact they do not even share a single code point.\footnote{The first words consists completely of Latin characters, namely U+0076, U+006F, U+0063, U+0065, U+0073, while the second is a mix of Cyrillic and Greek characters, namely U+03BD, U+03BF, U+0041, U+0435, U+0455.) For computers these two words are completely different entities.
+Although for most users it looks like the words and are identical, in actual fact they do not even share a single code point.\footnote{The first words consists completely of Latin characters, namely U+0076, U+006F, U+0063, U+0065, U+0073, while the second is a mix of Cyrillic and Greek characters, namely U+03BD, U+03BF, U+0041, U+0435, U+0455.} For computers these two words are completely different entities.
 Commonly, when users with Cyrillic or Greek keyboards have to type some Latin-based orthography, they mix similar looking Cyrillic or Greek characters into their text, because those characters are so much easier to type.
@@
 
-\subsection{Pitfall: Canonical equivalence}\label{pitfall-canonical-equivalence}
+\section{Pitfall: Canonical equivalence}
+\label{pitfall-canonical-equivalence}
 
@@
 Various players, like the Unicode Consortium, the W3C, or the TEI recommend NFC in most user-directed situations, and some software applications that we tested indeed seem to automatically convert strings into NFC.\footnote{See the summary of various recommendation here: http://www.win.tue.nl/\textasciitilde{}aeb/linux/uc/nfc\_vs\_nfd.html.
-) This means in practice that if a user, for example, enters <a> and <`>, i.e.~LATIN SMALL LETTER A at U+0061 and COMBINING GRAVE ACCENT at U+0300, this might be converted into <à>, i.e.~LATIN SMALL LETTER A WITH GRAVE at U+00E0.\footnote{The behaviour of software applications can be quite erratic in this respect.
+} This means in practice that if a user, for example, enters <a> and <`>, i.e.~LATIN SMALL LETTER A at U+0061 and COMBINING GRAVE ACCENT at U+0300, this might be converted into <à>, i.e.~LATIN SMALL LETTER A WITH GRAVE at U+00E0.\footnote{The behaviour of software applications can be quite erratic in this respect.
 For example, Apple's TextEdit does not do any conversion on text entry.
 However, when you copy and paste some text inside the same document in ``rich text mode'' (i.e.~RTF-format), it will be transformed into NFC on paste.
-Saving a document does not do any conversion to the glyphs on screen, but it will save the characters in NFC.)
+Saving a document does not do any conversion to the glyphs on screen, but it will save the characters in NFC.}
 
-\subsection{Pitfall: Characters are not glyphs}\label{pitfall-characters-are-not-glyphs}
+\section{Pitfall: Characters are not glyphs}
+\label{pitfall-characters-are-not-glyphs}
 
@@
 
-\subsection{Pitfall: File formats}\label{pitfall-file-formats}
+\section{Pitfall: File formats}
+\label{pitfall-file-formats}
 
@@
 Depending on different use cases, the Unicode Standard offers three different approaches, called UTF-32, UTF-16 and UTF-8.\footnote{The letters UTF stand for either ``Unicode Transformation Format'' or ``Universal Coded Character Set Transformation Format'', but the notion of `transformation' is a legacy notion that does not have meaning anymore.
-Nevertheless, the designation ``UTF'' (in capitals) has become an official standard designation, but should probably best be read as simply ``Unicode Format''.) The details of this issue is extensively explained in section 2.5 of the Unicode core specification.
+Nevertheless, the designation ``UTF'' (in capitals) has become an official standard designation, but should probably best be read as simply ``Unicode Format''.} The details of this issue is extensively explained in section 2.5 of the Unicode core specification.
 Basically, \textsc{UTF-32} encodes each character in 32 bits (32 `binary units', i.e.~32 zeros or ones) and is the most disk-space-consuming variant of the three.
@@
 Together with the topic of normalisation (NFC vs.~NFD, see Pitfall 5 above), this leads to the following nicely cryptic file-encoding recommendation, namely to preferably use \textsc{UTF-8 in NFC without BOM with LF line breaks}, whenever possible.
-
-\section{5. IPA meets Unicode}\label{ipa-meets-unicode}
-
-\subsection{Introducing the International Phonetic Alphabet (IPA)}\label{introducing-the-international-phonetic-alphabet-ipa}
-
-The International Phonetic Alphabet (IPA) is a common standard in linguistics to transcribe sounds of spoken language into some Latin-based characters (International Phonetic Association 1999).
-Although IPA is reasonably easily adhered to with pen and paper, it is not trivial to encode IPA characters electronically.
-Early work addressing the need for a universal computing environment for writing systems and their computational complexity is discussed in Simons 1989.
-For a long time, linguists (like all other computer users) were limited to ASCII-encoded 7-bit characters, which only includes Latin characters, numbers and some punctuation and symbols.
-Restricted to these standard character sets that lacked IPA support or other language-specific graphemes that they needed, linguists devised their own solutions (cf.~Bird and Simons 2003).
-For example, some chose to represent unavailable graphemes with substitutes, e.g.~the combination of to represent .
-Tech-savvy linguists redefined selected characters from a character encoding by mapping custom made fonts to those code points.
-However, one linguist's electronic text would not render properly on another linguist's computer without access to the same font.
-Further, if two character encodings defined two character sets differently, then data could not be reliably and correctly displayed.
-This is a common example of the non-interoperability of data and data formats.
-
-To alleviate this problem, during the late 1980s, SAMPA (Speech Assessment Methods Phonetic Alphabet) was created to represent IPA symbols with 7-bit printable ASCII character sequences, e.g.
-for {[}ɸ{]}.
-Two problems with SAMPA are that (i) it is only a partial encoding of the IPA and (ii) it encodes different languages in separate data tables, instead of a universal alphabet, like IPA.
-SAMPA tables are derived from phonemes appearing in several European languages that were developed as part of a European Commission-funded project to address technical problems like electronic mail exchange (what is now simply called email).
-SAMPA is essentially a hack to work around displaying IPA characters, but it provided speech technology and other fields a basis that has been widely adopted and used in code.
-So, SAMPA was a collection of tables to be compared, instead of a large universal table representing all languages.
-An extended version of SAMPA, called X-SAMPA, set out to include every symbol in the IPA chart including all diacritics (Wells nd.).
-X-SAMPA was considered more universally applicable because it consisted of one table that encoded the set of characters that represented phones/segments in IPA across languages.
-SAMPA and X-SAMPA have been widely used for speech technology and as an encoding system in computational linguistics.
-Eventually, ASCII-encoding of the IPA became depreciated through the advent of the Unicode Standard.
-Note however that many popular software packages used for linguistic analyses still require ASCII input, e.g.~RuG/L04\footnote{http://www.let.rug.nl/kleiweg/L04/) and SplitsTree4.\footnote{http://www.splitstree.org/)
-
-Still, there are a few pitfalls to be aware of when using the Unicode Standard to encode IPA.
-As we have said before, from a linguistic perspective it might look like the Unicode Consortium is making incomprehensible decisions, but it is important to realize that the consortium has tried and is continuing to try to be as consistent as possible across a wide range of use cases, and it does place linguistic traditions above other orthographic possibilities.
-In general, we strongly suggest linguists not to complain about any decisions in the Unicode Standard, but to try and understand the rationale of the Unicode Consortium (which in our experience is almost always well-conceived) and devise ways to work with any unexpected behaviour.
-Many of the current problems derive from the fact that the IPA is clearly historically based on the Latin script, but different enough from most other Latin-based writing systems to warrant special attention.
-This ambivalent status of the IPA glyphs (partly Latin, partly special) is unfortunately also attested in the treatment of IPA in the Unicode Standard.
-In retrospect, it might have been better to consider the IPA (and other transcription systems) to be a special or new kind of script within the Unicode Standard, and treat the obvious similarity to Latin glyphs as a historical relic.
-All IPA glyphs would then have their own code points, instead of the current situation in which some IPA glyphs have special code points, while others are treated as being identical to the `regular' Latin characters.
-Yet, the current situation, however unfortunate, is unlikely to change, so as linguists we will have to learn to deal with the specific pitfalls of IPA within the Unicode Standard.
-In this section, we will describe these pitfalls in some detail.
-
-\subsection{Pitfall: There is no single complete Unicode code block for IPA}\label{pitfall-there-is-no-single-complete-unicode-code-block-for-ipa}
-
-The ambivalent nature of IPA glyphs arises because, on the one hand, the IPA uses Latin-based glyphs like , or .
-From this perspective, the IPA seems to be just another orthographic tradition using Latin characters, all of which do not get a special treatment within the Unicode Standard (just like e.g.~the French, German, or Danish orthographic traditions do not have a special status).
-On the other hand, the IPA uses many special symbols (like turned , mirrored and/or extended Latin glyphs ) not found in any other Latin-based writing system.
-For this reason, and already in the first version of the Unicode Standard (Version 1.0 from 1991), a special block with code points, called ``IPA Extensions'' was included.
-As noted in Section 4, Pitfall 1, the Unicode Standard code space is subdivided into character blocks, which generally encode characters from a single script.
-However, as is illustrated by the IPA, characters that form a single writing system may be dispersed across several different character blocks.
-With its diverse collection of symbols from various scripts and diacritics, the IPA is spread across over 13 blocks in the Unicode Standard:\footnote{This number of blocks depends on whether only IPA-sanctioned symbols are counted or if the phonetic symbols commonly found in the literature are also included, see Moran 2012, Appendix C.)
-\begin{itemize}
-	\item Basic Latin (30 characters), e.g. <a, b, c, d, e> 
-	\item Latin-1 Supplement (4 characters): <æ, ç, ð, ø\textgreater{} 
-	\item Latin Extended-A (3 characters): <ħ, ŋ, œ> 
-	\item Latin Extended-B (5 characters): <ǀ, ǁ, ǂ, ǃ, ȵ> 
-	\item IPA Extensions (70 characters), e.g.: <ɐ, ɑ, ɔ> 
-	\item Spacing Modifier Letters (20 characters), e.g.: <ʰ ʷ ˥> 
-	\item Combining Diacritical Marks (33 characters), e.g.: <̝ ̥ ̪> 
-	\item Greek and Coptic (3 characters): <β, θ, χ> 
-	\item Phonetic Extensions (2 characters): <ᴅ, ᴴ> 
-	\item Phonetic Extensions Supplement (3 characters): <ᶑ , ᶾ, ᶣ> 
-	\item Superscripts and Subscripts (1 character): <ⁿ> 
-	\item Arrows (2 characters): \textless{}↑, ↓\textgreater{} 
-	\item Latin Extended-C (1 character): <ⱱ> 
-\end{itemize}
-
-\subsection{Pitfall: There are many IPA homoglyphs in Unicode}\label{pitfall-there-are-many-ipa-homoglyphs-in-unicode}
-
-Another problem is the large number of homoglyphs, i.e.~different characters that have highly similar glyphs (or even completely identical, depending on the font rendering).
-For example, a speaker of Russian should ideally not use the CYRILLIC SMALL LETTER A at code point U+0430 for IPA transcriptions, but instead the LATIN SMALL LETTER A at code point U+0061, although visually they are mostly indistinguishable, and the Cyrillic character is more easily typed on a Cyrillic keyboard.
-Another example we commonly encounter is the use of LATIN SMALL LETTER G at U+0067, instead of the sanctioned Unicode Standard IPA character for the voiced velar stop LATIN SMALL LETTER SCRIPT G at U+0261.
-One begins to question whether this issue is at all apparent to the working linguist, or if they simply use the U+0067 because it is easily keyboarded and thus saves time, whereas the latter must be cumbersomely inserted as a special symbol in most software.\footnote{This issue was recently addressed by the International Phonetic Association, which has taken the stance that both the keyboard LATIN SMALL LETTER G and the LATIN SMALL LETTER SCRIPT G are valid input characters for the voiced velar plosive.
-Unfortunately, this decision further introduces ambiguity for linguists trying to adhere to a strict Unicode Standard IPA encoding.)
-
-Furthermore, on the one hand even linguists are unlikely to distinguish between the LATIN SMALL LETTER SCHWA at code point U+0259 and LATIN SMALL LETTER TURNED E at U+01DD.
-On the other hand, non-linguists are unlikely to distinguish any semantic difference between an open back unrounded vowel , the LATIN SMALL LETTER ALPHA at U+0251, and the open front unrounded vowel , LATIN SMALL LETTER A at U+0061.
-But even among linguists this distinction leads to problems.
-For example, as pointed out by Mielke (2009), there is a problem stemming from the fact that about 75\% of languages are reported to have a five-vowel system (Maddieson 1984).
-Historically, linguistic descriptions tend not to include precise audio recording and measurements of formants, so this may lead one to ask if the many characters that are used in phonological description reflects a ``transcriptional bias''.
-The common use of in transcriptions could be in part due to the ease of typing the letter on an English keyboard (or for older descriptions, the typewriter).
-We found it to be exceedingly rare that a linguist uses for a low back unrounded vowel.\footnote{One example is Vidal 2001a:75, in which the author states: ``The definition of Pilagá /a/ as {[}+back{]} results from its behavior in certain phonological contexts.
-For instance, uvular and pharyngeal consonants only occur around /a/ and /o/.
-Hence, the characterization of /a/ and /o/ as a natural class of (i.e., {[}+back{]} vowels), as opposed to /i/ and /e/.'') They simply use as long as there is no opposition to .\footnote{See Thomason's Language Log post, ``Why I don't love the International Phonetic Alphabet'', at: http://itre.cis.upenn.edu/\textasciitilde{}myl/languagelog/archives/005287.html.) Making things even more problematic, there is an old typographic tradition that the double-story uses a single-story in italics.
-This leads to the unfortunate effect that in most well-designed fonts the italics of and use the same glyph.
-If this distinction has to be kept upright in italics, the only solution we can currently offer is to use `slanted' glyphs (i.e.~artificially italicised glyphs) instead of real italics (i.e.~special italics glyphs designed by a typographer).\footnote{For example, the widely used IPA font Doulos SIL (http://scripts.sil.org/cms/scripts/page.php?item\_id=DoulosSIL) does not have real italics.
-This leads some word-processing software, like Microsoft Word, to produce slanted glyphs instead.
-That particular combination of font and software application will thus lead to the desired effect.
-However, note that when the text is transferred to another font (i.e.~one that includes real italics) and/or to another software application (like Apple Pages, which does not perform slanting), then this visual appearance will be lost.
-In this case we are thus still in the pre-Unicode situation in which the choice of font and rendering software actually matters.
-The ideal solution from a linguistic point of view would be the introduction of a new IPA code point for a different kind of which explicitly specifies that it should still be rendered as a double-story character when italicized.
-After informal discussion with various Unicode players, our impression is that this highly restricted problem is not sufficiently urgent to introduce even more -like characters in Unicode (which already lead to much confusion, see Section 4, Pitfall 4).
-This is a clear situation in which the Unicode Consortium is not just thinking about linguists, but has a more wide-ranging practical view to consider.)
-
-Some other homoglyphs related to encoding IPA in the Unicode Standard are:
-\begin{itemize}
-	\item The uses of the apostrophe has led to long discussions on the Unicode Standard email list.
-An English keyboard inputs \textless{}`\textgreater{} APOSTROPHE at U+0027, although the `preferred' Unicode apostrophe is the \textless{}'\textgreater{} RIGHT SINGLE QUOTATION MARK at U+2019.
-Yet the glottal stop/glottalization/ejective marker is another completely different character, the MODIFIER LETTER APOSTROPHE at U+02BC, but unfortunately looks mostly highly similar to U+2019.
-	\item There is ambiguous encoding of IPA segments within the Unicode Standard.
-An example is the U+02C1 MODIFIER LETTER REVERSED GLOTTAL STOP vs the U+02E4 MODIFIER LETTER SMALL REVERSED GLOTTAL STOP .
-Both are denoted in the Unicode Standard as the `pharyngealized diacritic' and both appear in various resources representing phonetic data online.
-This is thus an example for which the Unicode Standard does not solve the linguistic standardization problem.
-	\item There is at least one case in which the character name assigned by the Unicode Consortium does not match the IPA's description: in the Unicode Standard at U+01C3 is labeled LATIN LETTER RETROFLEX CLICK, but in IPA is an alveolar or postalveolar click (not retroflex).
-This naming is probably best seen as a simple error in the Unicode Standard.
-Note that most linguists simply seem to use or <ou>) while ligatures are single characters (e.g.
-<ʧ> LATIN SMALL LETTER TESH DIGRAPH at U+02A7).
-Ligatures arose in the context of printing easier-to-read texts, and are included in the Unicode Standard for reasons of legacy encoding.
-However, their usage is discouraged by the Unicode core specification.
-Specifically related to IPA, various phonetic combinations of characters (typically affricates) are available as single code-points in the Unicode Standard, but are designated as ``ligatures'' or ``digraphs'' (confusingly both names appear interchangeably).
-Such glyphs might be used by software to produce a pleasing display, but they should not be hard-coded into the text itself.
-In the context of IPA, characters like the following ligatures should thus \emph{not} be used.
-Instead a combination of two characters is preferred:
-	\item <ʣ> LATIN SMALL LETTER DZ DIGRAPH at U+02A3 (use <d> + <z> instead) 
-	\item <ʧ> LATIN SMALL LETTER TESH DIGRAPH at U+02A7 (use <t> + <ʃ> instead) 
-	\item <ʩ> LATIN SMALL LETTER FENG DIGRAPH at U+02A9 (use <f> + <ŋ> instead) 
-\end{itemize}
-
-However, there are a few Unicode characters that are historically ligatures, but which are today considered as simple characters in the Unicode Standard and thus should be used when writing IPA, namely:
-\begin{itemize}
-	\item <ɮ> LATIN SMALL LETTER LEZH at U+026E 
-	\item <œ> LATIN SMALL LIGATURE OE at U+0153 
-	\item <ɶ> LATIN LETTER SMALL CAPITAL OE at U+0276 
-	\item <æ> LATIN SMALL LETTER AE at U+00E6 
-\end{itemize}
-
-\subsection{Pitfall: The IPA notion of diacritics is not the same as the Unicode Standard's notion of diacritics}\label{pitfall-the-ipa-notion-of-diacritics-is-not-the-same-as-the-unicode-standards-notion-of-diacritics}
-
-Another pitfall is diacritics.
-The problem is that the meaning of the term `diacritics' as used by the IPA is not the same as it used in the Unicode Standard.
-Specifically, diacritics in the IPA-sense are either so-called \textsc{Combining Diacritical Marks} or \textsc{Spacing Modifier Letters} in the Unicode Standard.
-Crucially, Combining Diacritical Marks are by definition combined with the character before them (to form so-called default grapheme clusters, see Section 3).
-In contrast, Spacing Modifier Letters are by definition \emph{not} combined into grapheme clusters with the preceding character, but simply treated as separate letters.
-In the context of the IPA, the following IPA-diacritics are actually Spacing Modifier Letters in the Unicode Standard:
-\begin{itemize}
-	\item Length marks, namely <ː> MODIFIER LETTER TRIANGULAR COLON at U+02D0 and <ˑ> MODIFIER LETTER HALF TRIANGULAR COLON at U+02D1. 
-	\item Tone letters, like <˥> MODIFIER LETTER EXTRA-HIGH TONE BAR at U+02E5, and others like this. 
-	\item Superscript letters, like <ʰ> MODIFIER LETTER SMALL H at U+02B0 or <ˤ> MODIFIER LETTER SMALL REVERSED GLOTTAL STOP at U+02E4, and others like this. 
-	\item <˞> MODIFIER LETTER RHOTIC HOOK at U+02DE. 
-\end{itemize}
-
-Although linguists might expect these characters to belong together with the character in front of them, at least for <ʰ> MODIFIER LETTER SMALL H at U+02B0 the Unicode Consortium's decision to treat it as a separate character is also linguistically correct, because according to the IPA it can be used both for aspiration (more precisely postaspiration following the base character) and preaspiration (preceding the base character).
-Note that there is a mechanism in Unicode to force separate characters to be combined (namely by using the ZERO WIDTH JOINER at U+200D), but this seems to be a rather impractical, and probably not enforceable solution to us.
-
-\subsection{Pitfall: Neither the IPA nor the Unicode Standard enforce a unique diacritic ordering}\label{pitfall-neither-the-ipa-nor-the-unicode-standard-enforce-a-unique-diacritic-ordering}
-
-Also related to diacritics is the question of ordering.
-To our knowledge, the International Phonetic Association does not specify a specific ordering for diacritics that combine with phonetic base symbols; this exercise is left to the reasoning of the transcriber.
-However, such marks have to be explicitly ordered if sequences of them are to be interoperable and compatible.
-An example is a labialized aspirated alveolar plosive: <tʷʰ>.
-There is nothing holding linguists back from using <tʰʷ> instead (with exactly the same intended meaning).
-However, from a technical standpoint, these two sequences are different, e.g.~if both sequences are used in a document, searching for <tʷʰ> will not find any instances of <tʰʷ>, and vice versa.
-Likewise, a creaky voiced syllabic dental nasal can be encoded in various orders, e.g.
-<n̪̰̩>, <n̩̰̪> or <n̩̪̰>.
-
-In accordance with the absence of any specification of ordering in the IPA, the Unicode Standard likewise does not propose any standard orderings.
-Both leave it to the user to be consistent.
-However, there is one aspect of ordering for which the Unicode Standard does present a canonical solution, which is uncontroversial from a linguistic perspective.
-Diacritics in the Unicode Standard (i.e.~Combining Diacritical Marks, see above) are classified in Canonical Combining Classes.
-In practice, the diacritics are distinguished by their position relative to the base character.\footnote{See http://unicode.org/reports/tr44/\#Canonical\_Combining\_Class\_Values for a detailed description.) When applying a Unicode normalization (NFC or NFD, see previous section), the diacritics in different positions are put in a specified order.
-This process therefore harmonizes the difference between different encodings, for example, of a midtone creaky voice vowel <ḛ̄>.
-This grapheme cluster can be encoded either as <e> + <̄> + <̰> or as <e> + <̰> + <̄> .
-To prevent this twofold encoding, the Unicode Standard specifies the second ordering as canonical (diacritics below before diacritics above).
-
-When encoding a string according to the Unicode Standard, it is possible to do this either using the NFC (``composition'') or NFD (``decomposition'') normalization.
-Decomposition implies that precomposed characters (like <á> LATIN SMALL LETTER A WITH ACUTE at U+00E1) will be split into its parts.
-This might sound preferable for a linguistic analysis, as the different diacritics are separated from the base characters.
-However, note that most attached elements like strokes (e.g.~in the <ɨ>), retroflex hooks (e.g.~in <ʐ>) or rhotic hooks (e.g.~in <ɝ>) will not be decomposed, but strangely enough a cedilla (like in <ç>) will be decomposed.
-In general, Unicode decomposition does not behave like a feature decomposition as expected from a linguistic perspective.
-It is thus important to consider Unicode decomposition only as a technical procedure, and not assume that it is linguistically sensible.
-
-Facing the problem of specifying a consistent ordering of diacritics while developing a large database of phonological inventories from the world's languages, Moran (2012: 540) defines a set of diacritic ordering conventions.\footnote{The most recent version of these conventions is online: http://phoible.github.io/conventions/) The conventions are influenced by the linguistic literature, though some ad-hoc decisions had to be taken.
-The goal was to explicitly define all character sequences so that the vast variety of phonemes found in descriptions of the world's language were normalized into consistent character sequences, e.g.~if one language description uses and another , when both are intended to be phonetically equivalent, then a decision to normalize to one form was taken.
-For example, when a character sequence contains more than one character in Spacing Modifier Letters, the order that is proposed is the following (where <\textbar{}> indicates ``or'' and <→> indicates ``precedes''):
-\begin{itemize}
-	\itemsep1pt\parskip0pt\parsep0pt 
-	\item \textsc{Spacing Modifier Letters ordering:} ( unreleased <̚> \textbar{} lateral release <ˡ> \textbar{} nasal release <ⁿ>) → ( palatalized <ʲ>) → ( labialized <ʷ>) → ( velarized <ˠ>) → ( pharyngealized <ˤ>) → ( aspirated <ʰ> \textbar{} ejective <ʼ>) → ( long <ː> \textbar{} half-long <ˑ>)
-\end{itemize}
-
-If a character sequence contains more than one diacritic below the base character, then the place feature is applied first (dental, laminal, apical, fronted, backed, lowered, raised), then the laryngeal setting (voiced, voiceless, creaky voice, breathy voice) and finally the syllabic or non-syllabic marker (for vowels, ATR gets put on between the place and laryngeal setting).
-So:
-\begin{itemize}
-	\itemsep1pt\parskip0pt\parsep0pt 
-	\item \textsc{Combining Diacritical Marks (below) ordering:} ( dental <t̪> \textbar{} laminal <t̻> \textbar{} apical <t̺>) → ( fronted <u̟> \textbar{} backed <e̠>) →( lowered <e̞> \textbar{} raised <e̝>) → ( ATR <e̘ e̙>) → ( voiced <s̬> \textbar{} voiceless <n̥> \textbar{} creaky voice <b̰> \textbar{} breathy voice <b̤>) → ( syllabic <n̩> \textbar{} non-syllabic <e̯>)
-\end{itemize}
-
-Character sequences with diacritics above the base character were not problematic in Moran 2012 because they include only the centralized, mid-centralized and nasalized combining characters.
-Moran (2012) marks tones as singletons with Space Modifier Letters, e.g.
-\textless{}˦\textgreater{} for a phonemic high tone, instead of accent diacritics, alleviating potential conflicts.
-Building on the work of Moran (2012), if a character sequence contains more than one diacritic above the base character, we propose:
-\begin{itemize}
-	\itemsep1pt\parskip0pt\parsep0pt 
-	\item \textsc{Combining Diacritical Marks (above) ordering:} (centralized <ë> \textbar{} mid-centralized <e̽>) → (extra short <ĕ>) →( tone accents, e.g.
-<è> ) → ( Spacing Modifier Letters ) → ( tone letters, e.g.
-<e˦>)
-\end{itemize}
-
-\section{6. Orthography profiles}\label{orthography-profiles}
-
-\subsection{Characterizing writing systems}\label{characterizing-writing-systems}
-
-At this point in the course of rapid ongoing developments, we are left with a situation in which the Unicode Standard offers a highly detailed and flexible approach to deal computationally with writing systems, but it has unfortunately not influenced the linguistic practice very much.
-In many practical situations, the Unicode Standard is far too complex for the day-to-day practice in linguistics because it does not offer practical solutions for the down-to-earth problems of many linguists.
-In this section, we propose some simple practical guidelines and methods to improve on this situation.
-
-Our central aims for linguistics, to be approached with a Unicode-based solution, are: (i) to improve the consistency of the encoding of sources, (ii) to transparently document knowledge about the writing system (including transliteration), and (iii) to do all of that in a way that is easy and quick to manage for many different sources with many different writing systems.
-The central concept in our proposal is the \textsc{orthography profile}, a simple tab-separated CSV text file, that characterizes and documents a writing system.
-We also offer basic implementations in Python and R to assist with the production of such files, and to apply orthography profiles for consistency testing, grapheme tokenization and transliteration.
-Not only can orthography profiles be helpful in the daily practice of linguistics, they also succinctly document the orthographic details of a specific source, and, as such, might fruitfully be published alongside sources (e.g.~in digital archives).
-Also, in high-level linguistic analyses in which the graphemic detail is of central importance (e.g.~phonotactic or comparative-historical studies), orthography profiles can transparently document the decisions that have been taken in the interpretation of the orthography in the sources used.
-
-Given these goals, Unicode locale descriptions (see Section 3) might seem like the ideal orthography profiles.
-However, there are various practical obstacles preventing the use of such locale descriptions in the daily linguistic practice, namely: (i) the XML-structure is too verbose to easily and quickly produce or correct manually, (ii) locale descriptions are designed for a wide scope on information (like date formats or names of weekdays) most of which is not applicable for documenting writing systems, and (iii) most crucially, even if someone made the effort to produce a technically correct locale description for a specific source at hand, then it is nigh impossible to deploy the description.
-This is because a locale description has to be submitted to and accepted by the Unicode Common Locale Data Repository.
-The repository is (rightly so) not interested in descriptions that only apply to a limited set of sources (e.g.~only a single specific dictionary).
-
-The major challenge then is developing an infrastructure to identify the elements that are individual graphemes in a source, specifically for the enormous variety of sources using some kind of alphabetic writing system.
-Authors of source documents (e.g.~dictionaries, wordlists, corpora) use a variety of writing systems that range from their own idiosyncratic transcriptions to already well-established practical or longstanding orthographies.
-Although the IPA is one practical choice as a sound-based normalization for writing systems (which can act as an interlingual pivot to attain interoperability across writing systems), graphemes in each writing system must also be identified and standardized if interoperability across different sources is to be achieved.
-In most cases, this amounts to more than simply mapping a grapheme to an IPA segment because graphemes must first be identified in context (e.g.~is the sequence one sound or two sounds or both?) and strings must be tokenized, which may include taking orthographic rules into account (e.g.
-between vowels is /n/ and after a vowel but before a consonant is a nasalized vowel /ṽ/).
-In our experience, data from each source must be individually tokenized into graphemes so that its orthographic structure is identified and its contents can be extracted.
-To extract data for analysis, a source-by-source approach is required before an orthography profile can be created.
-For example, almost each available lexicon on the world's languages is idiosyncratic in its orthography and thus requires lexicon-specific approaches to identify graphemes in the writing system and to map graphemes to phonemes, if desired.
-
-Thus, our key proposal for the characterization of a writing system is to use a grapheme tokenization as an inter-orthographic pivot.
-Basically, any source document is tokenized by graphemes, and only then a mapping to IPA (or any other orthographic conversion) is performed.
-An orthography profile then is a description of the units and rules that are needed to adequately model a graphemic tokenization for a language variety as described in a particular source document.
-An orthography profile summarizes the Unicode (tailored) graphemes and orthographic rules used to write a language (the details of the structure and assumptions of such a profile will be presented in the next section).
-
-As an example of graphemic tokenization, note the three different levels of technological and linguistic elements that interact in the hypothetical lexical form <tsʰṍ̰shi> :
-\begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item code points (10 text elements): t s ʰ o ̃ ̰ ´ s h i 
-	\item grapheme clusters (7 text elements): t s ʰ ṍ̰ s h i 
-	\item tailored grapheme clusters (4 text elements): tsʰ ṍ̰ sh i 
-\end{enumerate}
-
-In (1), the string <tsʰṍ̰shi> has been tokenized into ten Unicode code points (using NFD normalization), delimited here by space.
-Unicode tokenization is required because sequences of code points can differ in their visual and logical orders.
-For example, <õ̰> is ambiguous to whether it is the sequence of + <̰> + <̃> or + <̃> + <̰>.
-Although these two variants are visually homoglyphs, computationally they are different.
-Unicode normalization should be applied to this string to reorder the code points into a canonical order, allowing the data to be treated canonically equivalently for search and comparison.
-In (2), the Unicode code points have been logically normalized and visually organized into grapheme clusters, as specified by the Unicode Standard.
-The combining character sequence <õ̰> is normalized and visually grouped together.
-Note that , the MODIFIER LETTER SMALL H at U+02B0, is not grouped with .
-This is because it belongs to Spacing Modifier Letters category in the Unicode Standard.
-These characters are underspecified for the direction in which they modify a host character.
-For example, can indicate either pre- or post-aspiration (whereas the nasalization or creaky diacritic is defined in the Unicode Standard to apply to a specified base character).
-Finally, to arrive at the graphemic tokenization in (3), tailored grapheme clusters are needed (as for example specified in an orthography profile).
-For example, this orthography profile would specify that the sequence of characters , and form a single grapheme , and that and form a grapheme .
-The orthography profile could also specify orthographic rules, e.g.~when tokenization graphemes, in say English, the in the forms and should be treated as distinct sequences depending on their contexts.
-
-\subsection{Informal description of orthography profiles}\label{informal-description-of-orthography-profiles}
-
-An orthography profile describes the Unicode code points, characters, graphemes and orthographic rules in a writing system.
-An orthography profile is a language-specific (and often even resource-specific) description of the units and rules that are needed to adequately model a writing system.
-An important assumption of our work is that we assume a resource is encoded in Unicode (or has been converted to Unicode).
-Any data source that the Unicode Standard is unable to capture, will also not be captured by an orthography profile.
-
-Informally, an orthography profile specifies the graphemes (or, in Unicode parlance, ``tailored grapheme clusters'') that are expected to occur in any data to be analysed or checked for consistency.
-These graphemes are first identified throughout the whole data (a step which we call `tokenization'), and possibly simply returned as such, possibly including error messages about any parts of the data that are not specified by the orthography profile.
-Once the graphemes are identified, they might also be changed into other graphemes (a step which we call `transliteration').
-When a grapheme has different possible transliterations, then these differences should be separated by contextual specification, possibly down to listing individual exceptional cases.
-
-In practice, we foresee a workflow in which orthography profiles are iteratively refined, while at the same time inconsistencies and errors in the data to be tokenized are corrected.
-In some more complex use-cases there might even be a need for multiple different orthography profiles to be applied in sequence (see Section 8 on various exemplary use-cases).
-The result of any such workflow will normally be a cleaned dataset and an explicit description of the orthographic structure in the form of an orthography profile.
-Subsequently, the orthography profiles can be easily distributed in scholarly channels alongside the cleaned data, for example in supplementary material added to journal papers or in electronic archives.
-
-\subsection{Formal specification of orthography profiles}\label{formal-specification-of-orthography-profiles}
-
-The formal specifications of an orthography profile (or simply `profile' for short) are the following:
-\begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item \textsc{A profile is a} \textsc{Unicode UTF-8 encoded text file} (ideally using NFC, no-BOM, and LF; see Section 4, Pitfall ``File Formats'') that includes the information pertinent to the orthography.
-	\item \textsc{A profile is a} \textsc{tab-separated CSV file with an obligatory header line}.
-A minimal profile can have just a single column, in which case there will of course no tabs, but the first line will still be the header.
-For all columns we assume the name in the header of the CSV file to be crucial.
-The actual ordering of the columns is unimportant.
-	\item \textsc{Lines starting with a hash \textless{}\#\textgreater{} are ignored.} Comments and metadata can be included inside the file, but only as complete lines in the profile, to be marked by lines starting with hash ``\#'' (NUMBER SIGN at U+0023).
-Hashes somewhere else in the file are to be treated literally, i.e.~hashes are only to be ignored when they occur at the start of a line.\footnote{Comments that belong to specific lines will have to be put in a separate column of the CSV file, e.g.~add a column called `comments'.
-Further, if the content of a profile contains a hash at the start of a line, either reorder the columns so the hash does not occur at the start of the line, or add a dummy column in front of the data to not have the data start with a hash.)
-	\item \textsc{Metadata are given in commented lines at the beginning of the text file in a basic ``tag: value'' format.
-}Metadata about the orthographic description given in the orthography profile includes, minimally, (i) author, (ii) date, (iii) title, (iv) a stable language identifier encoded in BCP 47/ISO 639-3, and (v) bibliographic data for resource(s) that illustrate the orthography described in the profile.
-\end{enumerate}
-
-The content of a profile consists of lines, each describing a grapheme of the orthography, using the following columns:
-\begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item \textsc{A minimal profile consists of a single column with header `graphemes'}, listing each of the different graphemes in a separate line. 
-	\item \textsc{Optional columns called `left' and `right' can be used to specify the left and right context of the grapheme, respectively.} The same grapheme can occur multiple times with different contextual specifications, for example to distinguish different pronunciations depending on the context.
-	\item \textsc{The columns `grapheme', `left' and `right' can use regular expression metacharacters.} If regular expressions are used, then all literal usage of the special symbols, like full stops ``.'' or dollar signs ``\$'' (so-called ``metacharacters'') have to be explicitly escaped by adding a backslash before them (i.e.~use ``.'' or ``\$'').
-Note that any specification of context automatically expects regular expressions, so it is probably better to always escape all regular expression metacharacters when used literally in the orthography, i.e.~the following symbols will need to be preceded by a backslash: {[} {]} ( ) \{ \} ~+ * .
-- !
-?
-\^{} \$ .
-	\item \textsc{An optional column called `class' can be used to specify classes of graphemes}, for example to define a class of vowels.
-Users can simply add ad-hoc identifiers in this column to indicate a group of graphemes, which can then be used in the description of the graphemes or the context.
-The identifiers should of course be chosen such that they do not conflate with any symbols used in the orthography themselves.
-Note that such classes only refer to the graphemes, not to the context.
-	\item \textsc{Columns describing transliterations for each graphemes can be added and named at will}.
-Often more than a single possible transliteration will be of interest.
-Any software application using these profiles should use the names of these columns to select a specific transliteration column.
-	\item \textsc{Any other columns can be added freely, but will mostly be ignored by any software application using the profiles}.
-As orthography profiles are also intended to be read and interpreted by humans, it is often highly useful to add extra information on the graphemes in further columns, like for example Unicode codepoints, Unicode names, frequency of occurrence, examples of occurrence, explanation of the contextual restrictions, or comments.
-\end{enumerate}
-
-For the automatic processing of the profiles, the following technical standards will be expected:
-\begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item \textsc{Each line of a profile will be interpreted as a regular expression.
-}Software applications using profiles can also offer to interpret a profile in the literal sense to avoid the necessity for the user to escape regular expressions metacharacters in the profile.
-However, this only is possible when no contexts or classes are described, so this seems only useful in the most basic orthographies.
-	\item \textsc{The `class' column will be used to produce explicit ``or'' chains of regular expressions}, which will then be inserted in the `graphemes', `left' and `right' columns at the position indicated by the class-identifiers.
-For example, a class ``V'' as a context specification might be replaced by a regular expression like ``(a\textbar{}e\textbar{}i\textbar{}o\textbar{}u\textbar{}ei\textbar{}au)''.
-Only the graphemes themselves are included here, not any contexts specified for the elements of the class.
-	\item \textsc{The `left' and `right' contexts will be included into the regular expressions by using lookbehind and lookahead}.
-Basically, the actual regular expression syntax of lookbehind and lookahead is simply hidden to the users by allowing them to only specify the contexts themselves.
-Internally, the contexts in the columns ``left'' and ``right'' are combined with the column ``graphemes'' to form a complex regular expression like ``(?\textless{}=left)graphemes(?=right)''.
-	\item \textsc{The regular expressions will be applied in the order as specified in the profile, from top to bottom.} A software implementation can offer help in figuring out the optimal ordering of the regular expressions, but should then explicitly report on the order used.
-\end{enumerate}
-
-The actual implementation of the profile on some text-string will function as follows:
-\begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item \textsc{All graphemes are matched in the text before they are tokenized or transliterated}.
-In this way, there is no necessity for the user to consider `feeding' and `bleeding' situations, in which the application of a rule either changes the text so another rule suddenly applies (``feeding'') or prevents another rule to apply (`bleeding').
-	\item \textsc{The matching of the graphemes can occur either globally or linearly.
-}From a computer science perspective, the most natural way to match graphemes from a profile in some text is by walking linearly through the text-string from left to right, and at each position go through all graphemes in the profile to see which one matches, then go to the position at the end of the matched grapheme and start over.
-This is basically how a finite state transducer works, which is a well-established technique in computer science.
-However, from a linguistic point of view, our experience is that most linguists find it more natural to think from a global perspective.
-In this approach, the first grapheme in the profile is matched everywhere in the text-string first, before moving to the next grapheme in the profile.
-Theoretically, these approaches will lead to different results, though in practice of actual natural language orthographies they almost always lead to the same result.
-Still, we suggest that any software application using orthography profiles should offer both approaches (i.e.
-``global'' or ``linear'') to the user.
-The approach used should be documented in the metadata as ``tokenization method''.
-	\item \textsc{The matching of the graphemes can occur either in NFC or NFD.
-}By default, both the profile and the text-string to be tokenized should be treated as NFC (see section 4, Pitfall ``Canonical equivalence'' above).
-However, in some use-cases it turns out to be practical to treat both text and profile as NFD.
-This typically happens when very many different combinations of diacritics occur in the data.
-An NFD-profile can then be used to first check which individual diacritics are used, before turning to the more cumbersome inspection of all combinations.
-We suggest that any software application using orthography profiles should offer both approaches (i.e.
-``NFC'' or ``NFD'') to the user.
-The approach used should be documented in the metadata as ``unicode normalization''.
-	\item \textsc{The text-string is always returned in tokenized form} by separating the matched graphemes by a user-specified symbols-string.
-Any transliteration will be returned on top of the tokenization.
-	\item \textsc{Leftover characters (i.e.~characters that are not matched by the profile) should be reported to the user as errors.} Typically, the unmatched character are replaced in the tokenization by a user-specified symbol-string.
-\end{enumerate}
-
-Any software application offering to use orthography profile:
-\begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item \textsc{should offer user-options} to specify:
-	\begin{enumerate}
-		\def\labelenumii{\arabic{enumii}.} 
-		\item \textsc{the name of the column to be used for transliteration} (if any). 
-		\item \textsc{the symbol-string to be inserted between graphemes.} Optionally, a warning might be given if the chosen string includes characters from the orthography itself. 
-		\item \textsc{the symbol-string to be inserted for unmatched strings} in the tokenized and transliterated output. 
-		\item \textsc{the tokenization method}, i.e.~whether the tokenization should proceed ``global'' or ``linear''. 
-		\item \textsc{unicode normalization}, i.e.~whether the text-string and profile should use ``NFC'' or ``NFD''. 
-	\end{enumerate}
-	\item \textsc{might offer user-options }to:
-	\begin{enumerate}
-		\def\labelenumii{\arabic{enumii}.} \setcounter{enumii}{5} 
-		\item \textsc{assist in the ordering of the graphemes.} In our experience, it makes sense to apply larger graphemes before shorter graphemes, and to apply graphemes with context before graphemes without context.
-Further, frequently relevant rules might be applied after rarely relevant rules (though frequency is difficult to establish in practice, as it depends on the available data).
-Also, if this all fails to give any decisive ordering between rules, it seems useful to offer linguists the option to reverse the ordering from any manual specified ordering, because linguists tend to write the more general rule first, before turning to exceptions or special cases.
-		\item \textsc{assist in dealing with upper and lower case characters.} It seems practical to offer some basic case matching, so characters like <a> and <A> are treated equally.
-This will be useful in many concrete cases, although the user should be warned that case matching does not function universally in the same way across orthographies.
-Ideally, users should prepare orthography profiles with all lowercase and uppercase variants explicitly mentioned, so by default no case matching should be performed.
-		\item \textsc{treat the profile literal}, i.e.~to not interpret regular expression metacharacters.
-Matching graphemes literally often leads to strong speed increase, and would allow users to not needing to worry about escaping metacharacters.
-However, in our experience all actually interesting use-cases of orthography profiles include some contexts, which automatically prevents any literal interpretation, so by default the matching should not be literal.
-	\end{enumerate}
-	\item \textsc{should return the following information} to the user:
-	\begin{enumerate}
-		\def\labelenumii{\arabic{enumii}.} \setcounter{enumii}{8} 
-		\item \textsc{the original text-strings to be processed in the used Unicode normalization}, i.e.~in either NFC or NFD as specified by the user. 
-		\item \textsc{the tokenized strings}, with additionally any transliterated strings, if transliteration is requested. 
-		\item \textsc{a survey of all errors encountered}, ideally both in which text-strings any errors occurred and which characters in the text-strings lead to errors. 
-		\item \textsc{a reordered profile}, when any automatic reordering is offered 
-	\end{enumerate}
-\end{enumerate}
-
-\section{7. Use cases}\label{use-cases}
-
-We now present several use cases that have motivated the development of orthography profiles. These include:
-\begin{itemize}
-	\item grapheme tokenization 
-	\item normalization of orthographic systems to attain interoperability across different source documents, 
-	\item cognate identification for detecting the similarity between words from different languages 
-\end{itemize}
-
-An important assumption of our work is that input sources are encoded in Unicode (UTF-8 to be precise).
-Anything that the Unicode Standard is unable to capture, cannot be captured by an orthography profile.
-We also use Unicode normalization form NFD to decompose all incoming text input into normalized and (Unicode) logically ordered strings.
-This process is of fundamental importance when working with text data in the Unicode Standard.
-
-\emph{Grapheme tokenization}
-
-Once text has been normalized, an additional and straightforward process is Unicode grapheme tokenization.
-More recently this regular expression, often available as short cut ``\X'', identifies all sequences of ``base'' characters followed by 0 or more combining diacritics.
-For example, the sequence, <ŋ͡mṍraː> would tokenize as a space-delimited sequence as: <ŋ͡ m ṍ r a ː>.
-For a first pass at identifying grapheme clusters, this is a straightforward tokenization.
-However, the linguist will note that both the tie bar (COMBINING DOUBLE INVERTED BREVE, a Unicode Combining Modifier) and the length marker (MODIFIER LETTER TRIANGULAR COLON.
-a Unicode Letter Modifier) do not appear ``correctly'', i.e.~the tie bar is grouped with the first character in the sequence and the length marker appears singly.
-This is necessary, as defined by the Unicode Standard, because of these character's semantic properties -- with Unicode we cannot know if, for example, the MODIFIER LETTER TRIANGULAR COLON should appear before or after the base character that it modifies, cf.~the IPA aspiration marker, , the MODIFIER LETTER SMALL H, which is also a Unicode Letter Modifier and for linguists a diacritic that be be used for pre- or post-aspiration.
-These ambiguities of position in tokenization are not uncommon in IPA, thus orthographic (or source) normalization and tokenization is needed.
-
-*Cross-orthography **normalization*
-
-Orthographic tokenization requires a specification of the tailored grapheme clusters (linguists: graphemes) to separate orthographic units within an input source.
-In the simplest case, the Unicode graphemic tokenization with regular expression ``\X'' captures the grapheme clusters, i.e.~the input data is graphemically tokenized and there are no additional string sequences required to capture each grapheme, e.g.~no specification.
-This scenario is the most unlikely.
-
-The creation of an orthography typically requires that someone describes knowledge about the input data source's writing system.
-For example, when orthographically tokenizing a dictionary as input, the sequences of grapheme clusters need to be explicitly stated or a regular expression rule can be specified to transform the input.
-We illustrate these methods with examples from Thiesen 1998, a bilingual dictionary of Diccionario Bora-Castellano, and with Huber \& Huber 1992, a comparative vocabulary of sixty-nine indigenous languages of Columbia.
-
-Huber \& Huber 1992
-
-({[}a\textbar{}á\textbar{}e\textbar{}é\textbar{}i\textbar{}í\textbar{}o\textbar{}ó\textbar{}u\textbar{}ú{]})(n)(\s)({[}a\textbar{}á\textbar{}e\textbar{}é\textbar{}i\textbar{}í\textbar{}o\textbar{}ó\textbar{}u\textbar{}ú{]}), \1 \2 \4
-
-A preliminary orthography profile can be generated with code available online.\footnote{https://github.com/bambooforest/tokenizer) The Python script creates a unigram model for individual Unicode characters or Unicode grapheme clusters and their frequency in the source input.
-We find that this gives a broad overview of the input at the character and grapheme levels and it oftens finds typographic errors in the source input by identifying rarely occurring or one-off characters in the input.
-The models also provides the basis for the ``quick'' development of an initial orthography profile by saving the user keyboarding time.
-
-Note that these symbols are still phonetically unspecified, i.e.~symbol may or may not represent, say, a front open unrounded vowel.
-
-The orthography profile is designed so that specified strings of characters are captured for tokenization.
-
-\% how the tokenization works
-
-The graphemic cluster specifications in the orthography profile are read into a trie data structure, which is used to tokenize the incoming input source by identifying sentinels (tokenization boundaries) in the input stream, which we then separate grapheme clusters by spaces.
-
-An illustration is given in Figure N.
-
-Comparative wordlists often start as the compilation of different sources with different writing systems.
-The Dogon comparative wordlist is an example of a compilation of slightly divergent transcription practices, which we will use to illustrate our point and to orthographically normalize an input source by transforming different transcriptions into a single IPA-like transcription.
-
-*Orthographic normalization / cross-source **normalization*
-
-Once we have normalized and tokenized our text elements, we can apply various sorts of source normalization across resources to create interoperable and comparable data.
-
-Sound-based normalization is practical because \ldots{}
-
-We use orthography profiles in our work to describe cross-linguistic difference between writing systems.
-
-\emph{Cognate detection}
-
-The identification of cognates is a non-trivial task, which involves identifying the semantic and phonetic similarity of words.
-
-use grapheme correspondences to approximate cognacy and sound correspondences
-
-to identify cognates and sound changes
-
-Examples
-
-Orthographic similarity != Phonetic similarity != Regular sound change
-
-Cognate identification is further complicated by the fact that languages are written differently, and even more so for the low resource languages that we work with, linguists' phonetic transcriptions are heterogeneous and typically document-specific, i.e.~two linguists working on the same language variety most often develop different practical orthographies by using different symbols and different orthographic rules.
-Thus the essence of language documentation and description is diverse and ephemeral in nature.
-The encoding of linguistic data and linguistic data models is diverse and abstracting away from individual analytical preferences requires a document-by-document approach and a mechanism for standardization of writing to form an interlingual pivot so that different descriptions can be normalized into one encoding system to undertake analysis.
-
-\% Orthography profile implementation in Lingpy / generic tokenizer
-
-\% describe Tokenizer's function
-
-\% describe algorithm for tokenizing IPA (Unicode grapheme cluster tokenization and then alignment of Letter Modifiers and Tie Bars)
-
-\% insert an example of orthographic tokenization from PAD
-
-\section{move Letter Modifiers left}\label{move-letter-modifiers-left}
-
-(\s)(\p{Lm}), \2
-
-which catches:
-
-\section{Diacritics: Unicode letter modifiers (Lm)}\label{diacritics-unicode-letter-modifiers-lm}
-
-\section{({[}ː\textbar{}ʰ\textbar{}ʼ\textbar{}ʿ\textbar{}ʾ{]})}\label{ux2d0ux2b0ux2bcux2bfux2be}
-
-from the set of this:
-
-\section{Diacritics: all}\label{diacritics-all}
-
-\section{({[}ː\textbar{}ʰ\textbar{}ʼ\textbar{}ʿ\textbar{}ʾ\textbar{}̜\textbar{}̠\textbar{}̟\textbar{}͡\textbar{}̪\textbar{}̰\textbar{}̣\textbar{}̥\textbar{}̩\textbar{}‿\textbar{}.\textbar{}→\textbar{}˻\textbar{}+\textbar{}̃\textbar{}̫\textbar{}̤{]})}\label{ux2d0ux2b0ux2bcux2bfux2be.}
-
-and allows us to with one regular expression do this:
-
-PAD examples
-
-A side effect of character tokenization is identification of outliers, e.g.
-
-Some examples in Python
-
-Some examples in R
-
-{[}UNICODE USE-CASES{]}
-
-The most basic overall Unicode character property is the \textsc{General Category}, which categorizes Unicode characters into: \emph{Letters, Punctuation, Symbols, Marks, Numbers, Separators, }and* Other*.
-Each Unicode character property also has a character property value.
-So each General Category is denoted by a property value by a single letter abbreviation, e.g.
-``L'' for Letter.
-Each General Category property also has subcategories that are also properties with property values.
-For example, ``Letters'' is broken down into Uppercase Letter (Lu), Lowercase Letter (Ll), Titlecase Letter (Lt), Modifier Letter (Lm) and Other Letter (Lo).\footnote{The subcategory ``Other Letter'' includes characters such as uncased letters as in the Chinese writing system.) By defining character properties for each code point, algorithmic implementations such as text processing applications or regular expression engines that are conformant to the Unicode Standard can be used to recognize whole (sub)categories of characters or ranges of characters, e.g.~to match proper names in English in title case or more generally to match character types, such as scripts, or symbols and integers in email addresses, phone numbers, etc.
-
-Script information is useful for algorithmic processes, such as collation, searching and for multilingual text processing.
-For example, if a regular expression engine includes functionality to match characters at the level of script, then it provides the ability to identify characters, e.g.
-\match{Greek}, which tests whether letters in a text belong to the Greek script or not.
-This functionality can be used to determine the boundaries between different writing systems that appear in the same document.
-In Section 6 (Use cases of grapheme tokenization) we provide an example of a regular expression match at the level of character property.
-
-Unicode-aware regular expression engines may contain a meta-character, ``\X'', that matches this level of grapheme cluster, i.e.~a base character followed by some number of accents or diacritical marks.
-Grapheme clusters are important to determine word or line boundaries, for collation (`ordering') and for user interface interaction (e.g.~mouse selection, backspacing, cursor movement).
-
-\section{9. Conclusion}\label{conclusion}
-
-In our work we focus on the writing systems used in lesser-described and endangered languages and our orthography profiles are used to describe writing systems and to transpose them into phonetic transcription.
-Of course the International Phonetic Alphabet and other transcription systems are essentially just orthographies that are an approximation of sound.
-Nevertheless, sound-based normalization is practical in automatically identifying cognates and sound changes with quantitative methods.
-We leverage orthography profiles to enable comparative analysis of languages with different writing systems.
-
-An orthography profile lists the graphemes in a particular description of language data, e.g.~a wordlist, dictionary or corpus.
-Building on the knowledge that can be extracted from that description by tokenizing words by the graphemes made explicit in the orthography profile, it is straightforward to undertake other analyses of the data.
-For example, various ngram models of the data can be extracted with a few lines of code.
-A unigram model with counts, frequencies and positive log probability (``plogs''; ``entropy under this model'') (Goldsmith \& Riggle, 2010) provides a fair amount of information about a given data source.
-Essentially, the orthography profile provides the description that allows this information to be calculated based on the mapping of sequences of characters into graphemic units.
-Bigram models are also straightforwardly extracted.
-In the case of bigrams, mutual information can be captured and used in various other statistical analyses, such as quantitative language comparison, inferring phylogenetic trees, etc.
-
-\section{10. Nice text blocks to be included somewhere}\label{nice-text-blocks-to-be-included-somewhere}
-
-Looking at the variety of writing systems used for the world's languages, there are various challenges for the computational processing of resources in which such writing systems are used.
-
-A widespread mechanism to derive new symbols for a writing system is to use a combination of multiple symbols already available.
-That can lead to completely new symbols, like <æ> or <w>, which are still rather transparently derived from <a e> and <v v>, respectively.
-More typically, many orthographies use sequences of symbols as special characters, like <sch> or <gh>.
-To identify such `grapheme clusters' is one of the central problems to be tackled with orthography profiles.
-A further frequently used mechanism is to add extra material to an already available symbol, either to make it better visible (this is probably the reason for the addition of a dot on top of the <i>) or to derive new symbols.
-The widespread use of diacritics is the obvious result of such processes (Gaultney 2002).
-However, also the derivation of new symbols by added extra strokes is widespread (and not only in Far-Eastern character-based writing systems), like for example <ŋ> or <ç>.
-
-The use of punctuation (Daniëls 1994, Humez \& Humez 2008) and spaces (Saenger 1997) to improve readability. {[}e.g.~space after punctuation, but not before{]}
-
-Currently, the Unicode Standard is the standard character encoding for linguistic data (Anderson 2003).
-However, there are still hundreds of different encoding systems that were invented independently to capture orthographic diversity as different nations adopted and developed computer systems.
-These different encoding systems are problematic and in conflict with one another because different standards were formalized differently and for different purposes by different standards committees in different countries.
-No unified encoding scheme contained enough code points to encode all characters, so two different encoding schemes possibly used the same code point for different characters, or used different code points to represent the same character.
-Because computers support multiple character encoding schemes, data risked being corrupted when handled by different applications and encodings.
-The Unicode Standard was devised to alleviate these problems.
-
-For orthographies, identifying graphemes can be even more challenging than identifying phones and phonemes in phonetic transcription because although transcriptions may not adhere strictly to IPA, they tend to have straightforward mappings between sounds and symbols.
-On the other hand, orthographies can introduce orthographic rules, which add an additional challenges in identifying graphemes in words, as mentioned above.
-Thus for resources not in IPA or IPA-like transcriptions, graphemes often must be manually identified, whether they are encoded as singletons or multi-character sequences.
-The identification of graphemes and the formulation of orthographic rules are used to create an orthography profile.
-
-Character-to-glyph mappings are particularly not clear-cut for linguists using the IPA.\footnote{We give more linguistic examples in Section 5.) Just look in the IPA Extensions block.\footnote{Character ``blocks'' are described in Section 3 (The Unicode approach).) There are several characters that are glyphs (variants) of the same grapheme in the Latin block, e.g.
-and \& and .\footnote{LATIN SMALL LETTER ALPHA U+0251 vs LATIN SMALL LETTER A U+0061 , and LATIN SMALL LETTER SCRIPT G U+0261 vs LATIN SMALL LETTER G U+0067 .) Other characters like
-
-, , do not appear at all in the IPA Extensions block, but as separate phonetic symbols; they are already encoded in the Basic Latin, Latin Extended-A and Greek \& Coptic blocks.
-When a linguist transcribes an IPA
-
-on a QWERTY keyboard, if it is from the Basic Latin group, it is valid Unicode IPA (it is not valid IPA if it is from the Cyrillic Block).
-However, keyboard and and because they have been given specific code points in the IPA Extensions block, where they have been given special status as IPA characters, and which most often go unnoticed by linguists.\footnote{In Section 5 we discuss in detail the Unicode IPA alphabet and its characters dispersal across different Unicode Standards scripts and blocks.)
-
-The Unicode Standard divides text elements roughly into two classes of graphemes.
-The cover term \textsc{grapheme cluster}\footnote{Formerly referred to as ``locale-independent graphemes'' to emphasize that the term grapheme is used differently in linguistics.) is used to emphasize the difference between the term \emph{grapheme }as used by linguists and what the users perceive as a character (a basic unit of a writing system for a language), which may not be a single Unicode character.\footnote{http://www.unicode.org/reports/tr29/) The Unicode Consortium makes a further distinction between grapheme clusters and tailored grapheme clusters.
-
-The simplest example of a grapheme cluster is essentially a base character followed by a Letter Modifier character.
-For example, the sequence + \textless{}\textasciitilde{}\textgreater{} (COMBINING TILDE at U+0303) combines visually into , a user-percieved character in writing systems like that of Spanish.\footnote{Although the code point sequence can also be represented with a single code point, LATIN SMALL LETTER N WITH TILDE at U+00F1, it is not the case that all multi-code point character sequences have a single code point equivalent, which can be deduced from normalization.
-Often for commonly-used letters in European languages, a single code point form exists.)
-
-What the user perceives as a single character may actually involved a multi-code point sequence.\footnote{Unicode-aware regular expression engines may contain a meta-character, ``\X'', that matches this level of grapheme cluster, i.e.~a base character followed by some number of accents or diacritical marks.
-As mentioned, grapheme clusters are important for collation algorithms, but they are also used to determine other processes such as word or line boundaries and for user interface interaction (e.g.~mouse selection, backspacing, cursor movement).)
-
-More challenging for text processing is what the Unicode Consortium terms tailored grapheme clusters.
-For example, the sequence + for the Slovak digraph or the sequence .
-These grapheme clusters are ``tailored'' in the sense that they must be specified on a language-by-language basis.
-Algorithms can be adapted to use as input these tailored language-specific orthographies for transliteration, localization, etc.
-In fact, the Unicode Standard provides a method for defining language-specific tailored writing systems called the Unicode Locale Description.
-Unicode Locale Descriptions allow users to define a set of extended grapheme clusters.
-Unicode Locale Descriptions are saved in the Common Locale Data Repository (CLDR),\footnote{http://cldr.unicode.org/) a repository of language-specific definitions of writing system properties, each of which describes specific usages of characters.
-Each can be encoded in a Unicode Locale Data Markup Language (LDML) document.
-LDML is an XML format and vocabulary for the exchange of structured locale data.
-It is a format used not only for CLDR, but also for general interchange of locale data, such as in Microsoft's .NET.
-We return to LDML in Section 5 (Orthography profiles and locale descriptions).
-
-We have noted that the rendering of electronic text requires that characters are encoded and stored in a computer's memory.
-These character sequences are then mapped from character code points to glyphs by software that takes into account a character's context.
-Transparent to most users is the fact that different logical orders of characters may result in the same visual output.
-For example, phonetic characters with certain combinations of diacritics may be homoglyphs (see below), even while the logical order of their character sequences are not the same.\footnote{One example is a vowel that is both nasalized and creaky voice, e.g.
-{[}ḛ̃{]}.
-See discussion below.) Thus some type of standardization, what the Unicode Consortium calls ``normalization'' of the ordering of (some classes of) characters, is required to make sure that all data are logically consistent and therefore easily comparable and equally searchable.
-The aim of standardization is to attain syntactic \textsc{interoperability} of data.\footnote{By interoperability we mean the ability to ubiquitously exchange and merge disparate (textual) data sets, and data encoding formats, to facilitate data sharing and to ``effortlessly'' undertake comparison and analysis.
-For a discussion of syntactic versus semantic interoperability, see Ide and Pustejovsky, 2010.)
-
-The IPA is often used by linguists in undertaking initial phonetic and phonemic analyses of languages.
-However, once the distinctive sounds of a given language have been identified (and a phonological description of the target language has been described), a linguist may continue his or her documentation and description of higher-order structures like words and sentences with a broad phoneme-based alphabet.
-In fact, literary materials development often leapfrogs from a phonemic alphabet to locale-specific practical orthographies, in which numerous factors affect a language's orthography development, e.g.~politics, purpose, speaker communities familiarity with other writing systems, technological limitations.
-
-One simple example is illustrated with two genealogically related languages spoken in Northern Ghana: Western Sisaala\footnote{Western Sisaala lacks /kp/.) and Sisaala Tumulung.\footnote{Sisaala Tumulung (ISO 639-3: sil) is spoken around the city Tumu, which is about 50 kilometers geographically to the east of Western Sisaala (ssl) speaking areas.) Both languages contain the same basic set of contrastive sounds, but their orthographies encode affricates (and vowels) differently graphemically; the influence stem from orthographies of government-sponsored literacy languages, which are taught to children in schools in these areas.
-In Western Sisaala the sound /tʃ/ is written and /dʒ/ as due to familiarity of these graphemes by students learning Dagaare (Moran 2006).
-Spoken further to the east, Sisaala Tumuling encodes the sound /tʃ/ as and /dʒ/ as (Blass 1975).
-
-If encoded in the Unicode Standard, the design of both orthographies has to specify that certain sequences of characters, such as + , are single graphemes (compare English and which is ambiguous).
-Design is important, for example, for collation ordering (sorting), i.e.~the sequence of characters that is treated as a unit for ordering purposes, such as in dictionaries.
-Indeed, the characters , and are three unique graphemes consisting of two Unicode characters from the Latin script, but developers of a software program that recognizes the Western Sisaala alphabet might want to specify the sort order , , \ldots{} .
-Thus there is a need for orthography-specific specifications that define sequences of characters as graphemes, whether technologically, linguistically, or both, for multilingual text processing.
-
-A major challenge that the developers of the Unicode Standard faced is that there exists no universal set of fundamental units of text because the divisions of text into elements, such as graphemes, is dependent on language-specific writing systems and orthographies, as just shown.
-Moreover, the division of electronic text into elements is process dependent, e.g.~sorting and spell checking may or may not need to take capitalization into account, depending on the orthography in question.
-Thus the encoding and processing of text requires well defined \textsc{text elements} and rules for interpreting them; these elements and rules are language-specific when we talk about major languages or well established orthographies, but more often in our research we face input from lesser-studied languages documented in author-specific transcription or practical orthography.\footnote{And this is usually down to the document-specific level, the description of which has been coined ``doculect'' for purposes of discussing differences of opinion in language descriptions (Cysouw and Good, 2013).)
-
-In Section XXX we noted that the Unicode Consortium makes a crucial distinction between characters and glyphs.
-The relations between characters and glyphs is a complex one and it is typically hidden from the user.
-Technologically, we must distinguish between characters and glyphs because:
-\begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item The logical order of a sequence of characters (in computer memory) may not be the same as the visual order of their (rendered) glyphs. 
-	\item There is not always a one-to-one mapping between characters and glyphs. 
-\end{enumerate}
-
-Regarding the first point, a user cannot assume that contiguously rendered text on-screen actually reflects the computer's internal storage order of individual and sequential characters in memory, i.e.~the \textsc{logical order}.
-In other words, in the Unicode Standard the visual order of glyphs may not be the same as the logical order of their digitally encoded characters; contiguous display is not indicative of contiguous text.* *A nice example is Hebrew text.\footnote{http://www.macchiato.com/unicode/globalization-gotchas) Another example is the glyph <ṍ>, which may consist of one, two or three characters in encoded in various physical orders.
-The possible mismatch in sequences becomes clearer when we consider the second point.
-
-When a base character, i.e.~any graphic character not in the General Category of Combining Mark, is followed by one or more characters having the property General Category of Combining Mark, they can be combined visually into one or more glyphs using the Canonical Ordering Algorithm to create Unicode \textsc{grapheme clusters}.
-As we have shown, this input may be ambiguous from a technological perspective if normalization is not taken into account.
-For example, a linguist may insert a single character or a character sequence for to denote a high back unrounded vowel with high tone:
-\begin{itemize}
-	\item LATIN SMALL LETTER U WITH ACUTE (U+00FA) 
-	\item LATIN SMALL LETTER U (U+0075) + COMBINING ACUTE ACCENT (U+0301) 
-\end{itemize}
-
-The normalization algorithms in the Unicode Standard are used to make these character sequences canonically equivalent for purposes such as ubiquitous search and comparison.
-
-However, the standard does not specify many (or most) language-specific graphemes, or in Unicode parlance, \textsc{tailored grapheme clusters}, such as English .\footnote{Some (unfortunate) exceptions included for legacy purposes, and used incorrectly by linguists, include digraphs like LATIN SMALL LETTER TS DIGRAPH at U+02A6 or LATIN SMALL LETTER DEZH DIGRAPH at U+02A4.
-These and other digraphs are encoded in the Unicode Standard's IPA Extensions block.) For technological reasons, these language-specific tailored grapheme clusters must be defined outside of the Unicode Standard, in what are called (language-specific) Unicode Locale Descriptions.\footnote{http://www.unicode.org/reports/tr35/) The encoding of these multiple character sequences in Locale Descriptions is beyond the need (or technical ability) of many linguists; thus we have developed a simpler approach which we used in our own research and describe in detailed in Section XXX.
-Basically, the description of writing systems for lower resource languages, and in particular for document-specific descriptions (e.g.~different descriptions of the same language with different orthographies), requires an orthographic description at the (tailored) grapheme (cluster) level.
-
-An orthography profile is defined as a simple delimited flat text file encoded in the UTF-8, with character sequences in Unicode Normalization Form D.
-Metadata aside, the column delimited contents of an orthography profile are essential a state-table, i.e.~an implementation of a state transition table, but for ease of use encoded in a simple flat text file.
-
-In an orthography profile, all character strings are converted into NFD because each character in a grapheme may have phonetic value, and by using NFD, all graphemes are decomposed into a standardized order.
-For example, a vowel that is both nasalized and creaky looks like in IPA, where is some vowel.
-Although visually the same, a nasalized and creaky vowel can be composed of several different character sequences, as illustrated with in 1-3:
-\begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item <õ̰> LATIN SMALL LETTER O + COMBINING TILDE + COMBINING TILDE BELOW: U+006F + U+0303 + U+0330 
-	\item <õ̰> LATIN SMALL LETTER O + COMBINING TILDE BELOW + COMBINING TILDE: U+006F + U+0330 + U+0303 
-	\item <õ̰> LATIN SMALL LETTER O WITH TILDE + COMBINING TILDE BELOW: U+00F5 + U+0330 
-\end{enumerate}
-
-Applying NFD to these three character sequences results in one standard sequence; in this example the character sequence is that shown 1.
-NFD makes different sequences of input interoperable and it retains all of the phonetic information captured by the separate characters that combine to form a vowel with nasalization and creaky voice phonation.
-This is an important step to consider in linguistic research because each character has phonetic value, i.e.~the combining tilde denotes that the base character is produce by a lowering of the velum, which allows air to escape through the nose when the sound is produced.
-The tilde below also has meaning: it represents a phonation type in which the vocal cords are vibrating, but they are also being held slightly further apart, allowing more air to escape, which produces a murmured sound.
-To enable comparative analyses of languages at a deep phonetic level, retaining a code point for each character of phonetic value is deciduous.
-For example, if the sound systems of two languages are to be compared, one would not want to lose a possibly contrastive phonological feature like nasalization, which may have been collapsed into a single character in its logical representation.
-
-Our approach is similar in nature to the LDML, but we use a simpler state-table method, instead of generating an XML format for the ``exchange of structured locale data''.
-An orthography profile is a specification of the tailored grapheme clusters in a particular source document and any orthographic re-write rules needed after the application of Unicode normalization and orthographic profile tokenization.
-Orthography profiles can of course be used for transliteration between scripts, e.g.~from practical orthography to IPA and vice versa, by specifying mapping (replacements) from the list of tailored grapheme clusters in the source document and a set of replacement characters.
-This will be made clear in the next section in which we provide some use cases.
-
-To summarize, LDML is basically overkill for our intentions of identifying and delimiting grapheme boundaries (or in Unicode parlance ``tailored grapheme clusters'').
-LDML is appropriate, even necessary, for complex localization complexes for software applications -- localization of, say, an operating system or software application across cultures may require the instantiation of different date stamps an other local standards.
-For linguistic purposes of source document data tokenization, we therefore develop a simple-to-implement process of ``orthographic'' tokenization, using standard Unicode Standard-specified processes (e.g.~normalization and grapheme cluster tokenization) with an additional level of linguistic ``tailored grapheme cluster'' specification.
-
-\textsc{REFERENCES}
-
-Nancy Ide and James Pustejovsky.
-2010.
-What does inter- operability mean, anyway?
-toward an operational def- inition of interoperability for language technology.
-In Proceedings of the Second International Conference on Global Interoperability for Language Resources.
-Hong Kong, China.
-
-Huber, Randal Q. and Robert B. Reed (compilers).
-\begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} \setcounter{enumi}{1991} 
-	\itemsep1pt\parskip0pt\parsep0pt 
-	\item Vocabulario comparativo: Palabras selectas de lenguas indígenas de Colombia (Comparative vocabulary: Selected words in indigenous languages of Colombia).
-Bogota: Instituto Lingüísti- co de Verano.
-\end{enumerate}
-
-Thiesen, W.
-(1998).
-Diccionario bora-castellano, castellano-bora (1st ed.).
-Lima Perú: Instituto Linguístico de Verano.Retrieved from: http://www.worldcat.org/title/diccionario-bora-castellano-castellano-bora/oclc/40505215\&referer=brief\_results
-
-\textsc{APPENDIX - ORTHOGRAPHY PROFILE EXAMPLE}
-
-\textsc{APPENDIX - UNICODE IPA}
-
-Provide a clear and full Unicode IPA chart with additional information as an appendix
-\begin{verbatim}
-	- create this chart
-
-	- perhaps we should spell out conformance rules with regard to this chart? i.e. if a source claims conformance to our Unicode IPA chart, it should:
-
-	- identify the sections (or range) of the table that the data conforms to
-
-	- the source is in accordance with not only the Unicode IPA characters, but their intrinsic meanings (e.g.
-an Africanist might use <y> for /j/, but this level of conformance requires that <j> symbolizes /j/, etc.)
-
-	- someone who wants to use an Africanist transcription system can declare that they use the Unicode IPA chart, except for a range of characters (but they provide the mapping to Unicode IPA chart) 
-\end{verbatim}
-
-\textsc{Notes}
-
-\url{http://www.unicode.org/reports/tr29/}
-
-The default boundary determination mechanism specified in this annex provides a straightforward and efficient way to determine some of the most significant boundaries in text: user-perceived characters, words, and sentences.
-Boundaries used in line breaking (also called \emph{word wrapping}) are defined in {[}\href{http://www.unicode.org/reports/tr41/tr41-11.html\#UAX14}{UAX14}{]}.
-
-\textsc{Nice quotes}
-
-vanZundert: ``How does one design and plan a general unified infrastructure for a target that is all over the place?
-A `one size fits all' approach would be a disastrous underestimation of the specific needs of humanities research.
-The essence of humanities research is in its diverse, heterogeneous and ephemeral nature.
-But the need of big infrastructures to be designed well before implementation and use, means that every single design decision closes the resulting system to some category of users.
-Anything specific can potentially break the data model and the defined workflow.
-Of course big infrastructure architects also know this; to compensate, they look for possibilities to make more abstract models and malleable workflows.
-But abstract away far enough and the model becomes rather meaningless.
-If the design recommendation of big infrastructure is that we will use XML, what was the use of all that time and effort?
-Big infrastructure design seems to routinely underestimate this problem.
-To give but one simple example: the token seems such a logical, granular, easy to automate conceptual catch-all for text models.
-Can there possibly be a more generalizable model for text than as a series of tokens?
-You may think so, until someone asks how you handle languages that do not use spaces, or how you account for prefixes and suffixes, let alone ambiguous functions.
-Regarding text as a series of tokens denies in any case several other essential humanistic aspects of text: its typography, its materiality, its referentiality to name but a few.
-At second glance, from the humanities research perspective, there is pretty little actually generalizable about tokenization.''
-
-\% Ladefoged, Peter.
-1990.
-Some Reflections on the IPA.
-In UCLA Working Papers in Phonetics, 74.
-``The most interesting ambiguities in the Association's theoretical concepts concern what the symbols are symbolizing.''
-
-There are several ambiguous aspects of the IPA, but the most interesting concerns the theoretical concepts that its symbols are symbolizing \cite[64]{Ladefoged1990a}.
-Although the IPA is \texttt{intended to be a set of symbols for representing all the possible sounds of the world\textquotesingle{}s languages\textquotesingle{}\textquotesingle{} \textbackslash{}citep\{IPA2005\}, Ladefoged points out that the IPA also states that IPA symbols are}\textit{not} symbols for phones they are simply shorthand for what a phonologist would regard as a bundle of features'' \cite{Ladefoged1990a}.
-From the IPA:
-\begin{quote}
-
-	``The representation of these sounds uses a set of phonetic categories which describe how each sound is made.
-These categories define classes of sounds that operate in phonological rules and historical sounds changes.
-The symbols of the IPA are shorthand ways of indicating certain intersections of these categories.
-Thus [p] is a shorthand way of designating the intersection of the categories voiceless, bilabial, and plosive; [m] is the intersection of the categories voiced, bilabial, and nasal; and so on.''
-\end{quote}
-
-\noindent Thus, the ``association is deliberately not explicit about what is meant by a phonological contrast'' \cite[65]{Ladefoged1990a}.

### 2015-05-29 chapters/terminology.tex
initial commit paper

--- 2015-05-29 chapters/originaltext.tex
@@
-The Unicode Cookbook for Linguists Managing orthographies in multilingual datasets using orthography profiles
-
-Steven Moran \& Michael Cysouw
-
-\textsc{Abstract}
-
-In this paper, we define the notion of an orthography profile, which is a language-specific (or even resource-specific) description of the units and rules that are needed to adequately model a writing system.
-An orthography profile describes the Unicode code points, characters, graphemes and orthographic rules as used in a writing system.
-Our orthography profiles are used to describe writing systems and possibly even to transliterate them into another (for example IPA-like) transcriptions with the goal to harmonize orthographies across languages and resources on an ad-hoc basis.
-The central aspect of our approach is that we introduce an intermediate step of graphemic tokenization, which is closer to linguistic practice and easier to manage than transducer-based transliteration.
-We leverage orthography profiles to enable comparative analysis of languages with different writing systems and to perform quality control on datasets with highly specific orthographic systems.
-We also describe how our proposals relate to the current Unicode specification, and introduce some of the more delicate aspects of the Unicode Standard for a linguistic audience.
-
-\textsc{Contents}
-
-{[}{[}TOC{]}{]}
-
-\section{1. Introduction}\label{introduction}
-
-Writing systems arise and develop in a complex mixture of cultural, technological and practical pressures.
-They tend to be highly conservative, in that people who have learned to read and write in a specific way (however impractical or tedious) are mostly unwilling to change their habits, e.g.~they tend to resist spelling reforms.
-In all literate societies there exists a strong socio-political mainstream that tries to force unification of writing (for example by strongly enforcing `right' from `wrong' writing in schools).
-However, there is also a large community of users who take as many liberties in their writing as they can get away with.
-
-For example, the writing of tone diacritics in Yoruba is often proclaimed to be the `right' way to write, although many users of Yoruba writing seem to be perfectly fine with leaving them out.
-As pointed out by the proponents of the `official' rules, there are some homographs when leaving out the tone diacritics (Olúmúyìw 2013: 44).
-However, writing systems (and the languages they represent) are normally full of homophones, which is normally not a problem at all for speakers of the language.
-More importantly, writing is not just a purely functional tool, but just as importantly it is a mechanism to signal social affiliation.
-By showing that you `know the rules' of expressing yourself in writing, others will more easily accept you as a worthy participant in their group.
-And that just as well holds for obeying to the `official' rules when writing a job application, as for obeying to the `informal' rules when writing an SMS to classmates in school.
-The case of Yoruba writing is an exemplary case, as even after more than a century of efforts to standardize the writing systems, there is still a wide range of variation in daily use (Olúmúyìw 2013).
-
-The sometimes cumbersome and sometimes illogical structure, and the enormous variability of existing writing systems is a fact of life scholars have to accept and should try to adapt to as good as possible.
-Our plea here is a proposal for a formalization to do exactly that.
-
-When considering the worldwide linguistic diversity, including all lesser-studied and endangered languages, there exist numerous different orthographies using symbols from the same scripts.
-For example, there are hundreds of orthographies using Latin-based alphabetic scripts.
-All of these orthographies use the same symbols, but these symbols differ in meaning and usage throughout the various orthographies.
-To be able to computationally use and compare different orthographies, we need a way to specify all orthographic idiosyncrasies in a computer-readable format (a process called `tailoring' in Unicode parlance).
-We call such specifications \textsc{orthography profiles}.
-Ideally, these specifications have to be integrated into so-called Unicode locale descriptions, though we will argue that in practice this is often not the most useful solution for the kind of problems arising in the daily practice of linguistics.
-Consequently, a central goal of this paper is to flesh out the linguistic challenges for locale descriptions, and work out suggestions to improve their structure for usage in a linguistic context.
-Conversely, we also aim to improve linguists' understanding and appreciation for the accomplishments of the Unicode Consortium in the development of the Unicode Standard.
-
-The necessity to computationally use and compare different orthographies most forcefully arises in the context of language comparison.
-Concretely, in our current research our goal is to develop quantitative methods for language comparison and historical analysis in order to investigate worldwide linguistic variation and to model the historical and areal processes that underlie linguistic diversity (cf.~Steiner et al.
-2010, List 2012a, 2012b, List \& Moran 2013, Moran \& Prokic 2013).
-In this work, it is crucial to be able to flexibly process across numerous resources with different orthographies.
-In many cases even different resources on the `same' language use different orthographic conventions.
-Another orthographic challenge that we encounter regularly in our linguistic practice is electronic resources on a particular language that claim to follow a specific orthographic convention (often a resource-specific convention), but on closer inspection such resources are almost always not consistently encoded.
-Thus a second goal of our orthography profiles is to allow for an easy specification of orthographic conventions, and use such profiles to check consistency and to report errors to be corrected.
-
-A central step in our proposed solution to this problem is the tailored grapheme separation of strings of symbols, a process we call \textsc{grapheme tokenization}.
-Basically, given some strings of symbols (e.g.~morphemes, words, sentences) in a specific source, our first processing step is to specify how these strings have to be separated into graphemes, considering the specific orthographic conventions used in a particular source document.
-Our experience is that such a graphemic tokenization can be performed without extensive in-depth knowledge about the phonetic and phonological details of the language in question.
-For example, the specification that <ou> is a grapheme of English is a much easier task than to specify what exactly the phonetic values of this grapheme are in any specific occurrence in English words.
-Grapheme separation is a task that can be performed relatively reliably and with limited availability of time and resources (compare, for example, the task of creating a complete phonetic or phonological normalization).
-
-Although grapheme tokenization is only one part of the solution, it is an important and highly fruitful processing step.
-Given a grapheme tokenization, various subsequent tasks become easier, like (a) temporarily reducing the orthography in a processing pipeline, e.g.~only distinguishing high versus low vowels; (b) normalizing orthographies across sources (often including temporary reduction of oppositions), e.g.~specifying an (approximate) mapping to the International Phonetic Alphabet; (c) using co-occurrence statistics across different languages (or different sources in the same language) to estimate the probability of grapheme matches, e.g.~with the goal to find regular sound changes between related languages or transliterations between different sources in the same language.
-
-Before we deal with these proposals, in the first part of this paper (Sections 2 through 5) we give an extended introduction to the notion of encoding (Section 2) and writing systems, both from a linguistic perspective and from the perspective of the Unicode Consortium (Section 3).
-We consider the Unicode Standard to be a breakthrough (and ongoing) development that fundamentally changed the way we look at writing systems, and we aim to provide here a slightly more in-depth survey of the many techniques that are available in the standard.
-A good appreciation for the solutions that the Unicode Standard also allows for a thorough understanding of possible pitfalls that one might encounter when using it (Section 4).
-As an example of the current state-of-the-art, we discuss the rather problematic marriage of the International Phonetic Alphabet (IPA) with Unicode (Section 5).
-
-The second part of the paper (Sections 6 and 7) describes our proposals for how to deal with the Unicode Standard in the daily practice of (comparative) linguists.
-First, we discuss the challenges of characterizing a writing system.
-To solve these problems, we propose the notions of orthography profiles, closely related to Unicode locale descriptions (Section 6).
-Finally, we discuss practical issues with actual examples (Section 7).
-We provide reference implementation of our proposals in R and in Python, available as open-source libraries.
-
-The following conventions are followed in this paper.
-All phonemic and phonetic representations are given in the International Phonetic Alphabet (IPA), unless noted otherwise (International Phonetic Association, 2005).
-Standard conventions are used for distinguishing between graphemic <\textgreater{}, phonemic / / and phonetic {[} {]} representations.
-For character descriptions, we follow the notational conventions of the Unicode Standard (Unicode Consortium 2014).
-Character names are represented in small capital letters (e.g.~LATIN SMALL LETTER SCHWA) and code points are expressed as U\emph{+n} where \emph{n} is a four to six digit hexadecimal number, e.g.~U+0256, which can be rendered as the glyph <ə>.
-
-\section{2. Encoding}\label{encoding}
-
-There are many in-depth histories of the origin and development of writing systems (e.g.~Robinson 1997, Powell 2012), a story that we will not repeat here.
-However, the history of turning writing into computer readable code is not so often told, so we decided to offer a short survey of the major developments here.
-This history turns out to be intimately related to the history of telegraphic communication.\footnote{Because of the recent history as summarised in this section, we have used mostly rather ephemeral internet sources, specifically http://www.unicode.org/history/ and various Wikipedia pages for the information presented here (when not references by traditional literature in the bibliography).
-A useful survey of the historical development of the physical hardware of telegraphy and telecommunication is Huurdeman (2003).
-Most books that discuss the development of encoding of telegraphic communication focus of cryptography (e.g.~Singh 1999) and forego the rather interesting story of `open' encoding that is related here.)
-
-Writing systems have existed for roughly 6000 years, allowing people to exchange messages through time and space.
-Additionally, to bridge large geographic distances, \textsc{telegraphic systems} of communication (from Greek \emph{τῆλε γράφειν} ``distant writing'') have a long and widespread history since ancient times.
-The most widespread telegraphic systems worldwide are so-called whistled languages (Meyer 2015), but also drumming languages (Meyer et al.
-2012) and smoke, fire, hydraulic or flag signals are forms of telegraphy.
-Telegraphy was reinvigorated in the end of the eighteenth century through the introduction of semaphoric systems by Claude Chapelle (since then using flags, flashing lights, or various specially designed contraptions) to convey messages over large distances.
-The `innovation' of those systems was that all characters of the written language were replaced one-to-one by visual signals.
-Since then, all telegraphic systems have taken this principle, namely that any language to be transmitted first has to be turned into some orthographic system, which subsequently is encoded for transmission by the sender, and then turned back into orthographic representation at the receiver side.\footnote{Sound and video-based telecommunication of course takes a different approach by ignoring the written version of language and directly encode sound waves or light patterns.) This of course implies that the usefulness of any such telegraphic encoding completely depends on the often rather haphazard structure of orthographic systems.
-
-In the nineteenth century, \textsc{electric telegraphy} lead to a new approach in which written language characters were encoded by signals sent through a copper wire.
-Originally, `bisignal' codes were used, consisting of two different signals.
-For example, Carl Friedrich Gauss in 1833 used positive and negative current (Mania 2008: 282).
-More famous and influential, Samuel Morse in 1836 used long and short pulses.
-In those bisignal systems each character from the written language was encoded with a different number of signals (between one and five), so two different separators are needed: one between signals and one between characters (in Morse-code there is a short pause between signals and a long pause between characters).\footnote{Actually, Morse-code also includes an extra long pause between words.
-Interestingly, it took a long time to consider the written white-space as a bona-fide character that should simply be encoded with its own code point instead of with a different kind of pause.
-This happened only with the revision of the Baudot-code (see below) by Donald Murray in 1901, in which he finally introduced a white-space code.
-This principle has been followed ever since, and for example was included in the second version of the International Telegraph Alphabet (ITA2) in 1930.)
-
-From those encodings, true \textsc{binary codes} developed with a fixed length of signals per character.
-In such systems only a single separator between signals is needed, because the separation between characters can be established by counting until a fixed number of signals.\footnote{Of course, no explicit separator is needed when the timing of the signals is known, which is the principle used in all modern telecommunication systems.
-An important modern consideration is also how to know where to start counting when you did not catch the start of a message, something that is known in Unicode as `self synchronizing'.) In the context of electric telegraphy, such a system was first established by Émile Baudot in 1870, using a fixed combination of five signals for each written character.\footnote{True binary codes have a longer history, going at least back to the Baconian cipher devised by Francis Bacon in 1605.) There are 25 = 32 possible combination when using five binary signals; an encoding today designated as `5-bit'.
-These codes are sufficient for all Latin letters, but of course they do not suffice for all written symbols, including punctuation and digits.
-As a solution, the Baudot code uses a so-called `shift' character, which signifies that from that point onwards (until shifted back) a different encoding is used, allowing for yet another set of 32 codes.
-In effect, this means that the Baudot code, and the \textsc{International Telegraph Alphabet (ITA)} derived from it, had an extra `bit' of information, so the encoding is actually 6-bit (with 26 = 64 different possible characters).
-For decades, this encoding was the standard for all telegraphy and it is still in limited use today.
-
-To also allow for different uppercase and lowercase letters and a large variety of control characters, the American Standards Association decided to propose a new 7-bit encoding in 1963 (with 27 = 128 different possible characters), known as the \textsc{American Standard Code for Information Interchange(ASCII)}, geared towards the encoding of English orthography.
-With the ascent of other orthographies in computer usage, the wish to encode further variation of Latin letters (like German <ß> or various letters with diacritics like <è>) led the Digital Equipment Corporation to introduce an 8-bit \textsc{Multinational Character Set (MCS)} (with 28 = 256 different possible characters), first used with the introduction of the VT220 Terminal in 1983.
-Because 256 characters were clearly not enough for the many different characters needed in the world's writing systems, the ISO/IEC 8859 standard in 1987 extended the MCS to include 16 different 8-bit code pages.
-For example, part 5 was used for Cyrillic characters, part 6 for Arabic, and part 7 for Greek.\footnote{In effect, because 16 = 24, this means that ISO/IEC 8859 was actually a 8+4 = 12-bit encoding, though with very many duplicates by design (e.g.~all ASCII codes were repeated in each 8-bit code page).
-To be precise, ISO/IEC 8859 used the 7-bit ASCII as a basis for each code page, and defined 16 different 7-bit extensions, leading to (1+16) times 27 = 2,176 possible characters.)
-
-This system almost immediately was understood to be insufficient and impractical, so various initiatives to extend and reorganize the encoding started in the 1980s.
-This led, for example, to various proprietary encodings from Microsoft (e.g.~Windows Latin 1) and Apple (e.g.~Mac OS Roman), which one still sometimes encounters today.
-More wide-ranging, various people in the 1980s started to develop true international code sets.
-In the United States, a group of computer scientists formed the Unicode Consortium, proposing a 16-bit encoding in 1991 (with 216 = 65,536 different possible characters).
-At the same time in Europe, the International Organization for Standardization (ISO) was working on ISO 10646 to supplant the ISO 8859 standard.
-Their first draft of the \textsc{Universal Character Set (UCS)} in 1990 was 31-bit (with theoretically 231 = 2,147,483,648 possible characters, but because of some technical restrictions only 679,477,248 were allowed).
-Since 1991, the Unicode Consortium and the International Organization for Standardization have jointly developed the \textsc{Unicode Standard}, or ISO/IEC 10646, leading to the current system including the original 16-bit Unicode proposal as the \textsc{Basic Multilingual Plane}, and 16 additional planes of 16-bit for further extensions (with in total 1+16 times 216 = 1,114,112 possible characters).
-The most recent version of the Unicode Standard (7.0) was published in June 2014 and it defines 112,956 different characters (Unicode Consortium 2014).
-
-\section{3. Terminology}\label{terminology}
-
-\subsection{The linguistic tradition}\label{the-linguistic-tradition}
+\section{The linguistic terminological tradition}
+\label{the-linguistic-tradition}
 
 In this section we provide a very brief overview of the linguistic terminology concerning writing systems before turning to the slightly different computational terminology in the next section on the Unicode Standard.
+
 Linguistically speaking, a \textsc{writing system} is a symbolic system that uses visible or tactile signs to represent language in a systematic way.
@@
 First, it may refer to the way a particular language is written.
-In this sense the term refers to the writing system of a particular language, as, for example, in \emph{``the Serbian writing system uses two scripts: Latin and Cyrillic''}.
-Second, the term `writing system' may refer to a type of symbolic system as used among the world's languages to represent languages, as, for example, in \emph{``alphabetic writing system''}.
-In this latter sense the term refers to how scripts have been classified according to the way that they encode language, as in, for example, \emph{``the Latin and Cyrillic scripts are both alphabetic writing systems''}.
-To avoid confusion, this second notion of `writing system' would have been more aptly called \textsc{script system}.
+In this sense the term refers to the writing system of a particular language, as, for example, in \emph{`the Serbian writing system uses two scripts: Latin and Cyrillic'}.
+Second, the term `writing system' may also refer to a type of symbolic system as used among the world's languages to represent the language, as, for example, in \emph{`alphabetic writing system'}.
+In this latter sense the term refers to how scripts have been classified according to the way that they encode language, as in, for example, \emph{`the Latin and Cyrillic scripts are both alphabetic writing systems'}.
+To avoid confusion, this second notion of `writing system' would more aptly have been called `script system'.
+
 The term \textsc{script} refers to a collection of distinct symbols as employed by one or more orthographies.
@@
 
-Only focussing here on the first sense, we distinguish two different kinds of writing systems as used to write a particular language, namely transcriptions and orthographies.
+Only focussing here on the first sense of writing systems, we distinguish two different kinds of writing systems used to write a particular language, namely transcriptions and orthographies.
 First, \textsc{transcription} is a scientific procedure (and also the result of that procedure) for graphically representing the sounds of human speech at the phonetic level.
 It incorporates a set of unambiguous symbols to represent speech sounds, including conventions that specify how these symbols should be combined.
-A \textsc{transcription system} is a specific system of symbols and rules used for transcription of the sounds of a spoken language variety.
+A transcription system is a specific system of symbols and rules used for transcription of the sounds of a spoken language variety.
 In principle, a transcription system should be language-independent, in that it should be applicable to all spoken human languages.
 The International Phonetic Alphabet (IPA) is a commonly used transcription system that provides a medium for transcribing languages at the phonetic level.
-However, there is a long history of alternative kinds of transcription systems (see Kemp 1994) and today various alternatives are in widespread use (e.g.~X-SAMPA and Cyrillic-based phonetic transcription systems).\footnote{Most users of IPA do not follow the standard explicitly and many dialects based on it have emerged, e.g.~the Africanist and Americanist transcription systems.
-) Note that IPA symbols are also often used to represent language on a phonemic level.
+However, there is a long history of alternative kinds of transcription systems (see Kemp 1994) and today various alternatives are in widespread use (e.g.~X-SAMPA and Cyrillic-based phonetic transcription systems).\footnote{Most users of IPA do not follow the standard explicitly and many dialects based on it have emerged, e.g.~the Africanist and Americanist transcription systems.} Note that IPA symbols are also often used to represent language on a phonemic level.
 However, it is important to realize that in this usage the IPA symbols are not a transcription system, but rather an orthography (though with strong links to the pronunciation).
 Further, a transcription system does not need to be highly detailed, but it can also be a system of broad sound classes.
-Although such an approximative transcription is not normally used in linguistics, it is widespread in technological approaches (e.g.~Soundex, Beider \& Morse 2010)\footnote{http://stevemorse.org/phonetics/bmpm.htm , http://stevemorse.org/phonetics/bmpm2.htm) and it is sometimes fruitfully used in automatic approaches to historical linguistics (Dolgopolsky 1986, ASJP\footnote{http://asjp.clld.org/)).
+Although such an approximative transcription is not normally used in linguistics, it is widespread in technological approaches (e.g.~Soundex, Beider \& Morse 2010)\footnote{http://stevemorse.org/phonetics/bmpm.htm , http://stevemorse.org/phonetics/bmpm2.htm} and it is sometimes fruitfully used in automatic approaches to historical linguistics (Dolgopolsky 1986, ASJP).\footnote{http://asjp.clld.org/}
 
@@
 All orthographies are language-specific (and often even resource-specific), although individual symbols or rules might be shared between languages.
+
 A \textsc{practical orthography} is a strongly phoneme-based writing system designed for practical use by speakers.
@@
 Symbols from the IPA are often used by linguists in the development of practical orthographies for languages without writing systems, though this usage of IPA symbols should not be confused with transcription (as defined above).
+
 Further, a \textsc{transliteration} is a mapping between two different orthographies.
@@
 In the linguistic terminology of writing systems, a \textsc{character} is a general term for any self-contained element in a writing system.\footnote{There is a second interpretation of the term `character', i.e.~a conventional term for a unit in the Chinese writing system (Daniels 1996).
-This interpretation will not be further explored in this paper.) Although in literate societies most people have a strong intuition about what the characters are in their particular orthography or orthographies, it turns out that the separation of an orthography into separate characters is far from trivial.
+This interpretation will not be further explored in this paper.} Although in literate societies most people have a strong intuition about what the characters are in their particular orthography or orthographies, it turns out that the separation of an orthography into separate characters is far from trivial.
 The widespread intuitive notion of a character is strongly biased towards educational traditions, like the alphabet taught at schools, and technological possibilities, like the available type pieces in a printer's job case, the keys on a typewriter, or the symbols displayed in Microsoft Word's symbol browser.
@@
 In alphabetic scripts, this problem is most forcefully exemplified by diacritics.
+
 A \textsc{diacritic} is a mark, or series of marks, that may be above, below, or through other characters (Gaultney 2002).
 Diacritics are sometimes used to distinguish homophonous words, but they are more often used to indicate a modified pronunciation (Daniels \& Bright 1996, xli).
-The central question is whether, for example, , , and should be considered four characters, or different combinations of three characters.
+The central question is whether, for example, <e>, <è>, <a> and <à> should be considered four characters, or different combinations of three characters.
 In general, multiple characters together can form another character, and it is not always possible to decide on principled grounds what should be the basic building blocks of an orthography.
@@
 For example, <x> in English orthography represents a combination of the phonemes /k/ and /s/.
+
 Further, conditioned or free variants of a grapheme are called \textsc{allographs}.
@@
 
-\subsection{The Unicode approach}\label{the-unicode-approach}
+\section{The Unicode approach}
+\label{the-unicode-approach}
 
-The conceptualisation and terminology of writing systems was rejuvenated through the development of the Unicode Standard, with major input from Mark Davis, co-founder and long-term president of the Unicode Consortium.
-For many years, `exotic' writing systems and phonetic transcription systems on personal computers were constrained by the American Standard Code for Information Interchange* *(ASCII) character encoding scheme (based on the Latin script), which only allowed for a strongly limited number of different symbols to be encoded.
+The conceptualization and terminology of writing systems was rejuvenated through the development of the Unicode Standard, with major input from Mark Davis, co-founder and long-term president of the Unicode Consortium.
+For many years, `exotic' writing systems and phonetic transcription systems on personal computers were constrained by the American Standard Code for Information Interchange (ASCII) character encoding scheme (based on the Latin script), which only allowed for a strongly limited number of different symbols to be encoded.
 This implied that users could either use and adopt the (extended) Latin alphabet or they could assign new symbols to the small number of code points in the ASCII encoding scheme to be rendered by a specifically designed font (Bird \& Simons 2003).
@@
 For a general introduction to the principles of Unicode, Chapter 2 of the core specification, called ``General Structure'', is particularly insightful.
-Different from many other documents of the Unicode Standard, this general introduction is relatively easy to read and illustrated with many interesting examples from various orthography traditions all over the world.
-) The \textsc{core specification} is a book directed toward human readers that describes the formal standard for encoding multilingual text.
+Different from many other documents of the Unicode Standard, this general introduction is relatively easy to read and illustrated with many interesting examples from various orthography traditions all over the world.} The \textsc{core specification} is a book directed toward human readers that describes the formal standard for encoding multilingual text.
 The \textsc{code charts} provide a humanly readable online reference to the character contents of the Unicode Standard in the form of PDF files.
@@
 In the actual practice of applying this standard in a computational setting, a specific implementation is necessary.
-The most widely used implementation of the Unicode Standard is the \textsc{International Components for Unicode (ICU)}, which offers C/C++ and Java libraries trying to implement to Unicode Standard.\footnote{http://icu-project.org )
+The most widely used implementation of the Unicode Standard is the \textsc{International Components for Unicode (ICU)}, which offers C/C++ and Java libraries trying to implement to Unicode Standard.\footnote{http://icu-project.org}
 
@@
 A \textsc{code point} is a unique non-negative integer within a code space (i.e.~within a certain numerical range).
-In the Unicode Standard character encoding system, an \textsc{abstract character}, for example a ``LATIN SMALL LETTER P'', is mapped to a particular code point, such as the decimal value 112, normally represented in hexadecimal and in Unicode parlance as (U+)0070.\footnote{Hexadecimal (`base-16') 0070 is equivalent to decimal (`base-10') 112, which can be calculated by considering 0·163 + 0·162 + 7·161 + 0·160 = 7·16 = 112.
-Underlyingly, computers will of course treat this code point binary (`base-2') as 11100000, i.e.~112 = 1·27 + 1·26 + 1·25 + 0·24 + 0·23 + 0·22 + 0·21 + 0·20 = 64 + 32 + 16.) That encoded abstract character is rendered on a computer screen (or printed page) as a \textsc{glyph}, e.g.
+In the Unicode Standard character encoding system, an \textsc{abstract character}, for example a ``LATIN SMALL LETTER P'', is mapped to a particular code point, such as the decimal value 112, normally represented in hexadecimal and in Unicode parlance as (U+)0070.\footnote{Hexadecimal (`base-16') 0070 is equivalent to decimal (`base-10') 112, which can be calculated by considering that $(0\cdot16^3) + (0\cdot16^2) + (7\cdot16^1) + (0\cdot16^0) = 7\cdot16 = 112$.
+Underlyingly, computers will of course treat this code point binary (`base-2') as 11100000, as can be seen by calculating that $(1\cdot2^7) + (1\cdot2^6) + (1\cdot2^5) + (0\cdot2^4) + (0\cdot2^3) + (0\cdot2^2) + (0\cdot2^1) + (0\cdot2^0) = 64 + 32 + 16 = 112$.} That encoded abstract character is rendered on a computer screen (or printed page) as a \textsc{glyph}, e.g.
 <p>, depending on the \textsc{font} and the context in which that character appears.
@@
 In contrast, a \textsc{glyph} is a concrete graphical representation of a character as it appears when rendered (or rasterized) and displayed on an electronic device or on printed paper.
-For example, <reinstert Gs here> are different glyphs of the same character, i.e.~they may be rendered differently depending on the typography being used, but they all share the same code point.
+For example, <reinsert Gs here> are different glyphs of the same character, i.e.~they may be rendered differently depending on the typography being used, but they all share the same code point.
 From the perspective of Unicode they are `the same thing'.
@@
 A simplified overview of the properties is available at http://userguide.icu-project.org/strings/properties.
-The actual code tables listing all properties for all Unicode code points are available at http://www.unicode.org/Public/UCD/latest/ucd/.
-) Basically, those properties are just a long list of values for each character.
+The actual code tables listing all properties for all Unicode code points are available at http://www.unicode.org/Public/UCD/latest/ucd/.} Basically, those properties are just a long list of values for each character.
 For example, code point U+0047 has the following properties (among many others):
@@
 These character combinations are defined in the Unicode Standard as functioning as one \textsc{text element}.\footnote{The Unicode Standard defines text element as: ``A minimum unit of text in relation to a particular text process, in the context of a given writing system.
-In general, the mapping between text elements and code points is many-to-many.'') The simplest example of a grapheme cluster is a base character followed by a letter modifier character.
+In general, the mapping between text elements and code points is many-to-many.''} The simplest example of a grapheme cluster is a base character followed by a letter modifier character.
 For example, the sequence + <̃\textgreater{} (LATIN SMALL LETTER N at U+006E followed by COMBINING TILDE at U+0303) combines visually into , a user-percieved character in writing systems like that of Spanish.
@@
 The Unicode Standard provides technological specifications for creating locale specific data in so-called \textsc{Unicode Locale Descriptions}, i.e.~a set of specification that defines a set of language-specific elements (e.g.~tailored grapheme clusters, collation order, capitalization-equivalence), as well as other special information, like how to format numbers, dates, or currencies.
-Locale descriptions are saved in the \textsc{Common Locale Data Repository (CLDR)},\footnote{http://cldr.unicode.org/ ) a repository of language-specific definitions of writing system properties, each of which describes specific usages of characters.
+Locale descriptions are saved in the \textsc{Common Locale Data Repository (CLDR)},\footnote{http://cldr.unicode.org/} a repository of language-specific definitions of writing system properties, each of which describes specific usages of characters.
 Each locale can be encoded in a document using the \textsc{Locale Data Markup Language (LDML)}.
@@
 However, there are various drawbacks of locale descriptions for the daily practice of linguistic work in a multilingual setting (see Section 7).
-
-\section{4. Unicode Pitfalls}\label{unicode-pitfalls}
-
-\subsection{Wrong it is not}\label{wrong-it-is-not}
-
-In this section, we describe some of the most common pitfalls that we have encountered when using the Unicode Standard in our own work, or in discussion with other linguists.
-This section is not meant as a criticism of the decisions made by the Unicode Consortium; on the contrary, we aim to highlight where the technological aspects of the Unicode Standard diverge from many users' intuitions.
-What have sometimes been referred to as problems or inconsistencies in the Unicode Standard are mostly due to legacy compatibility issues, which can lead to unexpected behavior by linguists using the standard.
-However, there are also some cases in which the Unicode Standard has made decisions that theoretically could have been taken differently, but for some reason or another (mostly very good reasons) were accepted as they are now.
-We call behavior that executes without error but does something different than the user expected---often unknowingly---a ``pitfall''.
-
-In this context, it is important to realize that the Unicode Standard was not developed to solve linguistic problems per se, but to offer a consistent computational environment for written language.
-In those cases in which the Unicode Standard behaves differently as expected, we think it is important not to dismiss Unicode as `wrong' or `deficient', because our experience is that in almost all cases the behaviour of the Unicode Standard has been particularly well thought through.
-The Unicode Consortium considers important practical use-cases that from a linguistic point of view are normally not considered.
-Our general guideline for dealing with the Unicode Standard is to accept it as it is, and not to battle windmills.
-Alternatively, of course, it is possible to actively engage in the development of the standard itself, an effort that is highly appreciated by the Unicode Consortium.
-
-\subsection{Pitfall: Blocks}\label{pitfall-blocks}
-
-The Unicode code space is subdivided into blocks of contiguous code points.
-For example, the block called ``Cyrillic'' runs from U+0400 till U+04FF.
-These blocks arose as an attempt at ordering the enormous amount of characters in Unicode, but the ideas of blocks very quickly ran into problems.
-First, the size of a block is fixed, so when a block is full, a new block will have to be instantiated somewhere further in the code space.
-For example, this led to the blocks ``Cyrillic Supplement'', ``Cyrillic Extended-A'' (which are also already full) and ``Cyrillic Extended-B''.
-Second, when a specific character already exists, then it is not duplicated in another block, although the name of the block might indicate that a specific symbol should be available there.
-In general, names of blocks are just an approximate indication of the kind of characters that will be in the block.
-
-The problem with blocks arises because finding the right character among the thousands of Unicode characters is not easy.
-Many software applications present blocks as a primary search mechanism, because the block names suggest where to look for a particular character.
-However, when a user searches for an IPA character in the block ``IPA Extensions'' then many IPA characters will not be found there.
-For example, the velar nasal <ŋ> is not part of the block ``IPA Extensions'' because it was already included as LATIN SMALL LETTER ENG at U+014B in the block ``Latin Extensions-A''.
-
-In general, finding a specific character in the Unicode Standard is often not trivial.
-The names of the blocks can help, but it is not (and never was supposed to be) a foolproof structure.
-It is not the goal nor aim of the Unicode Consortium to provide a user interface to the Unicode Standard.
-If one often encounters the problem of needing to find a suitable character, there are various other useful services for end users available.\footnote{The Unicode website offers a basic interface to the code charts at http://www.unicode.org/charts/index.html.
-As a more flexible interface, we particularly like PopChar from Macility, available for both Macintosh and Windows.
-However, there are various free websites that offer search interfaces to the Unicode code tables, like http://unicode-search.net or http://unicode-search.net.
-A further useful approach for searching characters using shape matching is http://shapecatcher.com.
-)
-
-\subsection{Pitfall: Missing glyphs}\label{pitfall-missing-glyphs}
-
-The Unicode Standard is often praised (and deservedly so) for solving many of the perennial problems with the interchange and display of the world's writing systems.
-However, a common complaint from users is that, while the praise may be true, they mostly just see some boxes on their screen instead of those promised symbols.
-The problem of course is that users' computers do not have any glyphs installed matching the Unicode code points in the file they are trying to inspect.
-It is important to realize that internally in the computer everything still works as expected: any handling of Unicode code points works independently of how they are displayed on the screen.
-So, although a user might only see boxes being displayed, this user should be assured that everything is still in order.
-
-The central problem behind the missing glyphs is that designing actual glyphs includes a lot of different considerations and it is a time-consuming process.
-Many traditional expectations of how specific characters should look like have to be taken into account when designing glyphs.
-Those expectations are often not well documented, and it is mostly up to the knowledge and experience of the font designer to try and conform to them as good as possible.
-Therefore, most designers produce fonts only including glyphs for certains parts of the Unicode Standard, namely for those characters they feel comfortable with.
-At the same time, the number of characters defined by the Unicode Standard is growing with each new version, so it is neigh impossible for any designer to produce glyphs for all characters.
-The result of this is that, almost necessarily, each font only includes glyphs for a subset of the characters in the Unicode Standard.
-
-The simple solution to missing glyphs is thus to install additional fonts providing additional glyphs.
-For the more exotic characters there is often not much choice.
-There are a few particularly large fonts that might be considered.
-First, there is the \textsc{Everson Mono} font made by Michael Everson, which currently includes 9,756 different glyphs (not including Chinese) updated up to Unicode 7.0.\footnote{Everson Mono is available as shareware at http://www.evertype.com/emono/.
-) Already a bit older is the \textsc{Titus Cyberbit Basic} font made by Jost Gippert and Carl-Martin Bunz, which includes 10,044 different glyphs (not including Chinese), but not including newer characters added after Unicode 4.0.\footnote{Titus Cyberbit Basic is available at http://titus.fkidg1.uni-frankfurt.de/unicode/tituut.asp.
-)
-
-Further, we suggest to always install at least one so-called \textsc{fall-back font,} which provides glyphs that at least show the user some information about the underlyingly encoded character.
-Apple Macintoshes have such a font (which is invisible to the user), which is designed by Michael Everson and made available for other systems through the Unicode Consortium.\footnote{The Apple/Everson fallback font is available at http://www.unicode.org/policies/lastresortfont\_eula.html.
-) Further, the \textsc{GNU Unifont} is a clever way to produce bitmaps approximating the intended glyph of each available character, updated to Unicode 7.0.\footnote{The GNU Unifont is available at http://unifoundry.com/unifont.html.
-) Finally, the Summer Institute of Linguistics provides a \textsc{SIL Unicode BMP Fallback Font}, currently available up to Unicode version 6.1.
-This font does not even attempt to show a real glyph, but only shows the hexadecimal code inside a box for each character, so a user can at least see the Unicode codepoint of the character to be displayed.\footnote{The SIL Unicode BMP Fallback Font is available at http://scripts.sil.org/UnicodeBMPFallbackFont.
-)
-
-\subsection{Pitfall: Faulty rendering}\label{pitfall-faulty-rendering}
-
-A similar complaint to missing glyphs, discussed previously, is that while there might be a glyph being displayed, it does not `look right'.
-There are two reasons for unexpected visual display, namely automatic font substitution and faulty rendering.
-Like missing glyphs, any such problems are independent from the Unicode Standard.
-The Unicode Standard only includes very general information about characters and leaves the specific visual display to others to decide on.
-Any faulty display is thus not to be blamed on the Unicode Consortium, but on a complex interplay of different mechanisms happening in a computer to turn Unicode codepoints into visual symbols.
-We will only sketch a few aspects of this complex interplay here.
-
-Most modern software applications (like Microsoft Word) offer some approach to \textsc{automatic font substitution}.
-This means that when a text is written in a specific font (e.g.~Times New Roman) and a inserted Unicode character does not have a glyph within this font, then the software application will automatically search for another font to display the glyph.
-The result will be that some specific glyph will look slightly different from the others.
-This mechanism works differently depending on the software application, and mostly only limited user influence is expected and little feedback is given, which might be rather frustrating to font-aware users.\footnote{For example, Apple Pages does not give any feedback that a font is being replaced, and the user does not seem to have any influence on the choice of replacement (except by manually marking all occurrences).
-In contrast, Microsoft Word does indicate the font replacement by showing the name of the font replacement, but it simply changes the font, so any text written after the replacement is written in a different font as before.
-Both behaviors leave much to be desired.)
-
-The other problem with visual display is related to the so-called `font rendering'.
-\textsc{Font rendering }refers to the process of the actual positioning of Unicode characters on a page of written text.
-This positioning is actually a highly complex problem, and many things can go wrong in the process.
-Well-known rendering problems, like proportional glyph size or ligatures are reasonably well understood.
-However, even within the Latin script, the positioning of diacritics relative to a base character is still a widespread problem.
-Especially when more than one diacritic is supposed to be placed above (or below) each other, this often leads to unexpected effects in many modern software applications.
-The problems arising in Arabic and in many southeast Asian scripts (like Devanagari or Burmese) are even much more complex.
-There are basically three different approaches to font rendering.
-The most widespread is Adobe's and Microsoft's \emph{OpenType }system.
-This approach makes it relatively easy for font developers, as the font itself does not include all details about the precise placement of individual characters.
-For those details, additional script-descriptions are necessary.
-All of those systems can lead to unexpected behaviour.\footnote{For more details about OpenType, see http://www.adobe.com/products/type/opentype.html and http://www.microsoft.com/typography/otspec/.
-Additional systems for complex text layout are, among others, Microsoft's DirectWrite https://msdn.microsoft.com/library/dd368038.aspx and the open-source project HarfBuzz http://www.freedesktop.org/wiki/Software/HarfBuzz/.
-) Alternative systems are Apple's \emph{Apple Advanced Typography} (AAT) and the open-source \emph{Graphite} system from the Summer Institute of Linguistics (SIL).\footnote{More information about AAT can be found at https://developer.apple.com/fonts/.
-SIL's Graphite is described in detail at http://scripts.sil.org/cms/scripts/page.php?site\_id=projects\&item\_id=graphite\_home.
-) In both of these systems, a larger burden is placed on the description inside the font.
-
-\subsection{Pitfall: Homoglyphs}\label{pitfall-homoglyphs}
-
-Homoglyphs are visually indistinguishable glyphs (or highly similar glyphs) that have different code points in the Unicode Standard and thus different character semantics.
-As a principle, the Unicode Standard does not specify how a character appears visually on the page or the screen.
-So in most cases, a different appearance is caused by the specific design of a font, or by user-settings like size or boldface.
-Taking an example already discussed in the previous section, the following symbols <reinsert Gs here> are different glyphs of the same character, i.e.~they may be rendered differently depending on the typography being used, but they all share the same code point (viz.
-LATIN SMALL LETTER G at U+0067).
-In contrast, the symbols <A А Α Ꭺ ᗅ ᴀ ꓮＡ𐊠 𐌀 𝖠 𝙰> are all different code points, although they look highly similar -- in some cases even sharing exactly the same glyph in some fonts.
-All these different A-like characters include the following code points in the Unicode Standard:
-\begin{itemize}
-	\item <A\textgreater{} LATIN CAPITAL LETTER A, at U+0041 
-	\item <А> CYRILLIC CAPITAL LETTER A, at U+0410 
-	\item <Α> GREEK CAPITAL LETTER ALPHA, at U+0391 
-	\item <Ꭺ> CHEROKEE LETTER GO, at U+13AA 
-	\item <ᗅ> CANADIAN SYLLABICS CARRIER GHO, at U+15C5 
-	\item <ᴀ> LATIN LETTER SMALL CAPITAL A, at U+1D00 
-	\item <ꓮ> LISU LETTER A, at U+A4EE 
-	\item <Ａ> FULLWIDTH LATIN CAPITAL LETTER A, at U+FF21 
-	\item <𐊠> CARIAN LETTER A, at U+102A0 
-	\item <𐌀> OLD ITALIC LETTER A, at U+10300 
-	\item <𝖠> MATHEMATICAL SANS-SERIF CAPITAL A, U+1D5A0 
-	\item <𝙰> MATHEMATICAL MONOSPACE CAPITAL A, at U+1D670 
-\end{itemize}
-
-The existence of such homoglyphs is partly due to legacy compatibility, but for the most part these characters are simply different characters that happen to look similar.\footnote{A particularly nice interface to look for homoglyphs is http://shapecatcher.com, based on the principle of recognizing shapes (Belongie et al.
-2002).) Yet, they are supposed to behave different from the perspective of a font designer.
-For example, when designing a Cyrillic font, the will have different aesthetics and different traditional expectation compared to a Latin .
-
-Such homoglyphs are a widespread problem for consistent encoding.
-Although for most users it looks like the words and are identical, in actual fact they do not even share a single code point.\footnote{The first words consists completely of Latin characters, namely U+0076, U+006F, U+0063, U+0065, U+0073, while the second is a mix of Cyrillic and Greek characters, namely U+03BD, U+03BF, U+0041, U+0435, U+0455.) For computers these two words are completely different entities.
-Commonly, when users with Cyrillic or Greek keyboards have to type some Latin-based orthography, they mix similar looking Cyrillic or Greek characters into their text, because those characters are so much easier to type.
-Similarly, when users want to enter an unusual symbol, they normally search by visual impression in their favourite software application, and just pick something that looks reasonably alike to what they expect the glyph to look like.
-
-\subsection{Pitfall: Canonical equivalence}\label{pitfall-canonical-equivalence}
-
-For some characters, there is more than one possible encoding in the Unicode Standard.
-This is a possible pitfall, as this would mean that for the computer there exist multiple different entities that for a user are the same.
-This would, for example, lead to problems with searching, as the computer would search for specific encodings, and not find all expected characters.
-As a solution, the Unicode Standard includes a notion of \textsc{canonical equivalence}.
-Different encodings are explicitly declared as equivalent in the Unicode Standard code tables.
-Further, to harmonize all encodings in a specific piece of text, the Unicode Standard proposes a mechanism of \textsc{normalization}.
-
-Consider for example the characters and following Unicode code points:
-\begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item LATIN CAPITAL LETTER A WITH RING ABOVE (U+00C5) 
-	\item ANGSTROM SIGN (U+212B) 
-	\item LATIN CAPITAL LETTER A (U+0041) + COMBINING RING ABOVE (U+030A) 
-\end{enumerate}
-
-The character, represented here by glyph , is encoded in the Unicode Standard in the first two examples by a single-character sequence; each is assigned a different code point.
-In the third example, the glyph is encoded in a multiple-character sequence that is composed of two character code points.
-All three sequences are \textsc{canonically equivalent}, i.e.~they are strings that represent the same abstract character and because they are not distinguishable by the user, the Unicode Standard requires them to be treated the same in regards to their behavior and appearance.
-Nevertheless, they are encoded differently.
-For example, if one were to search an electronic text (with software that does not apply Unicode Standard normalization) for ANGSTROM SIGN (U+212B), instances of LATIN CAPITAL LETTER A WITH RING ABOVE (U+00C5) would not be found.
-
-In other words, there are equivalent sequences of Unicode characters that should be normalized, i.e.~transformed into a unique Unicode-sanctioned representation of a character sequence called a \textsc{normalization form}.
-Unicode provides a \emph{Unicode Normalization Algorithm}, which essentially puts combining marks into a specific logical order and it defines decomposition and composition transformation rules to convert each string into one of four normalization forms.
-We will discuss here the two most relevant normalization forms: NFC and NFD.
-
-The first of the three characters above is considered the \textsc{Normalization Form C (NFC)}, where ``C'' stands for composition.
-When the process of NFC normalization is applied to the character sequences in 2 and 3, both sequences are normalized into the `pre-composed' character sequence in 1.
-Thus all three canonical character sequences are standardized into one composition form in NFC.
-The other central Unicode normalization form is the \textsc{Normalization Form D (NFD)}, where ``D'' stands for decomposition.
-When NFD is applied to the three examples above, all three, including importantly the single-character sequences in 1 and 2, are normalized into the `decomposed' multiple-sequence of characters in 3.
-Again, all three are then logically equivalent and therefore comparable and syntactically interoperable.
-
-As illustrated, some characters in the Unicode Standard have alternative representations (in fact, many do), but the Unicode Normalization Algorithm can be used to transform certain sequences of characters into canonical (or compatible) forms to test for equivalency.
-To determine equivalence, each character in the Unicode Standard is associated with a combining class, which is formally defined as a character property called ``Canonical Combining Class'' which is specified in the Unicode Character Database.
-The combining class assigned to each code point is a numeric value between 0 and 254 and is used by the Canonical Ordering Algorithm to determine which sequences of combining marks are canonically equivalent.
-Normalization forms, as very briefly described above, can be used to ensure character equivalence by ordering character sequences so that they can be faithfully compared.
-
-It is very important to note that any software applications that is Unicode Standard compliant is free to change the character stream from one representation to another.
-This means that a software application may compose, decompose or reorder characters as its developers desire; as long as the resultant strings are canonically equivalent to the original.
-This might lead to unexpected behaviour for users.
-Various players, like the Unicode Consortium, the W3C, or the TEI recommend NFC in most user-directed situations, and some software applications that we tested indeed seem to automatically convert strings into NFC.\footnote{See the summary of various recommendation here: http://www.win.tue.nl/\textasciitilde{}aeb/linux/uc/nfc\_vs\_nfd.html.
-) This means in practice that if a user, for example, enters <a> and <`>, i.e.~LATIN SMALL LETTER A at U+0061 and COMBINING GRAVE ACCENT at U+0300, this might be converted into <à>, i.e.~LATIN SMALL LETTER A WITH GRAVE at U+00E0.\footnote{The behaviour of software applications can be quite erratic in this respect.
-For example, Apple's TextEdit does not do any conversion on text entry.
-However, when you copy and paste some text inside the same document in ``rich text mode'' (i.e.~RTF-format), it will be transformed into NFC on paste.
-Saving a document does not do any conversion to the glyphs on screen, but it will save the characters in NFC.)
-
-\subsection{Pitfall: Characters are not glyphs}\label{pitfall-characters-are-not-glyphs}
-
-A central principle of Unicode is the distinction between character and glyph.
-A character is the abstract notion of a symbol in a writing system, while a glyph is the concrete drawing of such a symbol.
-In practice, there is a complex interaction between characters and glyphs.
-A single Unicode character may of course be rendered as a single glyph.
-However, a character may also be a `piece' of a glyph, or vice-versa.
-Actually, all possible relations between glyphs and characters are attested.
-
-First, \textsc{a single character may have different contextually determined glyphs.} For example, characters in writing systems like Hebrew and Arabic have different glyphs depending on where they appear in a word.
-Some letters in Hebrew change their form at the end of the word, and in Arabic, primary letters have four contextually-sensitive variants (isolated, word initial, medial and final).
-Second, \textsc{a single character may be rendered as a sequence of multiple glyphs.} For example, in Tamil one Unicode character may result in a combination of a consonant and vowel, which are rendered as two adjacent glyphs by fonts that supports Tamil.
-Third, \textsc{a single glyph may be a combination of multiple characters.
-}For example, the ligature <ﬁ>, a single glyph, is the result of two characters, <f> and <i>, that have undergone glyph substitution by font rendering.
-Like contextually-determined glyphs, ligatures are (intended) artifacts of text processing instructions.
-Finally, \textsc{a single glyph may be a part of a character} (e.g.~diacritics).
-
-Further, the rendering of a glyph is also always dependent on the font being used.
-For example, the Unicode character LATIN SMALL LETTER G appears as and in the Computer Modern and Courier fonts because their typefaces are designed differently.
-Furthermore, font face may change the visual appearance of a character, for example Times New Roman two-story <a> changes to a single-story a in italics <\emph{a}>.
-This becomes a real problem for some phonetic typesetting (see Section 5, Pitfall 2).
-
-In sum, character-to-glyph mappings are complex technological issues that the Unicode Consortium has had to address in the development of the Unicode Standard, but for the lay user they can be utterly confusing because visual rendering does not indicate logical encoding, or, in other words, what you see isn't (necessarily) what you get.
-
-\subsection{Pitfall: File formats}\label{pitfall-file-formats}
-
-Unicode is a character encoding standard, but these characters of course actually appear inside some kind of file.
-The most basic Unicode-based file format is pure line-based text, i.e.~strings of Unicode-encoded characters separated by line breaks (note that these line breaks are what for most people intuitively corresponds to paragraph breaks).
-Unfortunately, even within this apparently basic setting there exist a multitude of variants.
-In general, these different possibilities are well-understood in the software industry, and nowadays they normally do not lead to any problems for the end user.
-However, there are some situations in which a user is suddenly confronted with cryptic questions in the user interface involving abbreviations like LF, CR, BE, LE or BOM.
-Most prominently this occurs with exporting or importing data in several software applications from Microsoft.
-Basically, there are two different issues involved.
-First, the encoding of** line breaks** and, second, the encoding of the Unicode characters into \textsc{code units} and the related issue of \textsc{endianness}.
-
-The issue with \textsc{line breaks} originated with the instructions necessary to direct a printing head to a new line.
-This involves two movements, known as \textsc{carriage return (CR)} (returning the printing head to the start of the line on the page) and \textsc{line feed (LF)} (moving the printing head to the next line on the page).
-Physically, these are two different events, but conceptually together they form one action.
-In the history of computing, various encodings of line breaks have been used (e.g.~CR+LF, LF+CR, only LF, or only CR).
-Currently, all Unix and Unix-derived systems use only LF as code for a line break, while software from Microsoft still uses a combination of CR and LF.
-Today, most software applications recognize both options, and are able to deal with either encoding of line breaks (until rather recently this was not the case, and using the wrong line breaks could lead to unexpected errors).
-Our impression is that there is a strong tendency in software development to standardize on the simpler ``only LF'' encoding for line breaks, and we suggest that everybody use this encoding whenever possible.
-
-A code unit is a sequence of bits used to encode character in a encoding.
-The issue with code units stems from the question how to separate a stream of binary ones and zero, i.e.~bits, into chunks representing Unicode characters.
-Depending on different use cases, the Unicode Standard offers three different approaches, called UTF-32, UTF-16 and UTF-8.\footnote{The letters UTF stand for either ``Unicode Transformation Format'' or ``Universal Coded Character Set Transformation Format'', but the notion of `transformation' is a legacy notion that does not have meaning anymore.
-Nevertheless, the designation ``UTF'' (in capitals) has become an official standard designation, but should probably best be read as simply ``Unicode Format''.) The details of this issue is extensively explained in section 2.5 of the Unicode core specification.
-Basically, \textsc{UTF-32} encodes each character in 32 bits (32 `binary units', i.e.~32 zeros or ones) and is the most disk-space-consuming variant of the three.
-However, it is the most efficient encoding processing-wise, because the computer simply has to separate each character after 32 bits.
-In contrast, \textsc{UTF-16} uses only 16 bits per character, which is sufficient for the large majority of Unicode characters, but not for all of them.
-A special system of `surrogates' is defined within the Unicode Standard to deal with these additional characters.
-The effect is a more disk-space efficient encoding (approximately half the size), while adding a limited computational overhead.
-Finally, \textsc{UTF-8} is a more complex system that dynamically encodes each character with the minimally necessary number of bits, choosing either 8, 16 or 32 bits depending on the character.
-This represents again a strong reduction in space (particularly due to the high frequency of data using ASCII characters that need only 8 bits) at the expense of even more computation necessary to process such strings.
-However, because of the ever growing computational power of modern machines, the processing overhead is in most practical situations a non-issue, while saving on space is still useful, particularly for sending texts over the Internet.
-As an effect, UTF-8 has become the dominant encoding on the World Wide Web.
-We suggest that everybody uses UTF-8 as their default encoding.
-
-A related problem is a general issue about how to store information in computer memory, which is known as \textsc{endianness}.
-The details of this issue go beyond the scope of this paper.
-It suffices to realize that there is a difference between \textsc{big-endian} (BE) storage and \textsc{little-endian} (LE) storage.
-The Unicode Standard offers a possibility to explicitly indicate what kind of storage is used by starting a file with a so-called `Byte Order Mark' (BOM).
-However, the Unicode Standard does not require the usage of BOM, preferring other non-Unicode methods to signal to computers which kind of endianness is used.
-This issue only arises with UTF-32 and UTF-16 encodings, and when using the preferred UTF-8, using a BOM is theoretically possible, but strongly dispreferred according to the Unicode Standard.
-We suggest that everyone tries to prevent the inclusion of BOM in your data.
-
-Together with the topic of normalisation (NFC vs.~NFD, see Pitfall 5 above), this leads to the following nicely cryptic file-encoding recommendation, namely to preferably use \textsc{UTF-8 in NFC without BOM with LF line breaks}, whenever possible.
-
-\section{5. IPA meets Unicode}\label{ipa-meets-unicode}
-
-\subsection{Introducing the International Phonetic Alphabet (IPA)}\label{introducing-the-international-phonetic-alphabet-ipa}
-
-The International Phonetic Alphabet (IPA) is a common standard in linguistics to transcribe sounds of spoken language into some Latin-based characters (International Phonetic Association 1999).
-Although IPA is reasonably easily adhered to with pen and paper, it is not trivial to encode IPA characters electronically.
-Early work addressing the need for a universal computing environment for writing systems and their computational complexity is discussed in Simons 1989.
-For a long time, linguists (like all other computer users) were limited to ASCII-encoded 7-bit characters, which only includes Latin characters, numbers and some punctuation and symbols.
-Restricted to these standard character sets that lacked IPA support or other language-specific graphemes that they needed, linguists devised their own solutions (cf.~Bird and Simons 2003).
-For example, some chose to represent unavailable graphemes with substitutes, e.g.~the combination of to represent .
-Tech-savvy linguists redefined selected characters from a character encoding by mapping custom made fonts to those code points.
-However, one linguist's electronic text would not render properly on another linguist's computer without access to the same font.
-Further, if two character encodings defined two character sets differently, then data could not be reliably and correctly displayed.
-This is a common example of the non-interoperability of data and data formats.
-
-To alleviate this problem, during the late 1980s, SAMPA (Speech Assessment Methods Phonetic Alphabet) was created to represent IPA symbols with 7-bit printable ASCII character sequences, e.g.
-for {[}ɸ{]}.
-Two problems with SAMPA are that (i) it is only a partial encoding of the IPA and (ii) it encodes different languages in separate data tables, instead of a universal alphabet, like IPA.
-SAMPA tables are derived from phonemes appearing in several European languages that were developed as part of a European Commission-funded project to address technical problems like electronic mail exchange (what is now simply called email).
-SAMPA is essentially a hack to work around displaying IPA characters, but it provided speech technology and other fields a basis that has been widely adopted and used in code.
-So, SAMPA was a collection of tables to be compared, instead of a large universal table representing all languages.
-An extended version of SAMPA, called X-SAMPA, set out to include every symbol in the IPA chart including all diacritics (Wells nd.).
-X-SAMPA was considered more universally applicable because it consisted of one table that encoded the set of characters that represented phones/segments in IPA across languages.
-SAMPA and X-SAMPA have been widely used for speech technology and as an encoding system in computational linguistics.
-Eventually, ASCII-encoding of the IPA became depreciated through the advent of the Unicode Standard.
-Note however that many popular software packages used for linguistic analyses still require ASCII input, e.g.~RuG/L04\footnote{http://www.let.rug.nl/kleiweg/L04/) and SplitsTree4.\footnote{http://www.splitstree.org/)
-
-Still, there are a few pitfalls to be aware of when using the Unicode Standard to encode IPA.
-As we have said before, from a linguistic perspective it might look like the Unicode Consortium is making incomprehensible decisions, but it is important to realize that the consortium has tried and is continuing to try to be as consistent as possible across a wide range of use cases, and it does place linguistic traditions above other orthographic possibilities.
-In general, we strongly suggest linguists not to complain about any decisions in the Unicode Standard, but to try and understand the rationale of the Unicode Consortium (which in our experience is almost always well-conceived) and devise ways to work with any unexpected behaviour.
-Many of the current problems derive from the fact that the IPA is clearly historically based on the Latin script, but different enough from most other Latin-based writing systems to warrant special attention.
-This ambivalent status of the IPA glyphs (partly Latin, partly special) is unfortunately also attested in the treatment of IPA in the Unicode Standard.
-In retrospect, it might have been better to consider the IPA (and other transcription systems) to be a special or new kind of script within the Unicode Standard, and treat the obvious similarity to Latin glyphs as a historical relic.
-All IPA glyphs would then have their own code points, instead of the current situation in which some IPA glyphs have special code points, while others are treated as being identical to the `regular' Latin characters.
-Yet, the current situation, however unfortunate, is unlikely to change, so as linguists we will have to learn to deal with the specific pitfalls of IPA within the Unicode Standard.
-In this section, we will describe these pitfalls in some detail.
-
-\subsection{Pitfall: There is no single complete Unicode code block for IPA}\label{pitfall-there-is-no-single-complete-unicode-code-block-for-ipa}
-
-The ambivalent nature of IPA glyphs arises because, on the one hand, the IPA uses Latin-based glyphs like , or .
-From this perspective, the IPA seems to be just another orthographic tradition using Latin characters, all of which do not get a special treatment within the Unicode Standard (just like e.g.~the French, German, or Danish orthographic traditions do not have a special status).
-On the other hand, the IPA uses many special symbols (like turned , mirrored and/or extended Latin glyphs ) not found in any other Latin-based writing system.
-For this reason, and already in the first version of the Unicode Standard (Version 1.0 from 1991), a special block with code points, called ``IPA Extensions'' was included.
-As noted in Section 4, Pitfall 1, the Unicode Standard code space is subdivided into character blocks, which generally encode characters from a single script.
-However, as is illustrated by the IPA, characters that form a single writing system may be dispersed across several different character blocks.
-With its diverse collection of symbols from various scripts and diacritics, the IPA is spread across over 13 blocks in the Unicode Standard:\footnote{This number of blocks depends on whether only IPA-sanctioned symbols are counted or if the phonetic symbols commonly found in the literature are also included, see Moran 2012, Appendix C.)
-\begin{itemize}
-	\item Basic Latin (30 characters), e.g. <a, b, c, d, e> 
-	\item Latin-1 Supplement (4 characters): <æ, ç, ð, ø\textgreater{} 
-	\item Latin Extended-A (3 characters): <ħ, ŋ, œ> 
-	\item Latin Extended-B (5 characters): <ǀ, ǁ, ǂ, ǃ, ȵ> 
-	\item IPA Extensions (70 characters), e.g.: <ɐ, ɑ, ɔ> 
-	\item Spacing Modifier Letters (20 characters), e.g.: <ʰ ʷ ˥> 
-	\item Combining Diacritical Marks (33 characters), e.g.: <̝ ̥ ̪> 
-	\item Greek and Coptic (3 characters): <β, θ, χ> 
-	\item Phonetic Extensions (2 characters): <ᴅ, ᴴ> 
-	\item Phonetic Extensions Supplement (3 characters): <ᶑ , ᶾ, ᶣ> 
-	\item Superscripts and Subscripts (1 character): <ⁿ> 
-	\item Arrows (2 characters): \textless{}↑, ↓\textgreater{} 
-	\item Latin Extended-C (1 character): <ⱱ> 
-\end{itemize}
-
-\subsection{Pitfall: There are many IPA homoglyphs in Unicode}\label{pitfall-there-are-many-ipa-homoglyphs-in-unicode}
-
-Another problem is the large number of homoglyphs, i.e.~different characters that have highly similar glyphs (or even completely identical, depending on the font rendering).
-For example, a speaker of Russian should ideally not use the CYRILLIC SMALL LETTER A at code point U+0430 for IPA transcriptions, but instead the LATIN SMALL LETTER A at code point U+0061, although visually they are mostly indistinguishable, and the Cyrillic character is more easily typed on a Cyrillic keyboard.
-Another example we commonly encounter is the use of LATIN SMALL LETTER G at U+0067, instead of the sanctioned Unicode Standard IPA character for the voiced velar stop LATIN SMALL LETTER SCRIPT G at U+0261.
-One begins to question whether this issue is at all apparent to the working linguist, or if they simply use the U+0067 because it is easily keyboarded and thus saves time, whereas the latter must be cumbersomely inserted as a special symbol in most software.\footnote{This issue was recently addressed by the International Phonetic Association, which has taken the stance that both the keyboard LATIN SMALL LETTER G and the LATIN SMALL LETTER SCRIPT G are valid input characters for the voiced velar plosive.
-Unfortunately, this decision further introduces ambiguity for linguists trying to adhere to a strict Unicode Standard IPA encoding.)
-
-Furthermore, on the one hand even linguists are unlikely to distinguish between the LATIN SMALL LETTER SCHWA at code point U+0259 and LATIN SMALL LETTER TURNED E at U+01DD.
-On the other hand, non-linguists are unlikely to distinguish any semantic difference between an open back unrounded vowel , the LATIN SMALL LETTER ALPHA at U+0251, and the open front unrounded vowel , LATIN SMALL LETTER A at U+0061.
-But even among linguists this distinction leads to problems.
-For example, as pointed out by Mielke (2009), there is a problem stemming from the fact that about 75\% of languages are reported to have a five-vowel system (Maddieson 1984).
-Historically, linguistic descriptions tend not to include precise audio recording and measurements of formants, so this may lead one to ask if the many characters that are used in phonological description reflects a ``transcriptional bias''.
-The common use of in transcriptions could be in part due to the ease of typing the letter on an English keyboard (or for older descriptions, the typewriter).
-We found it to be exceedingly rare that a linguist uses for a low back unrounded vowel.\footnote{One example is Vidal 2001a:75, in which the author states: ``The definition of Pilagá /a/ as {[}+back{]} results from its behavior in certain phonological contexts.
-For instance, uvular and pharyngeal consonants only occur around /a/ and /o/.
-Hence, the characterization of /a/ and /o/ as a natural class of (i.e., {[}+back{]} vowels), as opposed to /i/ and /e/.'') They simply use as long as there is no opposition to .\footnote{See Thomason's Language Log post, ``Why I don't love the International Phonetic Alphabet'', at: http://itre.cis.upenn.edu/\textasciitilde{}myl/languagelog/archives/005287.html.) Making things even more problematic, there is an old typographic tradition that the double-story uses a single-story in italics.
-This leads to the unfortunate effect that in most well-designed fonts the italics of and use the same glyph.
-If this distinction has to be kept upright in italics, the only solution we can currently offer is to use `slanted' glyphs (i.e.~artificially italicised glyphs) instead of real italics (i.e.~special italics glyphs designed by a typographer).\footnote{For example, the widely used IPA font Doulos SIL (http://scripts.sil.org/cms/scripts/page.php?item\_id=DoulosSIL) does not have real italics.
-This leads some word-processing software, like Microsoft Word, to produce slanted glyphs instead.
-That particular combination of font and software application will thus lead to the desired effect.
-However, note that when the text is transferred to another font (i.e.~one that includes real italics) and/or to another software application (like Apple Pages, which does not perform slanting), then this visual appearance will be lost.
-In this case we are thus still in the pre-Unicode situation in which the choice of font and rendering software actually matters.
-The ideal solution from a linguistic point of view would be the introduction of a new IPA code point for a different kind of which explicitly specifies that it should still be rendered as a double-story character when italicized.
-After informal discussion with various Unicode players, our impression is that this highly restricted problem is not sufficiently urgent to introduce even more -like characters in Unicode (which already lead to much confusion, see Section 4, Pitfall 4).
-This is a clear situation in which the Unicode Consortium is not just thinking about linguists, but has a more wide-ranging practical view to consider.)
-
-Some other homoglyphs related to encoding IPA in the Unicode Standard are:
-\begin{itemize}
-	\item The uses of the apostrophe has led to long discussions on the Unicode Standard email list.
-An English keyboard inputs \textless{}`\textgreater{} APOSTROPHE at U+0027, although the `preferred' Unicode apostrophe is the \textless{}'\textgreater{} RIGHT SINGLE QUOTATION MARK at U+2019.
-Yet the glottal stop/glottalization/ejective marker is another completely different character, the MODIFIER LETTER APOSTROPHE at U+02BC, but unfortunately looks mostly highly similar to U+2019.
-	\item There is ambiguous encoding of IPA segments within the Unicode Standard.
-An example is the U+02C1 MODIFIER LETTER REVERSED GLOTTAL STOP vs the U+02E4 MODIFIER LETTER SMALL REVERSED GLOTTAL STOP .
-Both are denoted in the Unicode Standard as the `pharyngealized diacritic' and both appear in various resources representing phonetic data online.
-This is thus an example for which the Unicode Standard does not solve the linguistic standardization problem.
-	\item There is at least one case in which the character name assigned by the Unicode Consortium does not match the IPA's description: in the Unicode Standard at U+01C3 is labeled LATIN LETTER RETROFLEX CLICK, but in IPA is an alveolar or postalveolar click (not retroflex).
-This naming is probably best seen as a simple error in the Unicode Standard.
-Note that most linguists simply seem to use or <ou>) while ligatures are single characters (e.g.
-<ʧ> LATIN SMALL LETTER TESH DIGRAPH at U+02A7).
-Ligatures arose in the context of printing easier-to-read texts, and are included in the Unicode Standard for reasons of legacy encoding.
-However, their usage is discouraged by the Unicode core specification.
-Specifically related to IPA, various phonetic combinations of characters (typically affricates) are available as single code-points in the Unicode Standard, but are designated as ``ligatures'' or ``digraphs'' (confusingly both names appear interchangeably).
-Such glyphs might be used by software to produce a pleasing display, but they should not be hard-coded into the text itself.
-In the context of IPA, characters like the following ligatures should thus \emph{not} be used.
-Instead a combination of two characters is preferred:
-	\item <ʣ> LATIN SMALL LETTER DZ DIGRAPH at U+02A3 (use <d> + <z> instead) 
-	\item <ʧ> LATIN SMALL LETTER TESH DIGRAPH at U+02A7 (use <t> + <ʃ> instead) 
-	\item <ʩ> LATIN SMALL LETTER FENG DIGRAPH at U+02A9 (use <f> + <ŋ> instead) 
-\end{itemize}
-
-However, there are a few Unicode characters that are historically ligatures, but which are today considered as simple characters in the Unicode Standard and thus should be used when writing IPA, namely:
-\begin{itemize}
-	\item <ɮ> LATIN SMALL LETTER LEZH at U+026E 
-	\item <œ> LATIN SMALL LIGATURE OE at U+0153 
-	\item <ɶ> LATIN LETTER SMALL CAPITAL OE at U+0276 
-	\item <æ> LATIN SMALL LETTER AE at U+00E6 
-\end{itemize}
-
-\subsection{Pitfall: The IPA notion of diacritics is not the same as the Unicode Standard's notion of diacritics}\label{pitfall-the-ipa-notion-of-diacritics-is-not-the-same-as-the-unicode-standards-notion-of-diacritics}
-
-Another pitfall is diacritics.
-The problem is that the meaning of the term `diacritics' as used by the IPA is not the same as it used in the Unicode Standard.
-Specifically, diacritics in the IPA-sense are either so-called \textsc{Combining Diacritical Marks} or \textsc{Spacing Modifier Letters} in the Unicode Standard.
-Crucially, Combining Diacritical Marks are by definition combined with the character before them (to form so-called default grapheme clusters, see Section 3).
-In contrast, Spacing Modifier Letters are by definition \emph{not} combined into grapheme clusters with the preceding character, but simply treated as separate letters.
-In the context of the IPA, the following IPA-diacritics are actually Spacing Modifier Letters in the Unicode Standard:
-\begin{itemize}
-	\item Length marks, namely <ː> MODIFIER LETTER TRIANGULAR COLON at U+02D0 and <ˑ> MODIFIER LETTER HALF TRIANGULAR COLON at U+02D1. 
-	\item Tone letters, like <˥> MODIFIER LETTER EXTRA-HIGH TONE BAR at U+02E5, and others like this. 
-	\item Superscript letters, like <ʰ> MODIFIER LETTER SMALL H at U+02B0 or <ˤ> MODIFIER LETTER SMALL REVERSED GLOTTAL STOP at U+02E4, and others like this. 
-	\item <˞> MODIFIER LETTER RHOTIC HOOK at U+02DE. 
-\end{itemize}
-
-Although linguists might expect these characters to belong together with the character in front of them, at least for <ʰ> MODIFIER LETTER SMALL H at U+02B0 the Unicode Consortium's decision to treat it as a separate character is also linguistically correct, because according to the IPA it can be used both for aspiration (more precisely postaspiration following the base character) and preaspiration (preceding the base character).
-Note that there is a mechanism in Unicode to force separate characters to be combined (namely by using the ZERO WIDTH JOINER at U+200D), but this seems to be a rather impractical, and probably not enforceable solution to us.
-
-\subsection{Pitfall: Neither the IPA nor the Unicode Standard enforce a unique diacritic ordering}\label{pitfall-neither-the-ipa-nor-the-unicode-standard-enforce-a-unique-diacritic-ordering}
-
-Also related to diacritics is the question of ordering.
-To our knowledge, the International Phonetic Association does not specify a specific ordering for diacritics that combine with phonetic base symbols; this exercise is left to the reasoning of the transcriber.
-However, such marks have to be explicitly ordered if sequences of them are to be interoperable and compatible.
-An example is a labialized aspirated alveolar plosive: <tʷʰ>.
-There is nothing holding linguists back from using <tʰʷ> instead (with exactly the same intended meaning).
-However, from a technical standpoint, these two sequences are different, e.g.~if both sequences are used in a document, searching for <tʷʰ> will not find any instances of <tʰʷ>, and vice versa.
-Likewise, a creaky voiced syllabic dental nasal can be encoded in various orders, e.g.
-<n̪̰̩>, <n̩̰̪> or <n̩̪̰>.
-
-In accordance with the absence of any specification of ordering in the IPA, the Unicode Standard likewise does not propose any standard orderings.
-Both leave it to the user to be consistent.
-However, there is one aspect of ordering for which the Unicode Standard does present a canonical solution, which is uncontroversial from a linguistic perspective.
-Diacritics in the Unicode Standard (i.e.~Combining Diacritical Marks, see above) are classified in Canonical Combining Classes.
-In practice, the diacritics are distinguished by their position relative to the base character.\footnote{See http://unicode.org/reports/tr44/\#Canonical\_Combining\_Class\_Values for a detailed description.) When applying a Unicode normalization (NFC or NFD, see previous section), the diacritics in different positions are put in a specified order.
-This process therefore harmonizes the difference between different encodings, for example, of a midtone creaky voice vowel <ḛ̄>.
-This grapheme cluster can be encoded either as <e> + <̄> + <̰> or as <e> + <̰> + <̄> .
-To prevent this twofold encoding, the Unicode Standard specifies the second ordering as canonical (diacritics below before diacritics above).
-
-When encoding a string according to the Unicode Standard, it is possible to do this either using the NFC (``composition'') or NFD (``decomposition'') normalization.
-Decomposition implies that precomposed characters (like <á> LATIN SMALL LETTER A WITH ACUTE at U+00E1) will be split into its parts.
-This might sound preferable for a linguistic analysis, as the different diacritics are separated from the base characters.
-However, note that most attached elements like strokes (e.g.~in the <ɨ>), retroflex hooks (e.g.~in <ʐ>) or rhotic hooks (e.g.~in <ɝ>) will not be decomposed, but strangely enough a cedilla (like in <ç>) will be decomposed.
-In general, Unicode decomposition does not behave like a feature decomposition as expected from a linguistic perspective.
-It is thus important to consider Unicode decomposition only as a technical procedure, and not assume that it is linguistically sensible.
-
-Facing the problem of specifying a consistent ordering of diacritics while developing a large database of phonological inventories from the world's languages, Moran (2012: 540) defines a set of diacritic ordering conventions.\footnote{The most recent version of these conventions is online: http://phoible.github.io/conventions/) The conventions are influenced by the linguistic literature, though some ad-hoc decisions had to be taken.
-The goal was to explicitly define all character sequences so that the vast variety of phonemes found in descriptions of the world's language were normalized into consistent character sequences, e.g.~if one language description uses and another , when both are intended to be phonetically equivalent, then a decision to normalize to one form was taken.
-For example, when a character sequence contains more than one character in Spacing Modifier Letters, the order that is proposed is the following (where <\textbar{}> indicates ``or'' and <→> indicates ``precedes''):
-\begin{itemize}
-	\itemsep1pt\parskip0pt\parsep0pt 
-	\item \textsc{Spacing Modifier Letters ordering:} ( unreleased <̚> \textbar{} lateral release <ˡ> \textbar{} nasal release <ⁿ>) → ( palatalized <ʲ>) → ( labialized <ʷ>) → ( velarized <ˠ>) → ( pharyngealized <ˤ>) → ( aspirated <ʰ> \textbar{} ejective <ʼ>) → ( long <ː> \textbar{} half-long <ˑ>)
-\end{itemize}
-
-If a character sequence contains more than one diacritic below the base character, then the place feature is applied first (dental, laminal, apical, fronted, backed, lowered, raised), then the laryngeal setting (voiced, voiceless, creaky voice, breathy voice) and finally the syllabic or non-syllabic marker (for vowels, ATR gets put on between the place and laryngeal setting).
-So:
-\begin{itemize}
-	\itemsep1pt\parskip0pt\parsep0pt 
-	\item \textsc{Combining Diacritical Marks (below) ordering:} ( dental <t̪> \textbar{} laminal <t̻> \textbar{} apical <t̺>) → ( fronted <u̟> \textbar{} backed <e̠>) →( lowered <e̞> \textbar{} raised <e̝>) → ( ATR <e̘ e̙>) → ( voiced <s̬> \textbar{} voiceless <n̥> \textbar{} creaky voice <b̰> \textbar{} breathy voice <b̤>) → ( syllabic <n̩> \textbar{} non-syllabic <e̯>)
-\end{itemize}
-
-Character sequences with diacritics above the base character were not problematic in Moran 2012 because they include only the centralized, mid-centralized and nasalized combining characters.
-Moran (2012) marks tones as singletons with Space Modifier Letters, e.g.
-\textless{}˦\textgreater{} for a phonemic high tone, instead of accent diacritics, alleviating potential conflicts.
-Building on the work of Moran (2012), if a character sequence contains more than one diacritic above the base character, we propose:
-\begin{itemize}
-	\itemsep1pt\parskip0pt\parsep0pt 
-	\item \textsc{Combining Diacritical Marks (above) ordering:} (centralized <ë> \textbar{} mid-centralized <e̽>) → (extra short <ĕ>) →( tone accents, e.g.
-<è> ) → ( Spacing Modifier Letters ) → ( tone letters, e.g.
-<e˦>)
-\end{itemize}
-
-\section{6. Orthography profiles}\label{orthography-profiles}
-
-\subsection{Characterizing writing systems}\label{characterizing-writing-systems}
-
-At this point in the course of rapid ongoing developments, we are left with a situation in which the Unicode Standard offers a highly detailed and flexible approach to deal computationally with writing systems, but it has unfortunately not influenced the linguistic practice very much.
-In many practical situations, the Unicode Standard is far too complex for the day-to-day practice in linguistics because it does not offer practical solutions for the down-to-earth problems of many linguists.
-In this section, we propose some simple practical guidelines and methods to improve on this situation.
-
-Our central aims for linguistics, to be approached with a Unicode-based solution, are: (i) to improve the consistency of the encoding of sources, (ii) to transparently document knowledge about the writing system (including transliteration), and (iii) to do all of that in a way that is easy and quick to manage for many different sources with many different writing systems.
-The central concept in our proposal is the \textsc{orthography profile}, a simple tab-separated CSV text file, that characterizes and documents a writing system.
-We also offer basic implementations in Python and R to assist with the production of such files, and to apply orthography profiles for consistency testing, grapheme tokenization and transliteration.
-Not only can orthography profiles be helpful in the daily practice of linguistics, they also succinctly document the orthographic details of a specific source, and, as such, might fruitfully be published alongside sources (e.g.~in digital archives).
-Also, in high-level linguistic analyses in which the graphemic detail is of central importance (e.g.~phonotactic or comparative-historical studies), orthography profiles can transparently document the decisions that have been taken in the interpretation of the orthography in the sources used.
-
-Given these goals, Unicode locale descriptions (see Section 3) might seem like the ideal orthography profiles.
-However, there are various practical obstacles preventing the use of such locale descriptions in the daily linguistic practice, namely: (i) the XML-structure is too verbose to easily and quickly produce or correct manually, (ii) locale descriptions are designed for a wide scope on information (like date formats or names of weekdays) most of which is not applicable for documenting writing systems, and (iii) most crucially, even if someone made the effort to produce a technically correct locale description for a specific source at hand, then it is nigh impossible to deploy the description.
-This is because a locale description has to be submitted to and accepted by the Unicode Common Locale Data Repository.
-The repository is (rightly so) not interested in descriptions that only apply to a limited set of sources (e.g.~only a single specific dictionary).
-
-The major challenge then is developing an infrastructure to identify the elements that are individual graphemes in a source, specifically for the enormous variety of sources using some kind of alphabetic writing system.
-Authors of source documents (e.g.~dictionaries, wordlists, corpora) use a variety of writing systems that range from their own idiosyncratic transcriptions to already well-established practical or longstanding orthographies.
-Although the IPA is one practical choice as a sound-based normalization for writing systems (which can act as an interlingual pivot to attain interoperability across writing systems), graphemes in each writing system must also be identified and standardized if interoperability across different sources is to be achieved.
-In most cases, this amounts to more than simply mapping a grapheme to an IPA segment because graphemes must first be identified in context (e.g.~is the sequence one sound or two sounds or both?) and strings must be tokenized, which may include taking orthographic rules into account (e.g.
-between vowels is /n/ and after a vowel but before a consonant is a nasalized vowel /ṽ/).
-In our experience, data from each source must be individually tokenized into graphemes so that its orthographic structure is identified and its contents can be extracted.
-To extract data for analysis, a source-by-source approach is required before an orthography profile can be created.
-For example, almost each available lexicon on the world's languages is idiosyncratic in its orthography and thus requires lexicon-specific approaches to identify graphemes in the writing system and to map graphemes to phonemes, if desired.
-
-Thus, our key proposal for the characterization of a writing system is to use a grapheme tokenization as an inter-orthographic pivot.
-Basically, any source document is tokenized by graphemes, and only then a mapping to IPA (or any other orthographic conversion) is performed.
-An orthography profile then is a description of the units and rules that are needed to adequately model a graphemic tokenization for a language variety as described in a particular source document.
-An orthography profile summarizes the Unicode (tailored) graphemes and orthographic rules used to write a language (the details of the structure and assumptions of such a profile will be presented in the next section).
-
-As an example of graphemic tokenization, note the three different levels of technological and linguistic elements that interact in the hypothetical lexical form <tsʰṍ̰shi> :
-\begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item code points (10 text elements): t s ʰ o ̃ ̰ ´ s h i 
-	\item grapheme clusters (7 text elements): t s ʰ ṍ̰ s h i 
-	\item tailored grapheme clusters (4 text elements): tsʰ ṍ̰ sh i 
-\end{enumerate}
-
-In (1), the string <tsʰṍ̰shi> has been tokenized into ten Unicode code points (using NFD normalization), delimited here by space.
-Unicode tokenization is required because sequences of code points can differ in their visual and logical orders.
-For example, <õ̰> is ambiguous to whether it is the sequence of + <̰> + <̃> or + <̃> + <̰>.
-Although these two variants are visually homoglyphs, computationally they are different.
-Unicode normalization should be applied to this string to reorder the code points into a canonical order, allowing the data to be treated canonically equivalently for search and comparison.
-In (2), the Unicode code points have been logically normalized and visually organized into grapheme clusters, as specified by the Unicode Standard.
-The combining character sequence <õ̰> is normalized and visually grouped together.
-Note that , the MODIFIER LETTER SMALL H at U+02B0, is not grouped with .
-This is because it belongs to Spacing Modifier Letters category in the Unicode Standard.
-These characters are underspecified for the direction in which they modify a host character.
-For example, can indicate either pre- or post-aspiration (whereas the nasalization or creaky diacritic is defined in the Unicode Standard to apply to a specified base character).
-Finally, to arrive at the graphemic tokenization in (3), tailored grapheme clusters are needed (as for example specified in an orthography profile).
-For example, this orthography profile would specify that the sequence of characters , and form a single grapheme , and that and form a grapheme .
-The orthography profile could also specify orthographic rules, e.g.~when tokenization graphemes, in say English, the in the forms and should be treated as distinct sequences depending on their contexts.
-
-\subsection{Informal description of orthography profiles}\label{informal-description-of-orthography-profiles}
-
-An orthography profile describes the Unicode code points, characters, graphemes and orthographic rules in a writing system.
-An orthography profile is a language-specific (and often even resource-specific) description of the units and rules that are needed to adequately model a writing system.
-An important assumption of our work is that we assume a resource is encoded in Unicode (or has been converted to Unicode).
-Any data source that the Unicode Standard is unable to capture, will also not be captured by an orthography profile.
-
-Informally, an orthography profile specifies the graphemes (or, in Unicode parlance, ``tailored grapheme clusters'') that are expected to occur in any data to be analysed or checked for consistency.
-These graphemes are first identified throughout the whole data (a step which we call `tokenization'), and possibly simply returned as such, possibly including error messages about any parts of the data that are not specified by the orthography profile.
-Once the graphemes are identified, they might also be changed into other graphemes (a step which we call `transliteration').
-When a grapheme has different possible transliterations, then these differences should be separated by contextual specification, possibly down to listing individual exceptional cases.
-
-In practice, we foresee a workflow in which orthography profiles are iteratively refined, while at the same time inconsistencies and errors in the data to be tokenized are corrected.
-In some more complex use-cases there might even be a need for multiple different orthography profiles to be applied in sequence (see Section 8 on various exemplary use-cases).
-The result of any such workflow will normally be a cleaned dataset and an explicit description of the orthographic structure in the form of an orthography profile.
-Subsequently, the orthography profiles can be easily distributed in scholarly channels alongside the cleaned data, for example in supplementary material added to journal papers or in electronic archives.
-
-\subsection{Formal specification of orthography profiles}\label{formal-specification-of-orthography-profiles}
-
-The formal specifications of an orthography profile (or simply `profile' for short) are the following:
-\begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item \textsc{A profile is a} \textsc{Unicode UTF-8 encoded text file} (ideally using NFC, no-BOM, and LF; see Section 4, Pitfall ``File Formats'') that includes the information pertinent to the orthography.
-	\item \textsc{A profile is a} \textsc{tab-separated CSV file with an obligatory header line}.
-A minimal profile can have just a single column, in which case there will of course no tabs, but the first line will still be the header.
-For all columns we assume the name in the header of the CSV file to be crucial.
-The actual ordering of the columns is unimportant.
-	\item \textsc{Lines starting with a hash \textless{}\#\textgreater{} are ignored.} Comments and metadata can be included inside the file, but only as complete lines in the profile, to be marked by lines starting with hash ``\#'' (NUMBER SIGN at U+0023).
-Hashes somewhere else in the file are to be treated literally, i.e.~hashes are only to be ignored when they occur at the start of a line.\footnote{Comments that belong to specific lines will have to be put in a separate column of the CSV file, e.g.~add a column called `comments'.
-Further, if the content of a profile contains a hash at the start of a line, either reorder the columns so the hash does not occur at the start of the line, or add a dummy column in front of the data to not have the data start with a hash.)
-	\item \textsc{Metadata are given in commented lines at the beginning of the text file in a basic ``tag: value'' format.
-}Metadata about the orthographic description given in the orthography profile includes, minimally, (i) author, (ii) date, (iii) title, (iv) a stable language identifier encoded in BCP 47/ISO 639-3, and (v) bibliographic data for resource(s) that illustrate the orthography described in the profile.
-\end{enumerate}
-
-The content of a profile consists of lines, each describing a grapheme of the orthography, using the following columns:
-\begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item \textsc{A minimal profile consists of a single column with header `graphemes'}, listing each of the different graphemes in a separate line. 
-	\item \textsc{Optional columns called `left' and `right' can be used to specify the left and right context of the grapheme, respectively.} The same grapheme can occur multiple times with different contextual specifications, for example to distinguish different pronunciations depending on the context.
-	\item \textsc{The columns `grapheme', `left' and `right' can use regular expression metacharacters.} If regular expressions are used, then all literal usage of the special symbols, like full stops ``.'' or dollar signs ``\$'' (so-called ``metacharacters'') have to be explicitly escaped by adding a backslash before them (i.e.~use ``.'' or ``\$'').
-Note that any specification of context automatically expects regular expressions, so it is probably better to always escape all regular expression metacharacters when used literally in the orthography, i.e.~the following symbols will need to be preceded by a backslash: {[} {]} ( ) \{ \} ~+ * .
-- !
-?
-\^{} \$ .
-	\item \textsc{An optional column called `class' can be used to specify classes of graphemes}, for example to define a class of vowels.
-Users can simply add ad-hoc identifiers in this column to indicate a group of graphemes, which can then be used in the description of the graphemes or the context.
-The identifiers should of course be chosen such that they do not conflate with any symbols used in the orthography themselves.
-Note that such classes only refer to the graphemes, not to the context.
-	\item \textsc{Columns describing transliterations for each graphemes can be added and named at will}.
-Often more than a single possible transliteration will be of interest.
-Any software application using these profiles should use the names of these columns to select a specific transliteration column.
-	\item \textsc{Any other columns can be added freely, but will mostly be ignored by any software application using the profiles}.
-As orthography profiles are also intended to be read and interpreted by humans, it is often highly useful to add extra information on the graphemes in further columns, like for example Unicode codepoints, Unicode names, frequency of occurrence, examples of occurrence, explanation of the contextual restrictions, or comments.
-\end{enumerate}
-
-For the automatic processing of the profiles, the following technical standards will be expected:
-\begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item \textsc{Each line of a profile will be interpreted as a regular expression.
-}Software applications using profiles can also offer to interpret a profile in the literal sense to avoid the necessity for the user to escape regular expressions metacharacters in the profile.
-However, this only is possible when no contexts or classes are described, so this seems only useful in the most basic orthographies.
-	\item \textsc{The `class' column will be used to produce explicit ``or'' chains of regular expressions}, which will then be inserted in the `graphemes', `left' and `right' columns at the position indicated by the class-identifiers.
-For example, a class ``V'' as a context specification might be replaced by a regular expression like ``(a\textbar{}e\textbar{}i\textbar{}o\textbar{}u\textbar{}ei\textbar{}au)''.
-Only the graphemes themselves are included here, not any contexts specified for the elements of the class.
-	\item \textsc{The `left' and `right' contexts will be included into the regular expressions by using lookbehind and lookahead}.
-Basically, the actual regular expression syntax of lookbehind and lookahead is simply hidden to the users by allowing them to only specify the contexts themselves.
-Internally, the contexts in the columns ``left'' and ``right'' are combined with the column ``graphemes'' to form a complex regular expression like ``(?\textless{}=left)graphemes(?=right)''.
-	\item \textsc{The regular expressions will be applied in the order as specified in the profile, from top to bottom.} A software implementation can offer help in figuring out the optimal ordering of the regular expressions, but should then explicitly report on the order used.
-\end{enumerate}
-
-The actual implementation of the profile on some text-string will function as follows:
-\begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item \textsc{All graphemes are matched in the text before they are tokenized or transliterated}.
-In this way, there is no necessity for the user to consider `feeding' and `bleeding' situations, in which the application of a rule either changes the text so another rule suddenly applies (``feeding'') or prevents another rule to apply (`bleeding').
-	\item \textsc{The matching of the graphemes can occur either globally or linearly.
-}From a computer science perspective, the most natural way to match graphemes from a profile in some text is by walking linearly through the text-string from left to right, and at each position go through all graphemes in the profile to see which one matches, then go to the position at the end of the matched grapheme and start over.
-This is basically how a finite state transducer works, which is a well-established technique in computer science.
-However, from a linguistic point of view, our experience is that most linguists find it more natural to think from a global perspective.
-In this approach, the first grapheme in the profile is matched everywhere in the text-string first, before moving to the next grapheme in the profile.
-Theoretically, these approaches will lead to different results, though in practice of actual natural language orthographies they almost always lead to the same result.
-Still, we suggest that any software application using orthography profiles should offer both approaches (i.e.
-``global'' or ``linear'') to the user.
-The approach used should be documented in the metadata as ``tokenization method''.
-	\item \textsc{The matching of the graphemes can occur either in NFC or NFD.
-}By default, both the profile and the text-string to be tokenized should be treated as NFC (see section 4, Pitfall ``Canonical equivalence'' above).
-However, in some use-cases it turns out to be practical to treat both text and profile as NFD.
-This typically happens when very many different combinations of diacritics occur in the data.
-An NFD-profile can then be used to first check which individual diacritics are used, before turning to the more cumbersome inspection of all combinations.
-We suggest that any software application using orthography profiles should offer both approaches (i.e.
-``NFC'' or ``NFD'') to the user.
-The approach used should be documented in the metadata as ``unicode normalization''.
-	\item \textsc{The text-string is always returned in tokenized form} by separating the matched graphemes by a user-specified symbols-string.
-Any transliteration will be returned on top of the tokenization.
-	\item \textsc{Leftover characters (i.e.~characters that are not matched by the profile) should be reported to the user as errors.} Typically, the unmatched character are replaced in the tokenization by a user-specified symbol-string.
-\end{enumerate}
-
-Any software application offering to use orthography profile:
-\begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item \textsc{should offer user-options} to specify:
-	\begin{enumerate}
-		\def\labelenumii{\arabic{enumii}.} 
-		\item \textsc{the name of the column to be used for transliteration} (if any). 
-		\item \textsc{the symbol-string to be inserted between graphemes.} Optionally, a warning might be given if the chosen string includes characters from the orthography itself. 
-		\item \textsc{the symbol-string to be inserted for unmatched strings} in the tokenized and transliterated output. 
-		\item \textsc{the tokenization method}, i.e.~whether the tokenization should proceed ``global'' or ``linear''. 
-		\item \textsc{unicode normalization}, i.e.~whether the text-string and profile should use ``NFC'' or ``NFD''. 
-	\end{enumerate}
-	\item \textsc{might offer user-options }to:
-	\begin{enumerate}
-		\def\labelenumii{\arabic{enumii}.} \setcounter{enumii}{5} 
-		\item \textsc{assist in the ordering of the graphemes.} In our experience, it makes sense to apply larger graphemes before shorter graphemes, and to apply graphemes with context before graphemes without context.
-Further, frequently relevant rules might be applied after rarely relevant rules (though frequency is difficult to establish in practice, as it depends on the available data).
-Also, if this all fails to give any decisive ordering between rules, it seems useful to offer linguists the option to reverse the ordering from any manual specified ordering, because linguists tend to write the more general rule first, before turning to exceptions or special cases.
-		\item \textsc{assist in dealing with upper and lower case characters.} It seems practical to offer some basic case matching, so characters like <a> and <A> are treated equally.
-This will be useful in many concrete cases, although the user should be warned that case matching does not function universally in the same way across orthographies.
-Ideally, users should prepare orthography profiles with all lowercase and uppercase variants explicitly mentioned, so by default no case matching should be performed.
-		\item \textsc{treat the profile literal}, i.e.~to not interpret regular expression metacharacters.
-Matching graphemes literally often leads to strong speed increase, and would allow users to not needing to worry about escaping metacharacters.
-However, in our experience all actually interesting use-cases of orthography profiles include some contexts, which automatically prevents any literal interpretation, so by default the matching should not be literal.
-	\end{enumerate}
-	\item \textsc{should return the following information} to the user:
-	\begin{enumerate}
-		\def\labelenumii{\arabic{enumii}.} \setcounter{enumii}{8} 
-		\item \textsc{the original text-strings to be processed in the used Unicode normalization}, i.e.~in either NFC or NFD as specified by the user. 
-		\item \textsc{the tokenized strings}, with additionally any transliterated strings, if transliteration is requested. 
-		\item \textsc{a survey of all errors encountered}, ideally both in which text-strings any errors occurred and which characters in the text-strings lead to errors. 
-		\item \textsc{a reordered profile}, when any automatic reordering is offered 
-	\end{enumerate}
-\end{enumerate}
-
-\section{7. Use cases}\label{use-cases}
-
-We now present several use cases that have motivated the development of orthography profiles. These include:
-\begin{itemize}
-	\item grapheme tokenization 
-	\item normalization of orthographic systems to attain interoperability across different source documents, 
-	\item cognate identification for detecting the similarity between words from different languages 
-\end{itemize}
-
-An important assumption of our work is that input sources are encoded in Unicode (UTF-8 to be precise).
-Anything that the Unicode Standard is unable to capture, cannot be captured by an orthography profile.
-We also use Unicode normalization form NFD to decompose all incoming text input into normalized and (Unicode) logically ordered strings.
-This process is of fundamental importance when working with text data in the Unicode Standard.
-
-\emph{Grapheme tokenization}
-
-Once text has been normalized, an additional and straightforward process is Unicode grapheme tokenization.
-More recently this regular expression, often available as short cut ``\X'', identifies all sequences of ``base'' characters followed by 0 or more combining diacritics.
-For example, the sequence, <ŋ͡mṍraː> would tokenize as a space-delimited sequence as: <ŋ͡ m ṍ r a ː>.
-For a first pass at identifying grapheme clusters, this is a straightforward tokenization.
-However, the linguist will note that both the tie bar (COMBINING DOUBLE INVERTED BREVE, a Unicode Combining Modifier) and the length marker (MODIFIER LETTER TRIANGULAR COLON.
-a Unicode Letter Modifier) do not appear ``correctly'', i.e.~the tie bar is grouped with the first character in the sequence and the length marker appears singly.
-This is necessary, as defined by the Unicode Standard, because of these character's semantic properties -- with Unicode we cannot know if, for example, the MODIFIER LETTER TRIANGULAR COLON should appear before or after the base character that it modifies, cf.~the IPA aspiration marker, , the MODIFIER LETTER SMALL H, which is also a Unicode Letter Modifier and for linguists a diacritic that be be used for pre- or post-aspiration.
-These ambiguities of position in tokenization are not uncommon in IPA, thus orthographic (or source) normalization and tokenization is needed.
-
-*Cross-orthography **normalization*
-
-Orthographic tokenization requires a specification of the tailored grapheme clusters (linguists: graphemes) to separate orthographic units within an input source.
-In the simplest case, the Unicode graphemic tokenization with regular expression ``\X'' captures the grapheme clusters, i.e.~the input data is graphemically tokenized and there are no additional string sequences required to capture each grapheme, e.g.~no specification.
-This scenario is the most unlikely.
-
-The creation of an orthography typically requires that someone describes knowledge about the input data source's writing system.
-For example, when orthographically tokenizing a dictionary as input, the sequences of grapheme clusters need to be explicitly stated or a regular expression rule can be specified to transform the input.
-We illustrate these methods with examples from Thiesen 1998, a bilingual dictionary of Diccionario Bora-Castellano, and with Huber \& Huber 1992, a comparative vocabulary of sixty-nine indigenous languages of Columbia.
-
-Huber \& Huber 1992
-
-({[}a\textbar{}á\textbar{}e\textbar{}é\textbar{}i\textbar{}í\textbar{}o\textbar{}ó\textbar{}u\textbar{}ú{]})(n)(\s)({[}a\textbar{}á\textbar{}e\textbar{}é\textbar{}i\textbar{}í\textbar{}o\textbar{}ó\textbar{}u\textbar{}ú{]}), \1 \2 \4
-
-A preliminary orthography profile can be generated with code available online.\footnote{https://github.com/bambooforest/tokenizer) The Python script creates a unigram model for individual Unicode characters or Unicode grapheme clusters and their frequency in the source input.
-We find that this gives a broad overview of the input at the character and grapheme levels and it oftens finds typographic errors in the source input by identifying rarely occurring or one-off characters in the input.
-The models also provides the basis for the ``quick'' development of an initial orthography profile by saving the user keyboarding time.
-
-Note that these symbols are still phonetically unspecified, i.e.~symbol may or may not represent, say, a front open unrounded vowel.
-
-The orthography profile is designed so that specified strings of characters are captured for tokenization.
-
-\% how the tokenization works
-
-The graphemic cluster specifications in the orthography profile are read into a trie data structure, which is used to tokenize the incoming input source by identifying sentinels (tokenization boundaries) in the input stream, which we then separate grapheme clusters by spaces.
-
-An illustration is given in Figure N.
-
-Comparative wordlists often start as the compilation of different sources with different writing systems.
-The Dogon comparative wordlist is an example of a compilation of slightly divergent transcription practices, which we will use to illustrate our point and to orthographically normalize an input source by transforming different transcriptions into a single IPA-like transcription.
-
-*Orthographic normalization / cross-source **normalization*
-
-Once we have normalized and tokenized our text elements, we can apply various sorts of source normalization across resources to create interoperable and comparable data.
-
-Sound-based normalization is practical because \ldots{}
-
-We use orthography profiles in our work to describe cross-linguistic difference between writing systems.
-
-\emph{Cognate detection}
-
-The identification of cognates is a non-trivial task, which involves identifying the semantic and phonetic similarity of words.
-
-use grapheme correspondences to approximate cognacy and sound correspondences
-
-to identify cognates and sound changes
-
-Examples
-
-Orthographic similarity != Phonetic similarity != Regular sound change
-
-Cognate identification is further complicated by the fact that languages are written differently, and even more so for the low resource languages that we work with, linguists' phonetic transcriptions are heterogeneous and typically document-specific, i.e.~two linguists working on the same language variety most often develop different practical orthographies by using different symbols and different orthographic rules.
-Thus the essence of language documentation and description is diverse and ephemeral in nature.
-The encoding of linguistic data and linguistic data models is diverse and abstracting away from individual analytical preferences requires a document-by-document approach and a mechanism for standardization of writing to form an interlingual pivot so that different descriptions can be normalized into one encoding system to undertake analysis.
-
-\% Orthography profile implementation in Lingpy / generic tokenizer
-
-\% describe Tokenizer's function
-
-\% describe algorithm for tokenizing IPA (Unicode grapheme cluster tokenization and then alignment of Letter Modifiers and Tie Bars)
-
-\% insert an example of orthographic tokenization from PAD
-
-\section{move Letter Modifiers left}\label{move-letter-modifiers-left}
-
-(\s)(\p{Lm}), \2
-
-which catches:
-
-\section{Diacritics: Unicode letter modifiers (Lm)}\label{diacritics-unicode-letter-modifiers-lm}
-
-\section{({[}ː\textbar{}ʰ\textbar{}ʼ\textbar{}ʿ\textbar{}ʾ{]})}\label{ux2d0ux2b0ux2bcux2bfux2be}
-
-from the set of this:
-
-\section{Diacritics: all}\label{diacritics-all}
-
-\section{({[}ː\textbar{}ʰ\textbar{}ʼ\textbar{}ʿ\textbar{}ʾ\textbar{}̜\textbar{}̠\textbar{}̟\textbar{}͡\textbar{}̪\textbar{}̰\textbar{}̣\textbar{}̥\textbar{}̩\textbar{}‿\textbar{}.\textbar{}→\textbar{}˻\textbar{}+\textbar{}̃\textbar{}̫\textbar{}̤{]})}\label{ux2d0ux2b0ux2bcux2bfux2be.}
-
-and allows us to with one regular expression do this:
-
-PAD examples
-
-A side effect of character tokenization is identification of outliers, e.g.
-
-Some examples in Python
-
-Some examples in R
-
-{[}UNICODE USE-CASES{]}
-
-The most basic overall Unicode character property is the \textsc{General Category}, which categorizes Unicode characters into: \emph{Letters, Punctuation, Symbols, Marks, Numbers, Separators, }and* Other*.
-Each Unicode character property also has a character property value.
-So each General Category is denoted by a property value by a single letter abbreviation, e.g.
-``L'' for Letter.
-Each General Category property also has subcategories that are also properties with property values.
-For example, ``Letters'' is broken down into Uppercase Letter (Lu), Lowercase Letter (Ll), Titlecase Letter (Lt), Modifier Letter (Lm) and Other Letter (Lo).\footnote{The subcategory ``Other Letter'' includes characters such as uncased letters as in the Chinese writing system.) By defining character properties for each code point, algorithmic implementations such as text processing applications or regular expression engines that are conformant to the Unicode Standard can be used to recognize whole (sub)categories of characters or ranges of characters, e.g.~to match proper names in English in title case or more generally to match character types, such as scripts, or symbols and integers in email addresses, phone numbers, etc.
-
-Script information is useful for algorithmic processes, such as collation, searching and for multilingual text processing.
-For example, if a regular expression engine includes functionality to match characters at the level of script, then it provides the ability to identify characters, e.g.
-\match{Greek}, which tests whether letters in a text belong to the Greek script or not.
-This functionality can be used to determine the boundaries between different writing systems that appear in the same document.
-In Section 6 (Use cases of grapheme tokenization) we provide an example of a regular expression match at the level of character property.
-
-Unicode-aware regular expression engines may contain a meta-character, ``\X'', that matches this level of grapheme cluster, i.e.~a base character followed by some number of accents or diacritical marks.
-Grapheme clusters are important to determine word or line boundaries, for collation (`ordering') and for user interface interaction (e.g.~mouse selection, backspacing, cursor movement).
-
-\section{9. Conclusion}\label{conclusion}
-
-In our work we focus on the writing systems used in lesser-described and endangered languages and our orthography profiles are used to describe writing systems and to transpose them into phonetic transcription.
-Of course the International Phonetic Alphabet and other transcription systems are essentially just orthographies that are an approximation of sound.
-Nevertheless, sound-based normalization is practical in automatically identifying cognates and sound changes with quantitative methods.
-We leverage orthography profiles to enable comparative analysis of languages with different writing systems.
-
-An orthography profile lists the graphemes in a particular description of language data, e.g.~a wordlist, dictionary or corpus.
-Building on the knowledge that can be extracted from that description by tokenizing words by the graphemes made explicit in the orthography profile, it is straightforward to undertake other analyses of the data.
-For example, various ngram models of the data can be extracted with a few lines of code.
-A unigram model with counts, frequencies and positive log probability (``plogs''; ``entropy under this model'') (Goldsmith \& Riggle, 2010) provides a fair amount of information about a given data source.
-Essentially, the orthography profile provides the description that allows this information to be calculated based on the mapping of sequences of characters into graphemic units.
-Bigram models are also straightforwardly extracted.
-In the case of bigrams, mutual information can be captured and used in various other statistical analyses, such as quantitative language comparison, inferring phylogenetic trees, etc.
-
-\section{10. Nice text blocks to be included somewhere}\label{nice-text-blocks-to-be-included-somewhere}
-
-Looking at the variety of writing systems used for the world's languages, there are various challenges for the computational processing of resources in which such writing systems are used.
-
-A widespread mechanism to derive new symbols for a writing system is to use a combination of multiple symbols already available.
-That can lead to completely new symbols, like <æ> or <w>, which are still rather transparently derived from <a e> and <v v>, respectively.
-More typically, many orthographies use sequences of symbols as special characters, like <sch> or <gh>.
-To identify such `grapheme clusters' is one of the central problems to be tackled with orthography profiles.
-A further frequently used mechanism is to add extra material to an already available symbol, either to make it better visible (this is probably the reason for the addition of a dot on top of the <i>) or to derive new symbols.
-The widespread use of diacritics is the obvious result of such processes (Gaultney 2002).
-However, also the derivation of new symbols by added extra strokes is widespread (and not only in Far-Eastern character-based writing systems), like for example <ŋ> or <ç>.
-
-The use of punctuation (Daniëls 1994, Humez \& Humez 2008) and spaces (Saenger 1997) to improve readability. {[}e.g.~space after punctuation, but not before{]}
-
-Currently, the Unicode Standard is the standard character encoding for linguistic data (Anderson 2003).
-However, there are still hundreds of different encoding systems that were invented independently to capture orthographic diversity as different nations adopted and developed computer systems.
-These different encoding systems are problematic and in conflict with one another because different standards were formalized differently and for different purposes by different standards committees in different countries.
-No unified encoding scheme contained enough code points to encode all characters, so two different encoding schemes possibly used the same code point for different characters, or used different code points to represent the same character.
-Because computers support multiple character encoding schemes, data risked being corrupted when handled by different applications and encodings.
-The Unicode Standard was devised to alleviate these problems.
-
-For orthographies, identifying graphemes can be even more challenging than identifying phones and phonemes in phonetic transcription because although transcriptions may not adhere strictly to IPA, they tend to have straightforward mappings between sounds and symbols.
-On the other hand, orthographies can introduce orthographic rules, which add an additional challenges in identifying graphemes in words, as mentioned above.
-Thus for resources not in IPA or IPA-like transcriptions, graphemes often must be manually identified, whether they are encoded as singletons or multi-character sequences.
-The identification of graphemes and the formulation of orthographic rules are used to create an orthography profile.
-
-Character-to-glyph mappings are particularly not clear-cut for linguists using the IPA.\footnote{We give more linguistic examples in Section 5.) Just look in the IPA Extensions block.\footnote{Character ``blocks'' are described in Section 3 (The Unicode approach).) There are several characters that are glyphs (variants) of the same grapheme in the Latin block, e.g.
-and \& and .\footnote{LATIN SMALL LETTER ALPHA U+0251 vs LATIN SMALL LETTER A U+0061 , and LATIN SMALL LETTER SCRIPT G U+0261 vs LATIN SMALL LETTER G U+0067 .) Other characters like
-
-, , do not appear at all in the IPA Extensions block, but as separate phonetic symbols; they are already encoded in the Basic Latin, Latin Extended-A and Greek \& Coptic blocks.
-When a linguist transcribes an IPA
-
-on a QWERTY keyboard, if it is from the Basic Latin group, it is valid Unicode IPA (it is not valid IPA if it is from the Cyrillic Block).
-However, keyboard and and because they have been given specific code points in the IPA Extensions block, where they have been given special status as IPA characters, and which most often go unnoticed by linguists.\footnote{In Section 5 we discuss in detail the Unicode IPA alphabet and its characters dispersal across different Unicode Standards scripts and blocks.)
-
-The Unicode Standard divides text elements roughly into two classes of graphemes.
-The cover term \textsc{grapheme cluster}\footnote{Formerly referred to as ``locale-independent graphemes'' to emphasize that the term grapheme is used differently in linguistics.) is used to emphasize the difference between the term \emph{grapheme }as used by linguists and what the users perceive as a character (a basic unit of a writing system for a language), which may not be a single Unicode character.\footnote{http://www.unicode.org/reports/tr29/) The Unicode Consortium makes a further distinction between grapheme clusters and tailored grapheme clusters.
-
-The simplest example of a grapheme cluster is essentially a base character followed by a Letter Modifier character.
-For example, the sequence + \textless{}\textasciitilde{}\textgreater{} (COMBINING TILDE at U+0303) combines visually into , a user-percieved character in writing systems like that of Spanish.\footnote{Although the code point sequence can also be represented with a single code point, LATIN SMALL LETTER N WITH TILDE at U+00F1, it is not the case that all multi-code point character sequences have a single code point equivalent, which can be deduced from normalization.
-Often for commonly-used letters in European languages, a single code point form exists.)
-
-What the user perceives as a single character may actually involved a multi-code point sequence.\footnote{Unicode-aware regular expression engines may contain a meta-character, ``\X'', that matches this level of grapheme cluster, i.e.~a base character followed by some number of accents or diacritical marks.
-As mentioned, grapheme clusters are important for collation algorithms, but they are also used to determine other processes such as word or line boundaries and for user interface interaction (e.g.~mouse selection, backspacing, cursor movement).)
-
-More challenging for text processing is what the Unicode Consortium terms tailored grapheme clusters.
-For example, the sequence + for the Slovak digraph or the sequence .
-These grapheme clusters are ``tailored'' in the sense that they must be specified on a language-by-language basis.
-Algorithms can be adapted to use as input these tailored language-specific orthographies for transliteration, localization, etc.
-In fact, the Unicode Standard provides a method for defining language-specific tailored writing systems called the Unicode Locale Description.
-Unicode Locale Descriptions allow users to define a set of extended grapheme clusters.
-Unicode Locale Descriptions are saved in the Common Locale Data Repository (CLDR),\footnote{http://cldr.unicode.org/) a repository of language-specific definitions of writing system properties, each of which describes specific usages of characters.
-Each can be encoded in a Unicode Locale Data Markup Language (LDML) document.
-LDML is an XML format and vocabulary for the exchange of structured locale data.
-It is a format used not only for CLDR, but also for general interchange of locale data, such as in Microsoft's .NET.
-We return to LDML in Section 5 (Orthography profiles and locale descriptions).
-
-We have noted that the rendering of electronic text requires that characters are encoded and stored in a computer's memory.
-These character sequences are then mapped from character code points to glyphs by software that takes into account a character's context.
-Transparent to most users is the fact that different logical orders of characters may result in the same visual output.
-For example, phonetic characters with certain combinations of diacritics may be homoglyphs (see below), even while the logical order of their character sequences are not the same.\footnote{One example is a vowel that is both nasalized and creaky voice, e.g.
-{[}ḛ̃{]}.
-See discussion below.) Thus some type of standardization, what the Unicode Consortium calls ``normalization'' of the ordering of (some classes of) characters, is required to make sure that all data are logically consistent and therefore easily comparable and equally searchable.
-The aim of standardization is to attain syntactic \textsc{interoperability} of data.\footnote{By interoperability we mean the ability to ubiquitously exchange and merge disparate (textual) data sets, and data encoding formats, to facilitate data sharing and to ``effortlessly'' undertake comparison and analysis.
-For a discussion of syntactic versus semantic interoperability, see Ide and Pustejovsky, 2010.)
-
-The IPA is often used by linguists in undertaking initial phonetic and phonemic analyses of languages.
-However, once the distinctive sounds of a given language have been identified (and a phonological description of the target language has been described), a linguist may continue his or her documentation and description of higher-order structures like words and sentences with a broad phoneme-based alphabet.
-In fact, literary materials development often leapfrogs from a phonemic alphabet to locale-specific practical orthographies, in which numerous factors affect a language's orthography development, e.g.~politics, purpose, speaker communities familiarity with other writing systems, technological limitations.
-
-One simple example is illustrated with two genealogically related languages spoken in Northern Ghana: Western Sisaala\footnote{Western Sisaala lacks /kp/.) and Sisaala Tumulung.\footnote{Sisaala Tumulung (ISO 639-3: sil) is spoken around the city Tumu, which is about 50 kilometers geographically to the east of Western Sisaala (ssl) speaking areas.) Both languages contain the same basic set of contrastive sounds, but their orthographies encode affricates (and vowels) differently graphemically; the influence stem from orthographies of government-sponsored literacy languages, which are taught to children in schools in these areas.
-In Western Sisaala the sound /tʃ/ is written and /dʒ/ as due to familiarity of these graphemes by students learning Dagaare (Moran 2006).
-Spoken further to the east, Sisaala Tumuling encodes the sound /tʃ/ as and /dʒ/ as (Blass 1975).
-
-If encoded in the Unicode Standard, the design of both orthographies has to specify that certain sequences of characters, such as + , are single graphemes (compare English and which is ambiguous).
-Design is important, for example, for collation ordering (sorting), i.e.~the sequence of characters that is treated as a unit for ordering purposes, such as in dictionaries.
-Indeed, the characters , and are three unique graphemes consisting of two Unicode characters from the Latin script, but developers of a software program that recognizes the Western Sisaala alphabet might want to specify the sort order , , \ldots{} .
-Thus there is a need for orthography-specific specifications that define sequences of characters as graphemes, whether technologically, linguistically, or both, for multilingual text processing.
-
-A major challenge that the developers of the Unicode Standard faced is that there exists no universal set of fundamental units of text because the divisions of text into elements, such as graphemes, is dependent on language-specific writing systems and orthographies, as just shown.
-Moreover, the division of electronic text into elements is process dependent, e.g.~sorting and spell checking may or may not need to take capitalization into account, depending on the orthography in question.
-Thus the encoding and processing of text requires well defined \textsc{text elements} and rules for interpreting them; these elements and rules are language-specific when we talk about major languages or well established orthographies, but more often in our research we face input from lesser-studied languages documented in author-specific transcription or practical orthography.\footnote{And this is usually down to the document-specific level, the description of which has been coined ``doculect'' for purposes of discussing differences of opinion in language descriptions (Cysouw and Good, 2013).)
-
-In Section XXX we noted that the Unicode Consortium makes a crucial distinction between characters and glyphs.
-The relations between characters and glyphs is a complex one and it is typically hidden from the user.
-Technologically, we must distinguish between characters and glyphs because:
-\begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item The logical order of a sequence of characters (in computer memory) may not be the same as the visual order of their (rendered) glyphs. 
-	\item There is not always a one-to-one mapping between characters and glyphs. 
-\end{enumerate}
-
-Regarding the first point, a user cannot assume that contiguously rendered text on-screen actually reflects the computer's internal storage order of individual and sequential characters in memory, i.e.~the \textsc{logical order}.
-In other words, in the Unicode Standard the visual order of glyphs may not be the same as the logical order of their digitally encoded characters; contiguous display is not indicative of contiguous text.* *A nice example is Hebrew text.\footnote{http://www.macchiato.com/unicode/globalization-gotchas) Another example is the glyph <ṍ>, which may consist of one, two or three characters in encoded in various physical orders.
-The possible mismatch in sequences becomes clearer when we consider the second point.
-
-When a base character, i.e.~any graphic character not in the General Category of Combining Mark, is followed by one or more characters having the property General Category of Combining Mark, they can be combined visually into one or more glyphs using the Canonical Ordering Algorithm to create Unicode \textsc{grapheme clusters}.
-As we have shown, this input may be ambiguous from a technological perspective if normalization is not taken into account.
-For example, a linguist may insert a single character or a character sequence for to denote a high back unrounded vowel with high tone:
-\begin{itemize}
-	\item LATIN SMALL LETTER U WITH ACUTE (U+00FA) 
-	\item LATIN SMALL LETTER U (U+0075) + COMBINING ACUTE ACCENT (U+0301) 
-\end{itemize}
-
-The normalization algorithms in the Unicode Standard are used to make these character sequences canonically equivalent for purposes such as ubiquitous search and comparison.
-
-However, the standard does not specify many (or most) language-specific graphemes, or in Unicode parlance, \textsc{tailored grapheme clusters}, such as English .\footnote{Some (unfortunate) exceptions included for legacy purposes, and used incorrectly by linguists, include digraphs like LATIN SMALL LETTER TS DIGRAPH at U+02A6 or LATIN SMALL LETTER DEZH DIGRAPH at U+02A4.
-These and other digraphs are encoded in the Unicode Standard's IPA Extensions block.) For technological reasons, these language-specific tailored grapheme clusters must be defined outside of the Unicode Standard, in what are called (language-specific) Unicode Locale Descriptions.\footnote{http://www.unicode.org/reports/tr35/) The encoding of these multiple character sequences in Locale Descriptions is beyond the need (or technical ability) of many linguists; thus we have developed a simpler approach which we used in our own research and describe in detailed in Section XXX.
-Basically, the description of writing systems for lower resource languages, and in particular for document-specific descriptions (e.g.~different descriptions of the same language with different orthographies), requires an orthographic description at the (tailored) grapheme (cluster) level.
-
-An orthography profile is defined as a simple delimited flat text file encoded in the UTF-8, with character sequences in Unicode Normalization Form D.
-Metadata aside, the column delimited contents of an orthography profile are essential a state-table, i.e.~an implementation of a state transition table, but for ease of use encoded in a simple flat text file.
-
-In an orthography profile, all character strings are converted into NFD because each character in a grapheme may have phonetic value, and by using NFD, all graphemes are decomposed into a standardized order.
-For example, a vowel that is both nasalized and creaky looks like in IPA, where is some vowel.
-Although visually the same, a nasalized and creaky vowel can be composed of several different character sequences, as illustrated with in 1-3:
-\begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item <õ̰> LATIN SMALL LETTER O + COMBINING TILDE + COMBINING TILDE BELOW: U+006F + U+0303 + U+0330 
-	\item <õ̰> LATIN SMALL LETTER O + COMBINING TILDE BELOW + COMBINING TILDE: U+006F + U+0330 + U+0303 
-	\item <õ̰> LATIN SMALL LETTER O WITH TILDE + COMBINING TILDE BELOW: U+00F5 + U+0330 
-\end{enumerate}
-
-Applying NFD to these three character sequences results in one standard sequence; in this example the character sequence is that shown 1.
-NFD makes different sequences of input interoperable and it retains all of the phonetic information captured by the separate characters that combine to form a vowel with nasalization and creaky voice phonation.
-This is an important step to consider in linguistic research because each character has phonetic value, i.e.~the combining tilde denotes that the base character is produce by a lowering of the velum, which allows air to escape through the nose when the sound is produced.
-The tilde below also has meaning: it represents a phonation type in which the vocal cords are vibrating, but they are also being held slightly further apart, allowing more air to escape, which produces a murmured sound.
-To enable comparative analyses of languages at a deep phonetic level, retaining a code point for each character of phonetic value is deciduous.
-For example, if the sound systems of two languages are to be compared, one would not want to lose a possibly contrastive phonological feature like nasalization, which may have been collapsed into a single character in its logical representation.
-
-Our approach is similar in nature to the LDML, but we use a simpler state-table method, instead of generating an XML format for the ``exchange of structured locale data''.
-An orthography profile is a specification of the tailored grapheme clusters in a particular source document and any orthographic re-write rules needed after the application of Unicode normalization and orthographic profile tokenization.
-Orthography profiles can of course be used for transliteration between scripts, e.g.~from practical orthography to IPA and vice versa, by specifying mapping (replacements) from the list of tailored grapheme clusters in the source document and a set of replacement characters.
-This will be made clear in the next section in which we provide some use cases.
-
-To summarize, LDML is basically overkill for our intentions of identifying and delimiting grapheme boundaries (or in Unicode parlance ``tailored grapheme clusters'').
-LDML is appropriate, even necessary, for complex localization complexes for software applications -- localization of, say, an operating system or software application across cultures may require the instantiation of different date stamps an other local standards.
-For linguistic purposes of source document data tokenization, we therefore develop a simple-to-implement process of ``orthographic'' tokenization, using standard Unicode Standard-specified processes (e.g.~normalization and grapheme cluster tokenization) with an additional level of linguistic ``tailored grapheme cluster'' specification.
-
-\textsc{REFERENCES}
-
-Nancy Ide and James Pustejovsky.
-2010.
-What does inter- operability mean, anyway?
-toward an operational def- inition of interoperability for language technology.
-In Proceedings of the Second International Conference on Global Interoperability for Language Resources.
-Hong Kong, China.
-
-Huber, Randal Q. and Robert B. Reed (compilers).
-\begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} \setcounter{enumi}{1991} 
-	\itemsep1pt\parskip0pt\parsep0pt 
-	\item Vocabulario comparativo: Palabras selectas de lenguas indígenas de Colombia (Comparative vocabulary: Selected words in indigenous languages of Colombia).
-Bogota: Instituto Lingüísti- co de Verano.
-\end{enumerate}
-
-Thiesen, W.
-(1998).
-Diccionario bora-castellano, castellano-bora (1st ed.).
-Lima Perú: Instituto Linguístico de Verano.Retrieved from: http://www.worldcat.org/title/diccionario-bora-castellano-castellano-bora/oclc/40505215\&referer=brief\_results
-
-\textsc{APPENDIX - ORTHOGRAPHY PROFILE EXAMPLE}
-
-\textsc{APPENDIX - UNICODE IPA}
-
-Provide a clear and full Unicode IPA chart with additional information as an appendix
-\begin{verbatim}
-	- create this chart
-
-	- perhaps we should spell out conformance rules with regard to this chart? i.e. if a source claims conformance to our Unicode IPA chart, it should:
-
-	- identify the sections (or range) of the table that the data conforms to
-
-	- the source is in accordance with not only the Unicode IPA characters, but their intrinsic meanings (e.g.
-an Africanist might use <y> for /j/, but this level of conformance requires that <j> symbolizes /j/, etc.)
-
-	- someone who wants to use an Africanist transcription system can declare that they use the Unicode IPA chart, except for a range of characters (but they provide the mapping to Unicode IPA chart) 
-\end{verbatim}
-
-\textsc{Notes}
-
-\url{http://www.unicode.org/reports/tr29/}
-
-The default boundary determination mechanism specified in this annex provides a straightforward and efficient way to determine some of the most significant boundaries in text: user-perceived characters, words, and sentences.
-Boundaries used in line breaking (also called \emph{word wrapping}) are defined in {[}\href{http://www.unicode.org/reports/tr41/tr41-11.html\#UAX14}{UAX14}{]}.
-
-\textsc{Nice quotes}
-
-vanZundert: ``How does one design and plan a general unified infrastructure for a target that is all over the place?
-A `one size fits all' approach would be a disastrous underestimation of the specific needs of humanities research.
-The essence of humanities research is in its diverse, heterogeneous and ephemeral nature.
-But the need of big infrastructures to be designed well before implementation and use, means that every single design decision closes the resulting system to some category of users.
-Anything specific can potentially break the data model and the defined workflow.
-Of course big infrastructure architects also know this; to compensate, they look for possibilities to make more abstract models and malleable workflows.
-But abstract away far enough and the model becomes rather meaningless.
-If the design recommendation of big infrastructure is that we will use XML, what was the use of all that time and effort?
-Big infrastructure design seems to routinely underestimate this problem.
-To give but one simple example: the token seems such a logical, granular, easy to automate conceptual catch-all for text models.
-Can there possibly be a more generalizable model for text than as a series of tokens?
-You may think so, until someone asks how you handle languages that do not use spaces, or how you account for prefixes and suffixes, let alone ambiguous functions.
-Regarding text as a series of tokens denies in any case several other essential humanistic aspects of text: its typography, its materiality, its referentiality to name but a few.
-At second glance, from the humanities research perspective, there is pretty little actually generalizable about tokenization.''
-
-\% Ladefoged, Peter.
-1990.
-Some Reflections on the IPA.
-In UCLA Working Papers in Phonetics, 74.
-``The most interesting ambiguities in the Association's theoretical concepts concern what the symbols are symbolizing.''
-
-There are several ambiguous aspects of the IPA, but the most interesting concerns the theoretical concepts that its symbols are symbolizing \cite[64]{Ladefoged1990a}.
-Although the IPA is \texttt{intended to be a set of symbols for representing all the possible sounds of the world\textquotesingle{}s languages\textquotesingle{}\textquotesingle{} \textbackslash{}citep\{IPA2005\}, Ladefoged points out that the IPA also states that IPA symbols are}\textit{not} symbols for phones they are simply shorthand for what a phonologist would regard as a bundle of features'' \cite{Ladefoged1990a}.
-From the IPA:
-\begin{quote}
-
-	``The representation of these sounds uses a set of phonetic categories which describe how each sound is made.
-These categories define classes of sounds that operate in phonological rules and historical sounds changes.
-The symbols of the IPA are shorthand ways of indicating certain intersections of these categories.
-Thus [p] is a shorthand way of designating the intersection of the categories voiceless, bilabial, and plosive; [m] is the intersection of the categories voiced, bilabial, and nasal; and so on.''
-\end{quote}
-
-\noindent Thus, the ``association is deliberately not explicit about what is meant by a phonological contrast'' \cite[65]{Ladefoged1990a}.

### 2015-05-29 chapters/use_cases.tex
initial commit paper

--- 2015-05-29 chapters/originaltext.tex
@@
-The Unicode Cookbook for Linguists Managing orthographies in multilingual datasets using orthography profiles
-
-Steven Moran \& Michael Cysouw
-
-\textsc{Abstract}
-
-In this paper, we define the notion of an orthography profile, which is a language-specific (or even resource-specific) description of the units and rules that are needed to adequately model a writing system.
-An orthography profile describes the Unicode code points, characters, graphemes and orthographic rules as used in a writing system.
-Our orthography profiles are used to describe writing systems and possibly even to transliterate them into another (for example IPA-like) transcriptions with the goal to harmonize orthographies across languages and resources on an ad-hoc basis.
-The central aspect of our approach is that we introduce an intermediate step of graphemic tokenization, which is closer to linguistic practice and easier to manage than transducer-based transliteration.
-We leverage orthography profiles to enable comparative analysis of languages with different writing systems and to perform quality control on datasets with highly specific orthographic systems.
-We also describe how our proposals relate to the current Unicode specification, and introduce some of the more delicate aspects of the Unicode Standard for a linguistic audience.
-
-\textsc{Contents}
-
-{[}{[}TOC{]}{]}
-
-\section{1. Introduction}\label{introduction}
-
-Writing systems arise and develop in a complex mixture of cultural, technological and practical pressures.
-They tend to be highly conservative, in that people who have learned to read and write in a specific way (however impractical or tedious) are mostly unwilling to change their habits, e.g.~they tend to resist spelling reforms.
-In all literate societies there exists a strong socio-political mainstream that tries to force unification of writing (for example by strongly enforcing `right' from `wrong' writing in schools).
-However, there is also a large community of users who take as many liberties in their writing as they can get away with.
-
-For example, the writing of tone diacritics in Yoruba is often proclaimed to be the `right' way to write, although many users of Yoruba writing seem to be perfectly fine with leaving them out.
-As pointed out by the proponents of the `official' rules, there are some homographs when leaving out the tone diacritics (Olúmúyìw 2013: 44).
-However, writing systems (and the languages they represent) are normally full of homophones, which is normally not a problem at all for speakers of the language.
-More importantly, writing is not just a purely functional tool, but just as importantly it is a mechanism to signal social affiliation.
-By showing that you `know the rules' of expressing yourself in writing, others will more easily accept you as a worthy participant in their group.
-And that just as well holds for obeying to the `official' rules when writing a job application, as for obeying to the `informal' rules when writing an SMS to classmates in school.
-The case of Yoruba writing is an exemplary case, as even after more than a century of efforts to standardize the writing systems, there is still a wide range of variation in daily use (Olúmúyìw 2013).
-
-The sometimes cumbersome and sometimes illogical structure, and the enormous variability of existing writing systems is a fact of life scholars have to accept and should try to adapt to as good as possible.
-Our plea here is a proposal for a formalization to do exactly that.
-
-When considering the worldwide linguistic diversity, including all lesser-studied and endangered languages, there exist numerous different orthographies using symbols from the same scripts.
-For example, there are hundreds of orthographies using Latin-based alphabetic scripts.
-All of these orthographies use the same symbols, but these symbols differ in meaning and usage throughout the various orthographies.
-To be able to computationally use and compare different orthographies, we need a way to specify all orthographic idiosyncrasies in a computer-readable format (a process called `tailoring' in Unicode parlance).
-We call such specifications \textsc{orthography profiles}.
-Ideally, these specifications have to be integrated into so-called Unicode locale descriptions, though we will argue that in practice this is often not the most useful solution for the kind of problems arising in the daily practice of linguistics.
-Consequently, a central goal of this paper is to flesh out the linguistic challenges for locale descriptions, and work out suggestions to improve their structure for usage in a linguistic context.
-Conversely, we also aim to improve linguists' understanding and appreciation for the accomplishments of the Unicode Consortium in the development of the Unicode Standard.
-
-The necessity to computationally use and compare different orthographies most forcefully arises in the context of language comparison.
-Concretely, in our current research our goal is to develop quantitative methods for language comparison and historical analysis in order to investigate worldwide linguistic variation and to model the historical and areal processes that underlie linguistic diversity (cf.~Steiner et al.
-2010, List 2012a, 2012b, List \& Moran 2013, Moran \& Prokic 2013).
-In this work, it is crucial to be able to flexibly process across numerous resources with different orthographies.
-In many cases even different resources on the `same' language use different orthographic conventions.
-Another orthographic challenge that we encounter regularly in our linguistic practice is electronic resources on a particular language that claim to follow a specific orthographic convention (often a resource-specific convention), but on closer inspection such resources are almost always not consistently encoded.
-Thus a second goal of our orthography profiles is to allow for an easy specification of orthographic conventions, and use such profiles to check consistency and to report errors to be corrected.
-
-A central step in our proposed solution to this problem is the tailored grapheme separation of strings of symbols, a process we call \textsc{grapheme tokenization}.
-Basically, given some strings of symbols (e.g.~morphemes, words, sentences) in a specific source, our first processing step is to specify how these strings have to be separated into graphemes, considering the specific orthographic conventions used in a particular source document.
-Our experience is that such a graphemic tokenization can be performed without extensive in-depth knowledge about the phonetic and phonological details of the language in question.
-For example, the specification that <ou> is a grapheme of English is a much easier task than to specify what exactly the phonetic values of this grapheme are in any specific occurrence in English words.
-Grapheme separation is a task that can be performed relatively reliably and with limited availability of time and resources (compare, for example, the task of creating a complete phonetic or phonological normalization).
-
-Although grapheme tokenization is only one part of the solution, it is an important and highly fruitful processing step.
-Given a grapheme tokenization, various subsequent tasks become easier, like (a) temporarily reducing the orthography in a processing pipeline, e.g.~only distinguishing high versus low vowels; (b) normalizing orthographies across sources (often including temporary reduction of oppositions), e.g.~specifying an (approximate) mapping to the International Phonetic Alphabet; (c) using co-occurrence statistics across different languages (or different sources in the same language) to estimate the probability of grapheme matches, e.g.~with the goal to find regular sound changes between related languages or transliterations between different sources in the same language.
-
-Before we deal with these proposals, in the first part of this paper (Sections 2 through 5) we give an extended introduction to the notion of encoding (Section 2) and writing systems, both from a linguistic perspective and from the perspective of the Unicode Consortium (Section 3).
-We consider the Unicode Standard to be a breakthrough (and ongoing) development that fundamentally changed the way we look at writing systems, and we aim to provide here a slightly more in-depth survey of the many techniques that are available in the standard.
-A good appreciation for the solutions that the Unicode Standard also allows for a thorough understanding of possible pitfalls that one might encounter when using it (Section 4).
-As an example of the current state-of-the-art, we discuss the rather problematic marriage of the International Phonetic Alphabet (IPA) with Unicode (Section 5).
-
-The second part of the paper (Sections 6 and 7) describes our proposals for how to deal with the Unicode Standard in the daily practice of (comparative) linguists.
-First, we discuss the challenges of characterizing a writing system.
-To solve these problems, we propose the notions of orthography profiles, closely related to Unicode locale descriptions (Section 6).
-Finally, we discuss practical issues with actual examples (Section 7).
-We provide reference implementation of our proposals in R and in Python, available as open-source libraries.
-
-The following conventions are followed in this paper.
-All phonemic and phonetic representations are given in the International Phonetic Alphabet (IPA), unless noted otherwise (International Phonetic Association, 2005).
-Standard conventions are used for distinguishing between graphemic <\textgreater{}, phonemic / / and phonetic {[} {]} representations.
-For character descriptions, we follow the notational conventions of the Unicode Standard (Unicode Consortium 2014).
-Character names are represented in small capital letters (e.g.~LATIN SMALL LETTER SCHWA) and code points are expressed as U\emph{+n} where \emph{n} is a four to six digit hexadecimal number, e.g.~U+0256, which can be rendered as the glyph <ə>.
-
-\section{2. Encoding}\label{encoding}
-
-There are many in-depth histories of the origin and development of writing systems (e.g.~Robinson 1997, Powell 2012), a story that we will not repeat here.
-However, the history of turning writing into computer readable code is not so often told, so we decided to offer a short survey of the major developments here.
-This history turns out to be intimately related to the history of telegraphic communication.\footnote{Because of the recent history as summarised in this section, we have used mostly rather ephemeral internet sources, specifically http://www.unicode.org/history/ and various Wikipedia pages for the information presented here (when not references by traditional literature in the bibliography).
-A useful survey of the historical development of the physical hardware of telegraphy and telecommunication is Huurdeman (2003).
-Most books that discuss the development of encoding of telegraphic communication focus of cryptography (e.g.~Singh 1999) and forego the rather interesting story of `open' encoding that is related here.)
-
-Writing systems have existed for roughly 6000 years, allowing people to exchange messages through time and space.
-Additionally, to bridge large geographic distances, \textsc{telegraphic systems} of communication (from Greek \emph{τῆλε γράφειν} ``distant writing'') have a long and widespread history since ancient times.
-The most widespread telegraphic systems worldwide are so-called whistled languages (Meyer 2015), but also drumming languages (Meyer et al.
-2012) and smoke, fire, hydraulic or flag signals are forms of telegraphy.
-Telegraphy was reinvigorated in the end of the eighteenth century through the introduction of semaphoric systems by Claude Chapelle (since then using flags, flashing lights, or various specially designed contraptions) to convey messages over large distances.
-The `innovation' of those systems was that all characters of the written language were replaced one-to-one by visual signals.
-Since then, all telegraphic systems have taken this principle, namely that any language to be transmitted first has to be turned into some orthographic system, which subsequently is encoded for transmission by the sender, and then turned back into orthographic representation at the receiver side.\footnote{Sound and video-based telecommunication of course takes a different approach by ignoring the written version of language and directly encode sound waves or light patterns.) This of course implies that the usefulness of any such telegraphic encoding completely depends on the often rather haphazard structure of orthographic systems.
-
-In the nineteenth century, \textsc{electric telegraphy} lead to a new approach in which written language characters were encoded by signals sent through a copper wire.
-Originally, `bisignal' codes were used, consisting of two different signals.
-For example, Carl Friedrich Gauss in 1833 used positive and negative current (Mania 2008: 282).
-More famous and influential, Samuel Morse in 1836 used long and short pulses.
-In those bisignal systems each character from the written language was encoded with a different number of signals (between one and five), so two different separators are needed: one between signals and one between characters (in Morse-code there is a short pause between signals and a long pause between characters).\footnote{Actually, Morse-code also includes an extra long pause between words.
-Interestingly, it took a long time to consider the written white-space as a bona-fide character that should simply be encoded with its own code point instead of with a different kind of pause.
-This happened only with the revision of the Baudot-code (see below) by Donald Murray in 1901, in which he finally introduced a white-space code.
-This principle has been followed ever since, and for example was included in the second version of the International Telegraph Alphabet (ITA2) in 1930.)
-
-From those encodings, true \textsc{binary codes} developed with a fixed length of signals per character.
-In such systems only a single separator between signals is needed, because the separation between characters can be established by counting until a fixed number of signals.\footnote{Of course, no explicit separator is needed when the timing of the signals is known, which is the principle used in all modern telecommunication systems.
-An important modern consideration is also how to know where to start counting when you did not catch the start of a message, something that is known in Unicode as `self synchronizing'.) In the context of electric telegraphy, such a system was first established by Émile Baudot in 1870, using a fixed combination of five signals for each written character.\footnote{True binary codes have a longer history, going at least back to the Baconian cipher devised by Francis Bacon in 1605.) There are 25 = 32 possible combination when using five binary signals; an encoding today designated as `5-bit'.
-These codes are sufficient for all Latin letters, but of course they do not suffice for all written symbols, including punctuation and digits.
-As a solution, the Baudot code uses a so-called `shift' character, which signifies that from that point onwards (until shifted back) a different encoding is used, allowing for yet another set of 32 codes.
-In effect, this means that the Baudot code, and the \textsc{International Telegraph Alphabet (ITA)} derived from it, had an extra `bit' of information, so the encoding is actually 6-bit (with 26 = 64 different possible characters).
-For decades, this encoding was the standard for all telegraphy and it is still in limited use today.
-
-To also allow for different uppercase and lowercase letters and a large variety of control characters, the American Standards Association decided to propose a new 7-bit encoding in 1963 (with 27 = 128 different possible characters), known as the \textsc{American Standard Code for Information Interchange(ASCII)}, geared towards the encoding of English orthography.
-With the ascent of other orthographies in computer usage, the wish to encode further variation of Latin letters (like German <ß> or various letters with diacritics like <è>) led the Digital Equipment Corporation to introduce an 8-bit \textsc{Multinational Character Set (MCS)} (with 28 = 256 different possible characters), first used with the introduction of the VT220 Terminal in 1983.
-Because 256 characters were clearly not enough for the many different characters needed in the world's writing systems, the ISO/IEC 8859 standard in 1987 extended the MCS to include 16 different 8-bit code pages.
-For example, part 5 was used for Cyrillic characters, part 6 for Arabic, and part 7 for Greek.\footnote{In effect, because 16 = 24, this means that ISO/IEC 8859 was actually a 8+4 = 12-bit encoding, though with very many duplicates by design (e.g.~all ASCII codes were repeated in each 8-bit code page).
-To be precise, ISO/IEC 8859 used the 7-bit ASCII as a basis for each code page, and defined 16 different 7-bit extensions, leading to (1+16) times 27 = 2,176 possible characters.)
-
-This system almost immediately was understood to be insufficient and impractical, so various initiatives to extend and reorganize the encoding started in the 1980s.
-This led, for example, to various proprietary encodings from Microsoft (e.g.~Windows Latin 1) and Apple (e.g.~Mac OS Roman), which one still sometimes encounters today.
-More wide-ranging, various people in the 1980s started to develop true international code sets.
-In the United States, a group of computer scientists formed the Unicode Consortium, proposing a 16-bit encoding in 1991 (with 216 = 65,536 different possible characters).
-At the same time in Europe, the International Organization for Standardization (ISO) was working on ISO 10646 to supplant the ISO 8859 standard.
-Their first draft of the \textsc{Universal Character Set (UCS)} in 1990 was 31-bit (with theoretically 231 = 2,147,483,648 possible characters, but because of some technical restrictions only 679,477,248 were allowed).
-Since 1991, the Unicode Consortium and the International Organization for Standardization have jointly developed the \textsc{Unicode Standard}, or ISO/IEC 10646, leading to the current system including the original 16-bit Unicode proposal as the \textsc{Basic Multilingual Plane}, and 16 additional planes of 16-bit for further extensions (with in total 1+16 times 216 = 1,114,112 possible characters).
-The most recent version of the Unicode Standard (7.0) was published in June 2014 and it defines 112,956 different characters (Unicode Consortium 2014).
-
-\section{3. Terminology}\label{terminology}
-
-\subsection{The linguistic tradition}\label{the-linguistic-tradition}
-
-In this section we provide a very brief overview of the linguistic terminology concerning writing systems before turning to the slightly different computational terminology in the next section on the Unicode Standard.
-Linguistically speaking, a \textsc{writing system} is a symbolic system that uses visible or tactile signs to represent language in a systematic way.
-The term `writing system' has two mutually exclusive meanings.
-First, it may refer to the way a particular language is written.
-In this sense the term refers to the writing system of a particular language, as, for example, in \emph{``the Serbian writing system uses two scripts: Latin and Cyrillic''}.
-Second, the term `writing system' may refer to a type of symbolic system as used among the world's languages to represent languages, as, for example, in \emph{``alphabetic writing system''}.
-In this latter sense the term refers to how scripts have been classified according to the way that they encode language, as in, for example, \emph{``the Latin and Cyrillic scripts are both alphabetic writing systems''}.
-To avoid confusion, this second notion of `writing system' would have been more aptly called \textsc{script system}.
-The term \textsc{script} refers to a collection of distinct symbols as employed by one or more orthographies.
-For example, both Serbian and Russian are written with subsets of the Cyrillic script.
-A single language, like Serbian or Japanese, can also be written using orthographies based on different scripts.
-Over the years linguists have typologized script systems in a variety of ways, with the tripartite classification of logographic, syllabic, and alphabetic remaining the most popular, even though there are at least half a dozen different types of script systems that can be distinguished (Daniels, 1990, 1996).
-
-Only focussing here on the first sense, we distinguish two different kinds of writing systems as used to write a particular language, namely transcriptions and orthographies.
-First, \textsc{transcription} is a scientific procedure (and also the result of that procedure) for graphically representing the sounds of human speech at the phonetic level.
-It incorporates a set of unambiguous symbols to represent speech sounds, including conventions that specify how these symbols should be combined.
-A \textsc{transcription system} is a specific system of symbols and rules used for transcription of the sounds of a spoken language variety.
-In principle, a transcription system should be language-independent, in that it should be applicable to all spoken human languages.
-The International Phonetic Alphabet (IPA) is a commonly used transcription system that provides a medium for transcribing languages at the phonetic level.
-However, there is a long history of alternative kinds of transcription systems (see Kemp 1994) and today various alternatives are in widespread use (e.g.~X-SAMPA and Cyrillic-based phonetic transcription systems).\footnote{Most users of IPA do not follow the standard explicitly and many dialects based on it have emerged, e.g.~the Africanist and Americanist transcription systems.
-) Note that IPA symbols are also often used to represent language on a phonemic level.
-However, it is important to realize that in this usage the IPA symbols are not a transcription system, but rather an orthography (though with strong links to the pronunciation).
-Further, a transcription system does not need to be highly detailed, but it can also be a system of broad sound classes.
-Although such an approximative transcription is not normally used in linguistics, it is widespread in technological approaches (e.g.~Soundex, Beider \& Morse 2010)\footnote{http://stevemorse.org/phonetics/bmpm.htm , http://stevemorse.org/phonetics/bmpm2.htm) and it is sometimes fruitfully used in automatic approaches to historical linguistics (Dolgopolsky 1986, ASJP\footnote{http://asjp.clld.org/)).
-
-Second, an \textsc{orthography} specifies the symbols, punctuations, and the rules in which a specific language is written in a standardized way.
-Orthographies are often based on a phonemic analysis, but they almost always include idiosyncrasies because of historical developments (like sound changes or loans) and because of the widely-followed principle of lexical integrity (i.e.~the attempt to write the same lexical root in a consistent way, also when synchronic phonemic rules change the pronunciation, for example with final devoicing in many Germanic languages).
-All orthographies are language-specific (and often even resource-specific), although individual symbols or rules might be shared between languages.
-A \textsc{practical orthography} is a strongly phoneme-based writing system designed for practical use by speakers.
-The mapping relation between phonemes and graphemes in practical orthographies is purposely shallow, i.e.~there is mostly a systematic and faithful mapping from a phoneme to a grapheme.
-Practical orthographies are intended to jumpstart written materials development by correlating a writing system with the sound units of a language (cf.~Meinhof \& Jones 1928).
-Symbols from the IPA are often used by linguists in the development of practical orthographies for languages without writing systems, though this usage of IPA symbols should not be confused with transcription (as defined above).
-Further, a \textsc{transliteration} is a mapping between two different orthographies.
-It is the process of ``recording the graphic symbols of one writing system in terms of the corresponding graphic symbols of a second writing system'' (Kemp 2006:396).
-In straightforward cases, such a transliteration is simply a matter of replacing one symbol with another.
-However, there are widespread complications, like one-to-many or many-to-many mappings, which are not always easy, or even possible, to solve without listing all cases individually (cf.~Moran 2012, chp 2).
-
-Breaking it down further, a script consists of \textsc{graphemes}, and graphemes consist of \textsc{characters}.
-In the linguistic terminology of writing systems, a \textsc{character} is a general term for any self-contained element in a writing system.\footnote{There is a second interpretation of the term `character', i.e.~a conventional term for a unit in the Chinese writing system (Daniels 1996).
-This interpretation will not be further explored in this paper.) Although in literate societies most people have a strong intuition about what the characters are in their particular orthography or orthographies, it turns out that the separation of an orthography into separate characters is far from trivial.
-The widespread intuitive notion of a character is strongly biased towards educational traditions, like the alphabet taught at schools, and technological possibilities, like the available type pieces in a printer's job case, the keys on a typewriter, or the symbols displayed in Microsoft Word's symbol browser.
-In practice, characters often consist of multiple building blocks, each of which could be considered a character in its own right.
-For example, although a Chinese character may be considered to be a single basic unanalyzable unit, at a more fine-grained level of analysis the internal structure of Chinese characters is often comprised of smaller semantic and phonetic units that should be considered characters (Sproat 2000).
-In alphabetic scripts, this problem is most forcefully exemplified by diacritics.
-A \textsc{diacritic} is a mark, or series of marks, that may be above, below, or through other characters (Gaultney 2002).
-Diacritics are sometimes used to distinguish homophonous words, but they are more often used to indicate a modified pronunciation (Daniels \& Bright 1996, xli).
-The central question is whether, for example, , , and should be considered four characters, or different combinations of three characters.
-In general, multiple characters together can form another character, and it is not always possible to decide on principled grounds what should be the basic building blocks of an orthography.
-
-For that reason, it is better to analyse an orthography as a collection of graphemes.
-A \textsc{grapheme} is the basic, minimally distinctive symbol of a particular writing system, alike to the phoneme is an abstract representation of a distinct sound in a specific language.
-The term `grapheme' was modeled after the term `phoneme' and represents a contrastive graphical unit in a writing system (see Kohrt 1986 for a historical overview of the term grapheme).
-Most importantly, a single grapheme regularly consists of multiple characters, like <th>, <ou> and <gh> in English (note that each character in these graphemes is also a separate grapheme in English).
-Such complex graphemes are often used to represent single phonemes.
-So, a combination of characters is used to represent a single phoneme.
-Note that the opposite is also found in writing systems, in cases in which a single character represents a combination of two or more phonemes.
-For example, <x> in English orthography represents a combination of the phonemes /k/ and /s/.
-Further, conditioned or free variants of a grapheme are called \textsc{allographs}.
-For example, the distinctive forms of Greek sigma are conditioned, with <σ> being used word-internally and <ς> being used at the end of a word.
-In sum, there are many-to-many relationships between phonemes and graphemes as they are expressed in the myriad of language- and resource-specific orthographies.
-
-This exposition of the linguistic terminology involved in describing writing systems has been purposely brief.
-We have highlighted some of the linguistic notions that are pertinent, yet sometimes confused with, the technological definitions developed for the computational processing of the world's writing systems, which we describe in the next section.
-
-\subsection{The Unicode approach}\label{the-unicode-approach}
-
-The conceptualisation and terminology of writing systems was rejuvenated through the development of the Unicode Standard, with major input from Mark Davis, co-founder and long-term president of the Unicode Consortium.
-For many years, `exotic' writing systems and phonetic transcription systems on personal computers were constrained by the American Standard Code for Information Interchange* *(ASCII) character encoding scheme (based on the Latin script), which only allowed for a strongly limited number of different symbols to be encoded.
-This implied that users could either use and adopt the (extended) Latin alphabet or they could assign new symbols to the small number of code points in the ASCII encoding scheme to be rendered by a specifically designed font (Bird \& Simons 2003).
-In this situation, it was necessary to specify the font together with each document to ensure the rightful display of its content.
-To alleviate this problem of assigning different symbols to the same code points, in the late 80's and early 90's the Unicode Consortium set itself the ambitious goal of developing a single universal character encoding to provide a unique number, a code point, for every character in the world's writing systems.
-Nowadays, the Unicode Standard is the default encoding of the technologies that support the World Wide Web and for all modern operating systems, software and programming languages.
-
-The Unicode Standard represents a massive step forward because it aims to eradicate the distinction between `universal' (ASCII) versus `language/resource-particular' (Font) by adding as much as possible language-specific information into the universal standard.
-However, there are still language/resource-specific specifications necessary for the proper usage of Unicode, as will be discussed below.
-Within the Unicode structure many of these specifications can be captured by so-called `Locales' or `Locale Descriptions', so we are moving to a new distinction of `universal' (Unicode Standard) versus `language-particular' (Locale Description).
-The major gain is a much larger compatibility on the universal level (because Unicode standardizes a much larger portion of writing system diversity), and much better possibilities for automatic processing on the language-particular level (because Locale Descriptions are computer readable specifications).
-
-Each version of the Unicode Standard (as of writing at version 7) consists of a set of specifications and guidelines that include (i) a core specification, (ii) code charts, (iii) standard annexes and (iv) a character database.\footnote{All documents of the Unicode Standard are available at http://www.unicode.org/versions/latest/.
-For a quick survey of the use of terminology inside the Unicode Standard, their glossary is particularly useful, available at http://www.unicode.org/glossary/.
-For a general introduction to the principles of Unicode, Chapter 2 of the core specification, called ``General Structure'', is particularly insightful.
-Different from many other documents of the Unicode Standard, this general introduction is relatively easy to read and illustrated with many interesting examples from various orthography traditions all over the world.
-) The \textsc{core specification} is a book directed toward human readers that describes the formal standard for encoding multilingual text.
-The \textsc{code charts} provide a humanly readable online reference to the character contents of the Unicode Standard in the form of PDF files.
-The \textsc{Unicode Standard Annexes (UAX)} are a set of technical standards that describe the implementation of the Unicode Standard for software development, Web standards, programming languages, etc.
-The \textsc{Unicode Character Database (UCD)} is a set of computer-readable text files that describe the character properties, including a set of rich character and writing system semantics, for each character in the Unicode Standard.
-In this section, we introduce the basic Unicode concepts, but we will leave out many details.
-Please consult the above mentioned full documentation for a more detailed discussion.
-Further note that the Unicode Standard is exactly that, namely a standard.
-It normatively describes notions and rules to be followed.
-In the actual practice of applying this standard in a computational setting, a specific implementation is necessary.
-The most widely used implementation of the Unicode Standard is the \textsc{International Components for Unicode (ICU)}, which offers C/C++ and Java libraries trying to implement to Unicode Standard.\footnote{http://icu-project.org )
-
-The Unicode Standard is a \textsc{character encoding system} which goal it is to support the interchange and processing of written characters and text in a computational setting.
-Underlyingly, the character encoding is represented by a range of numerical values called a \textsc{code space}, which is used to encode a set of characters.
-A \textsc{code point} is a unique non-negative integer within a code space (i.e.~within a certain numerical range).
-In the Unicode Standard character encoding system, an \textsc{abstract character}, for example a ``LATIN SMALL LETTER P'', is mapped to a particular code point, such as the decimal value 112, normally represented in hexadecimal and in Unicode parlance as (U+)0070.\footnote{Hexadecimal (`base-16') 0070 is equivalent to decimal (`base-10') 112, which can be calculated by considering 0·163 + 0·162 + 7·161 + 0·160 = 7·16 = 112.
-Underlyingly, computers will of course treat this code point binary (`base-2') as 11100000, i.e.~112 = 1·27 + 1·26 + 1·25 + 0·24 + 0·23 + 0·22 + 0·21 + 0·20 = 64 + 32 + 16.) That encoded abstract character is rendered on a computer screen (or printed page) as a \textsc{glyph}, e.g.
-<p>, depending on the \textsc{font} and the context in which that character appears.
-
-In Unicode Standard terminology, an (abstract) \textsc{character} is the basic encoding unit.
-The term `character' can be quite confusing due to its alternative definitions across different scientific disciplines and because in general the word `character' means many different things to different people.
-It is therefore often preferable to refer to Unicode characters simply as \textsc{code points}, because there is a one-to-one mapping between Unicode characters and their numeric representation.
-The Unicode Consortium (2012) defines a character as ``the smallest component of written language that has semantic value.
-It refers to the abstract meaning and/or general shape, rather than a specific shape, though in code tables some form of visual representation is essential for the reader's understanding.'' Unicode defines characters as abstractions of orthographic symbols, and it does not define visualizations for these characters (although it does presents examples).
-In contrast, a \textsc{glyph} is a concrete graphical representation of a character as it appears when rendered (or rasterized) and displayed on an electronic device or on printed paper.
-For example, <reinstert Gs here> are different glyphs of the same character, i.e.~they may be rendered differently depending on the typography being used, but they all share the same code point.
-From the perspective of Unicode they are `the same thing'.
-In this approach, a \textsc{font} is then simply a collection of glyphs linked to code points.
-Allography is not specified in Unicode (expect for a few exceptional cases, due to legacy encoding issues), but can be specified in a font as a \textsc{contextual variant} (aka presentation form).
-
-Each code point in the Unicode Standard is associated with a set of \textsc{character properties} as defined by the Unicode character property model.\footnote{The character property model is described in http://www.unicode.org/reports/tr23/, but the actual properties are described in http://www.unicode.org/reports/tr44/.
-A simplified overview of the properties is available at http://userguide.icu-project.org/strings/properties.
-The actual code tables listing all properties for all Unicode code points are available at http://www.unicode.org/Public/UCD/latest/ucd/.
-) Basically, those properties are just a long list of values for each character.
-For example, code point U+0047 has the following properties (among many others):
-\begin{itemize}
-	\item Name: LATIN CAPITAL LETTER G 
-	\item Alphabetic: YES 
-	\item Uppercase: YES 
-	\item Script: LATIN 
-	\item Extender: NO 
-	\item Simple\_Lowercase\_Mapping: 0067 
-\end{itemize}
-
-These properties contain the basic information of the Unicode Standard and they are necessary to define the correct behavior and conformance required for interoperability in and across different software implementations (as defined in the Unicode Standard Annexes).
-The character properties assigned to each code point is based on each character's behavior in the real-world writing traditions.
-For example, the corresponding lowercase character to U+0047 is U+0067 (though note that the relation between uppercase and lowercase is in many situations much more complex than this, and Unicode has further specifications for those cases).
-Another use of properties is to define the script of a character.
-The Unicode Standard defines the term \textsc{script} as, ``A collection of letters and other written signs used to represent textual information in one or more writing systems.
-For example, Russian is written with a subset of the Cyrillic script; Ukrainian is written with a different subset.
-The Japanese writing system uses several scripts.'' In practice, script is simply defined for each character as the explicit ``Script'' property in the Unicode Character Database.
-
-One frequently references property is the \textsc{block} property, which is often used in software applications to impose some structure to the large number of Unicode characters.
-Each character in Unicode belongs to a specific block.
-These blocks are basically an organisational structure to alleviate the administrative burden of keeping Unicode up-to-date.
-Blocks consist of characters that in some way belong together, so that characters are easier to find.
-Some blocks are connected with a specific script, like the Hebrew block or the Gujarati block.
-However, blocks are predefined ranges of code points, and often there will come a point after which the range is completely filled.
-Any extra characters will have to be assigned somewhere else.
-There is, for example, a block `Arabic', which contains most Arabic symbols.
-However, there is also a block `Arabic Supplement', `Arabic Presentation Forms-A' and `Arabic Presentation Form B'.
-The situation with Latin symbols is even more extra.
-In general, the names for block should be taken as a definitional statement.
-For example, many IPA symbols are not located in the aptly-names block ``IPA extensions'', but in other blocks (see Section 5).
-
-There are many cases in which a sequence of characters (i.e.~a sequence of more than one code point) represents what a user perceives as an individual unit in a particular orthographic writing system.
-For this reason the Unicode Standard differentiates between \textsc{abstract character} and \textsc{user-perceived character}.
-Sequences of multiple code points that correspond to a single user-perceived characters are called** grapheme clusters** in Unicode parlance.
-Grapheme clusters come in two flavors: (default) grapheme clusters and tailored grapheme clusters.
-
-The (default) \textsc{grapheme clusters} are locale-independent graphemes, i.e.~they always apply when a particular combination of characters occurs independent of the writing system in which they are used.
-These character combinations are defined in the Unicode Standard as functioning as one \textsc{text element}.\footnote{The Unicode Standard defines text element as: ``A minimum unit of text in relation to a particular text process, in the context of a given writing system.
-In general, the mapping between text elements and code points is many-to-many.'') The simplest example of a grapheme cluster is a base character followed by a letter modifier character.
-For example, the sequence + <̃\textgreater{} (LATIN SMALL LETTER N at U+006E followed by COMBINING TILDE at U+0303) combines visually into , a user-percieved character in writing systems like that of Spanish.
-So, what the user perceives as a single character actually involves a multi-code-point sequence.
-Note that this specific sequence can also be represented with a single \textsc{precomposed} code point, LATIN SMALL LETTER N WITH TILDE at U+00F1, but this is not the case for all multi-code point character sequences.
-The problem that there multiple encodings possible for the same text element has been acknowledged early on in the Unicode Standard (e.g.~for , the sequence U+006E U+0303 should in all situations be treated identically to the precomposed U+00F1), and a system of \textsc{canonical equivalence} is available for such situations.
-Basically, the Unicode Standard offers different kind of normalizations to either decompose all precomposed characters (called \textsc{NFD}, \textsc{Normalization Form Canonical Decomposition}), or compose as much as possible combinations (called \textsc{NFC}, \textsc{Normalization Form Canonical Composition}).
-In current practice of software development, NFC seems to be preferred in most situations and is widely proposed as the preferred canonical form.
-
-More difficult for text processing, because less standardized, is what the Unicode Standard terms \textsc{tailored grapheme clusters}.
-Tailored grapheme clusters are locale-dependent graphemes, i.e.~such combination of characters do not function as text elements in all situations.
-For example, the sequence + for the Slovak digraph or the sequence in the Sisaala practical orthography (pronounced as IPA /tʃ/; Moran 2006).
-These grapheme clusters are ``tailored'' in the sense that they must be specified on a language-by-language or a writing system-by-writing system basis.
-The Unicode Standard provides technological specifications for creating locale specific data in so-called \textsc{Unicode Locale Descriptions}, i.e.~a set of specification that defines a set of language-specific elements (e.g.~tailored grapheme clusters, collation order, capitalization-equivalence), as well as other special information, like how to format numbers, dates, or currencies.
-Locale descriptions are saved in the \textsc{Common Locale Data Repository (CLDR)},\footnote{http://cldr.unicode.org/ ) a repository of language-specific definitions of writing system properties, each of which describes specific usages of characters.
-Each locale can be encoded in a document using the \textsc{Locale Data Markup Language (LDML)}.
-LDML is an XML format and vocabulary for the exchange of structured locale data.
-
-Unicode Locale Descriptions allow users to define language- or resource-specific writing systems or orthographies.
-The Unicode Consortium defines \textsc{writing system} only very loosely, as it is not a central concept in the Unicode Standard.
-A writing system is, ``A set of rules for using one or more scripts to write a particular language.
-Examples include the American English writing system, the British English writing system, the French writing system, and the Japanese writing system.''.
-However, there are various drawbacks of locale descriptions for the daily practice of linguistic work in a multilingual setting (see Section 7).
-
-\section{4. Unicode Pitfalls}\label{unicode-pitfalls}
-
-\subsection{Wrong it is not}\label{wrong-it-is-not}
-
-In this section, we describe some of the most common pitfalls that we have encountered when using the Unicode Standard in our own work, or in discussion with other linguists.
-This section is not meant as a criticism of the decisions made by the Unicode Consortium; on the contrary, we aim to highlight where the technological aspects of the Unicode Standard diverge from many users' intuitions.
-What have sometimes been referred to as problems or inconsistencies in the Unicode Standard are mostly due to legacy compatibility issues, which can lead to unexpected behavior by linguists using the standard.
-However, there are also some cases in which the Unicode Standard has made decisions that theoretically could have been taken differently, but for some reason or another (mostly very good reasons) were accepted as they are now.
-We call behavior that executes without error but does something different than the user expected---often unknowingly---a ``pitfall''.
-
-In this context, it is important to realize that the Unicode Standard was not developed to solve linguistic problems per se, but to offer a consistent computational environment for written language.
-In those cases in which the Unicode Standard behaves differently as expected, we think it is important not to dismiss Unicode as `wrong' or `deficient', because our experience is that in almost all cases the behaviour of the Unicode Standard has been particularly well thought through.
-The Unicode Consortium considers important practical use-cases that from a linguistic point of view are normally not considered.
-Our general guideline for dealing with the Unicode Standard is to accept it as it is, and not to battle windmills.
-Alternatively, of course, it is possible to actively engage in the development of the standard itself, an effort that is highly appreciated by the Unicode Consortium.
-
-\subsection{Pitfall: Blocks}\label{pitfall-blocks}
-
-The Unicode code space is subdivided into blocks of contiguous code points.
-For example, the block called ``Cyrillic'' runs from U+0400 till U+04FF.
-These blocks arose as an attempt at ordering the enormous amount of characters in Unicode, but the ideas of blocks very quickly ran into problems.
-First, the size of a block is fixed, so when a block is full, a new block will have to be instantiated somewhere further in the code space.
-For example, this led to the blocks ``Cyrillic Supplement'', ``Cyrillic Extended-A'' (which are also already full) and ``Cyrillic Extended-B''.
-Second, when a specific character already exists, then it is not duplicated in another block, although the name of the block might indicate that a specific symbol should be available there.
-In general, names of blocks are just an approximate indication of the kind of characters that will be in the block.
-
-The problem with blocks arises because finding the right character among the thousands of Unicode characters is not easy.
-Many software applications present blocks as a primary search mechanism, because the block names suggest where to look for a particular character.
-However, when a user searches for an IPA character in the block ``IPA Extensions'' then many IPA characters will not be found there.
-For example, the velar nasal <ŋ> is not part of the block ``IPA Extensions'' because it was already included as LATIN SMALL LETTER ENG at U+014B in the block ``Latin Extensions-A''.
-
-In general, finding a specific character in the Unicode Standard is often not trivial.
-The names of the blocks can help, but it is not (and never was supposed to be) a foolproof structure.
-It is not the goal nor aim of the Unicode Consortium to provide a user interface to the Unicode Standard.
-If one often encounters the problem of needing to find a suitable character, there are various other useful services for end users available.\footnote{The Unicode website offers a basic interface to the code charts at http://www.unicode.org/charts/index.html.
-As a more flexible interface, we particularly like PopChar from Macility, available for both Macintosh and Windows.
-However, there are various free websites that offer search interfaces to the Unicode code tables, like http://unicode-search.net or http://unicode-search.net.
-A further useful approach for searching characters using shape matching is http://shapecatcher.com.
-)
-
-\subsection{Pitfall: Missing glyphs}\label{pitfall-missing-glyphs}
-
-The Unicode Standard is often praised (and deservedly so) for solving many of the perennial problems with the interchange and display of the world's writing systems.
-However, a common complaint from users is that, while the praise may be true, they mostly just see some boxes on their screen instead of those promised symbols.
-The problem of course is that users' computers do not have any glyphs installed matching the Unicode code points in the file they are trying to inspect.
-It is important to realize that internally in the computer everything still works as expected: any handling of Unicode code points works independently of how they are displayed on the screen.
-So, although a user might only see boxes being displayed, this user should be assured that everything is still in order.
-
-The central problem behind the missing glyphs is that designing actual glyphs includes a lot of different considerations and it is a time-consuming process.
-Many traditional expectations of how specific characters should look like have to be taken into account when designing glyphs.
-Those expectations are often not well documented, and it is mostly up to the knowledge and experience of the font designer to try and conform to them as good as possible.
-Therefore, most designers produce fonts only including glyphs for certains parts of the Unicode Standard, namely for those characters they feel comfortable with.
-At the same time, the number of characters defined by the Unicode Standard is growing with each new version, so it is neigh impossible for any designer to produce glyphs for all characters.
-The result of this is that, almost necessarily, each font only includes glyphs for a subset of the characters in the Unicode Standard.
-
-The simple solution to missing glyphs is thus to install additional fonts providing additional glyphs.
-For the more exotic characters there is often not much choice.
-There are a few particularly large fonts that might be considered.
-First, there is the \textsc{Everson Mono} font made by Michael Everson, which currently includes 9,756 different glyphs (not including Chinese) updated up to Unicode 7.0.\footnote{Everson Mono is available as shareware at http://www.evertype.com/emono/.
-) Already a bit older is the \textsc{Titus Cyberbit Basic} font made by Jost Gippert and Carl-Martin Bunz, which includes 10,044 different glyphs (not including Chinese), but not including newer characters added after Unicode 4.0.\footnote{Titus Cyberbit Basic is available at http://titus.fkidg1.uni-frankfurt.de/unicode/tituut.asp.
-)
-
-Further, we suggest to always install at least one so-called \textsc{fall-back font,} which provides glyphs that at least show the user some information about the underlyingly encoded character.
-Apple Macintoshes have such a font (which is invisible to the user), which is designed by Michael Everson and made available for other systems through the Unicode Consortium.\footnote{The Apple/Everson fallback font is available at http://www.unicode.org/policies/lastresortfont\_eula.html.
-) Further, the \textsc{GNU Unifont} is a clever way to produce bitmaps approximating the intended glyph of each available character, updated to Unicode 7.0.\footnote{The GNU Unifont is available at http://unifoundry.com/unifont.html.
-) Finally, the Summer Institute of Linguistics provides a \textsc{SIL Unicode BMP Fallback Font}, currently available up to Unicode version 6.1.
-This font does not even attempt to show a real glyph, but only shows the hexadecimal code inside a box for each character, so a user can at least see the Unicode codepoint of the character to be displayed.\footnote{The SIL Unicode BMP Fallback Font is available at http://scripts.sil.org/UnicodeBMPFallbackFont.
-)
-
-\subsection{Pitfall: Faulty rendering}\label{pitfall-faulty-rendering}
-
-A similar complaint to missing glyphs, discussed previously, is that while there might be a glyph being displayed, it does not `look right'.
-There are two reasons for unexpected visual display, namely automatic font substitution and faulty rendering.
-Like missing glyphs, any such problems are independent from the Unicode Standard.
-The Unicode Standard only includes very general information about characters and leaves the specific visual display to others to decide on.
-Any faulty display is thus not to be blamed on the Unicode Consortium, but on a complex interplay of different mechanisms happening in a computer to turn Unicode codepoints into visual symbols.
-We will only sketch a few aspects of this complex interplay here.
-
-Most modern software applications (like Microsoft Word) offer some approach to \textsc{automatic font substitution}.
-This means that when a text is written in a specific font (e.g.~Times New Roman) and a inserted Unicode character does not have a glyph within this font, then the software application will automatically search for another font to display the glyph.
-The result will be that some specific glyph will look slightly different from the others.
-This mechanism works differently depending on the software application, and mostly only limited user influence is expected and little feedback is given, which might be rather frustrating to font-aware users.\footnote{For example, Apple Pages does not give any feedback that a font is being replaced, and the user does not seem to have any influence on the choice of replacement (except by manually marking all occurrences).
-In contrast, Microsoft Word does indicate the font replacement by showing the name of the font replacement, but it simply changes the font, so any text written after the replacement is written in a different font as before.
-Both behaviors leave much to be desired.)
-
-The other problem with visual display is related to the so-called `font rendering'.
-\textsc{Font rendering }refers to the process of the actual positioning of Unicode characters on a page of written text.
-This positioning is actually a highly complex problem, and many things can go wrong in the process.
-Well-known rendering problems, like proportional glyph size or ligatures are reasonably well understood.
-However, even within the Latin script, the positioning of diacritics relative to a base character is still a widespread problem.
-Especially when more than one diacritic is supposed to be placed above (or below) each other, this often leads to unexpected effects in many modern software applications.
-The problems arising in Arabic and in many southeast Asian scripts (like Devanagari or Burmese) are even much more complex.
-There are basically three different approaches to font rendering.
-The most widespread is Adobe's and Microsoft's \emph{OpenType }system.
-This approach makes it relatively easy for font developers, as the font itself does not include all details about the precise placement of individual characters.
-For those details, additional script-descriptions are necessary.
-All of those systems can lead to unexpected behaviour.\footnote{For more details about OpenType, see http://www.adobe.com/products/type/opentype.html and http://www.microsoft.com/typography/otspec/.
-Additional systems for complex text layout are, among others, Microsoft's DirectWrite https://msdn.microsoft.com/library/dd368038.aspx and the open-source project HarfBuzz http://www.freedesktop.org/wiki/Software/HarfBuzz/.
-) Alternative systems are Apple's \emph{Apple Advanced Typography} (AAT) and the open-source \emph{Graphite} system from the Summer Institute of Linguistics (SIL).\footnote{More information about AAT can be found at https://developer.apple.com/fonts/.
-SIL's Graphite is described in detail at http://scripts.sil.org/cms/scripts/page.php?site\_id=projects\&item\_id=graphite\_home.
-) In both of these systems, a larger burden is placed on the description inside the font.
-
-\subsection{Pitfall: Homoglyphs}\label{pitfall-homoglyphs}
-
-Homoglyphs are visually indistinguishable glyphs (or highly similar glyphs) that have different code points in the Unicode Standard and thus different character semantics.
-As a principle, the Unicode Standard does not specify how a character appears visually on the page or the screen.
-So in most cases, a different appearance is caused by the specific design of a font, or by user-settings like size or boldface.
-Taking an example already discussed in the previous section, the following symbols <reinsert Gs here> are different glyphs of the same character, i.e.~they may be rendered differently depending on the typography being used, but they all share the same code point (viz.
-LATIN SMALL LETTER G at U+0067).
-In contrast, the symbols <A А Α Ꭺ ᗅ ᴀ ꓮＡ𐊠 𐌀 𝖠 𝙰> are all different code points, although they look highly similar -- in some cases even sharing exactly the same glyph in some fonts.
-All these different A-like characters include the following code points in the Unicode Standard:
-\begin{itemize}
-	\item <A\textgreater{} LATIN CAPITAL LETTER A, at U+0041 
-	\item <А> CYRILLIC CAPITAL LETTER A, at U+0410 
-	\item <Α> GREEK CAPITAL LETTER ALPHA, at U+0391 
-	\item <Ꭺ> CHEROKEE LETTER GO, at U+13AA 
-	\item <ᗅ> CANADIAN SYLLABICS CARRIER GHO, at U+15C5 
-	\item <ᴀ> LATIN LETTER SMALL CAPITAL A, at U+1D00 
-	\item <ꓮ> LISU LETTER A, at U+A4EE 
-	\item <Ａ> FULLWIDTH LATIN CAPITAL LETTER A, at U+FF21 
-	\item <𐊠> CARIAN LETTER A, at U+102A0 
-	\item <𐌀> OLD ITALIC LETTER A, at U+10300 
-	\item <𝖠> MATHEMATICAL SANS-SERIF CAPITAL A, U+1D5A0 
-	\item <𝙰> MATHEMATICAL MONOSPACE CAPITAL A, at U+1D670 
-\end{itemize}
-
-The existence of such homoglyphs is partly due to legacy compatibility, but for the most part these characters are simply different characters that happen to look similar.\footnote{A particularly nice interface to look for homoglyphs is http://shapecatcher.com, based on the principle of recognizing shapes (Belongie et al.
-2002).) Yet, they are supposed to behave different from the perspective of a font designer.
-For example, when designing a Cyrillic font, the will have different aesthetics and different traditional expectation compared to a Latin .
-
-Such homoglyphs are a widespread problem for consistent encoding.
-Although for most users it looks like the words and are identical, in actual fact they do not even share a single code point.\footnote{The first words consists completely of Latin characters, namely U+0076, U+006F, U+0063, U+0065, U+0073, while the second is a mix of Cyrillic and Greek characters, namely U+03BD, U+03BF, U+0041, U+0435, U+0455.) For computers these two words are completely different entities.
-Commonly, when users with Cyrillic or Greek keyboards have to type some Latin-based orthography, they mix similar looking Cyrillic or Greek characters into their text, because those characters are so much easier to type.
-Similarly, when users want to enter an unusual symbol, they normally search by visual impression in their favourite software application, and just pick something that looks reasonably alike to what they expect the glyph to look like.
-
-\subsection{Pitfall: Canonical equivalence}\label{pitfall-canonical-equivalence}
-
-For some characters, there is more than one possible encoding in the Unicode Standard.
-This is a possible pitfall, as this would mean that for the computer there exist multiple different entities that for a user are the same.
-This would, for example, lead to problems with searching, as the computer would search for specific encodings, and not find all expected characters.
-As a solution, the Unicode Standard includes a notion of \textsc{canonical equivalence}.
-Different encodings are explicitly declared as equivalent in the Unicode Standard code tables.
-Further, to harmonize all encodings in a specific piece of text, the Unicode Standard proposes a mechanism of \textsc{normalization}.
-
-Consider for example the characters and following Unicode code points:
-\begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item LATIN CAPITAL LETTER A WITH RING ABOVE (U+00C5) 
-	\item ANGSTROM SIGN (U+212B) 
-	\item LATIN CAPITAL LETTER A (U+0041) + COMBINING RING ABOVE (U+030A) 
-\end{enumerate}
-
-The character, represented here by glyph , is encoded in the Unicode Standard in the first two examples by a single-character sequence; each is assigned a different code point.
-In the third example, the glyph is encoded in a multiple-character sequence that is composed of two character code points.
-All three sequences are \textsc{canonically equivalent}, i.e.~they are strings that represent the same abstract character and because they are not distinguishable by the user, the Unicode Standard requires them to be treated the same in regards to their behavior and appearance.
-Nevertheless, they are encoded differently.
-For example, if one were to search an electronic text (with software that does not apply Unicode Standard normalization) for ANGSTROM SIGN (U+212B), instances of LATIN CAPITAL LETTER A WITH RING ABOVE (U+00C5) would not be found.
-
-In other words, there are equivalent sequences of Unicode characters that should be normalized, i.e.~transformed into a unique Unicode-sanctioned representation of a character sequence called a \textsc{normalization form}.
-Unicode provides a \emph{Unicode Normalization Algorithm}, which essentially puts combining marks into a specific logical order and it defines decomposition and composition transformation rules to convert each string into one of four normalization forms.
-We will discuss here the two most relevant normalization forms: NFC and NFD.
-
-The first of the three characters above is considered the \textsc{Normalization Form C (NFC)}, where ``C'' stands for composition.
-When the process of NFC normalization is applied to the character sequences in 2 and 3, both sequences are normalized into the `pre-composed' character sequence in 1.
-Thus all three canonical character sequences are standardized into one composition form in NFC.
-The other central Unicode normalization form is the \textsc{Normalization Form D (NFD)}, where ``D'' stands for decomposition.
-When NFD is applied to the three examples above, all three, including importantly the single-character sequences in 1 and 2, are normalized into the `decomposed' multiple-sequence of characters in 3.
-Again, all three are then logically equivalent and therefore comparable and syntactically interoperable.
-
-As illustrated, some characters in the Unicode Standard have alternative representations (in fact, many do), but the Unicode Normalization Algorithm can be used to transform certain sequences of characters into canonical (or compatible) forms to test for equivalency.
-To determine equivalence, each character in the Unicode Standard is associated with a combining class, which is formally defined as a character property called ``Canonical Combining Class'' which is specified in the Unicode Character Database.
-The combining class assigned to each code point is a numeric value between 0 and 254 and is used by the Canonical Ordering Algorithm to determine which sequences of combining marks are canonically equivalent.
-Normalization forms, as very briefly described above, can be used to ensure character equivalence by ordering character sequences so that they can be faithfully compared.
-
-It is very important to note that any software applications that is Unicode Standard compliant is free to change the character stream from one representation to another.
-This means that a software application may compose, decompose or reorder characters as its developers desire; as long as the resultant strings are canonically equivalent to the original.
-This might lead to unexpected behaviour for users.
-Various players, like the Unicode Consortium, the W3C, or the TEI recommend NFC in most user-directed situations, and some software applications that we tested indeed seem to automatically convert strings into NFC.\footnote{See the summary of various recommendation here: http://www.win.tue.nl/\textasciitilde{}aeb/linux/uc/nfc\_vs\_nfd.html.
-) This means in practice that if a user, for example, enters <a> and <`>, i.e.~LATIN SMALL LETTER A at U+0061 and COMBINING GRAVE ACCENT at U+0300, this might be converted into <à>, i.e.~LATIN SMALL LETTER A WITH GRAVE at U+00E0.\footnote{The behaviour of software applications can be quite erratic in this respect.
-For example, Apple's TextEdit does not do any conversion on text entry.
-However, when you copy and paste some text inside the same document in ``rich text mode'' (i.e.~RTF-format), it will be transformed into NFC on paste.
-Saving a document does not do any conversion to the glyphs on screen, but it will save the characters in NFC.)
-
-\subsection{Pitfall: Characters are not glyphs}\label{pitfall-characters-are-not-glyphs}
-
-A central principle of Unicode is the distinction between character and glyph.
-A character is the abstract notion of a symbol in a writing system, while a glyph is the concrete drawing of such a symbol.
-In practice, there is a complex interaction between characters and glyphs.
-A single Unicode character may of course be rendered as a single glyph.
-However, a character may also be a `piece' of a glyph, or vice-versa.
-Actually, all possible relations between glyphs and characters are attested.
-
-First, \textsc{a single character may have different contextually determined glyphs.} For example, characters in writing systems like Hebrew and Arabic have different glyphs depending on where they appear in a word.
-Some letters in Hebrew change their form at the end of the word, and in Arabic, primary letters have four contextually-sensitive variants (isolated, word initial, medial and final).
-Second, \textsc{a single character may be rendered as a sequence of multiple glyphs.} For example, in Tamil one Unicode character may result in a combination of a consonant and vowel, which are rendered as two adjacent glyphs by fonts that supports Tamil.
-Third, \textsc{a single glyph may be a combination of multiple characters.
-}For example, the ligature <ﬁ>, a single glyph, is the result of two characters, <f> and <i>, that have undergone glyph substitution by font rendering.
-Like contextually-determined glyphs, ligatures are (intended) artifacts of text processing instructions.
-Finally, \textsc{a single glyph may be a part of a character} (e.g.~diacritics).
-
-Further, the rendering of a glyph is also always dependent on the font being used.
-For example, the Unicode character LATIN SMALL LETTER G appears as and in the Computer Modern and Courier fonts because their typefaces are designed differently.
-Furthermore, font face may change the visual appearance of a character, for example Times New Roman two-story <a> changes to a single-story a in italics <\emph{a}>.
-This becomes a real problem for some phonetic typesetting (see Section 5, Pitfall 2).
-
-In sum, character-to-glyph mappings are complex technological issues that the Unicode Consortium has had to address in the development of the Unicode Standard, but for the lay user they can be utterly confusing because visual rendering does not indicate logical encoding, or, in other words, what you see isn't (necessarily) what you get.
-
-\subsection{Pitfall: File formats}\label{pitfall-file-formats}
-
-Unicode is a character encoding standard, but these characters of course actually appear inside some kind of file.
-The most basic Unicode-based file format is pure line-based text, i.e.~strings of Unicode-encoded characters separated by line breaks (note that these line breaks are what for most people intuitively corresponds to paragraph breaks).
-Unfortunately, even within this apparently basic setting there exist a multitude of variants.
-In general, these different possibilities are well-understood in the software industry, and nowadays they normally do not lead to any problems for the end user.
-However, there are some situations in which a user is suddenly confronted with cryptic questions in the user interface involving abbreviations like LF, CR, BE, LE or BOM.
-Most prominently this occurs with exporting or importing data in several software applications from Microsoft.
-Basically, there are two different issues involved.
-First, the encoding of** line breaks** and, second, the encoding of the Unicode characters into \textsc{code units} and the related issue of \textsc{endianness}.
-
-The issue with \textsc{line breaks} originated with the instructions necessary to direct a printing head to a new line.
-This involves two movements, known as \textsc{carriage return (CR)} (returning the printing head to the start of the line on the page) and \textsc{line feed (LF)} (moving the printing head to the next line on the page).
-Physically, these are two different events, but conceptually together they form one action.
-In the history of computing, various encodings of line breaks have been used (e.g.~CR+LF, LF+CR, only LF, or only CR).
-Currently, all Unix and Unix-derived systems use only LF as code for a line break, while software from Microsoft still uses a combination of CR and LF.
-Today, most software applications recognize both options, and are able to deal with either encoding of line breaks (until rather recently this was not the case, and using the wrong line breaks could lead to unexpected errors).
-Our impression is that there is a strong tendency in software development to standardize on the simpler ``only LF'' encoding for line breaks, and we suggest that everybody use this encoding whenever possible.
-
-A code unit is a sequence of bits used to encode character in a encoding.
-The issue with code units stems from the question how to separate a stream of binary ones and zero, i.e.~bits, into chunks representing Unicode characters.
-Depending on different use cases, the Unicode Standard offers three different approaches, called UTF-32, UTF-16 and UTF-8.\footnote{The letters UTF stand for either ``Unicode Transformation Format'' or ``Universal Coded Character Set Transformation Format'', but the notion of `transformation' is a legacy notion that does not have meaning anymore.
-Nevertheless, the designation ``UTF'' (in capitals) has become an official standard designation, but should probably best be read as simply ``Unicode Format''.) The details of this issue is extensively explained in section 2.5 of the Unicode core specification.
-Basically, \textsc{UTF-32} encodes each character in 32 bits (32 `binary units', i.e.~32 zeros or ones) and is the most disk-space-consuming variant of the three.
-However, it is the most efficient encoding processing-wise, because the computer simply has to separate each character after 32 bits.
-In contrast, \textsc{UTF-16} uses only 16 bits per character, which is sufficient for the large majority of Unicode characters, but not for all of them.
-A special system of `surrogates' is defined within the Unicode Standard to deal with these additional characters.
-The effect is a more disk-space efficient encoding (approximately half the size), while adding a limited computational overhead.
-Finally, \textsc{UTF-8} is a more complex system that dynamically encodes each character with the minimally necessary number of bits, choosing either 8, 16 or 32 bits depending on the character.
-This represents again a strong reduction in space (particularly due to the high frequency of data using ASCII characters that need only 8 bits) at the expense of even more computation necessary to process such strings.
-However, because of the ever growing computational power of modern machines, the processing overhead is in most practical situations a non-issue, while saving on space is still useful, particularly for sending texts over the Internet.
-As an effect, UTF-8 has become the dominant encoding on the World Wide Web.
-We suggest that everybody uses UTF-8 as their default encoding.
-
-A related problem is a general issue about how to store information in computer memory, which is known as \textsc{endianness}.
-The details of this issue go beyond the scope of this paper.
-It suffices to realize that there is a difference between \textsc{big-endian} (BE) storage and \textsc{little-endian} (LE) storage.
-The Unicode Standard offers a possibility to explicitly indicate what kind of storage is used by starting a file with a so-called `Byte Order Mark' (BOM).
-However, the Unicode Standard does not require the usage of BOM, preferring other non-Unicode methods to signal to computers which kind of endianness is used.
-This issue only arises with UTF-32 and UTF-16 encodings, and when using the preferred UTF-8, using a BOM is theoretically possible, but strongly dispreferred according to the Unicode Standard.
-We suggest that everyone tries to prevent the inclusion of BOM in your data.
-
-Together with the topic of normalisation (NFC vs.~NFD, see Pitfall 5 above), this leads to the following nicely cryptic file-encoding recommendation, namely to preferably use \textsc{UTF-8 in NFC without BOM with LF line breaks}, whenever possible.
-
-\section{5. IPA meets Unicode}\label{ipa-meets-unicode}
-
-\subsection{Introducing the International Phonetic Alphabet (IPA)}\label{introducing-the-international-phonetic-alphabet-ipa}
-
-The International Phonetic Alphabet (IPA) is a common standard in linguistics to transcribe sounds of spoken language into some Latin-based characters (International Phonetic Association 1999).
-Although IPA is reasonably easily adhered to with pen and paper, it is not trivial to encode IPA characters electronically.
-Early work addressing the need for a universal computing environment for writing systems and their computational complexity is discussed in Simons 1989.
-For a long time, linguists (like all other computer users) were limited to ASCII-encoded 7-bit characters, which only includes Latin characters, numbers and some punctuation and symbols.
-Restricted to these standard character sets that lacked IPA support or other language-specific graphemes that they needed, linguists devised their own solutions (cf.~Bird and Simons 2003).
-For example, some chose to represent unavailable graphemes with substitutes, e.g.~the combination of to represent .
-Tech-savvy linguists redefined selected characters from a character encoding by mapping custom made fonts to those code points.
-However, one linguist's electronic text would not render properly on another linguist's computer without access to the same font.
-Further, if two character encodings defined two character sets differently, then data could not be reliably and correctly displayed.
-This is a common example of the non-interoperability of data and data formats.
-
-To alleviate this problem, during the late 1980s, SAMPA (Speech Assessment Methods Phonetic Alphabet) was created to represent IPA symbols with 7-bit printable ASCII character sequences, e.g.
-for {[}ɸ{]}.
-Two problems with SAMPA are that (i) it is only a partial encoding of the IPA and (ii) it encodes different languages in separate data tables, instead of a universal alphabet, like IPA.
-SAMPA tables are derived from phonemes appearing in several European languages that were developed as part of a European Commission-funded project to address technical problems like electronic mail exchange (what is now simply called email).
-SAMPA is essentially a hack to work around displaying IPA characters, but it provided speech technology and other fields a basis that has been widely adopted and used in code.
-So, SAMPA was a collection of tables to be compared, instead of a large universal table representing all languages.
-An extended version of SAMPA, called X-SAMPA, set out to include every symbol in the IPA chart including all diacritics (Wells nd.).
-X-SAMPA was considered more universally applicable because it consisted of one table that encoded the set of characters that represented phones/segments in IPA across languages.
-SAMPA and X-SAMPA have been widely used for speech technology and as an encoding system in computational linguistics.
-Eventually, ASCII-encoding of the IPA became depreciated through the advent of the Unicode Standard.
-Note however that many popular software packages used for linguistic analyses still require ASCII input, e.g.~RuG/L04\footnote{http://www.let.rug.nl/kleiweg/L04/) and SplitsTree4.\footnote{http://www.splitstree.org/)
-
-Still, there are a few pitfalls to be aware of when using the Unicode Standard to encode IPA.
-As we have said before, from a linguistic perspective it might look like the Unicode Consortium is making incomprehensible decisions, but it is important to realize that the consortium has tried and is continuing to try to be as consistent as possible across a wide range of use cases, and it does place linguistic traditions above other orthographic possibilities.
-In general, we strongly suggest linguists not to complain about any decisions in the Unicode Standard, but to try and understand the rationale of the Unicode Consortium (which in our experience is almost always well-conceived) and devise ways to work with any unexpected behaviour.
-Many of the current problems derive from the fact that the IPA is clearly historically based on the Latin script, but different enough from most other Latin-based writing systems to warrant special attention.
-This ambivalent status of the IPA glyphs (partly Latin, partly special) is unfortunately also attested in the treatment of IPA in the Unicode Standard.
-In retrospect, it might have been better to consider the IPA (and other transcription systems) to be a special or new kind of script within the Unicode Standard, and treat the obvious similarity to Latin glyphs as a historical relic.
-All IPA glyphs would then have their own code points, instead of the current situation in which some IPA glyphs have special code points, while others are treated as being identical to the `regular' Latin characters.
-Yet, the current situation, however unfortunate, is unlikely to change, so as linguists we will have to learn to deal with the specific pitfalls of IPA within the Unicode Standard.
-In this section, we will describe these pitfalls in some detail.
-
-\subsection{Pitfall: There is no single complete Unicode code block for IPA}\label{pitfall-there-is-no-single-complete-unicode-code-block-for-ipa}
-
-The ambivalent nature of IPA glyphs arises because, on the one hand, the IPA uses Latin-based glyphs like , or .
-From this perspective, the IPA seems to be just another orthographic tradition using Latin characters, all of which do not get a special treatment within the Unicode Standard (just like e.g.~the French, German, or Danish orthographic traditions do not have a special status).
-On the other hand, the IPA uses many special symbols (like turned , mirrored and/or extended Latin glyphs ) not found in any other Latin-based writing system.
-For this reason, and already in the first version of the Unicode Standard (Version 1.0 from 1991), a special block with code points, called ``IPA Extensions'' was included.
-As noted in Section 4, Pitfall 1, the Unicode Standard code space is subdivided into character blocks, which generally encode characters from a single script.
-However, as is illustrated by the IPA, characters that form a single writing system may be dispersed across several different character blocks.
-With its diverse collection of symbols from various scripts and diacritics, the IPA is spread across over 13 blocks in the Unicode Standard:\footnote{This number of blocks depends on whether only IPA-sanctioned symbols are counted or if the phonetic symbols commonly found in the literature are also included, see Moran 2012, Appendix C.)
-\begin{itemize}
-	\item Basic Latin (30 characters), e.g. <a, b, c, d, e> 
-	\item Latin-1 Supplement (4 characters): <æ, ç, ð, ø\textgreater{} 
-	\item Latin Extended-A (3 characters): <ħ, ŋ, œ> 
-	\item Latin Extended-B (5 characters): <ǀ, ǁ, ǂ, ǃ, ȵ> 
-	\item IPA Extensions (70 characters), e.g.: <ɐ, ɑ, ɔ> 
-	\item Spacing Modifier Letters (20 characters), e.g.: <ʰ ʷ ˥> 
-	\item Combining Diacritical Marks (33 characters), e.g.: <̝ ̥ ̪> 
-	\item Greek and Coptic (3 characters): <β, θ, χ> 
-	\item Phonetic Extensions (2 characters): <ᴅ, ᴴ> 
-	\item Phonetic Extensions Supplement (3 characters): <ᶑ , ᶾ, ᶣ> 
-	\item Superscripts and Subscripts (1 character): <ⁿ> 
-	\item Arrows (2 characters): \textless{}↑, ↓\textgreater{} 
-	\item Latin Extended-C (1 character): <ⱱ> 
-\end{itemize}
-
-\subsection{Pitfall: There are many IPA homoglyphs in Unicode}\label{pitfall-there-are-many-ipa-homoglyphs-in-unicode}
-
-Another problem is the large number of homoglyphs, i.e.~different characters that have highly similar glyphs (or even completely identical, depending on the font rendering).
-For example, a speaker of Russian should ideally not use the CYRILLIC SMALL LETTER A at code point U+0430 for IPA transcriptions, but instead the LATIN SMALL LETTER A at code point U+0061, although visually they are mostly indistinguishable, and the Cyrillic character is more easily typed on a Cyrillic keyboard.
-Another example we commonly encounter is the use of LATIN SMALL LETTER G at U+0067, instead of the sanctioned Unicode Standard IPA character for the voiced velar stop LATIN SMALL LETTER SCRIPT G at U+0261.
-One begins to question whether this issue is at all apparent to the working linguist, or if they simply use the U+0067 because it is easily keyboarded and thus saves time, whereas the latter must be cumbersomely inserted as a special symbol in most software.\footnote{This issue was recently addressed by the International Phonetic Association, which has taken the stance that both the keyboard LATIN SMALL LETTER G and the LATIN SMALL LETTER SCRIPT G are valid input characters for the voiced velar plosive.
-Unfortunately, this decision further introduces ambiguity for linguists trying to adhere to a strict Unicode Standard IPA encoding.)
-
-Furthermore, on the one hand even linguists are unlikely to distinguish between the LATIN SMALL LETTER SCHWA at code point U+0259 and LATIN SMALL LETTER TURNED E at U+01DD.
-On the other hand, non-linguists are unlikely to distinguish any semantic difference between an open back unrounded vowel , the LATIN SMALL LETTER ALPHA at U+0251, and the open front unrounded vowel , LATIN SMALL LETTER A at U+0061.
-But even among linguists this distinction leads to problems.
-For example, as pointed out by Mielke (2009), there is a problem stemming from the fact that about 75\% of languages are reported to have a five-vowel system (Maddieson 1984).
-Historically, linguistic descriptions tend not to include precise audio recording and measurements of formants, so this may lead one to ask if the many characters that are used in phonological description reflects a ``transcriptional bias''.
-The common use of in transcriptions could be in part due to the ease of typing the letter on an English keyboard (or for older descriptions, the typewriter).
-We found it to be exceedingly rare that a linguist uses for a low back unrounded vowel.\footnote{One example is Vidal 2001a:75, in which the author states: ``The definition of Pilagá /a/ as {[}+back{]} results from its behavior in certain phonological contexts.
-For instance, uvular and pharyngeal consonants only occur around /a/ and /o/.
-Hence, the characterization of /a/ and /o/ as a natural class of (i.e., {[}+back{]} vowels), as opposed to /i/ and /e/.'') They simply use as long as there is no opposition to .\footnote{See Thomason's Language Log post, ``Why I don't love the International Phonetic Alphabet'', at: http://itre.cis.upenn.edu/\textasciitilde{}myl/languagelog/archives/005287.html.) Making things even more problematic, there is an old typographic tradition that the double-story uses a single-story in italics.
-This leads to the unfortunate effect that in most well-designed fonts the italics of and use the same glyph.
-If this distinction has to be kept upright in italics, the only solution we can currently offer is to use `slanted' glyphs (i.e.~artificially italicised glyphs) instead of real italics (i.e.~special italics glyphs designed by a typographer).\footnote{For example, the widely used IPA font Doulos SIL (http://scripts.sil.org/cms/scripts/page.php?item\_id=DoulosSIL) does not have real italics.
-This leads some word-processing software, like Microsoft Word, to produce slanted glyphs instead.
-That particular combination of font and software application will thus lead to the desired effect.
-However, note that when the text is transferred to another font (i.e.~one that includes real italics) and/or to another software application (like Apple Pages, which does not perform slanting), then this visual appearance will be lost.
-In this case we are thus still in the pre-Unicode situation in which the choice of font and rendering software actually matters.
-The ideal solution from a linguistic point of view would be the introduction of a new IPA code point for a different kind of which explicitly specifies that it should still be rendered as a double-story character when italicized.
-After informal discussion with various Unicode players, our impression is that this highly restricted problem is not sufficiently urgent to introduce even more -like characters in Unicode (which already lead to much confusion, see Section 4, Pitfall 4).
-This is a clear situation in which the Unicode Consortium is not just thinking about linguists, but has a more wide-ranging practical view to consider.)
-
-Some other homoglyphs related to encoding IPA in the Unicode Standard are:
-\begin{itemize}
-	\item The uses of the apostrophe has led to long discussions on the Unicode Standard email list.
-An English keyboard inputs \textless{}`\textgreater{} APOSTROPHE at U+0027, although the `preferred' Unicode apostrophe is the \textless{}'\textgreater{} RIGHT SINGLE QUOTATION MARK at U+2019.
-Yet the glottal stop/glottalization/ejective marker is another completely different character, the MODIFIER LETTER APOSTROPHE at U+02BC, but unfortunately looks mostly highly similar to U+2019.
-	\item There is ambiguous encoding of IPA segments within the Unicode Standard.
-An example is the U+02C1 MODIFIER LETTER REVERSED GLOTTAL STOP vs the U+02E4 MODIFIER LETTER SMALL REVERSED GLOTTAL STOP .
-Both are denoted in the Unicode Standard as the `pharyngealized diacritic' and both appear in various resources representing phonetic data online.
-This is thus an example for which the Unicode Standard does not solve the linguistic standardization problem.
-	\item There is at least one case in which the character name assigned by the Unicode Consortium does not match the IPA's description: in the Unicode Standard at U+01C3 is labeled LATIN LETTER RETROFLEX CLICK, but in IPA is an alveolar or postalveolar click (not retroflex).
-This naming is probably best seen as a simple error in the Unicode Standard.
-Note that most linguists simply seem to use or <ou>) while ligatures are single characters (e.g.
-<ʧ> LATIN SMALL LETTER TESH DIGRAPH at U+02A7).
-Ligatures arose in the context of printing easier-to-read texts, and are included in the Unicode Standard for reasons of legacy encoding.
-However, their usage is discouraged by the Unicode core specification.
-Specifically related to IPA, various phonetic combinations of characters (typically affricates) are available as single code-points in the Unicode Standard, but are designated as ``ligatures'' or ``digraphs'' (confusingly both names appear interchangeably).
-Such glyphs might be used by software to produce a pleasing display, but they should not be hard-coded into the text itself.
-In the context of IPA, characters like the following ligatures should thus \emph{not} be used.
-Instead a combination of two characters is preferred:
-	\item <ʣ> LATIN SMALL LETTER DZ DIGRAPH at U+02A3 (use <d> + <z> instead) 
-	\item <ʧ> LATIN SMALL LETTER TESH DIGRAPH at U+02A7 (use <t> + <ʃ> instead) 
-	\item <ʩ> LATIN SMALL LETTER FENG DIGRAPH at U+02A9 (use <f> + <ŋ> instead) 
-\end{itemize}
-
-However, there are a few Unicode characters that are historically ligatures, but which are today considered as simple characters in the Unicode Standard and thus should be used when writing IPA, namely:
-\begin{itemize}
-	\item <ɮ> LATIN SMALL LETTER LEZH at U+026E 
-	\item <œ> LATIN SMALL LIGATURE OE at U+0153 
-	\item <ɶ> LATIN LETTER SMALL CAPITAL OE at U+0276 
-	\item <æ> LATIN SMALL LETTER AE at U+00E6 
-\end{itemize}
-
-\subsection{Pitfall: The IPA notion of diacritics is not the same as the Unicode Standard's notion of diacritics}\label{pitfall-the-ipa-notion-of-diacritics-is-not-the-same-as-the-unicode-standards-notion-of-diacritics}
-
-Another pitfall is diacritics.
-The problem is that the meaning of the term `diacritics' as used by the IPA is not the same as it used in the Unicode Standard.
-Specifically, diacritics in the IPA-sense are either so-called \textsc{Combining Diacritical Marks} or \textsc{Spacing Modifier Letters} in the Unicode Standard.
-Crucially, Combining Diacritical Marks are by definition combined with the character before them (to form so-called default grapheme clusters, see Section 3).
-In contrast, Spacing Modifier Letters are by definition \emph{not} combined into grapheme clusters with the preceding character, but simply treated as separate letters.
-In the context of the IPA, the following IPA-diacritics are actually Spacing Modifier Letters in the Unicode Standard:
-\begin{itemize}
-	\item Length marks, namely <ː> MODIFIER LETTER TRIANGULAR COLON at U+02D0 and <ˑ> MODIFIER LETTER HALF TRIANGULAR COLON at U+02D1. 
-	\item Tone letters, like <˥> MODIFIER LETTER EXTRA-HIGH TONE BAR at U+02E5, and others like this. 
-	\item Superscript letters, like <ʰ> MODIFIER LETTER SMALL H at U+02B0 or <ˤ> MODIFIER LETTER SMALL REVERSED GLOTTAL STOP at U+02E4, and others like this. 
-	\item <˞> MODIFIER LETTER RHOTIC HOOK at U+02DE. 
-\end{itemize}
-
-Although linguists might expect these characters to belong together with the character in front of them, at least for <ʰ> MODIFIER LETTER SMALL H at U+02B0 the Unicode Consortium's decision to treat it as a separate character is also linguistically correct, because according to the IPA it can be used both for aspiration (more precisely postaspiration following the base character) and preaspiration (preceding the base character).
-Note that there is a mechanism in Unicode to force separate characters to be combined (namely by using the ZERO WIDTH JOINER at U+200D), but this seems to be a rather impractical, and probably not enforceable solution to us.
-
-\subsection{Pitfall: Neither the IPA nor the Unicode Standard enforce a unique diacritic ordering}\label{pitfall-neither-the-ipa-nor-the-unicode-standard-enforce-a-unique-diacritic-ordering}
-
-Also related to diacritics is the question of ordering.
-To our knowledge, the International Phonetic Association does not specify a specific ordering for diacritics that combine with phonetic base symbols; this exercise is left to the reasoning of the transcriber.
-However, such marks have to be explicitly ordered if sequences of them are to be interoperable and compatible.
-An example is a labialized aspirated alveolar plosive: <tʷʰ>.
-There is nothing holding linguists back from using <tʰʷ> instead (with exactly the same intended meaning).
-However, from a technical standpoint, these two sequences are different, e.g.~if both sequences are used in a document, searching for <tʷʰ> will not find any instances of <tʰʷ>, and vice versa.
-Likewise, a creaky voiced syllabic dental nasal can be encoded in various orders, e.g.
-<n̪̰̩>, <n̩̰̪> or <n̩̪̰>.
-
-In accordance with the absence of any specification of ordering in the IPA, the Unicode Standard likewise does not propose any standard orderings.
-Both leave it to the user to be consistent.
-However, there is one aspect of ordering for which the Unicode Standard does present a canonical solution, which is uncontroversial from a linguistic perspective.
-Diacritics in the Unicode Standard (i.e.~Combining Diacritical Marks, see above) are classified in Canonical Combining Classes.
-In practice, the diacritics are distinguished by their position relative to the base character.\footnote{See http://unicode.org/reports/tr44/\#Canonical\_Combining\_Class\_Values for a detailed description.) When applying a Unicode normalization (NFC or NFD, see previous section), the diacritics in different positions are put in a specified order.
-This process therefore harmonizes the difference between different encodings, for example, of a midtone creaky voice vowel <ḛ̄>.
-This grapheme cluster can be encoded either as <e> + <̄> + <̰> or as <e> + <̰> + <̄> .
-To prevent this twofold encoding, the Unicode Standard specifies the second ordering as canonical (diacritics below before diacritics above).
-
-When encoding a string according to the Unicode Standard, it is possible to do this either using the NFC (``composition'') or NFD (``decomposition'') normalization.
-Decomposition implies that precomposed characters (like <á> LATIN SMALL LETTER A WITH ACUTE at U+00E1) will be split into its parts.
-This might sound preferable for a linguistic analysis, as the different diacritics are separated from the base characters.
-However, note that most attached elements like strokes (e.g.~in the <ɨ>), retroflex hooks (e.g.~in <ʐ>) or rhotic hooks (e.g.~in <ɝ>) will not be decomposed, but strangely enough a cedilla (like in <ç>) will be decomposed.
-In general, Unicode decomposition does not behave like a feature decomposition as expected from a linguistic perspective.
-It is thus important to consider Unicode decomposition only as a technical procedure, and not assume that it is linguistically sensible.
-
-Facing the problem of specifying a consistent ordering of diacritics while developing a large database of phonological inventories from the world's languages, Moran (2012: 540) defines a set of diacritic ordering conventions.\footnote{The most recent version of these conventions is online: http://phoible.github.io/conventions/) The conventions are influenced by the linguistic literature, though some ad-hoc decisions had to be taken.
-The goal was to explicitly define all character sequences so that the vast variety of phonemes found in descriptions of the world's language were normalized into consistent character sequences, e.g.~if one language description uses and another , when both are intended to be phonetically equivalent, then a decision to normalize to one form was taken.
-For example, when a character sequence contains more than one character in Spacing Modifier Letters, the order that is proposed is the following (where <\textbar{}> indicates ``or'' and <→> indicates ``precedes''):
-\begin{itemize}
-	\itemsep1pt\parskip0pt\parsep0pt 
-	\item \textsc{Spacing Modifier Letters ordering:} ( unreleased <̚> \textbar{} lateral release <ˡ> \textbar{} nasal release <ⁿ>) → ( palatalized <ʲ>) → ( labialized <ʷ>) → ( velarized <ˠ>) → ( pharyngealized <ˤ>) → ( aspirated <ʰ> \textbar{} ejective <ʼ>) → ( long <ː> \textbar{} half-long <ˑ>)
-\end{itemize}
-
-If a character sequence contains more than one diacritic below the base character, then the place feature is applied first (dental, laminal, apical, fronted, backed, lowered, raised), then the laryngeal setting (voiced, voiceless, creaky voice, breathy voice) and finally the syllabic or non-syllabic marker (for vowels, ATR gets put on between the place and laryngeal setting).
-So:
-\begin{itemize}
-	\itemsep1pt\parskip0pt\parsep0pt 
-	\item \textsc{Combining Diacritical Marks (below) ordering:} ( dental <t̪> \textbar{} laminal <t̻> \textbar{} apical <t̺>) → ( fronted <u̟> \textbar{} backed <e̠>) →( lowered <e̞> \textbar{} raised <e̝>) → ( ATR <e̘ e̙>) → ( voiced <s̬> \textbar{} voiceless <n̥> \textbar{} creaky voice <b̰> \textbar{} breathy voice <b̤>) → ( syllabic <n̩> \textbar{} non-syllabic <e̯>)
-\end{itemize}
-
-Character sequences with diacritics above the base character were not problematic in Moran 2012 because they include only the centralized, mid-centralized and nasalized combining characters.
-Moran (2012) marks tones as singletons with Space Modifier Letters, e.g.
-\textless{}˦\textgreater{} for a phonemic high tone, instead of accent diacritics, alleviating potential conflicts.
-Building on the work of Moran (2012), if a character sequence contains more than one diacritic above the base character, we propose:
-\begin{itemize}
-	\itemsep1pt\parskip0pt\parsep0pt 
-	\item \textsc{Combining Diacritical Marks (above) ordering:} (centralized <ë> \textbar{} mid-centralized <e̽>) → (extra short <ĕ>) →( tone accents, e.g.
-<è> ) → ( Spacing Modifier Letters ) → ( tone letters, e.g.
-<e˦>)
-\end{itemize}
-
-\section{6. Orthography profiles}\label{orthography-profiles}
-
-\subsection{Characterizing writing systems}\label{characterizing-writing-systems}
-
-At this point in the course of rapid ongoing developments, we are left with a situation in which the Unicode Standard offers a highly detailed and flexible approach to deal computationally with writing systems, but it has unfortunately not influenced the linguistic practice very much.
-In many practical situations, the Unicode Standard is far too complex for the day-to-day practice in linguistics because it does not offer practical solutions for the down-to-earth problems of many linguists.
-In this section, we propose some simple practical guidelines and methods to improve on this situation.
-
-Our central aims for linguistics, to be approached with a Unicode-based solution, are: (i) to improve the consistency of the encoding of sources, (ii) to transparently document knowledge about the writing system (including transliteration), and (iii) to do all of that in a way that is easy and quick to manage for many different sources with many different writing systems.
-The central concept in our proposal is the \textsc{orthography profile}, a simple tab-separated CSV text file, that characterizes and documents a writing system.
-We also offer basic implementations in Python and R to assist with the production of such files, and to apply orthography profiles for consistency testing, grapheme tokenization and transliteration.
-Not only can orthography profiles be helpful in the daily practice of linguistics, they also succinctly document the orthographic details of a specific source, and, as such, might fruitfully be published alongside sources (e.g.~in digital archives).
-Also, in high-level linguistic analyses in which the graphemic detail is of central importance (e.g.~phonotactic or comparative-historical studies), orthography profiles can transparently document the decisions that have been taken in the interpretation of the orthography in the sources used.
-
-Given these goals, Unicode locale descriptions (see Section 3) might seem like the ideal orthography profiles.
-However, there are various practical obstacles preventing the use of such locale descriptions in the daily linguistic practice, namely: (i) the XML-structure is too verbose to easily and quickly produce or correct manually, (ii) locale descriptions are designed for a wide scope on information (like date formats or names of weekdays) most of which is not applicable for documenting writing systems, and (iii) most crucially, even if someone made the effort to produce a technically correct locale description for a specific source at hand, then it is nigh impossible to deploy the description.
-This is because a locale description has to be submitted to and accepted by the Unicode Common Locale Data Repository.
-The repository is (rightly so) not interested in descriptions that only apply to a limited set of sources (e.g.~only a single specific dictionary).
-
-The major challenge then is developing an infrastructure to identify the elements that are individual graphemes in a source, specifically for the enormous variety of sources using some kind of alphabetic writing system.
-Authors of source documents (e.g.~dictionaries, wordlists, corpora) use a variety of writing systems that range from their own idiosyncratic transcriptions to already well-established practical or longstanding orthographies.
-Although the IPA is one practical choice as a sound-based normalization for writing systems (which can act as an interlingual pivot to attain interoperability across writing systems), graphemes in each writing system must also be identified and standardized if interoperability across different sources is to be achieved.
-In most cases, this amounts to more than simply mapping a grapheme to an IPA segment because graphemes must first be identified in context (e.g.~is the sequence one sound or two sounds or both?) and strings must be tokenized, which may include taking orthographic rules into account (e.g.
-between vowels is /n/ and after a vowel but before a consonant is a nasalized vowel /ṽ/).
-In our experience, data from each source must be individually tokenized into graphemes so that its orthographic structure is identified and its contents can be extracted.
-To extract data for analysis, a source-by-source approach is required before an orthography profile can be created.
-For example, almost each available lexicon on the world's languages is idiosyncratic in its orthography and thus requires lexicon-specific approaches to identify graphemes in the writing system and to map graphemes to phonemes, if desired.
-
-Thus, our key proposal for the characterization of a writing system is to use a grapheme tokenization as an inter-orthographic pivot.
-Basically, any source document is tokenized by graphemes, and only then a mapping to IPA (or any other orthographic conversion) is performed.
-An orthography profile then is a description of the units and rules that are needed to adequately model a graphemic tokenization for a language variety as described in a particular source document.
-An orthography profile summarizes the Unicode (tailored) graphemes and orthographic rules used to write a language (the details of the structure and assumptions of such a profile will be presented in the next section).
-
-As an example of graphemic tokenization, note the three different levels of technological and linguistic elements that interact in the hypothetical lexical form <tsʰṍ̰shi> :
-\begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item code points (10 text elements): t s ʰ o ̃ ̰ ´ s h i 
-	\item grapheme clusters (7 text elements): t s ʰ ṍ̰ s h i 
-	\item tailored grapheme clusters (4 text elements): tsʰ ṍ̰ sh i 
-\end{enumerate}
-
-In (1), the string <tsʰṍ̰shi> has been tokenized into ten Unicode code points (using NFD normalization), delimited here by space.
-Unicode tokenization is required because sequences of code points can differ in their visual and logical orders.
-For example, <õ̰> is ambiguous to whether it is the sequence of + <̰> + <̃> or + <̃> + <̰>.
-Although these two variants are visually homoglyphs, computationally they are different.
-Unicode normalization should be applied to this string to reorder the code points into a canonical order, allowing the data to be treated canonically equivalently for search and comparison.
-In (2), the Unicode code points have been logically normalized and visually organized into grapheme clusters, as specified by the Unicode Standard.
-The combining character sequence <õ̰> is normalized and visually grouped together.
-Note that , the MODIFIER LETTER SMALL H at U+02B0, is not grouped with .
-This is because it belongs to Spacing Modifier Letters category in the Unicode Standard.
-These characters are underspecified for the direction in which they modify a host character.
-For example, can indicate either pre- or post-aspiration (whereas the nasalization or creaky diacritic is defined in the Unicode Standard to apply to a specified base character).
-Finally, to arrive at the graphemic tokenization in (3), tailored grapheme clusters are needed (as for example specified in an orthography profile).
-For example, this orthography profile would specify that the sequence of characters , and form a single grapheme , and that and form a grapheme .
-The orthography profile could also specify orthographic rules, e.g.~when tokenization graphemes, in say English, the in the forms and should be treated as distinct sequences depending on their contexts.
-
-\subsection{Informal description of orthography profiles}\label{informal-description-of-orthography-profiles}
-
-An orthography profile describes the Unicode code points, characters, graphemes and orthographic rules in a writing system.
-An orthography profile is a language-specific (and often even resource-specific) description of the units and rules that are needed to adequately model a writing system.
-An important assumption of our work is that we assume a resource is encoded in Unicode (or has been converted to Unicode).
-Any data source that the Unicode Standard is unable to capture, will also not be captured by an orthography profile.
-
-Informally, an orthography profile specifies the graphemes (or, in Unicode parlance, ``tailored grapheme clusters'') that are expected to occur in any data to be analysed or checked for consistency.
-These graphemes are first identified throughout the whole data (a step which we call `tokenization'), and possibly simply returned as such, possibly including error messages about any parts of the data that are not specified by the orthography profile.
-Once the graphemes are identified, they might also be changed into other graphemes (a step which we call `transliteration').
-When a grapheme has different possible transliterations, then these differences should be separated by contextual specification, possibly down to listing individual exceptional cases.
-
-In practice, we foresee a workflow in which orthography profiles are iteratively refined, while at the same time inconsistencies and errors in the data to be tokenized are corrected.
-In some more complex use-cases there might even be a need for multiple different orthography profiles to be applied in sequence (see Section 8 on various exemplary use-cases).
-The result of any such workflow will normally be a cleaned dataset and an explicit description of the orthographic structure in the form of an orthography profile.
-Subsequently, the orthography profiles can be easily distributed in scholarly channels alongside the cleaned data, for example in supplementary material added to journal papers or in electronic archives.
-
-\subsection{Formal specification of orthography profiles}\label{formal-specification-of-orthography-profiles}
-
-The formal specifications of an orthography profile (or simply `profile' for short) are the following:
-\begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item \textsc{A profile is a} \textsc{Unicode UTF-8 encoded text file} (ideally using NFC, no-BOM, and LF; see Section 4, Pitfall ``File Formats'') that includes the information pertinent to the orthography.
-	\item \textsc{A profile is a} \textsc{tab-separated CSV file with an obligatory header line}.
-A minimal profile can have just a single column, in which case there will of course no tabs, but the first line will still be the header.
-For all columns we assume the name in the header of the CSV file to be crucial.
-The actual ordering of the columns is unimportant.
-	\item \textsc{Lines starting with a hash \textless{}\#\textgreater{} are ignored.} Comments and metadata can be included inside the file, but only as complete lines in the profile, to be marked by lines starting with hash ``\#'' (NUMBER SIGN at U+0023).
-Hashes somewhere else in the file are to be treated literally, i.e.~hashes are only to be ignored when they occur at the start of a line.\footnote{Comments that belong to specific lines will have to be put in a separate column of the CSV file, e.g.~add a column called `comments'.
-Further, if the content of a profile contains a hash at the start of a line, either reorder the columns so the hash does not occur at the start of the line, or add a dummy column in front of the data to not have the data start with a hash.)
-	\item \textsc{Metadata are given in commented lines at the beginning of the text file in a basic ``tag: value'' format.
-}Metadata about the orthographic description given in the orthography profile includes, minimally, (i) author, (ii) date, (iii) title, (iv) a stable language identifier encoded in BCP 47/ISO 639-3, and (v) bibliographic data for resource(s) that illustrate the orthography described in the profile.
-\end{enumerate}
-
-The content of a profile consists of lines, each describing a grapheme of the orthography, using the following columns:
-\begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item \textsc{A minimal profile consists of a single column with header `graphemes'}, listing each of the different graphemes in a separate line. 
-	\item \textsc{Optional columns called `left' and `right' can be used to specify the left and right context of the grapheme, respectively.} The same grapheme can occur multiple times with different contextual specifications, for example to distinguish different pronunciations depending on the context.
-	\item \textsc{The columns `grapheme', `left' and `right' can use regular expression metacharacters.} If regular expressions are used, then all literal usage of the special symbols, like full stops ``.'' or dollar signs ``\$'' (so-called ``metacharacters'') have to be explicitly escaped by adding a backslash before them (i.e.~use ``.'' or ``\$'').
-Note that any specification of context automatically expects regular expressions, so it is probably better to always escape all regular expression metacharacters when used literally in the orthography, i.e.~the following symbols will need to be preceded by a backslash: {[} {]} ( ) \{ \} ~+ * .
-- !
-?
-\^{} \$ .
-	\item \textsc{An optional column called `class' can be used to specify classes of graphemes}, for example to define a class of vowels.
-Users can simply add ad-hoc identifiers in this column to indicate a group of graphemes, which can then be used in the description of the graphemes or the context.
-The identifiers should of course be chosen such that they do not conflate with any symbols used in the orthography themselves.
-Note that such classes only refer to the graphemes, not to the context.
-	\item \textsc{Columns describing transliterations for each graphemes can be added and named at will}.
-Often more than a single possible transliteration will be of interest.
-Any software application using these profiles should use the names of these columns to select a specific transliteration column.
-	\item \textsc{Any other columns can be added freely, but will mostly be ignored by any software application using the profiles}.
-As orthography profiles are also intended to be read and interpreted by humans, it is often highly useful to add extra information on the graphemes in further columns, like for example Unicode codepoints, Unicode names, frequency of occurrence, examples of occurrence, explanation of the contextual restrictions, or comments.
-\end{enumerate}
-
-For the automatic processing of the profiles, the following technical standards will be expected:
-\begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item \textsc{Each line of a profile will be interpreted as a regular expression.
-}Software applications using profiles can also offer to interpret a profile in the literal sense to avoid the necessity for the user to escape regular expressions metacharacters in the profile.
-However, this only is possible when no contexts or classes are described, so this seems only useful in the most basic orthographies.
-	\item \textsc{The `class' column will be used to produce explicit ``or'' chains of regular expressions}, which will then be inserted in the `graphemes', `left' and `right' columns at the position indicated by the class-identifiers.
-For example, a class ``V'' as a context specification might be replaced by a regular expression like ``(a\textbar{}e\textbar{}i\textbar{}o\textbar{}u\textbar{}ei\textbar{}au)''.
-Only the graphemes themselves are included here, not any contexts specified for the elements of the class.
-	\item \textsc{The `left' and `right' contexts will be included into the regular expressions by using lookbehind and lookahead}.
-Basically, the actual regular expression syntax of lookbehind and lookahead is simply hidden to the users by allowing them to only specify the contexts themselves.
-Internally, the contexts in the columns ``left'' and ``right'' are combined with the column ``graphemes'' to form a complex regular expression like ``(?\textless{}=left)graphemes(?=right)''.
-	\item \textsc{The regular expressions will be applied in the order as specified in the profile, from top to bottom.} A software implementation can offer help in figuring out the optimal ordering of the regular expressions, but should then explicitly report on the order used.
-\end{enumerate}
-
-The actual implementation of the profile on some text-string will function as follows:
-\begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item \textsc{All graphemes are matched in the text before they are tokenized or transliterated}.
-In this way, there is no necessity for the user to consider `feeding' and `bleeding' situations, in which the application of a rule either changes the text so another rule suddenly applies (``feeding'') or prevents another rule to apply (`bleeding').
-	\item \textsc{The matching of the graphemes can occur either globally or linearly.
-}From a computer science perspective, the most natural way to match graphemes from a profile in some text is by walking linearly through the text-string from left to right, and at each position go through all graphemes in the profile to see which one matches, then go to the position at the end of the matched grapheme and start over.
-This is basically how a finite state transducer works, which is a well-established technique in computer science.
-However, from a linguistic point of view, our experience is that most linguists find it more natural to think from a global perspective.
-In this approach, the first grapheme in the profile is matched everywhere in the text-string first, before moving to the next grapheme in the profile.
-Theoretically, these approaches will lead to different results, though in practice of actual natural language orthographies they almost always lead to the same result.
-Still, we suggest that any software application using orthography profiles should offer both approaches (i.e.
-``global'' or ``linear'') to the user.
-The approach used should be documented in the metadata as ``tokenization method''.
-	\item \textsc{The matching of the graphemes can occur either in NFC or NFD.
-}By default, both the profile and the text-string to be tokenized should be treated as NFC (see section 4, Pitfall ``Canonical equivalence'' above).
-However, in some use-cases it turns out to be practical to treat both text and profile as NFD.
-This typically happens when very many different combinations of diacritics occur in the data.
-An NFD-profile can then be used to first check which individual diacritics are used, before turning to the more cumbersome inspection of all combinations.
-We suggest that any software application using orthography profiles should offer both approaches (i.e.
-``NFC'' or ``NFD'') to the user.
-The approach used should be documented in the metadata as ``unicode normalization''.
-	\item \textsc{The text-string is always returned in tokenized form} by separating the matched graphemes by a user-specified symbols-string.
-Any transliteration will be returned on top of the tokenization.
-	\item \textsc{Leftover characters (i.e.~characters that are not matched by the profile) should be reported to the user as errors.} Typically, the unmatched character are replaced in the tokenization by a user-specified symbol-string.
-\end{enumerate}
-
-Any software application offering to use orthography profile:
-\begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item \textsc{should offer user-options} to specify:
-	\begin{enumerate}
-		\def\labelenumii{\arabic{enumii}.} 
-		\item \textsc{the name of the column to be used for transliteration} (if any). 
-		\item \textsc{the symbol-string to be inserted between graphemes.} Optionally, a warning might be given if the chosen string includes characters from the orthography itself. 
-		\item \textsc{the symbol-string to be inserted for unmatched strings} in the tokenized and transliterated output. 
-		\item \textsc{the tokenization method}, i.e.~whether the tokenization should proceed ``global'' or ``linear''. 
-		\item \textsc{unicode normalization}, i.e.~whether the text-string and profile should use ``NFC'' or ``NFD''. 
-	\end{enumerate}
-	\item \textsc{might offer user-options }to:
-	\begin{enumerate}
-		\def\labelenumii{\arabic{enumii}.} \setcounter{enumii}{5} 
-		\item \textsc{assist in the ordering of the graphemes.} In our experience, it makes sense to apply larger graphemes before shorter graphemes, and to apply graphemes with context before graphemes without context.
-Further, frequently relevant rules might be applied after rarely relevant rules (though frequency is difficult to establish in practice, as it depends on the available data).
-Also, if this all fails to give any decisive ordering between rules, it seems useful to offer linguists the option to reverse the ordering from any manual specified ordering, because linguists tend to write the more general rule first, before turning to exceptions or special cases.
-		\item \textsc{assist in dealing with upper and lower case characters.} It seems practical to offer some basic case matching, so characters like <a> and <A> are treated equally.
-This will be useful in many concrete cases, although the user should be warned that case matching does not function universally in the same way across orthographies.
-Ideally, users should prepare orthography profiles with all lowercase and uppercase variants explicitly mentioned, so by default no case matching should be performed.
-		\item \textsc{treat the profile literal}, i.e.~to not interpret regular expression metacharacters.
-Matching graphemes literally often leads to strong speed increase, and would allow users to not needing to worry about escaping metacharacters.
-However, in our experience all actually interesting use-cases of orthography profiles include some contexts, which automatically prevents any literal interpretation, so by default the matching should not be literal.
-	\end{enumerate}
-	\item \textsc{should return the following information} to the user:
-	\begin{enumerate}
-		\def\labelenumii{\arabic{enumii}.} \setcounter{enumii}{8} 
-		\item \textsc{the original text-strings to be processed in the used Unicode normalization}, i.e.~in either NFC or NFD as specified by the user. 
-		\item \textsc{the tokenized strings}, with additionally any transliterated strings, if transliteration is requested. 
-		\item \textsc{a survey of all errors encountered}, ideally both in which text-strings any errors occurred and which characters in the text-strings lead to errors. 
-		\item \textsc{a reordered profile}, when any automatic reordering is offered 
-	\end{enumerate}
-\end{enumerate}
-
-\section{7. Use cases}\label{use-cases}
+\section{Use cases}\label{use-cases}
 
@@
 Once text has been normalized, an additional and straightforward process is Unicode grapheme tokenization.
-More recently this regular expression, often available as short cut ``\X'', identifies all sequences of ``base'' characters followed by 0 or more combining diacritics.
-For example, the sequence, <ŋ͡mṍraː> would tokenize as a space-delimited sequence as: <ŋ͡ m ṍ r a ː>.
-For a first pass at identifying grapheme clusters, this is a straightforward tokenization.
+More recently this regular expression, often available as short cut ``\textbackslash{X}'', identifies all sequences of ``base'' characters followed by 0 or more combining diacritics.
+For example, the sequence, % <ŋ͡mṍraː> would tokenize as a space-delimited sequence as: <ŋ͡ m ṍ r a ː>.
+For a first pass at identifying grapheme clusters, this is a straightforward
+tokenization.
 However, the linguist will note that both the tie bar (COMBINING DOUBLE INVERTED BREVE, a Unicode Combining Modifier) and the length marker (MODIFIER LETTER TRIANGULAR COLON.
@@
 Orthographic tokenization requires a specification of the tailored grapheme clusters (linguists: graphemes) to separate orthographic units within an input source.
-In the simplest case, the Unicode graphemic tokenization with regular expression ``\X'' captures the grapheme clusters, i.e.~the input data is graphemically tokenized and there are no additional string sequences required to capture each grapheme, e.g.~no specification.
+In the simplest case, the Unicode graphemic tokenization with regular expression ``\textbackslash{X}'' captures the grapheme clusters, i.e.~the input data is graphemically tokenized and there are no additional string sequences required to capture each grapheme, e.g.~no specification.
 This scenario is the most unlikely.
@@
 
-({[}a\textbar{}á\textbar{}e\textbar{}é\textbar{}i\textbar{}í\textbar{}o\textbar{}ó\textbar{}u\textbar{}ú{]})(n)(\s)({[}a\textbar{}á\textbar{}e\textbar{}é\textbar{}i\textbar{}í\textbar{}o\textbar{}ó\textbar{}u\textbar{}ú{]}), \1 \2 \4
+% ({[}a\textbar{}á\textbar{}e\textbar{}é\textbar{}i\textbar{}í\textbar{}o\textbar{}ó\textbar{}u\textbar{}ú{]})(n)(\s)({[}a\textbar{}á\textbar{}e\textbar{}é\textbar{}i\textbar{}í\textbar{}o\textbar{}ó\textbar{}u\textbar{}ú{]}), \1 \2 \4
 
-A preliminary orthography profile can be generated with code available online.\footnote{https://github.com/bambooforest/tokenizer) The Python script creates a unigram model for individual Unicode characters or Unicode grapheme clusters and their frequency in the source input.
-We find that this gives a broad overview of the input at the character and grapheme levels and it oftens finds typographic errors in the source input by identifying rarely occurring or one-off characters in the input.
+A preliminary orthography profile can be generated with code available online.\footnote{https://github.com/bambooforest/tokenizer} The Python script creates a unigram model for individual Unicode characters or Unicode grapheme clusters and their frequency in the source input.
+We find that this gives a broad overview of the input at the character and grapheme levels and it often finds typographic errors in the source input by identifying rarely occurring or one-off characters in the input.
 The models also provides the basis for the ``quick'' development of an initial orthography profile by saving the user keyboarding time.
@@
 Each General Category property also has subcategories that are also properties with property values.
-For example, ``Letters'' is broken down into Uppercase Letter (Lu), Lowercase Letter (Ll), Titlecase Letter (Lt), Modifier Letter (Lm) and Other Letter (Lo).\footnote{The subcategory ``Other Letter'' includes characters such as uncased letters as in the Chinese writing system.) By defining character properties for each code point, algorithmic implementations such as text processing applications or regular expression engines that are conformant to the Unicode Standard can be used to recognize whole (sub)categories of characters or ranges of characters, e.g.~to match proper names in English in title case or more generally to match character types, such as scripts, or symbols and integers in email addresses, phone numbers, etc.
+For example, ``Letters'' is broken down into Uppercase Letter (Lu), Lowercase Letter (Ll), Titlecase Letter (Lt), Modifier Letter (Lm) and Other Letter (Lo).\footnote{The subcategory ``Other Letter'' includes characters such as uncased letters as in the Chinese writing system.} By defining character properties for each code point, algorithmic implementations such as text processing applications or regular expression engines that are conformant to the Unicode Standard can be used to recognize whole (sub)categories of characters or ranges of characters, e.g.~to match proper names in English in title case or more generally to match character types, such as scripts, or symbols and integers in email addresses, phone numbers, etc.
 
@@
 
-Unicode-aware regular expression engines may contain a meta-character, ``\X'', that matches this level of grapheme cluster, i.e.~a base character followed by some number of accents or diacritical marks.
+Unicode-aware regular expression engines may contain a meta-character, ``\textbackslash{X}'', that matches this level of grapheme cluster, i.e.~a base character followed by some number of accents or diacritical marks.
 Grapheme clusters are important to determine word or line boundaries, for collation (`ordering') and for user interface interaction (e.g.~mouse selection, backspacing, cursor movement).
-
-\section{9. Conclusion}\label{conclusion}
-
-In our work we focus on the writing systems used in lesser-described and endangered languages and our orthography profiles are used to describe writing systems and to transpose them into phonetic transcription.
-Of course the International Phonetic Alphabet and other transcription systems are essentially just orthographies that are an approximation of sound.
-Nevertheless, sound-based normalization is practical in automatically identifying cognates and sound changes with quantitative methods.
-We leverage orthography profiles to enable comparative analysis of languages with different writing systems.
-
-An orthography profile lists the graphemes in a particular description of language data, e.g.~a wordlist, dictionary or corpus.
-Building on the knowledge that can be extracted from that description by tokenizing words by the graphemes made explicit in the orthography profile, it is straightforward to undertake other analyses of the data.
-For example, various ngram models of the data can be extracted with a few lines of code.
-A unigram model with counts, frequencies and positive log probability (``plogs''; ``entropy under this model'') (Goldsmith \& Riggle, 2010) provides a fair amount of information about a given data source.
-Essentially, the orthography profile provides the description that allows this information to be calculated based on the mapping of sequences of characters into graphemic units.
-Bigram models are also straightforwardly extracted.
-In the case of bigrams, mutual information can be captured and used in various other statistical analyses, such as quantitative language comparison, inferring phylogenetic trees, etc.
-
-\section{10. Nice text blocks to be included somewhere}\label{nice-text-blocks-to-be-included-somewhere}
-
-Looking at the variety of writing systems used for the world's languages, there are various challenges for the computational processing of resources in which such writing systems are used.
-
-A widespread mechanism to derive new symbols for a writing system is to use a combination of multiple symbols already available.
-That can lead to completely new symbols, like <æ> or <w>, which are still rather transparently derived from <a e> and <v v>, respectively.
-More typically, many orthographies use sequences of symbols as special characters, like <sch> or <gh>.
-To identify such `grapheme clusters' is one of the central problems to be tackled with orthography profiles.
-A further frequently used mechanism is to add extra material to an already available symbol, either to make it better visible (this is probably the reason for the addition of a dot on top of the <i>) or to derive new symbols.
-The widespread use of diacritics is the obvious result of such processes (Gaultney 2002).
-However, also the derivation of new symbols by added extra strokes is widespread (and not only in Far-Eastern character-based writing systems), like for example <ŋ> or <ç>.
-
-The use of punctuation (Daniëls 1994, Humez \& Humez 2008) and spaces (Saenger 1997) to improve readability. {[}e.g.~space after punctuation, but not before{]}
-
-Currently, the Unicode Standard is the standard character encoding for linguistic data (Anderson 2003).
-However, there are still hundreds of different encoding systems that were invented independently to capture orthographic diversity as different nations adopted and developed computer systems.
-These different encoding systems are problematic and in conflict with one another because different standards were formalized differently and for different purposes by different standards committees in different countries.
-No unified encoding scheme contained enough code points to encode all characters, so two different encoding schemes possibly used the same code point for different characters, or used different code points to represent the same character.
-Because computers support multiple character encoding schemes, data risked being corrupted when handled by different applications and encodings.
-The Unicode Standard was devised to alleviate these problems.
-
-For orthographies, identifying graphemes can be even more challenging than identifying phones and phonemes in phonetic transcription because although transcriptions may not adhere strictly to IPA, they tend to have straightforward mappings between sounds and symbols.
-On the other hand, orthographies can introduce orthographic rules, which add an additional challenges in identifying graphemes in words, as mentioned above.
-Thus for resources not in IPA or IPA-like transcriptions, graphemes often must be manually identified, whether they are encoded as singletons or multi-character sequences.
-The identification of graphemes and the formulation of orthographic rules are used to create an orthography profile.
-
-Character-to-glyph mappings are particularly not clear-cut for linguists using the IPA.\footnote{We give more linguistic examples in Section 5.) Just look in the IPA Extensions block.\footnote{Character ``blocks'' are described in Section 3 (The Unicode approach).) There are several characters that are glyphs (variants) of the same grapheme in the Latin block, e.g.
-and \& and .\footnote{LATIN SMALL LETTER ALPHA U+0251 vs LATIN SMALL LETTER A U+0061 , and LATIN SMALL LETTER SCRIPT G U+0261 vs LATIN SMALL LETTER G U+0067 .) Other characters like
-
-, , do not appear at all in the IPA Extensions block, but as separate phonetic symbols; they are already encoded in the Basic Latin, Latin Extended-A and Greek \& Coptic blocks.
-When a linguist transcribes an IPA
-
-on a QWERTY keyboard, if it is from the Basic Latin group, it is valid Unicode IPA (it is not valid IPA if it is from the Cyrillic Block).
-However, keyboard and and because they have been given specific code points in the IPA Extensions block, where they have been given special status as IPA characters, and which most often go unnoticed by linguists.\footnote{In Section 5 we discuss in detail the Unicode IPA alphabet and its characters dispersal across different Unicode Standards scripts and blocks.)
-
-The Unicode Standard divides text elements roughly into two classes of graphemes.
-The cover term \textsc{grapheme cluster}\footnote{Formerly referred to as ``locale-independent graphemes'' to emphasize that the term grapheme is used differently in linguistics.) is used to emphasize the difference between the term \emph{grapheme }as used by linguists and what the users perceive as a character (a basic unit of a writing system for a language), which may not be a single Unicode character.\footnote{http://www.unicode.org/reports/tr29/) The Unicode Consortium makes a further distinction between grapheme clusters and tailored grapheme clusters.
-
-The simplest example of a grapheme cluster is essentially a base character followed by a Letter Modifier character.
-For example, the sequence + \textless{}\textasciitilde{}\textgreater{} (COMBINING TILDE at U+0303) combines visually into , a user-percieved character in writing systems like that of Spanish.\footnote{Although the code point sequence can also be represented with a single code point, LATIN SMALL LETTER N WITH TILDE at U+00F1, it is not the case that all multi-code point character sequences have a single code point equivalent, which can be deduced from normalization.
-Often for commonly-used letters in European languages, a single code point form exists.)
-
-What the user perceives as a single character may actually involved a multi-code point sequence.\footnote{Unicode-aware regular expression engines may contain a meta-character, ``\X'', that matches this level of grapheme cluster, i.e.~a base character followed by some number of accents or diacritical marks.
-As mentioned, grapheme clusters are important for collation algorithms, but they are also used to determine other processes such as word or line boundaries and for user interface interaction (e.g.~mouse selection, backspacing, cursor movement).)
-
-More challenging for text processing is what the Unicode Consortium terms tailored grapheme clusters.
-For example, the sequence + for the Slovak digraph or the sequence .
-These grapheme clusters are ``tailored'' in the sense that they must be specified on a language-by-language basis.
-Algorithms can be adapted to use as input these tailored language-specific orthographies for transliteration, localization, etc.
-In fact, the Unicode Standard provides a method for defining language-specific tailored writing systems called the Unicode Locale Description.
-Unicode Locale Descriptions allow users to define a set of extended grapheme clusters.
-Unicode Locale Descriptions are saved in the Common Locale Data Repository (CLDR),\footnote{http://cldr.unicode.org/) a repository of language-specific definitions of writing system properties, each of which describes specific usages of characters.
-Each can be encoded in a Unicode Locale Data Markup Language (LDML) document.
-LDML is an XML format and vocabulary for the exchange of structured locale data.
-It is a format used not only for CLDR, but also for general interchange of locale data, such as in Microsoft's .NET.
-We return to LDML in Section 5 (Orthography profiles and locale descriptions).
-
-We have noted that the rendering of electronic text requires that characters are encoded and stored in a computer's memory.
-These character sequences are then mapped from character code points to glyphs by software that takes into account a character's context.
-Transparent to most users is the fact that different logical orders of characters may result in the same visual output.
-For example, phonetic characters with certain combinations of diacritics may be homoglyphs (see below), even while the logical order of their character sequences are not the same.\footnote{One example is a vowel that is both nasalized and creaky voice, e.g.
-{[}ḛ̃{]}.
-See discussion below.) Thus some type of standardization, what the Unicode Consortium calls ``normalization'' of the ordering of (some classes of) characters, is required to make sure that all data are logically consistent and therefore easily comparable and equally searchable.
-The aim of standardization is to attain syntactic \textsc{interoperability} of data.\footnote{By interoperability we mean the ability to ubiquitously exchange and merge disparate (textual) data sets, and data encoding formats, to facilitate data sharing and to ``effortlessly'' undertake comparison and analysis.
-For a discussion of syntactic versus semantic interoperability, see Ide and Pustejovsky, 2010.)
-
-The IPA is often used by linguists in undertaking initial phonetic and phonemic analyses of languages.
-However, once the distinctive sounds of a given language have been identified (and a phonological description of the target language has been described), a linguist may continue his or her documentation and description of higher-order structures like words and sentences with a broad phoneme-based alphabet.
-In fact, literary materials development often leapfrogs from a phonemic alphabet to locale-specific practical orthographies, in which numerous factors affect a language's orthography development, e.g.~politics, purpose, speaker communities familiarity with other writing systems, technological limitations.
-
-One simple example is illustrated with two genealogically related languages spoken in Northern Ghana: Western Sisaala\footnote{Western Sisaala lacks /kp/.) and Sisaala Tumulung.\footnote{Sisaala Tumulung (ISO 639-3: sil) is spoken around the city Tumu, which is about 50 kilometers geographically to the east of Western Sisaala (ssl) speaking areas.) Both languages contain the same basic set of contrastive sounds, but their orthographies encode affricates (and vowels) differently graphemically; the influence stem from orthographies of government-sponsored literacy languages, which are taught to children in schools in these areas.
-In Western Sisaala the sound /tʃ/ is written and /dʒ/ as due to familiarity of these graphemes by students learning Dagaare (Moran 2006).
-Spoken further to the east, Sisaala Tumuling encodes the sound /tʃ/ as and /dʒ/ as (Blass 1975).
-
-If encoded in the Unicode Standard, the design of both orthographies has to specify that certain sequences of characters, such as + , are single graphemes (compare English and which is ambiguous).
-Design is important, for example, for collation ordering (sorting), i.e.~the sequence of characters that is treated as a unit for ordering purposes, such as in dictionaries.
-Indeed, the characters , and are three unique graphemes consisting of two Unicode characters from the Latin script, but developers of a software program that recognizes the Western Sisaala alphabet might want to specify the sort order , , \ldots{} .
-Thus there is a need for orthography-specific specifications that define sequences of characters as graphemes, whether technologically, linguistically, or both, for multilingual text processing.
-
-A major challenge that the developers of the Unicode Standard faced is that there exists no universal set of fundamental units of text because the divisions of text into elements, such as graphemes, is dependent on language-specific writing systems and orthographies, as just shown.
-Moreover, the division of electronic text into elements is process dependent, e.g.~sorting and spell checking may or may not need to take capitalization into account, depending on the orthography in question.
-Thus the encoding and processing of text requires well defined \textsc{text elements} and rules for interpreting them; these elements and rules are language-specific when we talk about major languages or well established orthographies, but more often in our research we face input from lesser-studied languages documented in author-specific transcription or practical orthography.\footnote{And this is usually down to the document-specific level, the description of which has been coined ``doculect'' for purposes of discussing differences of opinion in language descriptions (Cysouw and Good, 2013).)
-
-In Section XXX we noted that the Unicode Consortium makes a crucial distinction between characters and glyphs.
-The relations between characters and glyphs is a complex one and it is typically hidden from the user.
-Technologically, we must distinguish between characters and glyphs because:
-\begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item The logical order of a sequence of characters (in computer memory) may not be the same as the visual order of their (rendered) glyphs. 
-	\item There is not always a one-to-one mapping between characters and glyphs. 
-\end{enumerate}
-
-Regarding the first point, a user cannot assume that contiguously rendered text on-screen actually reflects the computer's internal storage order of individual and sequential characters in memory, i.e.~the \textsc{logical order}.
-In other words, in the Unicode Standard the visual order of glyphs may not be the same as the logical order of their digitally encoded characters; contiguous display is not indicative of contiguous text.* *A nice example is Hebrew text.\footnote{http://www.macchiato.com/unicode/globalization-gotchas) Another example is the glyph <ṍ>, which may consist of one, two or three characters in encoded in various physical orders.
-The possible mismatch in sequences becomes clearer when we consider the second point.
-
-When a base character, i.e.~any graphic character not in the General Category of Combining Mark, is followed by one or more characters having the property General Category of Combining Mark, they can be combined visually into one or more glyphs using the Canonical Ordering Algorithm to create Unicode \textsc{grapheme clusters}.
-As we have shown, this input may be ambiguous from a technological perspective if normalization is not taken into account.
-For example, a linguist may insert a single character or a character sequence for to denote a high back unrounded vowel with high tone:
-\begin{itemize}
-	\item LATIN SMALL LETTER U WITH ACUTE (U+00FA) 
-	\item LATIN SMALL LETTER U (U+0075) + COMBINING ACUTE ACCENT (U+0301) 
-\end{itemize}
-
-The normalization algorithms in the Unicode Standard are used to make these character sequences canonically equivalent for purposes such as ubiquitous search and comparison.
-
-However, the standard does not specify many (or most) language-specific graphemes, or in Unicode parlance, \textsc{tailored grapheme clusters}, such as English .\footnote{Some (unfortunate) exceptions included for legacy purposes, and used incorrectly by linguists, include digraphs like LATIN SMALL LETTER TS DIGRAPH at U+02A6 or LATIN SMALL LETTER DEZH DIGRAPH at U+02A4.
-These and other digraphs are encoded in the Unicode Standard's IPA Extensions block.) For technological reasons, these language-specific tailored grapheme clusters must be defined outside of the Unicode Standard, in what are called (language-specific) Unicode Locale Descriptions.\footnote{http://www.unicode.org/reports/tr35/) The encoding of these multiple character sequences in Locale Descriptions is beyond the need (or technical ability) of many linguists; thus we have developed a simpler approach which we used in our own research and describe in detailed in Section XXX.
-Basically, the description of writing systems for lower resource languages, and in particular for document-specific descriptions (e.g.~different descriptions of the same language with different orthographies), requires an orthographic description at the (tailored) grapheme (cluster) level.
-
-An orthography profile is defined as a simple delimited flat text file encoded in the UTF-8, with character sequences in Unicode Normalization Form D.
-Metadata aside, the column delimited contents of an orthography profile are essential a state-table, i.e.~an implementation of a state transition table, but for ease of use encoded in a simple flat text file.
-
-In an orthography profile, all character strings are converted into NFD because each character in a grapheme may have phonetic value, and by using NFD, all graphemes are decomposed into a standardized order.
-For example, a vowel that is both nasalized and creaky looks like in IPA, where is some vowel.
-Although visually the same, a nasalized and creaky vowel can be composed of several different character sequences, as illustrated with in 1-3:
-\begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item <õ̰> LATIN SMALL LETTER O + COMBINING TILDE + COMBINING TILDE BELOW: U+006F + U+0303 + U+0330 
-	\item <õ̰> LATIN SMALL LETTER O + COMBINING TILDE BELOW + COMBINING TILDE: U+006F + U+0330 + U+0303 
-	\item <õ̰> LATIN SMALL LETTER O WITH TILDE + COMBINING TILDE BELOW: U+00F5 + U+0330 
-\end{enumerate}
-
-Applying NFD to these three character sequences results in one standard sequence; in this example the character sequence is that shown 1.
-NFD makes different sequences of input interoperable and it retains all of the phonetic information captured by the separate characters that combine to form a vowel with nasalization and creaky voice phonation.
-This is an important step to consider in linguistic research because each character has phonetic value, i.e.~the combining tilde denotes that the base character is produce by a lowering of the velum, which allows air to escape through the nose when the sound is produced.
-The tilde below also has meaning: it represents a phonation type in which the vocal cords are vibrating, but they are also being held slightly further apart, allowing more air to escape, which produces a murmured sound.
-To enable comparative analyses of languages at a deep phonetic level, retaining a code point for each character of phonetic value is deciduous.
-For example, if the sound systems of two languages are to be compared, one would not want to lose a possibly contrastive phonological feature like nasalization, which may have been collapsed into a single character in its logical representation.
-
-Our approach is similar in nature to the LDML, but we use a simpler state-table method, instead of generating an XML format for the ``exchange of structured locale data''.
-An orthography profile is a specification of the tailored grapheme clusters in a particular source document and any orthographic re-write rules needed after the application of Unicode normalization and orthographic profile tokenization.
-Orthography profiles can of course be used for transliteration between scripts, e.g.~from practical orthography to IPA and vice versa, by specifying mapping (replacements) from the list of tailored grapheme clusters in the source document and a set of replacement characters.
-This will be made clear in the next section in which we provide some use cases.
-
-To summarize, LDML is basically overkill for our intentions of identifying and delimiting grapheme boundaries (or in Unicode parlance ``tailored grapheme clusters'').
-LDML is appropriate, even necessary, for complex localization complexes for software applications -- localization of, say, an operating system or software application across cultures may require the instantiation of different date stamps an other local standards.
-For linguistic purposes of source document data tokenization, we therefore develop a simple-to-implement process of ``orthographic'' tokenization, using standard Unicode Standard-specified processes (e.g.~normalization and grapheme cluster tokenization) with an additional level of linguistic ``tailored grapheme cluster'' specification.
-
-\textsc{REFERENCES}
-
-Nancy Ide and James Pustejovsky.
-2010.
-What does inter- operability mean, anyway?
-toward an operational def- inition of interoperability for language technology.
-In Proceedings of the Second International Conference on Global Interoperability for Language Resources.
-Hong Kong, China.
-
-Huber, Randal Q. and Robert B. Reed (compilers).
-\begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} \setcounter{enumi}{1991} 
-	\itemsep1pt\parskip0pt\parsep0pt 
-	\item Vocabulario comparativo: Palabras selectas de lenguas indígenas de Colombia (Comparative vocabulary: Selected words in indigenous languages of Colombia).
-Bogota: Instituto Lingüísti- co de Verano.
-\end{enumerate}
-
-Thiesen, W.
-(1998).
-Diccionario bora-castellano, castellano-bora (1st ed.).
-Lima Perú: Instituto Linguístico de Verano.Retrieved from: http://www.worldcat.org/title/diccionario-bora-castellano-castellano-bora/oclc/40505215\&referer=brief\_results
-
-\textsc{APPENDIX - ORTHOGRAPHY PROFILE EXAMPLE}
-
-\textsc{APPENDIX - UNICODE IPA}
-
-Provide a clear and full Unicode IPA chart with additional information as an appendix
-\begin{verbatim}
-	- create this chart
-
-	- perhaps we should spell out conformance rules with regard to this chart? i.e. if a source claims conformance to our Unicode IPA chart, it should:
-
-	- identify the sections (or range) of the table that the data conforms to
-
-	- the source is in accordance with not only the Unicode IPA characters, but their intrinsic meanings (e.g.
-an Africanist might use <y> for /j/, but this level of conformance requires that <j> symbolizes /j/, etc.)
-
-	- someone who wants to use an Africanist transcription system can declare that they use the Unicode IPA chart, except for a range of characters (but they provide the mapping to Unicode IPA chart) 
-\end{verbatim}
-
-\textsc{Notes}
-
-\url{http://www.unicode.org/reports/tr29/}
-
-The default boundary determination mechanism specified in this annex provides a straightforward and efficient way to determine some of the most significant boundaries in text: user-perceived characters, words, and sentences.
-Boundaries used in line breaking (also called \emph{word wrapping}) are defined in {[}\href{http://www.unicode.org/reports/tr41/tr41-11.html\#UAX14}{UAX14}{]}.
-
-\textsc{Nice quotes}
-
-vanZundert: ``How does one design and plan a general unified infrastructure for a target that is all over the place?
-A `one size fits all' approach would be a disastrous underestimation of the specific needs of humanities research.
-The essence of humanities research is in its diverse, heterogeneous and ephemeral nature.
-But the need of big infrastructures to be designed well before implementation and use, means that every single design decision closes the resulting system to some category of users.
-Anything specific can potentially break the data model and the defined workflow.
-Of course big infrastructure architects also know this; to compensate, they look for possibilities to make more abstract models and malleable workflows.
-But abstract away far enough and the model becomes rather meaningless.
-If the design recommendation of big infrastructure is that we will use XML, what was the use of all that time and effort?
-Big infrastructure design seems to routinely underestimate this problem.
-To give but one simple example: the token seems such a logical, granular, easy to automate conceptual catch-all for text models.
-Can there possibly be a more generalizable model for text than as a series of tokens?
-You may think so, until someone asks how you handle languages that do not use spaces, or how you account for prefixes and suffixes, let alone ambiguous functions.
-Regarding text as a series of tokens denies in any case several other essential humanistic aspects of text: its typography, its materiality, its referentiality to name but a few.
-At second glance, from the humanities research perspective, there is pretty little actually generalizable about tokenization.''
-
-\% Ladefoged, Peter.
-1990.
-Some Reflections on the IPA.
-In UCLA Working Papers in Phonetics, 74.
-``The most interesting ambiguities in the Association's theoretical concepts concern what the symbols are symbolizing.''
-
-There are several ambiguous aspects of the IPA, but the most interesting concerns the theoretical concepts that its symbols are symbolizing \cite[64]{Ladefoged1990a}.
-Although the IPA is \texttt{intended to be a set of symbols for representing all the possible sounds of the world\textquotesingle{}s languages\textquotesingle{}\textquotesingle{} \textbackslash{}citep\{IPA2005\}, Ladefoged points out that the IPA also states that IPA symbols are}\textit{not} symbols for phones they are simply shorthand for what a phonologist would regard as a bundle of features'' \cite{Ladefoged1990a}.
-From the IPA:
-\begin{quote}
-
-	``The representation of these sounds uses a set of phonetic categories which describe how each sound is made.
-These categories define classes of sounds that operate in phonological rules and historical sounds changes.
-The symbols of the IPA are shorthand ways of indicating certain intersections of these categories.
-Thus [p] is a shorthand way of designating the intersection of the categories voiceless, bilabial, and plosive; [m] is the intersection of the categories voiced, bilabial, and nasal; and so on.''
-\end{quote}
-
-\noindent Thus, the ``association is deliberately not explicit about what is meant by a phonological contrast'' \cite[65]{Ladefoged1990a}.

### 2015-05-29 localpackages.tex
initial commit paper

% add all extra packages you need to load to this file

\usepackage{amsmath} 
\usepackage{unicode-math}
\setmathfont{Asana-Math.otf}
\usepackage[libertine]{newtxmath}

\usepackage{tabularx}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%                                              %%%
%%%           Examples                           %%%
%%%                                              %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage{lsp-gb4e}

%% to add additional information to the right of examples, uncomment the following line
% \usepackage{jambox}
%% if you want the source line of examples to be in italics, uncomment the following line
% \renewcommand{\exfont}{\itshape}

\usepackage{listings}

\lstset{ %
  backgroundcolor=\color{white},   % choose the background color; you must add \usepackage{color} or \usepackage{xcolor}
  basicstyle=\footnotesize\ttfamily,        % the size of the fonts that are used for the code 
  keywordstyle=\color{blue!60!black},       % keyword style
  language=XML,                 % the language of the code 
  stringstyle=\color{green!60!black},     % string literal style 
  morekeywords={token,xlink:href, Action, Value, Cursor,LogEvent}
}

### 2015-05-31 chapters/encoding.tex
edits, bib, adherence to lang sci press formatting, etc.

@@
-\section{Encoding}
+\chapter{Encoding}
 \label{encoding}
 
-There are many in-depth histories of the origin and development of writing systems (e.g.~Robinson 1997, Powell 2012), a story that we will not repeat here.
+There are many in-depth histories of the origin and development of writing systems (e.g.~\citet{Robinson1997,Powell2012}), a story that we will not repeat here.
 However, the history of turning writing into computer readable code is not so often told, so we decided to offer a short survey of the major developments here.
-This history turns out to be intimately related to the history of telegraphic communication.\footnote{Because of the recent history as summarised in this section, we have used mostly rather ephemeral internet sources.
+This history turns out to be intimately related to the history of telegraphic communication.\footnote{Because of the recent history as summarized in this section, we have used mostly rather ephemeral internet sources.
 When not references by traditional literature in the bibliography, we have specifically used \url{http://www.unicode.org/history/} and various Wikipedia pages for the information presented here.
-A useful survey of the historical development of the physical hardware of telegraphy and telecommunication is Huurdeman (2003).
-Most books that discuss the development of encoding of telegraphic communication focus of cryptography (e.g.~Singh 1999) and forego the rather interesting story of `open' encoding that is related here.}
+A useful survey of the historical development of the physical hardware of telegraphy and telecommunication is \citet{Huurdeman2003}.
+Most books that discuss the development of encoding of telegraphic communication focus of cryptography, e.g.~\citet{Singh1999}, and forego the rather interesting story of `open' encoding that is related here.}
 
@@
 Additionally, to bridge large geographic distances, \textsc{telegraphic systems} of communication (from Greek \emph{τῆλε γράφειν} `distant writing') have a long and widespread history since ancient times.
-The most widespread telegraphic systems worldwide are so-called whistled languages (Meyer 2015), but also drumming languages (Meyer et al.
-2012) and smoke, fire, hydraulic or flag signals are forms of telegraphy.
+The most widespread telegraphic systems worldwide are so-called whistled languages \citep{Meyer2015}, but also drumming languages \citep{Meyer_etal2012} and smoke, fire, hydraulic or flag signals are forms of telegraphy.
 
 Telegraphy was reinvigorated in the end of the eighteenth century through the introduction of semaphoric systems by Claude Chapelle (since then using flags, flashing lights, or various specially designed contraptions) to convey messages over large distances.
-The ``innovation'' of those systems was that all characters of the written language were replaced one-to-one by visual signals.
+The \textit{innovation} of those systems was that all characters of the written language were replaced one-to-one by visual signals.
 Since then, all telegraphic systems have taken this principle, namely that any language to be transmitted first has to be turned into some orthographic system, which subsequently is encoded for transmission by the sender, and then turned back into orthographic representation at the receiver side.\footnote{Sound and video-based telecommunication of course takes a different approach by ignoring the written version of language and directly encode sound waves or light patterns.} This of course implies that the usefulness of any such telegraphic encoding completely depends on the sometimes rather haphazard structure of orthographic systems.
@@
 In the nineteenth century, \textsc{electric telegraphy} lead to a new approach in which written language characters were encoded by signals sent through a copper wire.
-Originally, ``bisignal'' codes were used, consisting of two different signals.
-For example, Carl Friedrich Gauss in 1833 used positive and negative current (Mania 2008:282).
+Originally, \textsc{bisignal} codes were used, consisting of two different signals.
+For example, Carl Friedrich Gauss in 1833 used positive and negative current \citep[282]{Mania2008}.
 More famous and influential, Samuel Morse in 1836 used long and short pulses.
@@
 In such systems only a single separator between signals is needed, because the separation between characters can be established by counting until a fixed number of signals.\footnote{Of course, no explicit separator is needed when the timing of the signals is known, which is the principle used in all modern telecommunication systems.
-An important modern consideration is also how to know where to start counting when you did not catch the start of a message, something that is known in Unicode as ``self synchronization''.}
+An important modern consideration is also how to know where to start counting when you did not catch the start of a message, something that is known in Unicode as \textsc{self synchronization}.}
 
 In the context of electric telegraphy, such a binary code system was first established by Émile Baudot in 1870, using a fixed combination of five signals for each written character.\footnote{True binary codes have a longer history, going at least back to the Baconian cipher devised by Francis Bacon in 1605.
-However, the proposal by Baudot was the quintessential proposal leading to all modern systems.} There are $2^5 = 32$ possible combination when using five binary signals; an encoding today designated as ``5-bit''.
+However, the proposal by Baudot was the quintessential proposal leading to all modern systems.} There are $2^5 = 32$ possible combination when using five binary signals; an encoding today designated as \textsc{5-bit}.
 These codes are sufficient for all Latin letters, but of course they do not suffice for all written symbols, including punctuation and digits.
-As a solution, the Baudot code uses a so-called ``shift'' character, which signifies that from that point onwards---until shifted back---a different encoding is used, allowing for yet another set of 32 codes.
+As a solution, the Baudot code uses a so-called \textsc{shift} character, which signifies that from that point onwards---until shifted back---a different encoding is used, allowing for yet another set of 32 codes.
 In effect, this means that the Baudot code, and the \textsc{International Telegraph Alphabet} (ITA) derived from it, had an extra `bit' of information, so the encoding is actually 6-bit (with $2^6 = 64$ different possible characters).
@@
 For example, part 5 was used for Cyrillic characters, part 6 for Arabic, and part 7 for Greek.\footnote{In effect, because $16 = 2^4$, this means that ISO/IEC~8859 was actually a $8+4=12$-bit encoding, though with very many duplicates by design (e.g.~all ASCII codes were repeated in each 8-bit code page).
-To be precise, ISO/IEC~8859 used the 7-bit ASCII as the basis for each code page, and defined 16 different 7-bit extensions, leading to $(1+16)\cdot{2^7} = 2.176$ possible characters.}
+To be precise, ISO/IEC~8859 used the 7-bit ASCII as the basis for each code page, and defined 16 different 7-bit extensions, leading to $(1+16)\cdot{2^7} = 2,176$ possible characters.}
 
@@
 More wide-ranging, various people in the 1980s started to develop true international code sets.
-In the United States, a group of computer scientists formed the \textsc{Unicode Consortium}, proposing a 16-bit encoding in 1991 (with $2^{16} = 65.536$ different possible characters).
+In the United States, a group of computer scientists formed the \textsc{Unicode Consortium}, proposing a 16-bit encoding in 1991 (with $2^{16} = 65,536$ different possible characters).
 At the same time in Europe, the \textsc{International Organization for Standardization} (ISO) was working on ISO~10646 to supplant the ISO/IEC~8859 standard.
-Their first draft of the \textsc{Universal Character Set} (UCS) in 1990 was 31-bit (with theoretically $2^{31} = 2.147.483.648$ possible characters, but because of some technical restrictions only 679.477.248 were allowed).
+Their first draft of the \textsc{Universal Character Set} (UCS) in 1990 was 31-bit (with theoretically $2^{31} = 2,147,483,648$ possible characters, but because of some technical restrictions only 679,477,248 were allowed).
 
-Since 1991, the Unicode Consortium and the ISO jointly develop the \textsc{unicode standard}, or ISO/IEC~10646, leading to the current system including the original 16-bit Unicode proposal as the \textsc{basic multilingual plane}, and 16 additional planes of 16-bit for further extensions (with in total $(1+16) \cdot 2^{16} = 1.114.112$ possible characters).
-The most recent version of the Unicode Standard (7.0) was published in June 2014 and it defines 112.956 different characters (Unicode Consortium 2014).
+Since 1991, the Unicode Consortium and the ISO jointly develop the \textsc{unicode standard}, or ISO/IEC~10646, leading to the current system including the original 16-bit Unicode proposal as the \textsc{basic multilingual plane}, and 16 additional planes of 16-bit for further extensions (with in total $(1+16) \cdot 2^{16} = 1,114,112$ possible characters).
+The most recent version of the Unicode Standard (7.0) was published in June 2014 and it defines 112,956 different characters \citep{Unicode2014}.

### 2015-05-31 chapters/introduction.tex
edits, bib, adherence to lang sci press formatting, etc.

@@
 
-\section{Introduction}
-\label{introduction}
+% \section{Introduction}
+% \label{introduction}
 
@@
 They tend to be highly conservative, in that people who have learned to read and write in a specific way (however impractical or tedious) are mostly unwilling to change their habits, e.g.~they tend to resist spelling reforms.
-In all literate societies there exists a strong socio-political mainstream that tries to force unification of writing (for example by strongly enforcing `right' from `wrong' writing in schools).
+In all literate societies there exists a strong socio-political mainstream that tries to force unification of writing (for example by strongly enforcing ``right'' from ``wrong'' writing in schools).
 However, there is also a large community of users who take as many liberties in their writing as they can get away with.
 
-For example, the writing of tone diacritics in Yoruba is often proclaimed to be the `right' way to write, although many users of Yoruba writing seem to be perfectly fine with leaving them out.
-As pointed out by the proponents of the `official' rules, there are some homographs when leaving out the tone diacritics (Olúmúyìw 2013: 44).
+For example, the writing of tone diacritics in Yoruba is often proclaimed to be the right way to write, although many users of Yoruba writing seem to be perfectly fine with leaving them out.
+As pointed out by the proponents of the official rules, there are some homographs when leaving out the tone diacritics \citet[44]{Olumuyiw2013}.
 However, writing systems (and the languages they represent) are normally full of homophones, which is normally not a problem at all for speakers of the language.
 More importantly, writing is not just a purely functional tool, but just as importantly it is a mechanism to signal social affiliation.
-By showing that you `know the rules' of expressing yourself in writing, others will more easily accept you as a worthy participant in their group.
-And that just as well holds for obeying to the `official' rules when writing a job application, as for obeying to the `informal' rules when writing an SMS to classmates in school.
-The case of Yoruba writing is an exemplary case, as even after more than a century of efforts to standardize the writing systems, there is still a wide range of variation in daily use (Olúmúyìw 2013).
+By showing that you \textit{know the rules} of expressing yourself in writing, others will more easily accept you as a worthy participant in their group.
+And that just as well holds for obeying to the official rules when writing a job application, as for obeying to the informal rules when writing an SMS to classmates in school.
+The case of Yoruba writing is an exemplary case, as even after more than a century of efforts to standardize the writing systems, there is still a wide range of variation in daily use \citet{Olumuyiw2013}.
 
@@
 The necessity to computationally use and compare different orthographies most forcefully arises in the context of language comparison.
-Concretely, in our current research our goal is to develop quantitative methods for language comparison and historical analysis in order to investigate worldwide linguistic variation and to model the historical and areal processes that underlie linguistic diversity (cf.~Steiner et al.
-2010, List 2012a, 2012b, List \& Moran 2013, Moran \& Prokic 2013).
+Concretely, in our current research our goal is to develop quantitative methods for language comparison and historical analysis in order to investigate worldwide linguistic variation and to model the historical and areal processes that underlie linguistic diversity, cf.~\citet{Steiner_etal2011,List2012,List2012a,ListMoran2013,MoranProkic2013}.
 In this work, it is crucial to be able to flexibly process across numerous resources with different orthographies.
@@
 Our experience is that such a graphemic tokenization can be performed without extensive in-depth knowledge about the phonetic and phonological details of the language in question.
-For example, the specification that <ou> is a grapheme of English is a much easier task than to specify what exactly the phonetic values of this grapheme are in any specific occurrence in English words.
+For example, the specification that $<$ou$>$ is a grapheme of English is a much easier task than to specify what exactly the phonetic values of this grapheme are in any specific occurrence in English words.
 Grapheme separation is a task that can be performed relatively reliably and with limited availability of time and resources (compare, for example, the task of creating a complete phonetic or phonological normalization).
@@
 
-Before we deal with these proposals, in the first part of this paper (Sections 2 through 5) we give an extended introduction to the notion of encoding (Section 2) and writing systems, both from a linguistic perspective and from the perspective of the Unicode Consortium (Section 3).
+Before we deal with these proposals, in the first part of this paper (Sections \ref{encoding} through \ref{ipa-meets-unicode}) we give an extended introduction to the notion of encoding (Section \ref{encoding}) and writing systems, both from a linguistic perspective and from the perspective of the Unicode Consortium (Section \ref{terminology}).
 We consider the Unicode Standard to be a breakthrough (and ongoing) development that fundamentally changed the way we look at writing systems, and we aim to provide here a slightly more in-depth survey of the many techniques that are available in the standard.
-A good appreciation for the solutions that the Unicode Standard also allows for a thorough understanding of possible pitfalls that one might encounter when using it (Section 4).
-As an example of the current state-of-the-art, we discuss the rather problematic marriage of the International Phonetic Alphabet (IPA) with Unicode (Section 5).
+A good appreciation for the solutions that the Unicode Standard also allows for a thorough understanding of possible pitfalls that one might encounter when using it (Section \ref{unicode-pitfalls}).
+As an example of the current state-of-the-art, we discuss the rather problematic marriage of the International Phonetic Alphabet (IPA) with the Unicode Standard (Section \ref{ipa-meets-unicode}).
 
-The second part of the paper (Sections 6 and 7) describes our proposals for how to deal with the Unicode Standard in the daily practice of (comparative) linguists.
+The second part of the paper (Sections \ref{orthography-profiles} and \ref{use-cases}) describes our proposals for how to deal with the Unicode Standard in the daily practice of (comparative) linguists.
 First, we discuss the challenges of characterizing a writing system.
-To solve these problems, we propose the notions of orthography profiles, closely related to Unicode locale descriptions (Section 6).
-Finally, we discuss practical issues with actual examples (Section 7).
+To solve these problems, we propose the notions of orthography profiles, closely related to Unicode locale descriptions (Section \ref{orthography-profiles}).
+Finally, we discuss practical issues with actual examples (Section \ref{use-cases}).
 We provide reference implementation of our proposals in R and in Python, available as open-source libraries.
@@
 The following conventions are followed in this paper.
-All phonemic and phonetic representations are given in the International Phonetic Alphabet (IPA), unless noted otherwise (International Phonetic Association 2005).
+All phonemic and phonetic representations are given in the International Phonetic Alphabet (IPA), unless noted otherwise \citep{IPA2005}.
 Standard conventions are used for distinguishing between graphemic < >, phonemic / / and phonetic [ ] representations.
-For character descriptions, we follow the notational conventions of the Unicode Standard (Unicode Consortium 2014).
+For character descriptions, we follow the notational conventions of the Unicode Standard \citep{Unicode2014}.
 Character names are represented in small capital letters (e.g.~\textsc{latin small letter schwa}) and code points are expressed as U\emph{+n} where \emph{n} is a four to six digit hexadecimal number, e.g.~U+0256, which can be rendered as the glyph <ə>.

### 2015-05-31 chapters/orthography_profiles.tex
edits, bib, adherence to lang sci press formatting, etc.

@@
 Not only can orthography profiles be helpful in the daily practice of linguistics, they also succinctly document the orthographic details of a specific source, and, as such, might fruitfully be published alongside sources (e.g.~in digital archives).
-Also, in high-level linguistic analyses in which the graphemic detail is of central importance (e.g.~phonotactic or comparative-historical studies), orthography profiles can transparently document the decisions that have been taken in the interpretation of the orthography in the sources used.
+Also, in high-level linguistic analyzes in which the graphemic detail is of central importance (e.g.~phonotactic or comparative-historical studies), orthography profiles can transparently document the decisions that have been taken in the interpretation of the orthography in the sources used.
 
-Given these goals, Unicode locale descriptions (see Section 3) might seem like the ideal orthography profiles.
+Given these goals, Unicode locale descriptions (see Section \ref{terminology}) might seem like the ideal orthography profiles.
 However, there are various practical obstacles preventing the use of such locale descriptions in the daily linguistic practice, namely: (i) the XML-structure is too verbose to easily and quickly produce or correct manually, (ii) locale descriptions are designed for a wide scope on information (like date formats or names of weekdays) most of which is not applicable for documenting writing systems, and (iii) most crucially, even if someone made the effort to produce a technically correct locale description for a specific source at hand, then it is nigh impossible to deploy the description.
@@
 As an example of graphemic tokenization, note the three different levels of technological and linguistic elements that interact in the hypothetical lexical form <tsʰṍ̰shi> :
+
 \begin{enumerate}
@@
 
-Informally, an orthography profile specifies the graphemes (or, in Unicode parlance, ``tailored grapheme clusters'') that are expected to occur in any data to be analysed or checked for consistency.
-These graphemes are first identified throughout the whole data (a step which we call `tokenization'), and possibly simply returned as such, possibly including error messages about any parts of the data that are not specified by the orthography profile.
-Once the graphemes are identified, they might also be changed into other graphemes (a step which we call `transliteration').
+Informally, an orthography profile specifies the graphemes (or, in Unicode parlance, \textsc{tailored grapheme clusters}) that are expected to occur in any data to be analyzed or checked for consistency.
+These graphemes are first identified throughout the whole data (a step which we call \textsc{tokenization}), and possibly simply returned as such, possibly including error messages about any parts of the data that are not specified by the orthography profile.
+Once the graphemes are identified, they might also be changed into other graphemes (a step which we call \textsc{transliteration}).
 When a grapheme has different possible transliterations, then these differences should be separated by contextual specification, possibly down to listing individual exceptional cases.
@@
 In practice, we foresee a workflow in which orthography profiles are iteratively refined, while at the same time inconsistencies and errors in the data to be tokenized are corrected.
-In some more complex use-cases there might even be a need for multiple different orthography profiles to be applied in sequence (see Section 8 on various exemplary use-cases).
+In some more complex use-cases there might even be a need for multiple different orthography profiles to be applied in sequence (see Section \ref{use-cases} on various exemplary use-cases).
 The result of any such workflow will normally be a cleaned dataset and an explicit description of the orthographic structure in the form of an orthography profile.
@@
 
-The formal specifications of an orthography profile (or simply `profile' for short) are the following:
+The formal specifications of an orthography profile (or simply \textsc{profile} for short) are the following:
+
 \begin{enumerate}
 	\def\labelenumi{\arabic{enumi}.} 
-	\item \textsc{A profile is a} \textsc{Unicode UTF-8 encoded text file} (ideally using NFC, no-BOM, and LF; see Section 4, Pitfall ``File Formats'') that includes the information pertinent to the orthography.
+	\item \textsc{A profile is a} \textsc{Unicode UTF-8 encoded text file} (ideally using NFC, no-BOM, and LF; see Section \ref{pitfall-file-formats}, Pitfall: File Formats) that includes the information pertinent to the orthography.
 	\item \textsc{A profile is a} \textsc{tab-separated CSV file with an obligatory header line}.
@@
 The actual ordering of the columns is unimportant.
-	\item \textsc{Lines starting with a hash \textless{}\#\textgreater{} are ignored.} Comments and metadata can be included inside the file, but only as complete lines in the profile, to be marked by lines starting with hash ``\#'' (NUMBER SIGN at U+0023).
-Hashes somewhere else in the file are to be treated literally, i.e.~hashes are only to be ignored when they occur at the start of a line.\footnote{Comments that belong to specific lines will have to be put in a separate column of the CSV file, e.g.~add a column called `comments'.
+	\item \textsc{Lines starting with a hash \textless{}\#\textgreater{} are ignored.} Comments and metadata can be included inside the file, but only as complete lines in the profile, to be marked by lines starting with hash \textsc{\#} (\textsc{number sign} at U+0023).
+Hashes somewhere else in the file are to be treated literally, i.e.~hashes are only to be ignored when they occur at the start of a line.\footnote{Comments that belong to specific lines will have to be put in a separate column of the CSV file, e.g.~add a column called \textsc{comments}.
 Further, if the content of a profile contains a hash at the start of a line, either reorder the columns so the hash does not occur at the start of the line, or add a dummy column in front of the data to not have the data start with a hash.}
-	\item \textsc{Metadata are given in commented lines at the beginning of the text file in a basic ``tag: value'' format.
+	\item \textsc{Metadata are given in commented lines at the beginning of the text file in a basic \textsc{tag: value} format.
 }Metadata about the orthographic description given in the orthography profile includes, minimally, (i) author, (ii) date, (iii) title, (iv) a stable language identifier encoded in BCP 47/ISO 639-3, and (v) bibliographic data for resource(s) that illustrate the orthography described in the profile.
@@
 The content of a profile consists of lines, each describing a grapheme of the orthography, using the following columns:
+
 \begin{enumerate}
 	\def\labelenumi{\arabic{enumi}.} 
-	\item \textsc{A minimal profile consists of a single column with header `graphemes'}, listing each of the different graphemes in a separate line. 
+	\item \textsc{A minimal profile consists of a single column with header \textsc{graphemes}}, listing each of the different graphemes in a separate line. 
 	\item \textsc{Optional columns called `left' and `right' can be used to specify the left and right context of the grapheme, respectively.} The same grapheme can occur multiple times with different contextual specifications, for example to distinguish different pronunciations depending on the context.
-	\item \textsc{The columns `grapheme', `left' and `right' can use regular expression metacharacters.} If regular expressions are used, then all literal usage of the special symbols, like full stops ``.'' or dollar signs ``\$'' (so-called ``metacharacters'') have to be explicitly escaped by adding a backslash before them (i.e.~use ``.'' or ``\$'').
+	\item \textsc{The columns `grapheme', `left' and `right' can use regular expression metacharacters.} If regular expressions are used, then all literal usage of the special symbols, like full stops <.> or dollar signs <\$> (so-called \textsc{metacharacters}) have to be explicitly escaped by adding a backslash before them (i.e.~use <.> or <\$>).
 Note that any specification of context automatically expects regular expressions, so it is probably better to always escape all regular expression metacharacters when used literally in the orthography, i.e.~the following symbols will need to be preceded by a backslash: {[} {]} ( ) \{ \} ~+ * .
@@
 For the automatic processing of the profiles, the following technical standards will be expected:
+
 \begin{enumerate}
@@
 However, this only is possible when no contexts or classes are described, so this seems only useful in the most basic orthographies.
-	\item \textsc{The `class' column will be used to produce explicit ``or'' chains of regular expressions}, which will then be inserted in the `graphemes', `left' and `right' columns at the position indicated by the class-identifiers.
-For example, a class ``V'' as a context specification might be replaced by a regular expression like ``(a\textbar{}e\textbar{}i\textbar{}o\textbar{}u\textbar{}ei\textbar{}au)''.
+	\item \textsc{The `class' column will be used to produce explicit \textsc{or} chains of regular expressions}, which will then be inserted in the `graphemes', `left' and `right' columns at the position indicated by the class-identifiers.
+For example, a class \textsc{V} as a context specification might be replaced by a regular expression like: (a\textbar{}e\textbar{}i\textbar{}o\textbar{}u\textbar{}ei\textbar{}au).
 Only the graphemes themselves are included here, not any contexts specified for the elements of the class.
@@
 Basically, the actual regular expression syntax of lookbehind and lookahead is simply hidden to the users by allowing them to only specify the contexts themselves.
-Internally, the contexts in the columns ``left'' and ``right'' are combined with the column ``graphemes'' to form a complex regular expression like ``(?\textless{}=left)graphemes(?=right)''.
+Internally, the contexts in the columns \textsc{left} and \textsc{right} are combined with the column \textsc{graphemes} to form a complex regular expression like: (?\textless{}=left)graphemes(?=right).
 	\item \textsc{The regular expressions will be applied in the order as specified in the profile, from top to bottom.} A software implementation can offer help in figuring out the optimal ordering of the regular expressions, but should then explicitly report on the order used.
@@
 The actual implementation of the profile on some text-string will function as follows:
+
 \begin{enumerate}
@@
 	\item \textsc{All graphemes are matched in the text before they are tokenized or transliterated}.
-In this way, there is no necessity for the user to consider `feeding' and `bleeding' situations, in which the application of a rule either changes the text so another rule suddenly applies (``feeding'') or prevents another rule to apply (`bleeding').
+In this way, there is no necessity for the user to consider `feeding' and `bleeding' situations, in which the application of a rule either changes the text so another rule suddenly applies (feeding) or prevents another rule to apply (`bleeding').
 	\item \textsc{The matching of the graphemes can occur either globally or linearly.
@@
 Still, we suggest that any software application using orthography profiles should offer both approaches (i.e.
-``global'' or ``linear'') to the user.
-The approach used should be documented in the metadata as ``tokenization method''.
+\textsc{global} or \textsc{linear}) to the user.
+The approach used should be documented in the metadata as \textsc{tokenization method}.
 	\item \textsc{The matching of the graphemes can occur either in NFC or NFD.
-}By default, both the profile and the text-string to be tokenized should be treated as NFC (see section 4, Pitfall ``Canonical equivalence'' above).
+}By default, both the profile and the text-string to be tokenized should be treated as NFC (see section \ref{pitfall-canonical-equivalence}, Pitfall: Canonical equivalence, above).
 However, in some use-cases it turns out to be practical to treat both text and profile as NFD.
@@
 We suggest that any software application using orthography profiles should offer both approaches (i.e.
-``NFC'' or ``NFD'') to the user.
-The approach used should be documented in the metadata as ``unicode normalization''.
+\textsc{NFC} or \textsc{NFD}) to the user.
+The approach used should be documented in the metadata as \textsc{unicode normalization}.
 	\item \textsc{The text-string is always returned in tokenized form} by separating the matched graphemes by a user-specified symbols-string.
@@
 Any software application offering to use orthography profile:
+
 \begin{enumerate}
@@
 		\item \textsc{the symbol-string to be inserted for unmatched strings} in the tokenized and transliterated output. 
-		\item \textsc{the tokenization method}, i.e.~whether the tokenization should proceed ``global'' or ``linear''. 
-		\item \textsc{unicode normalization}, i.e.~whether the text-string and profile should use ``NFC'' or ``NFD''. 
+		\item \textsc{the tokenization method}, i.e.~whether the tokenization should proceed \textsc{global} or \textsc{linear}. 
+		\item \textsc{unicode normalization}, i.e.~whether the text-string and profile should use \textsc{NFC} or \textsc{NFD}. 
 	\end{enumerate}

### 2015-05-31 chapters/use_cases.tex
edits, bib, adherence to lang sci press formatting, etc.

@@
-\section{Use cases}\label{use-cases}
+\chapter{Use cases}
+\label{use-cases}
+
+\section{Introduction}
 
@@
 \begin{itemize}
-	\item grapheme tokenization 
+	\item tokenization and error checking
 	\item normalization of orthographic systems to attain interoperability across different source documents, 
@@
 
-\emph{Grapheme tokenization}
+\section{Tokenization and error checking}
+The most basic use case is tokenization of language data. Tokenization comes in three types:
+
+- Unicode character tokenization
+- grapheme tokenization
+- tailored grapheme tokenization
+
+The simplest form of tokenization provided by any Unicode Standard-compliant software will split an input text stream on Unicode character code points.
+The input text is split on the character bit sequences as they have been encoded by the user.
+This may be in various Unicode Normalization Forms (see Section \ref{}) and tokenization split function returns a byte stream sequence given a particular encoding (see Section \ref{}).
+
+There is increasing support for the regular expression match, commonly ``\\X'', to identify Unicode graphemes (see Section \ref{}) and perform tokenization on these sets of Unicode characters.
+Examples are given in Section \ref{}.
+
+The orthography profile provides the mechanism for tokenizing an input stream on tailored grapheme clusters.
+The orthography's formal specification is given in Section \ref{}.
+To sum, an orthography profile is a list of tailored grapheme clusters and / or orthographic rules that specify how a specific resource of text input should be tokenized.
 
@@
 
-*Cross-orthography **normalization*
+\begin{comment}
+
+\section{Cross-orthography **normalization*}
 
@@
 
-A preliminary orthography profile can be generated with code available online.\footnote{https://github.com/bambooforest/tokenizer} The Python script creates a unigram model for individual Unicode characters or Unicode grapheme clusters and their frequency in the source input.
+A preliminary orthography profile can be generated with code available online.\footnote{\url{https://github.com/bambooforest/tokenizer}} The Python script creates a unigram model for individual Unicode characters or Unicode grapheme clusters and their frequency in the source input.
 We find that this gives a broad overview of the input at the character and grapheme levels and it often finds typographic errors in the source input by identifying rarely occurring or one-off characters in the input.
@@
 Grapheme clusters are important to determine word or line boundaries, for collation (`ordering') and for user interface interaction (e.g.~mouse selection, backspacing, cursor movement).
+
+\end{comment}

### 2015-06-05 chapters/encoding.tex
wraps line at character width 80 in all tex files

@@
-\chapter{Encoding}
-\label{encoding}
+\chapter{Encoding} \label{encoding}
 
-There are many in-depth histories of the origin and development of writing systems (e.g.~\citet{Robinson1997,Powell2012}), a story that we will not repeat here.
-However, the history of turning writing into computer readable code is not so often told, so we decided to offer a short survey of the major developments here.
-This history turns out to be intimately related to the history of telegraphic communication.\footnote{Because of the recent history as summarized in this section, we have used mostly rather ephemeral internet sources.
-When not references by traditional literature in the bibliography, we have specifically used \url{http://www.unicode.org/history/} and various Wikipedia pages for the information presented here.
-A useful survey of the historical development of the physical hardware of telegraphy and telecommunication is \citet{Huurdeman2003}.
-Most books that discuss the development of encoding of telegraphic communication focus of cryptography, e.g.~\citet{Singh1999}, and forego the rather interesting story of `open' encoding that is related here.}
+There are many in-depth histories of the origin and development of writing
+systems (e.g.~\citet{Robinson1997,Powell2012}), a story that we will not repeat
+here. However, the history of turning writing into computer readable code is not
+so often told, so we decided to offer a short survey of the major developments
+here. This history turns out to be intimately related to the history of
+telegraphic communication.\footnote{Because of the recent history as summarized
+in this section, we have used mostly rather ephemeral internet sources. When not
+references by traditional literature in the bibliography, we have specifically
+used \url{http://www.unicode.org/history/} and various Wikipedia pages for the
+information presented here. A useful survey of the historical development of the
+physical hardware of telegraphy and telecommunication is \citet{Huurdeman2003}.
+Most books that discuss the development of encoding of telegraphic communication
+focus of cryptography, e.g.~\citet{Singh1999}, and forego the rather interesting
+story of ``open encoding'' that is related here.}
 
-Writing systems have existed for roughly 6000 years, allowing people to exchange messages through time and space.
-Additionally, to bridge large geographic distances, \textsc{telegraphic systems} of communication (from Greek \emph{τῆλε γράφειν} `distant writing') have a long and widespread history since ancient times.
-The most widespread telegraphic systems worldwide are so-called whistled languages \citep{Meyer2015}, but also drumming languages \citep{Meyer_etal2012} and smoke, fire, hydraulic or flag signals are forms of telegraphy.
+Writing systems have existed for roughly 6000 years, allowing people to exchange
+messages through time and space. Additionally, to bridge large geographic
+distances, \textsc{telegraphic systems} of communication (from Greek \emph{τῆλε
+γράφειν} `distant writing') have a long and widespread history since ancient
+times. The most widespread telegraphic systems worldwide are so-called whistled
+languages \citep{Meyer2015}, but also drumming languages \citep{Meyer_etal2012}
+and smoke, fire, hydraulic or flag signals are forms of telegraphy.
 
-Telegraphy was reinvigorated in the end of the eighteenth century through the introduction of semaphoric systems by Claude Chapelle (since then using flags, flashing lights, or various specially designed contraptions) to convey messages over large distances.
-The \textit{innovation} of those systems was that all characters of the written language were replaced one-to-one by visual signals.
-Since then, all telegraphic systems have taken this principle, namely that any language to be transmitted first has to be turned into some orthographic system, which subsequently is encoded for transmission by the sender, and then turned back into orthographic representation at the receiver side.\footnote{Sound and video-based telecommunication of course takes a different approach by ignoring the written version of language and directly encode sound waves or light patterns.} This of course implies that the usefulness of any such telegraphic encoding completely depends on the sometimes rather haphazard structure of orthographic systems.
+Telegraphy was reinvigorated in the end of the eighteenth century through the
+introduction of semaphoric systems by Claude Chapelle (since then using flags,
+flashing lights, or various specially designed contraptions) to convey messages
+over large distances. The \textit{innovation} of those systems was that all
+characters of the written language were replaced one-to-one by visual signals.
+Since then, all telegraphic systems have taken this principle, namely that any
+language to be transmitted first has to be turned into some orthographic system,
+which subsequently is encoded for transmission by the sender, and then turned
+back into orthographic representation at the receiver side.\footnote{Sound and
+video-based telecommunication of course takes a different approach by ignoring
+the written version of language and directly encode sound waves or light
+patterns.} This of course implies that the usefulness of any such telegraphic
+encoding completely depends on the sometimes rather haphazard structure of
+orthographic systems.
 
-In the nineteenth century, \textsc{electric telegraphy} lead to a new approach in which written language characters were encoded by signals sent through a copper wire.
-Originally, \textsc{bisignal} codes were used, consisting of two different signals.
-For example, Carl Friedrich Gauss in 1833 used positive and negative current \citep[282]{Mania2008}.
-More famous and influential, Samuel Morse in 1836 used long and short pulses.
+In the nineteenth century, \textsc{electric telegraphy} lead to a new approach
+in which written language characters were encoded by signals sent through a
+copper wire. Originally, \textsc{bisignal} codes were used, consisting of two
+different signals. For example, Carl Friedrich Gauss in 1833 used positive and
+negative current \citep[282]{Mania2008}. More famous and influential, Samuel
+Morse in 1836 used long and short pulses.
 
-In those \textsc{bisignal codes} each character from the written language was encoded with a different number of signals (between one and five), so two different separators are needed: one between signals and one between characters.
-For example, in Morse-code there is a short pause between signals and a long pause between characters.
-Actually, Morse-code also includes an extra long pause between words.
-Interestingly, it took a long time to consider the written word boundary---using white-space---as a bona-fide character that should simply be encoded with its own code point.
-This happened only with the revision of the Baudot-code (see below) by Donald Murray in 1901, in which he introduced a specific white-space code.
-This principle has been followed ever since.
+In those \textsc{bisignal codes} each character from the written language was
+encoded with a different number of signals (between one and five), so two
+different separators are needed: one between signals and one between characters.
+For example, in Morse-code there is a short pause between signals and a long
+pause between characters. Actually, Morse-code also includes an extra long pause
+between words. Interestingly, it took a long time to consider the written word
+boundary---using white-space---as a bona-fide character that should simply be
+encoded with its own code point. This happened only with the revision of the
+Baudot-code (see below) by Donald Murray in 1901, in which he introduced a
+specific white-space code. This principle has been followed ever since.
 
-From those bisignal encodings, true \textsc{binary codes} developed with a fixed length of signals per character.
-In such systems only a single separator between signals is needed, because the separation between characters can be established by counting until a fixed number of signals.\footnote{Of course, no explicit separator is needed when the timing of the signals is known, which is the principle used in all modern telecommunication systems.
-An important modern consideration is also how to know where to start counting when you did not catch the start of a message, something that is known in Unicode as \textsc{self synchronization}.}
+From those bisignal encodings, true \textsc{binary codes} developed with a fixed
+length of signals per character. In such systems only a single separator between
+signals is needed, because the separation between characters can be established
+by counting until a fixed number of signals.\footnote{Of course, no explicit
+separator is needed when the timing of the signals is known, which is the
+principle used in all modern telecommunication systems. An important modern
+consideration is also how to know where to start counting when you did not catch
+the start of a message, something that is known in Unicode as \textsc{self
+synchronization}.}
 
-In the context of electric telegraphy, such a binary code system was first established by Émile Baudot in 1870, using a fixed combination of five signals for each written character.\footnote{True binary codes have a longer history, going at least back to the Baconian cipher devised by Francis Bacon in 1605.
-However, the proposal by Baudot was the quintessential proposal leading to all modern systems.} There are $2^5 = 32$ possible combination when using five binary signals; an encoding today designated as \textsc{5-bit}.
-These codes are sufficient for all Latin letters, but of course they do not suffice for all written symbols, including punctuation and digits.
-As a solution, the Baudot code uses a so-called \textsc{shift} character, which signifies that from that point onwards---until shifted back---a different encoding is used, allowing for yet another set of 32 codes.
-In effect, this means that the Baudot code, and the \textsc{International Telegraph Alphabet} (ITA) derived from it, had an extra `bit' of information, so the encoding is actually 6-bit (with $2^6 = 64$ different possible characters).
-For decades, this encoding was the standard for all telegraphy and it is still in limited use today.
+In the context of electric telegraphy, such a binary code system was first
+established by Émile Baudot in 1870, using a fixed combination of five signals
+for each written character.\footnote{True binary codes have a longer history,
+going at least back to the Baconian cipher devised by Francis Bacon in 1605.
+However, the proposal by Baudot was the quintessential proposal leading to all
+modern systems.} There are $2^5 = 32$ possible combination when using five
+binary signals; an encoding today designated as \textsc{5-bit}. These codes are
+sufficient for all Latin letters, but of course they do not suffice for all
+written symbols, including punctuation and digits. As a solution, the Baudot
+code uses a so-called \textsc{shift} character, which signifies that from that
+point onwards---until shifted back---a different encoding is used, allowing for
+yet another set of 32 codes. In effect, this means that the Baudot code, and the
+\textsc{International Telegraph Alphabet} (ITA) derived from it, had an extra
+\textsc{bit} of information, so the encoding is actually 6-bit (with $2^6 = 64$
+different possible characters). For decades, this encoding was the standard for
+all telegraphy and it is still in limited use today.
 
-To also allow for different uppercase and lowercase letters and a large variety of control characters to be used in the newly developing technology of computers, the American Standards Association decided to propose a new 7-bit encoding in 1963 (with $2^7 = 128$ different possible characters), known as the \textsc{American Standard Code for Information Interchange} (ASCII), geared towards the encoding of English orthography.
+To also allow for different uppercase and lowercase letters and a large variety
+of control characters to be used in the newly developing technology of
+computers, the American Standards Association decided to propose a new 7-bit
+encoding in 1963 (with $2^7 = 128$ different possible characters), known as the
+\textsc{American Standard Code for Information Interchange} (ASCII), geared
+towards the encoding of English orthography.
 
-With the ascent of other orthographies in computer usage, the wish to encode further variation of Latin letters (like German <ß> or various letters with diacritics like <è>) led the Digital Equipment Corporation to introduce an 8-bit \textsc{Multinational Character Set} (MCS, with $2^8 = 256$ different possible characters), first used with the introduction of the {\small VT}220 Terminal in 1983.
+With the ascent of other orthographies in computer usage, the wish to encode
+further variation of Latin letters (like German <ß> or various letters with
+diacritics like <è>) led the Digital Equipment Corporation to introduce an 8-bit
+\textsc{Multinational Character Set} (MCS, with $2^8 = 256$ different possible
+characters), first used with the introduction of the {\small VT}220 Terminal in
+1983.
 
-Because 256 characters were clearly not enough for the many different characters needed in the world's writing systems, the ISO/IEC~8859 standard in 1987 extended the MCS to include 16 different 8-bit code pages.
-For example, part 5 was used for Cyrillic characters, part 6 for Arabic, and part 7 for Greek.\footnote{In effect, because $16 = 2^4$, this means that ISO/IEC~8859 was actually a $8+4=12$-bit encoding, though with very many duplicates by design (e.g.~all ASCII codes were repeated in each 8-bit code page).
-To be precise, ISO/IEC~8859 used the 7-bit ASCII as the basis for each code page, and defined 16 different 7-bit extensions, leading to $(1+16)\cdot{2^7} = 2,176$ possible characters.}
+Because 256 characters were clearly not enough for the many different characters
+needed in the world's writing systems, the ISO/IEC~8859 standard in 1987
+extended the MCS to include 16 different 8-bit code pages. For example, part 5
+was used for Cyrillic characters, part 6 for Arabic, and part 7 for
+Greek.\footnote{In effect, because $16 = 2^4$, this means that ISO/IEC~8859 was
+actually a $8+4=12$-bit encoding, though with very many duplicates by design
+(e.g.~all ASCII codes were repeated in each 8-bit code page). To be precise,
+ISO/IEC~8859 used the 7-bit ASCII as the basis for each code page, and defined
+16 different 7-bit extensions, leading to $(1+16)\cdot{2^7} = 2,176$ possible
+characters.}
 
-This system almost immediately was understood to be insufficient and impractical, so various initiatives to extend and reorganize the encoding started in the 1980s.
-This led, for example, to various proprietary encodings from Microsoft (e.g.~Windows Latin 1) and Apple (e.g.~Mac OS Roman), which one still sometimes encounters today.
+This system almost immediately was understood to be insufficient and
+impractical, so various initiatives to extend and reorganize the encoding
+started in the 1980s. This led, for example, to various proprietary encodings
+from Microsoft (e.g.~Windows Latin 1) and Apple (e.g.~Mac OS Roman), which one
+still sometimes encounters today.
 
-More wide-ranging, various people in the 1980s started to develop true international code sets.
-In the United States, a group of computer scientists formed the \textsc{Unicode Consortium}, proposing a 16-bit encoding in 1991 (with $2^{16} = 65,536$ different possible characters).
-At the same time in Europe, the \textsc{International Organization for Standardization} (ISO) was working on ISO~10646 to supplant the ISO/IEC~8859 standard.
-Their first draft of the \textsc{Universal Character Set} (UCS) in 1990 was 31-bit (with theoretically $2^{31} = 2,147,483,648$ possible characters, but because of some technical restrictions only 679,477,248 were allowed).
+More wide-ranging, various people in the 1980s started to develop true
+international code sets. In the United States, a group of computer scientists
+formed the \textsc{Unicode Consortium}, proposing a 16-bit encoding in 1991
+(with $2^{16} = 65,536$ different possible characters). At the same time in
+Europe, the \textsc{International Organization for Standardization} (ISO) was
+working on ISO~10646 to supplant the ISO/IEC~8859 standard. Their first draft of
+the \textsc{Universal Character Set} (UCS) in 1990 was 31-bit (with
+theoretically $2^{31} = 2,147,483,648$ possible characters, but because of some
+technical restrictions only 679,477,248 were allowed).
 
-Since 1991, the Unicode Consortium and the ISO jointly develop the \textsc{unicode standard}, or ISO/IEC~10646, leading to the current system including the original 16-bit Unicode proposal as the \textsc{basic multilingual plane}, and 16 additional planes of 16-bit for further extensions (with in total $(1+16) \cdot 2^{16} = 1,114,112$ possible characters).
-The most recent version of the Unicode Standard (7.0) was published in June 2014 and it defines 112,956 different characters \citep{Unicode2014}.
+Since 1991, the Unicode Consortium and the ISO jointly develop the
+\textsc{unicode standard}, or ISO/IEC~10646, leading to the current system
+including the original 16-bit Unicode proposal as the \textsc{basic multilingual
+plane}, and 16 additional planes of 16-bit for further extensions (with in total
+$(1+16) \cdot 2^{16} = 1,114,112$ possible characters). The most recent version
+of the Unicode Standard (7.0) was published in June 2014 and it defines 112,956
+different characters \citep{Unicode2014}.

### 2015-06-05 chapters/introduction.tex
wraps line at character width 80 in all tex files

@@
-\chapter{Writing Systems}
-\label{writing_systems}
+\chapter{Writing Systems} \label{writing_systems}
 
-% \section{Introduction}
-% \label{introduction}
+% \section{Introduction} % \label{introduction}
 
-Writing systems arise and develop in a complex mixture of cultural, technological and practical pressures.
-They tend to be highly conservative, in that people who have learned to read and write in a specific way (however impractical or tedious) are mostly unwilling to change their habits, e.g.~they tend to resist spelling reforms.
-In all literate societies there exists a strong socio-political mainstream that tries to force unification of writing (for example by strongly enforcing ``right'' from ``wrong'' writing in schools).
-However, there is also a large community of users who take as many liberties in their writing as they can get away with.
+Writing systems arise and develop in a complex mixture of cultural,
+technological and practical pressures. They tend to be highly conservative, in
+that people who have learned to read and write in a specific way (however
+impractical or tedious) are mostly unwilling to change their habits, e.g.~they
+tend to resist spelling reforms. In all literate societies there exists a strong
+socio-political mainstream that tries to force unification of writing (for
+example by strongly enforcing ``right'' from ``wrong'' writing in schools).
+However, there is also a large community of users who take as many liberties in
+their writing as they can get away with.
 
-For example, the writing of tone diacritics in Yoruba is often proclaimed to be the right way to write, although many users of Yoruba writing seem to be perfectly fine with leaving them out.
-As pointed out by the proponents of the official rules, there are some homographs when leaving out the tone diacritics \citet[44]{Olumuyiw2013}.
-However, writing systems (and the languages they represent) are normally full of homophones, which is normally not a problem at all for speakers of the language.
-More importantly, writing is not just a purely functional tool, but just as importantly it is a mechanism to signal social affiliation.
-By showing that you \textit{know the rules} of expressing yourself in writing, others will more easily accept you as a worthy participant in their group.
-And that just as well holds for obeying to the official rules when writing a job application, as for obeying to the informal rules when writing an SMS to classmates in school.
-The case of Yoruba writing is an exemplary case, as even after more than a century of efforts to standardize the writing systems, there is still a wide range of variation in daily use \citet{Olumuyiw2013}.
+For example, the writing of tone diacritics in Yoruba is often proclaimed to be
+the right way to write, although many users of Yoruba writing seem to be
+perfectly fine with leaving them out. As pointed out by the proponents of the
+official rules, there are some homographs when leaving out the tone diacritics
+\citet[44]{Olumuyiw2013}. However, writing systems (and the languages they
+represent) are normally full of homophones, which is normally not a problem at
+all for speakers of the language. More importantly, writing is not just a purely
+functional tool, but just as importantly it is a mechanism to signal social
+affiliation. By showing that you \textit{know the rules} of expressing yourself
+in writing, others will more easily accept you as a worthy participant in their
+group. And that just as well holds for obeying to the official rules when
+writing a job application, as for obeying to the informal rules when writing an
+SMS to classmates in school. The case of Yoruba writing is an exemplary case, as
+even after more than a century of efforts to standardize the writing systems,
+there is still a wide range of variation in daily use \citet{Olumuyiw2013}.
 
-The sometimes cumbersome and sometimes illogical structure, and the enormous variability of existing writing systems is a fact of life scholars have to accept and should try to adapt to as good as possible.
-Our plea here is a proposal for a formalization to do exactly that.
+The sometimes cumbersome and sometimes illogical structure, and the enormous
+variability of existing writing systems is a fact of life scholars have to
+accept and should try to adapt to as good as possible. Our plea here is a
+proposal for a formalization to do exactly that.
 
-When considering the worldwide linguistic diversity, including all lesser-studied and endangered languages, there exist numerous different orthographies using symbols from the same scripts.
-For example, there are hundreds of orthographies using Latin-based alphabetic scripts.
-All of these orthographies use the same symbols, but these symbols differ in meaning and usage throughout the various orthographies.
-To be able to computationally use and compare different orthographies, we need a way to specify all orthographic idiosyncrasies in a computer-readable format (a process called `tailoring' in Unicode parlance).
-We call such specifications \textsc{orthography profiles}.
-Ideally, these specifications have to be integrated into so-called Unicode locale descriptions, though we will argue that in practice this is often not the most useful solution for the kind of problems arising in the daily practice of linguistics.
-Consequently, a central goal of this paper is to flesh out the linguistic challenges for locale descriptions, and work out suggestions to improve their structure for usage in a linguistic context.
-Conversely, we also aim to improve linguists' understanding and appreciation for the accomplishments of the Unicode Consortium in the development of the Unicode Standard.
+When considering the worldwide linguistic diversity, including all
+lesser-studied and endangered languages, there exist numerous different
+orthographies using symbols from the same scripts. For example, there are
+hundreds of orthographies using Latin-based alphabetic scripts. All of these
+orthographies use the same symbols, but these symbols differ in meaning and
+usage throughout the various orthographies. To be able to computationally use
+and compare different orthographies, we need a way to specify all orthographic
+idiosyncrasies in a computer-readable format (a process called
+\textsc{tailoring} in Unicode parlance). We call such specifications
+\textsc{orthography profiles}. Ideally, these specifications have to be
+integrated into so-called Unicode locale descriptions, though we will argue that
+in practice this is often not the most useful solution for the kind of problems
+arising in the daily practice of linguistics. Consequently, a central goal of
+this paper is to flesh out the linguistic challenges for locale descriptions,
+and work out suggestions to improve their structure for usage in a linguistic
+context. Conversely, we also aim to improve linguists' understanding and
+appreciation for the accomplishments of the Unicode Consortium in the
+development of the Unicode Standard.
 
-The necessity to computationally use and compare different orthographies most forcefully arises in the context of language comparison.
-Concretely, in our current research our goal is to develop quantitative methods for language comparison and historical analysis in order to investigate worldwide linguistic variation and to model the historical and areal processes that underlie linguistic diversity, cf.~\citet{Steiner_etal2011,List2012,List2012a,ListMoran2013,MoranProkic2013}.
-In this work, it is crucial to be able to flexibly process across numerous resources with different orthographies.
-In many cases even different resources on the `same' language use different orthographic conventions.
-Another orthographic challenge that we encounter regularly in our linguistic practice is electronic resources on a particular language that claim to follow a specific orthographic convention (often a resource-specific convention), but on closer inspection such resources are almost always not consistently encoded.
-Thus a second goal of our orthography profiles is to allow for an easy specification of orthographic conventions, and use such profiles to check consistency and to report errors to be corrected.
+The necessity to computationally use and compare different orthographies most
+forcefully arises in the context of language comparison. Concretely, in our
+current research our goal is to develop quantitative methods for language
+comparison and historical analysis in order to investigate worldwide linguistic
+variation and to model the historical and areal processes that underlie
+linguistic diversity,
+cf.~\citet{Steiner_etal2011,List2012,List2012a,ListMoran2013,MoranProkic2013}.
+In this work, it is crucial to be able to flexibly process across numerous
+resources with different orthographies. In many cases even different resources
+on the \textit{same} language use different orthographic conventions. Another
+orthographic challenge that we encounter regularly in our linguistic practice is
+electronic resources on a particular language that claim to follow a specific
+orthographic convention (often a resource-specific convention), but on closer
+inspection such resources are almost always not consistently encoded. Thus a
+second goal of our orthography profiles is to allow for an easy specification of
+orthographic conventions, and use such profiles to check consistency and to
+report errors to be corrected.
 
-A central step in our proposed solution to this problem is the tailored grapheme separation of strings of symbols, a process we call \textsc{grapheme tokenization}.
-Basically, given some strings of symbols (e.g.~morphemes, words, sentences) in a specific source, our first processing step is to specify how these strings have to be separated into graphemes, considering the specific orthographic conventions used in a particular source document.
-Our experience is that such a graphemic tokenization can be performed without extensive in-depth knowledge about the phonetic and phonological details of the language in question.
-For example, the specification that $<$ou$>$ is a grapheme of English is a much easier task than to specify what exactly the phonetic values of this grapheme are in any specific occurrence in English words.
-Grapheme separation is a task that can be performed relatively reliably and with limited availability of time and resources (compare, for example, the task of creating a complete phonetic or phonological normalization).
+A central step in our proposed solution to this problem is the tailored grapheme
+separation of strings of symbols, a process we call \textsc{grapheme
+tokenization}. Basically, given some strings of symbols (e.g.~morphemes, words,
+sentences) in a specific source, our first processing step is to specify how
+these strings have to be separated into graphemes, considering the specific
+orthographic conventions used in a particular source document. Our experience is
+that such a graphemic tokenization can be performed without extensive in-depth
+knowledge about the phonetic and phonological details of the language in
+question. For example, the specification that $<$ou$>$ is a grapheme of English
+is a much easier task than to specify what exactly the phonetic values of this
+grapheme are in any specific occurrence in English words. Grapheme separation is
+a task that can be performed relatively reliably and with limited availability
+of time and resources (compare, for example, the task of creating a complete
+phonetic or phonological normalization).
 
-Although grapheme tokenization is only one part of the solution, it is an important and highly fruitful processing step.
-Given a grapheme tokenization, various subsequent tasks become easier, like (a) temporarily reducing the orthography in a processing pipeline, e.g.~only distinguishing high versus low vowels; (b) normalizing orthographies across sources (often including temporary reduction of oppositions), e.g.~specifying an (approximate) mapping to the International Phonetic Alphabet; (c) using co-occurrence statistics across different languages (or different sources in the same language) to estimate the probability of grapheme matches, e.g.~with the goal to find regular sound changes between related languages or transliterations between different sources in the same language.
+Although grapheme tokenization is only one part of the solution, it is an
+important and highly fruitful processing step. Given a grapheme tokenization,
+various subsequent tasks become easier, like (a) temporarily reducing the
+orthography in a processing pipeline, e.g.~only distinguishing high versus low
+vowels; (b) normalizing orthographies across sources (often including temporary
+reduction of oppositions), e.g.~specifying an (approximate) mapping to the
+International Phonetic Alphabet; (c) using co-occurrence statistics across
+different languages (or different sources in the same language) to estimate the
+probability of grapheme matches, e.g.~with the goal to find regular sound
+changes between related languages or transliterations between different sources
+in the same language.
 
-Before we deal with these proposals, in the first part of this paper (Sections \ref{encoding} through \ref{ipa-meets-unicode}) we give an extended introduction to the notion of encoding (Section \ref{encoding}) and writing systems, both from a linguistic perspective and from the perspective of the Unicode Consortium (Section \ref{terminology}).
-We consider the Unicode Standard to be a breakthrough (and ongoing) development that fundamentally changed the way we look at writing systems, and we aim to provide here a slightly more in-depth survey of the many techniques that are available in the standard.
-A good appreciation for the solutions that the Unicode Standard also allows for a thorough understanding of possible pitfalls that one might encounter when using it (Section \ref{unicode-pitfalls}).
-As an example of the current state-of-the-art, we discuss the rather problematic marriage of the International Phonetic Alphabet (IPA) with the Unicode Standard (Section \ref{ipa-meets-unicode}).
+Before we deal with these proposals, in the first part of this paper (Sections
+\ref{encoding} through \ref{ipa-meets-unicode}) we give an extended introduction
+to the notion of encoding (Section \ref{encoding}) and writing systems, both
+from a linguistic perspective and from the perspective of the Unicode Consortium
+(Section \ref{terminology}). We consider the Unicode Standard to be a
+breakthrough (and ongoing) development that fundamentally changed the way we
+look at writing systems, and we aim to provide here a slightly more in-depth
+survey of the many techniques that are available in the standard. A good
+appreciation for the solutions that the Unicode Standard also allows for a
+thorough understanding of possible pitfalls that one might encounter when using
+it (Section \ref{unicode-pitfalls}). As an example of the current
+state-of-the-art, we discuss the rather problematic marriage of the
+International Phonetic Alphabet (IPA) with the Unicode Standard (Section
+\ref{ipa-meets-unicode}).
 
-The second part of the paper (Sections \ref{orthography-profiles} and \ref{use-cases}) describes our proposals for how to deal with the Unicode Standard in the daily practice of (comparative) linguists.
-First, we discuss the challenges of characterizing a writing system.
-To solve these problems, we propose the notions of orthography profiles, closely related to Unicode locale descriptions (Section \ref{orthography-profiles}).
-Finally, we discuss practical issues with actual examples (Section \ref{use-cases}).
-We provide reference implementation of our proposals in R and in Python, available as open-source libraries.
+The second part of the paper (Sections \ref{orthography-profiles} and
+\ref{use-cases}) describes our proposals for how to deal with the Unicode
+Standard in the daily practice of (comparative) linguists. First, we discuss the
+challenges of characterizing a writing system. To solve these problems, we
+propose the notions of orthography profiles, closely related to Unicode locale
+descriptions (Section \ref{orthography-profiles}). Finally, we discuss practical
+issues with actual examples (Section \ref{use-cases}). We provide reference
+implementation of our proposals in R and in Python, available as open-source
+libraries.
 
-The following conventions are followed in this paper.
-All phonemic and phonetic representations are given in the International Phonetic Alphabet (IPA), unless noted otherwise \citep{IPA2005}.
-Standard conventions are used for distinguishing between graphemic < >, phonemic / / and phonetic [ ] representations.
-For character descriptions, we follow the notational conventions of the Unicode Standard \citep{Unicode2014}.
-Character names are represented in small capital letters (e.g.~\textsc{latin small letter schwa}) and code points are expressed as U\emph{+n} where \emph{n} is a four to six digit hexadecimal number, e.g.~U+0256, which can be rendered as the glyph <ə>.
+The following conventions are followed in this paper. All phonemic and phonetic
+representations are given in the International Phonetic Alphabet (IPA), unless
+noted otherwise \citep{IPA2005}. Standard conventions are used for
+distinguishing between graphemic < >, phonemic / / and phonetic [ ]
+representations. For character descriptions, we follow the notational
+conventions of the Unicode Standard \citep{Unicode2014}. Character names are
+represented in small capital letters (e.g.~\textsc{latin small letter schwa})
+and code points are expressed as U\emph{+n} where \emph{n} is a four to six
+digit hexadecimal number, e.g.~U+0256, which can be rendered as the glyph <ə>.

### 2015-06-05 chapters/ipa_meets_unicode.tex
wraps line at character width 80 in all tex files

@@
-\chapter{IPA meets Unicode}
-\label{ipa-meets-unicode}
+\chapter{IPA meets Unicode} \label{ipa-meets-unicode}
 
@@
 
-The International Phonetic Alphabet (IPA) is a common standard in linguistics to transcribe sounds of spoken language into some Latin-based characters (International Phonetic Association 1999).
-Although IPA is reasonably easily adhered to with pen and paper, it is not trivial to encode IPA characters electronically.
-Early work addressing the need for a universal computing environment for writing systems and their computational complexity is discussed in Simons 1989.
-For a long time, linguists (like all other computer users) were limited to ASCII-encoded 7-bit characters, which only includes Latin characters, numbers and some punctuation and symbols.
-Restricted to these standard character sets that lacked IPA support or other language-specific graphemes that they needed, linguists devised their own solutions (cf.~Bird and Simons 2003).
-For example, some chose to represent unavailable graphemes with substitutes, e.g.~the combination of to represent .
-Tech-savvy linguists redefined selected characters from a character encoding by mapping custom made fonts to those code points.
-However, one linguist's electronic text would not render properly on another linguist's computer without access to the same font.
-Further, if two character encodings defined two character sets differently, then data could not be reliably and correctly displayed.
-This is a common example of the non-interoperability of data and data formats.
+The International Phonetic Alphabet (IPA) is a common standard in linguistics to
+transcribe sounds of spoken language into some Latin-based characters
+\citep{IPA2005}. Although IPA is reasonably easily adhered to with pen and
+paper, it is not trivial to encode IPA characters electronically. Early work
+addressing the need for a universal computing environment for writing systems
+and their computational complexity is discussed in Simons 1989. For a long time,
+linguists (like all other computer users) were limited to ASCII-encoded 7-bit
+characters, which only includes Latin characters, numbers and some punctuation
+and symbols. Restricted to these standard character sets that lacked IPA support
+or other language-specific graphemes that they needed, linguists devised their
+own solutions \citep{BirdSimons2003}. For example, some chose to represent
+unavailable graphemes with substitutes, e.g.~the combination of to represent .
+Tech-savvy linguists redefined selected characters from a character encoding by
+mapping custom made fonts to those code points. However, one linguist's
+electronic text would not render properly on another linguist's computer without
+access to the same font. Further, if two character encodings defined two
+character sets differently, then data could not be reliably and correctly
+displayed. This is a common example of the non-interoperability of data and data
+formats.
 
-To alleviate this problem, during the late 1980s, SAMPA (Speech Assessment Methods Phonetic Alphabet) was created to represent IPA symbols with 7-bit printable ASCII character sequences, e.g.
-for [ɸ].
-Two problems with SAMPA are that (i) it is only a partial encoding of the IPA and (ii) it encodes different languages in separate data tables, instead of a universal alphabet, like IPA.
-SAMPA tables are derived from phonemes appearing in several European languages that were developed as part of a European Commission-funded project to address technical problems like electronic mail exchange (what is now simply called email).
-SAMPA is essentially a hack to work around displaying IPA characters, but it provided speech technology and other fields a basis that has been widely adopted and used in code.
-So, SAMPA was a collection of tables to be compared, instead of a large universal table representing all languages.
-An extended version of SAMPA, called X-SAMPA, set out to include every symbol in the IPA chart including all diacritics (Wells nd.).
-X-SAMPA was considered more universally applicable because it consisted of one table that encoded the set of characters that represented phones/segments in IPA across languages.
-SAMPA and X-SAMPA have been widely used for speech technology and as an encoding system in computational linguistics.
-Eventually, ASCII-encoding of the IPA became depreciated through the advent of the Unicode Standard.
-Note however that many popular software packages used for linguistic analyses still require ASCII input, e.g.~RuG/L04\footnote{http://www.let.rug.nl/kleiweg/L04/} and SplitsTree4.\footnote{http://www.splitstree.org/}
+To alleviate this problem, during the late 1980s, SAMPA (Speech Assessment
+Methods Phonetic Alphabet) was created to represent IPA symbols with 7-bit
+printable ASCII character sequences, e.g. <p\textbackslash> for [ɸ]. Two
+problems with SAMPA are that (i) it is only a partial encoding of the IPA and
+(ii) it encodes different languages in separate data tables, instead of a
+universal alphabet, like IPA. SAMPA tables are derived from phonemes appearing
+in several European languages that were developed as part of a European
+Commission-funded project to address technical problems like electronic mail
+exchange (what is now simply called email). SAMPA is essentially a hack to work
+around displaying IPA characters, but it provided speech technology and other
+fields a basis that has been widely adopted and used in code. So, SAMPA was a
+collection of tables to be compared, instead of a large universal table
+representing all languages. An extended version of SAMPA, called X-SAMPA, set
+out to include every symbol in the IPA chart including all diacritics
+\citep{WellsND}. X-SAMPA was considered more universally applicable because it
+consisted of one table that encoded the set of characters that represented
+phones/segments in IPA across languages. SAMPA and X-SAMPA have been widely used
+for speech technology and as an encoding system in computational linguistics.
+Eventually, ASCII-encoding of the IPA became depreciated through the advent of
+the Unicode Standard. Note however that many popular software packages used for
+linguistic analyses still require ASCII input,
+e.g.~RuG/L04\footnote{\url{http://www.let.rug.nl/kleiweg/L04/}} and
+SplitsTree4.\footnote{\url{http://www.splitstree.org/}}
 
-Still, there are a few pitfalls to be aware of when using the Unicode Standard to encode IPA.
-As we have said before, from a linguistic perspective it might look like the Unicode Consortium is making incomprehensible decisions, but it is important to realize that the consortium has tried and is continuing to try to be as consistent as possible across a wide range of use cases, and it does place linguistic traditions above other orthographic possibilities.
-In general, we strongly suggest linguists not to complain about any decisions in the Unicode Standard, but to try and understand the rationale of the Unicode Consortium (which in our experience is almost always well-conceived) and devise ways to work with any unexpected behaviour.
-Many of the current problems derive from the fact that the IPA is clearly historically based on the Latin script, but different enough from most other Latin-based writing systems to warrant special attention.
-This ambivalent status of the IPA glyphs (partly Latin, partly special) is unfortunately also attested in the treatment of IPA in the Unicode Standard.
-In retrospect, it might have been better to consider the IPA (and other transcription systems) to be a special or new kind of script within the Unicode Standard, and treat the obvious similarity to Latin glyphs as a historical relic.
-All IPA glyphs would then have their own code points, instead of the current situation in which some IPA glyphs have special code points, while others are treated as being identical to the `regular' Latin characters.
-Yet, the current situation, however unfortunate, is unlikely to change, so as linguists we will have to learn to deal with the specific pitfalls of IPA within the Unicode Standard.
-In this section, we will describe these pitfalls in some detail.
+Still, there are a few pitfalls to be aware of when using the Unicode Standard
+to encode IPA. As we have said before, from a linguistic perspective it might
+look like the Unicode Consortium is making incomprehensible decisions, but it is
+important to realize that the consortium has tried and is continuing to try to
+be as consistent as possible across a wide range of use cases, and it does place
+linguistic traditions above other orthographic possibilities. In general, we
+strongly suggest linguists not to complain about any decisions in the Unicode
+Standard, but to try and understand the rationale of the Unicode Consortium
+(which in our experience is almost always well-conceived) and devise ways to
+work with any unexpected behavior. Many of the current problems derive from the
+fact that the IPA is clearly historically based on the Latin script, but
+different enough from most other Latin-based writing systems to warrant special
+attention. This ambivalent status of the IPA glyphs (partly Latin, partly
+special) is unfortunately also attested in the treatment of IPA in the Unicode
+Standard. In retrospect, it might have been better to consider the IPA (and
+other transcription systems) to be a special or new kind of script within the
+Unicode Standard, and treat the obvious similarity to Latin glyphs as a
+historical relic. All IPA glyphs would then have their own code points, instead
+of the current situation in which some IPA glyphs have special code points,
+while others are treated as being identical to the regular Latin characters.
+Yet, the current situation, however unfortunate, is unlikely to change, so as
+linguists we will have to learn to deal with the specific pitfalls of IPA within
+the Unicode Standard. In this section, we will describe these pitfalls in some
+detail.
 
@@
 
-The ambivalent nature of IPA glyphs arises because, on the one hand, the IPA uses Latin-based glyphs like , or .
-From this perspective, the IPA seems to be just another orthographic tradition using Latin characters, all of which do not get a special treatment within the Unicode Standard (just like e.g.~the French, German, or Danish orthographic traditions do not have a special status).
-On the other hand, the IPA uses many special symbols (like turned , mirrored and/or extended Latin glyphs ) not found in any other Latin-based writing system.
-For this reason, and already in the first version of the Unicode Standard (Version 1.0 from 1991), a special block with code points, called ``IPA Extensions'' was included.
-As noted in Section 4, Pitfall 1, the Unicode Standard code space is subdivided into character blocks, which generally encode characters from a single script.
-However, as is illustrated by the IPA, characters that form a single writing system may be dispersed across several different character blocks.
-With its diverse collection of symbols from various scripts and diacritics, the IPA is spread across over 13 blocks in the Unicode Standard:\footnote{This number of blocks depends on whether only IPA-sanctioned symbols are counted or if the phonetic symbols commonly found in the literature are also included, see Moran 2012, Appendix C.}
+The ambivalent nature of IPA glyphs arises because, on the one hand, the IPA
+uses Latin-based glyphs like , or . From this perspective, the IPA seems to be
+just another orthographic tradition using Latin characters, all of which do not
+get a special treatment within the Unicode Standard (just like e.g.~the French,
+German, or Danish orthographic traditions do not have a special status). On the
+other hand, the IPA uses many special symbols (like turned , mirrored and/or
+extended Latin glyphs ) not found in any other Latin-based writing system. For
+this reason, and already in the first version of the Unicode Standard (Version
+1.0 from 1991), a special block with code points, called \textsc{IPA Extensions}
+was included. As noted in Section 4, Pitfall 1, the Unicode Standard code space
+is subdivided into character blocks, which generally encode characters from a
+single script. However, as is illustrated by the IPA, characters that form a
+single writing system may be dispersed across several different character
+blocks. With its diverse collection of symbols from various scripts and
+diacritics, the IPA is spread across over 13 blocks in the Unicode
+Standard:\footnote{This number of blocks depends on whether only IPA-sanctioned
+symbols are counted or if the phonetic symbols commonly found in the literature
+are also included, see Moran 2012, Appendix C.}
+
 \begin{itemize}
-	\item Basic Latin (30 characters), e.g. <a, b, c, d, e> 
-	\item Latin-1 Supplement (4 characters): <æ, ç, ð, ø\textgreater{} 
-	\item Latin Extended-A (3 characters): <ħ, ŋ, œ> 
-	\item Latin Extended-B (5 characters): <ǀ, ǁ, ǂ, ǃ, ȵ> 
-	\item IPA Extensions (70 characters), e.g.: <ɐ, ɑ, ɔ> 
-	\item Spacing Modifier Letters (20 characters), e.g.: <ʰ ʷ ˥> 
-	\item Combining Diacritical Marks (33 characters), e.g.: <̝ ̥ ̪> 
-	\item Greek and Coptic (3 characters): <β, θ, χ> 
-	\item Phonetic Extensions (2 characters): <ᴅ, ᴴ> 
-	\item Phonetic Extensions Supplement (3 characters): <ᶑ , ᶾ, ᶣ> 
-	\item Superscripts and Subscripts (1 character): <ⁿ> 
-	\item Arrows (2 characters): \textless{}↑, ↓\textgreater{} 
-	\item Latin Extended-C (1 character): <ⱱ> 
+	\item Basic Latin (30 characters), e.g. <a, b, c, d, e> \item Latin-1
+	Supplement (4 characters): <æ, ç, ð, ø\textgreater{} \item Latin
+	Extended-A (3 characters): <ħ, ŋ, œ> \item Latin Extended-B (5
+	characters): <ǀ, ǁ, ǂ, ǃ, > \item IPA Extensions (70 characters), e.g.:
+	<ɐ, ɑ, ɔ> \item Spacing Modifier Letters (20 characters), e.g.: <ʰ ʷ ˥>
+	\item Combining Diacritical Marks (33 characters), e.g.: <̝ ̥ ̪> \item
+	Greek and Coptic (3 characters): <β, θ, χ> \item Phonetic Extensions (2
+	characters): <, > \item Phonetic Extensions Supplement (3 characters): <
+	, , > \item Superscripts and Subscripts (1 character): <ⁿ> \item Arrows
+	(2 characters): \textless{}↑, ↓\textgreater{} \item Latin Extended-C (1
+	character): <>
 \end{itemize}
@@
 
-Another problem is the large number of homoglyphs, i.e.~different characters that have highly similar glyphs (or even completely identical, depending on the font rendering).
-For example, a speaker of Russian should ideally not use the CYRILLIC SMALL LETTER A at code point U+0430 for IPA transcriptions, but instead the LATIN SMALL LETTER A at code point U+0061, although visually they are mostly indistinguishable, and the Cyrillic character is more easily typed on a Cyrillic keyboard.
-Another example we commonly encounter is the use of LATIN SMALL LETTER G at U+0067, instead of the sanctioned Unicode Standard IPA character for the voiced velar stop LATIN SMALL LETTER SCRIPT G at U+0261.
-One begins to question whether this issue is at all apparent to the working linguist, or if they simply use the U+0067 because it is easily keyboarded and thus saves time, whereas the latter must be cumbersomely inserted as a special symbol in most software.\footnote{This issue was recently addressed by the International Phonetic Association, which has taken the stance that both the keyboard LATIN SMALL LETTER G and the LATIN SMALL LETTER SCRIPT G are valid input characters for the voiced velar plosive.
-Unfortunately, this decision further introduces ambiguity for linguists trying to adhere to a strict Unicode Standard IPA encoding.}
+Another problem is the large number of homoglyphs, i.e.~different characters
+that have highly similar glyphs (or even completely identical, depending on the
+font rendering). For example, a speaker of Russian should ideally not use the
+\textsc{cyrillic small letter a} at code point U+0430 for IPA transcriptions,
+but instead the \textsc{latin small letter a} at code point U+0061, although
+visually they are mostly indistinguishable, and the Cyrillic character is more
+easily typed on a Cyrillic keyboard. Another example we commonly encounter is
+the use of \textsc{latin small letter g} at U+0067, instead of the sanctioned
+Unicode Standard IPA character for the voiced velar stop \textsc{latin small
+letter script g} at U+0261. One begins to question whether this issue is at all
+apparent to the working linguist, or if they simply use the U+0067 because it is
+easily keyboarded and thus saves time, whereas the latter must be cumbersomely
+inserted as a special symbol in most software.\footnote{This issue was recently
+addressed by the International Phonetic Association, which has taken the stance
+that both the keyboard \textsc{latin small letter g} and the \textsc{latin small
+letter script g} are valid input characters for the voiced velar plosive.
+Unfortunately, this decision further introduces ambiguity for linguists trying
+to adhere to a strict Unicode Standard IPA encoding.}
 
-Furthermore, on the one hand even linguists are unlikely to distinguish between the LATIN SMALL LETTER SCHWA at code point U+0259 and LATIN SMALL LETTER TURNED E at U+01DD.
-On the other hand, non-linguists are unlikely to distinguish any semantic difference between an open back unrounded vowel , the LATIN SMALL LETTER ALPHA at U+0251, and the open front unrounded vowel , LATIN SMALL LETTER A at U+0061.
-But even among linguists this distinction leads to problems.
-For example, as pointed out by Mielke (2009), there is a problem stemming from the fact that about 75\% of languages are reported to have a five-vowel system (Maddieson 1984).
-Historically, linguistic descriptions tend not to include precise audio recording and measurements of formants, so this may lead one to ask if the many characters that are used in phonological description reflects a ``transcriptional bias''.
-The common use of in transcriptions could be in part due to the ease of typing the letter on an English keyboard (or for older descriptions, the typewriter).
-We found it to be exceedingly rare that a linguist uses for a low back unrounded vowel.\footnote{One example is Vidal 2001a:75, in which the author states: ``The definition of Pilagá /a/ as {[}+back{]} results from its behavior in certain phonological contexts.
+Furthermore, on the one hand even linguists are unlikely to distinguish between
+the <ə> \textsc{latin small letter schwa} at code point U+0259 and <ǝ>
+\textsc{latin small letter turned e} at U+01DD. On the other hand, non-linguists
+are unlikely to distinguish any semantic difference between an open back
+unrounded vowel <ɑ>, the \textsc{latin small letter alpha} at U+0251, and the
+open front unrounded vowel <a>, \textsc{latin small letter a} at U+0061. But
+even among linguists this distinction leads to problems. For example, as pointed
+out by \citet{Mielke2009}, there is a problem stemming from the fact that about
+75\% of languages are reported to have a five-vowel system
+\citep{Maddieson1984}. Historically, linguistic descriptions tend not to include
+precise audio recording and measurements of formants, so this may lead one to
+ask if the many characters that are used in phonological description reflects a
+\textsc{transcriptional bias}. The common use of in transcriptions could be in
+part due to the ease of typing the letter on an English keyboard (or for older
+descriptions, the typewriter). We found it to be exceedingly rare that a
+linguist uses for a low back unrounded vowel.\footnote{One example is
+\citet[75]{Vidal2001a}, in which the author states: ``The definition of Pilagá
+/a/ as {[}+back{]} results from its behavior in certain phonological contexts.
 For instance, uvular and pharyngeal consonants only occur around /a/ and /o/.
-Hence, the characterization of /a/ and /o/ as a natural class of (i.e., {[}+back{]} vowels), as opposed to /i/ and /e/.''} They simply use $<$a$>$ as long as there is no opposition to $<$ɑ$>$.\footnote{See Thomason's Language Log post, ``Why I don't love the International Phonetic Alphabet'', at: http://itre.cis.upenn.edu/\textasciitilde{}myl/languagelog/archives/005287.html.} Making things even more problematic, there is an old typographic tradition that the double-story uses a single-story in italics.
-This leads to the unfortunate effect that in most well-designed fonts the italics of and use the same glyph.
-If this distinction has to be kept upright in italics, the only solution we can currently offer is to use `slanted' glyphs (i.e.~artificially italicised glyphs) instead of real italics (i.e.~special italics glyphs designed by a typographer).\footnote{For example, the widely used IPA font Doulos SIL (http://scripts.sil.org/cms/scripts/page.php?item\_id=DoulosSIL) does not have real italics.
-This leads some word-processing software, like Microsoft Word, to produce slanted glyphs instead.
-That particular combination of font and software application will thus lead to the desired effect.
-However, note that when the text is transferred to another font (i.e.~one that includes real italics) and/or to another software application (like Apple Pages, which does not perform slanting), then this visual appearance will be lost.
-In this case we are thus still in the pre-Unicode situation in which the choice of font and rendering software actually matters.
-The ideal solution from a linguistic point of view would be the introduction of a new IPA code point for a different kind of which explicitly specifies that it should still be rendered as a double-story character when italicized.
-After informal discussion with various Unicode players, our impression is that this highly restricted problem is not sufficiently urgent to introduce even more -like characters in Unicode (which already lead to much confusion, see Section 4, Pitfall 4).
-This is a clear situation in which the Unicode Consortium is not just thinking about linguists, but has a more wide-ranging practical view to consider.}
+Hence, the characterization of /a/ and /o/ as a natural class of (i.e.,
+{[}+back{]} vowels), as opposed to /i/ and /e/.''} They simply use $<$a$>$ as
+long as there is no opposition to $<$ɑ$>$.\footnote{See Thomason's Language Log
+post, ``Why I don't love the International Phonetic Alphabet'', at:
+\url{http://itre.cis.upenn.edu/\textasciitilde{}myl/languagelog/archives/005287.html}.}
+Making things even more problematic, there is an old typographic tradition that
+the double-story uses a single-story in italics. This leads to the unfortunate
+effect that in most well-designed fonts the italics of and use the same glyph.
+If this distinction has to be kept upright in italics, the only solution we can
+currently offer is to use \textit{slanted} glyphs (i.e.~artificially italicized
+glyphs) instead of real italics (i.e.~special italics glyphs designed by a
+typographer).\footnote{For example, the widely used IPA font Doulos SIL
+(\url{http://scripts.sil.org/cms/scripts/page.php?item\_id=DoulosSIL}) does not
+have real italics. This leads some word-processing software, like Microsoft
+Word, to produce slanted glyphs instead. That particular combination of font and
+software application will thus lead to the desired effect. However, note that
+when the text is transferred to another font (i.e.~one that includes real
+italics) and/or to another software application (like Apple Pages, which does
+not perform slanting), then this visual appearance will be lost. In this case we
+are thus still in the pre-Unicode situation in which the choice of font and
+rendering software actually matters. The ideal solution from a linguistic point
+of view would be the introduction of a new IPA code point for a different kind
+of which explicitly specifies that it should still be rendered as a double-story
+character when italicized. After informal discussion with various Unicode
+players, our impression is that this highly restricted problem is not
+sufficiently urgent to introduce even more <a>-like characters in Unicode (which
+already lead to much confusion, see Section \ref{pitfall-homoglyphs}). This is a
+clear situation in which the Unicode Consortium is not just thinking about
+linguists, but has a more wide-ranging practical view to consider.}
 
 Some other homoglyphs related to encoding IPA in the Unicode Standard are:
+
 \begin{itemize}
-	\item The uses of the apostrophe has led to long discussions on the Unicode Standard email list.
-An English keyboard inputs \textless{}`\textgreater{} APOSTROPHE at U+0027, although the `preferred' Unicode apostrophe is the \textless{}'\textgreater{} RIGHT SINGLE QUOTATION MARK at U+2019.
-Yet the glottal stop/glottalization/ejective marker is another completely different character, the MODIFIER LETTER APOSTROPHE at U+02BC, but unfortunately looks mostly highly similar to U+2019.
-	\item There is ambiguous encoding of IPA segments within the Unicode Standard.
-An example is the U+02C1 MODIFIER LETTER REVERSED GLOTTAL STOP vs the U+02E4 MODIFIER LETTER SMALL REVERSED GLOTTAL STOP .
-Both are denoted in the Unicode Standard as the `pharyngealized diacritic' and both appear in various resources representing phonetic data online.
-This is thus an example for which the Unicode Standard does not solve the linguistic standardization problem.
-	\item There is at least one case in which the character name assigned by the Unicode Consortium does not match the IPA's description: in the Unicode Standard at U+01C3 is labeled LATIN LETTER RETROFLEX CLICK, but in IPA is an alveolar or postalveolar click (not retroflex).
-This naming is probably best seen as a simple error in the Unicode Standard.
-Note that most linguists simply seem to use or <ou>) while ligatures are single characters (e.g.
-<ʧ> LATIN SMALL LETTER TESH DIGRAPH at U+02A7).
-Ligatures arose in the context of printing easier-to-read texts, and are included in the Unicode Standard for reasons of legacy encoding.
-However, their usage is discouraged by the Unicode core specification.
-Specifically related to IPA, various phonetic combinations of characters (typically affricates) are available as single code-points in the Unicode Standard, but are designated as ``ligatures'' or ``digraphs'' (confusingly both names appear interchangeably).
-Such glyphs might be used by software to produce a pleasing display, but they should not be hard-coded into the text itself.
-In the context of IPA, characters like the following ligatures should thus \emph{not} be used.
-Instead a combination of two characters is preferred:
-	\item <ʣ> LATIN SMALL LETTER DZ DIGRAPH at U+02A3 (use <d> + <z> instead) 
-	\item <ʧ> LATIN SMALL LETTER TESH DIGRAPH at U+02A7 (use <t> + <ʃ> instead) 
-	\item <ʩ> LATIN SMALL LETTER FENG DIGRAPH at U+02A9 (use <f> + <ŋ> instead) 
+	\item The uses of the apostrophe has led to long discussions on the
+	Unicode Standard email list. An English keyboard inputs
+	\textless{}`\textgreater{} APOSTROPHE at U+0027, although the
+	\textit{preferred} Unicode apostrophe is the \textless{}'\textgreater{}
+	\textsc{right single quotation mark} at U+2019. Yet the glottal
+	stop/glottalization/ejective marker is another completely different
+	character, the \textsc{modifier letter apostrophe} at U+02BC, but
+	unfortunately looks mostly highly similar to U+2019.  \item There is
+	ambiguous encoding of IPA segments within the Unicode Standard. An
+	example is the U+02C1 \textsc{modifier letter reversed glottal stop} vs
+	the U+02E4 \textsc{modifier letter small reversed glottal stop} . Both
+	are denoted in the Unicode Standard as the \textsc{pharyngealized
+	diacritic} and both appear in various resources representing phonetic
+	data online. This is thus an example for which the Unicode Standard does
+	not solve the linguistic standardization problem.  \item There is at
+	least one case in which the character name assigned by the Unicode
+	Consortium does not match the IPA's description: in the Unicode Standard
+	at U+01C3 is labeled \textsc{latin letter retroflex click}, but in IPA
+	is an alveolar or postalveolar click (not retroflex). This naming is
+	probably best seen as a simple error in the Unicode Standard. Note that
+	most linguists simply seem to use or <ou>) while ligatures are single
+	characters (e.g. <ʧ> \textsc{latin small letter tesh digraph} at
+	U+02A7). Ligatures arose in the context of printing easier-to-read
+	texts, and are included in the Unicode Standard for reasons of legacy
+	encoding. However, their usage is discouraged by the Unicode core
+	specification. Specifically related to IPA, various phonetic
+	combinations of characters (typically affricates) are available as
+	single code-points in the Unicode Standard, but are designated as
+	\textsc{ligatures} or \textsc{digraphs} (confusingly both names appear
+	interchangeably). Such glyphs might be used by software to produce a
+	pleasing display, but they should not be hard-coded into the text
+	itself. In the context of IPA, characters like the following ligatures
+	should thus \emph{not} be used. Instead a combination of two characters
+	is preferred: \item <ʣ> \textsc{latin small letter dz digraph} at U+02A3
+	(use <d> + <z> instead) \item <ʧ> \textsc{latin small letter tesh
+	digraph} at U+02A7 (use <t> + <ʃ> instead) \item <ʩ> \textsc{latin small
+	letter feng digraph} at U+02A9 (use <f> + <ŋ> instead)
 \end{itemize}
 
-However, there are a few Unicode characters that are historically ligatures, but which are today considered as simple characters in the Unicode Standard and thus should be used when writing IPA, namely:
+However, there are a few Unicode characters that are historically ligatures, but
+which are today considered as simple characters in the Unicode Standard and thus
+should be used when writing IPA, namely:
+
 \begin{itemize}
-	\item <ɮ> LATIN SMALL LETTER LEZH at U+026E 
-	\item <œ> LATIN SMALL LIGATURE OE at U+0153 
-	\item <ɶ> LATIN LETTER SMALL CAPITAL OE at U+0276 
-	\item <æ> LATIN SMALL LETTER AE at U+00E6 
+	\item <ɮ> \textsc{latin small letter lezh} at U+026E \item <œ>
+	\textsc{latin small ligature oe} at U+0153 \item <ɶ> \textsc{latin
+	letter small capital oe} at U+0276 \item <æ> \textsc{latin small letter
+	ae} at U+00E6
 \end{itemize}
@@
 
-Another pitfall is diacritics.
-The problem is that the meaning of the term `diacritics' as used by the IPA is not the same as it used in the Unicode Standard.
-Specifically, diacritics in the IPA-sense are either so-called \textsc{Combining Diacritical Marks} or \textsc{Spacing Modifier Letters} in the Unicode Standard.
-Crucially, Combining Diacritical Marks are by definition combined with the character before them (to form so-called default grapheme clusters, see Section 3).
-In contrast, Spacing Modifier Letters are by definition \emph{not} combined into grapheme clusters with the preceding character, but simply treated as separate letters.
-In the context of the IPA, the following IPA-diacritics are actually Spacing Modifier Letters in the Unicode Standard:
+Another pitfall is diacritics. The problem is that the meaning of the term
+\textsc{diacritics} as used by the IPA is not the same as it used in the Unicode
+Standard. Specifically, diacritics in the IPA-sense are either so-called
+\textsc{Combining Diacritical Marks} or \textsc{Spacing Modifier Letters} in the
+Unicode Standard. Crucially, Combining Diacritical Marks are by definition
+combined with the character before them (to form so-called default grapheme
+clusters, see Section 3). In contrast, Spacing Modifier Letters are by
+definition \emph{not} combined into grapheme clusters with the preceding
+character, but simply treated as separate letters. In the context of the IPA,
+the following IPA-diacritics are actually Spacing Modifier Letters in the
+Unicode Standard:
+
 \begin{itemize}
-	\item Length marks, namely <ː> MODIFIER LETTER TRIANGULAR COLON at U+02D0 and <ˑ> MODIFIER LETTER HALF TRIANGULAR COLON at U+02D1. 
-	\item Tone letters, like <˥> MODIFIER LETTER EXTRA-HIGH TONE BAR at U+02E5, and others like this. 
-	\item Superscript letters, like <ʰ> MODIFIER LETTER SMALL H at U+02B0 or <ˤ> MODIFIER LETTER SMALL REVERSED GLOTTAL STOP at U+02E4, and others like this. 
-	\item <˞> MODIFIER LETTER RHOTIC HOOK at U+02DE. 
+	\item Length marks, namely <ː> \textsc{modifier letter triangular colon}
+	at U+02D0 and <ˑ> \textsc{modifier letter half triangular colon} at
+	U+02D1.  \item Tone letters, like <˥> \textsc{modifier letter extra-high
+	tone bar} at U+02E5, and others like this.  \item Superscript letters,
+	like <ʰ> \textsc{modifier letter small h} at U+02B0 or <ˤ>
+	\textsc{modifier letter small reversed glottal stop} at U+02E4, and
+	others like this.  \item <˞> \textsc{modifier letter rhotic hook} at
+	U+02DE.
 \end{itemize}
 
-Although linguists might expect these characters to belong together with the character in front of them, at least for <ʰ> MODIFIER LETTER SMALL H at U+02B0 the Unicode Consortium's decision to treat it as a separate character is also linguistically correct, because according to the IPA it can be used both for aspiration (more precisely postaspiration following the base character) and preaspiration (preceding the base character).
-Note that there is a mechanism in Unicode to force separate characters to be combined (namely by using the ZERO WIDTH JOINER at U+200D), but this seems to be a rather impractical, and probably not enforceable solution to us.
+Although linguists might expect these characters to belong together with the
+character in front of them, at least for <ʰ> \textsc{modifier letter small h} at
+U+02B0 the Unicode Consortium's decision to treat it as a separate character is
+also linguistically correct, because according to the IPA it can be used both
+for aspiration (more precisely post-aspiration following the base character) and
+pre-aspiration (preceding the base character). Note that there is a mechanism in
+Unicode to force separate characters to be combined (namely by using the
+\textsc{zero width joiner} at U+200D), but this seems to be a rather
+impractical, and probably not enforceable solution to us.
 
@@
 
-Also related to diacritics is the question of ordering.
-To our knowledge, the International Phonetic Association does not specify a specific ordering for diacritics that combine with phonetic base symbols; this exercise is left to the reasoning of the transcriber.
-However, such marks have to be explicitly ordered if sequences of them are to be interoperable and compatible.
-An example is a labialized aspirated alveolar plosive: <tʷʰ>.
-There is nothing holding linguists back from using <tʰʷ> instead (with exactly the same intended meaning).
-However, from a technical standpoint, these two sequences are different, e.g.~if both sequences are used in a document, searching for <tʷʰ> will not find any instances of <tʰʷ>, and vice versa.
-Likewise, a creaky voiced syllabic dental nasal can be encoded in various orders, e.g.
-<n̪̰̩>, <n̩̰̪> or <n̩̪̰>.
+Also related to diacritics is the question of ordering. To our knowledge, the
+International Phonetic Association does not specify a specific ordering for
+diacritics that combine with phonetic base symbols; this exercise is left to the
+reasoning of the transcriber. However, such marks have to be explicitly ordered
+if sequences of them are to be interoperable and compatible. An example is a
+labialized aspirated alveolar plosive: <tʷʰ>. There is nothing holding linguists
+back from using <tʰʷ> instead (with exactly the same intended meaning). However,
+from a technical standpoint, these two sequences are different, e.g.~if both
+sequences are used in a document, searching for <tʷʰ> will not find any
+instances of <tʰʷ>, and vice versa. Likewise, a creaky voiced syllabic dental
+nasal can be encoded in various orders, e.g. <n̪̰̩>, <n̩̰̪> or <n̩̪̰>.
 
-In accordance with the absence of any specification of ordering in the IPA, the Unicode Standard likewise does not propose any standard orderings.
-Both leave it to the user to be consistent.
-However, there is one aspect of ordering for which the Unicode Standard does present a canonical solution, which is uncontroversial from a linguistic perspective.
-Diacritics in the Unicode Standard (i.e.~Combining Diacritical Marks, see above) are classified in Canonical Combining Classes.
-In practice, the diacritics are distinguished by their position relative to the base character.\footnote{See http://unicode.org/reports/tr44/\#Canonical\_Combining\_Class\_Values for a detailed description.} When applying a Unicode normalization (NFC or NFD, see previous section), the diacritics in different positions are put in a specified order.
-This process therefore harmonizes the difference between different encodings, for example, of a midtone creaky voice vowel <ḛ̄>.
-This grapheme cluster can be encoded either as <e> + <̄> + <̰> or as <e> + <̰> + <̄> .
-To prevent this twofold encoding, the Unicode Standard specifies the second ordering as canonical (diacritics below before diacritics above).
+In accordance with the absence of any specification of ordering in the IPA, the
+Unicode Standard likewise does not propose any standard orderings. Both leave it
+to the user to be consistent. However, there is one aspect of ordering for which
+the Unicode Standard does present a canonical solution, which is uncontroversial
+from a linguistic perspective. Diacritics in the Unicode Standard
+(i.e.~Combining Diacritical Marks, see above) are classified in Canonical
+Combining Classes. In practice, the diacritics are distinguished by their
+position relative to the base character.\footnote{See
+\url{http://unicode.org/reports/tr44/\#Canonical\_Combining\_Class\_Values for a
+detailed description}.} When applying a Unicode normalization (NFC or NFD, see
+previous section), the diacritics in different positions are put in a specified
+order. This process therefore harmonizes the difference between different
+encodings, for example, of a midtone creaky voice vowel <ḛ̄>. This grapheme
+cluster can be encoded either as <e> + <̄> + <̰> or as <e> + <̰> + <̄> . To
+prevent this twofold encoding, the Unicode Standard specifies the second
+ordering as canonical (diacritics below before diacritics above).
 
-When encoding a string according to the Unicode Standard, it is possible to do this either using the NFC (``composition'') or NFD (``decomposition'') normalization.
-Decomposition implies that precomposed characters (like <á> LATIN SMALL LETTER A WITH ACUTE at U+00E1) will be split into its parts.
-This might sound preferable for a linguistic analysis, as the different diacritics are separated from the base characters.
-However, note that most attached elements like strokes (e.g.~in the <ɨ>), retroflex hooks (e.g.~in <ʐ>) or rhotic hooks (e.g.~in <ɝ>) will not be decomposed, but strangely enough a cedilla (like in <ç>) will be decomposed.
-In general, Unicode decomposition does not behave like a feature decomposition as expected from a linguistic perspective.
-It is thus important to consider Unicode decomposition only as a technical procedure, and not assume that it is linguistically sensible.
+When encoding a string according to the Unicode Standard, it is possible to do
+this either using the NFC (composition) or NFD (decomposition) normalization.
+Decomposition implies that precomposed characters (like <á> \textsc{latin small
+letter a with acute} at U+00E1) will be split into its parts. This might sound
+preferable for a linguistic analysis, as the different diacritics are separated
+from the base characters. However, note that most attached elements like strokes
+(e.g.~in the <ɨ>), retroflex hooks (e.g.~in <ʐ>) or rhotic hooks (e.g.~in <ɝ>)
+will not be decomposed, but strangely enough a cedilla (like in <ç>) will be
+decomposed. In general, Unicode decomposition does not behave like a feature
+decomposition as expected from a linguistic perspective. It is thus important to
+consider Unicode decomposition only as a technical procedure, and not assume
+that it is linguistically sensible.
 
-Facing the problem of specifying a consistent ordering of diacritics while developing a large database of phonological inventories from the world's languages, Moran (2012: 540) defines a set of diacritic ordering conventions.\footnote{The most recent version of these conventions is online: http://phoible.github.io/conventions/} The conventions are influenced by the linguistic literature, though some ad-hoc decisions had to be taken.
-The goal was to explicitly define all character sequences so that the vast variety of phonemes found in descriptions of the world's language were normalized into consistent character sequences, e.g.~if one language description uses and another , when both are intended to be phonetically equivalent, then a decision to normalize to one form was taken.
-For example, when a character sequence contains more than one character in Spacing Modifier Letters, the order that is proposed is the following (where <\textbar{}> indicates ``or'' and <→> indicates ``precedes''):
+Facing the problem of specifying a consistent ordering of diacritics while
+developing a large database of phonological inventories from the world's
+languages, \citet[540]{Moran2012} defines a set of diacritic ordering
+conventions.\footnote{The most recent version of these conventions is online:
+\url{http://phoible.github.io/conventions/}} The conventions are influenced by
+the linguistic literature, though some ad-hoc decisions had to be taken. The
+goal was to explicitly define all character sequences so that the vast variety
+of phonemes found in descriptions of the world's language were normalized into
+consistent character sequences, e.g.~if one language description uses and
+another , when both are intended to be phonetically equivalent, then a decision
+to normalize to one form was taken. For example, when a character sequence
+contains more than one character in Spacing Modifier Letters, the order that is
+proposed is the following (where <\textbar{}> indicates \textit{or} and <→>
+indicates \textit{precedes}):
+
 \begin{itemize}
-	\itemsep1pt\parskip0pt\parsep0pt 
-	\item \textsc{Spacing Modifier Letters ordering:} ( unreleased <̚> \textbar{} lateral release <ˡ> \textbar{} nasal release <ⁿ>) → ( palatalized <ʲ>) → ( labialized <ʷ>) → ( velarized <ˠ>) → ( pharyngealized <ˤ>) → ( aspirated <ʰ> \textbar{} ejective <ʼ>) → ( long <ː> \textbar{} half-long <ˑ>)
+	\itemsep1pt\parskip0pt\parsep0pt \item \textsc{Spacing Modifier Letters
+	ordering:} ( unreleased <̚> \textbar{} lateral release <ˡ> \textbar{}
+	nasal release <ⁿ>) → ( palatalized <ʲ>) → ( labialized <ʷ>) → (
+	velarized <ˠ>) → ( pharyngealized <ˤ>) → ( aspirated <ʰ> \textbar{}
+	ejective <ʼ>) → ( long <ː> \textbar{} half-long <ˑ>)
 \end{itemize}
 
-If a character sequence contains more than one diacritic below the base character, then the place feature is applied first (dental, laminal, apical, fronted, backed, lowered, raised), then the laryngeal setting (voiced, voiceless, creaky voice, breathy voice) and finally the syllabic or non-syllabic marker (for vowels, ATR gets put on between the place and laryngeal setting).
+If a character sequence contains more than one diacritic below the base
+character, then the place feature is applied first (dental, laminal, apical,
+fronted, backed, lowered, raised), then the laryngeal setting (voiced,
+voiceless, creaky voice, breathy voice) and finally the syllabic or non-syllabic
+marker (for vowels, ATR gets put on between the place and laryngeal setting).
 So:
+
 \begin{itemize}
-	\itemsep1pt\parskip0pt\parsep0pt 
-	\item \textsc{Combining Diacritical Marks (below) ordering:} ( dental <t̪> \textbar{} laminal <t̻> \textbar{} apical <t̺>) → ( fronted <u̟> \textbar{} backed <e̠>) →( lowered <e̞> \textbar{} raised <e̝>) → ( ATR <e̘ e̙>) → ( voiced <s̬> \textbar{} voiceless <n̥> \textbar{} creaky voice <b̰> \textbar{} breathy voice <b̤>) → ( syllabic <n̩> \textbar{} non-syllabic <e̯>)
+	\itemsep1pt\parskip0pt\parsep0pt \item \textsc{Combining Diacritical
+	Marks (below) ordering:} ( dental <t̪> \textbar{} laminal <t̻>
+	\textbar{} apical <t̺>) → ( fronted <u̟> \textbar{} backed <e̠>) →(
+	lowered <e̞> \textbar{} raised <e̝>) → ( ATR <e̘ e̙>) → ( voiced <s̬>
+	\textbar{} voiceless <n̥> \textbar{} creaky voice <b̰> \textbar{}
+	breathy voice <b̤>) → ( syllabic <n̩> \textbar{} non-syllabic <e̯>)
 \end{itemize}
 
-Character sequences with diacritics above the base character were not problematic in Moran 2012 because they include only the centralized, mid-centralized and nasalized combining characters.
-Moran (2012) marks tones as singletons with Space Modifier Letters, e.g.
-\textless{}˦\textgreater{} for a phonemic high tone, instead of accent diacritics, alleviating potential conflicts.
-Building on the work of Moran (2012), if a character sequence contains more than one diacritic above the base character, we propose:
+Character sequences with diacritics above the base character were not
+problematic in \citet{Moran2012} because they include only the centralized,
+mid-centralized and nasalized combining characters. \citet{Moran2012} marks
+tones as singletons with Space Modifier Letters, e.g. \textless{}˦\textgreater{}
+for a phonemic high tone, instead of accent diacritics, alleviating potential
+conflicts. Building on the work of \citet{Moran2012}, if a character sequence
+contains more than one diacritic above the base character, we propose:
+
 \begin{itemize}
-	\itemsep1pt\parskip0pt\parsep0pt 
-	\item \textsc{Combining Diacritical Marks (above) ordering:} (centralized <ë> \textbar{} mid-centralized <e̽>) → (extra short <ĕ>) →( tone accents, e.g.
-<è> ) → ( Spacing Modifier Letters ) → ( tone letters, e.g.
-<e˦>)
+	\itemsep1pt\parskip0pt\parsep0pt \item \textsc{Combining Diacritical
+	Marks (above) ordering:} (centralized <ë> \textbar{} mid-centralized
+	<e̽>) → (extra short <ĕ>) →( tone accents, e.g. <è> ) → ( Spacing
+	Modifier Letters ) → ( tone letters, e.g. <e˦>)
 \end{itemize}

### 2015-06-05 chapters/orthography_profiles.tex
wraps line at character width 80 in all tex files

@@
-\chapter{Orthography profiles}
-\label{orthography-profiles}
+\chapter{Orthography profiles} \label{orthography-profiles}
 
-\section{Characterizing writing systems}
-\label{characterizing-writing-systems}
+\section{Characterizing writing systems} \label{characterizing-writing-systems}
 
-At this point in the course of rapid ongoing developments, we are left with a situation in which the Unicode Standard offers a highly detailed and flexible approach to deal computationally with writing systems, but it has unfortunately not influenced the linguistic practice very much.
-In many practical situations, the Unicode Standard is far too complex for the day-to-day practice in linguistics because it does not offer practical solutions for the down-to-earth problems of many linguists.
-In this section, we propose some simple practical guidelines and methods to improve on this situation.
+At this point in the course of rapid ongoing developments, we are left with a
+situation in which the Unicode Standard offers a highly detailed and flexible
+approach to deal computationally with writing systems, but it has unfortunately
+not influenced the linguistic practice very much. In many practical situations,
+the Unicode Standard is far too complex for the day-to-day practice in
+linguistics because it does not offer practical solutions for the down-to-earth
+problems of many linguists. In this section, we propose some simple practical
+guidelines and methods to improve on this situation.
 
-Our central aims for linguistics, to be approached with a Unicode-based solution, are: (i) to improve the consistency of the encoding of sources, (ii) to transparently document knowledge about the writing system (including transliteration), and (iii) to do all of that in a way that is easy and quick to manage for many different sources with many different writing systems.
-The central concept in our proposal is the \textsc{orthography profile}, a simple tab-separated CSV text file, that characterizes and documents a writing system.
-We also offer basic implementations in Python and R to assist with the production of such files, and to apply orthography profiles for consistency testing, grapheme tokenization and transliteration.
-Not only can orthography profiles be helpful in the daily practice of linguistics, they also succinctly document the orthographic details of a specific source, and, as such, might fruitfully be published alongside sources (e.g.~in digital archives).
-Also, in high-level linguistic analyzes in which the graphemic detail is of central importance (e.g.~phonotactic or comparative-historical studies), orthography profiles can transparently document the decisions that have been taken in the interpretation of the orthography in the sources used.
+Our central aims for linguistics, to be approached with a Unicode-based
+solution, are: (i) to improve the consistency of the encoding of sources, (ii)
+to transparently document knowledge about the writing system (including
+transliteration), and (iii) to do all of that in a way that is easy and quick to
+manage for many different sources with many different writing systems. The
+central concept in our proposal is the \textsc{orthography profile}, a simple
+tab-separated CSV text file, that characterizes and documents a writing system.
+We also offer basic implementations in Python and R to assist with the
+production of such files, and to apply orthography profiles for consistency
+testing, grapheme tokenization and transliteration. Not only can orthography
+profiles be helpful in the daily practice of linguistics, they also succinctly
+document the orthographic details of a specific source, and, as such, might
+fruitfully be published alongside sources (e.g.~in digital archives). Also, in
+high-level linguistic analyzes in which the graphemic detail is of central
+importance (e.g.~phonotactic or comparative-historical studies), orthography
+profiles can transparently document the decisions that have been taken in the
+interpretation of the orthography in the sources used.
 
-Given these goals, Unicode locale descriptions (see Section \ref{terminology}) might seem like the ideal orthography profiles.
-However, there are various practical obstacles preventing the use of such locale descriptions in the daily linguistic practice, namely: (i) the XML-structure is too verbose to easily and quickly produce or correct manually, (ii) locale descriptions are designed for a wide scope on information (like date formats or names of weekdays) most of which is not applicable for documenting writing systems, and (iii) most crucially, even if someone made the effort to produce a technically correct locale description for a specific source at hand, then it is nigh impossible to deploy the description.
-This is because a locale description has to be submitted to and accepted by the Unicode Common Locale Data Repository.
-The repository is (rightly so) not interested in descriptions that only apply to a limited set of sources (e.g.~only a single specific dictionary).
+Given these goals, Unicode locale descriptions (see Section \ref{terminology})
+might seem like the ideal orthography profiles. However, there are various
+practical obstacles preventing the use of such locale descriptions in the daily
+linguistic practice, namely: (i) the XML-structure is too verbose to easily and
+quickly produce or correct manually, (ii) locale descriptions are designed for a
+wide scope on information (like date formats or names of weekdays) most of which
+is not applicable for documenting writing systems, and (iii) most crucially,
+even if someone made the effort to produce a technically correct locale
+description for a specific source at hand, then it is nigh impossible to deploy
+the description. This is because a locale description has to be submitted to and
+accepted by the Unicode Common Locale Data Repository. The repository is
+(rightly so) not interested in descriptions that only apply to a limited set of
+sources (e.g.~only a single specific dictionary).
 
-The major challenge then is developing an infrastructure to identify the elements that are individual graphemes in a source, specifically for the enormous variety of sources using some kind of alphabetic writing system.
-Authors of source documents (e.g.~dictionaries, wordlists, corpora) use a variety of writing systems that range from their own idiosyncratic transcriptions to already well-established practical or longstanding orthographies.
-Although the IPA is one practical choice as a sound-based normalization for writing systems (which can act as an interlingual pivot to attain interoperability across writing systems), graphemes in each writing system must also be identified and standardized if interoperability across different sources is to be achieved.
-In most cases, this amounts to more than simply mapping a grapheme to an IPA segment because graphemes must first be identified in context (e.g.~is the sequence one sound or two sounds or both?) and strings must be tokenized, which may include taking orthographic rules into account (e.g.
-between vowels is /n/ and after a vowel but before a consonant is a nasalized vowel /ṽ/).
-In our experience, data from each source must be individually tokenized into graphemes so that its orthographic structure is identified and its contents can be extracted.
-To extract data for analysis, a source-by-source approach is required before an orthography profile can be created.
-For example, almost each available lexicon on the world's languages is idiosyncratic in its orthography and thus requires lexicon-specific approaches to identify graphemes in the writing system and to map graphemes to phonemes, if desired.
+The major challenge then is developing an infrastructure to identify the
+elements that are individual graphemes in a source, specifically for the
+enormous variety of sources using some kind of alphabetic writing system.
+Authors of source documents (e.g.~dictionaries, wordlists, corpora) use a
+variety of writing systems that range from their own idiosyncratic
+transcriptions to already well-established practical or longstanding
+orthographies. Although the IPA is one practical choice as a sound-based
+normalization for writing systems (which can act as an interlingual pivot to
+attain interoperability across writing systems), graphemes in each writing
+system must also be identified and standardized if interoperability across
+different sources is to be achieved. In most cases, this amounts to more than
+simply mapping a grapheme to an IPA segment because graphemes must first be
+identified in context (e.g.~is the sequence one sound or two sounds or both?)
+and strings must be tokenized, which may include taking orthographic rules into
+account (e.g. between vowels is /n/ and after a vowel but before a consonant is
+a nasalized vowel /ṽ/). In our experience, data from each source must be
+individually tokenized into graphemes so that its orthographic structure is
+identified and its contents can be extracted. To extract data for analysis, a
+source-by-source approach is required before an orthography profile can be
+created. For example, almost each available lexicon on the world's languages is
+idiosyncratic in its orthography and thus requires lexicon-specific approaches
+to identify graphemes in the writing system and to map graphemes to phonemes, if
+desired.
 
-Thus, our key proposal for the characterization of a writing system is to use a grapheme tokenization as an inter-orthographic pivot.
-Basically, any source document is tokenized by graphemes, and only then a mapping to IPA (or any other orthographic conversion) is performed.
-An orthography profile then is a description of the units and rules that are needed to adequately model a graphemic tokenization for a language variety as described in a particular source document.
-An orthography profile summarizes the Unicode (tailored) graphemes and orthographic rules used to write a language (the details of the structure and assumptions of such a profile will be presented in the next section).
+Thus, our key proposal for the characterization of a writing system is to use a
+grapheme tokenization as an inter-orthographic pivot. Basically, any source
+document is tokenized by graphemes, and only then a mapping to IPA (or any other
+orthographic conversion) is performed. An orthography profile then is a
+description of the units and rules that are needed to adequately model a
+graphemic tokenization for a language variety as described in a particular
+source document. An orthography profile summarizes the Unicode (tailored)
+graphemes and orthographic rules used to write a language (the details of the
+structure and assumptions of such a profile will be presented in the next
+section).
 
-As an example of graphemic tokenization, note the three different levels of technological and linguistic elements that interact in the hypothetical lexical form <tsʰṍ̰shi> :
+As an example of graphemic tokenization, note the three different levels of
+technological and linguistic elements that interact in the hypothetical lexical
+form <tsʰṍ̰shi> :
 
 \begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item code points (10 text elements): t s ʰ o ̃ ̰ ´ s h i 
-	\item grapheme clusters (7 text elements): t s ʰ ṍ̰ s h i 
-	\item tailored grapheme clusters (4 text elements): tsʰ ṍ̰ sh i 
+	\def\labelenumi{\arabic{enumi}.} \item code points (10 text elements): t
+	s ʰ o ̃ ̰ ´ s h i \item grapheme clusters (7 text elements): t s ʰ ṍ̰ s
+	h i \item tailored grapheme clusters (4 text elements): tsʰ ṍ̰ sh i
 \end{enumerate}
 
-In (1), the string <tsʰṍ̰shi> has been tokenized into ten Unicode code points (using NFD normalization), delimited here by space.
-Unicode tokenization is required because sequences of code points can differ in their visual and logical orders.
-For example, <õ̰> is ambiguous to whether it is the sequence of + <̰> + <̃> or + <̃> + <̰>.
-Although these two variants are visually homoglyphs, computationally they are different.
-Unicode normalization should be applied to this string to reorder the code points into a canonical order, allowing the data to be treated canonically equivalently for search and comparison.
-In (2), the Unicode code points have been logically normalized and visually organized into grapheme clusters, as specified by the Unicode Standard.
-The combining character sequence <õ̰> is normalized and visually grouped together.
-Note that , the MODIFIER LETTER SMALL H at U+02B0, is not grouped with .
-This is because it belongs to Spacing Modifier Letters category in the Unicode Standard.
-These characters are underspecified for the direction in which they modify a host character.
-For example, can indicate either pre- or post-aspiration (whereas the nasalization or creaky diacritic is defined in the Unicode Standard to apply to a specified base character).
-Finally, to arrive at the graphemic tokenization in (3), tailored grapheme clusters are needed (as for example specified in an orthography profile).
-For example, this orthography profile would specify that the sequence of characters , and form a single grapheme , and that and form a grapheme .
-The orthography profile could also specify orthographic rules, e.g.~when tokenization graphemes, in say English, the in the forms and should be treated as distinct sequences depending on their contexts.
+In (1), the string <tsʰṍ̰shi> has been tokenized into ten Unicode code points
+(using NFD normalization), delimited here by space. Unicode tokenization is
+required because sequences of code points can differ in their visual and logical
+orders. For example, <õ̰> is ambiguous to whether it is the sequence of + <̰> +
+<̃> or + <̃> + <̰>. Although these two variants are visually homoglyphs,
+computationally they are different. Unicode normalization should be applied to
+this string to reorder the code points into a canonical order, allowing the data
+to be treated canonically equivalently for search and comparison. In (2), the
+Unicode code points have been logically normalized and visually organized into
+grapheme clusters, as specified by the Unicode Standard. The combining character
+sequence <õ̰> is normalized and visually grouped together. Note that , the
+MODIFIER LETTER SMALL H at U+02B0, is not grouped with . This is because it
+belongs to Spacing Modifier Letters category in the Unicode Standard. These
+characters are underspecified for the direction in which they modify a host
+character. For example, can indicate either pre- or post-aspiration (whereas the
+nasalization or creaky diacritic is defined in the Unicode Standard to apply to
+a specified base character). Finally, to arrive at the graphemic tokenization in
+(3), tailored grapheme clusters are needed (as for example specified in an
+orthography profile). For example, this orthography profile would specify that
+the sequence of characters , and form a single grapheme , and that and form a
+grapheme . The orthography profile could also specify orthographic rules,
+e.g.~when tokenization graphemes, in say English, the in the forms and should be
+treated as distinct sequences depending on their contexts.
 
@@
 
-An orthography profile describes the Unicode code points, characters, graphemes and orthographic rules in a writing system.
-An orthography profile is a language-specific (and often even resource-specific) description of the units and rules that are needed to adequately model a writing system.
-An important assumption of our work is that we assume a resource is encoded in Unicode (or has been converted to Unicode).
-Any data source that the Unicode Standard is unable to capture, will also not be captured by an orthography profile.
+An orthography profile describes the Unicode code points, characters, graphemes
+and orthographic rules in a writing system. An orthography profile is a
+language-specific (and often even resource-specific) description of the units
+and rules that are needed to adequately model a writing system. An important
+assumption of our work is that we assume a resource is encoded in Unicode (or
+has been converted to Unicode). Any data source that the Unicode Standard is
+unable to capture, will also not be captured by an orthography profile.
 
-Informally, an orthography profile specifies the graphemes (or, in Unicode parlance, \textsc{tailored grapheme clusters}) that are expected to occur in any data to be analyzed or checked for consistency.
-These graphemes are first identified throughout the whole data (a step which we call \textsc{tokenization}), and possibly simply returned as such, possibly including error messages about any parts of the data that are not specified by the orthography profile.
-Once the graphemes are identified, they might also be changed into other graphemes (a step which we call \textsc{transliteration}).
-When a grapheme has different possible transliterations, then these differences should be separated by contextual specification, possibly down to listing individual exceptional cases.
+Informally, an orthography profile specifies the graphemes (or, in Unicode
+parlance, \textsc{tailored grapheme clusters}) that are expected to occur in any
+data to be analyzed or checked for consistency. These graphemes are first
+identified throughout the whole data (a step which we call
+\textsc{tokenization}), and possibly simply returned as such, possibly including
+error messages about any parts of the data that are not specified by the
+orthography profile. Once the graphemes are identified, they might also be
+changed into other graphemes (a step which we call \textsc{transliteration}).
+When a grapheme has different possible transliterations, then these differences
+should be separated by contextual specification, possibly down to listing
+individual exceptional cases.
 
-In practice, we foresee a workflow in which orthography profiles are iteratively refined, while at the same time inconsistencies and errors in the data to be tokenized are corrected.
-In some more complex use-cases there might even be a need for multiple different orthography profiles to be applied in sequence (see Section \ref{use-cases} on various exemplary use-cases).
-The result of any such workflow will normally be a cleaned dataset and an explicit description of the orthographic structure in the form of an orthography profile.
-Subsequently, the orthography profiles can be easily distributed in scholarly channels alongside the cleaned data, for example in supplementary material added to journal papers or in electronic archives.
+In practice, we foresee a workflow in which orthography profiles are iteratively
+refined, while at the same time inconsistencies and errors in the data to be
+tokenized are corrected. In some more complex use-cases there might even be a
+need for multiple different orthography profiles to be applied in sequence (see
+Section \ref{use-cases} on various exemplary use-cases). The result of any such
+workflow will normally be a cleaned dataset and an explicit description of the
+orthographic structure in the form of an orthography profile. Subsequently, the
+orthography profiles can be easily distributed in scholarly channels alongside
+the cleaned data, for example in supplementary material added to journal papers
+or in electronic archives.
 
@@
 
-The formal specifications of an orthography profile (or simply \textsc{profile} for short) are the following:
+The formal specifications of an orthography profile (or simply \textsc{profile}
+for short) are the following:
 
 \begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item \textsc{A profile is a} \textsc{Unicode UTF-8 encoded text file} (ideally using NFC, no-BOM, and LF; see Section \ref{pitfall-file-formats}, Pitfall: File Formats) that includes the information pertinent to the orthography.
-	\item \textsc{A profile is a} \textsc{tab-separated CSV file with an obligatory header line}.
-A minimal profile can have just a single column, in which case there will of course no tabs, but the first line will still be the header.
-For all columns we assume the name in the header of the CSV file to be crucial.
-The actual ordering of the columns is unimportant.
-	\item \textsc{Lines starting with a hash \textless{}\#\textgreater{} are ignored.} Comments and metadata can be included inside the file, but only as complete lines in the profile, to be marked by lines starting with hash \textsc{\#} (\textsc{number sign} at U+0023).
-Hashes somewhere else in the file are to be treated literally, i.e.~hashes are only to be ignored when they occur at the start of a line.\footnote{Comments that belong to specific lines will have to be put in a separate column of the CSV file, e.g.~add a column called \textsc{comments}.
-Further, if the content of a profile contains a hash at the start of a line, either reorder the columns so the hash does not occur at the start of the line, or add a dummy column in front of the data to not have the data start with a hash.}
-	\item \textsc{Metadata are given in commented lines at the beginning of the text file in a basic \textsc{tag: value} format.
-}Metadata about the orthographic description given in the orthography profile includes, minimally, (i) author, (ii) date, (iii) title, (iv) a stable language identifier encoded in BCP 47/ISO 639-3, and (v) bibliographic data for resource(s) that illustrate the orthography described in the profile.
+	\def\labelenumi{\arabic{enumi}.} \item \textsc{A profile is a}
+	\textsc{Unicode UTF-8 encoded text file} (ideally using NFC, no-BOM, and
+	LF; see Section \ref{pitfall-file-formats}, Pitfall: File Formats) that
+	includes the information pertinent to the orthography.  \item \textsc{A
+	profile is a} \textsc{tab-separated CSV file with an obligatory header
+	line}. A minimal profile can have just a single column, in which case
+	there will of course no tabs, but the first line will still be the
+	header. For all columns we assume the name in the header of the CSV file
+	to be crucial. The actual ordering of the columns is unimportant.  \item
+	\textsc{Lines starting with a hash \textless{}\#\textgreater{} are
+	ignored.} Comments and metadata can be included inside the file, but
+	only as complete lines in the profile, to be marked by lines starting
+	with hash \textsc{\#} (\textsc{number sign} at U+0023). Hashes somewhere
+	else in the file are to be treated literally, i.e.~hashes are only to be
+	ignored when they occur at the start of a line.\footnote{Comments that
+	belong to specific lines will have to be put in a separate column of the
+	CSV file, e.g.~add a column called \textsc{comments}. Further, if the
+	content of a profile contains a hash at the start of a line, either
+	reorder the columns so the hash does not occur at the start of the line,
+	or add a dummy column in front of the data to not have the data start
+	with a hash.} \item \textsc{Metadata are given in commented lines at the
+	beginning of the text file in a basic \textsc{tag: value} format.
+	}Metadata about the orthographic description given in the orthography
+	profile includes, minimally, (i) author, (ii) date, (iii) title, (iv) a
+	stable language identifier encoded in BCP 47/ISO 639-3, and (v)
+	bibliographic data for resource(s) that illustrate the orthography
+	described in the profile.
 \end{enumerate}
 
-The content of a profile consists of lines, each describing a grapheme of the orthography, using the following columns:
+The content of a profile consists of lines, each describing a grapheme of the
+orthography, using the following columns:
 
 \begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item \textsc{A minimal profile consists of a single column with header \textsc{graphemes}}, listing each of the different graphemes in a separate line. 
-	\item \textsc{Optional columns called `left' and `right' can be used to specify the left and right context of the grapheme, respectively.} The same grapheme can occur multiple times with different contextual specifications, for example to distinguish different pronunciations depending on the context.
-	\item \textsc{The columns `grapheme', `left' and `right' can use regular expression metacharacters.} If regular expressions are used, then all literal usage of the special symbols, like full stops <.> or dollar signs <\$> (so-called \textsc{metacharacters}) have to be explicitly escaped by adding a backslash before them (i.e.~use <.> or <\$>).
-Note that any specification of context automatically expects regular expressions, so it is probably better to always escape all regular expression metacharacters when used literally in the orthography, i.e.~the following symbols will need to be preceded by a backslash: {[} {]} ( ) \{ \} ~+ * .
-- !
-?
-\^{} \$ .
-	\item \textsc{An optional column called `class' can be used to specify classes of graphemes}, for example to define a class of vowels.
-Users can simply add ad-hoc identifiers in this column to indicate a group of graphemes, which can then be used in the description of the graphemes or the context.
-The identifiers should of course be chosen such that they do not conflate with any symbols used in the orthography themselves.
-Note that such classes only refer to the graphemes, not to the context.
-	\item \textsc{Columns describing transliterations for each graphemes can be added and named at will}.
-Often more than a single possible transliteration will be of interest.
-Any software application using these profiles should use the names of these columns to select a specific transliteration column.
-	\item \textsc{Any other columns can be added freely, but will mostly be ignored by any software application using the profiles}.
-As orthography profiles are also intended to be read and interpreted by humans, it is often highly useful to add extra information on the graphemes in further columns, like for example Unicode codepoints, Unicode names, frequency of occurrence, examples of occurrence, explanation of the contextual restrictions, or comments.
+	\def\labelenumi{\arabic{enumi}.} \item \textsc{A minimal profile
+	consists of a single column with header \textsc{graphemes}}, listing
+	each of the different graphemes in a separate line.  \item
+	\textsc{Optional columns called \textsc{left} and \textsc{right} can be
+	used to specify the left and right context of the grapheme,
+	respectively.} The same grapheme can occur multiple times with different
+	contextual specifications, for example to distinguish different
+	pronunciations depending on the context.  \item \textsc{The columns
+	\textsc{grapheme}, \textsc{left} and \textsc{right} can use regular
+	expression metacharacters.} If regular expressions are used, then all
+	literal usage of the special symbols, like full stops <.> or dollar
+	signs <\$> (so-called \textsc{metacharacters}) have to be explicitly
+	escaped by adding a backslash before them (i.e.~use <.> or <\$>). Note
+	that any specification of context automatically expects regular
+	expressions, so it is probably better to always escape all regular
+	expression metacharacters when used literally in the orthography,
+	i.e.~the following symbols will need to be preceded by a backslash: {[}
+	{]} ( ) \{ \} ~+ * . - ! ? \^{} \$ .  \item \textsc{An optional column
+	called \textsc{class} can be used to specify classes of graphemes}, for
+	example to define a class of vowels. Users can simply add ad-hoc
+	identifiers in this column to indicate a group of graphemes, which can
+	then be used in the description of the graphemes or the context. The
+	identifiers should of course be chosen such that they do not conflate
+	with any symbols used in the orthography themselves. Note that such
+	classes only refer to the graphemes, not to the context.  \item
+	\textsc{Columns describing transliterations for each graphemes can be
+	added and named at will}. Often more than a single possible
+	transliteration will be of interest. Any software application using
+	these profiles should use the names of these columns to select a
+	specific transliteration column.  \item \textsc{Any other columns can be
+	added freely, but will mostly be ignored by any software application
+	using the profiles}. As orthography profiles are also intended to be
+	read and interpreted by humans, it is often highly useful to add extra
+	information on the graphemes in further columns, like for example
+	Unicode codepoints, Unicode names, frequency of occurrence, examples of
+	occurrence, explanation of the contextual restrictions, or comments.
 \end{enumerate}
 
-For the automatic processing of the profiles, the following technical standards will be expected:
+For the automatic processing of the profiles, the following technical standards
+will be expected:
 
 \begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item \textsc{Each line of a profile will be interpreted as a regular expression.
-}Software applications using profiles can also offer to interpret a profile in the literal sense to avoid the necessity for the user to escape regular expressions metacharacters in the profile.
-However, this only is possible when no contexts or classes are described, so this seems only useful in the most basic orthographies.
-	\item \textsc{The `class' column will be used to produce explicit \textsc{or} chains of regular expressions}, which will then be inserted in the `graphemes', `left' and `right' columns at the position indicated by the class-identifiers.
-For example, a class \textsc{V} as a context specification might be replaced by a regular expression like: (a\textbar{}e\textbar{}i\textbar{}o\textbar{}u\textbar{}ei\textbar{}au).
-Only the graphemes themselves are included here, not any contexts specified for the elements of the class.
-	\item \textsc{The `left' and `right' contexts will be included into the regular expressions by using lookbehind and lookahead}.
-Basically, the actual regular expression syntax of lookbehind and lookahead is simply hidden to the users by allowing them to only specify the contexts themselves.
-Internally, the contexts in the columns \textsc{left} and \textsc{right} are combined with the column \textsc{graphemes} to form a complex regular expression like: (?\textless{}=left)graphemes(?=right).
-	\item \textsc{The regular expressions will be applied in the order as specified in the profile, from top to bottom.} A software implementation can offer help in figuring out the optimal ordering of the regular expressions, but should then explicitly report on the order used.
+	\def\labelenumi{\arabic{enumi}.} \item \textsc{Each line of a profile
+	will be interpreted as a regular expression. }Software applications
+	using profiles can also offer to interpret a profile in the literal
+	sense to avoid the necessity for the user to escape regular expressions
+	metacharacters in the profile. However, this only is possible when no
+	contexts or classes are described, so this seems only useful in the most
+	basic orthographies.  \item \textsc{The \textsc{class} column will be
+	used to produce explicit \textsc{or} chains of regular expressions},
+	which will then be inserted in the \textsc{graphemes}, \textsc{left} and
+	\textsc{right} columns at the position indicated by the
+	class-identifiers. For example, a class \textsc{V} as a context
+	specification might be replaced by a regular expression like:
+	(a\textbar{}e\textbar{}i\textbar{}o\textbar{}u\textbar{}ei\textbar{}au).
+	Only the graphemes themselves are included here, not any contexts
+	specified for the elements of the class.  \item \textsc{The
+	\textsc{left} and \textsc{right} contexts will be included into the
+	regular expressions by using lookbehind and lookahead}. Basically, the
+	actual regular expression syntax of lookbehind and lookahead is simply
+	hidden to the users by allowing them to only specify the contexts
+	themselves. Internally, the contexts in the columns \textsc{left} and
+	\textsc{right} are combined with the column \textsc{graphemes} to form a
+	complex regular expression like: (?\textless{}=left)graphemes(?=right).
+	\item \textsc{The regular expressions will be applied in the order as
+	specified in the profile, from top to bottom.} A software implementation
+	can offer help in figuring out the optimal ordering of the regular
+	expressions, but should then explicitly report on the order used.
 \end{enumerate}
 
-The actual implementation of the profile on some text-string will function as follows:
+The actual implementation of the profile on some text-string will function as
+follows:
 
 \begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item \textsc{All graphemes are matched in the text before they are tokenized or transliterated}.
-In this way, there is no necessity for the user to consider `feeding' and `bleeding' situations, in which the application of a rule either changes the text so another rule suddenly applies (feeding) or prevents another rule to apply (`bleeding').
-	\item \textsc{The matching of the graphemes can occur either globally or linearly.
-}From a computer science perspective, the most natural way to match graphemes from a profile in some text is by walking linearly through the text-string from left to right, and at each position go through all graphemes in the profile to see which one matches, then go to the position at the end of the matched grapheme and start over.
-This is basically how a finite state transducer works, which is a well-established technique in computer science.
-However, from a linguistic point of view, our experience is that most linguists find it more natural to think from a global perspective.
-In this approach, the first grapheme in the profile is matched everywhere in the text-string first, before moving to the next grapheme in the profile.
-Theoretically, these approaches will lead to different results, though in practice of actual natural language orthographies they almost always lead to the same result.
-Still, we suggest that any software application using orthography profiles should offer both approaches (i.e.
-\textsc{global} or \textsc{linear}) to the user.
-The approach used should be documented in the metadata as \textsc{tokenization method}.
-	\item \textsc{The matching of the graphemes can occur either in NFC or NFD.
-}By default, both the profile and the text-string to be tokenized should be treated as NFC (see section \ref{pitfall-canonical-equivalence}, Pitfall: Canonical equivalence, above).
-However, in some use-cases it turns out to be practical to treat both text and profile as NFD.
-This typically happens when very many different combinations of diacritics occur in the data.
-An NFD-profile can then be used to first check which individual diacritics are used, before turning to the more cumbersome inspection of all combinations.
-We suggest that any software application using orthography profiles should offer both approaches (i.e.
-\textsc{NFC} or \textsc{NFD}) to the user.
-The approach used should be documented in the metadata as \textsc{unicode normalization}.
-	\item \textsc{The text-string is always returned in tokenized form} by separating the matched graphemes by a user-specified symbols-string.
-Any transliteration will be returned on top of the tokenization.
-	\item \textsc{Leftover characters (i.e.~characters that are not matched by the profile) should be reported to the user as errors.} Typically, the unmatched character are replaced in the tokenization by a user-specified symbol-string.
+	\def\labelenumi{\arabic{enumi}.} \item \textsc{All graphemes are matched
+	in the text before they are tokenized or transliterated}. In this way,
+	there is no necessity for the user to consider `feeding' and `bleeding'
+	situations, in which the application of a rule either changes the text
+	so another rule suddenly applies (feeding) or prevents another rule to
+	apply (`bleeding').  \item \textsc{The matching of the graphemes can
+	occur either globally or linearly. }From a computer science perspective,
+	the most natural way to match graphemes from a profile in some text is
+	by walking linearly through the text-string from left to right, and at
+	each position go through all graphemes in the profile to see which one
+	matches, then go to the position at the end of the matched grapheme and
+	start over. This is basically how a finite state transducer works, which
+	is a well-established technique in computer science. However, from a
+	linguistic point of view, our experience is that most linguists find it
+	more natural to think from a global perspective. In this approach, the
+	first grapheme in the profile is matched everywhere in the text-string
+	first, before moving to the next grapheme in the profile. Theoretically,
+	these approaches will lead to different results, though in practice of
+	actual natural language orthographies they almost always lead to the
+	same result. Still, we suggest that any software application using
+	orthography profiles should offer both approaches (i.e. \textsc{global}
+	or \textsc{linear}) to the user. The approach used should be documented
+	in the metadata as \textsc{tokenization method}.  \item \textsc{The
+	matching of the graphemes can occur either in NFC or NFD. }By default,
+	both the profile and the text-string to be tokenized should be treated
+	as NFC (see section \ref{pitfall-canonical-equivalence}, Pitfall:
+	Canonical equivalence, above). However, in some use-cases it turns out
+	to be practical to treat both text and profile as NFD. This typically
+	happens when very many different combinations of diacritics occur in the
+	data. An NFD-profile can then be used to first check which individual
+	diacritics are used, before turning to the more cumbersome inspection of
+	all combinations. We suggest that any software application using
+	orthography profiles should offer both approaches (i.e. \textsc{NFC} or
+	\textsc{NFD}) to the user. The approach used should be documented in the
+	metadata as \textsc{unicode normalization}.  \item \textsc{The
+	text-string is always returned in tokenized form} by separating the
+	matched graphemes by a user-specified symbols-string. Any
+	transliteration will be returned on top of the tokenization.  \item
+	\textsc{Leftover characters (i.e.~characters that are not matched by the
+	profile) should be reported to the user as errors.} Typically, the
+	unmatched character are replaced in the tokenization by a user-specified
+	symbol-string.
 \end{enumerate}
@@
 \begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item \textsc{should offer user-options} to specify:
+	\def\labelenumi{\arabic{enumi}.} \item \textsc{should offer
+	user-options} to specify: \begin{enumerate}
+		\def\labelenumii{\arabic{enumii}.} \item \textsc{the name of the
+		column to be used for transliteration} (if any).  \item
+		\textsc{the symbol-string to be inserted between graphemes.}
+		Optionally, a warning might be given if the chosen string
+		includes characters from the orthography itself.  \item
+		\textsc{the symbol-string to be inserted for unmatched strings}
+		in the tokenized and transliterated output.  \item \textsc{the
+		tokenization method}, i.e.~whether the tokenization should
+		proceed \textsc{global} or \textsc{linear}.  \item
+		\textsc{unicode normalization}, i.e.~whether the text-string and
+		profile should use \textsc{NFC} or \textsc{NFD}.
+	\end{enumerate} \item \textsc{might offer user-options }to:
 	\begin{enumerate}
-		\def\labelenumii{\arabic{enumii}.} 
-		\item \textsc{the name of the column to be used for transliteration} (if any). 
-		\item \textsc{the symbol-string to be inserted between graphemes.} Optionally, a warning might be given if the chosen string includes characters from the orthography itself. 
-		\item \textsc{the symbol-string to be inserted for unmatched strings} in the tokenized and transliterated output. 
-		\item \textsc{the tokenization method}, i.e.~whether the tokenization should proceed \textsc{global} or \textsc{linear}. 
-		\item \textsc{unicode normalization}, i.e.~whether the text-string and profile should use \textsc{NFC} or \textsc{NFD}. 
-	\end{enumerate}
-	\item \textsc{might offer user-options }to:
-	\begin{enumerate}
-		\def\labelenumii{\arabic{enumii}.} \setcounter{enumii}{5} 
-		\item \textsc{assist in the ordering of the graphemes.} In our experience, it makes sense to apply larger graphemes before shorter graphemes, and to apply graphemes with context before graphemes without context.
-Further, frequently relevant rules might be applied after rarely relevant rules (though frequency is difficult to establish in practice, as it depends on the available data).
-Also, if this all fails to give any decisive ordering between rules, it seems useful to offer linguists the option to reverse the ordering from any manual specified ordering, because linguists tend to write the more general rule first, before turning to exceptions or special cases.
-		\item \textsc{assist in dealing with upper and lower case characters.} It seems practical to offer some basic case matching, so characters like <a> and <A> are treated equally.
-This will be useful in many concrete cases, although the user should be warned that case matching does not function universally in the same way across orthographies.
-Ideally, users should prepare orthography profiles with all lowercase and uppercase variants explicitly mentioned, so by default no case matching should be performed.
-		\item \textsc{treat the profile literal}, i.e.~to not interpret regular expression metacharacters.
-Matching graphemes literally often leads to strong speed increase, and would allow users to not needing to worry about escaping metacharacters.
-However, in our experience all actually interesting use-cases of orthography profiles include some contexts, which automatically prevents any literal interpretation, so by default the matching should not be literal.
-	\end{enumerate}
-	\item \textsc{should return the following information} to the user:
-	\begin{enumerate}
-		\def\labelenumii{\arabic{enumii}.} \setcounter{enumii}{8} 
-		\item \textsc{the original text-strings to be processed in the used Unicode normalization}, i.e.~in either NFC or NFD as specified by the user. 
-		\item \textsc{the tokenized strings}, with additionally any transliterated strings, if transliteration is requested. 
-		\item \textsc{a survey of all errors encountered}, ideally both in which text-strings any errors occurred and which characters in the text-strings lead to errors. 
-		\item \textsc{a reordered profile}, when any automatic reordering is offered 
+		\def\labelenumii{\arabic{enumii}.} \setcounter{enumii}{5} \item
+		\textsc{assist in the ordering of the graphemes.} In our
+		experience, it makes sense to apply larger graphemes before
+		shorter graphemes, and to apply graphemes with context before
+		graphemes without context. Further, frequently relevant rules
+		might be applied after rarely relevant rules (though frequency
+		is difficult to establish in practice, as it depends on the
+		available data). Also, if this all fails to give any decisive
+		ordering between rules, it seems useful to offer linguists the
+		option to reverse the ordering from any manual specified
+		ordering, because linguists tend to write the more general rule
+		first, before turning to exceptions or special cases.  \item
+		\textsc{assist in dealing with upper and lower case characters.}
+		It seems practical to offer some basic case matching, so
+		characters like <a> and <A> are treated equally. This will be
+		useful in many concrete cases, although the user should be
+		warned that case matching does not function universally in the
+		same way across orthographies. Ideally, users should prepare
+		orthography profiles with all lowercase and uppercase variants
+		explicitly mentioned, so by default no case matching should be
+		performed.  \item \textsc{treat the profile literal}, i.e.~to
+		not interpret regular expression metacharacters. Matching
+		graphemes literally often leads to strong speed increase, and
+		would allow users to not needing to worry about escaping
+		metacharacters. However, in our experience all actually
+		interesting use-cases of orthography profiles include some
+		contexts, which automatically prevents any literal
+		interpretation, so by default the matching should not be
+		literal.
+	\end{enumerate} \item \textsc{should return the following information}
+	to the user: \begin{enumerate}
+		\def\labelenumii{\arabic{enumii}.} \setcounter{enumii}{8} \item
+		\textsc{the original text-strings to be processed in the used
+		Unicode normalization}, i.e.~in either NFC or NFD as specified
+		by the user.  \item \textsc{the tokenized strings}, with
+		additionally any transliterated strings, if transliteration is
+		requested.  \item \textsc{a survey of all errors encountered},
+		ideally both in which text-strings any errors occurred and which
+		characters in the text-strings lead to errors.  \item \textsc{a
+		reordered profile}, when any automatic reordering is offered
 	\end{enumerate}

### 2015-06-05 chapters/terminology.tex
wraps line at character width 80 in all tex files

@@
+\chapter{Terminology} \label{terminology}
+
 \section{The linguistic terminological tradition}
@@
 
-In this section we provide a very brief overview of the linguistic terminology concerning writing systems before turning to the slightly different computational terminology in the next section on the Unicode Standard.
+In this section we provide a very brief overview of the linguistic terminology
+concerning writing systems before turning to the slightly different
+computational terminology in the next section on the Unicode Standard.
 
-Linguistically speaking, a \textsc{writing system} is a symbolic system that uses visible or tactile signs to represent language in a systematic way.
-The term `writing system' has two mutually exclusive meanings.
-First, it may refer to the way a particular language is written.
-In this sense the term refers to the writing system of a particular language, as, for example, in \emph{`the Serbian writing system uses two scripts: Latin and Cyrillic'}.
-Second, the term `writing system' may also refer to a type of symbolic system as used among the world's languages to represent the language, as, for example, in \emph{`alphabetic writing system'}.
-In this latter sense the term refers to how scripts have been classified according to the way that they encode language, as in, for example, \emph{`the Latin and Cyrillic scripts are both alphabetic writing systems'}.
-To avoid confusion, this second notion of `writing system' would more aptly have been called `script system'.
+Linguistically speaking, a \textsc{writing system} is a symbolic system that
+uses visible or tactile signs to represent language in a systematic way. The
+term \textsc{writing system} has two mutually exclusive meanings. First, it may
+refer to the way a particular language is written. In this sense the term refers
+to the writing system of a particular language, as, for example, in \emph{``the
+Serbian writing system uses two scripts: Latin and Cyrillic''}. Second, the term
+\textsc{writing system} may also refer to a type of symbolic system as used
+among the world's languages to represent the language, as, for example, in
+\emph{``alphabetic writing system''}. In this latter sense the term refers to
+how scripts have been classified according to the way that they encode language,
+as in, for example, \emph{``the Latin and Cyrillic scripts are both alphabetic
+writing systems''}. To avoid confusion, this second notion of \textsc{writing
+system} would more aptly have been called \textsc{script system}.
 
-The term \textsc{script} refers to a collection of distinct symbols as employed by one or more orthographies.
-For example, both Serbian and Russian are written with subsets of the Cyrillic script.
-A single language, like Serbian or Japanese, can also be written using orthographies based on different scripts.
-Over the years linguists have typologized script systems in a variety of ways, with the tripartite classification of logographic, syllabic, and alphabetic remaining the most popular, even though there are at least half a dozen different types of script systems that can be distinguished (Daniels, 1990, 1996).
+The term \textsc{script} refers to a collection of distinct symbols as employed
+by one or more orthographies. For example, both Serbian and Russian are written
+with subsets of the Cyrillic script. A single language, like Serbian or
+Japanese, can also be written using orthographies based on different scripts.
+Over the years linguists have typologized script systems in a variety of ways,
+with the tripartite classification of logographic, syllabic, and alphabetic
+remaining the most popular, even though there are at least half a dozen
+different types of script systems that can be distinguished
+\citep{Daniels1990,Daniels1996}.
 
-Only focussing here on the first sense of writing systems, we distinguish two different kinds of writing systems used to write a particular language, namely transcriptions and orthographies.
-First, \textsc{transcription} is a scientific procedure (and also the result of that procedure) for graphically representing the sounds of human speech at the phonetic level.
-It incorporates a set of unambiguous symbols to represent speech sounds, including conventions that specify how these symbols should be combined.
-A transcription system is a specific system of symbols and rules used for transcription of the sounds of a spoken language variety.
-In principle, a transcription system should be language-independent, in that it should be applicable to all spoken human languages.
-The International Phonetic Alphabet (IPA) is a commonly used transcription system that provides a medium for transcribing languages at the phonetic level.
-However, there is a long history of alternative kinds of transcription systems (see Kemp 1994) and today various alternatives are in widespread use (e.g.~X-SAMPA and Cyrillic-based phonetic transcription systems).\footnote{Most users of IPA do not follow the standard explicitly and many dialects based on it have emerged, e.g.~the Africanist and Americanist transcription systems.} Note that IPA symbols are also often used to represent language on a phonemic level.
-However, it is important to realize that in this usage the IPA symbols are not a transcription system, but rather an orthography (though with strong links to the pronunciation).
-Further, a transcription system does not need to be highly detailed, but it can also be a system of broad sound classes.
-Although such an approximative transcription is not normally used in linguistics, it is widespread in technological approaches (e.g.~Soundex, Beider \& Morse 2010)\footnote{http://stevemorse.org/phonetics/bmpm.htm , http://stevemorse.org/phonetics/bmpm2.htm} and it is sometimes fruitfully used in automatic approaches to historical linguistics (Dolgopolsky 1986, ASJP).\footnote{http://asjp.clld.org/}
+Only focussing here on the first sense of writing systems, we distinguish two
+different kinds of writing systems used to write a particular language, namely
+transcriptions and orthographies. First, \textsc{transcription} is a scientific
+procedure (and also the result of that procedure) for graphically representing
+the sounds of human speech at the phonetic level. It incorporates a set of
+unambiguous symbols to represent speech sounds, including conventions that
+specify how these symbols should be combined. A transcription system is a
+specific system of symbols and rules used for transcription of the sounds of a
+spoken language variety. In principle, a transcription system should be
+language-independent, in that it should be applicable to all spoken human
+languages. The International Phonetic Alphabet (IPA) is a commonly used
+transcription system that provides a medium for transcribing languages at the
+phonetic level. However, there is a long history of alternative kinds of
+transcription systems (see \citet{Kemp1994}) and today various alternatives are
+in widespread use (e.g.~X-SAMPA and Cyrillic-based phonetic transcription
+systems).\footnote{Most users of IPA do not follow the standard explicitly and
+many dialects based on it have emerged, e.g.~the Africanist and Americanist
+transcription systems.} Note that IPA symbols are also often used to represent
+language on a phonemic level. However, it is important to realize that in this
+usage the IPA symbols are not a transcription system, but rather an orthography
+(though with strong links to the pronunciation). Further, a transcription system
+does not need to be highly detailed, but it can also be a system of broad sound
+classes. Although such an approximative transcription is not normally used in
+linguistics, it is widespread in technological approaches
+(e.g.~\citet{SoundexBeiderMorse2010}\footnote{\url{http://stevemorse.org/phonetics/bmpm.htm},
+\url{http://stevemorse.org/phonetics/bmpm2.htm}} and it is sometimes fruitfully
+used in automatic approaches to historical linguistics
+\citep{Dolgopolsky1986,ASJP16}.\footnote{\url{http://asjp.clld.org/}}
 
-Second, an \textsc{orthography} specifies the symbols, punctuations, and the rules in which a specific language is written in a standardized way.
-Orthographies are often based on a phonemic analysis, but they almost always include idiosyncrasies because of historical developments (like sound changes or loans) and because of the widely-followed principle of lexical integrity (i.e.~the attempt to write the same lexical root in a consistent way, also when synchronic phonemic rules change the pronunciation, for example with final devoicing in many Germanic languages).
-All orthographies are language-specific (and often even resource-specific), although individual symbols or rules might be shared between languages.
+Second, an \textsc{orthography} specifies the symbols, punctuations, and the
+rules in which a specific language is written in a standardized way.
+Orthographies are often based on a phonemic analysis, but they almost always
+include idiosyncrasies because of historical developments (like sound changes or
+loans) and because of the widely-followed principle of lexical integrity
+(i.e.~the attempt to write the same lexical root in a consistent way, also when
+synchronic phonemic rules change the pronunciation, for example with final
+devoicing in many Germanic languages). All orthographies are language-specific
+(and often even resource-specific), although individual symbols or rules might
+be shared between languages.
 
-A \textsc{practical orthography} is a strongly phoneme-based writing system designed for practical use by speakers.
-The mapping relation between phonemes and graphemes in practical orthographies is purposely shallow, i.e.~there is mostly a systematic and faithful mapping from a phoneme to a grapheme.
-Practical orthographies are intended to jumpstart written materials development by correlating a writing system with the sound units of a language (cf.~Meinhof \& Jones 1928).
-Symbols from the IPA are often used by linguists in the development of practical orthographies for languages without writing systems, though this usage of IPA symbols should not be confused with transcription (as defined above).
+A \textsc{practical orthography} is a strongly phoneme-based writing system
+designed for practical use by speakers. The mapping relation between phonemes
+and graphemes in practical orthographies is purposely shallow, i.e.~there is
+mostly a systematic and faithful mapping from a phoneme to a grapheme. Practical
+orthographies are intended to jumpstart written materials development by
+correlating a writing system with the sound units of a language
+(cf.~\citet{MeinhofJones1928}). Symbols from the IPA are often used by linguists
+in the development of practical orthographies for languages without writing
+systems, though this usage of IPA symbols should not be confused with
+transcription (as defined above).
 
-Further, a \textsc{transliteration} is a mapping between two different orthographies.
-It is the process of ``recording the graphic symbols of one writing system in terms of the corresponding graphic symbols of a second writing system'' (Kemp 2006:396).
-In straightforward cases, such a transliteration is simply a matter of replacing one symbol with another.
-However, there are widespread complications, like one-to-many or many-to-many mappings, which are not always easy, or even possible, to solve without listing all cases individually (cf.~Moran 2012, chp 2).
+Further, a \textsc{transliteration} is a mapping between two different
+orthographies. It is the process of ``recording the graphic symbols of one
+writing system in terms of the corresponding graphic symbols of a second writing
+system'' \citep[396]{Kemp2006}. In straightforward cases, such a transliteration
+is simply a matter of replacing one symbol with another. However, there are
+widespread complications, like one-to-many or many-to-many mappings, which are
+not always easy, or even possible, to solve without listing all cases
+individually (cf.~\citet[chp 2]{Moran2012}).
 
-Breaking it down further, a script consists of \textsc{graphemes}, and graphemes consist of \textsc{characters}.
-In the linguistic terminology of writing systems, a \textsc{character} is a general term for any self-contained element in a writing system.\footnote{There is a second interpretation of the term `character', i.e.~a conventional term for a unit in the Chinese writing system (Daniels 1996).
-This interpretation will not be further explored in this paper.} Although in literate societies most people have a strong intuition about what the characters are in their particular orthography or orthographies, it turns out that the separation of an orthography into separate characters is far from trivial.
-The widespread intuitive notion of a character is strongly biased towards educational traditions, like the alphabet taught at schools, and technological possibilities, like the available type pieces in a printer's job case, the keys on a typewriter, or the symbols displayed in Microsoft Word's symbol browser.
-In practice, characters often consist of multiple building blocks, each of which could be considered a character in its own right.
-For example, although a Chinese character may be considered to be a single basic unanalyzable unit, at a more fine-grained level of analysis the internal structure of Chinese characters is often comprised of smaller semantic and phonetic units that should be considered characters (Sproat 2000).
-In alphabetic scripts, this problem is most forcefully exemplified by diacritics.
+Breaking it down further, a script consists of \textsc{graphemes}, and graphemes
+consist of \textsc{characters}. In the linguistic terminology of writing
+systems, a \textsc{character} is a general term for any self-contained element
+in a writing system.\footnote{There is a second interpretation of the term
+\textsc{character}, i.e.~a conventional term for a unit in the Chinese writing
+system \citep{Daniels1996}. This interpretation will not be further explored in
+this paper.} Although in literate societies most people have a strong intuition
+about what the characters are in their particular orthography or orthographies,
+it turns out that the separation of an orthography into separate characters is
+far from trivial. The widespread intuitive notion of a character is strongly
+biased towards educational traditions, like the alphabet taught at schools, and
+technological possibilities, like the available type pieces in a printer's job
+case, the keys on a typewriter, or the symbols displayed in Microsoft Word's
+symbol browser. In practice, characters often consist of multiple building
+blocks, each of which could be considered a character in its own right. For
+example, although a Chinese character may be considered to be a single basic
+unanalyzable unit, at a more fine-grained level of analysis the internal
+structure of Chinese characters is often comprised of smaller semantic and
+phonetic units that should be considered characters \citep{Sproat2000}. In
+alphabetic scripts, this problem is most forcefully exemplified by diacritics.
 
-A \textsc{diacritic} is a mark, or series of marks, that may be above, below, or through other characters (Gaultney 2002).
-Diacritics are sometimes used to distinguish homophonous words, but they are more often used to indicate a modified pronunciation (Daniels \& Bright 1996, xli).
-The central question is whether, for example, <e>, <è>, <a> and <à> should be considered four characters, or different combinations of three characters.
-In general, multiple characters together can form another character, and it is not always possible to decide on principled grounds what should be the basic building blocks of an orthography.
+A \textsc{diacritic} is a mark, or series of marks, that may be above, below, or
+through other characters \citep{Gaultney2002}. Diacritics are sometimes used to
+distinguish homophonous words, but they are more often used to indicate a
+modified pronunciation \citep[xli]{DanielsBright1996}. The central question is
+whether, for example, <e>, <è>, <a> and <à> should be considered four
+characters, or different combinations of three characters. In general, multiple
+characters together can form another character, and it is not always possible to
+decide on principled grounds what should be the basic building blocks of an
+orthography.
 
-For that reason, it is better to analyse an orthography as a collection of graphemes.
-A \textsc{grapheme} is the basic, minimally distinctive symbol of a particular writing system, alike to the phoneme is an abstract representation of a distinct sound in a specific language.
-The term `grapheme' was modeled after the term `phoneme' and represents a contrastive graphical unit in a writing system (see Kohrt 1986 for a historical overview of the term grapheme).
-Most importantly, a single grapheme regularly consists of multiple characters, like <th>, <ou> and <gh> in English (note that each character in these graphemes is also a separate grapheme in English).
-Such complex graphemes are often used to represent single phonemes.
-So, a combination of characters is used to represent a single phoneme.
-Note that the opposite is also found in writing systems, in cases in which a single character represents a combination of two or more phonemes.
-For example, <x> in English orthography represents a combination of the phonemes /k/ and /s/.
+For that reason, it is better to analyze an orthography as a collection of
+graphemes. A \textsc{grapheme} is the basic, minimally distinctive symbol of a
+particular writing system, alike to the phoneme is an abstract representation of
+a distinct sound in a specific language. The term \textsc{grapheme} was modeled
+after the term \textsc{phoneme} and represents a contrastive graphical unit in a
+writing system (see \citet{Kohrt1986} for a historical overview of the term
+grapheme). Most importantly, a single grapheme regularly consists of multiple
+characters, like <th>, <ou> and <gh> in English (note that each character in
+these graphemes is also a separate grapheme in English). Such complex graphemes
+are often used to represent single phonemes. So, a combination of characters is
+used to represent a single phoneme. Note that the opposite is also found in
+writing systems, in cases in which a single character represents a combination
+of two or more phonemes. For example, <x> in English orthography represents a
+combination of the phonemes /k/ and /s/.
 
-Further, conditioned or free variants of a grapheme are called \textsc{allographs}.
-For example, the distinctive forms of Greek sigma are conditioned, with <σ> being used word-internally and <ς> being used at the end of a word.
-In sum, there are many-to-many relationships between phonemes and graphemes as they are expressed in the myriad of language- and resource-specific orthographies.
+Further, conditioned or free variants of a grapheme are called
+\textsc{allographs}. For example, the distinctive forms of Greek sigma are
+conditioned, with <σ> being used word-internally and <ς> being used at the end
+of a word. In sum, there are many-to-many relationships between phonemes and
+graphemes as they are expressed in the myriad of language- and resource-specific
+orthographies.
 
-This exposition of the linguistic terminology involved in describing writing systems has been purposely brief.
-We have highlighted some of the linguistic notions that are pertinent, yet sometimes confused with, the technological definitions developed for the computational processing of the world's writing systems, which we describe in the next section.
+This exposition of the linguistic terminology involved in describing writing
+systems has been purposely brief. We have highlighted some of the linguistic
+notions that are pertinent, yet sometimes confused with, the technological
+definitions developed for the computational processing of the world's writing
+systems, which we describe in the next section.
 
-\section{The Unicode approach}
-\label{the-unicode-approach}
+\section{The Unicode approach} \label{the-unicode-approach}
 
-The conceptualization and terminology of writing systems was rejuvenated through the development of the Unicode Standard, with major input from Mark Davis, co-founder and long-term president of the Unicode Consortium.
-For many years, `exotic' writing systems and phonetic transcription systems on personal computers were constrained by the American Standard Code for Information Interchange (ASCII) character encoding scheme (based on the Latin script), which only allowed for a strongly limited number of different symbols to be encoded.
-This implied that users could either use and adopt the (extended) Latin alphabet or they could assign new symbols to the small number of code points in the ASCII encoding scheme to be rendered by a specifically designed font (Bird \& Simons 2003).
-In this situation, it was necessary to specify the font together with each document to ensure the rightful display of its content.
-To alleviate this problem of assigning different symbols to the same code points, in the late 80's and early 90's the Unicode Consortium set itself the ambitious goal of developing a single universal character encoding to provide a unique number, a code point, for every character in the world's writing systems.
-Nowadays, the Unicode Standard is the default encoding of the technologies that support the World Wide Web and for all modern operating systems, software and programming languages.
+The conceptualization and terminology of writing systems was rejuvenated through
+the development of the Unicode Standard, with major input from Mark Davis,
+co-founder and long-term president of the Unicode Consortium. For many years,
+\textit{exotic} writing systems and phonetic transcription systems on personal
+computers were constrained by the American Standard Code for Information
+Interchange (ASCII) character encoding scheme (based on the Latin script), which
+only allowed for a strongly limited number of different symbols to be encoded.
+This implied that users could either use and adopt the (extended) Latin alphabet
+or they could assign new symbols to the small number of code points in the ASCII
+encoding scheme to be rendered by a specifically designed font
+\citep{BirdSimons2003}. In this situation, it was necessary to specify the font
+together with each document to ensure the rightful display of its content. To
+alleviate this problem of assigning different symbols to the same code points,
+in the late 80's and early 90's the Unicode Consortium set itself the ambitious
+goal of developing a single universal character encoding to provide a unique
+number, a code point, for every character in the world's writing systems.
+Nowadays, the Unicode Standard is the default encoding of the technologies that
+support the World Wide Web and for all modern operating systems, software and
+programming languages.
 
-The Unicode Standard represents a massive step forward because it aims to eradicate the distinction between `universal' (ASCII) versus `language/resource-particular' (Font) by adding as much as possible language-specific information into the universal standard.
-However, there are still language/resource-specific specifications necessary for the proper usage of Unicode, as will be discussed below.
-Within the Unicode structure many of these specifications can be captured by so-called `Locales' or `Locale Descriptions', so we are moving to a new distinction of `universal' (Unicode Standard) versus `language-particular' (Locale Description).
-The major gain is a much larger compatibility on the universal level (because Unicode standardizes a much larger portion of writing system diversity), and much better possibilities for automatic processing on the language-particular level (because Locale Descriptions are computer readable specifications).
+The Unicode Standard represents a massive step forward because it aims to
+eradicate the distinction between \textsc{universal} (ASCII) versus
+\textsc{language/resource-particular} (Font) by adding as much as possible
+language-specific information into the universal standard. However, there are
+still language/resource-specific specifications necessary for the proper usage
+of Unicode, as will be discussed below. Within the Unicode structure many of
+these specifications can be captured by so-called \textsc{Locales} or
+\textsc{Locale Descriptions}, so we are moving to a new distinction of
+\textsc{universal} (Unicode Standard) versus \textsc{language-particular}
+(Locale Description). The major gain is a much larger compatibility on the
+universal level (because Unicode standardizes a much larger portion of writing
+system diversity), and much better possibilities for automatic processing on the
+language-particular level (because Locale Descriptions are computer readable
+specifications).
 
-Each version of the Unicode Standard (as of writing at version 7) consists of a set of specifications and guidelines that include (i) a core specification, (ii) code charts, (iii) standard annexes and (iv) a character database.\footnote{All documents of the Unicode Standard are available at http://www.unicode.org/versions/latest/.
-For a quick survey of the use of terminology inside the Unicode Standard, their glossary is particularly useful, available at http://www.unicode.org/glossary/.
-For a general introduction to the principles of Unicode, Chapter 2 of the core specification, called ``General Structure'', is particularly insightful.
-Different from many other documents of the Unicode Standard, this general introduction is relatively easy to read and illustrated with many interesting examples from various orthography traditions all over the world.} The \textsc{core specification} is a book directed toward human readers that describes the formal standard for encoding multilingual text.
-The \textsc{code charts} provide a humanly readable online reference to the character contents of the Unicode Standard in the form of PDF files.
-The \textsc{Unicode Standard Annexes (UAX)} are a set of technical standards that describe the implementation of the Unicode Standard for software development, Web standards, programming languages, etc.
-The \textsc{Unicode Character Database (UCD)} is a set of computer-readable text files that describe the character properties, including a set of rich character and writing system semantics, for each character in the Unicode Standard.
-In this section, we introduce the basic Unicode concepts, but we will leave out many details.
-Please consult the above mentioned full documentation for a more detailed discussion.
-Further note that the Unicode Standard is exactly that, namely a standard.
-It normatively describes notions and rules to be followed.
-In the actual practice of applying this standard in a computational setting, a specific implementation is necessary.
-The most widely used implementation of the Unicode Standard is the \textsc{International Components for Unicode (ICU)}, which offers C/C++ and Java libraries trying to implement to Unicode Standard.\footnote{http://icu-project.org}
+Each version of the Unicode Standard (as of writing at version 7) consists of a
+set of specifications and guidelines that include (i) a core specification, (ii)
+code charts, (iii) standard annexes and (iv) a character database.\footnote{All
+documents of the Unicode Standard are available at
+\url{http://www.unicode.org/versions/latest/}. For a quick survey of the use of
+terminology inside the Unicode Standard, their glossary is particularly useful,
+available at \url{http://www.unicode.org/glossary/}. For a general introduction
+to the principles of Unicode, Chapter 2 of the core specification, called
+\textsc{General Structure}, is particularly insightful. Different from many
+other documents of the Unicode Standard, this general introduction is relatively
+easy to read and illustrated with many interesting examples from various
+orthography traditions all over the world.} The \textsc{core specification} is a
+book directed toward human readers that describes the formal standard for
+encoding multilingual text. The \textsc{code charts} provide a humanly readable
+online reference to the character contents of the Unicode Standard in the form
+of PDF files. The \textsc{Unicode Standard Annexes (UAX)} are a set of technical
+standards that describe the implementation of the Unicode Standard for software
+development, Web standards, programming languages, etc. The \textsc{Unicode
+Character Database (UCD)} is a set of computer-readable text files that describe
+the character properties, including a set of rich character and writing system
+semantics, for each character in the Unicode Standard. In this section, we
+introduce the basic Unicode concepts, but we will leave out many details. Please
+consult the above mentioned full documentation for a more detailed discussion.
+Further note that the Unicode Standard is exactly that, namely a standard. It
+normatively describes notions and rules to be followed. In the actual practice
+of applying this standard in a computational setting, a specific implementation
+is necessary. The most widely used implementation of the Unicode Standard is the
+\textsc{International Components for Unicode (ICU)}, which offers C/C++ and Java
+libraries trying to implement to Unicode
+Standard.\footnote{\url{http://icu-project.org}}
 
-The Unicode Standard is a \textsc{character encoding system} which goal it is to support the interchange and processing of written characters and text in a computational setting.
-Underlyingly, the character encoding is represented by a range of numerical values called a \textsc{code space}, which is used to encode a set of characters.
-A \textsc{code point} is a unique non-negative integer within a code space (i.e.~within a certain numerical range).
-In the Unicode Standard character encoding system, an \textsc{abstract character}, for example a ``LATIN SMALL LETTER P'', is mapped to a particular code point, such as the decimal value 112, normally represented in hexadecimal and in Unicode parlance as (U+)0070.\footnote{Hexadecimal (`base-16') 0070 is equivalent to decimal (`base-10') 112, which can be calculated by considering that $(0\cdot16^3) + (0\cdot16^2) + (7\cdot16^1) + (0\cdot16^0) = 7\cdot16 = 112$.
-Underlyingly, computers will of course treat this code point binary (`base-2') as 11100000, as can be seen by calculating that $(1\cdot2^7) + (1\cdot2^6) + (1\cdot2^5) + (0\cdot2^4) + (0\cdot2^3) + (0\cdot2^2) + (0\cdot2^1) + (0\cdot2^0) = 64 + 32 + 16 = 112$.} That encoded abstract character is rendered on a computer screen (or printed page) as a \textsc{glyph}, e.g.
-<p>, depending on the \textsc{font} and the context in which that character appears.
+The Unicode Standard is a \textsc{character encoding system} which goal it is to
+support the interchange and processing of written characters and text in a
+computational setting. Underlyingly, the character encoding is represented by a
+range of numerical values called a \textsc{code space}, which is used to encode
+a set of characters. A \textsc{code point} is a unique non-negative integer
+within a code space (i.e.~within a certain numerical range). In the Unicode
+Standard character encoding system, an \textsc{abstract character}, for example
+a \textsc{LATIN SMALL LETTER P}, is mapped to a particular code point, such as
+the decimal value 112, normally represented in hexadecimal and in Unicode
+parlance as (U+)0070.\footnote{Hexadecimal (base-16) 0070 is equivalent to
+decimal (base-10) 112, which can be calculated by considering that $(0\cdot16^3)
++ (0\cdot16^2) + (7\cdot16^1) + (0\cdot16^0) = 7\cdot16 = 112$. Underlyingly,
+computers will of course treat this code point binary (base-2) as 11100000, as
+can be seen by calculating that $(1\cdot2^7) + (1\cdot2^6) + (1\cdot2^5) +
+(0\cdot2^4) + (0\cdot2^3) + (0\cdot2^2) + (0\cdot2^1) + (0\cdot2^0) = 64 + 32 +
+16 = 112$.} That encoded abstract character is rendered on a computer screen (or
+printed page) as a \textsc{glyph}, e.g. <p>, depending on the \textsc{font} and
+the context in which that character appears.
 
-In Unicode Standard terminology, an (abstract) \textsc{character} is the basic encoding unit.
-The term `character' can be quite confusing due to its alternative definitions across different scientific disciplines and because in general the word `character' means many different things to different people.
-It is therefore often preferable to refer to Unicode characters simply as \textsc{code points}, because there is a one-to-one mapping between Unicode characters and their numeric representation.
-The Unicode Consortium (2012) defines a character as ``the smallest component of written language that has semantic value.
-It refers to the abstract meaning and/or general shape, rather than a specific shape, though in code tables some form of visual representation is essential for the reader's understanding.'' Unicode defines characters as abstractions of orthographic symbols, and it does not define visualizations for these characters (although it does presents examples).
-In contrast, a \textsc{glyph} is a concrete graphical representation of a character as it appears when rendered (or rasterized) and displayed on an electronic device or on printed paper.
-For example, <reinsert Gs here> are different glyphs of the same character, i.e.~they may be rendered differently depending on the typography being used, but they all share the same code point.
-From the perspective of Unicode they are `the same thing'.
-In this approach, a \textsc{font} is then simply a collection of glyphs linked to code points.
-Allography is not specified in Unicode (expect for a few exceptional cases, due to legacy encoding issues), but can be specified in a font as a \textsc{contextual variant} (aka presentation form).
+In Unicode Standard terminology, an (abstract) \textsc{character} is the basic
+encoding unit. The term \textsc{character} can be quite confusing due to its
+alternative definitions across different scientific disciplines and because in
+general the word \textsc{character} means many different things to different
+people. It is therefore often preferable to refer to Unicode characters simply
+as \textsc{code points}, because there is a one-to-one mapping between Unicode
+characters and their numeric representation. The \textit{Unicode2012} defines a
+character as ``the smallest component of written language that has semantic
+value. It refers to the abstract meaning and/or general shape, rather than a
+specific shape, though in code tables some form of visual representation is
+essential for the reader's understanding.'' Unicode defines characters as
+abstractions of orthographic symbols, and it does not define visualizations for
+these characters (although it does presents examples). In contrast, a
+\textsc{glyph} is a concrete graphical representation of a character as it
+appears when rendered (or rasterized) and displayed on an electronic device or
+on printed paper. For example, <reinsert Gs here> are different glyphs of the
+same character, i.e.~they may be rendered differently depending on the
+typography being used, but they all share the same code point. From the
+perspective of Unicode they are \textit{the same thing}. In this approach, a
+\textsc{font} is then simply a collection of glyphs linked to code points.
+Allography is not specified in Unicode (expect for a few exceptional cases, due
+to legacy encoding issues), but can be specified in a font as a
+\textsc{contextual variant} (aka presentation form).
 
-Each code point in the Unicode Standard is associated with a set of \textsc{character properties} as defined by the Unicode character property model.\footnote{The character property model is described in http://www.unicode.org/reports/tr23/, but the actual properties are described in http://www.unicode.org/reports/tr44/.
-A simplified overview of the properties is available at http://userguide.icu-project.org/strings/properties.
-The actual code tables listing all properties for all Unicode code points are available at http://www.unicode.org/Public/UCD/latest/ucd/.} Basically, those properties are just a long list of values for each character.
-For example, code point U+0047 has the following properties (among many others):
-\begin{itemize}
-	\item Name: LATIN CAPITAL LETTER G 
-	\item Alphabetic: YES 
-	\item Uppercase: YES 
-	\item Script: LATIN 
-	\item Extender: NO 
-	\item Simple\_Lowercase\_Mapping: 0067 
+Each code point in the Unicode Standard is associated with a set of
+\textsc{character properties} as defined by the Unicode character property
+model.\footnote{The character property model is described in
+\url{http://www.unicode.org/reports/tr23/}, but the actual properties are
+described in \url{http://www.unicode.org/reports/tr44/}. A simplified overview
+of the properties is available at
+\url{http://userguide.icu-project.org/strings/properties}. The actual code
+tables listing all properties for all Unicode code points are available at
+\url{http://www.unicode.org/Public/UCD/latest/ucd/}.} Basically, those
+properties are just a long list of values for each character. For example, code
+point U+0047 has the following properties (among many others): \begin{itemize}
+	\item Name: LATIN CAPITAL LETTER G \item Alphabetic: YES \item
+	Uppercase: YES \item Script: LATIN \item Extender: NO \item
+	Simple\_Lowercase\_Mapping: 0067
 \end{itemize}
 
-These properties contain the basic information of the Unicode Standard and they are necessary to define the correct behavior and conformance required for interoperability in and across different software implementations (as defined in the Unicode Standard Annexes).
-The character properties assigned to each code point is based on each character's behavior in the real-world writing traditions.
-For example, the corresponding lowercase character to U+0047 is U+0067 (though note that the relation between uppercase and lowercase is in many situations much more complex than this, and Unicode has further specifications for those cases).
-Another use of properties is to define the script of a character.
-The Unicode Standard defines the term \textsc{script} as, ``A collection of letters and other written signs used to represent textual information in one or more writing systems.
-For example, Russian is written with a subset of the Cyrillic script; Ukrainian is written with a different subset.
-The Japanese writing system uses several scripts.'' In practice, script is simply defined for each character as the explicit ``Script'' property in the Unicode Character Database.
+These properties contain the basic information of the Unicode Standard and they
+are necessary to define the correct behavior and conformance required for
+interoperability in and across different software implementations (as defined in
+the Unicode Standard Annexes). The character properties assigned to each code
+point is based on each character's behavior in the real-world writing
+traditions. For example, the corresponding lowercase character to U+0047 is
+U+0067 (though note that the relation between uppercase and lowercase is in many
+situations much more complex than this, and Unicode has further specifications
+for those cases). Another use of properties is to define the script of a
+character. The Unicode Standard defines the term \textsc{script} as, ``A
+collection of letters and other written signs used to represent textual
+information in one or more writing systems. For example, Russian is written with
+a subset of the Cyrillic script; Ukrainian is written with a different subset.
+The Japanese writing system uses several scripts.'' In practice, script is
+simply defined for each character as the explicit \textsc{Script} property in
+the Unicode Character Database.
 
-One frequently references property is the \textsc{block} property, which is often used in software applications to impose some structure to the large number of Unicode characters.
-Each character in Unicode belongs to a specific block.
-These blocks are basically an organisational structure to alleviate the administrative burden of keeping Unicode up-to-date.
-Blocks consist of characters that in some way belong together, so that characters are easier to find.
-Some blocks are connected with a specific script, like the Hebrew block or the Gujarati block.
-However, blocks are predefined ranges of code points, and often there will come a point after which the range is completely filled.
-Any extra characters will have to be assigned somewhere else.
-There is, for example, a block `Arabic', which contains most Arabic symbols.
-However, there is also a block `Arabic Supplement', `Arabic Presentation Forms-A' and `Arabic Presentation Form B'.
-The situation with Latin symbols is even more extra.
-In general, the names for block should be taken as a definitional statement.
-For example, many IPA symbols are not located in the aptly-names block ``IPA extensions'', but in other blocks (see Section 5).
+One frequently references property is the \textsc{block} property, which is
+often used in software applications to impose some structure to the large number
+of Unicode characters. Each character in Unicode belongs to a specific block.
+These blocks are basically an organizational structure to alleviate the
+administrative burden of keeping Unicode up-to-date. Blocks consist of
+characters that in some way belong together, so that characters are easier to
+find. Some blocks are connected with a specific script, like the Hebrew block or
+the Gujarati block. However, blocks are predefined ranges of code points, and
+often there will come a point after which the range is completely filled. Any
+extra characters will have to be assigned somewhere else. There is, for example,
+a block \textsc{Arabic}, which contains most Arabic symbols. However, there is
+also a block \textsc{Arabic Supplement}, \textsc{Arabic Presentation Forms-A}
+and \textsc{Arabic Presentation Form B}. The situation with Latin symbols is
+even more extra. In general, the names for block should be taken as a
+definitional statement. For example, many IPA symbols are not located in the
+aptly-names block \textsc{IPA extensions}, but in other blocks (see Section
+\ref{ipa-meets-unicode}).
 
-There are many cases in which a sequence of characters (i.e.~a sequence of more than one code point) represents what a user perceives as an individual unit in a particular orthographic writing system.
-For this reason the Unicode Standard differentiates between \textsc{abstract character} and \textsc{user-perceived character}.
-Sequences of multiple code points that correspond to a single user-perceived characters are called** grapheme clusters** in Unicode parlance.
-Grapheme clusters come in two flavors: (default) grapheme clusters and tailored grapheme clusters.
+There are many cases in which a sequence of characters (i.e.~a sequence of more
+than one code point) represents what a user perceives as an individual unit in a
+particular orthographic writing system. For this reason the Unicode Standard
+differentiates between \textsc{abstract character} and \textsc{user-perceived
+character}. Sequences of multiple code points that correspond to a single
+user-perceived characters are called grapheme clusters in Unicode parlance.
+Grapheme clusters come in two flavors: (default) grapheme clusters and tailored
+grapheme clusters.
 
-The (default) \textsc{grapheme clusters} are locale-independent graphemes, i.e.~they always apply when a particular combination of characters occurs independent of the writing system in which they are used.
-These character combinations are defined in the Unicode Standard as functioning as one \textsc{text element}.\footnote{The Unicode Standard defines text element as: ``A minimum unit of text in relation to a particular text process, in the context of a given writing system.
-In general, the mapping between text elements and code points is many-to-many.''} The simplest example of a grapheme cluster is a base character followed by a letter modifier character.
-For example, the sequence + <̃\textgreater{} (LATIN SMALL LETTER N at U+006E followed by COMBINING TILDE at U+0303) combines visually into , a user-percieved character in writing systems like that of Spanish.
-So, what the user perceives as a single character actually involves a multi-code-point sequence.
-Note that this specific sequence can also be represented with a single \textsc{precomposed} code point, LATIN SMALL LETTER N WITH TILDE at U+00F1, but this is not the case for all multi-code point character sequences.
-The problem that there multiple encodings possible for the same text element has been acknowledged early on in the Unicode Standard (e.g.~for , the sequence U+006E U+0303 should in all situations be treated identically to the precomposed U+00F1), and a system of \textsc{canonical equivalence} is available for such situations.
-Basically, the Unicode Standard offers different kind of normalizations to either decompose all precomposed characters (called \textsc{NFD}, \textsc{Normalization Form Canonical Decomposition}), or compose as much as possible combinations (called \textsc{NFC}, \textsc{Normalization Form Canonical Composition}).
-In current practice of software development, NFC seems to be preferred in most situations and is widely proposed as the preferred canonical form.
+The (default) \textsc{grapheme clusters} are locale-independent graphemes,
+i.e.~they always apply when a particular combination of characters occurs
+independent of the writing system in which they are used. These character
+combinations are defined in the Unicode Standard as functioning as one
+\textsc{text element}.\footnote{The Unicode Standard defines text element as:
+``A minimum unit of text in relation to a particular text process, in the
+context of a given writing system. In general, the mapping between text elements
+and code points is many-to-many.''} The simplest example of a grapheme cluster
+is a base character followed by a letter modifier character. For example, the
+sequence + <̃\textgreater{} (LATIN SMALL LETTER N at U+006E followed by
+COMBINING TILDE at U+0303) combines visually into , a user-perceived character
+in writing systems like that of Spanish. So, what the user perceives as a single
+character actually involves a multi-code-point sequence. Note that this specific
+sequence can also be represented with a single \textsc{precomposed} code point,
+LATIN SMALL LETTER N WITH TILDE at U+00F1, but this is not the case for all
+multi-code point character sequences. The problem that there multiple encodings
+possible for the same text element has been acknowledged early on in the Unicode
+Standard (e.g.~for , the sequence U+006E U+0303 should in all situations be
+treated identically to the precomposed U+00F1), and a system of
+\textsc{canonical equivalence} is available for such situations. Basically, the
+Unicode Standard offers different kind of normalizations to either decompose all
+precomposed characters (called \textsc{NFD}, \textsc{Normalization Form
+Canonical Decomposition}), or compose as much as possible combinations (called
+\textsc{NFC}, \textsc{Normalization Form Canonical Composition}). In current
+practice of software development, NFC seems to be preferred in most situations
+and is widely proposed as the preferred canonical form.
 
-More difficult for text processing, because less standardized, is what the Unicode Standard terms \textsc{tailored grapheme clusters}.
-Tailored grapheme clusters are locale-dependent graphemes, i.e.~such combination of characters do not function as text elements in all situations.
-For example, the sequence + for the Slovak digraph or the sequence in the Sisaala practical orthography (pronounced as IPA /tʃ/; Moran 2006).
-These grapheme clusters are ``tailored'' in the sense that they must be specified on a language-by-language or a writing system-by-writing system basis.
-The Unicode Standard provides technological specifications for creating locale specific data in so-called \textsc{Unicode Locale Descriptions}, i.e.~a set of specification that defines a set of language-specific elements (e.g.~tailored grapheme clusters, collation order, capitalization-equivalence), as well as other special information, like how to format numbers, dates, or currencies.
-Locale descriptions are saved in the \textsc{Common Locale Data Repository (CLDR)},\footnote{http://cldr.unicode.org/} a repository of language-specific definitions of writing system properties, each of which describes specific usages of characters.
-Each locale can be encoded in a document using the \textsc{Locale Data Markup Language (LDML)}.
-LDML is an XML format and vocabulary for the exchange of structured locale data.
+More difficult for text processing, because less standardized, is what the
+Unicode Standard terms \textsc{tailored grapheme clusters}. Tailored grapheme
+clusters are locale-dependent graphemes, i.e.~such combination of characters do
+not function as text elements in all situations. For example, the sequence + for
+the Slovak digraph or the sequence in the Sisaala practical orthography
+(pronounced as IPA /tʃ/; \cite{Moran2006}). These grapheme clusters are
+\textsc{tailored} in the sense that they must be specified on a
+language-by-language or a writing system-by-writing system basis. The Unicode
+Standard provides technological specifications for creating locale specific data
+in so-called \textsc{Unicode Locale Descriptions}, i.e.~a set of specification
+that defines a set of language-specific elements (e.g.~tailored grapheme
+clusters, collation order, capitalization-equivalence), as well as other special
+information, like how to format numbers, dates, or currencies. Locale
+descriptions are saved in the \textsc{Common Locale Data Repository
+(CLDR)},\footnote{\url{http://cldr.unicode.org/}} a repository of
+language-specific definitions of writing system properties, each of which
+describes specific usages of characters. Each locale can be encoded in a
+document using the \textsc{Locale Data Markup Language (LDML)}. LDML is an XML
+format and vocabulary for the exchange of structured locale data.
 
-Unicode Locale Descriptions allow users to define language- or resource-specific writing systems or orthographies.
-The Unicode Consortium defines \textsc{writing system} only very loosely, as it is not a central concept in the Unicode Standard.
-A writing system is, ``A set of rules for using one or more scripts to write a particular language.
-Examples include the American English writing system, the British English writing system, the French writing system, and the Japanese writing system.''.
-However, there are various drawbacks of locale descriptions for the daily practice of linguistic work in a multilingual setting (see Section 7).
+Unicode Locale Descriptions allow users to define language- or resource-specific
+writing systems or orthographies. The Unicode Consortium defines \textsc{writing
+system} only very loosely, as it is not a central concept in the Unicode
+Standard. A writing system is, ``A set of rules for using one or more scripts to
+write a particular language. Examples include the American English writing
+system, the British English writing system, the French writing system, and the
+Japanese writing system.''. However, there are various drawbacks of locale
+descriptions for the daily practice of linguistic work in a multilingual setting
+(see Section \ref{use-cases}).

### 2015-06-05 chapters/use_cases.tex
wraps line at character width 80 in all tex files

@@
-\chapter{Use cases}
-\label{use-cases}
+\chapter{Use cases} \label{use-cases}
 
@@
 
-We now present several use cases that have motivated the development of orthography profiles. These include:
-\begin{itemize}
-	\item tokenization and error checking
-	\item normalization of orthographic systems to attain interoperability across different source documents, 
-	\item cognate identification for detecting the similarity between words from different languages 
+We now present several use cases that have motivated the development of
+orthography profiles. These include: \begin{itemize}
+	\item tokenization and error checking \item normalization of
+	orthographic systems to attain interoperability across different source
+	documents, \item cognate identification for detecting the similarity
+	between words from different languages
 \end{itemize}
 
-An important assumption of our work is that input sources are encoded in Unicode (UTF-8 to be precise).
-Anything that the Unicode Standard is unable to capture, cannot be captured by an orthography profile.
-We also use Unicode normalization form NFD to decompose all incoming text input into normalized and (Unicode) logically ordered strings.
-This process is of fundamental importance when working with text data in the Unicode Standard.
+An important assumption of our work is that input sources are encoded in Unicode
+(UTF-8 to be precise). Anything that the Unicode Standard is unable to capture,
+cannot be captured by an orthography profile. We also use Unicode normalization
+form NFD to decompose all incoming text input into normalized and (Unicode)
+logically ordered strings. This process is of fundamental importance when
+working with text data in the Unicode Standard.
 
-\section{Tokenization and error checking}
-The most basic use case is tokenization of language data. Tokenization comes in three types:
+\section{Tokenization and error checking} The most basic use case is
+tokenization of language data. Tokenization comes in three types:
 
-- Unicode character tokenization
-- grapheme tokenization
-- tailored grapheme tokenization
+- Unicode character tokenization - grapheme tokenization - tailored grapheme
+tokenization
 
-The simplest form of tokenization provided by any Unicode Standard-compliant software will split an input text stream on Unicode character code points.
-The input text is split on the character bit sequences as they have been encoded by the user.
-This may be in various Unicode Normalization Forms (see Section \ref{}) and tokenization split function returns a byte stream sequence given a particular encoding (see Section \ref{}).
+The simplest form of tokenization provided by any Unicode Standard-compliant
+software will split an input text stream on Unicode character code points. The
+input text is split on the character bit sequences as they have been encoded by
+the user. This may be in various Unicode Normalization Forms (see Section
+\ref{}) and tokenization split function returns a byte stream sequence given a
+particular encoding (see Section \ref{}).
 
-There is increasing support for the regular expression match, commonly ``\\X'', to identify Unicode graphemes (see Section \ref{}) and perform tokenization on these sets of Unicode characters.
-Examples are given in Section \ref{}.
+There is increasing support for the regular expression match, commonly ``\\X'',
+to identify Unicode graphemes (see Section \ref{}) and perform tokenization on
+these sets of Unicode characters. Examples are given in Section \ref{}.
 
-The orthography profile provides the mechanism for tokenizing an input stream on tailored grapheme clusters.
-The orthography's formal specification is given in Section \ref{}.
-To sum, an orthography profile is a list of tailored grapheme clusters and / or orthographic rules that specify how a specific resource of text input should be tokenized.
+The orthography profile provides the mechanism for tokenizing an input stream on
+tailored grapheme clusters. The orthography's formal specification is given in
+Section \ref{}. To sum, an orthography profile is a list of tailored grapheme
+clusters and / or orthographic rules that specify how a specific resource of
+text input should be tokenized.
 
-Once text has been normalized, an additional and straightforward process is Unicode grapheme tokenization.
-More recently this regular expression, often available as short cut ``\textbackslash{X}'', identifies all sequences of ``base'' characters followed by 0 or more combining diacritics.
-For example, the sequence, % <ŋ͡mṍraː> would tokenize as a space-delimited sequence as: <ŋ͡ m ṍ r a ː>.
-For a first pass at identifying grapheme clusters, this is a straightforward
-tokenization.
-However, the linguist will note that both the tie bar (COMBINING DOUBLE INVERTED BREVE, a Unicode Combining Modifier) and the length marker (MODIFIER LETTER TRIANGULAR COLON.
-a Unicode Letter Modifier) do not appear ``correctly'', i.e.~the tie bar is grouped with the first character in the sequence and the length marker appears singly.
-This is necessary, as defined by the Unicode Standard, because of these character's semantic properties -- with Unicode we cannot know if, for example, the MODIFIER LETTER TRIANGULAR COLON should appear before or after the base character that it modifies, cf.~the IPA aspiration marker, , the MODIFIER LETTER SMALL H, which is also a Unicode Letter Modifier and for linguists a diacritic that be be used for pre- or post-aspiration.
-These ambiguities of position in tokenization are not uncommon in IPA, thus orthographic (or source) normalization and tokenization is needed.
+Once text has been normalized, an additional and straightforward process is
+Unicode grapheme tokenization. More recently this regular expression, often
+available as short cut ``\textbackslash{X}'', identifies all sequences of
+``base'' characters followed by 0 or more combining diacritics. For example, the
+sequence, % <ŋ͡mṍraː> would tokenize as a space-delimited sequence as: <ŋ͡ m ṍ r
+a ː>. For a first pass at identifying grapheme clusters, this is a
+straightforward tokenization. However, the linguist will note that both the tie
+bar (COMBINING DOUBLE INVERTED BREVE, a Unicode Combining Modifier) and the
+length marker (MODIFIER LETTER TRIANGULAR COLON. a Unicode Letter Modifier) do
+not appear ``correctly'', i.e.~the tie bar is grouped with the first character
+in the sequence and the length marker appears singly. This is necessary, as
+defined by the Unicode Standard, because of these character's semantic
+properties -- with Unicode we cannot know if, for example, the MODIFIER LETTER
+TRIANGULAR COLON should appear before or after the base character that it
+modifies, cf.~the IPA aspiration marker, , the MODIFIER LETTER SMALL H, which is
+also a Unicode Letter Modifier and for linguists a diacritic that be be used for
+pre- or post-aspiration. These ambiguities of position in tokenization are not
+uncommon in IPA, thus orthographic (or source) normalization and tokenization is
+needed.
+
+\section{Cross-orthographic analysis of writing systems}
+
+Given that orthography profiles are stored in a standard CSV format, we can use
+tools for converting and working with CSV. One such tool is the command line
+utility \textsc{csvkit}.\footnote{\url{https://csvkit.readthedocs.org}}
+
+% insert some examples of creating a table of language | grapheme, language |
+orthographic contexts % any measure of writing systems complexity?
 
@@
 
-Orthographic tokenization requires a specification of the tailored grapheme clusters (linguists: graphemes) to separate orthographic units within an input source.
-In the simplest case, the Unicode graphemic tokenization with regular expression ``\textbackslash{X}'' captures the grapheme clusters, i.e.~the input data is graphemically tokenized and there are no additional string sequences required to capture each grapheme, e.g.~no specification.
-This scenario is the most unlikely.
+Orthographic tokenization requires a specification of the tailored grapheme
+clusters (linguists: graphemes) to separate orthographic units within an input
+source. In the simplest case, the Unicode graphemic tokenization with regular
+expression ``\textbackslash{X}'' captures the grapheme clusters, i.e.~the input
+data is graphemically tokenized and there are no additional string sequences
+required to capture each grapheme, e.g.~no specification. This scenario is the
+most unlikely.
 
-The creation of an orthography typically requires that someone describes knowledge about the input data source's writing system.
-For example, when orthographically tokenizing a dictionary as input, the sequences of grapheme clusters need to be explicitly stated or a regular expression rule can be specified to transform the input.
-We illustrate these methods with examples from Thiesen 1998, a bilingual dictionary of Diccionario Bora-Castellano, and with Huber \& Huber 1992, a comparative vocabulary of sixty-nine indigenous languages of Columbia.
+The creation of an orthography typically requires that someone describes
+knowledge about the input data source's writing system. For example, when
+orthographically tokenizing a dictionary as input, the sequences of grapheme
+clusters need to be explicitly stated or a regular expression rule can be
+specified to transform the input. We illustrate these methods with examples from
+Thiesen 1998, a bilingual dictionary of Diccionario Bora-Castellano, and with
+Huber \& Huber 1992, a comparative vocabulary of sixty-nine indigenous languages
+of Columbia.
 
@@
 
-% ({[}a\textbar{}á\textbar{}e\textbar{}é\textbar{}i\textbar{}í\textbar{}o\textbar{}ó\textbar{}u\textbar{}ú{]})(n)(\s)({[}a\textbar{}á\textbar{}e\textbar{}é\textbar{}i\textbar{}í\textbar{}o\textbar{}ó\textbar{}u\textbar{}ú{]}), \1 \2 \4
+%
+({[}a\textbar{}á\textbar{}e\textbar{}é\textbar{}i\textbar{}í\textbar{}o\textbar{}ó\textbar{}u\textbar{}ú{]})(n)(\s)({[}a\textbar{}á\textbar{}e\textbar{}é\textbar{}i\textbar{}í\textbar{}o\textbar{}ó\textbar{}u\textbar{}ú{]}),
+\1 \2 \4
 
-A preliminary orthography profile can be generated with code available online.\footnote{\url{https://github.com/bambooforest/tokenizer}} The Python script creates a unigram model for individual Unicode characters or Unicode grapheme clusters and their frequency in the source input.
-We find that this gives a broad overview of the input at the character and grapheme levels and it often finds typographic errors in the source input by identifying rarely occurring or one-off characters in the input.
-The models also provides the basis for the ``quick'' development of an initial orthography profile by saving the user keyboarding time.
+A preliminary orthography profile can be generated with code available
+online.\footnote{\url{https://github.com/bambooforest/tokenizer}} The Python
+script creates a unigram model for individual Unicode characters or Unicode
+grapheme clusters and their frequency in the source input. We find that this
+gives a broad overview of the input at the character and grapheme levels and it
+often finds typographic errors in the source input by identifying rarely
+occurring or one-off characters in the input. The models also provides the basis
+for the ``quick'' development of an initial orthography profile by saving the
+user keyboarding time.
 
-Note that these symbols are still phonetically unspecified, i.e.~symbol may or may not represent, say, a front open unrounded vowel.
+Note that these symbols are still phonetically unspecified, i.e.~symbol may or
+may not represent, say, a front open unrounded vowel.
 
-The orthography profile is designed so that specified strings of characters are captured for tokenization.
+The orthography profile is designed so that specified strings of characters are
+captured for tokenization.
 
@@
 
-The graphemic cluster specifications in the orthography profile are read into a trie data structure, which is used to tokenize the incoming input source by identifying sentinels (tokenization boundaries) in the input stream, which we then separate grapheme clusters by spaces.
+The graphemic cluster specifications in the orthography profile are read into a
+trie data structure, which is used to tokenize the incoming input source by
+identifying sentinels (tokenization boundaries) in the input stream, which we
+then separate grapheme clusters by spaces.
 
@@
 
-Comparative wordlists often start as the compilation of different sources with different writing systems.
-The Dogon comparative wordlist is an example of a compilation of slightly divergent transcription practices, which we will use to illustrate our point and to orthographically normalize an input source by transforming different transcriptions into a single IPA-like transcription.
+Comparative wordlists often start as the compilation of different sources with
+different writing systems. The Dogon comparative wordlist is an example of a
+compilation of slightly divergent transcription practices, which we will use to
+illustrate our point and to orthographically normalize an input source by
+transforming different transcriptions into a single IPA-like transcription.
 
@@
 
-Once we have normalized and tokenized our text elements, we can apply various sorts of source normalization across resources to create interoperable and comparable data.
+Once we have normalized and tokenized our text elements, we can apply various
+sorts of source normalization across resources to create interoperable and
+comparable data.
 
@@
 
-We use orthography profiles in our work to describe cross-linguistic difference between writing systems.
+We use orthography profiles in our work to describe cross-linguistic difference
+between writing systems.
 
@@
 
-The identification of cognates is a non-trivial task, which involves identifying the semantic and phonetic similarity of words.
+The identification of cognates is a non-trivial task, which involves identifying
+the semantic and phonetic similarity of words.
 
@@
 
-Cognate identification is further complicated by the fact that languages are written differently, and even more so for the low resource languages that we work with, linguists' phonetic transcriptions are heterogeneous and typically document-specific, i.e.~two linguists working on the same language variety most often develop different practical orthographies by using different symbols and different orthographic rules.
-Thus the essence of language documentation and description is diverse and ephemeral in nature.
-The encoding of linguistic data and linguistic data models is diverse and abstracting away from individual analytical preferences requires a document-by-document approach and a mechanism for standardization of writing to form an interlingual pivot so that different descriptions can be normalized into one encoding system to undertake analysis.
+Cognate identification is further complicated by the fact that languages are
+written differently, and even more so for the low resource languages that we
+work with, linguists' phonetic transcriptions are heterogeneous and typically
+document-specific, i.e.~two linguists working on the same language variety most
+often develop different practical orthographies by using different symbols and
+different orthographic rules. Thus the essence of language documentation and
+description is diverse and ephemeral in nature. The encoding of linguistic data
+and linguistic data models is diverse and abstracting away from individual
+analytical preferences requires a document-by-document approach and a mechanism
+for standardization of writing to form an interlingual pivot so that different
+descriptions can be normalized into one encoding system to undertake analysis.
 
@@
 
-\% describe algorithm for tokenizing IPA (Unicode grapheme cluster tokenization and then alignment of Letter Modifiers and Tie Bars)
+\% describe algorithm for tokenizing IPA (Unicode grapheme cluster tokenization
+and then alignment of Letter Modifiers and Tie Bars)
 
@@
 
-\section{Diacritics: Unicode letter modifiers (Lm)}\label{diacritics-unicode-letter-modifiers-lm}
+\section{Diacritics: Unicode letter modifiers
+(Lm)}\label{diacritics-unicode-letter-modifiers-lm}
 
@@
 
-\section{({[}ː\textbar{}ʰ\textbar{}ʼ\textbar{}ʿ\textbar{}ʾ\textbar{}̜\textbar{}̠\textbar{}̟\textbar{}͡\textbar{}̪\textbar{}̰\textbar{}̣\textbar{}̥\textbar{}̩\textbar{}‿\textbar{}.\textbar{}→\textbar{}˻\textbar{}+\textbar{}̃\textbar{}̫\textbar{}̤{]})}\label{ux2d0ux2b0ux2bcux2bfux2be.}
+\section{({[}ː\textbar{}ʰ\textbar{}ʼ\textbar{}ʿ\textbar{}ʾ\textbar{}̜\textbar{}̠\textbar{}̟\textbar{}͡\textbar{}̪\textbar{}̰\textbar{}̣\textbar{}̥\textbar{}̩\textbar{}‿\textbar{}.\textbar{}→\textbar{}\textbar{}+\textbar{}̃\textbar{}̫\textbar{}̤{]})}\label{ux2d0ux2b0ux2bcux2bfux2be.}
 
@@
 
-The most basic overall Unicode character property is the \textsc{General Category}, which categorizes Unicode characters into: \emph{Letters, Punctuation, Symbols, Marks, Numbers, Separators, }and* Other*.
-Each Unicode character property also has a character property value.
-So each General Category is denoted by a property value by a single letter abbreviation, e.g.
-``L'' for Letter.
-Each General Category property also has subcategories that are also properties with property values.
-For example, ``Letters'' is broken down into Uppercase Letter (Lu), Lowercase Letter (Ll), Titlecase Letter (Lt), Modifier Letter (Lm) and Other Letter (Lo).\footnote{The subcategory ``Other Letter'' includes characters such as uncased letters as in the Chinese writing system.} By defining character properties for each code point, algorithmic implementations such as text processing applications or regular expression engines that are conformant to the Unicode Standard can be used to recognize whole (sub)categories of characters or ranges of characters, e.g.~to match proper names in English in title case or more generally to match character types, such as scripts, or symbols and integers in email addresses, phone numbers, etc.
+The most basic overall Unicode character property is the \textsc{General
+Category}, which categorizes Unicode characters into: \emph{Letters,
+Punctuation, Symbols, Marks, Numbers, Separators, }and* Other*. Each Unicode
+character property also has a character property value. So each General Category
+is denoted by a property value by a single letter abbreviation, e.g. ``L'' for
+Letter. Each General Category property also has subcategories that are also
+properties with property values. For example, ``Letters'' is broken down into
+Uppercase Letter (Lu), Lowercase Letter (Ll), Titlecase Letter (Lt), Modifier
+Letter (Lm) and Other Letter (Lo).\footnote{The subcategory ``Other Letter''
+includes characters such as uncased letters as in the Chinese writing system.}
+By defining character properties for each code point, algorithmic
+implementations such as text processing applications or regular expression
+engines that are conformant to the Unicode Standard can be used to recognize
+whole (sub)categories of characters or ranges of characters, e.g.~to match
+proper names in English in title case or more generally to match character
+types, such as scripts, or symbols and integers in email addresses, phone
+numbers, etc.
 
-Script information is useful for algorithmic processes, such as collation, searching and for multilingual text processing.
-For example, if a regular expression engine includes functionality to match characters at the level of script, then it provides the ability to identify characters, e.g.
-\match{Greek}, which tests whether letters in a text belong to the Greek script or not.
-This functionality can be used to determine the boundaries between different writing systems that appear in the same document.
-In Section 6 (Use cases of grapheme tokenization) we provide an example of a regular expression match at the level of character property.
+Script information is useful for algorithmic processes, such as collation,
+searching and for multilingual text processing. For example, if a regular
+expression engine includes functionality to match characters at the level of
+script, then it provides the ability to identify characters, e.g. \match{Greek},
+which tests whether letters in a text belong to the Greek script or not. This
+functionality can be used to determine the boundaries between different writing
+systems that appear in the same document. In Section 6 (Use cases of grapheme
+tokenization) we provide an example of a regular expression match at the level
+of character property.
 
-Unicode-aware regular expression engines may contain a meta-character, ``\textbackslash{X}'', that matches this level of grapheme cluster, i.e.~a base character followed by some number of accents or diacritical marks.
-Grapheme clusters are important to determine word or line boundaries, for collation (`ordering') and for user interface interaction (e.g.~mouse selection, backspacing, cursor movement).
+Unicode-aware regular expression engines may contain a meta-character,
+``\textbackslash{X}'', that matches this level of grapheme cluster, i.e.~a base
+character followed by some number of accents or diacritical marks. Grapheme
+clusters are important to determine word or line boundaries, for collation
+(`ordering') and for user interface interaction (e.g.~mouse selection,
+backspacing, cursor movement).
 

### 2015-06-05 chapters/_remove/orthography_profiles.tex
major style revision

--- 2015-05-29 chapters/originaltext.tex
@@
-The Unicode Cookbook for Linguists Managing orthographies in multilingual datasets using orthography profiles
-
-Steven Moran \& Michael Cysouw
-
-\textsc{Abstract}
-
-In this paper, we define the notion of an orthography profile, which is a language-specific (or even resource-specific) description of the units and rules that are needed to adequately model a writing system.
-An orthography profile describes the Unicode code points, characters, graphemes and orthographic rules as used in a writing system.
-Our orthography profiles are used to describe writing systems and possibly even to transliterate them into another (for example IPA-like) transcriptions with the goal to harmonize orthographies across languages and resources on an ad-hoc basis.
-The central aspect of our approach is that we introduce an intermediate step of graphemic tokenization, which is closer to linguistic practice and easier to manage than transducer-based transliteration.
-We leverage orthography profiles to enable comparative analysis of languages with different writing systems and to perform quality control on datasets with highly specific orthographic systems.
-We also describe how our proposals relate to the current Unicode specification, and introduce some of the more delicate aspects of the Unicode Standard for a linguistic audience.
-
-\textsc{Contents}
-
-{[}{[}TOC{]}{]}
-
-\section{1. Introduction}\label{introduction}
-
-Writing systems arise and develop in a complex mixture of cultural, technological and practical pressures.
-They tend to be highly conservative, in that people who have learned to read and write in a specific way (however impractical or tedious) are mostly unwilling to change their habits, e.g.~they tend to resist spelling reforms.
-In all literate societies there exists a strong socio-political mainstream that tries to force unification of writing (for example by strongly enforcing `right' from `wrong' writing in schools).
-However, there is also a large community of users who take as many liberties in their writing as they can get away with.
-
-For example, the writing of tone diacritics in Yoruba is often proclaimed to be the `right' way to write, although many users of Yoruba writing seem to be perfectly fine with leaving them out.
-As pointed out by the proponents of the `official' rules, there are some homographs when leaving out the tone diacritics (Olúmúyìw 2013: 44).
-However, writing systems (and the languages they represent) are normally full of homophones, which is normally not a problem at all for speakers of the language.
-More importantly, writing is not just a purely functional tool, but just as importantly it is a mechanism to signal social affiliation.
-By showing that you `know the rules' of expressing yourself in writing, others will more easily accept you as a worthy participant in their group.
-And that just as well holds for obeying to the `official' rules when writing a job application, as for obeying to the `informal' rules when writing an SMS to classmates in school.
-The case of Yoruba writing is an exemplary case, as even after more than a century of efforts to standardize the writing systems, there is still a wide range of variation in daily use (Olúmúyìw 2013).
-
-The sometimes cumbersome and sometimes illogical structure, and the enormous variability of existing writing systems is a fact of life scholars have to accept and should try to adapt to as good as possible.
-Our plea here is a proposal for a formalization to do exactly that.
-
-When considering the worldwide linguistic diversity, including all lesser-studied and endangered languages, there exist numerous different orthographies using symbols from the same scripts.
-For example, there are hundreds of orthographies using Latin-based alphabetic scripts.
-All of these orthographies use the same symbols, but these symbols differ in meaning and usage throughout the various orthographies.
-To be able to computationally use and compare different orthographies, we need a way to specify all orthographic idiosyncrasies in a computer-readable format (a process called `tailoring' in Unicode parlance).
-We call such specifications \textsc{orthography profiles}.
-Ideally, these specifications have to be integrated into so-called Unicode locale descriptions, though we will argue that in practice this is often not the most useful solution for the kind of problems arising in the daily practice of linguistics.
-Consequently, a central goal of this paper is to flesh out the linguistic challenges for locale descriptions, and work out suggestions to improve their structure for usage in a linguistic context.
-Conversely, we also aim to improve linguists' understanding and appreciation for the accomplishments of the Unicode Consortium in the development of the Unicode Standard.
-
-The necessity to computationally use and compare different orthographies most forcefully arises in the context of language comparison.
-Concretely, in our current research our goal is to develop quantitative methods for language comparison and historical analysis in order to investigate worldwide linguistic variation and to model the historical and areal processes that underlie linguistic diversity (cf.~Steiner et al.
-2010, List 2012a, 2012b, List \& Moran 2013, Moran \& Prokic 2013).
-In this work, it is crucial to be able to flexibly process across numerous resources with different orthographies.
-In many cases even different resources on the `same' language use different orthographic conventions.
-Another orthographic challenge that we encounter regularly in our linguistic practice is electronic resources on a particular language that claim to follow a specific orthographic convention (often a resource-specific convention), but on closer inspection such resources are almost always not consistently encoded.
-Thus a second goal of our orthography profiles is to allow for an easy specification of orthographic conventions, and use such profiles to check consistency and to report errors to be corrected.
-
-A central step in our proposed solution to this problem is the tailored grapheme separation of strings of symbols, a process we call \textsc{grapheme tokenization}.
-Basically, given some strings of symbols (e.g.~morphemes, words, sentences) in a specific source, our first processing step is to specify how these strings have to be separated into graphemes, considering the specific orthographic conventions used in a particular source document.
-Our experience is that such a graphemic tokenization can be performed without extensive in-depth knowledge about the phonetic and phonological details of the language in question.
-For example, the specification that <ou> is a grapheme of English is a much easier task than to specify what exactly the phonetic values of this grapheme are in any specific occurrence in English words.
-Grapheme separation is a task that can be performed relatively reliably and with limited availability of time and resources (compare, for example, the task of creating a complete phonetic or phonological normalization).
-
-Although grapheme tokenization is only one part of the solution, it is an important and highly fruitful processing step.
-Given a grapheme tokenization, various subsequent tasks become easier, like (a) temporarily reducing the orthography in a processing pipeline, e.g.~only distinguishing high versus low vowels; (b) normalizing orthographies across sources (often including temporary reduction of oppositions), e.g.~specifying an (approximate) mapping to the International Phonetic Alphabet; (c) using co-occurrence statistics across different languages (or different sources in the same language) to estimate the probability of grapheme matches, e.g.~with the goal to find regular sound changes between related languages or transliterations between different sources in the same language.
-
-Before we deal with these proposals, in the first part of this paper (Sections 2 through 5) we give an extended introduction to the notion of encoding (Section 2) and writing systems, both from a linguistic perspective and from the perspective of the Unicode Consortium (Section 3).
-We consider the Unicode Standard to be a breakthrough (and ongoing) development that fundamentally changed the way we look at writing systems, and we aim to provide here a slightly more in-depth survey of the many techniques that are available in the standard.
-A good appreciation for the solutions that the Unicode Standard also allows for a thorough understanding of possible pitfalls that one might encounter when using it (Section 4).
-As an example of the current state-of-the-art, we discuss the rather problematic marriage of the International Phonetic Alphabet (IPA) with Unicode (Section 5).
-
-The second part of the paper (Sections 6 and 7) describes our proposals for how to deal with the Unicode Standard in the daily practice of (comparative) linguists.
-First, we discuss the challenges of characterizing a writing system.
-To solve these problems, we propose the notions of orthography profiles, closely related to Unicode locale descriptions (Section 6).
-Finally, we discuss practical issues with actual examples (Section 7).
-We provide reference implementation of our proposals in R and in Python, available as open-source libraries.
-
-The following conventions are followed in this paper.
-All phonemic and phonetic representations are given in the International Phonetic Alphabet (IPA), unless noted otherwise (International Phonetic Association, 2005).
-Standard conventions are used for distinguishing between graphemic <\textgreater{}, phonemic / / and phonetic {[} {]} representations.
-For character descriptions, we follow the notational conventions of the Unicode Standard (Unicode Consortium 2014).
-Character names are represented in small capital letters (e.g.~LATIN SMALL LETTER SCHWA) and code points are expressed as U\emph{+n} where \emph{n} is a four to six digit hexadecimal number, e.g.~U+0256, which can be rendered as the glyph <ə>.
-
-\section{2. Encoding}\label{encoding}
-
-There are many in-depth histories of the origin and development of writing systems (e.g.~Robinson 1997, Powell 2012), a story that we will not repeat here.
-However, the history of turning writing into computer readable code is not so often told, so we decided to offer a short survey of the major developments here.
-This history turns out to be intimately related to the history of telegraphic communication.\footnote{Because of the recent history as summarised in this section, we have used mostly rather ephemeral internet sources, specifically http://www.unicode.org/history/ and various Wikipedia pages for the information presented here (when not references by traditional literature in the bibliography).
-A useful survey of the historical development of the physical hardware of telegraphy and telecommunication is Huurdeman (2003).
-Most books that discuss the development of encoding of telegraphic communication focus of cryptography (e.g.~Singh 1999) and forego the rather interesting story of `open' encoding that is related here.)
-
-Writing systems have existed for roughly 6000 years, allowing people to exchange messages through time and space.
-Additionally, to bridge large geographic distances, \textsc{telegraphic systems} of communication (from Greek \emph{τῆλε γράφειν} ``distant writing'') have a long and widespread history since ancient times.
-The most widespread telegraphic systems worldwide are so-called whistled languages (Meyer 2015), but also drumming languages (Meyer et al.
-2012) and smoke, fire, hydraulic or flag signals are forms of telegraphy.
-Telegraphy was reinvigorated in the end of the eighteenth century through the introduction of semaphoric systems by Claude Chapelle (since then using flags, flashing lights, or various specially designed contraptions) to convey messages over large distances.
-The `innovation' of those systems was that all characters of the written language were replaced one-to-one by visual signals.
-Since then, all telegraphic systems have taken this principle, namely that any language to be transmitted first has to be turned into some orthographic system, which subsequently is encoded for transmission by the sender, and then turned back into orthographic representation at the receiver side.\footnote{Sound and video-based telecommunication of course takes a different approach by ignoring the written version of language and directly encode sound waves or light patterns.) This of course implies that the usefulness of any such telegraphic encoding completely depends on the often rather haphazard structure of orthographic systems.
-
-In the nineteenth century, \textsc{electric telegraphy} lead to a new approach in which written language characters were encoded by signals sent through a copper wire.
-Originally, `bisignal' codes were used, consisting of two different signals.
-For example, Carl Friedrich Gauss in 1833 used positive and negative current (Mania 2008: 282).
-More famous and influential, Samuel Morse in 1836 used long and short pulses.
-In those bisignal systems each character from the written language was encoded with a different number of signals (between one and five), so two different separators are needed: one between signals and one between characters (in Morse-code there is a short pause between signals and a long pause between characters).\footnote{Actually, Morse-code also includes an extra long pause between words.
-Interestingly, it took a long time to consider the written white-space as a bona-fide character that should simply be encoded with its own code point instead of with a different kind of pause.
-This happened only with the revision of the Baudot-code (see below) by Donald Murray in 1901, in which he finally introduced a white-space code.
-This principle has been followed ever since, and for example was included in the second version of the International Telegraph Alphabet (ITA2) in 1930.)
-
-From those encodings, true \textsc{binary codes} developed with a fixed length of signals per character.
-In such systems only a single separator between signals is needed, because the separation between characters can be established by counting until a fixed number of signals.\footnote{Of course, no explicit separator is needed when the timing of the signals is known, which is the principle used in all modern telecommunication systems.
-An important modern consideration is also how to know where to start counting when you did not catch the start of a message, something that is known in Unicode as `self synchronizing'.) In the context of electric telegraphy, such a system was first established by Émile Baudot in 1870, using a fixed combination of five signals for each written character.\footnote{True binary codes have a longer history, going at least back to the Baconian cipher devised by Francis Bacon in 1605.) There are 25 = 32 possible combination when using five binary signals; an encoding today designated as `5-bit'.
-These codes are sufficient for all Latin letters, but of course they do not suffice for all written symbols, including punctuation and digits.
-As a solution, the Baudot code uses a so-called `shift' character, which signifies that from that point onwards (until shifted back) a different encoding is used, allowing for yet another set of 32 codes.
-In effect, this means that the Baudot code, and the \textsc{International Telegraph Alphabet (ITA)} derived from it, had an extra `bit' of information, so the encoding is actually 6-bit (with 26 = 64 different possible characters).
-For decades, this encoding was the standard for all telegraphy and it is still in limited use today.
-
-To also allow for different uppercase and lowercase letters and a large variety of control characters, the American Standards Association decided to propose a new 7-bit encoding in 1963 (with 27 = 128 different possible characters), known as the \textsc{American Standard Code for Information Interchange(ASCII)}, geared towards the encoding of English orthography.
-With the ascent of other orthographies in computer usage, the wish to encode further variation of Latin letters (like German <ß> or various letters with diacritics like <è>) led the Digital Equipment Corporation to introduce an 8-bit \textsc{Multinational Character Set (MCS)} (with 28 = 256 different possible characters), first used with the introduction of the VT220 Terminal in 1983.
-Because 256 characters were clearly not enough for the many different characters needed in the world's writing systems, the ISO/IEC 8859 standard in 1987 extended the MCS to include 16 different 8-bit code pages.
-For example, part 5 was used for Cyrillic characters, part 6 for Arabic, and part 7 for Greek.\footnote{In effect, because 16 = 24, this means that ISO/IEC 8859 was actually a 8+4 = 12-bit encoding, though with very many duplicates by design (e.g.~all ASCII codes were repeated in each 8-bit code page).
-To be precise, ISO/IEC 8859 used the 7-bit ASCII as a basis for each code page, and defined 16 different 7-bit extensions, leading to (1+16) times 27 = 2,176 possible characters.)
-
-This system almost immediately was understood to be insufficient and impractical, so various initiatives to extend and reorganize the encoding started in the 1980s.
-This led, for example, to various proprietary encodings from Microsoft (e.g.~Windows Latin 1) and Apple (e.g.~Mac OS Roman), which one still sometimes encounters today.
-More wide-ranging, various people in the 1980s started to develop true international code sets.
-In the United States, a group of computer scientists formed the Unicode Consortium, proposing a 16-bit encoding in 1991 (with 216 = 65,536 different possible characters).
-At the same time in Europe, the International Organization for Standardization (ISO) was working on ISO 10646 to supplant the ISO 8859 standard.
-Their first draft of the \textsc{Universal Character Set (UCS)} in 1990 was 31-bit (with theoretically 231 = 2,147,483,648 possible characters, but because of some technical restrictions only 679,477,248 were allowed).
-Since 1991, the Unicode Consortium and the International Organization for Standardization have jointly developed the \textsc{Unicode Standard}, or ISO/IEC 10646, leading to the current system including the original 16-bit Unicode proposal as the \textsc{Basic Multilingual Plane}, and 16 additional planes of 16-bit for further extensions (with in total 1+16 times 216 = 1,114,112 possible characters).
-The most recent version of the Unicode Standard (7.0) was published in June 2014 and it defines 112,956 different characters (Unicode Consortium 2014).
-
-\section{3. Terminology}\label{terminology}
-
-\subsection{The linguistic tradition}\label{the-linguistic-tradition}
-
-In this section we provide a very brief overview of the linguistic terminology concerning writing systems before turning to the slightly different computational terminology in the next section on the Unicode Standard.
-Linguistically speaking, a \textsc{writing system} is a symbolic system that uses visible or tactile signs to represent language in a systematic way.
-The term `writing system' has two mutually exclusive meanings.
-First, it may refer to the way a particular language is written.
-In this sense the term refers to the writing system of a particular language, as, for example, in \emph{``the Serbian writing system uses two scripts: Latin and Cyrillic''}.
-Second, the term `writing system' may refer to a type of symbolic system as used among the world's languages to represent languages, as, for example, in \emph{``alphabetic writing system''}.
-In this latter sense the term refers to how scripts have been classified according to the way that they encode language, as in, for example, \emph{``the Latin and Cyrillic scripts are both alphabetic writing systems''}.
-To avoid confusion, this second notion of `writing system' would have been more aptly called \textsc{script system}.
-The term \textsc{script} refers to a collection of distinct symbols as employed by one or more orthographies.
-For example, both Serbian and Russian are written with subsets of the Cyrillic script.
-A single language, like Serbian or Japanese, can also be written using orthographies based on different scripts.
-Over the years linguists have typologized script systems in a variety of ways, with the tripartite classification of logographic, syllabic, and alphabetic remaining the most popular, even though there are at least half a dozen different types of script systems that can be distinguished (Daniels, 1990, 1996).
-
-Only focussing here on the first sense, we distinguish two different kinds of writing systems as used to write a particular language, namely transcriptions and orthographies.
-First, \textsc{transcription} is a scientific procedure (and also the result of that procedure) for graphically representing the sounds of human speech at the phonetic level.
-It incorporates a set of unambiguous symbols to represent speech sounds, including conventions that specify how these symbols should be combined.
-A \textsc{transcription system} is a specific system of symbols and rules used for transcription of the sounds of a spoken language variety.
-In principle, a transcription system should be language-independent, in that it should be applicable to all spoken human languages.
-The International Phonetic Alphabet (IPA) is a commonly used transcription system that provides a medium for transcribing languages at the phonetic level.
-However, there is a long history of alternative kinds of transcription systems (see Kemp 1994) and today various alternatives are in widespread use (e.g.~X-SAMPA and Cyrillic-based phonetic transcription systems).\footnote{Most users of IPA do not follow the standard explicitly and many dialects based on it have emerged, e.g.~the Africanist and Americanist transcription systems.
-) Note that IPA symbols are also often used to represent language on a phonemic level.
-However, it is important to realize that in this usage the IPA symbols are not a transcription system, but rather an orthography (though with strong links to the pronunciation).
-Further, a transcription system does not need to be highly detailed, but it can also be a system of broad sound classes.
-Although such an approximative transcription is not normally used in linguistics, it is widespread in technological approaches (e.g.~Soundex, Beider \& Morse 2010)\footnote{http://stevemorse.org/phonetics/bmpm.htm , http://stevemorse.org/phonetics/bmpm2.htm) and it is sometimes fruitfully used in automatic approaches to historical linguistics (Dolgopolsky 1986, ASJP\footnote{http://asjp.clld.org/)).
-
-Second, an \textsc{orthography} specifies the symbols, punctuations, and the rules in which a specific language is written in a standardized way.
-Orthographies are often based on a phonemic analysis, but they almost always include idiosyncrasies because of historical developments (like sound changes or loans) and because of the widely-followed principle of lexical integrity (i.e.~the attempt to write the same lexical root in a consistent way, also when synchronic phonemic rules change the pronunciation, for example with final devoicing in many Germanic languages).
-All orthographies are language-specific (and often even resource-specific), although individual symbols or rules might be shared between languages.
-A \textsc{practical orthography} is a strongly phoneme-based writing system designed for practical use by speakers.
-The mapping relation between phonemes and graphemes in practical orthographies is purposely shallow, i.e.~there is mostly a systematic and faithful mapping from a phoneme to a grapheme.
-Practical orthographies are intended to jumpstart written materials development by correlating a writing system with the sound units of a language (cf.~Meinhof \& Jones 1928).
-Symbols from the IPA are often used by linguists in the development of practical orthographies for languages without writing systems, though this usage of IPA symbols should not be confused with transcription (as defined above).
-Further, a \textsc{transliteration} is a mapping between two different orthographies.
-It is the process of ``recording the graphic symbols of one writing system in terms of the corresponding graphic symbols of a second writing system'' (Kemp 2006:396).
-In straightforward cases, such a transliteration is simply a matter of replacing one symbol with another.
-However, there are widespread complications, like one-to-many or many-to-many mappings, which are not always easy, or even possible, to solve without listing all cases individually (cf.~Moran 2012, chp 2).
-
-Breaking it down further, a script consists of \textsc{graphemes}, and graphemes consist of \textsc{characters}.
-In the linguistic terminology of writing systems, a \textsc{character} is a general term for any self-contained element in a writing system.\footnote{There is a second interpretation of the term `character', i.e.~a conventional term for a unit in the Chinese writing system (Daniels 1996).
-This interpretation will not be further explored in this paper.) Although in literate societies most people have a strong intuition about what the characters are in their particular orthography or orthographies, it turns out that the separation of an orthography into separate characters is far from trivial.
-The widespread intuitive notion of a character is strongly biased towards educational traditions, like the alphabet taught at schools, and technological possibilities, like the available type pieces in a printer's job case, the keys on a typewriter, or the symbols displayed in Microsoft Word's symbol browser.
-In practice, characters often consist of multiple building blocks, each of which could be considered a character in its own right.
-For example, although a Chinese character may be considered to be a single basic unanalyzable unit, at a more fine-grained level of analysis the internal structure of Chinese characters is often comprised of smaller semantic and phonetic units that should be considered characters (Sproat 2000).
-In alphabetic scripts, this problem is most forcefully exemplified by diacritics.
-A \textsc{diacritic} is a mark, or series of marks, that may be above, below, or through other characters (Gaultney 2002).
-Diacritics are sometimes used to distinguish homophonous words, but they are more often used to indicate a modified pronunciation (Daniels \& Bright 1996, xli).
-The central question is whether, for example, , , and should be considered four characters, or different combinations of three characters.
-In general, multiple characters together can form another character, and it is not always possible to decide on principled grounds what should be the basic building blocks of an orthography.
-
-For that reason, it is better to analyse an orthography as a collection of graphemes.
-A \textsc{grapheme} is the basic, minimally distinctive symbol of a particular writing system, alike to the phoneme is an abstract representation of a distinct sound in a specific language.
-The term `grapheme' was modeled after the term `phoneme' and represents a contrastive graphical unit in a writing system (see Kohrt 1986 for a historical overview of the term grapheme).
-Most importantly, a single grapheme regularly consists of multiple characters, like <th>, <ou> and <gh> in English (note that each character in these graphemes is also a separate grapheme in English).
-Such complex graphemes are often used to represent single phonemes.
-So, a combination of characters is used to represent a single phoneme.
-Note that the opposite is also found in writing systems, in cases in which a single character represents a combination of two or more phonemes.
-For example, <x> in English orthography represents a combination of the phonemes /k/ and /s/.
-Further, conditioned or free variants of a grapheme are called \textsc{allographs}.
-For example, the distinctive forms of Greek sigma are conditioned, with <σ> being used word-internally and <ς> being used at the end of a word.
-In sum, there are many-to-many relationships between phonemes and graphemes as they are expressed in the myriad of language- and resource-specific orthographies.
-
-This exposition of the linguistic terminology involved in describing writing systems has been purposely brief.
-We have highlighted some of the linguistic notions that are pertinent, yet sometimes confused with, the technological definitions developed for the computational processing of the world's writing systems, which we describe in the next section.
-
-\subsection{The Unicode approach}\label{the-unicode-approach}
-
-The conceptualisation and terminology of writing systems was rejuvenated through the development of the Unicode Standard, with major input from Mark Davis, co-founder and long-term president of the Unicode Consortium.
-For many years, `exotic' writing systems and phonetic transcription systems on personal computers were constrained by the American Standard Code for Information Interchange* *(ASCII) character encoding scheme (based on the Latin script), which only allowed for a strongly limited number of different symbols to be encoded.
-This implied that users could either use and adopt the (extended) Latin alphabet or they could assign new symbols to the small number of code points in the ASCII encoding scheme to be rendered by a specifically designed font (Bird \& Simons 2003).
-In this situation, it was necessary to specify the font together with each document to ensure the rightful display of its content.
-To alleviate this problem of assigning different symbols to the same code points, in the late 80's and early 90's the Unicode Consortium set itself the ambitious goal of developing a single universal character encoding to provide a unique number, a code point, for every character in the world's writing systems.
-Nowadays, the Unicode Standard is the default encoding of the technologies that support the World Wide Web and for all modern operating systems, software and programming languages.
-
-The Unicode Standard represents a massive step forward because it aims to eradicate the distinction between `universal' (ASCII) versus `language/resource-particular' (Font) by adding as much as possible language-specific information into the universal standard.
-However, there are still language/resource-specific specifications necessary for the proper usage of Unicode, as will be discussed below.
-Within the Unicode structure many of these specifications can be captured by so-called `Locales' or `Locale Descriptions', so we are moving to a new distinction of `universal' (Unicode Standard) versus `language-particular' (Locale Description).
-The major gain is a much larger compatibility on the universal level (because Unicode standardizes a much larger portion of writing system diversity), and much better possibilities for automatic processing on the language-particular level (because Locale Descriptions are computer readable specifications).
-
-Each version of the Unicode Standard (as of writing at version 7) consists of a set of specifications and guidelines that include (i) a core specification, (ii) code charts, (iii) standard annexes and (iv) a character database.\footnote{All documents of the Unicode Standard are available at http://www.unicode.org/versions/latest/.
-For a quick survey of the use of terminology inside the Unicode Standard, their glossary is particularly useful, available at http://www.unicode.org/glossary/.
-For a general introduction to the principles of Unicode, Chapter 2 of the core specification, called ``General Structure'', is particularly insightful.
-Different from many other documents of the Unicode Standard, this general introduction is relatively easy to read and illustrated with many interesting examples from various orthography traditions all over the world.
-) The \textsc{core specification} is a book directed toward human readers that describes the formal standard for encoding multilingual text.
-The \textsc{code charts} provide a humanly readable online reference to the character contents of the Unicode Standard in the form of PDF files.
-The \textsc{Unicode Standard Annexes (UAX)} are a set of technical standards that describe the implementation of the Unicode Standard for software development, Web standards, programming languages, etc.
-The \textsc{Unicode Character Database (UCD)} is a set of computer-readable text files that describe the character properties, including a set of rich character and writing system semantics, for each character in the Unicode Standard.
-In this section, we introduce the basic Unicode concepts, but we will leave out many details.
-Please consult the above mentioned full documentation for a more detailed discussion.
-Further note that the Unicode Standard is exactly that, namely a standard.
-It normatively describes notions and rules to be followed.
-In the actual practice of applying this standard in a computational setting, a specific implementation is necessary.
-The most widely used implementation of the Unicode Standard is the \textsc{International Components for Unicode (ICU)}, which offers C/C++ and Java libraries trying to implement to Unicode Standard.\footnote{http://icu-project.org )
-
-The Unicode Standard is a \textsc{character encoding system} which goal it is to support the interchange and processing of written characters and text in a computational setting.
-Underlyingly, the character encoding is represented by a range of numerical values called a \textsc{code space}, which is used to encode a set of characters.
-A \textsc{code point} is a unique non-negative integer within a code space (i.e.~within a certain numerical range).
-In the Unicode Standard character encoding system, an \textsc{abstract character}, for example a ``LATIN SMALL LETTER P'', is mapped to a particular code point, such as the decimal value 112, normally represented in hexadecimal and in Unicode parlance as (U+)0070.\footnote{Hexadecimal (`base-16') 0070 is equivalent to decimal (`base-10') 112, which can be calculated by considering 0·163 + 0·162 + 7·161 + 0·160 = 7·16 = 112.
-Underlyingly, computers will of course treat this code point binary (`base-2') as 11100000, i.e.~112 = 1·27 + 1·26 + 1·25 + 0·24 + 0·23 + 0·22 + 0·21 + 0·20 = 64 + 32 + 16.) That encoded abstract character is rendered on a computer screen (or printed page) as a \textsc{glyph}, e.g.
-<p>, depending on the \textsc{font} and the context in which that character appears.
-
-In Unicode Standard terminology, an (abstract) \textsc{character} is the basic encoding unit.
-The term `character' can be quite confusing due to its alternative definitions across different scientific disciplines and because in general the word `character' means many different things to different people.
-It is therefore often preferable to refer to Unicode characters simply as \textsc{code points}, because there is a one-to-one mapping between Unicode characters and their numeric representation.
-The Unicode Consortium (2012) defines a character as ``the smallest component of written language that has semantic value.
-It refers to the abstract meaning and/or general shape, rather than a specific shape, though in code tables some form of visual representation is essential for the reader's understanding.'' Unicode defines characters as abstractions of orthographic symbols, and it does not define visualizations for these characters (although it does presents examples).
-In contrast, a \textsc{glyph} is a concrete graphical representation of a character as it appears when rendered (or rasterized) and displayed on an electronic device or on printed paper.
-For example, <reinstert Gs here> are different glyphs of the same character, i.e.~they may be rendered differently depending on the typography being used, but they all share the same code point.
-From the perspective of Unicode they are `the same thing'.
-In this approach, a \textsc{font} is then simply a collection of glyphs linked to code points.
-Allography is not specified in Unicode (expect for a few exceptional cases, due to legacy encoding issues), but can be specified in a font as a \textsc{contextual variant} (aka presentation form).
-
-Each code point in the Unicode Standard is associated with a set of \textsc{character properties} as defined by the Unicode character property model.\footnote{The character property model is described in http://www.unicode.org/reports/tr23/, but the actual properties are described in http://www.unicode.org/reports/tr44/.
-A simplified overview of the properties is available at http://userguide.icu-project.org/strings/properties.
-The actual code tables listing all properties for all Unicode code points are available at http://www.unicode.org/Public/UCD/latest/ucd/.
-) Basically, those properties are just a long list of values for each character.
-For example, code point U+0047 has the following properties (among many others):
-\begin{itemize}
-	\item Name: LATIN CAPITAL LETTER G 
-	\item Alphabetic: YES 
-	\item Uppercase: YES 
-	\item Script: LATIN 
-	\item Extender: NO 
-	\item Simple\_Lowercase\_Mapping: 0067 
-\end{itemize}
-
-These properties contain the basic information of the Unicode Standard and they are necessary to define the correct behavior and conformance required for interoperability in and across different software implementations (as defined in the Unicode Standard Annexes).
-The character properties assigned to each code point is based on each character's behavior in the real-world writing traditions.
-For example, the corresponding lowercase character to U+0047 is U+0067 (though note that the relation between uppercase and lowercase is in many situations much more complex than this, and Unicode has further specifications for those cases).
-Another use of properties is to define the script of a character.
-The Unicode Standard defines the term \textsc{script} as, ``A collection of letters and other written signs used to represent textual information in one or more writing systems.
-For example, Russian is written with a subset of the Cyrillic script; Ukrainian is written with a different subset.
-The Japanese writing system uses several scripts.'' In practice, script is simply defined for each character as the explicit ``Script'' property in the Unicode Character Database.
-
-One frequently references property is the \textsc{block} property, which is often used in software applications to impose some structure to the large number of Unicode characters.
-Each character in Unicode belongs to a specific block.
-These blocks are basically an organisational structure to alleviate the administrative burden of keeping Unicode up-to-date.
-Blocks consist of characters that in some way belong together, so that characters are easier to find.
-Some blocks are connected with a specific script, like the Hebrew block or the Gujarati block.
-However, blocks are predefined ranges of code points, and often there will come a point after which the range is completely filled.
-Any extra characters will have to be assigned somewhere else.
-There is, for example, a block `Arabic', which contains most Arabic symbols.
-However, there is also a block `Arabic Supplement', `Arabic Presentation Forms-A' and `Arabic Presentation Form B'.
-The situation with Latin symbols is even more extra.
-In general, the names for block should be taken as a definitional statement.
-For example, many IPA symbols are not located in the aptly-names block ``IPA extensions'', but in other blocks (see Section 5).
-
-There are many cases in which a sequence of characters (i.e.~a sequence of more than one code point) represents what a user perceives as an individual unit in a particular orthographic writing system.
-For this reason the Unicode Standard differentiates between \textsc{abstract character} and \textsc{user-perceived character}.
-Sequences of multiple code points that correspond to a single user-perceived characters are called** grapheme clusters** in Unicode parlance.
-Grapheme clusters come in two flavors: (default) grapheme clusters and tailored grapheme clusters.
-
-The (default) \textsc{grapheme clusters} are locale-independent graphemes, i.e.~they always apply when a particular combination of characters occurs independent of the writing system in which they are used.
-These character combinations are defined in the Unicode Standard as functioning as one \textsc{text element}.\footnote{The Unicode Standard defines text element as: ``A minimum unit of text in relation to a particular text process, in the context of a given writing system.
-In general, the mapping between text elements and code points is many-to-many.'') The simplest example of a grapheme cluster is a base character followed by a letter modifier character.
-For example, the sequence + <̃\textgreater{} (LATIN SMALL LETTER N at U+006E followed by COMBINING TILDE at U+0303) combines visually into , a user-percieved character in writing systems like that of Spanish.
-So, what the user perceives as a single character actually involves a multi-code-point sequence.
-Note that this specific sequence can also be represented with a single \textsc{precomposed} code point, LATIN SMALL LETTER N WITH TILDE at U+00F1, but this is not the case for all multi-code point character sequences.
-The problem that there multiple encodings possible for the same text element has been acknowledged early on in the Unicode Standard (e.g.~for , the sequence U+006E U+0303 should in all situations be treated identically to the precomposed U+00F1), and a system of \textsc{canonical equivalence} is available for such situations.
-Basically, the Unicode Standard offers different kind of normalizations to either decompose all precomposed characters (called \textsc{NFD}, \textsc{Normalization Form Canonical Decomposition}), or compose as much as possible combinations (called \textsc{NFC}, \textsc{Normalization Form Canonical Composition}).
-In current practice of software development, NFC seems to be preferred in most situations and is widely proposed as the preferred canonical form.
-
-More difficult for text processing, because less standardized, is what the Unicode Standard terms \textsc{tailored grapheme clusters}.
-Tailored grapheme clusters are locale-dependent graphemes, i.e.~such combination of characters do not function as text elements in all situations.
-For example, the sequence + for the Slovak digraph or the sequence in the Sisaala practical orthography (pronounced as IPA /tʃ/; Moran 2006).
-These grapheme clusters are ``tailored'' in the sense that they must be specified on a language-by-language or a writing system-by-writing system basis.
-The Unicode Standard provides technological specifications for creating locale specific data in so-called \textsc{Unicode Locale Descriptions}, i.e.~a set of specification that defines a set of language-specific elements (e.g.~tailored grapheme clusters, collation order, capitalization-equivalence), as well as other special information, like how to format numbers, dates, or currencies.
-Locale descriptions are saved in the \textsc{Common Locale Data Repository (CLDR)},\footnote{http://cldr.unicode.org/ ) a repository of language-specific definitions of writing system properties, each of which describes specific usages of characters.
-Each locale can be encoded in a document using the \textsc{Locale Data Markup Language (LDML)}.
-LDML is an XML format and vocabulary for the exchange of structured locale data.
-
-Unicode Locale Descriptions allow users to define language- or resource-specific writing systems or orthographies.
-The Unicode Consortium defines \textsc{writing system} only very loosely, as it is not a central concept in the Unicode Standard.
-A writing system is, ``A set of rules for using one or more scripts to write a particular language.
-Examples include the American English writing system, the British English writing system, the French writing system, and the Japanese writing system.''.
-However, there are various drawbacks of locale descriptions for the daily practice of linguistic work in a multilingual setting (see Section 7).
-
-\section{4. Unicode Pitfalls}\label{unicode-pitfalls}
-
-\subsection{Wrong it is not}\label{wrong-it-is-not}
-
-In this section, we describe some of the most common pitfalls that we have encountered when using the Unicode Standard in our own work, or in discussion with other linguists.
-This section is not meant as a criticism of the decisions made by the Unicode Consortium; on the contrary, we aim to highlight where the technological aspects of the Unicode Standard diverge from many users' intuitions.
-What have sometimes been referred to as problems or inconsistencies in the Unicode Standard are mostly due to legacy compatibility issues, which can lead to unexpected behavior by linguists using the standard.
-However, there are also some cases in which the Unicode Standard has made decisions that theoretically could have been taken differently, but for some reason or another (mostly very good reasons) were accepted as they are now.
-We call behavior that executes without error but does something different than the user expected---often unknowingly---a ``pitfall''.
-
-In this context, it is important to realize that the Unicode Standard was not developed to solve linguistic problems per se, but to offer a consistent computational environment for written language.
-In those cases in which the Unicode Standard behaves differently as expected, we think it is important not to dismiss Unicode as `wrong' or `deficient', because our experience is that in almost all cases the behaviour of the Unicode Standard has been particularly well thought through.
-The Unicode Consortium considers important practical use-cases that from a linguistic point of view are normally not considered.
-Our general guideline for dealing with the Unicode Standard is to accept it as it is, and not to battle windmills.
-Alternatively, of course, it is possible to actively engage in the development of the standard itself, an effort that is highly appreciated by the Unicode Consortium.
-
-\subsection{Pitfall: Blocks}\label{pitfall-blocks}
-
-The Unicode code space is subdivided into blocks of contiguous code points.
-For example, the block called ``Cyrillic'' runs from U+0400 till U+04FF.
-These blocks arose as an attempt at ordering the enormous amount of characters in Unicode, but the ideas of blocks very quickly ran into problems.
-First, the size of a block is fixed, so when a block is full, a new block will have to be instantiated somewhere further in the code space.
-For example, this led to the blocks ``Cyrillic Supplement'', ``Cyrillic Extended-A'' (which are also already full) and ``Cyrillic Extended-B''.
-Second, when a specific character already exists, then it is not duplicated in another block, although the name of the block might indicate that a specific symbol should be available there.
-In general, names of blocks are just an approximate indication of the kind of characters that will be in the block.
-
-The problem with blocks arises because finding the right character among the thousands of Unicode characters is not easy.
-Many software applications present blocks as a primary search mechanism, because the block names suggest where to look for a particular character.
-However, when a user searches for an IPA character in the block ``IPA Extensions'' then many IPA characters will not be found there.
-For example, the velar nasal <ŋ> is not part of the block ``IPA Extensions'' because it was already included as LATIN SMALL LETTER ENG at U+014B in the block ``Latin Extensions-A''.
-
-In general, finding a specific character in the Unicode Standard is often not trivial.
-The names of the blocks can help, but it is not (and never was supposed to be) a foolproof structure.
-It is not the goal nor aim of the Unicode Consortium to provide a user interface to the Unicode Standard.
-If one often encounters the problem of needing to find a suitable character, there are various other useful services for end users available.\footnote{The Unicode website offers a basic interface to the code charts at http://www.unicode.org/charts/index.html.
-As a more flexible interface, we particularly like PopChar from Macility, available for both Macintosh and Windows.
-However, there are various free websites that offer search interfaces to the Unicode code tables, like http://unicode-search.net or http://unicode-search.net.
-A further useful approach for searching characters using shape matching is http://shapecatcher.com.
-)
-
-\subsection{Pitfall: Missing glyphs}\label{pitfall-missing-glyphs}
-
-The Unicode Standard is often praised (and deservedly so) for solving many of the perennial problems with the interchange and display of the world's writing systems.
-However, a common complaint from users is that, while the praise may be true, they mostly just see some boxes on their screen instead of those promised symbols.
-The problem of course is that users' computers do not have any glyphs installed matching the Unicode code points in the file they are trying to inspect.
-It is important to realize that internally in the computer everything still works as expected: any handling of Unicode code points works independently of how they are displayed on the screen.
-So, although a user might only see boxes being displayed, this user should be assured that everything is still in order.
-
-The central problem behind the missing glyphs is that designing actual glyphs includes a lot of different considerations and it is a time-consuming process.
-Many traditional expectations of how specific characters should look like have to be taken into account when designing glyphs.
-Those expectations are often not well documented, and it is mostly up to the knowledge and experience of the font designer to try and conform to them as good as possible.
-Therefore, most designers produce fonts only including glyphs for certains parts of the Unicode Standard, namely for those characters they feel comfortable with.
-At the same time, the number of characters defined by the Unicode Standard is growing with each new version, so it is neigh impossible for any designer to produce glyphs for all characters.
-The result of this is that, almost necessarily, each font only includes glyphs for a subset of the characters in the Unicode Standard.
-
-The simple solution to missing glyphs is thus to install additional fonts providing additional glyphs.
-For the more exotic characters there is often not much choice.
-There are a few particularly large fonts that might be considered.
-First, there is the \textsc{Everson Mono} font made by Michael Everson, which currently includes 9,756 different glyphs (not including Chinese) updated up to Unicode 7.0.\footnote{Everson Mono is available as shareware at http://www.evertype.com/emono/.
-) Already a bit older is the \textsc{Titus Cyberbit Basic} font made by Jost Gippert and Carl-Martin Bunz, which includes 10,044 different glyphs (not including Chinese), but not including newer characters added after Unicode 4.0.\footnote{Titus Cyberbit Basic is available at http://titus.fkidg1.uni-frankfurt.de/unicode/tituut.asp.
-)
-
-Further, we suggest to always install at least one so-called \textsc{fall-back font,} which provides glyphs that at least show the user some information about the underlyingly encoded character.
-Apple Macintoshes have such a font (which is invisible to the user), which is designed by Michael Everson and made available for other systems through the Unicode Consortium.\footnote{The Apple/Everson fallback font is available at http://www.unicode.org/policies/lastresortfont\_eula.html.
-) Further, the \textsc{GNU Unifont} is a clever way to produce bitmaps approximating the intended glyph of each available character, updated to Unicode 7.0.\footnote{The GNU Unifont is available at http://unifoundry.com/unifont.html.
-) Finally, the Summer Institute of Linguistics provides a \textsc{SIL Unicode BMP Fallback Font}, currently available up to Unicode version 6.1.
-This font does not even attempt to show a real glyph, but only shows the hexadecimal code inside a box for each character, so a user can at least see the Unicode codepoint of the character to be displayed.\footnote{The SIL Unicode BMP Fallback Font is available at http://scripts.sil.org/UnicodeBMPFallbackFont.
-)
-
-\subsection{Pitfall: Faulty rendering}\label{pitfall-faulty-rendering}
-
-A similar complaint to missing glyphs, discussed previously, is that while there might be a glyph being displayed, it does not `look right'.
-There are two reasons for unexpected visual display, namely automatic font substitution and faulty rendering.
-Like missing glyphs, any such problems are independent from the Unicode Standard.
-The Unicode Standard only includes very general information about characters and leaves the specific visual display to others to decide on.
-Any faulty display is thus not to be blamed on the Unicode Consortium, but on a complex interplay of different mechanisms happening in a computer to turn Unicode codepoints into visual symbols.
-We will only sketch a few aspects of this complex interplay here.
-
-Most modern software applications (like Microsoft Word) offer some approach to \textsc{automatic font substitution}.
-This means that when a text is written in a specific font (e.g.~Times New Roman) and a inserted Unicode character does not have a glyph within this font, then the software application will automatically search for another font to display the glyph.
-The result will be that some specific glyph will look slightly different from the others.
-This mechanism works differently depending on the software application, and mostly only limited user influence is expected and little feedback is given, which might be rather frustrating to font-aware users.\footnote{For example, Apple Pages does not give any feedback that a font is being replaced, and the user does not seem to have any influence on the choice of replacement (except by manually marking all occurrences).
-In contrast, Microsoft Word does indicate the font replacement by showing the name of the font replacement, but it simply changes the font, so any text written after the replacement is written in a different font as before.
-Both behaviors leave much to be desired.)
-
-The other problem with visual display is related to the so-called `font rendering'.
-\textsc{Font rendering }refers to the process of the actual positioning of Unicode characters on a page of written text.
-This positioning is actually a highly complex problem, and many things can go wrong in the process.
-Well-known rendering problems, like proportional glyph size or ligatures are reasonably well understood.
-However, even within the Latin script, the positioning of diacritics relative to a base character is still a widespread problem.
-Especially when more than one diacritic is supposed to be placed above (or below) each other, this often leads to unexpected effects in many modern software applications.
-The problems arising in Arabic and in many southeast Asian scripts (like Devanagari or Burmese) are even much more complex.
-There are basically three different approaches to font rendering.
-The most widespread is Adobe's and Microsoft's \emph{OpenType }system.
-This approach makes it relatively easy for font developers, as the font itself does not include all details about the precise placement of individual characters.
-For those details, additional script-descriptions are necessary.
-All of those systems can lead to unexpected behaviour.\footnote{For more details about OpenType, see http://www.adobe.com/products/type/opentype.html and http://www.microsoft.com/typography/otspec/.
-Additional systems for complex text layout are, among others, Microsoft's DirectWrite https://msdn.microsoft.com/library/dd368038.aspx and the open-source project HarfBuzz http://www.freedesktop.org/wiki/Software/HarfBuzz/.
-) Alternative systems are Apple's \emph{Apple Advanced Typography} (AAT) and the open-source \emph{Graphite} system from the Summer Institute of Linguistics (SIL).\footnote{More information about AAT can be found at https://developer.apple.com/fonts/.
-SIL's Graphite is described in detail at http://scripts.sil.org/cms/scripts/page.php?site\_id=projects\&item\_id=graphite\_home.
-) In both of these systems, a larger burden is placed on the description inside the font.
-
-\subsection{Pitfall: Homoglyphs}\label{pitfall-homoglyphs}
-
-Homoglyphs are visually indistinguishable glyphs (or highly similar glyphs) that have different code points in the Unicode Standard and thus different character semantics.
-As a principle, the Unicode Standard does not specify how a character appears visually on the page or the screen.
-So in most cases, a different appearance is caused by the specific design of a font, or by user-settings like size or boldface.
-Taking an example already discussed in the previous section, the following symbols <reinsert Gs here> are different glyphs of the same character, i.e.~they may be rendered differently depending on the typography being used, but they all share the same code point (viz.
-LATIN SMALL LETTER G at U+0067).
-In contrast, the symbols <A А Α Ꭺ ᗅ ᴀ ꓮＡ𐊠 𐌀 𝖠 𝙰> are all different code points, although they look highly similar -- in some cases even sharing exactly the same glyph in some fonts.
-All these different A-like characters include the following code points in the Unicode Standard:
-\begin{itemize}
-	\item <A\textgreater{} LATIN CAPITAL LETTER A, at U+0041 
-	\item <А> CYRILLIC CAPITAL LETTER A, at U+0410 
-	\item <Α> GREEK CAPITAL LETTER ALPHA, at U+0391 
-	\item <Ꭺ> CHEROKEE LETTER GO, at U+13AA 
-	\item <ᗅ> CANADIAN SYLLABICS CARRIER GHO, at U+15C5 
-	\item <ᴀ> LATIN LETTER SMALL CAPITAL A, at U+1D00 
-	\item <ꓮ> LISU LETTER A, at U+A4EE 
-	\item <Ａ> FULLWIDTH LATIN CAPITAL LETTER A, at U+FF21 
-	\item <𐊠> CARIAN LETTER A, at U+102A0 
-	\item <𐌀> OLD ITALIC LETTER A, at U+10300 
-	\item <𝖠> MATHEMATICAL SANS-SERIF CAPITAL A, U+1D5A0 
-	\item <𝙰> MATHEMATICAL MONOSPACE CAPITAL A, at U+1D670 
-\end{itemize}
-
-The existence of such homoglyphs is partly due to legacy compatibility, but for the most part these characters are simply different characters that happen to look similar.\footnote{A particularly nice interface to look for homoglyphs is http://shapecatcher.com, based on the principle of recognizing shapes (Belongie et al.
-2002).) Yet, they are supposed to behave different from the perspective of a font designer.
-For example, when designing a Cyrillic font, the will have different aesthetics and different traditional expectation compared to a Latin .
-
-Such homoglyphs are a widespread problem for consistent encoding.
-Although for most users it looks like the words and are identical, in actual fact they do not even share a single code point.\footnote{The first words consists completely of Latin characters, namely U+0076, U+006F, U+0063, U+0065, U+0073, while the second is a mix of Cyrillic and Greek characters, namely U+03BD, U+03BF, U+0041, U+0435, U+0455.) For computers these two words are completely different entities.
-Commonly, when users with Cyrillic or Greek keyboards have to type some Latin-based orthography, they mix similar looking Cyrillic or Greek characters into their text, because those characters are so much easier to type.
-Similarly, when users want to enter an unusual symbol, they normally search by visual impression in their favourite software application, and just pick something that looks reasonably alike to what they expect the glyph to look like.
-
-\subsection{Pitfall: Canonical equivalence}\label{pitfall-canonical-equivalence}
-
-For some characters, there is more than one possible encoding in the Unicode Standard.
-This is a possible pitfall, as this would mean that for the computer there exist multiple different entities that for a user are the same.
-This would, for example, lead to problems with searching, as the computer would search for specific encodings, and not find all expected characters.
-As a solution, the Unicode Standard includes a notion of \textsc{canonical equivalence}.
-Different encodings are explicitly declared as equivalent in the Unicode Standard code tables.
-Further, to harmonize all encodings in a specific piece of text, the Unicode Standard proposes a mechanism of \textsc{normalization}.
-
-Consider for example the characters and following Unicode code points:
-\begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item LATIN CAPITAL LETTER A WITH RING ABOVE (U+00C5) 
-	\item ANGSTROM SIGN (U+212B) 
-	\item LATIN CAPITAL LETTER A (U+0041) + COMBINING RING ABOVE (U+030A) 
-\end{enumerate}
-
-The character, represented here by glyph , is encoded in the Unicode Standard in the first two examples by a single-character sequence; each is assigned a different code point.
-In the third example, the glyph is encoded in a multiple-character sequence that is composed of two character code points.
-All three sequences are \textsc{canonically equivalent}, i.e.~they are strings that represent the same abstract character and because they are not distinguishable by the user, the Unicode Standard requires them to be treated the same in regards to their behavior and appearance.
-Nevertheless, they are encoded differently.
-For example, if one were to search an electronic text (with software that does not apply Unicode Standard normalization) for ANGSTROM SIGN (U+212B), instances of LATIN CAPITAL LETTER A WITH RING ABOVE (U+00C5) would not be found.
-
-In other words, there are equivalent sequences of Unicode characters that should be normalized, i.e.~transformed into a unique Unicode-sanctioned representation of a character sequence called a \textsc{normalization form}.
-Unicode provides a \emph{Unicode Normalization Algorithm}, which essentially puts combining marks into a specific logical order and it defines decomposition and composition transformation rules to convert each string into one of four normalization forms.
-We will discuss here the two most relevant normalization forms: NFC and NFD.
-
-The first of the three characters above is considered the \textsc{Normalization Form C (NFC)}, where ``C'' stands for composition.
-When the process of NFC normalization is applied to the character sequences in 2 and 3, both sequences are normalized into the `pre-composed' character sequence in 1.
-Thus all three canonical character sequences are standardized into one composition form in NFC.
-The other central Unicode normalization form is the \textsc{Normalization Form D (NFD)}, where ``D'' stands for decomposition.
-When NFD is applied to the three examples above, all three, including importantly the single-character sequences in 1 and 2, are normalized into the `decomposed' multiple-sequence of characters in 3.
-Again, all three are then logically equivalent and therefore comparable and syntactically interoperable.
-
-As illustrated, some characters in the Unicode Standard have alternative representations (in fact, many do), but the Unicode Normalization Algorithm can be used to transform certain sequences of characters into canonical (or compatible) forms to test for equivalency.
-To determine equivalence, each character in the Unicode Standard is associated with a combining class, which is formally defined as a character property called ``Canonical Combining Class'' which is specified in the Unicode Character Database.
-The combining class assigned to each code point is a numeric value between 0 and 254 and is used by the Canonical Ordering Algorithm to determine which sequences of combining marks are canonically equivalent.
-Normalization forms, as very briefly described above, can be used to ensure character equivalence by ordering character sequences so that they can be faithfully compared.
-
-It is very important to note that any software applications that is Unicode Standard compliant is free to change the character stream from one representation to another.
-This means that a software application may compose, decompose or reorder characters as its developers desire; as long as the resultant strings are canonically equivalent to the original.
-This might lead to unexpected behaviour for users.
-Various players, like the Unicode Consortium, the W3C, or the TEI recommend NFC in most user-directed situations, and some software applications that we tested indeed seem to automatically convert strings into NFC.\footnote{See the summary of various recommendation here: http://www.win.tue.nl/\textasciitilde{}aeb/linux/uc/nfc\_vs\_nfd.html.
-) This means in practice that if a user, for example, enters <a> and <`>, i.e.~LATIN SMALL LETTER A at U+0061 and COMBINING GRAVE ACCENT at U+0300, this might be converted into <à>, i.e.~LATIN SMALL LETTER A WITH GRAVE at U+00E0.\footnote{The behaviour of software applications can be quite erratic in this respect.
-For example, Apple's TextEdit does not do any conversion on text entry.
-However, when you copy and paste some text inside the same document in ``rich text mode'' (i.e.~RTF-format), it will be transformed into NFC on paste.
-Saving a document does not do any conversion to the glyphs on screen, but it will save the characters in NFC.)
-
-\subsection{Pitfall: Characters are not glyphs}\label{pitfall-characters-are-not-glyphs}
-
-A central principle of Unicode is the distinction between character and glyph.
-A character is the abstract notion of a symbol in a writing system, while a glyph is the concrete drawing of such a symbol.
-In practice, there is a complex interaction between characters and glyphs.
-A single Unicode character may of course be rendered as a single glyph.
-However, a character may also be a `piece' of a glyph, or vice-versa.
-Actually, all possible relations between glyphs and characters are attested.
-
-First, \textsc{a single character may have different contextually determined glyphs.} For example, characters in writing systems like Hebrew and Arabic have different glyphs depending on where they appear in a word.
-Some letters in Hebrew change their form at the end of the word, and in Arabic, primary letters have four contextually-sensitive variants (isolated, word initial, medial and final).
-Second, \textsc{a single character may be rendered as a sequence of multiple glyphs.} For example, in Tamil one Unicode character may result in a combination of a consonant and vowel, which are rendered as two adjacent glyphs by fonts that supports Tamil.
-Third, \textsc{a single glyph may be a combination of multiple characters.
-}For example, the ligature <ﬁ>, a single glyph, is the result of two characters, <f> and <i>, that have undergone glyph substitution by font rendering.
-Like contextually-determined glyphs, ligatures are (intended) artifacts of text processing instructions.
-Finally, \textsc{a single glyph may be a part of a character} (e.g.~diacritics).
-
-Further, the rendering of a glyph is also always dependent on the font being used.
-For example, the Unicode character LATIN SMALL LETTER G appears as and in the Computer Modern and Courier fonts because their typefaces are designed differently.
-Furthermore, font face may change the visual appearance of a character, for example Times New Roman two-story <a> changes to a single-story a in italics <\emph{a}>.
-This becomes a real problem for some phonetic typesetting (see Section 5, Pitfall 2).
-
-In sum, character-to-glyph mappings are complex technological issues that the Unicode Consortium has had to address in the development of the Unicode Standard, but for the lay user they can be utterly confusing because visual rendering does not indicate logical encoding, or, in other words, what you see isn't (necessarily) what you get.
-
-\subsection{Pitfall: File formats}\label{pitfall-file-formats}
-
-Unicode is a character encoding standard, but these characters of course actually appear inside some kind of file.
-The most basic Unicode-based file format is pure line-based text, i.e.~strings of Unicode-encoded characters separated by line breaks (note that these line breaks are what for most people intuitively corresponds to paragraph breaks).
-Unfortunately, even within this apparently basic setting there exist a multitude of variants.
-In general, these different possibilities are well-understood in the software industry, and nowadays they normally do not lead to any problems for the end user.
-However, there are some situations in which a user is suddenly confronted with cryptic questions in the user interface involving abbreviations like LF, CR, BE, LE or BOM.
-Most prominently this occurs with exporting or importing data in several software applications from Microsoft.
-Basically, there are two different issues involved.
-First, the encoding of** line breaks** and, second, the encoding of the Unicode characters into \textsc{code units} and the related issue of \textsc{endianness}.
-
-The issue with \textsc{line breaks} originated with the instructions necessary to direct a printing head to a new line.
-This involves two movements, known as \textsc{carriage return (CR)} (returning the printing head to the start of the line on the page) and \textsc{line feed (LF)} (moving the printing head to the next line on the page).
-Physically, these are two different events, but conceptually together they form one action.
-In the history of computing, various encodings of line breaks have been used (e.g.~CR+LF, LF+CR, only LF, or only CR).
-Currently, all Unix and Unix-derived systems use only LF as code for a line break, while software from Microsoft still uses a combination of CR and LF.
-Today, most software applications recognize both options, and are able to deal with either encoding of line breaks (until rather recently this was not the case, and using the wrong line breaks could lead to unexpected errors).
-Our impression is that there is a strong tendency in software development to standardize on the simpler ``only LF'' encoding for line breaks, and we suggest that everybody use this encoding whenever possible.
-
-A code unit is a sequence of bits used to encode character in a encoding.
-The issue with code units stems from the question how to separate a stream of binary ones and zero, i.e.~bits, into chunks representing Unicode characters.
-Depending on different use cases, the Unicode Standard offers three different approaches, called UTF-32, UTF-16 and UTF-8.\footnote{The letters UTF stand for either ``Unicode Transformation Format'' or ``Universal Coded Character Set Transformation Format'', but the notion of `transformation' is a legacy notion that does not have meaning anymore.
-Nevertheless, the designation ``UTF'' (in capitals) has become an official standard designation, but should probably best be read as simply ``Unicode Format''.) The details of this issue is extensively explained in section 2.5 of the Unicode core specification.
-Basically, \textsc{UTF-32} encodes each character in 32 bits (32 `binary units', i.e.~32 zeros or ones) and is the most disk-space-consuming variant of the three.
-However, it is the most efficient encoding processing-wise, because the computer simply has to separate each character after 32 bits.
-In contrast, \textsc{UTF-16} uses only 16 bits per character, which is sufficient for the large majority of Unicode characters, but not for all of them.
-A special system of `surrogates' is defined within the Unicode Standard to deal with these additional characters.
-The effect is a more disk-space efficient encoding (approximately half the size), while adding a limited computational overhead.
-Finally, \textsc{UTF-8} is a more complex system that dynamically encodes each character with the minimally necessary number of bits, choosing either 8, 16 or 32 bits depending on the character.
-This represents again a strong reduction in space (particularly due to the high frequency of data using ASCII characters that need only 8 bits) at the expense of even more computation necessary to process such strings.
-However, because of the ever growing computational power of modern machines, the processing overhead is in most practical situations a non-issue, while saving on space is still useful, particularly for sending texts over the Internet.
-As an effect, UTF-8 has become the dominant encoding on the World Wide Web.
-We suggest that everybody uses UTF-8 as their default encoding.
-
-A related problem is a general issue about how to store information in computer memory, which is known as \textsc{endianness}.
-The details of this issue go beyond the scope of this paper.
-It suffices to realize that there is a difference between \textsc{big-endian} (BE) storage and \textsc{little-endian} (LE) storage.
-The Unicode Standard offers a possibility to explicitly indicate what kind of storage is used by starting a file with a so-called `Byte Order Mark' (BOM).
-However, the Unicode Standard does not require the usage of BOM, preferring other non-Unicode methods to signal to computers which kind of endianness is used.
-This issue only arises with UTF-32 and UTF-16 encodings, and when using the preferred UTF-8, using a BOM is theoretically possible, but strongly dispreferred according to the Unicode Standard.
-We suggest that everyone tries to prevent the inclusion of BOM in your data.
-
-Together with the topic of normalisation (NFC vs.~NFD, see Pitfall 5 above), this leads to the following nicely cryptic file-encoding recommendation, namely to preferably use \textsc{UTF-8 in NFC without BOM with LF line breaks}, whenever possible.
-
-\section{5. IPA meets Unicode}\label{ipa-meets-unicode}
-
-\subsection{Introducing the International Phonetic Alphabet (IPA)}\label{introducing-the-international-phonetic-alphabet-ipa}
-
-The International Phonetic Alphabet (IPA) is a common standard in linguistics to transcribe sounds of spoken language into some Latin-based characters (International Phonetic Association 1999).
-Although IPA is reasonably easily adhered to with pen and paper, it is not trivial to encode IPA characters electronically.
-Early work addressing the need for a universal computing environment for writing systems and their computational complexity is discussed in Simons 1989.
-For a long time, linguists (like all other computer users) were limited to ASCII-encoded 7-bit characters, which only includes Latin characters, numbers and some punctuation and symbols.
-Restricted to these standard character sets that lacked IPA support or other language-specific graphemes that they needed, linguists devised their own solutions (cf.~Bird and Simons 2003).
-For example, some chose to represent unavailable graphemes with substitutes, e.g.~the combination of to represent .
-Tech-savvy linguists redefined selected characters from a character encoding by mapping custom made fonts to those code points.
-However, one linguist's electronic text would not render properly on another linguist's computer without access to the same font.
-Further, if two character encodings defined two character sets differently, then data could not be reliably and correctly displayed.
-This is a common example of the non-interoperability of data and data formats.
-
-To alleviate this problem, during the late 1980s, SAMPA (Speech Assessment Methods Phonetic Alphabet) was created to represent IPA symbols with 7-bit printable ASCII character sequences, e.g.
-for {[}ɸ{]}.
-Two problems with SAMPA are that (i) it is only a partial encoding of the IPA and (ii) it encodes different languages in separate data tables, instead of a universal alphabet, like IPA.
-SAMPA tables are derived from phonemes appearing in several European languages that were developed as part of a European Commission-funded project to address technical problems like electronic mail exchange (what is now simply called email).
-SAMPA is essentially a hack to work around displaying IPA characters, but it provided speech technology and other fields a basis that has been widely adopted and used in code.
-So, SAMPA was a collection of tables to be compared, instead of a large universal table representing all languages.
-An extended version of SAMPA, called X-SAMPA, set out to include every symbol in the IPA chart including all diacritics (Wells nd.).
-X-SAMPA was considered more universally applicable because it consisted of one table that encoded the set of characters that represented phones/segments in IPA across languages.
-SAMPA and X-SAMPA have been widely used for speech technology and as an encoding system in computational linguistics.
-Eventually, ASCII-encoding of the IPA became depreciated through the advent of the Unicode Standard.
-Note however that many popular software packages used for linguistic analyses still require ASCII input, e.g.~RuG/L04\footnote{http://www.let.rug.nl/kleiweg/L04/) and SplitsTree4.\footnote{http://www.splitstree.org/)
-
-Still, there are a few pitfalls to be aware of when using the Unicode Standard to encode IPA.
-As we have said before, from a linguistic perspective it might look like the Unicode Consortium is making incomprehensible decisions, but it is important to realize that the consortium has tried and is continuing to try to be as consistent as possible across a wide range of use cases, and it does place linguistic traditions above other orthographic possibilities.
-In general, we strongly suggest linguists not to complain about any decisions in the Unicode Standard, but to try and understand the rationale of the Unicode Consortium (which in our experience is almost always well-conceived) and devise ways to work with any unexpected behaviour.
-Many of the current problems derive from the fact that the IPA is clearly historically based on the Latin script, but different enough from most other Latin-based writing systems to warrant special attention.
-This ambivalent status of the IPA glyphs (partly Latin, partly special) is unfortunately also attested in the treatment of IPA in the Unicode Standard.
-In retrospect, it might have been better to consider the IPA (and other transcription systems) to be a special or new kind of script within the Unicode Standard, and treat the obvious similarity to Latin glyphs as a historical relic.
-All IPA glyphs would then have their own code points, instead of the current situation in which some IPA glyphs have special code points, while others are treated as being identical to the `regular' Latin characters.
-Yet, the current situation, however unfortunate, is unlikely to change, so as linguists we will have to learn to deal with the specific pitfalls of IPA within the Unicode Standard.
-In this section, we will describe these pitfalls in some detail.
-
-\subsection{Pitfall: There is no single complete Unicode code block for IPA}\label{pitfall-there-is-no-single-complete-unicode-code-block-for-ipa}
-
-The ambivalent nature of IPA glyphs arises because, on the one hand, the IPA uses Latin-based glyphs like , or .
-From this perspective, the IPA seems to be just another orthographic tradition using Latin characters, all of which do not get a special treatment within the Unicode Standard (just like e.g.~the French, German, or Danish orthographic traditions do not have a special status).
-On the other hand, the IPA uses many special symbols (like turned , mirrored and/or extended Latin glyphs ) not found in any other Latin-based writing system.
-For this reason, and already in the first version of the Unicode Standard (Version 1.0 from 1991), a special block with code points, called ``IPA Extensions'' was included.
-As noted in Section 4, Pitfall 1, the Unicode Standard code space is subdivided into character blocks, which generally encode characters from a single script.
-However, as is illustrated by the IPA, characters that form a single writing system may be dispersed across several different character blocks.
-With its diverse collection of symbols from various scripts and diacritics, the IPA is spread across over 13 blocks in the Unicode Standard:\footnote{This number of blocks depends on whether only IPA-sanctioned symbols are counted or if the phonetic symbols commonly found in the literature are also included, see Moran 2012, Appendix C.)
-\begin{itemize}
-	\item Basic Latin (30 characters), e.g. <a, b, c, d, e> 
-	\item Latin-1 Supplement (4 characters): <æ, ç, ð, ø\textgreater{} 
-	\item Latin Extended-A (3 characters): <ħ, ŋ, œ> 
-	\item Latin Extended-B (5 characters): <ǀ, ǁ, ǂ, ǃ, ȵ> 
-	\item IPA Extensions (70 characters), e.g.: <ɐ, ɑ, ɔ> 
-	\item Spacing Modifier Letters (20 characters), e.g.: <ʰ ʷ ˥> 
-	\item Combining Diacritical Marks (33 characters), e.g.: <̝ ̥ ̪> 
-	\item Greek and Coptic (3 characters): <β, θ, χ> 
-	\item Phonetic Extensions (2 characters): <ᴅ, ᴴ> 
-	\item Phonetic Extensions Supplement (3 characters): <ᶑ , ᶾ, ᶣ> 
-	\item Superscripts and Subscripts (1 character): <ⁿ> 
-	\item Arrows (2 characters): \textless{}↑, ↓\textgreater{} 
-	\item Latin Extended-C (1 character): <ⱱ> 
-\end{itemize}
-
-\subsection{Pitfall: There are many IPA homoglyphs in Unicode}\label{pitfall-there-are-many-ipa-homoglyphs-in-unicode}
-
-Another problem is the large number of homoglyphs, i.e.~different characters that have highly similar glyphs (or even completely identical, depending on the font rendering).
-For example, a speaker of Russian should ideally not use the CYRILLIC SMALL LETTER A at code point U+0430 for IPA transcriptions, but instead the LATIN SMALL LETTER A at code point U+0061, although visually they are mostly indistinguishable, and the Cyrillic character is more easily typed on a Cyrillic keyboard.
-Another example we commonly encounter is the use of LATIN SMALL LETTER G at U+0067, instead of the sanctioned Unicode Standard IPA character for the voiced velar stop LATIN SMALL LETTER SCRIPT G at U+0261.
-One begins to question whether this issue is at all apparent to the working linguist, or if they simply use the U+0067 because it is easily keyboarded and thus saves time, whereas the latter must be cumbersomely inserted as a special symbol in most software.\footnote{This issue was recently addressed by the International Phonetic Association, which has taken the stance that both the keyboard LATIN SMALL LETTER G and the LATIN SMALL LETTER SCRIPT G are valid input characters for the voiced velar plosive.
-Unfortunately, this decision further introduces ambiguity for linguists trying to adhere to a strict Unicode Standard IPA encoding.)
-
-Furthermore, on the one hand even linguists are unlikely to distinguish between the LATIN SMALL LETTER SCHWA at code point U+0259 and LATIN SMALL LETTER TURNED E at U+01DD.
-On the other hand, non-linguists are unlikely to distinguish any semantic difference between an open back unrounded vowel , the LATIN SMALL LETTER ALPHA at U+0251, and the open front unrounded vowel , LATIN SMALL LETTER A at U+0061.
-But even among linguists this distinction leads to problems.
-For example, as pointed out by Mielke (2009), there is a problem stemming from the fact that about 75\% of languages are reported to have a five-vowel system (Maddieson 1984).
-Historically, linguistic descriptions tend not to include precise audio recording and measurements of formants, so this may lead one to ask if the many characters that are used in phonological description reflects a ``transcriptional bias''.
-The common use of in transcriptions could be in part due to the ease of typing the letter on an English keyboard (or for older descriptions, the typewriter).
-We found it to be exceedingly rare that a linguist uses for a low back unrounded vowel.\footnote{One example is Vidal 2001a:75, in which the author states: ``The definition of Pilagá /a/ as {[}+back{]} results from its behavior in certain phonological contexts.
-For instance, uvular and pharyngeal consonants only occur around /a/ and /o/.
-Hence, the characterization of /a/ and /o/ as a natural class of (i.e., {[}+back{]} vowels), as opposed to /i/ and /e/.'') They simply use as long as there is no opposition to .\footnote{See Thomason's Language Log post, ``Why I don't love the International Phonetic Alphabet'', at: http://itre.cis.upenn.edu/\textasciitilde{}myl/languagelog/archives/005287.html.) Making things even more problematic, there is an old typographic tradition that the double-story uses a single-story in italics.
-This leads to the unfortunate effect that in most well-designed fonts the italics of and use the same glyph.
-If this distinction has to be kept upright in italics, the only solution we can currently offer is to use `slanted' glyphs (i.e.~artificially italicised glyphs) instead of real italics (i.e.~special italics glyphs designed by a typographer).\footnote{For example, the widely used IPA font Doulos SIL (http://scripts.sil.org/cms/scripts/page.php?item\_id=DoulosSIL) does not have real italics.
-This leads some word-processing software, like Microsoft Word, to produce slanted glyphs instead.
-That particular combination of font and software application will thus lead to the desired effect.
-However, note that when the text is transferred to another font (i.e.~one that includes real italics) and/or to another software application (like Apple Pages, which does not perform slanting), then this visual appearance will be lost.
-In this case we are thus still in the pre-Unicode situation in which the choice of font and rendering software actually matters.
-The ideal solution from a linguistic point of view would be the introduction of a new IPA code point for a different kind of which explicitly specifies that it should still be rendered as a double-story character when italicized.
-After informal discussion with various Unicode players, our impression is that this highly restricted problem is not sufficiently urgent to introduce even more -like characters in Unicode (which already lead to much confusion, see Section 4, Pitfall 4).
-This is a clear situation in which the Unicode Consortium is not just thinking about linguists, but has a more wide-ranging practical view to consider.)
-
-Some other homoglyphs related to encoding IPA in the Unicode Standard are:
-\begin{itemize}
-	\item The uses of the apostrophe has led to long discussions on the Unicode Standard email list.
-An English keyboard inputs \textless{}`\textgreater{} APOSTROPHE at U+0027, although the `preferred' Unicode apostrophe is the \textless{}'\textgreater{} RIGHT SINGLE QUOTATION MARK at U+2019.
-Yet the glottal stop/glottalization/ejective marker is another completely different character, the MODIFIER LETTER APOSTROPHE at U+02BC, but unfortunately looks mostly highly similar to U+2019.
-	\item There is ambiguous encoding of IPA segments within the Unicode Standard.
-An example is the U+02C1 MODIFIER LETTER REVERSED GLOTTAL STOP vs the U+02E4 MODIFIER LETTER SMALL REVERSED GLOTTAL STOP .
-Both are denoted in the Unicode Standard as the `pharyngealized diacritic' and both appear in various resources representing phonetic data online.
-This is thus an example for which the Unicode Standard does not solve the linguistic standardization problem.
-	\item There is at least one case in which the character name assigned by the Unicode Consortium does not match the IPA's description: in the Unicode Standard at U+01C3 is labeled LATIN LETTER RETROFLEX CLICK, but in IPA is an alveolar or postalveolar click (not retroflex).
-This naming is probably best seen as a simple error in the Unicode Standard.
-Note that most linguists simply seem to use or <ou>) while ligatures are single characters (e.g.
-<ʧ> LATIN SMALL LETTER TESH DIGRAPH at U+02A7).
-Ligatures arose in the context of printing easier-to-read texts, and are included in the Unicode Standard for reasons of legacy encoding.
-However, their usage is discouraged by the Unicode core specification.
-Specifically related to IPA, various phonetic combinations of characters (typically affricates) are available as single code-points in the Unicode Standard, but are designated as ``ligatures'' or ``digraphs'' (confusingly both names appear interchangeably).
-Such glyphs might be used by software to produce a pleasing display, but they should not be hard-coded into the text itself.
-In the context of IPA, characters like the following ligatures should thus \emph{not} be used.
-Instead a combination of two characters is preferred:
-	\item <ʣ> LATIN SMALL LETTER DZ DIGRAPH at U+02A3 (use <d> + <z> instead) 
-	\item <ʧ> LATIN SMALL LETTER TESH DIGRAPH at U+02A7 (use <t> + <ʃ> instead) 
-	\item <ʩ> LATIN SMALL LETTER FENG DIGRAPH at U+02A9 (use <f> + <ŋ> instead) 
-\end{itemize}
-
-However, there are a few Unicode characters that are historically ligatures, but which are today considered as simple characters in the Unicode Standard and thus should be used when writing IPA, namely:
-\begin{itemize}
-	\item <ɮ> LATIN SMALL LETTER LEZH at U+026E 
-	\item <œ> LATIN SMALL LIGATURE OE at U+0153 
-	\item <ɶ> LATIN LETTER SMALL CAPITAL OE at U+0276 
-	\item <æ> LATIN SMALL LETTER AE at U+00E6 
-\end{itemize}
-
-\subsection{Pitfall: The IPA notion of diacritics is not the same as the Unicode Standard's notion of diacritics}\label{pitfall-the-ipa-notion-of-diacritics-is-not-the-same-as-the-unicode-standards-notion-of-diacritics}
-
-Another pitfall is diacritics.
-The problem is that the meaning of the term `diacritics' as used by the IPA is not the same as it used in the Unicode Standard.
-Specifically, diacritics in the IPA-sense are either so-called \textsc{Combining Diacritical Marks} or \textsc{Spacing Modifier Letters} in the Unicode Standard.
-Crucially, Combining Diacritical Marks are by definition combined with the character before them (to form so-called default grapheme clusters, see Section 3).
-In contrast, Spacing Modifier Letters are by definition \emph{not} combined into grapheme clusters with the preceding character, but simply treated as separate letters.
-In the context of the IPA, the following IPA-diacritics are actually Spacing Modifier Letters in the Unicode Standard:
-\begin{itemize}
-	\item Length marks, namely <ː> MODIFIER LETTER TRIANGULAR COLON at U+02D0 and <ˑ> MODIFIER LETTER HALF TRIANGULAR COLON at U+02D1. 
-	\item Tone letters, like <˥> MODIFIER LETTER EXTRA-HIGH TONE BAR at U+02E5, and others like this. 
-	\item Superscript letters, like <ʰ> MODIFIER LETTER SMALL H at U+02B0 or <ˤ> MODIFIER LETTER SMALL REVERSED GLOTTAL STOP at U+02E4, and others like this. 
-	\item <˞> MODIFIER LETTER RHOTIC HOOK at U+02DE. 
-\end{itemize}
-
-Although linguists might expect these characters to belong together with the character in front of them, at least for <ʰ> MODIFIER LETTER SMALL H at U+02B0 the Unicode Consortium's decision to treat it as a separate character is also linguistically correct, because according to the IPA it can be used both for aspiration (more precisely postaspiration following the base character) and preaspiration (preceding the base character).
-Note that there is a mechanism in Unicode to force separate characters to be combined (namely by using the ZERO WIDTH JOINER at U+200D), but this seems to be a rather impractical, and probably not enforceable solution to us.
-
-\subsection{Pitfall: Neither the IPA nor the Unicode Standard enforce a unique diacritic ordering}\label{pitfall-neither-the-ipa-nor-the-unicode-standard-enforce-a-unique-diacritic-ordering}
-
-Also related to diacritics is the question of ordering.
-To our knowledge, the International Phonetic Association does not specify a specific ordering for diacritics that combine with phonetic base symbols; this exercise is left to the reasoning of the transcriber.
-However, such marks have to be explicitly ordered if sequences of them are to be interoperable and compatible.
-An example is a labialized aspirated alveolar plosive: <tʷʰ>.
-There is nothing holding linguists back from using <tʰʷ> instead (with exactly the same intended meaning).
-However, from a technical standpoint, these two sequences are different, e.g.~if both sequences are used in a document, searching for <tʷʰ> will not find any instances of <tʰʷ>, and vice versa.
-Likewise, a creaky voiced syllabic dental nasal can be encoded in various orders, e.g.
-<n̪̰̩>, <n̩̰̪> or <n̩̪̰>.
-
-In accordance with the absence of any specification of ordering in the IPA, the Unicode Standard likewise does not propose any standard orderings.
-Both leave it to the user to be consistent.
-However, there is one aspect of ordering for which the Unicode Standard does present a canonical solution, which is uncontroversial from a linguistic perspective.
-Diacritics in the Unicode Standard (i.e.~Combining Diacritical Marks, see above) are classified in Canonical Combining Classes.
-In practice, the diacritics are distinguished by their position relative to the base character.\footnote{See http://unicode.org/reports/tr44/\#Canonical\_Combining\_Class\_Values for a detailed description.) When applying a Unicode normalization (NFC or NFD, see previous section), the diacritics in different positions are put in a specified order.
-This process therefore harmonizes the difference between different encodings, for example, of a midtone creaky voice vowel <ḛ̄>.
-This grapheme cluster can be encoded either as <e> + <̄> + <̰> or as <e> + <̰> + <̄> .
-To prevent this twofold encoding, the Unicode Standard specifies the second ordering as canonical (diacritics below before diacritics above).
-
-When encoding a string according to the Unicode Standard, it is possible to do this either using the NFC (``composition'') or NFD (``decomposition'') normalization.
-Decomposition implies that precomposed characters (like <á> LATIN SMALL LETTER A WITH ACUTE at U+00E1) will be split into its parts.
-This might sound preferable for a linguistic analysis, as the different diacritics are separated from the base characters.
-However, note that most attached elements like strokes (e.g.~in the <ɨ>), retroflex hooks (e.g.~in <ʐ>) or rhotic hooks (e.g.~in <ɝ>) will not be decomposed, but strangely enough a cedilla (like in <ç>) will be decomposed.
-In general, Unicode decomposition does not behave like a feature decomposition as expected from a linguistic perspective.
-It is thus important to consider Unicode decomposition only as a technical procedure, and not assume that it is linguistically sensible.
-
-Facing the problem of specifying a consistent ordering of diacritics while developing a large database of phonological inventories from the world's languages, Moran (2012: 540) defines a set of diacritic ordering conventions.\footnote{The most recent version of these conventions is online: http://phoible.github.io/conventions/) The conventions are influenced by the linguistic literature, though some ad-hoc decisions had to be taken.
-The goal was to explicitly define all character sequences so that the vast variety of phonemes found in descriptions of the world's language were normalized into consistent character sequences, e.g.~if one language description uses and another , when both are intended to be phonetically equivalent, then a decision to normalize to one form was taken.
-For example, when a character sequence contains more than one character in Spacing Modifier Letters, the order that is proposed is the following (where <\textbar{}> indicates ``or'' and <→> indicates ``precedes''):
-\begin{itemize}
-	\itemsep1pt\parskip0pt\parsep0pt 
-	\item \textsc{Spacing Modifier Letters ordering:} ( unreleased <̚> \textbar{} lateral release <ˡ> \textbar{} nasal release <ⁿ>) → ( palatalized <ʲ>) → ( labialized <ʷ>) → ( velarized <ˠ>) → ( pharyngealized <ˤ>) → ( aspirated <ʰ> \textbar{} ejective <ʼ>) → ( long <ː> \textbar{} half-long <ˑ>)
-\end{itemize}
-
-If a character sequence contains more than one diacritic below the base character, then the place feature is applied first (dental, laminal, apical, fronted, backed, lowered, raised), then the laryngeal setting (voiced, voiceless, creaky voice, breathy voice) and finally the syllabic or non-syllabic marker (for vowels, ATR gets put on between the place and laryngeal setting).
-So:
-\begin{itemize}
-	\itemsep1pt\parskip0pt\parsep0pt 
-	\item \textsc{Combining Diacritical Marks (below) ordering:} ( dental <t̪> \textbar{} laminal <t̻> \textbar{} apical <t̺>) → ( fronted <u̟> \textbar{} backed <e̠>) →( lowered <e̞> \textbar{} raised <e̝>) → ( ATR <e̘ e̙>) → ( voiced <s̬> \textbar{} voiceless <n̥> \textbar{} creaky voice <b̰> \textbar{} breathy voice <b̤>) → ( syllabic <n̩> \textbar{} non-syllabic <e̯>)
-\end{itemize}
-
-Character sequences with diacritics above the base character were not problematic in Moran 2012 because they include only the centralized, mid-centralized and nasalized combining characters.
-Moran (2012) marks tones as singletons with Space Modifier Letters, e.g.
-\textless{}˦\textgreater{} for a phonemic high tone, instead of accent diacritics, alleviating potential conflicts.
-Building on the work of Moran (2012), if a character sequence contains more than one diacritic above the base character, we propose:
-\begin{itemize}
-	\itemsep1pt\parskip0pt\parsep0pt 
-	\item \textsc{Combining Diacritical Marks (above) ordering:} (centralized <ë> \textbar{} mid-centralized <e̽>) → (extra short <ĕ>) →( tone accents, e.g.
-<è> ) → ( Spacing Modifier Letters ) → ( tone letters, e.g.
-<e˦>)
-\end{itemize}
-
-\section{6. Orthography profiles}\label{orthography-profiles}
+\section{Orthography profiles}\label{orthography-profiles}
 

### 2015-06-05 chapters/_remove/terminology.tex
major style revision

--- 2015-06-05 chapters/terminology.tex
@@
-\chapter{Terminology} \label{terminology}
+% \chapter{Terminology}
+% \label{terminology}
 
-\section{The linguistic terminological tradition}
-\label{the-linguistic-tradition}
+\
 
-In this section we provide a very brief overview of the linguistic terminology
+\noindent In the next section we provide a very brief overview of the linguistic terminology
 concerning writing systems before turning to the slightly different
@@
 
+\section{Linguistic Terminology}
+\label{linguistic-terminology}
+
 Linguistically speaking, a \textsc{writing system} is a symbolic system that
 uses visible or tactile signs to represent language in a systematic way. The
-term \textsc{writing system} has two mutually exclusive meanings. First, it may
+term writing system has two mutually exclusive meanings. First, it may
 refer to the way a particular language is written. In this sense the term refers
-to the writing system of a particular language, as, for example, in \emph{``the
-Serbian writing system uses two scripts: Latin and Cyrillic''}. Second, the term
-\textsc{writing system} may also refer to a type of symbolic system as used
-among the world's languages to represent the language, as, for example, in
-\emph{``alphabetic writing system''}. In this latter sense the term refers to
-how scripts have been classified according to the way that they encode language,
-as in, for example, \emph{``the Latin and Cyrillic scripts are both alphabetic
-writing systems''}. To avoid confusion, this second notion of \textsc{writing
-system} would more aptly have been called \textsc{script system}.
+to the writing system of a particular language, as, for example, in \emph{`the
+Serbian writing system uses two scripts: Latin and Cyrillic.'} Second, the term
+writing system may also refer to a type of symbolic system as used among the
+world's languages to represent the language, as, for example, in
+\emph{`alphabetic writing system.'} In this latter sense the term refers to how
+scripts have been classified according to the way that they encode language, as
+in, for example, \emph{`the Latin and Cyrillic scripts are both alphabetic
+writing systems.'} To avoid confusion, this second notion of writing system
+would more aptly have been called \textsc{script system}.
 
-The term \textsc{script} refers to a collection of distinct symbols as employed
-by one or more orthographies. For example, both Serbian and Russian are written
-with subsets of the Cyrillic script. A single language, like Serbian or
-Japanese, can also be written using orthographies based on different scripts.
-Over the years linguists have typologized script systems in a variety of ways,
-with the tripartite classification of logographic, syllabic, and alphabetic
-remaining the most popular, even though there are at least half a dozen
-different types of script systems that can be distinguished
-\citep{Daniels1990,Daniels1996}.
+\subsubsection*{Writing systems}
 
-Only focussing here on the first sense of writing systems, we distinguish two
-different kinds of writing systems used to write a particular language, namely
-transcriptions and orthographies. First, \textsc{transcription} is a scientific
-procedure (and also the result of that procedure) for graphically representing
-the sounds of human speech at the phonetic level. It incorporates a set of
-unambiguous symbols to represent speech sounds, including conventions that
-specify how these symbols should be combined. A transcription system is a
-specific system of symbols and rules used for transcription of the sounds of a
-spoken language variety. In principle, a transcription system should be
-language-independent, in that it should be applicable to all spoken human
-languages. The International Phonetic Alphabet (IPA) is a commonly used
-transcription system that provides a medium for transcribing languages at the
-phonetic level. However, there is a long history of alternative kinds of
-transcription systems (see \citet{Kemp1994}) and today various alternatives are
-in widespread use (e.g.~X-SAMPA and Cyrillic-based phonetic transcription
-systems).\footnote{Most users of IPA do not follow the standard explicitly and
-many dialects based on it have emerged, e.g.~the Africanist and Americanist
-transcription systems.} Note that IPA symbols are also often used to represent
-language on a phonemic level. However, it is important to realize that in this
-usage the IPA symbols are not a transcription system, but rather an orthography
-(though with strong links to the pronunciation). Further, a transcription system
-does not need to be highly detailed, but it can also be a system of broad sound
+Focussing on the first sense, we distinguish two different kinds of 
+writing systems used for a particular language, namely transcriptions and
+orthographies. First, \textsc{transcription} is a scientific procedure (and also
+the result of that procedure) for graphically representing the sounds of human
+speech at the phonetic level. It incorporates a set of unambiguous symbols to
+represent speech sounds, including conventions that specify how these symbols
+should be combined. A transcription system is a specific system of symbols and
+rules used for transcription of the sounds of a spoken language variety. In
+principle, a transcription system should be language-independent, in that it
+should be applicable to all spoken human languages. The \textsc{International
+Phonetic Alphabet} (IPA) is a commonly used transcription system that provides a
+medium for transcribing languages at the phonetic level. However, there is a
+long history of alternative kinds of transcription systems
+\citep[see][]{Kemp2006} and today various alternatives are in widespread use
+(e.g.~X-SAMPA and Cyrillic-based phonetic transcription systems). Many
+users of IPA do not follow the standard to the letter, and many dialects based
+on the IPA have emerged, e.g.~the Africanist and Americanist transcription
+systems. Note that IPA symbols are also often used to represent language on a
+phonemic level. However, it is important to realize that in this usage the IPA
+symbols are not a transcription system, but rather an orthography (though with
+strong links to the pronunciation). Further, a transcription system does not
+need to be as highly detailed as the IPA.\ It can also be a system of broad sound
 classes. Although such an approximative transcription is not normally used in
 linguistics, it is widespread in technological approaches
-(e.g.~\citet{SoundexBeiderMorse2010}\footnote{\url{http://stevemorse.org/phonetics/bmpm.htm},
-\url{http://stevemorse.org/phonetics/bmpm2.htm}} and it is sometimes fruitfully
-used in automatic approaches to historical linguistics
-\citep{Dolgopolsky1986,ASJP16}.\footnote{\url{http://asjp.clld.org/}}
+\citetext{\citealp[Soundex and variants, e.g.~][391--392]{Knuth1973};
+\citealp{postel1969,Beider2008}}, and it is sometimes fruitfully used in
+automatic approaches to historical linguistics
+\citep{Dolgopolsky1986,List2012esslli,Brown2013}.
 
@@
 (i.e.~the attempt to write the same lexical root in a consistent way, also when
-synchronic phonemic rules change the pronunciation, for example with final
+synchronic phonemic rules change the pronunciation, as for example with final
 devoicing in many Germanic languages). All orthographies are language-specific
 (and often even resource-specific), although individual symbols or rules might
-be shared between languages.
-
-A \textsc{practical orthography} is a strongly phoneme-based writing system
-designed for practical use by speakers. The mapping relation between phonemes
-and graphemes in practical orthographies is purposely shallow, i.e.~there is
-mostly a systematic and faithful mapping from a phoneme to a grapheme. Practical
-orthographies are intended to jumpstart written materials development by
-correlating a writing system with the sound units of a language
-(cf.~\citet{MeinhofJones1928}). Symbols from the IPA are often used by linguists
-in the development of practical orthographies for languages without writing
-systems, though this usage of IPA symbols should not be confused with
-transcription (as defined above).
+be shared between languages. A \textsc{practical orthography} is a strongly
+phoneme-based writing system designed for practical use by speakers. The mapping
+relation between phonemes and graphemes in practical orthographies is purposely
+shallow, i.e.~there is mostly a systematic and faithful mapping from a phoneme
+to a grapheme. Practical orthographies are intended to jumpstart written
+materials development by correlating a writing system with the sound units of a
+language \citep[cf.~][]{MeinhofJones1928}. Symbols from the IPA are often used
+by linguists in the development of such practical orthographies for languages
+without writing systems, though this usage of IPA symbols should not be confused
+with transcription (as defined above).
 
@@
 not always easy, or even possible, to solve without listing all cases
-individually (cf.~\citet[chp 2]{Moran2012}).
+individually \citep[cf.~][Ch.~2]{Moran2012}.
+
+\subsubsection*{Script systems}
+
+Different kinds of writing systems are classified into script
+systems. A \textsc{script} is a collection of distinct symbols as
+employed by one or more orthographies. For example, both Serbian and Russian are
+written with subsets of the Cyrillic script. A single language, like Serbian or
+Japanese, can also be written using orthographies based on different scripts.
+Over the years linguists have classified script systems in a variety of ways,
+with the tripartite classification of logographic, syllabic, and alphabetic
+remaining the most popular, even though there are at least half a dozen
+different types of script systems that can be distinguished
+\citep{Daniels1990,Daniels1996}.
 
@@
 after the term \textsc{phoneme} and represents a contrastive graphical unit in a
-writing system (see \citet{Kohrt1986} for a historical overview of the term
+writing system (see \citealp{Kohrt1986} for a historical overview of the term
 grapheme). Most importantly, a single grapheme regularly consists of multiple
@@
 
-This exposition of the linguistic terminology involved in describing writing
+\
+
+\noindent This exposition of the linguistic terminology involved in describing writing
 systems has been purposely brief. We have highlighted some of the linguistic
@@
 
-\section{The Unicode approach} \label{the-unicode-approach}
+\section{The Unicode approach}
+\label{the-unicode-approach}
 
-The conceptualization and terminology of writing systems was rejuvenated through
+The conceptualization and terminology of writing systems was rejuvenated by
 the development of the Unicode Standard, with major input from Mark Davis,
 co-founder and long-term president of the Unicode Consortium. For many years,
-\textit{exotic} writing systems and phonetic transcription systems on personal
+``exotic'' writing systems and phonetic transcription systems on personal
 computers were constrained by the American Standard Code for Information
@@
 
+\subsubsection*{The Unicode Standard}
+
 The Unicode Standard represents a massive step forward because it aims to
@@
 of Unicode, as will be discussed below. Within the Unicode structure many of
-these specifications can be captured by so-called \textsc{Locales} or
-\textsc{Locale Descriptions}, so we are moving to a new distinction of
-\textsc{universal} (Unicode Standard) versus \textsc{language-particular}
-(Locale Description). The major gain is a much larger compatibility on the
-universal level (because Unicode standardizes a much larger portion of writing
-system diversity), and much better possibilities for automatic processing on the
-language-particular level (because Locale Descriptions are computer readable
-specifications).
+these specifications can be captured by so-called \textsc{Locale Descriptions}
+or simply locales, so we are moving to a new distinction of universal (Unicode
+Standard) versus language-particular (Locale Description). The major gain is a
+much larger compatibility on the universal level (because Unicode standardizes a
+much larger portion of writing system diversity), and much better possibilities
+for automatic processing on the language-particular level (because Locale
+Descriptions are computer readable specifications).
 
-Each version of the Unicode Standard (as of writing at version 7) consists of a
-set of specifications and guidelines that include (i) a core specification, (ii)
-code charts, (iii) standard annexes and (iv) a character database.\footnote{All
-documents of the Unicode Standard are available at
-\url{http://www.unicode.org/versions/latest/}. For a quick survey of the use of
-terminology inside the Unicode Standard, their glossary is particularly useful,
-available at \url{http://www.unicode.org/glossary/}. For a general introduction
-to the principles of Unicode, Chapter 2 of the core specification, called
-\textsc{General Structure}, is particularly insightful. Different from many
-other documents of the Unicode Standard, this general introduction is relatively
-easy to read and illustrated with many interesting examples from various
-orthography traditions all over the world.} The \textsc{core specification} is a
-book directed toward human readers that describes the formal standard for
-encoding multilingual text. The \textsc{code charts} provide a humanly readable
-online reference to the character contents of the Unicode Standard in the form
-of PDF files. The \textsc{Unicode Standard Annexes (UAX)} are a set of technical
-standards that describe the implementation of the Unicode Standard for software
-development, Web standards, programming languages, etc. The \textsc{Unicode
-Character Database (UCD)} is a set of computer-readable text files that describe
-the character properties, including a set of rich character and writing system
-semantics, for each character in the Unicode Standard. In this section, we
-introduce the basic Unicode concepts, but we will leave out many details. Please
-consult the above mentioned full documentation for a more detailed discussion.
-Further note that the Unicode Standard is exactly that, namely a standard. It
-normatively describes notions and rules to be followed. In the actual practice
-of applying this standard in a computational setting, a specific implementation
-is necessary. The most widely used implementation of the Unicode Standard is the
-\textsc{International Components for Unicode (ICU)}, which offers C/C++ and Java
-libraries trying to implement to Unicode
+Each version of the Unicode Standard (\citealp{Unicode2014}, as of writing at
+version 7) consists of a set of specifications and guidelines that include (i) a
+core specification, (ii) code charts, (iii) standard annexes and (iv) a
+character database.\footnote{All documents of the Unicode Standard are available
+at \url{http://www.unicode.org/versions/latest/}. For a quick survey of the use
+of terminology inside the Unicode Standard, their glossary is particularly
+useful, available at \url{http://www.unicode.org/glossary/}. For a general
+introduction to the principles of Unicode, Chapter 2 of the core specification,
+called \textsc{general structure}, is particularly insightful. Different from
+many other documents of the Unicode Standard, this general introduction is
+relatively easy to read and illustrated with many interesting examples from
+various orthography traditions all over the world.} The \textsc{core
+specification} is a book directed toward human readers that describes the formal
+standard for encoding multilingual text. The \textsc{code charts} provide a
+humanly readable online reference to the character contents of the Unicode
+Standard in the form of PDF files. The \textsc{Unicode Standard Annexes (UAX)}
+are a set of technical standards that describe the implementation of the Unicode
+Standard for software development, Web standards, and programming languages.
+The \textsc{Unicode Character Database (UCD)} is a set of computer-readable text
+files that describe the character properties, including a set of rich character
+and writing system semantics, for each character in the Unicode Standard. In
+this section, we introduce the basic Unicode concepts, but we will leave out
+many details. Please consult the above mentioned full documentation for a more
+detailed discussion. Further note that the Unicode Standard is exactly that,
+namely a standard. It normatively describes notions and rules to be followed. In
+the actual practice of applying this standard in a computational setting, a
+specific implementation is necessary. The most widely used implementation of the
+Unicode Standard is the \textsc{International Components for Unicode (ICU)},
+which offers C/C++ and Java libraries trying to implement the Unicode
 Standard.\footnote{\url{http://icu-project.org}}
 
-The Unicode Standard is a \textsc{character encoding system} which goal it is to
-support the interchange and processing of written characters and text in a
-computational setting. Underlyingly, the character encoding is represented by a
-range of numerical values called a \textsc{code space}, which is used to encode
-a set of characters. A \textsc{code point} is a unique non-negative integer
-within a code space (i.e.~within a certain numerical range). In the Unicode
-Standard character encoding system, an \textsc{abstract character}, for example
-a \textsc{LATIN SMALL LETTER P}, is mapped to a particular code point, such as
-the decimal value 112, normally represented in hexadecimal and in Unicode
-parlance as (U+)0070.\footnote{Hexadecimal (base-16) 0070 is equivalent to
-decimal (base-10) 112, which can be calculated by considering that $(0\cdot16^3)
-+ (0\cdot16^2) + (7\cdot16^1) + (0\cdot16^0) = 7\cdot16 = 112$. Underlyingly,
+\subsubsection*{Character encoding system}
+
+The Unicode Standard is a \textsc{character encoding system} which
+goal it is to support the interchange and processing of written characters and
+text in a computational setting. Underlyingly, the character encoding is
+represented by a range of numerical values called a \textsc{code space}, which
+is used to encode a set of characters. A \textsc{code point} is a unique
+non-negative integer within a code space (i.e.~within a certain numerical
+range). In the Unicode Standard character encoding system, an \textsc{abstract
+character}, for example the \textsc{latin small letter p}, is mapped to a
+particular code point, in this case the decimal value 112, normally represented in
+hexadecimal, which then looks in Unicode parlance as
+\uni{0070}.\footnote{Hexadecimal (base-16) 0070 is equivalent to decimal
+(base-10) 112, which can be calculated by considering that $(0\cdot16^3) +
+(0\cdot16^2) + (7\cdot16^1) + (0\cdot16^0) = 7\cdot16 = 112$. Underlyingly,
 computers will of course treat this code point binary (base-2) as 11100000, as
@@
 as \textsc{code points}, because there is a one-to-one mapping between Unicode
-characters and their numeric representation. The \textit{Unicode2012} defines a
-character as ``the smallest component of written language that has semantic
-value. It refers to the abstract meaning and/or general shape, rather than a
+characters and their numeric representation. In the Unicode approach, a
+character refers to the abstract meaning and/or general shape, rather than a
 specific shape, though in code tables some form of visual representation is
-essential for the reader's understanding.'' Unicode defines characters as
+essential for the reader's understanding. Unicode defines characters as
 abstractions of orthographic symbols, and it does not define visualizations for
@@
 appears when rendered (or rasterized) and displayed on an electronic device or
-on printed paper. For example, <reinsert Gs here> are different glyphs of the
+on printed paper. For example, <g {\large \textit{g}} \textbf{g}
+{\fontspec{ArialMT} {\small g} \textit{g} \textbf{g}}> are different glyphs of the
 same character, i.e.~they may be rendered differently depending on the
@@
 to legacy encoding issues), but can be specified in a font as a
-\textsc{contextual variant} (aka presentation form).
+\textsc{contextual variant} (a.k.a.~presentation form).
 
@@
 properties are just a long list of values for each character. For example, code
-point U+0047 has the following properties (among many others): \begin{itemize}
-	\item Name: LATIN CAPITAL LETTER G \item Alphabetic: YES \item
-	Uppercase: YES \item Script: LATIN \item Extender: NO \item
-	Simple\_Lowercase\_Mapping: 0067
+point \uni{0047} has the following properties (among many others): 
+\begin{itemize}
+	\item Name: LATIN CAPITAL LETTER G 
+	\item Alphabetic: YES 
+	\item Uppercase: YES 
+	\item Script: LATIN 
+	\item Extender: NO 
+	\item Simple\_Lowercase\_Mapping: 0067 
 \end{itemize}
@@
 point is based on each character's behavior in the real-world writing
-traditions. For example, the corresponding lowercase character to U+0047 is
-U+0067 (though note that the relation between uppercase and lowercase is in many
-situations much more complex than this, and Unicode has further specifications
-for those cases). Another use of properties is to define the script of a
-character. The Unicode Standard defines the term \textsc{script} as, ``A
-collection of letters and other written signs used to represent textual
+traditions. For example, the corresponding lowercase character to \uni{0047} is
+\uni{0067} (though note that the relation between uppercase and lowercase is in
+many situations much more complex than this, and Unicode has further
+specifications for those cases). Another use of properties is to define the
+script of a character.\footnote{The Unicode Glossary defines the term \textsc{script} as
+a ``collection of letters and other written signs used to represent textual
 information in one or more writing systems. For example, Russian is written with
 a subset of the Cyrillic script; Ukrainian is written with a different subset.
-The Japanese writing system uses several scripts.'' In practice, script is
-simply defined for each character as the explicit \textsc{Script} property in
+The Japanese writing system uses several scripts.''} In practice, script is
+simply defined for each character as the explicit \textsc{script} property in
 the Unicode Character Database.
@@
 also a block \textsc{Arabic Supplement}, \textsc{Arabic Presentation Forms-A}
-and \textsc{Arabic Presentation Form B}. The situation with Latin symbols is
+and \textsc{Arabic Presentation Form-B}. The situation with Latin symbols is
 even more extra. In general, the names for block should be taken as a
 definitional statement. For example, many IPA symbols are not located in the
-aptly-names block \textsc{IPA extensions}, but in other blocks (see Section
-\ref{ipa-meets-unicode}).
+aptly-names block \textsc{IPA extensions}, but in other blocks (see Section~\ref{ipa-meets-unicode}).
+
+\subsubsection*{Grapheme clusters}
 
@@
 combinations are defined in the Unicode Standard as functioning as one
-\textsc{text element}.\footnote{The Unicode Standard defines text element as:
+\textsc{text element}.\footnote{The Unicode Glossary defines text element as:
 ``A minimum unit of text in relation to a particular text process, in the
@@
 is a base character followed by a letter modifier character. For example, the
-sequence + <̃\textgreater{} (LATIN SMALL LETTER N at U+006E followed by
-COMBINING TILDE at U+0303) combines visually into , a user-perceived character
-in writing systems like that of Spanish. So, what the user perceives as a single
-character actually involves a multi-code-point sequence. Note that this specific
-sequence can also be represented with a single \textsc{precomposed} code point,
-LATIN SMALL LETTER N WITH TILDE at U+00F1, but this is not the case for all
-multi-code point character sequences. The problem that there multiple encodings
-possible for the same text element has been acknowledged early on in the Unicode
-Standard (e.g.~for , the sequence U+006E U+0303 should in all situations be
-treated identically to the precomposed U+00F1), and a system of
+sequence <n> + < ̃> (i.e.~\textsc{latin small letter n} at \uni{006E}, followed
+by \textsc{combining tilde} at \uni{0303}) combines visually into <ñ>, a
+user-perceived character in writing systems like that of Spanish. So, what the
+user perceives as a single character actually involves a multi-code-point
+sequence. Note that this specific sequence can also be represented with a single
+so-called \textsc{precomposed} code point, \textsc{latin small letter n with
+tilde} at \uni{00F1}, but this is not the case for all multi-code point
+character sequences. The problem that there multiple encodings possible for the
+same text element has been acknowledged early on in the Unicode Standard
+(e.g.~for <ñ>, the sequence \uni{006E} \uni{0303} should in all situations be
+treated identically to the precomposed \uni{00F1}), and a system of
 \textsc{canonical equivalence} is available for such situations. Basically, the
@@
 precomposed characters (called \textsc{NFD}, \textsc{Normalization Form
-Canonical Decomposition}), or compose as much as possible combinations (called
-\textsc{NFC}, \textsc{Normalization Form Canonical Composition}). In current
+canonical Decomposition}), or compose as much as possible combinations (called
+\textsc{NFC}, \textsc{Normalization Form canonical Composition}). In current
 practice of software development, NFC seems to be preferred in most situations
@@
 clusters are locale-dependent graphemes, i.e.~such combination of characters do
-not function as text elements in all situations. For example, the sequence + for
-the Slovak digraph or the sequence in the Sisaala practical orthography
-(pronounced as IPA /tʃ/; \cite{Moran2006}). These grapheme clusters are
-\textsc{tailored} in the sense that they must be specified on a
+not function as text elements in all situations. For example, the sequence <c>~+~<h> for the Slovak digraph <ch> or the sequence <ky> in the Sisaala practical
+orthography (pronounced as IPA /tʃ/,~\cite{Moran2006}). These grapheme clusters
+are \textsc{tailored} in the sense that they must be specified on a
 language-by-language or a writing system-by-writing system basis. The Unicode
@@
 document using the \textsc{Locale Data Markup Language (LDML)}. LDML is an XML
-format and vocabulary for the exchange of structured locale data.
-
-Unicode Locale Descriptions allow users to define language- or resource-specific
-writing systems or orthographies. The Unicode Consortium defines \textsc{writing
+format and vocabulary for the exchange of structured locale data. Unicode Locale
+Descriptions allow users to define language- or even resource-specific writing
+systems or orthographies.\footnote{The Unicode Glossary defines \textsc{writing
 system} only very loosely, as it is not a central concept in the Unicode
@@
 system, the British English writing system, the French writing system, and the
-Japanese writing system.''. However, there are various drawbacks of locale
+Japanese writing system.''} However, there are various drawbacks of locale
 descriptions for the daily practice of linguistic work in a multilingual setting
-(see Section \ref{use-cases}).
+(see Section~\ref{use-cases}).

### 2015-06-05 chapters/ipa_meets_unicode.tex
major style revision

@@
-\chapter{IPA meets Unicode} \label{ipa-meets-unicode}
+\chapter{IPA meets Unicode}
+\label{ipa-meets-unicode}
 
-\section{Introducing the International Phonetic Alphabet (IPA)}
-\label{introducing-the-international-phonetic-alphabet-ipa}
+\section{The International Phonetic Alphabet}
+\label{the-international-phonetic-alphabet}
 
@@
 
-\section{Pitfall: There is no single complete Unicode code block for IPA}
-\label{pitfall-there-is-no-single-complete-unicode-code-block-for-ipa}
+\section{Pitfall: No complete IPA code block}
+\label{pitfall-no-complete-ipa-block}
 
@@
 \begin{itemize}
-	\item Basic Latin (30 characters), e.g. <a, b, c, d, e> \item Latin-1
-	Supplement (4 characters): <æ, ç, ð, ø\textgreater{} \item Latin
-	Extended-A (3 characters): <ħ, ŋ, œ> \item Latin Extended-B (5
-	characters): <ǀ, ǁ, ǂ, ǃ, > \item IPA Extensions (70 characters), e.g.:
-	<ɐ, ɑ, ɔ> \item Spacing Modifier Letters (20 characters), e.g.: <ʰ ʷ ˥>
-	\item Combining Diacritical Marks (33 characters), e.g.: <̝ ̥ ̪> \item
-	Greek and Coptic (3 characters): <β, θ, χ> \item Phonetic Extensions (2
-	characters): <, > \item Phonetic Extensions Supplement (3 characters): <
-	, , > \item Superscripts and Subscripts (1 character): <ⁿ> \item Arrows
-	(2 characters): \textless{}↑, ↓\textgreater{} \item Latin Extended-C (1
-	character): <>
+	\item Basic Latin (30 characters), e.g. <a, b, c, d, e> 
+	\item Latin-1 Supplement (4 characters): <æ, ç, ð, ø\textgreater{} 
+	\item Latin Extended-A (3 characters): <ħ, ŋ, œ> 
+	\item Latin Extended-B (5 characters): <ǀ, ǁ, ǂ, ǃ, ȵ> 
+	\item IPA Extensions (70 characters), e.g.: <ɐ, ɑ, ɔ> 
+	\item Spacing Modifier Letters (20 characters), e.g.: <ʰ ʷ ˥> 
+	\item Combining Diacritical Marks (33 characters), e.g.: <̝ ̥ ̪> 
+	\item Greek and Coptic (3 characters): <β, θ, χ> 
+	\item Phonetic Extensions (2 characters): <ᴅ, ᴴ> 
+	\item Phonetic Extensions Supplement (3 characters): <ᶑ , ᶾ, ᶣ> 
+	\item Superscripts and Subscripts (1 character): <ⁿ> 
+	\item Arrows (2 characters): \textless{}↑, ↓\textgreater{} 
+	\item Latin Extended-C (1 character): <ⱱ> 
 \end{itemize}
 
-\section{Pitfall: There are many IPA homoglyphs in Unicode}
+\section{Pitfall: Many IPA homoglyphs}
 \label{pitfall-there-are-many-ipa-homoglyphs-in-unicode}
@@
 font rendering). For example, a speaker of Russian should ideally not use the
-\textsc{cyrillic small letter a} at code point U+0430 for IPA transcriptions,
-but instead the \textsc{latin small letter a} at code point U+0061, although
-visually they are mostly indistinguishable, and the Cyrillic character is more
-easily typed on a Cyrillic keyboard. Another example we commonly encounter is
-the use of \textsc{latin small letter g} at U+0067, instead of the sanctioned
-Unicode Standard IPA character for the voiced velar stop \textsc{latin small
-letter script g} at U+0261. One begins to question whether this issue is at all
-apparent to the working linguist, or if they simply use the U+0067 because it is
-easily keyboarded and thus saves time, whereas the latter must be cumbersomely
-inserted as a special symbol in most software.\footnote{This issue was recently
-addressed by the International Phonetic Association, which has taken the stance
-that both the keyboard \textsc{latin small letter g} and the \textsc{latin small
-letter script g} are valid input characters for the voiced velar plosive.
-Unfortunately, this decision further introduces ambiguity for linguists trying
-to adhere to a strict Unicode Standard IPA encoding.}
+\textsc{cyrillic small letter a} at code point \uni{0430} for IPA
+transcriptions, but instead the \textsc{latin small letter a} at code point
+\uni{0061}, although visually they are mostly indistinguishable, and the
+Cyrillic character is more easily typed on a Cyrillic keyboard. Another example
+we commonly encounter is the use of \textsc{latin small letter g} at \uni{0067},
+instead of the sanctioned Unicode Standard IPA character for the voiced velar
+stop \textsc{latin small letter script g} at \uni{0261}. One begins to question
+whether this issue is at all apparent to the working linguist, or if they simply
+use the \uni{0067} because it is easily keyboarded and thus saves time, whereas
+the latter must be cumbersomely inserted as a special symbol in most
+software.\footnote{This issue was recently addressed by the International
+Phonetic Association, which has taken the stance that both the keyboard
+\textsc{latin small letter g} and the \textsc{latin small letter script g} are
+valid input characters for the voiced velar plosive. Unfortunately, this
+decision further introduces ambiguity for linguists trying to adhere to a strict
+Unicode Standard IPA encoding.}
 
 Furthermore, on the one hand even linguists are unlikely to distinguish between
-the <ə> \textsc{latin small letter schwa} at code point U+0259 and <ǝ>
-\textsc{latin small letter turned e} at U+01DD. On the other hand, non-linguists
-are unlikely to distinguish any semantic difference between an open back
-unrounded vowel <ɑ>, the \textsc{latin small letter alpha} at U+0251, and the
-open front unrounded vowel <a>, \textsc{latin small letter a} at U+0061. But
-even among linguists this distinction leads to problems. For example, as pointed
-out by \citet{Mielke2009}, there is a problem stemming from the fact that about
-75\% of languages are reported to have a five-vowel system
-\citep{Maddieson1984}. Historically, linguistic descriptions tend not to include
-precise audio recording and measurements of formants, so this may lead one to
-ask if the many characters that are used in phonological description reflects a
-\textsc{transcriptional bias}. The common use of in transcriptions could be in
-part due to the ease of typing the letter on an English keyboard (or for older
-descriptions, the typewriter). We found it to be exceedingly rare that a
-linguist uses for a low back unrounded vowel.\footnote{One example is
+the <ə> \textsc{latin small letter schwa} at code point \uni{0259} and <ǝ>
+\textsc{latin small letter turned e} at \uni{01DD}. On the other hand,
+non-linguists are unlikely to distinguish any semantic difference between an
+open back unrounded vowel <ɑ>, the \textsc{latin small letter alpha} at
+\uni{0251}, and the open front unrounded vowel <a>, \textsc{latin small letter
+a} at \uni{0061}. But even among linguists this distinction leads to problems.
+For example, as pointed out by \citet{Mielke2009}, there is a problem stemming
+from the fact that about 75\% of languages are reported to have a five-vowel
+system \citep{Maddieson1984}. Historically, linguistic descriptions tend not to
+include precise audio recording and measurements of formants, so this may lead
+one to ask if the many characters that are used in phonological description
+reflects a \textsc{transcriptional bias}. The common use of in transcriptions
+could be in part due to the ease of typing the letter on an English keyboard (or
+for older descriptions, the typewriter). We found it to be exceedingly rare that
+a linguist uses for a low back unrounded vowel.\footnote{One example is
 \citet[75]{Vidal2001a}, in which the author states: ``The definition of Pilagá
@@
 \begin{itemize}
-	\item The uses of the apostrophe has led to long discussions on the
-	Unicode Standard email list. An English keyboard inputs
-	\textless{}`\textgreater{} APOSTROPHE at U+0027, although the
-	\textit{preferred} Unicode apostrophe is the \textless{}'\textgreater{}
-	\textsc{right single quotation mark} at U+2019. Yet the glottal
-	stop/glottalization/ejective marker is another completely different
-	character, the \textsc{modifier letter apostrophe} at U+02BC, but
-	unfortunately looks mostly highly similar to U+2019.  \item There is
-	ambiguous encoding of IPA segments within the Unicode Standard. An
-	example is the U+02C1 \textsc{modifier letter reversed glottal stop} vs
-	the U+02E4 \textsc{modifier letter small reversed glottal stop} . Both
-	are denoted in the Unicode Standard as the \textsc{pharyngealized
-	diacritic} and both appear in various resources representing phonetic
-	data online. This is thus an example for which the Unicode Standard does
-	not solve the linguistic standardization problem.  \item There is at
-	least one case in which the character name assigned by the Unicode
-	Consortium does not match the IPA's description: in the Unicode Standard
-	at U+01C3 is labeled \textsc{latin letter retroflex click}, but in IPA
-	is an alveolar or postalveolar click (not retroflex). This naming is
-	probably best seen as a simple error in the Unicode Standard. Note that
-	most linguists simply seem to use or <ou>) while ligatures are single
-	characters (e.g. <ʧ> \textsc{latin small letter tesh digraph} at
-	U+02A7). Ligatures arose in the context of printing easier-to-read
-	texts, and are included in the Unicode Standard for reasons of legacy
-	encoding. However, their usage is discouraged by the Unicode core
-	specification. Specifically related to IPA, various phonetic
-	combinations of characters (typically affricates) are available as
-	single code-points in the Unicode Standard, but are designated as
-	\textsc{ligatures} or \textsc{digraphs} (confusingly both names appear
-	interchangeably). Such glyphs might be used by software to produce a
-	pleasing display, but they should not be hard-coded into the text
-	itself. In the context of IPA, characters like the following ligatures
-	should thus \emph{not} be used. Instead a combination of two characters
-	is preferred: \item <ʣ> \textsc{latin small letter dz digraph} at U+02A3
-	(use <d> + <z> instead) \item <ʧ> \textsc{latin small letter tesh
-	digraph} at U+02A7 (use <t> + <ʃ> instead) \item <ʩ> \textsc{latin small
-	letter feng digraph} at U+02A9 (use <f> + <ŋ> instead)
+	\item The uses of the apostrophe has led to long discussions on the Unicode Standard email list.
+An English keyboard inputs \textless{}`\textgreater{} APOSTROPHE at \uni{0027}, although the \textit{preferred} Unicode apostrophe is the \textless{}'\textgreater{} \textsc{right single quotation mark} at \uni{2019}.
+Yet the glottal stop/glottalization/ejective marker is another completely different character, the \textsc{modifier letter apostrophe} at \uni{02BC}, but unfortunately looks mostly highly similar to \uni{2019}.
+	\item There is ambiguous encoding of IPA segments within the Unicode Standard.
+An example is the \uni{02C1} \textsc{modifier letter reversed glottal stop} vs the \uni{02E4} \textsc{modifier letter small reversed glottal stop} .
+Both are denoted in the Unicode Standard as the \textsc{pharyngealized diacritic} and both appear in various resources representing phonetic data online.
+This is thus an example for which the Unicode Standard does not solve the linguistic standardization problem.
+	\item There is at least one case in which the character name assigned by the Unicode Consortium does not match the IPA's description: in the Unicode Standard at \uni{01C3} is labeled \textsc{latin letter retroflex click}, but in IPA is an alveolar or postalveolar click (not retroflex).
+This naming is probably best seen as a simple error in the Unicode Standard.
+Note that most linguists simply seem to use or <ou>) while ligatures are single characters (e.g.
+<ʧ> \textsc{latin small letter tesh digraph} at \uni{02A7}).
+Ligatures arose in the context of printing easier-to-read texts, and are included in the Unicode Standard for reasons of legacy encoding.
+However, their usage is discouraged by the Unicode core specification.
+Specifically related to IPA, various phonetic combinations of characters (typically affricates) are available as single code-points in the Unicode Standard, but are designated as \textsc{ligatures} or \textsc{digraphs} (confusingly both names appear interchangeably).
+Such glyphs might be used by software to produce a pleasing display, but they should not be hard-coded into the text itself.
+In the context of IPA, characters like the following ligatures should thus \emph{not} be used.
+Instead a combination of two characters is preferred:
+	\item <ʣ> \textsc{latin small letter dz digraph} at \uni{02A3} (use <d> + <z> instead) 
+	\item <ʧ> \textsc{latin small letter tesh digraph} at \uni{02A7} (use <t> + <ʃ> instead) 
+	\item <ʩ> \textsc{latin small letter feng digraph} at \uni{02A9} (use <f> + <ŋ> instead) 
 \end{itemize}
@@
 \begin{itemize}
-	\item <ɮ> \textsc{latin small letter lezh} at U+026E \item <œ>
-	\textsc{latin small ligature oe} at U+0153 \item <ɶ> \textsc{latin
-	letter small capital oe} at U+0276 \item <æ> \textsc{latin small letter
-	ae} at U+00E6
+	\item <ɮ> \textsc{latin small letter lezh} at \uni{026E} 
+	\item <œ> \textsc{latin small ligature oe} at \uni{0153} 
+	\item <ɶ> \textsc{latin letter small capital oe} at \uni{0276} 
+	\item <æ> \textsc{latin small letter ae} at \uni{00E6} 
 \end{itemize}
 
-\section{Pitfall: The notion of diacritic differs between IPA and Unicode}
+\section{Pitfall: No decomposition of strikethrough}
+\label{pitfall-no-decomposition-of=strikethrough}
+
+\section{Pitfall: Different notions of diacritics}
 \label{pitfall-the-ipa-notion-of-diacritics-is-not-the-same-as-the-unicode-standards-notion-of-diacritics}
@@
 \begin{itemize}
-	\item Length marks, namely <ː> \textsc{modifier letter triangular colon}
-	at U+02D0 and <ˑ> \textsc{modifier letter half triangular colon} at
-	U+02D1.  \item Tone letters, like <˥> \textsc{modifier letter extra-high
-	tone bar} at U+02E5, and others like this.  \item Superscript letters,
-	like <ʰ> \textsc{modifier letter small h} at U+02B0 or <ˤ>
-	\textsc{modifier letter small reversed glottal stop} at U+02E4, and
-	others like this.  \item <˞> \textsc{modifier letter rhotic hook} at
-	U+02DE.
+	\item Length marks, namely <ː> \textsc{modifier letter triangular colon} at \uni{02D0} and <ˑ> \textsc{modifier letter half triangular colon} at \uni{02D1}. 
+	\item Tone letters, like <˥> \textsc{modifier letter extra-high tone bar} at \uni{02E5}, and others like this. 
+	\item Superscript letters, like <ʰ> \textsc{modifier letter small h} at \uni{02B0} or <ˤ> \textsc{modifier letter small reversed glottal stop} at \uni{02E4}, and others like this. 
+	\item <˞> \textsc{modifier letter rhotic hook} at \uni{02DE}. 
 \end{itemize}
@@
 character in front of them, at least for <ʰ> \textsc{modifier letter small h} at
-U+02B0 the Unicode Consortium's decision to treat it as a separate character is
-also linguistically correct, because according to the IPA it can be used both
+\uni{02B0} the Unicode Consortium's decision to treat it as a separate character
+is also linguistically correct, because according to the IPA it can be used both
 for aspiration (more precisely post-aspiration following the base character) and
@@
 Unicode to force separate characters to be combined (namely by using the
-\textsc{zero width joiner} at U+200D), but this seems to be a rather
+\textsc{zero width joiner} at \uni{200D}), but this seems to be a rather
 impractical, and probably not enforceable solution to us.
 
-\section{Pitfall: There is no unique diacritic ordering in IPA, nor in Unicode}
-\label{pitfall-neither-the-ipa-nor-the-unicode-standard-enforce-a-unique-diacritic-ordering}
+\section{Pitfall: No unique diacritic ordering}
+\label{pitfall-no-unique-diacritic-ordering}
 
@@
 Decomposition implies that precomposed characters (like <á> \textsc{latin small
-letter a with acute} at U+00E1) will be split into its parts. This might sound
-preferable for a linguistic analysis, as the different diacritics are separated
-from the base characters. However, note that most attached elements like strokes
-(e.g.~in the <ɨ>), retroflex hooks (e.g.~in <ʐ>) or rhotic hooks (e.g.~in <ɝ>)
-will not be decomposed, but strangely enough a cedilla (like in <ç>) will be
-decomposed. In general, Unicode decomposition does not behave like a feature
-decomposition as expected from a linguistic perspective. It is thus important to
-consider Unicode decomposition only as a technical procedure, and not assume
-that it is linguistically sensible.
+letter a with acute} at \uni{00E1}) will be split into its parts. This might
+sound preferable for a linguistic analysis, as the different diacritics are
+separated from the base characters. However, note that most attached elements
+like strokes (e.g.~in the <ɨ>), retroflex hooks (e.g.~in <ʐ>) or rhotic hooks
+(e.g.~in <ɝ>) will not be decomposed, but strangely enough a cedilla (like in
+<ç>) will be decomposed. In general, Unicode decomposition does not behave like
+a feature decomposition as expected from a linguistic perspective. It is thus
+important to consider Unicode decomposition only as a technical procedure, and
+not assume that it is linguistically sensible.
 
@@
 \begin{itemize}
-	\itemsep1pt\parskip0pt\parsep0pt \item \textsc{Spacing Modifier Letters
-	ordering:} ( unreleased <̚> \textbar{} lateral release <ˡ> \textbar{}
-	nasal release <ⁿ>) → ( palatalized <ʲ>) → ( labialized <ʷ>) → (
-	velarized <ˠ>) → ( pharyngealized <ˤ>) → ( aspirated <ʰ> \textbar{}
-	ejective <ʼ>) → ( long <ː> \textbar{} half-long <ˑ>)
+	\itemsep1pt\parskip0pt\parsep0pt 
+	\item \textsc{Spacing Modifier Letters ordering:} ( unreleased <̚> \textbar{} lateral release <ˡ> \textbar{} nasal release <ⁿ>) → ( palatalized <ʲ>) → ( labialized <ʷ>) → ( velarized <ˠ>) → ( pharyngealized <ˤ>) → ( aspirated <ʰ> \textbar{} ejective <ʼ>) → ( long <ː> \textbar{} half-long <ˑ>)
 \end{itemize}
@@
 \begin{itemize}
-	\itemsep1pt\parskip0pt\parsep0pt \item \textsc{Combining Diacritical
-	Marks (below) ordering:} ( dental <t̪> \textbar{} laminal <t̻>
-	\textbar{} apical <t̺>) → ( fronted <u̟> \textbar{} backed <e̠>) →(
-	lowered <e̞> \textbar{} raised <e̝>) → ( ATR <e̘ e̙>) → ( voiced <s̬>
-	\textbar{} voiceless <n̥> \textbar{} creaky voice <b̰> \textbar{}
-	breathy voice <b̤>) → ( syllabic <n̩> \textbar{} non-syllabic <e̯>)
+	\itemsep1pt\parskip0pt\parsep0pt 
+	\item \textsc{Combining Diacritical Marks (below) ordering:} ( dental <t̪> \textbar{} laminal <t̻> \textbar{} apical <t̺>) → ( fronted <u̟> \textbar{} backed <e̠>) →( lowered <e̞> \textbar{} raised <e̝>) → ( ATR <e̘ e̙>) → ( voiced <s̬> \textbar{} voiceless <n̥> \textbar{} creaky voice <b̰> \textbar{} breathy voice <b̤>) → ( syllabic <n̩> \textbar{} non-syllabic <e̯>)
 \end{itemize}
@@
 \begin{itemize}
-	\itemsep1pt\parskip0pt\parsep0pt \item \textsc{Combining Diacritical
-	Marks (above) ordering:} (centralized <ë> \textbar{} mid-centralized
-	<e̽>) → (extra short <ĕ>) →( tone accents, e.g. <è> ) → ( Spacing
-	Modifier Letters ) → ( tone letters, e.g. <e˦>)
+	\itemsep1pt\parskip0pt\parsep0pt 
+	\item \textsc{Combining Diacritical Marks (above) ordering:} (centralized <ë> \textbar{} mid-centralized <e̽>) → (extra short <ĕ>) →( tone accents, e.g.
+<è> ) → ( Spacing Modifier Letters ) → ( tone letters, e.g.
+<e˦>)
 \end{itemize}

### 2015-06-05 chapters/orthography_profiles.tex
major style revision

--- 2015-05-31 chapters/orthography_profiles.tex
@@
 
-At this point in the course of rapid ongoing developments, we are left with a situation in which the Unicode Standard offers a highly detailed and flexible approach to deal computationally with writing systems, but it has unfortunately not influenced the linguistic practice very much.
-In many practical situations, the Unicode Standard is far too complex for the day-to-day practice in linguistics because it does not offer practical solutions for the down-to-earth problems of many linguists.
-In this section, we propose some simple practical guidelines and methods to improve on this situation.
+At this point in the course of rapid ongoing developments, we are left with a
+situation in which the Unicode Standard offers a highly detailed and flexible
+approach to deal computationally with writing systems, but it has unfortunately
+not influenced the linguistic practice very much. In many practical situations,
+the Unicode Standard is far too complex for the day-to-day practice in
+linguistics because it does not offer practical solutions for the down-to-earth
+problems of many linguists. In this section, we propose some simple practical
+guidelines and methods to improve on this situation.
 
-Our central aims for linguistics, to be approached with a Unicode-based solution, are: (i) to improve the consistency of the encoding of sources, (ii) to transparently document knowledge about the writing system (including transliteration), and (iii) to do all of that in a way that is easy and quick to manage for many different sources with many different writing systems.
-The central concept in our proposal is the \textsc{orthography profile}, a simple tab-separated CSV text file, that characterizes and documents a writing system.
-We also offer basic implementations in Python and R to assist with the production of such files, and to apply orthography profiles for consistency testing, grapheme tokenization and transliteration.
-Not only can orthography profiles be helpful in the daily practice of linguistics, they also succinctly document the orthographic details of a specific source, and, as such, might fruitfully be published alongside sources (e.g.~in digital archives).
-Also, in high-level linguistic analyzes in which the graphemic detail is of central importance (e.g.~phonotactic or comparative-historical studies), orthography profiles can transparently document the decisions that have been taken in the interpretation of the orthography in the sources used.
+Our central aims for linguistics, to be approached with a Unicode-based
+solution, are: (i) to improve the consistency of the encoding of sources, (ii)
+to transparently document knowledge about the writing system (including
+transliteration), and (iii) to do all of that in a way that is easy and quick to
+manage for many different sources with many different writing systems. The
+central concept in our proposal is the \textsc{orthography profile}, a simple
+tab-separated CSV text file, that characterizes and documents a writing system.
+We also offer basic implementations in Python and R to assist with the
+production of such files, and to apply orthography profiles for consistency
+testing, grapheme tokenization and transliteration. Not only can orthography
+profiles be helpful in the daily practice of linguistics, they also succinctly
+document the orthographic details of a specific source, and, as such, might
+fruitfully be published alongside sources (e.g.~in digital archives). Also, in
+high-level linguistic analyzes in which the graphemic detail is of central
+importance (e.g.~phonotactic or comparative-historical studies), orthography
+profiles can transparently document the decisions that have been taken in the
+interpretation of the orthography in the sources used.
 
-Given these goals, Unicode locale descriptions (see Section \ref{terminology}) might seem like the ideal orthography profiles.
-However, there are various practical obstacles preventing the use of such locale descriptions in the daily linguistic practice, namely: (i) the XML-structure is too verbose to easily and quickly produce or correct manually, (ii) locale descriptions are designed for a wide scope on information (like date formats or names of weekdays) most of which is not applicable for documenting writing systems, and (iii) most crucially, even if someone made the effort to produce a technically correct locale description for a specific source at hand, then it is nigh impossible to deploy the description.
-This is because a locale description has to be submitted to and accepted by the Unicode Common Locale Data Repository.
-The repository is (rightly so) not interested in descriptions that only apply to a limited set of sources (e.g.~only a single specific dictionary).
+Given these goals, Unicode locale descriptions (see Section~\ref{terminology})
+might seem like the ideal orthography profiles. However, there are various
+practical obstacles preventing the use of such locale descriptions in the daily
+linguistic practice, namely: (i) the XML-structure is too verbose to easily and
+quickly produce or correct manually, (ii) locale descriptions are designed for a
+wide scope on information (like date formats or names of weekdays) most of which
+is not applicable for documenting writing systems, and (iii) most crucially,
+even if someone made the effort to produce a technically correct locale
+description for a specific source at hand, then it is nigh impossible to deploy
+the description. This is because a locale description has to be submitted to and
+accepted by the Unicode Common Locale Data Repository. The repository is
+(rightly so) not interested in descriptions that only apply to a limited set of
+sources (e.g.~only a single specific dictionary).
 
-The major challenge then is developing an infrastructure to identify the elements that are individual graphemes in a source, specifically for the enormous variety of sources using some kind of alphabetic writing system.
-Authors of source documents (e.g.~dictionaries, wordlists, corpora) use a variety of writing systems that range from their own idiosyncratic transcriptions to already well-established practical or longstanding orthographies.
-Although the IPA is one practical choice as a sound-based normalization for writing systems (which can act as an interlingual pivot to attain interoperability across writing systems), graphemes in each writing system must also be identified and standardized if interoperability across different sources is to be achieved.
-In most cases, this amounts to more than simply mapping a grapheme to an IPA segment because graphemes must first be identified in context (e.g.~is the sequence one sound or two sounds or both?) and strings must be tokenized, which may include taking orthographic rules into account (e.g.
-between vowels is /n/ and after a vowel but before a consonant is a nasalized vowel /ṽ/).
-In our experience, data from each source must be individually tokenized into graphemes so that its orthographic structure is identified and its contents can be extracted.
-To extract data for analysis, a source-by-source approach is required before an orthography profile can be created.
-For example, almost each available lexicon on the world's languages is idiosyncratic in its orthography and thus requires lexicon-specific approaches to identify graphemes in the writing system and to map graphemes to phonemes, if desired.
+The major challenge then is developing an infrastructure to identify the
+elements that are individual graphemes in a source, specifically for the
+enormous variety of sources using some kind of alphabetic writing system.
+Authors of source documents (e.g.~dictionaries, wordlists, corpora) use a
+variety of writing systems that range from their own idiosyncratic
+transcriptions to already well-established practical or longstanding
+orthographies. Although the IPA is one practical choice as a sound-based
+normalization for writing systems (which can act as an interlingual pivot to
+attain interoperability across writing systems), graphemes in each writing
+system must also be identified and standardized if interoperability across
+different sources is to be achieved. In most cases, this amounts to more than
+simply mapping a grapheme to an IPA segment because graphemes must first be
+identified in context (e.g.~is the sequence one sound or two sounds or both?)
+and strings must be tokenized, which may include taking orthographic rules into
+account (e.g.~between vowels is /n/ and after a vowel but before a consonant is
+a nasalized vowel /ṽ/). In our experience, data from each source must be
+individually tokenized into graphemes so that its orthographic structure is
+identified and its contents can be extracted. To extract data for analysis, a
+source-by-source approach is required before an orthography profile can be
+created. For example, almost each available lexicon on the world's languages is
+idiosyncratic in its orthography and thus requires lexicon-specific approaches
+to identify graphemes in the writing system and to map graphemes to phonemes, if
+desired.
 
-Thus, our key proposal for the characterization of a writing system is to use a grapheme tokenization as an inter-orthographic pivot.
-Basically, any source document is tokenized by graphemes, and only then a mapping to IPA (or any other orthographic conversion) is performed.
-An orthography profile then is a description of the units and rules that are needed to adequately model a graphemic tokenization for a language variety as described in a particular source document.
-An orthography profile summarizes the Unicode (tailored) graphemes and orthographic rules used to write a language (the details of the structure and assumptions of such a profile will be presented in the next section).
+Thus, our key proposal for the characterization of a writing system is to use a
+grapheme tokenization as an inter-orthographic pivot. Basically, any source
+document is tokenized by graphemes, and only then a mapping to IPA (or any other
+orthographic conversion) is performed. An orthography profile then is a
+description of the units and rules that are needed to adequately model a
+graphemic tokenization for a language variety as described in a particular
+source document. An orthography profile summarizes the Unicode (tailored)
+graphemes and orthographic rules used to write a language (the details of the
+structure and assumptions of such a profile will be presented in the next
+section).
 
-As an example of graphemic tokenization, note the three different levels of technological and linguistic elements that interact in the hypothetical lexical form <tsʰṍ̰shi> :
+As an example of graphemic tokenization, note the three different levels of
+technological and linguistic elements that interact in the hypothetical lexical
+form <tsʰṍ̰shi>:
 
@@
 
-In (1), the string <tsʰṍ̰shi> has been tokenized into ten Unicode code points (using NFD normalization), delimited here by space.
-Unicode tokenization is required because sequences of code points can differ in their visual and logical orders.
-For example, <õ̰> is ambiguous to whether it is the sequence of + <̰> + <̃> or + <̃> + <̰>.
-Although these two variants are visually homoglyphs, computationally they are different.
-Unicode normalization should be applied to this string to reorder the code points into a canonical order, allowing the data to be treated canonically equivalently for search and comparison.
-In (2), the Unicode code points have been logically normalized and visually organized into grapheme clusters, as specified by the Unicode Standard.
-The combining character sequence <õ̰> is normalized and visually grouped together.
-Note that , the MODIFIER LETTER SMALL H at U+02B0, is not grouped with .
-This is because it belongs to Spacing Modifier Letters category in the Unicode Standard.
-These characters are underspecified for the direction in which they modify a host character.
-For example, can indicate either pre- or post-aspiration (whereas the nasalization or creaky diacritic is defined in the Unicode Standard to apply to a specified base character).
-Finally, to arrive at the graphemic tokenization in (3), tailored grapheme clusters are needed (as for example specified in an orthography profile).
-For example, this orthography profile would specify that the sequence of characters , and form a single grapheme , and that and form a grapheme .
-The orthography profile could also specify orthographic rules, e.g.~when tokenization graphemes, in say English, the in the forms and should be treated as distinct sequences depending on their contexts.
+In (1), the string <tsʰṍ̰shi> has been tokenized into ten Unicode code points
+(using NFD normalization), delimited here by space. Unicode tokenization is
+required because sequences of code points can differ in their visual and logical
+orders. For example, <õ̰> is ambiguous to whether it is the sequence of + <̰> +
+<̃> or + <̃> + <̰>. Although these two variants are visually homoglyphs,
+computationally they are different. Unicode normalization should be applied to
+this string to reorder the code points into a canonical order, allowing the data
+to be treated canonically equivalently for search and comparison. In (2), the
+Unicode code points have been logically normalized and visually organized into
+grapheme clusters, as specified by the Unicode Standard. The combining character
+sequence <õ̰> is normalized and visually grouped together. Note that, the
+MODIFIER LETTER SMALL H at \uni{02B0}, is not grouped with. This is because it
+belongs to Spacing Modifier Letters category in the Unicode Standard. These
+characters are underspecified for the direction in which they modify a host
+character. For example, can indicate either pre- or post-aspiration (whereas the
+nasalization or creaky diacritic is defined in the Unicode Standard to apply to
+a specified base character). Finally, to arrive at the graphemic tokenization in
+(3), tailored grapheme clusters are needed (as for example specified in an
+orthography profile). For example, this orthography profile would specify that
+the sequence of characters, and form a single grapheme, and that and form a
+grapheme. The orthography profile could also specify orthographic rules,
+e.g.~when tokenization graphemes, in say English, the in the forms and should be
+treated as distinct sequences depending on their contexts.
 
-\section{Informal description of orthography profiles}
+\section{Informal description}
 \label{informal-description-of-orthography-profiles}
 
-An orthography profile describes the Unicode code points, characters, graphemes and orthographic rules in a writing system.
-An orthography profile is a language-specific (and often even resource-specific) description of the units and rules that are needed to adequately model a writing system.
-An important assumption of our work is that we assume a resource is encoded in Unicode (or has been converted to Unicode).
-Any data source that the Unicode Standard is unable to capture, will also not be captured by an orthography profile.
+An orthography profile describes the Unicode code points, characters, graphemes
+and orthographic rules in a writing system. An orthography profile is a
+language-specific (and often even resource-specific) description of the units
+and rules that are needed to adequately model a writing system. An important
+assumption of our work is that we assume a resource is encoded in Unicode (or
+has been converted to Unicode). Any data source that the Unicode Standard is
+unable to capture, will also not be captured by an orthography profile.
 
-Informally, an orthography profile specifies the graphemes (or, in Unicode parlance, \textsc{tailored grapheme clusters}) that are expected to occur in any data to be analyzed or checked for consistency.
-These graphemes are first identified throughout the whole data (a step which we call \textsc{tokenization}), and possibly simply returned as such, possibly including error messages about any parts of the data that are not specified by the orthography profile.
-Once the graphemes are identified, they might also be changed into other graphemes (a step which we call \textsc{transliteration}).
-When a grapheme has different possible transliterations, then these differences should be separated by contextual specification, possibly down to listing individual exceptional cases.
+Informally, an orthography profile specifies the graphemes (or, in Unicode
+parlance, \textsc{tailored grapheme clusters}) that are expected to occur in any
+data to be analyzed or checked for consistency. These graphemes are first
+identified throughout the whole data (a step which we call
+\textsc{tokenization}), and possibly simply returned as such, possibly including
+error messages about any parts of the data that are not specified by the
+orthography profile. Once the graphemes are identified, they might also be
+changed into other graphemes (a step which we call \textsc{transliteration}).
+When a grapheme has different possible transliterations, then these differences
+should be separated by contextual specification, possibly down to listing
+individual exceptional cases.
 
-In practice, we foresee a workflow in which orthography profiles are iteratively refined, while at the same time inconsistencies and errors in the data to be tokenized are corrected.
-In some more complex use-cases there might even be a need for multiple different orthography profiles to be applied in sequence (see Section \ref{use-cases} on various exemplary use-cases).
-The result of any such workflow will normally be a cleaned dataset and an explicit description of the orthographic structure in the form of an orthography profile.
-Subsequently, the orthography profiles can be easily distributed in scholarly channels alongside the cleaned data, for example in supplementary material added to journal papers or in electronic archives.
+In practice, we foresee a workflow in which orthography profiles are iteratively
+refined, while at the same time inconsistencies and errors in the data to be
+tokenized are corrected. In some more complex use-cases there might even be a
+need for multiple different orthography profiles to be applied in sequence (see
+Section~\ref{use-cases} on various exemplary use-cases). The result of any such
+workflow will normally be a cleaned dataset and an explicit description of the
+orthographic structure in the form of an orthography profile. Subsequently, the
+orthography profiles can be easily distributed in scholarly channels alongside
+the cleaned data, for example in supplementary material added to journal papers
+or in electronic archives.
 
-\section{Formal specification of orthography profiles}
+\section{Formal specification}
 \label{formal-specification-of-orthography-profiles}
 
-The formal specifications of an orthography profile (or simply \textsc{profile} for short) are the following:
+The formal specifications of an orthography profile (or simply \textsc{profile}
+for short) are the following:
 
@@
 	\def\labelenumi{\arabic{enumi}.} 
-	\item \textsc{A profile is a} \textsc{Unicode UTF-8 encoded text file} (ideally using NFC, no-BOM, and LF; see Section \ref{pitfall-file-formats}, Pitfall: File Formats) that includes the information pertinent to the orthography.
+	\item \textsc{A profile is a} \textsc{Unicode UTF-8 encoded text file} (ideally using NFC, no-BOM, and LF; see Section~\ref{pitfall-file-formats}, Pitfall: File Formats) that includes the information pertinent to the orthography.
 	\item \textsc{A profile is a} \textsc{tab-separated CSV file with an obligatory header line}.
@@
 The actual ordering of the columns is unimportant.
-	\item \textsc{Lines starting with a hash \textless{}\#\textgreater{} are ignored.} Comments and metadata can be included inside the file, but only as complete lines in the profile, to be marked by lines starting with hash \textsc{\#} (\textsc{number sign} at U+0023).
+	\item \textsc{Lines starting with a hash \textless{}\#\textgreater{} are ignored.} Comments and metadata can be included inside the file, but only as complete lines in the profile, to be marked by lines starting with hash \textsc{\#} (\textsc{number sign} at \uni{0023}).
 Hashes somewhere else in the file are to be treated literally, i.e.~hashes are only to be ignored when they occur at the start of a line.\footnote{Comments that belong to specific lines will have to be put in a separate column of the CSV file, e.g.~add a column called \textsc{comments}.
@@
 
-The content of a profile consists of lines, each describing a grapheme of the orthography, using the following columns:
+The content of a profile consists of lines, each describing a grapheme of the
+orthography, using the following columns:
 
@@
 	\item \textsc{A minimal profile consists of a single column with header \textsc{graphemes}}, listing each of the different graphemes in a separate line. 
-	\item \textsc{Optional columns called `left' and `right' can be used to specify the left and right context of the grapheme, respectively.} The same grapheme can occur multiple times with different contextual specifications, for example to distinguish different pronunciations depending on the context.
-	\item \textsc{The columns `grapheme', `left' and `right' can use regular expression metacharacters.} If regular expressions are used, then all literal usage of the special symbols, like full stops <.> or dollar signs <\$> (so-called \textsc{metacharacters}) have to be explicitly escaped by adding a backslash before them (i.e.~use <.> or <\$>).
+	\item \textsc{Optional columns called \textsc{left} and \textsc{right} can be used to specify the left and right context of the grapheme, respectively.} The same grapheme can occur multiple times with different contextual specifications, for example to distinguish different pronunciations depending on the context.
+	\item \textsc{The columns \textsc{grapheme}, \textsc{left} and \textsc{right} can use regular expression metacharacters.} If regular expressions are used, then all literal usage of the special symbols, like full stops <.> or dollar signs <\$> (so-called \textsc{metacharacters}) have to be explicitly escaped by adding a backslash before them (i.e.~use <.> or <\$>).
 Note that any specification of context automatically expects regular expressions, so it is probably better to always escape all regular expression metacharacters when used literally in the orthography, i.e.~the following symbols will need to be preceded by a backslash: {[} {]} ( ) \{ \} ~+ * .
@@
 \^{} \$ .
-	\item \textsc{An optional column called `class' can be used to specify classes of graphemes}, for example to define a class of vowels.
+	\item \textsc{An optional column called \textsc{class} can be used to specify classes of graphemes}, for example to define a class of vowels.
 Users can simply add ad-hoc identifiers in this column to indicate a group of graphemes, which can then be used in the description of the graphemes or the context.
@@
 
-For the automatic processing of the profiles, the following technical standards will be expected:
+For the automatic processing of the profiles, the following technical standards
+will be expected:
 
@@
 However, this only is possible when no contexts or classes are described, so this seems only useful in the most basic orthographies.
-	\item \textsc{The `class' column will be used to produce explicit \textsc{or} chains of regular expressions}, which will then be inserted in the `graphemes', `left' and `right' columns at the position indicated by the class-identifiers.
+	\item \textsc{The \textsc{class} column will be used to produce explicit \textsc{or} chains of regular expressions}, which will then be inserted in the \textsc{graphemes}, \textsc{left} and \textsc{right} columns at the position indicated by the class-identifiers.
 For example, a class \textsc{V} as a context specification might be replaced by a regular expression like: (a\textbar{}e\textbar{}i\textbar{}o\textbar{}u\textbar{}ei\textbar{}au).
 Only the graphemes themselves are included here, not any contexts specified for the elements of the class.
-	\item \textsc{The `left' and `right' contexts will be included into the regular expressions by using lookbehind and lookahead}.
+	\item \textsc{The \textsc{left} and \textsc{right} contexts will be included into the regular expressions by using lookbehind and lookahead}.
 Basically, the actual regular expression syntax of lookbehind and lookahead is simply hidden to the users by allowing them to only specify the contexts themselves.
@@
 
-The actual implementation of the profile on some text-string will function as follows:
+The actual implementation of the profile on some text-string will function as
+follows:
 
@@
 		\item \textsc{the original text-strings to be processed in the used Unicode normalization}, i.e.~in either NFC or NFD as specified by the user. 
-		\item \textsc{the tokenized strings}, with additionally any transliterated strings, if transliteration is requested. 
-		\item \textsc{a survey of all errors encountered}, ideally both in which text-strings any errors occurred and which characters in the text-strings lead to errors. 
+		\item \textsc{the tokenized strings}, with additionally any transliterated
+        strings, if transliteration is requested. 
+		\item \textsc{a survey of all errors encountered}, ideally both in which
+        text-strings any errors occurred and which characters in the
+        text-strings lead to errors. 
 		\item \textsc{a reordered profile}, when any automatic reordering is offered 

### 2015-06-05 chapters/pitfalls.tex
major style revision

@@
-\chapter{Unicode Pitfalls}
+\chapter{Unicode pitfalls}
 \label{unicode-pitfalls}
 
+% ==========================
 \section{Wrong it is not}
 \label{wrong-it-is-not}
+% ==========================
 
-In this section, we describe some of the most common pitfalls that we have encountered when using the Unicode Standard in our own work, or in discussion with other linguists.
-This section is not meant as a criticism of the decisions made by the Unicode Consortium; on the contrary, we aim to highlight where the technological aspects of the Unicode Standard diverge from many users' intuitions.
-What have sometimes been referred to as problems or inconsistencies in the Unicode Standard are mostly due to legacy compatibility issues, which can lead to unexpected behavior by linguists using the standard.
-However, there are also some cases in which the Unicode Standard has made decisions that theoretically could have been taken differently, but for some reason or another (mostly very good reasons) were accepted as they are now.
-We call behavior that executes without error but does something different than the user expected---often unknowingly---a ``pitfall''.
+In this chapter we describe some of the most common pitfalls that we have
+encountered when using the Unicode Standard in our own work, or in discussion
+with other linguists. This section is not meant as a criticism of the decisions
+made by the Unicode Consortium; on the contrary, we aim to highlight where the
+technological aspects of the Unicode Standard diverge from many users'
+intuitions. What have sometimes been referred to as problems or inconsistencies
+in the Unicode Standard are mostly due to legacy compatibility issues, which can
+lead to unexpected behavior by linguists using the standard. However, there are
+also some cases in which the Unicode Standard has made decisions that
+theoretically could have been taken differently, but for some reason or another
+(mostly very good reasons) were accepted as they are now. We call behavior that
+executes without error but does something different than the user
+expected---often unknowingly---a \textsc{pitfall}.
 
-In this context, it is important to realize that the Unicode Standard was not developed to solve linguistic problems per se, but to offer a consistent computational environment for written language.
-In those cases in which the Unicode Standard behaves differently as expected, we think it is important not to dismiss Unicode as `wrong' or `deficient', because our experience is that in almost all cases the behaviour of the Unicode Standard has been particularly well thought through.
-The Unicode Consortium considers important practical use-cases that from a linguistic point of view are normally not considered.
-Our general guideline for dealing with the Unicode Standard is to accept it as it is, and not to battle windmills.
-Alternatively, of course, it is possible to actively engage in the development of the standard itself, an effort that is highly appreciated by the Unicode Consortium.
+In this context, it is important to realize that the Unicode Standard was not
+developed to solve linguistic problems per se, but to offer a consistent
+computational environment for written language. In those cases in which the
+Unicode Standard behaves differently as expected, we think it is important not
+to dismiss Unicode as ``wrong'' or ``deficient'', because our
+experience is that in almost all cases the behavior of the Unicode Standard has
+been particularly well thought through. The Unicode Consortium has a more
+wide-ranging view of matters and often examines important practical use-cases
+that from a linguistic point of view are normally not considered. Our general
+guideline for dealing with the Unicode Standard is to accept it as it is, and
+not to battle windmills. Alternatively, of course, it is possible to actively
+engage in the development of the standard itself, an effort that is highly
+appreciated by the Unicode Consortium.
 
+% ==========================
 \section{Pitfall: Blocks}
 \label{pitfall-blocks}
+% ==========================
 
-The Unicode code space is subdivided into blocks of contiguous code points.
-For example, the block called ``Cyrillic'' runs from U+0400 till U+04FF.
-These blocks arose as an attempt at ordering the enormous amount of characters in Unicode, but the ideas of blocks very quickly ran into problems.
-First, the size of a block is fixed, so when a block is full, a new block will have to be instantiated somewhere further in the code space.
-For example, this led to the blocks ``Cyrillic Supplement'', ``Cyrillic Extended-A'' (which are also already full) and ``Cyrillic Extended-B''.
-Second, when a specific character already exists, then it is not duplicated in another block, although the name of the block might indicate that a specific symbol should be available there.
-In general, names of blocks are just an approximate indication of the kind of characters that will be in the block.
+The Unicode code space is subdivided into blocks of contiguous code points. For
+example, the block called \textsc{Cyrillic} runs from \uni{0400} till
+\uni{04FF}. These blocks arose as an attempt at ordering the enormous amount of
+characters in Unicode, but the ideas of blocks very quickly ran into problems.
+First, the size of a block is fixed, so when a block is full, a new block will
+have to be instantiated somewhere further in the code space. For example, this
+led to the blocks \textsc{Cyrillic Supplement}, \textsc{Cyrillic Extended-A}
+(both of which are also already full) and \textsc{Cyrillic Extended-B}. Second,
+when a specific character already exists, then it is not duplicated in another
+block, although the name of the block might indicate that a specific symbol
+should be available there. In general, names of blocks are just an approximate
+indication of the kind of characters that will be in the block.
 
-The problem with blocks arises because finding the right character among the thousands of Unicode characters is not easy.
-Many software applications present blocks as a primary search mechanism, because the block names suggest where to look for a particular character.
-However, when a user searches for an IPA character in the block ``IPA Extensions'' then many IPA characters will not be found there.
-For example, the velar nasal <ŋ> is not part of the block ``IPA Extensions'' because it was already included as LATIN SMALL LETTER ENG at U+014B in the block ``Latin Extensions-A''.
+The problem with blocks arises because finding the right character among the
+thousands of Unicode characters is not easy. Many software applications present
+blocks as a primary search mechanism, because the block names suggest where to
+look for a particular character. However, when a user searches for an IPA
+character in the block \textsc{IPA Extensions}, then many IPA characters will not
+be found there. For example, the velar nasal <ŋ> is not part of the block
+\textsc{IPA Extensions} because it was already included as \textsc{latin small letter
+eng} at \uni{014B} in the block \textsc{Latin Extensions-A}.
 
-In general, finding a specific character in the Unicode Standard is often not trivial.
-The names of the blocks can help, but it is not (and never was supposed to be) a foolproof structure.
-It is not the goal nor aim of the Unicode Consortium to provide a user interface to the Unicode Standard.
-If one often encounters the problem of needing to find a suitable character, there are various other useful services for end users available.\footnote{The Unicode website offers a basic interface to the code charts at http://www.unicode.org/charts/index.html.
-As a more flexible interface, we particularly like PopChar from Macility, available for both Macintosh and Windows.
-However, there are various free websites that offer search interfaces to the Unicode code tables, like http://unicode-search.net or http://unicode-search.net.
-A further useful approach for searching characters using shape matching is http://shapecatcher.com.}
+In general, finding a specific character in the Unicode Standard is often not
+trivial. The names of the blocks can help, but it is not (and never was supposed
+to be) a foolproof structure. It is not the goal nor aim of the Unicode
+Consortium to provide a user interface to the Unicode Standard. If one often
+encounters the problem of needing to find a suitable character, there are
+various other useful services for end-users available.\footnote{The Unicode
+website offers a basic interface to the code charts at
+\url{http://www.unicode.org/charts/index.html}. As a more flexible interface, we
+particularly like PopChar from Macility, available for both Macintosh and
+Windows. However, there are also various free websites that offer search interfaces
+to the Unicode code tables, like \url{http://unicode-search.net} or
+\url{http://unicode-search.net}. A further useful approach for searching characters
+using shape matching is \url{http://shapecatcher.com}.}
 
+% ==========================
 \section{Pitfall: Missing glyphs}
 \label{pitfall-missing-glyphs}
+% ==========================
 
-The Unicode Standard is often praised (and deservedly so) for solving many of the perennial problems with the interchange and display of the world's writing systems.
-However, a common complaint from users is that, while the praise may be true, they mostly just see some boxes on their screen instead of those promised symbols.
-The problem of course is that users' computers do not have any glyphs installed matching the Unicode code points in the file they are trying to inspect.
-It is important to realize that internally in the computer everything still works as expected: any handling of Unicode code points works independently of how they are displayed on the screen.
-So, although a user might only see boxes being displayed, this user should be assured that everything is still in order.
+The Unicode Standard is often praised (and deservedly so) for solving many of
+the perennial problems with the interchange and display of the world's writing
+systems. However, a common complaint from users is that, while the praise may be
+true, they mostly just see some boxes on their screen instead of those promised
+symbols. The problem of course is that users' computers do not have any glyphs
+installed matching the Unicode code points in the file they are trying to
+inspect. It is important to realize that internally in the computer everything
+still works as expected: any handling of Unicode code points works independently
+of how they are displayed on the screen. So, although a user might only see
+boxes being displayed, this user should be assured that everything is still in
+order.
 
-The central problem behind the missing glyphs is that designing actual glyphs includes a lot of different considerations and it is a time-consuming process.
-Many traditional expectations of how specific characters should look like have to be taken into account when designing glyphs.
-Those expectations are often not well documented, and it is mostly up to the knowledge and experience of the font designer to try and conform to them as good as possible.
-Therefore, most designers produce fonts only including glyphs for certains parts of the Unicode Standard, namely for those characters they feel comfortable with.
-At the same time, the number of characters defined by the Unicode Standard is growing with each new version, so it is neigh impossible for any designer to produce glyphs for all characters.
-The result of this is that, almost necessarily, each font only includes glyphs for a subset of the characters in the Unicode Standard.
+The central problem behind the missing glyphs is that designing actual glyphs
+includes a lot of different considerations and it is a time-consuming process.
+Many traditional expectations of how specific characters should look like have
+to be taken into account when designing glyphs. Those expectations are often not
+well documented, and it is mostly up to the knowledge and experience of the font
+designer to try and conform to them as good as possible. Therefore, most
+designers produce fonts only including glyphs for certain parts of the Unicode
+Standard, namely for those characters they feel comfortable with. At the same
+time, the number of characters defined by the Unicode Standard is growing with
+each new version, so it is neigh impossible for any designer to produce glyphs
+for all characters. The result of this is that, almost necessarily, each font
+only includes glyphs for a subset of the characters in the Unicode Standard.
 
-The simple solution to missing glyphs is thus to install additional fonts providing additional glyphs.
-For the more exotic characters there is often not much choice.
-There are a few particularly large fonts that might be considered.
-First, there is the \textsc{Everson Mono} font made by Michael Everson, which currently includes 9,756 different glyphs (not including Chinese) updated up to Unicode 7.0.\footnote{Everson Mono is available as shareware at http://www.evertype.com/emono/.} Already a bit older is the \textsc{Titus Cyberbit Basic} font made by Jost Gippert and Carl-Martin Bunz, which includes 10,044 different glyphs (not including Chinese), but not including newer characters added after Unicode 4.0.\footnote{Titus Cyberbit Basic is available at http://titus.fkidg1.uni-frankfurt.de/unicode/tituut.asp.}
+The simple solution to missing glyphs is thus to install additional fonts
+providing additional glyphs. For the more exotic characters there is often not
+much choice. There are a few particularly large fonts that might be considered.
+First, there is the \textsc{Everson Mono} font made by Michael Everson, which
+currently includes 9,756 different glyphs (not including Chinese) updated up to
+Unicode 7.0.\footnote{Everson Mono is available as shareware at
+\url{http://www.evertype.com/emono/}.} Already a bit older is the \textsc{Titus
+Cyberbit Basic} font made by Jost Gippert and Carl-Martin Bunz, which includes
+10,044 different glyphs (not including Chinese), but not including newer
+characters added after Unicode 4.0.\footnote{Titus Cyberbit Basic is available
+at \url{http://titus.fkidg1.uni-frankfurt.de/unicode/tituut.asp}.}
 
-Further, we suggest to always install at least one so-called \textsc{fall-back font,} which provides glyphs that at least show the user some information about the underlyingly encoded character.
-Apple Macintoshes have such a font (which is invisible to the user), which is designed by Michael Everson and made available for other systems through the Unicode Consortium.\footnote{The Apple/Everson fallback font is available at http://www.unicode.org/policies/lastresortfont\_eula.html.} Further, the \textsc{GNU Unifont} is a clever way to produce bitmaps approximating the intended glyph of each available character, updated to Unicode 7.0.\footnote{The GNU Unifont is available at http://unifoundry.com/unifont.html.} Finally, the Summer Institute of Linguistics provides a \textsc{SIL Unicode BMP Fallback Font}, currently available up to Unicode version 6.1.
-This font does not even attempt to show a real glyph, but only shows the hexadecimal code inside a box for each character, so a user can at least see the Unicode codepoint of the character to be displayed.\footnote{The SIL Unicode BMP Fallback Font is available at http://scripts.sil.org/UnicodeBMPFallbackFont.}
+Further, we suggest to always install at least one so-called \textsc{fall-back
+font}, which provides glyphs that at least show the user some information about
+the underlying encoded character. Apple Macintoshes have such a font (which is
+invisible to the user), which is designed by Michael Everson and made available
+for other systems through the Unicode Consortium.\footnote{The Apple/Everson
+fallback font is available for non-Macintosh users at \newline
+\url{http://www.unicode.org/policies/lastresortfont\_eula.html}.} Further, the
+\textsc{GNU Unifont} is a clever way to produce bitmaps approximating the
+intended glyph of each available character, updated to Unicode 7.0.\footnote{The
+GNU Unifont is available at \url{http://unifoundry.com/unifont.html}.} Finally,
+the Summer Institute of Linguistics provides a \textsc{SIL Unicode BMP Fallback
+Font}, currently available up to Unicode version 6.1. This font does not even
+attempt to show a real glyph, but only shows the hexadecimal code inside a box
+for each character, so a user can at least see the Unicode codepoint of the
+character to be displayed.\footnote{The SIL Unicode BMP Fallback Font is
+available at \newline \url{http://scripts.sil.org/UnicodeBMPFallbackFont}.}
 
+% ==========================
 \section{Pitfall: Faulty rendering}
 \label{pitfall-faulty-rendering}
+% ==========================
 
-A similar complaint to missing glyphs, discussed previously, is that while there might be a glyph being displayed, it does not `look right'.
-There are two reasons for unexpected visual display, namely automatic font substitution and faulty rendering.
-Like missing glyphs, any such problems are independent from the Unicode Standard.
-The Unicode Standard only includes very general information about characters and leaves the specific visual display to others to decide on.
-Any faulty display is thus not to be blamed on the Unicode Consortium, but on a complex interplay of different mechanisms happening in a computer to turn Unicode codepoints into visual symbols.
-We will only sketch a few aspects of this complex interplay here.
+A similar complaint to missing glyphs, discussed previously, is that while there
+might be a glyph being displayed, it does not look right. There are two
+reasons for unexpected visual display, namely automatic font substitution and
+faulty rendering. Like missing glyphs, any such problems are independent from
+the Unicode Standard. The Unicode Standard only includes very general
+information about characters and leaves the specific visual display to others to
+decide on. Any faulty display is thus not to be blamed on the Unicode
+Consortium, but on a complex interplay of different mechanisms happening in a
+computer to turn Unicode codepoints into visual symbols. We will only sketch a
+few aspects of this complex interplay here.
 
-Most modern software applications (like Microsoft Word) offer some approach to \textsc{automatic font substitution}.
-This means that when a text is written in a specific font (e.g.~Times New Roman) and a inserted Unicode character does not have a glyph within this font, then the software application will automatically search for another font to display the glyph.
-The result will be that some specific glyph will look slightly different from the others.
-This mechanism works differently depending on the software application, and mostly only limited user influence is expected and little feedback is given, which might be rather frustrating to font-aware users.\footnote{For example, Apple Pages does not give any feedback that a font is being replaced, and the user does not seem to have any influence on the choice of replacement (except by manually marking all occurrences).
-In contrast, Microsoft Word does indicate the font replacement by showing the name of the font replacement, but it simply changes the font, so any text written after the replacement is written in a different font as before.
-Both behaviors leave much to be desired.}
+Most modern software applications (like Microsoft Word) offer some approach to
+\textsc{automatic font substitution}. This means that when a text is written in
+a specific font (e.g.~Times New Roman) and an inserted Unicode character does not
+have a glyph within this font, then the software application will automatically
+search for another font to display the glyph. The result will be that this
+specific glyph will look slightly different from the others. This mechanism
+works differently depending on the software application, and mostly only limited
+user influence is expected and little feedback is given, which might be rather
+frustrating to font-aware users.\footnote{For example, Apple Pages does not give
+any feedback that a font is being replaced, and the user does not seem to have
+any influence on the choice of replacement (except by manually marking all
+occurrences). In contrast, Microsoft Word does indicate the font replacement by
+showing the name in the font menu of the font replacement. However, Word simply
+changes the font completely, so any text written after the replacement is written in a
+different font as before. Both behaviors leave much to be desired.}
 
-The other problem with visual display is related to the so-called `font rendering'.
-\textsc{Font rendering }refers to the process of the actual positioning of Unicode characters on a page of written text.
-This positioning is actually a highly complex problem, and many things can go wrong in the process.
-Well-known rendering problems, like proportional glyph size or ligatures are reasonably well understood.
-However, even within the Latin script, the positioning of diacritics relative to a base character is still a widespread problem.
-Especially when more than one diacritic is supposed to be placed above (or below) each other, this often leads to unexpected effects in many modern software applications.
-The problems arising in Arabic and in many southeast Asian scripts (like Devanagari or Burmese) are even much more complex.
-There are basically three different approaches to font rendering.
-The most widespread is Adobe's and Microsoft's \emph{OpenType }system.
-This approach makes it relatively easy for font developers, as the font itself does not include all details about the precise placement of individual characters.
-For those details, additional script-descriptions are necessary.
-All of those systems can lead to unexpected behaviour.\footnote{For more details about OpenType, see http://www.adobe.com/products/type/opentype.html and http://www.microsoft.com/typography/otspec/.
-Additional systems for complex text layout are, among others, Microsoft's DirectWrite https://msdn.microsoft.com/library/dd368038.aspx and the open-source project HarfBuzz http://www.freedesktop.org/wiki/Software/HarfBuzz/.} Alternative systems are Apple's \emph{Apple Advanced Typography} (AAT) and the open-source \emph{Graphite} system from the Summer Institute of Linguistics (SIL).\footnote{More information about AAT can be found at https://developer.apple.com/fonts/.
-SIL's Graphite is described in detail at http://scripts.sil.org/cms/scripts/page.php?site\_id=projects\&item\_id=graphite\_home.} In both of these systems, a larger burden is placed on the description inside the font.
+The other problem with visual display is related to the so-called \textsc{font
+rendering}. Font rendering refers to the process of the actual positioning of
+Unicode characters on a page of written text. This positioning is actually a
+highly complex problem, and many things can go wrong in the process. Well-known
+rendering problems, like proportional glyph size or ligatures are reasonably
+well understood. However, even within the Latin script, the positioning of
+diacritics relative to a base character is still a widespread problem.
+Especially when more than one diacritic is supposed to be placed above (or
+below) each other, this often leads to unexpected effects in many modern
+software applications. The problems arising in Arabic and in many southeast
+Asian scripts (like Devanagari or Burmese) are even much more complex. There are
+basically three different approaches to font rendering. The most widespread is
+Adobe's and Microsoft's \emph{OpenType }system. This approach makes it
+relatively easy for font developers, as the font itself does not include all
+details about the precise placement of individual characters. For those details,
+additional script-descriptions are necessary. All of those systems can lead to
+unexpected behavior.\footnote{For more details about OpenType, see
+\url{http://www.adobe.com/products/type/opentype.html} and
+\url{http://www.microsoft.com/typography/otspec/}. Additional systems for
+complex text layout are, among others, Microsoft's DirectWrite
+\url{https://msdn.microsoft.com/library/dd368038.aspx} and the open-source
+project HarfBuzz \url{http://www.freedesktop.org/wiki/Software/HarfBuzz/}.}
+Alternative systems are Apple's \emph{Apple Advanced Typography} (AAT) and the
+open-source \emph{Graphite} system from the Summer Institute of Linguistics
+(SIL).\footnote{More information about AAT can be found at
+\url{https://developer.apple.com/fonts/}. SIL's Graphite is described in detail
+at
+\url{http://scripts.sil.org/cms/scripts/page.php?site\_id=projects\&item\_id=graphite\_home}.}
+In both of these systems, a larger burden is placed on the description inside
+the font.
 
+% ==========================
 \section{Pitfall: Homoglyphs}
 \label{pitfall-homoglyphs}
+% ==========================
 
-Homoglyphs are visually indistinguishable glyphs (or highly similar glyphs) that have different code points in the Unicode Standard and thus different character semantics.
-As a principle, the Unicode Standard does not specify how a character appears visually on the page or the screen.
-So in most cases, a different appearance is caused by the specific design of a font, or by user-settings like size or boldface.
-Taking an example already discussed in the previous section, the following symbols <reinsert Gs here> are different glyphs of the same character, i.e.~they may be rendered differently depending on the typography being used, but they all share the same code point (viz.
-LATIN SMALL LETTER G at U+0067).
-In contrast, the symbols <> are all different code points, although they look highly similar -- in some cases even sharing exactly the same glyph in some fonts.
-All these different A-like characters include the following code points in the Unicode Standard:
+Homoglyphs are visually indistinguishable glyphs (or highly similar glyphs) that
+have different code points in the Unicode Standard and thus different character
+semantics. As a principle, the Unicode Standard does not specify how a character
+appears visually on the page or the screen. So in most cases, a different
+appearance is caused by the specific design of a font, or by user-settings like
+size or boldface. Taking an example already discussed in
+Section~\ref{pitfall-homoglyphs}, the following symbols <g {\large \textit{g}}
+\textbf{g} {\fontspec{ArialMT} {\small g} \textit{g} \textbf{g}}> are different
+glyphs of the same character, i.e.~they may be rendered differently depending on
+the typography being used, but they all share the same code point (viz.
+\textsc{latin small letter g} at \uni{0067}). In contrast, the symbols
+<{\fontspec{EversonMono}AАΑᎪᗅᴀꓮ𐊠𐌀𝖠𝙰}> are all different code points,
+although they look highly similar---in some cases even sharing exactly the same
+glyph in some fonts. All these different A-like characters include the following
+code points in the Unicode Standard:
+
 \begin{itemize}
-	\item <A> LATIN CAPITAL LETTER A, at U+0041 
-	\item <> CYRILLIC CAPITAL LETTER A, at U+0410 
-	\item <> GREEK CAPITAL LETTER ALPHA, at U+0391 
-	\item <> CHEROKEE LETTER GO, at U+13AA 
-	\item <> CANADIAN SYLLABICS CARRIER GHO, at U+15C5 
-	\item <> LATIN LETTER SMALL CAPITAL A, at U+1D00 
-	\item <> LISU LETTER A, at U+A4EE 
-	\item <> FULLWIDTH LATIN CAPITAL LETTER A, at U+FF21 
-	\item <> CARIAN LETTER A, at U+102A0 
-	\item <> OLD ITALIC LETTER A, at U+10300 
-	\item <> MATHEMATICAL SANS-SERIF CAPITAL A, U+1D5A0 
-	\item <> MATHEMATICAL MONOSPACE CAPITAL A, at U+1D670 
+	\item <{\fontspec{EversonMono}A}> \textsc{latin capital letter a}, at \uni{0041} 
+	\item <{\fontspec{EversonMono}А}> \textsc{cyrillic capital letter a}, at \uni{0410} 
+	\item <{\fontspec{EversonMono}Α}> \textsc{greek capital letter alpha}, at \uni{0391} 
+	\item <{\fontspec{EversonMono}Ꭺ}> \textsc{cherokee letter go}, at \uni{13AA} 
+	\item <{\fontspec{EversonMono}ᗅ}> \textsc{canadian syllabics carrier gho}, at \uni{15C5} 
+	\item <{\fontspec{EversonMono}ᴀ}> \textsc{latin small letter capital a}, at \uni{1D00} 
+	\item <{\fontspec{EversonMono}ꓮ}> \textsc{lisu letter a}, at \uni{A4EE} 
+%	\item <{\fontspec{EversonMono}Ａ}> \textsc{fullwidth latin capital letter a}, at \uni{FF21} 
+	\item <{\fontspec{EversonMono}𐊠}> \textsc{carian letter a}, at \uni{102A0} 
+	\item <{\fontspec{EversonMono}𐌀}> \textsc{old italic letter a}, at \uni{10300} 
+	\item <{\fontspec{EversonMono}𝖠}> \textsc{mathematical sans-serif capital a}, \uni{1D5A0} 
+	\item <{\fontspec{EversonMono}𝙰}> \textsc{mathematical monospace capital a}, at \uni{1D670} 
 \end{itemize}
 
-The existence of such homoglyphs is partly due to legacy compatibility, but for the most part these characters are simply different characters that happen to look similar.\footnote{A particularly nice interface to look for homoglyphs is http://shapecatcher.com, based on the principle of recognizing shapes (Belongie et al.
-2002).} Yet, they are supposed to behave different from the perspective of a font designer.
-For example, when designing a Cyrillic font, the will have different aesthetics and different traditional expectation compared to a Latin .
+The existence of such homoglyphs is partly due to legacy compatibility, but for
+the most part these characters are simply different characters that happen to
+look similar.\footnote{A particularly nice interface to look for homoglyphs is
+\url{http://shapecatcher.com}, based on the principle of recognizing shapes
+\citep{Belongie2002}.} Yet, they are supposed to behave different from the
+perspective of a font designer. For example, when designing a Cyrillic font, the
+<A> will have different aesthetics and different traditional expectation
+compared to a Latin <A>.
 
-Such homoglyphs are a widespread problem for consistent encoding.
-Although for most users it looks like the words and are identical, in actual fact they do not even share a single code point.\footnote{The first words consists completely of Latin characters, namely U+0076, U+006F, U+0063, U+0065, U+0073, while the second is a mix of Cyrillic and Greek characters, namely U+03BD, U+03BF, U+0041, U+0435, U+0455.} For computers these two words are completely different entities.
-Commonly, when users with Cyrillic or Greek keyboards have to type some Latin-based orthography, they mix similar looking Cyrillic or Greek characters into their text, because those characters are so much easier to type.
-Similarly, when users want to enter an unusual symbol, they normally search by visual impression in their favourite software application, and just pick something that looks reasonably alike to what they expect the glyph to look like.
+Such homoglyphs are a widespread problem for consistent encoding. Although for
+most users it looks like the words <voces> and <νοсеѕ> are identical, in actual
+fact they do not even share a single code point.\footnote{The first words
+consists completely of Latin characters, namely \unif{0076}, \unif{006F},
+\unif{0063}, \unif{0065} and \unif{0073}, while the second is a mix of Cyrillic and
+Greek characters, namely \unif{03BD}, \unif{03BF}, \unif{0041}, \unif{0435} and
+\unif{0455}.} For computers these two words are completely different entities.
+Commonly, when users with Cyrillic or Greek keyboards have to type some
+Latin-based orthography, they mix similar looking Cyrillic or Greek characters
+into their text, because those characters are so much easier to type. Similarly,
+when users want to enter an unusual symbol, they normally search by visual
+impression in their favorite software application, and just pick something that
+looks reasonably alike to what they expect the glyph to look like.
 
+% ==========================
 \section{Pitfall: Canonical equivalence}
 \label{pitfall-canonical-equivalence}
+% ==========================
 
-For some characters, there is more than one possible encoding in the Unicode Standard.
-This is a possible pitfall, as this would mean that for the computer there exist multiple different entities that for a user are the same.
-This would, for example, lead to problems with searching, as the computer would search for specific encodings, and not find all expected characters.
-As a solution, the Unicode Standard includes a notion of \textsc{canonical equivalence}.
-Different encodings are explicitly declared as equivalent in the Unicode Standard code tables.
-Further, to harmonize all encodings in a specific piece of text, the Unicode Standard proposes a mechanism of \textsc{normalization}.
+For some characters, there is more than one possible encoding in the Unicode
+Standard. This is a possible pitfall, as this would mean that for the computer
+there exist multiple different entities that for a user are the same. This
+would, for example, lead to problems with searching, as the computer would
+search for specific encodings, and not find all expected characters. As a
+solution, the Unicode Standard includes a notion of \textsc{canonical
+equivalence}. Different encodings are explicitly declared as equivalent in the
+Unicode Standard code tables. Further, to harmonize all encodings in a specific
+piece of text, the Unicode Standard proposes a mechanism of
+\textsc{normalization}.
 
 Consider for example the characters and following Unicode code points:
-\begin{enumerate}
+\begin{itemize}
 	\def\labelenumi{\arabic{enumi}.} 
-	\item LATIN CAPITAL LETTER A WITH RING ABOVE (U+00C5) 
-	\item ANGSTROM SIGN (U+212B) 
-	\item LATIN CAPITAL LETTER A (U+0041) + COMBINING RING ABOVE (U+030A) 
-\end{enumerate}
+	\item <Å> \textsc{latin capital letter a with ring above} \uni{00C5} 
+	\item <Å> \textsc{angstrom sign} \uni{212B}
+	\item <Å> \textsc{latin capital letter a} \uni{0041}
+	+ \textsc{combining ring above} \uni{030A}
+\end{itemize}
 
-The character, represented here by glyph , is encoded in the Unicode Standard in the first two examples by a single-character sequence; each is assigned a different code point.
-In the third example, the glyph is encoded in a multiple-character sequence that is composed of two character code points.
-All three sequences are \textsc{canonically equivalent}, i.e.~they are strings that represent the same abstract character and because they are not distinguishable by the user, the Unicode Standard requires them to be treated the same in regards to their behavior and appearance.
-Nevertheless, they are encoded differently.
-For example, if one were to search an electronic text (with software that does not apply Unicode Standard normalization) for ANGSTROM SIGN (U+212B), instances of LATIN CAPITAL LETTER A WITH RING ABOVE (U+00C5) would not be found.
+The character, represented here by glyph <Å>, is encoded in the Unicode Standard
+in the first two examples by a single-character sequence; each is assigned a
+different code point. In the third example, the glyph is encoded in a
+multiple-character sequence that is composed of two character code points. All
+three sequences are \textsc{canonically equivalent}, i.e.~they are strings that
+represent the same abstract character and because they are not distinguishable
+by the user, the Unicode Standard requires them to be treated the same in
+regards to their behavior and appearance. Nevertheless, they are encoded
+differently. For example, if one were to search an electronic text (with
+software that does not apply Unicode Standard normalization) for
+\textsc{angstrom sign} (\uni{212B}), instances of \textsc{latin capital letter a
+with ring above} (\uni{00C5}) would not be found.
 
-In other words, there are equivalent sequences of Unicode characters that should be normalized, i.e.~transformed into a unique Unicode-sanctioned representation of a character sequence called a \textsc{normalization form}.
-Unicode provides a \emph{Unicode Normalization Algorithm}, which essentially puts combining marks into a specific logical order and it defines decomposition and composition transformation rules to convert each string into one of four normalization forms.
-We will discuss here the two most relevant normalization forms: NFC and NFD.
+In other words, there are equivalent sequences of Unicode characters that should
+be normalized, i.e.~transformed into a unique Unicode-sanctioned representation
+of a character sequence called a \textsc{normalization form}. Unicode provides a
+Unicode Normalization Algorithm, which essentially puts combining marks
+into a specific logical order and it defines decomposition and composition
+transformation rules to convert each string into one of four normalization
+forms. We will discuss here the two most relevant normalization forms: NFC and
+NFD.\@
 
-The first of the three characters above is considered the \textsc{Normalization Form C (NFC)}, where ``C'' stands for composition.
-When the process of NFC normalization is applied to the character sequences in 2 and 3, both sequences are normalized into the `pre-composed' character sequence in 1.
-Thus all three canonical character sequences are standardized into one composition form in NFC.
-The other central Unicode normalization form is the \textsc{Normalization Form D (NFD)}, where ``D'' stands for decomposition.
-When NFD is applied to the three examples above, all three, including importantly the single-character sequences in 1 and 2, are normalized into the `decomposed' multiple-sequence of characters in 3.
-Again, all three are then logically equivalent and therefore comparable and syntactically interoperable.
+The first of the three characters above is considered the \textsc{Normalization
+Form C (NFC)}, where \textsc{C} stands for composition. When the process of NFC
+normalization is applied to the character sequences in 2 and 3, both sequences
+are normalized into the \textsc{pre-composed} character sequence in 1. Thus all
+three canonical character sequences are standardized into one composition form
+in NFC.\@The other central Unicode normalization form is the
+\textsc{Normalization Form D (NFD)}, where \textsc{D} stands for decomposition.
+When NFD is applied to the three examples above, all three, including
+importantly the single-character sequences in 1 and 2, are normalized into the
+\textsc{decomposed} multiple-sequence of characters in 3. Again, all three are
+then logically equivalent and therefore comparable and syntactically
+interoperable.
 
-As illustrated, some characters in the Unicode Standard have alternative representations (in fact, many do), but the Unicode Normalization Algorithm can be used to transform certain sequences of characters into canonical (or compatible) forms to test for equivalency.
-To determine equivalence, each character in the Unicode Standard is associated with a combining class, which is formally defined as a character property called ``Canonical Combining Class'' which is specified in the Unicode Character Database.
-The combining class assigned to each code point is a numeric value between 0 and 254 and is used by the Canonical Ordering Algorithm to determine which sequences of combining marks are canonically equivalent.
-Normalization forms, as very briefly described above, can be used to ensure character equivalence by ordering character sequences so that they can be faithfully compared.
+As illustrated, some characters in the Unicode Standard have alternative
+representations (in fact, many do), but the Unicode Normalization Algorithm can
+be used to transform certain sequences of characters into canonical (or
+compatible) forms to test for equivalency. To determine equivalence, each
+character in the Unicode Standard is associated with a combining class, which is
+formally defined as a character property called \textsc{canonical combining
+class} which is specified in the Unicode Character Database. The combining class
+assigned to each code point is a numeric value between 0 and 254 and is used by
+the Unicode Canonical Ordering Algorithm to determine which sequences of
+combining marks are canonically equivalent. Normalization forms, as very briefly
+described above, can be used to ensure character equivalence by ordering
+character sequences so that they can be faithfully compared.
 
-It is very important to note that any software applications that is Unicode Standard compliant is free to change the character stream from one representation to another.
-This means that a software application may compose, decompose or reorder characters as its developers desire; as long as the resultant strings are canonically equivalent to the original.
-This might lead to unexpected behaviour for users.
-Various players, like the Unicode Consortium, the W3C, or the TEI recommend NFC in most user-directed situations, and some software applications that we tested indeed seem to automatically convert strings into NFC.\footnote{See the summary of various recommendation here: http://www.win.tue.nl/\textasciitilde{}aeb/linux/uc/nfc\_vs\_nfd.html.
-} This means in practice that if a user, for example, enters <a> and <`>, i.e.~LATIN SMALL LETTER A at U+0061 and COMBINING GRAVE ACCENT at U+0300, this might be converted into <à>, i.e.~LATIN SMALL LETTER A WITH GRAVE at U+00E0.\footnote{The behaviour of software applications can be quite erratic in this respect.
-For example, Apple's TextEdit does not do any conversion on text entry.
-However, when you copy and paste some text inside the same document in ``rich text mode'' (i.e.~RTF-format), it will be transformed into NFC on paste.
-Saving a document does not do any conversion to the glyphs on screen, but it will save the characters in NFC.}
+It is very important to note that any software applications that is Unicode
+Standard compliant is free to change the character stream from one
+representation to another. This means that a software application may compose,
+decompose or reorder characters as its developers desire; as long as the
+resultant strings are canonically equivalent to the original. This might lead to
+unexpected behavior for users. Various players, like the Unicode Consortium, the
+W{\large 3}C, or the TEI recommend NFC in most user-directed situations, and some
+software applications that we tested indeed seem to automatically convert
+strings into NFC.\footnote{See the summary of various recommendation here:
+\url{http://www.win.tue.nl/~aeb/linux/uc/nfc_vs_nfd.html}.}
+This means in practice that if a user, for example, enters <a> and < ̀>,
+i.e.~\textsc{latin small letter a} at \uni{0061} and \textsc{combining grave
+accent} at \uni{0300}, this might be converted into <à>, i.e.~\textsc{latin
+small letter a with grave} at \uni{00E0}.\footnote{The behavior of software
+applications can be quite erratic in this respect. For example, Apple's TextEdit
+does not do any conversion on text entry. However, when you copy and paste some
+text inside the same document in rich text mode (i.e.~RTF-format), it
+will be transformed into NFC on paste. Saving a document does not do any
+conversion to the glyphs on screen, but it will save the characters in NFC.}
 
+% ==========================
 \section{Pitfall: Characters are not glyphs}
 \label{pitfall-characters-are-not-glyphs}
+% ==========================
 
-A central principle of Unicode is the distinction between character and glyph.
-A character is the abstract notion of a symbol in a writing system, while a glyph is the concrete drawing of such a symbol.
-In practice, there is a complex interaction between characters and glyphs.
-A single Unicode character may of course be rendered as a single glyph.
-However, a character may also be a `piece' of a glyph, or vice-versa.
-Actually, all possible relations between glyphs and characters are attested.
+A central principle of Unicode is the distinction between character and glyph. A
+character is the abstract notion of a symbol in a writing system, while a glyph
+is the concrete drawing of such a symbol. In practice, there is a complex
+interaction between characters and glyphs. A single Unicode character may of
+course be rendered as a single glyph. However, a character may also be a
+\textit{piece} of a glyph, or vice-versa. Actually, all possible relations
+between glyphs and characters are attested.
 
-First, \textsc{a single character may have different contextually determined glyphs.} For example, characters in writing systems like Hebrew and Arabic have different glyphs depending on where they appear in a word.
-Some letters in Hebrew change their form at the end of the word, and in Arabic, primary letters have four contextually-sensitive variants (isolated, word initial, medial and final).
-Second, \textsc{a single character may be rendered as a sequence of multiple glyphs.} For example, in Tamil one Unicode character may result in a combination of a consonant and vowel, which are rendered as two adjacent glyphs by fonts that supports Tamil.
-Third, \textsc{a single glyph may be a combination of multiple characters.
-}For example, the ligature <ﬁ>, a single glyph, is the result of two characters, <f> and <i>, that have undergone glyph substitution by font rendering.
-Like contextually-determined glyphs, ligatures are (intended) artifacts of text processing instructions.
-Finally, \textsc{a single glyph may be a part of a character} (e.g.~diacritics).
+First, \textit{a single character may have different contextually determined
+glyphs.} For example, characters in writing systems like Hebrew and Arabic have
+different glyphs depending on where they appear in a word. Some letters in
+Hebrew change their form at the end of the word, and in Arabic, primary letters
+have four contextually-sensitive variants (isolated, word initial, medial and
+final). Second, \textit{a single character may be rendered as a sequence of
+multiple glyphs.} For example, in Tamil one Unicode character may result in a
+combination of a consonant and vowel, which are rendered as two adjacent glyphs
+by fonts that supports Tamil. Third, \textit{a single glyph may be a combination
+of multiple characters.} For example, the ligature <ï¬>, a single glyph, is the
+result of two characters, <f> and <i>, that have undergone glyph substitution by
+font rendering. Like contextually-determined glyphs, ligatures are (intended)
+artifacts of text processing instructions. Finally, \textit{a single glyph may
+be a part of a character} (e.g.~diacritics).
 
-Further, the rendering of a glyph is also always dependent on the font being used.
-For example, the Unicode character LATIN SMALL LETTER G appears as and in the Computer Modern and Courier fonts because their typefaces are designed differently.
-Furthermore, font face may change the visual appearance of a character, for example Times New Roman two-story <a> changes to a single-story a in italics <\emph{a}>.
-This becomes a real problem for some phonetic typesetting (see Section 5, Pitfall 2).
+Further, the rendering of a glyph is also always dependent on the font being
+used. For example, the Unicode character \textsc{latin small letter g} appears
+as and in the Computer Modern and Courier fonts because their typefaces are
+designed differently. Furthermore, font face may change the visual appearance of
+a character, for example Times New Roman two-story <a> changes to a single-story
+a in italics <\emph{a}>. This becomes a real problem for some phonetic
+typesetting (see Section~\ref{pitfall-missing-glyphs}).
 
-In sum, character-to-glyph mappings are complex technological issues that the Unicode Consortium has had to address in the development of the Unicode Standard, but for the lay user they can be utterly confusing because visual rendering does not indicate logical encoding, or, in other words, what you see isn't (necessarily) what you get.
+In sum, character-to-glyph mappings are complex technological issues that the
+Unicode Consortium has had to address in the development of the Unicode
+Standard, but for the lay user they can be utterly confusing because visual
+rendering does not indicate logical encoding, or, in other words, what you see
+isn't (necessarily) what you get.
 
+% ==========================
 \section{Pitfall: File formats}
 \label{pitfall-file-formats}
+% ==========================
 
-Unicode is a character encoding standard, but these characters of course actually appear inside some kind of file.
-The most basic Unicode-based file format is pure line-based text, i.e.~strings of Unicode-encoded characters separated by line breaks (note that these line breaks are what for most people intuitively corresponds to paragraph breaks).
-Unfortunately, even within this apparently basic setting there exist a multitude of variants.
-In general, these different possibilities are well-understood in the software industry, and nowadays they normally do not lead to any problems for the end user.
-However, there are some situations in which a user is suddenly confronted with cryptic questions in the user interface involving abbreviations like LF, CR, BE, LE or BOM.
-Most prominently this occurs with exporting or importing data in several software applications from Microsoft.
-Basically, there are two different issues involved.
-First, the encoding of** line breaks** and, second, the encoding of the Unicode characters into \textsc{code units} and the related issue of \textsc{endianness}.
+Unicode is a character encoding standard, but these characters of course
+actually appear inside some kind of file. The most basic Unicode-based file
+format is pure line-based text, i.e.~strings of Unicode-encoded characters
+separated by line breaks (note that these line breaks are what for most people
+intuitively corresponds to paragraph breaks). Unfortunately, even within this
+apparently basic setting there exist a multitude of variants. In general, these
+different possibilities are well-understood in the software industry, and
+nowadays they normally do not lead to any problems for the end user. However,
+there are some situations in which a user is suddenly confronted with cryptic
+questions in the user interface involving abbreviations like \textsc{lf},
+\textsc{cr}, \textsc{be}, \textsc{le} or \textsc{bom}. Most prominently this
+occurs with exporting or importing data in several software applications from
+Microsoft. Basically, there are two different issues involved. First, the
+encoding of line breaks and, second, the encoding of the Unicode characters into
+\textsc{code units} and the related issue of \textsc{endianness}.
 
-The issue with \textsc{line breaks} originated with the instructions necessary to direct a printing head to a new line.
-This involves two movements, known as \textsc{carriage return (CR)} (returning the printing head to the start of the line on the page) and \textsc{line feed (LF)} (moving the printing head to the next line on the page).
-Physically, these are two different events, but conceptually together they form one action.
-In the history of computing, various encodings of line breaks have been used (e.g.~CR+LF, LF+CR, only LF, or only CR).
-Currently, all Unix and Unix-derived systems use only LF as code for a line break, while software from Microsoft still uses a combination of CR and LF.
-Today, most software applications recognize both options, and are able to deal with either encoding of line breaks (until rather recently this was not the case, and using the wrong line breaks could lead to unexpected errors).
-Our impression is that there is a strong tendency in software development to standardize on the simpler ``only LF'' encoding for line breaks, and we suggest that everybody use this encoding whenever possible.
+The issue with \textsc{line breaks} originated with the instructions necessary
+to direct a printing head to a new line. This involves two movements, known as
+\textsc{carriage return (cr)} (returning the printing head to the start of the
+line on the page) and \textsc{line feed (lf)} (moving the printing head to the
+next line on the page). Physically, these are two different events, but
+conceptually together they form one action. In the history of computing, various
+encodings of line breaks have been used (e.g.~\textsc{cr+lf}, \textsc{lf+cr},
+only \textsc{lf}, or only \textsc{cr}). Currently, all Unix and Unix-derived
+systems use only \textsc{lf} as code for a line break, while software from
+Microsoft still uses a combination of \textsc{cr} and \textsc{lf}. Today, most
+software applications recognize both options, and are able to deal with either
+encoding of line breaks (until rather recently this was not the case, and using
+the wrong line breaks could lead to unexpected errors). Our impression is that
+there is a strong tendency in software development to standardize on the simpler
+\textit{only \textsc{lf}} encoding for line breaks, and we suggest that
+everybody use this encoding whenever possible.
 
-A code unit is a sequence of bits used to encode character in a encoding.
-The issue with code units stems from the question how to separate a stream of binary ones and zero, i.e.~bits, into chunks representing Unicode characters.
-Depending on different use cases, the Unicode Standard offers three different approaches, called UTF-32, UTF-16 and UTF-8.\footnote{The letters UTF stand for either ``Unicode Transformation Format'' or ``Universal Coded Character Set Transformation Format'', but the notion of `transformation' is a legacy notion that does not have meaning anymore.
-Nevertheless, the designation ``UTF'' (in capitals) has become an official standard designation, but should probably best be read as simply ``Unicode Format''.} The details of this issue is extensively explained in section 2.5 of the Unicode core specification.
-Basically, \textsc{UTF-32} encodes each character in 32 bits (32 `binary units', i.e.~32 zeros or ones) and is the most disk-space-consuming variant of the three.
-However, it is the most efficient encoding processing-wise, because the computer simply has to separate each character after 32 bits.
-In contrast, \textsc{UTF-16} uses only 16 bits per character, which is sufficient for the large majority of Unicode characters, but not for all of them.
-A special system of `surrogates' is defined within the Unicode Standard to deal with these additional characters.
-The effect is a more disk-space efficient encoding (approximately half the size), while adding a limited computational overhead.
-Finally, \textsc{UTF-8} is a more complex system that dynamically encodes each character with the minimally necessary number of bits, choosing either 8, 16 or 32 bits depending on the character.
-This represents again a strong reduction in space (particularly due to the high frequency of data using ASCII characters that need only 8 bits) at the expense of even more computation necessary to process such strings.
-However, because of the ever growing computational power of modern machines, the processing overhead is in most practical situations a non-issue, while saving on space is still useful, particularly for sending texts over the Internet.
-As an effect, UTF-8 has become the dominant encoding on the World Wide Web.
-We suggest that everybody uses UTF-8 as their default encoding.
+A code unit is a sequence of bits used to encode character in a encoding. The
+issue with code units stems from the question how to separate a stream of binary
+ones and zero, i.e.~bits, into chunks representing Unicode characters. Depending
+on different use cases, the Unicode Standard offers three different approaches,
+called UTF-32, UTF-16 and UTF-8.\footnote{The letters UTF stand for either
+\textsc{Unicode Transformation Format} or \textsc{Universal Coded Character Set
+Transformation Format}, but the notion of \textit{transformation} is a legacy
+notion that does not have meaning anymore. Nevertheless, the designation
+\textsc{UTF} (in capitals) has become an official standard designation, but
+should probably best be read as simply \textit{Unicode Format}.} The details of
+this issue is extensively explained in section 2.5 of the Unicode core
+specification. Basically, \textsc{UTF-32} encodes each character in 32 bits (32
+\textit{bi}nary un\textit{ts}, i.e.~32 zeros or ones) and is the most
+disk-space-consuming variant of the three. However, it is the most efficient
+encoding processing-wise, because the computer simply has to separate each
+character after 32 bits. In contrast, \textsc{UTF-16} uses only 16 bits per
+character, which is sufficient for the large majority of Unicode characters, but
+not for all of them. A special system of `surrogates' is defined within the
+Unicode Standard to deal with these additional characters. The effect is a more
+disk-space efficient encoding (approximately half the size), while adding a
+limited computational overhead. Finally, \textsc{UTF-8} is a more complex system
+that dynamically encodes each character with the minimally necessary number of
+bits, choosing either 8, 16 or 32 bits depending on the character. This
+represents again a strong reduction in space (particularly due to the high
+frequency of data using ASCII characters that need only 8 bits) at the expense
+of even more computation necessary to process such strings. However, because of
+the ever growing computational power of modern machines, the processing overhead
+is in most practical situations a non-issue, while saving on space is still
+useful, particularly for sending texts over the Internet. As an effect, UTF-8
+has become the dominant encoding on the World Wide Web. We suggest that
+everybody uses UTF-8 as their default encoding.
 
-A related problem is a general issue about how to store information in computer memory, which is known as \textsc{endianness}.
-The details of this issue go beyond the scope of this paper.
-It suffices to realize that there is a difference between \textsc{big-endian} (BE) storage and \textsc{little-endian} (LE) storage.
-The Unicode Standard offers a possibility to explicitly indicate what kind of storage is used by starting a file with a so-called `Byte Order Mark' (BOM).
-However, the Unicode Standard does not require the usage of BOM, preferring other non-Unicode methods to signal to computers which kind of endianness is used.
-This issue only arises with UTF-32 and UTF-16 encodings, and when using the preferred UTF-8, using a BOM is theoretically possible, but strongly dispreferred according to the Unicode Standard.
-We suggest that everyone tries to prevent the inclusion of BOM in your data.
+A related problem is a general issue about how to store information in computer
+memory, which is known as \textsc{endianness}. The details of this issue go
+beyond the scope of this paper. It suffices to realize that there is a
+difference between \textsc{big-endian} (BE) storage and \textsc{little-endian}
+(LE) storage. The Unicode Standard offers a possibility to explicitly indicate
+what kind of storage is used by starting a file with a so-called `Byte Order
+Mark' (BOM). However, the Unicode Standard does not require the usage of BOM,
+preferring other non-Unicode methods to signal to computers which kind of
+endianness is used. This issue only arises with UTF-32 and UTF-16 encodings, and
+when using the preferred UTF-8, using a BOM is theoretically possible, but
+strongly dispreferred according to the Unicode Standard. We suggest that
+everyone tries to prevent the inclusion of BOM in your data.
 
-Together with the topic of normalisation (NFC vs.~NFD, see Pitfall 5 above), this leads to the following nicely cryptic file-encoding recommendation, namely to preferably use \textsc{UTF-8 in NFC without BOM with LF line breaks}, whenever possible.
+Together with the topic of normalization (NFC vs.~NFD, see Pitfall 5 above),
+this leads to the following nicely cryptic file-encoding recommendation, namely
+to preferably use \textsc{UTF-8 in NFC without BOM with LF line breaks},
+whenever possible.

### 2015-06-05 chapters/use_cases.tex
major style revision

--- 2015-05-31 chapters/use_cases.tex
@@
 
-We now present several use cases that have motivated the development of orthography profiles. These include:
+We now present several use cases that have motivated the development of
+orthography profiles. These include: 
 \begin{itemize}
@@
 
-An important assumption of our work is that input sources are encoded in Unicode (UTF-8 to be precise).
-Anything that the Unicode Standard is unable to capture, cannot be captured by an orthography profile.
-We also use Unicode normalization form NFD to decompose all incoming text input into normalized and (Unicode) logically ordered strings.
-This process is of fundamental importance when working with text data in the Unicode Standard.
+An important assumption of our work is that input sources are encoded in Unicode
+(UTF-8 to be precise). Anything that the Unicode Standard is unable to capture,
+cannot be captured by an orthography profile. We also use Unicode normalization
+form NFD to decompose all incoming text input into normalized and (Unicode)
+logically ordered strings. This process is of fundamental importance when
+working with text data in the Unicode Standard.
 
-\section{Tokenization and error checking}
-The most basic use case is tokenization of language data. Tokenization comes in three types:
+\section{Tokenization and error checking} The most basic use case is
+tokenization of language data. Tokenization comes in three types:
 
@@
 
-The simplest form of tokenization provided by any Unicode Standard-compliant software will split an input text stream on Unicode character code points.
-The input text is split on the character bit sequences as they have been encoded by the user.
-This may be in various Unicode Normalization Forms (see Section \ref{}) and tokenization split function returns a byte stream sequence given a particular encoding (see Section \ref{}).
+The simplest form of tokenization provided by any Unicode Standard-compliant
+software will split an input text stream on Unicode character code points. The
+input text is split on the character bit sequences as they have been encoded by
+the user. This may be in various Unicode Normalization Forms (see Section
+\ref{}) and tokenization split function returns a byte stream sequence given a
+particular encoding (see Section \ref{}).
 
-There is increasing support for the regular expression match, commonly ``\\X'', to identify Unicode graphemes (see Section \ref{}) and perform tokenization on these sets of Unicode characters.
-Examples are given in Section \ref{}.
+There is increasing support for the regular expression match, commonly ``\\X'',
+to identify Unicode graphemes (see Section \ref{}) and perform tokenization on
+these sets of Unicode characters. Examples are given in Section \ref{}.
 
-The orthography profile provides the mechanism for tokenizing an input stream on tailored grapheme clusters.
-The orthography's formal specification is given in Section \ref{}.
-To sum, an orthography profile is a list of tailored grapheme clusters and / or orthographic rules that specify how a specific resource of text input should be tokenized.
+The orthography profile provides the mechanism for tokenizing an input stream on
+tailored grapheme clusters. The orthography's formal specification is given in
+Section \ref{}. To sum, an orthography profile is a list of tailored grapheme
+clusters and / or orthographic rules that specify how a specific resource of
+text input should be tokenized.
 
-Once text has been normalized, an additional and straightforward process is Unicode grapheme tokenization.
-More recently this regular expression, often available as short cut ``\textbackslash{X}'', identifies all sequences of ``base'' characters followed by 0 or more combining diacritics.
-For example, the sequence, % <ŋ͡mṍraː> would tokenize as a space-delimited sequence as: <ŋ͡ m ṍ r a ː>.
-For a first pass at identifying grapheme clusters, this is a straightforward
-tokenization.
-However, the linguist will note that both the tie bar (COMBINING DOUBLE INVERTED BREVE, a Unicode Combining Modifier) and the length marker (MODIFIER LETTER TRIANGULAR COLON.
-a Unicode Letter Modifier) do not appear ``correctly'', i.e.~the tie bar is grouped with the first character in the sequence and the length marker appears singly.
-This is necessary, as defined by the Unicode Standard, because of these character's semantic properties -- with Unicode we cannot know if, for example, the MODIFIER LETTER TRIANGULAR COLON should appear before or after the base character that it modifies, cf.~the IPA aspiration marker, , the MODIFIER LETTER SMALL H, which is also a Unicode Letter Modifier and for linguists a diacritic that be be used for pre- or post-aspiration.
-These ambiguities of position in tokenization are not uncommon in IPA, thus orthographic (or source) normalization and tokenization is needed.
+Once text has been normalized, an additional and straightforward process is
+Unicode grapheme tokenization. More recently this regular expression, often
+available as short cut ``\textbackslash{X}'', identifies all sequences of
+``base'' characters followed by 0 or more combining diacritics. For example, the
+sequence, % <ŋ͡mṍraː> would tokenize as a space-delimited sequence as: <ŋ͡ m ṍ r
+a ː>. For a first pass at identifying grapheme clusters, this is a
+straightforward tokenization. However, the linguist will note that both the tie
+bar (COMBINING DOUBLE INVERTED BREVE, a Unicode Combining Modifier) and the
+length marker (MODIFIER LETTER TRIANGULAR COLON. a Unicode Letter Modifier) do
+not appear ``correctly'', i.e.~the tie bar is grouped with the first character
+in the sequence and the length marker appears singly. This is necessary, as
+defined by the Unicode Standard, because of these character's semantic
+properties -- with Unicode we cannot know if, for example, the MODIFIER LETTER
+TRIANGULAR COLON should appear before or after the base character that it
+modifies, cf.~the IPA aspiration marker, , the MODIFIER LETTER SMALL H, which is
+also a Unicode Letter Modifier and for linguists a diacritic that be be used for
+pre- or post-aspiration. These ambiguities of position in tokenization are not
+uncommon in IPA, thus orthographic (or source) normalization and tokenization is
+needed.
+
+\section{Cross-orthographic analysis of writing systems}
+
+Given that orthography profiles are stored in a standard CSV format, we can use tools for converting and working with CSV.
+One such tool is the command line utility \textsc{csvkit}.\footnote{\url{https://csvkit.readthedocs.org}}
+
+% insert some examples of creating a table of language | grapheme, language | orthographic contexts
+% any measure of writing systems complexity?
 
@@
 
-A preliminary orthography profile can be generated with code available online.\footnote{\url{https://github.com/bambooforest/tokenizer}} The Python script creates a unigram model for individual Unicode characters or Unicode grapheme clusters and their frequency in the source input.
-We find that this gives a broad overview of the input at the character and grapheme levels and it often finds typographic errors in the source input by identifying rarely occurring or one-off characters in the input.
-The models also provides the basis for the ``quick'' development of an initial orthography profile by saving the user keyboarding time.
+A preliminary orthography profile can be generated with code available
+online.\footnote{\url{https://github.com/bambooforest/tokenizer}} The Python
+script creates a unigram model for individual Unicode characters or Unicode
+grapheme clusters and their frequency in the source input. We find that this
+gives a broad overview of the input at the character and grapheme levels and it
+often finds typographic errors in the source input by identifying rarely
+occurring or one-off characters in the input. The models also provides the basis
+for the ``quick'' development of an initial orthography profile by saving the
+user keyboarding time.
 
-Note that these symbols are still phonetically unspecified, i.e.~symbol may or may not represent, say, a front open unrounded vowel.
+Note that these symbols are still phonetically unspecified, i.e.~symbol may or
+may not represent, say, a front open unrounded vowel.
 
-The orthography profile is designed so that specified strings of characters are captured for tokenization.
+The orthography profile is designed so that specified strings of characters are
+captured for tokenization.
 
@@
 
-\section{Diacritics: Unicode letter modifiers (Lm)}\label{diacritics-unicode-letter-modifiers-lm}
+\section{Diacritics: Unicode letter modifiers (Lm)}
+\label{diacritics-unicode-letter-modifiers-lm}
 

### 2015-06-05 chapters/encoding.tex
reverting this morning's width adjustments

--- 2015-05-31 chapters/encoding.tex
@@
 A useful survey of the historical development of the physical hardware of telegraphy and telecommunication is \citet{Huurdeman2003}.
-Most books that discuss the development of encoding of telegraphic communication focus of cryptography, e.g.~\citet{Singh1999}, and forego the rather interesting story of `open' encoding that is related here.}
+Most books that discuss the development of encoding of telegraphic communication focus of cryptography, e.g.~\citet{Singh1999}, and forego the rather interesting story of ``open encoding'' that is related here.}
 
@@
 As a solution, the Baudot code uses a so-called \textsc{shift} character, which signifies that from that point onwards---until shifted back---a different encoding is used, allowing for yet another set of 32 codes.
-In effect, this means that the Baudot code, and the \textsc{International Telegraph Alphabet} (ITA) derived from it, had an extra `bit' of information, so the encoding is actually 6-bit (with $2^6 = 64$ different possible characters).
+In effect, this means that the Baudot code, and the \textsc{International Telegraph Alphabet} (ITA) derived from it, had an extra \textsc{bit} of information, so the encoding is actually 6-bit (with $2^6 = 64$ different possible characters).
 For decades, this encoding was the standard for all telegraphy and it is still in limited use today.

### 2015-06-05 chapters/introduction.tex
reverting this morning's width adjustments

--- 2015-05-31 chapters/introduction.tex
@@
 All of these orthographies use the same symbols, but these symbols differ in meaning and usage throughout the various orthographies.
-To be able to computationally use and compare different orthographies, we need a way to specify all orthographic idiosyncrasies in a computer-readable format (a process called `tailoring' in Unicode parlance).
+To be able to computationally use and compare different orthographies, we need a way to specify all orthographic idiosyncrasies in a computer-readable format (a process called \textsc{tailoring} in Unicode parlance).
 We call such specifications \textsc{orthography profiles}.
@@
 In this work, it is crucial to be able to flexibly process across numerous resources with different orthographies.
-In many cases even different resources on the `same' language use different orthographic conventions.
+In many cases even different resources on the \textit{same} language use different orthographic conventions.
 Another orthographic challenge that we encounter regularly in our linguistic practice is electronic resources on a particular language that claim to follow a specific orthographic convention (often a resource-specific convention), but on closer inspection such resources are almost always not consistently encoded.

### 2015-06-05 chapters/ipa_meets_unicode.tex
reverting this morning's width adjustments

--- 2015-06-05 chapters/ipa_meets_unicode.tex
@@
-\chapter{IPA meets Unicode} \label{ipa-meets-unicode}
+\chapter{IPA meets Unicode}
+\label{ipa-meets-unicode}
 
@@
 
-The International Phonetic Alphabet (IPA) is a common standard in linguistics to
-transcribe sounds of spoken language into some Latin-based characters
-\citep{IPA2005}. Although IPA is reasonably easily adhered to with pen and
-paper, it is not trivial to encode IPA characters electronically. Early work
-addressing the need for a universal computing environment for writing systems
-and their computational complexity is discussed in Simons 1989. For a long time,
-linguists (like all other computer users) were limited to ASCII-encoded 7-bit
-characters, which only includes Latin characters, numbers and some punctuation
-and symbols. Restricted to these standard character sets that lacked IPA support
-or other language-specific graphemes that they needed, linguists devised their
-own solutions \citep{BirdSimons2003}. For example, some chose to represent
-unavailable graphemes with substitutes, e.g.~the combination of to represent .
-Tech-savvy linguists redefined selected characters from a character encoding by
-mapping custom made fonts to those code points. However, one linguist's
-electronic text would not render properly on another linguist's computer without
-access to the same font. Further, if two character encodings defined two
-character sets differently, then data could not be reliably and correctly
-displayed. This is a common example of the non-interoperability of data and data
-formats.
+The International Phonetic Alphabet (IPA) is a common standard in linguistics to transcribe sounds of spoken language into some Latin-based characters \citep{IPA2005}.
+Although IPA is reasonably easily adhered to with pen and paper, it is not trivial to encode IPA characters electronically.
+Early work addressing the need for a universal computing environment for writing systems and their computational complexity is discussed in Simons 1989.
+For a long time, linguists (like all other computer users) were limited to ASCII-encoded 7-bit characters, which only includes Latin characters, numbers and some punctuation and symbols.
+Restricted to these standard character sets that lacked IPA support or other language-specific graphemes that they needed, linguists devised their own solutions \citep{BirdSimons2003}.
+For example, some chose to represent unavailable graphemes with substitutes, e.g.~the combination of to represent .
+Tech-savvy linguists redefined selected characters from a character encoding by mapping custom made fonts to those code points.
+However, one linguist's electronic text would not render properly on another linguist's computer without access to the same font.
+Further, if two character encodings defined two character sets differently, then data could not be reliably and correctly displayed.
+This is a common example of the non-interoperability of data and data formats.
 
-To alleviate this problem, during the late 1980s, SAMPA (Speech Assessment
-Methods Phonetic Alphabet) was created to represent IPA symbols with 7-bit
-printable ASCII character sequences, e.g. <p\textbackslash> for [ɸ]. Two
-problems with SAMPA are that (i) it is only a partial encoding of the IPA and
-(ii) it encodes different languages in separate data tables, instead of a
-universal alphabet, like IPA. SAMPA tables are derived from phonemes appearing
-in several European languages that were developed as part of a European
-Commission-funded project to address technical problems like electronic mail
-exchange (what is now simply called email). SAMPA is essentially a hack to work
-around displaying IPA characters, but it provided speech technology and other
-fields a basis that has been widely adopted and used in code. So, SAMPA was a
-collection of tables to be compared, instead of a large universal table
-representing all languages. An extended version of SAMPA, called X-SAMPA, set
-out to include every symbol in the IPA chart including all diacritics
-\citep{WellsND}. X-SAMPA was considered more universally applicable because it
-consisted of one table that encoded the set of characters that represented
-phones/segments in IPA across languages. SAMPA and X-SAMPA have been widely used
-for speech technology and as an encoding system in computational linguistics.
-Eventually, ASCII-encoding of the IPA became depreciated through the advent of
-the Unicode Standard. Note however that many popular software packages used for
-linguistic analyses still require ASCII input,
-e.g.~RuG/L04\footnote{\url{http://www.let.rug.nl/kleiweg/L04/}} and
-SplitsTree4.\footnote{\url{http://www.splitstree.org/}}
+To alleviate this problem, during the late 1980s, SAMPA (Speech Assessment Methods Phonetic Alphabet) was created to represent IPA symbols with 7-bit printable ASCII character sequences, e.g.
+<p\textbackslash> for [ɸ].
+Two problems with SAMPA are that (i) it is only a partial encoding of the IPA and (ii) it encodes different languages in separate data tables, instead of a universal alphabet, like IPA.
+SAMPA tables are derived from phonemes appearing in several European languages that were developed as part of a European Commission-funded project to address technical problems like electronic mail exchange (what is now simply called email).
+SAMPA is essentially a hack to work around displaying IPA characters, but it provided speech technology and other fields a basis that has been widely adopted and used in code.
+So, SAMPA was a collection of tables to be compared, instead of a large universal table representing all languages.
+An extended version of SAMPA, called X-SAMPA, set out to include every symbol in the IPA chart including all diacritics \citep{WellsND}.
+X-SAMPA was considered more universally applicable because it consisted of one table that encoded the set of characters that represented phones/segments in IPA across languages.
+SAMPA and X-SAMPA have been widely used for speech technology and as an encoding system in computational linguistics.
+Eventually, ASCII-encoding of the IPA became depreciated through the advent of the Unicode Standard.
+Note however that many popular software packages used for linguistic analyses still require ASCII input, e.g.~RuG/L04\footnote{\url{http://www.let.rug.nl/kleiweg/L04/}} and SplitsTree4.\footnote{\url{http://www.splitstree.org/}}
 
-Still, there are a few pitfalls to be aware of when using the Unicode Standard
-to encode IPA. As we have said before, from a linguistic perspective it might
-look like the Unicode Consortium is making incomprehensible decisions, but it is
-important to realize that the consortium has tried and is continuing to try to
-be as consistent as possible across a wide range of use cases, and it does place
-linguistic traditions above other orthographic possibilities. In general, we
-strongly suggest linguists not to complain about any decisions in the Unicode
-Standard, but to try and understand the rationale of the Unicode Consortium
-(which in our experience is almost always well-conceived) and devise ways to
-work with any unexpected behavior. Many of the current problems derive from the
-fact that the IPA is clearly historically based on the Latin script, but
-different enough from most other Latin-based writing systems to warrant special
-attention. This ambivalent status of the IPA glyphs (partly Latin, partly
-special) is unfortunately also attested in the treatment of IPA in the Unicode
-Standard. In retrospect, it might have been better to consider the IPA (and
-other transcription systems) to be a special or new kind of script within the
-Unicode Standard, and treat the obvious similarity to Latin glyphs as a
-historical relic. All IPA glyphs would then have their own code points, instead
-of the current situation in which some IPA glyphs have special code points,
-while others are treated as being identical to the regular Latin characters.
-Yet, the current situation, however unfortunate, is unlikely to change, so as
-linguists we will have to learn to deal with the specific pitfalls of IPA within
-the Unicode Standard. In this section, we will describe these pitfalls in some
-detail.
+Still, there are a few pitfalls to be aware of when using the Unicode Standard to encode IPA.
+As we have said before, from a linguistic perspective it might look like the Unicode Consortium is making incomprehensible decisions, but it is important to realize that the consortium has tried and is continuing to try to be as consistent as possible across a wide range of use cases, and it does place linguistic traditions above other orthographic possibilities.
+In general, we strongly suggest linguists not to complain about any decisions in the Unicode Standard, but to try and understand the rationale of the Unicode Consortium (which in our experience is almost always well-conceived) and devise ways to work with any unexpected behavior.
+Many of the current problems derive from the fact that the IPA is clearly historically based on the Latin script, but different enough from most other Latin-based writing systems to warrant special attention.
+This ambivalent status of the IPA glyphs (partly Latin, partly special) is unfortunately also attested in the treatment of IPA in the Unicode Standard.
+In retrospect, it might have been better to consider the IPA (and other transcription systems) to be a special or new kind of script within the Unicode Standard, and treat the obvious similarity to Latin glyphs as a historical relic.
+All IPA glyphs would then have their own code points, instead of the current situation in which some IPA glyphs have special code points, while others are treated as being identical to the regular Latin characters.
+Yet, the current situation, however unfortunate, is unlikely to change, so as linguists we will have to learn to deal with the specific pitfalls of IPA within the Unicode Standard.
+In this section, we will describe these pitfalls in some detail.
 
@@
 
-The ambivalent nature of IPA glyphs arises because, on the one hand, the IPA
-uses Latin-based glyphs like , or . From this perspective, the IPA seems to be
-just another orthographic tradition using Latin characters, all of which do not
-get a special treatment within the Unicode Standard (just like e.g.~the French,
-German, or Danish orthographic traditions do not have a special status). On the
-other hand, the IPA uses many special symbols (like turned , mirrored and/or
-extended Latin glyphs ) not found in any other Latin-based writing system. For
-this reason, and already in the first version of the Unicode Standard (Version
-1.0 from 1991), a special block with code points, called \textsc{IPA Extensions}
-was included. As noted in Section 4, Pitfall 1, the Unicode Standard code space
-is subdivided into character blocks, which generally encode characters from a
-single script. However, as is illustrated by the IPA, characters that form a
-single writing system may be dispersed across several different character
-blocks. With its diverse collection of symbols from various scripts and
-diacritics, the IPA is spread across over 13 blocks in the Unicode
-Standard:\footnote{This number of blocks depends on whether only IPA-sanctioned
-symbols are counted or if the phonetic symbols commonly found in the literature
-are also included, see Moran 2012, Appendix C.}
+The ambivalent nature of IPA glyphs arises because, on the one hand, the IPA uses Latin-based glyphs like , or .
+From this perspective, the IPA seems to be just another orthographic tradition using Latin characters, all of which do not get a special treatment within the Unicode Standard (just like e.g.~the French, German, or Danish orthographic traditions do not have a special status).
+On the other hand, the IPA uses many special symbols (like turned , mirrored and/or extended Latin glyphs ) not found in any other Latin-based writing system.
+For this reason, and already in the first version of the Unicode Standard (Version 1.0 from 1991), a special block with code points, called \textsc{IPA Extensions} was included.
+As noted in Section 4, Pitfall 1, the Unicode Standard code space is subdivided into character blocks, which generally encode characters from a single script.
+However, as is illustrated by the IPA, characters that form a single writing system may be dispersed across several different character blocks.
+With its diverse collection of symbols from various scripts and diacritics, the IPA is spread across over 13 blocks in the Unicode Standard:\footnote{This number of blocks depends on whether only IPA-sanctioned symbols are counted or if the phonetic symbols commonly found in the literature are also included, see Moran 2012, Appendix C.}
 
 \begin{itemize}
-	\item Basic Latin (30 characters), e.g. <a, b, c, d, e> \item Latin-1
-	Supplement (4 characters): <æ, ç, ð, ø\textgreater{} \item Latin
-	Extended-A (3 characters): <ħ, ŋ, œ> \item Latin Extended-B (5
-	characters): <ǀ, ǁ, ǂ, ǃ, > \item IPA Extensions (70 characters), e.g.:
-	<ɐ, ɑ, ɔ> \item Spacing Modifier Letters (20 characters), e.g.: <ʰ ʷ ˥>
-	\item Combining Diacritical Marks (33 characters), e.g.: <̝ ̥ ̪> \item
-	Greek and Coptic (3 characters): <β, θ, χ> \item Phonetic Extensions (2
-	characters): <, > \item Phonetic Extensions Supplement (3 characters): <
-	, , > \item Superscripts and Subscripts (1 character): <ⁿ> \item Arrows
-	(2 characters): \textless{}↑, ↓\textgreater{} \item Latin Extended-C (1
-	character): <>
+	\item Basic Latin (30 characters), e.g. <a, b, c, d, e> 
+	\item Latin-1 Supplement (4 characters): <æ, ç, ð, ø\textgreater{} 
+	\item Latin Extended-A (3 characters): <ħ, ŋ, œ> 
+	\item Latin Extended-B (5 characters): <ǀ, ǁ, ǂ, ǃ, ȵ> 
+	\item IPA Extensions (70 characters), e.g.: <ɐ, ɑ, ɔ> 
+	\item Spacing Modifier Letters (20 characters), e.g.: <ʰ ʷ ˥> 
+	\item Combining Diacritical Marks (33 characters), e.g.: <̝ ̥ ̪> 
+	\item Greek and Coptic (3 characters): <β, θ, χ> 
+	\item Phonetic Extensions (2 characters): <ᴅ, ᴴ> 
+	\item Phonetic Extensions Supplement (3 characters): <ᶑ , ᶾ, ᶣ> 
+	\item Superscripts and Subscripts (1 character): <ⁿ> 
+	\item Arrows (2 characters): \textless{}↑, ↓\textgreater{} 
+	\item Latin Extended-C (1 character): <ⱱ> 
 \end{itemize}
@@
 
-Another problem is the large number of homoglyphs, i.e.~different characters
-that have highly similar glyphs (or even completely identical, depending on the
-font rendering). For example, a speaker of Russian should ideally not use the
-\textsc{cyrillic small letter a} at code point U+0430 for IPA transcriptions,
-but instead the \textsc{latin small letter a} at code point U+0061, although
-visually they are mostly indistinguishable, and the Cyrillic character is more
-easily typed on a Cyrillic keyboard. Another example we commonly encounter is
-the use of \textsc{latin small letter g} at U+0067, instead of the sanctioned
-Unicode Standard IPA character for the voiced velar stop \textsc{latin small
-letter script g} at U+0261. One begins to question whether this issue is at all
-apparent to the working linguist, or if they simply use the U+0067 because it is
-easily keyboarded and thus saves time, whereas the latter must be cumbersomely
-inserted as a special symbol in most software.\footnote{This issue was recently
-addressed by the International Phonetic Association, which has taken the stance
-that both the keyboard \textsc{latin small letter g} and the \textsc{latin small
-letter script g} are valid input characters for the voiced velar plosive.
-Unfortunately, this decision further introduces ambiguity for linguists trying
-to adhere to a strict Unicode Standard IPA encoding.}
+Another problem is the large number of homoglyphs, i.e.~different characters that have highly similar glyphs (or even completely identical, depending on the font rendering).
+For example, a speaker of Russian should ideally not use the \textsc{cyrillic small letter a} at code point U+0430 for IPA transcriptions, but instead the \textsc{latin small letter a} at code point U+0061, although visually they are mostly indistinguishable, and the Cyrillic character is more easily typed on a Cyrillic keyboard.
+Another example we commonly encounter is the use of \textsc{latin small letter g} at U+0067, instead of the sanctioned Unicode Standard IPA character for the voiced velar stop \textsc{latin small letter script g} at U+0261.
+One begins to question whether this issue is at all apparent to the working linguist, or if they simply use the U+0067 because it is easily keyboarded and thus saves time, whereas the latter must be cumbersomely inserted as a special symbol in most software.\footnote{This issue was recently addressed by the International Phonetic Association, which has taken the stance that both the keyboard \textsc{latin small letter g} and the \textsc{latin small letter script g} are valid input characters for the voiced velar plosive.
+Unfortunately, this decision further introduces ambiguity for linguists trying to adhere to a strict Unicode Standard IPA encoding.}
 
-Furthermore, on the one hand even linguists are unlikely to distinguish between
-the <ə> \textsc{latin small letter schwa} at code point U+0259 and <ǝ>
-\textsc{latin small letter turned e} at U+01DD. On the other hand, non-linguists
-are unlikely to distinguish any semantic difference between an open back
-unrounded vowel <ɑ>, the \textsc{latin small letter alpha} at U+0251, and the
-open front unrounded vowel <a>, \textsc{latin small letter a} at U+0061. But
-even among linguists this distinction leads to problems. For example, as pointed
-out by \citet{Mielke2009}, there is a problem stemming from the fact that about
-75\% of languages are reported to have a five-vowel system
-\citep{Maddieson1984}. Historically, linguistic descriptions tend not to include
-precise audio recording and measurements of formants, so this may lead one to
-ask if the many characters that are used in phonological description reflects a
-\textsc{transcriptional bias}. The common use of in transcriptions could be in
-part due to the ease of typing the letter on an English keyboard (or for older
-descriptions, the typewriter). We found it to be exceedingly rare that a
-linguist uses for a low back unrounded vowel.\footnote{One example is
-\citet[75]{Vidal2001a}, in which the author states: ``The definition of Pilagá
-/a/ as {[}+back{]} results from its behavior in certain phonological contexts.
+Furthermore, on the one hand even linguists are unlikely to distinguish between the <ə> \textsc{latin small letter schwa} at code point U+0259 and <ǝ>  \textsc{latin small letter turned e} at U+01DD.
+On the other hand, non-linguists are unlikely to distinguish any semantic difference between an open back unrounded vowel <ɑ>, the \textsc{latin small letter alpha} at U+0251, and the open front unrounded vowel <a>, \textsc{latin small letter a} at U+0061.
+But even among linguists this distinction leads to problems.
+For example, as pointed out by \citet{Mielke2009}, there is a problem stemming from the fact that about 75\% of languages are reported to have a five-vowel system \citep{Maddieson1984}.
+Historically, linguistic descriptions tend not to include precise audio recording and measurements of formants, so this may lead one to ask if the many characters that are used in phonological description reflects a \textsc{transcriptional bias}.
+The common use of in transcriptions could be in part due to the ease of typing the letter on an English keyboard (or for older descriptions, the typewriter).
+We found it to be exceedingly rare that a linguist uses for a low back unrounded vowel.\footnote{One example is \citet[75]{Vidal2001a}, in which the author states: ``The definition of Pilagá /a/ as {[}+back{]} results from its behavior in certain phonological contexts.
 For instance, uvular and pharyngeal consonants only occur around /a/ and /o/.
-Hence, the characterization of /a/ and /o/ as a natural class of (i.e.,
-{[}+back{]} vowels), as opposed to /i/ and /e/.''} They simply use $<$a$>$ as
-long as there is no opposition to $<$ɑ$>$.\footnote{See Thomason's Language Log
-post, ``Why I don't love the International Phonetic Alphabet'', at:
-\url{http://itre.cis.upenn.edu/\textasciitilde{}myl/languagelog/archives/005287.html}.}
-Making things even more problematic, there is an old typographic tradition that
-the double-story uses a single-story in italics. This leads to the unfortunate
-effect that in most well-designed fonts the italics of and use the same glyph.
-If this distinction has to be kept upright in italics, the only solution we can
-currently offer is to use \textit{slanted} glyphs (i.e.~artificially italicized
-glyphs) instead of real italics (i.e.~special italics glyphs designed by a
-typographer).\footnote{For example, the widely used IPA font Doulos SIL
-(\url{http://scripts.sil.org/cms/scripts/page.php?item\_id=DoulosSIL}) does not
-have real italics. This leads some word-processing software, like Microsoft
-Word, to produce slanted glyphs instead. That particular combination of font and
-software application will thus lead to the desired effect. However, note that
-when the text is transferred to another font (i.e.~one that includes real
-italics) and/or to another software application (like Apple Pages, which does
-not perform slanting), then this visual appearance will be lost. In this case we
-are thus still in the pre-Unicode situation in which the choice of font and
-rendering software actually matters. The ideal solution from a linguistic point
-of view would be the introduction of a new IPA code point for a different kind
-of which explicitly specifies that it should still be rendered as a double-story
-character when italicized. After informal discussion with various Unicode
-players, our impression is that this highly restricted problem is not
-sufficiently urgent to introduce even more <a>-like characters in Unicode (which
-already lead to much confusion, see Section \ref{pitfall-homoglyphs}). This is a
-clear situation in which the Unicode Consortium is not just thinking about
-linguists, but has a more wide-ranging practical view to consider.}
+Hence, the characterization of /a/ and /o/ as a natural class of (i.e., {[}+back{]} vowels), as opposed to /i/ and /e/.''} They simply use $<$a$>$ as long as there is no opposition to $<$ɑ$>$.\footnote{See Thomason's Language Log post, ``Why I don't love the International Phonetic Alphabet'', at: \url{http://itre.cis.upenn.edu/\textasciitilde{}myl/languagelog/archives/005287.html}.} Making things even more problematic, there is an old typographic tradition that the double-story uses a single-story in italics.
+This leads to the unfortunate effect that in most well-designed fonts the italics of and use the same glyph.
+If this distinction has to be kept upright in italics, the only solution we can currently offer is to use \textit{slanted} glyphs (i.e.~artificially italicized glyphs) instead of real italics (i.e.~special italics glyphs designed by a typographer).\footnote{For example, the widely used IPA font Doulos SIL (\url{http://scripts.sil.org/cms/scripts/page.php?item\_id=DoulosSIL}) does not have real italics.
+This leads some word-processing software, like Microsoft Word, to produce slanted glyphs instead.
+That particular combination of font and software application will thus lead to the desired effect.
+However, note that when the text is transferred to another font (i.e.~one that includes real italics) and/or to another software application (like Apple Pages, which does not perform slanting), then this visual appearance will be lost.
+In this case we are thus still in the pre-Unicode situation in which the choice of font and rendering software actually matters.
+The ideal solution from a linguistic point of view would be the introduction of a new IPA code point for a different kind of which explicitly specifies that it should still be rendered as a double-story character when italicized.
+After informal discussion with various Unicode players, our impression is that this highly restricted problem is not sufficiently urgent to introduce even more <a>-like characters in Unicode (which already lead to much confusion, see Section \ref{pitfall-homoglyphs}).
+This is a clear situation in which the Unicode Consortium is not just thinking about linguists, but has a more wide-ranging practical view to consider.}
 
@@
 \begin{itemize}
-	\item The uses of the apostrophe has led to long discussions on the
-	Unicode Standard email list. An English keyboard inputs
-	\textless{}`\textgreater{} APOSTROPHE at U+0027, although the
-	\textit{preferred} Unicode apostrophe is the \textless{}'\textgreater{}
-	\textsc{right single quotation mark} at U+2019. Yet the glottal
-	stop/glottalization/ejective marker is another completely different
-	character, the \textsc{modifier letter apostrophe} at U+02BC, but
-	unfortunately looks mostly highly similar to U+2019.  \item There is
-	ambiguous encoding of IPA segments within the Unicode Standard. An
-	example is the U+02C1 \textsc{modifier letter reversed glottal stop} vs
-	the U+02E4 \textsc{modifier letter small reversed glottal stop} . Both
-	are denoted in the Unicode Standard as the \textsc{pharyngealized
-	diacritic} and both appear in various resources representing phonetic
-	data online. This is thus an example for which the Unicode Standard does
-	not solve the linguistic standardization problem.  \item There is at
-	least one case in which the character name assigned by the Unicode
-	Consortium does not match the IPA's description: in the Unicode Standard
-	at U+01C3 is labeled \textsc{latin letter retroflex click}, but in IPA
-	is an alveolar or postalveolar click (not retroflex). This naming is
-	probably best seen as a simple error in the Unicode Standard. Note that
-	most linguists simply seem to use or <ou>) while ligatures are single
-	characters (e.g. <ʧ> \textsc{latin small letter tesh digraph} at
-	U+02A7). Ligatures arose in the context of printing easier-to-read
-	texts, and are included in the Unicode Standard for reasons of legacy
-	encoding. However, their usage is discouraged by the Unicode core
-	specification. Specifically related to IPA, various phonetic
-	combinations of characters (typically affricates) are available as
-	single code-points in the Unicode Standard, but are designated as
-	\textsc{ligatures} or \textsc{digraphs} (confusingly both names appear
-	interchangeably). Such glyphs might be used by software to produce a
-	pleasing display, but they should not be hard-coded into the text
-	itself. In the context of IPA, characters like the following ligatures
-	should thus \emph{not} be used. Instead a combination of two characters
-	is preferred: \item <ʣ> \textsc{latin small letter dz digraph} at U+02A3
-	(use <d> + <z> instead) \item <ʧ> \textsc{latin small letter tesh
-	digraph} at U+02A7 (use <t> + <ʃ> instead) \item <ʩ> \textsc{latin small
-	letter feng digraph} at U+02A9 (use <f> + <ŋ> instead)
+	\item The uses of the apostrophe has led to long discussions on the Unicode Standard email list.
+An English keyboard inputs \textless{}`\textgreater{} APOSTROPHE at U+0027, although the \textit{preferred} Unicode apostrophe is the \textless{}'\textgreater{} \textsc{right single quotation mark} at U+2019.
+Yet the glottal stop/glottalization/ejective marker is another completely different character, the \textsc{modifier letter apostrophe} at U+02BC, but unfortunately looks mostly highly similar to U+2019.
+	\item There is ambiguous encoding of IPA segments within the Unicode Standard.
+An example is the U+02C1 \textsc{modifier letter reversed glottal stop} vs the U+02E4 \textsc{modifier letter small reversed glottal stop} .
+Both are denoted in the Unicode Standard as the \textsc{pharyngealized diacritic} and both appear in various resources representing phonetic data online.
+This is thus an example for which the Unicode Standard does not solve the linguistic standardization problem.
+	\item There is at least one case in which the character name assigned by the Unicode Consortium does not match the IPA's description: in the Unicode Standard at U+01C3 is labeled \textsc{latin letter retroflex click}, but in IPA is an alveolar or postalveolar click (not retroflex).
+This naming is probably best seen as a simple error in the Unicode Standard.
+Note that most linguists simply seem to use or <ou>) while ligatures are single characters (e.g.
+<ʧ> \textsc{latin small letter tesh digraph} at U+02A7).
+Ligatures arose in the context of printing easier-to-read texts, and are included in the Unicode Standard for reasons of legacy encoding.
+However, their usage is discouraged by the Unicode core specification.
+Specifically related to IPA, various phonetic combinations of characters (typically affricates) are available as single code-points in the Unicode Standard, but are designated as \textsc{ligatures} or \textsc{digraphs} (confusingly both names appear interchangeably).
+Such glyphs might be used by software to produce a pleasing display, but they should not be hard-coded into the text itself.
+In the context of IPA, characters like the following ligatures should thus \emph{not} be used.
+Instead a combination of two characters is preferred:
+	\item <ʣ> \textsc{latin small letter dz digraph} at U+02A3 (use <d> + <z> instead) 
+	\item <ʧ> \textsc{latin small letter tesh digraph} at U+02A7 (use <t> + <ʃ> instead) 
+	\item <ʩ> \textsc{latin small letter feng digraph} at U+02A9 (use <f> + <ŋ> instead) 
 \end{itemize}
 
-However, there are a few Unicode characters that are historically ligatures, but
-which are today considered as simple characters in the Unicode Standard and thus
-should be used when writing IPA, namely:
+However, there are a few Unicode characters that are historically ligatures, but which are today considered as simple characters in the Unicode Standard and thus should be used when writing IPA, namely:
 
 \begin{itemize}
-	\item <ɮ> \textsc{latin small letter lezh} at U+026E \item <œ>
-	\textsc{latin small ligature oe} at U+0153 \item <ɶ> \textsc{latin
-	letter small capital oe} at U+0276 \item <æ> \textsc{latin small letter
-	ae} at U+00E6
+	\item <ɮ> \textsc{latin small letter lezh} at U+026E 
+	\item <œ> \textsc{latin small ligature oe} at U+0153 
+	\item <ɶ> \textsc{latin letter small capital oe} at U+0276 
+	\item <æ> \textsc{latin small letter ae} at U+00E6 
 \end{itemize}
@@
 
-Another pitfall is diacritics. The problem is that the meaning of the term
-\textsc{diacritics} as used by the IPA is not the same as it used in the Unicode
-Standard. Specifically, diacritics in the IPA-sense are either so-called
-\textsc{Combining Diacritical Marks} or \textsc{Spacing Modifier Letters} in the
-Unicode Standard. Crucially, Combining Diacritical Marks are by definition
-combined with the character before them (to form so-called default grapheme
-clusters, see Section 3). In contrast, Spacing Modifier Letters are by
-definition \emph{not} combined into grapheme clusters with the preceding
-character, but simply treated as separate letters. In the context of the IPA,
-the following IPA-diacritics are actually Spacing Modifier Letters in the
-Unicode Standard:
+Another pitfall is diacritics.
+The problem is that the meaning of the term \textsc{diacritics} as used by the IPA is not the same as it used in the Unicode Standard.
+Specifically, diacritics in the IPA-sense are either so-called \textsc{Combining Diacritical Marks} or \textsc{Spacing Modifier Letters} in the Unicode Standard.
+Crucially, Combining Diacritical Marks are by definition combined with the character before them (to form so-called default grapheme clusters, see Section 3).
+In contrast, Spacing Modifier Letters are by definition \emph{not} combined into grapheme clusters with the preceding character, but simply treated as separate letters.
+In the context of the IPA, the following IPA-diacritics are actually Spacing Modifier Letters in the Unicode Standard:
 
 \begin{itemize}
-	\item Length marks, namely <ː> \textsc{modifier letter triangular colon}
-	at U+02D0 and <ˑ> \textsc{modifier letter half triangular colon} at
-	U+02D1.  \item Tone letters, like <˥> \textsc{modifier letter extra-high
-	tone bar} at U+02E5, and others like this.  \item Superscript letters,
-	like <ʰ> \textsc{modifier letter small h} at U+02B0 or <ˤ>
-	\textsc{modifier letter small reversed glottal stop} at U+02E4, and
-	others like this.  \item <˞> \textsc{modifier letter rhotic hook} at
-	U+02DE.
+	\item Length marks, namely <ː> \textsc{modifier letter triangular colon} at U+02D0 and <ˑ> \textsc{modifier letter half triangular colon} at U+02D1. 
+	\item Tone letters, like <˥> \textsc{modifier letter extra-high tone bar} at U+02E5, and others like this. 
+	\item Superscript letters, like <ʰ> \textsc{modifier letter small h} at U+02B0 or <ˤ> \textsc{modifier letter small reversed glottal stop} at U+02E4, and others like this. 
+	\item <˞> \textsc{modifier letter rhotic hook} at U+02DE. 
 \end{itemize}
 
-Although linguists might expect these characters to belong together with the
-character in front of them, at least for <ʰ> \textsc{modifier letter small h} at
-U+02B0 the Unicode Consortium's decision to treat it as a separate character is
-also linguistically correct, because according to the IPA it can be used both
-for aspiration (more precisely post-aspiration following the base character) and
-pre-aspiration (preceding the base character). Note that there is a mechanism in
-Unicode to force separate characters to be combined (namely by using the
-\textsc{zero width joiner} at U+200D), but this seems to be a rather
-impractical, and probably not enforceable solution to us.
+Although linguists might expect these characters to belong together with the character in front of them, at least for <ʰ> \textsc{modifier letter small h} at U+02B0 the Unicode Consortium's decision to treat it as a separate character is also linguistically correct, because according to the IPA it can be used both for aspiration (more precisely post-aspiration following the base character) and pre-aspiration (preceding the base character).
+Note that there is a mechanism in Unicode to force separate characters to be combined (namely by using the \textsc{zero width joiner} at U+200D), but this seems to be a rather impractical, and probably not enforceable solution to us.
 
@@
 
-Also related to diacritics is the question of ordering. To our knowledge, the
-International Phonetic Association does not specify a specific ordering for
-diacritics that combine with phonetic base symbols; this exercise is left to the
-reasoning of the transcriber. However, such marks have to be explicitly ordered
-if sequences of them are to be interoperable and compatible. An example is a
-labialized aspirated alveolar plosive: <tʷʰ>. There is nothing holding linguists
-back from using <tʰʷ> instead (with exactly the same intended meaning). However,
-from a technical standpoint, these two sequences are different, e.g.~if both
-sequences are used in a document, searching for <tʷʰ> will not find any
-instances of <tʰʷ>, and vice versa. Likewise, a creaky voiced syllabic dental
-nasal can be encoded in various orders, e.g. <n̪̰̩>, <n̩̰̪> or <n̩̪̰>.
+Also related to diacritics is the question of ordering.
+To our knowledge, the International Phonetic Association does not specify a specific ordering for diacritics that combine with phonetic base symbols; this exercise is left to the reasoning of the transcriber.
+However, such marks have to be explicitly ordered if sequences of them are to be interoperable and compatible.
+An example is a labialized aspirated alveolar plosive: <tʷʰ>.
+There is nothing holding linguists back from using <tʰʷ> instead (with exactly the same intended meaning).
+However, from a technical standpoint, these two sequences are different, e.g.~if both sequences are used in a document, searching for <tʷʰ> will not find any instances of <tʰʷ>, and vice versa.
+Likewise, a creaky voiced syllabic dental nasal can be encoded in various orders, e.g.
+<n̪̰̩>, <n̩̰̪> or <n̩̪̰>.
 
-In accordance with the absence of any specification of ordering in the IPA, the
-Unicode Standard likewise does not propose any standard orderings. Both leave it
-to the user to be consistent. However, there is one aspect of ordering for which
-the Unicode Standard does present a canonical solution, which is uncontroversial
-from a linguistic perspective. Diacritics in the Unicode Standard
-(i.e.~Combining Diacritical Marks, see above) are classified in Canonical
-Combining Classes. In practice, the diacritics are distinguished by their
-position relative to the base character.\footnote{See
-\url{http://unicode.org/reports/tr44/\#Canonical\_Combining\_Class\_Values for a
-detailed description}.} When applying a Unicode normalization (NFC or NFD, see
-previous section), the diacritics in different positions are put in a specified
-order. This process therefore harmonizes the difference between different
-encodings, for example, of a midtone creaky voice vowel <ḛ̄>. This grapheme
-cluster can be encoded either as <e> + <̄> + <̰> or as <e> + <̰> + <̄> . To
-prevent this twofold encoding, the Unicode Standard specifies the second
-ordering as canonical (diacritics below before diacritics above).
+In accordance with the absence of any specification of ordering in the IPA, the Unicode Standard likewise does not propose any standard orderings.
+Both leave it to the user to be consistent.
+However, there is one aspect of ordering for which the Unicode Standard does present a canonical solution, which is uncontroversial from a linguistic perspective.
+Diacritics in the Unicode Standard (i.e.~Combining Diacritical Marks, see above) are classified in Canonical Combining Classes.
+In practice, the diacritics are distinguished by their position relative to the base character.\footnote{See \url{http://unicode.org/reports/tr44/\#Canonical\_Combining\_Class\_Values for a detailed description}.} When applying a Unicode normalization (NFC or NFD, see previous section), the diacritics in different positions are put in a specified order.
+This process therefore harmonizes the difference between different encodings, for example, of a midtone creaky voice vowel <ḛ̄>.
+This grapheme cluster can be encoded either as <e> + <̄> + <̰> or as <e> + <̰> + <̄> .
+To prevent this twofold encoding, the Unicode Standard specifies the second ordering as canonical (diacritics below before diacritics above).
 
-When encoding a string according to the Unicode Standard, it is possible to do
-this either using the NFC (composition) or NFD (decomposition) normalization.
-Decomposition implies that precomposed characters (like <á> \textsc{latin small
-letter a with acute} at U+00E1) will be split into its parts. This might sound
-preferable for a linguistic analysis, as the different diacritics are separated
-from the base characters. However, note that most attached elements like strokes
-(e.g.~in the <ɨ>), retroflex hooks (e.g.~in <ʐ>) or rhotic hooks (e.g.~in <ɝ>)
-will not be decomposed, but strangely enough a cedilla (like in <ç>) will be
-decomposed. In general, Unicode decomposition does not behave like a feature
-decomposition as expected from a linguistic perspective. It is thus important to
-consider Unicode decomposition only as a technical procedure, and not assume
-that it is linguistically sensible.
+When encoding a string according to the Unicode Standard, it is possible to do this either using the NFC (composition) or NFD (decomposition) normalization.
+Decomposition implies that precomposed characters (like <á> \textsc{latin small letter a with acute} at U+00E1) will be split into its parts.
+This might sound preferable for a linguistic analysis, as the different diacritics are separated from the base characters.
+However, note that most attached elements like strokes (e.g.~in the <ɨ>), retroflex hooks (e.g.~in <ʐ>) or rhotic hooks (e.g.~in <ɝ>) will not be decomposed, but strangely enough a cedilla (like in <ç>) will be decomposed.
+In general, Unicode decomposition does not behave like a feature decomposition as expected from a linguistic perspective.
+It is thus important to consider Unicode decomposition only as a technical procedure, and not assume that it is linguistically sensible.
 
-Facing the problem of specifying a consistent ordering of diacritics while
-developing a large database of phonological inventories from the world's
-languages, \citet[540]{Moran2012} defines a set of diacritic ordering
-conventions.\footnote{The most recent version of these conventions is online:
-\url{http://phoible.github.io/conventions/}} The conventions are influenced by
-the linguistic literature, though some ad-hoc decisions had to be taken. The
-goal was to explicitly define all character sequences so that the vast variety
-of phonemes found in descriptions of the world's language were normalized into
-consistent character sequences, e.g.~if one language description uses and
-another , when both are intended to be phonetically equivalent, then a decision
-to normalize to one form was taken. For example, when a character sequence
-contains more than one character in Spacing Modifier Letters, the order that is
-proposed is the following (where <\textbar{}> indicates \textit{or} and <→>
-indicates \textit{precedes}):
+Facing the problem of specifying a consistent ordering of diacritics while developing a large database of phonological inventories from the world's languages, \citet[540]{Moran2012} defines a set of diacritic ordering conventions.\footnote{The most recent version of these conventions is online: \url{http://phoible.github.io/conventions/}} The conventions are influenced by the linguistic literature, though some ad-hoc decisions had to be taken.
+The goal was to explicitly define all character sequences so that the vast variety of phonemes found in descriptions of the world's language were normalized into consistent character sequences, e.g.~if one language description uses and another , when both are intended to be phonetically equivalent, then a decision to normalize to one form was taken.
+For example, when a character sequence contains more than one character in Spacing Modifier Letters, the order that is proposed is the following (where <\textbar{}> indicates \textit{or} and <→> indicates \textit{precedes}):
 
 \begin{itemize}
-	\itemsep1pt\parskip0pt\parsep0pt \item \textsc{Spacing Modifier Letters
-	ordering:} ( unreleased <̚> \textbar{} lateral release <ˡ> \textbar{}
-	nasal release <ⁿ>) → ( palatalized <ʲ>) → ( labialized <ʷ>) → (
-	velarized <ˠ>) → ( pharyngealized <ˤ>) → ( aspirated <ʰ> \textbar{}
-	ejective <ʼ>) → ( long <ː> \textbar{} half-long <ˑ>)
+	\itemsep1pt\parskip0pt\parsep0pt 
+	\item \textsc{Spacing Modifier Letters ordering:} ( unreleased <̚> \textbar{} lateral release <ˡ> \textbar{} nasal release <ⁿ>) → ( palatalized <ʲ>) → ( labialized <ʷ>) → ( velarized <ˠ>) → ( pharyngealized <ˤ>) → ( aspirated <ʰ> \textbar{} ejective <ʼ>) → ( long <ː> \textbar{} half-long <ˑ>)
 \end{itemize}
 
-If a character sequence contains more than one diacritic below the base
-character, then the place feature is applied first (dental, laminal, apical,
-fronted, backed, lowered, raised), then the laryngeal setting (voiced,
-voiceless, creaky voice, breathy voice) and finally the syllabic or non-syllabic
-marker (for vowels, ATR gets put on between the place and laryngeal setting).
+If a character sequence contains more than one diacritic below the base character, then the place feature is applied first (dental, laminal, apical, fronted, backed, lowered, raised), then the laryngeal setting (voiced, voiceless, creaky voice, breathy voice) and finally the syllabic or non-syllabic marker (for vowels, ATR gets put on between the place and laryngeal setting).
 So:
@@
 \begin{itemize}
-	\itemsep1pt\parskip0pt\parsep0pt \item \textsc{Combining Diacritical
-	Marks (below) ordering:} ( dental <t̪> \textbar{} laminal <t̻>
-	\textbar{} apical <t̺>) → ( fronted <u̟> \textbar{} backed <e̠>) →(
-	lowered <e̞> \textbar{} raised <e̝>) → ( ATR <e̘ e̙>) → ( voiced <s̬>
-	\textbar{} voiceless <n̥> \textbar{} creaky voice <b̰> \textbar{}
-	breathy voice <b̤>) → ( syllabic <n̩> \textbar{} non-syllabic <e̯>)
+	\itemsep1pt\parskip0pt\parsep0pt 
+	\item \textsc{Combining Diacritical Marks (below) ordering:} ( dental <t̪> \textbar{} laminal <t̻> \textbar{} apical <t̺>) → ( fronted <u̟> \textbar{} backed <e̠>) →( lowered <e̞> \textbar{} raised <e̝>) → ( ATR <e̘ e̙>) → ( voiced <s̬> \textbar{} voiceless <n̥> \textbar{} creaky voice <b̰> \textbar{} breathy voice <b̤>) → ( syllabic <n̩> \textbar{} non-syllabic <e̯>)
 \end{itemize}
 
-Character sequences with diacritics above the base character were not
-problematic in \citet{Moran2012} because they include only the centralized,
-mid-centralized and nasalized combining characters. \citet{Moran2012} marks
-tones as singletons with Space Modifier Letters, e.g. \textless{}˦\textgreater{}
-for a phonemic high tone, instead of accent diacritics, alleviating potential
-conflicts. Building on the work of \citet{Moran2012}, if a character sequence
-contains more than one diacritic above the base character, we propose:
+Character sequences with diacritics above the base character were not problematic in \citet{Moran2012} because they include only the centralized, mid-centralized and nasalized combining characters.
+\citet{Moran2012} marks tones as singletons with Space Modifier Letters, e.g.
+\textless{}˦\textgreater{} for a phonemic high tone, instead of accent diacritics, alleviating potential conflicts.
+Building on the work of \citet{Moran2012}, if a character sequence contains more than one diacritic above the base character, we propose:
 
 \begin{itemize}
-	\itemsep1pt\parskip0pt\parsep0pt \item \textsc{Combining Diacritical
-	Marks (above) ordering:} (centralized <ë> \textbar{} mid-centralized
-	<e̽>) → (extra short <ĕ>) →( tone accents, e.g. <è> ) → ( Spacing
-	Modifier Letters ) → ( tone letters, e.g. <e˦>)
+	\itemsep1pt\parskip0pt\parsep0pt 
+	\item \textsc{Combining Diacritical Marks (above) ordering:} (centralized <ë> \textbar{} mid-centralized <e̽>) → (extra short <ĕ>) →( tone accents, e.g.
+<è> ) → ( Spacing Modifier Letters ) → ( tone letters, e.g.
+<e˦>)
 \end{itemize}

### 2015-06-05 chapters/terminology.tex
reverting this morning's width adjustments

@@
-\chapter{Terminology} \label{terminology}
+\chapter{Terminology}
+\label{terminology}
 
@@
 
-In this section we provide a very brief overview of the linguistic terminology
-concerning writing systems before turning to the slightly different
-computational terminology in the next section on the Unicode Standard.
+In this section we provide a very brief overview of the linguistic terminology concerning writing systems before turning to the slightly different computational terminology in the next section on the Unicode Standard.
 
-Linguistically speaking, a \textsc{writing system} is a symbolic system that
-uses visible or tactile signs to represent language in a systematic way. The
-term \textsc{writing system} has two mutually exclusive meanings. First, it may
-refer to the way a particular language is written. In this sense the term refers
-to the writing system of a particular language, as, for example, in \emph{``the
-Serbian writing system uses two scripts: Latin and Cyrillic''}. Second, the term
-\textsc{writing system} may also refer to a type of symbolic system as used
-among the world's languages to represent the language, as, for example, in
-\emph{``alphabetic writing system''}. In this latter sense the term refers to
-how scripts have been classified according to the way that they encode language,
-as in, for example, \emph{``the Latin and Cyrillic scripts are both alphabetic
-writing systems''}. To avoid confusion, this second notion of \textsc{writing
-system} would more aptly have been called \textsc{script system}.
+Linguistically speaking, a \textsc{writing system} is a symbolic system that uses visible or tactile signs to represent language in a systematic way.
+The term \textsc{writing system} has two mutually exclusive meanings.
+First, it may refer to the way a particular language is written.
+In this sense the term refers to the writing system of a particular language, as, for example, in \emph{``the Serbian writing system uses two scripts: Latin and Cyrillic''}.
+Second, the term \textsc{writing system} may also refer to a type of symbolic system as used among the world's languages to represent the language, as, for example, in \emph{``alphabetic writing system''}.
+In this latter sense the term refers to how scripts have been classified according to the way that they encode language, as in, for example, \emph{``the Latin and Cyrillic scripts are both alphabetic writing systems''}.
+To avoid confusion, this second notion of \textsc{writing system} would more aptly have been called \textsc{script system}.
 
-The term \textsc{script} refers to a collection of distinct symbols as employed
-by one or more orthographies. For example, both Serbian and Russian are written
-with subsets of the Cyrillic script. A single language, like Serbian or
-Japanese, can also be written using orthographies based on different scripts.
-Over the years linguists have typologized script systems in a variety of ways,
-with the tripartite classification of logographic, syllabic, and alphabetic
-remaining the most popular, even though there are at least half a dozen
-different types of script systems that can be distinguished
-\citep{Daniels1990,Daniels1996}.
+The term \textsc{script} refers to a collection of distinct symbols as employed by one or more orthographies.
+For example, both Serbian and Russian are written with subsets of the Cyrillic script.
+A single language, like Serbian or Japanese, can also be written using orthographies based on different scripts.
+Over the years linguists have typologized script systems in a variety of ways, with the tripartite classification of logographic, syllabic, and alphabetic remaining the most popular, even though there are at least half a dozen different types of script systems that can be distinguished \citep{Daniels1990,Daniels1996}.
 
-Only focussing here on the first sense of writing systems, we distinguish two
-different kinds of writing systems used to write a particular language, namely
-transcriptions and orthographies. First, \textsc{transcription} is a scientific
-procedure (and also the result of that procedure) for graphically representing
-the sounds of human speech at the phonetic level. It incorporates a set of
-unambiguous symbols to represent speech sounds, including conventions that
-specify how these symbols should be combined. A transcription system is a
-specific system of symbols and rules used for transcription of the sounds of a
-spoken language variety. In principle, a transcription system should be
-language-independent, in that it should be applicable to all spoken human
-languages. The International Phonetic Alphabet (IPA) is a commonly used
-transcription system that provides a medium for transcribing languages at the
-phonetic level. However, there is a long history of alternative kinds of
-transcription systems (see \citet{Kemp1994}) and today various alternatives are
-in widespread use (e.g.~X-SAMPA and Cyrillic-based phonetic transcription
-systems).\footnote{Most users of IPA do not follow the standard explicitly and
-many dialects based on it have emerged, e.g.~the Africanist and Americanist
-transcription systems.} Note that IPA symbols are also often used to represent
-language on a phonemic level. However, it is important to realize that in this
-usage the IPA symbols are not a transcription system, but rather an orthography
-(though with strong links to the pronunciation). Further, a transcription system
-does not need to be highly detailed, but it can also be a system of broad sound
-classes. Although such an approximative transcription is not normally used in
-linguistics, it is widespread in technological approaches
-(e.g.~\citet{SoundexBeiderMorse2010}\footnote{\url{http://stevemorse.org/phonetics/bmpm.htm},
-\url{http://stevemorse.org/phonetics/bmpm2.htm}} and it is sometimes fruitfully
-used in automatic approaches to historical linguistics
-\citep{Dolgopolsky1986,ASJP16}.\footnote{\url{http://asjp.clld.org/}}
+Only focussing here on the first sense of writing systems, we distinguish two different kinds of writing systems used to write a particular language, namely transcriptions and orthographies.
+First, \textsc{transcription} is a scientific procedure (and also the result of that procedure) for graphically representing the sounds of human speech at the phonetic level.
+It incorporates a set of unambiguous symbols to represent speech sounds, including conventions that specify how these symbols should be combined.
+A transcription system is a specific system of symbols and rules used for transcription of the sounds of a spoken language variety.
+In principle, a transcription system should be language-independent, in that it should be applicable to all spoken human languages.
+The International Phonetic Alphabet (IPA) is a commonly used transcription system that provides a medium for transcribing languages at the phonetic level.
+However, there is a long history of alternative kinds of transcription systems (see \citet{Kemp1994}) and today various alternatives are in widespread use (e.g.~X-SAMPA and Cyrillic-based phonetic transcription systems).\footnote{Most users of IPA do not follow the standard explicitly and many dialects based on it have emerged, e.g.~the Africanist and Americanist transcription systems.} Note that IPA symbols are also often used to represent language on a phonemic level.
+However, it is important to realize that in this usage the IPA symbols are not a transcription system, but rather an orthography (though with strong links to the pronunciation).
+Further, a transcription system does not need to be highly detailed, but it can also be a system of broad sound classes.
+Although such an approximative transcription is not normally used in linguistics, it is widespread in technological approaches (e.g.~\citet{SoundexBeiderMorse2010}\footnote{\url{http://stevemorse.org/phonetics/bmpm.htm}, \url{http://stevemorse.org/phonetics/bmpm2.htm}} and it is sometimes fruitfully used in automatic approaches to historical linguistics \citep{Dolgopolsky1986,ASJP16}.\footnote{\url{http://asjp.clld.org/}}
 
-Second, an \textsc{orthography} specifies the symbols, punctuations, and the
-rules in which a specific language is written in a standardized way.
-Orthographies are often based on a phonemic analysis, but they almost always
-include idiosyncrasies because of historical developments (like sound changes or
-loans) and because of the widely-followed principle of lexical integrity
-(i.e.~the attempt to write the same lexical root in a consistent way, also when
-synchronic phonemic rules change the pronunciation, for example with final
-devoicing in many Germanic languages). All orthographies are language-specific
-(and often even resource-specific), although individual symbols or rules might
-be shared between languages.
+Second, an \textsc{orthography} specifies the symbols, punctuations, and the rules in which a specific language is written in a standardized way.
+Orthographies are often based on a phonemic analysis, but they almost always include idiosyncrasies because of historical developments (like sound changes or loans) and because of the widely-followed principle of lexical integrity (i.e.~the attempt to write the same lexical root in a consistent way, also when synchronic phonemic rules change the pronunciation, for example with final devoicing in many Germanic languages).
+All orthographies are language-specific (and often even resource-specific), although individual symbols or rules might be shared between languages.
 
-A \textsc{practical orthography} is a strongly phoneme-based writing system
-designed for practical use by speakers. The mapping relation between phonemes
-and graphemes in practical orthographies is purposely shallow, i.e.~there is
-mostly a systematic and faithful mapping from a phoneme to a grapheme. Practical
-orthographies are intended to jumpstart written materials development by
-correlating a writing system with the sound units of a language
-(cf.~\citet{MeinhofJones1928}). Symbols from the IPA are often used by linguists
-in the development of practical orthographies for languages without writing
-systems, though this usage of IPA symbols should not be confused with
-transcription (as defined above).
+A \textsc{practical orthography} is a strongly phoneme-based writing system designed for practical use by speakers.
+The mapping relation between phonemes and graphemes in practical orthographies is purposely shallow, i.e.~there is mostly a systematic and faithful mapping from a phoneme to a grapheme.
+Practical orthographies are intended to jumpstart written materials development by correlating a writing system with the sound units of a language (cf.~\citet{MeinhofJones1928}).
+Symbols from the IPA are often used by linguists in the development of practical orthographies for languages without writing systems, though this usage of IPA symbols should not be confused with transcription (as defined above).
 
-Further, a \textsc{transliteration} is a mapping between two different
-orthographies. It is the process of ``recording the graphic symbols of one
-writing system in terms of the corresponding graphic symbols of a second writing
-system'' \citep[396]{Kemp2006}. In straightforward cases, such a transliteration
-is simply a matter of replacing one symbol with another. However, there are
-widespread complications, like one-to-many or many-to-many mappings, which are
-not always easy, or even possible, to solve without listing all cases
-individually (cf.~\citet[chp 2]{Moran2012}).
+Further, a \textsc{transliteration} is a mapping between two different orthographies.
+It is the process of ``recording the graphic symbols of one writing system in terms of the corresponding graphic symbols of a second writing system'' \citep[396]{Kemp2006}.
+In straightforward cases, such a transliteration is simply a matter of replacing one symbol with another.
+However, there are widespread complications, like one-to-many or many-to-many mappings, which are not always easy, or even possible, to solve without listing all cases individually (cf.~\citet[chp 2]{Moran2012}).
 
-Breaking it down further, a script consists of \textsc{graphemes}, and graphemes
-consist of \textsc{characters}. In the linguistic terminology of writing
-systems, a \textsc{character} is a general term for any self-contained element
-in a writing system.\footnote{There is a second interpretation of the term
-\textsc{character}, i.e.~a conventional term for a unit in the Chinese writing
-system \citep{Daniels1996}. This interpretation will not be further explored in
-this paper.} Although in literate societies most people have a strong intuition
-about what the characters are in their particular orthography or orthographies,
-it turns out that the separation of an orthography into separate characters is
-far from trivial. The widespread intuitive notion of a character is strongly
-biased towards educational traditions, like the alphabet taught at schools, and
-technological possibilities, like the available type pieces in a printer's job
-case, the keys on a typewriter, or the symbols displayed in Microsoft Word's
-symbol browser. In practice, characters often consist of multiple building
-blocks, each of which could be considered a character in its own right. For
-example, although a Chinese character may be considered to be a single basic
-unanalyzable unit, at a more fine-grained level of analysis the internal
-structure of Chinese characters is often comprised of smaller semantic and
-phonetic units that should be considered characters \citep{Sproat2000}. In
-alphabetic scripts, this problem is most forcefully exemplified by diacritics.
+Breaking it down further, a script consists of \textsc{graphemes}, and graphemes consist of \textsc{characters}.
+In the linguistic terminology of writing systems, a \textsc{character} is a general term for any self-contained element in a writing system.\footnote{There is a second interpretation of the term \textsc{character}, i.e.~a conventional term for a unit in the Chinese writing system \citep{Daniels1996}.
+This interpretation will not be further explored in this paper.} Although in literate societies most people have a strong intuition about what the characters are in their particular orthography or orthographies, it turns out that the separation of an orthography into separate characters is far from trivial.
+The widespread intuitive notion of a character is strongly biased towards educational traditions, like the alphabet taught at schools, and technological possibilities, like the available type pieces in a printer's job case, the keys on a typewriter, or the symbols displayed in Microsoft Word's symbol browser.
+In practice, characters often consist of multiple building blocks, each of which could be considered a character in its own right.
+For example, although a Chinese character may be considered to be a single basic unanalyzable unit, at a more fine-grained level of analysis the internal structure of Chinese characters is often comprised of smaller semantic and phonetic units that should be considered characters \citep{Sproat2000}.
+In alphabetic scripts, this problem is most forcefully exemplified by diacritics.
 
-A \textsc{diacritic} is a mark, or series of marks, that may be above, below, or
-through other characters \citep{Gaultney2002}. Diacritics are sometimes used to
-distinguish homophonous words, but they are more often used to indicate a
-modified pronunciation \citep[xli]{DanielsBright1996}. The central question is
-whether, for example, <e>, <è>, <a> and <à> should be considered four
-characters, or different combinations of three characters. In general, multiple
-characters together can form another character, and it is not always possible to
-decide on principled grounds what should be the basic building blocks of an
-orthography.
+A \textsc{diacritic} is a mark, or series of marks, that may be above, below, or through other characters \citep{Gaultney2002}.
+Diacritics are sometimes used to distinguish homophonous words, but they are more often used to indicate a modified pronunciation \citep[xli]{DanielsBright1996}.
+The central question is whether, for example, <e>, <è>, <a> and <à> should be considered four characters, or different combinations of three characters.
+In general, multiple characters together can form another character, and it is not always possible to decide on principled grounds what should be the basic building blocks of an orthography.
 
-For that reason, it is better to analyze an orthography as a collection of
-graphemes. A \textsc{grapheme} is the basic, minimally distinctive symbol of a
-particular writing system, alike to the phoneme is an abstract representation of
-a distinct sound in a specific language. The term \textsc{grapheme} was modeled
-after the term \textsc{phoneme} and represents a contrastive graphical unit in a
-writing system (see \citet{Kohrt1986} for a historical overview of the term
-grapheme). Most importantly, a single grapheme regularly consists of multiple
-characters, like <th>, <ou> and <gh> in English (note that each character in
-these graphemes is also a separate grapheme in English). Such complex graphemes
-are often used to represent single phonemes. So, a combination of characters is
-used to represent a single phoneme. Note that the opposite is also found in
-writing systems, in cases in which a single character represents a combination
-of two or more phonemes. For example, <x> in English orthography represents a
-combination of the phonemes /k/ and /s/.
+For that reason, it is better to analyze an orthography as a collection of graphemes.
+A \textsc{grapheme} is the basic, minimally distinctive symbol of a particular writing system, alike to the phoneme is an abstract representation of a distinct sound in a specific language.
+The term \textsc{grapheme} was modeled after the term \textsc{phoneme} and represents a contrastive graphical unit in a writing system (see \citet{Kohrt1986} for a historical overview of the term grapheme).
+Most importantly, a single grapheme regularly consists of multiple characters, like <th>, <ou> and <gh> in English (note that each character in these graphemes is also a separate grapheme in English).
+Such complex graphemes are often used to represent single phonemes.
+So, a combination of characters is used to represent a single phoneme.
+Note that the opposite is also found in writing systems, in cases in which a single character represents a combination of two or more phonemes.
+For example, <x> in English orthography represents a combination of the phonemes /k/ and /s/.
 
-Further, conditioned or free variants of a grapheme are called
-\textsc{allographs}. For example, the distinctive forms of Greek sigma are
-conditioned, with <σ> being used word-internally and <ς> being used at the end
-of a word. In sum, there are many-to-many relationships between phonemes and
-graphemes as they are expressed in the myriad of language- and resource-specific
-orthographies.
+Further, conditioned or free variants of a grapheme are called \textsc{allographs}.
+For example, the distinctive forms of Greek sigma are conditioned, with <σ> being used word-internally and <ς> being used at the end of a word.
+In sum, there are many-to-many relationships between phonemes and graphemes as they are expressed in the myriad of language- and resource-specific orthographies.
 
-This exposition of the linguistic terminology involved in describing writing
-systems has been purposely brief. We have highlighted some of the linguistic
-notions that are pertinent, yet sometimes confused with, the technological
-definitions developed for the computational processing of the world's writing
-systems, which we describe in the next section.
+This exposition of the linguistic terminology involved in describing writing systems has been purposely brief.
+We have highlighted some of the linguistic notions that are pertinent, yet sometimes confused with, the technological definitions developed for the computational processing of the world's writing systems, which we describe in the next section.
 
-\section{The Unicode approach} \label{the-unicode-approach}
+\section{The Unicode approach}
+\label{the-unicode-approach}
 
-The conceptualization and terminology of writing systems was rejuvenated through
-the development of the Unicode Standard, with major input from Mark Davis,
-co-founder and long-term president of the Unicode Consortium. For many years,
-\textit{exotic} writing systems and phonetic transcription systems on personal
-computers were constrained by the American Standard Code for Information
-Interchange (ASCII) character encoding scheme (based on the Latin script), which
-only allowed for a strongly limited number of different symbols to be encoded.
-This implied that users could either use and adopt the (extended) Latin alphabet
-or they could assign new symbols to the small number of code points in the ASCII
-encoding scheme to be rendered by a specifically designed font
-\citep{BirdSimons2003}. In this situation, it was necessary to specify the font
-together with each document to ensure the rightful display of its content. To
-alleviate this problem of assigning different symbols to the same code points,
-in the late 80's and early 90's the Unicode Consortium set itself the ambitious
-goal of developing a single universal character encoding to provide a unique
-number, a code point, for every character in the world's writing systems.
-Nowadays, the Unicode Standard is the default encoding of the technologies that
-support the World Wide Web and for all modern operating systems, software and
-programming languages.
+The conceptualization and terminology of writing systems was rejuvenated through the development of the Unicode Standard, with major input from Mark Davis, co-founder and long-term president of the Unicode Consortium.
+For many years, \textit{exotic} writing systems and phonetic transcription systems on personal computers were constrained by the American Standard Code for Information Interchange (ASCII) character encoding scheme (based on the Latin script), which only allowed for a strongly limited number of different symbols to be encoded.
+This implied that users could either use and adopt the (extended) Latin alphabet or they could assign new symbols to the small number of code points in the ASCII encoding scheme to be rendered by a specifically designed font \citep{BirdSimons2003}.
+In this situation, it was necessary to specify the font together with each document to ensure the rightful display of its content.
+To alleviate this problem of assigning different symbols to the same code points, in the late 80's and early 90's the Unicode Consortium set itself the ambitious goal of developing a single universal character encoding to provide a unique number, a code point, for every character in the world's writing systems.
+Nowadays, the Unicode Standard is the default encoding of the technologies that support the World Wide Web and for all modern operating systems, software and programming languages.
 
-The Unicode Standard represents a massive step forward because it aims to
-eradicate the distinction between \textsc{universal} (ASCII) versus
-\textsc{language/resource-particular} (Font) by adding as much as possible
-language-specific information into the universal standard. However, there are
-still language/resource-specific specifications necessary for the proper usage
-of Unicode, as will be discussed below. Within the Unicode structure many of
-these specifications can be captured by so-called \textsc{Locales} or
-\textsc{Locale Descriptions}, so we are moving to a new distinction of
-\textsc{universal} (Unicode Standard) versus \textsc{language-particular}
-(Locale Description). The major gain is a much larger compatibility on the
-universal level (because Unicode standardizes a much larger portion of writing
-system diversity), and much better possibilities for automatic processing on the
-language-particular level (because Locale Descriptions are computer readable
-specifications).
+The Unicode Standard represents a massive step forward because it aims to eradicate the distinction between \textsc{universal} (ASCII) versus \textsc{language/resource-particular} (Font) by adding as much as possible language-specific information into the universal standard.
+However, there are still language/resource-specific specifications necessary for the proper usage of Unicode, as will be discussed below.
+Within the Unicode structure many of these specifications can be captured by so-called \textsc{Locales} or \textsc{Locale Descriptions}, so we are moving to a new distinction of \textsc{universal} (Unicode Standard) versus \textsc{language-particular} (Locale Description).
+The major gain is a much larger compatibility on the universal level (because Unicode standardizes a much larger portion of writing system diversity), and much better possibilities for automatic processing on the language-particular level (because Locale Descriptions are computer readable specifications).
 
-Each version of the Unicode Standard (as of writing at version 7) consists of a
-set of specifications and guidelines that include (i) a core specification, (ii)
-code charts, (iii) standard annexes and (iv) a character database.\footnote{All
-documents of the Unicode Standard are available at
-\url{http://www.unicode.org/versions/latest/}. For a quick survey of the use of
-terminology inside the Unicode Standard, their glossary is particularly useful,
-available at \url{http://www.unicode.org/glossary/}. For a general introduction
-to the principles of Unicode, Chapter 2 of the core specification, called
-\textsc{General Structure}, is particularly insightful. Different from many
-other documents of the Unicode Standard, this general introduction is relatively
-easy to read and illustrated with many interesting examples from various
-orthography traditions all over the world.} The \textsc{core specification} is a
-book directed toward human readers that describes the formal standard for
-encoding multilingual text. The \textsc{code charts} provide a humanly readable
-online reference to the character contents of the Unicode Standard in the form
-of PDF files. The \textsc{Unicode Standard Annexes (UAX)} are a set of technical
-standards that describe the implementation of the Unicode Standard for software
-development, Web standards, programming languages, etc. The \textsc{Unicode
-Character Database (UCD)} is a set of computer-readable text files that describe
-the character properties, including a set of rich character and writing system
-semantics, for each character in the Unicode Standard. In this section, we
-introduce the basic Unicode concepts, but we will leave out many details. Please
-consult the above mentioned full documentation for a more detailed discussion.
-Further note that the Unicode Standard is exactly that, namely a standard. It
-normatively describes notions and rules to be followed. In the actual practice
-of applying this standard in a computational setting, a specific implementation
-is necessary. The most widely used implementation of the Unicode Standard is the
-\textsc{International Components for Unicode (ICU)}, which offers C/C++ and Java
-libraries trying to implement to Unicode
-Standard.\footnote{\url{http://icu-project.org}}
+Each version of the Unicode Standard (as of writing at version 7) consists of a set of specifications and guidelines that include (i) a core specification, (ii) code charts, (iii) standard annexes and (iv) a character database.\footnote{All documents of the Unicode Standard are available at \url{http://www.unicode.org/versions/latest/}.
+For a quick survey of the use of terminology inside the Unicode Standard, their glossary is particularly useful, available at \url{http://www.unicode.org/glossary/}.
+For a general introduction to the principles of Unicode, Chapter 2 of the core specification, called \textsc{General Structure}, is particularly insightful.
+Different from many other documents of the Unicode Standard, this general introduction is relatively easy to read and illustrated with many interesting examples from various orthography traditions all over the world.} The \textsc{core specification} is a book directed toward human readers that describes the formal standard for encoding multilingual text.
+The \textsc{code charts} provide a humanly readable online reference to the character contents of the Unicode Standard in the form of PDF files.
+The \textsc{Unicode Standard Annexes (UAX)} are a set of technical standards that describe the implementation of the Unicode Standard for software development, Web standards, programming languages, etc.
+The \textsc{Unicode Character Database (UCD)} is a set of computer-readable text files that describe the character properties, including a set of rich character and writing system semantics, for each character in the Unicode Standard.
+In this section, we introduce the basic Unicode concepts, but we will leave out many details.
+Please consult the above mentioned full documentation for a more detailed discussion.
+Further note that the Unicode Standard is exactly that, namely a standard.
+It normatively describes notions and rules to be followed.
+In the actual practice of applying this standard in a computational setting, a specific implementation is necessary.
+The most widely used implementation of the Unicode Standard is the \textsc{International Components for Unicode (ICU)}, which offers C/C++ and Java libraries trying to implement to Unicode Standard.\footnote{\url{http://icu-project.org}}
 
-The Unicode Standard is a \textsc{character encoding system} which goal it is to
-support the interchange and processing of written characters and text in a
-computational setting. Underlyingly, the character encoding is represented by a
-range of numerical values called a \textsc{code space}, which is used to encode
-a set of characters. A \textsc{code point} is a unique non-negative integer
-within a code space (i.e.~within a certain numerical range). In the Unicode
-Standard character encoding system, an \textsc{abstract character}, for example
-a \textsc{LATIN SMALL LETTER P}, is mapped to a particular code point, such as
-the decimal value 112, normally represented in hexadecimal and in Unicode
-parlance as (U+)0070.\footnote{Hexadecimal (base-16) 0070 is equivalent to
-decimal (base-10) 112, which can be calculated by considering that $(0\cdot16^3)
-+ (0\cdot16^2) + (7\cdot16^1) + (0\cdot16^0) = 7\cdot16 = 112$. Underlyingly,
-computers will of course treat this code point binary (base-2) as 11100000, as
-can be seen by calculating that $(1\cdot2^7) + (1\cdot2^6) + (1\cdot2^5) +
-(0\cdot2^4) + (0\cdot2^3) + (0\cdot2^2) + (0\cdot2^1) + (0\cdot2^0) = 64 + 32 +
-16 = 112$.} That encoded abstract character is rendered on a computer screen (or
-printed page) as a \textsc{glyph}, e.g. <p>, depending on the \textsc{font} and
-the context in which that character appears.
+The Unicode Standard is a \textsc{character encoding system} which goal it is to support the interchange and processing of written characters and text in a computational setting.
+Underlyingly, the character encoding is represented by a range of numerical values called a \textsc{code space}, which is used to encode a set of characters.
+A \textsc{code point} is a unique non-negative integer within a code space (i.e.~within a certain numerical range).
+In the Unicode Standard character encoding system, an \textsc{abstract character}, for example a \textsc{LATIN SMALL LETTER P}, is mapped to a particular code point, such as the decimal value 112, normally represented in hexadecimal and in Unicode parlance as (U+)0070.\footnote{Hexadecimal (base-16) 0070 is equivalent to decimal (base-10) 112, which can be calculated by considering that $(0\cdot16^3) + (0\cdot16^2) + (7\cdot16^1) + (0\cdot16^0) = 7\cdot16 = 112$.
+Underlyingly, computers will of course treat this code point binary (base-2) as 11100000, as can be seen by calculating that $(1\cdot2^7) + (1\cdot2^6) + (1\cdot2^5) + (0\cdot2^4) + (0\cdot2^3) + (0\cdot2^2) + (0\cdot2^1) + (0\cdot2^0) = 64 + 32 + 16 = 112$.} That encoded abstract character is rendered on a computer screen (or printed page) as a \textsc{glyph}, e.g.
+<p>, depending on the \textsc{font} and the context in which that character appears.
 
-In Unicode Standard terminology, an (abstract) \textsc{character} is the basic
-encoding unit. The term \textsc{character} can be quite confusing due to its
-alternative definitions across different scientific disciplines and because in
-general the word \textsc{character} means many different things to different
-people. It is therefore often preferable to refer to Unicode characters simply
-as \textsc{code points}, because there is a one-to-one mapping between Unicode
-characters and their numeric representation. The \textit{Unicode2012} defines a
-character as ``the smallest component of written language that has semantic
-value. It refers to the abstract meaning and/or general shape, rather than a
-specific shape, though in code tables some form of visual representation is
-essential for the reader's understanding.'' Unicode defines characters as
-abstractions of orthographic symbols, and it does not define visualizations for
-these characters (although it does presents examples). In contrast, a
-\textsc{glyph} is a concrete graphical representation of a character as it
-appears when rendered (or rasterized) and displayed on an electronic device or
-on printed paper. For example, <reinsert Gs here> are different glyphs of the
-same character, i.e.~they may be rendered differently depending on the
-typography being used, but they all share the same code point. From the
-perspective of Unicode they are \textit{the same thing}. In this approach, a
-\textsc{font} is then simply a collection of glyphs linked to code points.
-Allography is not specified in Unicode (expect for a few exceptional cases, due
-to legacy encoding issues), but can be specified in a font as a
-\textsc{contextual variant} (aka presentation form).
+In Unicode Standard terminology, an (abstract) \textsc{character} is the basic encoding unit.
+The term \textsc{character} can be quite confusing due to its alternative definitions across different scientific disciplines and because in general the word \textsc{character} means many different things to different people.
+It is therefore often preferable to refer to Unicode characters simply as \textsc{code points}, because there is a one-to-one mapping between Unicode characters and their numeric representation.
+The \textit{Unicode2012} defines a character as ``the smallest component of written language that has semantic value.
+It refers to the abstract meaning and/or general shape, rather than a specific shape, though in code tables some form of visual representation is essential for the reader's understanding.'' Unicode defines characters as abstractions of orthographic symbols, and it does not define visualizations for these characters (although it does presents examples).
+In contrast, a \textsc{glyph} is a concrete graphical representation of a character as it appears when rendered (or rasterized) and displayed on an electronic device or on printed paper.
+For example, <reinsert Gs here> are different glyphs of the same character, i.e.~they may be rendered differently depending on the typography being used, but they all share the same code point.
+From the perspective of Unicode they are \textit{the same thing}.
+In this approach, a \textsc{font} is then simply a collection of glyphs linked to code points.
+Allography is not specified in Unicode (expect for a few exceptional cases, due to legacy encoding issues), but can be specified in a font as a \textsc{contextual variant} (aka presentation form).
 
-Each code point in the Unicode Standard is associated with a set of
-\textsc{character properties} as defined by the Unicode character property
-model.\footnote{The character property model is described in
-\url{http://www.unicode.org/reports/tr23/}, but the actual properties are
-described in \url{http://www.unicode.org/reports/tr44/}. A simplified overview
-of the properties is available at
-\url{http://userguide.icu-project.org/strings/properties}. The actual code
-tables listing all properties for all Unicode code points are available at
-\url{http://www.unicode.org/Public/UCD/latest/ucd/}.} Basically, those
-properties are just a long list of values for each character. For example, code
-point U+0047 has the following properties (among many others): \begin{itemize}
-	\item Name: LATIN CAPITAL LETTER G \item Alphabetic: YES \item
-	Uppercase: YES \item Script: LATIN \item Extender: NO \item
-	Simple\_Lowercase\_Mapping: 0067
+Each code point in the Unicode Standard is associated with a set of \textsc{character properties} as defined by the Unicode character property model.\footnote{The character property model is described in \url{http://www.unicode.org/reports/tr23/}, but the actual properties are described in \url{http://www.unicode.org/reports/tr44/}.
+A simplified overview of the properties is available at \url{http://userguide.icu-project.org/strings/properties}.
+The actual code tables listing all properties for all Unicode code points are available at \url{http://www.unicode.org/Public/UCD/latest/ucd/}.} Basically, those properties are just a long list of values for each character.
+For example, code point U+0047 has the following properties (among many others):
+\begin{itemize}
+	\item Name: LATIN CAPITAL LETTER G 
+	\item Alphabetic: YES 
+	\item Uppercase: YES 
+	\item Script: LATIN 
+	\item Extender: NO 
+	\item Simple\_Lowercase\_Mapping: 0067 
 \end{itemize}
 
-These properties contain the basic information of the Unicode Standard and they
-are necessary to define the correct behavior and conformance required for
-interoperability in and across different software implementations (as defined in
-the Unicode Standard Annexes). The character properties assigned to each code
-point is based on each character's behavior in the real-world writing
-traditions. For example, the corresponding lowercase character to U+0047 is
-U+0067 (though note that the relation between uppercase and lowercase is in many
-situations much more complex than this, and Unicode has further specifications
-for those cases). Another use of properties is to define the script of a
-character. The Unicode Standard defines the term \textsc{script} as, ``A
-collection of letters and other written signs used to represent textual
-information in one or more writing systems. For example, Russian is written with
-a subset of the Cyrillic script; Ukrainian is written with a different subset.
-The Japanese writing system uses several scripts.'' In practice, script is
-simply defined for each character as the explicit \textsc{Script} property in
-the Unicode Character Database.
+These properties contain the basic information of the Unicode Standard and they are necessary to define the correct behavior and conformance required for interoperability in and across different software implementations (as defined in the Unicode Standard Annexes).
+The character properties assigned to each code point is based on each character's behavior in the real-world writing traditions.
+For example, the corresponding lowercase character to U+0047 is U+0067 (though note that the relation between uppercase and lowercase is in many situations much more complex than this, and Unicode has further specifications for those cases).
+Another use of properties is to define the script of a character.
+The Unicode Standard defines the term \textsc{script} as, ``A collection of letters and other written signs used to represent textual information in one or more writing systems.
+For example, Russian is written with a subset of the Cyrillic script; Ukrainian is written with a different subset.
+The Japanese writing system uses several scripts.'' In practice, script is simply defined for each character as the explicit \textsc{Script} property in the Unicode Character Database.
 
-One frequently references property is the \textsc{block} property, which is
-often used in software applications to impose some structure to the large number
-of Unicode characters. Each character in Unicode belongs to a specific block.
-These blocks are basically an organizational structure to alleviate the
-administrative burden of keeping Unicode up-to-date. Blocks consist of
-characters that in some way belong together, so that characters are easier to
-find. Some blocks are connected with a specific script, like the Hebrew block or
-the Gujarati block. However, blocks are predefined ranges of code points, and
-often there will come a point after which the range is completely filled. Any
-extra characters will have to be assigned somewhere else. There is, for example,
-a block \textsc{Arabic}, which contains most Arabic symbols. However, there is
-also a block \textsc{Arabic Supplement}, \textsc{Arabic Presentation Forms-A}
-and \textsc{Arabic Presentation Form B}. The situation with Latin symbols is
-even more extra. In general, the names for block should be taken as a
-definitional statement. For example, many IPA symbols are not located in the
-aptly-names block \textsc{IPA extensions}, but in other blocks (see Section
-\ref{ipa-meets-unicode}).
+One frequently references property is the \textsc{block} property, which is often used in software applications to impose some structure to the large number of Unicode characters.
+Each character in Unicode belongs to a specific block.
+These blocks are basically an organizational structure to alleviate the administrative burden of keeping Unicode up-to-date.
+Blocks consist of characters that in some way belong together, so that characters are easier to find.
+Some blocks are connected with a specific script, like the Hebrew block or the Gujarati block.
+However, blocks are predefined ranges of code points, and often there will come a point after which the range is completely filled.
+Any extra characters will have to be assigned somewhere else.
+There is, for example, a block \textsc{Arabic}, which contains most Arabic symbols.
+However, there is also a block \textsc{Arabic Supplement}, \textsc{Arabic Presentation Forms-A} and \textsc{Arabic Presentation Form B}.
+The situation with Latin symbols is even more extra.
+In general, the names for block should be taken as a definitional statement.
+For example, many IPA symbols are not located in the aptly-names block \textsc{IPA extensions}, but in other blocks (see Section \ref{ipa-meets-unicode}).
 
-There are many cases in which a sequence of characters (i.e.~a sequence of more
-than one code point) represents what a user perceives as an individual unit in a
-particular orthographic writing system. For this reason the Unicode Standard
-differentiates between \textsc{abstract character} and \textsc{user-perceived
-character}. Sequences of multiple code points that correspond to a single
-user-perceived characters are called grapheme clusters in Unicode parlance.
-Grapheme clusters come in two flavors: (default) grapheme clusters and tailored
-grapheme clusters.
+There are many cases in which a sequence of characters (i.e.~a sequence of more than one code point) represents what a user perceives as an individual unit in a particular orthographic writing system.
+For this reason the Unicode Standard differentiates between \textsc{abstract character} and \textsc{user-perceived character}.
+Sequences of multiple code points that correspond to a single user-perceived characters are called grapheme clusters in Unicode parlance.
+Grapheme clusters come in two flavors: (default) grapheme clusters and tailored grapheme clusters.
 
-The (default) \textsc{grapheme clusters} are locale-independent graphemes,
-i.e.~they always apply when a particular combination of characters occurs
-independent of the writing system in which they are used. These character
-combinations are defined in the Unicode Standard as functioning as one
-\textsc{text element}.\footnote{The Unicode Standard defines text element as:
-``A minimum unit of text in relation to a particular text process, in the
-context of a given writing system. In general, the mapping between text elements
-and code points is many-to-many.''} The simplest example of a grapheme cluster
-is a base character followed by a letter modifier character. For example, the
-sequence + <̃\textgreater{} (LATIN SMALL LETTER N at U+006E followed by
-COMBINING TILDE at U+0303) combines visually into , a user-perceived character
-in writing systems like that of Spanish. So, what the user perceives as a single
-character actually involves a multi-code-point sequence. Note that this specific
-sequence can also be represented with a single \textsc{precomposed} code point,
-LATIN SMALL LETTER N WITH TILDE at U+00F1, but this is not the case for all
-multi-code point character sequences. The problem that there multiple encodings
-possible for the same text element has been acknowledged early on in the Unicode
-Standard (e.g.~for , the sequence U+006E U+0303 should in all situations be
-treated identically to the precomposed U+00F1), and a system of
-\textsc{canonical equivalence} is available for such situations. Basically, the
-Unicode Standard offers different kind of normalizations to either decompose all
-precomposed characters (called \textsc{NFD}, \textsc{Normalization Form
-Canonical Decomposition}), or compose as much as possible combinations (called
-\textsc{NFC}, \textsc{Normalization Form Canonical Composition}). In current
-practice of software development, NFC seems to be preferred in most situations
-and is widely proposed as the preferred canonical form.
+The (default) \textsc{grapheme clusters} are locale-independent graphemes, i.e.~they always apply when a particular combination of characters occurs independent of the writing system in which they are used.
+These character combinations are defined in the Unicode Standard as functioning as one \textsc{text element}.\footnote{The Unicode Standard defines text element as: ``A minimum unit of text in relation to a particular text process, in the context of a given writing system.
+In general, the mapping between text elements and code points is many-to-many.''} The simplest example of a grapheme cluster is a base character followed by a letter modifier character.
+For example, the sequence + <̃\textgreater{} (LATIN SMALL LETTER N at U+006E followed by COMBINING TILDE at U+0303) combines visually into , a user-perceived character in writing systems like that of Spanish.
+So, what the user perceives as a single character actually involves a multi-code-point sequence.
+Note that this specific sequence can also be represented with a single \textsc{precomposed} code point, LATIN SMALL LETTER N WITH TILDE at U+00F1, but this is not the case for all multi-code point character sequences.
+The problem that there multiple encodings possible for the same text element has been acknowledged early on in the Unicode Standard (e.g.~for , the sequence U+006E U+0303 should in all situations be treated identically to the precomposed U+00F1), and a system of \textsc{canonical equivalence} is available for such situations.
+Basically, the Unicode Standard offers different kind of normalizations to either decompose all precomposed characters (called \textsc{NFD}, \textsc{Normalization Form Canonical Decomposition}), or compose as much as possible combinations (called \textsc{NFC}, \textsc{Normalization Form Canonical Composition}).
+In current practice of software development, NFC seems to be preferred in most situations and is widely proposed as the preferred canonical form.
 
-More difficult for text processing, because less standardized, is what the
-Unicode Standard terms \textsc{tailored grapheme clusters}. Tailored grapheme
-clusters are locale-dependent graphemes, i.e.~such combination of characters do
-not function as text elements in all situations. For example, the sequence + for
-the Slovak digraph or the sequence in the Sisaala practical orthography
-(pronounced as IPA /tʃ/; \cite{Moran2006}). These grapheme clusters are
-\textsc{tailored} in the sense that they must be specified on a
-language-by-language or a writing system-by-writing system basis. The Unicode
-Standard provides technological specifications for creating locale specific data
-in so-called \textsc{Unicode Locale Descriptions}, i.e.~a set of specification
-that defines a set of language-specific elements (e.g.~tailored grapheme
-clusters, collation order, capitalization-equivalence), as well as other special
-information, like how to format numbers, dates, or currencies. Locale
-descriptions are saved in the \textsc{Common Locale Data Repository
-(CLDR)},\footnote{\url{http://cldr.unicode.org/}} a repository of
-language-specific definitions of writing system properties, each of which
-describes specific usages of characters. Each locale can be encoded in a
-document using the \textsc{Locale Data Markup Language (LDML)}. LDML is an XML
-format and vocabulary for the exchange of structured locale data.
+More difficult for text processing, because less standardized, is what the Unicode Standard terms \textsc{tailored grapheme clusters}.
+Tailored grapheme clusters are locale-dependent graphemes, i.e.~such combination of characters do not function as text elements in all situations.
+For example, the sequence + for the Slovak digraph or the sequence in the Sisaala practical orthography (pronounced as IPA /tʃ/; \cite{Moran2006}).
+These grapheme clusters are \textsc{tailored} in the sense that they must be specified on a language-by-language or a writing system-by-writing system basis.
+The Unicode Standard provides technological specifications for creating locale specific data in so-called \textsc{Unicode Locale Descriptions}, i.e.~a set of specification that defines a set of language-specific elements (e.g.~tailored grapheme clusters, collation order, capitalization-equivalence), as well as other special information, like how to format numbers, dates, or currencies.
+Locale descriptions are saved in the \textsc{Common Locale Data Repository (CLDR)},\footnote{\url{http://cldr.unicode.org/}} a repository of language-specific definitions of writing system properties, each of which describes specific usages of characters.
+Each locale can be encoded in a document using the \textsc{Locale Data Markup Language (LDML)}.
+LDML is an XML format and vocabulary for the exchange of structured locale data.
 
-Unicode Locale Descriptions allow users to define language- or resource-specific
-writing systems or orthographies. The Unicode Consortium defines \textsc{writing
-system} only very loosely, as it is not a central concept in the Unicode
-Standard. A writing system is, ``A set of rules for using one or more scripts to
-write a particular language. Examples include the American English writing
-system, the British English writing system, the French writing system, and the
-Japanese writing system.''. However, there are various drawbacks of locale
-descriptions for the daily practice of linguistic work in a multilingual setting
-(see Section \ref{use-cases}).
+Unicode Locale Descriptions allow users to define language- or resource-specific writing systems or orthographies.
+The Unicode Consortium defines \textsc{writing system} only very loosely, as it is not a central concept in the Unicode Standard.
+A writing system is, ``A set of rules for using one or more scripts to write a particular language.
+Examples include the American English writing system, the British English writing system, the French writing system, and the Japanese writing system.''.
+However, there are various drawbacks of locale descriptions for the daily practice of linguistic work in a multilingual setting (see Section \ref{use-cases}).

### 2015-06-06 chapters/pitfalls.tex
adding pitfall absence of equivalence

@@
 % ==========================
+\section{Pitfall: Absence of canonical equivalence}
+\label{pitfall-absence-of-equivalence}
+% ==========================
+
+Although in most cases canonical equivalence will take care of alternative
+encodings of the same character, there are a some cases in which the Unicode
+Standard decided against equivalence. This leads to identical characters that
+are not equivalent, like <ø> \textsc{latin small letter o with stroke} at
+\uni{00F8} and <o̷> a combination of \textsc{latin small letter o} at \uni{006F}
+with \textsc{combining short solidus overlay} at \uni{0037}.
+
+The general rule followed is that extensions of Latin characters that are
+connected to the character are not separated as combining diacritics. For
+example, characters like <ƞ ɲ ɳ> or <ɰ ɱ> are obviously derived from <n> and <m>
+respectively, but they are treated like new separate characters in the Unicode
+Standard. Likewise, characters like <ø> and <ƈ> are not separated into a base 
+character <o> and <c> with an attached combining diacritic.
+
+Interestingly, and somewhat illogically, there are three elements attached to
+other characters that are still treated as separable in the Unicode Standard. First,
+the cedilla, like in <ç> \textsc{latin small letter c with cedilla} at
+\uni{00E7}, second the ogonek, like in <ą> \textsc{latin small letter a with
+ogonek} at \uni{0105} and, third, the horn, like in <ơ> \textsc{latin small
+letter o with horn} at \uni{01A1}. These characters are indeed decomposed in a 
+base character with a combining diacritic. However, it is these cases that should be 
+considered the exceptions to the rule.
+
+There are two combinations that deserve specific care because it is actually 
+possible to produce identical characters in different ways, without them being 
+canonical equivalent. First, there are the
+combining characters designated as ``overlay'' in the Unicode Standard, like    
+<\ \ {\large  ̴}> \textsc{combining tilde overlay} at \uni{0334} or 
+<\ \ {\large  ̵}> \textsc{combining short stroke overlay} at \uni{0335}. 
+There are many characters that look like they are 
+precomposed with such an overlay, for example 
+<{\small \fontspec{CharisSIL}ɫ ᵬ ᵭ ᵱ}> or 
+<{\small \fontspec{CharisSIL}ƚ ɨ ɉ ɍ}>, or also the 
+example of <ø> given at the start of this section. Second, the same situation also 
+occurs with combining characters designated as ``hook'', like < {\large  ̡}> 
+\textsc{combining palatalized hook below} at \uni{0321}. This element seems to
+occur in precomposed characters like <{\small \fontspec{CharisSIL}ᶀ ᶁ ᶂ ᶄ}>. 
+However, all these 
+characters are not canonical equivalent with a combination of their base 
+characters with the combining overlay or combining hook.
+
+% ==========================
 \section{Pitfall: Characters are not glyphs}

### 2015-06-06 chapters/introduction.tex
adding IPA-pitfall about non-decomposition

--- 2015-06-05 chapters/introduction.tex
@@
-\chapter{Writing Systems} \label{writing_systems}
+\chapter{Writing Systems}
+\label{writing_systems}
 
-% \section{Introduction} % \label{introduction}
+% ==========================
+\section{Introduction}
+\label{introduction}
+% ==========================
 
@@
 technological and practical pressures. They tend to be highly conservative, in
-that people who have learned to read and write in a specific way (however
-impractical or tedious) are mostly unwilling to change their habits, e.g.~they
+that people who have learned to read and write in a specific way---however
+impractical or tedious---are mostly unwilling to change their habits. Writers
 tend to resist spelling reforms. In all literate societies there exists a strong
@@
 For example, the writing of tone diacritics in Yoruba is often proclaimed to be
-the right way to write, although many users of Yoruba writing seem to be
+the right way to write, although many users of Yoruba orthography seem to be
 perfectly fine with leaving them out. As pointed out by the proponents of the
 official rules, there are some homographs when leaving out the tone diacritics
-\citet[44]{Olumuyiw2013}. However, writing systems (and the languages they
+\citep[44]{Olumuyiw2013}. However, writing systems (and the languages they
 represent) are normally full of homophones, which is normally not a problem at
@@
 even after more than a century of efforts to standardize the writing systems,
-there is still a wide range of variation in daily use \citet{Olumuyiw2013}.
+there is still a wide range of variation in daily use \citep{Olumuyiw2013}.
 
@@
 Before we deal with these proposals, in the first part of this paper (Sections
-\ref{encoding} through \ref{ipa-meets-unicode}) we give an extended introduction
-to the notion of encoding (Section \ref{encoding}) and writing systems, both
+~\ref{encoding} through~\ref{ipa-meets-unicode}) we give an extended introduction
+to the notion of encoding (Section~\ref{encoding}) and writing systems, both
 from a linguistic perspective and from the perspective of the Unicode Consortium
-(Section \ref{terminology}). We consider the Unicode Standard to be a
+(Section~\ref{terminology}). We consider the Unicode Standard to be a
 breakthrough (and ongoing) development that fundamentally changed the way we
@@
 thorough understanding of possible pitfalls that one might encounter when using
-it (Section \ref{unicode-pitfalls}). As an example of the current
+it (Section~\ref{unicode-pitfalls}). As an example of the current
 state-of-the-art, we discuss the rather problematic marriage of the
@@
 
-The second part of the paper (Sections \ref{orthography-profiles} and
+The second part of the paper (Sections~\ref{orthography-profiles} and
 \ref{use-cases}) describes our proposals for how to deal with the Unicode
@@
 propose the notions of orthography profiles, closely related to Unicode locale
-descriptions (Section \ref{orthography-profiles}). Finally, we discuss practical
-issues with actual examples (Section \ref{use-cases}). We provide reference
+descriptions (Section~\ref{orthography-profiles}). Finally, we discuss practical
+issues with actual examples (Section~\ref{use-cases}). We provide reference
 implementation of our proposals in R and in Python, available as open-source
@@
 and code points are expressed as U\emph{+n} where \emph{n} is a four to six
-digit hexadecimal number, e.g.~U+0256, which can be rendered as the glyph <ə>.
+digit hexadecimal number, e.g.~\uni{0256}, which can be rendered as the glyph <ə>.
+
+% ==========================
+\section{Encoding}
+\label{encoding}
+% ==========================
+
+There are many in-depth histories of the origin and development of writing
+systems \citep[e.g.~][]{Robinson1995,Powell2012}, a story that we therefore will
+not repeat here. However, the history of turning writing into computer readable
+code is not so often told, so we decided to offer a short survey of the major
+developments of such \textsc{encoding} here.\footnote{Because of the recent
+history as summarized in this section, we have used mostly rather ephemeral
+internet sources. When not references by traditional literature in the
+bibliography, we have used \url{http://www.unicode.org/history/} and various
+Wikipedia pages for the information presented here. A useful survey of the
+historical development of the physical hardware of telegraphy and
+telecommunication is \citet{Huurdeman2003}. Most books that discuss the
+development of encoding of telegraphic communication focus of cryptography,
+e.g.~\citet{Singh1999}, and forego the rather interesting story of ``open'',
+i.e.~non-cryptographic, encoding that is related here.} This history turns out
+to be intimately related to the history of telegraphic communication.
+
+\subsubsection*{Telegraphy}
+
+Writing systems have existed for roughly 6000 years, allowing people to exchange
+messages through time and space. Additionally, to quickly bridge large geographic
+distances, telegraphic systems of communication (from Greek \emph{τῆλε
+γράφειν} `distant writing') have a long and widespread history since ancient
+times. The most common telegraphic systems worldwide are so-called whistled
+languages \citep{Meyer2015}, but also drumming languages \citep{Meyer2012} and
+signalling by smoke, fire, flags or even change in water levels through
+hydraulic pressure have been used as forms of telegraphy.
+
+Telegraphy was reinvigorated in the end of the eighteenth century through the
+introduction of so-called semaphoric systems by Claude Chapelle to convey
+messages over large distances. Originally, various specially designed
+contraptions were used to send messages. Today, descendants of these systems
+are still in use utilizing flags or flashing lights. The ``innovation'' of those
+semaphoric systems was that all characters of the written language were replaced
+one-to-one by visual signals. Since then, all telegraphic systems have taken
+this principle, namely that any language to be transmitted first has to be
+turned into some orthographic system, which subsequently is encoded for
+transmission by the sender, and then turned back into orthographic
+representation at the receiver side.\footnote{Sound and video-based
+telecommunication of course takes a different approach by ignoring the written
+version of language and directly encode sound waves or light patterns.} This of
+course implies that the usefulness of any such telegraphic encoding completely
+depends on the sometimes rather haphazard structure of orthographic systems.
+
+In the nineteenth century, electric telegraphy lead to a new approach
+in which written language characters were encoded by signals sent through a
+copper wire. Originally, \textsc{bisignal codes} were used, consisting of two
+different signals. For example, Carl Friedrich Gauss in 1833 used positive and
+negative current \citep[282]{Mania2008}. More famous and influential, Samuel
+Morse in 1836 used long and short pulses. In those bisignal codes each character
+from the written language was encoded with a different number of signals
+(between one and five), so two different separators are needed: one between
+signals and one between characters. For example, in Morse-code there is a short
+pause between signals and a long pause between characters.\footnote{Actually,
+Morse-code also includes an extra long pause between words. Interestingly, it
+took a long time to consider the written word boundary---using white-space---as
+a bona-fide character that should simply be encoded with its own code point.
+This happened only with the revision of the Baudot-code (see below) by Donald
+Murray in 1901, in which he introduced a specific white-space code. This
+principle has been followed ever since.}
+
+\subsubsection*{Binary encoding}
+
+From those bisignal encodings, true \textsc{binary codes} developed with a fixed
+length of signals per character. In such systems only a single separator between
+signals is needed, because the separation between characters can be established
+by counting until a fixed number of signals has passed.\footnote{Of course, no
+explicit separator is needed when the timing of the signals is known, which is
+the principle used in all modern telecommunication systems. An important modern
+consideration is also how to know where to start counting when you did not catch
+the start of a message, something that is known in Unicode as \textsc{self
+synchronization}.} In the context of electric telegraphy, such a binary code
+system was first established by Émile Baudot in 1870, using a fixed combination
+of five signals for each written character.\footnote{True binary codes have a
+longer history, going at least back to the Baconian cipher devised by Francis
+Bacon in 1605. However, the proposal by Baudot was the quintessential proposal
+leading to all modern systems.} There are $2^5 = 32$ possible combination when
+using five binary signals; an encoding today designated as 5-bit. These
+codes are sufficient for all Latin letters, but of course they do not suffice
+for all written symbols, including punctuation and digits. As a solution, the
+Baudot code uses a so-called ``shift'' character, which signifies that from
+that point onwards---until shifted back---a different encoding is used, allowing
+for yet another set of 32 codes. In effect, this means that the Baudot code, and
+the \textsc{International Telegraph Alphabet} (ITA) derived from it, had an
+extra bit of information, so the encoding is actually 6-bit (with $2^6
+= 64$ different possible characters). For decades, this encoding was the
+standard for all telegraphy and it is still in limited use today.
+
+To also allow for different uppercase and lowercase letters and for a large
+variety of control characters to be used in the newly developing technology of
+computers, the American Standards Association decided to propose a new 7-bit
+encoding in 1963 (with $2^7 = 128$ different possible characters), known as the
+\textsc{American Standard Code for Information Interchange} (ASCII), geared
+towards the encoding of English orthography. With the ascent of other
+orthographies in computer usage, the wish to encode further variation of Latin
+letters (like German <ß> or various letters with diacritics, like <è>) led the
+Digital Equipment Corporation to introduce an 8-bit \textsc{Multinational
+Character Set} (MCS, with $2^8 = 256$ different possible characters), first used
+with the introduction of the VT{\large 220} Terminal in 1983.
+
+Because 256 characters were clearly not enough for the many different characters
+needed in the world's writing systems, the ISO/IEC~8859 standard in 1987
+extended the MCS to include 16 different 8-bit code pages. For example, part 5
+was used for Cyrillic characters, part 6 for Arabic, and part 7 for
+Greek.\footnote{In effect, because $16 = 2^4$, this means that ISO/IEC~8859 was
+actually an $8+4=12$-bit encoding, though with very many duplicates by design, namely 
+all ASCII codes were repeated in each 8-bit code page. To be precise,
+ISO/IEC~8859 used the 7-bit ASCII as the basis for each code page, and defined
+16 different 7-bit extensions, leading to $(1+16)\cdot{2^7} = 2,176$ possible
+characters.} This system almost immediately was understood to be insufficient and
+impractical, so various initiatives to extend and reorganize the encoding
+started in the 1980s. This led, for example, to various proprietary encodings
+from Microsoft (e.g.~Windows Latin 1) and Apple (e.g.~Mac OS Roman), which one
+still sometimes encounters today.
+
+More wide-ranging, various people in the 1980s started to develop true
+international code sets. In the United States, a group of computer scientists
+formed the \textsc{Unicode Consortium}, proposing a 16-bit encoding in 1991
+(with $2^{16} = 65,536$ different possible characters). At the same time in
+Europe, the \textsc{International Organization for Standardization} (ISO) was
+working on ISO~10646 to supplant the ISO/IEC~8859 standard. Their first draft of
+the \textsc{Universal Character Set} (UCS) in 1990 was 31-bit (with
+theoretically $2^{31} = 2,147,483,648$ possible characters, but because of some
+technical restrictions only 679,477,248 were allowed). Since 1991, the Unicode
+Consortium and the ISO jointly develop the \textsc{unicode standard}, or
+ISO/IEC~10646, leading to the current system including the original 16-bit
+Unicode proposal as the \textsc{basic multilingual plane}, and 16 additional
+planes of 16-bit for further extensions (with in total $(1+16) \cdot 2^{16} =
+1,114,112$ possible characters). The most recent version of the Unicode Standard
+(currently at version number 7.0) was published in June 2014 and it defines
+112,956 different characters \citep{Unicode2014}.
+
+\
+
+\noindent In the next section we provide a very brief overview of the linguistic terminology
+concerning writing systems before turning to the slightly different
+computational terminology in the next section on the Unicode Standard.
+
+% ==========================
+\section{Linguistic terminology}
+\label{linguistic-terminology}
+% ==========================
+
+Linguistically speaking, a \textsc{writing system} is a symbolic system that
+uses visible or tactile signs to represent language in a systematic way. The
+term writing system has two mutually exclusive meanings. First, it may
+refer to the way a particular language is written. In this sense the term refers
+to the writing system of a particular language, as, for example, in \emph{`the
+Serbian writing system uses two scripts: Latin and Cyrillic.'} Second, the term
+writing system may also refer to a type of symbolic system as used among the
+world's languages to represent the language, as, for example, in
+\emph{`alphabetic writing system.'} In this latter sense the term refers to how
+scripts have been classified according to the way that they encode language, as
+in, for example, \emph{`the Latin and Cyrillic scripts are both alphabetic
+writing systems.'} To avoid confusion, this second notion of writing system
+would more aptly have been called \textsc{script system}.
+
+\subsubsection*{Writing systems}
+
+Focussing on the first sense, we distinguish two different kinds of 
+writing systems used for a particular language, namely transcriptions and
+orthographies. First, \textsc{transcription} is a scientific procedure (and also
+the result of that procedure) for graphically representing the sounds of human
+speech at the phonetic level. It incorporates a set of unambiguous symbols to
+represent speech sounds, including conventions that specify how these symbols
+should be combined. A transcription system is a specific system of symbols and
+rules used for transcription of the sounds of a spoken language variety. In
+principle, a transcription system should be language-independent, in that it
+should be applicable to all spoken human languages. The \textsc{International
+Phonetic Alphabet} (IPA) is a commonly used transcription system that provides a
+medium for transcribing languages at the phonetic level. However, there is a
+long history of alternative kinds of transcription systems
+\citep[see][]{Kemp2006} and today various alternatives are in widespread use
+(e.g.~X-SAMPA and Cyrillic-based phonetic transcription systems). Many
+users of IPA do not follow the standard to the letter, and many dialects based
+on the IPA have emerged, e.g.~the Africanist and Americanist transcription
+systems. Note that IPA symbols are also often used to represent language on a
+phonemic level. However, it is important to realize that in this usage the IPA
+symbols are not a transcription system, but rather an orthography (though with
+strong links to the pronunciation). Further, a transcription system does not
+need to be as highly detailed as the IPA.\ It can also be a system of broad sound
+classes. Although such an approximative transcription is not normally used in
+linguistics, it is widespread in technological approaches
+\citetext{\citealp[Soundex and variants, e.g.~][391--392]{Knuth1973};
+\citealp{postel1969,Beider2008}}, and it is sometimes fruitfully used in
+automatic approaches to historical linguistics
+\citep{Dolgopolsky1986,List2012esslli,Brown2013}.
+
+Second, an \textsc{orthography} specifies the symbols, punctuations, and the
+rules in which a specific language is written in a standardized way.
+Orthographies are often based on a phonemic analysis, but they almost always
+include idiosyncrasies because of historical developments (like sound changes or
+loans) and because of the widely-followed principle of lexical integrity
+(i.e.~the attempt to write the same lexical root in a consistent way, also when
+synchronic phonemic rules change the pronunciation, as for example with final
+devoicing in many Germanic languages). All orthographies are language-specific
+(and often even resource-specific), although individual symbols or rules might
+be shared between languages. A \textsc{practical orthography} is a strongly
+phoneme-based writing system designed for practical use by speakers. The mapping
+relation between phonemes and graphemes in practical orthographies is purposely
+shallow, i.e.~there is mostly a systematic and faithful mapping from a phoneme
+to a grapheme. Practical orthographies are intended to jumpstart written
+materials development by correlating a writing system with the sound units of a
+language \citep[cf.~][]{MeinhofJones1928}. Symbols from the IPA are often used
+by linguists in the development of such practical orthographies for languages
+without writing systems, though this usage of IPA symbols should not be confused
+with transcription (as defined above).
+
+Further, a \textsc{transliteration} is a mapping between two different
+orthographies. It is the process of ``recording the graphic symbols of one
+writing system in terms of the corresponding graphic symbols of a second writing
+system'' \citep[396]{Kemp2006}. In straightforward cases, such a transliteration
+is simply a matter of replacing one symbol with another. However, there are
+widespread complications, like one-to-many or many-to-many mappings, which are
+not always easy, or even possible, to solve without listing all cases
+individually \citep[cf.~][Ch.~2]{Moran2012}.
+
+\subsubsection*{Script systems}
+
+Different kinds of writing systems are classified into script
+systems. A \textsc{script} is a collection of distinct symbols as
+employed by one or more orthographies. For example, both Serbian and Russian are
+written with subsets of the Cyrillic script. A single language, like Serbian or
+Japanese, can also be written using orthographies based on different scripts.
+Over the years linguists have classified script systems in a variety of ways,
+with the tripartite classification of logographic, syllabic, and alphabetic
+remaining the most popular, even though there are at least half a dozen
+different types of script systems that can be distinguished
+\citep{Daniels1990,Daniels1996}.
+
+Breaking it down further, a script consists of \textsc{graphemes}, and graphemes
+consist of \textsc{characters}. In the linguistic terminology of writing
+systems, a \textsc{character} is a general term for any self-contained element
+in a writing system.\footnote{There is a second interpretation of the term
+\textsc{character}, i.e.~a conventional term for a unit in the Chinese writing
+system \citep{Daniels1996}. This interpretation will not be further explored in
+this paper.} Although in literate societies most people have a strong intuition
+about what the characters are in their particular orthography or orthographies,
+it turns out that the separation of an orthography into separate characters is
+far from trivial. The widespread intuitive notion of a character is strongly
+biased towards educational traditions, like the alphabet taught at schools, and
+technological possibilities, like the available type pieces in a printer's job
+case, the keys on a typewriter, or the symbols displayed in Microsoft Word's
+symbol browser. In practice, characters often consist of multiple building
+blocks, each of which could be considered a character in its own right. For
+example, although a Chinese character may be considered to be a single basic
+unanalyzable unit, at a more fine-grained level of analysis the internal
+structure of Chinese characters is often comprised of smaller semantic and
+phonetic units that should be considered characters \citep{Sproat2000}. In
+alphabetic scripts, this problem is most forcefully exemplified by diacritics.
+
+A \textsc{diacritic} is a mark, or series of marks, that may be above, below, or
+through other characters \citep{Gaultney2002}. Diacritics are sometimes used to
+distinguish homophonous words, but they are more often used to indicate a
+modified pronunciation \citep[xli]{DanielsBright1996}. The central question is
+whether, for example, <e>, <è>, <a> and <à> should be considered four
+characters, or different combinations of three characters. In general, multiple
+characters together can form another character, and it is not always possible to
+decide on principled grounds what should be the basic building blocks of an
+orthography.
+
+For that reason, it is better to analyze an orthography as a collection of
+graphemes. A \textsc{grapheme} is the basic, minimally distinctive symbol of a
+particular writing system, alike to the phoneme is an abstract representation of
+a distinct sound in a specific language. The term \textsc{grapheme} was modeled
+after the term \textsc{phoneme} and represents a contrastive graphical unit in a
+writing system (see \citealp{Kohrt1986} for a historical overview of the term
+grapheme). Most importantly, a single grapheme regularly consists of multiple
+characters, like <th>, <ou> and <gh> in English (note that each character in
+these graphemes is also a separate grapheme in English). Such complex graphemes
+are often used to represent single phonemes. So, a combination of characters is
+used to represent a single phoneme. Note that the opposite is also found in
+writing systems, in cases in which a single character represents a combination
+of two or more phonemes. For example, <x> in English orthography represents a
+combination of the phonemes /k/ and /s/.
+
+Further, conditioned or free variants of a grapheme are called
+\textsc{allographs}. For example, the distinctive forms of Greek sigma are
+conditioned, with <σ> being used word-internally and <ς> being used at the end
+of a word. In sum, there are many-to-many relationships between phonemes and
+graphemes as they are expressed in the myriad of language- and resource-specific
+orthographies.
+
+\
+
+\noindent This exposition of the linguistic terminology involved in describing writing
+systems has been purposely brief. We have highlighted some of the linguistic
+notions that are pertinent, yet sometimes confused with, the technological
+definitions developed for the computational processing of the world's writing
+systems, which we describe in the next section.
+
+% ==========================
+\section{The Unicode approach}
+\label{the-unicode-approach}
+% ==========================
+
+The conceptualization and terminology of writing systems was rejuvenated by
+the development of the Unicode Standard, with major input from Mark Davis,
+co-founder and long-term president of the Unicode Consortium. For many years,
+``exotic'' writing systems and phonetic transcription systems on personal
+computers were constrained by the American Standard Code for Information
+Interchange (ASCII) character encoding scheme (based on the Latin script), which
+only allowed for a strongly limited number of different symbols to be encoded.
+This implied that users could either use and adopt the (extended) Latin alphabet
+or they could assign new symbols to the small number of code points in the ASCII
+encoding scheme to be rendered by a specifically designed font
+\citep{BirdSimons2003}. In this situation, it was necessary to specify the font
+together with each document to ensure the rightful display of its content. To
+alleviate this problem of assigning different symbols to the same code points,
+in the late 80's and early 90's the Unicode Consortium set itself the ambitious
+goal of developing a single universal character encoding to provide a unique
+number, a code point, for every character in the world's writing systems.
+Nowadays, the Unicode Standard is the default encoding of the technologies that
+support the World Wide Web and for all modern operating systems, software and
+programming languages.
+
+\subsubsection*{The Unicode Standard}
+
+The Unicode Standard represents a massive step forward because it aims to
+eradicate the distinction between \textsc{universal} (ASCII) versus
+\textsc{language/resource-particular} (Font) by adding as much as possible
+language-specific information into the universal standard. However, there are
+still language/resource-specific specifications necessary for the proper usage
+of Unicode, as will be discussed below. Within the Unicode structure many of
+these specifications can be captured by so-called \textsc{Locale Descriptions}
+or simply locales, so we are moving to a new distinction of universal (Unicode
+Standard) versus language-particular (Locale Description). The major gain is a
+much larger compatibility on the universal level (because Unicode standardizes a
+much larger portion of writing system diversity), and much better possibilities
+for automatic processing on the language-particular level (because Locale
+Descriptions are computer readable specifications).
+
+Each version of the Unicode Standard (\citealp{Unicode2014}, as of writing at
+version 7) consists of a set of specifications and guidelines that include (i) a
+core specification, (ii) code charts, (iii) standard annexes and (iv) a
+character database.\footnote{All documents of the Unicode Standard are available
+at \url{http://www.unicode.org/versions/latest/}. For a quick survey of the use
+of terminology inside the Unicode Standard, their glossary is particularly
+useful, available at \url{http://www.unicode.org/glossary/}. For a general
+introduction to the principles of Unicode, Chapter 2 of the core specification,
+called \textsc{general structure}, is particularly insightful. Different from
+many other documents of the Unicode Standard, this general introduction is
+relatively easy to read and illustrated with many interesting examples from
+various orthography traditions all over the world.} The \textsc{core
+specification} is a book directed toward human readers that describes the formal
+standard for encoding multilingual text. The \textsc{code charts} provide a
+humanly readable online reference to the character contents of the Unicode
+Standard in the form of PDF files. The \textsc{Unicode Standard Annexes (UAX)}
+are a set of technical standards that describe the implementation of the Unicode
+Standard for software development, Web standards, and programming languages.
+The \textsc{Unicode Character Database (UCD)} is a set of computer-readable text
+files that describe the character properties, including a set of rich character
+and writing system semantics, for each character in the Unicode Standard. In
+this section, we introduce the basic Unicode concepts, but we will leave out
+many details. Please consult the above mentioned full documentation for a more
+detailed discussion. Further note that the Unicode Standard is exactly that,
+namely a standard. It normatively describes notions and rules to be followed. In
+the actual practice of applying this standard in a computational setting, a
+specific implementation is necessary. The most widely used implementation of the
+Unicode Standard is the \textsc{International Components for Unicode (ICU)},
+which offers C/C++ and Java libraries trying to implement the Unicode
+Standard.\footnote{\url{http://icu-project.org}}
+
+\subsubsection*{Character encoding system}
+
+The Unicode Standard is a \textsc{character encoding system} which
+goal it is to support the interchange and processing of written characters and
+text in a computational setting. Underlyingly, the character encoding is
+represented by a range of numerical values called a \textsc{code space}, which
+is used to encode a set of characters. A \textsc{code point} is a unique
+non-negative integer within a code space (i.e.~within a certain numerical
+range). In the Unicode Standard character encoding system, an \textsc{abstract
+character}, for example the \textsc{latin small letter p}, is mapped to a
+particular code point, in this case the decimal value 112, normally represented in
+hexadecimal, which then looks in Unicode parlance as
+\uni{0070}.\footnote{Hexadecimal (base-16) 0070 is equivalent to decimal
+(base-10) 112, which can be calculated by considering that $(0\cdot16^3) +
+(0\cdot16^2) + (7\cdot16^1) + (0\cdot16^0) = 7\cdot16 = 112$. Underlyingly,
+computers will of course treat this code point binary (base-2) as 11100000, as
+can be seen by calculating that $(1\cdot2^7) + (1\cdot2^6) + (1\cdot2^5) +
+(0\cdot2^4) + (0\cdot2^3) + (0\cdot2^2) + (0\cdot2^1) + (0\cdot2^0) = 64 + 32 +
+16 = 112$.} That encoded abstract character is rendered on a computer screen (or
+printed page) as a \textsc{glyph}, e.g. <p>, depending on the \textsc{font} and
+the context in which that character appears.
+
+In Unicode Standard terminology, an (abstract) \textsc{character} is the basic
+encoding unit. The term \textsc{character} can be quite confusing due to its
+alternative definitions across different scientific disciplines and because in
+general the word \textsc{character} means many different things to different
+people. It is therefore often preferable to refer to Unicode characters simply
+as \textsc{code points}, because there is a one-to-one mapping between Unicode
+characters and their numeric representation. In the Unicode approach, a
+character refers to the abstract meaning and/or general shape, rather than a
+specific shape, though in code tables some form of visual representation is
+essential for the reader's understanding. Unicode defines characters as
+abstractions of orthographic symbols, and it does not define visualizations for
+these characters (although it does presents examples). In contrast, a
+\textsc{glyph} is a concrete graphical representation of a character as it
+appears when rendered (or rasterized) and displayed on an electronic device or
+on printed paper. For example, <g {\large \textit{g}} \textbf{g}
+{\fontspec{ArialMT} {\small g} \textit{g} \textbf{g}}> are different glyphs of the
+same character, i.e.~they may be rendered differently depending on the
+typography being used, but they all share the same code point. From the
+perspective of Unicode they are \textit{the same thing}. In this approach, a
+\textsc{font} is then simply a collection of glyphs linked to code points.
+Allography is not specified in Unicode (expect for a few exceptional cases, due
+to legacy encoding issues), but can be specified in a font as a
+\textsc{contextual variant} (a.k.a.~presentation form).
+
+Each code point in the Unicode Standard is associated with a set of
+\textsc{character properties} as defined by the Unicode character property
+model.\footnote{The character property model is described in
+\url{http://www.unicode.org/reports/tr23/}, but the actual properties are
+described in \url{http://www.unicode.org/reports/tr44/}. A simplified overview
+of the properties is available at
+\url{http://userguide.icu-project.org/strings/properties}. The actual code
+tables listing all properties for all Unicode code points are available at
+\url{http://www.unicode.org/Public/UCD/latest/ucd/}.} Basically, those
+properties are just a long list of values for each character. For example, code
+point \uni{0047} has the following properties (among many others): 
+\begin{itemize}
+	\item Name: LATIN CAPITAL LETTER G 
+	\item Alphabetic: YES 
+	\item Uppercase: YES 
+	\item Script: LATIN 
+	\item Extender: NO 
+	\item Simple\_Lowercase\_Mapping: 0067 
+\end{itemize}
+
+These properties contain the basic information of the Unicode Standard and they
+are necessary to define the correct behavior and conformance required for
+interoperability in and across different software implementations (as defined in
+the Unicode Standard Annexes). The character properties assigned to each code
+point is based on each character's behavior in the real-world writing
+traditions. For example, the corresponding lowercase character to \uni{0047} is
+\uni{0067} (though note that the relation between uppercase and lowercase is in
+many situations much more complex than this, and Unicode has further
+specifications for those cases). Another use of properties is to define the
+script of a character.\footnote{The Unicode Glossary defines the term \textsc{script} as
+a ``collection of letters and other written signs used to represent textual
+information in one or more writing systems. For example, Russian is written with
+a subset of the Cyrillic script; Ukrainian is written with a different subset.
+The Japanese writing system uses several scripts.''} In practice, script is
+simply defined for each character as the explicit \textsc{script} property in
+the Unicode Character Database.
+
+One frequently references property is the \textsc{block} property, which is
+often used in software applications to impose some structure to the large number
+of Unicode characters. Each character in Unicode belongs to a specific block.
+These blocks are basically an organizational structure to alleviate the
+administrative burden of keeping Unicode up-to-date. Blocks consist of
+characters that in some way belong together, so that characters are easier to
+find. Some blocks are connected with a specific script, like the Hebrew block or
+the Gujarati block. However, blocks are predefined ranges of code points, and
+often there will come a point after which the range is completely filled. Any
+extra characters will have to be assigned somewhere else. There is, for example,
+a block \textsc{Arabic}, which contains most Arabic symbols. However, there is
+also a block \textsc{Arabic Supplement}, \textsc{Arabic Presentation Forms-A}
+and \textsc{Arabic Presentation Form-B}. The situation with Latin symbols is
+even more extra. In general, the names for block should be taken as a
+definitional statement. For example, many IPA symbols are not located in the
+aptly-names block \textsc{IPA extensions}, but in other blocks (see Section~\ref{ipa-meets-unicode}).
+
+\subsubsection*{Grapheme clusters}
+
+There are many cases in which a sequence of characters (i.e.~a sequence of more
+than one code point) represents what a user perceives as an individual unit in a
+particular orthographic writing system. For this reason the Unicode Standard
+differentiates between \textsc{abstract character} and \textsc{user-perceived
+character}. Sequences of multiple code points that correspond to a single
+user-perceived characters are called grapheme clusters in Unicode parlance.
+Grapheme clusters come in two flavors: (default) grapheme clusters and tailored
+grapheme clusters.
+
+The (default) \textsc{grapheme clusters} are locale-independent graphemes,
+i.e.~they always apply when a particular combination of characters occurs
+independent of the writing system in which they are used. These character
+combinations are defined in the Unicode Standard as functioning as one
+\textsc{text element}.\footnote{The Unicode Glossary defines text element as:
+``A minimum unit of text in relation to a particular text process, in the
+context of a given writing system. In general, the mapping between text elements
+and code points is many-to-many.''} The simplest example of a grapheme cluster
+is a base character followed by a letter modifier character. For example, the
+sequence <n> + < ̃> (i.e.~\textsc{latin small letter n} at \uni{006E}, followed
+by \textsc{combining tilde} at \uni{0303}) combines visually into <ñ>, a
+user-perceived character in writing systems like that of Spanish. So, what the
+user perceives as a single character actually involves a multi-code-point
+sequence. Note that this specific sequence can also be represented with a single
+so-called \textsc{precomposed} code point, \textsc{latin small letter n with
+tilde} at \uni{00F1}, but this is not the case for all multi-code point
+character sequences. The problem that there multiple encodings possible for the
+same text element has been acknowledged early on in the Unicode Standard
+(e.g.~for <ñ>, the sequence \uni{006E} \uni{0303} should in all situations be
+treated identically to the precomposed \uni{00F1}), and a system of
+\textsc{canonical equivalence} is available for such situations. Basically, the
+Unicode Standard offers different kind of normalizations to either decompose all
+precomposed characters (called \textsc{NFD}, \textsc{Normalization Form
+canonical Decomposition}), or compose as much as possible combinations (called
+\textsc{NFC}, \textsc{Normalization Form canonical Composition}). In current
+practice of software development, NFC seems to be preferred in most situations
+and is widely proposed as the preferred canonical form.
+
+More difficult for text processing, because less standardized, is what the
+Unicode Standard terms \textsc{tailored grapheme clusters}. Tailored grapheme
+clusters are locale-dependent graphemes, i.e.~such combination of characters do
+not function as text elements in all situations. For example, the sequence <c>~+~<h> for the Slovak digraph <ch> or the sequence <ky> in the Sisaala practical
+orthography (pronounced as IPA /tʃ/,~\cite{Moran2006}). These grapheme clusters
+are \textsc{tailored} in the sense that they must be specified on a
+language-by-language or a writing system-by-writing system basis. The Unicode
+Standard provides technological specifications for creating locale specific data
+in so-called \textsc{Unicode Locale Descriptions}, i.e.~a set of specification
+that defines a set of language-specific elements (e.g.~tailored grapheme
+clusters, collation order, capitalization-equivalence), as well as other special
+information, like how to format numbers, dates, or currencies. Locale
+descriptions are saved in the \textsc{Common Locale Data Repository
+(CLDR)},\footnote{\url{http://cldr.unicode.org/}} a repository of
+language-specific definitions of writing system properties, each of which
+describes specific usages of characters. Each locale can be encoded in a
+document using the \textsc{Locale Data Markup Language (LDML)}. LDML is an XML
+format and vocabulary for the exchange of structured locale data. Unicode Locale
+Descriptions allow users to define language- or even resource-specific writing
+systems or orthographies.\footnote{The Unicode Glossary defines \textsc{writing
+system} only very loosely, as it is not a central concept in the Unicode
+Standard. A writing system is, ``A set of rules for using one or more scripts to
+write a particular language. Examples include the American English writing
+system, the British English writing system, the French writing system, and the
+Japanese writing system.''} However, there are various drawbacks of locale
+descriptions for the daily practice of linguistic work in a multilingual setting
+(see Section~\ref{use-cases}).

### 2015-06-06 chapters/ipa_meets_unicode.tex
adding pitfall absence of equivalence
adding IPA-pitfall about non-decomposition

--- 2015-06-05 chapters/ipa_meets_unicode.tex
@@
 
-\section{Pitfall: No decomposition of strikethrough}
-\label{pitfall-no-decomposition-of=strikethrough}
+\section{Pitfall: No decomposition with some diacritics}
+\label{pitfall-no-decomposition-of-overlay}
+
+Although many combinations of base character with diacritic are treated as 
+canonical equivalent with precomposed characters, there are a few combinations 
+in IPA that allow for multiple, apparently identical, encodings that are not 
+canonical equivalent. The following elements should not be treated as diacritics 
+when encoding IPA in Unicode:
+\begin{itemize}
+  \item <\ {\large  ̡}\ > \textsc{combining palatalized hook below} at \uni{0321}
+  \item <\ \ {\large  ̢}> \textsc{combining retroflex hook below} at \uni{0322}
+  \item <\ \ {\large  ̵}> \textsc{combining short stroke overlay} at \uni{0335}
+  \item <\ \ {\large  ̷}> \textsc{combining short solidus overlay} at \uni{0337}
+  \item <\ \ {\large  ̴}> \textsc{combining tilde overlay} at \uni{0334}
+  \item <{\large ˞}> \textsc{modifier letter rhotic hook} at \uni{02DE}
+\end{itemize}
+
+There turn out to be a lot of characters in the IPA that could be conceived as 
+using any of these elements, like <ɲ>, <ɳ>, <ɨ>, <ø>, <ɫ> or <ɚ>. However, all 
+such characters exist as well as ``precomposed'' combination in Unicode, and these 
+precomposed characters should preferably be used. When combinations of a base 
+character with diacritic are used, then these combinations are not canonical 
+equivalent to the precomposed combinations. This means that any search will not 
+find both at the same time.
+
+Reversely, <ç> \textsc{latin small letter c with cedilla} at \uni{00E7} will be
+decomposed into <c> \textsc{latin small letter c} at \uni{0063} with 
+<\ \ {\large  ̧}> \textsc{combining cedilla} at \uni{0327}, 
+also if this is not intended.
 

### 2015-06-06 chapters/ipa_meets_unicode.tex
restoring pitfall on digraphs

@@
 Some other homoglyphs related to encoding IPA in the Unicode Standard are:
+\begin{itemize}
+	\item The uses of the apostrophe has led to long discussions on the Unicode
+       Standard email list. An English keyboard inputs
+       \textless{}`\textgreater{} APOSTROPHE at \uni{0027}, although the
+       \textit{preferred} Unicode apostrophe is the \textless{}'\textgreater{}
+       \textsc{right single quotation mark} at \uni{2019}. Yet the glottal
+       stop/glottalization/ejective marker is another completely different
+       character, the \textsc{modifier letter apostrophe} at \uni{02BC}, but
+       unfortunately looks mostly highly similar to \uni{2019}. 
+	\item There is ambiguous encoding of IPA segments within the Unicode Standard.
+       An example is the \uni{02C1} \textsc{modifier letter reversed glottal
+       stop} vs the \uni{02E4} \textsc{modifier letter small reversed glottal
+       stop} . Both are denoted in the Unicode Standard as the
+       \textsc{pharyngealized diacritic} and both appear in various resources
+       representing phonetic data online. This is thus an example for which the
+       Unicode Standard does not solve the linguistic standardization problem. 
+	\item There is at least one case in which the character name assigned by the
+       Unicode Consortium does not match the IPA's description: in the Unicode
+       Standard at \uni{01C3} is labeled \textsc{latin letter retroflex click},
+       but in IPA is an alveolar or postalveolar click (not retroflex). This
+       naming is probably best seen as a simple error in the Unicode Standard.
+\end{itemize}
+
+\section{Pitfall: Ligature and digraphs}
+\label{pitfall-ligatures-digraphs}
+
+One important distinction to acknowledge is the difference between multigraphs
+and ligatures. Multigraphs are groups of characters (in the context of IPA e.g.
+< tʃ > or < ou >) while ligatures are single characters (e.g. <ʧ> \textsc{latin
+small letter tesh digraph} at \uni{02A7}). Ligatures arose in the context of
+printing easier-to-read texts, and are included in the Unicode Standard for
+reasons of legacy encoding. However, their usage is discouraged by the Unicode
+core specification. Specifically related to IPA, various phonetic combinations
+of characters (typically affricates) are available as single code-points in the
+Unicode Standard, but are designated as \textsc{ligatures} or \textsc{digraphs}
+(confusingly both names appear interchangeably). Such glyphs might be used by
+software to produce a pleasing display, but they should not be hard-coded into
+the text itself. In the context of IPA, characters like the following ligatures
+should thus \emph{not} be used. Instead a combination of two characters is
+preferred:
 
 \begin{itemize} 
-	\item The uses of the apostrophe has led to long discussions on the Unicode Standard email list.
-An English keyboard inputs \textless{}`\textgreater{} APOSTROPHE at \uni{0027}, although the \textit{preferred} Unicode apostrophe is the \textless{}'\textgreater{} \textsc{right single quotation mark} at \uni{2019}.
-Yet the glottal stop/glottalization/ejective marker is another completely different character, the \textsc{modifier letter apostrophe} at \uni{02BC}, but unfortunately looks mostly highly similar to \uni{2019}.
-	\item There is ambiguous encoding of IPA segments within the Unicode Standard.
-An example is the \uni{02C1} \textsc{modifier letter reversed glottal stop} vs the \uni{02E4} \textsc{modifier letter small reversed glottal stop} .
-Both are denoted in the Unicode Standard as the \textsc{pharyngealized diacritic} and both appear in various resources representing phonetic data online.
-This is thus an example for which the Unicode Standard does not solve the linguistic standardization problem.
-	\item There is at least one case in which the character name assigned by the Unicode Consortium does not match the IPA's description: in the Unicode Standard at \uni{01C3} is labeled \textsc{latin letter retroflex click}, but in IPA is an alveolar or postalveolar click (not retroflex).
-This naming is probably best seen as a simple error in the Unicode Standard.
-Note that most linguists simply seem to use or <ou>) while ligatures are single characters (e.g.
-<ʧ> \textsc{latin small letter tesh digraph} at \uni{02A7}).
-Ligatures arose in the context of printing easier-to-read texts, and are included in the Unicode Standard for reasons of legacy encoding.
-However, their usage is discouraged by the Unicode core specification.
-Specifically related to IPA, various phonetic combinations of characters (typically affricates) are available as single code-points in the Unicode Standard, but are designated as \textsc{ligatures} or \textsc{digraphs} (confusingly both names appear interchangeably).
-Such glyphs might be used by software to produce a pleasing display, but they should not be hard-coded into the text itself.
-In the context of IPA, characters like the following ligatures should thus \emph{not} be used.
-Instead a combination of two characters is preferred:
 	\item <ʣ> \textsc{latin small letter dz digraph} at \uni{02A3} (use <d> + <z> instead) 
@@
 <\ \ {\large  ̧}> \textsc{combining cedilla} at \uni{0327}, 
-also if this is not intended.
+also if such a decompotion is not intended, because it is meaningless in IPA.
 

### 2015-06-06 chapters/ipa_meets_unicode.tex
adding pitfall multiple options in IPA

@@
 
-\section{Pitfall: Many IPA homoglyphs}
-\label{pitfall-there-are-many-ipa-homoglyphs-in-unicode}
+\section{Pitfall: IPA homoglyphs in Unicode}
+\label{pitfall-ipa-homoglyphs}
 
@@
 
+\section{Pitfall: Homoglyphs in IPA}
+
+It is not just the Unicode Standard that offers multiple options for encoding 
+the IPA. Even the IPA itself is not consistent in how specific meaningful 
+transcriptions have to be encoded. There are a few cases in which the IPA 
+explicitly allows for different options of transcribing the same phonetic 
+content
+\begin{itemize}
+  \item tone marks (prefer tone symbols)
+  \item marking of <g>
+  \item marking of voiceless with ring diacritic
+  \item tie bar
+\end{itemize}
+
 \section{Pitfall: Ligature and digraphs}
@@
 
-\section{Pitfall: No decomposition with some diacritics}
-\label{pitfall-no-decomposition-of-overlay}
+\section{Pitfall: Missing decomposition}
+\label{pitfall-missing-decomposition}
 

### 2015-06-06 chapters/pitfalls.tex
adding IPA-pitfall about non-decomposition
reformatted pitfall file formats

@@
 with \textsc{combining short solidus overlay} at \uni{0037}.
-
 The general rule followed is that extensions of Latin characters that are
@@
 occur in precomposed characters like <{\small \fontspec{CharisSIL}ᶀ ᶁ ᶂ ᶄ}>. 
-However, all these 
-characters are not canonical equivalent with a combination of their base 
-characters with the combining overlay or combining hook.
+However, all these characters are not canonical equivalent with a combination of 
+their base characters with the combining overlay or combining hook.
 
@@
 there are some situations in which a user is suddenly confronted with cryptic
-questions in the user interface involving abbreviations like \textsc{lf},
-\textsc{cr}, \textsc{be}, \textsc{le} or \textsc{bom}. Most prominently this
-occurs with exporting or importing data in several software applications from
-Microsoft. Basically, there are two different issues involved. First, the
-encoding of line breaks and, second, the encoding of the Unicode characters into
-\textsc{code units} and the related issue of \textsc{endianness}.
+questions in the user interface involving abbreviations like LF, CR, BE, LE or
+BOM. Most prominently this occurs with exporting or importing data in several
+software applications from Microsoft. Basically, there are two different issues
+involved. First, the encoding of line breaks and, second, the encoding of the
+Unicode characters into \textsc{code units} and the related issue of
+\textsc{endianness}.
+
+\subsubsection*{Line breaks}
 
@@
 to direct a printing head to a new line. This involves two movements, known as
-\textsc{carriage return (cr)} (returning the printing head to the start of the
-line on the page) and \textsc{line feed (lf)} (moving the printing head to the
+\textsc{carriage return} (CR, returning the printing head to the start of the
+line on the page) and \textsc{line feed} (LF, moving the printing head to the
 next line on the page). Physically, these are two different events, but
 conceptually together they form one action. In the history of computing, various
-encodings of line breaks have been used (e.g.~\textsc{cr+lf}, \textsc{lf+cr},
-only \textsc{lf}, or only \textsc{cr}). Currently, all Unix and Unix-derived
-systems use only \textsc{lf} as code for a line break, while software from
-Microsoft still uses a combination of \textsc{cr} and \textsc{lf}. Today, most
-software applications recognize both options, and are able to deal with either
-encoding of line breaks (until rather recently this was not the case, and using
-the wrong line breaks could lead to unexpected errors). Our impression is that
-there is a strong tendency in software development to standardize on the simpler
-\textit{only \textsc{lf}} encoding for line breaks, and we suggest that
-everybody use this encoding whenever possible.
+encodings of line breaks have been used (e.g.~CR+LF, LF+CR, only LF, or only CR.
+Currently, all Unix and Unix-derived systems use only LF as code for a
+line break, while software from Microsoft still uses a combination of
+CR and LF. Today, most software applications recognize both
+options, and are able to deal with either encoding of line breaks (until rather
+recently this was not the case, and using the wrong line breaks could lead to
+unexpected errors). Our impression is that there is a strong tendency in
+software development to standardize on the simpler \textsc{only LF}
+encoding for line breaks, and we suggest that everybody use this encoding
+whenever possible.
 
-A code unit is a sequence of bits used to encode character in a encoding. The
-issue with code units stems from the question how to separate a stream of binary
-ones and zero, i.e.~bits, into chunks representing Unicode characters. Depending
-on different use cases, the Unicode Standard offers three different approaches,
-called UTF-32, UTF-16 and UTF-8.\footnote{The letters UTF stand for either
-\textsc{Unicode Transformation Format} or \textsc{Universal Coded Character Set
-Transformation Format}, but the notion of \textit{transformation} is a legacy
-notion that does not have meaning anymore. Nevertheless, the designation
-\textsc{UTF} (in capitals) has become an official standard designation, but
-should probably best be read as simply \textit{Unicode Format}.} The details of
-this issue is extensively explained in section 2.5 of the Unicode core
-specification. Basically, \textsc{UTF-32} encodes each character in 32 bits (32
-\textit{bi}nary un\textit{ts}, i.e.~32 zeros or ones) and is the most
-disk-space-consuming variant of the three. However, it is the most efficient
-encoding processing-wise, because the computer simply has to separate each
-character after 32 bits. In contrast, \textsc{UTF-16} uses only 16 bits per
-character, which is sufficient for the large majority of Unicode characters, but
-not for all of them. A special system of `surrogates' is defined within the
-Unicode Standard to deal with these additional characters. The effect is a more
-disk-space efficient encoding (approximately half the size), while adding a
-limited computational overhead. Finally, \textsc{UTF-8} is a more complex system
-that dynamically encodes each character with the minimally necessary number of
-bits, choosing either 8, 16 or 32 bits depending on the character. This
-represents again a strong reduction in space (particularly due to the high
-frequency of data using ASCII characters that need only 8 bits) at the expense
-of even more computation necessary to process such strings. However, because of
-the ever growing computational power of modern machines, the processing overhead
-is in most practical situations a non-issue, while saving on space is still
-useful, particularly for sending texts over the Internet. As an effect, UTF-8
-has become the dominant encoding on the World Wide Web. We suggest that
-everybody uses UTF-8 as their default encoding.
+\subsubsection*{Code units}
+
+The issue with \textsc{code units} stems from the question how to separate a
+stream of binary ones and zero, i.e.~bits, into chunks representing Unicode
+characters. A code unit is the sequence of bits used to encode a single
+character in an encoding. Depending on different use cases, the Unicode Standard
+offers three different approaches, called UTF-32, UTF-16 and UTF-8.\footnote{The
+letters UTF stand for either \textsc{Unicode Transformation Format} or
+\textsc{Universal Coded Character Set Transformation Format}, but the notion of
+\textit{transformation} is a legacy notion that does not have meaning anymore.
+Nevertheless, the designation \textsc{UTF} (in capitals) has become an official
+standard designation, but should probably best be read as simply Unicode
+Format.} The details of this issue is extensively explained in section 2.5 of
+the Unicode Core Specification \cite{Unicode2014}.
+
+Basically, \textsc{UTF-32} encodes each character in 32 bits (32 \textit{bi}nary
+uni\textit{ts}, i.e.~32 zeros or ones) and is the most disk-space-consuming
+variant of the three. However, it is the most efficient encoding
+processing-wise, because the computer simply has to separate each character
+after 32 bits.
+
+In contrast, \textsc{UTF-16} uses only 16 bits per character, which is
+sufficient for the large majority of Unicode characters, but not for all of
+them. A special system of \textsc{surrogates} is defined within the Unicode Standard to
+deal with these additional characters. The effect is a more disk-space efficient
+encoding (approximately half the size), while adding a limited computational
+overhead to manage the surrogates.
+
+Finally, \textsc{UTF-8} is a more complex system that dynamically
+encodes each character with the minimally necessary number of bits, choosing
+either 8, 16 or 32 bits depending on the character. This represents again a
+strong reduction in space (particularly due to the high frequency of data using
+ASCII characters, which need only 8 bits) at the expense of even more computation
+necessary to process such strings. However, because of the ever growing
+computational power of modern machines, the processing overhead is in most
+practical situations a non-issue, while saving on space is still useful,
+particularly for sending texts over the Internet. As an effect, UTF-8 has become
+the dominant encoding on the World Wide Web. We suggest that everybody uses
+UTF-8 as their default encoding.
 
@@
 memory, which is known as \textsc{endianness}. The details of this issue go
-beyond the scope of this paper. It suffices to realize that there is a
+beyond the scope of this book. It suffices to realize that there is a
 difference between \textsc{big-endian} (BE) storage and \textsc{little-endian}
 (LE) storage. The Unicode Standard offers a possibility to explicitly indicate
-what kind of storage is used by starting a file with a so-called `Byte Order
-Mark' (BOM). However, the Unicode Standard does not require the usage of BOM,
+what kind of storage is used by starting a file with a so-called \textsc{Byte Order
+Mark} (BOM). However, the Unicode Standard does not require the usage of BOM,
 preferring other non-Unicode methods to signal to computers which kind of
-endianness is used. This issue only arises with UTF-32 and UTF-16 encodings, and
-when using the preferred UTF-8, using a BOM is theoretically possible, but
+endianness is used. This issue only arises with UTF-32 and UTF-16 encodings.
+When using the preferred UTF-8, using a BOM is theoretically possible, but
 strongly dispreferred according to the Unicode Standard. We suggest that
@@
 
-Together with the topic of normalization (NFC vs.~NFD, see Pitfall 5 above),
+\
+
+\noindent Together with the topic of normalization (NFC vs.~NFD, see 
+Section~\ref{pitfall-canonical-equivalence} above),
 this leads to the following nicely cryptic file-encoding recommendation, namely
-to preferably use \textsc{UTF-8 in NFC without BOM with LF line breaks},
+to preferably use UTF-8 in NFC without BOM with LF line breaks,
 whenever possible.

### 2015-06-06 chapters/ipa_meets_unicode.tex
update section names
updated IPA code block pitfall

@@
 
-\section{The International Phonetic Alphabet}
+\section{The International Phonetic Alphabet (IPA)}
 \label{the-international-phonetic-alphabet}
@@
 The ambivalent nature of IPA glyphs arises because, on the one hand, the IPA
-uses Latin-based glyphs like , or . From this perspective, the IPA seems to be
-just another orthographic tradition using Latin characters, all of which do not
-get a special treatment within the Unicode Standard (just like e.g.~the French,
-German, or Danish orthographic traditions do not have a special status). On the
-other hand, the IPA uses many special symbols (like turned , mirrored and/or
-extended Latin glyphs ) not found in any other Latin-based writing system. For
-this reason, and already in the first version of the Unicode Standard (Version
-1.0 from 1991), a special block with code points, called \textsc{IPA Extensions}
-was included. As noted in Section 4, Pitfall 1, the Unicode Standard code space
-is subdivided into character blocks, which generally encode characters from a
+uses Latin-based glyphs like <a>, <b> or <p>. From this perspective, the IPA
+seems to be just another orthographic tradition using Latin characters, all of
+which do not get a special treatment within the Unicode Standard (just like
+e.g.~the French, German, or Danish orthographic traditions do not have a special
+status). On the other hand, the IPA uses many special symbols (like turned,
+mirrored and/or extended Latin glyphs) not found in any other Latin-based
+writing system. For this reason, and already in the first version of the Unicode
+Standard (Version 1.0 from 1991), a special block with code points, called
+\textsc{IPA Extensions} was included.
+
+As explained in Section~\ref{pitfall-blocks}, the Unicode Standard code space is
+subdivided into character blocks, which generally encode characters from a
 single script. However, as is illustrated by the IPA, characters that form a
@@
 blocks. With its diverse collection of symbols from various scripts and
-diacritics, the IPA is spread across over 13 blocks in the Unicode
+diacritics, the IPA is spread across 13 blocks in the Unicode
 Standard:\footnote{This number of blocks depends on whether only IPA-sanctioned
 symbols are counted or if the phonetic symbols commonly found in the literature
-are also included, see Moran 2012, Appendix C.}
-
+are also included, see \cite[Appendix C]{Moran2012}}
 \begin{itemize}
-	\item Basic Latin (30 characters), e.g. <a, b, c, d, e> 
-	\item Latin-1 Supplement (4 characters): <æ, ç, ð, ø\textgreater{} 
-	\item Latin Extended-A (3 characters): <ħ, ŋ, œ> 
-	\item Latin Extended-B (5 characters): <ǀ, ǁ, ǂ, ǃ, ȵ> 
-	\item IPA Extensions (70 characters), e.g.: <ɐ, ɑ, ɔ> 
-	\item Spacing Modifier Letters (20 characters), e.g.: <ʰ ʷ ˥> 
-	\item Combining Diacritical Marks (33 characters), e.g.: <̝ ̥ ̪> 
-	\item Greek and Coptic (3 characters): <β, θ, χ> 
-	\item Phonetic Extensions (2 characters): <ᴅ, ᴴ> 
-	\item Phonetic Extensions Supplement (3 characters): <ᶑ , ᶾ, ᶣ> 
+	\item Basic Latin (30 characters), e.g. <a b c d e \ldots> 
+	\item Latin-1 Supplement (4 characters): <æ ç ð ø> 
+	\item Latin Extended-A (3 characters): <ħ ŋ œ> 
+	\item Latin Extended-B (5 characters): <ǀ ǁ ǂ ǃ ȵ> 
+	\item IPA Extensions (70 characters), e.g. <ɐ ɑ ɔ \ldots> 
+	\item Spacing Modifier Letters (20 characters), e.g. <ʰ ʷ ˥ \ldots> 
+	\item Combining Diacritical Marks (33 characters), e.g. <{\large \ \ ̝\ \ ̥\ \ ̪ }\ldots> 
+	\item Greek and Coptic (3 characters): <β θ χ> 
+	\item Phonetic Extensions (2 characters): <{\small \fontspec{CharisSIL}ᴅ ᴴ}> 
+	\item Phonetic Extensions Supplement (3 characters): <{\small \fontspec{CharisSIL}ᶑ ᶾ ᶣ}> 
 	\item Superscripts and Subscripts (1 character): <ⁿ> 
-	\item Arrows (2 characters): \textless{}↑, ↓\textgreater{} 
-	\item Latin Extended-C (1 character): <ⱱ> 
+	\item Arrows (4 characters): <↑ ↓ ↗ ↘>
+	\item Latin Extended-C (1 character): <{\small \fontspec{CharisSIL}ⱱ}> 
 \end{itemize}
@@
 
-\section{Pitfall: Ligature and digraphs}
+\section{Pitfall: Ligatures and digraphs}
 \label{pitfall-ligatures-digraphs}

### 2015-06-06 chapters/ipa_meets_unicode.tex
Updated Pitfall IPA homoglyphs

@@
 symbols are counted or if the phonetic symbols commonly found in the literature
-are also included, see \cite[Appendix C]{Moran2012}}
+are also included, see \cite[Appendix C]{Moran2012}.}
+
 \begin{itemize}
-	\item Basic Latin (30 characters), e.g. <a b c d e \ldots> 
+	\item Basic Latin (30 characters), e.g. <a b c d e> 
 	\item Latin-1 Supplement (4 characters): <æ ç ð ø> 
@@
 	\item Latin Extended-B (5 characters): <ǀ ǁ ǂ ǃ ȵ> 
-	\item IPA Extensions (70 characters), e.g. <ɐ ɑ ɔ \ldots> 
-	\item Spacing Modifier Letters (20 characters), e.g. <ʰ ʷ ˥ \ldots> 
-	\item Combining Diacritical Marks (33 characters), e.g. <{\large \ \ ̝\ \ ̥\ \ ̪ }\ldots> 
+	\item IPA Extensions (70 characters), e.g. <ɐ ɑ ɔ> 
+	\item Spacing Modifier Letters (20 characters), e.g. <ʰ ʷ ˥> 
+	\item Combining Diacritical Marks (33 characters), e.g. <{\large \ \ ̝\ \ ̥\ \ ̪ }> 
 	\item Greek and Coptic (3 characters): <β θ χ> 
@@
 font rendering). For example, a speaker of Russian should ideally not use the
-\textsc{cyrillic small letter a} at code point \uni{0430} for IPA
-transcriptions, but instead the \textsc{latin small letter a} at code point
+<а> \textsc{cyrillic small letter a} at code point \uni{0430} for IPA
+transcriptions, but instead use the <a> \textsc{latin small letter a} at code point
 \uni{0061}, although visually they are mostly indistinguishable, and the
-Cyrillic character is more easily typed on a Cyrillic keyboard. Another example
-we commonly encounter is the use of \textsc{latin small letter g} at \uni{0067},
-instead of the sanctioned Unicode Standard IPA character for the voiced velar
-stop \textsc{latin small letter script g} at \uni{0261}. One begins to question
-whether this issue is at all apparent to the working linguist, or if they simply
-use the \uni{0067} because it is easily keyboarded and thus saves time, whereas
-the latter must be cumbersomely inserted as a special symbol in most
-software.\footnote{This issue was recently addressed by the International
-Phonetic Association, which has taken the stance that both the keyboard
-\textsc{latin small letter g} and the \textsc{latin small letter script g} are
-valid input characters for the voiced velar plosive. Unfortunately, this
-decision further introduces ambiguity for linguists trying to adhere to a strict
-Unicode Standard IPA encoding.}
+Cyrillic character is more easily typed on a Cyrillic keyboard.
 
-Furthermore, on the one hand even linguists are unlikely to distinguish between
+Furthermore, even linguists are unlikely to distinguish between
 the <ə> \textsc{latin small letter schwa} at code point \uni{0259} and <ǝ>
-\textsc{latin small letter turned e} at \uni{01DD}. On the other hand,
+\textsc{latin small letter turned e} at \uni{01DD}. Conversely,
 non-linguists are unlikely to distinguish any semantic difference between an
-open back unrounded vowel <ɑ>, the \textsc{latin small letter alpha} at
-\uni{0251}, and the open front unrounded vowel <a>, \textsc{latin small letter
+open back unrounded vowel <ɑ> \textsc{latin small letter alpha} at
+\uni{0251}, and the open front unrounded vowel <a> \textsc{latin small letter
 a} at \uni{0061}. But even among linguists this distinction leads to problems.
@@
 one to ask if the many characters that are used in phonological description
-reflects a \textsc{transcriptional bias}. The common use of in transcriptions
+reflects a transcriptional bias. The common use of <a> in transcriptions
 could be in part due to the ease of typing the letter on an English keyboard (or
 for older descriptions, the typewriter). We found it to be exceedingly rare that
-a linguist uses for a low back unrounded vowel.\footnote{One example is
+a linguist uses <ɑ> for a low back unrounded vowel.\footnote{One example is
 \citet[75]{Vidal2001a}, in which the author states: ``The definition of Pilagá
@@
 Hence, the characterization of /a/ and /o/ as a natural class of (i.e.,
-{[}+back{]} vowels), as opposed to /i/ and /e/.''} They simply use $<$a$>$ as
-long as there is no opposition to $<$ɑ$>$.\footnote{See Thomason's Language Log
-post, ``Why I don't love the International Phonetic Alphabet'', at:
-\url{http://itre.cis.upenn.edu/\textasciitilde{}myl/languagelog/archives/005287.html}.}
+{[}+back{]} vowels), as opposed to /i/ and /e/.''} They simply use <a> as
+long as there is no opposition to <ɑ>.\footnote{See Thomason's Language Log
+post, ``Why I don't love the International Phonetic Alphabet'' at:
+\url{http://itre.cis.upenn.edu/~myl/languagelog/archives/005287.html}.}
+
 Making things even more problematic, there is an old typographic tradition that
-the double-story uses a single-story in italics. This leads to the unfortunate
-effect that in most well-designed fonts the italics of and use the same glyph.
-If this distinction has to be kept upright in italics, the only solution we can
-currently offer is to use \textit{slanted} glyphs (i.e.~artificially italicized
-glyphs) instead of real italics (i.e.~special italics glyphs designed by a
-typographer).\footnote{For example, the widely used IPA font Doulos SIL
+the double-story <a> uses a single-story <ɑ> in italics. This leads to the
+unfortunate effect that in most well-designed fonts the italics of <a> and <ɑ>
+use the same glyph. If this distinction has to be kept upright in italics, the
+only solution we can currently offer is to use \textsc{slanted} glyphs
+(i.e.~artificially italicized glyphs) instead of real italics (i.e.~special
+italics glyphs designed by a typographer).\footnote{For example, the widely used
+IPA font Doulos SIL
 (\url{http://scripts.sil.org/cms/scripts/page.php?item\_id=DoulosSIL}) does not
@@
 Word, to produce slanted glyphs instead. That particular combination of font and
-software application will thus lead to the desired effect. However, note that
-when the text is transferred to another font (i.e.~one that includes real
-italics) and/or to another software application (like Apple Pages, which does
-not perform slanting), then this visual appearance will be lost. In this case we
-are thus still in the pre-Unicode situation in which the choice of font and
-rendering software actually matters. The ideal solution from a linguistic point
-of view would be the introduction of a new IPA code point for a different kind
-of which explicitly specifies that it should still be rendered as a double-story
-character when italicized. After informal discussion with various Unicode
-players, our impression is that this highly restricted problem is not
-sufficiently urgent to introduce even more <a>-like characters in Unicode (which
-already lead to much confusion, see Section \ref{pitfall-homoglyphs}). This is a
-clear situation in which the Unicode Consortium is not just thinking about
-linguists, but has a more wide-ranging practical view to consider.}
+software application will thus lead to the desired effect distinguishing <a>
+from <ɑ>. However, note that when the text is transferred to another font
+(i.e.~one that includes real italics) and/or to another software application
+(like Apple Pages, which does not perform slanting), then this visual appearance
+will be lost. In this case we are thus still in the pre-Unicode situation in
+which the choice of font and rendering software actually matters. The ideal
+solution from a linguistic point of view would be the introduction of a new IPA
+code point for a different kind of which explicitly specifies that it should
+still be rendered as a double-story character when italicized. After informal
+discussion with various Unicode players, our impression is that this highly
+restricted problem is not sufficiently urgent to introduce even more <a>-like
+characters in Unicode (which already lead to much confusion, see Section~
+\ref{pitfall-homoglyphs}).}
 
 Some other homoglyphs related to encoding IPA in the Unicode Standard are:
+
 \begin{itemize}
 	\item The uses of the apostrophe has led to long discussions on the Unicode
-       Standard email list. An English keyboard inputs
-       \textless{}`\textgreater{} APOSTROPHE at \uni{0027}, although the
-       \textit{preferred} Unicode apostrophe is the \textless{}'\textgreater{}
-       \textsc{right single quotation mark} at \uni{2019}. Yet the glottal
-       stop/glottalization/ejective marker is another completely different
-       character, the \textsc{modifier letter apostrophe} at \uni{02BC}, but
-       unfortunately looks mostly highly similar to \uni{2019}. 
-	\item There is ambiguous encoding of IPA segments within the Unicode Standard.
-       An example is the \uni{02C1} \textsc{modifier letter reversed glottal
-       stop} vs the \uni{02E4} \textsc{modifier letter small reversed glottal
-       stop} . Both are denoted in the Unicode Standard as the
-       \textsc{pharyngealized diacritic} and both appear in various resources
-       representing phonetic data online. This is thus an example for which the
-       Unicode Standard does not solve the linguistic standardization problem. 
+       Standard email list. An English keyboard inputs <{\fontspec{Monaco}'}> \textsc{apostrophe}
+       at \uni{0027}, although the preferred Unicode apostrophe is the <\ ' >
+       \textsc{right single quotation mark} at \uni{2019}.
+     \item The glottal stop/glottalization/ejective marker is another completely
+        different character <{\large ʼ}>, the \textsc{modifier letter apostrophe} at
+        \uni{02BC}, but unfortunately looks mostly highly similar to
+        \uni{2019}. 
+	\item Another problem is the <ˁ> \textsc{modifier letter reversed
+       glottal stop} at \uni{02C1} vs.\@ the <ˤ> \textsc{modifier
+       letter small reversed glottal stop} at \uni{02E4}. Both are denoted in
+       the Unicode Standard as the \textsc{pharyngealized diacritic} and both
+       appear in various resources representing phonetic data online. This is
+       thus an example for which the Unicode Standard does not solve the
+       linguistic standardization problem. 
 	\item There is at least one case in which the character name assigned by the
-       Unicode Consortium does not match the IPA's description: in the Unicode
-       Standard at \uni{01C3} is labeled \textsc{latin letter retroflex click},
-       but in IPA is an alveolar or postalveolar click (not retroflex). This
-       naming is probably best seen as a simple error in the Unicode Standard.
+       Unicode Consortium does not match the IPA's description. In the Unicode
+       Standard the <ǃ> at \uni{01C3} is labeled \textsc{latin letter retroflex
+       click}, but in IPA is an alveolar or postalveolar click (not retroflex).
+       This naming is probably best seen as a simple error in the Unicode
+       Standard. This character is of course often simply typed as <!>
+       \textsc{exclamation mark} at \uni{0021}.
 \end{itemize}
@@
 content
+
+Another example
+we commonly encounter is the use of <g> \textsc{latin small letter g} at \uni{0067},
+instead of the Unicode Standard IPA character for the voiced velar
+stop <ɡ> \textsc{latin small letter script g} at \uni{0261}. One begins to question
+whether this issue is at all apparent to the working linguist, or if they simply
+use the \uni{0067} because it is easily keyboarded and thus saves time, whereas
+the latter must be cumbersomely inserted as a special symbol in most
+software.
+
 \begin{itemize}
   \item tone marks (prefer tone symbols)
-  \item marking of <g>
+  \item marking of <g>\footnote{This issue was recently addressed by the International
+Phonetic Association, which has taken the stance that both the keyboard
+\textsc{latin small letter g} and the \textsc{latin small letter script g} are
+valid input characters for the voiced velar plosive. Unfortunately, this
+decision further introduces ambiguity for linguists trying to adhere to a strict
+Unicode Standard IPA encoding.}
   \item marking of voiceless with ring diacritic

### 2015-06-06 chapters/ipa_meets_unicode.tex
started working on pitfall IPA homoglyphs

@@
 restricted problem is not sufficiently urgent to introduce even more <a>-like
-characters in Unicode (which already lead to much confusion, see Section~
-\ref{pitfall-homoglyphs}).}
+characters in Unicode (which already lead to much confusion, see Section~\ref{pitfall-homoglyphs}).}
 
@@
        glottal stop} at \uni{02C1} vs.\@ the <ˤ> \textsc{modifier
-       letter small reversed glottal stop} at \uni{02E4}. Both are denoted in
-       the Unicode Standard as the \textsc{pharyngealized diacritic} and both
+       letter small reversed glottal stop} at \uni{02E4}. Both 
        appear in various resources representing phonetic data online. This is
-       thus an example for which the Unicode Standard does not solve the
+       thus a clear example for which the Unicode Standard does not solve the
        linguistic standardization problem. 
@@
        Standard the <ǃ> at \uni{01C3} is labeled \textsc{latin letter retroflex
-       click}, but in IPA is an alveolar or postalveolar click (not retroflex).
-       This naming is probably best seen as a simple error in the Unicode
-       Standard. This character is of course often simply typed as <!>
-       \textsc{exclamation mark} at \uni{0021}.
-\end{itemize}
+       click}, but in IPA that glyph is used for an alveolar or postalveolar
+       click (not retroflex). This naming is probably best seen as a simple
+       error in the Unicode Standard. This character is of course often simply
+       typed as <!> \textsc{exclamation mark} at \uni{0021}. \end{itemize}
 
@@
 It is not just the Unicode Standard that offers multiple options for encoding 
-the IPA. Even the IPA itself is not consistent in how specific meaningful 
+the IPA. Even the IPA itself offers some flexibility in how specific meaningful 
 transcriptions have to be encoded. There are a few cases in which the IPA 
 explicitly allows for different options of transcribing the same phonetic 
-content
+content. This is understandable from a transcriber's point of view, but it is 
+not good interoperability between resources supposedly written in IPA. We 
+consider it crucial to distinguish between ``lax'' IPA, in which the 
+phonetically-trained reader should be able to understand the transcription, and 
+``strict'' IPA, which should be standardized on a single encoding so search will 
+work across resources.
 
-Another example
-we commonly encounter is the use of <g> \textsc{latin small letter g} at \uni{0067},
-instead of the Unicode Standard IPA character for the voiced velar
-stop <ɡ> \textsc{latin small letter script g} at \uni{0261}. One begins to question
-whether this issue is at all apparent to the working linguist, or if they simply
-use the \uni{0067} because it is easily keyboarded and thus saves time, whereas
-the latter must be cumbersomely inserted as a special symbol in most
-software.
-
-\begin{itemize}
-  \item tone marks (prefer tone symbols)
-  \item marking of <g>\footnote{This issue was recently addressed by the International
+Another example we commonly encounter is the use of <g> \textsc{latin small
+letter g} at \uni{0067}, instead of the Unicode Standard IPA character for the
+voiced velar stop <ɡ> \textsc{latin small letter script g} at \uni{0261}. One
+begins to question whether this issue is at all apparent to the working
+linguist, or if they simply use the \uni{0067} because it is easily keyboarded
+and thus saves time, whereas the latter must be cumbersomely inserted as a
+special symbol in most software.\footnote{This issue was recently addressed by the International
 Phonetic Association, which has taken the stance that both the keyboard
@@
 Unicode Standard IPA encoding.}
-  \item marking of voiceless with ring diacritic
+
+\begin{itemize}
+  \item tone marks (prefer tone symbols)
+  \item marking of <g>  \item marking of voiceless with ring diacritic
   \item tie bar

### 2015-06-06 chapters/ipa_meets_unicode.tex
finish pitfall IPA homoglyphs

@@
 \section{Pitfall: Homoglyphs in IPA}
+\label{pitfall-homoglyphs-in-IPA}
 
 It is not just the Unicode Standard that offers multiple options for encoding
-the IPA. Even the IPA itself offers some flexibility in how specific meaningful 
+the IPA. Even the IPA specification itself offers some flexibility in how
 transcriptions have to be encoded. There are a few cases in which the IPA
@@
 content. This is understandable from a transcriber's point of view, but it is
-not good interoperability between resources supposedly written in IPA. We 
-consider it crucial to distinguish between ``lax'' IPA, in which the 
-phonetically-trained reader should be able to understand the transcription, and 
-``strict'' IPA, which should be standardized on a single encoding so search will 
-work across resources.
+not acceptable for interoperability between resources written in IPA. We
+consider it crucial to distinguish between ``lax'' IPA, for which it is
+sufficient that any phonetically-trained reader is able to understand the
+transcription, and ``strict'' IPA, which should be standardized on a single
+unique encoding for each sound, so search will work across resources.
 
-Another example we commonly encounter is the use of <g> \textsc{latin small
-letter g} at \uni{0067}, instead of the Unicode Standard IPA character for the
-voiced velar stop <ɡ> \textsc{latin small letter script g} at \uni{0261}. One
-begins to question whether this issue is at all apparent to the working
-linguist, or if they simply use the \uni{0067} because it is easily keyboarded
-and thus saves time, whereas the latter must be cumbersomely inserted as a
-special symbol in most software.\footnote{This issue was recently addressed by the International
-Phonetic Association, which has taken the stance that both the keyboard
-\textsc{latin small letter g} and the \textsc{latin small letter script g} are
-valid input characters for the voiced velar plosive. Unfortunately, this
-decision further introduces ambiguity for linguists trying to adhere to a strict
-Unicode Standard IPA encoding.}
+The first case in which the IPA allows for different encodings is the question
+of how to transcribe tone. There is an old tradition to use diacritics on vowels
+to mark different tone levels, e.g. <ȅèée̋>.\footnote{There are at least two
+different Unicode homoglyphs for the low and high level tones, namely <\ ̀~>
+\textsc{combining grave tone mark} at \unif{0340} vs.\@ <\ ̀~> \textsc{combining
+grave accent} at \unif{0300} for low tone, and <\ ́~> \textsc{combining acute tone
+mark} at \unif{0341} vs.\@ <\ ́~> \textsc{combining acute accent} at \unif{0301}
+for high tone.} The IPA also proposes a less widespread used, but more
+consistent option of tone letters, e.g. <˥˦˧˨˩>. Tone letters in the IPA have
+five different levels, and sequences of these letters can be used to indicate
+contours. Well-designed fonts will even merge a sequence of tone letters into a
+contour. For example, compare the font Linux Libertine, which does not merge
+tone letters <{\fontspec{LinLibertineO}˥˨˧˩}>, with the font CharisSIL, which
+merges this sequence of four tone letters into a single contour <\charis{˥˨˧˩}>.
+For strict IPA encoding we propose to standardize on tone letters.
 
-\begin{itemize}
-  \item tone marks (prefer tone symbols)
-  \item marking of <g>  \item marking of voiceless with ring diacritic
-  \item tie bar
-\end{itemize}
+Second, we commonly encounter the use of <g> \textsc{latin small letter g} at
+\uni{0067}, instead of the Unicode Standard IPA character for the voiced velar
+stop <ɡ> \textsc{latin small letter script g} at \uni{0261}. One begins to
+question whether this issue is at all apparent to the working linguist, or if
+they simply use the \uni{0067} because it is easily keyboarded and thus saves
+time, whereas the latter must be cumbersomely inserted as a special symbol in
+most software. This issue was recently addressed by The International Phonetic
+Association has taken the stance that both the keyboard \textsc{latin small
+letter g} and the \textsc{latin small letter script g} are valid input
+characters for the voiced velar plosive. Unfortunately, this decision further
+introduces ambiguity for linguists trying to adhere to a strict Unicode Standard
+IPA encoding. For strict IPA encoding we propose to standardize on the more 
+idiosyncratic \textsc{latin small letter script g} at \uni{0261}.
+
+Third, for marking marking of voiceless pronunciation of voiced segments the IPA
+uses the ring diacritic. Originally, the ring should be placed below the 
+base character, like in <m̥>, using the \textsc{combining ring below} at \uni{0325}. 
+However, in letters with long descenders the IPA also allows to put the ring 
+above the base, like in <ŋ̊>, using the \textsc{combining ring above} at 
+\uni{030A}. Yet, proper font design does not have any problem with rendering 
+the ring below the base character, like in <ŋ̥>, so for strict IPA encoding we 
+propose to standardize on the ring below.
+
+Likewise, the tie bar to indicate affricates, doubly articulated consonants or
+diphthongs can be placed either below or on top of the base characters, e.g.
+<\charis{t͡s}> or <\charis{t͜s}>. For strict IPA encoding we propose to standardize on the tie bar
+above the base characters, using Unicode \textsc{combining double inverted
+breve} at \uni{0361}. Note that this Unicode character is placed between the two 
+base characters to be combined. In principle, it is possible to combine more 
+than two base characters by repeating the tie bar, like in <a͡o͡u>. If really
+necessary, we consider this possible, even though the rendering will never look 
+good.
 
@@
 and ligatures. Multigraphs are groups of characters (in the context of IPA e.g.
-< tʃ > or < ou >) while ligatures are single characters (e.g. <ʧ> \textsc{latin
+<tʃ> or <ou>) while ligatures are single characters (e.g. <ʧ> \textsc{latin
 small letter tesh digraph} at \uni{02A7}). Ligatures arose in the context of

### 2015-06-06 chapters/ipa_meets_unicode.tex
updated pitfall ligatures

@@
 \begin{itemize} 
-	\item <ʣ> \textsc{latin small letter dz digraph} at \uni{02A3} (use <d> + <z> instead) 
-	\item <ʧ> \textsc{latin small letter tesh digraph} at \uni{02A7} (use <t> + <ʃ> instead) 
-	\item <ʩ> \textsc{latin small letter feng digraph} at \uni{02A9} (use <f> + <ŋ> instead) 
+	\item <ʣ> \textsc{latin small letter dz digraph} at \uni{02A3} 
+	  (use <dz> instead) 
+    \item <ʤ> \textsc{latin small letter dezh digraph} at \uni{02A4}
+      (use <dʒ> instead)
+    \item <ʥ> \textsc{latin small letter dz digraph with curl} at \uni{02A5}
+      (use <dʑ>)
+    \item <ʦ> \textsc{latin small letter ts digraph} at \uni{02A6} 
+      (use <ts> instead)
+	\item <ʧ> \textsc{latin small letter tesh digraph} at \uni{02A7} 
+	  (use <tʃ> instead) 
+    \item <ʨ> \textsc{latin small letter tc digraph with curl} at \uni{02A8}
+      (use <tɕ>)
+   	\item <ʩ> \textsc{latin small letter feng digraph} at \uni{02A9}
+	  (use <fŋ> instead) 
 \end{itemize}

### 2015-06-06 chapters/ipa_meets_unicode.tex
update IPA intro

@@
 addressing the need for a universal computing environment for writing systems
-and their computational complexity is discussed in Simons 1989. For a long time,
-linguists (like all other computer users) were limited to ASCII-encoded 7-bit
-characters, which only includes Latin characters, numbers and some punctuation
-and symbols. Restricted to these standard character sets that lacked IPA support
-or other language-specific graphemes that they needed, linguists devised their
-own solutions \citep{BirdSimons2003}. For example, some chose to represent
-unavailable graphemes with substitutes, e.g.~the combination of to represent .
-Tech-savvy linguists redefined selected characters from a character encoding by
-mapping custom made fonts to those code points. However, one linguist's
-electronic text would not render properly on another linguist's computer without
-access to the same font. Further, if two character encodings defined two
-character sets differently, then data could not be reliably and correctly
-displayed. This is a common example of the non-interoperability of data and data
-formats.
+and their computational complexity is discussed in \citet{Simons1989}. For a long
+time, linguists (like all other computer users) were limited to ASCII-encoded
+7-bit characters, which only includes Latin characters, numbers and some
+punctuation and symbols. Restricted to these standard character sets that lacked
+IPA support or other language-specific graphemes that they needed, linguists
+devised their own solutions \citep{BirdSimons2003}. For example, some chose to
+represent unavailable graphemes with substitutes, e.g.~the combination of <ng> to
+represent <ŋ>. Tech-savvy linguists redefined selected characters from a character
+encoding by mapping custom made fonts to those code points. However, one
+linguist's electronic text would not render properly on another linguist's
+computer without access to the same font. Further, if two character encodings
+defined two character sets differently, then data could not be reliably and
+correctly displayed. This is a common example of the non-interoperability of
+data and data formats.
 
@@
 for speech technology and as an encoding system in computational linguistics.
-Eventually, ASCII-encoding of the IPA became depreciated through the advent of
+Eventually, ASCII-encoding of the IPA became deprecated through the advent of
 the Unicode Standard. Note however that many popular software packages used for
-linguistic analyses still require ASCII input,
-e.g.~RuG/L04\footnote{\url{http://www.let.rug.nl/kleiweg/L04/}} and
-SplitsTree4.\footnote{\url{http://www.splitstree.org/}}
+linguistic analyses still require ASCII input, e.g.~RuG/L04 and
+SplitsTree4.\footnote{See \url{http://www.let.rug.nl/kleiweg/L04/} and
+\url{http://www.splitstree.org/}, respectively}
 
-Still, there are a few pitfalls to be aware of when using the Unicode Standard
+There are a few pitfalls to be aware of when using the Unicode Standard
 to encode IPA. As we have said before, from a linguistic perspective it might

### 2015-06-06 chapters/ipa_meets_unicode.tex
updating IPA diacritics pitfall

@@
 transcription, and ``strict'' IPA, which should be standardized on a single
-unique encoding for each sound, so search will work across resources.
+unique encoding for each sound, so search will work across resources. We are 
+aware of the following double options in the IPA, which will be discussed in 
+turn below:
 
-The first case in which the IPA allows for different encodings is the question
+\begin{itemize}
+  \item The marking of tone
+  \item The marking of <g>
+  \item The diacritic for voiceless
+  \item The tie bar to indicate a close bond between sounds
+\end{itemize}
+
+\
+
+\noindent The first case in which the IPA allows for different encodings is the question
 of how to transcribe tone. There is an old tradition to use diacritics on vowels
@@
 <\ \ {\large  ̧}> \textsc{combining cedilla} at \uni{0327}, 
-also if such a decompotion is not intended, because it is meaningless in IPA.
+also if such a decomposition is not intended, because it is meaningless in IPA.
 
 \section{Pitfall: Different notions of diacritics}
-\label{pitfall-the-ipa-notion-of-diacritics-is-not-the-same-as-the-unicode-standards-notion-of-diacritics}
+\label{pitfall-different-notions-of-diacritics}
 
-Another pitfall is diacritics. The problem is that the meaning of the term
-\textsc{diacritics} as used by the IPA is not the same as it used in the Unicode
-Standard. Specifically, diacritics in the IPA-sense are either so-called
-\textsc{Combining Diacritical Marks} or \textsc{Spacing Modifier Letters} in the
-Unicode Standard. Crucially, Combining Diacritical Marks are by definition
-combined with the character before them (to form so-called default grapheme
-clusters, see Section 3). In contrast, Spacing Modifier Letters are by
-definition \emph{not} combined into grapheme clusters with the preceding
-character, but simply treated as separate letters. In the context of the IPA,
-the following IPA-diacritics are actually Spacing Modifier Letters in the
-Unicode Standard:
+Another pitfall relates the concept of what are diacritics. The problem is that
+the meaning of the term diacritics as used by the IPA is not the same as is used
+in the Unicode Standard. Specifically, diacritics in the IPA-sense are either
+so-called \textsc{combining diacritical marks} or \textsc{spacing modifier
+letters} in the Unicode Standard. Crucially, Combining Diacritical Marks are by
+definition combined with the character before them (to form so-called default
+grapheme clusters, see Section~\ref{the-unicode-approach}). In contrast, Spacing
+Modifier Letters are by definition \emph{not} combined into grapheme clusters
+with the preceding character, but simply treated as separate letters. In the
+context of the IPA, the following IPA-diacritics are actually Spacing Modifier
+Letters in the Unicode Standard:
 
 \begin{itemize}
-	\item Length marks, namely <ː> \textsc{modifier letter triangular colon} at \uni{02D0} and <ˑ> \textsc{modifier letter half triangular colon} at \uni{02D1}. 
-	\item Tone letters, like <˥> \textsc{modifier letter extra-high tone bar} at \uni{02E5}, and others like this. 
-	\item Superscript letters, like <ʰ> \textsc{modifier letter small h} at \uni{02B0} or <ˤ> \textsc{modifier letter small reversed glottal stop} at \uni{02E4}, and others like this. 
-	\item <˞> \textsc{modifier letter rhotic hook} at \uni{02DE}. 
+	\item[] Length marks, namely 
+	\begin{itemize}
+	  \item[] <ː> \textsc{modifier letter triangular colon} at \uni{02D0}
+	  \item[] <ˑ> \textsc{modifier letter half triangular colon} at \uni{02D1}
+	\end{itemize}
+
+	\item[] Tone letters, like 
+	\begin{itemize} 
+	  \item[] <˥> \textsc{modifier letter extra-high tone bar} at \uni{02E5}
+	  \item[] <˧> \textsc{modifier letter mid tone bar} at \uni{02E7}
+	  \item[] and others like this
+	\end{itemize}
+
+	\item[] Superscript letters, like 
+	\begin{itemize}
+	  \item[] <ʰ> \textsc{modifier letter small h} at \uni{02B0}
+	  \item[] <ˤ> \textsc{modifier letter small reversed glottal stop} at \uni{02E4}
+	  \item[] and many more like this
+	\end{itemize}
+
+	\item[] the rhotic hook <˞> \textsc{modifier letter rhotic hook} at \uni{02DE}
 \end{itemize}
@@
 for aspiration (more precisely post-aspiration following the base character) and
-pre-aspiration (preceding the base character). Note that there is a mechanism in
+pre-aspiration (preceding the base character). Note that there exists a mechanism in
 Unicode to force separate characters to be combined (namely by using the
 \textsc{zero width joiner} at \uni{200D}), but this seems to be a rather
-impractical, and probably not enforceable solution to us.
+impractical, and probably not an enforceable solution to us.
 

### 2015-06-06 chapters/ipa_meets_unicode.tex
tried a different itemize formatting

@@
 (ii) it encodes different languages in separate data tables, instead of a
-universal alphabet, like IPA. SAMPA tables are derived from phonemes appearing
+universal alphabet, like IPA.\@ SAMPA tables are derived from phonemes appearing
 in several European languages that were developed as part of a European
@@
 There are a few pitfalls to be aware of when using the Unicode Standard
-to encode IPA. As we have said before, from a linguistic perspective it might
+to encode IPA.\@ As we have said before, from a linguistic perspective it might
 look like the Unicode Consortium is making incomprehensible decisions, but it is
@@
 symbols are counted or if the phonetic symbols commonly found in the literature
-are also included, see \cite[Appendix C]{Moran2012}.}
+are also included, see~\cite[Appendix~C]{Moran2012}.}
 
@@
 \citet[75]{Vidal2001a}, in which the author states: ``The definition of Pilagá
-/a/ as {[}+back{]} results from its behavior in certain phonological contexts.
+/a/ as [+back] results from its behavior in certain phonological contexts.
 For instance, uvular and pharyngeal consonants only occur around /a/ and /o/.
 Hence, the characterization of /a/ and /o/ as a natural class of (i.e.,
-{[}+back{]} vowels), as opposed to /i/ and /e/.''} They simply use <a> as
+[+back] vowels), as opposed to /i/ and /e/.''} They simply use <a> as
 long as there is no opposition to <ɑ>.\footnote{See Thomason's Language Log
@@
 It is not just the Unicode Standard that offers multiple options for encoding
-the IPA. Even the IPA specification itself offers some flexibility in how
+the IPA.\@ Even the IPA specification itself offers some flexibility in how
 transcriptions have to be encoded. There are a few cases in which the IPA
@@
 content. This is understandable from a transcriber's point of view, but it is
-not acceptable for interoperability between resources written in IPA. We
+not acceptable for interoperability between resources written in IPA.\@ We
 consider it crucial to distinguish between ``lax'' IPA, for which it is
@@
 
-\
-
-\noindent The first case in which the IPA allows for different encodings is the question
+The first case in which the IPA allows for different encodings is the question
 of how to transcribe tone. There is an old tradition to use diacritics on vowels
@@
 
-Likewise, the tie bar to indicate affricates, doubly articulated consonants or
+Finally, the tie bar to indicate affricates, doubly articulated consonants or
 diphthongs can be placed either below or on top of the base characters, e.g.
@@
 \begin{itemize} 
-	\item <ʣ> \textsc{latin small letter dz digraph} at \uni{02A3} 
-	  (use <dz> instead) 
-    \item <ʤ> \textsc{latin small letter dezh digraph} at \uni{02A4}
-      (use <dʒ> instead)
-    \item <ʥ> \textsc{latin small letter dz digraph with curl} at \uni{02A5}
+	\item[] <ʣ> \textsc{latin small letter dz digraph} at \uni{02A3} 
+	  (use <dz>) 
+    \item[] <ʤ> \textsc{latin small letter dezh digraph} at \uni{02A4}
+      (use <dʒ>)
+    \item[] <ʥ> \textsc{latin small letter dz digraph with curl} at \uni{02A5}
       (use <dʑ>)
-    \item <ʦ> \textsc{latin small letter ts digraph} at \uni{02A6} 
-      (use <ts> instead)
-	\item <ʧ> \textsc{latin small letter tesh digraph} at \uni{02A7} 
-	  (use <tʃ> instead) 
-    \item <ʨ> \textsc{latin small letter tc digraph with curl} at \uni{02A8}
+    \item[] <ʦ> \textsc{latin small letter ts digraph} at \uni{02A6} 
+      (use <ts>)
+	\item[] <ʧ> \textsc{latin small letter tesh digraph} at \uni{02A7} 
+	  (use <tʃ>) 
+    \item[] <ʨ> \textsc{latin small letter tc digraph with curl} at \uni{02A8}
       (use <tɕ>)
-   	\item <ʩ> \textsc{latin small letter feng digraph} at \uni{02A9}
-	  (use <fŋ> instead) 
+   	\item[] <ʩ> \textsc{latin small letter feng digraph} at \uni{02A9}
+	  (use <fŋ>) 
 \end{itemize}
@@
 \begin{itemize}
-	\item <ɮ> \textsc{latin small letter lezh} at \uni{026E} 
-	\item <œ> \textsc{latin small ligature oe} at \uni{0153} 
-	\item <ɶ> \textsc{latin letter small capital oe} at \uni{0276} 
-	\item <æ> \textsc{latin small letter ae} at \uni{00E6} 
+	\item[] <ɮ> \textsc{latin small letter lezh} at \uni{026E} 
+	\item[] <œ> \textsc{latin small ligature oe} at \uni{0153} 
+	\item[] <ɶ> \textsc{latin letter small capital oe} at \uni{0276} 
+	\item[] <æ> \textsc{latin small letter ae} at \uni{00E6} 
 \end{itemize}
@@
 \begin{itemize}
-  \item <\ {\large  ̡}\ > \textsc{combining palatalized hook below} at \uni{0321}
-  \item <\ \ {\large  ̢}> \textsc{combining retroflex hook below} at \uni{0322}
-  \item <\ \ {\large  ̵}> \textsc{combining short stroke overlay} at \uni{0335}
-  \item <\ \ {\large  ̷}> \textsc{combining short solidus overlay} at \uni{0337}
-  \item <\ \ {\large  ̴}> \textsc{combining tilde overlay} at \uni{0334}
-  \item <{\large ˞}> \textsc{modifier letter rhotic hook} at \uni{02DE}
+  \item[] <\ {\large  ̡}\ > \textsc{combining palatalized hook below} at \uni{0321}
+  \item[] <\ \ {\large  ̢}> \textsc{combining retroflex hook below} at \uni{0322}
+  \item[] <\ \ {\large  ̵}> \textsc{combining short stroke overlay} at \uni{0335}
+  \item[] <\ \ {\large  ̷}> \textsc{combining short solidus overlay} at \uni{0337}
+  \item[] <\ \ {\large  ̴}> \textsc{combining tilde overlay} at \uni{0334}
+  \item[] <{\large ˞}> \textsc{modifier letter rhotic hook} at \uni{02DE}
 \end{itemize}
@@
 <\ \ {\large  ̧}> \textsc{combining cedilla} at \uni{0327}, 
-also if such a decomposition is not intended, because it is meaningless in IPA.
+also if such a decomposition is not intended, because it is meaningless in 
+IPA.\@
 
@@
 \begin{itemize}
-	\item[] Length marks, namely 
+
+	\item[] Length marks, namely: 
 	\begin{itemize}
@@
 
-	\item[] Tone letters, like 
+	\item[] Tone letters, like: 
 	\begin{itemize} 
@@
 
-	\item[] Superscript letters, like 
+	\item[] Superscript letters, like:
 	\begin{itemize}
@@
 
-	\item[] the rhotic hook <˞> \textsc{modifier letter rhotic hook} at \uni{02DE}
+	\item[] The rhotic hook:
+	\begin{itemize}
+	  \item[] <˞> \textsc{modifier letter rhotic hook} at \uni{02DE}
+	\end{itemize}
+
 \end{itemize}
@@
 
+\subsubsection*{Canonical combining classes}
+
 In accordance with the absence of any specification of ordering in the IPA, the
@@
 position relative to the base character.\footnote{See
-\url{http://unicode.org/reports/tr44/\#Canonical\_Combining\_Class\_Values for a
-detailed description}.} When applying a Unicode normalization (NFC or NFD, see
-previous section), the diacritics in different positions are put in a specified
-order. This process therefore harmonizes the difference between different
-encodings, for example, of a midtone creaky voice vowel <ḛ̄>. This grapheme
-cluster can be encoded either as <e> + <̄> + <̰> or as <e> + <̰> + <̄> . To
-prevent this twofold encoding, the Unicode Standard specifies the second
-ordering as canonical (diacritics below before diacritics above).
+\url{http://unicode.org/reports/tr44/\#Canonical\_Combining\_Class\_Values} for a
+detailed description.} When applying a Unicode normalization (NFC or NFD, see
+Section~\ref{pitfall-canonical-equivalence}), the diacritics in different
+positions are put in a specified order. This process therefore harmonizes the
+difference between different encodings, for example, of a midtone creaky voice
+vowel <ḛ̄>. This grapheme cluster can be encoded either as <e> + <\ \ ̄> + <\ \
+̰> or as <e> + <\ \ ̰> + <\ \ ̄>. To prevent this twofold encoding, the Unicode
+Standard specifies the second ordering as canonical (in this case, diacritics
+below are put before diacritics above).
 
@@
 (e.g.~in <ɝ>) will not be decomposed, but strangely enough a cedilla (like in
-<ç>) will be decomposed. In general, Unicode decomposition does not behave like
-a feature decomposition as expected from a linguistic perspective. It is thus
-important to consider Unicode decomposition only as a technical procedure, and
-not assume that it is linguistically sensible.
+<ç>) will be decomposed (see Section~\ref{pitfall-missing-decomposition}). In
+general, Unicode decomposition does not behave like a feature decomposition as
+expected from a linguistic perspective. It is thus important to consider Unicode
+decomposition only as a technical procedure, and not assume that it is
+linguistically sensible.
+
+\subsubsection*{Diacritic ordering}
 
@@
 languages, \citet[540]{Moran2012} defines a set of diacritic ordering
-conventions.\footnote{The most recent version of these conventions is online:
+conventions.\footnote{The most recent version of these conventions is online at
 \url{http://phoible.github.io/conventions/}} The conventions are influenced by
@@
 consistent character sequences, e.g.~if one language description uses and
-another , when both are intended to be phonetically equivalent, then a decision
+another, when both are intended to be phonetically equivalent, then a decision
 to normalize to one form was taken. For example, when a character sequence
@@
 \begin{itemize}
-	\itemsep1pt\parskip0pt\parsep0pt 
-	\item \textsc{Spacing Modifier Letters ordering:} ( unreleased <̚> \textbar{} lateral release <ˡ> \textbar{} nasal release <ⁿ>) → ( palatalized <ʲ>) → ( labialized <ʷ>) → ( velarized <ˠ>) → ( pharyngealized <ˤ>) → ( aspirated <ʰ> \textbar{} ejective <ʼ>) → ( long <ː> \textbar{} half-long <ˑ>)
+	\item[] \textsc{Spacing Modifier Letters ordering:}
+	\begin{itemize}
+	  \item[→] (unreleased <\ \ ̚ > \textbar{} lateral release <ˡ> \textbar{} nasal release <ⁿ>)
+	  \item[→] (palatalized <ʲ>)
+	  \item[→] (labialized <ʷ>)
+	  \item[→] (velarized <ˠ>)
+	  \item[→] (pharyngealized <ˤ>)
+	  \item[→] (aspirated <ʰ> \textbar{} ejective <ʼ>)
+	  \item[→] (long <ː> \textbar{} half-long <ˑ>)
+	\end{itemize}
 \end{itemize}
@@
 \begin{itemize}
-	\itemsep1pt\parskip0pt\parsep0pt 
-	\item \textsc{Combining Diacritical Marks (below) ordering:} ( dental <t̪> \textbar{} laminal <t̻> \textbar{} apical <t̺>) → ( fronted <u̟> \textbar{} backed <e̠>) →( lowered <e̞> \textbar{} raised <e̝>) → ( ATR <e̘ e̙>) → ( voiced <s̬> \textbar{} voiceless <n̥> \textbar{} creaky voice <b̰> \textbar{} breathy voice <b̤>) → ( syllabic <n̩> \textbar{} non-syllabic <e̯>)
+	\item[] \textsc{Combining Diacritical Marks (below) ordering:}
+	\begin{itemize}	
+	  \item[→] (dental <t̪> \textbar{} laminal <t̻> \textbar{} apical <t̺>) 
+	  \item[→] (fronted <u̟> \textbar{} backed <e̠>)
+	  \item[→] (lowered <e̞> \textbar{} raised <e̝>) 
+	  \item[→] (ATR <e̘ e̙>)
+	  \item[→] (voiced <s̬> \textbar{} voiceless <n̥> \textbar{} creaky voice <b̰> \textbar{} breathy voice <b̤>)
+	  \item[→] (syllabic <n̩> \textbar{} non-syllabic <e̯>)
+	\end{itemize}
  \end{itemize}
@@
 \begin{itemize}
-	\itemsep1pt\parskip0pt\parsep0pt 
-	\item \textsc{Combining Diacritical Marks (above) ordering:} (centralized <ë> \textbar{} mid-centralized <e̽>) → (extra short <ĕ>) →( tone accents, e.g.
-<è> ) → ( Spacing Modifier Letters ) → ( tone letters, e.g.
-<e˦>)
+	\item[] \textsc{Combining Diacritical Marks (above) ordering:}
+	\begin{itemize}
+	  \item[→] (centralized <ë> \textbar{} mid-centralized <e̽>)
+	  \item[→] (extra short <ĕ>)
+	  \item[→] (tone accents, e.g. <è>)
+	  \item[→] (Spacing Modifier Letters, see above)
+	  \item[→] (tone letters, e.g. <˦>)
 	\end{itemize}	 
+\end{itemize}

### 2015-06-06 chapters/pitfalls.tex
minor updates to chapter 2

@@
 basically three different approaches to font rendering. The most widespread is
-Adobe's and Microsoft's \emph{OpenType }system. This approach makes it
+Adobe's and Microsoft's \textsc{OpenType} system. This approach makes it
 relatively easy for font developers, as the font itself does not include all
@@
 project HarfBuzz \url{http://www.freedesktop.org/wiki/Software/HarfBuzz/}.}
-Alternative systems are Apple's \emph{Apple Advanced Typography} (AAT) and the
-open-source \emph{Graphite} system from the Summer Institute of Linguistics
+Alternative systems are Apple's \textsc{Apple Advanced Typography} (AAT) and the
+open-source \textsc{Graphite} system from the Summer Institute of Linguistics
 (SIL).\footnote{More information about AAT can be found at
@@
 \begin{itemize}
-	\item <{\fontspec{EversonMono}A}> \textsc{latin capital letter a}, at \uni{0041} 
-	\item <{\fontspec{EversonMono}А}> \textsc{cyrillic capital letter a}, at \uni{0410} 
-	\item <{\fontspec{EversonMono}Α}> \textsc{greek capital letter alpha}, at \uni{0391} 
-	\item <{\fontspec{EversonMono}Ꭺ}> \textsc{cherokee letter go}, at \uni{13AA} 
-	\item <{\fontspec{EversonMono}ᗅ}> \textsc{canadian syllabics carrier gho}, at \uni{15C5} 
-	\item <{\fontspec{EversonMono}ᴀ}> \textsc{latin small letter capital a}, at \uni{1D00} 
-	\item <{\fontspec{EversonMono}ꓮ}> \textsc{lisu letter a}, at \uni{A4EE} 
-%	\item <{\fontspec{EversonMono}Ａ}> \textsc{fullwidth latin capital letter a}, at \uni{FF21} 
-	\item <{\fontspec{EversonMono}𐊠}> \textsc{carian letter a}, at \uni{102A0} 
-	\item <{\fontspec{EversonMono}𐌀}> \textsc{old italic letter a}, at \uni{10300} 
-	\item <{\fontspec{EversonMono}𝖠}> \textsc{mathematical sans-serif capital a}, \uni{1D5A0} 
-	\item <{\fontspec{EversonMono}𝙰}> \textsc{mathematical monospace capital a}, at \uni{1D670} 
+	\item[] <{\fontspec{EversonMono}A}> \textsc{latin capital letter a}, at \uni{0041} 
+	\item[] <{\fontspec{EversonMono}А}> \textsc{cyrillic capital letter a}, at \uni{0410} 
+	\item[] <{\fontspec{EversonMono}Α}> \textsc{greek capital letter alpha}, at \uni{0391} 
+	\item[] <{\fontspec{EversonMono}Ꭺ}> \textsc{cherokee letter go}, at \uni{13AA} 
+	\item[] <{\fontspec{EversonMono}ᗅ}> \textsc{canadian syllabics carrier gho}, at \uni{15C5} 
+	\item[] <{\fontspec{EversonMono}ᴀ}> \textsc{latin small letter capital a}, at \uni{1D00} 
+	\item[] <{\fontspec{EversonMono}ꓮ}> \textsc{lisu letter a}, at \uni{A4EE} 
+%	\item[] <{\fontspec{EversonMono}Ａ}> \textsc{fullwidth latin capital letter a}, at \uni{FF21} 
+	\item[] <{\fontspec{EversonMono}𐊠}> \textsc{carian letter a}, at \uni{102A0} 
+	\item[] <{\fontspec{EversonMono}𐌀}> \textsc{old italic letter a}, at \uni{10300} 
+	\item[] <{\fontspec{EversonMono}𝖠}> \textsc{mathematical sans-serif capital a}, \uni{1D5A0} 
+	\item[] <{\fontspec{EversonMono}𝙰}> \textsc{mathematical monospace capital a}, at \uni{1D670} 
 \end{itemize}
@@
 	\def\labelenumi{\arabic{enumi}.} 
-	\item <Å> \textsc{latin capital letter a with ring above} \uni{00C5} 
-	\item <Å> \textsc{angstrom sign} \uni{212B}
-	\item <Å> \textsc{latin capital letter a} \uni{0041}
+	\item[] <Å> \textsc{latin capital letter a with ring above} \uni{00C5} 
+	\item[] <Å> \textsc{angstrom sign} \uni{212B}
+	\item[] <Å> \textsc{latin capital letter a} \uni{0041}
 	+ \textsc{combining ring above} \uni{030A}
@@
 precomposed with such an overlay, for example 
-<{\small \fontspec{CharisSIL}ɫ ᵬ ᵭ ᵱ}> or 
-<{\small \fontspec{CharisSIL}ƚ ɨ ɉ ɍ}>, or also the 
+<\charis{ɫ ᵬ ᵭ ᵱ}> or 
+<\charis{ƚ ɨ ɉ ɍ}>, or also the 
 example of <ø> given at the start of this section. Second, the same situation also 
@@
 \textsc{combining palatalized hook below} at \uni{0321}. This element seems to
-occur in precomposed characters like <{\small \fontspec{CharisSIL}ᶀ ᶁ ᶂ ᶄ}>. 
+occur in precomposed characters like <\charis{ᶀ ᶁ ᶂ ᶄ}>. 
 However, all these characters are not canonical equivalent with a combination of 
@@
 interaction between characters and glyphs. A single Unicode character may of
-course be rendered as a single glyph. However, a character may also be a
-\textit{piece} of a glyph, or vice-versa. Actually, all possible relations
-between glyphs and characters are attested.
+course be rendered as a single glyph. However, a character may also be a piece
+of a glyph, or vice-versa. Actually, all possible relations between glyphs and
+characters are attested.
 
-First, \textit{a single character may have different contextually determined
-glyphs.} For example, characters in writing systems like Hebrew and Arabic have
-different glyphs depending on where they appear in a word. Some letters in
-Hebrew change their form at the end of the word, and in Arabic, primary letters
-have four contextually-sensitive variants (isolated, word initial, medial and
-final). Second, \textit{a single character may be rendered as a sequence of
-multiple glyphs.} For example, in Tamil one Unicode character may result in a
-combination of a consonant and vowel, which are rendered as two adjacent glyphs
-by fonts that supports Tamil. Third, \textit{a single glyph may be a combination
-of multiple characters.} For example, the ligature <ï¬>, a single glyph, is the
-result of two characters, <f> and <i>, that have undergone glyph substitution by
-font rendering. Like contextually-determined glyphs, ligatures are (intended)
-artifacts of text processing instructions. Finally, \textit{a single glyph may
-be a part of a character} (e.g.~diacritics).
+First, a single character may have different contextually determined glyphs. For
+example, characters in writing systems like Hebrew and Arabic have different
+glyphs depending on where they appear in a word. Some letters in Hebrew change
+their form at the end of the word, and in Arabic, primary letters have four
+contextually-sensitive variants (isolated, word initial, medial and final).
+Second, a single character may be rendered as a sequence of multiple glyphs. For
+example, in Tamil one Unicode character may result in a combination of a
+consonant and vowel, which are rendered as two adjacent glyphs by fonts that
+supports Tamil. Third, a single glyph may be a combination of multiple
+characters. For example, the ligature <ﬁ>, a single glyph, is the result of two
+characters, <f> and <i>, that have undergone glyph substitution by font
+rendering (see also Section~\ref{pitfall-faulty-rendering}). Like
+contextually-determined glyphs, ligatures are (intended) artifacts of text
+processing instructions. Finally, a single glyph may be a part of a
+character, as exemplified by diacritics.
 
@@
 used. For example, the Unicode character \textsc{latin small letter g} appears
-as and in the Computer Modern and Courier fonts because their typefaces are
-designed differently. Furthermore, font face may change the visual appearance of
-a character, for example Times New Roman two-story <a> changes to a single-story
-a in italics <\emph{a}>. This becomes a real problem for some phonetic
-typesetting (see Section~\ref{pitfall-missing-glyphs}).
+as <g> and <{\fontspec{Courier}g}> in the Linux Libertine and Courier fonts,
+respectively, because their typefaces are designed differently. Furthermore,
+font face may change the visual appearance of a character, for example Times New
+Roman two-story <a> changes to a single-story a in italics <\emph{a}>. This
+becomes a real problem for some phonetic typesetting (see
+Section~\ref{pitfall-ipa-homoglyphs}).
 
@@
 Standard, but for the lay user they can be utterly confusing because visual
-rendering does not indicate logical encoding, or, in other words, what you see
-isn't (necessarily) what you get.
+rendering does not (necessarily) indicate logical encoding.
 

### 2015-06-06 chapters/pitfalls.tex
correction some latex errors

@@
 questions in the user interface involving abbreviations like LF, CR, BE, LE or
-BOM. Most prominently this occurs with exporting or importing data in several
+BOM.\@ Most prominently this occurs with exporting or importing data in several
 software applications from Microsoft. Basically, there are two different issues
@@
 conceptually together they form one action. In the history of computing, various
-encodings of line breaks have been used (e.g.~CR+LF, LF+CR, only LF, or only CR.
-Currently, all Unix and Unix-derived systems use only LF as code for a
-line break, while software from Microsoft still uses a combination of
-CR and LF. Today, most software applications recognize both
-options, and are able to deal with either encoding of line breaks (until rather
-recently this was not the case, and using the wrong line breaks could lead to
-unexpected errors). Our impression is that there is a strong tendency in
-software development to standardize on the simpler \textsc{only LF}
-encoding for line breaks, and we suggest that everybody use this encoding
-whenever possible.
+encodings of line breaks have been used (e.g.~CR+LF, LF+CR, only LF, or only
+CR). Currently, all Unix and Unix-derived systems use only LF as code for a
+line break, while software from Microsoft still uses a combination of CR and
+LF.\@ Today, most software applications recognize both options, and are able to
+deal with either encoding of line breaks (until rather recently this was not the
+case, and using the wrong line breaks could lead to unexpected errors). Our
+impression is that there is a strong tendency in software development to
+standardize on the simpler \textsc{only LF} encoding for line breaks, and we
+suggest that everybody use this encoding whenever possible.
 
@@
 Format.} The details of this issue is extensively explained in section 2.5 of
-the Unicode Core Specification \cite{Unicode2014}.
+the Unicode Core Specification~\cite{Unicode2014}.
 

### 2015-06-07 chapters/pitfalls.tex
Rearrangement Chapter 2

@@
 % ==========================
-\section{Pitfall: Blocks}
-\label{pitfall-blocks}
+\section{Pitfall: Characters are not glyphs}
+\label{pitfall-characters-are-not-glyphs}
 % ==========================
 
-The Unicode code space is subdivided into blocks of contiguous code points. For
-example, the block called \textsc{Cyrillic} runs from \uni{0400} till
-\uni{04FF}. These blocks arose as an attempt at ordering the enormous amount of
-characters in Unicode, but the ideas of blocks very quickly ran into problems.
-First, the size of a block is fixed, so when a block is full, a new block will
-have to be instantiated somewhere further in the code space. For example, this
-led to the blocks \textsc{Cyrillic Supplement}, \textsc{Cyrillic Extended-A}
-(both of which are also already full) and \textsc{Cyrillic Extended-B}. Second,
-when a specific character already exists, then it is not duplicated in another
-block, although the name of the block might indicate that a specific symbol
-should be available there. In general, names of blocks are just an approximate
-indication of the kind of characters that will be in the block.
+A central principle of Unicode is the distinction between character and glyph. A
+character is the abstract notion of a symbol in a writing system, while a glyph
+is the concrete drawing of such a symbol. In practice, there is a complex
+interaction between characters and glyphs. A single Unicode character may of
+course be rendered as a single glyph. However, a character may also be a piece
+of a glyph, or vice-versa. Actually, all possible relations between glyphs and
+characters are attested.
 
-The problem with blocks arises because finding the right character among the
-thousands of Unicode characters is not easy. Many software applications present
-blocks as a primary search mechanism, because the block names suggest where to
-look for a particular character. However, when a user searches for an IPA
-character in the block \textsc{IPA Extensions}, then many IPA characters will not
-be found there. For example, the velar nasal <ŋ> is not part of the block
-\textsc{IPA Extensions} because it was already included as \textsc{latin small letter
-eng} at \uni{014B} in the block \textsc{Latin Extensions-A}.
+First, a single character may have different contextually determined glyphs. For
+example, characters in writing systems like Hebrew and Arabic have different
+glyphs depending on where they appear in a word. Some letters in Hebrew change
+their form at the end of the word, and in Arabic, primary letters have four
+contextually-sensitive variants (isolated, word initial, medial and final).
+Second, a single character may be rendered as a sequence of multiple glyphs. For
+example, in Tamil one Unicode character may result in a combination of a
+consonant and vowel, which are rendered as two adjacent glyphs by fonts that
+supports Tamil. Third, a single glyph may be a combination of multiple
+characters. For example, the ligature <ﬁ>, a single glyph, is the result of two
+characters, <f> and <i>, that have undergone glyph substitution by font
+rendering (see also Section~\ref{pitfall-faulty-rendering}). Like
+contextually-determined glyphs, ligatures are (intended) artifacts of text
+processing instructions. Finally, a single glyph may be a part of a
+character, as exemplified by diacritics.
 
-In general, finding a specific character in the Unicode Standard is often not
-trivial. The names of the blocks can help, but it is not (and never was supposed
-to be) a foolproof structure. It is not the goal nor aim of the Unicode
-Consortium to provide a user interface to the Unicode Standard. If one often
-encounters the problem of needing to find a suitable character, there are
-various other useful services for end-users available.\footnote{The Unicode
-website offers a basic interface to the code charts at
-\url{http://www.unicode.org/charts/index.html}. As a more flexible interface, we
-particularly like PopChar from Macility, available for both Macintosh and
-Windows. However, there are also various free websites that offer search interfaces
-to the Unicode code tables, like \url{http://unicode-search.net} or
-\url{http://unicode-search.net}. A further useful approach for searching characters
-using shape matching is \url{http://shapecatcher.com}.}
+Further, the rendering of a glyph is also always dependent on the font being
+used. For example, the Unicode character \textsc{latin small letter g} appears
+as <g> and <{\fontspec{Courier}g}> in the Linux Libertine and Courier fonts,
+respectively, because their typefaces are designed differently. Furthermore,
+font face may change the visual appearance of a character, for example Times New
+Roman two-story <a> changes to a single-story a in italics <\emph{a}>. This
+becomes a real problem for some phonetic typesetting (see
+Section~\ref{pitfall-ipa-homoglyphs}).
+
+In sum, character-to-glyph mappings are complex technological issues that the
+Unicode Consortium has had to address in the development of the Unicode
+Standard, but for the lay user they can be utterly confusing because visual
+rendering does not (necessarily) indicate logical encoding.
 
@@
 the font.
+
+% ==========================
+\section{Pitfall: Blocks}
+\label{pitfall-blocks}
+% ==========================
+
+The Unicode code space is subdivided into blocks of contiguous code points. For
+example, the block called \textsc{Cyrillic} runs from \uni{0400} till
+\uni{04FF}. These blocks arose as an attempt at ordering the enormous amount of
+characters in Unicode, but the ideas of blocks very quickly ran into problems.
+First, the size of a block is fixed, so when a block is full, a new block will
+have to be instantiated somewhere further in the code space. For example, this
+led to the blocks \textsc{Cyrillic Supplement}, \textsc{Cyrillic Extended-A}
+(both of which are also already full) and \textsc{Cyrillic Extended-B}. Second,
+when a specific character already exists, then it is not duplicated in another
+block, although the name of the block might indicate that a specific symbol
+should be available there. In general, names of blocks are just an approximate
+indication of the kind of characters that will be in the block.
+
+The problem with blocks arises because finding the right character among the
+thousands of Unicode characters is not easy. Many software applications present
+blocks as a primary search mechanism, because the block names suggest where to
+look for a particular character. However, when a user searches for an IPA
+character in the block \textsc{IPA Extensions}, then many IPA characters will not
+be found there. For example, the velar nasal <ŋ> is not part of the block
+\textsc{IPA Extensions} because it was already included as \textsc{latin small letter
+eng} at \uni{014B} in the block \textsc{Latin Extensions-A}.
+
+In general, finding a specific character in the Unicode Standard is often not
+trivial. The names of the blocks can help, but it is not (and never was supposed
+to be) a foolproof structure. It is not the goal nor aim of the Unicode
+Consortium to provide a user interface to the Unicode Standard. If one often
+encounters the problem of needing to find a suitable character, there are
+various other useful services for end-users available.\footnote{The Unicode
+website offers a basic interface to the code charts at
+\url{http://www.unicode.org/charts/index.html}. As a more flexible interface, we
+particularly like PopChar from Macility, available for both Macintosh and
+Windows. However, there are also various free websites that offer search interfaces
+to the Unicode code tables, like \url{http://unicode-search.net} or
+\url{http://unicode-search.net}. A further useful approach for searching characters
+using shape matching is \url{http://shapecatcher.com}.}
 
@@
 % ==========================
-\section{Pitfall: Characters are not glyphs}
-\label{pitfall-characters-are-not-glyphs}
-% ==========================
-
-A central principle of Unicode is the distinction between character and glyph. A
-character is the abstract notion of a symbol in a writing system, while a glyph
-is the concrete drawing of such a symbol. In practice, there is a complex
-interaction between characters and glyphs. A single Unicode character may of
-course be rendered as a single glyph. However, a character may also be a piece
-of a glyph, or vice-versa. Actually, all possible relations between glyphs and
-characters are attested.
-
-First, a single character may have different contextually determined glyphs. For
-example, characters in writing systems like Hebrew and Arabic have different
-glyphs depending on where they appear in a word. Some letters in Hebrew change
-their form at the end of the word, and in Arabic, primary letters have four
-contextually-sensitive variants (isolated, word initial, medial and final).
-Second, a single character may be rendered as a sequence of multiple glyphs. For
-example, in Tamil one Unicode character may result in a combination of a
-consonant and vowel, which are rendered as two adjacent glyphs by fonts that
-supports Tamil. Third, a single glyph may be a combination of multiple
-characters. For example, the ligature <ﬁ>, a single glyph, is the result of two
-characters, <f> and <i>, that have undergone glyph substitution by font
-rendering (see also Section~\ref{pitfall-faulty-rendering}). Like
-contextually-determined glyphs, ligatures are (intended) artifacts of text
-processing instructions. Finally, a single glyph may be a part of a
-character, as exemplified by diacritics.
-
-Further, the rendering of a glyph is also always dependent on the font being
-used. For example, the Unicode character \textsc{latin small letter g} appears
-as <g> and <{\fontspec{Courier}g}> in the Linux Libertine and Courier fonts,
-respectively, because their typefaces are designed differently. Furthermore,
-font face may change the visual appearance of a character, for example Times New
-Roman two-story <a> changes to a single-story a in italics <\emph{a}>. This
-becomes a real problem for some phonetic typesetting (see
-Section~\ref{pitfall-ipa-homoglyphs}).
-
-In sum, character-to-glyph mappings are complex technological issues that the
-Unicode Consortium has had to address in the development of the Unicode
-Standard, but for the lay user they can be utterly confusing because visual
-rendering does not (necessarily) indicate logical encoding.
-
-% ==========================
 \section{Pitfall: File formats}
@@
 
-\
+% ==========================
+\section{Recommendations}
+\label{recommendations}
+% ==========================
 
-\noindent Together with the topic of normalization (NFC vs.~NFD, see 
+Together with the topic of normalization (NFC vs.~NFD, see 
 Section~\ref{pitfall-canonical-equivalence} above),

### 2015-06-07 chapters/pitfalls.tex
corrections to pitfall faulty rendering

@@
 rendering problems, like proportional glyph size or ligatures are reasonably
-well understood. However, even within the Latin script, the positioning of
-diacritics relative to a base character is still a widespread problem.
+well understood. In contrast, the positioning of multiple diacritics relative to
+a base character is still a widespread problem, even within the Latin script.
 Especially when more than one diacritic is supposed to be placed above (or
@@
 software applications. The problems arising in Arabic and in many southeast
-Asian scripts (like Devanagari or Burmese) are even much more complex. There are
+Asian scripts (like Devanagari or Burmese) are even more complex.
+
+To understand where any problems arise it is important to realize that there are
 basically three different approaches to font rendering. The most widespread is
@@
 project HarfBuzz \url{http://www.freedesktop.org/wiki/Software/HarfBuzz/}.}
-Alternative systems are Apple's \textsc{Apple Advanced Typography} (AAT) and the
+Alternative systems are \textsc{Apple Advanced Typography} (AAT) and the
 open-source \textsc{Graphite} system from the Summer Institute of Linguistics
 (SIL).\footnote{More information about AAT can be found at
-\url{https://developer.apple.com/fonts/}. SIL's Graphite is described in detail
-at
+\url{https://developer.apple.com/fonts/}. \newline SIL's Graphite is described
+in detail at
 \url{http://scripts.sil.org/cms/scripts/page.php?site\_id=projects\&item\_id=graphite\_home}.}
@@
 the font.
+
+There is mostly no real solution to problems arising from faulty font rendering.
+Switching to another software application that offers better handling is the
+only real alternative, but this is normally not an option for daily work. The 
+experience with rendering on the side of the software industry is developing 
+quickly, so we can expect the situation only to get better. In the meantime one 
+can try to correct faulty layout by tweaking baseline and/or kerning (when such 
+option are available).
 
@@
 In general, finding a specific character in the Unicode Standard is often not
-trivial. The names of the blocks can help, but it is not (and never was supposed
+trivial. The names of the blocks can help, but they are not (and never were supposed
 to be) a foolproof structure. It is not the goal nor aim of the Unicode
@@
 particularly like PopChar from Macility, available for both Macintosh and
-Windows. However, there are also various free websites that offer search interfaces
+Windows. There are also various free websites that offer search interfaces
 to the Unicode code tables, like \url{http://unicode-search.net} or

### 2015-06-07 chapters/pitfalls.tex
minor edits to pitfall homoglyphs

@@
 \textsc{latin small letter g} at \uni{0067}). In contrast, the symbols
-<{\fontspec{EversonMono}AАΑᎪᗅᴀꓮ𐊠𐌀𝖠𝙰}> are all different code points,
+<{\fontspec{EversonMono}AАΑᎪᗅᴀꓮ𐊠𝖠𝙰}> are all different code points,
 although they look highly similar---in some cases even sharing exactly the same
@@
 	\item[] <{\fontspec{EversonMono}𐊠}> \textsc{carian letter a}, at \uni{102A0} 
-	\item[] <{\fontspec{EversonMono}𐌀}> \textsc{old italic letter a}, at \uni{10300} 
+%	\item[] <{\fontspec{EversonMono}𐌀}> \textsc{old italic letter a}, at \uni{10300} 
 	\item[] <{\fontspec{EversonMono}𝖠}> \textsc{mathematical sans-serif capital a}, \uni{1D5A0} 
@@
 Such homoglyphs are a widespread problem for consistent encoding. Although for
-most users it looks like the words <voces> and <νοсеѕ> are identical, in actual
-fact they do not even share a single code point.\footnote{The first words
+most users it looks like the words <voces> and <νοсеѕ> are almost identical, in
+actual fact they do not even share a single code point.\footnote{The first words
 consists completely of Latin characters, namely \unif{0076}, \unif{006F},
-\unif{0063}, \unif{0065} and \unif{0073}, while the second is a mix of Cyrillic and
-Greek characters, namely \unif{03BD}, \unif{03BF}, \unif{0041}, \unif{0435} and
-\unif{0455}.} For computers these two words are completely different entities.
-Commonly, when users with Cyrillic or Greek keyboards have to type some
-Latin-based orthography, they mix similar looking Cyrillic or Greek characters
-into their text, because those characters are so much easier to type. Similarly,
-when users want to enter an unusual symbol, they normally search by visual
-impression in their favorite software application, and just pick something that
-looks reasonably alike to what they expect the glyph to look like.
+\unif{0063}, \unif{0065} and \unif{0073}, while the second is a mix of Cyrillic
+and Greek characters, namely \unif{03BD}, \unif{03BF}, \unif{0041}, \unif{0435}
+and \unif{0455}.} For computers these two words are completely different
+entities. Commonly, when users with Cyrillic or Greek keyboards have to type
+some Latin-based orthography, they mix similar looking Cyrillic or Greek
+characters into their text, because those characters are so much easier to type.
+Similarly, when users want to enter an unusual symbol, they normally search by
+visual impression in their favorite software application, and just pick
+something that looks reasonably alike to what they expect the glyph to look
+like.
+
+It is really easy to make errors at text entry and add characters that are 
+not supposed to be included. Our proposals for orthography profiles (see
+Chapter~\ref{orthography-profiles}) are a method for checking the consistency of 
+any text. In situations in which interoperability is important, we consider it 
+crucial to add such checks in any workflow.
 

### 2015-06-07 chapters/pitfalls.tex
updated pitfall absence equivalence

@@
 software that does not apply Unicode Standard normalization) for
-\textsc{angstrom sign} (\uni{212B}), instances of \textsc{latin capital letter a
-with ring above} (\uni{00C5}) would not be found.
+\textsc{angstrom sign} (\uni{212B}), then the instances of \textsc{latin 
+capital letter a with ring above} (\uni{00C5}) would not be found.
 
@@
 representations (in fact, many do), but the Unicode Normalization Algorithm can
-be used to transform certain sequences of characters into canonical (or
-compatible) forms to test for equivalency. To determine equivalence, each
+be used to transform certain sequences of characters into canonical
+forms to test for equivalency. To determine equivalence, each
 character in the Unicode Standard is associated with a combining class, which is
@@
 the Unicode Canonical Ordering Algorithm to determine which sequences of
-combining marks are canonically equivalent. Normalization forms, as very briefly
+characters are canonically equivalent. Normalization forms, as very briefly
 described above, can be used to ensure character equivalence by ordering
@@
 unexpected behavior for users. Various players, like the Unicode Consortium, the
-W{\large 3}C, or the TEI recommend NFC in most user-directed situations, and some
-software applications that we tested indeed seem to automatically convert
+W{\large 3}C, or the TEI recommend NFC in most user-directed situations, and
+some software applications that we tested indeed seem to automatically convert
 strings into NFC.\footnote{See the summary of various recommendation here:
-\url{http://www.win.tue.nl/~aeb/linux/uc/nfc_vs_nfd.html}.}
-This means in practice that if a user, for example, enters <a> and < ̀>,
-i.e.~\textsc{latin small letter a} at \uni{0061} and \textsc{combining grave
-accent} at \uni{0300}, this might be converted into <à>, i.e.~\textsc{latin
-small letter a with grave} at \uni{00E0}.\footnote{The behavior of software
-applications can be quite erratic in this respect. For example, Apple's TextEdit
-does not do any conversion on text entry. However, when you copy and paste some
-text inside the same document in rich text mode (i.e.~RTF-format), it
-will be transformed into NFC on paste. Saving a document does not do any
-conversion to the glyphs on screen, but it will save the characters in NFC.}
+\url{http://www.win.tue.nl/~aeb/linux/uc/nfc_vs_nfd.html}.} This means in
+practice that if a user, for example, enters <a> and < ̀>, i.e.~\textsc{latin
+small letter a} at \uni{0061} and \textsc{combining grave accent} at \uni{0300},
+this might be automatically converted into <à>, i.e.~\textsc{latin small letter
+a with grave} at \uni{00E0}.\footnote{The behavior of software applications can
+be quite erratic in this respect. For example, Apple's TextEdit does not do any
+conversion on text entry. However, when you copy and paste some text inside the
+same document in rich text mode (i.e.~RTF-format), it will be transformed into
+NFC on paste. Saving a document does not do any conversion to the glyphs on
+screen, but it will save the characters in NFC.}
 
@@
 The general rule followed is that extensions of Latin characters that are
-connected to the character are not separated as combining diacritics. For
-example, characters like <ƞ ɲ ɳ> or <ɰ ɱ> are obviously derived from <n> and <m>
+connected to the base character are not separated as combining diacritics. For
+example, characters like <ŋ ɲ ɳ> or <ɖ ɗ> are obviously derived from <n> and <d>
 respectively, but they are treated like new separate characters in the Unicode
@@
 
-Interestingly, and somewhat illogically, there are three elements attached to
-other characters that are still treated as separable in the Unicode Standard. First,
-the cedilla, like in <ç> \textsc{latin small letter c with cedilla} at
-\uni{00E7}, second the ogonek, like in <ą> \textsc{latin small letter a with
-ogonek} at \uni{0105} and, third, the horn, like in <ơ> \textsc{latin small
-letter o with horn} at \uni{01A1}. These characters are indeed decomposed in a 
-base character with a combining diacritic. However, it is these cases that should be 
-considered the exceptions to the rule.
+Interestingly, and somewhat illogically, there are three elements, which are
+directly attached to their base characters, but which are still treated as
+separable in the Unicode Standard. Such characters are decomposed (in NFD
+normalization) in a base character with a combining diacritic. However, it is
+these cases that should be considered the exceptions to the rule. These three 
+elements are the following:
 
-There are two combinations that deserve specific care because it is actually 
-possible to produce identical characters in different ways, without them being 
-canonical equivalent. First, there are the
-combining characters designated as ``overlay'' in the Unicode Standard, like    
-<\ \ {\large  ̴}> \textsc{combining tilde overlay} at \uni{0334} or 
-<\ \ {\large  ̵}> \textsc{combining short stroke overlay} at \uni{0335}. 
-There are many characters that look like they are 
-precomposed with such an overlay, for example 
-<\charis{ɫ ᵬ ᵭ ᵱ}> or 
-<\charis{ƚ ɨ ɉ ɍ}>, or also the 
-example of <ø> given at the start of this section. Second, the same situation also 
-occurs with combining characters designated as ``hook'', like < {\large  ̡}> 
-\textsc{combining palatalized hook below} at \uni{0321}. This element seems to
-occur in precomposed characters like <\charis{ᶀ ᶁ ᶂ ᶄ}>. 
-However, all these characters are not canonical equivalent with a combination of 
-their base characters with the combining overlay or combining hook.
+\begin{itemize}
+
+  \item the \textsc{combining cedilla} at \uni{0327} \newline This diacritic is
+        for example attested in the precomposed character <ç> \textsc{latin
+        small letter c with cedilla} at \uni{00E7}. This <ç> will thus be
+        decomposed in NFC normalization.
+  \item the \textsc{combining ogonek} at \uni{0328} \newline This diacritic is
+        for example attested in precomposed <ą> \textsc{latin small letter a
+        with ogonek} at \uni{0105}. This <ą> will thus be decomposed in NFC
+        normalization.
+  \item the \textsc{combining horn} at \uni{031B} \newline This diacritic is for
+        example attested in precomposed <ơ> \textsc{latin small letter o with
+        horn} at \uni{01A1}. This <ơ> will thus be decomposed in NFC
+        normalization.
+
+\end{itemize}
+
+There are further combinations that deserve specific care because it is actually
+possible to produce identical characters in different ways without them being
+canonically equivalent. In these situations, the general rule holds, namely that
+characters with attached extras are not decomposed. However, in the following
+cases the ``extras'' actually exist as combining diacritics, so there is also 
+the possibility to construct a character by using a base character with those 
+combining diacritics.
+
+\begin{itemize}
+
+  \item First, there are the combining characters designated as ``combining
+        overlay'' in the Unicode Standard, like <\ \ {\large ̴}>
+        \textsc{combining tilde overlay} at \uni{0334} or <\ \ {\large ̵}>
+        \textsc{combining short stroke overlay} at \uni{0335}. There are many
+        characters that look like they are precomposed with such an overlay,
+        for example <\charis{ɫ~ᵬ~ᵭ~ᵱ}> or <\charis{ƚ~ɨ~ɉ~ɍ}>, or also the
+        example of <ø> given at the start of this section. However, they are 
+        not decomposed in NFD normalization.
+  \item Second, the same situation also occurs with combining characters
+        designated as ``combining hook'', like < {\large ̡}> \textsc{combining
+        palatalized hook below} at \uni{0321}. This element seems to occur in
+        precomposed characters like <\charis{ᶀ~ᶁ~ᶂ~ᶄ}>. However, they are 
+        not decomposed in NFD normalization.
+
+\end{itemize}
+
+To harmonize the encoding in these cases it is not sufficient to use Unicode 
+normalization. Additional checks are necessary, for example by using orthography 
+profiles (see Chapter~\ref{orthography-profiles}).
 

### 2015-06-07 chapters/pitfalls.tex
updated pitfall file format

@@
 
-Further, the rendering of a glyph is also always dependent on the font being
-used. For example, the Unicode character \textsc{latin small letter g} appears
-as <g> and <{\fontspec{Courier}g}> in the Linux Libertine and Courier fonts,
-respectively, because their typefaces are designed differently. Furthermore,
-font face may change the visual appearance of a character, for example Times New
-Roman two-story <a> changes to a single-story a in italics <\emph{a}>. This
-becomes a real problem for some phonetic typesetting (see
-Section~\ref{pitfall-ipa-homoglyphs}).
+Further, the rendering of a glyph is dependent on the font being used. For
+example, the Unicode character \textsc{latin small letter g} appears as <g> and
+<{\fontspec{Courier}g}> in the Linux Libertine and Courier fonts, respectively,
+because their typefaces are designed differently. Furthermore, font face may
+change the visual appearance of a character, for example Times New Roman
+two-story <{\fontspec{Times New Roman}a}> changes to a single-story a in italics
+<\emph{\fontspec{Times New Roman}a}>. This becomes a real problem for some
+phonetic typesetting (see Section~\ref{pitfall-ipa-homoglyphs}).
 
@@
 Unicode is a character encoding standard, but these characters of course
-actually appear inside some kind of file. The most basic Unicode-based file
+actually appear inside some kind of computer file. The most basic Unicode-based file
 format is pure line-based text, i.e.~strings of Unicode-encoded characters
@@
 involved. First, the encoding of line breaks and, second, the encoding of the
-Unicode characters into \textsc{code units} and the related issue of
-\textsc{endianness}.
+Unicode characters into code units and the related issue of endianness.
 
@@
 The issue with \textsc{line breaks} originated with the instructions necessary
-to direct a printing head to a new line. This involves two movements, known as
-\textsc{carriage return} (CR, returning the printing head to the start of the
-line on the page) and \textsc{line feed} (LF, moving the printing head to the
-next line on the page). Physically, these are two different events, but
-conceptually together they form one action. In the history of computing, various
-encodings of line breaks have been used (e.g.~CR+LF, LF+CR, only LF, or only
-CR). Currently, all Unix and Unix-derived systems use only LF as code for a
-line break, while software from Microsoft still uses a combination of CR and
-LF.\@ Today, most software applications recognize both options, and are able to
-deal with either encoding of line breaks (until rather recently this was not the
-case, and using the wrong line breaks could lead to unexpected errors). Our
-impression is that there is a strong tendency in software development to
-standardize on the simpler \textsc{only LF} encoding for line breaks, and we
-suggest that everybody use this encoding whenever possible.
+to direct a printing head of a physical printer to a new line. This involves two
+movements, known as \textsc{carriage return} (CR, returning the printing head to
+the start of the line on the page) and \textsc{line feed} (LF, moving the
+printing head to the next line on the page). Physically, these are two different
+events, but conceptually together they form one action. In the history of
+computing, various encodings of line breaks have been used (e.g.~CR+LF, LF+CR,
+only LF, or only CR). Currently, all Unix and Unix-derived systems use only LF
+as code for a line break, while software from Microsoft still uses a combination
+of CR+LF.\@ Today, most software applications recognize both options, and
+are able to deal with either encoding of line breaks (until rather recently this
+was not the case, and using the wrong line breaks would lead to unexpected
+errors). Our impression is that there is a strong tendency in software
+development to standardize on the simpler ``only LF'' encoding for line
+breaks, and we suggest that everybody use this encoding whenever possible.
 
@@
 offers three different approaches, called UTF-32, UTF-16 and UTF-8.\footnote{The
-letters UTF stand for either \textsc{Unicode Transformation Format} or
-\textsc{Universal Coded Character Set Transformation Format}, but the notion of
-\textit{transformation} is a legacy notion that does not have meaning anymore.
-Nevertheless, the designation \textsc{UTF} (in capitals) has become an official
-standard designation, but should probably best be read as simply Unicode
-Format.} The details of this issue is extensively explained in section 2.5 of
-the Unicode Core Specification~\cite{Unicode2014}.
+letters UTF stand for \textsc{Unicode Transformation Format}, but the notion of
+``transformation'' is a legacy notion that does not have meaning anymore.
+Nevertheless, the designation UTF (in capitals) has become an official
+standard designation, but should probably best be read as simply ``Unicode
+Format.''} The details of this issue is extensively explained in section 2.5 of
+the Unicode Core Specification~\citet{Unicode2014}.
 
@@
 
-Finally, \textsc{UTF-8} is a more complex system that dynamically
-encodes each character with the minimally necessary number of bits, choosing
-either 8, 16 or 32 bits depending on the character. This represents again a
-strong reduction in space (particularly due to the high frequency of data using
-ASCII characters, which need only 8 bits) at the expense of even more computation
+Finally, \textsc{UTF-8} is a more complex system that dynamically encodes each
+character with the minimally necessary number of bits, choosing either 8, 16 or
+32 bits depending on the character. This represents again a strong reduction in
+space (particularly due to the high frequency of data using erstwhile ASCII
+characters, which need only 8 bits) at the expense of even more computation
 necessary to process such strings. However, because of the ever growing
@@
 (LE) storage. The Unicode Standard offers a possibility to explicitly indicate
-what kind of storage is used by starting a file with a so-called \textsc{Byte Order
-Mark} (BOM). However, the Unicode Standard does not require the usage of BOM,
+what kind of storage is used by starting a file with a so-called \textsc{byte order
+mark} (BOM). However, the Unicode Standard does not require the usage of BOM,
 preferring other non-Unicode methods to signal to computers which kind of

### 2015-06-07 chapters/pitfalls.tex
adding section summary to chapter 2

@@
 
-Together with the topic of normalization (NFC vs.~NFD, see 
-Section~\ref{pitfall-canonical-equivalence} above),
-this leads to the following nicely cryptic file-encoding recommendation, namely
-to preferably use UTF-8 in NFC without BOM with LF line breaks,
-whenever possible.
+Summerizing the pitfalls discussed in this chapter, we propose the following 
+recommendations:
+
+\begin{itemize}
+   \item To prevent strange boxes instead of nice glyphs, always install a few
+         fonts with a large glyph collection and at least one fall-back font
+         (see Section~\ref{pitfall-missing-glyphs}).
+   \item Unexpected visual impressions of symbols does not necessarily mean that
+         the actual encoding is wrong. It is mostly a problem of faulty
+         rendering (see Section~\ref{pitfall-faulty-rendering}).
+   \item Do not trust the names of Unicode blocks as a strategy to find specific 
+         characters (see Section~\ref{pitfall-blocks})
+   \item To ensure consistent encoding of texts, apply Unicode normalization
+         (NFC or NFD, see Section~\ref{pitfall-canonical-equivalence}).
+   \item To prevent remaining inconsistencies after normalization, for example 
+         stemming from homoglyphs (see Section~\ref{pitfall-homoglyphs}) 
+         or from missing canonical equivalence in the Unicode Standard
+         (see Section~\ref{pitfall-absence-of-equivalence})
+         use orthography profiles (see Chapter~\ref{orthography-profiles}).
+   \item As a preferred file format, use UTF-8 in NFC without BOM with 
+         LF line breaks, whenever possible (see 
+         Section~\ref{pitfall-file-formats}). This nicely cryptic recommendation 
+         has the T-shirt potential:
+
+\end{itemize}
+
+\begin{center}
+  I prefer it
+
+  \textbf{UTF-8 NFC LF no BOM}
+\end{center}

### 2015-06-07 chapters/pitfalls.tex
typos

@@
 
-Summerizing the pitfalls discussed in this chapter, we propose the following 
+Summarizing the pitfalls discussed in this chapter, we propose the following 
 recommendations:
@@
          use orthography profiles (see Chapter~\ref{orthography-profiles}).
-   \item As a preferred file format, use UTF-8 in NFC without BOM with 
-         LF line breaks, whenever possible (see 
-         Section~\ref{pitfall-file-formats}). This nicely cryptic recommendation 
-         has the T-shirt potential:
+   \item As a preferred file format, use Unicode Format UTF-8 in 
+         Normalization Form Composition (NFC) with LF line endings, 
+         but without byte order mark (BOM), whenever possible (see 
+         Section~\ref{pitfall-file-formats}). This last nicely cryptic 
+         recommendation has T-shirt potential:
 

### 2015-06-08 chapters/introduction.tex
minor corrections in section encoding

@@
 code is not so often told, so we decided to offer a short survey of the major
-developments of such \textsc{encoding} here.\footnote{Because of the recent
+developments of such encoding here.\footnote{Because of the recent
 history as summarized in this section, we have used mostly rather ephemeral
@@
 messages over large distances. Originally, various specially designed
-contraptions were used to send messages. Today, descendants of these systems
-are still in use utilizing flags or flashing lights. The ``innovation'' of those
-semaphoric systems was that all characters of the written language were replaced
-one-to-one by visual signals. Since then, all telegraphic systems have taken
-this principle, namely that any language to be transmitted first has to be
-turned into some orthographic system, which subsequently is encoded for
-transmission by the sender, and then turned back into orthographic
-representation at the receiver side.\footnote{Sound and video-based
-telecommunication of course takes a different approach by ignoring the written
-version of language and directly encode sound waves or light patterns.} This of
-course implies that the usefulness of any such telegraphic encoding completely
-depends on the sometimes rather haphazard structure of orthographic systems.
+contraptions were used to send messages. Today, descendants of these systems are
+still in limited use, for example utilizing flags or flashing lights. The
+``innovation'' of those semaphoric systems was that all characters of the
+written language were replaced one-to-one by visual signals. Since then, all
+telegraphic systems have taken this principle, namely that any language to be
+transmitted first has to be turned into some orthographic system, which
+subsequently is encoded for transmission by the sender, and then turned back
+into orthographic representation at the receiver side.\footnote{Sound and
+video-based telecommunication of course takes a different approach by ignoring
+the written version of language and directly encode sound waves or light
+patterns.} This of course implies that the usefulness of any such telegraphic
+encoding completely depends on the sometimes rather haphazard structure of
+orthographic systems.
 
@@
 by counting until a fixed number of signals has passed.\footnote{Of course, no
-explicit separator is needed when the timing of the signals is known, which is
+explicit separator is needed at all when the timing of the signals is known, which is
 the principle used in all modern telecommunication systems. An important modern
@@
 Greek.\footnote{In effect, because $16 = 2^4$, this means that ISO/IEC~8859 was
-actually an $8+4=12$-bit encoding, though with very many duplicates by design, namely 
-all ASCII codes were repeated in each 8-bit code page. To be precise,
+actually an $8+4=12$-bit encoding, though with very many duplicates by design,
+namely all ASCII codes were repeated in each 8-bit code page. To be precise,
 ISO/IEC~8859 used the 7-bit ASCII as the basis for each code page, and defined
 16 different 7-bit extensions, leading to $(1+16)\cdot{2^7} = 2,176$ possible
-characters.} This system almost immediately was understood to be insufficient and
-impractical, so various initiatives to extend and reorganize the encoding
-started in the 1980s. This led, for example, to various proprietary encodings
-from Microsoft (e.g.~Windows Latin 1) and Apple (e.g.~Mac OS Roman), which one
-still sometimes encounters today.
+characters. However, because of overlap and not-assigned codes points the actual
+number of symbols was much smaller.} This system almost immediately was
+understood to be insufficient and impractical, so various initiatives to extend
+and reorganize the encoding started in the 1980s. This led, for example, to
+various proprietary encodings from Microsoft (e.g.~Windows Latin 1) and Apple
+(e.g.~Mac OS Roman), which one still sometimes encounters today.
 

### 2015-06-08 chapters/introduction.tex
minor corrections to terminology/unicode approach

@@
 writing systems used for a particular language, namely transcriptions and
-orthographies. First, \textsc{transcription} is a scientific procedure (and also
+orthographies.
+
+First, \textsc{transcription} is a scientific procedure (and also
 the result of that procedure) for graphically representing the sounds of human
@@
 systems. Note that IPA symbols are also often used to represent language on a
-phonemic level. However, it is important to realize that in this usage the IPA
+phonemic level. It is important to realize that in this usage the IPA
 symbols are not a transcription system, but rather an orthography (though with
@@
 computers were constrained by the American Standard Code for Information
-Interchange (ASCII) character encoding scheme (based on the Latin script), which
+Interchange (ASCII) character encoding scheme, based on the Latin script, which
 only allowed for a strongly limited number of different symbols to be encoded.
@@
 The Unicode Standard represents a massive step forward because it aims to
-eradicate the distinction between \textsc{universal} (ASCII) versus
-\textsc{language/resource-particular} (Font) by adding as much as possible
-language-specific information into the universal standard. However, there are
-still language/resource-specific specifications necessary for the proper usage
-of Unicode, as will be discussed below. Within the Unicode structure many of
-these specifications can be captured by so-called \textsc{Locale Descriptions}
-or simply locales, so we are moving to a new distinction of universal (Unicode
-Standard) versus language-particular (Locale Description). The major gain is a
-much larger compatibility on the universal level (because Unicode standardizes a
-much larger portion of writing system diversity), and much better possibilities
-for automatic processing on the language-particular level (because Locale
-Descriptions are computer readable specifications).
+eradicate the distinction between universal (ASCII) versus language-particular
+(font) by adding as much as possible language-specific information into the
+universal standard. However, there are still language/resource-specific
+specifications necessary for the proper usage of Unicode, as will be discussed
+below. Within the Unicode structure many of these specifications can be captured
+by so-called \textsc{locale descriptions}, so we are moving to a new distinction
+of universal (Unicode Standard) versus language-particular (locale description).
+The major gain is a much larger compatibility on the universal level (because
+Unicode standardizes a much larger portion of writing system diversity), and
+much better possibilities for automatic processing on the language-particular
+level (because locale descriptions are computer readable specifications).
 
@@
 Unicode Standard is the \textsc{International Components for Unicode (ICU)},
-which offers C/C++ and Java libraries trying to implement the Unicode
-Standard.\footnote{\url{http://icu-project.org}}
+which offers C/C++ and Java libraries implementing the Unicode
+Standard.\footnote{More information about the ICU is available here:
+\url{http://icu-project.org}.}
 
@@
 
-One frequently references property is the \textsc{block} property, which is
+One frequently referenced property is the \textsc{block} property, which is
 often used in software applications to impose some structure to the large number
@@
 and \textsc{Arabic Presentation Form-B}. The situation with Latin symbols is
-even more extra. In general, the names for block should be taken as a
+even more extreme. In general, the names for block should not be taken as a
 definitional statement. For example, many IPA symbols are not located in the
-aptly-names block \textsc{IPA extensions}, but in other blocks (see Section~\ref{ipa-meets-unicode}).
+aptly-names block \textsc{IPA extensions}, but in other blocks
+(see Section~\ref{pitfall-no-complete-ipa-block}).
 
@@
 Unicode Standard offers different kind of normalizations to either decompose all
-precomposed characters (called \textsc{NFD}, \textsc{Normalization Form
-canonical Decomposition}), or compose as much as possible combinations (called
-\textsc{NFC}, \textsc{Normalization Form canonical Composition}). In current
+precomposed characters (called \textsc{NFD}, \textsc{normalization form
+canonical decomposition}), or compose as much as possible combinations (called
+\textsc{NFC}, \textsc{normalization form canonical composition}). In current
 practice of software development, NFC seems to be preferred in most situations
@@
 clusters are locale-dependent graphemes, i.e.~such combination of characters do
-not function as text elements in all situations. For example, the sequence <c>~+~<h> for the Slovak digraph <ch> or the sequence <ky> in the Sisaala practical
-orthography (pronounced as IPA /tʃ/,~\cite{Moran2006}). These grapheme clusters
-are \textsc{tailored} in the sense that they must be specified on a
-language-by-language or a writing system-by-writing system basis. The Unicode
+not function as text elements in all situations. For example, the sequence
+<c>~+~<h> for the Slovak digraph <ch> or the sequence <ky> in the Sisaala
+practical orthography (pronounced as IPA /tʃ/,~\cite{Moran2006}). These grapheme
+clusters are \textsc{tailored} in the sense that they must be specified on a
+language-by-language or writing-system-by-writing-system basis. The Unicode
 Standard provides technological specifications for creating locale specific data
@@
 descriptions are saved in the \textsc{Common Locale Data Repository
-(CLDR)},\footnote{\url{http://cldr.unicode.org/}} a repository of
+(CLDR)},\footnote{More information about the CLDR can be found here:
+\url{http://cldr.unicode.org/}.} a repository of
 language-specific definitions of writing system properties, each of which
@@
 Japanese writing system.''} However, there are various drawbacks of locale
-descriptions for the daily practice of linguistic work in a multilingual setting
-(see Section~\ref{use-cases}).
+descriptions for the daily practice of linguistic work in a multilingual setting.

### 2015-06-08 chapters/ipa_meets_unicode.tex
Minor edits intro chapter 3

@@
 
+% ==========================
 \section{The International Phonetic Alphabet (IPA)}
 \label{the-international-phonetic-alphabet}
+% ==========================
 
 The International Phonetic Alphabet (IPA) is a common standard in linguistics to
-transcribe sounds of spoken language into some Latin-based characters
+transcribe sounds of spoken language into Latin-based characters
 \citep{IPA2005}. Although IPA is reasonably easily adhered to with pen and
-paper, it is not trivial to encode IPA characters electronically. Early work
-addressing the need for a universal computing environment for writing systems
-and their computational complexity is discussed in \citet{Simons1989}. For a long
-time, linguists (like all other computer users) were limited to ASCII-encoded
-7-bit characters, which only includes Latin characters, numbers and some
-punctuation and symbols. Restricted to these standard character sets that lacked
-IPA support or other language-specific graphemes that they needed, linguists
-devised their own solutions \citep{BirdSimons2003}. For example, some chose to
-represent unavailable graphemes with substitutes, e.g.~the combination of <ng> to
-represent <ŋ>. Tech-savvy linguists redefined selected characters from a character
-encoding by mapping custom made fonts to those code points. However, one
-linguist's electronic text would not render properly on another linguist's
+paper, it is not trivial to encode IPA characters electronically. Similar to the
+previous chapter, this chapter will discuss various pitfalls with the encoding
+of the IPA.\@ The details of the encoding are unimportant as long as the
+transcription is only directed towards phonetically trained eyes. For a linguist
+reading an IPA transcription, many of the details that will be discussed in this
+chapter might seem like hair-splitting trivialities. However, if IPA
+transcriptions are intended to be used across resources (e.g.~searching similar
+phenomena across different languages) then it becomes crucial that there are strict
+encoding guidelines. Our main goal in this chapter is to present the encoding
+issues and propose recommendations for a ``strict'' IPA encoding.
+
+\
+
+\noindent For a long time, linguists, like all other computer users, were limited to
+ASCII-encoded 7-bit characters, which only includes Latin characters, numbers
+and some punctuation and symbols. Restricted to these standard character sets
+that lacked IPA support or other language-specific graphemes that they needed,
+linguists devised their own solutions.\footnote{Early work addressing the need
+for a universal computing environment for writing systems and their
+computational complexity is discussed in \citet{Simons1989}. A survey of
+practical recommendations for language resources, including notes on encoding,
+can be found in \citet{BirdSimons2003}} For example, some chose to represent
+unavailable graphemes with substitutes, e.g.~the combination of <ng> to
+represent <ŋ>. Tech-savvy linguists redefined selected characters from a
+character encoding by mapping custom made fonts to those code points. However,
+one linguist's electronic text would not render properly on another linguist's
 computer without access to the same font. Further, if two character encodings
@@
 To alleviate this problem, during the late 1980s, SAMPA (Speech Assessment
-Methods Phonetic Alphabet) was created to represent IPA symbols with 7-bit
-printable ASCII character sequences, e.g. <p\textbackslash> for [ɸ]. Two
-problems with SAMPA are that (i) it is only a partial encoding of the IPA and
-(ii) it encodes different languages in separate data tables, instead of a
-universal alphabet, like IPA.\@ SAMPA tables are derived from phonemes appearing
-in several European languages that were developed as part of a European
-Commission-funded project to address technical problems like electronic mail
-exchange (what is now simply called email). SAMPA is essentially a hack to work
-around displaying IPA characters, but it provided speech technology and other
-fields a basis that has been widely adopted and used in code. So, SAMPA was a
-collection of tables to be compared, instead of a large universal table
-representing all languages. An extended version of SAMPA, called X-SAMPA, set
-out to include every symbol in the IPA chart including all diacritics
-\citep{WellsND}. X-SAMPA was considered more universally applicable because it
-consisted of one table that encoded the set of characters that represented
-phones/segments in IPA across languages. SAMPA and X-SAMPA have been widely used
-for speech technology and as an encoding system in computational linguistics.
-Eventually, ASCII-encoding of the IPA became deprecated through the advent of
-the Unicode Standard. Note however that many popular software packages used for
-linguistic analyses still require ASCII input, e.g.~RuG/L04 and
-SplitsTree4.\footnote{See \url{http://www.let.rug.nl/kleiweg/L04/} and
-\url{http://www.splitstree.org/}, respectively}
+Methods Phonetic Alphabet) was created to represent IPA symbols with ASCII
+character sequences, e.g. <p\textbackslash> for [ɸ]. Two problems with SAMPA are
+that (i) it is only a partial encoding of the IPA and (ii) it encodes different
+languages in separate data tables, instead of using a universal alphabet, like
+IPA.\@ SAMPA tables are derived from phonemes appearing in several European
+languages that were developed as part of a European Commission-funded project to
+address technical problems like electronic mail exchange (what is now simply
+called email). SAMPA is essentially a hack to work around displaying IPA
+characters, but it provided speech technology and other fields a basis that has
+been widely adopted and used in code. So, SAMPA was a collection of tables to be
+compared, instead of a large universal table representing all languages. An
+extended version of SAMPA, called X-SAMPA, set out to include every symbol in
+the IPA chart including all diacritics \citep{WellsND}. X-SAMPA was considered
+more universally applicable because it consisted of one table that encoded all
+characters that represented phones/segments in IPA across languages. SAMPA and
+X-SAMPA have been widely used for speech technology and as an encoding system in
+computational linguistics. Eventually, ASCII-encoding of the IPA became
+deprecated through the advent of the Unicode Standard. Note however that many
+popular software packages used for linguistic analyses still require ASCII
+input, e.g.~RuG/L04 and SplitsTree4.\footnote{See
+\url{http://www.let.rug.nl/kleiweg/L04/} and \url{http://www.splitstree.org/},
+respectively}
 
-There are a few pitfalls to be aware of when using the Unicode Standard
-to encode IPA.\@ As we have said before, from a linguistic perspective it might
-look like the Unicode Consortium is making incomprehensible decisions, but it is
-important to realize that the consortium has tried and is continuing to try to
-be as consistent as possible across a wide range of use cases, and it does place
-linguistic traditions above other orthographic possibilities. In general, we
-strongly suggest linguists not to complain about any decisions in the Unicode
-Standard, but to try and understand the rationale of the Unicode Consortium
-(which in our experience is almost always well-conceived) and devise ways to
-work with any unexpected behavior. Many of the current problems derive from the
-fact that the IPA is clearly historically based on the Latin script, but
-different enough from most other Latin-based writing systems to warrant special
-attention. This ambivalent status of the IPA glyphs (partly Latin, partly
-special) is unfortunately also attested in the treatment of IPA in the Unicode
-Standard. In retrospect, it might have been better to consider the IPA (and
-other transcription systems) to be a special or new kind of script within the
-Unicode Standard, and treat the obvious similarity to Latin glyphs as a
+There are a few pitfalls to be aware of when using the Unicode Standard to
+encode IPA.\@ As we have said before, from a linguistic perspective it might
+sometimes look like the Unicode Consortium is making incomprehensible decisions,
+but it is important to realize that the consortium has tried and is continuing
+to try to be as consistent as possible across a wide range of use cases, and it
+does place linguistic traditions above other orthographic possibilities. In
+general, we strongly suggest linguists not to complain about any decisions in
+the Unicode Standard, but to try and understand the rationale of the Unicode
+Consortium (which in our experience is almost always well-conceived) and devise
+ways to work with any unexpected behavior. Many of the current problems derive
+from the fact that the IPA is clearly historically based on the Latin script,
+but different enough from most other Latin-based writing systems to warrant
+special attention. This ambivalent status of the IPA glyphs (partly Latin,
+partly special) is unfortunately also attested in the treatment of IPA in the
+Unicode Standard. In retrospect, it might have been better to consider the IPA
+(and other transcription systems) to be a special kind of script within
+the Unicode Standard, and treat the obvious similarity to Latin glyphs as a
 historical relic. All IPA glyphs would then have their own code points, instead
@@
 
+% ==========================
 \section{Pitfall: No complete IPA code block}
 \label{pitfall-no-complete-ipa-block}
+% ==========================
 
@@
 
+% ==========================
 \section{Pitfall: IPA homoglyphs in Unicode}
 \label{pitfall-ipa-homoglyphs}
+% ==========================
 
@@
 
+% ==========================
 \section{Pitfall: Homoglyphs in IPA}
 \label{pitfall-homoglyphs-in-IPA}
+% ==========================
 
@@
 
+% ==========================
 \section{Pitfall: Ligatures and digraphs}
 \label{pitfall-ligatures-digraphs}     
+% ==========================
 
@@
 
+% ==========================
 \section{Pitfall: Missing decomposition}
 \label{pitfall-missing-decomposition}
+% ==========================
 
@@
 
+% ==========================
 \section{Pitfall: Different notions of diacritics}
 \label{pitfall-different-notions-of-diacritics}
+% ==========================
 
@@
 
+% ==========================
 \section{Pitfall: No unique diacritic ordering}
 \label{pitfall-no-unique-diacritic-ordering}
+% ==========================
 
@@
 \end{itemize}
+
+% ==========================
+\section{Recommendations}
+\label{ipa-recommendations}
+% ==========================
+
+Summary

### 2015-06-08 chapters/ipa_meets_unicode.tex
starting to edit chapter 3

@@
 e.g.~the French, German, or Danish orthographic traditions do not have a special
-status). On the other hand, the IPA uses many special symbols (like turned,
-mirrored and/or extended Latin glyphs) not found in any other Latin-based
-writing system. For this reason, and already in the first version of the Unicode
-Standard (Version 1.0 from 1991), a special block with code points, called
-\textsc{IPA Extensions} was included.
+status). On the other hand, the IPA uses many special symbols (like turned <ɐ>,
+mirrored <ɘ> and/or extended <ɧ> Latin glyphs) not found in any other Latin-based
+writing system. For this reason a special block with code points, called
+\textsc{IPA Extensions} was included already in the first version of the Unicode
+Standard (Version 1.0 from 1991).
 
@@
 Another problem is the large number of homoglyphs, i.e.~different characters
-that have highly similar glyphs (or even completely identical, depending on the
-font rendering). For example, a speaker of Russian should ideally not use the
-<а> \textsc{cyrillic small letter a} at code point \uni{0430} for IPA
-transcriptions, but instead use the <a> \textsc{latin small letter a} at code point
-\uni{0061}, although visually they are mostly indistinguishable, and the
-Cyrillic character is more easily typed on a Cyrillic keyboard.
+that have highly similar glyphs (or even completely identical glyphs, depending
+on the font rendering). For example, a user of a Cyrillic computer keyboard
+should ideally not use the <а> \textsc{cyrillic small letter a} at code point
+\uni{0430} for IPA transcriptions, but instead use the <a> \textsc{latin small
+letter a} at code point \uni{0061}, although visually they are mostly
+indistinguishable, and the Cyrillic character is more easily typed on a Cyrillic
+keyboard.
 
@@
 the <ə> \textsc{latin small letter schwa} at code point \uni{0259} and <ǝ>
-\textsc{latin small letter turned e} at \uni{01DD}. Conversely,
+\textsc{latin small letter turned e} at \uni{01DD}.
+
+Conversely,
 non-linguists are unlikely to distinguish any semantic difference between an
@@
 include precise audio recording and measurements of formants, so this may lead
-one to ask if the many characters that are used in phonological description
+one to ask if the many <a> characters that are used in phonological description
 reflects a transcriptional bias. The common use of <a> in transcriptions
@@
 software application will thus lead to the desired effect distinguishing <a>
-from <ɑ>. However, note that when the text is transferred to another font
-(i.e.~one that includes real italics) and/or to another software application
-(like Apple Pages, which does not perform slanting), then this visual appearance
-will be lost. In this case we are thus still in the pre-Unicode situation in
-which the choice of font and rendering software actually matters. The ideal
-solution from a linguistic point of view would be the introduction of a new IPA
-code point for a different kind of which explicitly specifies that it should
-still be rendered as a double-story character when italicized. After informal
-discussion with various Unicode players, our impression is that this highly
-restricted problem is not sufficiently urgent to introduce even more <a>-like
-characters in Unicode (which already lead to much confusion, see Section~\ref{pitfall-homoglyphs}).}
+from <ɑ> in italics. However, note that when the text is transferred to another
+font (i.e.~one that includes real italics) and/or to another software
+application (like Apple Pages, which does not perform slanting), then this
+visual appearance will be lost. In this case we are thus still in the
+pre-Unicode situation in which the choice of font and rendering software
+actually matters. The ideal solution from a linguistic point of view would be
+the introduction of a new IPA code point for a different kind of <a>, which
+explicitly specifies that it should still be rendered as a double-story
+character when italicized. After informal discussion with various Unicode
+players, our impression is that this highly restricted problem is not
+sufficiently urgent to introduce even more <a> homographs in Unicode (which
+already lead to much confusion, see Section~\ref{pitfall-homoglyphs}).}
 

### 2015-06-15 chapters/ipa_meets_unicode.tex
minor correction
updated diacritic ordering section and committing initial IPA profile withnames.tsv

@@
 
-Furthermore, even linguists are unlikely to distinguish between
-the <ə> \textsc{latin small letter schwa} at code point \uni{0259} and <ǝ>
+Some other homoglyphs related to encoding IPA in the Unicode Standard are:
+
+\begin{itemize}
+
+  \item The uses of the apostrophe has led to long discussions on the Unicode
+     Standard email list. An English keyboard inputs <{\fontspec{Monaco}'}> \textsc{apostrophe}
+     at \uni{0027}, although the preferred Unicode apostrophe is the <\ ' >
+     \textsc{right single quotation mark} at \uni{2019}.
+  \item The glottal stop/glottalization/ejective marker is another completely
+      different character <{\large ʼ}>, the \textsc{modifier letter apostrophe} at
+      \uni{02BC}, but unfortunately looks mostly highly similar to
+      \uni{2019}. 
+  \item Another problem is the <ˁ> \textsc{modifier letter reversed
+     glottal stop} at \uni{02C1} vs.\@ the <ˤ> \textsc{modifier
+     letter small reversed glottal stop} at \uni{02E4}. Both 
+     appear in various resources representing phonetic data online. This is
+     thus a clear example for which the Unicode Standard does not solve the
+     linguistic standardization problem.
+  \item even linguists are unlikely to distinguish between the <ə>
+     \textsc{latin small letter schwa} at code point \uni{0259} and <ǝ>
      \textsc{latin small letter turned e} at \uni{01DD}.
+  \item There is at least one case in which the character name assigned by the
+     Unicode Consortium does not match the IPA's description. In the Unicode
+     Standard the <ǃ> at \uni{01C3} is labeled \textsc{latin letter retroflex
+     click}, but in IPA that glyph is used for an alveolar or postalveolar
+     click (not retroflex). This naming is probably best seen as a simple
+     error in the Unicode Standard. This character is of course often simply
+     typed as <!> \textsc{exclamation mark} at \uni{0021}.
+
+\end{itemize}
 
@@
 already lead to much confusion, see Section~\ref{pitfall-homoglyphs}).}
-
-Some other homoglyphs related to encoding IPA in the Unicode Standard are:
-
-\begin{itemize}
-	\item The uses of the apostrophe has led to long discussions on the Unicode
-       Standard email list. An English keyboard inputs <{\fontspec{Monaco}'}> \textsc{apostrophe}
-       at \uni{0027}, although the preferred Unicode apostrophe is the <\ ' >
-       \textsc{right single quotation mark} at \uni{2019}.
-     \item The glottal stop/glottalization/ejective marker is another completely
-        different character <{\large ʼ}>, the \textsc{modifier letter apostrophe} at
-        \uni{02BC}, but unfortunately looks mostly highly similar to
-        \uni{2019}. 
-	\item Another problem is the <ˁ> \textsc{modifier letter reversed
-       glottal stop} at \uni{02C1} vs.\@ the <ˤ> \textsc{modifier
-       letter small reversed glottal stop} at \uni{02E4}. Both 
-       appear in various resources representing phonetic data online. This is
-       thus a clear example for which the Unicode Standard does not solve the
-       linguistic standardization problem. 
-	\item There is at least one case in which the character name assigned by the
-       Unicode Consortium does not match the IPA's description. In the Unicode
-       Standard the <ǃ> at \uni{01C3} is labeled \textsc{latin letter retroflex
-       click}, but in IPA that glyph is used for an alveolar or postalveolar
-       click (not retroflex). This naming is probably best seen as a simple
-       error in the Unicode Standard. This character is of course often simply
-       typed as <!> \textsc{exclamation mark} at \uni{0021}. \end{itemize}
 
@@
 reasoning of the transcriber. However, such marks have to be explicitly ordered
-if sequences of them are to be interoperable and compatible. An example is a
+if sequences of them are to be interoperable and compatible computationally. An example is a
 labialized aspirated alveolar plosive: <tʷʰ>. There is nothing holding linguists
@@
 In accordance with the absence of any specification of ordering in the IPA, the
-Unicode Standard likewise does not propose any standard orderings. Both leave it
-to the user to be consistent. However, there is one aspect of ordering for which
-the Unicode Standard does present a canonical solution, which is uncontroversial
+Unicode Standard likewise does not propose any standardized orders. Both leave it
+to the user to be consistent; this approach naturally invites inconsistency across 
+different authored resources.
+
+There is one aspect of ordering for which
+the Unicode Standard does present a canonical solution. However, it is uncontroversial
 from a linguistic perspective. Diacritics in the Unicode Standard
@@
 positions are put in a specified order. This process therefore harmonizes the
-difference between different encodings, for example, of a midtone creaky voice
+difference between different encodings, e.g.\ of a midtone creaky voice
 vowel <ḛ̄>. This grapheme cluster can be encoded either as <e> + <\ \ ̄> + <\ \
@@
 languages, \citet[540]{Moran2012} defines a set of diacritic ordering
-conventions.\footnote{The most recent version of these conventions is online at
+conventions. As new languages' phonological inventories are added to the 
+database,\footnote{\url{http://phoible.org/}} updates and additions to the 
+conventions are published online by \cite{MoranMcCloy2014}.\footnote{
 \url{http://phoible.github.io/conventions/}} The conventions are influenced by
-the linguistic literature, though some ad-hoc decisions had to be taken. The
-goal was to explicitly define all character sequences so that the vast variety
-of phonemes found in descriptions of the world's language were normalized into
-consistent character sequences, e.g.~if one language description uses and
-another, when both are intended to be phonetically equivalent, then a decision
+the linguistic literature, though some ad-hoc decisions had to be taken, like 
+introducing several symbols for widely attested sounds that lack representation 
+in the current version of the IPA. Whenever possible, additions were drawn from 
+the extIPA symbols for disordered speech.\footnote{\url{https://www.internationalphoneticassociation.org/sites/default/files/extIPAChart2008.pdf}}
+
+\cite{MoranMcCloy2014} explicitly define all character sequences so that the vast variety
+of phonemes found in a large sample of descriptions of the world's language were normalized into
+consistent character sequences, e.g.~if one language description uses <kʲʼ> and
+another <kʼʲ>, when both are intended to be phonetically equivalent, then a decision
 to normalize to one form was taken. For example, when a character sequence
@@
 	\begin{itemize}
-	  \item[→] (unreleased <\ \ ̚ > \textbar{} lateral release <ˡ> \textbar{} nasal release <ⁿ>)
+	  \item[→] (unreleased <\ \ ̚ >)
+	  \item[→] (rhotic hook <\ \ ˞ >)
+	  \item[→] (lateral release <ˡ> \textbar{} nasal release <ⁿ>)
+	  \item[→] (labialized <ʷ>)
 	  \item[→] (palatalized <ʲ>)
-	  \item[→] (labialized <ʷ>)
 	  \item[→] (velarized <ˠ>)
 	  \item[→] (pharyngealized <ˤ>)
+	  \item[→] (glottalized <ˀ>)
 	  \item[→] (aspirated <ʰ> \textbar{} ejective <ʼ>)
@@
 If a character sequence contains more than one diacritic below the base
-character, then the place feature is applied first (dental, laminal, apical,
-fronted, backed, lowered, raised), then the laryngeal setting (voiced,
-voiceless, creaky voice, breathy voice) and finally the syllabic or non-syllabic
-marker (for vowels, ATR gets put on between the place and laryngeal setting).
+character, then the place features are applied first (linguolabial, dental, apical, laminal, advanced, 
+retracted), followed by the manner features (raised, lowered, advanced tongue root, 
+retracted tongue root, frictionalized), then secondary articulations (more round, less round, 
+derhoticized), laryngeal settings (creaky, breathy, voiced, devoiced (below), devoiced (above)), and finally the syllabic or non-syllabic
+marker (for vowels, advanced and retracted tongue root markers gets applied between the place and laryngeal setting).
 So:
@@
 	\begin{itemize}	
-	  \item[→] (dental <t̪> \textbar{} laminal <t̻> \textbar{} apical <t̺>) 
-	  \item[→] (fronted <u̟> \textbar{} backed <e̠>)
-	  \item[→] (lowered <e̞> \textbar{} raised <e̝>) 
-	  \item[→] (ATR <e̘ e̙>)
-	  \item[→] (voiced <s̬> \textbar{} voiceless <n̥> \textbar{} creaky voice <b̰> \textbar{} breathy voice <b̤>)
+	  \item[→] (linguolabial <t̼> \textbar{} dental <t̪> \textbar{} apical <t̺> \textbar{} laminal <t̻>)
+	  \item[→] (advanced <u̟> \textbar{} retracted <e̠>)	 
+	  \item[→] (raised <e̝> \textbar{} lowered <e̞>) 
+	  \item[→] (advanced tongue root <e̘> \textbar{} retracted tongue root <e̙>)
+	  \item[→] (frictionalized <ɾ͓>)
+	  \item[→] (more round <ɔ̹> \textbar{} less round <ɔ̜>)
+	  \item[→] (creaky voice <b̰> \textbar{} breathy voice <b̤> \textbar{} voiced <s̬> \textbar{} devoiced <n̥>)
 	  \item[→] (syllabic <n̩> \textbar{} non-syllabic <e̯>)
@@
 problematic in \citet{Moran2012} because they include only the centralized,
-mid-centralized and nasalized combining characters. \citet{Moran2012} marks
+mid-centralized and nasalized combining characters.
+
+\citet{Moran2012,MoranMcCloy2014} mark
 tones as singletons with Space Modifier Letters, e.g. \textless{}˦\textgreater{}
 for a phonemic high tone, instead of accent diacritics, alleviating potential
-conflicts. Building on the work of \citet{Moran2012}, if a character sequence
+conflicts. Building on these works, if a character sequence
 contains more than one diacritic above the base character, we propose:

### 2015-06-17 chapters/ipa_meets_unicode.tex
adapted block list to new strict IPA profile

@@
 \begin{itemize}
-	\item Basic Latin (30 characters), e.g. <a b c d e> 
-	\item Latin-1 Supplement (4 characters): <æ ç ð ø> 
-	\item Latin Extended-A (3 characters): <ħ ŋ œ> 
-	\item Latin Extended-B (5 characters): <ǀ ǁ ǂ ǃ ȵ> 
-	\item IPA Extensions (70 characters), e.g. <ɐ ɑ ɔ> 
-	\item Spacing Modifier Letters (20 characters), e.g. <ʰ ʷ ˥> 
-	\item Combining Diacritical Marks (33 characters), e.g. <{\large \ \ ̝\ \ ̥\ \ ̪ }> 
-	\item Greek and Coptic (3 characters): <β θ χ> 
-	\item Phonetic Extensions (2 characters): <{\small \fontspec{CharisSIL}ᴅ ᴴ}> 
-	\item Phonetic Extensions Supplement (3 characters): <{\small \fontspec{CharisSIL}ᶑ ᶾ ᶣ}> 
-	\item Superscripts and Subscripts (1 character): <ⁿ> 
-	\item Arrows (4 characters): <↑ ↓ ↗ ↘>
-	\item Latin Extended-C (1 character): <{\small \fontspec{CharisSIL}ⱱ}> 
+	\item Basic Latin (27 characters): \newline < a~b~c~d~e~f~h~i~j~k~l~m~n~o~p~q~r~s~t~u~v~w~x~y~z~.~| >
+	\item Latin-1 Supplement (4 characters): \newline < æ ç ð ø >
+	\item Latin Extended-A (3 characters): \newline < ħ ŋ œ >
+	\item Latin Extended-B (4 characters): \newline < ǀ ǁ ǂ ǃ >
+	\item IPA Extensions (68 characters): \newline < ɐ ɑ ɒ ɓ ɔ ɕ ɖ ɗ ɘ ə ɛ ɜ ɞ ɟ ɠ ɡ ɢ ɣ ɤ ɥ ɦ ɧ ɨ ɪ ɫ ɬ ɭ ɮ ɯ ɰ ɱ ɲ ɳ ɴ ɵ ɶ ɸ ɹ ɺ ɻ ɽ ɾ ʀ ʁ ʂ ʃ ʄ ʈ ʉ ʊ ʋ ʌ ʍ ʎ ʏ ʐ ʑ ʒ ʔ ʕ ʘ ʙ ʛ ʜ ʝ ʟ ʡ ʢ > 
+	\item Spacing Modifier Letters (17 characters): \newline < ʰ ʲ ʷ ʼ ˈ ˌ ː ˑ ˞ ˠ ˡ ˤ ˥ ˦ ˧ ˨ ˩
+	\item Combining Diacritical Marks (29 characters): \newline < {\large \  ̃\ \ ̆\ \ ̈\ ̚\ \ ̘\ \ ̙\ \ \ ̜\ \ ̝\ \ ̞\ \ ̟\ \ ̠\ \ ̤\ \ ̥\ \ ̩\ \ ̪\ \ ̬\ \ ̯\ \ ̰\ \ ̹\ \ ̺\ \ ̻\ \ ̼\ \ ̽\ \ ͜\  ̋\ \ ́\ \ ̄\ \ ̀\ \ ̏\ } >
+	\item Greek and Coptic (3 characters): \newline < β θ χ >
+%	\item Phonetic Extensions (2 characters): \charis{ᴅ ᴴ} 
+	\item Phonetic Extensions Supplement (1 characters): \newline < \charis{ᶑ} >
+    \item General Punctuation (2 characters): \newline < ‖ \charis{‿} >
+	\item Superscripts and Subscripts (1 character): \newline < ⁿ > 
+	\item Arrows (4 characters): \newline < ↑ ↓ ↗ ↘ >
+	\item Latin Extended-C (1 character): \newline < \charis{ⱱ} >
 \end{itemize}
@@
 database,\footnote{\url{http://phoible.org/}} updates and additions to the 
-conventions are published online by \cite{MoranMcCloy2014}.\footnote{
+conventions are published online by~\cite{MoranMcCloy2014}.\footnote{
 \url{http://phoible.github.io/conventions/}} The conventions are influenced by
@@
 introducing several symbols for widely attested sounds that lack representation 
-in the current version of the IPA. Whenever possible, additions were drawn from 
+in the current version of the IPA.\@ Whenever possible, additions were drawn from 
 the extIPA symbols for disordered speech.\footnote{\url{https://www.internationalphoneticassociation.org/sites/default/files/extIPAChart2008.pdf}}

### 2015-06-17 chapters/ipa_meets_unicode.tex
update IPA homoglyphs section

@@
 
-Some other homoglyphs related to encoding IPA in the Unicode Standard are:
+\
 
+\noindent Some problematic homoglyphs related to encoding IPA in the Unicode Standard are 
+the following.
 \begin{itemize}
@@
      at \uni{0027}, although the preferred Unicode apostrophe is the <\ ' >
-     \textsc{right single quotation mark} at \uni{2019}.
-  \item The glottal stop/glottalization/ejective marker is another completely
-      different character <{\large ʼ}>, the \textsc{modifier letter apostrophe} at
-      \uni{02BC}, but unfortunately looks mostly highly similar to
-      \uni{2019}. 
+     \textsc{right single quotation mark} at \uni{2019}.\footnote{Note that many word 
+     processors (like Microsoft Word) by default will replace straight quotes by 
+     curly quotes, depending on the whitespace around it. Such a replacement is 
+     a real character replacement, meaning that another Unicode code point is 
+     inserted.} However, the glottal stop/glottalization/ejective marker is yet 
+     another completely different character <{\large ʼ}>, the 
+     \textsc{modifier letter apostrophe} at \uni{02BC}, which unfortunately looks 
+     mostly extremely similar to \uni{2019}. 
   \item Another problem is the <ˁ> \textsc{modifier letter reversed
@@
      \textsc{latin small letter turned e} at \uni{01DD}.
-  \item There is at least one case in which the character name assigned by the
-     Unicode Consortium does not match the IPA's description. In the Unicode
-     Standard the <ǃ> at \uni{01C3} is labeled \textsc{latin letter retroflex
+  \item  The alveolar click <ǃ> at \uni{01C3} is of course often simply
+     typed as <!> \textsc{exclamation mark} at \uni{0021}.\footnote{In the Unicode
+     Standard the <ǃ> at \unif{01C3} is labeled \textsc{latin letter retroflex
      click}, but in IPA that glyph is used for an alveolar or postalveolar
-     click (not retroflex). This naming is probably best seen as a simple
-     error in the Unicode Standard. This character is of course often simply
-     typed as <!> \textsc{exclamation mark} at \uni{0021}.
+     click (not retroflex). This naming is probably best seen as an 
+     error in the Unicode Standard.}
 
@@
 
-Conversely,
-non-linguists are unlikely to distinguish any semantic difference between an
-open back unrounded vowel <ɑ> \textsc{latin small letter alpha} at
-\uni{0251}, and the open front unrounded vowel <a> \textsc{latin small letter
-a} at \uni{0061}. But even among linguists this distinction leads to problems.
-For example, as pointed out by \citet{Mielke2009}, there is a problem stemming
-from the fact that about 75\% of languages are reported to have a five-vowel
-system \citep{Maddieson1984}. Historically, linguistic descriptions tend not to
-include precise audio recording and measurements of formants, so this may lead
-one to ask if the many <a> characters that are used in phonological description
-reflects a transcriptional bias. The common use of <a> in transcriptions
-could be in part due to the ease of typing the letter on an English keyboard (or
-for older descriptions, the typewriter). We found it to be exceedingly rare that
-a linguist uses <ɑ> for a low back unrounded vowel.\footnote{One example is
+Conversely, non-linguists are unlikely to distinguish any semantic difference
+between an open back unrounded vowel <ɑ> \textsc{latin small letter alpha} at
+\uni{0251}, and the open front unrounded vowel <a> \textsc{latin small letter a}
+at \uni{0061}, basically treating them as homoglyphs, although they are
+different phonetic symbols.
+
+But even among linguists this distinction leads to problems. For example, as
+pointed out by \citet{Mielke2009}, there is a problem stemming from the fact
+that about 75\% of languages are reported to have a five-vowel system
+\citep{Maddieson1984}. Historically, linguistic descriptions tend not to include
+precise audio recording and measurements of formants, so this may lead one to
+ask if the many <a> characters that are used in phonological description
+reflects a transcriptional bias. The common use of <a> in transcriptions could
+be in part due to the ease of typing the letter on an English keyboard (or for
+older descriptions, the typewriter). We found it to be exceedingly rare that a
+linguist uses <ɑ> for a low back unrounded vowel.\footnote{One example is
 \citet[75]{Vidal2001a}, in which the author states: ``The definition of Pilagá
-/a/ as [+back] results from its behavior in certain phonological contexts.
-For instance, uvular and pharyngeal consonants only occur around /a/ and /o/.
-Hence, the characterization of /a/ and /o/ as a natural class of (i.e.,
-[+back] vowels), as opposed to /i/ and /e/.''} They simply use <a> as
-long as there is no opposition to <ɑ>.\footnote{See Thomason's Language Log
-post, ``Why I don't love the International Phonetic Alphabet'' at:
-\url{http://itre.cis.upenn.edu/~myl/languagelog/archives/005287.html}.}
+/a/ as [+back] results from its behavior in certain phonological contexts. For
+instance, uvular and pharyngeal consonants only occur around /a/ and /o/. Hence,
+the characterization of /a/ and /o/ as a natural class of (i.e., [+back]
+vowels), as opposed to /i/ and /e/.''} They simply use <a> as long as there is
+no opposition to <ɑ>.
+
+%\footnote{See Thomason's Language Log post, ``Why I don't love the International Phonetic Alphabet'' at:\url{http://itre.cis.upenn.edu/~myl/languagelog/archives/005287.html}.}
 

### 2015-06-17 chapters/ipa_meets_unicode.tex
update to homoglyphs in IPA section

@@
 indistinguishable, and the Cyrillic character is more easily typed on a Cyrillic
-keyboard.
-
-\
-
-\noindent Some problematic homoglyphs related to encoding IPA in the Unicode Standard are 
+keyboard. Some problematic homoglyphs related to encoding IPA in the Unicode Standard are 
 the following.
@@
      processors (like Microsoft Word) by default will replace straight quotes by 
-     curly quotes, depending on the whitespace around it. Such a replacement is 
-     a real character replacement, meaning that another Unicode code point is 
-     inserted.} However, the glottal stop/glottalization/ejective marker is yet 
+     curly quotes, depending on the whitespace around it.} However, the glottal stop/glottalization/ejective marker is yet 
      another completely different character <{\large ʼ}>, the 
@@
 at \uni{0061}, basically treating them as homoglyphs, although they are
-different phonetic symbols.
-
-But even among linguists this distinction leads to problems. For example, as
+different phonetic symbols. But even among linguists this distinction leads to problems. For example, as
 pointed out by \citet{Mielke2009}, there is a problem stemming from the fact
@@
   \item The marking of <g>
-  \item The diacritic for voiceless
+  \item The placement of diacritics
   \item The tie bar to indicate a close bond between sounds
@@
 of how to transcribe tone. There is an old tradition to use diacritics on vowels
-to mark different tone levels, e.g. <ȅèée̋>.\footnote{There are at least two
-different Unicode homoglyphs for the low and high level tones, namely <\ ̀~>
-\textsc{combining grave tone mark} at \unif{0340} vs.\@ <\ ̀~> \textsc{combining
-grave accent} at \unif{0300} for low tone, and <\ ́~> \textsc{combining acute tone
-mark} at \unif{0341} vs.\@ <\ ́~> \textsc{combining acute accent} at \unif{0301}
-for high tone.} The IPA also proposes a less widespread used, but more
-consistent option of tone letters, e.g. <˥˦˧˨˩>. Tone letters in the IPA have
-five different levels, and sequences of these letters can be used to indicate
-contours. Well-designed fonts will even merge a sequence of tone letters into a
-contour. For example, compare the font Linux Libertine, which does not merge
-tone letters <{\fontspec{LinLibertineO}˥˨˧˩}>, with the font CharisSIL, which
-merges this sequence of four tone letters into a single contour <\charis{˥˨˧˩}>.
-For strict IPA encoding we propose to standardize on tone letters.
+to mark different tone levels, e.g. <ȅèée̋>.\footnote{To make things even more
+complexe, there are at least two different Unicode homoglyphs for the low and
+high level tones, namely <\ ̀~> \textsc{combining grave tone mark} at
+\unif{0340} vs.\@ <\ ̀~> \textsc{combining grave accent} at \unif{0300} for low
+tone, and <\ ́~> \textsc{combining acute tone mark} at \unif{0341} vs.\@ <\ ́~>
+\textsc{combining acute accent} at \unif{0301} for high tone.} The IPA also
+proposes the option of tone letters, e.g. <˥˦˧˨˩>, which are used much less, but
+are more consistent for contours. Tone letters in the IPA have five different levels, and
+sequences of these letters can be used to indicate contours. Well-designed fonts
+will even merge a sequence of tone letters into a contour. For example, compare
+the font Linux Libertine, which does not merge tone letters
+<{\fontspec{LinLibertineO}˥˨˧˩}>, with the font CharisSIL, which merges this
+sequence of four tone letters into a single contour <\charis{˥˨˧˩}>. For strict
+IPA encoding we propose to standardize on tone letters.
 
@@
 time, whereas the latter must be cumbersomely inserted as a special symbol in
-most software. This issue was recently addressed by The International Phonetic
+most software. The International Phonetic
 Association has taken the stance that both the keyboard \textsc{latin small

### 2015-06-17 chapters/ipa_meets_unicode.tex
update section missing decomposition

@@
 of characters (typically affricates) are available as single code-points in the
-Unicode Standard, but are designated as \textsc{ligatures} or \textsc{digraphs}
-(confusingly both names appear interchangeably). Such glyphs might be used by
+Unicode Standard, but are designated \textsc{digraphs}. Such glyphs might be used by
 software to produce a pleasing display, but they should not be hard-coded into
@@
 in IPA that allow for multiple, apparently identical, encodings that are not
-canonical equivalent. The following elements should not be treated as diacritics 
-when encoding IPA in Unicode:
+canonical equivalent (see Section~\ref{pitfall-canonical-equivalence}),. The
+following elements should not be treated as diacritics when encoding IPA in
+Unicode: 
 \begin{itemize}
@@
   \item[] <\ \ {\large  ̴}> \textsc{combining tilde overlay} at \uni{0334}
-  \item[] <{\large ˞}> \textsc{modifier letter rhotic hook} at \uni{02DE}
 \end{itemize}
@@
 There turn out to be a lot of characters in the IPA that could be conceived as 
-using any of these elements, like <ɲ>, <ɳ>, <ɨ>, <ø>, <ɫ> or <ɚ>. However, all 
-such characters exist as well as ``precomposed'' combination in Unicode, and these 
+using any of these elements, like <ɲ>, <ɳ>, <ɨ>, <ø> or <ɫ>. However, all 
+such characters exist as well as precomposed combination in Unicode, and these 
 precomposed characters should preferably be used. When combinations of a base 
@@
 
-Reversely, <ç> \textsc{latin small letter c with cedilla} at \uni{00E7} will be
-decomposed into <c> \textsc{latin small letter c} at \uni{0063} with 
-<\ \ {\large  ̧}> \textsc{combining cedilla} at \uni{0327}, 
-also if such a decomposition is not intended, because it is meaningless in 
-IPA.\@
+A similar problem arises with the rhotic hook. There are two precomposed
+characters in Unicode with a rhotic hook, which are not canonical equivalent 
+with a combination of the vowel with a separately encoded hook:
+\begin{itemize}
+  \item[] <ɚ> \textsc{latin small letter schwa with hook} at \uni{025A}
+  \item[] <ɝ> \textsc{latin small letter reversed open e with hook}at \uni{025D}
+\end{itemize}
+
+All other combinations of vowels with rhotic hooks will have to be made by using
+<{\large ˞}> \textsc{modifier letter rhotic hook} at \uni{02DE}, because there
+is not complete set of precomposed characters with rhotic hooks. For that reason
+we propose to not use the two precomposed characters with hooks mentioned above,
+but always use the separate rhotic hook at \uni{02DE} in IPA.\@
+
+Reversely, note that the <ç> \textsc{latin small letter c with cedilla} at
+\uni{00E7} will be decomposed into <c> with <\ \ {\large ̧}> \textsc{combining
+cedilla} at \uni{0327} by Unicode canonical decomposition, also if such a
+decomposition is not intended in the IPA.\@
 

### 2015-06-17 chapters/ipa_meets_unicode.tex
update section diacritic ordering
I removed most references to PHOIBLE here. That is more a use case for
later. This section should be about “strict” IPA

@@
   \item[] <ɚ> \textsc{latin small letter schwa with hook} at \uni{025A}
-  \item[] <ɝ> \textsc{latin small letter reversed open e with hook}at \uni{025D}
+  \item[] <ɝ> \textsc{latin small letter reversed open e with hook} at \uni{025D}
 \end{itemize}
@@
 
-Another pitfall relates the concept of what are diacritics. The problem is that
+Another pitfall relates to the question of what are diacritics. The problem is that
 the meaning of the term diacritics as used by the IPA is not the same as is used
@@
 	  \item[] <ˤ> \textsc{modifier letter small reversed glottal stop} at \uni{02E4}
+	  \item[] <ⁿ> \textsc{superscript latin small letter n} at \uni{207F}
 	  \item[] and many more like this
@@
 languages, \citet[540]{Moran2012} defines a set of diacritic ordering
-conventions. As new languages' phonological inventories are added to the 
-database,\footnote{\url{http://phoible.org/}} updates and additions to the 
-conventions are published online by~\cite{MoranMcCloy2014}.\footnote{
-\url{http://phoible.github.io/conventions/}} The conventions are influenced by
-the linguistic literature, though some ad-hoc decisions had to be taken, like 
-introducing several symbols for widely attested sounds that lack representation 
-in the current version of the IPA.\@ Whenever possible, additions were drawn from 
-the extIPA symbols for disordered speech.\footnote{\url{https://www.internationalphoneticassociation.org/sites/default/files/extIPAChart2008.pdf}}
-
-\cite{MoranMcCloy2014} explicitly define all character sequences so that the vast variety
-of phonemes found in a large sample of descriptions of the world's language were normalized into
-consistent character sequences, e.g.~if one language description uses <kʲʼ> and
-another <kʼʲ>, when both are intended to be phonetically equivalent, then a decision
-to normalize to one form was taken. For example, when a character sequence
+conventions. The conventions are influenced by the linguistic literature, though
+some ad-hoc decisions had to be taken. For example, when a character sequence
 contains more than one character in Spacing Modifier Letters, the order that is
@@
 	\begin{itemize}
-	  \item[→] (unreleased <\ \ ̚ >)
-	  \item[→] (rhotic hook <\ \ ˞ >)
-	  \item[→] (lateral release <ˡ> \textbar{} nasal release <ⁿ>)
-	  \item[→] (labialized <ʷ>)
-	  \item[→] (palatalized <ʲ>)
-	  \item[→] (velarized <ˠ>)
-	  \item[→] (pharyngealized <ˤ>)
-	  \item[→] (glottalized <ˀ>)
-	  \item[→] (aspirated <ʰ> \textbar{} ejective <ʼ>)
-	  \item[→] (long <ː> \textbar{} half-long <ˑ>)
+	  \item[→] rhotic hook <{\large \ ˞} >
+	  \item[→] lateral release <ˡ> \textbar{} nasal release <ⁿ>
+	  \item[→] labialized <ʷ>
+	  \item[→] palatalized <ʲ>
+	  \item[→] velarized <ˠ>
+	  \item[→] pharyngealized <ˤ>
+	  \item[→] glottalized <ˀ>
+	  \item[→] aspirated <ʰ> \textbar{} ejective <ʼ>
+	  \item[→] long <ː> \textbar{} half-long <ˑ>
 	\end{itemize}
@@
 If a character sequence contains more than one diacritic below the base
-character, then the place features are applied first (linguolabial, dental, apical, laminal, advanced, 
-retracted), followed by the manner features (raised, lowered, advanced tongue root, 
-retracted tongue root, frictionalized), then secondary articulations (more round, less round, 
-derhoticized), laryngeal settings (creaky, breathy, voiced, devoiced (below), devoiced (above)), and finally the syllabic or non-syllabic
-marker (for vowels, advanced and retracted tongue root markers gets applied between the place and laryngeal setting).
-So:
+character, then the place features are applied first (linguolabial, dental,
+apical, laminal, advanced, retracted), followed by the manner features (raised,
+lowered, advanced tongue root, retracted tongue root, frictionalized), then
+secondary articulations (more round, less round, derhoticized), laryngeal
+settings (creaky, breathy, voiced, devoiced (below), devoiced (above)), and
+finally the syllabic or non-syllabic marker (for vowels, advanced and retracted
+tongue root markers gets applied between the place and laryngeal setting). So:
 
@@
 	\begin{itemize}	
-	  \item[→] (linguolabial <t̼> \textbar{} dental <t̪> \textbar{} apical <t̺> \textbar{} laminal <t̻>)
-	  \item[→] (advanced <u̟> \textbar{} retracted <e̠>)	 
-	  \item[→] (raised <e̝> \textbar{} lowered <e̞>) 
-	  \item[→] (advanced tongue root <e̘> \textbar{} retracted tongue root <e̙>)
-	  \item[→] (frictionalized <ɾ͓>)
-	  \item[→] (more round <ɔ̹> \textbar{} less round <ɔ̜>)
-	  \item[→] (creaky voice <b̰> \textbar{} breathy voice <b̤> \textbar{} voiced <s̬> \textbar{} devoiced <n̥>)
-	  \item[→] (syllabic <n̩> \textbar{} non-syllabic <e̯>)
+	  \item[→] linguolabial <t̼> \textbar{} dental <t̪> \textbar{} apical <t̺> \textbar{} laminal <t̻>
+	  \item[→] advanced <u̟> \textbar{} retracted <e̠> 
+	  \item[→] raised <e̝> \textbar{} lowered <e̞>
+	  \item[→] advanced tongue root <e̘> \textbar{} retracted tongue root <e̙>
+	  \item[→] frictionalized <ɾ͓>
+	  \item[→] more round <ɔ̹> \textbar{} less round <ɔ̜>
+	  \item[→] creaky voice <b̰> \textbar{} breathy voice <b̤> \textbar{} voiced <s̬> \textbar{} devoiced <n̥>
+	  \item[→] syllabic <n̩> \textbar{} non-syllabic <e̯>
 	\end{itemize}
@@
 
-Character sequences with diacritics above the base character were not
-problematic in \citet{Moran2012} because they include only the centralized,
-mid-centralized and nasalized combining characters.
-
-\citet{Moran2012,MoranMcCloy2014} mark
-tones as singletons with Space Modifier Letters, e.g. \textless{}˦\textgreater{}
-for a phonemic high tone, instead of accent diacritics, alleviating potential
-conflicts. Building on these works, if a character sequence
-contains more than one diacritic above the base character, we propose:
+Finally, if a character sequence contains more than one diacritic above the base
+character, we propose the following order, with Spacing Modifiers and tone
+letters being added at the end:
 
@@
 	\begin{itemize}
-	  \item[→] (centralized <ë> \textbar{} mid-centralized <e̽>)
-	  \item[→] (extra short <ĕ>)
-	  \item[→] (tone accents, e.g. <è>)
-	  \item[→] (Spacing Modifier Letters, see above)
-	  \item[→] (tone letters, e.g. <˦>)
+	  \item[→] centralized <ë> \textbar{} mid-centralized <e̽>
+	  \item[→] extra short <ĕ>
+%	  \item[→] (tone accents, e.g. <è>)
+	  \item[→] unreleased <e̚>
+	  \item[→] Spacing Modifier Letters, see above
+	  \item[→] tone letters <˥ ˦ ˧ ˨ ˩>
 	\end{itemize}	 

### 2015-06-17 chapters/use_cases.tex
reverting this morning's width adjustments
update section diacritic ordering
I removed most references to PHOIBLE here. That is more a use case for
later. This section should be about “strict” IPA

@@
 \end{comment}
+
+\section{PHOIBLE}
+
+As new languages' phonological inventories are added to the 
+database,\footnote{\url{http://phoible.org/}} updates and additions to the 
+conventions are published online by~\cite{MoranMcCloy2014}.\footnote{
+\url{http://phoible.github.io/conventions/}} , like 
+introducing several symbols for widely attested sounds that lack representation 
+in the current version of the IPA.\@ Whenever possible, additions were drawn from 
+the extIPA symbols for disordered speech.\footnote{\url{https://www.internationalphoneticassociation.org/sites/default/files/extIPAChart2008.pdf}}
+
+\cite{MoranMcCloy2014} explicitly define all character sequences so that the
+vast variety of phonemes found in a large sample of descriptions of the world's
+language were normalized into consistent character sequences, e.g.~if one
+language description uses <kʲʼ> and another <kʼʲ>, when both are intended to be
+phonetically equivalent, then a decision to normalize to one form was taken.
+
+Character sequences with diacritics above the base character were not
+problematic in \citet{Moran2012} because they include only the centralized,
+mid-centralized and nasalized combining characters.
+
+\citet{Moran2012;MoranMcCloy2014} mark tones as singletons with Space Modifier
+Letters, e.g. <˦> for a phonemic high tone, instead of accent diacritics,
+alleviating potential conflicts. Building on these works,

### 2015-06-17 chapters/ipa_meets_unicode.tex
update profile, first attempt at large table

@@
 	\item IPA Extensions (68 characters): \newline < ɐ ɑ ɒ ɓ ɔ ɕ ɖ ɗ ɘ ə ɛ ɜ ɞ ɟ ɠ ɡ ɢ ɣ ɤ ɥ ɦ ɧ ɨ ɪ ɫ ɬ ɭ ɮ ɯ ɰ ɱ ɲ ɳ ɴ ɵ ɶ ɸ ɹ ɺ ɻ ɽ ɾ ʀ ʁ ʂ ʃ ʄ ʈ ʉ ʊ ʋ ʌ ʍ ʎ ʏ ʐ ʑ ʒ ʔ ʕ ʘ ʙ ʛ ʜ ʝ ʟ ʡ ʢ > 
-	\item Spacing Modifier Letters (17 characters): \newline < ʰ ʲ ʷ ʼ ˈ ˌ ː ˑ ˞ ˠ ˡ ˤ ˥ ˦ ˧ ˨ ˩
+	\item Spacing Modifier Letters (17 characters): \newline < ʰ ʲ ʷ ʼ ˈ ˌ ː ˑ ˞ ˠ ˡ ˤ ˥ ˦ ˧ ˨ ˩ >
 	\item Combining Diacritical Marks (29 characters): \newline < {\large \  ̃\ \ ̆\ \ ̈\ ̚\ \ ̘\ \ ̙\ \ \ ̜\ \ ̝\ \ ̞\ \ ̟\ \ ̠\ \ ̤\ \ ̥\ \ ̩\ \ ̪\ \ ̬\ \ ̯\ \ ̰\ \ ̹\ \ ̺\ \ ̻\ \ ̼\ \ ̽\ \ ͜\  ̋\ \ ́\ \ ̄\ \ ̀\ \ ̏\ } >
@@
 There turn out to be a lot of characters in the IPA that could be conceived as
-using any of these elements, like <ɲ>, <ɳ>, <ɨ>, <ø> or <ɫ>. However, all 
-such characters exist as well as precomposed combination in Unicode, and these 
-precomposed characters should preferably be used. When combinations of a base 
-character with diacritic are used, then these combinations are not canonical 
-equivalent to the precomposed combinations. This means that any search will not 
-find both at the same time.
+using any of these elements, like <ɲ>, <ɳ>, <ɨ>, <ø> or <ɫ>. However, all such
+characters exist as well as precomposed combination in Unicode, and these
+precomposed characters should preferably be used.\footnote{The IPA does not
+describe any character for a voiced retroflex implosive, which would
+transparently be \charis{ᶑ}. We propose to add this character to the IPA.} When
+combinations of a base character with diacritic are used, then these
+combinations are not canonical equivalent to the precomposed combinations. This
+means that any search will not find both at the same time.
 
@@
 conventions. The conventions are influenced by the linguistic literature, though
-some ad-hoc decisions had to be taken. For example, when a character sequence
-contains more than one character in Spacing Modifier Letters, the order that is
-proposed is the following (where <\textbar{}> indicates \textit{or} and <→>
-indicates \textit{precedes}):
-
-\begin{itemize}
-	\item[] \textsc{Spacing Modifier Letters ordering:}
-	\begin{itemize}
-	  \item[→] rhotic hook <{\large \ ˞} >
-	  \item[→] lateral release <ˡ> \textbar{} nasal release <ⁿ>
-	  \item[→] labialized <ʷ>
-	  \item[→] palatalized <ʲ>
-	  \item[→] velarized <ˠ>
-	  \item[→] pharyngealized <ˤ>
-	  \item[→] glottalized <ˀ>
-	  \item[→] aspirated <ʰ> \textbar{} ejective <ʼ>
-	  \item[→] long <ː> \textbar{} half-long <ˑ>
-	\end{itemize}
-\end{itemize}
+some ad-hoc decisions had to be taken.
 
@@
 finally the syllabic or non-syllabic marker (for vowels, advanced and retracted
-tongue root markers gets applied between the place and laryngeal setting). So:
+tongue root markers gets applied between the place and laryngeal setting). So
+the order that is proposed is the following (where <\textbar{}> indicates
+\textit{or} and <→> indicates \textit{precedes}):
 
@@
 
-Finally, if a character sequence contains more than one diacritic above the base
-character, we propose the following order, with Spacing Modifiers and tone
-letters being added at the end:
+Further, if a character sequence contains more than one diacritic above the base
+character, we propose the following order. Note that this ordering is only
+important when the extra short superscript is used. All other combinations will 
+in practice never lead to problems.
 
@@
 %	  \item[→] (tone accents, e.g. <è>)
-	  \item[→] unreleased <e̚>
-	  \item[→] Spacing Modifier Letters, see above
+	  \item[→] unreleased <p̚>
+%	  \item[→] Spacing Modifier Letters, see above	  
+ \end{itemize} \end{itemize}
+
+Finally, when a character sequence contains more than one character in Spacing
+Modifier Letters, these will be placed after all combining diacritic marks in the
+following order:
+
+\begin{itemize}
+	\item[] \textsc{Spacing Modifier Letters ordering:}
+	\begin{itemize}
+	  \item[→] rhotic hook <{\large \ ˞} >
+	  \item[→] lateral release <ˡ> \textbar{} nasal release <ⁿ>
+	  \item[→] labialized <ʷ>
+	  \item[→] palatalized <ʲ>
+	  \item[→] velarized <ˠ>
+	  \item[→] pharyngealized <ˤ>
+	  \item[→] glottalized <ˀ>
+	  \item[→] aspirated <ʰ> \textbar{} ejective <ʼ>
+	  \item[→] long <ː> \textbar{} half-long <ˑ>
 	  \item[→] tone letters <˥ ˦ ˧ ˨ ˩>
@@
 
-Summary
+Summarizing the pitfalls discussed in this chapter, we propose to define a 
+``strict'' IPA encoding to be used when interoperability of phonetic resources 
+is intended. For regular research papers, when only the visual outline is 
+important,
+
+\begin{table}[htdp]
+
+ \topcaption{IPA letters with Unicode encodings}
+ \label{tab:ipa_letters}
+ \tablehead{
+   \toprule
+   & Code & Unicode name & IPA name \\
+  \midrule
+  }
+ \tabletail{\endline}
+
+\begin{center}
+\begin{xtabular}{ll>{\small}p{4cm}>{\small}l}
+a & \uni{0061} & \textsc{latin small letter a} & open front unrounded \\ 
+æ & \uni{00E6} & \textsc{latin small letter ae} & raised open front unrounded \\ 
+ɐ & \uni{0250} & \textsc{latin small letter turned a} & lowered schwa \\ 
+ɑ & \uni{0251} & \textsc{latin small letter alpha} & open back unrounded \\ 
+ɒ & \uni{0252} & \textsc{latin small letter turned alpha} & open back rounded \\ 
+b & \uni{0062} & \textsc{latin small letter b} & voiced bilabial plosive \\ 
+ʙ & \uni{0299} & \textsc{latin letter small capital b} & voiced bilabial trill \\ 
+ɓ & \uni{0253} & \textsc{latin small letter b with hook} & voiced bilabial implosive \\ 
+c & \uni{0063} & \textsc{latin small letter c} & voiceless palatal plosive \\ 
+ç & \uni{00E7} & \textsc{latin small letter c with cedilla} & voiceless palatal fricative \\ 
+ɕ & \uni{0255} & \textsc{latin small letter c with curl} & voiceless alveolo-palatal fricative \\ 
+d & \uni{0064} & \textsc{latin small letter d} & voiced alveolar plosive \\ 
+ð & \uni{00F0} & \textsc{latin small letter eth} & voiced dental fricative \\ 
+ɖ & \uni{0256} & \textsc{latin small letter d with tail} & voiced retroflex plosive \\ 
+ɗ & \uni{0257} & \textsc{latin small letter d with hook} & voiced dental/alveolar implosive \\ 
+\charis{ᶑ} & \uni{1D91} & \textsc{latin small letter d with hook and tail} & voiced retroflex implosive \\ 
+e & \uni{0065} & \textsc{latin small letter e} & close-mid front unrounded \\ 
+ə & \uni{0259} & \textsc{latin small letter schwa} & mid-central schwa \\ 
+ɛ & \uni{025B} & \textsc{latin small letter open e} & open-mid front unrounded \\ 
+ɘ & \uni{0258} & \textsc{latin small letter reversed e} & close-mid central unrounded \\ 
+ɜ & \uni{025C} & \textsc{latin small letter reversed open e} & open-mid central unrounded \\ 
+ɞ & \uni{025E} & \textsc{latin small letter closed reversed open e} & open-mid central rounded \\ 
+ɤ & \uni{0264} & \textsc{latin small letter rams horn} & close-mid back unrounded \\ 
+f & \uni{0066} & \textsc{latin small letter f} & voiceless labiodental fricative \\ 
+ɡ & \uni{0261} & \textsc{latin small letter script g} & voiced velar plosive \\ 
+ɢ & \uni{0262} & \textsc{latin letter small capital g} & voiced uvular plosive \\ 
+ɠ & \uni{0260} & \textsc{latin small letter g with hook} & voiced velar implosive \\ 
+ʛ & \uni{029B} & \textsc{latin letter small capital g with hook} & voiced uvular implosive \\ 
+ɣ & \uni{0263} & \textsc{latin small letter gamma} & voiced velar fricative \\ 
+h & \uni{0068} & \textsc{latin small letter h} & voiceless glottal fricative \\ 
+ħ & \uni{0127} & \textsc{latin small letter h with stroke} & voiceless pharyngeal fricative \\ 
+ʜ & \uni{029C} & \textsc{latin letter small capital h} & voiceless epiglottal fricative \\ 
+ɦ & \uni{0266} & \textsc{latin small letter h with hook} & voiced glottal fricative \\ 
+ɧ & \uni{0267} & \textsc{latin small letter heng with hook} & voiceless postalveolar+velar fricative \\ 
+i & \uni{0069} & \textsc{latin small letter i} & close front unrounded \\ 
+ɪ & \uni{026A} & \textsc{latin letter small capital i} & lax close front unrounded \\ 
+ɨ & \uni{0268} & \textsc{latin small letter i with stroke} & close central unrounded \\ 
+j & \uni{006A} & \textsc{latin small letter j} & voiced palatal approximant \\ 
+ʝ & \uni{029D} & \textsc{latin small letter j with crossed tail} & voiced palatal fricative \\ 
+ɟ & \uni{025F} & \textsc{latin small letter dotless j with stroke} & voiced palatal plosive \\ 
+ʄ & \uni{0284} & \textsc{latin small letter dotless j with stroke and hook} & voiced palatal implosive \\ 
+k & \uni{006B} & \textsc{latin small letter k} & voiceless velar plosive \\ 
+l & \uni{006C} & \textsc{latin small letter l} & voiced alveolar lateral approximant \\ 
+ʟ & \uni{029F} & \textsc{latin letter small capital l} & voiced velar lateral approximant \\ 
+ɫ & \uni{026B} & \textsc{latin small letter l with middle tilde} & voiced pharyngealized alveolar lateral approximant \\ 
+ɬ & \uni{026C} & \textsc{latin small letter l with belt} & voiceless alveolar lateral fricative \\ 
+ɭ & \uni{026D} & \textsc{latin small letter l with retroflex hook} & voiced retroflex lateral approximant \\ 
+ɮ & \uni{026E} & \textsc{latin small letter lezh} & voiced alveolar lateral fricative \\ 
+ʎ & \uni{028E} & \textsc{latin small letter turned y} & voiced palatal lateral approximant \\ 
+m & \uni{006D} & \textsc{latin small letter m} & voiced bilabial nasal \\ 
+ɱ & \uni{0271} & \textsc{latin small letter m with hook} & voiced labiodental nasal \\ 
+n & \uni{006E} & \textsc{latin small letter n} & voiced alveolar nasal \\ 
+ɴ & \uni{0274} & \textsc{latin letter small capital n} & voiced uvular nasal \\ 
+ɲ & \uni{0272} & \textsc{latin small letter n with left hook} & voiced palatal nasal \\ 
+ɳ & \uni{0273} & \textsc{latin small letter n with retroflex hook} & voiced retroflex nasal \\ 
+ŋ & \uni{014B} & \textsc{latin small letter eng} & voiced velar nasal \\ 
+o & \uni{006F} & \textsc{latin small letter o} & close-mid back rounded \\ 
+ø & \uni{00F8} & \textsc{latin small letter o with stroke} & close-mid front rounded \\ 
+œ & \uni{0153} & \textsc{latin small ligature oe} & open-mid front rounded \\ 
+ɶ & \uni{0276} & \textsc{latin letter small capital oe} & open front rounded \\ 
+ɔ & \uni{0254} & \textsc{latin small letter open o} & open-mid back rounded \\ 
+ɵ & \uni{0275} & \textsc{latin small letter barred o} & close-mid central rounded \\ 
+p & \uni{0070} & \textsc{latin small letter p} & voiceless bilabial plosive \\ 
+ɸ & \uni{0278} & \textsc{latin small letter phi} & voiceless bilabial fricative \\ 
+q & \uni{0071} & \textsc{latin small letter q} & voiceless uvular plosive \\ 
+r & \uni{0072} & \textsc{latin small letter r} & voiced alveolar trill \\ 
+ʀ & \uni{0280} & \textsc{latin letter small capital r} & voiced uvular trill \\ 
+ɹ & \uni{0279} & \textsc{latin small letter turned r} & voiced alveolar approximant \\ 
+ɺ & \uni{027A} & \textsc{latin small letter turned r with long leg} & voiced alveolar lateral flap \\ 
+ɻ & \uni{027B} & \textsc{latin small letter turned r with hook} & voiced retroflex approximant \\ 
+ɽ & \uni{027D} & \textsc{latin small letter r with tail} & voiced retroflex tap \\ 
+ɾ & \uni{027E} & \textsc{latin small letter r with fishhook} & voiced alveolar tap \\ 
+ʁ & \uni{0281} & \textsc{latin letter small capital inverted r} & voiced uvular fricative \\ 
+s & \uni{0073} & \textsc{latin small letter s} & voiceless alveolar fricative \\ 
+ʂ & \uni{0282} & \textsc{latin small letter s with hook} & voiceless retroflex fricative \\ 
+ʃ & \uni{0283} & \textsc{latin small letter esh} & voiceless postalveolar fricative \\ 
+t & \uni{0074} & \textsc{latin small letter t} & voiceless alveolar plosive \\ 
+ʈ & \uni{0288} & \textsc{latin small letter t with retroflex hook} & voiceless retroflex plosive \\ 
+u & \uni{0075} & \textsc{latin small letter u} & close back rounded \\ 
+ʉ & \uni{0289} & \textsc{latin small letter u bar} & close central rounded \\ 
+ɥ & \uni{0265} & \textsc{latin small letter turned h} & voiced labial-palatal approximant \\ 
+ɯ & \uni{026F} & \textsc{latin small letter turned m} & close back unrounded \\ 
+ɰ & \uni{0270} & \textsc{latin small letter turned m with long leg} & voiced velar approximant \\ 
+ʊ & \uni{028A} & \textsc{latin small letter upsilon} & lax close back rounded \\ 
+v & \uni{0076} & \textsc{latin small letter v} & voiced labiodental fricative \\ 
+ʋ & \uni{028B} & \textsc{latin small letter v with hook} & voiced labiodental approximant \\ 
+ⱱ & \uni{2C71} & \textsc{latin small letter v with right hook} & voiced labiodental tap \\ 
+ʌ & \uni{028C} & \textsc{latin small letter turned v} & open-mid back unrounded \\ 
+w & \uni{0077} & \textsc{latin small letter w} & voiced labial-velar approximant \\ 
+ʍ & \uni{028D} & \textsc{latin small letter turned w} & voiceless labial-velar fricative \\ 
+x & \uni{0078} & \textsc{latin small letter x} & voiceless velar fricative \\ 
+y & \uni{0079} & \textsc{latin small letter y} & close front rounded \\ 
+ʏ & \uni{028F} & \textsc{latin letter small capital y} & lax close front rounded \\ 
+z & \uni{007A} & \textsc{latin small letter z} & voiced alveolar fricative \\ 
+ʐ & \uni{0290} & \textsc{latin small letter z with retroflex hook} & voiced retroflex fricative \\ 
+ʑ & \uni{0291} & \textsc{latin small letter z with curl} & voiced alveolo-palatal fricative \\ 
+ʒ & \uni{0292} & \textsc{latin small letter ezh} & voiced postalveolar fricative \\ 
+ʔ & \uni{0294} & \textsc{latin letter glottal stop} & voiceless glottal plosive \\ 
+ʕ & \uni{0295} & \textsc{latin letter pharyngeal voiced fricative} & voiced pharyngeal fricative \\ 
+ʡ & \uni{02A1} & \textsc{latin letter glottal stop with stroke} & epiglottal plosive \\ 
+ʢ & \uni{02A2} & \textsc{latin letter reversed glottal stop with stroke} & voiced epiglottal fricative \\ 
+ǀ & \uni{01C0} & \textsc{latin letter dental click} & voiceless dental click \\ 
+ǁ & \uni{01C1} & \textsc{latin letter lateral click} & voiceless alveolar lateral click \\ 
+ǂ & \uni{01C2} & \textsc{latin letter alveolar click} & voiceless palatoalveolar click \\ 
+ǃ & \uni{01C3} & \textsc{latin letter retroflex click} & voiceless (post)alveolar click \\ 
+ʘ & \uni{0298} & \textsc{latin letter bilabial click} & voiceless bilabial click \\ 
+β & \uni{03B2} & \textsc{greek small letter beta} & voiced bilabial fricative \\ 
+θ & \uni{03B8} & \textsc{greek small letter theta} & voiceless dental fricative \\ 
+χ & \uni{03C7} & \textsc{greek small letter chi} & voiceless uvular fricative \\
+\end{xtabular}
+\end{center}
+\end{table}

### 2015-06-18 chapters/ipa_meets_unicode.tex
further testing large table
adds nasalization in diacritic ordering

@@
 conventions. The conventions are influenced by the linguistic literature, though
-some ad-hoc decisions had to be taken.
+some ad-hoc decisions had to be taken given the vast variability of phonological segments 
+described by linguists.
 
@@
 	\begin{itemize}
+	  \item[→] nasalized <ẽ>
 	  \item[→] centralized <ë> \textbar{} mid-centralized <e̽>
@@
 
-\begin{table}[htdp]
+%\begin{table}[htdp]
 
@@
  \label{tab:ipa_letters}
- \tablehead{
-   \toprule
-   & Code & Unicode name & IPA name \\
-  \midrule
-  }
- \tabletail{\endline}
+ \tablehead{\toprule & Code & Unicode name & IPA name \\ \midrule }
+ \tabletail{\bottomrule}
 
 \begin{center}
-\begin{xtabular}{ll>{\small}p{4cm}>{\small}l}
+\begin{xtabular}{ l l L{4.5cm} L{4.5cm} }
 a & \uni{0061} & \textsc{latin small letter a} & open front unrounded \\ 
@@
 ʋ & \uni{028B} & \textsc{latin small letter v with hook} & voiced labiodental approximant \\ 
-ⱱ & \uni{2C71} & \textsc{latin small letter v with right hook} & voiced labiodental tap \\ 
+\charis{ⱱ} & \uni{2C71} & \textsc{latin small letter v with right hook} & voiced labiodental tap \\ 
 ʌ & \uni{028C} & \textsc{latin small letter turned v} & open-mid back unrounded \\ 
@@
 \end{center}
-\end{table}
+%\end{table}

### 2015-06-18 chapters/ipa_meets_unicode.tex
added text on tie-bar; updated gitignore

@@
 Finally, the tie bar to indicate affricates, doubly articulated consonants or
+diphthongs can be placed above or below the base characters, e.g.
+<\charis{t͡s}> or <\charis{t͜s}>. IPA allows both the \textsc{combining double inverted
+breve} at \uni{0361} and the \textsc{combining double breve below} at \uni{035C}. The  
+choice between the two symbols is purely for legible rendering; there is no difference 
+in semantics between the two symbols. In fact, rendering is such a problematic issue 
+for tie bars in general, that many linguists simply do not use them. Further, tie bars 
+are a special type of character in the sense that they bind two graphemes (singleton base characters 
+or tailored grapheme clusters) together. In Section \ref{pitfall-different-notions-of-diacritics}, 
+we highlight the division of linguistic diacritics 
+into two categories in the Unicode Standard: Spacing Modifier Letters and Combining 
+Diacritical Marks.\footnote{The former contain characters like aspiration < ʰ > which is user 
+specified (before or after the base character) and the latter which combines 
+with the base character.} The tie-bar is somewhere in between these two character 
+classes, i.e. when the combining breve (above or below) is told to combine within 
+a sequence of Unicode characters, it will bind to the character immediately 
+preceding it -- the so-called base character -- regardless if that base 
+character is a Spacing Modifier Letter \textit{or} Combining Diacritical Mark. \footnote{In some cases 
+conversion of a string sequence into Unicode Normalization Form C alleviates 
+the problem by composing sequences of characters like <e> + < ́ > into single character < é > 
+(see however diacritics like the rhotic hook in ).} In principle, 
+it is possible to combine more than two base characters by repeating the tie bar, 
+like in <a͡o͡u>. If really necessary, we consider this possible, even though the 
+rendering will never look good.
+
+The tie-bar is a convenient diacritic for visually tokenizing input strings into chunks of 
+phonetically salient groups and both the above and below tie-bar were introduced in electronic 
+text to deal with difficult to render character sequences. The tie-bar is unnecessary for 
+computational tokenization given the design of orthography profiles, described in 
+Section \ref{usecase-tokenization}. However, to adhere to strict IPA encoding users 
+may use either tie bar, or none whatsoever.
+
+\begin{comment} 
+Finally, the tie bar to indicate affricates, doubly articulated consonants or
 diphthongs can be placed either below or on top of the base characters, e.g.
@@
 good.
+
+For strict IPA encoding we propose to standardize on the tie bar
+above the base characters, using Unicode \textsc{combining double inverted
+breve} at \uni{0361}. Note that this Unicode character is placed between the two 
+base characters to be combined. In principle, it is possible to combine more 
+than two base characters by repeating the tie bar, like in <a͡o͡u>. If really
+necessary, we consider this possible, even though the rendering will never look 
+good.\end{comment}
 

### 2015-06-18 chapters/use_cases.tex
added text on tie-bar; updated gitignore

@@
 orthography profiles. These include:
+
 \begin{itemize}
@@
 	\item cognate identification for detecting the similarity between words from different languages 
+	\item ...
 \end{itemize}
@@
 
-- Unicode character tokenization
-- grapheme tokenization
-- tailored grapheme tokenization
+\begin{itemize}
+	\item Unicode character tokenization
+	\item grapheme tokenization
+	\item tailored grapheme tokenization
+\end{itemize}
 
@@
 Letters, e.g. <˦> for a phonemic high tone, instead of accent diacritics,
-alleviating potential conflicts. Building on these works,
+alleviating potential conflicts.
+
+\citet{Moran2012;MoranMcCloy2014} represent clicks as a combination of anterior and 
+posterior articulations, where the set < k q ɡ ɢ ŋ ɴ > denote place of articulation 
+and phonation type (voiced, voiceless, nasal). The anterior articulation click characters 
+are < ʘ ǀ ǁ ǃ ǂ > with the addition of non-standard IPA < ‼ > \textsc{DOUBLE EXCLAMATION MARK)} 
+\uni{203C} to represent the retroflex click. Note that the (post)alveolar click < ǃ > at \uni{01C3} 
+is called incorrectly \textsc{LATIN LETTER RETROFLEX CLICK} (it is also a different character than 
+the keyboard < ! > \textsc{EXCLAMATION MARK} at \uni{0021}). All sources that contain clicks in 
+PHOIBLE are standardized to contain both the anterior and posterior characters and diacritics 
+are added to the posterior glyph, e.g.\ < ŋ̊ǃ >.\footnote{http://phoible.github.io/conventions/}
+
+Building on these works...

### 2015-06-18 chapters/ipa_meets_unicode.tex
update discussion tie bar

--- 2015-06-18 chapters/ipa_meets_unicode.tex
@@
   \item The placement of diacritics
-  \item The tie bar to indicate a close bond between sounds
 \end{itemize}
@@
 time, whereas the latter must be cumbersomely inserted as a special symbol in
-most software. The International Phonetic
-Association has taken the stance that both the keyboard \textsc{latin small
-letter g} and the \textsc{latin small letter script g} are valid input
-characters for the voiced velar plosive. Unfortunately, this decision further
-introduces ambiguity for linguists trying to adhere to a strict Unicode Standard
-IPA encoding. For strict IPA encoding we propose to standardize on the more 
-idiosyncratic \textsc{latin small letter script g} at \uni{0261}.
+most software. The International Phonetic Association has taken the stance that
+both the keyboard \textsc{latin small letter g} and the \textsc{latin small
+letter script g} are valid input characters for the voiced velar plosive.
+Unfortunately, this decision further introduces ambiguity for linguists trying
+to adhere to a strict Unicode Standard IPA encoding. For strict IPA encoding we
+propose to standardize on the more idiosyncratic \textsc{latin small letter
+script g} at \uni{0261}.
 
@@
 the ring below the base character, like in <ŋ̥>, so for strict IPA encoding we 
-propose to standardize on the ring below.
+propose to standardize on the ring below. As a principle, for strict IPA 
+encoding only one options should be allowed.
 
-Finally, the tie bar to indicate affricates, doubly articulated consonants or
-diphthongs can be placed either below or on top of the base characters, e.g.
-<\charis{t͡s}> or <\charis{t͜s}>. For strict IPA encoding we propose to standardize on the tie bar
-above the base characters, using Unicode \textsc{combining double inverted
-breve} at \uni{0361}. Note that this Unicode character is placed between the two 
-base characters to be combined. In principle, it is possible to combine more 
-than two base characters by repeating the tie bar, like in <a͡o͡u>. If really
-necessary, we consider this possible, even though the rendering will never look 
-good.
+% ==========================
+\section{Pitfall: Tie bar}
+\label{pitfall-tie-bar}
+% ==========================
+
+The IPA includes a tie bar to indicate affricates, doubly articulated consonants
+or diphthongs. This character can be placed above or below the base characters,
+e.g. <\charis{t͡s}> or <\charis{t͜s}>. IPA allows both options. The choice
+between the two symbols is purely for legible rendering; there is no difference
+in semantics between the two symbols. The tie-bar is a convenient diacritic for
+visually tokenizing input strings into chunks of phonetically salient groups.
+However, rendering is such a problematic issue for tie bars in general that many
+linguists simply do not use them. Just looking at a few different fonts already
+indicates that actually no font designer really gets the placement right in
+combination with superscripts and subscripts. If really necessary, we propose to
+standardize on the tie bar above the base characters, using Unicode
+\textsc{combining double inverted breve} at \uni{0361}.
+
+\begin{itemize}
+  \item[] {\fontspec{Times New Roman}Times new Roman: <t̥ʰ͡s> <t̥ʰ͜s>}
+  \item[] {\small \fontspec{CharisSIL}CharisSIL: <t̥ʰ͡s> <t̥ʰ͜s>}
+  \item[] {\footnotesize \fontspec{Monaco}Monaco: <t̥ʰ͡s> <t̥ʰ͜s>}
+  \item[] {\fontspec{DoulosSIL}DoulosSIL: <t̥ʰ͡s> <t̥ʰ͜s>}
+  \item[] Linux Libertine: <t̥ʰ͡s> <t̥ʰ͜s>
+\end{itemize}
+
+Tie bars are a special type of character in the sense that they do not belong to
+a segment, but bind two graphemes together. This actually turns out to be rather
+different from the Unicode conceptions, see
+Section~\ref{pitfall-different-notions-of-diacritics}. According to the spirit
+of the IPA, it is possible to combine more than two base characters into one tie
+bar, but this is not possible with Unicode (i.e. there is no possibility to draw
+a tie bar over three of four characters). It is possible to indicate such larger
+groups by repeating the tie bar, like in <a͡o͡u>. If really necessary, we
+consider this possible, even though the rendering will never look good.
+
+Most importantly though, the tie-bar is unnecessary for computational
+tokenization given the design of orthography profiles, as described in
+Section~\ref{usecase-tokenization}. The principle idea of such tokenization is 
+that groups of characters that belong together are defined in an orthography 
+profile, making it superfluous to indicate them explicitly by tie bars. For this 
+reason we actually would like to suggest to remove the tie bar from any 
+``strict'' IPA encoding.
 
@@
 impractical, and probably not an enforceable solution to us.
+
+% problem with tie bar!!!
+
+In Section~\ref{pitfall-different-notions-of-diacritics}, we highlight
+the division of linguistic diacritics into two categories in the Unicode
+Standard: Spacing Modifier Letters and Combining Diacritical Marks.\footnote{The
+former contain characters like aspiration < ʰ > which is user specified (before
+or after the base character) and the latter which combines with the base
+character.} The tie-bar is somewhere in between these two character classes,
+i.e.~when the combining breve (above or below) is told to combine within a
+sequence of Unicode characters, it will bind to the character immediately
+preceding it -- the so-called base character -- regardless if that base
+character is a Spacing Modifier Letter \textit{or} Combining Diacritical Mark.
+\footnote{In some cases conversion of a string sequence into Unicode
+Normalization Form C alleviates the problem by composing sequences of characters
+like <e> + < ́ > into single character < é > (see however diacritics like the
+rhotic hook in ).}
 

### 2015-06-18 chapters/ipa_meets_unicode.tex
update homoglyphs in Unicode

@@
    \item The uses of the apostrophe has led to long discussions on the Unicode
-     Standard email list. An English keyboard inputs <{\fontspec{Monaco}'}> \textsc{apostrophe}
-     at \uni{0027}, although the preferred Unicode apostrophe is the <\ ' >
-     \textsc{right single quotation mark} at \uni{2019}.\footnote{Note that many word 
-     processors (like Microsoft Word) by default will replace straight quotes by 
-     curly quotes, depending on the whitespace around it.} However, the glottal stop/glottalization/ejective marker is yet 
-     another completely different character <{\large ʼ}>, the 
-     \textsc{modifier letter apostrophe} at \uni{02BC}, which unfortunately looks 
-     mostly extremely similar to \uni{2019}. 
-  \item Another problem is the <ˁ> \textsc{modifier letter reversed
-     glottal stop} at \uni{02C1} vs.\@ the <ˤ> \textsc{modifier
-     letter small reversed glottal stop} at \uni{02E4}. Both 
-     appear in various resources representing phonetic data online. This is
-     thus a clear example for which the Unicode Standard does not solve the
-     linguistic standardization problem.
-  \item even linguists are unlikely to distinguish between the <ə>
+        Standard email list. An English keyboard inputs <{\fontspec{Monaco}'}>
+        \textsc{apostrophe} at \uni{0027}, although the preferred Unicode
+        apostrophe is the <\ ' > \textsc{right single quotation mark} at
+        \uni{2019}.\footnote{Note that many word processors (like Microsoft
+        Word) by default will replace straight quotes by curly quotes,
+        depending on the whitespace around it.} However, the glottal
+        stop/glottalization/ejective marker is yet another completely different
+        character <{\large ʼ}>, the \textsc{modifier letter apostrophe} at
+        \uni{02BC}, which unfortunately looks mostly extremely similar to
+        \uni{2019}. 
+  \item Another problem is the <ˁ> \textsc{modifier letter reversed glottal
+        stop} at \uni{02C1} vs.\@ the <ˤ> \textsc{modifier letter small reversed
+        glottal stop} at \uni{02E4}. Both appear in various resources
+        representing phonetic data online. This is thus a clear example for
+        which the Unicode Standard does not solve the linguistic standardization
+        problem.
+  \item Even linguists are unlikely to distinguish between the <ə>
         \textsc{latin small letter schwa} at code point \uni{0259} and <ǝ>
@@
   \item The alveolar click <ǃ> at \uni{01C3} is of course often simply
-     typed as <!> \textsc{exclamation mark} at \uni{0021}.\footnote{In the Unicode
-     Standard the <ǃ> at \unif{01C3} is labeled \textsc{latin letter retroflex
-     click}, but in IPA that glyph is used for an alveolar or postalveolar
-     click (not retroflex). This naming is probably best seen as an 
-     error in the Unicode Standard.}
+        typed as <!> \textsc{exclamation mark} at \uni{0021}.\footnote{In the
+        Unicode Standard the <ǃ> at \unif{01C3} is labeled \textsc{latin letter
+        retroflex click}, but in IPA that glyph is used for an alveolar or
+        postalveolar click (not retroflex). This naming is probably best seen as
+        an error in the Unicode Standard.}
+  \item The dental click <ǀ>, in Unicode known as \textsc{latin letter dental
+        click} at \uni{01C0}, is of course often simply typed as <|>
+        \textsc{vertical line} at \uni{007C}.
+  \item For the marking of length there is a special Unicode characters, namely
+        <ː> \textsc{modifier letter triangular colon} at \uni{02D0}. However,
+        typing <:> \textsc{colon} at \uni{003A} is of course much easier.
 
@@
   \item[] {\fontspec{Times New Roman}Times new Roman: <t̥ʰ͡s> <t̥ʰ͜s>}
-  \item[] {\small \fontspec{CharisSIL}CharisSIL: <t̥ʰ͡s> <t̥ʰ͜s>}
+  \item[] {\small \fontspec{CharisSIL}CharisSIL:\@ <t̥ʰ͡s> <t̥ʰ͜s>}
   \item[] {\footnotesize \fontspec{Monaco}Monaco: <t̥ʰ͡s> <t̥ʰ͜s>}
-  \item[] {\fontspec{DoulosSIL}DoulosSIL: <t̥ʰ͡s> <t̥ʰ͜s>}
+  \item[] {\fontspec{DoulosSIL}DoulosSIL:\@ <t̥ʰ͡s> <t̥ʰ͜s>}
   \item[] Linux Libertine: <t̥ʰ͡s> <t̥ʰ͜s>
@@
 of the IPA, it is possible to combine more than two base characters into one tie
-bar, but this is not possible with Unicode (i.e. there is no possibility to draw
+bar, but this is not possible with Unicode (i.e.~there is no possibility to draw
 a tie bar over three of four characters). It is possible to indicate such larger

### 2015-06-18 chapters/ipa_meets_unicode.tex
update different notions of diacritics

@@
         stop/glottalization/ejective marker is yet another completely different
-        character <{\large ʼ}>, the \textsc{modifier letter apostrophe} at
+        character <{\large ʼ}>, namely the \textsc{modifier letter apostrophe} at
         \uni{02BC}, which unfortunately looks mostly extremely similar to
@@
 % ==========================
+
+% this has to be revised: there are two issues: homoglyphs and alternative 
+encoding options. that is not necessarily the same. Also include minor group break 
+vs. dental click problem. and stress vs. ejective?
 
@@
 
-% problem with tie bar!!!
-
-In Section~\ref{pitfall-different-notions-of-diacritics}, we highlight
-the division of linguistic diacritics into two categories in the Unicode
-Standard: Spacing Modifier Letters and Combining Diacritical Marks.\footnote{The
-former contain characters like aspiration < ʰ > which is user specified (before
-or after the base character) and the latter which combines with the base
-character.} The tie-bar is somewhere in between these two character classes,
-i.e.~when the combining breve (above or below) is told to combine within a
-sequence of Unicode characters, it will bind to the character immediately
-preceding it -- the so-called base character -- regardless if that base
-character is a Spacing Modifier Letter \textit{or} Combining Diacritical Mark.
-\footnote{In some cases conversion of a string sequence into Unicode
-Normalization Form C alleviates the problem by composing sequences of characters
-like <e> + < ́ > into single character < é > (see however diacritics like the
-rhotic hook in ).}
+Further, the IPA tie bar (see Section~\ref{pitfall-tie-bar}) is somewhere in
+between these two character classes. The Unicode encoding of this character is
+uses a Combining Diacritical Mark, namely either \textsc{combining double
+inverted breve} at \uni{0361} or \textsc{combining double breve below} at
+\uni{035C}. Such a combining mark is by definition tied to the character in
+front, but not the character following it. The Unicode treatment of this
+character thus only partly corresponds to the IPA conception, which ideally
+would have the tie bar linked both to the character in front and to the 
+character following.
 

### 2015-06-18 chapters/ipa_meets_unicode.tex
different organisation of IPA homoglyph section

@@
 
-% this has to be revised: there are two issues: homoglyphs and alternative 
-encoding options. that is not necessarily the same. Also include minor group break 
-vs. dental click problem. and stress vs. ejective?
+Reversely, there are a few cases in which the IPA distinguishes different
+concepts, but the visual marking proposed by the IPA look very much alike. Such
+cases are thus homoglyphs in the IPA itself, which of course need different
+encodings.
+
+\begin{itemize}
+
+   \item The dental click <ǀ> and the indication of a minor group break <|>
+           (proposed to be used for separating feet) look almost the same in
+           most fonts. For a proper encoding, the \textsc{latin letter dental
+           click} at \uni{01C0} and the \textsc{vertical line }at \uni{007C} 
+           should be strictly distinguished.
+   \item Similarly, the alveolar lateral click <ǁ>	should be encoded with a
+           \textsc{latin letter lateral click} at \uni{01C1}, different from <‖>, 
+           which according to the IPA is a major group break (by intonation), to be
+           encoded by \textsc{double vertical line} at \uni{2016}.
+   \item The marking of primary stress < ˈ > looks like an apostrophe, and
+           is often typed with the same symbol as the ejective < ʼ >. For a
+           proper encoding, these two symbols should be typed as 
+           \textsc{modifier letter vertical line} at \uni{02C8} and
+           \textsc{modifier letter apostrophe} at \uni{02BC}, respectively.
+
+\end{itemize}
+
+% ==========================
+\section{Pitfall: Multiple options in IPA}
+\label{pitfall-multiple-options-ipa}     
+% ==========================
 

### 2015-06-18 chapters/ipa_meets_unicode.tex
update treatment of middle tilde

@@
 blocks. With its diverse collection of symbols from various scripts and
-diacritics, the IPA is spread across 13 blocks in the Unicode
+diacritics, the IPA is spread across 12 blocks in the Unicode
 Standard:\footnote{This number of blocks depends on whether only IPA-sanctioned
@@
 \begin{itemize}
-	\item Basic Latin (27 characters): \newline < a~b~c~d~e~f~h~i~j~k~l~m~n~o~p~q~r~s~t~u~v~w~x~y~z~.~| >
+	\item Basic Latin (27 characters): \newline < a b c d e f h i j k l m n o p q r s t u v w x y z~.~| >
 	\item Latin-1 Supplement (4 characters): \newline < æ ç ð ø >
@@
 	\item Latin Extended-B (4 characters): \newline < ǀ ǁ ǂ ǃ >
-	\item IPA Extensions (68 characters): \newline < ɐ ɑ ɒ ɓ ɔ ɕ ɖ ɗ ɘ ə ɛ ɜ ɞ ɟ ɠ ɡ ɢ ɣ ɤ ɥ ɦ ɧ ɨ ɪ ɫ ɬ ɭ ɮ ɯ ɰ ɱ ɲ ɳ ɴ ɵ ɶ ɸ ɹ ɺ ɻ ɽ ɾ ʀ ʁ ʂ ʃ ʄ ʈ ʉ ʊ ʋ ʌ ʍ ʎ ʏ ʐ ʑ ʒ ʔ ʕ ʘ ʙ ʛ ʜ ʝ ʟ ʡ ʢ > 
+	\item IPA Extensions (67 characters): \newline < ɐ ɑ ɒ ɓ ɔ ɕ ɖ ɗ ɘ ə ɛ ɜ ɞ ɟ ɠ ɡ ɢ ɣ ɤ ɥ ɦ ɧ ɨ ɪ ɬ ɭ ɮ ɯ ɰ ɱ ɲ ɳ ɴ ɵ ɶ ɸ ɹ ɺ ɻ ɽ ɾ ʀ ʁ ʂ ʃ ʄ ʈ ʉ ʊ ʋ ʌ ʍ ʎ ʏ ʐ ʑ ʒ ʔ ʕ ʘ ʙ ʛ ʜ ʝ ʟ ʡ ʢ > 
 	\item Spacing Modifier Letters (17 characters): \newline < ʰ ʲ ʷ ʼ ˈ ˌ ː ˑ ˞ ˠ ˡ ˤ ˥ ˦ ˧ ˨ ˩ >
@@
 %	\item Phonetic Extensions (2 characters): \charis{ᴅ ᴴ} 
-	\item Phonetic Extensions Supplement (1 characters): \newline < \charis{ᶑ} >
+%	\item Phonetic Extensions Supplement (1 characters): \newline < \charis{ᶑ} >
     \item General Punctuation (2 characters): \newline < ‖ \charis{‿} >
@@
 at \uni{0061}, basically treating them as homoglyphs, although they are
-different phonetic symbols. But even among linguists this distinction leads to problems. For example, as
-pointed out by \citet{Mielke2009}, there is a problem stemming from the fact
-that about 75\% of languages are reported to have a five-vowel system
-\citep{Maddieson1984}. Historically, linguistic descriptions tend not to include
-precise audio recording and measurements of formants, so this may lead one to
-ask if the many <a> characters that are used in phonological description
-reflects a transcriptional bias. The common use of <a> in transcriptions could
-be in part due to the ease of typing the letter on an English keyboard (or for
-older descriptions, the typewriter). We found it to be exceedingly rare that a
-linguist uses <ɑ> for a low back unrounded vowel.\footnote{One example is
-\citet[75]{Vidal2001a}, in which the author states: ``The definition of Pilagá
-/a/ as [+back] results from its behavior in certain phonological contexts. For
-instance, uvular and pharyngeal consonants only occur around /a/ and /o/. Hence,
-the characterization of /a/ and /o/ as a natural class of (i.e., [+back]
-vowels), as opposed to /i/ and /e/.''} They simply use <a> as long as there is
-no opposition to <ɑ>.
+different phonetic symbols. But even among linguists this distinction leads to
+problems. For example, as pointed out by \citet{Mielke2009}, there is a problem
+stemming from the fact that about 75\% of languages are reported to have a
+five-vowel system \citep{Maddieson1984}. Historically, linguistic descriptions
+tend not to include precise audio recording and measurements of formants, so
+this may lead one to ask if the many <a> characters that are used in
+phonological description reflects a transcriptional bias. The common use of <a>
+in transcriptions could be in part due to the ease of typing the letter on an
+English keyboard (or for older descriptions, the typewriter). We found it to be
+exceedingly rare that a linguist uses <ɑ> for a low back unrounded
+vowel.\footnote{One example is \citet[75]{Vidal2001a}, in which the author
+states: ``The definition of Pilagá /a/ as [+back] results from its behavior in
+certain phonological contexts. For instance, uvular and pharyngeal consonants
+only occur around /a/ and /o/. Hence, the characterization of /a/ and /o/ as a
+natural class of (i.e., [+back] vowels), as opposed to /i/ and /e/.''} They
+simply use <a> as long as there is no opposition to <ɑ>.
 
@@
            \textsc{modifier letter apostrophe} at \uni{02BC}, respectively. 
+   \item There are two different ``dashed''-l characters in IPA, namely the <ɫ> 	      \textsc{latin small letter l with middle tilde} at \uni{026B} and the <ɬ>	      \textsc{latin small letter l with belt} at \uni{026C}.
+These of course
+          look highly similar, although they are different sounds. As a solution, 
+          we will actually propose to not use the middle tilde at all 
+          (see Section~\ref{pitfall-multiple-options-ipa}).
 
@@
   \item The marking of <g>
+  \item The marking of velarization ad pharyngealization
   \item The placement of diacritics
@@
 
-Third, for marking marking of voiceless pronunciation of voiced segments the IPA
+Third, the IPA has different markers for velarization <ˠ> and pharyngealization
+<ˤ>. Confusingly, there is also a marker for ``velarized or pharyngealized'',
+using the <\ \ {\large ̴}> \textsc{combining tilde overlay} at \uni{0334}. The
+tilde overlay seems to be extremely rarely used. For strict IPA encoding we
+propose to not include the combining tilde overlay at all, but only use the 
+other two Spacing Modifiers.
+
+Finally, for marking marking of voiceless pronunciation of voiced segments the IPA
 uses the ring diacritic. Originally, the ring should be placed below the 

### 2015-06-18 chapters/ipa_meets_unicode.tex
update strict IPA tables

@@
 	  \item[→] advanced tongue root <e̘> \textbar{} retracted tongue root <e̙>
-	  \item[→] frictionalized <ɾ͓>
-	  \item[→] more round <ɔ̹> \textbar{} less round <ɔ̜>
-	  \item[→] creaky voice <b̰> \textbar{} breathy voice <b̤> \textbar{} voiced <s̬> \textbar{} devoiced <n̥>
+%	  \item[→] frictionalized <ɾ͓>
+	  \item[→] more rounded <ɔ̹> \textbar{} less rounded <ɔ̜>
+	  \item[→] creaky voiced <b̰> \textbar{} breathy voiced <b̤> \textbar{} voiced <s̬> \textbar{} voiceless <n̥>
 	  \item[→] syllabic <n̩> \textbar{} non-syllabic <e̯>
@@
 %	  \item[→] (tone accents, e.g. <è>)
-	  \item[→] unreleased <p̚>
+	  \item[→] no audible release <p̚>
 %	  \item[→] Spacing Modifier Letters, see above	  
@@
 
-Finally, when a character sequence contains more than one character in Spacing
+Then, when a character sequence contains more than one character in Spacing
 Modifier Letters, these will be placed after all combining diacritic marks in the
@@
 	  \item[→] pharyngealized <ˤ>
-	  \item[→] glottalized <ˀ>
+%	  \item[→] glottalized <ˀ>
 	  \item[→] aspirated <ʰ> \textbar{} ejective <ʼ>
 	  \item[→] long <ː> \textbar{} half-long <ˑ>
-	  \item[→] tone letters <˥ ˦ ˧ ˨ ˩>
+%	  \item[→] tone letters <˥ ˦ ˧ ˨ ˩>
 	\end{itemize}
 \end{itemize}
+
+Finally, Tone!
 
@@
 
-%\begin{table}[htdp]
+\input{chapters/tab:unicode_letters}
 
- \topcaption{IPA letters with Unicode encodings}
- \label{tab:ipa_letters}
- \tablehead{\toprule & Code & Unicode name & IPA name \\ \midrule }
- \tabletail{\bottomrule}
-
-\begin{center}
-\begin{xtabular}{ l l L{4.5cm} L{4.5cm} }
-a & \uni{0061} & \textsc{latin small letter a} & open front unrounded \\ 
-æ & \uni{00E6} & \textsc{latin small letter ae} & raised open front unrounded \\ 
-ɐ & \uni{0250} & \textsc{latin small letter turned a} & lowered schwa \\ 
-ɑ & \uni{0251} & \textsc{latin small letter alpha} & open back unrounded \\ 
-ɒ & \uni{0252} & \textsc{latin small letter turned alpha} & open back rounded \\ 
-b & \uni{0062} & \textsc{latin small letter b} & voiced bilabial plosive \\ 
-ʙ & \uni{0299} & \textsc{latin letter small capital b} & voiced bilabial trill \\ 
-ɓ & \uni{0253} & \textsc{latin small letter b with hook} & voiced bilabial implosive \\ 
-c & \uni{0063} & \textsc{latin small letter c} & voiceless palatal plosive \\ 
-ç & \uni{00E7} & \textsc{latin small letter c with cedilla} & voiceless palatal fricative \\ 
-ɕ & \uni{0255} & \textsc{latin small letter c with curl} & voiceless alveolo-palatal fricative \\ 
-d & \uni{0064} & \textsc{latin small letter d} & voiced alveolar plosive \\ 
-ð & \uni{00F0} & \textsc{latin small letter eth} & voiced dental fricative \\ 
-ɖ & \uni{0256} & \textsc{latin small letter d with tail} & voiced retroflex plosive \\ 
-ɗ & \uni{0257} & \textsc{latin small letter d with hook} & voiced dental/alveolar implosive \\ 
-\charis{ᶑ} & \uni{1D91} & \textsc{latin small letter d with hook and tail} & voiced retroflex implosive \\ 
-e & \uni{0065} & \textsc{latin small letter e} & close-mid front unrounded \\ 
-ə & \uni{0259} & \textsc{latin small letter schwa} & mid-central schwa \\ 
-ɛ & \uni{025B} & \textsc{latin small letter open e} & open-mid front unrounded \\ 
-ɘ & \uni{0258} & \textsc{latin small letter reversed e} & close-mid central unrounded \\ 
-ɜ & \uni{025C} & \textsc{latin small letter reversed open e} & open-mid central unrounded \\ 
-ɞ & \uni{025E} & \textsc{latin small letter closed reversed open e} & open-mid central rounded \\ 
-ɤ & \uni{0264} & \textsc{latin small letter rams horn} & close-mid back unrounded \\ 
-f & \uni{0066} & \textsc{latin small letter f} & voiceless labiodental fricative \\ 
-ɡ & \uni{0261} & \textsc{latin small letter script g} & voiced velar plosive \\ 
-ɢ & \uni{0262} & \textsc{latin letter small capital g} & voiced uvular plosive \\ 
-ɠ & \uni{0260} & \textsc{latin small letter g with hook} & voiced velar implosive \\ 
-ʛ & \uni{029B} & \textsc{latin letter small capital g with hook} & voiced uvular implosive \\ 
-ɣ & \uni{0263} & \textsc{latin small letter gamma} & voiced velar fricative \\ 
-h & \uni{0068} & \textsc{latin small letter h} & voiceless glottal fricative \\ 
-ħ & \uni{0127} & \textsc{latin small letter h with stroke} & voiceless pharyngeal fricative \\ 
-ʜ & \uni{029C} & \textsc{latin letter small capital h} & voiceless epiglottal fricative \\ 
-ɦ & \uni{0266} & \textsc{latin small letter h with hook} & voiced glottal fricative \\ 
-ɧ & \uni{0267} & \textsc{latin small letter heng with hook} & voiceless postalveolar+velar fricative \\ 
-i & \uni{0069} & \textsc{latin small letter i} & close front unrounded \\ 
-ɪ & \uni{026A} & \textsc{latin letter small capital i} & lax close front unrounded \\ 
-ɨ & \uni{0268} & \textsc{latin small letter i with stroke} & close central unrounded \\ 
-j & \uni{006A} & \textsc{latin small letter j} & voiced palatal approximant \\ 
-ʝ & \uni{029D} & \textsc{latin small letter j with crossed tail} & voiced palatal fricative \\ 
-ɟ & \uni{025F} & \textsc{latin small letter dotless j with stroke} & voiced palatal plosive \\ 
-ʄ & \uni{0284} & \textsc{latin small letter dotless j with stroke and hook} & voiced palatal implosive \\ 
-k & \uni{006B} & \textsc{latin small letter k} & voiceless velar plosive \\ 
-l & \uni{006C} & \textsc{latin small letter l} & voiced alveolar lateral approximant \\ 
-ʟ & \uni{029F} & \textsc{latin letter small capital l} & voiced velar lateral approximant \\ 
-ɫ & \uni{026B} & \textsc{latin small letter l with middle tilde} & voiced pharyngealized alveolar lateral approximant \\ 
-ɬ & \uni{026C} & \textsc{latin small letter l with belt} & voiceless alveolar lateral fricative \\ 
-ɭ & \uni{026D} & \textsc{latin small letter l with retroflex hook} & voiced retroflex lateral approximant \\ 
-ɮ & \uni{026E} & \textsc{latin small letter lezh} & voiced alveolar lateral fricative \\ 
-ʎ & \uni{028E} & \textsc{latin small letter turned y} & voiced palatal lateral approximant \\ 
-m & \uni{006D} & \textsc{latin small letter m} & voiced bilabial nasal \\ 
-ɱ & \uni{0271} & \textsc{latin small letter m with hook} & voiced labiodental nasal \\ 
-n & \uni{006E} & \textsc{latin small letter n} & voiced alveolar nasal \\ 
-ɴ & \uni{0274} & \textsc{latin letter small capital n} & voiced uvular nasal \\ 
-ɲ & \uni{0272} & \textsc{latin small letter n with left hook} & voiced palatal nasal \\ 
-ɳ & \uni{0273} & \textsc{latin small letter n with retroflex hook} & voiced retroflex nasal \\ 
-ŋ & \uni{014B} & \textsc{latin small letter eng} & voiced velar nasal \\ 
-o & \uni{006F} & \textsc{latin small letter o} & close-mid back rounded \\ 
-ø & \uni{00F8} & \textsc{latin small letter o with stroke} & close-mid front rounded \\ 
-œ & \uni{0153} & \textsc{latin small ligature oe} & open-mid front rounded \\ 
-ɶ & \uni{0276} & \textsc{latin letter small capital oe} & open front rounded \\ 
-ɔ & \uni{0254} & \textsc{latin small letter open o} & open-mid back rounded \\ 
-ɵ & \uni{0275} & \textsc{latin small letter barred o} & close-mid central rounded \\ 
-p & \uni{0070} & \textsc{latin small letter p} & voiceless bilabial plosive \\ 
-ɸ & \uni{0278} & \textsc{latin small letter phi} & voiceless bilabial fricative \\ 
-q & \uni{0071} & \textsc{latin small letter q} & voiceless uvular plosive \\ 
-r & \uni{0072} & \textsc{latin small letter r} & voiced alveolar trill \\ 
-ʀ & \uni{0280} & \textsc{latin letter small capital r} & voiced uvular trill \\ 
-ɹ & \uni{0279} & \textsc{latin small letter turned r} & voiced alveolar approximant \\ 
-ɺ & \uni{027A} & \textsc{latin small letter turned r with long leg} & voiced alveolar lateral flap \\ 
-ɻ & \uni{027B} & \textsc{latin small letter turned r with hook} & voiced retroflex approximant \\ 
-ɽ & \uni{027D} & \textsc{latin small letter r with tail} & voiced retroflex tap \\ 
-ɾ & \uni{027E} & \textsc{latin small letter r with fishhook} & voiced alveolar tap \\ 
-ʁ & \uni{0281} & \textsc{latin letter small capital inverted r} & voiced uvular fricative \\ 
-s & \uni{0073} & \textsc{latin small letter s} & voiceless alveolar fricative \\ 
-ʂ & \uni{0282} & \textsc{latin small letter s with hook} & voiceless retroflex fricative \\ 
-ʃ & \uni{0283} & \textsc{latin small letter esh} & voiceless postalveolar fricative \\ 
-t & \uni{0074} & \textsc{latin small letter t} & voiceless alveolar plosive \\ 
-ʈ & \uni{0288} & \textsc{latin small letter t with retroflex hook} & voiceless retroflex plosive \\ 
-u & \uni{0075} & \textsc{latin small letter u} & close back rounded \\ 
-ʉ & \uni{0289} & \textsc{latin small letter u bar} & close central rounded \\ 
-ɥ & \uni{0265} & \textsc{latin small letter turned h} & voiced labial-palatal approximant \\ 
-ɯ & \uni{026F} & \textsc{latin small letter turned m} & close back unrounded \\ 
-ɰ & \uni{0270} & \textsc{latin small letter turned m with long leg} & voiced velar approximant \\ 
-ʊ & \uni{028A} & \textsc{latin small letter upsilon} & lax close back rounded \\ 
-v & \uni{0076} & \textsc{latin small letter v} & voiced labiodental fricative \\ 
-ʋ & \uni{028B} & \textsc{latin small letter v with hook} & voiced labiodental approximant \\ 
-\charis{ⱱ} & \uni{2C71} & \textsc{latin small letter v with right hook} & voiced labiodental tap \\ 
-ʌ & \uni{028C} & \textsc{latin small letter turned v} & open-mid back unrounded \\ 
-w & \uni{0077} & \textsc{latin small letter w} & voiced labial-velar approximant \\ 
-ʍ & \uni{028D} & \textsc{latin small letter turned w} & voiceless labial-velar fricative \\ 
-x & \uni{0078} & \textsc{latin small letter x} & voiceless velar fricative \\ 
-y & \uni{0079} & \textsc{latin small letter y} & close front rounded \\ 
-ʏ & \uni{028F} & \textsc{latin letter small capital y} & lax close front rounded \\ 
-z & \uni{007A} & \textsc{latin small letter z} & voiced alveolar fricative \\ 
-ʐ & \uni{0290} & \textsc{latin small letter z with retroflex hook} & voiced retroflex fricative \\ 
-ʑ & \uni{0291} & \textsc{latin small letter z with curl} & voiced alveolo-palatal fricative \\ 
-ʒ & \uni{0292} & \textsc{latin small letter ezh} & voiced postalveolar fricative \\ 
-ʔ & \uni{0294} & \textsc{latin letter glottal stop} & voiceless glottal plosive \\ 
-ʕ & \uni{0295} & \textsc{latin letter pharyngeal voiced fricative} & voiced pharyngeal fricative \\ 
-ʡ & \uni{02A1} & \textsc{latin letter glottal stop with stroke} & epiglottal plosive \\ 
-ʢ & \uni{02A2} & \textsc{latin letter reversed glottal stop with stroke} & voiced epiglottal fricative \\ 
-ǀ & \uni{01C0} & \textsc{latin letter dental click} & voiceless dental click \\ 
-ǁ & \uni{01C1} & \textsc{latin letter lateral click} & voiceless alveolar lateral click \\ 
-ǂ & \uni{01C2} & \textsc{latin letter alveolar click} & voiceless palatoalveolar click \\ 
-ǃ & \uni{01C3} & \textsc{latin letter retroflex click} & voiceless (post)alveolar click \\ 
-ʘ & \uni{0298} & \textsc{latin letter bilabial click} & voiceless bilabial click \\ 
-β & \uni{03B2} & \textsc{greek small letter beta} & voiced bilabial fricative \\ 
-θ & \uni{03B8} & \textsc{greek small letter theta} & voiceless dental fricative \\ 
-χ & \uni{03C7} & \textsc{greek small letter chi} & voiceless uvular fricative \\
-\end{xtabular}
-\end{center}
-%\end{table}
+\input{chapters/tab:unicode_diacritics}

### 2015-06-18 chapters/tab:unicode_diacritics.tex
update strict IPA tables

\tablecaption{Strict-IPA diacritics with Unicode encodings}\label{tab:ipa_diacritics}
 \tablefirsthead{
   \toprule
   & Code & Unicode name & IPA name \\ 
   \midrule
   }
 \tablehead{
   \multicolumn{4}{c}{
     \captionsize\tablename\ \thetable{} 
     Strict-IPA diacritics with Unicode encodings --- \textit{continued}
     } \\
   \toprule
   & Code & Unicode name & IPA name \\ 
   \midrule
   }
 \tabletail{
   \bottomrule
   \multicolumn{4}{r}{
     \captionsize\textit{continued on next page}
     } \\
   }
 \tablelasttail{\bottomrule}

\begin{center}
\begin{xtabular}{ l L{1.2cm} L{5.6cm} L{4cm} }
{\large \ ̼} & \unif{033C} & \textsc{combining seagull below} & linguolabial \\
{\large \ ̪} & \unif{032A} & \textsc{combining bridge below} & dental \\
{\large \ ̻} & \unif{033B} & \textsc{combining square below} & laminal \\
{\large \ ̺} & \unif{033A} & \textsc{combining inverted bridge below} & apical \\
{\large \ ̟} & \unif{031F} & \textsc{combining plus sign below} & advanced \\
{\large \ ̠} & \unif{0320} & \textsc{combining minus sign below} & retracted \\
{\large \ ̝} & \unif{031D} & \textsc{combining up tack below} & raised \\
{\large \ ̞} & \unif{031E} & \textsc{combining down tack below} & lowered \\
{\large \ ̘} & \unif{0318} & \textsc{combining left tack below} & advanced tongue root \\
{\large \ ̙} & \unif{0319} & \textsc{combining right tack below} & retracted tongue root \\
{\large \ ̜} & \unif{031C} & \textsc{combining left half ring below} & less rounded \\
{\large \ ̹} & \unif{0339} & \textsc{combining right half ring below} & more rounded \\
{\large \ ̬} & \unif{032C} & \textsc{combining caron below} & voiced \\
{\large \ ̥} & \unif{0325} & \textsc{combining ring below} & voiceless \\
{\large \ ̰} & \unif{0330} & \textsc{combining tilde below} & creaky voiced \\
{\large \ ̤} & \unif{0324} & \textsc{combining diaeresis below} & breathy voiced \\
{\large \ ̩} & \unif{0329} & \textsc{combining vertical line below} & syllabic \\
{\large \ ̯} & \unif{032F} & \textsc{combining inverted breve below} & non-syllabic \\
\midrule
{\large \ ̃} & \unif{0303} & \textsc{combining tilde} & nasalized \\
{\large \ ̈} & \unif{0308} & \textsc{combining diaeresis} & centralized \\
{\large \ ̽} & \unif{033D} & \textsc{combining x above} & mid-centralized \\
{\large \ ̆} & \unif{0306} & \textsc{combining breve} & extra-short \\
{\large ̚} & \unif{031A} & \textsc{combining left angle above} & no audible release \\
\midrule
{\large ˞} & \unif{02DE} & \textsc{modifier letter rhotic hook} & rhotacized \\
{\large ˡ} & \unif{02E1} & \textsc{modifier letter small l} & lateral release \\
{\large ⁿ} & \unif{207F} & \textsc{superscript latin small letter n} & nasal release \\
{\large ʷ} & \unif{02B7} & \textsc{modifier letter small w} & labialized \\
{\large ʲ} & \unif{02B2} & \textsc{modifier letter small j} & palatalized \\
{\large ˠ} & \unif{02E0} & \textsc{modifier letter small gamma} & velarized \\
{\large ˤ} & \unif{02E4} & \textsc{modifier letter small reversed glottal stop} & pharyngealized \\
{\large ʰ} & \unif{02B0} & \textsc{modifier letter small h} & aspirated \\
{\large ʼ} & \unif{02BC} & \textsc{modifier letter apostrophe} & ejective \\
{\large ː} & \unif{02D0} & \textsc{modifier letter triangular colon} & long \\
{\large ˑ} & \unif{02D1} & \textsc{modifier letter half triangular colon} & half-long \\
\midrule
{˥} & \unif{02E5} & \textsc{modifier letter extra-high tone bar} & extra high tone \\
{˦} & \unif{02E6} & \textsc{modifier letter high tone bar} & high tone \\
{˧} & \unif{02E7} & \textsc{modifier letter mid tone bar} & mid tone \\
{˨} & \unif{02E8} & \textsc{modifier letter low tone bar} & low tone \\
{˩} & \unif{02E9} & \textsc{modifier letter extra-low tone bar} & extra low tone \\
{↑} & \unif{2191} & \textsc{upwards arrow} & upstep \\
{↓} & \unif{2193} & \textsc{downwards arrow} & downstep \\
{↗} & \unif{2197} & \textsc{north east arrow} & global rise \\
{↘} & \unif{2198} & \textsc{south east arrow} & global fall \\
\midrule
{\large ˈ} & \unif{02C8} & \textsc{modifier letter vertical line} & primary stress \\
{\large ˌ} & \unif{02CC} & \textsc{modifier letter low vertical line} & secondary stress \\
{\large.} & \unif{002E} & \textsc{full stop} & syllable break \\
{|} & \unif{007C} & \textsc{vertical line} & minor group break (foot) \\
{‖} & \unif{2016} & \textsc{double vertical line} & major group break (intonation) \\
{\fontspec{CharisSIL} ‿} & \unif{203F} & \textsc{undertie} & linking (absence of a break) \\
\end{xtabular}
\end{center}

### 2015-06-18 chapters/tab:unicode_letters.tex
update strict IPA tables

\tablecaption{Strict-IPA letters with Unicode encodings}\label{tab:ipa_letters}
 \tablefirsthead{
   \toprule
   & Code & Unicode name & IPA name \\ 
   \midrule
   }
 \tablehead{
   \multicolumn{4}{c}{
     \captionsize\tablename\ \thetable{} 
     Strict-IPA letters with Unicode encodings --- \textit{continued}
     } \\
   \toprule
   & Code & Unicode name & IPA name \\ 
   \midrule
   }
 \tabletail{
   \bottomrule
   \multicolumn{4}{r}{
     \captionsize\textit{continued on next page}
     } \\
   }
 \tablelasttail{\bottomrule}

\begin{center}
\begin{xtabular}{ l L{1.2cm} L{5.4cm} L{4.2cm} }
a & \unif{0061} & \textsc{latin small letter a} & open front unrounded \\ 
æ & \unif{00E6} & \textsc{latin small letter ae} & raised open front unrounded \\ 
ɐ & \unif{0250} & \textsc{latin small letter turned a} & lowered schwa \\ 
ɑ & \unif{0251} & \textsc{latin small letter alpha} & open back unrounded \\ 
ɒ & \unif{0252} & \textsc{latin small letter turned alpha} & open back rounded \\ 
b & \unif{0062} & \textsc{latin small letter b} & voiced bilabial plosive \\ 
ʙ & \unif{0299} & \textsc{latin letter small capital b} & voiced bilabial trill \\ 
ɓ & \unif{0253} & \textsc{latin small letter b with hook} & voiced bilabial implosive \\ 
c & \unif{0063} & \textsc{latin small letter c} & voiceless palatal plosive \\ 
ç & \unif{00E7} & \textsc{latin small letter c with cedilla} & voiceless palatal fricative \\ 
ɕ & \unif{0255} & \textsc{latin small letter c with curl} & voiceless alveolo-palatal fricative \\ 
d & \unif{0064} & \textsc{latin small letter d} & voiced alveolar plosive \\ 
ð & \unif{00F0} & \textsc{latin small letter eth} & voiced dental fricative \\ 
ɖ & \unif{0256} & \textsc{latin small letter d with tail} & voiced retroflex plosive \\ 
ɗ & \unif{0257} & \textsc{latin small letter d with hook} & voiced dental/alveolar implosive \\ 
% \charis{ᶑ} & \unif{1D91} & \textsc{latin small letter d with hook and tail} & voiced retroflex implosive \\ 
e & \unif{0065} & \textsc{latin small letter e} & close-mid front unrounded \\ 
ə & \unif{0259} & \textsc{latin small letter schwa} & mid-central schwa \\ 
ɛ & \unif{025B} & \textsc{latin small letter open e} & open-mid front unrounded \\ 
ɘ & \unif{0258} & \textsc{latin small letter reversed e} & close-mid central unrounded \\ 
ɜ & \unif{025C} & \textsc{latin small letter reversed open e} & open-mid central unrounded \\ 
ɞ & \unif{025E} & \textsc{latin small letter closed reversed open e} & open-mid central rounded \\ 
f & \unif{0066} & \textsc{latin small letter f} & voiceless labiodental fricative \\ 
ɡ & \unif{0261} & \textsc{latin small letter script g} & voiced velar plosive \\ 
ɢ & \unif{0262} & \textsc{latin letter small capital g} & voiced uvular plosive \\ 
ɠ & \unif{0260} & \textsc{latin small letter g with hook} & voiced velar implosive \\ 
ʛ & \unif{029B} & \textsc{latin letter small capital g with hook} & voiced uvular implosive \\ 
ɤ & \unif{0264} & \textsc{latin small letter rams horn} & close-mid back unrounded \\ 
ɣ & \unif{0263} & \textsc{latin small letter gamma} & voiced velar fricative \\ 
h & \unif{0068} & \textsc{latin small letter h} & voiceless glottal fricative \\ 
ħ & \unif{0127} & \textsc{latin small letter h with stroke} & voiceless pharyngeal fricative \\ 
ʜ & \unif{029C} & \textsc{latin letter small capital h} & voiceless epiglottal fricative \\ 
ɦ & \unif{0266} & \textsc{latin small letter h with hook} & voiced glottal fricative \\ 
ɧ & \unif{0267} & \textsc{latin small letter heng with hook} & simultaneous voiceless postalveolar+velar fricative \\ 
ɥ & \unif{0265} & \textsc{latin small letter turned h} & voiced labial-palatal approximant \\ 
i & \unif{0069} & \textsc{latin small letter i} & close front unrounded \\ 
ɪ & \unif{026A} & \textsc{latin letter small capital i} & lax close front unrounded \\ 
ɨ & \unif{0268} & \textsc{latin small letter i with stroke} & close central unrounded \\ 
j & \unif{006A} & \textsc{latin small letter j} & voiced palatal approximant \\ 
ʝ & \unif{029D} & \textsc{latin small letter j with crossed tail} & voiced palatal fricative \\ 
ɟ & \unif{025F} & \textsc{latin small letter dotless j with stroke} & voiced palatal plosive \\ 
ʄ & \unif{0284} & \textsc{latin small letter dotless j with stroke and hook} & voiced palatal implosive \\ 
k & \unif{006B} & \textsc{latin small letter k} & voiceless velar plosive \\ 
l & \unif{006C} & \textsc{latin small letter l} & voiced alveolar lateral approximant \\ 
ʟ & \unif{029F} & \textsc{latin letter small capital l} & voiced velar lateral approximant \\ 
% ɫ & \unif{026B} & \textsc{latin small letter l with middle tilde} & voiced pharyngealized alveolar lateral approximant \\ 
ɬ & \unif{026C} & \textsc{latin small letter l with belt} & voiceless alveolar lateral fricative \\ 
ɭ & \unif{026D} & \textsc{latin small letter l with retroflex hook} & voiced retroflex lateral approximant \\ 
ɮ & \unif{026E} & \textsc{latin small letter lezh} & voiced alveolar lateral fricative \\ 
ʎ & \unif{028E} & \textsc{latin small letter turned y} & voiced palatal lateral approximant \\ 
m & \unif{006D} & \textsc{latin small letter m} & voiced bilabial nasal \\ 
ɱ & \unif{0271} & \textsc{latin small letter m with hook} & voiced labiodental nasal \\ 
n & \unif{006E} & \textsc{latin small letter n} & voiced alveolar nasal \\ 
ɴ & \unif{0274} & \textsc{latin letter small capital n} & voiced uvular nasal \\ 
ɲ & \unif{0272} & \textsc{latin small letter n with left hook} & voiced palatal nasal \\ 
ɳ & \unif{0273} & \textsc{latin small letter n with retroflex hook} & voiced retroflex nasal \\ 
ŋ & \unif{014B} & \textsc{latin small letter eng} & voiced velar nasal \\ 
o & \unif{006F} & \textsc{latin small letter o} & close-mid back rounded \\ 
ø & \unif{00F8} & \textsc{latin small letter o with stroke} & close-mid front rounded \\ 
œ & \unif{0153} & \textsc{latin small ligature oe} & open-mid front rounded \\ 
ɶ & \unif{0276} & \textsc{latin letter small capital oe} & open front rounded \\ 
ɔ & \unif{0254} & \textsc{latin small letter open o} & open-mid back rounded \\ 
ɵ & \unif{0275} & \textsc{latin small letter barred o} & close-mid central rounded \\ 
p & \unif{0070} & \textsc{latin small letter p} & voiceless bilabial plosive \\ 
ɸ & \unif{0278} & \textsc{latin small letter phi} & voiceless bilabial fricative \\ 
q & \unif{0071} & \textsc{latin small letter q} & voiceless uvular plosive \\ 
r & \unif{0072} & \textsc{latin small letter r} & voiced alveolar trill \\ 
ʀ & \unif{0280} & \textsc{latin letter small capital r} & voiced uvular trill \\ 
ɹ & \unif{0279} & \textsc{latin small letter turned r} & voiced alveolar approximant \\ 
ɺ & \unif{027A} & \textsc{latin small letter turned r with long leg} & voiced alveolar lateral flap \\ 
ɻ & \unif{027B} & \textsc{latin small letter turned r with hook} & voiced retroflex approximant \\ 
ɽ & \unif{027D} & \textsc{latin small letter r with tail} & voiced retroflex tap \\ 
ɾ & \unif{027E} & \textsc{latin small letter r with fishhook} & voiced alveolar tap \\ 
ʁ & \unif{0281} & \textsc{latin letter small capital inverted r} & voiced uvular fricative \\ 
s & \unif{0073} & \textsc{latin small letter s} & voiceless alveolar fricative \\ 
ʂ & \unif{0282} & \textsc{latin small letter s with hook} & voiceless retroflex fricative \\ 
ʃ & \unif{0283} & \textsc{latin small letter esh} & voiceless postalveolar fricative \\ 
t & \unif{0074} & \textsc{latin small letter t} & voiceless alveolar plosive \\ 
ʈ & \unif{0288} & \textsc{latin small letter t with retroflex hook} & voiceless retroflex plosive \\ 
u & \unif{0075} & \textsc{latin small letter u} & close back rounded \\ 
ʉ & \unif{0289} & \textsc{latin small letter u bar} & close central rounded \\ 
ɯ & \unif{026F} & \textsc{latin small letter turned m} & close back unrounded \\ 
ɰ & \unif{0270} & \textsc{latin small letter turned m with long leg} & voiced velar approximant \\ 
ʊ & \unif{028A} & \textsc{latin small letter upsilon} & lax close back rounded \\ 
v & \unif{0076} & \textsc{latin small letter v} & voiced labiodental fricative \\ 
ʋ & \unif{028B} & \textsc{latin small letter v with hook} & voiced labiodental approximant \\ 
\charis{ⱱ} & \unif{2C71} & \textsc{latin small letter v with right hook} & voiced labiodental tap \\ 
ʌ & \unif{028C} & \textsc{latin small letter turned v} & open-mid back unrounded \\ 
w & \unif{0077} & \textsc{latin small letter w} & voiced labial-velar approximant \\ 
ʍ & \unif{028D} & \textsc{latin small letter turned w} & voiceless labial-velar fricative \\ 
x & \unif{0078} & \textsc{latin small letter x} & voiceless velar fricative \\ 
y & \unif{0079} & \textsc{latin small letter y} & close front rounded \\ 
ʏ & \unif{028F} & \textsc{latin letter small capital y} & lax close front rounded \\ 
z & \unif{007A} & \textsc{latin small letter z} & voiced alveolar fricative \\ 
ʐ & \unif{0290} & \textsc{latin small letter z with retroflex hook} & voiced retroflex fricative \\ 
ʑ & \unif{0291} & \textsc{latin small letter z with curl} & voiced alveolo-palatal fricative \\ 
ʒ & \unif{0292} & \textsc{latin small letter ezh} & voiced postalveolar fricative \\ 
ʔ & \unif{0294} & \textsc{latin letter glottal stop} & voiceless glottal plosive \\ 
ʕ & \unif{0295} & \textsc{latin letter pharyngeal voiced fricative} & voiced pharyngeal fricative \\ 
ʡ & \unif{02A1} & \textsc{latin letter glottal stop with stroke} & epiglottal plosive \\ 
ʢ & \unif{02A2} & \textsc{latin letter reversed glottal stop with stroke} & voiced epiglottal fricative \\ 
ǀ & \unif{01C0} & \textsc{latin letter dental click} & voiceless dental click \\ 
ǁ & \unif{01C1} & \textsc{latin letter lateral click} & voiceless alveolar lateral click \\ 
ǂ & \unif{01C2} & \textsc{latin letter alveolar click} & voiceless palatoalveolar click \\ 
ǃ & \unif{01C3} & \textsc{latin letter retroflex click} & voiceless (post)alveolar click \\ 
ʘ & \unif{0298} & \textsc{latin letter bilabial click} & voiceless bilabial click \\ 
β & \unif{03B2} & \textsc{greek small letter beta} & voiced bilabial fricative \\ 
θ & \unif{03B8} & \textsc{greek small letter theta} & voiceless dental fricative \\ 
χ & \unif{03C7} & \textsc{greek small letter chi} & voiceless uvular fricative \\
\end{xtabular}
\end{center}

### 2015-06-19 chapters/ipa_meets_unicode.tex
major update of IPA chapter

@@
 The International Phonetic Alphabet (IPA) is a common standard in linguistics to
-transcribe sounds of spoken language into Latin-based characters
-\citep{IPA2005}. Although IPA is reasonably easily adhered to with pen and
-paper, it is not trivial to encode IPA characters electronically. Similar to the
-previous chapter, this chapter will discuss various pitfalls with the encoding
-of the IPA.\@ The details of the encoding are unimportant as long as the
+transcribe sounds of spoken language into Latin-based characters. Although IPA
+is reasonably easily adhered to with pen and paper, it is not trivial to encode
+IPA characters electronically. Similar to the previous chapter, this chapter
+will discuss various pitfalls with the Unicode encoding of the IPA.\@ We will
+specifically refer to the 2005 version of the IPA \citep{IPA2005} and the 7.0 
+version of Unicode \citet{Unicode2014}.
+
+The details of the encoding are unimportant as long as the
 transcription is only directed towards phonetically trained eyes. For a linguist
@@
 
-\noindent For a long time, linguists, like all other computer users, were limited to
-ASCII-encoded 7-bit characters, which only includes Latin characters, numbers
-and some punctuation and symbols. Restricted to these standard character sets
-that lacked IPA support or other language-specific graphemes that they needed,
-linguists devised their own solutions.\footnote{Early work addressing the need
-for a universal computing environment for writing systems and their
+\noindent For a long time, linguists, like all other computer users, were
+limited to ASCII-encoded 7-bit characters, which only includes Latin characters,
+numbers and some punctuation and symbols. Restricted to these standard character
+sets that lacked IPA support or other language-specific graphemes that they
+needed, linguists devised their own solutions.\footnote{Early work addressing
+the need for a universal computing environment for writing systems and their
 computational complexity is discussed in \citet{Simons1989}. A survey of
@@
 symbols are counted or if the phonetic symbols commonly found in the literature
-are also included, see~\cite[Appendix~C]{Moran2012}.}
+are also included, see~\cite[Appendix~C]{Moran2012}. The 107 characters from 12 
+code blocks shown here are the characters proposed for ``strict'' IPA encoding, 
+as discussed in Section~\ref{ipa-recommendations}.}
 
-\begin{itemize}
-	\item Basic Latin (27 characters): \newline < a b c d e f h i j k l m n o p q r s t u v w x y z~.~| >
-	\item Latin-1 Supplement (4 characters): \newline < æ ç ð ø >
-	\item Latin Extended-A (3 characters): \newline < ħ ŋ œ >
-	\item Latin Extended-B (4 characters): \newline < ǀ ǁ ǂ ǃ >
-	\item IPA Extensions (67 characters): \newline < ɐ ɑ ɒ ɓ ɔ ɕ ɖ ɗ ɘ ə ɛ ɜ ɞ ɟ ɠ ɡ ɢ ɣ ɤ ɥ ɦ ɧ ɨ ɪ ɬ ɭ ɮ ɯ ɰ ɱ ɲ ɳ ɴ ɵ ɶ ɸ ɹ ɺ ɻ ɽ ɾ ʀ ʁ ʂ ʃ ʄ ʈ ʉ ʊ ʋ ʌ ʍ ʎ ʏ ʐ ʑ ʒ ʔ ʕ ʘ ʙ ʛ ʜ ʝ ʟ ʡ ʢ > 
-	\item Spacing Modifier Letters (17 characters): \newline < ʰ ʲ ʷ ʼ ˈ ˌ ː ˑ ˞ ˠ ˡ ˤ ˥ ˦ ˧ ˨ ˩ >
-	\item Combining Diacritical Marks (29 characters): \newline < {\large \  ̃\ \ ̆\ \ ̈\ ̚\ \ ̘\ \ ̙\ \ \ ̜\ \ ̝\ \ ̞\ \ ̟\ \ ̠\ \ ̤\ \ ̥\ \ ̩\ \ ̪\ \ ̬\ \ ̯\ \ ̰\ \ ̹\ \ ̺\ \ ̻\ \ ̼\ \ ̽\ \ ͜\  ̋\ \ ́\ \ ̄\ \ ̀\ \ ̏\ } >
-	\item Greek and Coptic (3 characters): \newline < β θ χ >
-%	\item Phonetic Extensions (2 characters): \charis{ᴅ ᴴ} 
-%	\item Phonetic Extensions Supplement (1 characters): \newline < \charis{ᶑ} >
-    \item General Punctuation (2 characters): \newline < ‖ \charis{‿} >
-	\item Superscripts and Subscripts (1 character): \newline < ⁿ > 
-	\item Arrows (4 characters): \newline < ↑ ↓ ↗ ↘ >
-	\item Latin Extended-C (1 character): \newline < \charis{ⱱ} >
+\begin{itemize}[itemsep=4pt]
+	\item \textsc{Basic Latin }(27 characters) \newline 
+	a b c d e f h i j k l m n o p q r s t u v w x y z~.~|
+	\item \textsc{Latin-1 Supplement} (4 characters) \newline 
+	æ ç ð ø
+	\item \textsc{Latin Extended-A} (3 characters) \newline 
+	ħ ŋ œ
+	\item \textsc{Latin Extended-B} (4 characters) \newline 
+	ǀ ǁ ǂ ǃ
+	\item \textsc{IPA Extensions} (67 characters) \newline 
+	ɐ ɑ ɒ ɓ ɔ ɕ ɖ ɗ ɘ ə ɛ ɜ ɞ ɟ ɠ ɡ ɢ ɣ ɤ ɥ ɦ ɧ ɨ ɪ ɬ ɭ ɮ ɯ ɰ ɱ ɲ ɳ ɴ \newline
+	ɵ ɶ ɸ ɹ ɺ ɻ ɽ ɾ ʀ ʁ ʂ ʃ ʄ ʈ ʉ ʊ ʋ ʌ ʍ ʎ ʏ ʐ ʑ ʒ ʔ ʕ ʘ ʙ ʛ ʜ ʝ ʟ ʡ ʢ 
+	\item \textsc{Spacing Modifier Letters} (17 characters) \newline
+	\dia{02DE} \dia{02E1} \dia{02B7} \dia{02B2} \dia{02E0} \dia{02E4} 
+	\dia{02B0} \dia{02BC} \dia{02D0} \dia{02D1} ˥ ˦ ˧ ˨ ˩ {\large ˈ ˌ}
+	\item \textsc{Combining Diacritical Marks} (23 characters) \newline 
+	\dia{033C} \dia{032A} \dia{033B} \dia{033A} \dia{031F} \dia{0320} \dia{031D} 
+	\dia{031E} \dia{0318} \dia{0319} \dia{031C} \dia{0339} \dia{032C} \dia{0325} 
+	\dia{0330} \dia{0324} \dia{0329} \dia{032F} \dia{0303} \dia{0308} \dia{033D} 
+	\dia{0306} \dia{031A} 
+	\item \textsc{Greek and Coptic} (3 characters) \newline 
+	β θ χ
+    \item \textsc{General Punctuation} (2 characters) \newline 
+    ‖ \charis{‿}
+	\item \textsc{Superscripts and Subscripts} (1 character) \newline
+	\dia{207F} 
+	\item \textsc{Arrows} (4 characters) \newline 
+	↑ ↓ ↗ ↘
+	\item \textsc{Latin Extended-C} (1 character): \newline
+	\charis{ⱱ}
 \end{itemize}
@@
 indistinguishable, and the Cyrillic character is more easily typed on a Cyrillic
-keyboard. Some problematic homoglyphs related to encoding IPA in the Unicode Standard are 
-the following.
+keyboard. Some further problematic homoglyphs related to encoding IPA in the
+Unicode Standard are the following.
+
 \begin{itemize}
@@
   \item The uses of the apostrophe has led to long discussions on the Unicode
-        Standard email list. An English keyboard inputs <{\fontspec{Monaco}'}>
-        \textsc{apostrophe} at \uni{0027}, although the preferred Unicode
-        apostrophe is the <\ ' > \textsc{right single quotation mark} at
+        Standard email list. An English keyboard inputs the symbol
+        <\dia{0027}> \textsc{apostrophe} at \uni{0027}, although the preferred Unicode
+        apostrophe is the <\dia{2019}> \textsc{right single quotation mark} at
         \uni{2019}.\footnote{Note that many word processors (like Microsoft
@@
         stop/glottalization/ejective marker is yet another completely different
-        character <{\large ʼ}>, namely the \textsc{modifier letter apostrophe} at
-        \uni{02BC}, which unfortunately looks mostly extremely similar to
+        character, namely <\dia{02BC}>, i.e.~the \textsc{modifier letter apostrophe} 
+        at \uni{02BC}, which unfortunately looks mostly extremely similar to
         \uni{2019}. 
-  \item Another problem is the <ˁ> \textsc{modifier letter reversed glottal
-        stop} at \uni{02C1} vs.\@ the <ˤ> \textsc{modifier letter small reversed
+  \item Another problem is the <\dia{02C1}> \textsc{modifier letter reversed glottal
+        stop} at \uni{02C1} vs.\@ the <\dia{02E4}> \textsc{modifier letter small reversed
         glottal stop} at \uni{02E4}. Both appear in various resources
@@
         problem.
-  \item Even linguists are unlikely to distinguish between the <ə>
+  \item Linguists are also unlikely to distinguish between the <ə>
         \textsc{latin small letter schwa} at code point \uni{0259} and <ǝ>
@@
         postalveolar click (not retroflex). This naming is probably best seen as
-        an error in the Unicode Standard.}
+        an error in the Unicode Standard. For the ``real'' retroflex click, see 
+        Section~\ref{ipa-additions}.}
   \item The dental click <ǀ>, in Unicode known as \textsc{latin letter dental
-        click} at \uni{01C0}, is of course often simply typed as <|>
-        \textsc{vertical line} at \uni{007C}.
+        click} at \uni{01C0}, is often simply typed as <|> \textsc{vertical
+        line} at \uni{007C}.
   \item For the marking of length there is a special Unicode characters, namely
-        <ː> \textsc{modifier letter triangular colon} at \uni{02D0}. However,
-        typing <:> \textsc{colon} at \uni{003A} is of course much easier.
+        <\dia{02D0}> \textsc{modifier letter triangular colon} at \uni{02D0}. However,
+        typing <\dia{003A}> \textsc{colon} at \uni{003A} is of course much easier.
 
@@
 the double-story <a> uses a single-story <ɑ> in italics. This leads to the
-unfortunate effect that in most well-designed fonts the italics of <a> and <ɑ>
-use the same glyph. If this distinction has to be kept upright in italics, the
-only solution we can currently offer is to use \textsc{slanted} glyphs
+unfortunate effect that even in many well-designed fonts the italics of <a> and
+<ɑ> use the same glyph. For example, in Linux Libertine (the font of this book)
+the italics of these characters are highly similar <\textit{a}> and
+<\textit{ɑ}>, while in Charis SIL they are identical: <\textit{\charis{a}}> and
+<\textit{\charis{ɑ}}>. If this distinction has to be kept upright in italics,
+the only solution we can currently offer is to use \textsc{slanted} glyphs
 (i.e.~artificially italicized glyphs) instead of real italics (i.e.~special
@@
 Reversely, there are a few cases in which the IPA distinguishes different
-concepts, but the visual marking proposed by the IPA look very much alike. Such
-cases are thus homoglyphs in the IPA itself, which of course need different
-encodings.
+phonetic concepts, but the visual characters used by the IPA look very much
+alike. Such cases are thus homoglyphs in the IPA itself, which of course need
+different encodings.
 
@@
    \item The dental click <ǀ> and the indication of a minor group break <|>
-           (proposed to be used for separating feet) look almost the same in
+           look almost the same in
            most fonts. For a proper encoding, the \textsc{latin letter dental
            click} at \uni{01C0} and the \textsc{vertical line }at \uni{007C} 
-           should be strictly distinguished.
+           should be used, respectively.
    \item Similarly, the alveolar lateral click <ǁ>	should be encoded with a
            \textsc{latin letter lateral click} at \uni{01C1}, different from <‖>, 
-           which according to the IPA is a major group break (by intonation), to be
-           encoded by \textsc{double vertical line} at \uni{2016}.
+           which according to the IPA is the character to by used for a major group 
+           break (by intonation), to be encoded by \textsc{double vertical line} 
+           at \uni{2016}.
    \item The marking of primary stress < ˈ > looks like an apostrophe, and
-           is often typed with the same symbol as the ejective < ʼ >. For a
+           is often typed with the same symbol as the ejective <\dia{02BC}>. For a
            proper encoding, these two symbols should be typed as 
@@
 % ==========================
-\section{Pitfall: Multiple options in IPA}
+\section{Pitfall: Multiple encoding options in IPA}
 \label{pitfall-multiple-options-ipa}     
@@
 content. This is understandable from a transcriber's point of view, but it is
-not acceptable for interoperability between resources written in IPA.\@ We
-consider it crucial to distinguish between ``lax'' IPA, for which it is
-sufficient that any phonetically-trained reader is able to understand the
+not acceptable when the goal is interoperability between resources written in
+IPA.\@ We consider it crucial to distinguish between ``lax'' IPA, for which it
+is sufficient that any phonetically-trained reader is able to understand the
 transcription, and ``strict'' IPA, which should be standardized on a single
 unique encoding for each sound, so search will work across resources. We are
-aware of the following double options in the IPA, which will be discussed in 
-turn below:
+aware of the following non-unique encoding options in the IPA, which will be
+discussed in turn below:
 
@@
   \item The marking of <g>
-  \item The marking of velarization ad pharyngealization
+  \item The marking of velarization and pharyngealization
   \item The placement of diacritics
@@
 complexe, there are at least two different Unicode homoglyphs for the low and
-high level tones, namely <\ ̀~> \textsc{combining grave tone mark} at
-\unif{0340} vs.\@ <\ ̀~> \textsc{combining grave accent} at \unif{0300} for low
-tone, and <\ ́~> \textsc{combining acute tone mark} at \unif{0341} vs.\@ <\ ́~>
-\textsc{combining acute accent} at \unif{0301} for high tone.} The IPA also
-proposes the option of tone letters, e.g. <˥˦˧˨˩>, which are used much less, but
-are more consistent for contours. Tone letters in the IPA have five different levels, and
-sequences of these letters can be used to indicate contours. Well-designed fonts
-will even merge a sequence of tone letters into a contour. For example, compare
-the font Linux Libertine, which does not merge tone letters
-<{\fontspec{LinLibertineO}˥˨˧˩}>, with the font CharisSIL, which merges this
-sequence of four tone letters into a single contour <\charis{˥˨˧˩}>. For strict
-IPA encoding we propose to standardize on tone letters.
+high level tones, namely <\diaf{0340}> \textsc{combining grave tone mark} at
+\unif{0340} vs.~<\diaf{0300}> \textsc{combining grave accent} at \unif{0300} for
+low tone, and <\diaf{0341}> \textsc{combining acute tone mark} at \unif{0341}
+vs.<\diaf{0301}> \textsc{combining acute accent} at \unif{0301} for high tone.}
+The IPA also proposes the option of tone letters, e.g. <˥˦˧˨˩>, which are much
+less often used, but are more consistent for contours. Tone letters in the IPA
+have five different levels, and sequences of these letters can be used to
+indicate contours. Well-designed fonts will even merge a sequence of tone
+letters into a contour. For example, compare the font Linux Libertine, which
+does not merge tone letters <{\fontspec{LinLibertineO}˥˨˧˩}>, with the font
+CharisSIL, which merges this sequence of four tone letters into a single contour
+<\charis{˥˨˧˩}>. For strict IPA encoding we propose to standardize on tone
+letters.
 
@@
 
-Third, the IPA has different markers for velarization <ˠ> and pharyngealization
-<ˤ>. Confusingly, there is also a marker for ``velarized or pharyngealized'',
-using the <\ \ {\large ̴}> \textsc{combining tilde overlay} at \uni{0334}. The
-tilde overlay seems to be extremely rarely used. For strict IPA encoding we
-propose to not include the combining tilde overlay at all, but only use the 
-other two Spacing Modifiers.
+Third, the IPA has special markers for velarization <\dia{02E0}> and
+pharyngealization <\dia{02E4}>. Confusingly, there is also a marker for
+``velarized or pharyngealized'', using the <\dia{0334}> \textsc{combining tilde
+overlay} at \uni{0334}. The tilde overlay seems to be extremely rarely used. For
+strict IPA encoding we propose to not include the combining tilde overlay at
+all, but only use the other two special markers for velarization and
+phayngaealization.
 
-Finally, for marking marking of voiceless pronunciation of voiced segments the IPA
-uses the ring diacritic. Originally, the ring should be placed below the 
-base character, like in <m̥>, using the \textsc{combining ring below} at \uni{0325}. 
-However, in letters with long descenders the IPA also allows to put the ring 
-above the base, like in <ŋ̊>, using the \textsc{combining ring above} at 
-\uni{030A}. Yet, proper font design does not have any problem with rendering 
-the ring below the base character, like in <ŋ̥>, so for strict IPA encoding we 
-propose to standardize on the ring below. As a principle, for strict IPA 
-encoding only one options should be allowed.
+Finally, the IPA states that ``diacritics may be placed above a symbol with a
+descender''. For example, for marking marking of voiceless pronunciation of
+voiced segments the IPA uses the ring diacritic. Originally, the ring should be
+placed below the base character, like in <m̥>, using the \textsc{combining ring
+below} at \uni{0325}. However, in letters with long descenders the IPA also
+allows to put the ring above the base, like in <ŋ̊>, using the \textsc{combining
+ring above} at \uni{030A}. Yet, proper font design does not have any problem
+with rendering the ring below the base character, like in <ŋ̥>, so for strict
+IPA encoding we propose to standardize on the ring below. As a principle, for
+strict IPA encoding only one options should be allowed for all diacritics.
+
+The variable encoding as allowed by the IPA becomes even more troublesome for
+the tilde and diaeresis diacritics. In these cases, the IPA itself attaches
+different semantics to the symbols above and below a base characters. The tilde
+above a character (like in <ã>, using the \textsc{combining tilde} at
+\uni{0303}) is used for nasalisation, while the tilde below a character (like in
+<a̰>, using the \textsc{combining tilde below} at \uni{0330}) indicates creaky
+voice. Likewise, the diaeresis above (like in <ä>, using the \textsc{combining
+diaeresis} at \uni{0308}) is used for centralization, while the diaeresis below
+a character (like in <a̤>, using the \textsc{combining diaeresis below} at
+\uni{0324}) indicates breathy voice. These cases strengthen our plea to not
+allow diacritics to switch position for typographic convenience.
 
@@
 
-The IPA includes a tie bar to indicate affricates, doubly articulated consonants
-or diphthongs. This character can be placed above or below the base characters,
-e.g. <\charis{t͡s}> or <\charis{t͜s}>. IPA allows both options. The choice
-between the two symbols is purely for legible rendering; there is no difference
-in semantics between the two symbols. The tie-bar is a convenient diacritic for
-visually tokenizing input strings into chunks of phonetically salient groups.
-However, rendering is such a problematic issue for tie bars in general that many
-linguists simply do not use them. Just looking at a few different fonts already
-indicates that actually no font designer really gets the placement right in
-combination with superscripts and subscripts. If really necessary, we propose to
-standardize on the tie bar above the base characters, using Unicode
-\textsc{combining double inverted breve} at \uni{0361}.
+The IPA includes a tie bar. The tie-bar is a convenient diacritic for visually
+tokenizing input strings into chunks of phonetically salient groups, for example
+to indicate affricates, doubly articulated consonants or diphthongs. This
+character can be placed above or below the base characters, e.g. <\charis{t͡s}>
+or <\charis{t͜s}>. IPA allows both options. The choice between the two symbols
+is purely for legible rendering; there is no difference in semantics between the
+two symbols. However, rendering is such a problematic issue for tie bars in
+general that many linguists simply do not use them. Just looking at a few
+different fonts already indicates that actually no font designer really gets the
+placement right in combination with superscripts and subscripts. If really
+necessary, we propose to standardize on the tie bar above the base characters,
+using a \textsc{combining double inverted breve} at \uni{0361}.
 
-\begin{itemize}
-  \item[] {\fontspec{Times New Roman}Times new Roman: <t̥ʰ͡s> <t̥ʰ͜s>}
-  \item[] {\small \fontspec{CharisSIL}CharisSIL:\@ <t̥ʰ͡s> <t̥ʰ͜s>}
-  \item[] {\footnotesize \fontspec{Monaco}Monaco: <t̥ʰ͡s> <t̥ʰ͜s>}
-  \item[] {\fontspec{DoulosSIL}DoulosSIL:\@ <t̥ʰ͡s> <t̥ʰ͜s>}
-  \item[] Linux Libertine: <t̥ʰ͡s> <t̥ʰ͜s>
+\begin{itemize}[itemsep=6pt]
+  \item[] {\fontspec{Times New Roman}Times new Roman: t̥ʰ͡s t̥ʰ͜s}
+  \item[] {\small \fontspec{CharisSIL}CharisSIL:\@ t̥ʰ͡s t̥ʰ͜s}
+  \item[] {\footnotesize \fontspec{Monaco}Monaco: t̥ʰ͡s t̥ʰ͜s}
+  \item[] {\fontspec{DoulosSIL}DoulosSIL:\@ t̥ʰ͡s t̥ʰ͜s}
+  \item[] Linux Libertine: t̥ʰ͡s t̥ʰ͜s
 \end{itemize}
@@
 different from the Unicode conceptions, see
-Section~\ref{pitfall-different-notions-of-diacritics}. According to the spirit
-of the IPA, it is possible to combine more than two base characters into one tie
-bar, but this is not possible with Unicode (i.e.~there is no possibility to draw
-a tie bar over three of four characters). It is possible to indicate such larger
-groups by repeating the tie bar, like in <a͡o͡u>. If really necessary, we
-consider this possible, even though the rendering will never look good.
+Section~\ref{pitfall-different-notions-of-diacritics}. Further, according to the
+spirit of the IPA, it would also be possible to combine more than two base
+characters into one tie bar, but this is not possible with Unicode (i.e.~there
+is no possibility to draw a tie bar over three of four characters). It is
+possible to indicate such larger groups by repeating the tie bar, like in
+<a͡o͡u>. If really necessary, we consider this possible, even though the
+rendering will never look good.
 
-Most importantly though, the tie-bar is unnecessary for computational
-tokenization given the design of orthography profiles, as described in
-Section~\ref{usecase-tokenization}. The principle idea of such tokenization is 
+Most importantly though, the tie-bar is unnecessary for computational processing
+given the design of orthography profiles, as described in
+Section~\ref{usecase-tokenization}. The principle idea of such processing is
 that groups of characters that belong together are defined in an orthography
 profile, making it superfluous to indicate them explicitly by tie bars. For this
-reason we actually would like to suggest to remove the tie bar from any 
-``strict'' IPA encoding.
+reason we suggest to remove the tie bar from any ``strict'' IPA encoding, and
+instead add descriptions of all so-called ``tailored grapheme clusters'' (see
+Section~\ref{the-unicode-approach}) to an orthography profile.
 
@@
 \begin{itemize}
-  \item[] <\ {\large  ̡}\ > \textsc{combining palatalized hook below} at \uni{0321}
-  \item[] <\ \ {\large  ̢}> \textsc{combining retroflex hook below} at \uni{0322}
-  \item[] <\ \ {\large  ̵}> \textsc{combining short stroke overlay} at \uni{0335}
-  \item[] <\ \ {\large  ̷}> \textsc{combining short solidus overlay} at \uni{0337}
-  \item[] <\ \ {\large  ̴}> \textsc{combining tilde overlay} at \uni{0334}
+  \item[] <{\fontspec{CharisSIL}{\large ◌}}\symbol{"0321}> \textsc{combining palatalized hook below} at \uni{0321}
+  \item[] <{\fontspec{CharisSIL}{\large ◌}}\symbol{"0322}> \textsc{combining retroflex hook below} at \uni{0322}
+  \item[] <\dia{0335}> \textsc{combining short stroke overlay} at \uni{0335}
+  \item[] <\dia{0337}> \textsc{combining short solidus overlay} at \uni{0337}
+  \item[] <\dia{0334}> \textsc{combining tilde overlay} at \uni{0334}
 \end{itemize}
@@
 describe any character for a voiced retroflex implosive, which would
-transparently be \charis{ᶑ}. We propose to add this character to the IPA.} When
-combinations of a base character with diacritic are used, then these
-combinations are not canonical equivalent to the precomposed combinations. This
-means that any search will not find both at the same time.
+transparently be \charis{ᶑ}. We propose to add this character to the IPA, see
+Section~\ref{ipa-additions}.} When instead combinations of a base character with
+diacritic are used, then these combinations are not canonical equivalent to the
+precomposed combinations. This means that any search will not find both at the
+same time.
 
@@
 Reversely, note that the <ç> \textsc{latin small letter c with cedilla} at
-\uni{00E7} will be decomposed into <c> with <\ \ {\large ̧}> \textsc{combining
+\uni{00E7} will be decomposed into <c> with <\dia{0327}> \textsc{combining
 cedilla} at \uni{0327} by Unicode canonical decomposition, also if such a
@@
 	\begin{itemize}
-	  \item[] <ː> \textsc{modifier letter triangular colon} at \uni{02D0}
-	  \item[] <ˑ> \textsc{modifier letter half triangular colon} at \uni{02D1}
+	  \item[] <\dia{02D0}> \textsc{modifier letter triangular colon} at \uni{02D0}
+	  \item[] <\dia{02D1}> \textsc{modifier letter half triangular colon} at \uni{02D1}
 	\end{itemize}
@@
 	\begin{itemize} 
-	  \item[] <˥> \textsc{modifier letter extra-high tone bar} at \uni{02E5}
-	  \item[] <˧> \textsc{modifier letter mid tone bar} at \uni{02E7}
+	  \item[] <\dia{02E5}> \textsc{modifier letter extra-high tone bar} at \uni{02E5}
+	  \item[] <\dia{02E8}> \textsc{modifier letter low tone bar} at \uni{02E8}
 	  \item[] and others like this
@@
 	\begin{itemize}
-	  \item[] <ʰ> \textsc{modifier letter small h} at \uni{02B0}
-	  \item[] <ˤ> \textsc{modifier letter small reversed glottal stop} at \uni{02E4}
-	  \item[] <ⁿ> \textsc{superscript latin small letter n} at \uni{207F}
+	  \item[] <\dia{02B0}> \textsc{modifier letter small h} at \uni{02B0}
+	  \item[] <\dia{02E4}> \textsc{modifier letter small reversed glottal stop} at \uni{02E4}
+	  \item[] <\dia{207F}> \textsc{superscript latin small letter n} at \uni{207F}
 	  \item[] and many more like this
@@
 
-	\item[] The rhotic hook:
+	\item[] The rhotic hook:\footnote{It is really unfortunate that the rhotic hook
+         in Unicode is classified as a Spacing Modifier, and not as a Combining 
+         Diacritical Mark. Although the rhotic hook is placed to the right of its 
+         base character (and not above or below), it still is always connected 
+         to the character in front, even physically connected to it. We cannot 
+         find any reason for this treatment, and consider it an error in 
+         Unicode. We hope it will be possible to change this classification in 
+         the future.}
 	\begin{itemize}
-	  \item[] <˞> \textsc{modifier letter rhotic hook} at \uni{02DE}
+	  \item[] <\dia{02DE}> \textsc{modifier letter rhotic hook} at \uni{02DE}
 	\end{itemize}
@@
 Further, the IPA tie bar (see Section~\ref{pitfall-tie-bar}) is somewhere in
-between these two character classes. The Unicode encoding of this character is
-uses a Combining Diacritical Mark, namely either \textsc{combining double
+between these two character classes. The Unicode encoding of this character
+belong to the Combining Diacritical Marks, namely either \textsc{combining double
 inverted breve} at \uni{0361} or \textsc{combining double breve below} at
@@
 
-There is one aspect of ordering for which
-the Unicode Standard does present a canonical solution. However, it is uncontroversial
-from a linguistic perspective. Diacritics in the Unicode Standard
-(i.e.~Combining Diacritical Marks, see above) are classified in Canonical
-Combining Classes. In practice, the diacritics are distinguished by their
-position relative to the base character.\footnote{See
-\url{http://unicode.org/reports/tr44/\#Canonical\_Combining\_Class\_Values} for a
-detailed description.} When applying a Unicode normalization (NFC or NFD, see
+There is one aspect of ordering for which the Unicode Standard does present a
+canonical solution. However, it is uncontroversial from a linguistic
+perspective. Diacritics in the Unicode Standard (i.e.~Combining Diacritical
+Marks, see above) are classified in Canonical Combining Classes. In practice,
+the diacritics are distinguished by their position relative to the base
+character.\footnote{See
+\url{http://unicode.org/reports/tr44/\#Canonical\_Combining\_Class\_Values} for
+a detailed description.} When applying a Unicode normalization (NFC or NFD, see
 Section~\ref{pitfall-canonical-equivalence}), the diacritics in different
 positions are put in a specified order. This process therefore harmonizes the
-difference between different encodings, e.g.\ of a midtone creaky voice
-vowel <ḛ̄>. This grapheme cluster can be encoded either as <e> + <\ \ ̄> + <\ \
-̰> or as <e> + <\ \ ̰> + <\ \ ̄>. To prevent this twofold encoding, the Unicode
-Standard specifies the second ordering as canonical (in this case, diacritics
-below are put before diacritics above).
+difference between different encodings, e.g.\ of an extra-short creaky voice vowel
+<ḛ̆>. This grapheme cluster can be encoded either as <e> + <\dia{0306}> +
+<\dia{0330}> or as <e> + <\dia{0330}> + <\dia{0306}>. To prevent this twofold encoding,
+the Unicode Standard specifies the second ordering as canonical (in this case,
+diacritics below are put before diacritics above).
 
@@
 
-\subsubsection*{Diacritic ordering}
+\subsubsection*{Proposal for diacritic ordering}
 
@@
 developing a large database of phonological inventories from the world's
-languages, \citet[540]{Moran2012} defines a set of diacritic ordering
+languages, \citet[540]{Moran2012} defined a set of diacritic ordering
 conventions. The conventions are influenced by the linguistic literature, though
@@
 apical, laminal, advanced, retracted), followed by the manner features (raised,
-lowered, advanced tongue root, retracted tongue root, frictionalized), then
-secondary articulations (more round, less round, derhoticized), laryngeal
-settings (creaky, breathy, voiced, devoiced (below), devoiced (above)), and
-finally the syllabic or non-syllabic marker (for vowels, advanced and retracted
-tongue root markers gets applied between the place and laryngeal setting). So
-the order that is proposed is the following (where <\textbar{}> indicates
-\textit{or} and <→> indicates \textit{precedes}):
+lowered, advanced and retracted tongue root), then
+secondary articulations (more round, less round), laryngeal settings (creaky,
+breathy, voiced, devoiced), and finally the syllabic or non-syllabic marker. So,
+the order that is proposed is the following, where <\textbar{}> indicates
+\textit{or} and <→> indicates \textit{precedes}. Note that the groups of 
+alternatives (as marked by <\textbar{}>) are supposed never to occur together 
+with the same base character. In effect, this represents yet another restriction 
+on possible diacritic sequences.
 
@@
 	\begin{itemize}	
-	  \item[→] linguolabial <t̼> \textbar{} dental <t̪> \textbar{} apical <t̺> \textbar{} laminal <t̻>
-	  \item[→] advanced <u̟> \textbar{} retracted <e̠> 
-	  \item[→] raised <e̝> \textbar{} lowered <e̞>
-	  \item[→] advanced tongue root <e̘> \textbar{} retracted tongue root <e̙>
-%	  \item[→] frictionalized <ɾ͓>
-	  \item[→] more rounded <ɔ̹> \textbar{} less rounded <ɔ̜>
-	  \item[→] creaky voiced <b̰> \textbar{} breathy voiced <b̤> \textbar{} voiced <s̬> \textbar{} voiceless <n̥>
-	  \item[→] syllabic <n̩> \textbar{} non-syllabic <e̯>
+	  \item[→] linguolabial <\dia{033C}> \textbar{} dental <\dia{032A}> \textbar{} apical <\dia{033A}> \textbar{} laminal <\dia{033B}>
+	  \item[→] advanced <\dia{031F}> \textbar{} retracted <\dia{0320}> 
+	  \item[→] raised <\dia{031D}> \textbar{} lowered <\dia{031E}>
+	  \item[→] advanced tongue root <\dia{0318}> \textbar{} retracted tongue root <\dia{0319}>
+	  \item[→] more rounded <\dia{0339}> \textbar{} less rounded <\dia{031C}>
+	  \item[→] creaky voiced <\dia{0330}> \textbar{} breathy voiced <\dia{0324}> \textbar{} voiced <\dia{032C}> \textbar{} voiceless <\dia{0325}>
+	  \item[→] syllabic <\dia{0329}> \textbar{} non-syllabic <\dia{032F}>
 	\end{itemize}
@@
 
-Further, if a character sequence contains more than one diacritic above the base
-character, we propose the following order. Note that this ordering is only
-important when the extra short superscript is used. All other combinations will 
-in practice never lead to problems.
+\noindent Next, if a character sequence contains more than one diacritic above the base
+character, we propose the following order:
 
@@
 	\begin{itemize}
-	  \item[→] nasalized <ẽ>
-	  \item[→] centralized <ë> \textbar{} mid-centralized <e̽>
-	  \item[→] extra short <ĕ>
-%	  \item[→] (tone accents, e.g. <è>)
-	  \item[→] no audible release <p̚>
-%	  \item[→] Spacing Modifier Letters, see above	  
+	  \item[→] nasalized <\dia{0303}>
+	  \item[→] centralized <\dia{0308}> \textbar{} mid-centralized <\dia{033D}>
+	  \item[→] extra short <\dia{0306}>
+	  \item[→] no audible release <\dia{031A}>
  \end{itemize} \end{itemize}
 
-Then, when a character sequence contains more than one character in Spacing
+\noindent Then, when a character sequence contains more than one character of the Spacing
 Modifier Letters, these will be placed after all combining diacritic marks in the
@@
 	\begin{itemize}
-	  \item[→] rhotic hook <{\large \ ˞} >
-	  \item[→] lateral release <ˡ> \textbar{} nasal release <ⁿ>
-	  \item[→] labialized <ʷ>
-	  \item[→] palatalized <ʲ>
-	  \item[→] velarized <ˠ>
-	  \item[→] pharyngealized <ˤ>
-%	  \item[→] glottalized <ˀ>
-	  \item[→] aspirated <ʰ> \textbar{} ejective <ʼ>
-	  \item[→] long <ː> \textbar{} half-long <ˑ>
+	  \item[→] rhotic hook <\dia{02DE}>
+	  \item[→] lateral release <\dia{02E1}> \textbar{} nasal release <\dia{207F}>
+	  \item[→] labialized <\dia{02B7}>
+	  \item[→] palatalized <\dia{02B2}>
+	  \item[→] velarized <\dia{02E0}>
+	  \item[→] pharyngealized <\dia{02E4}>
+	  \item[→] aspirated <\dia{02B0}> \textbar{} ejective <\dia{02BC}>
+	  \item[→] long <\dia{02D0}> \textbar{} half-long <\dia{02D1}>
 %	  \item[→] tone letters <˥ ˦ ˧ ˨ ˩>
@@
 
-Finally, Tone!
+\noindent Finally, Tone markers are put at the end of any such sequence:
+
+\begin{itemize}
+  \item[] \textsc{Tone Marks:}
+  \begin{itemize}
+    \item[→] tone letters <˥ ˦ ˧ ˨ ˩> \textbar{} tone movement <↑ ↓ ↗ ↘>
+  \end{itemize}
+\end{itemize}
+
+% ==========================
+\section{Additions to the IPA}
+\label{ipa-additions}
+% ==========================
+
+In the course of collecting a large sample of phoneme systems across the world's
+languages \citep{Moran2012}, we found that in order to preserve distinctions
+both within and across language descriptions, additions to the approved IPA
+glyph set were needed. Wherever possible these additions were drawn from the
+extIPA symbols for disordered
+speech.\footnote{\textbf{Add better reference here: }\url{https://www.internationalphoneticassociation.org/sites/default/files/extIPAChart2008.pdf}}
+This section describes our proposed additions to the IPA glyph set. These
+additions are not part of the official IPA recommendations, so they should be 
+used with care.
+
+\begin{itemize}
+
+\item \textsc{Retroflex click} \newline
+      Retroflex clicks can be represented by <‼> \textsc{double exclamation
+      mark} at \uni{203C}. Note that the (post-)alveolar click <ǃ> at \uni{01C3}
+      is confusingly referred to as \textsc{latin letter retroflex click} in the
+      Unicode standard, which is probably best seen as an error.
+
+\item \textsc{Voiced retroflex implosive} \newline 
+      Although the IPA includes a
+      series of voiced implosives (marked with a hook on top, see
+      Section~\ref{pitfall-missing-decomposition}), there is no voiced retroflex
+      implosive. Following the spirit of the IPA, we propose to use <\charis{ᶑ}>
+      \textsc{latin small letter d with hook and tail} at \uni{1D91} for this
+      sound.
+
+\item \textsc{Fortis/lenis} \newline
+      Languages described as having a fortis/plain/lenis distinction that
+      corresponds poorly with the traditional
+      voiced/voiceless-unaspirated/voiceless-aspirated continuum can be marked
+      using the voiceless glyph for the plain phoneme, and then 
+      <\dia{0348}> \textsc{combining double vertical line below} at
+      \uni{0348} to mark the fortis articulation, and/or <\dia{0349}>
+      \textsc{combining left angle below} at \uni{0349} for the lenis
+      articulation.
+
+\item \textsc{Frictionalization} \newline 
+      The diacritic <\dia{0353}>
+      \textsc{combining x below} at \uni{0353} can be used to represent three
+      types of frictionalized sounds: First, click consonants where the release
+      of the anterior closure involves an ingressive ``sucking'' sound similar
+      to a fricative, for example <kǃ͓ʰ>; second, ``frictionalized'' vowels
+      (sounds that are phonologically vocalic, but with sufficiently close
+      closures to create buzzing); and third, fricative sounds at places of
+      articulation that do not have dedicated fricative glyphs, for example
+      sounds with voiceless velar lateral frication, like <ʟ̥͓>.
+
+\item \textsc{Derhoticization} \newline 
+      For derhoticization we propose to use
+      <\dia{032E}> \textsc{combining breve below} at \uni{032E}. 
+      \textbf{WHAT EXACTLY IS THIS?}
+
+\item \textsc{Coronal non-sibilant} \newline
+      Languages described as having a sibilant/non-sibilant distinction among
+      coronal fricatives and affricates can be handled using the subscript
+      <\dia{0347}> \textsc{combining equals sign below} at \uni{0347} to mark
+      the non-sibilant phoneme.
+
+\item \textsc{Glottalization} \newline 
+      Glottalized sounds can be indicated using
+      <\dia{02C0}> \textsc{modifier letter glottal stop} at \uni{02C0}, unless
+      it is clear that either ``ejective'' or ``creaky voicing'' are the
+      intended sounds (in which cases the standard IPA diacritics should be
+      used). Pre-glottalized sounds can be marked with
+      <\diareverse{02C0}> to the left of the base
+      glyph, for example <\charis{ˀt}>.
+
+\item \textsc{Voiced pre-aspiration} \newline Voiced sounds having
+      pre-aspiration can be marked with
+      <\diareverse{02B1}> \textsc{modifier letter
+      small h with hook} at \uni{02B1} to the left of the base glyph, for
+      example <\charis{ʱd}>.
+
+\item \textsc{Epilaryngeal phonation} \newline 
+      There are some rare articulations that make
+      use of an epilaryngeal phonation mechanism (e.g.,~the “sphincteric vowels”
+      of~!Xóõ). To represent these vowels, we propose to use the modifier <\dia{1D31}>
+      \textsc{modifier letter capital e} at \uni{1D31} to denote such sphincteric
+      phonation.
+
+\end{itemize}
 
@@
 
-\input{chapters/tab:unicode_letters}
+\input{tables/unicode_letters}
 
-\input{chapters/tab:unicode_diacritics}
+\input{tables/unicode_diacritics}
+
+\input{tables/unicode_boundaries}
+
+\input{tables/unicode_additions}

### 2015-06-19 tables/unicode_additions.tex
major update of IPA chapter

\tablecaption{Additions to the IPA with Unicode encodings}\label{tab:ipa_additions}
 \tablefirsthead{
   \toprule
   & Code & Unicode name & Phonetic description \\ 
   \midrule
   }
 \tablehead{
   \multicolumn{4}{c}{
     \captionsize\tablename\ \thetable{} 
     Additions to the IPA with Unicode encodings --- \textit{continued}
     } \\
   \toprule
   & Code & Unicode name & Phonetic description \\ 
   \midrule
   }
 \tabletail{
   \bottomrule
   \multicolumn{4}{r}{
     \captionsize\textit{continued on next page}
     } \\
   }
 \tablelasttail{\bottomrule}

\begin{center}
\begin{xtabular}{ l L{1.2cm} L{6.1cm} L{3.5cm} }
‼ & \unif{203C} & \textsc{double exclamation mark} & retroflex click \\
\charis{ᶑ} & \unif{1D91} & \textsc{latin small letter d with hook and tail} & voiced retroflex implosive \\
\dia{0348} & \unif{0348} & \textsc{combining double vertical line below} & fortis \\
\dia{0349} & \unif{0349} & \textsc{combining left angle below} & lenis \\
\dia{0353} & \unif{0353} & \textsc{combining x below} & frictionalized \\
\dia{032E} & \unif{032E} & \textsc{combining breve below} & derhoticized \\
\dia{0347} & \unif{0347} & \textsc{combining equals sign below} & non-sibilant \\
\dia{02C0} & \unif{02C0} & \textsc{modifier letter glottal stop} & glottalized \\
\diareverse{02B1} & \unif{02B1} & \textsc{modifier letter small h with hook} & voiced pre-aspirated \\
\dia{1D31} & \unif{1D31} & \textsc{modifier letter capital e} & epilaryngeal phonation \\
\end{xtabular}
\end{center}

### 2015-06-19 tables/unicode_boundaries.tex
major update of IPA chapter

\tablecaption{Strict-IPA boundaries with Unicode encodings}\label{tab:ipa_boundaries}
 \tablefirsthead{
   \toprule
   & Code & Unicode name & IPA name \\ 
   \midrule
   }
 \tablehead{
   \multicolumn{4}{c}{
     \captionsize\tablename\ \thetable{} 
     Strict-IPA boundaries with Unicode encodings --- \textit{continued}
     } \\
   \toprule
   & Code & Unicode name & IPA name \\ 
   \midrule
   }
 \tabletail{
   \bottomrule
   \multicolumn{4}{r}{
     \captionsize\textit{continued on next page}
     } \\
   }
 \tablelasttail{\bottomrule}

\begin{center}
\begin{xtabular}{ l L{1.2cm} L{5.6cm} L{4cm} }
{\large ˈ} & \unif{02C8} & \textsc{modifier letter vertical line} & primary stress \\
{\large ˌ} & \unif{02CC} & \textsc{modifier letter low vertical line} & secondary stress \\
{\large.} & \unif{002E} & \textsc{full stop} & syllable break \\
{|} & \unif{007C} & \textsc{vertical line} & minor group break (foot) \\
{‖} & \unif{2016} & \textsc{double vertical line} & major group break (intonation) \\
\charis{‿} & \unif{203F} & \textsc{undertie} & linking (absence of a break) \\
\end{xtabular}
\end{center}

### 2015-06-19 tables/unicode_diacritics.tex
major update of IPA chapter

\tablecaption{Strict-IPA diacritics with Unicode encodings}\label{tab:ipa_diacritics}
 \tablefirsthead{
   \toprule
   & Code & Unicode name & IPA name \\ 
   \midrule
   }
 \tablehead{
   \multicolumn{4}{c}{
     \captionsize\tablename\ \thetable{} 
     Strict-IPA diacritics with Unicode encodings --- \textit{continued}
     } \\
   \toprule
   & Code & Unicode name & IPA name \\ 
   \midrule
   }
 \tabletail{
   \bottomrule
   \multicolumn{4}{r}{
     \captionsize\textit{continued on next page}
     } \\
   }
 \tablelasttail{\bottomrule}

\begin{center}
\begin{xtabular}{ l L{1.2cm} L{6.6cm} L{3cm} }
\dia{033C} & \unif{033C} & \textsc{combining seagull below} & linguolabial \\
\dia{032A} & \unif{032A} & \textsc{combining bridge below} & dental \\
\dia{033B} & \unif{033B} & \textsc{combining square below} & laminal \\
\dia{033A} & \unif{033A} & \textsc{combining inverted bridge below} & apical \\
\dia{031F} & \unif{031F} & \textsc{combining plus sign below} & advanced \\
\dia{0320} & \unif{0320} & \textsc{combining minus sign below} & retracted \\
\dia{031D} & \unif{031D} & \textsc{combining up tack below} & raised \\
\dia{031E} & \unif{031E} & \textsc{combining down tack below} & lowered \\
\dia{0318} & \unif{0318} & \textsc{combining left tack below} & advanced tongue root \\
\dia{0319} & \unif{0319} & \textsc{combining right tack below} & retracted tongue root \\
\dia{031C} & \unif{031C} & \textsc{combining left half ring below} & less rounded \\
\dia{0339} & \unif{0339} & \textsc{combining right half ring below} & more rounded \\
\dia{032C} & \unif{032C} & \textsc{combining caron below} & voiced \\
\dia{0325} & \unif{0325} & \textsc{combining ring below} & voiceless \\
\dia{0330} & \unif{0330} & \textsc{combining tilde below} & creaky voiced \\
\dia{0324} & \unif{0324} & \textsc{combining diaeresis below} & breathy voiced \\
\dia{0329} & \unif{0329} & \textsc{combining vertical line below} & syllabic \\
\dia{032F} & \unif{032F} & \textsc{combining inverted breve below} & non-syllabic \\
\dia{0303} & \unif{0303} & \textsc{combining tilde} & nasalized \\
\dia{0308} & \unif{0308} & \textsc{combining diaeresis} & centralized \\
\dia{033D} & \unif{033D} & \textsc{combining x above} & mid-centralized \\
\dia{0306} & \unif{0306} & \textsc{combining breve} & extra-short \\
\dia{031A} & \unif{031A} & \textsc{combining left angle above} & no audible release \\
\dia{02DE} & \unif{02DE} & \textsc{modifier letter rhotic hook} & rhotacized \\
\dia{02E1} & \unif{02E1} & \textsc{modifier letter small l} & lateral release \\
\dia{207F} & \unif{207F} & \textsc{superscript latin small letter n} & nasal release \\
\dia{02B7} & \unif{02B7} & \textsc{modifier letter small w} & labialized \\
\dia{02B2} & \unif{02B2} & \textsc{modifier letter small j} & palatalized \\
\dia{02E0} & \unif{02E0} & \textsc{modifier letter small gamma} & velarized \\
\dia{02E4} & \unif{02E4} & \textsc{modifier letter small reversed glottal stop} & pharyngealized \\
\dia{02B0} & \unif{02B0} & \textsc{modifier letter small h} & aspirated \\
\dia{02BC} & \unif{02BC} & \textsc{modifier letter apostrophe} & ejective \\
\dia{02D0} & \unif{02D0} & \textsc{modifier letter triangular colon} & long \\
\dia{02D1} & \unif{02D1} & \textsc{modifier letter half triangular colon} & half-long \\{˥} & \unif{02E5} & \textsc{modifier letter extra-high tone bar} & extra high tone \\
{˦} & \unif{02E6} & \textsc{modifier letter high tone bar} & high tone \\
{˧} & \unif{02E7} & \textsc{modifier letter mid tone bar} & mid tone \\
{˨} & \unif{02E8} & \textsc{modifier letter low tone bar} & low tone \\
{˩} & \unif{02E9} & \textsc{modifier letter extra-low tone bar} & extra low tone \\
{↑} & \unif{2191} & \textsc{upwards arrow} & upstep \\
{↓} & \unif{2193} & \textsc{downwards arrow} & downstep \\
{↗} & \unif{2197} & \textsc{north east arrow} & global rise \\
{↘} & \unif{2198} & \textsc{south east arrow} & global fall \\
\end{xtabular}
\end{center}

### 2015-06-19 chapters/ipa_meets_unicode.tex
refinements of IPA chapter

@@
       Unicode standard, which is probably best seen as an error.
-
 \item \textsc{Voiced retroflex implosive} \newline 
@@
       sound.
-
 \item \textsc{Fortis/lenis} \newline
@@
       articulation.
-
 \item \textsc{Frictionalization} \newline 
@@
       sounds with voiceless velar lateral frication, like <ʟ̥͓>.
-
 \item \textsc{Derhoticization} \newline 
@@
       \textbf{WHAT EXACTLY IS THIS?}
-
 \item \textsc{Coronal non-sibilant} \newline
@@
       the non-sibilant phoneme.
-
 \item \textsc{Glottalization} \newline 
@@
       glyph, for example <\charis{ˀt}>.
-
 \item \textsc{Voiced pre-aspiration} \newline Voiced sounds having
@@
       example <\charis{ʱd}>.
-
 \item \textsc{Epilaryngeal phonation} \newline 
@@
 
-Summarizing the pitfalls discussed in this chapter, we propose to define a 
+Summarizing the pitfalls as discussed in this chapter, we propose to define a 
 ``strict'' IPA encoding to be used when interoperability of phonetic resources 
 is intended. For regular research papers, when only the visual outline is 
-important,
+important, the details encoding of IPA in electronic format is rather 
+unimportant, as long as it looks as expected. So for those regular usage these 
+proposals are irrelevant.
 
+[TO DO: different kind of proposals]
+
+At the end of this chapter we have added a few longish tables summarizing all
+155 different unicode codepoints that form the basis of strict-IPA encoding
+(107 letters, 41 diacritics and 7 remaining symbols). Each of
+these tables shows a typical glyph, and then lists the Unicode Codepoint,
+Unicode Name and IPA description for each symbol.
+
+TO DO:
+[ordering]
+[removed characters: tie bar, tilde overlay]
+[additions]
+
+\begin{itemize}[itemsep=6pt]
+
+  \item \textsc{IPA letters} \newline
+        The 107 different IPA letters as allowed in strict-IPA encoding are
+        listed in Table~\ref{tab:ipa_letters} starting on
+        page~\pageref{tab:ipa_letters}.
+  \item \textsc{IPA diacritics} \newline The 41 different IPA diacritics and
+        tone markers (both Unicode Modifier Letters and Combining Diacritical
+        Marks) as allowed in strict-IPA encoding are listed in
+        Table~\ref{tab:ipa_diacritics} starting on
+        page~\pageref{tab:ipa_diacritics}.
+  \item \textsc{IPA remainders} \newline
+        The 7 remaining IPA symbols (boundary and stress markers) as allowed in
+        strict-IPA encoding are listed in Table~\ref{tab:ipa_boundaries} on
+        page~\pageref{tab:ipa_boundaries}. 
+  \item \textsc{IPA additions} \newline
+        The 10 proposed additions to the IPA are listed in
+        Table~\ref{tab:ipa_additions} on page~\pageref{tab:ipa_additions}.
+
+\end{itemize}
+
+\newpage
 \input{tables/unicode_letters}
-
 \input{tables/unicode_diacritics} 
-
 \input{tables/unicode_boundaries}
-
 \input{tables/unicode_additions}

### 2015-06-19 tables/unicode_boundaries.tex
refinements of IPA chapter

@@
 {\large ˌ} & \unif{02CC} & \textsc{modifier letter low vertical line} & secondary stress \\
+ & \unif{0020} & \textsc{space} & word break \\
 {\large.} & \unif{002E} & \textsc{full stop} & syllable break \\

### 2015-06-20 chapters/introduction.tex
adding two new unicode pittfalls

@@
 is a base character followed by a letter modifier character. For example, the
-sequence <n> + < ̃> (i.e.~\textsc{latin small letter n} at \uni{006E}, followed
+sequence <n> + <\dia{0303}> (i.e.~\textsc{latin small letter n} at \uni{006E}, followed
 by \textsc{combining tilde} at \uni{0303}) combines visually into <ñ>, a

### 2015-06-20 chapters/ipa_meets_unicode.tex
adding two new unicode pittfalls

@@
 specifically refer to the 2005 version of the IPA \citep{IPA2005} and the 7.0 
-version of Unicode \citet{Unicode2014}.
+version of Unicode \citep{Unicode2014}.
 
@@
 a segment, but bind two graphemes together. This actually turns out to be rather
-different from the Unicode conceptions, see
-Section~\ref{pitfall-different-notions-of-diacritics}. Further, according to the
+different from the Unicode conceptions. The Unicode encoding of this character
+belong to the Combining Diacritical Marks, namely either \textsc{combining double
+inverted breve} at \uni{0361} or \textsc{combining double breve below} at
+\uni{035C}. Such a combining mark is by definition tied to the character in
+front, but not the character following it. The Unicode treatment of this
+character thus only partly corresponds to the IPA conception, which ideally
+would have the tie bar linked both to the character in front and to the
+character following. The solution to this is to always use the tie bar in 
+combination with
+
+Further, according to the
 spirit of the IPA, it would also be possible to combine more than two base
@@
 for aspiration (more precisely post-aspiration following the base character) and
-pre-aspiration (preceding the base character). Note that there exists a mechanism in
-Unicode to force separate characters to be combined (namely by using the
-\textsc{zero width joiner} at \uni{200D}), but this seems to be a rather
-impractical, and probably not an enforceable solution to us.
-
-Further, the IPA tie bar (see Section~\ref{pitfall-tie-bar}) is somewhere in
-between these two character classes. The Unicode encoding of this character
-belong to the Combining Diacritical Marks, namely either \textsc{combining double
-inverted breve} at \uni{0361} or \textsc{combining double breve below} at
-\uni{035C}. Such a combining mark is by definition tied to the character in
-front, but not the character following it. The Unicode treatment of this
-character thus only partly corresponds to the IPA conception, which ideally
-would have the tie bar linked both to the character in front and to the
-character following.
+pre-aspiration (preceding the base character). Note that there exists a
+mechanism in Unicode to force separate characters to be combined (namely by
+using the \textsc{zero width joiner} at \uni{200D}), but to us this seems to be
+a rather impractical, and probably not an enforceable solution to the general
+problem of complex graphemes. The default combination of Spacing Modifiers with
+the preceding character can be specified in orthography profiles (see
+Chapter~\ref{orthography-profiles}). Only in case variable directions occur in 
+the same transcription (e.g. both pre-aspiration and poast-aspiration occur), 
+then each case has to be explicitly marked with a zero-width joiner.
 

### 2015-06-20 chapters/pitfalls.tex
adding two new unicode pittfalls

@@
 change the visual appearance of a character, for example Times New Roman
-two-story <{\fontspec{Times New Roman}a}> changes to a single-story a in italics
+two-story <{\fontspec{Times New Roman}a}> changes to a single-story glyph in italics
 <\emph{\fontspec{Times New Roman}a}>. This becomes a real problem for some
@@
 rendering does not (necessarily) indicate logical encoding.
+
+% ==========================
+\section{Pitfall: Characters are not graphemes}
+\label{pitfall-characters-are-not-graphemes}
+% ==========================
+
+The Unicode Standard is a character encoding system, and not a writing system
+encoding system. This most forcefully becomes clear with the notion of grapheme.
+From a linguistic point of view, graphemes are the basic building blocks of a
+writing system (see Section~\ref{linguistic-terminology}). It is extremely
+common, up to the point of being universally attested, that writing systems use
+combinations of multiple symbols as a single grapheme, like <sch>, <th> or <ei>.
+There is no possibility to encode such complex graphemes using the Unicode Standard.
+
+The Unicode Standard deals with complex graphemes only inasmuch they consist of
+base characters with diacritics (see
+Section~\ref{pitfall-different-notions-of-diacritics} for a discussion of the
+notion of diacritic). The Unicode Standard calls such combination ``grapheme
+clusters''. Complex graphemes consisting of multiple base characters,
+like <sch>, are called ``tailored grapheme clusters'' in Unicode parlance (see
+Section~\ref{the-unicode-approach}).
+
+Inspecting the Unicode Standard, there appear to be special Unicode characters
+to ``glue'' together characters into larger tailored grapheme clusters,
+specifically the \textsc{zero width joiner} at \uni{200D} and the
+\textsc{combining grapheme joiner} at \uni{034F}. However, these characters are
+confusingly named (cf.~Section~\ref{pitfall-names}). Both codepoints actually do
+not join characters, but explicitly separate them. The zero-width joiner (ZWJ)
+can be used to solve special problems related to ordering (called ``collation''
+in Unicode parlance). The combining grapheme joiner (CGJ) can be used to
+separate characters that are not supposed to form ligatures. To solve the issue
+of tailored grapheme clusters, Unicode offers some assistance in the form of the
+Unicode Locale Descriptions. However, in the practice of linguistic research,
+this is not a real solution. For that reason we propose to use orthography
+profiles (see Chapter~\ref{orthography-profiles}).
 
@@
 % ==========================
+\section{Pitfall: Names}
+\label{pitfall-names}
+% ==========================
+
+The names of characters in the Unicode Standard are sometimes misnomers and
+should not be misinterpreted as definitions. For example, the \textsc{combining
+grapheme joiner} at \uni{034F} does not join characters into larger graphemes
+(see Section~\ref{pitfall-characters-are-not-graphemes}) and the \textsc{latin
+letter retroflex click} \uni{01C3} is actually not the IPA symbol for a
+retroflex click, but for an alveolar click (see
+Section~\ref{pitfall-ipa-homoglyphs}). In a sense, these names can be seen as
+``errors.'' However, it is probably better to realize that such names are just
+convenience labels that are not going to be changed. Just like the block names
+(Section~\ref{pitfall-blocks}), the character names are often helpful, but they
+are not supposed to be definitions.
+
+The actual intended ``meaning'' of a Unicode codepoint is a combination of the
+name, the block and the character properties (see
+Section~\ref{the-unicode-approach}). Further details about the underlying intentions 
+with which a character should be used
+are only accessible by perusing the actual decisions of the Unicode Consortium.
+All proposals, discussions and decisions of the Unicode Consortium are publicly
+available. Unfortunately there is not (yet) any way to easily find everything
+that is ever proposed, discussed and decided in relation to a specific
+codepoint of interest, so many of the details are often somewhat
+hidden.\footnote{All proposals and other documents that are the basis of Unicode
+decisions are avaialbe at \url{http://www.unicode.org/L2/all-docs.html}. The
+actual decisions that make up the Unicode Standard are documented in the minutes
+of the Unicode Technical Committee, available at
+\url{http://www.unicode.org/consortium/utc-minutes.html}.}
+
+% ==========================
 \section{Pitfall: Homoglyphs}
@@
 	\def\labelenumi{\arabic{enumi}.} 
-	\item[] <Å> \textsc{latin capital letter a with ring above} \uni{00C5} 
-	\item[] <Å> \textsc{angstrom sign} \uni{212B}
-	\item[] <Å> \textsc{latin capital letter a} \uni{0041}
+	\item <Å> \textsc{latin capital letter a with ring above} \uni{00C5} 
+	\item <Å> \textsc{angstrom sign} \uni{212B}
+	\item <Å> \textsc{latin capital letter a} \uni{0041}
 	+ \textsc{combining ring above} \uni{030A}
@@
 \url{http://www.win.tue.nl/~aeb/linux/uc/nfc_vs_nfd.html}.} This means in
-practice that if a user, for example, enters <a> and < ̀>, i.e.~\textsc{latin
+practice that if a user, for example, enters <a> and <\dia{0300}>, i.e.~\textsc{latin
 small letter a} at \uni{0061} and \textsc{combining grave accent} at \uni{0300},
@@
 
-  \item the \textsc{combining cedilla} at \uni{0327} \newline This diacritic is
+  \item <\dia{0327}>: the \textsc{combining cedilla} at \uni{0327} \newline 
+        This diacritic is
         for example attested in the precomposed character <ç> \textsc{latin
@@
         decomposed in NFC normalization.
-  \item the \textsc{combining ogonek} at \uni{0328} \newline This diacritic is
+  \item <\dia{0328}>: the \textsc{combining ogonek} at \uni{0328} \newline 
+        This diacritic is
         for example attested in precomposed <ą> \textsc{latin small letter a
@@
         normalization.
-  \item the \textsc{combining horn} at \uni{031B} \newline This diacritic is for
+  \item <\dia{031B}>: the \textsc{combining horn} at \uni{031B} \newline 
+        This diacritic is for
         example attested in precomposed <ơ> \textsc{latin small letter o with
@@
   \item First, there are the combining characters designated as ``combining
-        overlay'' in the Unicode Standard, like <\ \ {\large ̴}>
-        \textsc{combining tilde overlay} at \uni{0334} or <\ \ {\large ̵}>
+        overlay'' in the Unicode Standard, like <\dia{0334}>
+        \textsc{combining tilde overlay} at \uni{0334} or <\dia{0335}>
         \textsc{combining short stroke overlay} at \uni{0335}. There are many
@@
   \item Second, the same situation also occurs with combining characters
-        designated as ``combining hook'', like < {\large ̡}> \textsc{combining
+        designated as ``combining hook'', like 
+        <{\fontspec{CharisSIL}{\large ◌}}\symbol{"0321}> \textsc{combining
         palatalized hook below} at \uni{0321}. This element seems to occur in
@@
          rendering (see Section~\ref{pitfall-faulty-rendering}).
-   \item Do not trust the names of Unicode blocks as a strategy to find specific 
-         characters (see Section~\ref{pitfall-blocks})
+   \item Do not trust the names of codepoints as a definition of the character
+         (see Section~\ref{pitfall-names}). Also do not trust Unicode blocks as
+         a strategy to find specific characters (see
+         Section~\ref{pitfall-blocks})
    \item To ensure consistent encoding of texts, apply Unicode normalization
@@
          use orthography profiles (see Chapter~\ref{orthography-profiles}).
+   \item To deal with ``tailored'' grapheme clusters
+         (Section~\ref{pitfall-characters-are-not-graphemes}), use Unicode Locale 
+         Descriptions, or orthography profiles 
+         (see Chapter~\ref{orthography-profiles}).
    \item As a preferred file format, use Unicode Format UTF-8 in 

### 2015-06-20 chapters/ipa_meets_unicode.tex
adding tilde overlay and tie bar to strict IPA

@@
 symbols are counted or if the phonetic symbols commonly found in the literature
-are also included, see~\cite[Appendix~C]{Moran2012}. The 107 characters from 12 
+are also included, see~\cite[Appendix~C]{Moran2012}. The 157 characters from 12 
 code blocks shown here are the characters proposed for ``strict'' IPA encoding, 
@@
 	\dia{02B0} \dia{02BC} \dia{02D0} \dia{02D1} ˥ ˦ ˧ ˨ ˩ {\large ˈ ˌ}
-	\item \textsc{Combining Diacritical Marks} (23 characters) \newline 
+	\item \textsc{Combining Diacritical Marks} (25 characters) \newline 
 	\dia{033C} \dia{032A} \dia{033B} \dia{033A} \dia{031F} \dia{0320} \dia{031D} 
 	\dia{031E} \dia{0318} \dia{0319} \dia{031C} \dia{0339} \dia{032C} \dia{0325} 
-	\dia{0330} \dia{0324} \dia{0329} \dia{032F} \dia{0303} \dia{0308} \dia{033D} 
-	\dia{0306} \dia{031A} 
+	\dia{0330} \dia{0324} \dia{0329} \dia{032F} \dia{0334} \dia{0303} \dia{0308} 
+	\dia{033D} \dia{0306} \dia{031A} \ \dia{0361}◌
 	\item \textsc{Greek and Coptic} (3 characters) \newline 
@@
 ``velarized or pharyngealized'', using the <\dia{0334}> \textsc{combining tilde
-overlay} at \uni{0334}. The tilde overlay seems to be extremely rarely used. For
-strict IPA encoding we propose to not include the combining tilde overlay at
-all, but only use the other two special markers for velarization and
-phayngaealization.
+overlay} at \uni{0334}. The tilde overlay seems to be extremely rarely used. We 
+suggest to try and avoid using the tilde overlay, though for reasons of backward 
+compatibility we will allow it in strict-IPA.\@
 
@@
 a segment, but bind two graphemes together. This actually turns out to be rather
-different from the Unicode conceptions. The Unicode encoding of this character
+different from Unicode conceptions. The Unicode encoding of this character
 belong to the Combining Diacritical Marks, namely either \textsc{combining double
@@
 would have the tie bar linked both to the character in front and to the
-character following. The solution to this is to always use the tie bar in 
-combination with
+character following.
 
-Further, according to the
-spirit of the IPA, it would also be possible to combine more than two base
-characters into one tie bar, but this is not possible with Unicode (i.e.~there
-is no possibility to draw a tie bar over three of four characters). It is
-possible to indicate such larger groups by repeating the tie bar, like in
-<a͡o͡u>. If really necessary, we consider this possible, even though the
-rendering will never look good.
+Further, according to the spirit of the IPA, it would also be possible to
+combine more than two base characters into one tie bar, but this is not possible
+with Unicode (i.e.~there is no possibility to draw a tie bar over three of four
+characters). It is possible to indicate such larger groups by repeating the tie
+bar, like for a triphthong <a͡ʊ͡ə> in the English word \textit{hour}. If really
+necessary, we consider this possible, even though the rendering will never look
+good.
 
-Most importantly though, the tie-bar is unnecessary for computational processing
-given the design of orthography profiles, as described in
-Section~\ref{usecase-tokenization}. The principle idea of such processing is
-that groups of characters that belong together are defined in an orthography
-profile, making it superfluous to indicate them explicitly by tie bars. For this
-reason we suggest to remove the tie bar from any ``strict'' IPA encoding, and
-instead add descriptions of all so-called ``tailored grapheme clusters'' (see
-Section~\ref{the-unicode-approach}) to an orthography profile.
+Most importantly though, in comparison to normal Unicode processing, the tie-bar
+actually takes a reversed approach to complex graphemes. Basically, the Unicode
+principle (see Section~\ref{pitfall-characters-are-not-graphemes}) is that fixed
+sequences in a writing system have to be specified as ``tailored'' grapheme
+clusters. In case the sequence is not a cluster, this has to be explicitly
+indicated. IPA takes a different approach. In IPA by default different base
+letters are not connected into larger clusters; only when it is specified in the
+string itself (using the tie bar).
 
@@
 in IPA that allow for multiple, apparently identical, encodings that are not
-canonical equivalent (see Section~\ref{pitfall-canonical-equivalence}),. The
+canonical equivalent (see Section~\ref{pitfall-canonical-equivalence}). The
 following elements should not be treated as diacritics when encoding IPA in
@@
   \item[] <\dia{0337}> \textsc{combining short solidus overlay} at \uni{0337}
-  \item[] <\dia{0334}> \textsc{combining tilde overlay} at \uni{0334}
 \end{itemize}
@@
 There turn out to be a lot of characters in the IPA that could be conceived as
-using any of these elements, like <ɲ>, <ɳ>, <ɨ>, <ø> or <ɫ>. However, all such
+using any of these elements, like <ɲ>, <ɳ>, <ɨ> or <ø>. However, all such
 characters exist as well as precomposed combination in Unicode, and these
@@
 \end{itemize}
-
 All other combinations of vowels with rhotic hooks will have to be made by using
-<{\large ˞}> \textsc{modifier letter rhotic hook} at \uni{02DE}, because there
+<\dia{02DE}> \textsc{modifier letter rhotic hook} at \uni{02DE}, because there
 is not complete set of precomposed characters with rhotic hooks. For that reason
@@
 
+A similar situation arises with <\dia{0334}> \textsc{combining tilde overlay} at
+\uni{0334}. The main reason some phoneticians like to use this in IPA is to mark
+the ``dark'' <l> in English codas, using the character <ɫ> \textsc{latin small
+letter l with middle tilde} at \uni{026B}. This character is not canonically
+equivalent to the combination <l> + <\dia{0334}>, so one of the two possible
+encodings has to be chosen. Because the tilde overlay is described as a general
+mechanism by the IPA, we propose to use the separated <\dia{0334}>
+\textsc{combining tilde overlay} at \uni{0334}. However, note that phonetically 
+this seems to be (almost) superfluous (see Section~\ref{pitfall-multiple-options-ipa}) 
+and the typical usage in the form of <ɫ> is (almost) a homoglyph with <ɬ> (see 
+Section~\ref{pitfall-homoglyphs-in-IPA}). For these reasons we suggest to try 
+and avoid the tilde overlay.
+
 Reversely, note that the <ç> \textsc{latin small letter c with cedilla} at
-\uni{00E7} will be decomposed into <c> with <\dia{0327}> \textsc{combining
-cedilla} at \uni{0327} by Unicode canonical decomposition, also if such a
-decomposition is not intended in the IPA.\@
+\uni{00E7} is canonically equivalent with <c> with <\dia{0327}>
+\textsc{combining cedilla} at \uni{0327}, so it will be separated into two
+characters by Unicode canonical decomposition, also if such a decomposition is
+not intended in the IPA.\@ However, because of the nature of canonical
+equivalence (see Section~\ref{pitfall-canonical-equivalence}), these two
+encodings are completely identical in any computational treatment, so this
+decomposition does not have any practical consequences.
 
@@
 	\begin{itemize} 
-	  \item[] <\dia{02E5}> \textsc{modifier letter extra-high tone bar} at \uni{02E5}
-	  \item[] <\dia{02E8}> \textsc{modifier letter low tone bar} at \uni{02E8}
+	  \item[] <˥> \textsc{modifier letter extra-high tone bar} at \uni{02E5}
+	  \item[] <˨> \textsc{modifier letter low tone bar} at \uni{02E8}
 	  \item[] and others like this
@@
 for aspiration (more precisely post-aspiration following the base character) and
-pre-aspiration (preceding the base character). Note that there exists a
-mechanism in Unicode to force separate characters to be combined (namely by
-using the \textsc{zero width joiner} at \uni{200D}), but to us this seems to be
-a rather impractical, and probably not an enforceable solution to the general
-problem of complex graphemes. The default combination of Spacing Modifiers with
+pre-aspiration (preceding the base character). The default combination of Spacing Modifiers with
 the preceding character can be specified in orthography profiles (see
-Chapter~\ref{orthography-profiles}). Only in case variable directions occur in 
-the same transcription (e.g. both pre-aspiration and poast-aspiration occur), 
-then each case has to be explicitly marked with a zero-width joiner.
+Chapter~\ref{orthography-profiles}).
 
@@
 character.\footnote{See
-\url{http://unicode.org/reports/tr44/\#Canonical\_Combining\_Class\_Values} for
-a detailed description.} When applying a Unicode normalization (NFC or NFD, see
+\url{http://unicode.org/reports/tr44/#Canonical_Combining_Class_Values} for a
+detailed description.} When applying a Unicode normalization (NFC or NFD, see
 Section~\ref{pitfall-canonical-equivalence}), the diacritics in different
 positions are put in a specified order. This process therefore harmonizes the
-difference between different encodings, e.g.\ of an extra-short creaky voice vowel
-<ḛ̆>. This grapheme cluster can be encoded either as <e> + <\dia{0306}> +
-<\dia{0330}> or as <e> + <\dia{0330}> + <\dia{0306}>. To prevent this twofold encoding,
-the Unicode Standard specifies the second ordering as canonical (in this case,
-diacritics below are put before diacritics above).
+difference between different encodings, e.g.\ of an extra-short creaky voice
+vowel <ḛ̆>. This grapheme cluster can be encoded either as
+<e>+<\dia{0306}>+<\dia{0330}> or as <e>+<\dia{0330}>+<\dia{0306}>. To
+prevent this twofold encoding, the Unicode Standard specifies the second
+ordering as canonical (in this case, diacritics below are put before diacritics
+above).
 
@@
 conventions. The conventions are influenced by the linguistic literature, though
-some ad-hoc decisions had to be taken given the vast variability of phonological segments 
-described by linguists.
+some ad-hoc decisions had to be taken given the vast variability of phonological
+segments described by linguists.
 
-If a character sequence contains more than one diacritic below the base
-character, then the place features are applied first (linguolabial, dental,
-apical, laminal, advanced, retracted), followed by the manner features (raised,
-lowered, advanced and retracted tongue root), then
+By Unicode Canonical Combining Classes, the diacritics on top of a character,
+like <\dia{0334}>, always come before diacritics below, which in turn always
+come before diacritics above. We follow this order and add the other IPA
+diacritics (which are not diacritics in the Unicode sense) after these
+diacritics. Within these classes of diacritics there is no canonical ordering
+specified by Unicode, so we propose an explicit ordering here.
+
+Starting with the diacritics below: if a character sequence contains more than
+one diacritic below the base character, then the place features are applied
+first (linguolabial, dental, apical, laminal, advanced, retracted), followed by
+the manner features (raised, lowered, advanced and retracted tongue root), then
 secondary articulations (more round, less round), laryngeal settings (creaky,
@@
 
-[TO DO: different kind of proposals]
+TO DO: 
+[different kind of profiles]
 
 At the end of this chapter we have added a few longish tables summarizing all
-155 different unicode codepoints that form the basis of strict-IPA encoding
-(107 letters, 41 diacritics and 7 remaining symbols). Each of
+157 different unicode codepoints that form the basis of strict-IPA encoding
+(107 letters, 43 diacritics and 7 remaining symbols). Each of
 these tables shows a typical glyph, and then lists the Unicode Codepoint,
@@
 [ordering]
-[removed characters: tie bar, tilde overlay]
 [additions]
@@
         page~\pageref{tab:ipa_letters}.
-  \item \textsc{IPA diacritics} \newline The 41 different IPA diacritics and
+  \item \textsc{IPA diacritics} \newline The 43 different IPA diacritics and
         tone markers (both Unicode Modifier Letters and Combining Diacritical

### 2015-06-21 chapters/ipa_meets_unicode.tex

@@
 By Unicode Canonical Combining Classes, the diacritics on top of a character,
-like <\dia{0334}>, always come before diacritics below, which in turn always
-come before diacritics above. We follow this order and add the other IPA
-diacritics (which are not diacritics in the Unicode sense) after these
-diacritics. Within these classes of diacritics there is no canonical ordering
-specified by Unicode, so we propose an explicit ordering here.
+like <\dia{0334}> (Combining Class number 1), always come before diacritics
+below (Combining Class number 220), which in turn always come before diacritics
+above (Combining Class number 230), which in turn come before diacritics over
+multiple characters like the tie bar <\dia{0361}{\large\fontspec{CharisSIL}◌}>
+(Combining Class number 233). We follow this order, but add the other IPA
+diacritics (which are not diacritics in the Unicode sense) between diacritics
+below and the tie bar. Within all these classes of diacritics there is no canonical
+ordering specified by Unicode, so we propose an explicit ordering here.
 
@@
 	  \item[→] extra short <\dia{0306}>
-	  \item[→] no audible release <\dia{031A}>
+	  \item[→] no audible release <\dia{031A}\ >
  \end{itemize} \end{itemize}
@@
 	  \item[→] long <\dia{02D0}> \textbar{} half-long <\dia{02D1}>
-%	  \item[→] tone letters <˥ ˦ ˧ ˨ ˩>
 	\end{itemize}
@@
 
-\noindent Finally, Tone markers are put at the end of any such sequence:
+\noindent Finally, Tone markers are put at the end of any such sequence, and the 
+tie bar follows at the very end:
 
@@
     \item[→] tone letters <˥ ˦ ˧ ˨ ˩> \textbar{} tone movement <↑ ↓ ↗ ↘>
+    \item[→] tie bar <\dia{0361}{\large\fontspec{CharisSIL}◌}>
   \end{itemize}
@@
 At the end of this chapter we have added a few longish tables summarizing all
-157 different unicode codepoints that form the basis of strict-IPA encoding
-(107 letters, 43 diacritics and 7 remaining symbols). Each of
+159 different unicode codepoints that form the basis of strict-IPA encoding
+(107 letters, 45 diacritics and 7 remaining symbols). Each of
 these tables shows a typical glyph, and then lists the Unicode Codepoint,
@@
         page~\pageref{tab:ipa_letters}.
-  \item \textsc{IPA diacritics} \newline The 43 different IPA diacritics and
+  \item \textsc{IPA diacritics} \newline The 45 different IPA diacritics and
         tone markers (both Unicode Modifier Letters and Combining Diacritical

### 2015-06-21 chapters/orthography_profiles.tex

--- 2015-06-05 chapters/orthography_profiles.tex
@@
 \end{enumerate}
+
+\section{Examples}
+
+[Here should a few abstract short simple examples be added]
+
+\
+
+Note that to deal with ambiguous parsing cases, we can use the Unicode approach
+using the zero width joiner. This is actually a non-joiner (the name is
+confusing): the idea is to add this character into the text to identify cases in
+which a sequence of characters is not supposed to be a complex grapheme (even
+though the sequence is in the orthography profile)

### 2015-06-21 chapters/pitfalls.tex

@@
 in Unicode parlance). The combining grapheme joiner (CGJ) can be used to
-separate characters that are not supposed to form ligatures. To solve the issue
-of tailored grapheme clusters, Unicode offers some assistance in the form of the
-Unicode Locale Descriptions. However, in the practice of linguistic research,
-this is not a real solution. For that reason we propose to use orthography
-profiles (see Chapter~\ref{orthography-profiles}).
+separate characters that are not supposed to form ligatures.
+
+To solve the issue of tailored grapheme clusters, Unicode offers some assistance
+in the form of the Unicode Locale Descriptions. However, in the practice of
+linguistic research, this is not a real solution. For that reason we propose to
+use orthography profiles (see Chapter~\ref{orthography-profiles}). Basically,
+both orthography profiles and locale descriptions offer a way to specify
+tailored grapheme clusters. For example, for English one could specify that <sh>
+is such a cluster. Consequently, this sequence of characters is then always
+interpreted as a complex grapheme. For cases in which this is not the right
+decision, like in the English word \textit{mishap}, the \textsc{zero width
+joiner} at \uni{200D} has to be entered between <s> and <h>.
 

### 2015-06-21 tables/unicode_boundaries.tex

--- 2015-06-19 tables/unicode_boundaries.tex
@@
 \begin{center}
-\begin{xtabular}{ l L{1.2cm} L{5.6cm} L{4cm} }
+\begin{xtabular}{ l L{1.2cm} L{5.4cm} L{4cm} }
 {\large ˈ} & \unif{02C8} & \textsc{modifier letter vertical line} & primary stress \\
 {\large ˌ} & \unif{02CC} & \textsc{modifier letter low vertical line} & secondary stress \\
+ & \unif{0020} & \textsc{space} & word break \\
 {\large.} & \unif{002E} & \textsc{full stop} & syllable break \\

### 2015-06-22 chapters/ipa_meets_unicode.tex
updated handling of arrows, and changed codepoints
see wikipedia proposals: https://en.wikipedia.org/wiki/Upstep

@@
 blocks. With its diverse collection of symbols from various scripts and
-diacritics, the IPA is spread across 12 blocks in the Unicode
+diacritics, the IPA is spread across 13 blocks in the Unicode
 Standard:\footnote{This number of blocks depends on whether only IPA-sanctioned
 symbols are counted or if the phonetic symbols commonly found in the literature
-are also included, see~\cite[Appendix~C]{Moran2012}. The 157 characters from 12 
+are also included, see~\cite[Appendix~C]{Moran2012}. The 159 characters from 13 
 code blocks shown here are the characters proposed for ``strict'' IPA encoding, 
@@
 	\dia{207F} 
-	\item \textsc{Arrows} (4 characters) \newline 
-	↑ ↓ ↗ ↘
+	\item \textsc{Modifier Tone Letters} (2 characters) \newline
+	ꜛ ꜜ
+	\item \textsc{Arrows} (2 characters) \newline 
+	↗ ↘
 	\item \textsc{Latin Extended-C} (1 character): \newline
@@
 159 different unicode codepoints that form the basis of strict-IPA encoding
-(107 letters, 45 diacritics and 7 remaining symbols). Each of
+(107 letters, 41 diacritics and 11 remaining symbols). Each of
 these tables shows a typical glyph, and then lists the Unicode Codepoint,
@@
         page~\pageref{tab:ipa_letters}.
-  \item \textsc{IPA diacritics} \newline The 45 different IPA diacritics and
+  \item \textsc{IPA diacritics} \newline The 41 different IPA diacritics and
         tone markers (both Unicode Modifier Letters and Combining Diacritical
@@
         page~\pageref{tab:ipa_diacritics}.
-  \item \textsc{IPA remainders} \newline
-        The 7 remaining IPA symbols (boundary and stress markers) as allowed in
-        strict-IPA encoding are listed in Table~\ref{tab:ipa_boundaries} on
+  \item \textsc{IPA remainders} \newline The 11 remaining IPA symbols
+        (boundary, stress and intonation markers) as allowed in strict-IPA
+        encoding are listed in Table~\ref{tab:ipa_boundaries} on
         page~\pageref{tab:ipa_boundaries}. 

### 2015-06-22 tables/unicode_boundaries.tex
updated handling of arrows, and changed codepoints
see wikipedia proposals: https://en.wikipedia.org/wiki/Upstep

--- 2015-06-19 tables/unicode_boundaries.tex
@@
 \begin{center}
-\begin{xtabular}{ l L{1.2cm} L{5.6cm} L{4cm} }
+\begin{xtabular}{ l L{1.2cm} L{5.4cm} L{4cm} }
 {\large ˈ} & \unif{02C8} & \textsc{modifier letter vertical line} & primary stress \\
 {\large ˌ} & \unif{02CC} & \textsc{modifier letter low vertical line} & secondary stress \\
+{ꜛ} & \unif{A71B} & \textsc{modifier letter raised up arrow} & upstep \\
+{ꜜ} & \unif{A71C} & \textsc{modifier letter raised down arrow} & downstep \\
+{↗} & \unif{2197} & \textsc{north east arrow} & global rise \\
+{↘} & \unif{2198} & \textsc{south east arrow} & global fall \\
+ & \unif{0020} & \textsc{space} & word break \\
 {\large.} & \unif{002E} & \textsc{full stop} & syllable break \\

### 2015-06-22 tables/unicode_diacritics.tex
adding tilde overlay and tie bar to strict IPA
updated handling of arrows, and changed codepoints
see wikipedia proposals: https://en.wikipedia.org/wiki/Upstep

@@
 \begin{center}
-\begin{xtabular}{ l L{1.2cm} L{6.6cm} L{3cm} }
+\begin{xtabular}{ l L{1.1cm} L{6.2cm} L{3.6cm} }
+\dia{0334} & \unif{0334} & \textsc{combining tilde overlay} & velarized or pharyngealized \\
 \dia{033C} & \unif{033C} & \textsc{combining seagull below} & linguolabial \\
@@
 {˩} & \unif{02E9} & \textsc{modifier letter extra-low tone bar} & extra low tone \\
-{↑} & \unif{2191} & \textsc{upwards arrow} & upstep \\
-{↓} & \unif{2193} & \textsc{downwards arrow} & downstep \\
-{↗} & \unif{2197} & \textsc{north east arrow} & global rise \\
-{↘} & \unif{2198} & \textsc{south east arrow} & global fall \\
+\dia{0361}{\large\fontspec{CharisSIL}◌} & \unif{0361} & \textsc{combining double inverted breve} & tie bar \\
 \end{xtabular}

### 2015-06-22 tables/unicode_boundaries.tex
minor layout update

--- 2015-06-21 tables/unicode_boundaries.tex
@@
 {\large ˌ} & \unif{02CC} & \textsc{modifier letter low vertical line} & secondary stress \\
+{\large\fontspec{CharisSIL}ꜛ} & \unif{A71B} & \textsc{modifier letter raised up arrow} & upstep \\
+{\large\fontspec{CharisSIL}ꜜ} & \unif{A71C} & \textsc{modifier letter raised down arrow} & downstep \\
+{↗} & \unif{2197} & \textsc{north east arrow} & global rise \\
+{↘} & \unif{2198} & \textsc{south east arrow} & global fall \\
  & \unif{0020} & \textsc{space} & word break \\

### 2015-06-22 chapters/ipa_meets_unicode.tex
minor layout update
another minor layout update
update tone handling and reverse arrow code points

--- 2015-06-21 chapters/ipa_meets_unicode.tex
@@
 symbols are counted or if the phonetic symbols commonly found in the literature
-are also included, see~\cite[Appendix~C]{Moran2012}. The 157 characters from 12 
+are also included, see~\cite[Appendix~C]{Moran2012}. The 159 characters from 12 
 code blocks shown here are the characters proposed for ``strict'' IPA encoding, 
@@
 	ǀ ǁ ǂ ǃ
+	\item \textsc{Latin Extended-C} (1 character): \newline
+	\charis{ⱱ}
 	\item \textsc{IPA Extensions} (67 characters) \newline 
@@
 	ɵ ɶ ɸ ɹ ɺ ɻ ɽ ɾ ʀ ʁ ʂ ʃ ʄ ʈ ʉ ʊ ʋ ʌ ʍ ʎ ʏ ʐ ʑ ʒ ʔ ʕ ʘ ʙ ʛ ʜ ʝ ʟ ʡ ʢ 
+	\item \textsc{Greek and Coptic} (3 characters) \newline 
+	β θ χ
 	\item \textsc{Spacing Modifier Letters} (17 characters) \newline
@@
 	\dia{02B0} \dia{02BC} \dia{02D0} \dia{02D1} ˥ ˦ ˧ ˨ ˩ {\large ˈ ˌ}
+	\item \textsc{Superscripts and Subscripts} (1 character) \newline
+	\dia{207F} 
 	\item \textsc{Combining Diacritical Marks} (25 characters) \newline 
@@
 	\dia{0330} \dia{0324} \dia{0329} \dia{032F} \dia{0334} \dia{0303} \dia{0308} 
-	\dia{033D} \dia{0306} \dia{031A} \ \dia{0361}◌
-	\item \textsc{Greek and Coptic} (3 characters) \newline 
-	β θ χ
+	\dia{033D} \dia{0306} \dia{031A} \ \dia{0361}{\large\fontspec{CharisSIL}◌}
     \item \textsc{General Punctuation} (2 characters) \newline 
     ‖ \charis{‿}
-	\item \textsc{Superscripts and Subscripts} (1 character) \newline
-	\dia{207F} 
+%	\item \textsc{Modifier Tone Letters} (2 characters) \newline
+%	{\large\fontspec{CharisSIL}ꜛ ꜜ}
 	\item \textsc{Arrows} (4 characters) \newline 
 	↑ ↓ ↗ ↘
-	\item \textsc{Latin Extended-C} (1 character): \newline
-	\charis{ⱱ}
+
 \end{itemize}
@@
 
-\noindent Finally, Tone markers are put at the end of any such sequence, and the 
-tie bar follows at the very end:
+\noindent Finally, the tie bar follows at the very end of any such sequence:
 
 \begin{itemize}
-  \item[] \textsc{Tone Marks:}
+  \item[] \textsc{Tie bar:}
   \begin{itemize}
-    \item[→] tone letters <˥ ˦ ˧ ˨ ˩> \textbar{} tone movement <↑ ↓ ↗ ↘>
+%    \item[→] tone letters <˥ ˦ ˧ ˨ ˩>
     \item[→] tie bar <\dia{0361}{\large\fontspec{CharisSIL}◌}>
@@
 159 different unicode codepoints that form the basis of strict-IPA encoding
-(107 letters, 45 diacritics and 7 remaining symbols). Each of
+(107 letters, 36 diacritics and 16 remaining symbols). Each of
 these tables shows a typical glyph, and then lists the Unicode Codepoint,
@@
 [ordering]
+[boundaries, stress, intonation]
 [additions]
@@
         page~\pageref{tab:ipa_letters}.
-  \item \textsc{IPA diacritics} \newline The 45 different IPA diacritics and
+  \item \textsc{IPA diacritics} \newline The 36 different IPA diacritics and
         tone markers (both Unicode Modifier Letters and Combining Diacritical
@@
         page~\pageref{tab:ipa_diacritics}.
-  \item \textsc{IPA remainders} \newline
-        The 7 remaining IPA symbols (boundary and stress markers) as allowed in
-        strict-IPA encoding are listed in Table~\ref{tab:ipa_boundaries} on
-        page~\pageref{tab:ipa_boundaries}. 
+  \item \textsc{IPA remainders} \newline The 16 remaining IPA symbols
+        (boundary, stress, tone letters and intonation markers) as allowed in strict-IPA
+        encoding are listed in Table~\ref{tab:ipa_leftovers} on
+        page~\pageref{tab:ipa_leftovers}. 
   \item \textsc{IPA additions} \newline

### 2015-06-22 tables/unicode_boundaries.tex
update tone handling and reverse arrow code points

--- 2015-06-21 tables/unicode_boundaries.tex
@@
-\tablecaption{Strict-IPA boundaries with Unicode encodings}\label{tab:ipa_boundaries}
+\tablecaption{Other Strict-IPA symbols with Unicode encodings}\label{tab:ipa_leftovers}
  \tablefirsthead{
@@
      \captionsize\tablename\ \thetable{} 
-     Strict-IPA boundaries with Unicode encodings --- \textit{continued}
+     Other Strict-IPA symbols with Unicode encodings --- \textit{continued}
      } \\
@@
 {\large ˌ} & \unif{02CC} & \textsc{modifier letter low vertical line} & secondary stress \\
+{˥} & \unif{02E5} & \textsc{modifier letter extra-high tone bar} & extra high tone \\
+{˦} & \unif{02E6} & \textsc{modifier letter high tone bar} & high tone \\
+{˧} & \unif{02E7} & \textsc{modifier letter mid tone bar} & mid tone \\
+{˨} & \unif{02E8} & \textsc{modifier letter low tone bar} & low tone \\
+{˩} & \unif{02E9} & \textsc{modifier letter extra-low tone bar} & extra low tone \\
+%{\large\fontspec{CharisSIL}ꜛ} & \unif{A71B} & \textsc{modifier letter raised up arrow} & upstep \\
+%{\large\fontspec{CharisSIL}ꜜ} & \unif{A71C} & \textsc{modifier letter raised down arrow} & downstep \\
+{↑} & \unif{2191} & \textsc{upwards arrow} & global rise \\
+{↓} & \unif{2193} & \textsc{downwards arrow} & global fall \\
+{↗} & \unif{2197} & \textsc{north east arrow} & global rise \\
+{↘} & \unif{2198} & \textsc{south east arrow} & global fall \\
  & \unif{0020} & \textsc{space} & word break \\

### 2015-06-22 chapters/ipa_meets_unicode.tex
adding (lax) IPA sanctioned symbols

@@
 [boundaries, stress, intonation]
+[laxIPA]
 [additions]
@@
         page~\pageref{tab:ipa_leftovers}.
-  \item \textsc{IPA additions} \newline
+  \item \textsc{IPA alternatives} \newline The 16 additional symbols as allowed in
+        lax-IPA encoding are listed in Table~\ref{tab:ipa_lax} on
+        page~\pageref{tab:ipa_lax}.     
+  \item \textsc{optional IPA additions} \newline
         The 10 proposed additions to the IPA are listed in
@@
 \input{tables/unicode_boundaries}
+\input{tables/unicode_lax}
 \input{tables/unicode_additions}

### 2015-06-22 tables/unicode_lax.tex
adding (lax) IPA sanctioned symbols

\tablecaption{Alternatives within IPA with Unicode encodings}\label{tab:ipa_lax}
 \tablefirsthead{
   \toprule
   & Code & Unicode name & Phonetic description \\ 
   \midrule
   }
 \tablehead{
   \multicolumn{4}{c}{
     \captionsize\tablename\ \thetable{} 
     Alternatives within IPA with Unicode encodings --- \textit{continued}
     } \\
   \toprule
   & Code & Unicode name & Phonetic description \\ 
   \midrule
   }
 \tabletail{
   \bottomrule
   \multicolumn{4}{r}{
     \captionsize\textit{continued on next page}
     } \\
   }
 \tablelasttail{\bottomrule}

\begin{center}
\begin{xtabular}{ l L{1.2cm} L{6.1cm} L{3.5cm} }
g & \unif{0067} & \textsc{latin small letter g} & voiced velar plosive \\
\dia{030A} & \unif{030A} & \textsc{combining ring above} & voiceless (above) \\
\dia{030B} & \unif{030B} & \textsc{combining double acute accent} & extra high tone \\
\dia{0301} & \unif{0301} & \textsc{combining acute accent} & high tone \\
\dia{0304} & \unif{0304} & \textsc{combining macron} & mid tone \\
\dia{0300} & \unif{0300} & \textsc{combining grave accent} & low tone \\
\dia{030F} & \unif{030F} & \textsc{combining double grave accent} & extra low tone \\
\dia{0302} & \unif{0302} & \textsc{combining circumflex accent} & falling \\
\dia{030C} & \unif{030C} & \textsc{combining caron} & rising \\
\dia{1DC4} & \unif{1DC4} & \textsc{combining macron-acute} & high rising \\
\dia{1DC5} & \unif{1DC5} & \textsc{combining grave-macron} & low rising \\
\dia{1DC6} & \unif{1DC6} & \textsc{combining macron-grave} & low falling \\
\dia{1DC7} & \unif{1DC7} & \textsc{combining acute-macron} & high falling \\
\dia{1DC8} & \unif{1DC8} & \textsc{combining grave-acute-grave} & rising-falling \\
\dia{1DC9} & \unif{1DC9} & \textsc{combining acute-grave-acute} & falling-rising \\ 
\dia{035C}{\large\fontspec{CharisSIL}◌} & \unif{035C} & \textsc{combining double breve below} & tie bar (below) \\
\end{xtabular}
\end{center}

### 2015-06-23 chapters/ipa_meets_unicode.tex
edits

@@
 IPA characters electronically. Similar to the previous chapter, this chapter
-will discuss various pitfalls with the Unicode encoding of the IPA.\@ We will
+will discuss various pitfalls with the encoding of IPA in the Unicode Standard.\@ We will
 specifically refer to the 2005 version of the IPA \citep{IPA2005} and the 7.0 
@@
 
-There are a few pitfalls to be aware of when using the Unicode Standard to
+There are several pitfalls to be aware of when using the Unicode Standard to
 encode IPA.\@ As we have said before, from a linguistic perspective it might
@@
 does place linguistic traditions above other orthographic possibilities. In
-general, we strongly suggest linguists not to complain about any decisions in
+general, we strongly suggest to linguists to not complain about any decisions in
 the Unicode Standard, but to try and understand the rationale of the Unicode
@@
 writing system. For this reason a special block with code points, called
-\textsc{IPA Extensions} was included already in the first version of the Unicode
+\textsc{IPA Extensions} was already included in the first version of the Unicode
 Standard (Version 1.0 from 1991).

### 2015-06-23 chapters/ipa_meets_unicode.tex
First draft of IPA recommendations

@@
 
-Summarizing the pitfalls as discussed in this chapter, we propose to define a 
-``strict'' IPA encoding to be used when interoperability of phonetic resources 
-is intended. For regular research papers, when only the visual outline is 
-important, the details encoding of IPA in electronic format is rather 
-unimportant, as long as it looks as expected. So for those regular usage these 
-proposals are irrelevant.
+Summarizing the pitfalls as discussed in this chapter, we propose to define 
+three different IPA encodings: strict-IPA, valid-IPA and widened-IPA.\@ These 
+three encodings are subsets of each other, i.e.~strict-IPA is more restricted 
+than valid-IPA, which it turn is more restricted than widened-IPA.\@ Informally 
+speaking, valid-IPA represents the current state of the IPA \citep{IPA2005}. 
+Strict-IPA represents a more constrained version of IPA, while widened-IPA is a 
+slightly extended version of IPA, allowing a few more symbols.
 
-TO DO: 
-[different kind of profiles]
+\
 
-At the end of this chapter we have added a few longish tables summarizing all
+\noindent Strict-IPA encoding is supposed to be used when interoperability of
+phonetic resources is intended. It is a strongly constrained subset of IPA
+geared towards uniqueness of encoding. Ideally, for each transcription there
+should be exactly one possible strict-IPA encoding. For each phonetic feature
+there is only one possibility (see Section~\ref{pitfall-multiple-options-ipa})
+and the IPA diacritics are forced into a canonical ordering (see
+Section~\ref{pitfall-no-unique-diacritic-ordering}).
+
+Valid-IPA does allow alternative symbols with the same phonetic meaning, as 
+specified in the official IPA specifications. Also, valid-IPA does not enforce a 
+specific ordering of diacritics, because the IPA does not propose any such 
+ordering. This means that in valid-IPA the same phonetic intention can be 
+encoded in multiple ways. This is sufficient for phonetically trained human 
+eyes, but it is not sufficient for automatic interoperability.
+
+Finally, widened-IPA includes a few more symbols which seem to be useful for
+various special cases (see Section~\ref{ipa-additions}).
+
+\
+
+\noindent At the end of this chapter we have added a few longish tables summarizing all
 159 different unicode codepoints that form the basis of strict-IPA encoding
@@
 these tables shows a typical glyph, and then lists the Unicode Codepoint,
-Unicode Name and IPA description for each symbol.
+Unicode Name and IPA description for each symbol. Further, there is a table with 
+the additional options for valid-IPA and a table with the additional options for 
+widened-IPA.\@
 
-TO DO:
-[ordering]
-[boundaries, stress, intonation]
-[laxIPA]
-[additions]
+textbf{NOTE: I HAVE ADDED VARIOUS TONE DIACRITICS TO VALID-IPA THAT ARE NOT IN
+THE OFFICIAL IPA SUMMARY. SHOULD THEY BE INCLUDED?}
 
@@
 
-  \item \textsc{IPA letters} \newline
+  \item \textsc{strict-IPA letters} \newline
         The 107 different IPA letters as allowed in strict-IPA encoding are
@@
         page~\pageref{tab:ipa_letters}.
-  \item \textsc{IPA diacritics} \newline The 36 different IPA diacritics and
+  \item \textsc{strict-IPA diacritics} \newline The 36 different IPA diacritics and
         tone markers (both Unicode Modifier Letters and Combining Diacritical
@@
         page~\pageref{tab:ipa_diacritics}.
-  \item \textsc{IPA remainders} \newline The 16 remaining IPA symbols
+  \item \textsc{strict-IPA remainders} \newline The 16 remaining IPA symbols
         (boundary, stress, tone letters and intonation markers) as allowed in strict-IPA
@@
         page~\pageref{tab:ipa_leftovers}.
-  \item \textsc{IPA alternatives} \newline The 16 additional symbols as allowed in
-        lax-IPA encoding are listed in Table~\ref{tab:ipa_lax} on
+  \item \textsc{valid-IPA additions} \newline The 16 additional symbols as allowed in
+        valid-IPA encoding are listed in Table~\ref{tab:ipa_lax} on
         page~\pageref{tab:ipa_lax}.     
-  \item \textsc{optional IPA additions} \newline
+  \item \textsc{widened-IPA additions} \newline
         The 10 proposed additions to the IPA are listed in
@@
 \input{tables/unicode_diacritics}
-\input{tables/unicode_boundaries}
-\input{tables/unicode_lax}
+\input{tables/unicode_remaining}
+\input{tables/unicode_valid}
 \input{tables/unicode_additions}

### 2015-06-23 tables/unicode_valid.tex
First draft of IPA recommendations

--- 2015-06-22 tables/unicode_lax.tex
@@
-\tablecaption{Alternatives within IPA with Unicode encodings}\label{tab:ipa_lax}
+\tablecaption{Additional characters for valid-IPA with Unicode encodings}\label{tab:ipa_lax}
  \tablefirsthead{
@@
      \captionsize\tablename\ \thetable{} 
-     Alternatives within IPA with Unicode encodings --- \textit{continued}
+     Additional characters for valid-IPA with Unicode encodings --- \textit{continued}
      } \\
@@
 \begin{xtabular}{ l L{1.2cm} L{6.1cm} L{3.5cm} }
+\dia{030A} & \unif{030A} & \textsc{combining ring above} & voiceless (above) \\
 g & \unif{0067} & \textsc{latin small letter g} & voiced velar plosive \\
-\dia{030A} & \unif{030A} & \textsc{combining ring above} & voiceless (above) \\
 \dia{030B} & \unif{030B} & \textsc{combining double acute accent} & extra high tone \\

### 2015-07-06 chapters/pitfalls.tex
updated bib; updated pitfalls text

@@
 \textsc{latin small letter g} at \uni{0067}). In contrast, the symbols
-<{\fontspec{EversonMono}AАΑᎪᗅᴀꓮ𐊠𝖠𝙰}> are all different code points,
+<{\fontspec{EversonMono}AАΑᎪᗅᴀꓮʠ햠홰}> are all different code points,
 although they look highly similar---in some cases even sharing exactly the same
@@
 %	\item[] <{\fontspec{EversonMono}Ａ}> \textsc{fullwidth latin capital letter a}, at \uni{FF21} 
-	\item[] <{\fontspec{EversonMono}𐊠}> \textsc{carian letter a}, at \uni{102A0} 
-%	\item[] <{\fontspec{EversonMono}𐌀}> \textsc{old italic letter a}, at \uni{10300} 
-	\item[] <{\fontspec{EversonMono}𝖠}> \textsc{mathematical sans-serif capital a}, \uni{1D5A0} 
-	\item[] <{\fontspec{EversonMono}𝙰}> \textsc{mathematical monospace capital a}, at \uni{1D670} 
+	\item[] <{\fontspec{EversonMono}ʠ}> \textsc{carian letter a}, at \uni{102A0} 
+%	\item[] <{\fontspec{EversonMono}̀}> \textsc{old italic letter a}, at \uni{10300} 
+	\item[] <{\fontspec{EversonMono}햠}> \textsc{mathematical sans-serif capital a}, \uni{1D5A0} 
+	\item[] <{\fontspec{EversonMono}홰}> \textsc{mathematical monospace capital a}, at \uni{1D670} 
 \end{itemize}
@@
 % ==========================
+\section{Pitfall: Software}
+\label{software}
+% ==========================
+
+% issues we've encountered in our computing environments
+%  save as UTF-8
+%  conversion between software programs (or when is something really UTF-8?)
+%  NFC / NFD in copy & paste(!)
+
+% ==========================
 \section{Recommendations}

### 2015-07-06 chapters/ipa_meets_unicode.tex
adds rich history of IPA and how it was incorporated into Unicode

@@
 
-% ==========================
-\section{The International Phonetic Alphabet (IPA)}
-\label{the-international-phonetic-alphabet}
-% ==========================
-
 The International Phonetic Alphabet (IPA) is a common standard in linguistics to
-transcribe sounds of spoken language into Latin-based characters. Although IPA
-is reasonably easily adhered to with pen and paper, it is not trivial to encode
-IPA characters electronically. Similar to the previous chapter, this chapter
-will discuss various pitfalls with the encoding of IPA in the Unicode Standard.\@ We will
+transcribe sounds of spoken language into discrete segments using a Latin-based 
+alphabet. Although IPA is reasonably easily adhered to with pen and paper, it is not trivial to encode
+IPA characters electronically. Similar to the previous chapter, in this chapter
+we discuss various pitfalls with the encoding of IPA in the Unicode Standard.\@ We will
 specifically refer to the 2005 version of the IPA \citep{IPA2005} and the 7.0 
@@
 
-\
-
-\noindent For a long time, linguists, like all other computer users, were
-limited to ASCII-encoded 7-bit characters, which only includes Latin characters,
-numbers and some punctuation and symbols. Restricted to these standard character
-sets that lacked IPA support or other language-specific graphemes that they
-needed, linguists devised their own solutions.\footnote{Early work addressing
-the need for a universal computing environment for writing systems and their
-computational complexity is discussed in \citet{Simons1989}. A survey of
-practical recommendations for language resources, including notes on encoding,
-can be found in \citet{BirdSimons2003}} For example, some chose to represent
-unavailable graphemes with substitutes, e.g.~the combination of <ng> to
-represent <ŋ>. Tech-savvy linguists redefined selected characters from a
-character encoding by mapping custom made fonts to those code points. However,
-one linguist's electronic text would not render properly on another linguist's
-computer without access to the same font. Further, if two character encodings
-defined two character sets differently, then data could not be reliably and
-correctly displayed. This is a common example of the non-interoperability of
-data and data formats.
-
-To alleviate this problem, during the late 1980s, SAMPA (Speech Assessment
-Methods Phonetic Alphabet) was created to represent IPA symbols with ASCII
-character sequences, e.g. <p\textbackslash> for [ɸ]. Two problems with SAMPA are
-that (i) it is only a partial encoding of the IPA and (ii) it encodes different
-languages in separate data tables, instead of using a universal alphabet, like
-IPA.\@ SAMPA tables are derived from phonemes appearing in several European
-languages that were developed as part of a European Commission-funded project to
-address technical problems like electronic mail exchange (what is now simply
-called email). SAMPA is essentially a hack to work around displaying IPA
-characters, but it provided speech technology and other fields a basis that has
-been widely adopted and used in code. So, SAMPA was a collection of tables to be
-compared, instead of a large universal table representing all languages. An
-extended version of SAMPA, called X-SAMPA, set out to include every symbol in
-the IPA chart including all diacritics \citep{WellsND}. X-SAMPA was considered
-more universally applicable because it consisted of one table that encoded all
-characters that represented phones/segments in IPA across languages. SAMPA and
-X-SAMPA have been widely used for speech technology and as an encoding system in
-computational linguistics. Eventually, ASCII-encoding of the IPA became
-deprecated through the advent of the Unicode Standard. Note however that many
-popular software packages used for linguistic analyses still require ASCII
-input, e.g.~RuG/L04 and SplitsTree4.\footnote{See
-\url{http://www.let.rug.nl/kleiweg/L04/} and \url{http://www.splitstree.org/},
-respectively}
-
 There are several pitfalls to be aware of when using the Unicode Standard to
@@
 to try to be as consistent as possible across a wide range of use cases, and it
-does place linguistic traditions above other orthographic possibilities. In
-general, we strongly suggest to linguists to not complain about any decisions in
-the Unicode Standard, but to try and understand the rationale of the Unicode
-Consortium (which in our experience is almost always well-conceived) and devise
+does place linguistic traditions above other orthographic possibilities. Furthermore, 
+when we look at the history of how the IPA meet Unicode, we see that many of the 
+decisions for IPA symbols in the Unicode Standard come directly from the 
+International Phonetic Association itself. Therefore as users, many pitfalls that we 
+encounter have their grounding in the historical principles of the IPA, as 
+well as the technological considerations involved in creating a single 
+multilingual computing encoding.
+
+In general, we strongly suggest to linguists to not complain about any decisions in
+the Unicode Standard, but to try and understand the rationale of the International Phonetic Association 
+and the Unicode Consortium (which in our experience is almost always well-conceived) and devise
 ways to work with any unexpected behavior. Many of the current problems derive
@@
 Yet, the current situation, however unfortunate, is unlikely to change, so as
-linguists we will have to learn to deal with the specific pitfalls of IPA within
-the Unicode Standard. In this section, we will describe these pitfalls in some
-detail.
+linguists we must learn to deal with the specific pitfalls of IPA within
+the Unicode Standard.
+
+In the following sections, we will describe these pitfalls 
+in detail. But first we present a brief history of IPA, its principles, 
+its computational representation and the need for and challenge in creating 
+a single multilingual computer encoding, and then how IPA meet Unicode.
+
+% ==========================
+\section{The International Phonetic Alphabet (IPA)}
+\label{the-international-phonetic-alphabet}
+% ==========================
+
+\subsubsection*{Brief history}
+
+Established in 1886, the International Phonetic Association (henceforth Association) 
+has long maintained a standard alphabet, the International Phonetic Alphabet (IPA),
+\footnote{Also referred to as API, for \textit{Association Phonétique Internationale}.} 
+which is a common standard in linguistics to transcribe sounds of spoken languages. 
+It was first published in 1888 as an international system of phonetic transcription 
+for oral languages and for pedagogical purposes. The draft of 1887 contained phonetic 
+values for English, French and German. Diacritics for length and nasalization were 
+already present in this draft and the same symbols are still used today.
+
+Although the original IPA alphabet was a list of symbols with pronunciation examples 
+from words in different languages, in 1900 the symbols were organized into chart and 
+given phonetic feature labels for consonants (e.g.\ for manner of articulation: `plosives', 
+`nasales', `fricatives'; for place of articulation: `bronchials', `laryngeales', `labiales') 
+and for vowels (`fermeées', `mi-fermeées', `mi-overtes', `overtes'). Over the last century 
+or so, the structure of the chart has changed with increases in phonetic knowledge, and 
+thus like notational systems in other scientific disciplines, the IPA reflects facts and 
+theories (of phonetic knowledge) that have developed over time. It is natural then that 
+the IPA is modified occasionally to accommodate scientific innovations and discoveries. 
+(In fact, this is part of the Association's mandate.) These changes are captured in 
+the revisions to the IPA.\footnote{For a detailed history see: 
+\url{https://en.wikipedia.org/wiki/History\_of\_the\_International\_Phonetic_Alphabet}}
+
+Over the years there has been several revisions, many minor. For example, 
+articulation labels -- what we now call features even though the IPA avoids this term -- 
+have changed (e.g.\ `lips', `throat', `rolled'). Phonetic symbol values have changed 
+(e.g.\ voiceless is no longer marked by <h>). Symbols have been dropped (e.g.\ the 
+caret diacritic denoting `long and narrow' is longer used). And many symbols have 
+been added to reflect contrastive sounds found in the world's very diverse 
+phonological systems.
+
+Although IPA began its life as a pedagogical tool, from its earliest days the 
+Association aimed to provide ``a separate sign for each distinctive sound; 
+that is, for each sound which, being used instead of another, in the same 
+language, can change the meaning of a word'' \citep[27]{IPA1999}. Distinctive 
+sounds became later known as \textsc{phonemes} and the IPA has developed historically 
+into a notational devise with a strictly segmented phonemic view. A phoneme is an 
+abstract theoretical notion derived from an acoustic signal as produced by speakers 
+in the real world. Therefore the IPA contains a number of theoretical assumptions 
+about speech and how to analyze speech in written form.
+
+Phonetic analysis is based on two premises: that it is possible to describe 
+the acoustic speech signal (i.e.\ sound waves) in terms of sequentially ordered 
+discrete segments, and, that each segment can be characterized by an articulatory 
+target.\footnote{A purely phonetic description is only derivable from instrumental 
+data in high quality sound recordings. Once spoken language data are segmented, 
+phonological consideration inextricably play a roll in transcription. In other 
+words, phonetic observations beyond quantitative acoustic analysis are made in 
+terms of some phonological framework.} Today, the IPA chart reflects a linguistic 
+basis grounded in principles of phonological contrast. This fact is stated 
+explicitly in several places, including in the \textit{Report on the 1989 Kiel 
+convention} published in the \textit{Journal of the International Phonetic 
+Association} \citep[67-68]{International1989report}:
+
+\begin{quote}
+The IPA is intended to be a set of symbols for representing all the possible 
+sounds of the world's languages. The representation of these sounds uses a set 
+of phonetic categories which describe how each sound is made. These categories 
+define a number of natural classes of sounds that operate in phonological rules 
+and historical sound changes. The symbols of the IPA are shorthand ways of 
+indicating certain intersections of these categories.
+\end{quote}
+
+\noindent and in the \textit{Handbook of the International Phonetic Association} \citep[18]{IPA1993}:
+
+\begin{quote}
+% The general value of the symbols in the chart is listed below. In each case 
+... a symbol can be regarded as a shorthand equivalent to a phonetic description, 
+and a way of representing the contrasting sounds that occur in a language. Thus 
+[m] is equivalent to `voiced bilabial nasal', and is also a way of representing 
+one of the contrasting nasal sounds that occur in English and other languages. 
+[...] When a symbol is said to be suitable for the representation of sounds in 
+two languages, it does not necessarily mean that the sounds in the two languages 
+are identical.
+\end{quote}
+
+\noindent Although the IPA provides symbols to unambiguously represent phonemes, 
+it also aims to represent phonetic details. Since phonetic detail could potentially 
+include things like `deep voice', the IPA restricts phonetic detail to linguistically 
+relevant aspects of speech. These principles and conventions for using the IPA are 
+outlined in the Association's Handbook.
+
+\subsubsection*{Principles}
+
+IPA transcription has essentially two parts. The first is a text containing IPA 
+symbols and the second is a set of conventions (rules) for interpreting those 
+symbols (and their combinations). The IPA is designed to meet practical linguistic 
+needs and is used to transcribe the phonetic or phonological structure of languages. 
+It is also used increasingly as a foreign language learning tool, as a standard 
+pronunciation guide and as a tool for creating practical orthographies of 
+previously unwritten languages.
+
+The current construction and use of the IPA are guided by principles outlined 
+in the \textit{Handbook of the International Phonetic Association: A guide 
+to the use of the International Phonetic Alphabet} \citep[159]{IPA1999}, henceforth 
+HB. The use of the symbols to represent a language's phonological system is guided 
+by the principle of contrast; where two words are distinguishable by phonetic contrast, 
+those contrasts should be transcribed with different symbols (graphemes not diacritics). 
+Allophonic distinction falls under the rubric of diacritically-distinguished symbols, 
+e.g. [stop̚] vs [spʰot]. In sum:
+
+\begin{itemize}
+	\item different symbols (without diacritics) should be used whenever a language employs two contrastive sounds
+	\item when two sounds in a language are not known to be contrastive, the same symbol should be used to represent these sounds; however, diacritics may be used to distinguish such sounds when necessary
+	\item diacritics cannot be dispensed with entirely, so the Association recommends to limit their use to:
+	\begin{itemize}
+		\item denoting length, stress and pitch
+		\item representing minute shades of sounds
+		\item obviating the design of a (large) number of new symbols when a single diacritic suffices (e.g. nasalized vowels, aspirated stops)
+	\end{itemize}
+\end{itemize}
+
+\noindent Thus, an IPA transcription \textit{always} consists of ``a set of 
+symbols and a set of conventions for their interpretation''.
+
+In systematic transcription (as opposed to impressionistic), there is a division 
+between phonemic and allophonic transcription. The terms phoneme and allophone 
+contain theoretical baggage, but the basic goal of a phonemic transcription is to 
+distinguish all words in a language with the minimal number of transcription 
+symbols \citep[19]{Abercrombie1964}. Allophonic transcription uses a broader 
+set of distinct symbols to describe systematic allophonic differences in sounds 
+in words. These two systematic transcriptions are related to each other by a 
+set of conventions in the IPA tradition, so that they can be converted between 
+one and another.
+
+An IPA transcription is connected to a speech event by a set of conventions. 
+A phonetic (or impressionistic) transcription may use the conventions implicit 
+in the IPA chart, i.e. the transcriber can indicate that the phonetic value of 
+<ŋ͡m> is a simultaneous labial and velar closure which is voiced and contains 
+nasal airflow.
+
+A phonemic transcription includes the conventions of a particular language's 
+phonological rules. These rules determine the realization of that language's 
+phonemes. However, there can be different systems of phonemic transcription 
+for the same variety of a language. The differences may result from the fact 
+that more than one phonetic symbol may be appropriate for a phoneme (see Section 
+\ref{}). Or the differences may be due to different phonemic analyses, e.g.\ 
+Standard German's vowel system is arguably contrastive in length or tenseness.
+
+An important principle of the IPA is that different representations resulting 
+from different symbols and different analyses are in line with the IPA's aims. 
+In other words, the IPA does not provide phonological analyses for specific 
+languages and the IPA does not define a single ``correct'' transcription system. 
+Rather, the IPA aims to provide a resource that allows users to express any 
+analysis so that it is widely understood. Thus the IPA suits many linguists' 
+needs because:
+
+\begin{itemize}
+	\item it is intended to be a set of symbols for representing all possible sounds in the world's (spoken) languages
+	\item its chart has a linguistic basis (specifically a phonological bias) rather than a general phonetic notation scheme
+	\item its symbols can be used to represent distinctive feature combinations\footnote{Although the chart uses traditional manner and place of articulation labels, the symbols can nevertheless represent any defined bundle of features, binary or otherwise, to define phonetic dimensions.}
+	\item its chart provides a summary of linguists' agreed-upon phonetic knowledge (a common denominator of phonological ``facts'')
+\end{itemize}
+
+Several styles of transcription with IPA are possible and the HB illustrates 
+these and notes that they are all valid (see the 29 languages and their transcriptions 
+in the original and initial \textit{Illustrations of the IPA} \citep[41--154]{IPA2007}). 
+Therefore, there are different but equivalent transcriptions, or as \cite[64]{Ladefoged1990a} 
+captures it, ``Perhaps now that the Association has been explicit in its eclectic 
+approach, outsiders to the Association will no longer speak of \textit{the} IPA 
+transcription of a given phenomenon, as if there were only one approved style.''.
+
+Clearly not all phoneticians agree (or will likely ever agree of course) on all 
+aspects of the IPA or on transcription practices. As noted above, there have 
+been several revisions in the IPA's long history, but the current version (2005) 
+is strikingly similar to the 1926 version. In 1989 an IPA revision convention 
+was held in Kiel, Germany. As per other revisions, there was expansion and revisions 
+to phonetic symbols in the IPA chart. Notably the marking of tone was extended 
+with the addition of a second system for marking linguistic tones (Chao tones). 
+Importantly, however, this was the first revision to address issues of 
+computational representations for the IPA -- the principles above of which 
+have had several ramifications that make the interoperability of electronic 
+linguistic data extremely difficult.
+
+\subsubsection*{Computational representation}
+
+Prior to the Kiel Convention for the modern revision of the IPA in 1989, 
+\cite{Wells1987} collected and published practical approaches to coded 
+representations of the IPA, which dealt mainly with the assignment of 
+characters on the keyboard. The process of assigning standardized ``computer 
+codes'' to phonetic symbols was assigned to the Workgroup on Computer 
+Coding (henceforth working group) at the Kiel Convention. This working 
+group was tasked with \citep{Esling1990,EslingGaylord1993}:
+
+\begin{itemize}
+	\item determining how to represent the IPA numerically
+	\item developing a set of numbers to refer to the IPA symbols unambiguously
+	\item providing each symbol a unique name (intended to provide a mnemonic description of that character's shape)
+\end{itemize}
+
+\noindent The identification of IPA symbols with unique identifiers was 
+a first step in formalizing the IPA computationally because it would give 
+each symbol an unambiguous numerical identifier called an \textsc{IPA Number}. 
+The numbering system was to be comprehensive enough to support future revisions 
+of the IPA, including symbol specifications and diacritic placement. The 
+application of diacritics was also to be made explicit.
+
+Although the Association had never officially approved a set of names 
+for the IPA symbols, each IPA symbol received a unique \textsc{IPA Name}. 
+Many symbols already had an informal name (or two) used by linguists, but 
+consensus on symbol names was growing due to the recent publication of the 
+\textit{Phonetic Symbol Guide} \citep{PullumLadusaw1986}. Thus most of the 
+IPA symbol names were taken from \cite{PullumLadusaw1986} \citep[31]{IPA1999}.
+
+The working group insightfully decided that the computing-coding convention 
+for the IPA should be independent of computer environments or formats (e.g.\ ASCII), 
+i.e.\ the IPA Number was not meant to be implemented directly in a computer 
+encoding. The working group report's declaration includes the explanatory 
+remarks \citep[82]{International1989report}:
+
+\begin{quote}
+The recommendation of a 7-bit ASCII or 8-bit extended-ASCII coding system 
+would be short-sighted in view of development towards 16-bit and 32-bit 
+processors. In fact, any specific recommendations would tie the Association 
+to a stage of technological development which is bound to be outdated long 
+before the next revision of the handbook.
+\end{quote}
+
+\noindent Thus the coding convention was not meant to address the engineering 
+aspects of the actual encoding in computers (cf.\ \cite{Anderson1984}). However, 
+it was meant to serve as a basis for a communication-interchange standard for 
+creating mapping tables from various computer encodings, fonts, phonetic-character-set 
+software, etc., to common IPA Numbers, and thus symbols.\footnote{Remember, at 
+this time in the late 1980s there was no stable multilingual computing environment. 
+But some solution was needed because scholars were increasingly using personal 
+computers for their research and many were quickly adopting electronic mail or 
+discussion boards like Usenet as a medium for international exchanges. This was 
+before the Internet as we know it today. Most of these systems ran on 8-bit 
+hardware systems using a 7-bit ASCII character encoding.}
+
+Furthermore, the assignment of computer codes to IPA symbols was meant to 
+represent an unbiased formulation. The Association plays the role of an international 
+advisory body and it stated that it should not recommend a particular existing 
+system of encoding. In fact, during this time there were a number of coding 
+systems used, but none of them had a dominant international position. The 
+differences between systems were also either too great or too subtle to warrant 
+an attempt at combining them \citep{International1989report}.
+
+The working group assigned each IPA symbol to a unique three-digit number, 
+i.e.\ an IPA Number. Encoded in this number scheme is information about the 
+status of each symbol (see below). The IPA numbers are listed with the IPA 
+symbols and they are also illustrated in IPA chart form (see \cite[84]{EslingGaylord1993} 
+or \cite[App. 2]{IPA2007}). The numbers were assigned in linear order (e.g.\ 
+[p] 101, [b] 102, [t] 103...) following the IPA revision of 1989 and its update in 1996.
+
+The working group made the decision that no IPA symbol, past or present, 
+could be ignored. The comprehensive inclusion of all IPA symbols was to 
+anticipate the possibility that some symbols might be added, withdrawn, 
+or reintroduced into current or future usage. For example, in the 1989 
+revision voiceless implosives < ƥ, ƭ, ƈ, ƙ, ʠ > were added; in the 1993 
+revision they were removed. Ligatures like < ʧ, ʤ > are included as formerly 
+recognized IPA symbols; they are assigned to the ``200 series'' of IPA numbers 
+as members of the group of symbols formerly recognized by the IPA. To ensure 
+backwards compatibility, legacy IPA symbols would retain an IPA Number and 
+an IPA Name for reference purposes. As we discuss below, this decision is 
+later reflected in the Unicode Standard; many legacy IPA symbols reside in 
+the \textsc{IPA Extensions} block.
+
+The IPA Number is simply expressed as a ``three-digit number numerical 
+directory of digit triples'' \cite{}.\footnote{For practical purposes, 
+the IPA Number also served as a typesetter's guide to the IPA chart.} 
+The numbering scheme specifies three-digit codes, the first digit of which 
+indicates the symbol's category \cite{Esling1990,EslingGaylord1993}:
+
+\begin{itemize}
+	\item 100s for accepted IPA consonants
+	\item 200s for former IPA consonants and non-IPA symbols
+	\item 300s for vowels
+	\item 400s for segmental diacritics
+	\item 500s for suprasegmental symbols
+	\item 600s-800s for future specifications
+	\item 900s for escape sequences
+\end{itemize}
+
+After a symbol is categorized, it is assigned a number sequentially, 
+e.g.\ [i] 301, [e] 302, [ɛ] 303. The system allows for the addition 
+of new symbols within the various series by appending them, e.g. [ⱱ] 
+184. Former or often used but non-IPA symbols for consonants, vowels 
+and diacritics are numbered from x99 backwards. For example, the voiceless 
+and voiced postalveolar affricates and fricatives < č, ǰ, š, ž > are 
+assigned the IPA numbers 299, 298, 297 and 296, respectively, because 
+they are not sanctioned IPA symbols.
+
+The assignment of the IPA numbers to IPA symbols provided the basis for 
+uniquely identifying the set of past and present IPA symbols as a type of 
+computational representational standard of the IPA. Within each revision 
+of the IPA, the coding defines a closed and clearly defined set of characters. 
+The benefits of this standardization are clear in at least two ways: it is 
+used in translation tables that reference ASCII representations of the IPA, 
+and this early computational representation of the IPA became the basis for 
+its inclusion into the Unicode Standard version 1.0.
+
+\subsubsection*{SAMPA and X-SAMPA}
+
+True to the working group's aim, the IPA numbers provided a mechanism for 
+a communication interchange standard for creating mapping tables to various 
+computer encodings. For example, the IPA coding system was used as a mapping 
+system in the creation of SAMPA \citep{Wells_etal1992}, an ASCII representation 
+of the IPA symbols.
+
+For a long time, linguists, like all other computer users, were
+limited to ASCII-encoded 7-bit characters, which only includes Latin characters,
+numbers and some punctuation and symbols. Restricted to these standard character
+sets that lacked IPA support or other language-specific graphemes that they
+needed, linguists devised their own solutions.\footnote{Early work addressing
+the need for a universal computing environment for writing systems and their
+computational complexity is discussed in \citet{Simons1989}. A survey of
+practical recommendations for language resources, including notes on encoding,
+can be found in \citet{BirdSimons2003}} For example, some chose to represent
+unavailable graphemes with substitutes, e.g.~the combination of <ng> to
+represent <ŋ>. Tech-savvy linguists redefined selected characters from a
+character encoding by mapping custom made fonts to specific code points.\footnote{For 
+example, SIL's popular font SIL IPA 1990}. However,
+one linguist's electronic text would not render properly on another linguist's
+computer without access to the same font. Furthermore, if two character encodings
+defined two character sets differently, then data could not be reliably and
+correctly displayed. This is a commonly encountered example of the non-interoperability of
+data and data formats.
+
+One solution was the ASCII-ification of the IPA, which simply involved 
+defining keyboard-able sequences as IPA symbol codings.\footnote{\cite{Wells1987} 
+provides an in-depth description of IPA codings from country-to-country. 
+Later ASCII-IPAs include Kirshenbaum (created in 1992 in a Usenet group and 
+named after its lead developer who was at Hewlett-Packard Laboratories) and 
+Worldbet (published in 1993 by \cite{Hieronymus1993}, who was at AT\&T Laboratories).} 
+A successful effort was SAMPA (Speech Assessment
+Methods Phonetic Alphabet), which was created between 1988--1991 in Europe to 
+represent IPA symbols with ASCII
+character sequences \citep{Wells1987,Wells_etal1992}, e.g. <p\textbackslash> 
+for [ɸ]. SAMPA was developed by a group of speech scientists from nine countries 
+in Europe and it constituted the ASCII-IPA symbols needed for phonemic transcription 
+of the principal European Union languages \citep{Wells1995}. It is still widely 
+used in language technology.
+
+Two problems with SAMPA are that (i) it is only a partial encoding of the IPA and (ii) it encodes different
+languages in separate data tables, instead of using a universal alphabet, like
+IPA.\@ SAMPA tables were developed as part of a European Commission-funded project to
+address technical problems like electronic mail exchange (what is now simply
+called email). SAMPA is essentially a hack to work around displaying IPA
+characters, but it provided speech technology and other fields a basis that has
+been widely adopted and often still used in code. So, SAMPA is a collection of tables to be
+compared, instead of a large universal table representing all languages.
+
+An extended version of SAMPA, called X-SAMPA, set out to include every symbol, 
+including all diacritics, in the IPA chart \citep{Wells1995}. X-SAMPA is considered
+more universally applicable because it consists of one table that encodes all
+characters in IPA. In line with the principles of the IPA, SAMPA and X-SAMPA include a 
+repertoire of symbols. These symbols are intended to represent phonemes rather than 
+all allophonic distinctions. Additionally, both ASCII-ifications of IPA are useful because 
+strings of SAMPA or X-SAMPA are (reportedly) uniquely parsable \citep{Wells1995}. However, 
+like the IPA, X-SAMPA has different notations for encoding the same phonetic phenomena 
+(see Pitfall \ref{}).
+
+SAMPA and X-SAMPA have been widely used for speech technology and as an encoding system in
+computational linguistics. In fact, they are stilled used in popular software packages 
+that require ASCII input and some of which have been co-opted for linguistic analyses, 
+e.g.~RuG/L04 and SplitsTree4.\footnote{See \url{http://www.let.rug.nl/kleiweg/L04/} and 
+\url{http://www.splitstree.org/}, respectively}
+
+\subsubsection*{The need for a single multilingual environment}
+
+During the 1980's, it became increasingly clear that an adequate solution 
+to the problem of multilingual computing environments was needed. Linguists 
+were on the forefront of addressing this issue because they faced these 
+challenges head-on by wishing to publish and communicate electronic text 
+with phonetic symbols which were not included in basic ASCII.\footnote{One 
+only needs to look at facsimiles of older electronic documents to see exotic 
+symbols written in by hand.}
+
+Long familiar were linguists already with the distinction between function 
+and form. Even in the context of the computer implementation of writing systems, 
+the necessity to distinguish form and function had been made \citep{Becker1984}. 
+The computer industry, on the other hand, did not consider, ignored, or simply 
+did not encode this principle when creating operating systems like MS-DOS, which 
+were limited to 256 code points (due to computer hardware architecture) and 
+encoded with one-to-one mappings from character codes to graphemes.
+
+Industry was starting to tackle the issues involved in developing a single 
+multilingual computing environment on a variety of fronts, including the then 
+new technology of bitmap fonts and the creation of Font Manager and Script 
+Manager by Apple \citep{Apple1985,Apple1986,Apple1988}. As noted above, around 
+this time linguists were developing work-arounds such as SAMPA, so that they 
+could communicate IPA transcription and use ASCII-based software. Some linguists 
+formalized the issues of multilingual text processing from a computational 
+perspective \citep{Anderson1984,Becker1984,Simons1989}. The study of writing 
+systems was also being invigorated by the computational challenges in making 
+computers work in a multilingual environment.\footnote{\cite[11--15]{Sampson1985} 
+urges linguists to view the study of writing systems as legitimate scientific enquiry.}
+
+% This standard became the basis for a proposal to include the IPA in the first version of the Unicode Standard.
+Decisions by the Computer Coding working group and work they continued after the 1989 Kiel Convention were adopted by the International Phonetic Association.
+These decisions are directly reflected in the Unicode Standard's encoding of IPA, seeing as it was the Association who submitted the script proposal to the Unicode Consortium.
+
+The second major benefit of the standardization of the IPA in a computational 
+representation by the Kiel working group is that it provided the basis for a 
+formal proposal to be submitted to various international standards organizations, 
+several of which were trying to tackle (and in a sense `win') the multilingual 
+computing environment problem. Basically, everyone from corporations to governments 
+to language scientists (for lack of a better term) wanted a single unified multilingual 
+character encoding set for all the world's writing systems, even if they did not 
+understand or appreciate the challenges involved in creating and adopting a solution. 
+Additionally, advancements in computer hardware were making the solutions easier 
+to implement in software.
+
+The engineering problems and solutions had been spelled out years before, e.g.\ 
+a two-byte encoding for multilingual text \citep{Anderson1984}. 
+Although languages vary to an astounding extent (cf.\ \cite{EvansLevinson2009}), 
+writing systems are quite similar formally and the issue of formal representation 
+of the world's orthographic systems had also been addressed \citep{Simons1989}.
+
+A major obstacle in creating a single encoding multilingual environment from 
+the perspective of writing systems involves the distinction between function 
+and form \citep{Becker1984} This distinction is so central to basic linguistic 
+theory and that trained linguists and semiologists take it as second nature. 
+A central challenge in developing a universal character set was to combine a 
+technological solution with a formalization of writing systems proper.\footnote{Of 
+course there were additional practical issues to overcome, e.g.\ funding, creating 
+the formal and technological proposal, deciding which characters and writing systems 
+to include initially, while setting precedence of how to add new ones in the future.}
+
+In hindsight it is easy to lose sight of how impactful 30 years of technological 
+development have been on linguistics, from theory development using quantitative 
+means to pure data collection and dissemination (which as fed back into the former). 
+But at the end of the 1970s, virtually no ordinary working linguist was using a 
+person computer \cite{Simons1996}. Personal computer usage, however, dramatically 
+increased throughout the 1980s. By 1990, dozens of character sets were in common use. 
+They varied in their architecture and in their character repertoires, which 
+made things a mess. There were two major players in the universal character 
+set race: Unicode and the International Organization for Standardization (ISO).
+
+\subsubsection*{Unicode and ISO 10646}
+
+In the late 1980s, a universal character set was being developed by what 
+is now referred to as the Unicode Consortium.\footnote{The Unicode 
+Consortium was officially incorporated in January 1991.} This consortium 
+consisted largely, although not entirely, of major US corporations, with 
+the aim of overcoming the inoperability of different coded character sets 
+and their costly hinderance for developing multilingual software development 
+and for internationalization efforts. Commercial importance of course drove 
+the early inclusion of Latin, non-Latin, and some exotic scripts; see the 
+table of commercial importance as measured by GDP of countries using certain 
+writing systems \citep[2]{unicode88}.
+
+The original Unicode manifesto is \cite{unicode88.pdf}.\footnote{http://www.unicode.org/history/unicode88.pdf} 
+Its aim was for a reliable international multilingual text encoding standard 
+that would encompass all scripts of the world, or in the author's own words, 
+``a new, world-wide ASCII''. An in-depth history of Unicode, highlighting 
+interesting facts like its first text prototypes at Apple and its incorporation 
+into TrueType, is retold online.\footnote{\url{http://www.unicode.org/history/earlyyears.html}}
+
+Unicode 88 provided the basic principles for the Unicode Standard's design -- 
+pushing for 16 bit representations of characters with a clear distinction 
+between characters and glyphs. Some of the contents of this status proposal 
+of 1988 were reworked for inclusion in the early Unicode Standard pre-publication 
+drafts and by August 1990, the proposal was in a (very) rough draft format. Its 
+editors and the Unicode Working Group (the predecessor to the Unicode Technical 
+Committee) worked together to lay out the the proposed standard's structure and 
+content. At this time, the proposal contained no code charts nor block descriptions.
+
+% http://www.unicode.org/history/earlyyears.html
+% During this period of time, in addition to his co-authoring of Apple KanjiTalk, Davis was involved in further discussions of a universal character set which were being prompted by the development of the Apple File Exchange.
+
+The other major player in developing a universal character set was the ISO 
+working group from the International Standards Organization (ISO), based 
+in Europe, which was responsible for ISO/IEC 10646. This character set 
+standard was composed in 1989 and a draft was published in 1990.\footnote{\url{http://www.iso.org/iso/catalogue_detail.htm?csnumber=56921}} 
+The `Universal Multiple-Octet Coded Character Set' or simply UCS was the first 
+officially standardized character encoding with the aim of including all characters from all writing systems.\footnote{\url{http://www.nada.kth.se/i18n/ucs/unicode-iso10646-oview.html}}
+
+ISO/IEC 10646 is partly based on ISO/IEC 8859, a series of of ASCII-based 
+standard character encodings published in 1987 that use a single bit 8-byte 
+character set. Each part of the standard, e.g.\ 8859-1, 8859-5, 8859-6, 
+encodes characters to support different languages' writing systems, e.g.\ 
+Latin-1 Western European, Latin/Cyrillic, Latin/Arabic, respectively. Being 
+a joint effort by the International Organization for Standardization (ISO) 
+and the International Electrotechnical Commission (IEC), the aim of the 
+standard is reliable information exchange. So, again, issues of phonetic 
+symbol encoding, typography, etc., were ignored -- or perhaps more properly 
+put, not commercially driven at this early stage.
+
+Intended for the major Western European languages, ISO/IEC 8859 was an 
+extension of the ASCII character encoding standard, which included the 
+English alphabet, numerals and computer control characters (e.g.\ beep, 
+space, carriage return). By extending ASCII's 7-bit system to 8-bit, the 
+character repertoire of each of ISO/IEC 8859 character set was doubled 
+from 128 to 256 characters. Each character set defined a mapping between 
+digital bit patterns and characters, which are visually rendered on screen 
+as graphic symbols. ASCII was shared between ISO/IEC 8859 character sets, 
+but the characters in the extra bit patterns were different. Thus an aim 
+of the ISO working group responsible for ISO/IEC 10646 was to bring all 
+characters in all writings systems into a single unified encoding.
+
+In 1991, the Unicode Consortium and the ISO Working Group for ISO/IEC 10646 
+decided to create a single universal standard for encoding multilingual text.\footnote{http://unicode.org/book/appC.pdf} 
+The two character sets converged, resulting in mutually acceptable changes 
+to both, and each group keeps versions of their respective character codes 
+and encoding forms synchronized.\footnote{http://www.unicode.org/versions/} 
+Although each standard has its own form of reference and the terminology in 
+each may differ slightly, the practical difference is that the Unicode Standard 
+is a formal implementation of ISO/IEC 10646 and imposes additional constraints 
+on its implementation. The Unicode Standard includes character data, algorithms 
+and specifications, outside the scope of ISO/IEC 10646, which ensure, when 
+properly implemented in software applications and platforms, that characters 
+are treated uniformly.
+
+The incorporation of the Unicode Standard into the international encoding 
+standard ISO 10646 was approved by ISO as an International Standard in June 
+1992.\footnote{http://www.unicode.org/versions/Unicode1.0.0/Notice.pdf} 
+The joint Unicode and ISO/IEC 10646 standard has become \textit{the} universal 
+character set and it is a single multilingual environment for the majority 
+of the world's written languages. Its formal implementation has also been 
+vital to the rise of a multi-lingual Internet.
+
+\subsubsection*{IPA and Unicode}
+
+It was a long journey, but the goal of achieving a single multilingual 
+computing environment has largely been accomplished. We users, however, 
+must cope with the pitfalls that were dug along the way (see Pitfall sections below). 
+Some linguists, including your humble authors, are particularly sensitive 
+to these issues. So we provide practical advice and approaches in the rest 
+of this chapter. But first, we explain how the IPA became incorporated 
+into the Unicode Standard via ISO/IEC 10646.
+
+% "The set of IPA symbols and their numbers were used to draw up an entity set within SGML by the Text Encoding Initiative (TEl).
+The name of each entity is formed by 'IPA' preceding the number, e.g.
+IPA304 is the rEIentity name of lower-case A.
+These symbols can be processed as IPA symbols and represented on paper and screen with the appropriate local font by modifying the :entity replacement text.
+The advantage of the SGML entity set is that it is independent or the character set being used."
+% "A TEl writing system declaration (wsd) has been drawn up for the IPA symbols."
+% A TEl writing system declaration (wsd) has been drawn up for the IPA symbols.
+This document gives information about the symbol and its IPA function, as well as its encoding in the accompanying SGML document and in UnicodelUCS and in AFII.
+The writing system declaration can be read as a text d9cument or processed by machines in an SMGL process.
+
+After the Kiel Convention in 1989, the Computer Coding of the IPA working 
+group assisted the International Phonetic Association in representing the 
+IPA to ISO and to the Text Encoding Initiative (TEI) \citep{EslingGaylord1993}. 
+The working group's formalization of the IPA, i.e.\ a full listing of agreed 
+upon computer codings for phonetic symbols, was used in developing writing 
+systems descriptions which were at the time being solicited for scripts to 
+be included in initiatives for new multilingual international character 
+encoding standards. The working group for ISO/IEC 10646 and Unicode were 
+two such initiatives.
+
+In the historical context of the IPA being considered for inclusion in 
+ISO/IEC 10646, it is important to realize that there were a variety of 
+sources (i.e.\ not just from the Association) which submitted character 
+proposals for phonetic alphabets. These proposals, including from the 
+Association via the Kiel working group, were considered as a whole by 
+the ISO working groups which were responsible for incorporating a phonetic 
+script into the universal character set (UCS). The ISO working groups that 
+were responsible for assigning a phonetic character set then made their 
+own submissions as part of a review process by ISO for approval based on 
+both ``informatic'' and phonetic criteria \citep[86]{EslingGaylord1993}.
+
+Character set ISO/IEC 10646 was approved by ISO, including the phonetic 
+characters submitted to them in May 1993. The set of IPA characters were 
+assigned UCS codes in 16 bit representation (in hexadecimal) and were 
+published Tables 2 and 3 in \cite{EslingGaylord1993}, which include a 
+graphical representation of the IPA symbol, its IPA Name, phonetic description, 
+IPA Number, UCS Code and AFII Code.\footnote{The Association for Font 
+Information Interchange (AFII) was an international database of glyphs 
+created to promote the standardization of font data required to produce 
+ISO/IEC 10646.} Because the character sets of ISO/IEC 10646 and the 
+Unicode Standard converged, the IPA as submitted by the Association 
+and reviewed and further submitted by the ISO working group, was included 
+in the Unicode Standard Version 1.0 -- largely as we know it today.\footnote{The 
+Association later made the foresightful remark, ``When this character set 
+is in wide use, it will be the normal way to encode IPA symbols.'' \citep[164]{IPA1999}.}
+
+With subsequent revisions to the IPA, one might expect the Unicode 
+Consortium would update the Unicode Standard in a way that is inline 
+with linguists' or other language scientists' intuitions. However, 
+updates that go against the ISO's and the Unicode Standard's principles 
+of maintaining backwards compatibility lose out, i.e.\ it is more important 
+to deal with the pitfalls created along the way then it is to change the 
+standard. Therefore, many of the pitfalls we encounter when using Unicode 
+IPA are historic relics that we have to come to grips with.
+
+% https://en.wikipedia.org/wiki/Uralic_Phonetic_Alphabet
+% http://www.unicode.org/conference/bulldog.html
+% http://www-01.sil.org/computing/computing_environment.html
 
@@
 to mark different tone levels, e.g. <ȅèée̋>.\footnote{To make things even more
-complexe, there are at least two different Unicode homoglyphs for the low and
+complex, there are at least two different Unicode homoglyphs for the low and
 high level tones, namely <\diaf{0340}> \textsc{combining grave tone mark} at
@@
 
+% IPA1999 pg 19
+% "Either of the variant letter shapes [g] and [g] may be used to represent the voiced velar plosive."
+
 Second, we commonly encounter the use of <g> \textsc{latin small letter g} at
@@
 
+% length IPA 1999:22
+% "Note: as in Finnish orthography, length can also be indicated in phonetic transcription by double letters: e.g . Finnish maatto [rncctto] 'electrical earth/ground'."
+
+% syllable breaks and word boundaries <\s, ., |, ||, tie-bar below>
+% White spaces can be used to indicate word boundaries.
+Syllable breaks can be marked when required.
+The other two boundary symbols are used to mark the domain of larger prosodic units.
+There is also a linking symbol that can be used for explicitly indicating the lack of a boundary.
+
+%  (In all these transcriptions, no spaces between words have been included.
+This is inevitable in an impressionistic transcription where it is not yet known how the utterance divides into words.
+In phonemic and allophonic transcriptions it is common to include spaces to aid legibility, but their theoretical validity is problematic.)
+
 % ==========================
@@
 % ==========================
+
+In the major revision of the IPA in 1932, affricates were represented by two consonants (e.g.\ tʃ), or when possible ligatures (ʧ), or with the tie-bar 
+(t͡s). In the 1938 revision the tie-bar's semantics were broadened to indicate simultaneous articulation, e.g.\ in labial velars such as k͡p.
+
+% Most consonants that involve two simultaneous places of articulation are written by combining two symbols with the 'tie bar' [~), for example [kp) which represents a voiceless labial-velar plosive.
+% In other cases, such as [w), the sound involves two places of articulation simultaneously, which makes it inconvenient to display in the table.Ifseparate columns for all consonants with two places of articulation were provided, the size of the grid would become unmanageable.
+% "Affricates can be written as digraphs, as ligatures, or with slur marks; thus ts, tS, dZ, ts, tS, dZ...
+
+% Wells1995:10 "The underscore could in principle also be pressed into service to represent the IPA tie bar.
+The current chart mentions its use only for affricates and double articulations, and then only "if necessary"."
 
@@
 string itself (using the tie bar).
+
+% the \textsc{undertie} at U+203F is used as a linking symbol to indicate the lack of a boundary, e.g.\ petit ami [pətit‿ami] `boyfriend' (French).
+This symbol looks like and can be confused with the tie-bar below.
 
@@
 
+% this section is wrong wrt IPA diacritics and needs updating @SM
+
 Another pitfall relates to the question of what are diacritics. The problem is that
@@
 character.\footnote{See
-\url{http://unicode.org/reports/tr44/#Canonical_Combining_Class_Values} for a
+\url{http://unicode.org/reports/tr44/\#Canonical\_Combining\_Class\_Values} for a
 detailed description.} When applying a Unicode normalization (NFC or NFD, see
@@
 % ==========================
+\section{Pitfall: Future revisions to the IPA}
+\label{ipa-revisions}
+% ==========================
+
+With each revision of the IPA, many decisions need to be made by 
+the Association as to which symbols should be added, removed or 
+changed. (Perhaps that's why the IPA is updated infrequently.) 
+For example, in the 1989 revision of the IPA at the Kiel Convention, 
+changes to specific symbols (in previous charts) were debated and 
+the Association's members certain decisions. The prevailing mood at 
+the convention was not to change specific symbols unless a strong 
+case was made \cite{Ladefoged1990a}. For example, two such decisions 
+included \cite[62]{Ladefoged1990a}:
+
+\begin{itemize}
+	\item symbols for clicks were changed from < ʇ, ʖ, ʗ > and replaced by < ǀ, ǁ, ǃ > because the latter were the symbols used by nearly all Khoisanists and Bantuists
+	\item the Americanist tradition of < š, ž, č, ǰ > were not adopted because the Association members at the convention ``were not sufficiently impressed by arguments ...
+to the effect that these sounds formed a natural class, and thus is would be appropriate to recognize this by maintaining a common aspect to their symbolism''
+\end{itemize}
+
+\noindent These decisions have practical consequences for transcribers 
+of IPA, particularly those who wish to follow recommended practices of 
+encoding electronic text in the Unicode Standard. For example, the Unicode 
+Standard contains removed symbols as labels them as clicks, e.g. <ʇ> 
+\textsc{latin small letter turned t} at U+0287 in the \textsc{IPA Extensions} 
+block has the comment `dental click (sound of ``tsk tsk'')'. In this case, 
+the IPA transcriber must know the status of click symbols in the current 
+version of the IPA and then identify those characters within the Unicode Standard.
+
+The most controversial issue regarding symbols debated at the convention 
+was the representation for voiceless implosives \cite[62]{Ladefoged1990a}. 
+Recall the principles of the IPA outlined in Section \ref{the-international-phonetic-alphabet}, 
+such as using distinct symbols for phonological contrast versus convenience of 
+display in the chart, must be taken into account when arguing for or against 
+the inclusion or deletion of IPA symbols in the IPA chart. Furthermore, the 
+inclusion of deletion of symbols should incorporate the phonetic knowledge 
+of the world's languages.
+
+\cite{Ladefoged1990a} argued against the inclusion of these symbols < ƥ, ƭ, ƈ, ƙ, ʠ > 
+voiceless implosives, noting that: they are not contrastive (e.g.\ in Mayan 
+languages); there is no instrumental evidence supporting voiceless implosives 
+in Africa; the sounds are sufficiently rare as to not need a whole new row 
+of symbols in the chart. Ladefoged favored symbolizing the sounds using a 
+voiceless diacritic ring below voiced implosives, e.g.\ <ɓ̥>.
+
+Nevertheless, in the 1989 IPA chart there is indeed a row for implosives 
+containing voiceless and voiced pairs.\footnote{https://en.wikipedia.org/wiki/File:IPA\_as\_of\_1989.png} 
+But already in the next revision, in 1993 (with an update in 1996), the voiceless 
+implosives were dropped. The implosives row from the IPA consonantal chart 
+disappeared and voiced implosives were given a column in the non-pulmonic 
+consonants table (which is still reflected in the latest revision to date, IPA 2005).
+
+The Journal of the International Phonetic Association follows its own 
+published standard for the IPA at the time of publication, even when it 
+may conflict with the Association's principle of using different symbols 
+for contrastive sounds and diacritics for phonetic variation. For example, 
+in the case of voiceless implosives, \cite{McLaughlin2005} shows that 
+Seereer-Siin (Niger-Congo; Atlantic; ISO 639-3: srr) has a phonologically 
+contrastive set of voiced and voiceless implosive stops at the labial, 
+coronal and palatal places of articulation. These symbols are transcribed 
+in an \textit{Illustrations of the IPA} article in the IPA journal as 
+< ɓ̥, ɗ̥, ʄ̥ >.\footnote{Note that the IPA chart states that diacritics 
+can also be placed above a symbol, e.g.\ reportedly for legibility's 
+sake < ɠ̥> vs < ɠ̊ >. This of course leads to the variability in encoding 
+that we are trying to address on a computational level in this work. 
+See Section \ref{}.}
+
+The point of this pitfall is to highlight that revisions to the IPA will 
+continue into the future, albeit they are infrequent. Nevertheless, 
+given the Unicode Standard's principle of maintaining backwards compatibility 
+(at all costs), transcribers and consumers of IPA cannot rely solely on 
+remarks in the Unicode Standard to reflect current standard IPA usage. 
+There is the possibility that at a later revision of the IPA, symbols that 
+are not currently encoded in the Unicode Standard are added to the IPA -- 
+although we think this is unlikely.
+
+% Great quote for standardization -- last sentence(s) in Ladefoged 1990:552
+
+% ==========================
 \section{Recommendations}

### 2015-07-07 chapters/ipa_meets_unicode.tex
minor edits, additions

@@
 above a character (like in <ã>, using the \textsc{combining tilde} at
-\uni{0303}) is used for nasalisation, while the tilde below a character (like in
+\uni{0303}) is used for nasalization, while the tilde below a character (like in
 <a̰>, using the \textsc{combining tilde below} at \uni{0330}) indicates creaky
@@
 
-In the major revision of the IPA in 1932, affricates were represented by two consonants (e.g.\ tʃ), or when possible ligatures (ʧ), or with the tie-bar 
-(t͡s). In the 1938 revision the tie-bar's semantics were broadened to indicate simultaneous articulation, e.g.\ in labial velars such as k͡p.
-
-% Most consonants that involve two simultaneous places of articulation are written by combining two symbols with the 'tie bar' [~), for example [kp) which represents a voiceless labial-velar plosive.
-% In other cases, such as [w), the sound involves two places of articulation simultaneously, which makes it inconvenient to display in the table.Ifseparate columns for all consonants with two places of articulation were provided, the size of the grid would become unmanageable.
-% "Affricates can be written as digraphs, as ligatures, or with slur marks; thus ts, tS, dZ, ts, tS, dZ...
-
 % Wells1995:10 "The underscore could in principle also be pressed into service to represent the IPA tie bar.
@@
 
-The IPA includes a tie bar. The tie-bar is a convenient diacritic for visually
-tokenizing input strings into chunks of phonetically salient groups, for example
-to indicate affricates, doubly articulated consonants or diphthongs. This
-character can be placed above or below the base characters, e.g. <\charis{t͡s}>
-or <\charis{t͜s}>. IPA allows both options. The choice between the two symbols
+In the major revision of the IPA in 1932, affricates were represented by two consonants 
+(e.g.\ <tʃ>), or ligatures <ʧ>, or with the tie-bar <t͡s>. In the 1938 revision the tie-bar's 
+semantics were broadened to indicate simultaneous articulation, e.g.\ in labial velars 
+such as <k͡p>. Thus the tie-bar is a convenient diacritic for visually tokenizing input 
+strings into chunks of phonetically salient groups, including affricates, doubly articulated 
+consonants or diphthongs.
+
+The tie bar can be placed above or below the base characters, e.g. <\charis{t͡s}>
+or <\charis{t͜s}>.\footnote{Note that the \textsc{undertie} at \uni{203F} symbol 
+looks like and is easily confused with the tie-bar below. However, it is a different 
+character and has a different function in IPA. The undertie is used as a linking 
+symbol to indicate the lack of a boundary, e.g.\ petit ami [pətit‿ami] `boyfriend' 
+($<$ French).} IPA allows both options. The choice between the two symbols
 is purely for legible rendering; there is no difference in semantics between the
@@
 
-% the \textsc{undertie} at U+203F is used as a linking symbol to indicate the lack of a boundary, e.g.\ petit ami [pətit‿ami] `boyfriend' (French).
-This symbol looks like and can be confused with the tie-bar below.
-
 % ==========================
@@
 % ==========================
+
+% TODO: link back to discussion on IPA principles and how ligatures got brought along historically
 
@@
 
-% this section is wrong wrt IPA diacritics and needs updating @SM
+% TODO: this section is wrong wrt IPA diacritics and needs updating @SM
 
@@
 
-% Great quote for standardization -- last sentence(s) in Ladefoged 1990:552
-
 % ==========================
@@
 % ==========================
+
+% Great quote for standardization -- last sentence(s) in Ladefoged 1990:552
 

### 2015-07-07 chapters/ipa_meets_unicode.tex
initial notes for IPA extensions

@@
 
+\begin{comment}
+\subsection{IPA (practical) extensions}
+
+The IPA extensions cover speech sounds beyond the sound systems of the world's languages, including symbols for pathological speech and paralinguistic functions (e.g.\ Braille?).
+
+- pathological speech
+- paralinguistic functions (Braille?)
+
+As noted in Section \ref{}, extensions to the IPA -- in particular ``large-scale'' extensions --
+
+% Englebretson2009.pdf
+http://www.ruf.rice.edu/~reng/englebretson2009.pdf
+
+X-SAMPA was an ASCII transliteration and is still purposeful for several programs
+
+There is also the extensions of the IPA designed for disordered speech, although the symbols are occasionally used for transcribing phonetic detail in ``normal'' speech or also cross-linguistic rarities.\footnote{\url{http://phoible.github.io/conventions/}}
+
+% Unicode has its own history; for a summary: \url{http://www.unicode.org/history/summary.html}
+
+https://www.phon.ucl.ac.uk/home/wells/ipa-unicode.htm
+
+Computer codes for phonetic symbols
+EslingGaylord1993
+
+\end{comment}
+
 % ==========================

### 2015-07-14 chapters/ipa_meets_unicode.tex
typo
reordered last pitfall; added subsections to IPA

@@
 
-\subsubsection*{Brief history}
+\subsection{Brief history}
 
@@
 
-\subsubsection*{Principles}
+\subsection{Principles}
 
@@
 
-\subsubsection*{Computational representation}
+\subsection{Computational representation}
 
@@
 
-\subsubsection*{SAMPA and X-SAMPA}
+\subsection{SAMPA and X-SAMPA}
 
@@
 
-\subsubsection*{The need for a single multilingual environment}
+\subsection{The need for a single multilingual environment}
 
@@
 
-\subsubsection*{Unicode and ISO 10646}
+\subsection{Unicode and ISO 10646}
 
@@
 
-\subsubsection*{IPA and Unicode}
+\subsection{IPA and Unicode}
 
@@
 In the major revision of the IPA in 1932, affricates were represented by two consonants 
-(e.g.\ <tʃ>), or ligatures <ʧ>, or with the tie-bar <t͡s>. In the 1938 revision the tie-bar's 
+(e.g.\ <tʃ>), or ligatures <ʧ>, or with the tie-bar <t͡ʃ>. In the 1938 revision the tie-bar's 
 semantics were broadened to indicate simultaneous articulation, e.g.\ in labial velars 
@@
 \end{itemize}
+
+% ==========================
+\section{Pitfall: Future revisions to the IPA}
+\label{ipa-revisions}
+% ==========================
+
+With each revision of the IPA, many decisions need to be made by 
+the Association as to which symbols should be added, removed or 
+changed. (Perhaps that's why the IPA is updated infrequently.) 
+For example, in the 1989 revision of the IPA at the Kiel Convention, 
+changes to specific symbols (in previous charts) were debated and 
+the Association's members certain decisions. The prevailing mood at 
+the convention was not to change specific symbols unless a strong 
+case was made \cite{Ladefoged1990a}. For example, two such decisions 
+included \cite[62]{Ladefoged1990a}:
+
+\begin{itemize}
+	\item symbols for clicks were changed from < ʇ, ʖ, ʗ > and replaced by < ǀ, ǁ, ǃ > because the latter were the symbols used by nearly all Khoisanists and Bantuists
+	\item the Americanist tradition of < š, ž, č, ǰ > were not adopted because the Association members at the convention ``were not sufficiently impressed by arguments ...
+to the effect that these sounds formed a natural class, and thus is would be appropriate to recognize this by maintaining a common aspect to their symbolism''
+\end{itemize}
+
+\noindent These decisions have practical consequences for transcribers 
+of IPA, particularly those who wish to follow recommended practices of 
+encoding electronic text in the Unicode Standard. For example, the Unicode 
+Standard contains removed symbols as labels them as clicks, e.g. <ʇ> 
+\textsc{latin small letter turned t} at U+0287 in the \textsc{IPA Extensions} 
+block has the comment `dental click (sound of ``tsk tsk'')'. In this case, 
+the IPA transcriber must know the status of click symbols in the current 
+version of the IPA and then identify those characters within the Unicode Standard.
+
+The most controversial issue regarding symbols debated at the convention 
+was the representation for voiceless implosives \cite[62]{Ladefoged1990a}. 
+Recall the principles of the IPA outlined in Section \ref{the-international-phonetic-alphabet}, 
+such as using distinct symbols for phonological contrast versus convenience of 
+display in the chart, must be taken into account when arguing for or against 
+the inclusion or deletion of IPA symbols in the IPA chart. Furthermore, the 
+inclusion of deletion of symbols should incorporate the phonetic knowledge 
+of the world's languages.
+
+\cite{Ladefoged1990a} argued against the inclusion of these symbols < ƥ, ƭ, ƈ, ƙ, ʠ > 
+voiceless implosives, noting that: they are not contrastive (e.g.\ in Mayan 
+languages); there is no instrumental evidence supporting voiceless implosives 
+in Africa; the sounds are sufficiently rare as to not need a whole new row 
+of symbols in the chart. Ladefoged favored symbolizing the sounds using a 
+voiceless diacritic ring below voiced implosives, e.g.\ <ɓ̥>.
+
+Nevertheless, in the 1989 IPA chart there is indeed a row for implosives 
+containing voiceless and voiced pairs.\footnote{https://en.wikipedia.org/wiki/File:IPA\_as\_of\_1989.png} 
+But already in the next revision, in 1993 (with an update in 1996), the voiceless 
+implosives were dropped. The implosives row from the IPA consonantal chart 
+disappeared and voiced implosives were given a column in the non-pulmonic 
+consonants table (which is still reflected in the latest revision to date, IPA 2005).
+
+The Journal of the International Phonetic Association follows its own 
+published standard for the IPA at the time of publication, even when it 
+may conflict with the Association's principle of using different symbols 
+for contrastive sounds and diacritics for phonetic variation. For example, 
+in the case of voiceless implosives, \cite{McLaughlin2005} shows that 
+Seereer-Siin (Niger-Congo; Atlantic; ISO 639-3: srr) has a phonologically 
+contrastive set of voiced and voiceless implosive stops at the labial, 
+coronal and palatal places of articulation. These symbols are transcribed 
+in an \textit{Illustrations of the IPA} article in the IPA journal as 
+< ɓ̥, ɗ̥, ʄ̥ >.\footnote{Note that the IPA chart states that diacritics 
+can also be placed above a symbol, e.g.\ reportedly for legibility's 
+sake < ɠ̥> vs < ɠ̊ >. This of course leads to the variability in encoding 
+that we are trying to address on a computational level in this work. 
+See Section \ref{}.}
+
+The point of this pitfall is to highlight that revisions to the IPA will 
+continue into the future, albeit they are infrequent. Nevertheless, 
+given the Unicode Standard's principle of maintaining backwards compatibility 
+(at all costs), transcribers and consumers of IPA cannot rely solely on 
+remarks in the Unicode Standard to reflect current standard IPA usage. 
+There is the possibility that at a later revision of the IPA, symbols that 
+are not currently encoded in the Unicode Standard are added to the IPA -- 
+although we think this is unlikely.
+
+\begin{comment}
+\subsection{IPA (practical) extensions}
+
+The IPA extensions cover speech sounds beyond the sound systems of the world's languages, including symbols for pathological speech and paralinguistic functions (e.g.\ Braille?).
+
+- pathological speech
+- paralinguistic functions (Braille?)
+
+As noted in Section \ref{}, extensions to the IPA -- in particular ``large-scale'' extensions --
+
+% Englebretson2009.pdf
+http://www.ruf.rice.edu/~reng/englebretson2009.pdf
+
+X-SAMPA was an ASCII transliteration and is still purposeful for several programs
+
+There is also the extensions of the IPA designed for disordered speech, although the symbols are occasionally used for transcribing phonetic detail in ``normal'' speech or also cross-linguistic rarities.\footnote{\url{http://phoible.github.io/conventions/}}
+
+% Unicode has its own history; for a summary: \url{http://www.unicode.org/history/summary.html}
+
+https://www.phon.ucl.ac.uk/home/wells/ipa-unicode.htm
+
+Computer codes for phonetic symbols
+EslingGaylord1993
+
+\end{comment}
 
@@
 % ==========================
-\section{Pitfall: Future revisions to the IPA}
-\label{ipa-revisions}
-% ==========================
-
-With each revision of the IPA, many decisions need to be made by 
-the Association as to which symbols should be added, removed or 
-changed. (Perhaps that's why the IPA is updated infrequently.) 
-For example, in the 1989 revision of the IPA at the Kiel Convention, 
-changes to specific symbols (in previous charts) were debated and 
-the Association's members certain decisions. The prevailing mood at 
-the convention was not to change specific symbols unless a strong 
-case was made \cite{Ladefoged1990a}. For example, two such decisions 
-included \cite[62]{Ladefoged1990a}:
-
-\begin{itemize}
-	\item symbols for clicks were changed from < ʇ, ʖ, ʗ > and replaced by < ǀ, ǁ, ǃ > because the latter were the symbols used by nearly all Khoisanists and Bantuists
-	\item the Americanist tradition of < š, ž, č, ǰ > were not adopted because the Association members at the convention ``were not sufficiently impressed by arguments ...
-to the effect that these sounds formed a natural class, and thus is would be appropriate to recognize this by maintaining a common aspect to their symbolism''
-\end{itemize}
-
-\noindent These decisions have practical consequences for transcribers 
-of IPA, particularly those who wish to follow recommended practices of 
-encoding electronic text in the Unicode Standard. For example, the Unicode 
-Standard contains removed symbols as labels them as clicks, e.g. <ʇ> 
-\textsc{latin small letter turned t} at U+0287 in the \textsc{IPA Extensions} 
-block has the comment `dental click (sound of ``tsk tsk'')'. In this case, 
-the IPA transcriber must know the status of click symbols in the current 
-version of the IPA and then identify those characters within the Unicode Standard.
-
-The most controversial issue regarding symbols debated at the convention 
-was the representation for voiceless implosives \cite[62]{Ladefoged1990a}. 
-Recall the principles of the IPA outlined in Section \ref{the-international-phonetic-alphabet}, 
-such as using distinct symbols for phonological contrast versus convenience of 
-display in the chart, must be taken into account when arguing for or against 
-the inclusion or deletion of IPA symbols in the IPA chart. Furthermore, the 
-inclusion of deletion of symbols should incorporate the phonetic knowledge 
-of the world's languages.
-
-\cite{Ladefoged1990a} argued against the inclusion of these symbols < ƥ, ƭ, ƈ, ƙ, ʠ > 
-voiceless implosives, noting that: they are not contrastive (e.g.\ in Mayan 
-languages); there is no instrumental evidence supporting voiceless implosives 
-in Africa; the sounds are sufficiently rare as to not need a whole new row 
-of symbols in the chart. Ladefoged favored symbolizing the sounds using a 
-voiceless diacritic ring below voiced implosives, e.g.\ <ɓ̥>.
-
-Nevertheless, in the 1989 IPA chart there is indeed a row for implosives 
-containing voiceless and voiced pairs.\footnote{https://en.wikipedia.org/wiki/File:IPA\_as\_of\_1989.png} 
-But already in the next revision, in 1993 (with an update in 1996), the voiceless 
-implosives were dropped. The implosives row from the IPA consonantal chart 
-disappeared and voiced implosives were given a column in the non-pulmonic 
-consonants table (which is still reflected in the latest revision to date, IPA 2005).
-
-The Journal of the International Phonetic Association follows its own 
-published standard for the IPA at the time of publication, even when it 
-may conflict with the Association's principle of using different symbols 
-for contrastive sounds and diacritics for phonetic variation. For example, 
-in the case of voiceless implosives, \cite{McLaughlin2005} shows that 
-Seereer-Siin (Niger-Congo; Atlantic; ISO 639-3: srr) has a phonologically 
-contrastive set of voiced and voiceless implosive stops at the labial, 
-coronal and palatal places of articulation. These symbols are transcribed 
-in an \textit{Illustrations of the IPA} article in the IPA journal as 
-< ɓ̥, ɗ̥, ʄ̥ >.\footnote{Note that the IPA chart states that diacritics 
-can also be placed above a symbol, e.g.\ reportedly for legibility's 
-sake < ɠ̥> vs < ɠ̊ >. This of course leads to the variability in encoding 
-that we are trying to address on a computational level in this work. 
-See Section \ref{}.}
-
-The point of this pitfall is to highlight that revisions to the IPA will 
-continue into the future, albeit they are infrequent. Nevertheless, 
-given the Unicode Standard's principle of maintaining backwards compatibility 
-(at all costs), transcribers and consumers of IPA cannot rely solely on 
-remarks in the Unicode Standard to reflect current standard IPA usage. 
-There is the possibility that at a later revision of the IPA, symbols that 
-are not currently encoded in the Unicode Standard are added to the IPA -- 
-although we think this is unlikely.
-
-\begin{comment}
-\subsection{IPA (practical) extensions}
-
-The IPA extensions cover speech sounds beyond the sound systems of the world's languages, including symbols for pathological speech and paralinguistic functions (e.g.\ Braille?).
-
-- pathological speech
-- paralinguistic functions (Braille?)
-
-As noted in Section \ref{}, extensions to the IPA -- in particular ``large-scale'' extensions --
-
-% Englebretson2009.pdf
-http://www.ruf.rice.edu/~reng/englebretson2009.pdf
-
-X-SAMPA was an ASCII transliteration and is still purposeful for several programs
-
-There is also the extensions of the IPA designed for disordered speech, although the symbols are occasionally used for transcribing phonetic detail in ``normal'' speech or also cross-linguistic rarities.\footnote{\url{http://phoible.github.io/conventions/}}
-
-% Unicode has its own history; for a summary: \url{http://www.unicode.org/history/summary.html}
-
-https://www.phon.ucl.ac.uk/home/wells/ipa-unicode.htm
-
-Computer codes for phonetic symbols
-EslingGaylord1993
-
-\end{comment}
-
-% ==========================
 \section{Recommendations}

### 2015-07-23 chapters/ipa_meets_unicode.tex
readability edits

@@
 
+\section{Introduction}
+
 The International Phonetic Alphabet (IPA) is a common standard in linguistics to
 transcribe sounds of spoken language into discrete segments using a Latin-based
-alphabet. Although IPA is reasonably easily adhered to with pen and paper, it is not trivial to encode
-IPA characters electronically. Similar to the previous chapter, in this chapter
-we discuss various pitfalls with the encoding of IPA in the Unicode Standard.\@ We will
-specifically refer to the 2005 version of the IPA \citep{IPA2005} and the 7.0 
-version of Unicode \citep{Unicode2014}.
+alphabet. Although IPA is reasonably easily adhered to with pen and paper, it is
+not trivial to encode IPA characters electronically. Similar to the previous
+chapter, in this chapter we discuss various pitfalls with the encoding of IPA in
+the Unicode Standard. We will specifically refer to the 2005 version of the
+IPA \citep{IPA2005} and the 7.0 version of Unicode \citep{Unicode2014}.
 
@@
 to try to be as consistent as possible across a wide range of use cases, and it
-does place linguistic traditions above other orthographic possibilities. Furthermore, 
-when we look at the history of how the IPA meet Unicode, we see that many of the 
+does place linguistic traditions above other orthographic choices. Furthermore,
+when we look at the history of how the IPA met Unicode, we see that many of the
 decisions for IPA symbols in the Unicode Standard come directly from the
-International Phonetic Association itself. Therefore as users, many pitfalls that we 
-encounter have their grounding in the historical principles of the IPA, as 
-well as the technological considerations involved in creating a single 
-multilingual computing encoding.
-
-In general, we strongly suggest to linguists to not complain about any decisions in
-the Unicode Standard, but to try and understand the rationale of the International Phonetic Association 
-and the Unicode Consortium (which in our experience is almost always well-conceived) and devise
-ways to work with any unexpected behavior. Many of the current problems derive
-from the fact that the IPA is clearly historically based on the Latin script,
-but different enough from most other Latin-based writing systems to warrant
-special attention. This ambivalent status of the IPA glyphs (partly Latin,
-partly special) is unfortunately also attested in the treatment of IPA in the
-Unicode Standard. In retrospect, it might have been better to consider the IPA
-(and other transcription systems) to be a special kind of script within
-the Unicode Standard, and treat the obvious similarity to Latin glyphs as a
-historical relic. All IPA glyphs would then have their own code points, instead
-of the current situation in which some IPA glyphs have special code points,
-while others are treated as being identical to the regular Latin characters.
-Yet, the current situation, however unfortunate, is unlikely to change, so as
-linguists we must learn to deal with the specific pitfalls of IPA within
-the Unicode Standard.
+International Phonetic Association itself. Therefore, many pitfalls that we will
+encounter have their grounding in the history of the principles of the IPA, as
+well as in the technological considerations involved in creating a single
+multilingual encoding. In general, we strongly suggest to linguists to not
+complain about any decisions in the Unicode Standard, but to try and understand
+the rationale of the International Phonetic Association and the Unicode
+Consortium (both of which are almost always well-conceived in our experience)
+and devise ways to work with any unexpected behavior.
+
+Many of the current problems derive from the fact that the
+IPA is clearly historically based on the Latin script, but different enough from
+most other Latin-based writing systems to warrant special attention. This
+ambivalent status of the IPA glyphs (partly Latin, partly special) is
+unfortunately also attested in the treatment of IPA in the Unicode Standard. In
+retrospect, it might have been better to consider the IPA (and other
+transcription systems) to be a special kind of script within the Unicode
+Standard, and treat the obvious similarity to Latin glyphs as a historical
+relic. All IPA glyphs would then have their own code points, instead of the
+current situation in which some IPA glyphs have special code points, while
+others are treated as being identical to the regular Latin characters. Yet, the
+current situation, however unfortunate, is unlikely to change, so as linguists
+we must learn to deal with the specific pitfalls of IPA within the Unicode
+Standard.
 
@@
 has long maintained a standard alphabet, the International Phonetic Alphabet (IPA),
-\footnote{Also referred to as API, for \textit{Association Phonétique Internationale}.} 
-which is a common standard in linguistics to transcribe sounds of spoken languages. 
-It was first published in 1888 as an international system of phonetic transcription 
-for oral languages and for pedagogical purposes. The draft of 1887 contained phonetic 
-values for English, French and German. Diacritics for length and nasalization were 
-already present in this draft and the same symbols are still used today.
-
-Although the original IPA alphabet was a list of symbols with pronunciation examples 
-from words in different languages, in 1900 the symbols were organized into chart and 
-given phonetic feature labels for consonants (e.g.\ for manner of articulation: `plosives', 
-`nasales', `fricatives'; for place of articulation: `bronchials', `laryngeales', `labiales') 
-and for vowels (`fermeées', `mi-fermeées', `mi-overtes', `overtes'). Over the last century 
-or so, the structure of the chart has changed with increases in phonetic knowledge, and 
-thus like notational systems in other scientific disciplines, the IPA reflects facts and 
-theories (of phonetic knowledge) that have developed over time. It is natural then that 
-the IPA is modified occasionally to accommodate scientific innovations and discoveries. 
-(In fact, this is part of the Association's mandate.) These changes are captured in 
-the revisions to the IPA.\footnote{For a detailed history see: 
-\url{https://en.wikipedia.org/wiki/History\_of\_the\_International\_Phonetic_Alphabet}}
+%\footnote{Also referred to as API, for \textit{Association Phonétique Internationale}.} 
+which is a common standard in linguistics to transcribe sounds of spoken
+languages. It was first published in 1888 as an international system of phonetic
+transcription for oral languages and for pedagogical purposes. The draft of 1887
+contained phonetic values for English, French and German. Diacritics for length
+and nasalization were already present in this draft and the same symbols are
+still used today.
+
+Originally, the IPA alphabet was a list of symbols with pronunciation examples
+from words in different languages. In 1900 the symbols were first organized into
+chart and were given phonetic feature labels for consonants (e.g.\ for manner of
+articulation: `plosives', `nasales', `fricatives'; for place of articulation:
+`bronchials', `laryngeales', `labiales') and for vowels (e.g.\ `fermeées',
+`mi-fermeées', `mi-overtes', `overtes'). Throughout the last centory, the
+structure of the chart has changed with increases in phonetic knowledge, and
+thus, like notational systems in other scientific disciplines, the IPA reflects
+facts and theories (of phonetic knowledge) that have developed over time. It is
+natural then that the IPA is modified occasionally to accommodate scientific
+innovations and discoveries. (In fact, this is part of the Association's
+mandate.) These changes are captured in the revisions to the IPA.
+%\footnote{For a detailed history see:
+%\url{https://en.wikipedia.org/wiki/History\_of\_the\_International\_Phonetic_Alphabet}}
 
 Over the years there has been several revisions, many minor. For example,
-articulation labels -- what we now call features even though the IPA avoids this term -- 
-have changed (e.g.\ `lips', `throat', `rolled'). Phonetic symbol values have changed 
-(e.g.\ voiceless is no longer marked by <h>). Symbols have been dropped (e.g.\ the 
-caret diacritic denoting `long and narrow' is longer used). And many symbols have 
-been added to reflect contrastive sounds found in the world's very diverse 
-phonological systems.
+articulation labels -- what often are call features even though the IPA avoids
+this term -- have changed (e.g.\ terms like `lips', `throat' or `rolled' are not
+used anymore). Phonetic symbol values have changed (e.g.\ voiceless is no longer
+marked by <h>). Symbols have been dropped (e.g.\ the caret diacritic denoting
+`long and narrow' is no longer used). And many symbols have been added to reflect
+contrastive sounds found in the world's very diverse phonological systems.
 
@@
 suggest to try and avoid using the tilde overlay, though for reasons of backward 
-compatibility we will allow it in strict-IPA.\@
+compatibility we will allow it in valid-IPA.\@
 
@@
 % length IPA 1999:22
-% "Note: as in Finnish orthography, length can also be indicated in phonetic transcription by double letters: e.g . Finnish maatto [rncctto] 'electrical earth/ground'."
+% "Note: as in Finnish orthography, length can also be indicated in phonetic
+% transcription by double letters: e.g . Finnish maatto [rncctto] 'electrical
+% earth/ground'."
 
 % syllable breaks and word boundaries <\s, ., |, ||, tie-bar below>
-% White spaces can be used to indicate word boundaries.
-Syllable breaks can be marked when required.
-The other two boundary symbols are used to mark the domain of larger prosodic units.
-There is also a linking symbol that can be used for explicitly indicating the lack of a boundary.
-
-%  (In all these transcriptions, no spaces between words have been included.
-This is inevitable in an impressionistic transcription where it is not yet known how the utterance divides into words.
-In phonemic and allophonic transcriptions it is common to include spaces to aid legibility, but their theoretical validity is problematic.)
+% White spaces can be used to indicate word boundaries. Syllable breaks can be
+% marked when required. The other two boundary symbols are used to mark the domain
+% of larger prosodic units. There is also a linking symbol that can be used for
+% explicitly indicating the lack of a boundary.
+
+% (In all these transcriptions, no spaces between words have been included. This
+% is inevitable in an impressionistic transcription where it is not yet known how
+% the utterance divides into words. In phonemic and allophonic transcriptions it
+% is common to include spaces to aid legibility, but their theoretical validity is
+% problematic.)
 
@@
 
-% Wells1995:10 "The underscore could in principle also be pressed into service to represent the IPA tie bar.
-The current chart mentions its use only for affricates and double articulations, and then only "if necessary"."
-
-In the major revision of the IPA in 1932, affricates were represented by two consonants 
-(e.g.\ <tʃ>), or ligatures <ʧ>, or with the tie-bar <t͡ʃ>. In the 1938 revision the tie-bar's 
-semantics were broadened to indicate simultaneous articulation, e.g.\ in labial velars 
-such as <k͡p>. Thus the tie-bar is a convenient diacritic for visually tokenizing input 
-strings into chunks of phonetically salient groups, including affricates, doubly articulated 
-consonants or diphthongs.
-
-The tie bar can be placed above or below the base characters, e.g. <\charis{t͡s}>
-or <\charis{t͜s}>.\footnote{Note that the \textsc{undertie} at \uni{203F} symbol 
-looks like and is easily confused with the tie-bar below. However, it is a different 
-character and has a different function in IPA. The undertie is used as a linking 
-symbol to indicate the lack of a boundary, e.g.\ petit ami [pətit‿ami] `boyfriend' 
-($<$ French).} IPA allows both options. The choice between the two symbols
-is purely for legible rendering; there is no difference in semantics between the
-two symbols. However, rendering is such a problematic issue for tie bars in
-general that many linguists simply do not use them. Just looking at a few
-different fonts already indicates that actually no font designer really gets the
-placement right in combination with superscripts and subscripts. If really
-necessary, we propose to standardize on the tie bar above the base characters,
-using a \textsc{combining double inverted breve} at \uni{0361}.
+% Wells1995:10 "The underscore could in principle also be pressed into service
+% to represent the IPA tie bar. The current chart mentions its use only for
+% affricates and double articulations, and then only "if necessary"."
+
+In the major revision of the IPA in 1932, affricates were represented by two
+consonants <tʃ>, ligatures <ʧ>, or with the tie-bar <t͡ʃ>. In the
+1938 revision the tie-bar's semantics were broadened to indicate simultaneous
+articulation, as for example in labial velars such as <k͡p>. Thus, the tie-bar is a
+convenient diacritic for visually tokenizing input strings into chunks of
+phonetically salient groups, including affricates, doubly articulated consonants
+or diphthongs.
+
+The tie bar can be placed above or below the base characters, e.g.
+<\charis{t͡s}> or <\charis{t͜s}>. IPA allows both options. The choice between
+the two symbols is purely for legible rendering; there is no difference in
+semantics between the two symbols. However, rendering is such a problematic
+issue for tie bars in general that many linguists simply do not use them. Just
+looking at a few different fonts already indicates that actually no font
+designer really gets the placement right in combination with superscripts and
+subscripts. If really necessary, we propose to standardize on the tie bar above
+the base characters, using a \textsc{combining double inverted breve} at
+\uni{0361}.\footnote{Also note that the \textsc{undertie} at \unif{203F} looks
+like the tie bar below and is easily confused with it. However, it is a
+different character and has a different function in IPA. The undertie is used as
+a linking symbol to indicate the lack of a boundary, e.g.\ French \textit{petit
+ami} [pətit\charis{‿}ami] `boyfriend'.}
 
@@
 
-% TODO: link back to discussion on IPA principles and how ligatures got brought along historically
+% TODO: link back to discussion on IPA principles and how ligatures got brought
+% along historically
 
@@
 characters exist as well as precomposed combination in Unicode, and these
-precomposed characters should preferably be used.\footnote{The IPA does not
-describe any character for a voiced retroflex implosive, which would
-transparently be \charis{ᶑ}. We propose to add this character to the IPA, see
-Section~\ref{ipa-additions}.} When instead combinations of a base character with
+precomposed characters should preferably be used. When instead combinations of a base character with
 diacritic are used, then these combinations are not canonical equivalent to the
@@
 
+% \footnote{The IPA does not
+% describe any character for a voiced retroflex implosive, which would
+% transparently be \charis{ᶑ}. We propose to add this character to the IPA, see
+% Section~\ref{ipa-additions}.}
+
 A similar problem arises with the rhotic hook. There are two precomposed
@@
 <\dia{02DE}> \textsc{modifier letter rhotic hook} at \uni{02DE}, because there
-is not complete set of precomposed characters with rhotic hooks. For that reason
-we propose to not use the two precomposed characters with hooks mentioned above,
-but always use the separate rhotic hook at \uni{02DE} in IPA.\@
+is no complete set of precomposed characters with rhotic hooks in Unicode. For
+that reason we propose to not use the two precomposed characters with hooks
+mentioned above, but always use the separate rhotic hook at \uni{02DE} in IPA.\@
 
@@
 letter l with middle tilde} at \uni{026B}. This character is not canonically
-equivalent to the combination <l> + <\dia{0334}>, so one of the two possible
+equivalent to the combination <l>~+~<\dia{0334}>, so one of the two possible
 encodings has to be chosen. Because the tilde overlay is described as a general
@@
 and the typical usage in the form of <ɫ> is (almost) a homoglyph with <ɬ> (see 
-Section~\ref{pitfall-homoglyphs-in-IPA}). For these reasons we suggest to try 
-and avoid the tilde overlay.
+Section~\ref{pitfall-homoglyphs-in-IPA}). For these reasons we also suggest to try 
+and avoid the tilde overlay completely.
 
@@
 is also linguistically correct, because according to the IPA it can be used both
-for aspiration (more precisely post-aspiration following the base character) and
-pre-aspiration (preceding the base character). The default combination of Spacing Modifiers with
-the preceding character can be specified in orthography profiles (see
+for post-aspiration (following the base character) and pre-aspiration (preceding
+the base character). The default combination of Spacing Modifiers with the
+preceding character can be specified in orthography profiles (see
 Chapter~\ref{orthography-profiles}).
@@
 
-There is one aspect of ordering for which the Unicode Standard does present a
-canonical solution. However, it is uncontroversial from a linguistic
-perspective. Diacritics in the Unicode Standard (i.e.~Combining Diacritical
-Marks, see above) are classified in Canonical Combining Classes. In practice,
+There is one (minor) aspect of ordering for which the Unicode Standard does
+present a canonical solution. Fortunately, this is uncontroversial from a
+linguistic perspective. Diacritics in the Unicode Standard (i.e.~Combining
+Diacritical Marks, see Section~\ref{pitfall-different-notions-of-diacritics})
+are classified in so-called \textsc{canonical combining classes}. In practice,
 the diacritics are distinguished by their position relative to the base
 character.\footnote{See
-\url{http://unicode.org/reports/tr44/\#Canonical\_Combining\_Class\_Values} for a
-detailed description.} When applying a Unicode normalization (NFC or NFD, see
+\url{http://unicode.org/reports/tr44/\#Canonical\_Combining\_Class\_Values} for
+a detailed description.} When applying a Unicode normalization (NFC or NFD, see
 Section~\ref{pitfall-canonical-equivalence}), the diacritics in different
 positions are put in a specified order. This process therefore harmonizes the
-difference between different encodings, e.g.\ of an extra-short creaky voice
-vowel <ḛ̆>. This grapheme cluster can be encoded either as
-<e>+<\dia{0306}>+<\dia{0330}> or as <e>+<\dia{0330}>+<\dia{0306}>. To
-prevent this twofold encoding, the Unicode Standard specifies the second
-ordering as canonical (in this case, diacritics below are put before diacritics
-above).
+difference between different encodings in some situations, for example in the
+case of an extra-short creaky voice vowel <ḛ̆>. This grapheme cluster can be
+encoded either as <e>+<\dia{0306}>+<\dia{0330}> or as
+<e>+<\dia{0330}>+<\dia{0306}>. To prevent this twofold encoding, the Unicode
+Standard specifies the second ordering as canonical (namely, diacritics
+below are put before diacritics above).
+
+% The next paragraph does not belong here - we should put it somewhere else
 
 When encoding a string according to the Unicode Standard, it is possible to do
-this either using the NFC (composition) or NFD (decomposition) normalization.
+this either using the NFC (composition) or NFD (decomposition) normalization (see
+Section~\ref{pitfall-canonical-equivalence}).
 Decomposition implies that precomposed characters (like <á> \textsc{latin small
@@
 diacritics (which are not diacritics in the Unicode sense) between diacritics
-below and the tie bar. Within all these classes of diacritics there is no canonical
-ordering specified by Unicode, so we propose an explicit ordering here.
+below and the tie bar. Further, within all these classes of diacritics there is
+no canonical ordering specified by Unicode, so we propose an explicit ordering
+here.
 
@@
 the Association as to which symbols should be added, removed or 
-changed. (Perhaps that's why the IPA is updated infrequently.) 
-For example, in the 1989 revision of the IPA at the Kiel Convention, 
+changed. For example, in the 1989 revision of the IPA at the Kiel Convention, 
 changes to specific symbols (in previous charts) were debated and 
@@
 case was made \cite{Ladefoged1990a}. For example, two such decisions 
-included \cite[62]{Ladefoged1990a}:
-
-\begin{itemize}
-	\item symbols for clicks were changed from < ʇ, ʖ, ʗ > and replaced by < ǀ, ǁ, ǃ > because the latter were the symbols used by nearly all Khoisanists and Bantuists
-	\item the Americanist tradition of < š, ž, č, ǰ > were not adopted because the Association members at the convention ``were not sufficiently impressed by arguments ...
-to the effect that these sounds formed a natural class, and thus is would be appropriate to recognize this by maintaining a common aspect to their symbolism''
+included:
+
+\begin{itemize}
+	\item symbols for clicks were changed from <ʇ~ʖ~ʗ> to <ǀ~ǁ~ǃ>
+       because the latter were the symbols used by nearly all Khoisanists and
+       Bantuists.
+	\item the Americanist tradition of using using <\dia{030C}>, a
+       \textsc{combining caron} at \uni{030C} for all post-alveolar sounds, like
+       in <š~ž~č~ǰ>, was not adopted because the Association members at the
+       convention ``were not sufficiently impressed by arguments ... to the
+       effect that these sounds formed a natural class, and thus is would be
+       appropriate to recognize this by maintaining a common aspect to their
+       symbolism'' \citep[62]{Ladefoged1990a}. 
 \end{itemize}

### 2015-08-01 chapters/ipa_meets_unicode.tex
adding note on superscript h

@@
         typing <\dia{003A}> \textsc{colon} at \uni{003A} is of course much easier.        
+  \item There are two mostly identical-looking Unicode characters for the superscript
+        <ʰ>: the \textsc{combining latin small letter h} at \uni{036A} and the
+        \textsc{modifier letter small h} at \uni{02B0}. Making the situation 
+        even more problematic is that they have different behavior (see 
+        Section~\ref{pitfall-different-notions-of-diacritics}). To harmonize the 
+        bevavior of <ʰ> with other superscript letters, we propose to 
+        standardize on the modifier letter at \uni{02B0} (see 
+        Section~\ref{pitfall-no-unique-diacritic-ordering}).
 
@@
 preceding character can be specified in orthography profiles (see
-Chapter~\ref{orthography-profiles}).
+Chapter~\ref{orthography-profiles}). Note that there is an alternative 
+homoglyph, namely the \textsc{combining latin small letter h} at \uni{036A}. We 
+propose not to use this combining diacritical mark, but to standardize on 
+Unicode modifier letters for all superscript letters in IPA.
 
@@
 
-The IPA extensions cover speech sounds beyond the sound systems of the world's languages, including symbols for pathological speech and paralinguistic functions (e.g.\ Braille?).
+The IPA extensions cover speech sounds beyond the sound systems of the world's
+languages, including symbols for pathological speech and paralinguistic
+functions (e.g.\ Braille?).
 
@@
 
-As noted in Section \ref{}, extensions to the IPA -- in particular ``large-scale'' extensions --
+As noted in Section \ref{}, extensions to the IPA -- in particular
+``large-scale'' extensions --
 
@@
 
-X-SAMPA was an ASCII transliteration and is still purposeful for several programs
-
-There is also the extensions of the IPA designed for disordered speech, although the symbols are occasionally used for transcribing phonetic detail in ``normal'' speech or also cross-linguistic rarities.\footnote{\url{http://phoible.github.io/conventions/}}
-
-% Unicode has its own history; for a summary: \url{http://www.unicode.org/history/summary.html}
+X-SAMPA was an ASCII transliteration and is still purposeful for several
+programs
+
+There is also the extensions of the IPA designed for disordered speech, although
+the symbols are occasionally used for transcribing phonetic detail in ``normal''
+speech or also cross-linguistic
+rarities.\footnote{\url{http://phoible.github.io/conventions/}}
+
+% Unicode has its own history; for a summary:
+\url{http://www.unicode.org/history/summary.html}
 

### 2015-08-03 chapters/ipa_meets_unicode.tex
edits

--- 2015-07-14 chapters/ipa_meets_unicode.tex
@@
 Although the original IPA alphabet was a list of symbols with pronunciation examples 
-from words in different languages, in 1900 the symbols were organized into chart and 
+from words in different languages, in 1900 the symbols were organized into chart form and 
 given phonetic feature labels for consonants (e.g.\ for manner of articulation: `plosives', 
@@
 
-Over the years there has been several revisions, many minor. For example, 
+Over the years there has been many revisions, most minor. For example, 
 articulation labels -- what we now call features even though the IPA avoids this term -- 

### 2015-09-28 chapters/pitfalls.tex
minor additions to pitfall char are not glyphs

@@
 theoretically could have been taken differently, but for some reason or another
-(mostly very good reasons) were accepted as they are now. We call behavior that
+(mostly very good reasons) were accepted as they are now. We call such behavior that
 executes without error but does something different than the user
@@
 rendering (see also Section~\ref{pitfall-faulty-rendering}). Like
-contextually-determined glyphs, ligatures are (intended) artifacts of text
+contextually-determined glyphs, ligatures are (intended) artefacts of text
 processing instructions. Finally, a single glyph may be a part of a
-character, as exemplified by diacritics.
+character, as exemplified by diacritics like the diaeresis <\dia{0308}> in <ë>.
+
+test {\fontspec[Alternative=02]{AlBayan}ب}
 

### 2015-09-28 chapters/orthography_profiles.tex
added some TODOs
minor correction 4.1

@@
 (rightly so) not interested in descriptions that only apply to a limited set of
-sources (e.g.~only a single specific dictionary).
+sources (e.g.~descriptions for only a single dictionary).
 
@@
 account (e.g.~between vowels is /n/ and after a vowel but before a consonant is
-a nasalized vowel /ṽ/). In our experience, data from each source must be
-individually tokenized into graphemes so that its orthographic structure is
+a nasalized vowel /ṽ/).
+
+In our experience, data from each source must be
+individually tokenized into graphemes so that its orthographic structure can be 
 identified and its contents can be extracted. To extract data for analysis, a
@@
 
-Thus, our key proposal for the characterization of a writing system is to use a
+Our key proposal for the characterization of a writing system is to use a
 grapheme tokenization as an inter-orthographic pivot. Basically, any source
 document is tokenized by graphemes, and only then a mapping to IPA (or any other
-orthographic conversion) is performed. An orthography profile then is a
+orthographic transliteration) is performed. An \textsc{orthography profile} then is a
 description of the units and rules that are needed to adequately model a
@@
 
+% TODO: add tsoshi figure here?
+
 As an example of graphemic tokenization, note the three different levels of
@@
 	\def\labelenumi{\arabic{enumi}.} 
-	\item code points (10 text elements): t s ʰ o ̃ ̰ ´ s h i 
+	\item code points (10 text elements): t s ʰ o \dia{0303} \dia{0330} \dia{0301} s h i 
 	\item grapheme clusters (7 text elements): t s ʰ ṍ̰ s h i 
@@
 required because sequences of code points can differ in their visual and logical
-orders. For example, <õ̰> is ambiguous to whether it is the sequence of + <̰> +
-<̃> or + <̃> + <̰>. Although these two variants are visually homoglyphs,
+orders. For example, <õ̰> is ambiguous to whether it is the sequence of <o> +
+<\dia{0303}> + <\dia{0330}> or <o> + <\dia{0330}> + <\dia{0303}>. Although these two variants are visually homoglyphs,
 computationally they are different. Unicode normalization should be applied to
 this string to reorder the code points into a canonical order, allowing the data
-to be treated canonically equivalently for search and comparison. In (2), the
+to be treated canonically equivalently for search and comparison.
+
+In (2), the
 Unicode code points have been logically normalized and visually organized into
@@
 sequence <õ̰> is normalized and visually grouped together. Note that, the
-MODIFIER LETTER SMALL H at \uni{02B0}, is not grouped with. This is because it
+\textsc{modifier letter small h} at \uni{02B0}, is not grouped with any 
+other character. This is because it
 belongs to Spacing Modifier Letters category in the Unicode Standard. These
@@
 nasalization or creaky diacritic is defined in the Unicode Standard to apply to
-a specified base character). Finally, to arrive at the graphemic tokenization in
+a specified base character).
+
+Finally, to arrive at the graphemic tokenization in
 (3), tailored grapheme clusters are needed (as for example specified in an

### 2015-09-28 chapters/orthography_profiles.tex
update formal specifications

@@
 
+\subsection*{File Format}
 The formal specifications of an orthography profile (or simply \textsc{profile}
@@
 \begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item \textsc{A profile is a} \textsc{Unicode UTF-8 encoded text file} (ideally using NFC, no-BOM, and LF; see Section~\ref{pitfall-file-formats}, Pitfall: File Formats) that includes the information pertinent to the orthography.
-	\item \textsc{A profile is a} \textsc{tab-separated CSV file with an obligatory header line}.
-A minimal profile can have just a single column, in which case there will of course no tabs, but the first line will still be the header.
-For all columns we assume the name in the header of the CSV file to be crucial.
-The actual ordering of the columns is unimportant.
-	\item \textsc{Lines starting with a hash \textless{}\#\textgreater{} are ignored.} Comments and metadata can be included inside the file, but only as complete lines in the profile, to be marked by lines starting with hash \textsc{\#} (\textsc{number sign} at \uni{0023}).
-Hashes somewhere else in the file are to be treated literally, i.e.~hashes are only to be ignored when they occur at the start of a line.\footnote{Comments that belong to specific lines will have to be put in a separate column of the CSV file, e.g.~add a column called \textsc{comments}.
-Further, if the content of a profile contains a hash at the start of a line, either reorder the columns so the hash does not occur at the start of the line, or add a dummy column in front of the data to not have the data start with a hash.}
-	\item \textsc{Metadata are given in commented lines at the beginning of the text file in a basic \textsc{tag: value} format.
-}Metadata about the orthographic description given in the orthography profile includes, minimally, (i) author, (ii) date, (iii) title, (iv) a stable language identifier encoded in BCP 47/ISO 639-3, and (v) bibliographic data for resource(s) that illustrate the orthography described in the profile.
+	\def\labelenumi{A\arabic{enumi}.} 
+	\item \textsc{A profile is a unicode UTF-8 encoded text file}, ideally
+       using NFC, no-BOM, and LF (see Section~\ref{pitfall-file-formats}), that
+       includes the information pertinent to the orthography. 
+	\item \textsc{A profile is a tab-separated CSV file with an obligatory header
+       line}. A minimal profile can have just a single column, in which case
+       there will of course be no tabs, but the first line will still be the
+       header. For all columns we assume the name in the header of the CSV file
+       to be crucial. The actual ordering of the columns is unimportant. 
+	\item \textsc{Metadata are added in a separate UTF-8 text file} using a basic
+       \textsc{tag: value} format. Metadata about the orthographic description
+       given in the orthography profile includes, minimally, (i) author, (ii)
+       date, (iii) title of the profile, (iv) a stable language identifier
+       encoded in BCP 47/ISO 639-3 of the target language of the profile, and (v)
+       bibliographic data for resource(s) that illustrate the orthography
+       described in the profile. Further, (vi) the tokenization method and (vii)
+       the unicode normalisation used should be documented here (see below).
+    	\item \textsc{Separate lines with comments are not allowed}. Comments that
+       belong to specific lines will have to be put in a separate column of
+       the CSV file, e.g.~add a column called \textsc{comments}.            
 \end{enumerate}
 
-The content of a profile consists of lines, each describing a grapheme of the
-orthography, using the following columns:
+\noindent The content of a profile consists of lines, each describing a grapheme
+of the orthography, using the following columns:
 
 \begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item \textsc{A minimal profile consists of a single column with header \textsc{graphemes}}, listing each of the different graphemes in a separate line. 
-	\item \textsc{Optional columns called \textsc{left} and \textsc{right} can be used to specify the left and right context of the grapheme, respectively.} The same grapheme can occur multiple times with different contextual specifications, for example to distinguish different pronunciations depending on the context.
-	\item \textsc{The columns \textsc{grapheme}, \textsc{left} and \textsc{right} can use regular expression metacharacters.} If regular expressions are used, then all literal usage of the special symbols, like full stops <.> or dollar signs <\$> (so-called \textsc{metacharacters}) have to be explicitly escaped by adding a backslash before them (i.e.~use <.> or <\$>).
-Note that any specification of context automatically expects regular expressions, so it is probably better to always escape all regular expression metacharacters when used literally in the orthography, i.e.~the following symbols will need to be preceded by a backslash: {[} {]} ( ) \{ \} ~+ * .
-- !
-?
-\^{} \$ .
-	\item \textsc{An optional column called \textsc{class} can be used to specify classes of graphemes}, for example to define a class of vowels.
-Users can simply add ad-hoc identifiers in this column to indicate a group of graphemes, which can then be used in the description of the graphemes or the context.
-The identifiers should of course be chosen such that they do not conflate with any symbols used in the orthography themselves.
-Note that such classes only refer to the graphemes, not to the context.
-	\item \textsc{Columns describing transliterations for each graphemes can be added and named at will}.
-Often more than a single possible transliteration will be of interest.
-Any software application using these profiles should use the names of these columns to select a specific transliteration column.
-	\item \textsc{Any other columns can be added freely, but will mostly be ignored by any software application using the profiles}.
-As orthography profiles are also intended to be read and interpreted by humans, it is often highly useful to add extra information on the graphemes in further columns, like for example Unicode codepoints, Unicode names, frequency of occurrence, examples of occurrence, explanation of the contextual restrictions, or comments.
+	\def\labelenumi{A\arabic{enumi}.} \setcounter{enumi}{4} 
+	\item \textsc{A minimal profile consists of a single column} with a header
+       called \texttt{Grapheme}, listing each of the different graphemes in a
+       separate line. The name of this column is crucial for automatic 
+       processing.
+	\item \textsc{Optional columns can be used to specify the left and right
+       context of the grapheme}, to be designated with the headers \texttt{Left}
+       and \texttt{Right}, respectively. The same grapheme can occur multiple
+       times with different contextual specifications, for example to
+       distinguish different pronunciations depending on the context. 
+	\item \textsc{The columns \texttt{Grapheme}, \texttt{Left} and \texttt{Right}
+       can use regular expression metacharacters.} If regular expressions are
+       used, then all literal usage of the special symbols, like full stops <.>
+       or dollar signs <\$> (so-called \textsc{metacharacters}) have to be
+       explicitly escaped by adding a backslash before them (i.e.~use
+       <\textbackslash.> or <\textbackslash\$>). Note that any specification of
+       context automatically expects regular expressions, so it is probably
+       better to always escape all regular expression metacharacters when used
+       literally in the orthography. The following symbols will need to be
+       preceded by a backslash: {[} {]} ( ) \{ \} ~+ * . - ! ? \^{} \$ and the
+       backslash \textbackslash~itself. 
+	\item \textsc{An optional column can be used to specify classes of graphemes},
+       to be identified by the header \texttt{Class}. For example, this column
+       can be used to define a class of vowels. Users can simply add ad-hoc
+       identifiers in this column to indicate a group of graphemes, which can
+       then be used in the description of the graphemes or the context. The
+       identifiers should of course be chosen such that they do not conflate
+       with any symbols used in the orthography themselves. Note that such
+       classes only refer to the graphemes, not to the context. 
+	\item \textsc{Columns describing transliterations for each graphemes can be
+       added and named at will}. Often more than a single possible
+       transliteration will be of interest. Any software application using these
+       profiles should prompt the user to name any of these columns to select a
+       specific transliteration. 
+	\item \textsc{Any other columns can be added freely, but will mostly be ignored
+       by any software application using the profiles}. As orthography profiles
+       are also intended to be read and interpreted by humans, it is often
+       highly useful to add extra information on the graphemes in further
+       columns, like for example Unicode codepoints, Unicode names, frequency of
+       occurrence, examples of occurrence, explanation of the contextual
+       restrictions, or comments. 
  \end{enumerate}
 
+\subsection*{Implementation}
 For the automatic processing of the profiles, the following technical standards
@@
 \begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item \textsc{Each line of a profile will be interpreted as a regular expression.
-}Software applications using profiles can also offer to interpret a profile in the literal sense to avoid the necessity for the user to escape regular expressions metacharacters in the profile.
-However, this only is possible when no contexts or classes are described, so this seems only useful in the most basic orthographies.
-	\item \textsc{The \textsc{class} column will be used to produce explicit \textsc{or} chains of regular expressions}, which will then be inserted in the \textsc{graphemes}, \textsc{left} and \textsc{right} columns at the position indicated by the class-identifiers.
-For example, a class \textsc{V} as a context specification might be replaced by a regular expression like: (a\textbar{}e\textbar{}i\textbar{}o\textbar{}u\textbar{}ei\textbar{}au).
-Only the graphemes themselves are included here, not any contexts specified for the elements of the class.
-	\item \textsc{The \textsc{left} and \textsc{right} contexts will be included into the regular expressions by using lookbehind and lookahead}.
-Basically, the actual regular expression syntax of lookbehind and lookahead is simply hidden to the users by allowing them to only specify the contexts themselves.
-Internally, the contexts in the columns \textsc{left} and \textsc{right} are combined with the column \textsc{graphemes} to form a complex regular expression like: (?\textless{}=left)graphemes(?=right).
-	\item \textsc{The regular expressions will be applied in the order as specified in the profile, from top to bottom.} A software implementation can offer help in figuring out the optimal ordering of the regular expressions, but should then explicitly report on the order used.
+	\def\labelenumi{B\arabic{enumi}.} 
+	\item \textsc{Each line of a profile will be interpreted as a regular
+       expression. }Software applications using profiles can also offer to
+       interpret a profile in the literal sense to avoid the necessity for the
+       user to escape regular expressions metacharacters in the profile.
+       However, this only is possible when no contexts or classes are described,
+       so this seems only useful in the most basic orthographies. 
+	\item \textsc{The \textsc{class} column will be used to produce explicit
+       \textsc{or} chains of regular expressions}, which will then be inserted
+       in the \texttt{Grapheme}, \texttt{Left} and \texttt{Right} columns at
+       the position indicated by the class-identifiers. For example, a class
+       called \texttt{V} as a context specification might be replaced by a regular
+       expression like:
+       (a\textbar{}e\textbar{}i\textbar{}o\textbar{}u\textbar{}ei\textbar{}au).
+       Only the graphemes themselves are included here, not any contexts
+       specified for the elements of the class. Note that in some cases the 
+       ordering inside this regular expression might be crucial.
+	\item \textsc{The \textsc{left} and \textsc{right} contexts will be included
+       into the regular expressions by using lookbehind and lookahead}.
+       Basically, the actual regular expression syntax of lookbehind and
+       lookahead is simply hidden to the users by allowing them to only specify
+       the contexts themselves. Internally, the contexts in the columns
+       \texttt{Left} and \texttt{Right} are combined with the column
+       \texttt{Grapheme} to form a complex regular expression like:\ 
+       \texttt{(?\textless{}=Left)Grapheme(?=Right)}. 
+	\item \textsc{The regular expressions will be applied in the order as specified
+       in the profile, from top to bottom.} A software implementation can offer
+       help in figuring out the optimal ordering of the regular expressions, but
+       should then explicitly report on the order used.      
 \end{enumerate}
 
-The actual implementation of the profile on some text-string will function as
+\noindent The actual implementation of the profile on some text-string will function as
 follows:
@@
 \begin{enumerate}
-	\def\labelenumi{\arabic{enumi}.} 
-	\item \textsc{All graphemes are matched in the text before they are tokenized or transliterated}.
-In this way, there is no necessity for the user to consider `feeding' and `bleeding' situations, in which the application of a rule either changes the text so another rule suddenly applies (feeding) or prevents another rule to apply (`bleeding').
-	\item \textsc{The matching of the graphemes can occur either globally or linearly.
-}From a computer science perspective, the most natural way to match graphemes from a profile in some text is by walking linearly through the text-string from left to right, and at each position go through all graphemes in the profile to see which one matches, then go to the position at the end of the matched grapheme and start over.
-This is basically how a finite state transducer works, which is a well-established technique in computer science.
-However, from a linguistic point of view, our experience is that most linguists find it more natural to think from a global perspective.
-In this approach, the first grapheme in the profile is matched everywhere in the text-string first, before moving to the next grapheme in the profile.
-Theoretically, these approaches will lead to different results, though in practice of actual natural language orthographies they almost always lead to the same result.
-Still, we suggest that any software application using orthography profiles should offer both approaches (i.e.
-\textsc{global} or \textsc{linear}) to the user.
-The approach used should be documented in the metadata as \textsc{tokenization method}.
-	\item \textsc{The matching of the graphemes can occur either in NFC or NFD.
-}By default, both the profile and the text-string to be tokenized should be treated as NFC (see section \ref{pitfall-canonical-equivalence}, Pitfall: Canonical equivalence, above).
-However, in some use-cases it turns out to be practical to treat both text and profile as NFD.
-This typically happens when very many different combinations of diacritics occur in the data.
-An NFD-profile can then be used to first check which individual diacritics are used, before turning to the more cumbersome inspection of all combinations.
-We suggest that any software application using orthography profiles should offer both approaches (i.e.
-\textsc{NFC} or \textsc{NFD}) to the user.
-The approach used should be documented in the metadata as \textsc{unicode normalization}.
-	\item \textsc{The text-string is always returned in tokenized form} by separating the matched graphemes by a user-specified symbols-string.
-Any transliteration will be returned on top of the tokenization.
-	\item \textsc{Leftover characters (i.e.~characters that are not matched by the profile) should be reported to the user as errors.} Typically, the unmatched character are replaced in the tokenization by a user-specified symbol-string.
+	\def\labelenumi{B\arabic{enumi}.} \setcounter{enumi}{4} 
+	\item \textsc{All graphemes are matched in the text before they are tokenized
+       or transliterated}. In this way, there is no necessity for the user to
+       consider `feeding' and `bleeding' situations, in which the application of
+       a rule either changes the text so another rule suddenly applies (feeding)
+       or prevents another rule to apply (`bleeding'). 
+	\item \textsc{The matching of the graphemes can occur either globally or
+       linearly. }From a computer science perspective, the most natural way to
+       match graphemes from a profile in some text is by walking linearly
+       through the text-string from left to right, and at each position go
+       through all graphemes in the profile to see which one matches, then go to
+       the position at the end of the matched grapheme and start over. This is
+       basically how a finite state transducer works, which is a
+       well-established technique in computer science. However, from a
+       linguistic point of view, our experience is that most linguists find it
+       more natural to think from a global perspective. In this approach, the
+       first grapheme in the profile is matched everywhere in the text-string
+       first, before moving to the next grapheme in the profile. Theoretically,
+       these approaches will lead to different results, though in practice of
+       actual natural language orthographies they almost always lead to the same
+       result. Still, we suggest that any software application using orthography
+       profiles should offer both approaches (i.e. \textsc{global} or
+       \textsc{linear}) to the user. The approach used should be documented in
+       the metadata as \textsc{tokenization method}. 
+	\item \textsc{The matching of the graphemes can occur either in NFC or NFD.} By
+       default, both the profile and the text-string to be tokenized should be
+       treated as NFC (see Section~\ref{pitfall-canonical-equivalence}). However, in some use-cases it turns out to
+       be practical to treat both text and profile as NFD. This typically
+       happens when very many different combinations of diacritics occur in the
+       data. An NFD-profile can then be used to first check which individual
+       diacritics are used, before turning to the more cumbersome inspection of
+       all combinations. We suggest that any software application using
+       orthography profiles should offer both approaches (i.e. \textsc{NFC} or
+       \textsc{NFD}) to the user. The approach used should be documented in the
+       metadata as \textsc{unicode normalization}. 
+	\item \textsc{The text-string is always returned in tokenized form} by
+       separating the matched graphemes by a user-specified symbols-string. Any
+       transliteration will be returned on top of the tokenization. 
+	\item \textsc{Leftover characters}, i.e.~\textsc{characters that are not
+       matched by the profile, should be reported to the user as errors.}
+       Typically, the unmatched character are replaced in the tokenization by a
+       user-specified symbol-string.       
  \end{enumerate}
+
+\subsection*{Software applications}
 
@@
 	\begin{enumerate}
-		\def\labelenumii{\arabic{enumii}.} 
+		\def\labelenumii{C\arabic{enumii}.} 
 		\item \textsc{the name of the column to be used for transliteration} (if any). 
-		\item \textsc{the symbol-string to be inserted between graphemes.} Optionally, a warning might be given if the chosen string includes characters from the orthography itself. 
-		\item \textsc{the symbol-string to be inserted for unmatched strings} in the tokenized and transliterated output. 
-		\item \textsc{the tokenization method}, i.e.~whether the tokenization should proceed \textsc{global} or \textsc{linear}. 
-		\item \textsc{unicode normalization}, i.e.~whether the text-string and profile should use \textsc{NFC} or \textsc{NFD}. 
+		\item \textsc{the symbol-string to be inserted between graphemes.} Optionally,
+        a warning might be given if the chosen string includes characters from
+        the orthography itself. 
+		\item \textsc{the symbol-string to be inserted for unmatched strings} in the
+        tokenized and transliterated output. 
+		\item \textsc{the tokenization method}, i.e.~whether the tokenization should
+        proceed \textsc{global} or \textsc{linear}. 
+		\item \textsc{unicode normalization}, i.e.~whether the text-string and profile
+        should use \textsc{NFC} or \textsc{NFD}. 
     \end{enumerate}
@@
 	\begin{enumerate}
-		\def\labelenumii{\arabic{enumii}.} \setcounter{enumii}{5} 
-		\item \textsc{assist in the ordering of the graphemes.} In our experience, it makes sense to apply larger graphemes before shorter graphemes, and to apply graphemes with context before graphemes without context.
-Further, frequently relevant rules might be applied after rarely relevant rules (though frequency is difficult to establish in practice, as it depends on the available data).
-Also, if this all fails to give any decisive ordering between rules, it seems useful to offer linguists the option to reverse the ordering from any manual specified ordering, because linguists tend to write the more general rule first, before turning to exceptions or special cases.
-		\item \textsc{assist in dealing with upper and lower case characters.} It seems practical to offer some basic case matching, so characters like <a> and <A> are treated equally.
-This will be useful in many concrete cases, although the user should be warned that case matching does not function universally in the same way across orthographies.
-Ideally, users should prepare orthography profiles with all lowercase and uppercase variants explicitly mentioned, so by default no case matching should be performed.
-		\item \textsc{treat the profile literal}, i.e.~to not interpret regular expression metacharacters.
-Matching graphemes literally often leads to strong speed increase, and would allow users to not needing to worry about escaping metacharacters.
-However, in our experience all actually interesting use-cases of orthography profiles include some contexts, which automatically prevents any literal interpretation, so by default the matching should not be literal.
+		\def\labelenumii{C\arabic{enumii}.} \setcounter{enumii}{5} 
+		\item \textsc{assist in the ordering of the graphemes.} In our experience, it
+        makes sense to apply larger graphemes before shorter graphemes, and to
+        apply graphemes with context before graphemes without context. Further,
+        frequently relevant rules might be applied after rarely relevant rules
+        (though frequency is difficult to establish in practice, as it depends
+        on the available data). Also, if this all fails to give any decisive
+        ordering between rules, it seems useful to offer linguists the option to
+        reverse the ordering from any manual specified ordering, because
+        linguists tend to write the more general rule first, before turning to
+        exceptions or special cases. 
+		\item \textsc{assist in dealing with upper and lower case characters.} It
+        seems practical to offer some basic case matching, so characters like
+        <a> and <A> are treated equally. This will be useful in many concrete
+        cases, although the user should be warned that case matching does not
+        function universally in the same way across orthographies. Ideally,
+        users should prepare orthography profiles with all lowercase and
+        uppercase variants explicitly mentioned, so by default no case matching
+        should be performed. 
+		\item \textsc{treat the profile literal}, i.e.~to not interpret regular
+        expression metacharacters. Matching graphemes literally often leads to
+        strong speed increase, and would allow users to not needing to worry
+        about escaping metacharacters. However, in our experience all actually
+        interesting use-cases of orthography profiles include some contexts,
+        which automatically prevents any literal interpretation.
     \end{enumerate}
@@
 	\begin{enumerate}
-		\def\labelenumii{\arabic{enumii}.} \setcounter{enumii}{8} 
-		\item \textsc{the original text-strings to be processed in the used Unicode normalization}, i.e.~in either NFC or NFD as specified by the user. 
+		\def\labelenumii{C\arabic{enumii}.} \setcounter{enumii}{8} 
+		\item \textsc{the original text-strings to be processed in the specified
+        unicode normalization}, i.e.~in either NFC or NFD as specified by the
+        user. 
 		\item \textsc{the tokenized strings}, with additionally any transliterated

### 2015-09-29 book/LSP/lsp-seriesinfo/cfls-info.tex
moves book contents into book directory

%% -*- coding:utf-8 -*-
{\large Conceptual Foundations of Language Science}

\bigskip

\textbf{Series editors}\\
Mark Dingemanse \textit{Max Planck Institute for Psycholinguistics}   \\
Nick J. Enfield \textit{University of Sydney}

\bigskip

\textbf{Editorial board} \\
Balthasar Bickel \textit{University of Zürich}   \\
Claire Bowern \textit{Yale University}          \\
Elizabeth Couper-Kuhlen \textit{University of Helsinki}    \\
William Croft \textit{University of New Mexico}            \\
Rose-Marie Déchaine \textit{University of British Columbia}      \\
William A. Foley \textit{University of Sydney}                   \\
William F. Hanks \textit{University of California at Berkeley}   \\
Paul Kockelman \textit{Yale University}           \\
Keren Rice \textit{University of Toronto}         \\
Sharon Rose \textit{University of California at San Diego }       \\
Frederick J. Newmeyer \textit{University of Washington}        \\
Wendy Sandler \textit{University of Haifa }            \\
Dan Sperber \textit{Jean Nicod Institute}     \\

\bigskip

\begin{minipage}{\textwidth}% for undoing \raggedright and justify the following two paragraphs

No scientific work proceeds without conceptual foundations. In language science, our concepts about language determine our assumptions, direct our attention, and guide our hypotheses and our reasoning. Only with clarity about conceptual foundations can we pose coherent research questions, design critical experiments, and collect crucial data.

\medskip

This series publishes short and accessible books that explore well-defined topics in the conceptual foundations of language science. The series provides a venue for conceptual arguments and explorations that do not require the traditional book-length treatment, yet that demand more space than a typical journal article allows.  
\end{minipage}

\bigskip

In this series:

\begin{enumerate}
\item Nick J. Enfield. \textit{Natural Causes of Language}.

\end{enumerate}

% 
% \vfill
% 
% \gdef\lsISSN{2363-5568}%
% \hfill ISSN: \lsISSN

### 2015-09-29 book/LSP/lsp-seriesinfo/sidl-info.tex
moves book contents into book directory

%% -*- coding:utf-8 -*-
{\large Studies in Diversity Linguistics}

\bigskip

Chief Editor: Martin Haspelmath \\
Consulting Editors: Fernando Zúñiga, Peter Arkadiev, Ruth Singer, Pilar Valen­zuela

\bigskip

In this series:

\begin{enumerate}
\item Handschuh, Corinna. A typology of marked-S languages
\item Rießler, Michael. Adjective attribution
\item Klamer, Marian (ed.). The Alor-Pantar languages: History and typology
\item Berghäll, Liisa. A grammar of Mauwake (Papua New Guinea)
\item Wilbur, Joshua. A grammar of Pite Saami
\end{enumerate}

\vfill

\gdef\lsISSN{2363-5568}%
\hfill ISSN: \lsISSN

### 2015-09-29 book/chapters/_remove/encoding.tex
moves book contents into book directory

--- 2015-06-08 chapters/introduction.tex
@@
-\chapter{Writing Systems}
-\label{writing_systems}
-
-% ==========================
-\section{Introduction}
-\label{introduction}
-% ==========================
-
-Writing systems arise and develop in a complex mixture of cultural,
-technological and practical pressures. They tend to be highly conservative, in
-that people who have learned to read and write in a specific way---however
-impractical or tedious---are mostly unwilling to change their habits. Writers
-tend to resist spelling reforms. In all literate societies there exists a strong
-socio-political mainstream that tries to force unification of writing (for
-example by strongly enforcing ``right'' from ``wrong'' writing in schools).
-However, there is also a large community of users who take as many liberties in
-their writing as they can get away with.
-
-For example, the writing of tone diacritics in Yoruba is often proclaimed to be
-the right way to write, although many users of Yoruba orthography seem to be
-perfectly fine with leaving them out. As pointed out by the proponents of the
-official rules, there are some homographs when leaving out the tone diacritics
-\citep[44]{Olumuyiw2013}. However, writing systems (and the languages they
-represent) are normally full of homophones, which is normally not a problem at
-all for speakers of the language. More importantly, writing is not just a purely
-functional tool, but just as importantly it is a mechanism to signal social
-affiliation. By showing that you \textit{know the rules} of expressing yourself
-in writing, others will more easily accept you as a worthy participant in their
-group. And that just as well holds for obeying to the official rules when
-writing a job application, as for obeying to the informal rules when writing an
-SMS to classmates in school. The case of Yoruba writing is an exemplary case, as
-even after more than a century of efforts to standardize the writing systems,
-there is still a wide range of variation in daily use \citep{Olumuyiw2013}.
-
-The sometimes cumbersome and sometimes illogical structure, and the enormous
-variability of existing writing systems is a fact of life scholars have to
-accept and should try to adapt to as good as possible. Our plea here is a
-proposal for a formalization to do exactly that.
-
-When considering the worldwide linguistic diversity, including all
-lesser-studied and endangered languages, there exist numerous different
-orthographies using symbols from the same scripts. For example, there are
-hundreds of orthographies using Latin-based alphabetic scripts. All of these
-orthographies use the same symbols, but these symbols differ in meaning and
-usage throughout the various orthographies. To be able to computationally use
-and compare different orthographies, we need a way to specify all orthographic
-idiosyncrasies in a computer-readable format (a process called
-\textsc{tailoring} in Unicode parlance). We call such specifications
-\textsc{orthography profiles}. Ideally, these specifications have to be
-integrated into so-called Unicode locale descriptions, though we will argue that
-in practice this is often not the most useful solution for the kind of problems
-arising in the daily practice of linguistics. Consequently, a central goal of
-this paper is to flesh out the linguistic challenges for locale descriptions,
-and work out suggestions to improve their structure for usage in a linguistic
-context. Conversely, we also aim to improve linguists' understanding and
-appreciation for the accomplishments of the Unicode Consortium in the
-development of the Unicode Standard.
-
-The necessity to computationally use and compare different orthographies most
-forcefully arises in the context of language comparison. Concretely, in our
-current research our goal is to develop quantitative methods for language
-comparison and historical analysis in order to investigate worldwide linguistic
-variation and to model the historical and areal processes that underlie
-linguistic diversity,
-cf.~\citet{Steiner_etal2011,List2012,List2012a,ListMoran2013,MoranProkic2013}.
-In this work, it is crucial to be able to flexibly process across numerous
-resources with different orthographies. In many cases even different resources
-on the \textit{same} language use different orthographic conventions. Another
-orthographic challenge that we encounter regularly in our linguistic practice is
-electronic resources on a particular language that claim to follow a specific
-orthographic convention (often a resource-specific convention), but on closer
-inspection such resources are almost always not consistently encoded. Thus a
-second goal of our orthography profiles is to allow for an easy specification of
-orthographic conventions, and use such profiles to check consistency and to
-report errors to be corrected.
-
-A central step in our proposed solution to this problem is the tailored grapheme
-separation of strings of symbols, a process we call \textsc{grapheme
-tokenization}. Basically, given some strings of symbols (e.g.~morphemes, words,
-sentences) in a specific source, our first processing step is to specify how
-these strings have to be separated into graphemes, considering the specific
-orthographic conventions used in a particular source document. Our experience is
-that such a graphemic tokenization can be performed without extensive in-depth
-knowledge about the phonetic and phonological details of the language in
-question. For example, the specification that $<$ou$>$ is a grapheme of English
-is a much easier task than to specify what exactly the phonetic values of this
-grapheme are in any specific occurrence in English words. Grapheme separation is
-a task that can be performed relatively reliably and with limited availability
-of time and resources (compare, for example, the task of creating a complete
-phonetic or phonological normalization).
-
-Although grapheme tokenization is only one part of the solution, it is an
-important and highly fruitful processing step. Given a grapheme tokenization,
-various subsequent tasks become easier, like (a) temporarily reducing the
-orthography in a processing pipeline, e.g.~only distinguishing high versus low
-vowels; (b) normalizing orthographies across sources (often including temporary
-reduction of oppositions), e.g.~specifying an (approximate) mapping to the
-International Phonetic Alphabet; (c) using co-occurrence statistics across
-different languages (or different sources in the same language) to estimate the
-probability of grapheme matches, e.g.~with the goal to find regular sound
-changes between related languages or transliterations between different sources
-in the same language.
-
-Before we deal with these proposals, in the first part of this paper (Sections
-~\ref{encoding} through~\ref{ipa-meets-unicode}) we give an extended introduction
-to the notion of encoding (Section~\ref{encoding}) and writing systems, both
-from a linguistic perspective and from the perspective of the Unicode Consortium
-(Section~\ref{terminology}). We consider the Unicode Standard to be a
-breakthrough (and ongoing) development that fundamentally changed the way we
-look at writing systems, and we aim to provide here a slightly more in-depth
-survey of the many techniques that are available in the standard. A good
-appreciation for the solutions that the Unicode Standard also allows for a
-thorough understanding of possible pitfalls that one might encounter when using
-it (Section~\ref{unicode-pitfalls}). As an example of the current
-state-of-the-art, we discuss the rather problematic marriage of the
-International Phonetic Alphabet (IPA) with the Unicode Standard (Section
-\ref{ipa-meets-unicode}).
-
-The second part of the paper (Sections~\ref{orthography-profiles} and
-\ref{use-cases}) describes our proposals for how to deal with the Unicode
-Standard in the daily practice of (comparative) linguists. First, we discuss the
-challenges of characterizing a writing system. To solve these problems, we
-propose the notions of orthography profiles, closely related to Unicode locale
-descriptions (Section~\ref{orthography-profiles}). Finally, we discuss practical
-issues with actual examples (Section~\ref{use-cases}). We provide reference
-implementation of our proposals in R and in Python, available as open-source
-libraries.
-
-The following conventions are followed in this paper. All phonemic and phonetic
-representations are given in the International Phonetic Alphabet (IPA), unless
-noted otherwise \citep{IPA2005}. Standard conventions are used for
-distinguishing between graphemic < >, phonemic / / and phonetic [ ]
-representations. For character descriptions, we follow the notational
-conventions of the Unicode Standard \citep{Unicode2014}. Character names are
-represented in small capital letters (e.g.~\textsc{latin small letter schwa})
-and code points are expressed as U\emph{+n} where \emph{n} is a four to six
-digit hexadecimal number, e.g.~\uni{0256}, which can be rendered as the glyph <ə>.
-
-% ==========================
 \section{Encoding}
 \label{encoding}
-% ==========================
 
@@
 code is not so often told, so we decided to offer a short survey of the major
-developments of such encoding here.\footnote{Because of the recent
+developments of such \textsc{encoding} here.\footnote{Because of the recent
 history as summarized in this section, we have used mostly rather ephemeral
@@
 messages over large distances. Originally, various specially designed
-contraptions were used to send messages. Today, descendants of these systems are
-still in limited use, for example utilizing flags or flashing lights. The
-``innovation'' of those semaphoric systems was that all characters of the
-written language were replaced one-to-one by visual signals. Since then, all
-telegraphic systems have taken this principle, namely that any language to be
-transmitted first has to be turned into some orthographic system, which
-subsequently is encoded for transmission by the sender, and then turned back
-into orthographic representation at the receiver side.\footnote{Sound and
-video-based telecommunication of course takes a different approach by ignoring
-the written version of language and directly encode sound waves or light
-patterns.} This of course implies that the usefulness of any such telegraphic
-encoding completely depends on the sometimes rather haphazard structure of
-orthographic systems.
+contraptions were used to send messages. Today, descendants of these systems
+are still in use utilizing flags or flashing lights. The ``innovation'' of those
+semaphoric systems was that all characters of the written language were replaced
+one-to-one by visual signals. Since then, all telegraphic systems have taken
+this principle, namely that any language to be transmitted first has to be
+turned into some orthographic system, which subsequently is encoded for
+transmission by the sender, and then turned back into orthographic
+representation at the receiver side.\footnote{Sound and video-based
+telecommunication of course takes a different approach by ignoring the written
+version of language and directly encode sound waves or light patterns.} This of
+course implies that the usefulness of any such telegraphic encoding completely
+depends on the sometimes rather haphazard structure of orthographic systems.
 
@@
 by counting until a fixed number of signals has passed.\footnote{Of course, no
-explicit separator is needed at all when the timing of the signals is known, which is
+explicit separator is needed when the timing of the signals is known, which is
 the principle used in all modern telecommunication systems. An important modern
@@
 Character Set} (MCS, with $2^8 = 256$ different possible characters), first used
-with the introduction of the VT{\large 220} Terminal in 1983.
+with the introduction of the {\small VT}220 Terminal in 1983.
 
@@
 Greek.\footnote{In effect, because $16 = 2^4$, this means that ISO/IEC~8859 was
-actually an $8+4=12$-bit encoding, though with very many duplicates by design,
-namely all ASCII codes were repeated in each 8-bit code page. To be precise,
+actually an $8+4=12$-bit encoding, though with very many duplicates by design, namely 
+all ASCII codes were repeated in each 8-bit code page. To be precise,
 ISO/IEC~8859 used the 7-bit ASCII as the basis for each code page, and defined
 16 different 7-bit extensions, leading to $(1+16)\cdot{2^7} = 2,176$ possible
-characters. However, because of overlap and not-assigned codes points the actual
-number of symbols was much smaller.} This system almost immediately was
-understood to be insufficient and impractical, so various initiatives to extend
-and reorganize the encoding started in the 1980s. This led, for example, to
-various proprietary encodings from Microsoft (e.g.~Windows Latin 1) and Apple
-(e.g.~Mac OS Roman), which one still sometimes encounters today.
+characters.} This system almost immediately was understood to be insufficient and
+impractical, so various initiatives to extend and reorganize the encoding
+started in the 1980s. This led, for example, to various proprietary encodings
+from Microsoft (e.g.~Windows Latin 1) and Apple (e.g.~Mac OS Roman), which one
+still sometimes encounters today.
 
@@
 112,956 different characters \citep{Unicode2014}.
-
-\
-
-\noindent In the next section we provide a very brief overview of the linguistic terminology
-concerning writing systems before turning to the slightly different
-computational terminology in the next section on the Unicode Standard.
-
-% ==========================
-\section{Linguistic terminology}
-\label{linguistic-terminology}
-% ==========================
-
-Linguistically speaking, a \textsc{writing system} is a symbolic system that
-uses visible or tactile signs to represent language in a systematic way. The
-term writing system has two mutually exclusive meanings. First, it may
-refer to the way a particular language is written. In this sense the term refers
-to the writing system of a particular language, as, for example, in \emph{`the
-Serbian writing system uses two scripts: Latin and Cyrillic.'} Second, the term
-writing system may also refer to a type of symbolic system as used among the
-world's languages to represent the language, as, for example, in
-\emph{`alphabetic writing system.'} In this latter sense the term refers to how
-scripts have been classified according to the way that they encode language, as
-in, for example, \emph{`the Latin and Cyrillic scripts are both alphabetic
-writing systems.'} To avoid confusion, this second notion of writing system
-would more aptly have been called \textsc{script system}.
-
-\subsubsection*{Writing systems}
-
-Focussing on the first sense, we distinguish two different kinds of 
-writing systems used for a particular language, namely transcriptions and
-orthographies. First, \textsc{transcription} is a scientific procedure (and also
-the result of that procedure) for graphically representing the sounds of human
-speech at the phonetic level. It incorporates a set of unambiguous symbols to
-represent speech sounds, including conventions that specify how these symbols
-should be combined. A transcription system is a specific system of symbols and
-rules used for transcription of the sounds of a spoken language variety. In
-principle, a transcription system should be language-independent, in that it
-should be applicable to all spoken human languages. The \textsc{International
-Phonetic Alphabet} (IPA) is a commonly used transcription system that provides a
-medium for transcribing languages at the phonetic level. However, there is a
-long history of alternative kinds of transcription systems
-\citep[see][]{Kemp2006} and today various alternatives are in widespread use
-(e.g.~X-SAMPA and Cyrillic-based phonetic transcription systems). Many
-users of IPA do not follow the standard to the letter, and many dialects based
-on the IPA have emerged, e.g.~the Africanist and Americanist transcription
-systems. Note that IPA symbols are also often used to represent language on a
-phonemic level. However, it is important to realize that in this usage the IPA
-symbols are not a transcription system, but rather an orthography (though with
-strong links to the pronunciation). Further, a transcription system does not
-need to be as highly detailed as the IPA.\ It can also be a system of broad sound
-classes. Although such an approximative transcription is not normally used in
-linguistics, it is widespread in technological approaches
-\citetext{\citealp[Soundex and variants, e.g.~][391--392]{Knuth1973};
-\citealp{postel1969,Beider2008}}, and it is sometimes fruitfully used in
-automatic approaches to historical linguistics
-\citep{Dolgopolsky1986,List2012esslli,Brown2013}.
-
-Second, an \textsc{orthography} specifies the symbols, punctuations, and the
-rules in which a specific language is written in a standardized way.
-Orthographies are often based on a phonemic analysis, but they almost always
-include idiosyncrasies because of historical developments (like sound changes or
-loans) and because of the widely-followed principle of lexical integrity
-(i.e.~the attempt to write the same lexical root in a consistent way, also when
-synchronic phonemic rules change the pronunciation, as for example with final
-devoicing in many Germanic languages). All orthographies are language-specific
-(and often even resource-specific), although individual symbols or rules might
-be shared between languages. A \textsc{practical orthography} is a strongly
-phoneme-based writing system designed for practical use by speakers. The mapping
-relation between phonemes and graphemes in practical orthographies is purposely
-shallow, i.e.~there is mostly a systematic and faithful mapping from a phoneme
-to a grapheme. Practical orthographies are intended to jumpstart written
-materials development by correlating a writing system with the sound units of a
-language \citep[cf.~][]{MeinhofJones1928}. Symbols from the IPA are often used
-by linguists in the development of such practical orthographies for languages
-without writing systems, though this usage of IPA symbols should not be confused
-with transcription (as defined above).
-
-Further, a \textsc{transliteration} is a mapping between two different
-orthographies. It is the process of ``recording the graphic symbols of one
-writing system in terms of the corresponding graphic symbols of a second writing
-system'' \citep[396]{Kemp2006}. In straightforward cases, such a transliteration
-is simply a matter of replacing one symbol with another. However, there are
-widespread complications, like one-to-many or many-to-many mappings, which are
-not always easy, or even possible, to solve without listing all cases
-individually \citep[cf.~][Ch.~2]{Moran2012}.
-
-\subsubsection*{Script systems}
-
-Different kinds of writing systems are classified into script
-systems. A \textsc{script} is a collection of distinct symbols as
-employed by one or more orthographies. For example, both Serbian and Russian are
-written with subsets of the Cyrillic script. A single language, like Serbian or
-Japanese, can also be written using orthographies based on different scripts.
-Over the years linguists have classified script systems in a variety of ways,
-with the tripartite classification of logographic, syllabic, and alphabetic
-remaining the most popular, even though there are at least half a dozen
-different types of script systems that can be distinguished
-\citep{Daniels1990,Daniels1996}.
-
-Breaking it down further, a script consists of \textsc{graphemes}, and graphemes
-consist of \textsc{characters}. In the linguistic terminology of writing
-systems, a \textsc{character} is a general term for any self-contained element
-in a writing system.\footnote{There is a second interpretation of the term
-\textsc{character}, i.e.~a conventional term for a unit in the Chinese writing
-system \citep{Daniels1996}. This interpretation will not be further explored in
-this paper.} Although in literate societies most people have a strong intuition
-about what the characters are in their particular orthography or orthographies,
-it turns out that the separation of an orthography into separate characters is
-far from trivial. The widespread intuitive notion of a character is strongly
-biased towards educational traditions, like the alphabet taught at schools, and
-technological possibilities, like the available type pieces in a printer's job
-case, the keys on a typewriter, or the symbols displayed in Microsoft Word's
-symbol browser. In practice, characters often consist of multiple building
-blocks, each of which could be considered a character in its own right. For
-example, although a Chinese character may be considered to be a single basic
-unanalyzable unit, at a more fine-grained level of analysis the internal
-structure of Chinese characters is often comprised of smaller semantic and
-phonetic units that should be considered characters \citep{Sproat2000}. In
-alphabetic scripts, this problem is most forcefully exemplified by diacritics.
-
-A \textsc{diacritic} is a mark, or series of marks, that may be above, below, or
-through other characters \citep{Gaultney2002}. Diacritics are sometimes used to
-distinguish homophonous words, but they are more often used to indicate a
-modified pronunciation \citep[xli]{DanielsBright1996}. The central question is
-whether, for example, <e>, <è>, <a> and <à> should be considered four
-characters, or different combinations of three characters. In general, multiple
-characters together can form another character, and it is not always possible to
-decide on principled grounds what should be the basic building blocks of an
-orthography.
-
-For that reason, it is better to analyze an orthography as a collection of
-graphemes. A \textsc{grapheme} is the basic, minimally distinctive symbol of a
-particular writing system, alike to the phoneme is an abstract representation of
-a distinct sound in a specific language. The term \textsc{grapheme} was modeled
-after the term \textsc{phoneme} and represents a contrastive graphical unit in a
-writing system (see \citealp{Kohrt1986} for a historical overview of the term
-grapheme). Most importantly, a single grapheme regularly consists of multiple
-characters, like <th>, <ou> and <gh> in English (note that each character in
-these graphemes is also a separate grapheme in English). Such complex graphemes
-are often used to represent single phonemes. So, a combination of characters is
-used to represent a single phoneme. Note that the opposite is also found in
-writing systems, in cases in which a single character represents a combination
-of two or more phonemes. For example, <x> in English orthography represents a
-combination of the phonemes /k/ and /s/.
-
-Further, conditioned or free variants of a grapheme are called
-\textsc{allographs}. For example, the distinctive forms of Greek sigma are
-conditioned, with <σ> being used word-internally and <ς> being used at the end
-of a word. In sum, there are many-to-many relationships between phonemes and
-graphemes as they are expressed in the myriad of language- and resource-specific
-orthographies.
-
-\
-
-\noindent This exposition of the linguistic terminology involved in describing writing
-systems has been purposely brief. We have highlighted some of the linguistic
-notions that are pertinent, yet sometimes confused with, the technological
-definitions developed for the computational processing of the world's writing
-systems, which we describe in the next section.
-
-% ==========================
-\section{The Unicode approach}
-\label{the-unicode-approach}
-% ==========================
-
-The conceptualization and terminology of writing systems was rejuvenated by
-the development of the Unicode Standard, with major input from Mark Davis,
-co-founder and long-term president of the Unicode Consortium. For many years,
-``exotic'' writing systems and phonetic transcription systems on personal
-computers were constrained by the American Standard Code for Information
-Interchange (ASCII) character encoding scheme (based on the Latin script), which
-only allowed for a strongly limited number of different symbols to be encoded.
-This implied that users could either use and adopt the (extended) Latin alphabet
-or they could assign new symbols to the small number of code points in the ASCII
-encoding scheme to be rendered by a specifically designed font
-\citep{BirdSimons2003}. In this situation, it was necessary to specify the font
-together with each document to ensure the rightful display of its content. To
-alleviate this problem of assigning different symbols to the same code points,
-in the late 80's and early 90's the Unicode Consortium set itself the ambitious
-goal of developing a single universal character encoding to provide a unique
-number, a code point, for every character in the world's writing systems.
-Nowadays, the Unicode Standard is the default encoding of the technologies that
-support the World Wide Web and for all modern operating systems, software and
-programming languages.
-
-\subsubsection*{The Unicode Standard}
-
-The Unicode Standard represents a massive step forward because it aims to
-eradicate the distinction between \textsc{universal} (ASCII) versus
-\textsc{language/resource-particular} (Font) by adding as much as possible
-language-specific information into the universal standard. However, there are
-still language/resource-specific specifications necessary for the proper usage
-of Unicode, as will be discussed below. Within the Unicode structure many of
-these specifications can be captured by so-called \textsc{Locale Descriptions}
-or simply locales, so we are moving to a new distinction of universal (Unicode
-Standard) versus language-particular (Locale Description). The major gain is a
-much larger compatibility on the universal level (because Unicode standardizes a
-much larger portion of writing system diversity), and much better possibilities
-for automatic processing on the language-particular level (because Locale
-Descriptions are computer readable specifications).
-
-Each version of the Unicode Standard (\citealp{Unicode2014}, as of writing at
-version 7) consists of a set of specifications and guidelines that include (i) a
-core specification, (ii) code charts, (iii) standard annexes and (iv) a
-character database.\footnote{All documents of the Unicode Standard are available
-at \url{http://www.unicode.org/versions/latest/}. For a quick survey of the use
-of terminology inside the Unicode Standard, their glossary is particularly
-useful, available at \url{http://www.unicode.org/glossary/}. For a general
-introduction to the principles of Unicode, Chapter 2 of the core specification,
-called \textsc{general structure}, is particularly insightful. Different from
-many other documents of the Unicode Standard, this general introduction is
-relatively easy to read and illustrated with many interesting examples from
-various orthography traditions all over the world.} The \textsc{core
-specification} is a book directed toward human readers that describes the formal
-standard for encoding multilingual text. The \textsc{code charts} provide a
-humanly readable online reference to the character contents of the Unicode
-Standard in the form of PDF files. The \textsc{Unicode Standard Annexes (UAX)}
-are a set of technical standards that describe the implementation of the Unicode
-Standard for software development, Web standards, and programming languages.
-The \textsc{Unicode Character Database (UCD)} is a set of computer-readable text
-files that describe the character properties, including a set of rich character
-and writing system semantics, for each character in the Unicode Standard. In
-this section, we introduce the basic Unicode concepts, but we will leave out
-many details. Please consult the above mentioned full documentation for a more
-detailed discussion. Further note that the Unicode Standard is exactly that,
-namely a standard. It normatively describes notions and rules to be followed. In
-the actual practice of applying this standard in a computational setting, a
-specific implementation is necessary. The most widely used implementation of the
-Unicode Standard is the \textsc{International Components for Unicode (ICU)},
-which offers C/C++ and Java libraries trying to implement the Unicode
-Standard.\footnote{\url{http://icu-project.org}}
-
-\subsubsection*{Character encoding system}
-
-The Unicode Standard is a \textsc{character encoding system} which
-goal it is to support the interchange and processing of written characters and
-text in a computational setting. Underlyingly, the character encoding is
-represented by a range of numerical values called a \textsc{code space}, which
-is used to encode a set of characters. A \textsc{code point} is a unique
-non-negative integer within a code space (i.e.~within a certain numerical
-range). In the Unicode Standard character encoding system, an \textsc{abstract
-character}, for example the \textsc{latin small letter p}, is mapped to a
-particular code point, in this case the decimal value 112, normally represented in
-hexadecimal, which then looks in Unicode parlance as
-\uni{0070}.\footnote{Hexadecimal (base-16) 0070 is equivalent to decimal
-(base-10) 112, which can be calculated by considering that $(0\cdot16^3) +
-(0\cdot16^2) + (7\cdot16^1) + (0\cdot16^0) = 7\cdot16 = 112$. Underlyingly,
-computers will of course treat this code point binary (base-2) as 11100000, as
-can be seen by calculating that $(1\cdot2^7) + (1\cdot2^6) + (1\cdot2^5) +
-(0\cdot2^4) + (0\cdot2^3) + (0\cdot2^2) + (0\cdot2^1) + (0\cdot2^0) = 64 + 32 +
-16 = 112$.} That encoded abstract character is rendered on a computer screen (or
-printed page) as a \textsc{glyph}, e.g. <p>, depending on the \textsc{font} and
-the context in which that character appears.
-
-In Unicode Standard terminology, an (abstract) \textsc{character} is the basic
-encoding unit. The term \textsc{character} can be quite confusing due to its
-alternative definitions across different scientific disciplines and because in
-general the word \textsc{character} means many different things to different
-people. It is therefore often preferable to refer to Unicode characters simply
-as \textsc{code points}, because there is a one-to-one mapping between Unicode
-characters and their numeric representation. In the Unicode approach, a
-character refers to the abstract meaning and/or general shape, rather than a
-specific shape, though in code tables some form of visual representation is
-essential for the reader's understanding. Unicode defines characters as
-abstractions of orthographic symbols, and it does not define visualizations for
-these characters (although it does presents examples). In contrast, a
-\textsc{glyph} is a concrete graphical representation of a character as it
-appears when rendered (or rasterized) and displayed on an electronic device or
-on printed paper. For example, <g {\large \textit{g}} \textbf{g}
-{\fontspec{ArialMT} {\small g} \textit{g} \textbf{g}}> are different glyphs of the
-same character, i.e.~they may be rendered differently depending on the
-typography being used, but they all share the same code point. From the
-perspective of Unicode they are \textit{the same thing}. In this approach, a
-\textsc{font} is then simply a collection of glyphs linked to code points.
-Allography is not specified in Unicode (expect for a few exceptional cases, due
-to legacy encoding issues), but can be specified in a font as a
-\textsc{contextual variant} (a.k.a.~presentation form).
-
-Each code point in the Unicode Standard is associated with a set of
-\textsc{character properties} as defined by the Unicode character property
-model.\footnote{The character property model is described in
-\url{http://www.unicode.org/reports/tr23/}, but the actual properties are
-described in \url{http://www.unicode.org/reports/tr44/}. A simplified overview
-of the properties is available at
-\url{http://userguide.icu-project.org/strings/properties}. The actual code
-tables listing all properties for all Unicode code points are available at
-\url{http://www.unicode.org/Public/UCD/latest/ucd/}.} Basically, those
-properties are just a long list of values for each character. For example, code
-point \uni{0047} has the following properties (among many others): 
-\begin{itemize}
-	\item Name: LATIN CAPITAL LETTER G 
-	\item Alphabetic: YES 
-	\item Uppercase: YES 
-	\item Script: LATIN 
-	\item Extender: NO 
-	\item Simple\_Lowercase\_Mapping: 0067 
-\end{itemize}
-
-These properties contain the basic information of the Unicode Standard and they
-are necessary to define the correct behavior and conformance required for
-interoperability in and across different software implementations (as defined in
-the Unicode Standard Annexes). The character properties assigned to each code
-point is based on each character's behavior in the real-world writing
-traditions. For example, the corresponding lowercase character to \uni{0047} is
-\uni{0067} (though note that the relation between uppercase and lowercase is in
-many situations much more complex than this, and Unicode has further
-specifications for those cases). Another use of properties is to define the
-script of a character.\footnote{The Unicode Glossary defines the term \textsc{script} as
-a ``collection of letters and other written signs used to represent textual
-information in one or more writing systems. For example, Russian is written with
-a subset of the Cyrillic script; Ukrainian is written with a different subset.
-The Japanese writing system uses several scripts.''} In practice, script is
-simply defined for each character as the explicit \textsc{script} property in
-the Unicode Character Database.
-
-One frequently references property is the \textsc{block} property, which is
-often used in software applications to impose some structure to the large number
-of Unicode characters. Each character in Unicode belongs to a specific block.
-These blocks are basically an organizational structure to alleviate the
-administrative burden of keeping Unicode up-to-date. Blocks consist of
-characters that in some way belong together, so that characters are easier to
-find. Some blocks are connected with a specific script, like the Hebrew block or
-the Gujarati block. However, blocks are predefined ranges of code points, and
-often there will come a point after which the range is completely filled. Any
-extra characters will have to be assigned somewhere else. There is, for example,
-a block \textsc{Arabic}, which contains most Arabic symbols. However, there is
-also a block \textsc{Arabic Supplement}, \textsc{Arabic Presentation Forms-A}
-and \textsc{Arabic Presentation Form-B}. The situation with Latin symbols is
-even more extra. In general, the names for block should be taken as a
-definitional statement. For example, many IPA symbols are not located in the
-aptly-names block \textsc{IPA extensions}, but in other blocks (see Section~\ref{ipa-meets-unicode}).
-
-\subsubsection*{Grapheme clusters}
-
-There are many cases in which a sequence of characters (i.e.~a sequence of more
-than one code point) represents what a user perceives as an individual unit in a
-particular orthographic writing system. For this reason the Unicode Standard
-differentiates between \textsc{abstract character} and \textsc{user-perceived
-character}. Sequences of multiple code points that correspond to a single
-user-perceived characters are called grapheme clusters in Unicode parlance.
-Grapheme clusters come in two flavors: (default) grapheme clusters and tailored
-grapheme clusters.
-
-The (default) \textsc{grapheme clusters} are locale-independent graphemes,
-i.e.~they always apply when a particular combination of characters occurs
-independent of the writing system in which they are used. These character
-combinations are defined in the Unicode Standard as functioning as one
-\textsc{text element}.\footnote{The Unicode Glossary defines text element as:
-``A minimum unit of text in relation to a particular text process, in the
-context of a given writing system. In general, the mapping between text elements
-and code points is many-to-many.''} The simplest example of a grapheme cluster
-is a base character followed by a letter modifier character. For example, the
-sequence <n> + < ̃> (i.e.~\textsc{latin small letter n} at \uni{006E}, followed
-by \textsc{combining tilde} at \uni{0303}) combines visually into <ñ>, a
-user-perceived character in writing systems like that of Spanish. So, what the
-user perceives as a single character actually involves a multi-code-point
-sequence. Note that this specific sequence can also be represented with a single
-so-called \textsc{precomposed} code point, \textsc{latin small letter n with
-tilde} at \uni{00F1}, but this is not the case for all multi-code point
-character sequences. The problem that there multiple encodings possible for the
-same text element has been acknowledged early on in the Unicode Standard
-(e.g.~for <ñ>, the sequence \uni{006E} \uni{0303} should in all situations be
-treated identically to the precomposed \uni{00F1}), and a system of
-\textsc{canonical equivalence} is available for such situations. Basically, the
-Unicode Standard offers different kind of normalizations to either decompose all
-precomposed characters (called \textsc{NFD}, \textsc{Normalization Form
-canonical Decomposition}), or compose as much as possible combinations (called
-\textsc{NFC}, \textsc{Normalization Form canonical Composition}). In current
-practice of software development, NFC seems to be preferred in most situations
-and is widely proposed as the preferred canonical form.
-
-More difficult for text processing, because less standardized, is what the
-Unicode Standard terms \textsc{tailored grapheme clusters}. Tailored grapheme
-clusters are locale-dependent graphemes, i.e.~such combination of characters do
-not function as text elements in all situations. For example, the sequence <c>~+~<h> for the Slovak digraph <ch> or the sequence <ky> in the Sisaala practical
-orthography (pronounced as IPA /tʃ/,~\cite{Moran2006}). These grapheme clusters
-are \textsc{tailored} in the sense that they must be specified on a
-language-by-language or a writing system-by-writing system basis. The Unicode
-Standard provides technological specifications for creating locale specific data
-in so-called \textsc{Unicode Locale Descriptions}, i.e.~a set of specification
-that defines a set of language-specific elements (e.g.~tailored grapheme
-clusters, collation order, capitalization-equivalence), as well as other special
-information, like how to format numbers, dates, or currencies. Locale
-descriptions are saved in the \textsc{Common Locale Data Repository
-(CLDR)},\footnote{\url{http://cldr.unicode.org/}} a repository of
-language-specific definitions of writing system properties, each of which
-describes specific usages of characters. Each locale can be encoded in a
-document using the \textsc{Locale Data Markup Language (LDML)}. LDML is an XML
-format and vocabulary for the exchange of structured locale data. Unicode Locale
-Descriptions allow users to define language- or even resource-specific writing
-systems or orthographies.\footnote{The Unicode Glossary defines \textsc{writing
-system} only very loosely, as it is not a central concept in the Unicode
-Standard. A writing system is, ``A set of rules for using one or more scripts to
-write a particular language. Examples include the American English writing
-system, the British English writing system, the French writing system, and the
-Japanese writing system.''} However, there are various drawbacks of locale
-descriptions for the daily practice of linguistic work in a multilingual setting
-(see Section~\ref{use-cases}).

### 2015-09-29 book/chapters/_remove/originaltext.tex
moves book contents into book directory

--- 2015-05-29 chapters/originaltext.tex
@@
 For character descriptions, we follow the notational conventions of the Unicode Standard (Unicode Consortium 2014).
-Character names are represented in small capital letters (e.g.~LATIN SMALL LETTER SCHWA) and code points are expressed as U\emph{+n} where \emph{n} is a four to six digit hexadecimal number, e.g.~U+0256, which can be rendered as the glyph <ə>.
+Character names are represented in small capital letters (e.g.~LATIN SMALL LETTER SCHWA) and code points are expressed as U\emph{+n} where \emph{n} is a four to six digit hexadecimal number, e.g.~\uni{0256}, which can be rendered as the glyph <ə>.
 
@@
 ) Basically, those properties are just a long list of values for each character.
-For example, code point U+0047 has the following properties (among many others):
+For example, code point \uni{0047} has the following properties (among many others):
 \begin{itemize}
@@
 The character properties assigned to each code point is based on each character's behavior in the real-world writing traditions.
-For example, the corresponding lowercase character to U+0047 is U+0067 (though note that the relation between uppercase and lowercase is in many situations much more complex than this, and Unicode has further specifications for those cases).
+For example, the corresponding lowercase character to \uni{0047} is \uni{0067} (though note that the relation between uppercase and lowercase is in many situations much more complex than this, and Unicode has further specifications for those cases).
 Another use of properties is to define the script of a character.
@@
 In general, the mapping between text elements and code points is many-to-many.'') The simplest example of a grapheme cluster is a base character followed by a letter modifier character.
-For example, the sequence + <̃\textgreater{} (LATIN SMALL LETTER N at U+006E followed by COMBINING TILDE at U+0303) combines visually into , a user-percieved character in writing systems like that of Spanish.
+For example, the sequence + <̃\textgreater{} (LATIN SMALL LETTER N at \uni{006E} followed by COMBINING TILDE at \uni{0303}) combines visually into , a user-percieved character in writing systems like that of Spanish.
 So, what the user perceives as a single character actually involves a multi-code-point sequence.
-Note that this specific sequence can also be represented with a single \textsc{precomposed} code point, LATIN SMALL LETTER N WITH TILDE at U+00F1, but this is not the case for all multi-code point character sequences.
-The problem that there multiple encodings possible for the same text element has been acknowledged early on in the Unicode Standard (e.g.~for , the sequence U+006E U+0303 should in all situations be treated identically to the precomposed U+00F1), and a system of \textsc{canonical equivalence} is available for such situations.
+Note that this specific sequence can also be represented with a single \textsc{precomposed} code point, LATIN SMALL LETTER N WITH TILDE at \uni{00F1}, but this is not the case for all multi-code point character sequences.
+The problem that there multiple encodings possible for the same text element has been acknowledged early on in the Unicode Standard (e.g.~for , the sequence \uni{006E} \uni{0303} should in all situations be treated identically to the precomposed \uni{00F1}), and a system of \textsc{canonical equivalence} is available for such situations.
 Basically, the Unicode Standard offers different kind of normalizations to either decompose all precomposed characters (called \textsc{NFD}, \textsc{Normalization Form Canonical Decomposition}), or compose as much as possible combinations (called \textsc{NFC}, \textsc{Normalization Form Canonical Composition}).
@@
 The Unicode code space is subdivided into blocks of contiguous code points.
-For example, the block called ``Cyrillic'' runs from U+0400 till U+04FF.
+For example, the block called ``Cyrillic'' runs from \uni{0400} till \uni{04FF}.
 These blocks arose as an attempt at ordering the enormous amount of characters in Unicode, but the ideas of blocks very quickly ran into problems.
@@
 However, when a user searches for an IPA character in the block ``IPA Extensions'' then many IPA characters will not be found there.
-For example, the velar nasal <ŋ> is not part of the block ``IPA Extensions'' because it was already included as LATIN SMALL LETTER ENG at U+014B in the block ``Latin Extensions-A''.
+For example, the velar nasal <ŋ> is not part of the block ``IPA Extensions'' because it was already included as LATIN SMALL LETTER ENG at \uni{014B} in the block ``Latin Extensions-A''.
 
@@
 Taking an example already discussed in the previous section, the following symbols <reinsert Gs here> are different glyphs of the same character, i.e.~they may be rendered differently depending on the typography being used, but they all share the same code point (viz.
-LATIN SMALL LETTER G at U+0067).
+LATIN SMALL LETTER G at \uni{0067}).
 In contrast, the symbols <A А Α Ꭺ ᗅ ᴀ ꓮＡ𐊠 𐌀 𝖠 𝙰> are all different code points, although they look highly similar -- in some cases even sharing exactly the same glyph in some fonts.
@@
 \begin{itemize}
-	\item <A\textgreater{} LATIN CAPITAL LETTER A, at U+0041 
-	\item <А> CYRILLIC CAPITAL LETTER A, at U+0410 
-	\item <Α> GREEK CAPITAL LETTER ALPHA, at U+0391 
-	\item <Ꭺ> CHEROKEE LETTER GO, at U+13AA 
-	\item <ᗅ> CANADIAN SYLLABICS CARRIER GHO, at U+15C5 
-	\item <ᴀ> LATIN LETTER SMALL CAPITAL A, at U+1D00 
-	\item <ꓮ> LISU LETTER A, at U+A4EE 
-	\item <Ａ> FULLWIDTH LATIN CAPITAL LETTER A, at U+FF21 
-	\item <𐊠> CARIAN LETTER A, at U+102A0 
-	\item <𐌀> OLD ITALIC LETTER A, at U+10300 
-	\item <𝖠> MATHEMATICAL SANS-SERIF CAPITAL A, U+1D5A0 
-	\item <𝙰> MATHEMATICAL MONOSPACE CAPITAL A, at U+1D670 
+	\item <A\textgreater{} LATIN CAPITAL LETTER A, at \uni{0041} 
+	\item <А> CYRILLIC CAPITAL LETTER A, at \uni{0410} 
+	\item <Α> GREEK CAPITAL LETTER ALPHA, at \uni{0391} 
+	\item <Ꭺ> CHEROKEE LETTER GO, at \uni{13AA} 
+	\item <ᗅ> CANADIAN SYLLABICS CARRIER GHO, at \uni{15C5} 
+	\item <ᴀ> LATIN LETTER SMALL CAPITAL A, at \uni{1D00} 
+	\item <ꓮ> LISU LETTER A, at \uni{A4EE} 
+	\item <Ａ> FULLWIDTH LATIN CAPITAL LETTER A, at \uni{FF21} 
+	\item <𐊠> CARIAN LETTER A, at \uni{102A}0 
+	\item <𐌀> OLD ITALIC LETTER A, at \uni{1030}0 
+	\item <𝖠> MATHEMATICAL SANS-SERIF CAPITAL A, \uni{1D5A}0 
+	\item <𝙰> MATHEMATICAL MONOSPACE CAPITAL A, at \uni{1D67}0 
 \end{itemize}
@@
 Such homoglyphs are a widespread problem for consistent encoding.
-Although for most users it looks like the words and are identical, in actual fact they do not even share a single code point.\footnote{The first words consists completely of Latin characters, namely U+0076, U+006F, U+0063, U+0065, U+0073, while the second is a mix of Cyrillic and Greek characters, namely U+03BD, U+03BF, U+0041, U+0435, U+0455.) For computers these two words are completely different entities.
+Although for most users it looks like the words and are identical, in actual fact they do not even share a single code point.\footnote{The first words consists completely of Latin characters, namely \uni{0076}, \uni{006F}, \uni{0063}, \uni{0065}, \uni{0073}, while the second is a mix of Cyrillic and Greek characters, namely \uni{03BD}, \uni{03BF}, \uni{0041}, \uni{0435}, \uni{0455}.) For computers these two words are completely different entities.
 Commonly, when users with Cyrillic or Greek keyboards have to type some Latin-based orthography, they mix similar looking Cyrillic or Greek characters into their text, because those characters are so much easier to type.
@@
 	\def\labelenumi{\arabic{enumi}.} 
-	\item LATIN CAPITAL LETTER A WITH RING ABOVE (U+00C5) 
-	\item ANGSTROM SIGN (U+212B) 
-	\item LATIN CAPITAL LETTER A (U+0041) + COMBINING RING ABOVE (U+030A) 
+	\item LATIN CAPITAL LETTER A WITH RING ABOVE (\uni{00C5}) 
+	\item ANGSTROM SIGN (\uni{212B}) 
+	\item LATIN CAPITAL LETTER A (\uni{0041}) + COMBINING RING ABOVE (\uni{030A}) 
 \end{enumerate}
@@
 Nevertheless, they are encoded differently.
-For example, if one were to search an electronic text (with software that does not apply Unicode Standard normalization) for ANGSTROM SIGN (U+212B), instances of LATIN CAPITAL LETTER A WITH RING ABOVE (U+00C5) would not be found.
+For example, if one were to search an electronic text (with software that does not apply Unicode Standard normalization) for ANGSTROM SIGN (\uni{212B}), instances of LATIN CAPITAL LETTER A WITH RING ABOVE (\uni{00C5}) would not be found.
 
@@
 Various players, like the Unicode Consortium, the W3C, or the TEI recommend NFC in most user-directed situations, and some software applications that we tested indeed seem to automatically convert strings into NFC.\footnote{See the summary of various recommendation here: http://www.win.tue.nl/\textasciitilde{}aeb/linux/uc/nfc\_vs\_nfd.html.
-) This means in practice that if a user, for example, enters <a> and <`>, i.e.~LATIN SMALL LETTER A at U+0061 and COMBINING GRAVE ACCENT at U+0300, this might be converted into <à>, i.e.~LATIN SMALL LETTER A WITH GRAVE at U+00E0.\footnote{The behaviour of software applications can be quite erratic in this respect.
+) This means in practice that if a user, for example, enters <a> and <`>, i.e.~LATIN SMALL LETTER A at \uni{0061} and COMBINING GRAVE ACCENT at \uni{0300}, this might be converted into <à>, i.e.~LATIN SMALL LETTER A WITH GRAVE at \uni{00E0}.\footnote{The behaviour of software applications can be quite erratic in this respect.
 For example, Apple's TextEdit does not do any conversion on text entry.
@@
 Another problem is the large number of homoglyphs, i.e.~different characters that have highly similar glyphs (or even completely identical, depending on the font rendering).
-For example, a speaker of Russian should ideally not use the CYRILLIC SMALL LETTER A at code point U+0430 for IPA transcriptions, but instead the LATIN SMALL LETTER A at code point U+0061, although visually they are mostly indistinguishable, and the Cyrillic character is more easily typed on a Cyrillic keyboard.
-Another example we commonly encounter is the use of LATIN SMALL LETTER G at U+0067, instead of the sanctioned Unicode Standard IPA character for the voiced velar stop LATIN SMALL LETTER SCRIPT G at U+0261.
-One begins to question whether this issue is at all apparent to the working linguist, or if they simply use the U+0067 because it is easily keyboarded and thus saves time, whereas the latter must be cumbersomely inserted as a special symbol in most software.\footnote{This issue was recently addressed by the International Phonetic Association, which has taken the stance that both the keyboard LATIN SMALL LETTER G and the LATIN SMALL LETTER SCRIPT G are valid input characters for the voiced velar plosive.
+For example, a speaker of Russian should ideally not use the CYRILLIC SMALL LETTER A at code point \uni{0430} for IPA transcriptions, but instead the LATIN SMALL LETTER A at code point \uni{0061}, although visually they are mostly indistinguishable, and the Cyrillic character is more easily typed on a Cyrillic keyboard.
+Another example we commonly encounter is the use of LATIN SMALL LETTER G at \uni{0067}, instead of the sanctioned Unicode Standard IPA character for the voiced velar stop LATIN SMALL LETTER SCRIPT G at \uni{0261}.
+One begins to question whether this issue is at all apparent to the working linguist, or if they simply use the \uni{0067} because it is easily keyboarded and thus saves time, whereas the latter must be cumbersomely inserted as a special symbol in most software.\footnote{This issue was recently addressed by the International Phonetic Association, which has taken the stance that both the keyboard LATIN SMALL LETTER G and the LATIN SMALL LETTER SCRIPT G are valid input characters for the voiced velar plosive.
 Unfortunately, this decision further introduces ambiguity for linguists trying to adhere to a strict Unicode Standard IPA encoding.)
 
-Furthermore, on the one hand even linguists are unlikely to distinguish between the LATIN SMALL LETTER SCHWA at code point U+0259 and LATIN SMALL LETTER TURNED E at U+01DD.
-On the other hand, non-linguists are unlikely to distinguish any semantic difference between an open back unrounded vowel , the LATIN SMALL LETTER ALPHA at U+0251, and the open front unrounded vowel , LATIN SMALL LETTER A at U+0061.
+Furthermore, on the one hand even linguists are unlikely to distinguish between the LATIN SMALL LETTER SCHWA at code point \uni{0259} and LATIN SMALL LETTER TURNED E at \uni{01DD}.
+On the other hand, non-linguists are unlikely to distinguish any semantic difference between an open back unrounded vowel , the LATIN SMALL LETTER ALPHA at \uni{0251}, and the open front unrounded vowel , LATIN SMALL LETTER A at \uni{0061}.
 But even among linguists this distinction leads to problems.
@@
 	\item The uses of the apostrophe has led to long discussions on the Unicode Standard email list.
-An English keyboard inputs \textless{}`\textgreater{} APOSTROPHE at U+0027, although the `preferred' Unicode apostrophe is the \textless{}'\textgreater{} RIGHT SINGLE QUOTATION MARK at U+2019.
-Yet the glottal stop/glottalization/ejective marker is another completely different character, the MODIFIER LETTER APOSTROPHE at U+02BC, but unfortunately looks mostly highly similar to U+2019.
+An English keyboard inputs \textless{}`\textgreater{} APOSTROPHE at \uni{0027}, although the `preferred' Unicode apostrophe is the \textless{}'\textgreater{} RIGHT SINGLE QUOTATION MARK at \uni{2019}.
+Yet the glottal stop/glottalization/ejective marker is another completely different character, the MODIFIER LETTER APOSTROPHE at \uni{02BC}, but unfortunately looks mostly highly similar to \uni{2019}.
 	\item There is ambiguous encoding of IPA segments within the Unicode Standard.
-An example is the U+02C1 MODIFIER LETTER REVERSED GLOTTAL STOP vs the U+02E4 MODIFIER LETTER SMALL REVERSED GLOTTAL STOP .
+An example is the \uni{02C1} MODIFIER LETTER REVERSED GLOTTAL STOP vs the \uni{02E4} MODIFIER LETTER SMALL REVERSED GLOTTAL STOP .
 Both are denoted in the Unicode Standard as the `pharyngealized diacritic' and both appear in various resources representing phonetic data online.
 This is thus an example for which the Unicode Standard does not solve the linguistic standardization problem.
-	\item There is at least one case in which the character name assigned by the Unicode Consortium does not match the IPA's description: in the Unicode Standard at U+01C3 is labeled LATIN LETTER RETROFLEX CLICK, but in IPA is an alveolar or postalveolar click (not retroflex).
+	\item There is at least one case in which the character name assigned by the Unicode Consortium does not match the IPA's description: in the Unicode Standard at \uni{01C3} is labeled LATIN LETTER RETROFLEX CLICK, but in IPA is an alveolar or postalveolar click (not retroflex).
 This naming is probably best seen as a simple error in the Unicode Standard.
 Note that most linguists simply seem to use or <ou>) while ligatures are single characters (e.g.
-<ʧ> LATIN SMALL LETTER TESH DIGRAPH at U+02A7).
+<ʧ> LATIN SMALL LETTER TESH DIGRAPH at \uni{02A7}).
 Ligatures arose in the context of printing easier-to-read texts, and are included in the Unicode Standard for reasons of legacy encoding.
@@
 Instead a combination of two characters is preferred:
-	\item <ʣ> LATIN SMALL LETTER DZ DIGRAPH at U+02A3 (use <d> + <z> instead) 
-	\item <ʧ> LATIN SMALL LETTER TESH DIGRAPH at U+02A7 (use <t> + <ʃ> instead) 
-	\item <ʩ> LATIN SMALL LETTER FENG DIGRAPH at U+02A9 (use <f> + <ŋ> instead) 
+	\item <ʣ> LATIN SMALL LETTER DZ DIGRAPH at \uni{02A3} (use <d> + <z> instead) 
+	\item <ʧ> LATIN SMALL LETTER TESH DIGRAPH at \uni{02A7} (use <t> + <ʃ> instead) 
+	\item <ʩ> LATIN SMALL LETTER FENG DIGRAPH at \uni{02A9} (use <f> + <ŋ> instead) 
 \end{itemize}
@@
 \begin{itemize}
-	\item <ɮ> LATIN SMALL LETTER LEZH at U+026E 
-	\item <œ> LATIN SMALL LIGATURE OE at U+0153 
-	\item <ɶ> LATIN LETTER SMALL CAPITAL OE at U+0276 
-	\item <æ> LATIN SMALL LETTER AE at U+00E6 
+	\item <ɮ> LATIN SMALL LETTER LEZH at \uni{026E} 
+	\item <œ> LATIN SMALL LIGATURE OE at \uni{0153} 
+	\item <ɶ> LATIN LETTER SMALL CAPITAL OE at \uni{0276} 
+	\item <æ> LATIN SMALL LETTER AE at \uni{00E6} 
 \end{itemize}
@@
 \begin{itemize}
-	\item Length marks, namely <ː> MODIFIER LETTER TRIANGULAR COLON at U+02D0 and <ˑ> MODIFIER LETTER HALF TRIANGULAR COLON at U+02D1. 
-	\item Tone letters, like <˥> MODIFIER LETTER EXTRA-HIGH TONE BAR at U+02E5, and others like this. 
-	\item Superscript letters, like <ʰ> MODIFIER LETTER SMALL H at U+02B0 or <ˤ> MODIFIER LETTER SMALL REVERSED GLOTTAL STOP at U+02E4, and others like this. 
-	\item <˞> MODIFIER LETTER RHOTIC HOOK at U+02DE. 
+	\item Length marks, namely <ː> MODIFIER LETTER TRIANGULAR COLON at \uni{02D0} and <ˑ> MODIFIER LETTER HALF TRIANGULAR COLON at \uni{02D1}. 
+	\item Tone letters, like <˥> MODIFIER LETTER EXTRA-HIGH TONE BAR at \uni{02E5}, and others like this. 
+	\item Superscript letters, like <ʰ> MODIFIER LETTER SMALL H at \uni{02B0} or <ˤ> MODIFIER LETTER SMALL REVERSED GLOTTAL STOP at \uni{02E4}, and others like this. 
+	\item <˞> MODIFIER LETTER RHOTIC HOOK at \uni{02DE}. 
 \end{itemize}
 
-Although linguists might expect these characters to belong together with the character in front of them, at least for <ʰ> MODIFIER LETTER SMALL H at U+02B0 the Unicode Consortium's decision to treat it as a separate character is also linguistically correct, because according to the IPA it can be used both for aspiration (more precisely postaspiration following the base character) and preaspiration (preceding the base character).
-Note that there is a mechanism in Unicode to force separate characters to be combined (namely by using the ZERO WIDTH JOINER at U+200D), but this seems to be a rather impractical, and probably not enforceable solution to us.
+Although linguists might expect these characters to belong together with the character in front of them, at least for <ʰ> MODIFIER LETTER SMALL H at \uni{02B0} the Unicode Consortium's decision to treat it as a separate character is also linguistically correct, because according to the IPA it can be used both for aspiration (more precisely postaspiration following the base character) and preaspiration (preceding the base character).
+Note that there is a mechanism in Unicode to force separate characters to be combined (namely by using the ZERO WIDTH JOINER at \uni{200D}), but this seems to be a rather impractical, and probably not enforceable solution to us.
 
@@
 When encoding a string according to the Unicode Standard, it is possible to do this either using the NFC (``composition'') or NFD (``decomposition'') normalization.
-Decomposition implies that precomposed characters (like <á> LATIN SMALL LETTER A WITH ACUTE at U+00E1) will be split into its parts.
+Decomposition implies that precomposed characters (like <á> LATIN SMALL LETTER A WITH ACUTE at \uni{00E1}) will be split into its parts.
 This might sound preferable for a linguistic analysis, as the different diacritics are separated from the base characters.
@@
 The combining character sequence <õ̰> is normalized and visually grouped together.
-Note that , the MODIFIER LETTER SMALL H at U+02B0, is not grouped with .
+Note that , the MODIFIER LETTER SMALL H at \uni{02B0}, is not grouped with .
 This is because it belongs to Spacing Modifier Letters category in the Unicode Standard.
@@
 The actual ordering of the columns is unimportant.
-	\item \textsc{Lines starting with a hash \textless{}\#\textgreater{} are ignored.} Comments and metadata can be included inside the file, but only as complete lines in the profile, to be marked by lines starting with hash ``\#'' (NUMBER SIGN at U+0023).
+	\item \textsc{Lines starting with a hash \textless{}\#\textgreater{} are ignored.} Comments and metadata can be included inside the file, but only as complete lines in the profile, to be marked by lines starting with hash ``\#'' (NUMBER SIGN at \uni{0023}).
 Hashes somewhere else in the file are to be treated literally, i.e.~hashes are only to be ignored when they occur at the start of a line.\footnote{Comments that belong to specific lines will have to be put in a separate column of the CSV file, e.g.~add a column called `comments'.
@@
 Character-to-glyph mappings are particularly not clear-cut for linguists using the IPA.\footnote{We give more linguistic examples in Section 5.) Just look in the IPA Extensions block.\footnote{Character ``blocks'' are described in Section 3 (The Unicode approach).) There are several characters that are glyphs (variants) of the same grapheme in the Latin block, e.g.
-and \& and .\footnote{LATIN SMALL LETTER ALPHA U+0251 vs LATIN SMALL LETTER A U+0061 , and LATIN SMALL LETTER SCRIPT G U+0261 vs LATIN SMALL LETTER G U+0067 .) Other characters like
+and \& and .\footnote{LATIN SMALL LETTER ALPHA \uni{0251} vs LATIN SMALL LETTER A \uni{0061} , and LATIN SMALL LETTER SCRIPT G \uni{0261} vs LATIN SMALL LETTER G \uni{0067} .) Other characters like
 
@@
 The simplest example of a grapheme cluster is essentially a base character followed by a Letter Modifier character.
-For example, the sequence + \textless{}\textasciitilde{}\textgreater{} (COMBINING TILDE at U+0303) combines visually into , a user-percieved character in writing systems like that of Spanish.\footnote{Although the code point sequence can also be represented with a single code point, LATIN SMALL LETTER N WITH TILDE at U+00F1, it is not the case that all multi-code point character sequences have a single code point equivalent, which can be deduced from normalization.
+For example, the sequence + \textless{}\textasciitilde{}\textgreater{} (COMBINING TILDE at \uni{0303}) combines visually into , a user-percieved character in writing systems like that of Spanish.\footnote{Although the code point sequence can also be represented with a single code point, LATIN SMALL LETTER N WITH TILDE at \uni{00F1}, it is not the case that all multi-code point character sequences have a single code point equivalent, which can be deduced from normalization.
 Often for commonly-used letters in European languages, a single code point form exists.)
@@
 \begin{itemize}
-	\item LATIN SMALL LETTER U WITH ACUTE (U+00FA) 
-	\item LATIN SMALL LETTER U (U+0075) + COMBINING ACUTE ACCENT (U+0301) 
+	\item LATIN SMALL LETTER U WITH ACUTE (\uni{00FA}) 
+	\item LATIN SMALL LETTER U (\uni{0075}) + COMBINING ACUTE ACCENT (\uni{0301}) 
 \end{itemize}
@@
 
-However, the standard does not specify many (or most) language-specific graphemes, or in Unicode parlance, \textsc{tailored grapheme clusters}, such as English .\footnote{Some (unfortunate) exceptions included for legacy purposes, and used incorrectly by linguists, include digraphs like LATIN SMALL LETTER TS DIGRAPH at U+02A6 or LATIN SMALL LETTER DEZH DIGRAPH at U+02A4.
+However, the standard does not specify many (or most) language-specific graphemes, or in Unicode parlance, \textsc{tailored grapheme clusters}, such as English .\footnote{Some (unfortunate) exceptions included for legacy purposes, and used incorrectly by linguists, include digraphs like LATIN SMALL LETTER TS DIGRAPH at \uni{02A6} or LATIN SMALL LETTER DEZH DIGRAPH at \uni{02A4}.
 These and other digraphs are encoded in the Unicode Standard's IPA Extensions block.) For technological reasons, these language-specific tailored grapheme clusters must be defined outside of the Unicode Standard, in what are called (language-specific) Unicode Locale Descriptions.\footnote{http://www.unicode.org/reports/tr35/) The encoding of these multiple character sequences in Locale Descriptions is beyond the need (or technical ability) of many linguists; thus we have developed a simpler approach which we used in our own research and describe in detailed in Section XXX.
@@
 	\def\labelenumi{\arabic{enumi}.} 
-	\item <õ̰> LATIN SMALL LETTER O + COMBINING TILDE + COMBINING TILDE BELOW: U+006F + U+0303 + U+0330 
-	\item <õ̰> LATIN SMALL LETTER O + COMBINING TILDE BELOW + COMBINING TILDE: U+006F + U+0330 + U+0303 
-	\item <õ̰> LATIN SMALL LETTER O WITH TILDE + COMBINING TILDE BELOW: U+00F5 + U+0330 
+	\item <õ̰> LATIN SMALL LETTER O + COMBINING TILDE + COMBINING TILDE BELOW: \uni{006F} + \uni{0303} + \uni{0330} 
+	\item <õ̰> LATIN SMALL LETTER O + COMBINING TILDE BELOW + COMBINING TILDE: \uni{006F} + \uni{0330} + \uni{0303} 
+	\item <õ̰> LATIN SMALL LETTER O WITH TILDE + COMBINING TILDE BELOW: \uni{00F5} + \uni{0330} 
 \end{enumerate}

### 2015-09-29 book/localmetadata.tex
moves book contents into book directory

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%                                              %%%
%%%                 Metadata                     %%%
%%%          fill in as appropriate              %%%
%%%                                              %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{The Unicode Cookbook \linebreak for Linguists}
\subtitle{Managing writing systems using orthography profiles}
\BackTitle{Change backtitle in  localmetadata.tex}
\BackBody{Change backbody in  localmetadata.tex}
\dedication{Change dedication in localmetadata.tex}
\typesetter{Change typesetter in localmetadata.tex}
\proofreader{Change proofreaders in localmetadata.tex}
\author{Steven Moran \& Michael Cysouw}
\renewcommand{\lsISBN}{978--3--000000--00--0}                     
\renewcommand{\lsSeries}{dummyseries} 
% use lowercase acronym, e.g. sidl, eotms, tgdi
\renewcommand{\lsSeriesNumber}{99} 
%will be assigned when the book enters the proofreading stage
\renewcommand{\lsURL}{http://langsci-press.org/catalog/book/0} 
% contact the coordinator for the right number

### 2015-09-29 book/tables/unicode_additions.tex
moves book contents into book directory

--- 2015-06-19 tables/unicode_additions.tex
@@
-\tablecaption{Additions to the IPA with Unicode encodings}\label{tab:ipa_additions}
+\tablecaption{Additions to widened-IPA with Unicode encodings}\label{tab:ipa_additions}
  \tablefirsthead{
@@
      \captionsize\tablename\ \thetable{} 
-     Additions to the IPA with Unicode encodings --- \textit{continued}
+     Additions to widened-IPA with Unicode encodings --- \textit{continued}
      } \\
@@
 \begin{center}
-\begin{xtabular}{ l L{1.2cm} L{6.1cm} L{3.5cm} }
+\begin{xtabular}{ l L{1.1cm} L{6.1cm} L{3.5cm} }
 ‼ & \unif{203C} & \textsc{double exclamation mark} & retroflex click \\

### 2015-09-29 book/tables/unicode_diacritics.tex
moves book contents into book directory

--- 2015-06-22 tables/unicode_diacritics.tex
@@
 \dia{02D0} & \unif{02D0} & \textsc{modifier letter triangular colon} & long \\
-\dia{02D1} & \unif{02D1} & \textsc{modifier letter half triangular colon} & half-long \\{˥} & \unif{02E5} & \textsc{modifier letter extra-high tone bar} & extra high tone \\
-{˦} & \unif{02E6} & \textsc{modifier letter high tone bar} & high tone \\
-{˧} & \unif{02E7} & \textsc{modifier letter mid tone bar} & mid tone \\
-{˨} & \unif{02E8} & \textsc{modifier letter low tone bar} & low tone \\
-{˩} & \unif{02E9} & \textsc{modifier letter extra-low tone bar} & extra low tone \\
+\dia{02D1} & \unif{02D1} & \textsc{modifier letter half triangular colon} & half-long \\
 \dia{0361}{\large\fontspec{CharisSIL}◌} & \unif{0361} & \textsc{combining double inverted breve} & tie bar \\

### 2015-09-29 book/unicode.tex
moves book contents into book directory

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%                                              %%%
%%%     Language Science Press Master File       %%%
%%%         follow the instructions below        %%%
%%%                                              %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Everything following a % is ignored
% Some lines start with %. Remove the % to include them

\documentclass[output=inprep,
%  long|short|inprep
%  ,blackandwhite
%  ,smallfont
%  ,draftmode  
		biblatex
		]{LSP/langsci}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%                                              %%%
%%%          additional packages                 %%%
%%%                                              %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% put all additional commands you need in the 
% following files. If you do not know what this might 
% mean, you can safely ignore this section
\usepackage{verbatim}
\input{localmetadata.tex}
\input{localpackages.tex}
\input{localhyphenation.tex}
\input{localcommands.tex}
\bibliography{localbibliography}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%                                              %%%
%%%             Frontmatter                      %%%
%%%                                              %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}
\maketitle
\frontmatter
% %% uncomment if you have preface and/or acknowledgements
% \include{chapters/preface}
% \include{chapters/acknowledgments}
% \include{chapters/abbreviations}
\tableofcontents
\mainmatter%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%                                              %%%
%%%             Chapters                         %%%
%%%                                              %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\include{chapters/introduction}
\include{chapters/pitfalls}
\include{chapters/ipa_meets_unicode}
\include{chapters/orthography_profiles}
\include{chapters/use_cases}

% TODOS:

% Lots of characters within <> were lost in translation...
% some things to note in the translation to LaTeX:
% tildes in URLs broken -- turned into \textasciitilde{}
% <a> etc. seem to sometimes be lost
% section in some cases should be "chapter"

% other stuff to fix / proof read
% references
% sections, e.g. Section 4, Pitfall 4)
% URLs
% `', ``''
% reinsertion of a lot of the glyphs, etc.
% fix ligatures
% all glyphs need to be checked
% all <>s need to be checked and/or readded

\begin{comment}
===
Double quotation marks are generally used for distancing, in particular in the following situations:

1. when a passage from another work is cited in the text (e.g. According to Takahashi (2009: 33), “quotatives were never used in subordinate clauses in Old Japanese”); but block quotations do not have quotation marks;
2. when a technical term is mentioned that the author does not want to adopt, but wants to mention, e.g. This is sometimes called “pseudo-conservatism”, but I will not use this term here, as it could lead to confusion.

Single quotation marks are used exclusively for linguistic meanings, as in the following: Latin habere ‘have’ is not cognate with Old English hafian ‘have’.

===
so, we should normally use double quotation in most of our cases :-). In general though, I would like us to try and remove double quotations as much as possible. Mostly is thus signals uncertainty on our part :-).

\end{comment}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%                                              %%%
%%%             Backmatter                       %%%
%%%                                              %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% There is normally no need to change the backmatter section
\input{backmatter.tex}
\end{document}

% you can create your book by running
% xelatex lsp-skeleton.tex
%
% you can also try a simple 
% make
% on the commandline

### 2015-09-29 book/autoindex.py

@@ -1,0 +1,29 @@
+#!/usr/bin/python3
+
+import glob
+import re
+
+lgs=open("locallanguages.txt").read().split('\n')
+terms=open("localsubjectterms.txt").read().split('\n')[::-1]#reverse to avoid double indexing
+print("found %i language names for autoindexing" % len(lgs))
+print("found %i subject terms for autoindexing" % len(terms))
+
+files = glob.glob('chapters/*tex')
+for f in files:
+  print("indexing %s" % f)
+  c = open(f).read()  
+  for lg in lgs:
+    c = re.sub('(?<!ili{)%s(?![\w}])'%lg, '\ili{%s}'%lg, c)
+  for term in terms:
+    c = re.sub('(?<!isi{)%s(?![\w}])'%term, '\isi{%s}'%term, c)
+  nlg = len(re.findall('\\ili{',c))
+  nt = len(re.findall('\\isi{',c))
+  outfile = open(f.replace('chapters','indexed'), 'w')
+  outfile.write(c)
+  outfile.close()
+  print(" %s now contains %i indexed languages and %i indexed subject terms"%(f.split('/')[-1],nlg,nt))
+  
+print("indexed files are in the folder 'indexed'")
+  
+  
+  

### 2015-09-29 autoindex.py
rms more

@@ -1,29 +1,0 @@
-#!/usr/bin/python3
-
-import glob
-import re
-
-lgs=open("locallanguages.txt").read().split('\n')
-terms=open("localsubjectterms.txt").read().split('\n')[::-1]#reverse to avoid double indexing
-print("found %i language names for autoindexing" % len(lgs))
-print("found %i subject terms for autoindexing" % len(terms))
-
-files = glob.glob('chapters/*tex')
-for f in files:
-  print("indexing %s" % f)
-  c = open(f).read()  
-  for lg in lgs:
-    c = re.sub('(?<!ili{)%s(?![\w}])'%lg, '\ili{%s}'%lg, c)
-  for term in terms:
-    c = re.sub('(?<!isi{)%s(?![\w}])'%term, '\isi{%s}'%term, c)
-  nlg = len(re.findall('\\ili{',c))
-  nt = len(re.findall('\\isi{',c))
-  outfile = open(f.replace('chapters','indexed'), 'w')
-  outfile.write(c)
-  outfile.close()
-  print(" %s now contains %i indexed languages and %i indexed subject terms"%(f.split('/')[-1],nlg,nt))
-  
-print("indexed files are in the folder 'indexed'")
-  
-  
-  

### 2015-09-30 use_cases/ASJP/manual.R
use case ASJP

@@ -1,0 +1,42 @@
+#' ---
+#' title: "Analysis of ASJP data"
+#' author: "Michael Cysouw"
+#' date: "30 September 2015"
+#' ---
+
+# install library from github
+# only necessary once
+# install.packages("devtools")
+# devtools::install_github("cysouw/qlcTokenize")
+library(qlcTokenize)
+
+# read ASJP data
+# this is pretty large, so it might take a bit
+asjp <- read.csv("data/ASJP.csv")
+
+# first make draft profile using default settings
+write.profile(asjp$Value, file = "sandbox/asjp_draft_profile.csv")
+
+# There is a warning because there are tabs in the input data
+# This is an error in the ASJP data, but we leave it in for now
+# It has to be changed in the orthography profile though!
+# Check the manually changed profile in profiles/asjp_corrected_profile.csv
+
+
+# correct this manually in the file profile and do tokenization
+# the code is not very fast, might take about half a minute for all of ASJP
+system.time(
+	tokens <- tokenize(asjp$Value
+						, profile = "profiles/asjp_corrected_profile.csv"
+						, regex = T
+						, file.out = "sandbox/")
+	)
+
+# check errors
+tokens$errors
+err_lines <- as.numeric(rownames(tokens$errors))
+asjp[err_lines,]
+no_errors <- tokens$strings$tokenized[-err_lines]
+
+# make "empirical" profile of graphemes
+write.profile(no_errors, sep = " ", file = "asjp_emprical_profile.csv")

### 2015-09-30 use_cases/ASJP/manual.R
ASJP use case for R ready

@@ -14,6 +14,7 @@ library(qlcTokenize)
 # this is pretty large, so it might take a bit
 asjp <- read.csv("data/ASJP.csv")
 
+# the actual words are in the fifth column of this csv-file ("Value")
 # first make draft profile using default settings
 write.profile(asjp$Value, file = "sandbox/asjp_draft_profile.csv")
 
@@ -23,20 +24,26 @@ write.profile(asjp$Value, file = "sandbox/asjp_draft_profile.csv")
 # Check the manually changed profile in profiles/asjp_corrected_profile.csv
 
 
-# correct this manually in the file profile and do tokenization
+# after manually correcting the profile we can use it for tokenization
+# there are regular expressions in the corrected profile, so "regex = TRUE"
 # the code is not very fast, might take about half a minute for all of ASJP
-system.time(
-	tokens <- tokenize(asjp$Value
-						, profile = "profiles/asjp_corrected_profile.csv"
-						, regex = T
-						, file.out = "sandbox/")
-	)
-
-# check errors
+tokens <- tokenize(asjp$Value
+					, profile = "profiles/asjp_corrected_profile.csv"
+					, regex = TRUE
+					)
+					
+# there are errors in the data. Let's check the errors
 tokens$errors
-err_lines <- as.numeric(rownames(tokens$errors))
-asjp[err_lines,]
-no_errors <- tokens$strings$tokenized[-err_lines]
 
-# make "empirical" profile of graphemes
-write.profile(no_errors, sep = " ", file = "asjp_emprical_profile.csv")
+# and look at the characters that are not in the profile
+tokens$missing
+
+# look at the errors in this original context
+error_lines <- as.numeric(rownames(tokens$errors))
+asjp[error_lines,]
+
+# remove errors for now (this has to be done in the original data of course!)
+# make "empirical" profile of graphemes, showing all tailored graphemes in ASJP
+# there are 793 different tailored graphemes in the data !!!
+no_errors <- tokens$strings$tokenized[-error_lines]
+write.profile(no_errors, sep = " ", file = "sandbox/asjp_emprical_profile.csv")

### 2015-09-30 use_cases/ASJP/manual.R
minor corrections to ASJP manual

@@ -46,4 +46,9 @@ asjp[error_lines,]
 # make "empirical" profile of graphemes, showing all tailored graphemes in ASJP
 # there are 793 different tailored graphemes in the data !!!
 no_errors <- tokens$strings$tokenized[-error_lines]
+tailored <- write.profile(no_errors, sep = " ")
+str(tailored)
+head(tailored, 20)
+
+# save the profile as a file
 write.profile(no_errors, sep = " ", file = "sandbox/asjp_emprical_profile.csv")

### 2015-10-01 use_cases/ASJP/manual.R
usability update to ASJP use case

@@ -4,8 +4,8 @@
 #' date: "30 September 2015"
 #' ---
 
+#' === STEP 0: Preparation ===
 # install library from github
-# only necessary once
 # install.packages("devtools")
 # devtools::install_github("cysouw/qlcTokenize")
 library(qlcTokenize)
@@ -14,41 +14,47 @@ library(qlcTokenize)
 # this is pretty large, so it might take a bit
 asjp <- read.csv("data/ASJP.csv")
 
+#' === STEP 1: making a draft profile ===
 # the actual words are in the fifth column of this csv-file ("Value")
 # first make draft profile using default settings
-write.profile(asjp$Value, file = "sandbox/asjp_draft_profile.csv")
+write.profile(asjp$Value, file = "sandbox/1_asjp_draft_profile.tsv")
 
+#' === STEP 2: correcting the draft profile ===
 # There is a warning because there are tabs in the input data
 # This is an error in the ASJP data, but we leave it in for now
 # It has to be changed in the orthography profile though!
-# Check the manually changed profile in profiles/asjp_corrected_profile.csv
+# Check the manually changed profile in data/2_asjp_corrected_profile.tsv
+# This profile should be capturing the ASJP orthography
 
-
+#' === STEP 3: tokenize the data and inspect errors ===
 # after manually correcting the profile we can use it for tokenization
 # there are regular expressions in the corrected profile, so "regex = TRUE"
+# save all the information from the tokenization in the sandbox
 # the code is not very fast, might take about half a minute for all of ASJP
 tokens <- tokenize(asjp$Value
-					, profile = "profiles/asjp_corrected_profile.csv"
+					, profile = "data/2_asjp_corrected_profile.tsv"
 					, regex = TRUE
+					, file.out = "sandbox/3_asjp_result"
 					)
-					
 # there are errors in the data. Let's check the errors
 tokens$errors
-
 # and look at the characters that are not in the profile
 tokens$missing
 
-# look at the errors in this original context
-error_lines <- as.numeric(rownames(tokens$errors))
+# look at the errors in their original context
+error_lines <- rownames(tokens$errors)
 asjp[error_lines,]
 
+#' === STEP 4: error correction and investigate tailored graphemes ===
 # remove errors for now (this has to be done in the original data of course!)
-# make "empirical" profile of graphemes, showing all tailored graphemes in ASJP
[…]

### 2015-10-01 use_cases/dutch/manual (Autosaved).R
first draft dutch

@@ -1,0 +1,32 @@
+
+
+library(qlcTokenize)
+
+# get examples from profile
+examples <- read.delim("data/dutch_IPA_profile.tsv", quote = "")$Examples
+examples <- unlist(strsplit(as.character(examples), ", "))
+
+# get examples from corpus
+nld10K <- read.delim("data/nld10Kwords.txt", header = FALSE, quote = "")
+corpus_words <- as.character(nld10K[,2])
+
+# write draft profile
+write.profile(corpus_words, file = "sandbox/nld10K_draft_profile.tsv", editing =  TRUE)
+
+# remove all words with obviously non-dutch characters
+tokens <- tokenize(corpus_words, profile = "data/nld10K_corrected_profile.tsv", transliterate = "Replacement", sep = "")
+
+# just remove all words that include some strange characters
+# and everything that starts or ends with a dash
+
+errors <- as.numeric(rownames(tokens$errors))
+corpus_words <- tokens$strings[-errors,]$transliterated
+
+
+
+words <- sort(unique(c(examples, corpus_words)))
+write(words, file = "sandbox/dutch_words.txt")
+
+words <- scan(file = "data/dutch_words.txt", what = "character")
+
+dutch <- tokenize(words, profile = "data/dutch_IPA_profile.tsv", transliterate = "IPA", file = "tmp", regex = TRUE)

### 2015-10-01 use_cases/dutch/manual.R

@@ -1,0 +1,28 @@
+
+
+library(qlcTokenize)
+
+# get examples from profile
+examples <- read.delim("data/dutch_IPA_profile.tsv", quote = "")$Examples
+examples <- unlist(strsplit(as.character(examples), ", "))
+
+# get examples from corpus
+nld10K <- read.delim("data/nld10Kwords.txt", header = FALSE, quote = "")
+corpus_words <- as.character(nld10K[,2])
+
+# write draft profile
+write.profile(corpus_words, file = "sandbox/nld10K_draft_profile.tsv", editing =  TRUE)
+
+# remove all obviously non-dutch characters
+tokens <- tokenize(corpus_words, profile = "data/nld10K_corrected_profile.tsv", transliterate = "Replacement", sep = "")
+
+# just remove all words that include some strange characters
+errors <- as.numeric(rownames(tokens$errors))
+corpus_words <- tokens$strings[-errors,]$transliterated
+
+words <- sort(unique(c(examples, corpus_words)))
+write(words, file = "sandbox/dutch_words.txt")
+
+words <- scan(file = "data/dutch_words.txt", what = "character")
+
+dutch <- tokenize(words, profile = "data/dutch_IPA_profile.tsv", transliterate = "IPA", file = "tmp", regex = TRUE)

### 2015-10-01 use_cases/dutch/code/correctwordlist.R
update dutch

@@ -1,0 +1,41 @@
+#' ---
+#' title: "Preparation of dutch wordlist"
+#' author: "Michael Cysouw"
+#' date: "1 October 2015"
+#' ---
+
+library(qlcTokenize)
+
+# get examples from handmade profile
+examples <- read.delim("../data/dutch_IPA_profile.tsv", quote = "")$Examples
+examples <- unlist(strsplit(as.character(examples), ", "))
+
+# get examples from corpus
+nld10K <- read.delim("../sources/nld_mixed_2012_10K-words.txt", header = FALSE, quote = "")
+corpus_words <- as.character(nld10K[,2])
+
+# write draft profile of corpus
+write.profile(corpus_words
+			, file = "../sandbox/nld10K_draft_profile.tsv"
+			, editing =  TRUE
+			)
+
+# edit the profile by hand, removing strange characters
+# and adding some conditional constraints of character occurrence
+# this leads to the profile in "data/nld10K_corrected_profile.tsv"
+
+# remove all words with obviously non-dutch characters
+# just remove all words that include some strange characters
+tokens <- tokenize(corpus_words
+					, profile = "nld10K_corrected_profile.tsv"
+					, transliterate = "Replacement"
+					, sep = ""
+					, regex = TRUE
+					)
+errors <- as.numeric(rownames(tokens$errors))
+corpus_words <- tokens$strings[-errors,]$transliterated
+
+# combine the corpus with the hand-selected examples
+# write these to a file
+words <- sort(unique(c(examples, corpus_words)))
+write(words, file = "../sandbox/dutch_words.txt")

### 2015-10-01 use_cases/dutch/manual (Autosaved).R

@@ -1,32 +1,0 @@
-
-
-library(qlcTokenize)
-
-# get examples from profile
-examples <- read.delim("data/dutch_IPA_profile.tsv", quote = "")$Examples
-examples <- unlist(strsplit(as.character(examples), ", "))
-
-# get examples from corpus
-nld10K <- read.delim("data/nld10Kwords.txt", header = FALSE, quote = "")
-corpus_words <- as.character(nld10K[,2])
-
-# write draft profile
-write.profile(corpus_words, file = "sandbox/nld10K_draft_profile.tsv", editing =  TRUE)
-
-# remove all words with obviously non-dutch characters
-tokens <- tokenize(corpus_words, profile = "data/nld10K_corrected_profile.tsv", transliterate = "Replacement", sep = "")
-
-# just remove all words that include some strange characters
-# and everything that starts or ends with a dash
-
-errors <- as.numeric(rownames(tokens$errors))
-corpus_words <- tokens$strings[-errors,]$transliterated
-
-
-
-words <- sort(unique(c(examples, corpus_words)))
-write(words, file = "sandbox/dutch_words.txt")
-
-words <- scan(file = "data/dutch_words.txt", what = "character")
-
-dutch <- tokenize(words, profile = "data/dutch_IPA_profile.tsv", transliterate = "IPA", file = "tmp", regex = TRUE)

### 2015-10-01 use_cases/dutch/manual.R

@@ -1,28 +1,18 @@
-
+#' ---
+#' title: "IPA-transliteration of dutch words"
+#' author: "Michael Cysouw"
+#' date: "1 October 2015"
+#' ---
 
 library(qlcTokenize)
 
-# get examples from profile
-examples <- read.delim("data/dutch_IPA_profile.tsv", quote = "")$Examples
-examples <- unlist(strsplit(as.character(examples), ", "))
 
-# get examples from corpus
-nld10K <- read.delim("data/nld10Kwords.txt", header = FALSE, quote = "")
-corpus_words <- as.character(nld10K[,2])
-
-# write draft profile
-write.profile(corpus_words, file = "sandbox/nld10K_draft_profile.tsv", editing =  TRUE)
-
-# remove all obviously non-dutch characters
-tokens <- tokenize(corpus_words, profile = "data/nld10K_corrected_profile.tsv", transliterate = "Replacement", sep = "")
-
-# just remove all words that include some strange characters
-errors <- as.numeric(rownames(tokens$errors))
-corpus_words <- tokens$strings[-errors,]$transliterated
-
-words <- sort(unique(c(examples, corpus_words)))
-write(words, file = "sandbox/dutch_words.txt")
-
+# read the wordlist with dutch words
 words <- scan(file = "data/dutch_words.txt", what = "character")
 
-dutch <- tokenize(words, profile = "data/dutch_IPA_profile.tsv", transliterate = "IPA", file = "tmp", regex = TRUE)
+dutch <- tokenize(words
+				, profile = "data/dutch_IPA_profile.tsv"
+				, transliterate = "IPA"
+				, file = "sandbox/dutch"
+				, regex = TRUE
+				)

### 2015-10-01 use_cases/dutch/code/correctwordlist.R
finished dutch use case

@@ -7,7 +7,7 @@
 library(qlcTokenize)
 
 # get examples from handmade profile
-examples <- read.delim("../data/dutch_IPA_profile.tsv", quote = "")$Examples
+examples <- read.delim("../data/1_dutch_IPA_profile.tsv", quote = "")$Examples
 examples <- unlist(strsplit(as.character(examples), ", "))
 
 # get examples from corpus
@@ -38,4 +38,4 @@ corpus_words <- tokens$strings[-errors,]$transliterated
 # combine the corpus with the hand-selected examples
 # write these to a file
 words <- sort(unique(c(examples, corpus_words)))
-write(words, file = "../sandbox/dutch_words.txt")
+write(words, file = "../sandbox/2_dutch_words.txt")

### 2015-10-01 use_cases/dutch/manual.R

@@ -5,14 +5,24 @@
 #' ---
 
 library(qlcTokenize)
-
-
 # read the wordlist with dutch words
-words <- scan(file = "data/dutch_words.txt", what = "character")
+words <- scan(file = "data/2_dutch_words.txt", what = "character")
 
+# use the hand-made orthography profile in `data/dutch_IPA_profile.tsv`
 dutch <- tokenize(words
-				, profile = "data/dutch_IPA_profile.tsv"
+				, profile = "data/1_dutch_IPA_profile.tsv"
 				, transliterate = "IPA"
-				, file = "sandbox/dutch"
+				, file = "sandbox/3_dutch"
 				, regex = TRUE
+				, sep = ""
 				)
+
+# there are only a errors coming from the tokenization
+dutch$errors
+
+# the IPA transliterations seems to be fine
+# there are some errors, but only a few
+# let's just check a random selection 
+# (note that this selection will change every time this code is executed)
+selection <- sample(1:nrow(dutch$strings), 50)
+dutch$strings[selection, c("originals", "transliterated")]

### 2015-10-02 use_cases/JIPA/manual.R
JIPA testing with R

@@ -1,0 +1,33 @@
+#' ---
+#' title: "Tokenization of JIPA transcriptions"
+#' author: "Michael Cysouw"
+#' date: "2 October 2015"
+#' ---
+
+library(qlcTokenize)
+
+# adhoc function to compare tokenization with hand-prepared results
+compare <- function(jipa) {
+	text_in <- paste0("data/", jipa, "_input.txt")
+	text_out <- paste0("data/", jipa, "_output.txt")
+	
+	input <- scan(text_in, what = "character", sep = "\n")
+	output <- scan(text_out, what = "character", sep = "\n")
+	
+	tokenized <- tokenize(input
+						, profile = "profiles/validIPA2005_graphemes.tsv"
+						, sep = " "
+						, sep.replace = "#"
+						, file = "tmp"
+						, regex = TRUE
+						, normalize = "NFD"
+						)
+	all.equal(output, tokenized$strings$tokenized)
+
+}
+
+# all true!
+compare("Brazilian_Portuguese")
+compare("Kabiye")
+compare("Vietnamese")
+compare("Zurich_German")

### 2015-10-02 use_cases/ASJP/manual.R
update ASJP manual.R

@@ -1,32 +1,34 @@
 #' ---
 #' title: "Analysis of ASJP data"
 #' author: "Michael Cysouw"
-#' date: "30 September 2015"
+#' date: "`r Sys.Date()`"
 #' ---
 
-#' === STEP 0: Preparation ===
+#' ### STEP 0: Preparation
 # install library from github
 # install.packages("devtools")
 # devtools::install_github("cysouw/qlcTokenize")
 library(qlcTokenize)
+# for the nice display of tables
+require(knitr)
 
 # read ASJP data
 # this is pretty large, so it might take a bit
 asjp <- read.csv("data/ASJP.csv")
 
-#' === STEP 1: making a draft profile ===
+#' ### STEP 1: making a draft profile
 # the actual words are in the fifth column of this csv-file ("Value")
 # first make draft profile using default settings
 write.profile(asjp$Value, file = "sandbox/1_asjp_draft_profile.tsv")
 
-#' === STEP 2: correcting the draft profile ===
+#' ### STEP 2: correcting the draft profile
 # There is a warning because there are tabs in the input data
 # This is an error in the ASJP data, but we leave it in for now
 # It has to be changed in the orthography profile though!
 # Check the manually changed profile in data/2_asjp_corrected_profile.tsv
 # This profile should be capturing the ASJP orthography
 
-#' === STEP 3: tokenize the data and inspect errors ===
+#' ### STEP 3: tokenize the data and inspect errors
 # after manually correcting the profile we can use it for tokenization
 # there are regular expressions in the corrected profile, so "regex = TRUE"
 # save all the information from the tokenization in the sandbox
@@ -37,15 +39,15 @@ tokens <- tokenize(asjp$Value
 					, file.out = "sandbox/3_asjp_result"
 					)
 # there are errors in the data. Let's check the errors
-tokens$errors
+knitr::kable(tokens$errors)
 # and look at the characters that are not in the profile
-tokens$missing
+knitr::kable(tokens$missing)
 
 # look at the errors in their original context
 error_lines <- rownames(tokens$errors)
-asjp[error_lines,]
+knitr::kable(asjp[error_lines,])
 
-#' === STEP 4: error correction and investigate tailored graphemes ===
+#' ### STEP 4: error correction and investigate tailored graphemes
 # remove errors for now (this has to be done in the original data of course!)
 no_errors <- tokens$strings[-as.numeric(error_lines),]
[…]

### 2015-10-02 use_cases/dutch/manual.R
update dutch manual.R

@@ -1,10 +1,14 @@
 #' ---
 #' title: "IPA-transliteration of dutch words"
 #' author: "Michael Cysouw"
-#' date: "1 October 2015"
+#' date: "`r Sys.Date()`"
 #' ---
 
+# load libraries
+# knitr is needed for a nice display of the tables
 library(qlcTokenize)
+require(knitr)
+
 # read the wordlist with dutch words
 words <- scan(file = "data/2_dutch_words.txt", what = "character")
 
@@ -17,12 +21,12 @@ dutch <- tokenize(words
 				, sep = ""
 				)
 
-# there are only a errors coming from the tokenization
-dutch$errors
+# there are only a few errors coming from the tokenization
+knitr::kable(dutch$errors)
 
 # the IPA transliterations seems to be fine
 # there are some errors, but only a few
 # let's just check a random selection 
 # (note that this selection will change every time this code is executed)
 selection <- sample(1:nrow(dutch$strings), 50)
-dutch$strings[selection, c("originals", "transliterated")]
+knitr::kable(dutch$strings[selection, c("originals", "transliterated")])

### 2015-10-02 use_cases/JIPA/manual.R
update JIPA manual.R

@@ -1,13 +1,16 @@
 #' ---
 #' title: "Tokenization of JIPA transcriptions"
 #' author: "Michael Cysouw"
-#' date: "2 October 2015"
+#' date: "`r Sys.Date()`"
 #' ---
 
 library(qlcTokenize)
 
+#' ### Check automatic tokenization
+# check it against hand-made tokenization
 # adhoc function to compare tokenization with hand-prepared results
-compare <- function(jipa) {
+
+compare <- function(jipa, profile) {
 	text_in <- paste0("data/", jipa, "_input.txt")
 	text_out <- paste0("data/", jipa, "_output.txt")
 	
@@ -15,19 +18,63 @@ compare <- function(jipa) {
 	output <- scan(text_out, what = "character", sep = "\n")
 	
 	tokenized <- tokenize(input
-						, profile = "profiles/validIPA2005_graphemes.tsv"
+						, profile = profile
 						, sep = " "
 						, sep.replace = "#"
-						, file = "tmp"
 						, regex = TRUE
 						, normalize = "NFD"
-						)
+						)						
+
 	all.equal(output, tokenized$strings$tokenized)
 
 }
 
-# all true!
-compare("Brazilian_Portuguese")
-compare("Kabiye")
-compare("Vietnamese")
-compare("Zurich_German")
+# all true as valid IPA!
+profile <- "profiles/validIPA2005_graphemes.tsv"
+compare("Brazilian_Portuguese", profile)
+compare("Kabiye", profile)
+compare("Zurich_German", profile)
+compare("Vietnamese", profile)
+
+# not all conform to the strict IPA interpretation
+profile <- "profiles/strictIPA2005_graphemes.tsv"
+compare("Brazilian_Portuguese", profile)
+compare("Kabiye", profile)
+compare("Zurich_German", profile)
+compare("Vietnamese", profile)
+
+#' ### Tailored grapheme clusters
+# list all different tailored graphemes with frequencies
+# write the profiles with the tailored graphemes to the sandbox
+# disaply requires the knitr library
+require("knitr")	
+
+tailored <- function(jipa) {
+	input <- scan(paste0("data/", jipa, "_input.txt")
+				, what = "character"
+				, sep = "\n"
+				)
+	
+	tokenized <- tokenize(input
+						, profile = "profiles/validIPA2005_graphemes.tsv"
+						, sep = " "
+						, sep.replace = ""
+						, regex = TRUE
[…]

### 2015-10-02 book/chapters/orthography_profiles.tex
update formal specifications 4.3

--- 2015-09-28 chapters/orthography_profiles.tex
@@
        header. For all columns we assume the name in the header of the CSV file
-       to be crucial. The actual ordering of the columns is unimportant. 
+       to be crucial. The actual ordering of the columns is unimportant. Because 
+       tabs and newlines are field separators of the profile, these characters 
+       will lead to problems when they occur in the original data.\footnote{The main
+       reason to choose for tab-delimited, and not for quoted comma-separated files
+       is that we want the profiles to be easily manipulated by hand. Tab-separation 
+       is strongly preferred for this reason.}
 	\item \textsc{Metadata are added in a separate UTF-8 text file} using a basic
@@
        literally in the orthography. The following symbols will need to be
-       preceded by a backslash: {[} {]} ( ) \{ \} ~+ * . - ! ? \^{} \$ and the
+       preceded by a backslash: {[} {]} ( ) \{ \} | ~+ * . - ! ? \^{} \$ and the
        backslash \textbackslash~itself. 
@@
 		\item \textsc{the tokenization method}, i.e.~whether the tokenization should
-        proceed \textsc{global} or \textsc{linear}. 
+        proceed \textsc{global} or \textsc{linear} (see B6 above). 
 		\item \textsc{unicode normalization}, i.e.~whether the text-string and profile
@@
     \end{enumerate}
-	\item \textsc{might offer user-options }to:
+	\item \textsc{might offer user-options}:
 	\begin{enumerate}
 		\def\labelenumii{C\arabic{enumii}.} \setcounter{enumii}{5} 
-		\item \textsc{assist in the ordering of the graphemes.} In our experience, it
+		\item \textsc{to assist in the ordering of the graphemes.} In our experience, it
         makes sense to apply larger graphemes before shorter graphemes, and to
@@
         exceptions or special cases. 
-		\item \textsc{assist in dealing with upper and lower case characters.} It
+		\item \textsc{to assist in dealing with upper and lower case characters.} It
         seems practical to offer some basic case matching, so characters like
@@
         should be performed. 
-		\item \textsc{treat the profile literal}, i.e.~to not interpret regular
+		\item \textsc{to treat the profile literal}, i.e.~to not interpret regular
         expression metacharacters. Matching graphemes literally often leads to
@@
 		\def\labelenumii{C\arabic{enumii}.} \setcounter{enumii}{8} 
-		\item \textsc{the original text-strings to be processed in the specified
-        unicode normalization}, i.e.~in either NFC or NFD as specified by the
+		\item \textsc{the original text-strings} to be processed in the specified
+        unicode normalization, i.e.~in either NFC or NFD as specified by the
         user. 

### 2015-10-15 use_cases/ASJP/manual.R
update manual to new R package

@@ -7,8 +7,8 @@
 #' ### STEP 0: Preparation
 # install library from github
 # install.packages("devtools")
-# devtools::install_github("cysouw/qlcTokenize")
-library(qlcTokenize)
+# devtools::install_github("cysouw/qlcData")
+library(qlcData)
 # for the nice display of tables
 require(knitr)
 

### 2015-10-15 use_cases/JIPA/manual.R

@@ -4,7 +4,7 @@
 #' date: "`r Sys.Date()`"
 #' ---
 
-library(qlcTokenize)
+library(qlcData)
 
 #' ### Check automatic tokenization
 # check it against hand-made tokenization

### 2015-10-15 use_cases/dutch/manual.R

@@ -6,7 +6,7 @@
 
 # load libraries
 # knitr is needed for a nice display of the tables
-library(qlcTokenize)
+library(qlcData)
 require(knitr)
 
 # read the wordlist with dutch words

### 2015-10-17 book/chapters/codetesting.tex
testing code inclusion

\chapter{Testing code inclusing}

Trying to include code through knitr. Different languages can be easily 
included!

<<bash-test, engine='bash'>>=
echo 'this is a simple bash example'
ls | wc
@

<<>>=
'R works out of the box'
(x <- 3+4)
Sys.getlocale("LC_CTYPE")
'aɽɮz'
@

You can refer to variables by using \textbackslash Sexpr{}. 
I think \Sexpr{x} is an interesting number!

<<python-test, engine='python'>>=
'python also works!'
print('bla' + 'bla')
print(3+4)
@

You have to add the option `engine.paht='python3'' to get Python r

<<python3-test, engine='python', engine.path='python3'>>=
print('python3 needs an extra tag')
uni = 'aɽɮz'
print(uni)
@

Testing: does referencing to variables also work in Python? It does give an error, so no... !

### 2015-10-17 book/unicode.tex
some general knitr settings added

@@
 		biblatex
-		]{LSP/langsci}
+		]{LSP/langsci}\usepackage[]{graphicx}\usepackage[]{color}
+%% maxwidth is the original width if it is less than linewidth
+%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+\definecolor{fgcolor}{rgb}{0, 0, 0}
+\newcommand{\hlnum}[1]{\textcolor[rgb]{0.533,0,0.133}{#1}}%
+\newcommand{\hlstr}[1]{\textcolor[rgb]{0.667,0.267,0}{#1}}%
+\newcommand{\hlcom}[1]{\textcolor[rgb]{1,0.533,0}{#1}}%
+\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{\textbf{#1}}}%
+\newcommand{\hlstd}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.4,0.067,0.067}{\textbf{#1}}}%
+\newcommand{\hlkwb}[1]{\textcolor[rgb]{0,0,0.4}{\textbf{#1}}}%
+\newcommand{\hlkwc}[1]{\textcolor[rgb]{0,0,0.4}{#1}}%
+\newcommand{\hlkwd}[1]{\textcolor[rgb]{0,0.267,0.4}{#1}}%
+
+\usepackage{framed}
+\makeatletter
+\newenvironment{kframe}{%
+ \def\at@end@of@kframe{}%
+ \ifinner\ifhmode%
+  \def\at@end@of@kframe{\end{minipage}}%
+  \begin{minipage}{\columnwidth}%
+ \fi\fi%
+ \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+ \colorbox{shadecolor}{##1}\hskip-\fboxsep
+     % There is no \\@totalrightmargin, so:
+     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+ \MakeFramed {\advance\hsize-\width
+   \@totalleftmargin\z@ \linewidth\hsize
+   \@setminipage}}%
+ {\par\unskip\endMakeFramed%
+ \at@end@of@kframe}
+\makeatother
+
+\definecolor{shadecolor}{rgb}{.97, .97, .97}
+\definecolor{messagecolor}{rgb}{0, 0, 0}
+\definecolor{warningcolor}{rgb}{1, 0, 1}
+\definecolor{errorcolor}{rgb}{1, 0, 0}
+\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+
+\usepackage{alltt}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
@@
 %%%                                              %%%
+%%%             knitr settings                   %%%
+%%%                                              %%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+% for themes, see:
+% https://rclickhandbuch.files.wordpress.com/2014/09/knitrthemesoverview.pdf
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%                                              %%%
 %%%             Chapters                         %%%
@@
 \include{chapters/use_cases}
+
+\chapter{Testing code inclusing}
+
+Trying to include code through knitr. Different languages can be easily 
+included!
+
+\begin{knitrout}\scriptsize
+\definecolor{shadecolor}{rgb}{0.965, 0.965, 0.965}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+echo 'this is a simple bash example'
+ls | wc
+\end{alltt}
+
+\begin{verbatim}
+## this is a simple bash example
+##       10      10     165
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+\begin{knitrout}\scriptsize
+\definecolor{shadecolor}{rgb}{0.965, 0.965, 0.965}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlstr{'R works out of the box'}
+\end{alltt}
+\begin{verbatim}
+## [1] "R works out of the box"
+\end{verbatim}
+\begin{alltt}
+\hlstd{(x} \hlkwb{<-} \hlnum{3}\hlopt{+}\hlnum{4}\hlstd{)}
+\end{alltt}
+\begin{verbatim}
+## [1] 7
+\end{verbatim}
+\begin{alltt}
+\hlkwd{Sys.getlocale}\hlstd{(}\hlstr{"LC_CTYPE"}\hlstd{)}
+\end{alltt}
+\begin{verbatim}
+## [1] "en_US.UTF-8"
+\end{verbatim}
+\begin{alltt}
+\hlstr{'aɽɮz'}
+\end{alltt}
+\begin{verbatim}
+## [1] "aɽɮz"
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+You can refer to variables by using \textbackslash Sexpr{}. 
+I think 7 is an interesting number!
+
+\begin{knitrout}\scriptsize
+\definecolor{shadecolor}{rgb}{0.965, 0.965, 0.965}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+'python also works!'
+print('bla' + 'bla')
+print(3+4)
+\end{alltt}
+
+\begin{verbatim}
+## blabla
+## 7
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+You have to add the option `engine.paht='python3'' to get Python r
+
+\begin{knitrout}\scriptsize
+\definecolor{shadecolor}{rgb}{0.965, 0.965, 0.965}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+print('python3 needs an extra tag')
+uni = 'aɽɮz'
+print(uni)
+\end{alltt}
+
+\begin{verbatim}
+## python3 needs an extra tag
+## aɽɮz
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+Testing: does referencing to variables also work in Python? It does give an error, so no... !
 

### 2015-10-18 book/chapters/codetesting.tex
some more knitr testing

@@
-\chapter{Testing code inclusing}
+\chapter{Testing code inclusion}
 
-Trying to include code through knitr. Different languages can be easily 
+Trying to include code through knitr. 
+Different languages can be easily 
 included!
 
+There is a bug for non-R engines: the fontsize settings and line spaceing are not correct. I have 
+filed an issue, hopefully this will be fixed soon.
+
+For syntax colouring, install the package `highlight' and for nice tables 
+install the package `xtable'.
+
 <<bash-test, engine='bash'>>=
+echo 'this is a simple bash example'
 echo 'this is a simple bash example'
@@
 <<>>=
-'R works out of the box'
-(x <- 3+4)
-Sys.getlocale("LC_CTYPE")
-'aɽɮz'
+'this is a simple R example'
+test = 3+4
+test
+@
+
+<<>>=
+library(qlcData)
+example  = 'AΑА'
+profile = write.profile(example)
+@
+
+<<Rtest2, tidy = TRUE, results = 'asis', echo = FALSE>>=
+library(xtable)
+print(xtable(profile, caption = 'test', label = 'bla')
+  , include.rownames = FALSE
+  , booktabs = TRUE
+  , size = 'small'
+  , table.placement = 'H'
+  )
 @
@@
 You can refer to variables by using \textbackslash Sexpr{}. 
-I think \Sexpr{x} is an interesting number!
+For example, the length of the example string is \Sexpr{nchar(example)}.
+Crossreferencing also works, see Table~\ref{bla}.
 
 <<python-test, engine='python'>>=
-'python also works!'
 print('bla' + 'bla')
@@
 
-You have to add the option `engine.paht='python3'' to get Python r
+You have to add the option ``engine.path='python3'' to get Python 3.
 

### 2015-10-18 book/unicode.tex
some more knitr testing

@@
 \definecolor{fgcolor}{rgb}{0, 0, 0}
-\newcommand{\hlnum}[1]{\textcolor[rgb]{0.533,0,0.133}{#1}}%
-\newcommand{\hlstr}[1]{\textcolor[rgb]{0.667,0.267,0}{#1}}%
-\newcommand{\hlcom}[1]{\textcolor[rgb]{1,0.533,0}{#1}}%
-\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{\textbf{#1}}}%
+\newcommand{\hlnum}[1]{\textcolor[rgb]{0.502,0,0.502}{\textbf{#1}}}%
+\newcommand{\hlstr}[1]{\textcolor[rgb]{0.651,0.522,0}{#1}}%
+\newcommand{\hlcom}[1]{\textcolor[rgb]{1,0.502,0}{#1}}%
+\newcommand{\hlopt}[1]{\textcolor[rgb]{1,0,0.502}{\textbf{#1}}}%
 \newcommand{\hlstd}[1]{\textcolor[rgb]{0,0,0}{#1}}%
-\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.4,0.067,0.067}{\textbf{#1}}}%
-\newcommand{\hlkwb}[1]{\textcolor[rgb]{0,0,0.4}{\textbf{#1}}}%
-\newcommand{\hlkwc}[1]{\textcolor[rgb]{0,0,0.4}{#1}}%
+\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.733,0.475,0.467}{\textbf{#1}}}%
+\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.502,0.502,0.753}{\textbf{#1}}}%
+\newcommand{\hlkwc}[1]{\textcolor[rgb]{0,0.502,0.753}{#1}}%
 \newcommand{\hlkwd}[1]{\textcolor[rgb]{0,0.267,0.4}{#1}}%
@@
 
-\chapter{Testing code inclusing}
+\chapter{Testing code inclusion}
 
-Trying to include code through knitr. Different languages can be easily 
+Trying to include code through knitr. 
+Different languages can be easily 
 included!
 
+There is a bug for non-R engines: the fontsize settings are not correct. I have 
+filed an issue, hopefully this will be fixed soon.
+
+For syntax colouring, install the package `highlight' and for nice tables 
+install the package `xtable'.
+
 \begin{knitrout}\scriptsize
-\definecolor{shadecolor}{rgb}{0.965, 0.965, 0.965}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-echo 'this is a simple bash example'
-ls | wc
-\end{alltt}
+\definecolor{shadecolor}{rgb}{0.933, 0.933, 0.933}\color{fgcolor}\begin{kframe}
+\noindent
+\ttfamily
+\hlstd{}\hlkwb{echo\ }\hlstd{}\hlstr{'this\ is\ a\ simple\ bash\ example'}\hlstd{}\hspace*{\fill}\\
+\hlkwb{echo\ }\hlstd{}\hlstr{'this\ is\ a\ simple\ bash\ example'}\hlstd{}\hspace*{\fill}\\
+\hlkwc{ls\ }\hlstd{\textbar \ }\hlkwc{wc}\hlstd{}\hspace*{\fill}
+\mbox{}
+\normalfont
+\normalsize
 
 \begin{verbatim}
+## this is a simple bash example
 ## this is a simple bash example
@@
 \begin{knitrout}\scriptsize
-\definecolor{shadecolor}{rgb}{0.965, 0.965, 0.965}\color{fgcolor}\begin{kframe}
+\definecolor{shadecolor}{rgb}{0.933, 0.933, 0.933}\color{fgcolor}\begin{kframe}
 \begin{alltt}
-\hlstr{'R works out of the box'}
+\hlstr{"this is a simple R example"}
 \end{alltt}
 \begin{verbatim}
-## [1] "R works out of the box"
+## [1] "this is a simple R example"
 \end{verbatim}
 \begin{alltt}
-\hlstd{(x} \hlkwb{<-} \hlnum{3}\hlopt{+}\hlnum{4}\hlstd{)}
+\hlstd{test} \hlkwb{<-} \hlnum{3} \hlopt{+} \hlnum{4}
+\hlstd{test}
 \end{alltt}
@@
 \end{verbatim}
-\begin{alltt}
-\hlkwd{Sys.getlocale}\hlstd{(}\hlstr{"LC_CTYPE"}\hlstd{)}
-\end{alltt}
-\begin{verbatim}
-## [1] "en_US.UTF-8"
-\end{verbatim}
-\begin{alltt}
-\hlstr{'aɽɮz'}
-\end{alltt}
-\begin{verbatim}
-## [1] "aɽɮz"
-\end{verbatim}
 \end{kframe}
@@
 
+\begin{knitrout}\scriptsize
+\definecolor{shadecolor}{rgb}{0.933, 0.933, 0.933}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlkwd{library}\hlstd{(qlcData)}
+\hlstd{example} \hlkwb{<-} \hlstr{"AΑА"}
+\hlstd{profile} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+% latex table generated in R 3.2.2 by xtable 1.7-4 package
+% Mon Oct 19 00:52:15 2015
+\begin{table}[H]
+\centering
+{\small
+\begin{tabular}{llll}
+  \toprule
+Grapheme & Frequency & Codepoint & UnicodeName \\ 
+  \midrule
+A & 1 & U+0041 & LATIN CAPITAL LETTER A \\ 
+  Α & 1 & U+0391 & GREEK CAPITAL LETTER ALPHA \\ 
+  А & 1 & U+0410 & CYRILLIC CAPITAL LETTER A \\ 
+   \bottomrule
+\end{tabular}
+}
+\caption{test} 
+\label{bla}
+\end{table}
+
 You can refer to variables by using \textbackslash Sexpr{}. 
-I think 7 is an interesting number!
+For example, the length of the example string is 3.
+Crossreferencing also works, see Table~\ref{bla}.
 
 \begin{knitrout}\scriptsize
-\definecolor{shadecolor}{rgb}{0.965, 0.965, 0.965}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-'python also works!'
-print('bla' + 'bla')
-print(3+4)
-\end{alltt}
+\definecolor{shadecolor}{rgb}{0.933, 0.933, 0.933}\color{fgcolor}\begin{kframe}
+\noindent
+\ttfamily
+\hlstd{}\hlkwa{print}\hlstd{}\hlopt{(}\hlstd{}\hlstr{'bla'}\hlstd{\ }\hlopt{+\ }\hlstd{}\hlstr{'bla'}\hlstd{}\hlopt{)}\hspace*{\fill}\\
+\hlstd{}\hlkwa{print}\hlstd{}\hlopt{(}\hlstd{}\hlnum{3}\hlstd{}\hlopt{+}\hlstd{}\hlnum{4}\hlstd{}\hlopt{)}\hlstd{}\hspace*{\fill}
+\mbox{}
+\normalfont
+\normalsize
 
@@
 
-You have to add the option `engine.paht='python3'' to get Python r
+You have to add the option ``engine.path='python3'' to get Python 3.
 
 \begin{knitrout}\scriptsize
-\definecolor{shadecolor}{rgb}{0.965, 0.965, 0.965}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-print('python3 needs an extra tag')
-uni = 'aɽɮz'
-print(uni)
-\end{alltt}
+\definecolor{shadecolor}{rgb}{0.933, 0.933, 0.933}\color{fgcolor}\begin{kframe}
+\noindent
+\ttfamily
+\hlstd{}\hlkwa{print}\hlstd{}\hlopt{(}\hlstd{}\hlstr{'python3\ needs\ an\ extra\ tag'}\hlstd{}\hlopt{)}\hspace*{\fill}\\
+\hlstd{uni\ }\hlopt{=\ }\hlstd{}\hlstr{'aɽɮz'}\hlstd{}\hspace*{\fill}\\
+\hlkwa{print}\hlstd{}\hlopt{(}\hlstd{uni}\hlopt{)}\hlstd{}\hspace*{\fill}
+\mbox{}
+\normalfont
+\normalsize
 

### 2015-10-19 book/chapters/codetesting.tex
some more knitr testing

@@
 'this is a simple R example'
+
 test = 3+4
@@
 library(qlcData) 
-example  = 'AΑА'
-profile = write.profile(example)
+profile = write.profile('AΑА')
 @
@@
 <<Rtest2, tidy = TRUE, results = 'asis', echo = FALSE>>=
-library(xtable)
 print(xtable(profile, caption = 'test', label = 'bla')

### 2015-10-19 book/unicode.tex
some more knitr testing

@@
 
-\definecolor{fgcolor}{rgb}{0, 0, 0}
-\newcommand{\hlnum}[1]{\textcolor[rgb]{0.502,0,0.502}{\textbf{#1}}}%
-\newcommand{\hlstr}[1]{\textcolor[rgb]{0.651,0.522,0}{#1}}%
-\newcommand{\hlcom}[1]{\textcolor[rgb]{1,0.502,0}{#1}}%
-\newcommand{\hlopt}[1]{\textcolor[rgb]{1,0,0.502}{\textbf{#1}}}%
-\newcommand{\hlstd}[1]{\textcolor[rgb]{0,0,0}{#1}}%
-\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.733,0.475,0.467}{\textbf{#1}}}%
-\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.502,0.502,0.753}{\textbf{#1}}}%
-\newcommand{\hlkwc}[1]{\textcolor[rgb]{0,0.502,0.753}{#1}}%
-\newcommand{\hlkwd}[1]{\textcolor[rgb]{0,0.267,0.4}{#1}}%
+\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
+\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
+\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
+\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
+\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
+\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
+\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
+\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
+\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
 
@@
 
-There is a bug for non-R engines: the fontsize settings are not correct. I have 
+There is a bug for non-R engines: the fontsize settings and line spaceing are not correct. I have 
 filed an issue, hopefully this will be fixed soon.
@@
 \begin{knitrout}\scriptsize
-\definecolor{shadecolor}{rgb}{0.933, 0.933, 0.933}\color{fgcolor}\begin{kframe}
-\noindent
-\ttfamily
-\hlstd{}\hlkwb{echo\ }\hlstd{}\hlstr{'this\ is\ a\ simple\ bash\ example'}\hlstd{}\hspace*{\fill}\\
-\hlkwb{echo\ }\hlstd{}\hlstr{'this\ is\ a\ simple\ bash\ example'}\hlstd{}\hspace*{\fill}\\
-\hlkwc{ls\ }\hlstd{\textbar \ }\hlkwc{wc}\hlstd{}\hspace*{\fill}
-\mbox{}
-\normalfont
-\normalsize
-
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
 \begin{verbatim}
+echo 'this is a simple bash example'
+echo 'this is a simple bash example'
+ls | wc
 ## this is a simple bash example
@@
 \begin{knitrout}\scriptsize
-\definecolor{shadecolor}{rgb}{0.933, 0.933, 0.933}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlstr{"this is a simple R example"}
-\end{alltt}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
 \begin{verbatim}
+"this is a simple R example"
 ## [1] "this is a simple R example"
-\end{verbatim}
-\begin{alltt}
-\hlstd{test} \hlkwb{<-} \hlnum{3} \hlopt{+} \hlnum{4}
-\hlstd{test}
-\end{alltt}
-\begin{verbatim}
+
+test <- 3 + 4
+test
 ## [1] 7
@@
 \begin{knitrout}\scriptsize
-\definecolor{shadecolor}{rgb}{0.933, 0.933, 0.933}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlkwd{library}\hlstd{(qlcData)}
-\hlstd{example} \hlkwb{<-} \hlstr{"AΑА"}
-\hlstd{profile} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example)}
-\end{alltt}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+library(qlcData)
+example <- "AΑА"
+profile <- write.profile(example)
+\end{verbatim}
 \end{kframe}
@@
 % latex table generated in R 3.2.2 by xtable 1.7-4 package
-% Mon Oct 19 00:52:15 2015
+% Mon Oct 19 18:08:51 2015
 \begin{table}[H]
@@
 \begin{knitrout}\scriptsize
-\definecolor{shadecolor}{rgb}{0.933, 0.933, 0.933}\color{fgcolor}\begin{kframe}
-\noindent
-\ttfamily
-\hlstd{}\hlkwa{print}\hlstd{}\hlopt{(}\hlstd{}\hlstr{'bla'}\hlstd{\ }\hlopt{+\ }\hlstd{}\hlstr{'bla'}\hlstd{}\hlopt{)}\hspace*{\fill}\\
-\hlstd{}\hlkwa{print}\hlstd{}\hlopt{(}\hlstd{}\hlnum{3}\hlstd{}\hlopt{+}\hlstd{}\hlnum{4}\hlstd{}\hlopt{)}\hlstd{}\hspace*{\fill}
-\mbox{}
-\normalfont
-\normalsize
-
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
 \begin{verbatim}
+print('bla' + 'bla')
+print(3+4)
 ## blabla
@@
 \begin{knitrout}\scriptsize
-\definecolor{shadecolor}{rgb}{0.933, 0.933, 0.933}\color{fgcolor}\begin{kframe}
-\noindent
-\ttfamily
-\hlstd{}\hlkwa{print}\hlstd{}\hlopt{(}\hlstd{}\hlstr{'python3\ needs\ an\ extra\ tag'}\hlstd{}\hlopt{)}\hspace*{\fill}\\
-\hlstd{uni\ }\hlopt{=\ }\hlstd{}\hlstr{'aɽɮz'}\hlstd{}\hspace*{\fill}\\
-\hlkwa{print}\hlstd{}\hlopt{(}\hlstd{uni}\hlopt{)}\hlstd{}\hspace*{\fill}
-\mbox{}
-\normalfont
-\normalsize
-
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
 \begin{verbatim}
+print('python3 needs an extra tag')
+uni = 'aɽɮz'
+print(uni)
 ## python3 needs an extra tag

### 2015-11-13 book/chapters/codetesting.tex
some more code inclusion testing and fine-tuning

@@
 echo 'this is a simple bash example'
-echo 'this is a simple bash example'
 ls | wc
@@
 
+<<engine='bash', echo = FALSE>>=
+echo -e 'note that bash chunks are executed as one, with all output at the end.
+You will have to make different chunks to get output separated.
+And add manual linebreaks!'
+@
+
+Spacing of lines is fine. I think without background looks better
+
 <<>>=
-'this is a simple R example'
+# adding comments inside code? Probably not a good idea
+(example <- 'this is a simple R example')
 

### 2015-11-13 book/unicode.tex
some more code inclusion testing and fine-tuning

@@
 
-\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
-\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
-\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
-\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
-\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
-\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
-\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
-\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
-\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
-\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+\definecolor{fgcolor}{rgb}{0, 0, 0}
+\newcommand{\hlnum}[1]{\textcolor[rgb]{0.502,0,0.502}{\textbf{#1}}}%
+\newcommand{\hlstr}[1]{\textcolor[rgb]{0.651,0.522,0}{#1}}%
+\newcommand{\hlcom}[1]{\textcolor[rgb]{1,0.502,0}{#1}}%
+\newcommand{\hlopt}[1]{\textcolor[rgb]{1,0,0.502}{\textbf{#1}}}%
+\newcommand{\hlstd}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.733,0.475,0.467}{\textbf{#1}}}%
+\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.502,0.502,0.753}{\textbf{#1}}}%
+\newcommand{\hlkwc}[1]{\textcolor[rgb]{0,0.502,0.753}{#1}}%
+\newcommand{\hlkwd}[1]{\textcolor[rgb]{0,0.267,0.4}{#1}}%
 
@@
 
-% for themes, see:
-% https://rclickhandbuch.files.wordpress.com/2014/09/knitrthemesoverview.pdf
+% Here are settings for knitr, which will be removed in the .tex file
+% spacing of code-chunks is set with the "knitrout" environment in localcommands.tex
 
@@
 \include{chapters/introduction}
+\include{chapters/unicode_background}
 \include{chapters/pitfalls}
+\include{chapters/ipa_background}
 \include{chapters/ipa_meets_unicode}
@@
 \begin{knitrout}\scriptsize
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
 \begin{verbatim}
 echo 'this is a simple bash example'
-echo 'this is a simple bash example'
 ls | wc
-## this is a simple bash example
 ## this is a simple bash example
@@
 \begin{knitrout}\scriptsize
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
 \begin{verbatim}
-"this is a simple R example"
+## note that bash chunks are executed as one, with all output at the end.
+## You will have to make different chunks to get output separated.
+## And add manual linebreaks!
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+Spacing of lines is fine. I think without background looks better
+
+\begin{knitrout}\scriptsize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+# adding comments inside code? Probably not a good idea
+(example <- "this is a simple R example")
 ## [1] "this is a simple R example"
@@
 \begin{knitrout}\scriptsize
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
 \begin{verbatim}
 library(qlcData)
-example <- "AΑА"
-profile <- write.profile(example)
+profile <- write.profile("AΑА")
 \end{verbatim}
@@
 % latex table generated in R 3.2.2 by xtable 1.7-4 package
-% Mon Oct 19 18:08:51 2015
+% Fri Nov 13 10:29:19 2015
 \begin{table}[H]
@@
 You can refer to variables by using \textbackslash Sexpr{}. 
-For example, the length of the example string is 3.
+For example, the length of the example string is 26.
 Crossreferencing also works, see Table~\ref{bla}.
@@
 \begin{knitrout}\scriptsize
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
 \begin{verbatim}

### 2015-11-13 book/chapters/ipa_background.tex
splitting chapters

--- 2015-07-23 chapters/ipa_meets_unicode.tex
@@
-\chapter{IPA meets Unicode}
-\label{ipa-meets-unicode}
-
-\section{Introduction}
-
-The International Phonetic Alphabet (IPA) is a common standard in linguistics to
-transcribe sounds of spoken language into discrete segments using a Latin-based
-alphabet. Although IPA is reasonably easily adhered to with pen and paper, it is
-not trivial to encode IPA characters electronically. Similar to the previous
-chapter, in this chapter we discuss various pitfalls with the encoding of IPA in
-the Unicode Standard. We will specifically refer to the 2005 version of the
-IPA \citep{IPA2005} and the 7.0 version of Unicode \citep{Unicode2014}.
-
-The details of the encoding are unimportant as long as the
-transcription is only directed towards phonetically trained eyes. For a linguist
-reading an IPA transcription, many of the details that will be discussed in this
-chapter might seem like hair-splitting trivialities. However, if IPA
-transcriptions are intended to be used across resources (e.g.~searching similar
-phenomena across different languages) then it becomes crucial that there are strict
-encoding guidelines. Our main goal in this chapter is to present the encoding
-issues and propose recommendations for a ``strict'' IPA encoding.
-
-There are several pitfalls to be aware of when using the Unicode Standard to
-encode IPA.\@ As we have said before, from a linguistic perspective it might
-sometimes look like the Unicode Consortium is making incomprehensible decisions,
-but it is important to realize that the consortium has tried and is continuing
-to try to be as consistent as possible across a wide range of use cases, and it
-does place linguistic traditions above other orthographic choices. Furthermore,
-when we look at the history of how the IPA met Unicode, we see that many of the
-decisions for IPA symbols in the Unicode Standard come directly from the
-International Phonetic Association itself. Therefore, many pitfalls that we will
-encounter have their grounding in the history of the principles of the IPA, as
-well as in the technological considerations involved in creating a single
-multilingual encoding. In general, we strongly suggest to linguists to not
-complain about any decisions in the Unicode Standard, but to try and understand
-the rationale of the International Phonetic Association and the Unicode
-Consortium (both of which are almost always well-conceived in our experience)
-and devise ways to work with any unexpected behavior.
-
-Many of the current problems derive from the fact that the
-IPA is clearly historically based on the Latin script, but different enough from
-most other Latin-based writing systems to warrant special attention. This
-ambivalent status of the IPA glyphs (partly Latin, partly special) is
-unfortunately also attested in the treatment of IPA in the Unicode Standard. In
-retrospect, it might have been better to consider the IPA (and other
-transcription systems) to be a special kind of script within the Unicode
-Standard, and treat the obvious similarity to Latin glyphs as a historical
-relic. All IPA glyphs would then have their own code points, instead of the
-current situation in which some IPA glyphs have special code points, while
-others are treated as being identical to the regular Latin characters. Yet, the
-current situation, however unfortunate, is unlikely to change, so as linguists
-we must learn to deal with the specific pitfalls of IPA within the Unicode
-Standard.
-
-In the following sections, we will describe these pitfalls 
-in detail. But first we present a brief history of IPA, its principles, 
-its computational representation and the need for and challenge in creating 
-a single multilingual computer encoding, and then how IPA meet Unicode.
-
 % ==========================
-\section{The International Phonetic Alphabet (IPA)}
+\chapter{The International Phonetic Alphabet (IPA)}
 \label{the-international-phonetic-alphabet}
@@
 
-\subsection{Brief history}
+\section{Brief history}
 
-Established in 1886, the International Phonetic Association (henceforth Association) 
-has long maintained a standard alphabet, the International Phonetic Alphabet (IPA),
+Established in 1886, the International Phonetic Association (henceforth
+Association) has long maintained a standard alphabet, the International Phonetic
+Alphabet (IPA), which is a common standard in linguistics to transcribe sounds
+of spoken languages. It was first published in 1888 as an international system
+of phonetic transcription for oral languages and for pedagogical purposes. It
+contained phonetic values for English, French and German. Diacritics for length
+and nasalization were already present in this version, and the same symbols are
+still used today. 
 %\footnote{Also referred to as API, for \textit{Association Phonétique Internationale}.}
-which is a common standard in linguistics to transcribe sounds of spoken
-languages. It was first published in 1888 as an international system of phonetic
-transcription for oral languages and for pedagogical purposes. The draft of 1887
-contained phonetic values for English, French and German. Diacritics for length
-and nasalization were already present in this draft and the same symbols are
-still used today.
 
@@
 from words in different languages. In 1900 the symbols were first organized into
-chart and were given phonetic feature labels for consonants (e.g.\ for manner of
-articulation: `plosives', `nasales', `fricatives'; for place of articulation:
-`bronchials', `laryngeales', `labiales') and for vowels (e.g.\ `fermeées',
-`mi-fermeées', `mi-overtes', `overtes'). Throughout the last centory, the
-structure of the chart has changed with increases in phonetic knowledge, and
-thus, like notational systems in other scientific disciplines, the IPA reflects
-facts and theories (of phonetic knowledge) that have developed over time. It is
-natural then that the IPA is modified occasionally to accommodate scientific
-innovations and discoveries. (In fact, this is part of the Association's
-mandate.) These changes are captured in the revisions to the IPA.
+chart and were given phonetic feature labels for consonants (for manner of
+articulation e.g.\ `plosives', `nasales', `fricatives'; for place of
+articulation, e.g.\ `bronchiales', `laryngales', `labiales') and for vowels
+(e.g.\ `fermées', `mi-fermées', `mi-ouvertes', `ouvertes'). Throughout the last
+century, the structure of the chart has changed with increases in phonetic
+knowledge, and thus, like notational systems in other scientific disciplines,
+the IPA reflects facts and theories of phonetic knowledge that have developed
+over time. It is natural then that the IPA is modified occasionally to
+accommodate scientific innovations and discoveries. In fact, this is part of
+the Association's mandate. These changes are captured in the revisions to the
+IPA.
 %\footnote{For a detailed history see:
@@
 
-Phonetic analysis is based on two premises: that it is possible to describe 
+Phonetic analysis is based on two premises: (i) that it is possible to describe
 the acoustic speech signal (i.e.\ sound waves) in terms of sequentially ordered
-discrete segments, and, that each segment can be characterized by an articulatory 
-target.\footnote{A purely phonetic description is only derivable from instrumental 
-data in high quality sound recordings. Once spoken language data are segmented, 
-phonological consideration inextricably play a roll in transcription. In other 
-words, phonetic observations beyond quantitative acoustic analysis are made in 
-terms of some phonological framework.} Today, the IPA chart reflects a linguistic 
-basis grounded in principles of phonological contrast. This fact is stated 
-explicitly in several places, including in the \textit{Report on the 1989 Kiel 
-convention} published in the \textit{Journal of the International Phonetic 
-Association} \citep[67-68]{International1989report}:
+discrete segments, and, (ii) that each segment can be characterized by an
+articulatory target.\footnote{A purely phonetic description is only derivable
+from instrumental data in high quality sound recordings. Once spoken language
+data are segmented, phonological consideration inextricably play a roll in
+transcription. In other words, phonetic observations beyond quantitative
+acoustic analysis are always made in terms of some phonological framework.}
+Today, the IPA chart reflects a linguistic theory grounded in principles of
+phonological contrast and in knowledge about the attested linguistic variation.
+This fact is stated explicitly in several places, including in the
+\textit{Report on the 1989 Kiel convention} published in the \textit{Journal of
+the International Phonetic Association} \citep[67-68]{International1989report}:
 
@@
 % The general value of the symbols in the chart is listed below. In each case 
-... a symbol can be regarded as a shorthand equivalent to a phonetic description, 
-and a way of representing the contrasting sounds that occur in a language. Thus 
-[m] is equivalent to `voiced bilabial nasal', and is also a way of representing 
-one of the contrasting nasal sounds that occur in English and other languages. 
-[...] When a symbol is said to be suitable for the representation of sounds in 
-two languages, it does not necessarily mean that the sounds in the two languages 
-are identical.
-\end{quote}
+[...] a symbol can be regarded as a shorthand equivalent to a phonetic
+description, and a way of representing the contrasting sounds that occur in a
+language. Thus [m] is equivalent to `voiced bilabial nasal', and is also a way
+of representing one of the contrasting nasal sounds that occur in English and
+other languages. [...] When a symbol is said to be suitable for the
+representation of sounds in two languages, it does not necessarily mean that the
+sounds in the two languages are identical. \end{quote}
 
 \noindent Although the IPA provides symbols to unambiguously represent phonemes,
-it also aims to represent phonetic details. Since phonetic detail could potentially 
-include things like `deep voice', the IPA restricts phonetic detail to linguistically 
-relevant aspects of speech. These principles and conventions for using the IPA are 
-outlined in the Association's Handbook.
+it also aims to represent phonetic details. Since phonetic detail could
+potentially include anything, e.g. something like `deep voice', the IPA
+restricts phonetic detail to linguistically relevant aspects of speech. These
+principles and conventions for using the IPA are outlined in the Association's
+Handbook.
 
-\subsection{Principles}
+\section{Principles}
 
@@
 
-\subsection{Computational representation}
+\section{Computational representation}
 
@@
 
-\subsection{SAMPA and X-SAMPA}
+\section{SAMPA and X-SAMPA}
 
@@
 
-\subsection{The need for a single multilingual environment}
+\section{The need for a single multilingual environment}
 
@@
 
-\subsection{Unicode and ISO 10646}
+\section{Unicode and ISO 10646}
 
@@
 
-\subsection{IPA and Unicode}
+\section{IPA and Unicode}
 
@@
 % http://www-01.sil.org/computing/computing_environment.html
-
-% ==========================
-\section{Pitfall: No complete IPA code block}
-\label{pitfall-no-complete-ipa-block}
-% ==========================
-
-The ambivalent nature of IPA glyphs arises because, on the one hand, the IPA
-uses Latin-based glyphs like <a>, <b> or <p>. From this perspective, the IPA
-seems to be just another orthographic tradition using Latin characters, all of
-which do not get a special treatment within the Unicode Standard (just like
-e.g.~the French, German, or Danish orthographic traditions do not have a special
-status). On the other hand, the IPA uses many special symbols (like turned <ɐ>,
-mirrored <ɘ> and/or extended <ɧ> Latin glyphs) not found in any other Latin-based
-writing system. For this reason a special block with code points, called
-\textsc{IPA Extensions} was already included in the first version of the Unicode
-Standard (Version 1.0 from 1991).
-
-As explained in Section~\ref{pitfall-blocks}, the Unicode Standard code space is
-subdivided into character blocks, which generally encode characters from a
-single script. However, as is illustrated by the IPA, characters that form a
-single writing system may be dispersed across several different character
-blocks. With its diverse collection of symbols from various scripts and
-diacritics, the IPA is spread across 12 blocks in the Unicode
-Standard:\footnote{This number of blocks depends on whether only IPA-sanctioned
-symbols are counted or if the phonetic symbols commonly found in the literature
-are also included, see~\cite[Appendix~C]{Moran2012}. The 159 characters from 12 
-code blocks shown here are the characters proposed for ``strict'' IPA encoding, 
-as discussed in Section~\ref{ipa-recommendations}.}
-
-\begin{itemize}[itemsep=4pt]
-	\item \textsc{Basic Latin }(27 characters) \newline 
-	a b c d e f h i j k l m n o p q r s t u v w x y z~.~|
-	\item \textsc{Latin-1 Supplement} (4 characters) \newline 
-	æ ç ð ø
-	\item \textsc{Latin Extended-A} (3 characters) \newline 
-	ħ ŋ œ
-	\item \textsc{Latin Extended-B} (4 characters) \newline 
-	ǀ ǁ ǂ ǃ
-	\item \textsc{Latin Extended-C} (1 character): \newline
-	\charis{ⱱ}
-	\item \textsc{IPA Extensions} (67 characters) \newline 
-	ɐ ɑ ɒ ɓ ɔ ɕ ɖ ɗ ɘ ə ɛ ɜ ɞ ɟ ɠ ɡ ɢ ɣ ɤ ɥ ɦ ɧ ɨ ɪ ɬ ɭ ɮ ɯ ɰ ɱ ɲ ɳ ɴ \newline
-	ɵ ɶ ɸ ɹ ɺ ɻ ɽ ɾ ʀ ʁ ʂ ʃ ʄ ʈ ʉ ʊ ʋ ʌ ʍ ʎ ʏ ʐ ʑ ʒ ʔ ʕ ʘ ʙ ʛ ʜ ʝ ʟ ʡ ʢ 
-	\item \textsc{Greek and Coptic} (3 characters) \newline 
-	β θ χ
-	\item \textsc{Spacing Modifier Letters} (17 characters) \newline
-	\dia{02DE} \dia{02E1} \dia{02B7} \dia{02B2} \dia{02E0} \dia{02E4} 
-	\dia{02B0} \dia{02BC} \dia{02D0} \dia{02D1} ˥ ˦ ˧ ˨ ˩ {\large ˈ ˌ}
-	\item \textsc{Superscripts and Subscripts} (1 character) \newline
-	\dia{207F} 
-	\item \textsc{Combining Diacritical Marks} (25 characters) \newline 
-	\dia{033C} \dia{032A} \dia{033B} \dia{033A} \dia{031F} \dia{0320} \dia{031D} 
-	\dia{031E} \dia{0318} \dia{0319} \dia{031C} \dia{0339} \dia{032C} \dia{0325} 
-	\dia{0330} \dia{0324} \dia{0329} \dia{032F} \dia{0334} \dia{0303} \dia{0308} 
-	\dia{033D} \dia{0306} \dia{031A} \ \dia{0361}{\large\fontspec{CharisSIL}◌}
-    \item \textsc{General Punctuation} (2 characters) \newline 
-    ‖ \charis{‿}
-%	\item \textsc{Modifier Tone Letters} (2 characters) \newline
-%	{\large\fontspec{CharisSIL}ꜛ ꜜ}
-	\item \textsc{Arrows} (4 characters) \newline 
-	↑ ↓ ↗ ↘
-
-\end{itemize}
-
-% ==========================
-\section{Pitfall: IPA homoglyphs in Unicode}
-\label{pitfall-ipa-homoglyphs}
-% ==========================
-
-Another problem is the large number of homoglyphs, i.e.~different characters
-that have highly similar glyphs (or even completely identical glyphs, depending
-on the font rendering). For example, a user of a Cyrillic computer keyboard
-should ideally not use the <а> \textsc{cyrillic small letter a} at code point
-\uni{0430} for IPA transcriptions, but instead use the <a> \textsc{latin small
-letter a} at code point \uni{0061}, although visually they are mostly
-indistinguishable, and the Cyrillic character is more easily typed on a Cyrillic
-keyboard. Some further problematic homoglyphs related to encoding IPA in the
-Unicode Standard are the following.
-
-\begin{itemize}
-
-  \item The uses of the apostrophe has led to long discussions on the Unicode
-        Standard email list. An English keyboard inputs the symbol
-        <\dia{0027}> \textsc{apostrophe} at \uni{0027}, although the preferred Unicode
-        apostrophe is the <\dia{2019}> \textsc{right single quotation mark} at
-        \uni{2019}.\footnote{Note that many word processors (like Microsoft
-        Word) by default will replace straight quotes by curly quotes,
-        depending on the whitespace around it.} However, the glottal
-        stop/glottalization/ejective marker is yet another completely different
-        character, namely <\dia{02BC}>, i.e.~the \textsc{modifier letter apostrophe} 
-        at \uni{02BC}, which unfortunately looks mostly extremely similar to
-        \uni{2019}. 
-  \item Another problem is the <\dia{02C1}> \textsc{modifier letter reversed glottal
-        stop} at \uni{02C1} vs.\@ the <\dia{02E4}> \textsc{modifier letter small reversed
-        glottal stop} at \uni{02E4}. Both appear in various resources
-        representing phonetic data online. This is thus a clear example for
-        which the Unicode Standard does not solve the linguistic standardization
-        problem.
-  \item Linguists are also unlikely to distinguish between the <ə>
-        \textsc{latin small letter schwa} at code point \uni{0259} and <ǝ>
-        \textsc{latin small letter turned e} at \uni{01DD}.
-  \item The alveolar click <ǃ> at \uni{01C3} is of course often simply
-        typed as <!> \textsc{exclamation mark} at \uni{0021}.\footnote{In the
-        Unicode Standard the <ǃ> at \unif{01C3} is labeled \textsc{latin letter
-        retroflex click}, but in IPA that glyph is used for an alveolar or
-        postalveolar click (not retroflex). This naming is probably best seen as
-        an error in the Unicode Standard. For the ``real'' retroflex click, see 
-        Section~\ref{ipa-additions}.}
-  \item The dental click <ǀ>, in Unicode known as \textsc{latin letter dental
-        click} at \uni{01C0}, is often simply typed as <|> \textsc{vertical
-        line} at \uni{007C}.
-  \item For the marking of length there is a special Unicode characters, namely
-        <\dia{02D0}> \textsc{modifier letter triangular colon} at \uni{02D0}. However,
-        typing <\dia{003A}> \textsc{colon} at \uni{003A} is of course much easier.
-
-\end{itemize}
-
-Conversely, non-linguists are unlikely to distinguish any semantic difference
-between an open back unrounded vowel <ɑ> \textsc{latin small letter alpha} at
-\uni{0251}, and the open front unrounded vowel <a> \textsc{latin small letter a}
-at \uni{0061}, basically treating them as homoglyphs, although they are
-different phonetic symbols. But even among linguists this distinction leads to
-problems. For example, as pointed out by \citet{Mielke2009}, there is a problem
-stemming from the fact that about 75\% of languages are reported to have a
-five-vowel system \citep{Maddieson1984}. Historically, linguistic descriptions
-tend not to include precise audio recording and measurements of formants, so
-this may lead one to ask if the many <a> characters that are used in
-phonological description reflects a transcriptional bias. The common use of <a>
-in transcriptions could be in part due to the ease of typing the letter on an
-English keyboard (or for older descriptions, the typewriter). We found it to be
-exceedingly rare that a linguist uses <ɑ> for a low back unrounded
-vowel.\footnote{One example is \citet[75]{Vidal2001a}, in which the author
-states: ``The definition of Pilagá /a/ as [+back] results from its behavior in
-certain phonological contexts. For instance, uvular and pharyngeal consonants
-only occur around /a/ and /o/. Hence, the characterization of /a/ and /o/ as a
-natural class of (i.e., [+back] vowels), as opposed to /i/ and /e/.''} They
-simply use <a> as long as there is no opposition to <ɑ>.
-
-%\footnote{See Thomason's Language Log post, ``Why I don't love the International Phonetic Alphabet'' at:\url{http://itre.cis.upenn.edu/~myl/languagelog/archives/005287.html}.}
-
-Making things even more problematic, there is an old typographic tradition that
-the double-story <a> uses a single-story <ɑ> in italics. This leads to the
-unfortunate effect that even in many well-designed fonts the italics of <a> and
-<ɑ> use the same glyph. For example, in Linux Libertine (the font of this book)
-the italics of these characters are highly similar <\textit{a}> and
-<\textit{ɑ}>, while in Charis SIL they are identical: <\textit{\charis{a}}> and
-<\textit{\charis{ɑ}}>. If this distinction has to be kept upright in italics,
-the only solution we can currently offer is to use \textsc{slanted} glyphs
-(i.e.~artificially italicized glyphs) instead of real italics (i.e.~special
-italics glyphs designed by a typographer).\footnote{For example, the widely used
-IPA font Doulos SIL
-(\url{http://scripts.sil.org/cms/scripts/page.php?item\_id=DoulosSIL}) does not
-have real italics. This leads some word-processing software, like Microsoft
-Word, to produce slanted glyphs instead. That particular combination of font and
-software application will thus lead to the desired effect distinguishing <a>
-from <ɑ> in italics. However, note that when the text is transferred to another
-font (i.e.~one that includes real italics) and/or to another software
-application (like Apple Pages, which does not perform slanting), then this
-visual appearance will be lost. In this case we are thus still in the
-pre-Unicode situation in which the choice of font and rendering software
-actually matters. The ideal solution from a linguistic point of view would be
-the introduction of a new IPA code point for a different kind of <a>, which
-explicitly specifies that it should still be rendered as a double-story
-character when italicized. After informal discussion with various Unicode
-players, our impression is that this highly restricted problem is not
-sufficiently urgent to introduce even more <a> homographs in Unicode (which
-already lead to much confusion, see Section~\ref{pitfall-homoglyphs}).}
-
-% ==========================
-\section{Pitfall: Homoglyphs in IPA}
-\label{pitfall-homoglyphs-in-IPA}
-% ==========================
-
-Reversely, there are a few cases in which the IPA distinguishes different
-phonetic concepts, but the visual characters used by the IPA look very much
-alike. Such cases are thus homoglyphs in the IPA itself, which of course need
-different encodings.
-
-\begin{itemize}
-
-   \item The dental click <ǀ> and the indication of a minor group break <|>
-           look almost the same in
-           most fonts. For a proper encoding, the \textsc{latin letter dental
-           click} at \uni{01C0} and the \textsc{vertical line }at \uni{007C} 
-           should be used, respectively.
-   \item Similarly, the alveolar lateral click <ǁ>	should be encoded with a
-           \textsc{latin letter lateral click} at \uni{01C1}, different from <‖>, 
-           which according to the IPA is the character to by used for a major group 
-           break (by intonation), to be encoded by \textsc{double vertical line} 
-           at \uni{2016}.
-   \item The marking of primary stress < ˈ > looks like an apostrophe, and
-           is often typed with the same symbol as the ejective <\dia{02BC}>. For a
-           proper encoding, these two symbols should be typed as 
-           \textsc{modifier letter vertical line} at \uni{02C8} and
-           \textsc{modifier letter apostrophe} at \uni{02BC}, respectively. 
-   \item There are two different ``dashed''-l characters in IPA, namely the <ɫ> 	      \textsc{latin small letter l with middle tilde} at \uni{026B} and the <ɬ>	      \textsc{latin small letter l with belt} at \uni{026C}.
-These of course
-          look highly similar, although they are different sounds. As a solution, 
-          we will actually propose to not use the middle tilde at all 
-          (see Section~\ref{pitfall-multiple-options-ipa}).
-
-\end{itemize}
-
-% ==========================
-\section{Pitfall: Multiple encoding options in IPA}
-\label{pitfall-multiple-options-ipa}     
-% ==========================
-
-It is not just the Unicode Standard that offers multiple options for encoding
-the IPA.\@ Even the IPA specification itself offers some flexibility in how
-transcriptions have to be encoded. There are a few cases in which the IPA
-explicitly allows for different options of transcribing the same phonetic
-content. This is understandable from a transcriber's point of view, but it is
-not acceptable when the goal is interoperability between resources written in
-IPA.\@ We consider it crucial to distinguish between ``lax'' IPA, for which it
-is sufficient that any phonetically-trained reader is able to understand the
-transcription, and ``strict'' IPA, which should be standardized on a single
-unique encoding for each sound, so search will work across resources. We are
-aware of the following non-unique encoding options in the IPA, which will be
-discussed in turn below:
-
-\begin{itemize}
-  \item The marking of tone
-  \item The marking of <g>
-  \item The marking of velarization and pharyngealization
-  \item The placement of diacritics
-\end{itemize}
-
-The first case in which the IPA allows for different encodings is the question
-of how to transcribe tone. There is an old tradition to use diacritics on vowels
-to mark different tone levels, e.g. <ȅèée̋>.\footnote{To make things even more
-complex, there are at least two different Unicode homoglyphs for the low and
-high level tones, namely <\diaf{0340}> \textsc{combining grave tone mark} at
-\unif{0340} vs.~<\diaf{0300}> \textsc{combining grave accent} at \unif{0300} for
-low tone, and <\diaf{0341}> \textsc{combining acute tone mark} at \unif{0341}
-vs.<\diaf{0301}> \textsc{combining acute accent} at \unif{0301} for high tone.}
-The IPA also proposes the option of tone letters, e.g. <˥˦˧˨˩>, which are much
-less often used, but are more consistent for contours. Tone letters in the IPA
-have five different levels, and sequences of these letters can be used to
-indicate contours. Well-designed fonts will even merge a sequence of tone
-letters into a contour. For example, compare the font Linux Libertine, which
-does not merge tone letters <{\fontspec{LinLibertineO}˥˨˧˩}>, with the font
-CharisSIL, which merges this sequence of four tone letters into a single contour
-<\charis{˥˨˧˩}>. For strict IPA encoding we propose to standardize on tone
-letters.
-
-% IPA1999 pg 19
-% "Either of the variant letter shapes [g] and [g] may be used to represent the voiced velar plosive."
-
-Second, we commonly encounter the use of <g> \textsc{latin small letter g} at
-\uni{0067}, instead of the Unicode Standard IPA character for the voiced velar
-stop <ɡ> \textsc{latin small letter script g} at \uni{0261}. One begins to
-question whether this issue is at all apparent to the working linguist, or if
-they simply use the \uni{0067} because it is easily keyboarded and thus saves
-time, whereas the latter must be cumbersomely inserted as a special symbol in
-most software. The International Phonetic Association has taken the stance that
-both the keyboard \textsc{latin small letter g} and the \textsc{latin small
-letter script g} are valid input characters for the voiced velar plosive.
-Unfortunately, this decision further introduces ambiguity for linguists trying
-to adhere to a strict Unicode Standard IPA encoding. For strict IPA encoding we
-propose to standardize on the more idiosyncratic \textsc{latin small letter
-script g} at \uni{0261}.
-
-Third, the IPA has special markers for velarization <\dia{02E0}> and
-pharyngealization <\dia{02E4}>. Confusingly, there is also a marker for
-``velarized or pharyngealized'', using the <\dia{0334}> \textsc{combining tilde
-overlay} at \uni{0334}. The tilde overlay seems to be extremely rarely used. We 
-suggest to try and avoid using the tilde overlay, though for reasons of backward 
-compatibility we will allow it in valid-IPA.\@
-
-Finally, the IPA states that ``diacritics may be placed above a symbol with a
-descender''. For example, for marking marking of voiceless pronunciation of
-voiced segments the IPA uses the ring diacritic. Originally, the ring should be
-placed below the base character, like in <m̥>, using the \textsc{combining ring
-below} at \uni{0325}. However, in letters with long descenders the IPA also
-allows to put the ring above the base, like in <ŋ̊>, using the \textsc{combining
-ring above} at \uni{030A}. Yet, proper font design does not have any problem
-with rendering the ring below the base character, like in <ŋ̥>, so for strict
-IPA encoding we propose to standardize on the ring below. As a principle, for
-strict IPA encoding only one options should be allowed for all diacritics.
-
-The variable encoding as allowed by the IPA becomes even more troublesome for
-the tilde and diaeresis diacritics. In these cases, the IPA itself attaches
-different semantics to the symbols above and below a base characters. The tilde
-above a character (like in <ã>, using the \textsc{combining tilde} at
-\uni{0303}) is used for nasalization, while the tilde below a character (like in
-<a̰>, using the \textsc{combining tilde below} at \uni{0330}) indicates creaky
-voice. Likewise, the diaeresis above (like in <ä>, using the \textsc{combining
-diaeresis} at \uni{0308}) is used for centralization, while the diaeresis below
-a character (like in <a̤>, using the \textsc{combining diaeresis below} at
-\uni{0324}) indicates breathy voice. These cases strengthen our plea to not
-allow diacritics to switch position for typographic convenience.
-
-% length IPA 1999:22
-% "Note: as in Finnish orthography, length can also be indicated in phonetic
-% transcription by double letters: e.g . Finnish maatto [rncctto] 'electrical
-% earth/ground'."
-
-% syllable breaks and word boundaries <\s, ., |, ||, tie-bar below>
-% White spaces can be used to indicate word boundaries. Syllable breaks can be
-% marked when required. The other two boundary symbols are used to mark the domain
-% of larger prosodic units. There is also a linking symbol that can be used for
-% explicitly indicating the lack of a boundary.
-
-% (In all these transcriptions, no spaces between words have been included. This
-% is inevitable in an impressionistic transcription where it is not yet known how
-% the utterance divides into words. In phonemic and allophonic transcriptions it
-% is common to include spaces to aid legibility, but their theoretical validity is
-% problematic.)
-
-% ==========================
-\section{Pitfall: Tie bar}
-\label{pitfall-tie-bar}
-% ==========================
-
-% Wells1995:10 "The underscore could in principle also be pressed into service
-% to represent the IPA tie bar. The current chart mentions its use only for
-% affricates and double articulations, and then only "if necessary"."
-
-In the major revision of the IPA in 1932, affricates were represented by two
-consonants <tʃ>, ligatures <ʧ>, or with the tie-bar <t͡ʃ>. In the
-1938 revision the tie-bar's semantics were broadened to indicate simultaneous
-articulation, as for example in labial velars such as <k͡p>. Thus, the tie-bar is a
-convenient diacritic for visually tokenizing input strings into chunks of
-phonetically salient groups, including affricates, doubly articulated consonants
-or diphthongs.
-
-The tie bar can be placed above or below the base characters, e.g.
-<\charis{t͡s}> or <\charis{t͜s}>. IPA allows both options. The choice between
-the two symbols is purely for legible rendering; there is no difference in
-semantics between the two symbols. However, rendering is such a problematic
-issue for tie bars in general that many linguists simply do not use them. Just
-looking at a few different fonts already indicates that actually no font
-designer really gets the placement right in combination with superscripts and
-subscripts. If really necessary, we propose to standardize on the tie bar above
-the base characters, using a \textsc{combining double inverted breve} at
-\uni{0361}.\footnote{Also note that the \textsc{undertie} at \unif{203F} looks
-like the tie bar below and is easily confused with it. However, it is a
-different character and has a different function in IPA. The undertie is used as
-a linking symbol to indicate the lack of a boundary, e.g.\ French \textit{petit
-ami} [pətit\charis{‿}ami] `boyfriend'.}
-
-\begin{itemize}[itemsep=6pt]
-  \item[] {\fontspec{Times New Roman}Times new Roman: t̥ʰ͡s t̥ʰ͜s}
-  \item[] {\small \fontspec{CharisSIL}CharisSIL:\@ t̥ʰ͡s t̥ʰ͜s}
-  \item[] {\footnotesize \fontspec{Monaco}Monaco: t̥ʰ͡s t̥ʰ͜s}
-  \item[] {\fontspec{DoulosSIL}DoulosSIL:\@ t̥ʰ͡s t̥ʰ͜s}
-  \item[] Linux Libertine: t̥ʰ͡s t̥ʰ͜s
-\end{itemize}
-
-Tie bars are a special type of character in the sense that they do not belong to
-a segment, but bind two graphemes together. This actually turns out to be rather
-different from Unicode conceptions. The Unicode encoding of this character
-belong to the Combining Diacritical Marks, namely either \textsc{combining double
-inverted breve} at \uni{0361} or \textsc{combining double breve below} at
-\uni{035C}. Such a combining mark is by definition tied to the character in
-front, but not the character following it. The Unicode treatment of this
-character thus only partly corresponds to the IPA conception, which ideally
-would have the tie bar linked both to the character in front and to the
-character following.
-
-Further, according to the spirit of the IPA, it would also be possible to
-combine more than two base characters into one tie bar, but this is not possible
-with Unicode (i.e.~there is no possibility to draw a tie bar over three of four
-characters). It is possible to indicate such larger groups by repeating the tie
-bar, like for a triphthong <a͡ʊ͡ə> in the English word \textit{hour}. If really
-necessary, we consider this possible, even though the rendering will never look
-good.
-
-Most importantly though, in comparison to normal Unicode processing, the tie-bar
-actually takes a reversed approach to complex graphemes. Basically, the Unicode
-principle (see Section~\ref{pitfall-characters-are-not-graphemes}) is that fixed
-sequences in a writing system have to be specified as ``tailored'' grapheme
-clusters. In case the sequence is not a cluster, this has to be explicitly
-indicated. IPA takes a different approach. In IPA by default different base
-letters are not connected into larger clusters; only when it is specified in the
-string itself (using the tie bar).
-
-% ==========================
-\section{Pitfall: Ligatures and digraphs}
-\label{pitfall-ligatures-digraphs}     
-% ==========================
-
-% TODO: link back to discussion on IPA principles and how ligatures got brought
-% along historically
-
-One important distinction to acknowledge is the difference between multigraphs
-and ligatures. Multigraphs are groups of characters (in the context of IPA e.g.
-<tʃ> or <ou>) while ligatures are single characters (e.g. <ʧ> \textsc{latin
-small letter tesh digraph} at \uni{02A7}). Ligatures arose in the context of
-printing easier-to-read texts, and are included in the Unicode Standard for
-reasons of legacy encoding. However, their usage is discouraged by the Unicode
-core specification. Specifically related to IPA, various phonetic combinations
-of characters (typically affricates) are available as single code-points in the
-Unicode Standard, but are designated \textsc{digraphs}. Such glyphs might be used by
-software to produce a pleasing display, but they should not be hard-coded into
-the text itself. In the context of IPA, characters like the following ligatures
-should thus \emph{not} be used. Instead a combination of two characters is
-preferred:
-
-\begin{itemize} 
-	\item[] <ʣ> \textsc{latin small letter dz digraph} at \uni{02A3} 
-	  (use <dz>) 
-    \item[] <ʤ> \textsc{latin small letter dezh digraph} at \uni{02A4}
-      (use <dʒ>)
-    \item[] <ʥ> \textsc{latin small letter dz digraph with curl} at \uni{02A5}
-      (use <dʑ>)
-    \item[] <ʦ> \textsc{latin small letter ts digraph} at \uni{02A6} 
-      (use <ts>)
-	\item[] <ʧ> \textsc{latin small letter tesh digraph} at \uni{02A7} 
-	  (use <tʃ>) 
-    \item[] <ʨ> \textsc{latin small letter tc digraph with curl} at \uni{02A8}
-      (use <tɕ>)
-   	\item[] <ʩ> \textsc{latin small letter feng digraph} at \uni{02A9}
-	  (use <fŋ>) 
-\end{itemize}
-
-However, there are a few Unicode characters that are historically ligatures, but
-which are today considered as simple characters in the Unicode Standard and thus
-should be used when writing IPA, namely:
-
-\begin{itemize}
-	\item[] <ɮ> \textsc{latin small letter lezh} at \uni{026E} 
-	\item[] <œ> \textsc{latin small ligature oe} at \uni{0153} 
-	\item[] <ɶ> \textsc{latin letter small capital oe} at \uni{0276} 
-	\item[] <æ> \textsc{latin small letter ae} at \uni{00E6} 
-\end{itemize}
-
-% ==========================
-\section{Pitfall: Missing decomposition}
-\label{pitfall-missing-decomposition}
-% ==========================
-
-Although many combinations of base character with diacritic are treated as
-canonical equivalent with precomposed characters, there are a few combinations
-in IPA that allow for multiple, apparently identical, encodings that are not
-canonical equivalent (see Section~\ref{pitfall-canonical-equivalence}). The
-following elements should not be treated as diacritics when encoding IPA in
-Unicode: 
-\begin{itemize}
-  \item[] <{\fontspec{CharisSIL}{\large ◌}}\symbol{"0321}> \textsc{combining palatalized hook below} at \uni{0321}
-  \item[] <{\fontspec{CharisSIL}{\large ◌}}\symbol{"0322}> \textsc{combining retroflex hook below} at \uni{0322}
-  \item[] <\dia{0335}> \textsc{combining short stroke overlay} at \uni{0335}
-  \item[] <\dia{0337}> \textsc{combining short solidus overlay} at \uni{0337}
-\end{itemize}
-
-There turn out to be a lot of characters in the IPA that could be conceived as
-using any of these elements, like <ɲ>, <ɳ>, <ɨ> or <ø>. However, all such
-characters exist as well as precomposed combination in Unicode, and these
-precomposed characters should preferably be used. When instead combinations of a base character with
-diacritic are used, then these combinations are not canonical equivalent to the
-precomposed combinations. This means that any search will not find both at the
-same time.
-
-% \footnote{The IPA does not
-% describe any character for a voiced retroflex implosive, which would
-% transparently be \charis{ᶑ}. We propose to add this character to the IPA, see
-% Section~\ref{ipa-additions}.}
-
-A similar problem arises with the rhotic hook. There are two precomposed
-characters in Unicode with a rhotic hook, which are not canonical equivalent 
-with a combination of the vowel with a separately encoded hook:
-\begin{itemize}
-  \item[] <ɚ> \textsc{latin small letter schwa with hook} at \uni{025A}
-  \item[] <ɝ> \textsc{latin small letter reversed open e with hook} at \uni{025D}
-\end{itemize}
-All other combinations of vowels with rhotic hooks will have to be made by using
-<\dia{02DE}> \textsc{modifier letter rhotic hook} at \uni{02DE}, because there
-is no complete set of precomposed characters with rhotic hooks in Unicode. For
-that reason we propose to not use the two precomposed characters with hooks
-mentioned above, but always use the separate rhotic hook at \uni{02DE} in IPA.\@
-
-A similar situation arises with <\dia{0334}> \textsc{combining tilde overlay} at
-\uni{0334}. The main reason some phoneticians like to use this in IPA is to mark
-the ``dark'' <l> in English codas, using the character <ɫ> \textsc{latin small
-letter l with middle tilde} at \uni{026B}. This character is not canonically
-equivalent to the combination <l>~+~<\dia{0334}>, so one of the two possible
-encodings has to be chosen. Because the tilde overlay is described as a general
-mechanism by the IPA, we propose to use the separated <\dia{0334}>
-\textsc{combining tilde overlay} at \uni{0334}. However, note that phonetically 
-this seems to be (almost) superfluous (see Section~\ref{pitfall-multiple-options-ipa}) 
-and the typical usage in the form of <ɫ> is (almost) a homoglyph with <ɬ> (see 
-Section~\ref{pitfall-homoglyphs-in-IPA}). For these reasons we also suggest to try 
-and avoid the tilde overlay completely.
-
-Reversely, note that the <ç> \textsc{latin small letter c with cedilla} at
-\uni{00E7} is canonically equivalent with <c> with <\dia{0327}>
-\textsc{combining cedilla} at \uni{0327}, so it will be separated into two
-characters by Unicode canonical decomposition, also if such a decomposition is
-not intended in the IPA.\@ However, because of the nature of canonical
-equivalence (see Section~\ref{pitfall-canonical-equivalence}), these two
-encodings are completely identical in any computational treatment, so this
-decomposition does not have any practical consequences.
-
-% ==========================
-\section{Pitfall: Different notions of diacritics}
-\label{pitfall-different-notions-of-diacritics}
-% ==========================
-
-% TODO: this section is wrong wrt IPA diacritics and needs updating @SM
-
-Another pitfall relates to the question of what are diacritics. The problem is that
-the meaning of the term diacritics as used by the IPA is not the same as is used
-in the Unicode Standard. Specifically, diacritics in the IPA-sense are either
-so-called \textsc{combining diacritical marks} or \textsc{spacing modifier
-letters} in the Unicode Standard. Crucially, Combining Diacritical Marks are by
-definition combined with the character before them (to form so-called default
-grapheme clusters, see Section~\ref{the-unicode-approach}). In contrast, Spacing
-Modifier Letters are by definition \emph{not} combined into grapheme clusters
-with the preceding character, but simply treated as separate letters. In the
-context of the IPA, the following IPA-diacritics are actually Spacing Modifier
-Letters in the Unicode Standard:
-
-\begin{itemize}
-
-	\item[] Length marks, namely: 
-	\begin{itemize}
-	  \item[] <\dia{02D0}> \textsc{modifier letter triangular colon} at \uni{02D0}
-	  \item[] <\dia{02D1}> \textsc{modifier letter half triangular colon} at \uni{02D1}
-	\end{itemize}
-
-	\item[] Tone letters, like: 
-	\begin{itemize} 
-	  \item[] <˥> \textsc{modifier letter extra-high tone bar} at \uni{02E5}
-	  \item[] <˨> \textsc{modifier letter low tone bar} at \uni{02E8}
-	  \item[] and others like this
-	\end{itemize}
-
-	\item[] Superscript letters, like:
-	\begin{itemize}
-	  \item[] <\dia{02B0}> \textsc{modifier letter small h} at \uni{02B0}
-	  \item[] <\dia{02E4}> \textsc{modifier letter small reversed glottal stop} at \uni{02E4}
-	  \item[] <\dia{207F}> \textsc{superscript latin small letter n} at \uni{207F}
-	  \item[] and many more like this
-	\end{itemize}
-
-	\item[] The rhotic hook:\footnote{It is really unfortunate that the rhotic hook
-         in Unicode is classified as a Spacing Modifier, and not as a Combining 
-         Diacritical Mark. Although the rhotic hook is placed to the right of its 
-         base character (and not above or below), it still is always connected 
-         to the character in front, even physically connected to it. We cannot 
-         find any reason for this treatment, and consider it an error in 
-         Unicode. We hope it will be possible to change this classification in 
-         the future.}
-	\begin{itemize}
-	  \item[] <\dia{02DE}> \textsc{modifier letter rhotic hook} at \uni{02DE}
-	\end{itemize}
-
-\end{itemize}
-
-Although linguists might expect these characters to belong together with the
-character in front of them, at least for <ʰ> \textsc{modifier letter small h} at
-\uni{02B0} the Unicode Consortium's decision to treat it as a separate character
-is also linguistically correct, because according to the IPA it can be used both
-for post-aspiration (following the base character) and pre-aspiration (preceding
-the base character). The default combination of Spacing Modifiers with the
-preceding character can be specified in orthography profiles (see
-Chapter~\ref{orthography-profiles}).
-
-% ==========================
-\section{Pitfall: No unique diacritic ordering}
-\label{pitfall-no-unique-diacritic-ordering}
-% ==========================
-
-Also related to diacritics is the question of ordering. To our knowledge, the
-International Phonetic Association does not specify a specific ordering for
-diacritics that combine with phonetic base symbols; this exercise is left to the
-reasoning of the transcriber. However, such marks have to be explicitly ordered
-if sequences of them are to be interoperable and compatible computationally. An example is a
-labialized aspirated alveolar plosive: <tʷʰ>. There is nothing holding linguists
-back from using <tʰʷ> instead (with exactly the same intended meaning). However,
-from a technical standpoint, these two sequences are different, e.g.~if both
-sequences are used in a document, searching for <tʷʰ> will not find any
-instances of <tʰʷ>, and vice versa. Likewise, a creaky voiced syllabic dental
-nasal can be encoded in various orders, e.g. <n̪̰̩>, <n̩̰̪> or <n̩̪̰>.
-
-\subsubsection*{Canonical combining classes}
-
-In accordance with the absence of any specification of ordering in the IPA, the
-Unicode Standard likewise does not propose any standardized orders. Both leave it
-to the user to be consistent; this approach naturally invites inconsistency across 
-different authored resources.
-
-There is one (minor) aspect of ordering for which the Unicode Standard does
-present a canonical solution. Fortunately, this is uncontroversial from a
-linguistic perspective. Diacritics in the Unicode Standard (i.e.~Combining
-Diacritical Marks, see Section~\ref{pitfall-different-notions-of-diacritics})
-are classified in so-called \textsc{canonical combining classes}. In practice,
-the diacritics are distinguished by their position relative to the base
-character.\footnote{See
-\url{http://unicode.org/reports/tr44/\#Canonical\_Combining\_Class\_Values} for
-a detailed description.} When applying a Unicode normalization (NFC or NFD, see
-Section~\ref{pitfall-canonical-equivalence}), the diacritics in different
-positions are put in a specified order. This process therefore harmonizes the
-difference between different encodings in some situations, for example in the
-case of an extra-short creaky voice vowel <ḛ̆>. This grapheme cluster can be
-encoded either as <e>+<\dia{0306}>+<\dia{0330}> or as
-<e>+<\dia{0330}>+<\dia{0306}>. To prevent this twofold encoding, the Unicode
-Standard specifies the second ordering as canonical (namely, diacritics
-below are put before diacritics above).
-
-% The next paragraph does not belong here - we should put it somewhere else
-
-When encoding a string according to the Unicode Standard, it is possible to do
-this either using the NFC (composition) or NFD (decomposition) normalization (see
-Section~\ref{pitfall-canonical-equivalence}).
-Decomposition implies that precomposed characters (like <á> \textsc{latin small
-letter a with acute} at \uni{00E1}) will be split into its parts. This might
-sound preferable for a linguistic analysis, as the different diacritics are
-separated from the base characters. However, note that most attached elements
-like strokes (e.g.~in the <ɨ>), retroflex hooks (e.g.~in <ʐ>) or rhotic hooks
-(e.g.~in <ɝ>) will not be decomposed, but strangely enough a cedilla (like in
-<ç>) will be decomposed (see Section~\ref{pitfall-missing-decomposition}). In
-general, Unicode decomposition does not behave like a feature decomposition as
-expected from a linguistic perspective. It is thus important to consider Unicode
-decomposition only as a technical procedure, and not assume that it is
-linguistically sensible.
-
-\subsubsection*{Proposal for diacritic ordering}
-
-Facing the problem of specifying a consistent ordering of diacritics while
-developing a large database of phonological inventories from the world's
-languages, \citet[540]{Moran2012} defined a set of diacritic ordering
-conventions. The conventions are influenced by the linguistic literature, though
-some ad-hoc decisions had to be taken given the vast variability of phonological
-segments described by linguists.
-
-By Unicode Canonical Combining Classes, the diacritics on top of a character,
-like <\dia{0334}> (Combining Class number 1), always come before diacritics
-below (Combining Class number 220), which in turn always come before diacritics
-above (Combining Class number 230), which in turn come before diacritics over
-multiple characters like the tie bar <\dia{0361}{\large\fontspec{CharisSIL}◌}>
-(Combining Class number 233). We follow this order, but add the other IPA
-diacritics (which are not diacritics in the Unicode sense) between diacritics
-below and the tie bar. Further, within all these classes of diacritics there is
-no canonical ordering specified by Unicode, so we propose an explicit ordering
-here.
-
-Starting with the diacritics below: if a character sequence contains more than
-one diacritic below the base character, then the place features are applied
-first (linguolabial, dental, apical, laminal, advanced, retracted), followed by
-the manner features (raised, lowered, advanced and retracted tongue root), then
-secondary articulations (more round, less round), laryngeal settings (creaky,
-breathy, voiced, devoiced), and finally the syllabic or non-syllabic marker. So,
-the order that is proposed is the following, where <\textbar{}> indicates
-\textit{or} and <→> indicates \textit{precedes}. Note that the groups of
-alternatives (as marked by <\textbar{}>) are supposed never to occur together
-with the same base character. In effect, this represents yet another restriction
-on possible diacritic sequences.
-
-\begin{itemize}
-	\item[] \textsc{Combining Diacritical Marks (below) ordering:}
-	\begin{itemize}	
-	  \item[→] linguolabial <\dia{033C}> \textbar{} dental <\dia{032A}> \textbar{} apical <\dia{033A}> \textbar{} laminal <\dia{033B}>
-	  \item[→] advanced <\dia{031F}> \textbar{} retracted <\dia{0320}> 
-	  \item[→] raised <\dia{031D}> \textbar{} lowered <\dia{031E}>
-	  \item[→] advanced tongue root <\dia{0318}> \textbar{} retracted tongue root <\dia{0319}>
-	  \item[→] more rounded <\dia{0339}> \textbar{} less rounded <\dia{031C}>
-	  \item[→] creaky voiced <\dia{0330}> \textbar{} breathy voiced <\dia{0324}> \textbar{} voiced <\dia{032C}> \textbar{} voiceless <\dia{0325}>
-	  \item[→] syllabic <\dia{0329}> \textbar{} non-syllabic <\dia{032F}>
-	\end{itemize}
- \end{itemize}
-
-\noindent Next, if a character sequence contains more than one diacritic above the base
-character, we propose the following order:
-
-\begin{itemize}
-	\item[] \textsc{Combining Diacritical Marks (above) ordering:}
-	\begin{itemize}
-	  \item[→] nasalized <\dia{0303}>
-	  \item[→] centralized <\dia{0308}> \textbar{} mid-centralized <\dia{033D}>
-	  \item[→] extra short <\dia{0306}>
-	  \item[→] no audible release <\dia{031A}\ >
- \end{itemize} \end{itemize}
-
-\noindent Then, when a character sequence contains more than one character of the Spacing
-Modifier Letters, these will be placed after all combining diacritic marks in the
-following order:
-
-\begin{itemize}
-	\item[] \textsc{Spacing Modifier Letters ordering:}
-	\begin{itemize}
-	  \item[→] rhotic hook <\dia{02DE}>
-	  \item[→] lateral release <\dia{02E1}> \textbar{} nasal release <\dia{207F}>
-	  \item[→] labialized <\dia{02B7}>
-	  \item[→] palatalized <\dia{02B2}>
-	  \item[→] velarized <\dia{02E0}>
-	  \item[→] pharyngealized <\dia{02E4}>
-	  \item[→] aspirated <\dia{02B0}> \textbar{} ejective <\dia{02BC}>
-	  \item[→] long <\dia{02D0}> \textbar{} half-long <\dia{02D1}>
-	\end{itemize}
-\end{itemize}
-
-\noindent Finally, the tie bar follows at the very end of any such sequence:
-
-\begin{itemize}
-  \item[] \textsc{Tie bar:}
-  \begin{itemize}
-%    \item[→] tone letters <˥ ˦ ˧ ˨ ˩>
-    \item[→] tie bar <\dia{0361}{\large\fontspec{CharisSIL}◌}>
-  \end{itemize}
-\end{itemize}
-
-% ==========================
-\section{Pitfall: Future revisions to the IPA}
-\label{ipa-revisions}
-% ==========================
-
-With each revision of the IPA, many decisions need to be made by 
-the Association as to which symbols should be added, removed or 
-changed. For example, in the 1989 revision of the IPA at the Kiel Convention, 
-changes to specific symbols (in previous charts) were debated and 
-the Association's members certain decisions. The prevailing mood at 
-the convention was not to change specific symbols unless a strong 
-case was made \cite{Ladefoged1990a}. For example, two such decisions 
-included:
-
-\begin{itemize}
-	\item symbols for clicks were changed from <ʇ~ʖ~ʗ> to <ǀ~ǁ~ǃ>
-       because the latter were the symbols used by nearly all Khoisanists and
-       Bantuists.
-	\item the Americanist tradition of using using <\dia{030C}>, a
-       \textsc{combining caron} at \uni{030C} for all post-alveolar sounds, like
-       in <š~ž~č~ǰ>, was not adopted because the Association members at the
-       convention ``were not sufficiently impressed by arguments ... to the
-       effect that these sounds formed a natural class, and thus is would be
-       appropriate to recognize this by maintaining a common aspect to their
-       symbolism'' \citep[62]{Ladefoged1990a}. 
-\end{itemize}
-
-\noindent These decisions have practical consequences for transcribers 
-of IPA, particularly those who wish to follow recommended practices of 
-encoding electronic text in the Unicode Standard. For example, the Unicode 
-Standard contains removed symbols as labels them as clicks, e.g. <ʇ> 
-\textsc{latin small letter turned t} at U+0287 in the \textsc{IPA Extensions} 
-block has the comment `dental click (sound of ``tsk tsk'')'. In this case, 
-the IPA transcriber must know the status of click symbols in the current 
-version of the IPA and then identify those characters within the Unicode Standard.
-
-The most controversial issue regarding symbols debated at the convention 
-was the representation for voiceless implosives \cite[62]{Ladefoged1990a}. 
-Recall the principles of the IPA outlined in Section \ref{the-international-phonetic-alphabet}, 
-such as using distinct symbols for phonological contrast versus convenience of 
-display in the chart, must be taken into account when arguing for or against 
-the inclusion or deletion of IPA symbols in the IPA chart. Furthermore, the 
-inclusion of deletion of symbols should incorporate the phonetic knowledge 
-of the world's languages.
-
-\cite{Ladefoged1990a} argued against the inclusion of these symbols < ƥ, ƭ, ƈ, ƙ, ʠ > 
-voiceless implosives, noting that: they are not contrastive (e.g.\ in Mayan 
-languages); there is no instrumental evidence supporting voiceless implosives 
-in Africa; the sounds are sufficiently rare as to not need a whole new row 
-of symbols in the chart. Ladefoged favored symbolizing the sounds using a 
-voiceless diacritic ring below voiced implosives, e.g.\ <ɓ̥>.
-
-Nevertheless, in the 1989 IPA chart there is indeed a row for implosives 
-containing voiceless and voiced pairs.\footnote{https://en.wikipedia.org/wiki/File:IPA\_as\_of\_1989.png} 
-But already in the next revision, in 1993 (with an update in 1996), the voiceless 
-implosives were dropped. The implosives row from the IPA consonantal chart 
-disappeared and voiced implosives were given a column in the non-pulmonic 
-consonants table (which is still reflected in the latest revision to date, IPA 2005).
-
-The Journal of the International Phonetic Association follows its own 
-published standard for the IPA at the time of publication, even when it 
-may conflict with the Association's principle of using different symbols 
-for contrastive sounds and diacritics for phonetic variation. For example, 
-in the case of voiceless implosives, \cite{McLaughlin2005} shows that 
-Seereer-Siin (Niger-Congo; Atlantic; ISO 639-3: srr) has a phonologically 
-contrastive set of voiced and voiceless implosive stops at the labial, 
-coronal and palatal places of articulation. These symbols are transcribed 
-in an \textit{Illustrations of the IPA} article in the IPA journal as 
-< ɓ̥, ɗ̥, ʄ̥ >.\footnote{Note that the IPA chart states that diacritics 
-can also be placed above a symbol, e.g.\ reportedly for legibility's 
-sake < ɠ̥> vs < ɠ̊ >. This of course leads to the variability in encoding 
-that we are trying to address on a computational level in this work. 
-See Section \ref{}.}
-
-The point of this pitfall is to highlight that revisions to the IPA will 
-continue into the future, albeit they are infrequent. Nevertheless, 
-given the Unicode Standard's principle of maintaining backwards compatibility 
-(at all costs), transcribers and consumers of IPA cannot rely solely on 
-remarks in the Unicode Standard to reflect current standard IPA usage. 
-There is the possibility that at a later revision of the IPA, symbols that 
-are not currently encoded in the Unicode Standard are added to the IPA -- 
-although we think this is unlikely.
-
-\begin{comment}
-\subsection{IPA (practical) extensions}
-
-The IPA extensions cover speech sounds beyond the sound systems of the world's languages, including symbols for pathological speech and paralinguistic functions (e.g.\ Braille?).
-
-- pathological speech
-- paralinguistic functions (Braille?)
-
-As noted in Section \ref{}, extensions to the IPA -- in particular ``large-scale'' extensions --
-
-% Englebretson2009.pdf
-http://www.ruf.rice.edu/~reng/englebretson2009.pdf
-
-X-SAMPA was an ASCII transliteration and is still purposeful for several programs
-
-There is also the extensions of the IPA designed for disordered speech, although the symbols are occasionally used for transcribing phonetic detail in ``normal'' speech or also cross-linguistic rarities.\footnote{\url{http://phoible.github.io/conventions/}}
-
-% Unicode has its own history; for a summary: \url{http://www.unicode.org/history/summary.html}
-
-https://www.phon.ucl.ac.uk/home/wells/ipa-unicode.htm
-
-Computer codes for phonetic symbols
-EslingGaylord1993
-
-\end{comment}
-
-% ==========================
-\section{Additions to the IPA}
-\label{ipa-additions}
-% ==========================
-
-In the course of collecting a large sample of phoneme systems across the world's
-languages \citep{Moran2012}, we found that in order to preserve distinctions
-both within and across language descriptions, additions to the approved IPA
-glyph set were needed. Wherever possible these additions were drawn from the
-extIPA symbols for disordered
-speech.\footnote{\textbf{Add better reference here: }\url{https://www.internationalphoneticassociation.org/sites/default/files/extIPAChart2008.pdf}}
-This section describes our proposed additions to the IPA glyph set. These
-additions are not part of the official IPA recommendations, so they should be 
-used with care.
-
-\begin{itemize}
-
-\item \textsc{Retroflex click} \newline
-      Retroflex clicks can be represented by <‼> \textsc{double exclamation
-      mark} at \uni{203C}. Note that the (post-)alveolar click <ǃ> at \uni{01C3}
-      is confusingly referred to as \textsc{latin letter retroflex click} in the
-      Unicode standard, which is probably best seen as an error.
-\item \textsc{Voiced retroflex implosive} \newline 
-      Although the IPA includes a
-      series of voiced implosives (marked with a hook on top, see
-      Section~\ref{pitfall-missing-decomposition}), there is no voiced retroflex
-      implosive. Following the spirit of the IPA, we propose to use <\charis{ᶑ}>
-      \textsc{latin small letter d with hook and tail} at \uni{1D91} for this
-      sound.
-\item \textsc{Fortis/lenis} \newline
-      Languages described as having a fortis/plain/lenis distinction that
-      corresponds poorly with the traditional
-      voiced/voiceless-unaspirated/voiceless-aspirated continuum can be marked
-      using the voiceless glyph for the plain phoneme, and then 
-      <\dia{0348}> \textsc{combining double vertical line below} at
-      \uni{0348} to mark the fortis articulation, and/or <\dia{0349}>
-      \textsc{combining left angle below} at \uni{0349} for the lenis
-      articulation.
-\item \textsc{Frictionalization} \newline 
-      The diacritic <\dia{0353}>
-      \textsc{combining x below} at \uni{0353} can be used to represent three
-      types of frictionalized sounds: First, click consonants where the release
-      of the anterior closure involves an ingressive ``sucking'' sound similar
-      to a fricative, for example <kǃ͓ʰ>; second, ``frictionalized'' vowels
-      (sounds that are phonologically vocalic, but with sufficiently close
-      closures to create buzzing); and third, fricative sounds at places of
-      articulation that do not have dedicated fricative glyphs, for example
-      sounds with voiceless velar lateral frication, like <ʟ̥͓>.
-\item \textsc{Derhoticization} \newline 
-      For derhoticization we propose to use
-      <\dia{032E}> \textsc{combining breve below} at \uni{032E}. 
-      \textbf{WHAT EXACTLY IS THIS?}
-\item \textsc{Coronal non-sibilant} \newline
-      Languages described as having a sibilant/non-sibilant distinction among
-      coronal fricatives and affricates can be handled using the subscript
-      <\dia{0347}> \textsc{combining equals sign below} at \uni{0347} to mark
-      the non-sibilant phoneme.
-\item \textsc{Glottalization} \newline 
-      Glottalized sounds can be indicated using
-      <\dia{02C0}> \textsc{modifier letter glottal stop} at \uni{02C0}, unless
-      it is clear that either ``ejective'' or ``creaky voicing'' are the
-      intended sounds (in which cases the standard IPA diacritics should be
-      used). Pre-glottalized sounds can be marked with
-      <\diareverse{02C0}> to the left of the base
-      glyph, for example <\charis{ˀt}>.
-\item \textsc{Voiced pre-aspiration} \newline Voiced sounds having
-      pre-aspiration can be marked with
-      <\diareverse{02B1}> \textsc{modifier letter
-      small h with hook} at \uni{02B1} to the left of the base glyph, for
-      example <\charis{ʱd}>.
-\item \textsc{Epilaryngeal phonation} \newline 
-      There are some rare articulations that make
-      use of an epilaryngeal phonation mechanism (e.g.,~the “sphincteric vowels”
-      of~!Xóõ). To represent these vowels, we propose to use the modifier <\dia{1D31}>
-      \textsc{modifier letter capital e} at \uni{1D31} to denote such sphincteric
-      phonation.
-
-\end{itemize}
-
-% ==========================
-\section{Recommendations}
-\label{ipa-recommendations}
-% ==========================
-
-% Great quote for standardization -- last sentence(s) in Ladefoged 1990:552
-
-Summarizing the pitfalls as discussed in this chapter, we propose to define 
-three different IPA encodings: strict-IPA, valid-IPA and widened-IPA.\@ These 
-three encodings are subsets of each other, i.e.~strict-IPA is more restricted 
-than valid-IPA, which it turn is more restricted than widened-IPA.\@ Informally 
-speaking, valid-IPA represents the current state of the IPA \citep{IPA2005}. 
-Strict-IPA represents a more constrained version of IPA, while widened-IPA is a 
-slightly extended version of IPA, allowing a few more symbols.
-
-\
-
-\noindent Strict-IPA encoding is supposed to be used when interoperability of
-phonetic resources is intended. It is a strongly constrained subset of IPA
-geared towards uniqueness of encoding. Ideally, for each transcription there
-should be exactly one possible strict-IPA encoding. For each phonetic feature
-there is only one possibility (see Section~\ref{pitfall-multiple-options-ipa})
-and the IPA diacritics are forced into a canonical ordering (see
-Section~\ref{pitfall-no-unique-diacritic-ordering}).
-
-Valid-IPA does allow alternative symbols with the same phonetic meaning, as 
-specified in the official IPA specifications. Also, valid-IPA does not enforce a 
-specific ordering of diacritics, because the IPA does not propose any such 
-ordering. This means that in valid-IPA the same phonetic intention can be 
-encoded in multiple ways. This is sufficient for phonetically trained human 
-eyes, but it is not sufficient for automatic interoperability.
-
-Finally, widened-IPA includes a few more symbols which seem to be useful for
-various special cases (see Section~\ref{ipa-additions}).
-
-\
-
-\noindent At the end of this chapter we have added a few longish tables summarizing all
-159 different unicode codepoints that form the basis of strict-IPA encoding
-(107 letters, 36 diacritics and 16 remaining symbols). Each of
-these tables shows a typical glyph, and then lists the Unicode Codepoint,
-Unicode Name and IPA description for each symbol. Further, there is a table with 
-the additional options for valid-IPA and a table with the additional options for 
-widened-IPA.\@
-
-textbf{NOTE: I HAVE ADDED VARIOUS TONE DIACRITICS TO VALID-IPA THAT ARE NOT IN
-THE OFFICIAL IPA SUMMARY. SHOULD THEY BE INCLUDED?}
-
-\begin{itemize}[itemsep=6pt]
-
-  \item \textsc{strict-IPA letters} \newline
-        The 107 different IPA letters as allowed in strict-IPA encoding are
-        listed in Table~\ref{tab:ipa_letters} starting on
-        page~\pageref{tab:ipa_letters}.
-  \item \textsc{strict-IPA diacritics} \newline The 36 different IPA diacritics and
-        tone markers (both Unicode Modifier Letters and Combining Diacritical
-        Marks) as allowed in strict-IPA encoding are listed in
-        Table~\ref{tab:ipa_diacritics} starting on
-        page~\pageref{tab:ipa_diacritics}.
-  \item \textsc{strict-IPA remainders} \newline The 16 remaining IPA symbols
-        (boundary, stress, tone letters and intonation markers) as allowed in strict-IPA
-        encoding are listed in Table~\ref{tab:ipa_leftovers} on
-        page~\pageref{tab:ipa_leftovers}.
-  \item \textsc{valid-IPA additions} \newline The 16 additional symbols as allowed in
-        valid-IPA encoding are listed in Table~\ref{tab:ipa_lax} on
-        page~\pageref{tab:ipa_lax}.     
-  \item \textsc{widened-IPA additions} \newline
-        The 10 proposed additions to the IPA are listed in
-        Table~\ref{tab:ipa_additions} on page~\pageref{tab:ipa_additions}.
-
-\end{itemize}
-
-\newpage
-\input{tables/unicode_letters}
-\input{tables/unicode_diacritics}
-\input{tables/unicode_remaining}
-\input{tables/unicode_valid}
-\input{tables/unicode_additions}

### 2015-11-13 book/chapters/ipa_meets_unicode.tex
minor readability corrections

--- 2015-08-01 chapters/ipa_meets_unicode.tex
@@
 
-\section{Introduction}
+\section{The twain shall meet}
 
@@
 alphabet. Although IPA is reasonably easily adhered to with pen and paper, it is
-not trivial to encode IPA characters electronically. Similar to the previous
-chapter, in this chapter we discuss various pitfalls with the encoding of IPA in
-the Unicode Standard. We will specifically refer to the 2005 version of the
-IPA \citep{IPA2005} and the 7.0 version of Unicode \citep{Unicode2014}.
-
-The details of the encoding are unimportant as long as the
-transcription is only directed towards phonetically trained eyes. For a linguist
+not trivial to encode IPA characters electronically. In this chapter we discuss
+various pitfalls with the encoding of IPA in the Unicode Standard. We will
+specifically refer to the 2005 version of the IPA \citep{IPA2005} and the 7.0
+version of Unicode \citep{Unicode2014}.
+
+As long as a transcription is only directed towards phonetically trained eyes,
+then all the details of the Unicode-encoding are unimportant. For a linguist
 reading an IPA transcription, many of the details that will be discussed in this
@@
 transcriptions are intended to be used across resources (e.g.~searching similar
-phenomena across different languages) then it becomes crucial that there are strict
-encoding guidelines. Our main goal in this chapter is to present the encoding
-issues and propose recommendations for a ``strict'' IPA encoding.
+phenomena across different languages) then it becomes crucial that there are
+strict encoding guidelines. Our main goal in this chapter is to present the
+encoding issues and propose recommendations for a ``strict'' IPA encoding for
+situations in which cross-resource consistency is crucial.
 
@@
 
-Many of the current problems derive from the fact that the
-IPA is clearly historically based on the Latin script, but different enough from
-most other Latin-based writing systems to warrant special attention. This
-ambivalent status of the IPA glyphs (partly Latin, partly special) is
-unfortunately also attested in the treatment of IPA in the Unicode Standard. In
-retrospect, it might have been better to consider the IPA (and other
-transcription systems) to be a special kind of script within the Unicode
-Standard, and treat the obvious similarity to Latin glyphs as a historical
-relic. All IPA glyphs would then have their own code points, instead of the
-current situation in which some IPA glyphs have special code points, while
-others are treated as being identical to the regular Latin characters. Yet, the
-current situation, however unfortunate, is unlikely to change, so as linguists
-we must learn to deal with the specific pitfalls of IPA within the Unicode
-Standard.
-
-In the following sections, we will describe these pitfalls 
-in detail. But first we present a brief history of IPA, its principles, 
-its computational representation and the need for and challenge in creating 
-a single multilingual computer encoding, and then how IPA meet Unicode.
-
-% ==========================
-\section{The International Phonetic Alphabet (IPA)}
-\label{the-international-phonetic-alphabet}
-% ==========================
-
-\subsection{Brief history}
-
-Established in 1886, the International Phonetic Association (henceforth Association) 
-has long maintained a standard alphabet, the International Phonetic Alphabet (IPA),
-%\footnote{Also referred to as API, for \textit{Association Phonétique Internationale}.} 
-which is a common standard in linguistics to transcribe sounds of spoken
-languages. It was first published in 1888 as an international system of phonetic
-transcription for oral languages and for pedagogical purposes. The draft of 1887
-contained phonetic values for English, French and German. Diacritics for length
-and nasalization were already present in this draft and the same symbols are
-still used today.
-
-Originally, the IPA alphabet was a list of symbols with pronunciation examples
-from words in different languages. In 1900 the symbols were first organized into
-chart and were given phonetic feature labels for consonants (e.g.\ for manner of
-articulation: `plosives', `nasales', `fricatives'; for place of articulation:
-`bronchials', `laryngeales', `labiales') and for vowels (e.g.\ `fermeées',
-`mi-fermeées', `mi-overtes', `overtes'). Throughout the last centory, the
-structure of the chart has changed with increases in phonetic knowledge, and
-thus, like notational systems in other scientific disciplines, the IPA reflects
-facts and theories (of phonetic knowledge) that have developed over time. It is
-natural then that the IPA is modified occasionally to accommodate scientific
-innovations and discoveries. (In fact, this is part of the Association's
-mandate.) These changes are captured in the revisions to the IPA.
-%\footnote{For a detailed history see:
-%\url{https://en.wikipedia.org/wiki/History\_of\_the\_International\_Phonetic_Alphabet}}
-
-Over the years there has been several revisions, many minor. For example,
-articulation labels -- what often are call features even though the IPA avoids
-this term -- have changed (e.g.\ terms like `lips', `throat' or `rolled' are not
-used anymore). Phonetic symbol values have changed (e.g.\ voiceless is no longer
-marked by <h>). Symbols have been dropped (e.g.\ the caret diacritic denoting
-`long and narrow' is no longer used). And many symbols have been added to reflect
-contrastive sounds found in the world's very diverse phonological systems.
-
-Although IPA began its life as a pedagogical tool, from its earliest days the 
-Association aimed to provide ``a separate sign for each distinctive sound; 
-that is, for each sound which, being used instead of another, in the same 
-language, can change the meaning of a word'' \citep[27]{IPA1999}. Distinctive 
-sounds became later known as \textsc{phonemes} and the IPA has developed historically 
-into a notational devise with a strictly segmented phonemic view. A phoneme is an 
-abstract theoretical notion derived from an acoustic signal as produced by speakers 
-in the real world. Therefore the IPA contains a number of theoretical assumptions 
-about speech and how to analyze speech in written form.
-
-Phonetic analysis is based on two premises: that it is possible to describe 
-the acoustic speech signal (i.e.\ sound waves) in terms of sequentially ordered 
-discrete segments, and, that each segment can be characterized by an articulatory 
-target.\footnote{A purely phonetic description is only derivable from instrumental 
-data in high quality sound recordings. Once spoken language data are segmented, 
-phonological consideration inextricably play a roll in transcription. In other 
-words, phonetic observations beyond quantitative acoustic analysis are made in 
-terms of some phonological framework.} Today, the IPA chart reflects a linguistic 
-basis grounded in principles of phonological contrast. This fact is stated 
-explicitly in several places, including in the \textit{Report on the 1989 Kiel 
-convention} published in the \textit{Journal of the International Phonetic 
-Association} \citep[67-68]{International1989report}:
-
-\begin{quote}
-The IPA is intended to be a set of symbols for representing all the possible 
-sounds of the world's languages. The representation of these sounds uses a set 
-of phonetic categories which describe how each sound is made. These categories 
-define a number of natural classes of sounds that operate in phonological rules 
-and historical sound changes. The symbols of the IPA are shorthand ways of 
-indicating certain intersections of these categories.
-\end{quote}
-
-\noindent and in the \textit{Handbook of the International Phonetic Association} \citep[18]{IPA1993}:
-
-\begin{quote}
-% The general value of the symbols in the chart is listed below. In each case 
-... a symbol can be regarded as a shorthand equivalent to a phonetic description, 
-and a way of representing the contrasting sounds that occur in a language. Thus 
-[m] is equivalent to `voiced bilabial nasal', and is also a way of representing 
-one of the contrasting nasal sounds that occur in English and other languages. 
-[...] When a symbol is said to be suitable for the representation of sounds in 
-two languages, it does not necessarily mean that the sounds in the two languages 
-are identical.
-\end{quote}
-
-\noindent Although the IPA provides symbols to unambiguously represent phonemes, 
-it also aims to represent phonetic details. Since phonetic detail could potentially 
-include things like `deep voice', the IPA restricts phonetic detail to linguistically 
-relevant aspects of speech. These principles and conventions for using the IPA are 
-outlined in the Association's Handbook.
-
-\subsection{Principles}
-
-IPA transcription has essentially two parts. The first is a text containing IPA 
-symbols and the second is a set of conventions (rules) for interpreting those 
-symbols (and their combinations). The IPA is designed to meet practical linguistic 
-needs and is used to transcribe the phonetic or phonological structure of languages. 
-It is also used increasingly as a foreign language learning tool, as a standard 
-pronunciation guide and as a tool for creating practical orthographies of 
-previously unwritten languages.
-
-The current construction and use of the IPA are guided by principles outlined 
-in the \textit{Handbook of the International Phonetic Association: A guide 
-to the use of the International Phonetic Alphabet} \citep[159]{IPA1999}, henceforth 
-HB. The use of the symbols to represent a language's phonological system is guided 
-by the principle of contrast; where two words are distinguishable by phonetic contrast, 
-those contrasts should be transcribed with different symbols (graphemes not diacritics). 
-Allophonic distinction falls under the rubric of diacritically-distinguished symbols, 
-e.g. [stop̚] vs [spʰot]. In sum:
-
-\begin{itemize}
-	\item different symbols (without diacritics) should be used whenever a language employs two contrastive sounds
-	\item when two sounds in a language are not known to be contrastive, the same symbol should be used to represent these sounds; however, diacritics may be used to distinguish such sounds when necessary
-	\item diacritics cannot be dispensed with entirely, so the Association recommends to limit their use to:
-	\begin{itemize}
-		\item denoting length, stress and pitch
-		\item representing minute shades of sounds
-		\item obviating the design of a (large) number of new symbols when a single diacritic suffices (e.g. nasalized vowels, aspirated stops)
-	\end{itemize}
-\end{itemize}
-
-\noindent Thus, an IPA transcription \textit{always} consists of ``a set of 
-symbols and a set of conventions for their interpretation''.
-
-In systematic transcription (as opposed to impressionistic), there is a division 
-between phonemic and allophonic transcription. The terms phoneme and allophone 
-contain theoretical baggage, but the basic goal of a phonemic transcription is to 
-distinguish all words in a language with the minimal number of transcription 
-symbols \citep[19]{Abercrombie1964}. Allophonic transcription uses a broader 
-set of distinct symbols to describe systematic allophonic differences in sounds 
-in words. These two systematic transcriptions are related to each other by a 
-set of conventions in the IPA tradition, so that they can be converted between 
-one and another.
-
-An IPA transcription is connected to a speech event by a set of conventions. 
-A phonetic (or impressionistic) transcription may use the conventions implicit 
-in the IPA chart, i.e. the transcriber can indicate that the phonetic value of 
-<ŋ͡m> is a simultaneous labial and velar closure which is voiced and contains 
-nasal airflow.
-
-A phonemic transcription includes the conventions of a particular language's 
-phonological rules. These rules determine the realization of that language's 
-phonemes. However, there can be different systems of phonemic transcription 
-for the same variety of a language. The differences may result from the fact 
-that more than one phonetic symbol may be appropriate for a phoneme (see Section 
-\ref{}). Or the differences may be due to different phonemic analyses, e.g.\ 
-Standard German's vowel system is arguably contrastive in length or tenseness.
-
-An important principle of the IPA is that different representations resulting 
-from different symbols and different analyses are in line with the IPA's aims. 
-In other words, the IPA does not provide phonological analyses for specific 
-languages and the IPA does not define a single ``correct'' transcription system. 
-Rather, the IPA aims to provide a resource that allows users to express any 
-analysis so that it is widely understood. Thus the IPA suits many linguists' 
-needs because:
-
-\begin{itemize}
-	\item it is intended to be a set of symbols for representing all possible sounds in the world's (spoken) languages
-	\item its chart has a linguistic basis (specifically a phonological bias) rather than a general phonetic notation scheme
-	\item its symbols can be used to represent distinctive feature combinations\footnote{Although the chart uses traditional manner and place of articulation labels, the symbols can nevertheless represent any defined bundle of features, binary or otherwise, to define phonetic dimensions.}
-	\item its chart provides a summary of linguists' agreed-upon phonetic knowledge (a common denominator of phonological ``facts'')
-\end{itemize}
-
-Several styles of transcription with IPA are possible and the HB illustrates 
-these and notes that they are all valid (see the 29 languages and their transcriptions 
-in the original and initial \textit{Illustrations of the IPA} \citep[41--154]{IPA2007}). 
-Therefore, there are different but equivalent transcriptions, or as \cite[64]{Ladefoged1990a} 
-captures it, ``Perhaps now that the Association has been explicit in its eclectic 
-approach, outsiders to the Association will no longer speak of \textit{the} IPA 
-transcription of a given phenomenon, as if there were only one approved style.''.
-
-Clearly not all phoneticians agree (or will likely ever agree of course) on all 
-aspects of the IPA or on transcription practices. As noted above, there have 
-been several revisions in the IPA's long history, but the current version (2005) 
-is strikingly similar to the 1926 version. In 1989 an IPA revision convention 
-was held in Kiel, Germany. As per other revisions, there was expansion and revisions 
-to phonetic symbols in the IPA chart. Notably the marking of tone was extended 
-with the addition of a second system for marking linguistic tones (Chao tones). 
-Importantly, however, this was the first revision to address issues of 
-computational representations for the IPA -- the principles above of which 
-have had several ramifications that make the interoperability of electronic 
-linguistic data extremely difficult.
-
-\subsection{Computational representation}
-
-Prior to the Kiel Convention for the modern revision of the IPA in 1989, 
-\cite{Wells1987} collected and published practical approaches to coded 
-representations of the IPA, which dealt mainly with the assignment of 
-characters on the keyboard. The process of assigning standardized ``computer 
-codes'' to phonetic symbols was assigned to the Workgroup on Computer 
-Coding (henceforth working group) at the Kiel Convention. This working 
-group was tasked with \citep{Esling1990,EslingGaylord1993}:
-
-\begin{itemize}
-	\item determining how to represent the IPA numerically
-	\item developing a set of numbers to refer to the IPA symbols unambiguously
-	\item providing each symbol a unique name (intended to provide a mnemonic description of that character's shape)
-\end{itemize}
-
-\noindent The identification of IPA symbols with unique identifiers was 
-a first step in formalizing the IPA computationally because it would give 
-each symbol an unambiguous numerical identifier called an \textsc{IPA Number}. 
-The numbering system was to be comprehensive enough to support future revisions 
-of the IPA, including symbol specifications and diacritic placement. The 
-application of diacritics was also to be made explicit.
-
-Although the Association had never officially approved a set of names 
-for the IPA symbols, each IPA symbol received a unique \textsc{IPA Name}. 
-Many symbols already had an informal name (or two) used by linguists, but 
-consensus on symbol names was growing due to the recent publication of the 
-\textit{Phonetic Symbol Guide} \citep{PullumLadusaw1986}. Thus most of the 
-IPA symbol names were taken from \cite{PullumLadusaw1986} \citep[31]{IPA1999}.
-
-The working group insightfully decided that the computing-coding convention 
-for the IPA should be independent of computer environments or formats (e.g.\ ASCII), 
-i.e.\ the IPA Number was not meant to be implemented directly in a computer 
-encoding. The working group report's declaration includes the explanatory 
-remarks \citep[82]{International1989report}:
-
-\begin{quote}
-The recommendation of a 7-bit ASCII or 8-bit extended-ASCII coding system 
-would be short-sighted in view of development towards 16-bit and 32-bit 
-processors. In fact, any specific recommendations would tie the Association 
-to a stage of technological development which is bound to be outdated long 
-before the next revision of the handbook.
-\end{quote}
-
-\noindent Thus the coding convention was not meant to address the engineering 
-aspects of the actual encoding in computers (cf.\ \cite{Anderson1984}). However, 
-it was meant to serve as a basis for a communication-interchange standard for 
-creating mapping tables from various computer encodings, fonts, phonetic-character-set 
-software, etc., to common IPA Numbers, and thus symbols.\footnote{Remember, at 
-this time in the late 1980s there was no stable multilingual computing environment. 
-But some solution was needed because scholars were increasingly using personal 
-computers for their research and many were quickly adopting electronic mail or 
-discussion boards like Usenet as a medium for international exchanges. This was 
-before the Internet as we know it today. Most of these systems ran on 8-bit 
-hardware systems using a 7-bit ASCII character encoding.}
-
-Furthermore, the assignment of computer codes to IPA symbols was meant to 
-represent an unbiased formulation. The Association plays the role of an international 
-advisory body and it stated that it should not recommend a particular existing 
-system of encoding. In fact, during this time there were a number of coding 
-systems used, but none of them had a dominant international position. The 
-differences between systems were also either too great or too subtle to warrant 
-an attempt at combining them \citep{International1989report}.
-
-The working group assigned each IPA symbol to a unique three-digit number, 
-i.e.\ an IPA Number. Encoded in this number scheme is information about the 
-status of each symbol (see below). The IPA numbers are listed with the IPA 
-symbols and they are also illustrated in IPA chart form (see \cite[84]{EslingGaylord1993} 
-or \cite[App. 2]{IPA2007}). The numbers were assigned in linear order (e.g.\ 
-[p] 101, [b] 102, [t] 103...) following the IPA revision of 1989 and its update in 1996.
-
-The working group made the decision that no IPA symbol, past or present, 
-could be ignored. The comprehensive inclusion of all IPA symbols was to 
-anticipate the possibility that some symbols might be added, withdrawn, 
-or reintroduced into current or future usage. For example, in the 1989 
-revision voiceless implosives < ƥ, ƭ, ƈ, ƙ, ʠ > were added; in the 1993 
-revision they were removed. Ligatures like < ʧ, ʤ > are included as formerly 
-recognized IPA symbols; they are assigned to the ``200 series'' of IPA numbers 
-as members of the group of symbols formerly recognized by the IPA. To ensure 
-backwards compatibility, legacy IPA symbols would retain an IPA Number and 
-an IPA Name for reference purposes. As we discuss below, this decision is 
-later reflected in the Unicode Standard; many legacy IPA symbols reside in 
-the \textsc{IPA Extensions} block.
-
-The IPA Number is simply expressed as a ``three-digit number numerical 
-directory of digit triples'' \cite{}.\footnote{For practical purposes, 
-the IPA Number also served as a typesetter's guide to the IPA chart.} 
-The numbering scheme specifies three-digit codes, the first digit of which 
-indicates the symbol's category \cite{Esling1990,EslingGaylord1993}:
-
-\begin{itemize}
-	\item 100s for accepted IPA consonants
-	\item 200s for former IPA consonants and non-IPA symbols
-	\item 300s for vowels
-	\item 400s for segmental diacritics
-	\item 500s for suprasegmental symbols
-	\item 600s-800s for future specifications
-	\item 900s for escape sequences
-\end{itemize}
-
-After a symbol is categorized, it is assigned a number sequentially, 
-e.g.\ [i] 301, [e] 302, [ɛ] 303. The system allows for the addition 
-of new symbols within the various series by appending them, e.g. [ⱱ] 
-184. Former or often used but non-IPA symbols for consonants, vowels 
-and diacritics are numbered from x99 backwards. For example, the voiceless 
-and voiced postalveolar affricates and fricatives < č, ǰ, š, ž > are 
-assigned the IPA numbers 299, 298, 297 and 296, respectively, because 
-they are not sanctioned IPA symbols.
-
-The assignment of the IPA numbers to IPA symbols provided the basis for 
-uniquely identifying the set of past and present IPA symbols as a type of 
-computational representational standard of the IPA. Within each revision 
-of the IPA, the coding defines a closed and clearly defined set of characters. 
-The benefits of this standardization are clear in at least two ways: it is 
-used in translation tables that reference ASCII representations of the IPA, 
-and this early computational representation of the IPA became the basis for 
-its inclusion into the Unicode Standard version 1.0.
-
-\subsection{SAMPA and X-SAMPA}
-
-True to the working group's aim, the IPA numbers provided a mechanism for 
-a communication interchange standard for creating mapping tables to various 
-computer encodings. For example, the IPA coding system was used as a mapping 
-system in the creation of SAMPA \citep{Wells_etal1992}, an ASCII representation 
-of the IPA symbols.
-
-For a long time, linguists, like all other computer users, were
-limited to ASCII-encoded 7-bit characters, which only includes Latin characters,
-numbers and some punctuation and symbols. Restricted to these standard character
-sets that lacked IPA support or other language-specific graphemes that they
-needed, linguists devised their own solutions.\footnote{Early work addressing
-the need for a universal computing environment for writing systems and their
-computational complexity is discussed in \citet{Simons1989}. A survey of
-practical recommendations for language resources, including notes on encoding,
-can be found in \citet{BirdSimons2003}} For example, some chose to represent
-unavailable graphemes with substitutes, e.g.~the combination of <ng> to
-represent <ŋ>. Tech-savvy linguists redefined selected characters from a
-character encoding by mapping custom made fonts to specific code points.\footnote{For 
-example, SIL's popular font SIL IPA 1990}. However,
-one linguist's electronic text would not render properly on another linguist's
-computer without access to the same font. Furthermore, if two character encodings
-defined two character sets differently, then data could not be reliably and
-correctly displayed. This is a commonly encountered example of the non-interoperability of
-data and data formats.
-
-One solution was the ASCII-ification of the IPA, which simply involved 
-defining keyboard-able sequences as IPA symbol codings.\footnote{\cite{Wells1987} 
-provides an in-depth description of IPA codings from country-to-country. 
-Later ASCII-IPAs include Kirshenbaum (created in 1992 in a Usenet group and 
-named after its lead developer who was at Hewlett-Packard Laboratories) and 
-Worldbet (published in 1993 by \cite{Hieronymus1993}, who was at AT\&T Laboratories).} 
-A successful effort was SAMPA (Speech Assessment
-Methods Phonetic Alphabet), which was created between 1988--1991 in Europe to 
-represent IPA symbols with ASCII
-character sequences \citep{Wells1987,Wells_etal1992}, e.g. <p\textbackslash> 
-for [ɸ]. SAMPA was developed by a group of speech scientists from nine countries 
-in Europe and it constituted the ASCII-IPA symbols needed for phonemic transcription 
-of the principal European Union languages \citep{Wells1995}. It is still widely 
-used in language technology.
-
-Two problems with SAMPA are that (i) it is only a partial encoding of the IPA and (ii) it encodes different
-languages in separate data tables, instead of using a universal alphabet, like
-IPA.\@ SAMPA tables were developed as part of a European Commission-funded project to
-address technical problems like electronic mail exchange (what is now simply
-called email). SAMPA is essentially a hack to work around displaying IPA
-characters, but it provided speech technology and other fields a basis that has
-been widely adopted and often still used in code. So, SAMPA is a collection of tables to be
-compared, instead of a large universal table representing all languages.
-
-An extended version of SAMPA, called X-SAMPA, set out to include every symbol, 
-including all diacritics, in the IPA chart \citep{Wells1995}. X-SAMPA is considered
-more universally applicable because it consists of one table that encodes all
-characters in IPA. In line with the principles of the IPA, SAMPA and X-SAMPA include a 
-repertoire of symbols. These symbols are intended to represent phonemes rather than 
-all allophonic distinctions. Additionally, both ASCII-ifications of IPA are useful because 
-strings of SAMPA or X-SAMPA are (reportedly) uniquely parsable \citep{Wells1995}. However, 
-like the IPA, X-SAMPA has different notations for encoding the same phonetic phenomena 
-(see Pitfall \ref{}).
-
-SAMPA and X-SAMPA have been widely used for speech technology and as an encoding system in
-computational linguistics. In fact, they are stilled used in popular software packages 
-that require ASCII input and some of which have been co-opted for linguistic analyses, 
-e.g.~RuG/L04 and SplitsTree4.\footnote{See \url{http://www.let.rug.nl/kleiweg/L04/} and 
-\url{http://www.splitstree.org/}, respectively}
-
-\subsection{The need for a single multilingual environment}
-
-During the 1980's, it became increasingly clear that an adequate solution 
-to the problem of multilingual computing environments was needed. Linguists 
-were on the forefront of addressing this issue because they faced these 
-challenges head-on by wishing to publish and communicate electronic text 
-with phonetic symbols which were not included in basic ASCII.\footnote{One 
-only needs to look at facsimiles of older electronic documents to see exotic 
-symbols written in by hand.}
-
-Long familiar were linguists already with the distinction between function 
-and form. Even in the context of the computer implementation of writing systems, 
-the necessity to distinguish form and function had been made \citep{Becker1984}. 
-The computer industry, on the other hand, did not consider, ignored, or simply 
-did not encode this principle when creating operating systems like MS-DOS, which 
-were limited to 256 code points (due to computer hardware architecture) and 
-encoded with one-to-one mappings from character codes to graphemes.
-
-Industry was starting to tackle the issues involved in developing a single 
-multilingual computing environment on a variety of fronts, including the then 
-new technology of bitmap fonts and the creation of Font Manager and Script 
-Manager by Apple \citep{Apple1985,Apple1986,Apple1988}. As noted above, around 
-this time linguists were developing work-arounds such as SAMPA, so that they 
-could communicate IPA transcription and use ASCII-based software. Some linguists 
-formalized the issues of multilingual text processing from a computational 
-perspective \citep{Anderson1984,Becker1984,Simons1989}. The study of writing 
-systems was also being invigorated by the computational challenges in making 
-computers work in a multilingual environment.\footnote{\cite[11--15]{Sampson1985} 
-urges linguists to view the study of writing systems as legitimate scientific enquiry.}
-
-% This standard became the basis for a proposal to include the IPA in the first version of the Unicode Standard.
-Decisions by the Computer Coding working group and work they continued after the 1989 Kiel Convention were adopted by the International Phonetic Association.
-These decisions are directly reflected in the Unicode Standard's encoding of IPA, seeing as it was the Association who submitted the script proposal to the Unicode Consortium.
-
-The second major benefit of the standardization of the IPA in a computational 
-representation by the Kiel working group is that it provided the basis for a 
-formal proposal to be submitted to various international standards organizations, 
-several of which were trying to tackle (and in a sense `win') the multilingual 
-computing environment problem. Basically, everyone from corporations to governments 
-to language scientists (for lack of a better term) wanted a single unified multilingual 
-character encoding set for all the world's writing systems, even if they did not 
-understand or appreciate the challenges involved in creating and adopting a solution. 
-Additionally, advancements in computer hardware were making the solutions easier 
-to implement in software.
-
-The engineering problems and solutions had been spelled out years before, e.g.\ 
-a two-byte encoding for multilingual text \citep{Anderson1984}. 
-Although languages vary to an astounding extent (cf.\ \cite{EvansLevinson2009}), 
-writing systems are quite similar formally and the issue of formal representation 
-of the world's orthographic systems had also been addressed \citep{Simons1989}.
-
-A major obstacle in creating a single encoding multilingual environment from 
-the perspective of writing systems involves the distinction between function 
-and form \citep{Becker1984} This distinction is so central to basic linguistic 
-theory and that trained linguists and semiologists take it as second nature. 
-A central challenge in developing a universal character set was to combine a 
-technological solution with a formalization of writing systems proper.\footnote{Of 
-course there were additional practical issues to overcome, e.g.\ funding, creating 
-the formal and technological proposal, deciding which characters and writing systems 
-to include initially, while setting precedence of how to add new ones in the future.}
-
-In hindsight it is easy to lose sight of how impactful 30 years of technological 
-development have been on linguistics, from theory development using quantitative 
-means to pure data collection and dissemination (which as fed back into the former). 
-But at the end of the 1970s, virtually no ordinary working linguist was using a 
-person computer \cite{Simons1996}. Personal computer usage, however, dramatically 
-increased throughout the 1980s. By 1990, dozens of character sets were in common use. 
-They varied in their architecture and in their character repertoires, which 
-made things a mess. There were two major players in the universal character 
-set race: Unicode and the International Organization for Standardization (ISO).
-
-\subsection{Unicode and ISO 10646}
-
-In the late 1980s, a universal character set was being developed by what 
-is now referred to as the Unicode Consortium.\footnote{The Unicode 
-Consortium was officially incorporated in January 1991.} This consortium 
-consisted largely, although not entirely, of major US corporations, with 
-the aim of overcoming the inoperability of different coded character sets 
-and their costly hinderance for developing multilingual software development 
-and for internationalization efforts. Commercial importance of course drove 
-the early inclusion of Latin, non-Latin, and some exotic scripts; see the 
-table of commercial importance as measured by GDP of countries using certain 
-writing systems \citep[2]{unicode88}.
-
-The original Unicode manifesto is \cite{unicode88.pdf}.\footnote{http://www.unicode.org/history/unicode88.pdf} 
-Its aim was for a reliable international multilingual text encoding standard 
-that would encompass all scripts of the world, or in the author's own words, 
-``a new, world-wide ASCII''. An in-depth history of Unicode, highlighting 
-interesting facts like its first text prototypes at Apple and its incorporation 
-into TrueType, is retold online.\footnote{\url{http://www.unicode.org/history/earlyyears.html}}
-
-Unicode 88 provided the basic principles for the Unicode Standard's design -- 
-pushing for 16 bit representations of characters with a clear distinction 
-between characters and glyphs. Some of the contents of this status proposal 
-of 1988 were reworked for inclusion in the early Unicode Standard pre-publication 
-drafts and by August 1990, the proposal was in a (very) rough draft format. Its 
-editors and the Unicode Working Group (the predecessor to the Unicode Technical 
-Committee) worked together to lay out the the proposed standard's structure and 
-content. At this time, the proposal contained no code charts nor block descriptions.
-
-% http://www.unicode.org/history/earlyyears.html
-% During this period of time, in addition to his co-authoring of Apple KanjiTalk, Davis was involved in further discussions of a universal character set which were being prompted by the development of the Apple File Exchange.
-
-The other major player in developing a universal character set was the ISO 
-working group from the International Standards Organization (ISO), based 
-in Europe, which was responsible for ISO/IEC 10646. This character set 
-standard was composed in 1989 and a draft was published in 1990.\footnote{\url{http://www.iso.org/iso/catalogue_detail.htm?csnumber=56921}} 
-The `Universal Multiple-Octet Coded Character Set' or simply UCS was the first 
-officially standardized character encoding with the aim of including all characters from all writing systems.\footnote{\url{http://www.nada.kth.se/i18n/ucs/unicode-iso10646-oview.html}}
-
-ISO/IEC 10646 is partly based on ISO/IEC 8859, a series of of ASCII-based 
-standard character encodings published in 1987 that use a single bit 8-byte 
-character set. Each part of the standard, e.g.\ 8859-1, 8859-5, 8859-6, 
-encodes characters to support different languages' writing systems, e.g.\ 
-Latin-1 Western European, Latin/Cyrillic, Latin/Arabic, respectively. Being 
-a joint effort by the International Organization for Standardization (ISO) 
-and the International Electrotechnical Commission (IEC), the aim of the 
-standard is reliable information exchange. So, again, issues of phonetic 
-symbol encoding, typography, etc., were ignored -- or perhaps more properly 
-put, not commercially driven at this early stage.
-
-Intended for the major Western European languages, ISO/IEC 8859 was an 
-extension of the ASCII character encoding standard, which included the 
-English alphabet, numerals and computer control characters (e.g.\ beep, 
-space, carriage return). By extending ASCII's 7-bit system to 8-bit, the 
-character repertoire of each of ISO/IEC 8859 character set was doubled 
-from 128 to 256 characters. Each character set defined a mapping between 
-digital bit patterns and characters, which are visually rendered on screen 
-as graphic symbols. ASCII was shared between ISO/IEC 8859 character sets, 
-but the characters in the extra bit patterns were different. Thus an aim 
-of the ISO working group responsible for ISO/IEC 10646 was to bring all 
-characters in all writings systems into a single unified encoding.
-
-In 1991, the Unicode Consortium and the ISO Working Group for ISO/IEC 10646 
-decided to create a single universal standard for encoding multilingual text.\footnote{http://unicode.org/book/appC.pdf} 
-The two character sets converged, resulting in mutually acceptable changes 
-to both, and each group keeps versions of their respective character codes 
-and encoding forms synchronized.\footnote{http://www.unicode.org/versions/} 
-Although each standard has its own form of reference and the terminology in 
-each may differ slightly, the practical difference is that the Unicode Standard 
-is a formal implementation of ISO/IEC 10646 and imposes additional constraints 
-on its implementation. The Unicode Standard includes character data, algorithms 
-and specifications, outside the scope of ISO/IEC 10646, which ensure, when 
-properly implemented in software applications and platforms, that characters 
-are treated uniformly.
-
-The incorporation of the Unicode Standard into the international encoding 
-standard ISO 10646 was approved by ISO as an International Standard in June 
-1992.\footnote{http://www.unicode.org/versions/Unicode1.0.0/Notice.pdf} 
-The joint Unicode and ISO/IEC 10646 standard has become \textit{the} universal 
-character set and it is a single multilingual environment for the majority 
-of the world's written languages. Its formal implementation has also been 
-vital to the rise of a multi-lingual Internet.
-
-\subsection{IPA and Unicode}
-
-It was a long journey, but the goal of achieving a single multilingual 
-computing environment has largely been accomplished. We users, however, 
-must cope with the pitfalls that were dug along the way (see Pitfall sections below). 
-Some linguists, including your humble authors, are particularly sensitive 
-to these issues. So we provide practical advice and approaches in the rest 
-of this chapter. But first, we explain how the IPA became incorporated 
-into the Unicode Standard via ISO/IEC 10646.
-
-% "The set of IPA symbols and their numbers were used to draw up an entity set within SGML by the Text Encoding Initiative (TEl).
-The name of each entity is formed by 'IPA' preceding the number, e.g.
-IPA304 is the rEIentity name of lower-case A.
-These symbols can be processed as IPA symbols and represented on paper and screen with the appropriate local font by modifying the :entity replacement text.
-The advantage of the SGML entity set is that it is independent or the character set being used."
-% "A TEl writing system declaration (wsd) has been drawn up for the IPA symbols."
-% A TEl writing system declaration (wsd) has been drawn up for the IPA symbols.
-This document gives information about the symbol and its IPA function, as well as its encoding in the accompanying SGML document and in UnicodelUCS and in AFII.
-The writing system declaration can be read as a text d9cument or processed by machines in an SMGL process.
-
-After the Kiel Convention in 1989, the Computer Coding of the IPA working 
-group assisted the International Phonetic Association in representing the 
-IPA to ISO and to the Text Encoding Initiative (TEI) \citep{EslingGaylord1993}. 
-The working group's formalization of the IPA, i.e.\ a full listing of agreed 
-upon computer codings for phonetic symbols, was used in developing writing 
-systems descriptions which were at the time being solicited for scripts to 
-be included in initiatives for new multilingual international character 
-encoding standards. The working group for ISO/IEC 10646 and Unicode were 
-two such initiatives.
-
-In the historical context of the IPA being considered for inclusion in 
-ISO/IEC 10646, it is important to realize that there were a variety of 
-sources (i.e.\ not just from the Association) which submitted character 
-proposals for phonetic alphabets. These proposals, including from the 
-Association via the Kiel working group, were considered as a whole by 
-the ISO working groups which were responsible for incorporating a phonetic 
-script into the universal character set (UCS). The ISO working groups that 
-were responsible for assigning a phonetic character set then made their 
-own submissions as part of a review process by ISO for approval based on 
-both ``informatic'' and phonetic criteria \citep[86]{EslingGaylord1993}.
-
-Character set ISO/IEC 10646 was approved by ISO, including the phonetic 
-characters submitted to them in May 1993. The set of IPA characters were 
-assigned UCS codes in 16 bit representation (in hexadecimal) and were 
-published Tables 2 and 3 in \cite{EslingGaylord1993}, which include a 
-graphical representation of the IPA symbol, its IPA Name, phonetic description, 
-IPA Number, UCS Code and AFII Code.\footnote{The Association for Font 
-Information Interchange (AFII) was an international database of glyphs 
-created to promote the standardization of font data required to produce 
-ISO/IEC 10646.} Because the character sets of ISO/IEC 10646 and the 
-Unicode Standard converged, the IPA as submitted by the Association 
-and reviewed and further submitted by the ISO working group, was included 
-in the Unicode Standard Version 1.0 -- largely as we know it today.\footnote{The 
-Association later made the foresightful remark, ``When this character set 
-is in wide use, it will be the normal way to encode IPA symbols.'' \citep[164]{IPA1999}.}
-
-With subsequent revisions to the IPA, one might expect the Unicode 
-Consortium would update the Unicode Standard in a way that is inline 
-with linguists' or other language scientists' intuitions. However, 
-updates that go against the ISO's and the Unicode Standard's principles 
-of maintaining backwards compatibility lose out, i.e.\ it is more important 
-to deal with the pitfalls created along the way then it is to change the 
-standard. Therefore, many of the pitfalls we encounter when using Unicode 
-IPA are historic relics that we have to come to grips with.
-
-% https://en.wikipedia.org/wiki/Uralic_Phonetic_Alphabet
-% http://www.unicode.org/conference/bulldog.html
-% http://www-01.sil.org/computing/computing_environment.html
+Many of the current problems derive from the fact that the IPA is clearly
+historically based on the Latin script, but different enough from most other
+Latin-based writing systems to warrant special attention. This ambivalent status
+of the IPA glyphs (partly Latin, partly special) is unfortunately also attested
+in the treatment of IPA in the Unicode Standard. In retrospect, it might have
+been better to consider the IPA (and other transcription systems) to be a
+special kind of script within the Unicode Standard, and treat the obvious
+similarity to Latin glyphs as a historical relic. All IPA glyphs would then have
+their own code points, instead of the current situation in which some IPA glyphs
+have special code points, while others are treated as being identical to the
+regular Latin characters. Yet, the current situation, however unfortunate, is
+unlikely to change, so as linguists we must learn to deal with the specific
+pitfalls of IPA within the Unicode Standard.
 
@@
 Conversely, non-linguists are unlikely to distinguish any semantic difference
-between an open back unrounded vowel <ɑ> \textsc{latin small letter alpha} at
-\uni{0251}, and the open front unrounded vowel <a> \textsc{latin small letter a}
-at \uni{0061}, basically treating them as homoglyphs, although they are
-different phonetic symbols. But even among linguists this distinction leads to
-problems. For example, as pointed out by \citet{Mielke2009}, there is a problem
-stemming from the fact that about 75\% of languages are reported to have a
-five-vowel system \citep{Maddieson1984}. Historically, linguistic descriptions
-tend not to include precise audio recording and measurements of formants, so
-this may lead one to ask if the many <a> characters that are used in
-phonological description reflects a transcriptional bias. The common use of <a>
-in transcriptions could be in part due to the ease of typing the letter on an
-English keyboard (or for older descriptions, the typewriter). We found it to be
-exceedingly rare that a linguist uses <ɑ> for a low back unrounded
-vowel.\footnote{One example is \citet[75]{Vidal2001a}, in which the author
-states: ``The definition of Pilagá /a/ as [+back] results from its behavior in
-certain phonological contexts. For instance, uvular and pharyngeal consonants
-only occur around /a/ and /o/. Hence, the characterization of /a/ and /o/ as a
-natural class of (i.e., [+back] vowels), as opposed to /i/ and /e/.''} They
-simply use <a> as long as there is no opposition to <ɑ>.
+between an open back unrounded vowel, which is encoded in IPA with a
+`single-story' <ɑ> \textsc{latin small letter alpha} at \uni{0251}, and the open
+front unrounded vowel, which is encoded in IPA with a `double-story' <a>
+\textsc{latin small letter a} at \uni{0061}, basically treating them as
+homoglyphs, although they are different phonetic symbols. But even among
+linguists this distinction leads to problems. For example, as pointed out by
+\citet{Mielke2009}, there is a problem stemming from the fact that about 75\% of
+languages are reported to have a five-vowel system \citep{Maddieson1984}.
+Historically, linguistic descriptions tend not to include precise audio
+recording and measurements of formants, so this may lead one to ask if the many
+<a> characters that are used in phonological description reflects a
+transcriptional bias. The common use of <a> in transcriptions could be in part
+due to the ease of typing the letter on an English keyboard (or for older
+descriptions, the typewriter). We found it to be exceedingly rare that a
+linguist uses <ɑ> for a low back unrounded vowel.\footnote{One example is
+\citet[75]{Vidal2001a}, in which the author states: ``The definition of Pilagá
+/a/ as [+back] results from its behavior in certain phonological contexts. For
+instance, uvular and pharyngeal consonants only occur around /a/ and /o/. Hence,
+the characterization of /a/ and /o/ as a natural class of (i.e., [+back]
+vowels), as opposed to /i/ and /e/.''} They simply use <a> as long as there is
+no opposition to <ɑ>.
 
@@
 not acceptable when the goal is interoperability between resources written in
-IPA.\@ We consider it crucial to distinguish between ``lax'' IPA, for which it
+IPA.\@ We consider it crucial to distinguish between ``valid'' IPA, for which it
 is sufficient that any phonetically-trained reader is able to understand the
@@
 The IPA also proposes the option of tone letters, e.g. <˥˦˧˨˩>, which are much
-less often used, but are more consistent for contours. Tone letters in the IPA
-have five different levels, and sequences of these letters can be used to
-indicate contours. Well-designed fonts will even merge a sequence of tone
-letters into a contour. For example, compare the font Linux Libertine, which
-does not merge tone letters <{\fontspec{LinLibertineO}˥˨˧˩}>, with the font
-CharisSIL, which merges this sequence of four tone letters into a single contour
-<\charis{˥˨˧˩}>. For strict IPA encoding we propose to standardize on tone
-letters.
+less often used, but are more consistent for contours.\footnote{Not sanctioned
+by the IPA, but nevertheless widely attested, is the usage of superscript
+numbers for marking tones.} Tone letters in the IPA have five different levels,
+and sequences of these letters can be used to indicate contours. Well-designed
+fonts will even merge a sequence of tone letters into a contour. For example,
+compare the font Linux Libertine, which does not merge tone letters
+<{\fontspec{LinLibertineO}˥˨˧˩}>, with the font CharisSIL, which merges this
+sequence of four tone letters into a single contour <\charis{˥˨˧˩}>. For strict
+IPA encoding we propose to standardize on tone letters.
 
@@
 sequences in a writing system have to be specified as ``tailored'' grapheme
-clusters. In case the sequence is not a cluster, this has to be explicitly
-indicated. IPA takes a different approach. In IPA by default different base
-letters are not connected into larger clusters; only when it is specified in the
-string itself (using the tie bar).
+clusters. Only in case the sequence is not a cluster, then this has to be
+explicitly indicated. IPA takes a different approach. In IPA by default
+different base letters are not connected into larger clusters; only when it is
+specified in the string itself (using the tie bar).
 
@@
 in IPA that allow for multiple, apparently identical, encodings that are not
-canonical equivalent (see Section~\ref{pitfall-canonical-equivalence}). The
-following elements should not be treated as diacritics when encoding IPA in
-Unicode: 
+canonical equivalent (see Section~\ref{pitfall-canonical-equivalence} on the
+principle of canonical equivalence). For that reason, the following elements
+should not be treated as diacritics when encoding IPA in Unicode:
 \begin{itemize}
@@
 Also related to diacritics is the question of ordering. To our knowledge, the
-International Phonetic Association does not specify a specific ordering for
+International Phonetic Association does not specify an ordering for
 diacritics that combine with phonetic base symbols; this exercise is left to the
@@
 diacritics (which are not diacritics in the Unicode sense) between diacritics
-below and the tie bar. Further, within all these classes of diacritics there is
-no canonical ordering specified by Unicode, so we propose an explicit ordering
-here.
+below and the tie bar. Further, \textit{within} all these classes of diacritics
+there is no canonical ordering specified by Unicode, so we propose an explicit
+ordering here.
 
@@
 the convention was not to change specific symbols unless a strong 
-case was made \cite{Ladefoged1990a}. For example, two such decisions 
+case was made \citep{Ladefoged1990a}. For example, two such decisions 
 included:
@@
 
-The most controversial issue regarding symbols debated at the convention 
-was the representation for voiceless implosives \cite[62]{Ladefoged1990a}. 
-Recall the principles of the IPA outlined in Section \ref{the-international-phonetic-alphabet}, 
-such as using distinct symbols for phonological contrast versus convenience of 
-display in the chart, must be taken into account when arguing for or against 
-the inclusion or deletion of IPA symbols in the IPA chart. Furthermore, the 
-inclusion of deletion of symbols should incorporate the phonetic knowledge 
-of the world's languages.
-
-\cite{Ladefoged1990a} argued against the inclusion of these symbols < ƥ, ƭ, ƈ, ƙ, ʠ > 
-voiceless implosives, noting that: they are not contrastive (e.g.\ in Mayan 
-languages); there is no instrumental evidence supporting voiceless implosives 
-in Africa; the sounds are sufficiently rare as to not need a whole new row 
-of symbols in the chart. Ladefoged favored symbolizing the sounds using a 
-voiceless diacritic ring below voiced implosives, e.g.\ <ɓ̥>.
-
-Nevertheless, in the 1989 IPA chart there is indeed a row for implosives 
-containing voiceless and voiced pairs.\footnote{https://en.wikipedia.org/wiki/File:IPA\_as\_of\_1989.png} 
-But already in the next revision, in 1993 (with an update in 1996), the voiceless 
+The most controversial issue regarding symbols debated at the convention was the
+representation for voiceless implosives \citep[62]{Ladefoged1990a}. In
+accordance with the principles of the IPA, as outlined in Section
+\ref{the-international-phonetic-alphabet}, distinct symbols are favoured for
+cases of phonological contrast. Further, also convenience of display in the chart
+must be taken into account when arguing for or against the inclusion or deletion
+of IPA symbols in the IPA chart. Finally, the inclusion or deletion of symbols
+should consider the current state of phonetic knowledge of the world's
+languages.
+
+\citet{Ladefoged1990a} argued against the inclusion of the symbols < ƥ, ƭ, ƈ, ƙ,
+ʠ > for voiceless implosives, noting (i) that they are not contrastive (e.g.\ in
+Mayan languages); (ii) that there is no instrumental evidence supporting
+voiceless implosives in Africa; and (iii) that the sounds are sufficiently rare
+as to not need a whole new row of symbols in the chart. Ladefoged favored
+symbolizing the sounds using a voiceless diacritic ring below voiced implosives,
+e.g.\ <ɓ̥>. Nevertheless, in the 1989 IPA chart there is indeed a row for
+implosives containing voiceless and voiced
+pairs.\footnote{https://en.wikipedia.org/wiki/File:IPA\_as\_of\_1989.png} But
+already in the next revision, in 1993 (with an update in 1996), the voiceless
 implosives were dropped. The implosives row from the IPA consonantal chart
 disappeared and voiced implosives were given a column in the non-pulmonic
-consonants table (which is still reflected in the latest revision to date, IPA 2005).
-
-The Journal of the International Phonetic Association follows its own 
+consonants table (which is still reflected in the latest revision to date, IPA
+2005).
+
+The \textit{Journal of the International Phonetic Association} follows its own 
 published standard for the IPA at the time of publication, even when it 
@@
 for contrastive sounds and diacritics for phonetic variation. For example, 
-in the case of voiceless implosives, \cite{McLaughlin2005} shows that 
-Seereer-Siin (Niger-Congo; Atlantic; ISO 639-3: srr) has a phonologically 
+in the case of voiceless implosives, \citet{McLaughlin2005} shows that 
+Seereer-Siin (Niger-Congo, Atlantic; ISO 639-3 srr) has a phonologically 
 contrastive set of voiced and voiceless implosive stops at the labial, 
@@
 in an \textit{Illustrations of the IPA} article in the IPA journal as 
-< ɓ̥, ɗ̥, ʄ̥ >.\footnote{Note that the IPA chart states that diacritics 
-can also be placed above a symbol, e.g.\ reportedly for legibility's 
-sake < ɠ̥> vs < ɠ̊ >. This of course leads to the variability in encoding 
-that we are trying to address on a computational level in this work. 
-See Section \ref{}.}
+< ɓ̥, ɗ̥, ʄ̥ >.
 

### 2015-11-13 book/chapters/ipa_background.tex
splitting and readability revisions

@@
 % ==========================
-\chapter{The International Phonetic Alphabet (IPA)}
+\chapter{The International Phonetic Alphabet}
 \label{the-international-phonetic-alphabet}
@@
 contained phonetic values for English, French and German. Diacritics for length
-and nasalization were already present in this version, and the same symbols are
-still used today. 
+and nasalization were already present in this first version, and the same
+symbols are still used today. 
 %\footnote{Also referred to as API, for \textit{Association Phonétique Internationale}.}
@@
 from words in different languages. In 1900 the symbols were first organized into
-chart and were given phonetic feature labels for consonants (for manner of
-articulation e.g.\ `plosives', `nasales', `fricatives'; for place of
-articulation, e.g.\ `bronchiales', `laryngales', `labiales') and for vowels
-(e.g.\ `fermées', `mi-fermées', `mi-ouvertes', `ouvertes'). Throughout the last
-century, the structure of the chart has changed with increases in phonetic
-knowledge, and thus, like notational systems in other scientific disciplines,
-the IPA reflects facts and theories of phonetic knowledge that have developed
-over time. It is natural then that the IPA is modified occasionally to
-accommodate scientific innovations and discoveries. In fact, this is part of
-the Association's mandate. These changes are captured in the revisions to the
-IPA.
+a chart and were given phonetic feature labels for consonants (e.g.\ for manner
+of articulation among others `plosives', `nasales', `fricatives' and for place
+of articulation among others `bronchiales', `laryngales', `labiales') and for
+vowels (e.g.\ `fermées', `mi-fermées', `mi-ouvertes', `ouvertes'). Throughout
+the last century, the structure of the chart has changed with increases in
+phonetic knowledge. Thus, similar to notational systems in other scientific
+disciplines, the IPA reflects facts and theories of phonetic knowledge that have
+developed over time. It is natural then that the IPA is modified occasionally to
+accommodate scientific innovations and discoveries. In fact, this is part of the
+Association's mandate. These changes are captured in the recurring revisions to
+the IPA.
 %\footnote{For a detailed history see:
@@
 
-Over the years there has been several revisions, many minor. For example,
-articulation labels -- what often are call features even though the IPA avoids
-this term -- have changed (e.g.\ terms like `lips', `throat' or `rolled' are not
-used anymore). Phonetic symbol values have changed (e.g.\ voiceless is no longer
-marked by <h>). Symbols have been dropped (e.g.\ the caret diacritic denoting
-`long and narrow' is no longer used). And many symbols have been added to reflect
-contrastive sounds found in the world's very diverse phonological systems.
+Over the years there have been several revisions, many minor. For example, the
+articulation labels -- what often are call `features' even though the IPA
+deliberately avoids this term -- have changed (e.g.\ terms like `lips', `throat'
+or `rolled' are not used anymore). Phonetic symbol values have changed (e.g.\
+voiceless is no longer marked by <h>). Symbols have been dropped (e.g.\ the
+caret diacritic denoting `long and narrow' is no longer used). And many symbols
+have been added to reflect contrastive sounds found in the world's very diverse
+phonological systems.
 
-Although IPA began its life as a pedagogical tool, from its earliest days the 
+Although the IPA began its life as a pedagogical tool, from its earliest days the 
 Association aimed to provide ``a separate sign for each distinctive sound; 

### 2015-11-13 book/chapters/ipa_meets_unicode.tex
splitting and readability revisions

@@
 although we think this is unlikely.
-
-\begin{comment}
-\subsection{IPA (practical) extensions}
-
-The IPA extensions cover speech sounds beyond the sound systems of the world's
-languages, including symbols for pathological speech and paralinguistic
-functions (e.g.\ Braille?).
-
-- pathological speech
-- paralinguistic functions (Braille?)
-
-As noted in Section \ref{}, extensions to the IPA -- in particular
-``large-scale'' extensions --
-
-% Englebretson2009.pdf
-http://www.ruf.rice.edu/~reng/englebretson2009.pdf
-
-X-SAMPA was an ASCII transliteration and is still purposeful for several
-programs
-
-There is also the extensions of the IPA designed for disordered speech, although
-the symbols are occasionally used for transcribing phonetic detail in ``normal''
-speech or also cross-linguistic
-rarities.\footnote{\url{http://phoible.github.io/conventions/}}
-
-% Unicode has its own history; for a summary:
-\url{http://www.unicode.org/history/summary.html}
-
-https://www.phon.ucl.ac.uk/home/wells/ipa-unicode.htm
-
-Computer codes for phonetic symbols
-EslingGaylord1993
-
-\end{comment}
 
@@
 Summarizing the pitfalls as discussed in this chapter, we propose to define
-three different IPA encodings: strict-IPA, valid-IPA and widened-IPA.\@ These 
-three encodings are subsets of each other, i.e.~strict-IPA is more restricted 
-than valid-IPA, which it turn is more restricted than widened-IPA.\@ Informally 
-speaking, valid-IPA represents the current state of the IPA \citep{IPA2005}. 
-Strict-IPA represents a more constrained version of IPA, while widened-IPA is a 
-slightly extended version of IPA, allowing a few more symbols.
+three different IPA encodings: strict-IPA, valid-IPA and widened-IPA.\@
+Informally speaking, valid-IPA represents the current state of the IPA
+\citep{IPA2005}. Strict-IPA represents a more constrained version of IPA, while
+widened-IPA is a slightly extended version of IPA, allowing a few more symbols.
 
@@
 widened-IPA.\@
-
-textbf{NOTE: I HAVE ADDED VARIOUS TONE DIACRITICS TO VALID-IPA THAT ARE NOT IN
-THE OFFICIAL IPA SUMMARY. SHOULD THEY BE INCLUDED?}
 

### 2015-11-13 book/chapters/ipa_meets_unicode.tex
adding note on tone and stress markers

@@
 Although linguists might expect these characters to belong together with the
-character in front of them, at least for <ʰ> \textsc{modifier letter small h} at
-\uni{02B0} the Unicode Consortium's decision to treat it as a separate character
-is also linguistically correct, because according to the IPA it can be used both
-for post-aspiration (following the base character) and pre-aspiration (preceding
-the base character). The default combination of Spacing Modifiers with the
-preceding character can be specified in orthography profiles (see
-Chapter~\ref{orthography-profiles}). Note that there is an alternative 
-homoglyph, namely the \textsc{combining latin small letter h} at \uni{036A}. We 
-propose not to use this combining diacritical mark, but to standardize on 
-Unicode modifier letters for all superscript letters in IPA.
+character in front of them, at least for tone letters, stress symbols and <ʰ>
+\textsc{modifier letter small h} at \uni{02B0} the Unicode Consortium's decision
+to treat it as a separate character is also linguistically correct.
+
+\begin{itemize}
+
+   \item According to the IPA, <ʰ> can be used both as <\dia{02B0}> for
+         post-aspiration (following the base character) and as
+         <\diareverse{02B0}> for pre-aspiration (preceding the base character).
+         Note that there is yet another homoglyph, namely the
+         \textsc{combining latin small letter h} at \uni{036A}. We propose not
+         to use this combining diacritical mark, but to standardize on Unicode
+         modifier letters for all superscript letters in IPA.
+
+   \item Tone letters are normally written at the end of the syllable,
+         possibly occurring immediately adjacent to a consonant in the coda of
+         the syllable. Such tone markers should of course not be treated as
+         belonging to this consonant, so we propose to treat tone letters as 
+         separate segments.
+
+   \item Stress markers <\diareverse{02C8}> at \uni{02C8} and
+         <\diareverse{02CC}> at \uni{02CC} have a very similar
+         distribution in that they normally are written at the start of
+         the stressed syllable. In a sense, they thus belong to the
+         characters following the stress marker, but it would be wrong to
+         cluster them together with whatever segment is at the start of
+         the syllable. So, like tone letters, we propose to treat stress
+         markers as separate segments.
+
+\end{itemize}
+
+If intended, then any default combination of Spacing Modifiers
+with the preceding character can be specified in orthography profiles (see
+Chapter~\ref{orthography-profiles}).
 

### 2015-11-13 book/chapters/pitfalls.tex
correction wrong unicode symbols :-)

--- 2015-09-28 chapters/pitfalls.tex
@@
 character, as exemplified by diacritics like the diaeresis <\dia{0308}> in <ë>.
-
-test {\fontspec[Alternative=02]{AlBayan}ب}
 
@@
 \textsc{latin small letter g} at \uni{0067}). In contrast, the symbols
-<{\fontspec{EversonMono}AАΑᎪᗅᴀꓮʠ햠홰}> are all different code points,
+<{\fontspec{EversonMono}AАΑᎪᗅᴀꓮ𝖠𝙰}> are all different code points,
 although they look highly similar---in some cases even sharing exactly the same
@@
 %	\item[] <{\fontspec{EversonMono}̀}> \textsc{old italic letter a}, at \uni{10300} 
-	\item[] <{\fontspec{EversonMono}햠}> \textsc{mathematical sans-serif capital a}, \uni{1D5A0} 
-	\item[] <{\fontspec{EversonMono}홰}> \textsc{mathematical monospace capital a}, at \uni{1D670} 
+	\item[] <{\fontspec{EversonMono}𝖠}> \textsc{mathematical sans-serif capital a}, \uni{1D5A0} 
+	\item[] <{\fontspec{EversonMono}𝙰}> \textsc{mathematical monospace capital a}, at \uni{1D670} 
 \end{itemize}
@@
 memory, which is known as \textsc{endianness}. The details of this issue go
-beyond the scope of this book. It suffices to realize that there is a
-difference between \textsc{big-endian} (BE) storage and \textsc{little-endian}
-(LE) storage. The Unicode Standard offers a possibility to explicitly indicate
-what kind of storage is used by starting a file with a so-called \textsc{byte order
+beyond the scope of this book. It suffices to realize that there is a difference
+between \textsc{big-endian} (BE) storage and \textsc{little-endian} (LE)
+storage. The Unicode Standard offers a possibility to explicitly indicate what
+kind of storage is used by starting a file with a so-called \textsc{byte order
 mark} (BOM). However, the Unicode Standard does not require the usage of BOM,
@@
 % ==========================
-\section{Pitfall: Software}
-\label{software}
+%\section{Pitfall: Software}
+%\label{software}
 % ==========================

### 2015-11-15 book/chapters/ipa_background.tex
revision of first two sections of IPA introduction

@@
 
+% ==========================
 \section{Brief history}
+\label{IPAhistory}
+% ==========================
 
@@
 have been added to reflect contrastive sounds found in the world's very diverse
-phonological systems.
+phonological systems. The current construction and use of the IPA are guided by
+principles outlined in the \textit{Handbook of the International Phonetic
+Association: A guide to the use of the International Phonetic Alphabet}
+\citep[159]{IPA1999}, henceforth simply called \textit{Handbook}.
 
-Although the IPA began its life as a pedagogical tool, from its earliest days the 
-Association aimed to provide ``a separate sign for each distinctive sound; 
-that is, for each sound which, being used instead of another, in the same 
-language, can change the meaning of a word'' \citep[27]{IPA1999}. Distinctive 
-sounds became later known as \textsc{phonemes} and the IPA has developed historically 
-into a notational devise with a strictly segmented phonemic view. A phoneme is an 
-abstract theoretical notion derived from an acoustic signal as produced by speakers 
-in the real world. Therefore the IPA contains a number of theoretical assumptions 
-about speech and how to analyze speech in written form.
+Today, the IPA is designed to meet practical linguistic needs and is used to
+transcribe the phonetic or phonological structure of languages. It is also used
+increasingly as a foreign language learning tool, as a standard pronunciation
+guide and as a tool for creating practical orthographies of previously unwritten
+languages. The IPA suits many linguists' needs because:
 
-Phonetic analysis is based on two premises: (i) that it is possible to describe
-the acoustic speech signal (i.e.\ sound waves) in terms of sequentially ordered
-discrete segments, and, (ii) that each segment can be characterized by an
-articulatory target.\footnote{A purely phonetic description is only derivable
+\begin{itemize}
+
+	\item it is intended to be a set of symbols for representing all possible
+       sounds in the world's (spoken) languages
+	\item its chart has a linguistic basis (specifically a phonological bias)
+       rather than a general phonetic notation scheme
+	\item its symbols can be used to represent distinctive feature
+       combinations\footnote{Although the chart uses traditional manner and
+       place of articulation labels, the symbols can nevertheless represent any
+       defined bundle of features, binary or otherwise, to define phonetic
+       dimensions.}
+	\item its chart provides a summary of linguists' agreed-upon phonetic knowledge
+       (a common denominator of phonological ``facts'')
+
+\end{itemize}
+
+Several styles of transcription with IPA are possible, as illustrated in the
+\textit{Handbook}, and they are all valid (see the 29 languages and their
+transcriptions in the original and initial \textit{Illustrations of the IPA}
+\citep[41--154]{IPA2007}). Therefore, there are different but equivalent
+transcriptions, or as \citet[64]{Ladefoged1990a} captures it, ``Perhaps now that
+the Association has been explicit in its eclectic approach, outsiders to the
+Association will no longer speak of \textit{the} IPA transcription of a given
+phenomenon, as if there were only one approved style.'' Clearly not all phoneticians agree (or will likely ever agree of course) on all 
+aspects of the IPA or on transcription practices. As noted above, there have 
+been several revisions in the IPA's long history, but the current version (2005) 
+is strikingly similar to the 1926 version, showing the viability of the IPA as a 
+common standard for linguistic transcription.
+
+% ==========================
+\section{Premises}
+\label{IPApremises}
+% ==========================
+
+Any IPA transcription is based on two premises: (i) that it is possible to
+describe the acoustic speech signal (i.e.\ sound waves) in terms of sequentially
+ordered discrete segments, and, (ii) that each segment can be characterized by
+an articulatory target.\footnote{A purely phonetic description is only derivable
 from instrumental data in high quality sound recordings. Once spoken language
@@
 
-\noindent and in the \textit{Handbook of the International Phonetic Association} \citep[18]{IPA1993}:
+\noindent and in the \textit{Handbook} \citep[18]{IPA1999}:
 
@@
 representation of sounds in two languages, it does not necessarily mean that the
-sounds in the two languages are identical. \end{quote}
+sounds in the two languages are identical.
+\end{quote}
 
-\noindent Although the IPA provides symbols to unambiguously represent phonemes,
-it also aims to represent phonetic details. Since phonetic detail could
-potentially include anything, e.g. something like `deep voice', the IPA
-restricts phonetic detail to linguistically relevant aspects of speech. These
-principles and conventions for using the IPA are outlined in the Association's
-Handbook.
+Although the IPA provides symbols to unambiguously represent phonemes, it also
+aims to represent phonetic details. Since phonetic detail could potentially
+include anything, e.g. something like `deep voice', the IPA restricts phonetic
+detail to linguistically relevant aspects of speech. From its earliest days the
+Association aimed to provide ``a separate sign for each distinctive sound; that
+is, for each sound which, being used instead of another, in the same language,
+can change the meaning of a word'' \citep[27]{IPA1999}. Distinctive sounds
+became later known as \textsc{phonemes} and the IPA has developed historically
+into a notational devise with a strictly segmental phonemic view. A phoneme is
+an abstract theoretical notion derived from an acoustic signal as produced by
+speakers in the real world. Therefore the IPA contains a number of theoretical
+assumptions about speech and how to analyze speech in written form.
 
+% ==========================
 \section{Principles}
+\label{IPAprinciples}
+% ==========================
 
-IPA transcription has essentially two parts. The first is a text containing IPA 
-symbols and the second is a set of conventions (rules) for interpreting those 
-symbols (and their combinations). The IPA is designed to meet practical linguistic 
-needs and is used to transcribe the phonetic or phonological structure of languages. 
-It is also used increasingly as a foreign language learning tool, as a standard 
-pronunciation guide and as a tool for creating practical orthographies of 
-previously unwritten languages.
+Transcription has essentially two parts. The first is a text containing IPA 
+symbols and the second is a set of conventions for interpreting those 
+symbols (and their combinations).
 
-The current construction and use of the IPA are guided by principles outlined 
-in the \textit{Handbook of the International Phonetic Association: A guide 
-to the use of the International Phonetic Alphabet} \citep[159]{IPA1999}, henceforth 
-HB. The use of the symbols to represent a language's phonological system is guided 
-by the principle of contrast; where two words are distinguishable by phonetic contrast, 
-those contrasts should be transcribed with different symbols (graphemes not diacritics). 
-Allophonic distinction falls under the rubric of diacritically-distinguished symbols, 
-e.g. [stop̚] vs [spʰot]. In sum:
+First, the symbols of IPA crucially distinguish between letter-like symbols and
+diacritics.\footnote{see Section~\ref{pitfall-different-notions-of-diacritics} for a more
+detailed discussion about the notion of diacritics} The use of the letter-like
+symbols to represent a language's sounds is guided by the principle of contrast;
+where two words are distinguishable by phonemic contrast, those contrasts should
+be transcribed with different letter symbols (not diacritics). Allophonic
+distinction falls under the rubric of diacritically-distinguished symbols, e.g.
+the difference in English between and unrealized /p/ in [stop̚] vs. an aspirated
+/p/ in [spʰot] should be marked with a diacritic. In sum:
 
 \begin{itemize}
-	\item different symbols (without diacritics) should be used whenever a language employs two contrastive sounds
-	\item when two sounds in a language are not known to be contrastive, the same symbol should be used to represent these sounds; however, diacritics may be used to distinguish such sounds when necessary
-	\item diacritics cannot be dispensed with entirely, so the Association recommends to limit their use to:
-	\begin{itemize}
-		\item denoting length, stress and pitch
-		\item representing minute shades of sounds
-		\item obviating the design of a (large) number of new symbols when a single diacritic suffices (e.g. nasalized vowels, aspirated stops)
-	\end{itemize}
+
+	\item Different letter-like symbols should be used whenever
+          a language employs two sound contrastively.
+	\item When two sounds in a language are not known to be contrastive, the same
+          symbol should be used to represent these sounds. Diacritics may
+          be used to distinguish different articulations when necessary.
 \end{itemize}
 
-\noindent Thus, an IPA transcription \textit{always} consists of ``a set of 
-symbols and a set of conventions for their interpretation''.
-
-In systematic transcription (as opposed to impressionistic), there is a division 
-between phonemic and allophonic transcription. The terms phoneme and allophone 
-contain theoretical baggage, but the basic goal of a phonemic transcription is to 
-distinguish all words in a language with the minimal number of transcription 
-symbols \citep[19]{Abercrombie1964}. Allophonic transcription uses a broader 
-set of distinct symbols to describe systematic allophonic differences in sounds 
-in words. These two systematic transcriptions are related to each other by a 
-set of conventions in the IPA tradition, so that they can be converted between 
-one and another.
-
-An IPA transcription is connected to a speech event by a set of conventions. 
-A phonetic (or impressionistic) transcription may use the conventions implicit 
-in the IPA chart, i.e. the transcriber can indicate that the phonetic value of 
-<ŋ͡m> is a simultaneous labial and velar closure which is voiced and contains 
-nasal airflow.
-
-A phonemic transcription includes the conventions of a particular language's 
-phonological rules. These rules determine the realization of that language's 
-phonemes. However, there can be different systems of phonemic transcription 
-for the same variety of a language. The differences may result from the fact 
-that more than one phonetic symbol may be appropriate for a phoneme (see Section 
-\ref{}). Or the differences may be due to different phonemic analyses, e.g.\ 
-Standard German's vowel system is arguably contrastive in length or tenseness.
-
-An important principle of the IPA is that different representations resulting 
-from different symbols and different analyses are in line with the IPA's aims. 
-In other words, the IPA does not provide phonological analyses for specific 
-languages and the IPA does not define a single ``correct'' transcription system. 
-Rather, the IPA aims to provide a resource that allows users to express any 
-analysis so that it is widely understood. Thus the IPA suits many linguists' 
-needs because:
+\noindent Yet, in some situations diacritics can be used to mark contrasts. The
+\textit{Handbook} recommends to limit the use of phonemic diacritics to the
+following situations:
 
 \begin{itemize}
-	\item it is intended to be a set of symbols for representing all possible sounds in the world's (spoken) languages
-	\item its chart has a linguistic basis (specifically a phonological bias) rather than a general phonetic notation scheme
-	\item its symbols can be used to represent distinctive feature combinations\footnote{Although the chart uses traditional manner and place of articulation labels, the symbols can nevertheless represent any defined bundle of features, binary or otherwise, to define phonetic dimensions.}
-	\item its chart provides a summary of linguists' agreed-upon phonetic knowledge (a common denominator of phonological ``facts'')
+
+ 	\item denoting length, stress and pitch;
+	\item obviating the design of a (large) number of new symbols when a 
+		  single diacritic suffices (e.g. nasalized vowels, aspirated stops).
+
 \end{itemize}
 
-Several styles of transcription with IPA are possible and the HB illustrates 
-these and notes that they are all valid (see the 29 languages and their transcriptions 
-in the original and initial \textit{Illustrations of the IPA} \citep[41--154]{IPA2007}). 
-Therefore, there are different but equivalent transcriptions, or as \cite[64]{Ladefoged1990a} 
-captures it, ``Perhaps now that the Association has been explicit in its eclectic 
-approach, outsiders to the Association will no longer speak of \textit{the} IPA 
-transcription of a given phenomenon, as if there were only one approved style.''.
+The interpretation of the IPA symbols in specific usage is not trivial. Although
+the articulatory properties of the IPA symbols themselves are rather succinctly
+summarised by the normative description in the \textit{Handbook}, it is common
+in practical applications that the transcribed symbols do not precisely
+represent the phonetic value of the sounds that they represent. So an IPA symbol
+<t> in one transcription is not always the same as an IPA <t> in another
+transcription (or even within a single transcription). The interpretation of any
+particular <t> is mostly a language-specific convention (or even
+resource-specific and possibly even context-specific), which -- unfortunately --
+is in most cases not made explicit by users of the IPA.
 
-Clearly not all phoneticians agree (or will likely ever agree of course) on all 
-aspects of the IPA or on transcription practices. As noted above, there have 
-been several revisions in the IPA's long history, but the current version (2005) 
-is strikingly similar to the 1926 version. In 1989 an IPA revision convention 
-was held in Kiel, Germany. As per other revisions, there was expansion and revisions 
-to phonetic symbols in the IPA chart. Notably the marking of tone was extended 
-with the addition of a second system for marking linguistic tones (Chao tones). 
-Importantly, however, this was the first revision to address issues of 
-computational representations for the IPA -- the principles above of which 
-have had several ramifications that make the interoperability of electronic 
-linguistic data extremely difficult.
+There are different reasons for this difficulty to interpret IPA symbols,
+both officially part of the IPA specifications. Basically, the IPA allows for
+both a `narrow' phonetic transcription and a `broad' phonological transcription.
+An important principle of the IPA is that different representations resulting
+from different underlying analyses are allowed. Because the IPA does not provide
+phonological analyses for specific languages, the IPA does not define a single
+``correct'' transcription system. Rather, the IPA aims to provide a resource
+that allows users to express any analysis so that it is widely understood.
 
+A `narrow' phonetic transcription may freely use the all symbols in the IPA
+chart with direct reference to their phonetic value, i.e. the transcriber can
+indicate with the symbols <ŋ͡m> that the phonetic value of the attested sound is
+a simultaneous labial and velar closure which is voiced and contains nasal
+airflow, independently of the phonemic status of this sound in the language in
+question.
+
+% examples from Wells \url{https://www.phon.ucl.ac.uk/home/wells/transcription-ELL.pdf}
+
+In contrast, the basic goal of a phonemic transcription is to distinguish all
+words in a language with a minimal number of transcription symbols
+\citep[19]{Abercrombie1964}. A phonemic transcription includes the conventions
+of a particular language's phonological rules. These rules determine the
+realization of that language's phonemes. For example, in the transcription of
+German, Dutch, English and French a symbol <t> might be used for the voiceless
+plosive in the alveolar area. This symbol is sufficient for a succinct
+transcription of these languages because there is no further phonemic
+subdivisions in this domain in either of these languages. However, the
+language-specific realization of this consonant is closer to [t̪ʰ], [t], [tʰ]
+and [t̪], respectively. Similarly, the five vowels of Greek can be represented
+phonemically in IPA as /ieaou/, though phonetically they are closer to [iεauɔ].
+Similarly, the Japanese five vowel system can also be transcribed in IPA as
+/ieaou/, while the phonetic targets in this case are closer to [ieaoɯ].
+
+Note also that there can be different systems of phonemic transcription for the
+same variety of a language, so two different resources on the `same' language
+might use different symbols that represent the same sound. The differences may
+result from the fact that more than one phonetic symbol may be appropriate for a
+phoneme, or the differences may be due to different phonemic analyses, e.g.\
+Standard German's vowel system is arguably contrastive in length, tenseness or
+vowel quality. Finally, even within a single phonemic transcription a specific
+symbol can have different realizations because of allophonic variation which 
+might not be explicitly transcribed.
+
+In sum, there are three different reasons why phonemically-based IPA 
+transcription is difficult to interpret:
+
+\begin{itemize}
+
+   \item A symbol used represents the phonemic status, and not necessarily the
+         precise phonetic realization. So, different transcriptions might use 
+         the same symbol for different underlying sounds.
+   \item The symbol that is used for a specific phoneme is not necessarily
+         unique, so different transcriptions might use different symbols for the
+         same underlying sound.
+   \item Allophonic variation can be disregarded in phonemic transcription, so
+         the same symbol within a single transcription can represent different
+         sounds.
+
+\end{itemize}
+
+Ideally, all such implicit conventions of a phonemic transcription would be
+explicitly codified. This could very well be performed by using an orthography
+profile (see Chapter~\ref{orthography-profiles}), linking the selected phonemic 
+transcription symbols to `narrow' phonetic transcriptions, possibly also including 
+specifications of contextual interpretation.
+
+% ==========================
 \section{Computational representation}
+% ==========================
+
+In 1989 an IPA revision convention was held in Kiel, Germany. As per other
+revisions, there was expansion and revisions to phonetic symbols in the IPA
+chart. Notably the marking of tone was extended with the addition of a second
+system for marking linguistic tones (Chao tones). Importantly, however, this was
+the first revision to address issues of computational representations for the
+IPA -- the principles above of which have had several ramifications that make
+the interoperability of electronic linguistic data extremely difficult.
 

### 2015-11-15 book/chapters/ipa_meets_unicode.tex
revision of first two sections of IPA introduction

@@
          post-aspiration (following the base character) and as
-         <\diareverse{02B0}> for pre-aspiration (preceding the base character).
-         Note that there is yet another homoglyph, namely the
-         \textsc{combining latin small letter h} at \uni{036A}. We propose not
-         to use this combining diacritical mark, but to standardize on Unicode
-         modifier letters for all superscript letters in IPA.
+         <\diareverse{02B0}> for pre-aspiration (preceding the base
+         character), so there is no consistent direction in which this
+         diactritic should bind. Note that there is yet another homoglyph,
+         namely the \textsc{combining latin small letter h} at \uni{036A}.
+         We propose not to use this combining diacritical mark, but to
+         standardize on Unicode modifier letters for all superscript
+         letters in IPA.
 

### 2015-11-15 book/chapters/ipa_background.tex
readability update

@@
 
-Established in 1886, the International Phonetic Association (henceforth
-Association) has long maintained a standard alphabet, the International Phonetic
-Alphabet (IPA), which is a common standard in linguistics to transcribe sounds
-of spoken languages. It was first published in 1888 as an international system
-of phonetic transcription for oral languages and for pedagogical purposes. It
-contained phonetic values for English, French and German. Diacritics for length
-and nasalization were already present in this first version, and the same
-symbols are still used today. 
+Established in 1886, the \textit{International Phonetic Association} (henceforth
+\textit{Association}) has long maintained a standard alphabet, \textit{the
+International Phonetic Alphabet} (henceforth \textit{IPA}), which is a common
+standard in linguistics to transcribe sounds of spoken languages. It was first
+published in 1888 as an international system of phonetic transcription for oral
+languages and for pedagogical purposes. It contained phonetic values for
+English, French and German. Diacritics for length and nasalization were already
+present in this first version, and the same symbols are still used today. 
 %\footnote{Also referred to as API, for \textit{Association Phonétique Internationale}.}
@@
 from words in different languages. In 1900 the symbols were first organized into
-a chart and were given phonetic feature labels for consonants (e.g.\ for manner
-of articulation among others `plosives', `nasales', `fricatives' and for place
-of articulation among others `bronchiales', `laryngales', `labiales') and for
-vowels (e.g.\ `fermées', `mi-fermées', `mi-ouvertes', `ouvertes'). Throughout
-the last century, the structure of the chart has changed with increases in
-phonetic knowledge. Thus, similar to notational systems in other scientific
-disciplines, the IPA reflects facts and theories of phonetic knowledge that have
-developed over time. It is natural then that the IPA is modified occasionally to
+a chart and were given phonetic feature labels (e.g.\ for manner of consonant
+articulation among others `plosives', `nasales', `fricatives', for place of
+articulation among others `bronchiales', `laryngales', `labiales' and for vowels
+e.g.\ `fermées', `mi-fermées', `mi-ouvertes', `ouvertes'). Throughout the last
+century, the structure of the chart has changed with increases in phonetic
+knowledge. Thus, similar to notational systems in other scientific disciplines,
+the IPA reflects facts and theories of phonetic knowledge that have developed
+over time. It is natural then that the IPA is modified occasionally to
 accommodate scientific innovations and discoveries. In fact, this is part of the
@@
 Association: A guide to the use of the International Phonetic Alphabet}
-\citep[159]{IPA1999}, henceforth simply called \textit{Handbook}.
+\citep{IPA1999}, henceforth simply called \textit{Handbook}.
 
@@
 	\item it is intended to be a set of symbols for representing all possible
-       sounds in the world's (spoken) languages
+       sounds in the world's (spoken) languages;
 	\item its chart has a linguistic basis (specifically a phonological bias)
-       rather than a general phonetic notation scheme
+       rather than a general phonetic notation scheme;
 	\item its symbols can be used to represent distinctive feature
-       combinations\footnote{Although the chart uses traditional manner and
-       place of articulation labels, the symbols can nevertheless represent any
-       defined bundle of features, binary or otherwise, to define phonetic
-       dimensions.}
-	\item its chart provides a summary of linguists' agreed-upon phonetic knowledge
-       (a common denominator of phonological ``facts'')
+       combinations;\footnote{Although the chart uses traditional manner and
+       place of articulation labels, the symbols can be used as a representation
+       of any defined bundle of features, binary or otherwise, to define
+       phonetic dimensions.}
+	\item its chart provides a summary of linguists' agreed-upon phonetic 
+	knowledge.
 
@@
 
-Several styles of transcription with IPA are possible, as illustrated in the
-\textit{Handbook}, and they are all valid (see the 29 languages and their
-transcriptions in the original and initial \textit{Illustrations of the IPA}
-\citep[41--154]{IPA2007}). Therefore, there are different but equivalent
-transcriptions, or as \citet[64]{Ladefoged1990a} captures it, ``Perhaps now that
+Several styles of transcription with the IPA are possible, as illustrated in the
+\textit{Handbook}, and they are all valid.\footnote{For illustration of
+different possibilities, see the 29 languages and their transcriptions in the
+original and initial \textit{Illustrations of the IPA}
+\citep[41--154]{IPA2007}} Therefore, there are different but equivalent
+transcriptions, or as \citet[64]{Ladefoged1990a} captures it, ``perhaps now that
 the Association has been explicit in its eclectic approach, outsiders to the
 Association will no longer speak of \textit{the} IPA transcription of a given
-phenomenon, as if there were only one approved style.'' Clearly not all phoneticians agree (or will likely ever agree of course) on all 
-aspects of the IPA or on transcription practices. As noted above, there have 
-been several revisions in the IPA's long history, but the current version (2005) 
-is strikingly similar to the 1926 version, showing the viability of the IPA as a 
+phenomenon, as if there were only one approved style.'' Clearly not all
+phoneticians agree (or will likely ever agree of course) on all aspects of the
+IPA or on transcription practices. As noted above, there have been several
+revisions in the IPA's long history, but the current version (2005) is
+strikingly similar to the 1926 version, showing the viability of the IPA as a
 common standard for linguistic transcription.
@@
 Any IPA transcription is based on two premises: (i) that it is possible to
-describe the acoustic speech signal (i.e.\ sound waves) in terms of sequentially
+describe the acoustic speech signal (sound waves) in terms of sequentially
 ordered discrete segments, and, (ii) that each segment can be characterized by
-an articulatory target.\footnote{A purely phonetic description is only derivable
-from instrumental data in high quality sound recordings. Once spoken language
-data are segmented, phonological consideration inextricably play a roll in
-transcription. In other words, phonetic observations beyond quantitative
-acoustic analysis are always made in terms of some phonological framework.}
+an articulatory target.
+
+Once spoken language data are segmented, the IPA provides symbols to
+unambiguously represent phonetic details. However, since phonetic detail could
+potentially include anything, e.g. something like `deep voice', the IPA
+restricts phonetic detail to linguistically relevant aspects of speech.
+Phonological consideration thus inextricably play a roll in transcription. In
+other words, phonetic observations beyond quantitative acoustic analysis are
+always made in terms of some phonological framework.
+
 Today, the IPA chart reflects a linguistic theory grounded in principles of
@@
 
-Although the IPA provides symbols to unambiguously represent phonemes, it also
-aims to represent phonetic details. Since phonetic detail could potentially
-include anything, e.g. something like `deep voice', the IPA restricts phonetic
-detail to linguistically relevant aspects of speech. From its earliest days the
-Association aimed to provide ``a separate sign for each distinctive sound; that
-is, for each sound which, being used instead of another, in the same language,
-can change the meaning of a word'' \citep[27]{IPA1999}. Distinctive sounds
-became later known as \textsc{phonemes} and the IPA has developed historically
-into a notational devise with a strictly segmental phonemic view. A phoneme is
-an abstract theoretical notion derived from an acoustic signal as produced by
-speakers in the real world. Therefore the IPA contains a number of theoretical
-assumptions about speech and how to analyze speech in written form.
+From its earliest days the Association aimed to provide ``a separate sign for
+each distinctive sound; that is, for each sound which, being used instead of
+another, in the same language, can change the meaning of a word''
+\citep[27]{IPA1999}. Distinctive sounds became later known as \textsc{phonemes}
+and the IPA has developed historically into a notational devise with a strictly
+segmental phonemic view. A phoneme is an abstract theoretical notion derived
+from an acoustic signal as produced by speakers in the real world. Therefore the
+IPA contains a number of theoretical assumptions about speech and how to
+transcribe speech in written form.
 
@@
 First, the symbols of IPA crucially distinguish between letter-like symbols and
-diacritics.\footnote{see Section~\ref{pitfall-different-notions-of-diacritics} for a more
-detailed discussion about the notion of diacritics} The use of the letter-like
-symbols to represent a language's sounds is guided by the principle of contrast;
-where two words are distinguishable by phonemic contrast, those contrasts should
-be transcribed with different letter symbols (not diacritics). Allophonic
-distinction falls under the rubric of diacritically-distinguished symbols, e.g.
-the difference in English between and unrealized /p/ in [stop̚] vs. an aspirated
-/p/ in [spʰot] should be marked with a diacritic. In sum:
+diacritics. The use of the letter-like symbols to represent a language's sounds
+is guided by the principle of contrast; where two words are distinguishable by
+phonemic contrast, those contrasts should be transcribed with different letter
+symbols (not diacritics). Allophonic distinction falls under the rubric of
+diacritically-distinguished symbols, e.g. the difference in English between and
+unrealized /p/ in [stop̚] vs. an aspirated /p/ in [spʰot] should be marked with
+a diacritic. In sum:
 
@@
 
-There are different reasons for this difficulty to interpret IPA symbols,
-both officially part of the IPA specifications. Basically, the IPA allows for
-both a `narrow' phonetic transcription and a `broad' phonological transcription.
-An important principle of the IPA is that different representations resulting
-from different underlying analyses are allowed. Because the IPA does not provide
-phonological analyses for specific languages, the IPA does not define a single
-``correct'' transcription system. Rather, the IPA aims to provide a resource
-that allows users to express any analysis so that it is widely understood.
+There are different reasons for this difficulty to interpret IPA symbols, all
+officially sanctioned by the IPA. An important principle of the IPA is that
+different representations resulting from different underlying analyses are
+allowed. Because the IPA does not provide phonological analyses for specific
+languages, the IPA does not define a single ``correct'' transcription system.
+Rather, the IPA aims to provide a resource that allows users to express any
+analysis so that it is widely understood.
 
-A `narrow' phonetic transcription may freely use the all symbols in the IPA
-chart with direct reference to their phonetic value, i.e. the transcriber can
-indicate with the symbols <ŋ͡m> that the phonetic value of the attested sound is
-a simultaneous labial and velar closure which is voiced and contains nasal
-airflow, independently of the phonemic status of this sound in the language in
-question.
+Basically, the IPA allows for both a `narrow' phonetic transcription and a
+`broad' phonological transcription. A `narrow' phonetic transcription may freely
+use the all symbols in the IPA chart with direct reference to their phonetic
+value, i.e. the transcriber can indicate with the symbols <ŋ͡m> that the
+phonetic value of the attested sound is a simultaneous labial and velar closure
+which is voiced and contains nasal airflow, independently of the phonemic status
+of this sound in the language in question.
 

### 2015-11-15 book/chapters/introduction.tex
minor changes

--- 2015-06-20 chapters/introduction.tex
@@
 ISO/IEC~8859 used the 7-bit ASCII as the basis for each code page, and defined
-16 different 7-bit extensions, leading to $(1+16)\cdot{2^7} = 2,176$ possible
+16 different 7-bit extensions, leading to $(1+16) \cdot {2^7} = 2,176$ possible
 characters. However, because of overlap and not-assigned codes points the actual
@@
 international code sets. In the United States, a group of computer scientists
-formed the \textsc{Unicode Consortium}, proposing a 16-bit encoding in 1991
+formed the \textsc{unicode consortium}, proposing a 16-bit encoding in 1991
 (with $2^{16} = 65,536$ different possible characters). At the same time in
-Europe, the \textsc{International Organization for Standardization} (ISO) was
+Europe, the \textsc{international organization for standardization} (ISO) was
 working on ISO~10646 to supplant the ISO/IEC~8859 standard. Their first draft of
-the \textsc{Universal Character Set} (UCS) in 1990 was 31-bit (with
+the \textsc{universal character set} (UCS) in 1990 was 31-bit (with
 theoretically $2^{31} = 2,147,483,648$ possible characters, but because of some
@@
 
-\noindent In the next section we provide a very brief overview of the linguistic terminology
-concerning writing systems before turning to the slightly different
-computational terminology in the next section on the Unicode Standard.
+\noindent In the next section we provide a very brief overview of the linguistic
+terminology concerning writing systems before turning to the slightly different
+computational terminology in the subsequent section on the Unicode Standard.
 

### 2015-11-15 book/chapters/ipa_background.tex
minor changes

@@
 
-Established in 1886, the \textit{International Phonetic Association} (henceforth
-\textit{Association}) has long maintained a standard alphabet, \textit{the
-International Phonetic Alphabet} (henceforth \textit{IPA}), which is a common
-standard in linguistics to transcribe sounds of spoken languages. It was first
-published in 1888 as an international system of phonetic transcription for oral
-languages and for pedagogical purposes. It contained phonetic values for
+Established in 1886, the \textsc{international phonetic association} (henceforth
+\textit{Association}) has long maintained a standard alphabet, the
+\texsct{international phonetic alphabet} (henceforth \textit{IPA}), which is a
+common standard in linguistics to transcribe sounds of spoken languages. It was
+first published in 1888 as an international system of phonetic transcription for
+oral languages and for pedagogical purposes. It contained phonetic values for
 English, French and German. Diacritics for length and nasalization were already
@@
 phonological systems. The current construction and use of the IPA are guided by
-principles outlined in the \textit{Handbook of the International Phonetic
-Association: A guide to the use of the International Phonetic Alphabet}
-\citep{IPA1999}, henceforth simply called \textit{Handbook}.
+principles outlined in the \textsc{handbook of the international phonetic
+association} \citep{IPA1999}, henceforth simply called \textit{Handbook}.
 
@@
 chart. Notably the marking of tone was extended with the addition of a second
-system for marking linguistic tones (Chao tones). Importantly, however, this was
-the first revision to address issues of computational representations for the
-IPA -- the principles above of which have had several ramifications that make
-the interoperability of electronic linguistic data extremely difficult.
+system for marking linguistic tones (the so-called `Chao' tone letters).
+Importantly, however, this was also the first revision to address issues of
+computational representations for the IPA -- the principles above of which have
+had several ramifications that make the interoperability of electronic
+linguistic data extremely difficult.
 

### 2015-11-16 book/chapters/introduction.tex
minor revision intro

@@
 \citep[44]{Olumuyiw2013}. However, writing systems (and the languages they
-represent) are normally full of homophones, which is normally not a problem at
-all for speakers of the language. More importantly, writing is not just a purely
-functional tool, but just as importantly it is a mechanism to signal social
-affiliation. By showing that you \textit{know the rules} of expressing yourself
-in writing, others will more easily accept you as a worthy participant in their
-group. And that just as well holds for obeying to the official rules when
-writing a job application, as for obeying to the informal rules when writing an
-SMS to classmates in school. The case of Yoruba writing is an exemplary case, as
-even after more than a century of efforts to standardize the writing systems,
-there is still a wide range of variation in daily use \citep{Olumuyiw2013}.
+represent) are normally full of homographs (and homophones), which is normally
+not a problem at all for speakers of the language. More importantly, writing is
+not just a purely functional tool, but just as importantly it is a mechanism to
+signal social affiliation. By showing that you \textit{know the rules} of
+expressing yourself in writing, others will more easily accept you as a worthy
+participant in their group. And that just as well holds for obeying to the
+official rules when writing a job application, as for obeying to the informal
+rules when writing an SMS to classmates in school. The case of Yoruba writing is
+an exemplary case, as even after more than a century of efforts to standardize
+the writing systems, there is still a wide range of variation of writing in
+daily use \citep{Olumuyiw2013}.
 
-The sometimes cumbersome and sometimes illogical structure, and the enormous
-variability of existing writing systems is a fact of life scholars have to
-accept and should try to adapt to as good as possible. Our plea here is a
-proposal for a formalization to do exactly that.
+\subsubsection*{Formalizing orthographic structure}
 
-When considering the worldwide linguistic diversity, including all
+The resulting cumbersome and often illogical structure of writing systems, and
+the enormous variability of existing writing systems for the world's languages
+is a fact of life scholars have to accept and should try to adapt to as good as
+possible. Our plea in this book is a proposal how to do exactly that: formalize 
+knowledge about individual writing systems in a form that is both easy to use 
+for linguists in daily practice, and at the same time computer-readable for 
+automatic processing.
+
+When considering the worldwide linguistic diversity, including the many
 lesser-studied and endangered languages, there exist numerous different
@@
 arising in the daily practice of linguistics. Consequently, a central goal of
-this paper is to flesh out the linguistic challenges for locale descriptions,
+this book is to flesh out the linguistic challenges for locale descriptions,
 and work out suggestions to improve their structure for usage in a linguistic
@@
 The necessity to computationally use and compare different orthographies most
-forcefully arises in the context of language comparison. Concretely, in our
-current research our goal is to develop quantitative methods for language
-comparison and historical analysis in order to investigate worldwide linguistic
-variation and to model the historical and areal processes that underlie
-linguistic diversity,
-cf.~\citet{Steiner_etal2011,List2012,List2012a,ListMoran2013,MoranProkic2013}.
+forcefully arises in the context of language comparison. Concretely, the proper
+processing of orthographies and transcriptions becomes crucial for the
+development of quantitative methods for language comparison and historical
+reconstruction in order to investigate worldwide linguistic variation and to
+model the historical and areal processes that underlie linguistic
+diversity.\footnote{
+cf.~\citet{Steiner_etal2011,List2012,List2012a,ListMoran2013,MoranProkic2013}}
 In this work, it is crucial to be able to flexibly process across numerous
 resources with different orthographies. In many cases even different resources
-on the \textit{same} language use different orthographic conventions. Another
+on the same language use different orthographic conventions. Another
 orthographic challenge that we encounter regularly in our linguistic practice is
@@
 orthographic conventions used in a particular source document. Our experience is
-that such a graphemic tokenization can be performed without extensive in-depth
-knowledge about the phonetic and phonological details of the language in
-question. For example, the specification that $<$ou$>$ is a grapheme of English
-is a much easier task than to specify what exactly the phonetic values of this
-grapheme are in any specific occurrence in English words. Grapheme separation is
-a task that can be performed relatively reliably and with limited availability
-of time and resources (compare, for example, the task of creating a complete
-phonetic or phonological normalization).
+that such a graphemic tokenization can often be performed reasonably accurately
+without extensive in-depth knowledge about the phonetic and phonological details
+of the language in question. For example, the specification that <ou> is a
+grapheme of English is a much easier task than to specify what exactly the
+phonetic values of this grapheme are in any specific occurrence in English
+words. Grapheme separation is a task that can be performed relatively reliably
+and with limited availability of time and resources (compare, for example, the
+daunting task of creating a complete phonetic or phonological normalization).
 
@@
 
-Before we deal with these proposals, in the first part of this paper (Sections
-~\ref{encoding} through~\ref{ipa-meets-unicode}) we give an extended introduction
-to the notion of encoding (Section~\ref{encoding}) and writing systems, both
-from a linguistic perspective and from the perspective of the Unicode Consortium
-(Section~\ref{terminology}). We consider the Unicode Standard to be a
-breakthrough (and ongoing) development that fundamentally changed the way we
-look at writing systems, and we aim to provide here a slightly more in-depth
-survey of the many techniques that are available in the standard. A good
-appreciation for the solutions that the Unicode Standard also allows for a
-thorough understanding of possible pitfalls that one might encounter when using
-it (Section~\ref{unicode-pitfalls}). As an example of the current
-state-of-the-art, we discuss the rather problematic marriage of the
-International Phonetic Alphabet (IPA) with the Unicode Standard (Section
+\subsubsection*{Structure of this bok}
+
+Before we deal with these proposals we will extensively discuss the theoretical
+background on text encoding, on the Unicode standard, and on the International
+Phonetic Alphabet. In the remainder of this chapter, we give an extended
+introduction to the notion of encoding (Section~\ref{encoding}) and writing
+systems, both from a linguistic perspective
+(Section~\ref{linguistic-terminology}) and from the perspective of the Unicode
+Consortium (Section~\ref{the-unicode-approach}). We consider the Unicode
+Standard to be a breakthrough (and ongoing) development that fundamentally
+changed the way we look at writing systems, and we aim to provide here a
+slightly more in-depth survey of the many techniques that are available in the
+standard. A good appreciation for the solutions that the Unicode Standard also
+allows for a thorough understanding of possible pitfalls that one might
+encounter when using it (Chapter~\ref{unicode-pitfalls}). As an example of the
+current state-of-the-art, we discuss the rather problematic marriage of the
+International Phonetic Alphabet (IPA) with the Unicode Standard (Chapter
 \ref{ipa-meets-unicode}).
 
-The second part of the paper (Sections~\ref{orthography-profiles} and
+The second part of the book (Sections~\ref{orthography-profiles} and
 \ref{use-cases}) describes our proposals for how to deal with the Unicode
-Standard in the daily practice of (comparative) linguists. First, we discuss the
+Standard in the daily practice of (comparative) linguistics. First, we discuss the
 challenges of characterizing a writing system. To solve these problems, we
 propose the notions of orthography profiles, closely related to Unicode locale
-descriptions (Section~\ref{orthography-profiles}). Finally, we discuss practical
-issues with actual examples (Section~\ref{use-cases}). We provide reference
+descriptions (Chapter~\ref{orthography-profiles}). Finally, we discuss practical
+issues with actual examples (Chapter~\ref{use-cases}). We provide reference
 implementation of our proposals in R and in Python, available as open-source

### 2015-12-10 book/chapters/ipa_background.tex
typo

@@
 \textit{Association}) has long maintained a standard alphabet, the
-\texsct{international phonetic alphabet} (henceforth \textit{IPA}), which is a
+\texsc{international phonetic alphabet} (henceforth \textit{IPA}), which is a
 common standard in linguistics to transcribe sounds of spoken languages. It was
@@
 	\item its chart has a linguistic basis (specifically a phonological bias)
-       rather than a general phonetic notation scheme;
+       rather than just being a general phonetic notation scheme;
 	\item its symbols can be used to represent distinctive feature

### 2015-12-10 book/chapters/pitfalls.tex
revision of first two sections of IPA introduction
section title change

--- 2015-09-28 chapters/pitfalls.tex
@@
 % ==========================
-\section{Wrong it is not}
+\section{Wrong it ain't}
 \label{wrong-it-is-not}
@@
 \textsc{latin small letter g} at \uni{0067}). In contrast, the symbols
-<{\fontspec{EversonMono}AАΑᎪᗅᴀꓮʠ햠홰}> are all different code points,
+<{\fontspec{EversonMono}AАΑᎪᗅᴀꓮ햠홰}> are all different code points,
 although they look highly similar---in some cases even sharing exactly the same

### 2015-12-10 book/chapters/ipa_background.tex
revise IPA background

@@
 \textit{Association}) has long maintained a standard alphabet, the
-\texsc{international phonetic alphabet} (henceforth \textit{IPA}), which is a
+\textsc{international phonetic alphabet} (henceforth \textit{IPA}), which is a
 common standard in linguistics to transcribe sounds of spoken languages. It was
@@
 
-Originally, the IPA alphabet was a list of symbols with pronunciation examples
-from words in different languages. In 1900 the symbols were first organized into
-a chart and were given phonetic feature labels (e.g.\ for manner of consonant
+Originally, the IPA was a list of symbols with pronunciation examples
+using words in different languages. In 1900 the symbols were first organized into
+a chart and were given phonetic feature labels (e.g.\ for manner of
 articulation among others `plosives', `nasales', `fricatives', for place of
@@
 
-From its earliest days the Association aimed to provide ``a separate sign for
+\noindent From its earliest days the Association aimed to provide ``a separate sign for
 each distinctive sound; that is, for each sound which, being used instead of
@@
 
-\noindent Yet, in some situations diacritics can be used to mark contrasts. The
-\textit{Handbook} recommends to limit the use of phonemic diacritics to the
-following situations:
+\noindent Yet, in some situations diacritics can be used to mark phonemic
+contrasts. The \textit{Handbook} recommends to limit the use of phonemic
+diacritics to the following situations:
 
@@
 particular <t> is mostly a language-specific convention (or even
-resource-specific and possibly even context-specific), which -- unfortunately --
-is in most cases not made explicit by users of the IPA.
+resource-specific and possibly even context-specific), a fact which --
+unfortunately -- is in most cases not made explicit by users of the IPA.
 
@@
 Rather, the IPA aims to provide a resource that allows users to express any
-analysis so that it is widely understood.
-
-Basically, the IPA allows for both a `narrow' phonetic transcription and a
+analysis so that it is widely understood. Basically, the IPA allows for both a `narrow' phonetic transcription and a
 `broad' phonological transcription. A `narrow' phonetic transcription may freely
@@
 
-In contrast, the basic goal of a phonemic transcription is to distinguish all
+In contrast, the basic goal of a phonemic ``broad'' transcription is to distinguish all
 words in a language with a minimal number of transcription symbols
@@
 
-   \item A symbol used represents the phonemic status, and not necessarily the
+   \item A symbol represents the phonemic status, and not necessarily the
          precise phonetic realization. So, different transcriptions might use 
          the same symbol for different underlying sounds.
-   \item The symbol that is used for a specific phoneme is not necessarily
+   \item Any symbol that is used for a specific phoneme is not necessarily
          unique, so different transcriptions might use different symbols for the
@@
 Importantly, however, this was also the first revision to address issues of
-computational representations for the IPA -- the principles above of which have
-had several ramifications that make the interoperability of electronic
-linguistic data extremely difficult.
+computational representations for the IPA.
 
 Prior to the Kiel Convention for the modern revision of the IPA in 1989,
-\cite{Wells1987} collected and published practical approaches to coded 
-representations of the IPA, which dealt mainly with the assignment of 
-characters on the keyboard. The process of assigning standardized ``computer 
-codes'' to phonetic symbols was assigned to the Workgroup on Computer 
-Coding (henceforth working group) at the Kiel Convention. This working 
-group was tasked with \citep{Esling1990,EslingGaylord1993}:
+\citet{Wells1987} collected and published practical approaches to coded
+representations of the IPA, which dealt mainly with the assignment of characters
+on the keyboard to IPA symbols. The process of assigning standardized ``computer
+codes'' to phonetic symbols was assigned to the \textsc{workgroup on computer
+coding} (henceforth \textit{working group}) at the Kiel Convention. This working
+group was assigned to consider the following tasks
+\citep{Esling1990,EslingGaylord1993}:
 
@@
 a first step in formalizing the IPA computationally because it would give 
-each symbol an unambiguous numerical identifier called an \textsc{IPA Number}. 
+each symbol an unambiguous numerical identifier called an \textsc{IPA number}. 
 The numbering system was to be comprehensive enough to support future revisions 
@@
 Although the Association had never officially approved a set of names 
-for the IPA symbols, each IPA symbol received a unique \textsc{IPA Name}. 
+for the IPA symbols, each IPA symbol received a unique \textsc{IPA name}. 
 Many symbols already had an informal name (or two) used by linguists, but 
@@
 \textit{Phonetic Symbol Guide} \citep{PullumLadusaw1986}. Thus most of the 
-IPA symbol names were taken from \cite{PullumLadusaw1986} \citep[31]{IPA1999}.
+IPA symbol names were taken from that source \citep[31]{IPA1999}.
 
@@
 i.e.\ the IPA Number was not meant to be implemented directly in a computer 
-encoding. The working group report's declaration includes the explanatory 
+encoding. The working group report's declaration includes the following explanatory 
 remarks \citep[82]{International1989report}:
@@
 computers for their research and many were quickly adopting electronic mail or 
-discussion boards like Usenet as a medium for international exchanges. This was 
-before the Internet as we know it today. Most of these systems ran on 8-bit 
+discussion boards like Usenet as a medium for international exchanges. Most of these systems ran on 8-bit 
 hardware systems using a 7-bit ASCII character encoding.}
@@
 Furthermore, the assignment of computer codes to IPA symbols was meant to
-represent an unbiased formulation. The Association plays the role of an international 
-advisory body and it stated that it should not recommend a particular existing 
-system of encoding. In fact, during this time there were a number of coding 
-systems used, but none of them had a dominant international position. The 
-differences between systems were also either too great or too subtle to warrant 
-an attempt at combining them \citep{International1989report}.
+represent an unbiased formulation. The Association here played the role of an
+international advisory body and it stated that it should not recommend a
+particular existing system of encoding. In fact, during this time there were a
+number of coding systems used (see Section~\ref{encoding}), but none of them had
+a dominant international position. The differences between systems were also
+either too great or too subtle to warrant an attempt at combining them
+\citep{International1989report}.
 
-The working group assigned each IPA symbol to a unique three-digit number, 
-i.e.\ an IPA Number. Encoded in this number scheme is information about the 
-status of each symbol (see below). The IPA numbers are listed with the IPA 
-symbols and they are also illustrated in IPA chart form (see \cite[84]{EslingGaylord1993} 
-or \cite[App. 2]{IPA2007}). The numbers were assigned in linear order (e.g.\ 
-[p] 101, [b] 102, [t] 103...) following the IPA revision of 1989 and its update in 1996.
+The working group assigned each IPA symbol to a unique three-digit number, i.e.\
+an IPA Number. Encoded in this number scheme implicitly includes information
+about the status of each symbol (see below). The IPA numbers were listed with the
+IPA symbols and they were also illustrated in IPA chart form (see
+\cite[84]{EslingGaylord1993} or \cite[App. 2]{IPA2007}). The numbers were
+assigned in linear order (e.g.\ [p] 101, [b] 102, [t] 103...) following the IPA
+revision of 1989 and its update in 1996. Although the numbering scheme still 
+exists, in practice is it superseded by the Unicode codification of symbols.
 
@@
 or reintroduced into current or future usage. For example, in the 1989 
-revision voiceless implosives < ƥ, ƭ, ƈ, ƙ, ʠ > were added; in the 1993 
-revision they were removed. Ligatures like < ʧ, ʤ > are included as formerly 
+revision voiceless implosives <~ƥ,~ƭ,~ƈ,~ƙ,~ʠ~> were added; in the 1993 
+revision they were removed. Ligatures like <~ʧ,~ʤ~> are included as formerly 
 recognized IPA symbols; they are assigned to the ``200 series'' of IPA numbers 
@@
 an IPA Name for reference purposes. As we discuss below, this decision is 
-later reflected in the Unicode Standard; many legacy IPA symbols reside in 
-the \textsc{IPA Extensions} block.
+later reflected in the Unicode Standard as many legacy IPA symbols still reside in 
+the \textsc{IPA extensions} block.
 
-The IPA Number is simply expressed as a ``three-digit number numerical 
-directory of digit triples'' \cite{}.\footnote{For practical purposes, 
-the IPA Number also served as a typesetter's guide to the IPA chart.} 
-The numbering scheme specifies three-digit codes, the first digit of which 
-indicates the symbol's category \cite{Esling1990,EslingGaylord1993}:
+The IPA Number is simply expressed as a ``three-digit number numerical directory
+of digit triples'' \cite{}. The numbering scheme specifies three-digit codes,
+the first digit of which indicates the symbol's category
+\citep{Esling1990,EslingGaylord1993}:
 
@@
 
-After a symbol is categorized, it is assigned a number sequentially, 
-e.g.\ [i] 301, [e] 302, [ɛ] 303. The system allows for the addition 
-of new symbols within the various series by appending them, e.g. [ⱱ] 
-184. Former or often used but non-IPA symbols for consonants, vowels 
-and diacritics are numbered from x99 backwards. For example, the voiceless 
-and voiced postalveolar affricates and fricatives < č, ǰ, š, ž > are 
-assigned the IPA numbers 299, 298, 297 and 296, respectively, because 
-they are not sanctioned IPA symbols.
+After a symbol is categorized, it is assigned a number sequentially, e.g.\ [i]
+301, [e] 302, [ɛ] 303. The system allows for the addition of new symbols within
+the various series by appending them, e.g. [\charis{ⱱ}] 184. Former non-IPA
+symbols (or often-used but non-offical IPA symbols) for consonants, vowels and
+diacritics are numbered from x99 backwards. For example, the voiceless and
+voiced postalveolar affricates and fricatives <~č,~ǰ,~š,~ž~> are assigned the
+IPA numbers 299, 298, 297 and 296, respectively, because they are not sanctioned
+IPA symbols.
 
-The assignment of the IPA numbers to IPA symbols provided the basis for 
-uniquely identifying the set of past and present IPA symbols as a type of 
-computational representational standard of the IPA. Within each revision 
-of the IPA, the coding defines a closed and clearly defined set of characters. 
-The benefits of this standardization are clear in at least two ways: it is 
-used in translation tables that reference ASCII representations of the IPA, 
-and this early computational representation of the IPA became the basis for 
-its inclusion into the Unicode Standard version 1.0.
+The assignment of the IPA numbers to IPA symbols provided the basis for uniquely
+identifying the set of past and present IPA symbols as a type of computational
+representational standard of the IPA. Within each revision of the IPA, the
+coding defines a closed and clearly defined set of characters. The benefits of
+this standardization are clear in at least two ways: it is used in translation
+tables that reference ASCII representations of the IPA, and this early
+computational representation of the IPA became the basis X-SAMPA and for the
+inclusion of the IPA into the Unicode Standard version 1.0.
 
+% ==========================
 \section{SAMPA and X-SAMPA}
+% ==========================
 
@@
 the need for a universal computing environment for writing systems and their
-computational complexity is discussed in \citet{Simons1989}. A survey of
+computational complexity is discussed in \citet{Simons1989}. A more recent survey of
 practical recommendations for language resources, including notes on encoding,
@@
 One solution was the ASCII-ification of the IPA, which simply involved 
-defining keyboard-able sequences as IPA symbol codings.\footnote{\cite{Wells1987} 
-provides an in-depth description of IPA codings from country-to-country. 
-Later ASCII-IPAs include Kirshenbaum (created in 1992 in a Usenet group and 
-named after its lead developer who was at Hewlett-Packard Laboratories) and 
-Worldbet (published in 1993 by \cite{Hieronymus1993}, who was at AT\&T Laboratories).} 
-A successful effort was SAMPA (Speech Assessment
+defining keyboard-able sequences consisting of ASCII combinations as an IPA symbol
+codings. For example, \cite{Wells1987} provides an in-depth description of IPA
+codings from country-to-country. Later ASCII-IPAs include Kirshenbaum (created
+in 1992 in a Usenet group and named after its lead developer who was at
+Hewlett-Packard Laboratories) and Worldbet (published in 1993 by
+\cite{Hieronymus1993}, who was at AT\&T Laboratories). 
+The most successful effort was SAMPA (Speech Assessment
 Methods Phonetic Alphabet), which was created between 1988--1991 in Europe to 
 represent IPA symbols with ASCII
-character sequences \citep{Wells1987,Wells_etal1992}, e.g. <p\textbackslash> 
+character sequences \citep{Wells1987,Wells_etal1992}, using e.g. <p\textbackslash> 
 for [ɸ]. SAMPA was developed by a group of speech scientists from nine countries 
 in Europe and it constituted the ASCII-IPA symbols needed for phonemic transcription 
-of the principal European Union languages \citep{Wells1995}. It is still widely 
+of the principal European languages \citep{Wells1995}. It is still widely 
 used in language technology.
 
-Two problems with SAMPA are that (i) it is only a partial encoding of the IPA and (ii) it encodes different
-languages in separate data tables, instead of using a universal alphabet, like
-IPA.\@ SAMPA tables were developed as part of a European Commission-funded project to
-address technical problems like electronic mail exchange (what is now simply
-called email). SAMPA is essentially a hack to work around displaying IPA
-characters, but it provided speech technology and other fields a basis that has
-been widely adopted and often still used in code. So, SAMPA is a collection of tables to be
-compared, instead of a large universal table representing all languages.
+Two problems with SAMPA are that (i) it is only a partial encoding of the IPA
+and (ii) it encodes different languages in separate data tables, instead of
+using a universal alphabet, like IPA.\@ SAMPA tables were developed as part of a
+European Commission-funded project to address technical problems like electronic
+mail exchange (what is now simply called email). SAMPA is essentially a hack to
+work around displaying IPA characters, but it provided speech technology and
+other fields a basis that has been widely adopted and often still used in code.
+So, SAMPA is a collection of tables to be compared, instead of a large universal
+table representing all languages.
 
 An extended version of SAMPA, called X-SAMPA, set out to include every symbol,
-including all diacritics, in the IPA chart \citep{Wells1995}. X-SAMPA is considered
-more universally applicable because it consists of one table that encodes all
-characters in IPA. In line with the principles of the IPA, SAMPA and X-SAMPA include a 
-repertoire of symbols. These symbols are intended to represent phonemes rather than 
-all allophonic distinctions. Additionally, both ASCII-ifications of IPA are useful because 
-strings of SAMPA or X-SAMPA are (reportedly) uniquely parsable \citep{Wells1995}. However, 
-like the IPA, X-SAMPA has different notations for encoding the same phonetic phenomena 
-(see Pitfall \ref{}).
+including all diacritics, in the IPA chart \citep{Wells1995}. X-SAMPA is
+considered more universally applicable because it consists of one table that
+encodes all characters in IPA. In line with the principles of the IPA, SAMPA and
+X-SAMPA include a repertoire of symbols. These symbols are intended to represent
+phonemes rather than all allophonic distinctions. Additionally, both
+ASCII-ifications of IPA -- SAMPA and X-SAMPA -- are
+(reportedly) uniquely parsable \citep{Wells1995}. However, like the IPA, X-SAMPA
+has different notations for encoding the same phonetic phenomena (cf. Section~
+\ref{pitfall-multiple-options-ipa}).
 
-SAMPA and X-SAMPA have been widely used for speech technology and as an encoding system in
-computational linguistics. In fact, they are stilled used in popular software packages 
-that require ASCII input and some of which have been co-opted for linguistic analyses, 
-e.g.~RuG/L04 and SplitsTree4.\footnote{See \url{http://www.let.rug.nl/kleiweg/L04/} and 
-\url{http://www.splitstree.org/}, respectively}
+SAMPA and X-SAMPA have been widely used for speech technology and as an encoding
+system in computational linguistics. In fact, they are still used in popular
+software packages that require ASCII input, e.g.~RuG/L04 and SplitsTree4.\footnote{See
+\url{http://www.let.rug.nl/kleiweg/L04/} and \url{http://www.splitstree.org/},
+respectively}
 
+% ==========================
 \section{The need for a single multilingual environment}
+% ==========================
+
+In hindsight it is easy to lose sight of how impactful 30 years of technological
+development have been on linguistics, from theory development using quantitative
+means to pure data collection and dissemination. But at the end of the 1970s,
+virtually no ordinary working linguist was using a personal computer
+\citep{Simons1996}. Personal computer usage, however, dramatically increased
+throughout the 1980s. By 1990, dozens of character sets were in common use. They
+varied in their architecture and in their character repertoires, which made
+things a mess.
 
@@
 challenges head-on by wishing to publish and communicate electronic text 
-with phonetic symbols which were not included in basic ASCII.\footnote{One 
+with phonetic symbols which were not included in basic ASCII. One 
 only needs to look at facsimiles of older electronic documents to see exotic 
-symbols written in by hand.}
+symbols written in by hand after the preparation of typed version.
 
-Long familiar were linguists already with the distinction between function 
-and form. Even in the context of the computer implementation of writing systems, 
-the necessity to distinguish form and function had been made \citep{Becker1984}. 
-The computer industry, on the other hand, did not consider, ignored, or simply 
-did not encode this principle when creating operating systems like MS-DOS, which 
-were limited to 256 code points (due to computer hardware architecture) and 
-encoded with one-to-one mappings from character codes to graphemes.
+%There were two major players in the universal character set race:
+%Unicode and the International Organization for Standardization (ISO).
+
+%Long familiar were linguists already with the distinction between function 
+%and form. Even in the context of the computer implementation of writing systems, 
+%the necessity to distinguish form and function had been made \citep{Becker1984}. 
+%The computer industry, on the other hand, did not consider, ignored, or simply 
+%did not encode this principle when creating operating systems like MS-DOS, which 
+%were limited to 256 code points (due to computer hardware architecture) and 
+%encoded with one-to-one mappings from character codes to graphemes.
+
+% This standard became the basis for a proposal to include the IPA in the first version of the Unicode Standard.
+Decisions by the Computer Coding working group and work they continued after the 1989 Kiel Convention were adopted by the International Phonetic Association.
+These decisions are directly reflected in the Unicode Standard's encoding of IPA, seeing as it was the Association who submitted the script proposal to the Unicode Consortium.
+
+A major benefit of the standardization of the IPA in a computational
+representation by the Kiel working group is that it provided the basis for a
+formal proposal to be submitted to various international standards
+organizations, several of which were trying to tackle (and in a sense `win') the
+multilingual computing environment problem (cf. Section~\ref{encoding}).
+Basically, everyone -- from corporations to governments to language scientists
+-- wanted a single unified multilingual character encoding set for all the
+world's writing systems, even if they did not understand or appreciate the
+challenges involved in creating and adopting a solution.
 
@@
 perspective \citep{Anderson1984,Becker1984,Simons1989}. The study of writing
-systems was also being invigorated by the computational challenges in making 
-computers work in a multilingual environment.\footnote{\cite[11--15]{Sampson1985} 
-urges linguists to view the study of writing systems as legitimate scientific enquiry.}
+systems was also being invigorated \citep[11--15]{Sampson1985} by the
+computational challenges in making computers work in a multilingual environment.
+The engineering problems and solutions had been spelled out years before, e.g.\
+a two-byte encoding for multilingual text \citep{Anderson1984}. Although
+languages vary to an astounding extent (cf.\ \cite{EvansLevinson2009}), writing
+systems are quite similar formally and the issue of formal representation of the
+world's orthographic systems had already been addressed \citep{Simons1989}.
 
-% This standard became the basis for a proposal to include the IPA in the first version of the Unicode Standard.
-Decisions by the Computer Coding working group and work they continued after the 1989 Kiel Convention were adopted by the International Phonetic Association.
-These decisions are directly reflected in the Unicode Standard's encoding of IPA, seeing as it was the Association who submitted the script proposal to the Unicode Consortium.
+%A major obstacle in creating a single encoding multilingual environment from 
+%the perspective of writing systems involves the distinction between function 
+%and form \citep{Becker1984} This distinction is so central to basic linguistic 
+%theory and that trained linguists and semiologists take it as second nature. 
+%A central challenge in developing a universal character set was to combine a 
+%technological solution with a formalization of writing systems proper.\footnote{Of 
+%course there were additional practical issues to overcome, e.g.\ funding, creating 
+%the formal and technological proposal, deciding which characters and writing systems 
+%to include initially, while setting precedence of how to add new ones in the future.}
 
-The second major benefit of the standardization of the IPA in a computational 
-representation by the Kiel working group is that it provided the basis for a 
-formal proposal to be submitted to various international standards organizations, 
-several of which were trying to tackle (and in a sense `win') the multilingual 
-computing environment problem. Basically, everyone from corporations to governments 
-to language scientists (for lack of a better term) wanted a single unified multilingual 
-character encoding set for all the world's writing systems, even if they did not 
-understand or appreciate the challenges involved in creating and adopting a solution. 
-Additionally, advancements in computer hardware were making the solutions easier 
-to implement in software.
+% "The set of IPA symbols and their numbers were used to draw up an entity set within SGML by the Text Encoding Initiative (TEl).
+The name of each entity is formed by 'IPA' preceding the number, e.g.
+IPA304 is the rEIentity name of lower-case A.
+These symbols can be processed as IPA symbols and represented on paper and screen with the appropriate local font by modifying the :entity replacement text.
+The advantage of the SGML entity set is that it is independent or the character set being used."
+% "A TEl writing system declaration (wsd) has been drawn up for the IPA symbols."
+% A TEl writing system declaration (wsd) has been drawn up for the IPA symbols.
+This document gives information about the symbol and its IPA function, as well as its encoding in the accompanying SGML document and in UnicodelUCS and in AFII.
+The writing system declaration can be read as a text d9cument or processed by machines in an SMGL process.
 
-The engineering problems and solutions had been spelled out years before, e.g.\ 
-a two-byte encoding for multilingual text \citep{Anderson1984}. 
-Although languages vary to an astounding extent (cf.\ \cite{EvansLevinson2009}), 
-writing systems are quite similar formally and the issue of formal representation 
-of the world's orthographic systems had also been addressed \citep{Simons1989}.
+After the Kiel Convention in 1989, the working group assisted
+the International Phonetic Association in representing the IPA to the
+\textsc{international standards organisation} (ISO) and to the \textsc{text
+encoding initiative} (TEI) \citep{EslingGaylord1993}. The working group's
+formalization of the IPA, i.e.\ a full listing of agreed upon computer codings
+for phonetic symbols, was used in developing writing system descriptions, which
+were at the time being solicited for scripts to be included in the new
+multilingual international character encoding standards. The working group for
+ISO/IEC 10646 and Unicode were two such initiatives.
 
-A major obstacle in creating a single encoding multilingual environment from 
-the perspective of writing systems involves the distinction between function 
-and form \citep{Becker1984} This distinction is so central to basic linguistic 
-theory and that trained linguists and semiologists take it as second nature. 
-A central challenge in developing a universal character set was to combine a 
-technological solution with a formalization of writing systems proper.\footnote{Of 
-course there were additional practical issues to overcome, e.g.\ funding, creating 
-the formal and technological proposal, deciding which characters and writing systems 
-to include initially, while setting precedence of how to add new ones in the future.}
+In the historical context of the IPA being considered for inclusion in 
+ISO/IEC 10646, it is important to realize that there were a variety of 
+sources (i.e.\ not just from the Association) which submitted character 
+proposals for phonetic alphabets. These proposals, including the one from the 
+Association via the Kiel working group, were considered as a whole by 
+the ISO working groups that were responsible for incorporating a phonetic 
+script into the universal character set (UCS). The ISO working groups that 
+were responsible for assigning a phonetic character set then made their 
+own submissions as part of a review process by ISO for approval based on 
+both ``informatic'' and phonetic criteria \citep[86]{EslingGaylord1993}.
 
-In hindsight it is easy to lose sight of how impactful 30 years of technological 
-development have been on linguistics, from theory development using quantitative 
-means to pure data collection and dissemination (which as fed back into the former). 
-But at the end of the 1970s, virtually no ordinary working linguist was using a 
-person computer \cite{Simons1996}. Personal computer usage, however, dramatically 
-increased throughout the 1980s. By 1990, dozens of character sets were in common use. 
-They varied in their architecture and in their character repertoires, which 
-made things a mess. There were two major players in the universal character 
-set race: Unicode and the International Organization for Standardization (ISO).
+Character set ISO/IEC 10646 was approved by ISO, including the phonetic
+characters submitted to them in May 1993. The set of IPA characters were
+assigned UCS codes in 16 bit representation (in hexadecimal) and were published
+Tables 2 and 3 in \cite{EslingGaylord1993}, which include a graphical
+representation of the IPA Symbol, its IPA Name, phonetic description, IPA
+Number, UCS Code and AFII Code.\footnote{The \textsc{association for font
+information interchange} (AFII) was an international database of glyphs created
+to promote the standardization of font data required to produce ISO/IEC 10646.} When the
+character sets of ISO/IEC 10646 and the Unicode Standard later converged (see
+Section~\ref{the-unicode-approach}), the IPA proposal was
+included in the Unicode Standard Version 1.0 -- largely as we know it
+today.
 
+With subsequent revisions to the IPA, one might have expected that the Unicode
+Consortium would update the Unicode Standard in a way that is inline with the
+development linguistic insights. However, updates that go against the principle
+of maintaining backwards compatibility lose out, i.e.\ it is more important to
+deal with the pitfalls created along the way then it is to change the standard.
+Therefore, many of the pitfalls we encounter today when using Unicode IPA are
+historic relics that we have to come to grips with.
+
+% https://en.wikipedia.org/wiki/Uralic_Phonetic_Alphabet
+% http://www.unicode.org/conference/bulldog.html
+% http://www-01.sil.org/computing/computing_environment.html
+
+It was a long journey, but the goal of achieving a single multilingual computing
+environment has largely been accomplished. Current users, however, must cope
+with the pitfalls that were dug along the way, as will be discussed in the next
+chapter. As the Association foresightfully remarked about Unicode:
+
+\begin{center}
+
+\textit{``When this character set is in wide use, \\
+it will be the normal way to encode IPA symbols.''}
+
+\ \\
+
+\citep[164]{IPA1999}.}
+
+\end{center}
+
+% ===============================================================
+
+\begin(comment)
+
+\subsection{IPA (practical) extensions}
+
+The IPA extensions cover speech sounds beyond the sound systems of the world's
+languages, including symbols for pathological speech and paralinguistic
+functions (e.g.\ Braille?).
+
+- pathological speech
+- paralinguistic functions (Braille?)
+
+As noted in Section \ref{}, extensions to the IPA -- in particular
+``large-scale'' extensions --
+
+% Englebretson2009.pdf
+http://www.ruf.rice.edu/~reng/englebretson2009.pdf
+
+X-SAMPA was an ASCII transliteration and is still purposeful for several
+programs
+
+There is also the extensions of the IPA designed for disordered speech, although
+the symbols are occasionally used for transcribing phonetic detail in ``normal''
+speech or also cross-linguistic
+rarities.\footnote{\url{http://phoible.github.io/conventions/}}
+
+% Unicode has its own history; for a summary:
+\url{http://www.unicode.org/history/summary.html}
+
+https://www.phon.ucl.ac.uk/home/wells/ipa-unicode.htm
+
+Computer codes for phonetic symbols
+EslingGaylord1993
+
+% ==========================
 \section{Unicode and ISO 10646}
+% ==========================
 
 In the late 1980s, a universal character set was being developed by what 
-is now referred to as the Unicode Consortium.\footnote{The Unicode 
-Consortium was officially incorporated in January 1991.} This consortium 
+is now referred to as the Unicode Consortium (officially incorporated in January 1991).
+This consortium 
 consisted largely, although not entirely, of major US corporations, with 
@@
 in Europe, which was responsible for ISO/IEC 10646. This character set 
-standard was composed in 1989 and a draft was published in 1990.\footnote{\url{http://www.iso.org/iso/catalogue_detail.htm?csnumber=56921}} 
+standard was composed in 1989 and a draft was published in 
+1990.\footnote{\url{http://www.iso.org/iso/catalogue_detail.htm?csnumber=56921}} 
 The `Universal Multiple-Octet Coded Character Set' or simply UCS was the first 
-officially standardized character encoding with the aim of including all characters from all writing systems.\footnote{\url{http://www.nada.kth.se/i18n/ucs/unicode-iso10646-oview.html}}
+officially standardized character encoding with the aim of including all 
+characters from all writing 
+systems.\footnote{\url{http://www.nada.kth.se/i18n/ucs/unicode-iso10646-oview.html}}
 
@@
 In 1991, the Unicode Consortium and the ISO Working Group for ISO/IEC 10646 
-decided to create a single universal standard for encoding multilingual text.\footnote{http://unicode.org/book/appC.pdf} 
+decided to create a single universal standard for encoding multilingual 
+text.\footnote{http://unicode.org/book/appC.pdf} 
 The two character sets converged, resulting in mutually acceptable changes 
@@
 
-\section{IPA and Unicode}
-
-It was a long journey, but the goal of achieving a single multilingual 
-computing environment has largely been accomplished. We users, however, 
-must cope with the pitfalls that were dug along the way (see Pitfall sections below). 
-Some linguists, including your humble authors, are particularly sensitive 
-to these issues. So we provide practical advice and approaches in the rest 
-of this chapter. But first, we explain how the IPA became incorporated 
-into the Unicode Standard via ISO/IEC 10646.
-
-% "The set of IPA symbols and their numbers were used to draw up an entity set within SGML by the Text Encoding Initiative (TEl).
-The name of each entity is formed by 'IPA' preceding the number, e.g.
-IPA304 is the rEIentity name of lower-case A.
-These symbols can be processed as IPA symbols and represented on paper and screen with the appropriate local font by modifying the :entity replacement text.
-The advantage of the SGML entity set is that it is independent or the character set being used."
-% "A TEl writing system declaration (wsd) has been drawn up for the IPA symbols."
-% A TEl writing system declaration (wsd) has been drawn up for the IPA symbols.
-This document gives information about the symbol and its IPA function, as well as its encoding in the accompanying SGML document and in UnicodelUCS and in AFII.
-The writing system declaration can be read as a text d9cument or processed by machines in an SMGL process.
-
-After the Kiel Convention in 1989, the Computer Coding of the IPA working 
-group assisted the International Phonetic Association in representing the 
-IPA to ISO and to the Text Encoding Initiative (TEI) \citep{EslingGaylord1993}. 
-The working group's formalization of the IPA, i.e.\ a full listing of agreed 
-upon computer codings for phonetic symbols, was used in developing writing 
-systems descriptions which were at the time being solicited for scripts to 
-be included in initiatives for new multilingual international character 
-encoding standards. The working group for ISO/IEC 10646 and Unicode were 
-two such initiatives.
-
-In the historical context of the IPA being considered for inclusion in 
-ISO/IEC 10646, it is important to realize that there were a variety of 
-sources (i.e.\ not just from the Association) which submitted character 
-proposals for phonetic alphabets. These proposals, including from the 
-Association via the Kiel working group, were considered as a whole by 
-the ISO working groups which were responsible for incorporating a phonetic 
-script into the universal character set (UCS). The ISO working groups that 
-were responsible for assigning a phonetic character set then made their 
-own submissions as part of a review process by ISO for approval based on 
-both ``informatic'' and phonetic criteria \citep[86]{EslingGaylord1993}.
-
-Character set ISO/IEC 10646 was approved by ISO, including the phonetic 
-characters submitted to them in May 1993. The set of IPA characters were 
-assigned UCS codes in 16 bit representation (in hexadecimal) and were 
-published Tables 2 and 3 in \cite{EslingGaylord1993}, which include a 
-graphical representation of the IPA symbol, its IPA Name, phonetic description, 
-IPA Number, UCS Code and AFII Code.\footnote{The Association for Font 
-Information Interchange (AFII) was an international database of glyphs 
-created to promote the standardization of font data required to produce 
-ISO/IEC 10646.} Because the character sets of ISO/IEC 10646 and the 
-Unicode Standard converged, the IPA as submitted by the Association 
-and reviewed and further submitted by the ISO working group, was included 
-in the Unicode Standard Version 1.0 -- largely as we know it today.\footnote{The 
-Association later made the foresightful remark, ``When this character set 
-is in wide use, it will be the normal way to encode IPA symbols.'' \citep[164]{IPA1999}.}
-
-With subsequent revisions to the IPA, one might expect the Unicode 
-Consortium would update the Unicode Standard in a way that is inline 
-with linguists' or other language scientists' intuitions. However, 
-updates that go against the ISO's and the Unicode Standard's principles 
-of maintaining backwards compatibility lose out, i.e.\ it is more important 
-to deal with the pitfalls created along the way then it is to change the 
-standard. Therefore, many of the pitfalls we encounter when using Unicode 
-IPA are historic relics that we have to come to grips with.
-
-% https://en.wikipedia.org/wiki/Uralic_Phonetic_Alphabet
-% http://www.unicode.org/conference/bulldog.html
-% http://www-01.sil.org/computing/computing_environment.html
+\end{comment}

### 2015-12-10 book/unicode.tex
minor changes
adding examples

--- 2015-10-19 book/unicode.tex
@@
 \include{chapters/pitfalls}
+\include{chapters/ipa_background}
 \include{chapters/ipa_meets_unicode}
-\include{chapters/orthography_profiles}
+
+%\include{chapters/orthography_profiles}
+
+\chapter{Orthography profiles}
+\label{orthography-profiles}
+
+\section{Characterizing writing systems}
+\label{characterizing-writing-systems}
+
+At this point in the course of rapid ongoing developments, we are left with a
+situation in which the Unicode Standard offers a highly detailed and flexible
+approach to deal computationally with writing systems, but it has unfortunately
+not influenced the linguistic practice very much. In many practical situations,
+the Unicode Standard is far too complex for the day-to-day practice in
+linguistics because it does not offer practical solutions for the down-to-earth
+problems of many linguists. In this section, we propose some simple practical
+guidelines and methods to improve on this situation.
+
+Our central aims for linguistics, to be approached with a Unicode-based
+solution, are: (i) to improve the consistency of the encoding of sources, (ii)
+to transparently document knowledge about the writing system (including
+transliteration), and (iii) to do all of that in a way that is easy and quick to
+manage for many different sources with many different writing systems. The
+central concept in our proposal is the \textsc{orthography profile}, a simple
+tab-separated CSV text file, that characterizes and documents a writing system.
+We also offer basic implementations in Python and R to assist with the
+production of such files, and to apply orthography profiles for consistency
+testing, grapheme tokenization and transliteration. Not only can orthography
+profiles be helpful in the daily practice of linguistics, they also succinctly
+document the orthographic details of a specific source, and, as such, might
+fruitfully be published alongside sources (e.g.~in digital archives). Also, in
+high-level linguistic analyzes in which the graphemic detail is of central
+importance (e.g.~phonotactic or comparative-historical studies), orthography
+profiles can transparently document the decisions that have been taken in the
+interpretation of the orthography in the sources used.
+
+Given these goals, Unicode locale descriptions (see Section~\ref{terminology})
+might seem like the ideal orthography profiles. However, there are various
+practical obstacles preventing the use of such locale descriptions in the daily
+linguistic practice, namely: (i) the XML-structure is too verbose to easily and
+quickly produce or correct manually, (ii) locale descriptions are designed for a
+wide scope on information (like date formats or names of weekdays) most of which
+is not applicable for documenting writing systems, and (iii) most crucially,
+even if someone made the effort to produce a technically correct locale
+description for a specific source at hand, then it is nigh impossible to deploy
+the description. This is because a locale description has to be submitted to and
+accepted by the Unicode Common Locale Data Repository. The repository is
+(rightly so) not interested in descriptions that only apply to a limited set of
+sources (e.g.~descriptions for only a single dictionary).
+
+The major challenge then is developing an infrastructure to identify the
+elements that are individual graphemes in a source, specifically for the
+enormous variety of sources using some kind of alphabetic writing system.
+Authors of source documents (e.g.~dictionaries, wordlists, corpora) use a
+variety of writing systems that range from their own idiosyncratic
+transcriptions to already well-established practical or longstanding
+orthographies. Although the IPA is one practical choice as a sound-based
+normalization for writing systems (which can act as an interlingual pivot to
+attain interoperability across writing systems), graphemes in each writing
+system must also be identified and standardized if interoperability across
+different sources is to be achieved. In most cases, this amounts to more than
+simply mapping a grapheme to an IPA segment because graphemes must first be
+identified in context (e.g.~is the sequence one sound or two sounds or both?)
+and strings must be tokenized, which may include taking orthographic rules into
+account (e.g.~between vowels is /n/ and after a vowel but before a consonant is
+a nasalized vowel /Ṽ/).
+
+In our experience, data from each source must be
+individually tokenized into graphemes so that its orthographic structure can be 
+identified and its contents can be extracted. To extract data for analysis, a
+source-by-source approach is required before an orthography profile can be
+created. For example, almost each available lexicon on the world's languages is
+idiosyncratic in its orthography and thus requires lexicon-specific approaches
+to identify graphemes in the writing system and to map graphemes to phonemes, if
+desired.
+
+Our key proposal for the characterization of a writing system is to use a
+grapheme tokenization as an inter-orthographic pivot. Basically, any source
+document is tokenized by graphemes, and only then a mapping to IPA (or any other
+orthographic transliteration) is performed. An \textsc{orthography profile} then is a
+description of the units and rules that are needed to adequately model a
+graphemic tokenization for a language variety as described in a particular
+source document. An orthography profile summarizes the Unicode (tailored)
+graphemes and orthographic rules used to write a language (the details of the
+structure and assumptions of such a profile will be presented in the next
+section).
+
+% TODO: add tsoshi figure here?
+
+As an example of graphemic tokenization, note the three different levels of
+technological and linguistic elements that interact in the hypothetical lexical
+form <tsʰṍ̰shi>:
+
+\begin{enumerate}
+	\def\labelenumi{\arabic{enumi}.} 
+	\item code points (10 text elements): t s ʰ o \dia{0303} \dia{0330} \dia{0301} s h i 
+	\item grapheme clusters (7 text elements): t s ʰ ṍ̰ s h i 
+	\item tailored grapheme clusters (4 text elements): tsʰ ṍ̰ sh i 
+\end{enumerate}
+
+In (1), the string <tsʰṍ̰shi> has been tokenized into ten Unicode code points
+(using NFD normalization), delimited here by space. Unicode tokenization is
+required because sequences of code points can differ in their visual and logical
+orders. For example, <õ̰> is ambiguous to whether it is the sequence of <o> +
+<\dia{0303}> + <\dia{0330}> or <o> + <\dia{0330}> + <\dia{0303}>. Although these two variants are visually homoglyphs,
+computationally they are different. Unicode normalization should be applied to
+this string to reorder the code points into a canonical order, allowing the data
+to be treated canonically equivalently for search and comparison.
+
+In (2), the
+Unicode code points have been logically normalized and visually organized into
+grapheme clusters, as specified by the Unicode Standard. The combining character
+sequence <õ̰> is normalized and visually grouped together. Note that, the
+\textsc{modifier letter small h} at \uni{02B0}, is not grouped with any 
+other character. This is because it
+belongs to Spacing Modifier Letters category in the Unicode Standard. These
+characters are underspecified for the direction in which they modify a host
+character. For example, can indicate either pre- or post-aspiration (whereas the
+nasalization or creaky diacritic is defined in the Unicode Standard to apply to
+a specified base character).
+
+Finally, to arrive at the graphemic tokenization in
+(3), tailored grapheme clusters are needed (as for example specified in an
+orthography profile). For example, this orthography profile would specify that
+the sequence of characters, and form a single grapheme, and that and form a
+grapheme. The orthography profile could also specify orthographic rules,
+e.g.~when tokenization graphemes, in say English, the in the forms and should be
+treated as distinct sequences depending on their contexts.
+
+\section{Informal description}
+\label{informal-description-of-orthography-profiles}
+
+An orthography profile describes the Unicode code points, characters, graphemes
+and orthographic rules in a writing system. An orthography profile is a
+language-specific (and often even resource-specific) description of the units
+and rules that are needed to adequately model a writing system. An important
+assumption of our work is that we assume a resource is encoded in Unicode (or
+has been converted to Unicode). Any data source that the Unicode Standard is
+unable to capture, will also not be captured by an orthography profile.
+
+Informally, an orthography profile specifies the graphemes (or, in Unicode
+parlance, \textsc{tailored grapheme clusters}) that are expected to occur in any
+data to be analyzed or checked for consistency. These graphemes are first
+identified throughout the whole data (a step which we call
+\textsc{tokenization}), and possibly simply returned as such, possibly including
+error messages about any parts of the data that are not specified by the
+orthography profile. Once the graphemes are identified, they might also be
+changed into other graphemes (a step which we call \textsc{transliteration}).
+When a grapheme has different possible transliterations, then these differences
+should be separated by contextual specification, possibly down to listing
+individual exceptional cases.
+
+The logic of orthographic parsing ...
+
+Note that to deal with ambiguous parsing cases, we can use the Unicode approach
+using the zero width joiner. This is actually a non-joiner (the name is
+confusing): the idea is to add this character into the text to identify cases in
+which a sequence of characters is not supposed to be a complex grapheme (even
+though the sequence is in the orthography profile)
+
+In practice, we foresee a workflow in which orthography profiles are iteratively
+refined, while at the same time inconsistencies and errors in the data to be
+tokenized are corrected. In some more complex use-cases there might even be a
+need for multiple different orthography profiles to be applied in sequence (see
+Section~\ref{use-cases} on various exemplary use-cases). The result of any such
+workflow will normally be a cleaned dataset and an explicit description of the
+orthographic structure in the form of an orthography profile. Subsequently, the
+orthography profiles can be easily distributed in scholarly channels alongside
+the cleaned data, for example in supplementary material added to journal papers
+or in electronic archives.
+
+\section{Formal specification}
+\label{formal-specification-of-orthography-profiles}
+
+\subsection*{File Format}
+The formal specifications of an orthography profile (or simply \textsc{profile}
+for short) are the following:
+
+\begin{enumerate}
+	\def\labelenumi{A\arabic{enumi}.} 
+	\item \textsc{A profile is a unicode UTF-8 encoded text file}, ideally
+       using NFC, no-BOM, and LF (see Section~\ref{pitfall-file-formats}), that
+       includes the information pertinent to the orthography. 
+	\item \textsc{A profile is a tab-separated CSV file with an obligatory header
+       line}. A minimal profile can have just a single column, in which case
+       there will of course be no tabs, but the first line will still be the
+       header. For all columns we assume the name in the header of the CSV file
+       to be crucial. The actual ordering of the columns is unimportant. Because 
+       tabs and newlines are field separators of the profile, these characters 
+       will lead to problems when they occur in the original data.\footnote{The main
+       reason to choose for tab-delimited, and not for quoted comma-separated files
+       is that we want the profiles to be easily manipulated by hand. Tab-separation 
+       is strongly preferred for this reason.}
+	\item \textsc{Metadata are added in a separate UTF-8 text file} using a basic
+       \textsc{tag: value} format. Metadata about the orthographic description
+       given in the orthography profile includes, minimally, (i) author, (ii)
+       date, (iii) title of the profile, (iv) a stable language identifier
+       encoded in BCP 47/ISO 639-3 of the target language of the profile, and (v)
+       bibliographic data for resource(s) that illustrate the orthography
+       described in the profile. Further, (vi) the tokenization method and (vii)
+       the unicode normalisation used should be documented here (see below).
+    	\item \textsc{Separate lines with comments are not allowed}. Comments that
+       belong to specific lines will have to be put in a separate column of
+       the CSV file, e.g.~add a column called \textsc{comments}.            
+\end{enumerate}
+
+\noindent The content of a profile consists of lines, each describing a grapheme
+of the orthography, using the following columns:
+
+\begin{enumerate}
+	\def\labelenumi{A\arabic{enumi}.} \setcounter{enumi}{4} 
+	\item \textsc{A minimal profile consists of a single column} with a header
+       called \texttt{Grapheme}, listing each of the different graphemes in a
+       separate line. The name of this column is crucial for automatic 
+       processing.
+	\item \textsc{Optional columns can be used to specify the left and right
+       context of the grapheme}, to be designated with the headers \texttt{Left}
+       and \texttt{Right}, respectively. The same grapheme can occur multiple
+       times with different contextual specifications, for example to
+       distinguish different pronunciations depending on the context. 
+	\item \textsc{The columns \texttt{Grapheme}, \texttt{Left} and \texttt{Right}
+       can use regular expression metacharacters.} If regular expressions are
+       used, then all literal usage of the special symbols, like full stops <.>
+       or dollar signs <\$> (so-called \textsc{metacharacters}) have to be
+       explicitly escaped by adding a backslash before them (i.e.~use
+       <\textbackslash.> or <\textbackslash\$>). Note that any specification of
+       context automatically expects regular expressions, so it is probably
+       better to always escape all regular expression metacharacters when used
+       literally in the orthography. The following symbols will need to be
+       preceded by a backslash: {[} {]} ( ) \{ \} | ~+ * . - ! ? \^{} \$ and the
+       backslash \textbackslash~itself. 
+	\item \textsc{An optional column can be used to specify classes of graphemes},
+       to be identified by the header \texttt{Class}. For example, this column
+       can be used to define a class of vowels. Users can simply add ad-hoc
+       identifiers in this column to indicate a group of graphemes, which can
+       then be used in the description of the graphemes or the context. The
+       identifiers should of course be chosen such that they do not conflate
+       with any symbols used in the orthography themselves. Note that such
+       classes only refer to the graphemes, not to the context. 
+	\item \textsc{Columns describing transliterations for each graphemes can be
+       added and named at will}. Often more than a single possible
+       transliteration will be of interest. Any software application using these
+       profiles should prompt the user to name any of these columns to select a
+       specific transliteration. 
+	\item \textsc{Any other columns can be added freely, but will mostly be ignored
+       by any software application using the profiles}. As orthography profiles
+       are also intended to be read and interpreted by humans, it is often
+       highly useful to add extra information on the graphemes in further
+       columns, like for example Unicode codepoints, Unicode names, frequency of
+       occurrence, examples of occurrence, explanation of the contextual
+       restrictions, or comments. 
+ \end{enumerate}
+
+\subsection*{Implementation}
+For the automatic processing of the profiles, the following technical standards
+will be expected:
+
+\begin{enumerate}
+	\def\labelenumi{B\arabic{enumi}.} 
+	\item \textsc{Each line of a profile will be interpreted as a regular
+       expression. }Software applications using profiles can also offer to
+       interpret a profile in the literal sense to avoid the necessity for the
+       user to escape regular expressions metacharacters in the profile.
+       However, this only is possible when no contexts or classes are described,
+       so this seems only useful in the most basic orthographies. 
+	\item \textsc{The \textsc{class} column will be used to produce explicit
+       \textsc{or} chains of regular expressions}, which will then be inserted
+       in the \texttt{Grapheme}, \texttt{Left} and \texttt{Right} columns at
+       the position indicated by the class-identifiers. For example, a class
+       called \texttt{V} as a context specification might be replaced by a regular
+       expression like:
+       \texttt{(a\textbar{}e\textbar{}i\textbar{}o\textbar{}u\textbar{}ei\textbar{}au)}.
+       Only the graphemes themselves are included here, not any contexts
+       specified for the elements of the class. Note that in some cases the 
+       ordering inside this regular expression might be crucial.
+	\item \textsc{The \textsc{left} and \textsc{right} contexts will be included
+       into the regular expressions by using lookbehind and lookahead}.
+       Basically, the actual regular expression syntax of lookbehind and
+       lookahead is simply hidden to the users by allowing them to only specify
+       the contexts themselves. Internally, the contexts in the columns
+       \texttt{Left} and \texttt{Right} are combined with the column
+       \texttt{Grapheme} to form a complex regular expression like:\ 
+       \texttt{(?\textless{}=Left)Grapheme(?=Right)}. 
+	\item \textsc{The regular expressions will be applied in the order as specified
+       in the profile, from top to bottom.} A software implementation can offer
+       help in figuring out the optimal ordering of the regular expressions, but
+       should then explicitly report on the order used.      
+\end{enumerate}
+
+\noindent The actual implementation of the profile on some text-string will function as
+follows:
+
+\begin{enumerate}
+	\def\labelenumi{B\arabic{enumi}.} \setcounter{enumi}{4} 
+	\item \textsc{All graphemes are matched in the text before they are tokenized
+       or transliterated}. In this way, there is no necessity for the user to
+       consider `feeding' and `bleeding' situations, in which the application of
+       a rule either changes the text so another rule suddenly applies (feeding)
+       or prevents another rule to apply (`bleeding'). 
+	\item \textsc{The matching of the graphemes can occur either globally or
+       linearly. }From a computer science perspective, the most natural way to
+       match graphemes from a profile in some text is by walking linearly
+       through the text-string from left to right, and at each position go
+       through all graphemes in the profile to see which one matches, then go to
+       the position at the end of the matched grapheme and start over. This is
+       basically how a finite state transducer works, which is a
+       well-established technique in computer science. However, from a
+       linguistic point of view, our experience is that most linguists find it
+       more natural to think from a global perspective. In this approach, the
+       first grapheme in the profile is matched everywhere in the text-string
+       first, before moving to the next grapheme in the profile. Theoretically,
+       these approaches will lead to different results, though in practice of
+       actual natural language orthographies they almost always lead to the same
+       result. Still, we suggest that any software application using orthography
+       profiles should offer both approaches (i.e. \textsc{global} or
+       \textsc{linear}) to the user. The approach used should be documented in
+       the metadata as \textsc{tokenization method}. 
+	\item \textsc{The matching of the graphemes can occur either in NFC or NFD.} By
+       default, both the profile and the text-string to be tokenized should be
+       treated as NFC (see Section~\ref{pitfall-canonical-equivalence}). However, in some use-cases it turns out to
+       be practical to treat both text and profile as NFD. This typically
+       happens when very many different combinations of diacritics occur in the
+       data. An NFD-profile can then be used to first check which individual
+       diacritics are used, before turning to the more cumbersome inspection of
+       all combinations. We suggest that any software application using
+       orthography profiles should offer both approaches (i.e. \textsc{NFC} or
+       \textsc{NFD}) to the user. The approach used should be documented in the
+       metadata as \textsc{unicode normalization}. 
+	\item \textsc{The text-string is always returned in tokenized form} by
+       separating the matched graphemes by a user-specified symbols-string. Any
+       transliteration will be returned on top of the tokenization. 
+	\item \textsc{Leftover characters}, i.e.~\textsc{characters that are not
+       matched by the profile, should be reported to the user as errors.}
+       Typically, the unmatched character are replaced in the tokenization by a
+       user-specified symbol-string.       
+ \end{enumerate}
+
+\subsection*{Software applications}
+
+Any software application offering to use orthography profile:
+
+\begin{enumerate}
+	\def\labelenumi{\arabic{enumi}.} 
+	\item \textsc{should offer user-options} to specify:
+	\begin{enumerate}
+		\def\labelenumii{C\arabic{enumii}.} 
+		\item \textsc{the name of the column to be used for transliteration} (if any). 
+		\item \textsc{the symbol-string to be inserted between graphemes.} Optionally,
+        a warning might be given if the chosen string includes characters from
+        the orthography itself. 
+		\item \textsc{the symbol-string to be inserted for unmatched strings} in the
+        tokenized and transliterated output. 
+		\item \textsc{the tokenization method}, i.e.~whether the tokenization should
+        proceed \textsc{global} or \textsc{linear} (see B6 above). 
+		\item \textsc{unicode normalization}, i.e.~whether the text-string and profile
+        should use \textsc{NFC} or \textsc{NFD}. 
+    \end{enumerate}
+	\item \textsc{might offer user-options}:
+	\begin{enumerate}
+		\def\labelenumii{C\arabic{enumii}.} \setcounter{enumii}{5} 
+		\item \textsc{to assist in the ordering of the graphemes.} In our experience, it
+        makes sense to apply larger graphemes before shorter graphemes, and to
+        apply graphemes with context before graphemes without context. Further,
+        frequently relevant rules might be applied after rarely relevant rules
+        (though frequency is difficult to establish in practice, as it depends
+        on the available data). Also, if this all fails to give any decisive
+        ordering between rules, it seems useful to offer linguists the option to
+        reverse the ordering from any manual specified ordering, because
+        linguists tend to write the more general rule first, before turning to
+        exceptions or special cases. 
+		\item \textsc{to assist in dealing with upper and lower case characters.} It
+        seems practical to offer some basic case matching, so characters like
+        <a> and <A> are treated equally. This will be useful in many concrete
+        cases, although the user should be warned that case matching does not
+        function universally in the same way across orthographies. Ideally,
+        users should prepare orthography profiles with all lowercase and
+        uppercase variants explicitly mentioned, so by default no case matching
+        should be performed. 
+		\item \textsc{to treat the profile literal}, i.e.~to not interpret regular
+        expression metacharacters. Matching graphemes literally often leads to
+        strong speed increase, and would allow users to not needing to worry
+        about escaping metacharacters. However, in our experience all actually
+        interesting use-cases of orthography profiles include some contexts,
+        which automatically prevents any literal interpretation.
+    \end{enumerate}
+	\item \textsc{should return the following information} to the user:
+	\begin{enumerate}
+		\def\labelenumii{C\arabic{enumii}.} \setcounter{enumii}{8} 
+		\item \textsc{the original text-strings} to be processed in the specified
+        unicode normalization, i.e.~in either NFC or NFD as specified by the
+        user. 
+		\item \textsc{the tokenized strings}, with additionally any transliterated
+        strings, if transliteration is requested. 
+		\item \textsc{a survey of all errors encountered}, ideally both in which
+        text-strings any errors occurred and which characters in the
+        text-strings lead to errors. 
+		\item \textsc{a reordered profile}, when any automatic reordering is offered 
+	\end{enumerate}
+\end{enumerate}
+
+\section{Available implementations}
+
+\section{Examples}
+
+In the following example, it looks as if we have two identical strings,
+\texttt{AABB}. However, this is just an surface-impression delivered by the
+current font, which renders Latin and Cyrillic capitals identically. We can
+identify this problem when we produce an orthography profile from the strings.
+Using here the R implementation of orthography profiles, we first assign the two
+strings to a variable \texttt{test}, and then produce an orthography profile
+with the function \texttt{write.profile}. As it turns out, some of the letters 
+are Cyrillic.
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlstd{test} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"AABB"}\hlstd{,} \hlstr{"AАBВ"}\hlstd{)}
+\hlkwd{write.profile}\hlstd{(test)}
+\end{alltt}
+\begin{verbatim}
+##   Grapheme Frequency Codepoint                UnicodeName
+## 1        A         3    U+0041     LATIN CAPITAL LETTER A
+## 2        B         3    U+0042     LATIN CAPITAL LETTER B
+## 3        А         1    U+0410  CYRILLIC CAPITAL LETTER A
+## 4        В         1    U+0412 CYRILLIC CAPITAL LETTER VE
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+\noindent The working of error-message reporting can also nicely be illustrated with this
+example. Suppose we made an orthography profile with just the Latin <A> and <B> 
+as possible graphemes, then this profile would not be sufficient to tokenize the 
+strings. There are graphemes in the data that are not in the profile, so the 
+tokenization produces an error, which can be used to fix the encoding. In the 
+example below, we can see that the Cyrillic encoding is found in the second 
+string of the \texttt{test} input.
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlstd{test} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"AABB"}\hlstd{,} \hlstr{"AАBВ"}\hlstd{)}
+\hlkwd{tokenize}\hlstd{(test,} \hlkwc{profile} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{"A"}\hlstd{,} \hlstr{"B"}\hlstd{))}
+\end{alltt}
+
+{\ttfamily\noindent\color{warningcolor}{\#\# Warning in tokenize(test, profile = c("{}A"{}, "{}B"{})): \\\#\# There were unknown characters found in the input data.\\\#\# Check output\$errors for a table with all problematic strings.}}\begin{verbatim}
+## $strings
+##   originals tokenized
+## 1      AABB   A A B B
+## 2      AАBВ   A ⁇ B ⁇
+## 
+## $profile
+##   matched_rules Grapheme
+## 1             3        B
+## 2             3        A
+## 
+## $errors
+##   originals  errors
+## 2      AАBВ A ⁇ B ⁇
+## 
+## $missing
+##   Grapheme Frequency Codepoint                UnicodeName
+## 1        А         1    U+0410  CYRILLIC CAPITAL LETTER A
+## 2        В         1    U+0412 CYRILLIC CAPITAL LETTER VE
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
 \include{chapters/use_cases}
@@
 
-\begin{knitrout}\scriptsize
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
-echo 'this is a simple bash example'
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
 echo 'this is a simple bash example'
 ls | wc
+\end{alltt}
+
+\begin{verbatim}
 ## this is a simple bash example
-## this is a simple bash example
-##       10      10     165
+##       11      11     184
 \end{verbatim}
@@
 
-\begin{knitrout}\scriptsize
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
 \begin{verbatim}
-"this is a simple R example"
+## note that bash chunks are executed as one, with all output at the end.
+## You will have to make different chunks to get output separated.
+## And add manual linebreaks!
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+Spacing of lines is fine. I think without background looks better
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# adding comments inside code? Probably not a good idea}
+\hlstd{(example} \hlkwb{<-} \hlstr{"this is a simple R example"}\hlstd{)}
+\end{alltt}
+\begin{verbatim}
 ## [1] "this is a simple R example"
-
-test <- 3 + 4
-test
+\end{verbatim}
+\begin{alltt}
+\hlstd{test} \hlkwb{<-} \hlnum{3} \hlopt{+} \hlnum{4}
+\hlstd{test}
+\end{alltt}
+\begin{verbatim}
 ## [1] 7
@@
 
-\begin{knitrout}\scriptsize
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
-library(qlcData)
-example <- "AΑА"
-profile <- write.profile(example)
-\end{verbatim}
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlkwd{library}\hlstd{(qlcData)}
+\hlstd{profile} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(}\hlstr{"AΑА"}\hlstd{)}
+\end{alltt}
 \end{kframe}
@@
 % latex table generated in R 3.2.2 by xtable 1.7-4 package
-% Mon Oct 19 18:08:51 2015
+% Thu Dec 10 18:07:34 2015
 \begin{table}[H]
@@
 You can refer to variables by using \textbackslash Sexpr{}. 
-For example, the length of the example string is 3.
+For example, the length of the example string is 26.
 Crossreferencing also works, see Table~\ref{bla}.
 
-\begin{knitrout}\scriptsize
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
 print('bla' + 'bla')
 print(3+4)
+\end{alltt}
+
+\begin{verbatim}
 ## blabla
@@
 
-\begin{knitrout}\scriptsize
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
 print('python3 needs an extra tag')
@@
 print(uni)
+\end{alltt}
+
+\begin{verbatim}
 ## python3 needs an extra tag

### 2015-12-15 book/unicode.tex
adds preface

--- 2015-10-19 book/unicode.tex
@@
 
+\include{chapters/preface}
 \include{chapters/introduction}
 \include{chapters/pitfalls}
+\include{chapters/ipa_background}
 \include{chapters/ipa_meets_unicode}
@@
 \begin{knitrout}\scriptsize
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
 \begin{verbatim}
-echo 'this is a simple bash example'
 echo 'this is a simple bash example'
@@
 ## this is a simple bash example
-## this is a simple bash example
-##       10      10     165
+##       18      18     306
 \end{verbatim}
@@
 \begin{knitrout}\scriptsize
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
 \begin{verbatim}
 library(qlcData)
-example <- "AΑА"
-profile <- write.profile(example)
 \end{verbatim}
-\end{kframe}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in library(qlcData): there is no package called 'qlcData'}}\begin{verbatim}
+profile <- write.profile("AΑА")
+\end{verbatim}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(expr, envir, enclos): could not find function "{}write.profile"{}}}\end{kframe}
 \end{knitrout}
 
-% latex table generated in R 3.2.2 by xtable 1.7-4 package
-% Mon Oct 19 18:08:51 2015
-\begin{table}[H]
-\centering
-{\small
-\begin{tabular}{llll}
-  \toprule
-Grapheme & Frequency & Codepoint & UnicodeName \\ 
-  \midrule
-A & 1 & U+0041 & LATIN CAPITAL LETTER A \\ 
-  Α & 1 & U+0391 & GREEK CAPITAL LETTER ALPHA \\ 
-  А & 1 & U+0410 & CYRILLIC CAPITAL LETTER A \\ 
-   \bottomrule
-\end{tabular}
-}
-\caption{test} 
-\label{bla}
-\end{table}
+\begin{kframe}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in print(xtable(profile, caption = "{}test"{}, label = "{}bla"{}), include.rownames = FALSE, : could not find function "{}xtable"{}}}\end{kframe}
 
 You can refer to variables by using \textbackslash Sexpr{}. 
-For example, the length of the example string is 3.
+For example, the length of the example string is 26.
 Crossreferencing also works, see Table~\ref{bla}.

### 2015-12-15 book/chapters/orthography_profiles.tex

@@
 account (e.g.~between vowels is /n/ and after a vowel but before a consonant is
-a nasalized vowel /ṽ/).
+a nasalized vowel /Ṽ/).
 
@@
 individual exceptional cases.
+
+The logic of orthographic parsing ...
+
+Note that to deal with ambiguous parsing cases, we can use the Unicode approach
+using the zero width joiner. This is actually a non-joiner (the name is
+confusing): the idea is to add this character into the text to identify cases in
+which a sequence of characters is not supposed to be a complex grapheme (even
+though the sequence is in the orthography profile)
 
@@
        expression like:
-       (a\textbar{}e\textbar{}i\textbar{}o\textbar{}u\textbar{}ei\textbar{}au).
+       \texttt{(a\textbar{}e\textbar{}i\textbar{}o\textbar{}u\textbar{}ei\textbar{}au)}.
        Only the graphemes themselves are included here, not any contexts
@@
 
+\section{Available implementations}
+
 \section{Examples}
 
-[Here should a few abstract short simple examples be added]
+In the following example, it looks as if we have two identical strings,
+\texttt{AABB}. However, this is just an surface-impression delivered by the
+current font, which renders Latin and Cyrillic capitals identically. We can
+identify this problem when we produce an orthography profile from the strings.
+Using here the R implementation of orthography profiles, we first assign the two
+strings to a variable \texttt{test}, and then produce an orthography profile
+with the function \texttt{write.profile}. As it turns out, some of the letters 
+are Cyrillic.
 
-\
+<<>>==
+test <- c('AABB', '\u0041\u0410\u0042\u0412')
+write.profile(test)
+@
 
-Note that to deal with ambiguous parsing cases, we can use the Unicode approach
-using the zero width joiner. This is actually a non-joiner (the name is
-confusing): the idea is to add this character into the text to identify cases in
-which a sequence of characters is not supposed to be a complex grapheme (even
-though the sequence is in the orthography profile)
+\noindent The working of error-message reporting can also nicely be illustrated with this
+example. Suppose we made an orthography profile with just the Latin <A> and <B> 
+as possible graphemes, then this profile would not be sufficient to tokenize the 
+strings. There are graphemes in the data that are not in the profile, so the 
+tokenization produces an error, which can be used to fix the encoding. In the 
+example below, we can see that the Cyrillic encoding is found in the second 
+string of the \texttt{test} input.
+
+<<>>==
+test <- c('AABB', '\u0041\u0410\u0042\u0412')
+tokenize( test, profile = c('A', 'B') )
+@

### 2015-12-15 book/chapters/preface.tex
adds preface file

\chapter{Preface}
\label{preface}

% ==========================
% \section{Preface}
% \label{preface}
% ==========================

% Early phoneticians: Richard Lepsius, Isaac Pitman, Alexander J. Ellis, and Alexander Melville Bell
% Later ones: Henry Sweet, L'abbé Jean-Pierre Rousselot, Eduard Sievers, Paul Passy, and Jens Otto Harry Jesperson

In the 1840s, several researchers and teachers began developing a notation to accurately transcribe articulatory gestures and their acoustic correlates. Their goal was not only transcription, they also wanted to use their notation for spelling reform (especially English), for teaching literacy and foreign languages, and in particular for developing writing systems for unwritten languages.

At first, the challenge of devising a universal phonetic notation seemed straightforward: normalize the relationship between graphemes and sounds (<c> in `cake' -> /k/; <c> in `cell' -> /s/) to remove inconsistencies introduced by the historical divergence of written and spoken forms (e.g.\ in English and French orthographies), and keep adding new symbol-sound mappings until all sounds are accounted for.

% These early transcription system pioneers found success with the first few languages that they encountered. Grapheme replacement and normalization of symbols across different orthographies seemed to work. But as they were confronted with more articulations and different notational devices, early phoneticians encountered a wealth of potentially important data that they had not expected.

This approach was successful for the first few languages investigated. But as the early phoneticians were confronted with more and more articulations, they encountered a wealth of data that they had not expected. The in-depth exploration of articulatory phonetics led to the discovery that the number of possible spoken sounds is infinite. \cite{Hockett1995} notes:

\begin{quote}
``But as they observed articulation more and more closely, and as they took more and more languages into consideration, the number of sounds they could discriminate and the number of symbols needed to denote them both increased seemingly without limit. In fact, as early as 1848, Ellis asserted that in principle the number of different sounds is endless. With the development of instrumental phonetics, largely by L'abbe Rousselt, it soon turned out that, given sufficiently accurate measuring devices, that is in fact the case.''
\end{quote}

% Hockett 1995: "And so the phoneticians began to develop notations that would show only the constant and distinctive features and let the transient and predictable ones be provided for by cover statements"

% The phonemic principle emerged and the model for it was alphabetic writing.

The first instrumental measurements in the 1870-80s showed unequivocally that spoken language is comprised of sounds waves; it is dynamic and not discrete. Given the fluid nature of the speech stream, sounds differ from one utterance to the next and the same sound may differ in different environments. However, many of these changes are predictable in the context of those environments and they need not necessarily form distinctive lexical contrasts.

These findings left the early phoneticians in a conundrum given the aims of developing precise phonetic notion, but also practical alphabetic-based writing. It is in this vein that the phonetic and phonemic distinction was introduced very early into transcription. The etic vs emic principle took hold, in essence, even before terminology for the distinction was introduced and adopted.

A transcription system that showed both contrastive and non-contrastive elements emerged, taking form notably in the International Phonetic Alphabet (which itself has been updated from time-to-time over the last nearly 120 years). The association between phonetic and phonemic would be maintained in the International Phonetic Alphabet and through the various instantiations of phonological theory, which developed from the work of the early phoneticians.

The model for the principle of contrast was alphabetic writing, in particular shallow orthography, i.e.\ a one-to-one mapping between sounds and symbol. Traditional alphabetic orthography, however, did not scale to precise phonetic notation, so the former was sprinkled with diacritics. This dichotomy  causes many issues -- linguistic, theoretic, and recently technological -- which we bring to light in this book.

### 2015-12-15 book/chapters/preface.tex
edits in preface

@@
 
-In the 1840s, several researchers and teachers began developing a notation to accurately transcribe articulatory gestures and their acoustic correlates.
-Their goal was not only transcription, they also wanted to use their notation for spelling reform (especially English), for teaching literacy and foreign languages, and in particular for developing writing systems for unwritten languages.
+In the 1840s, several researchers and teachers began developing notation to accurately transcribe articulatory gestures and their acoustic correlates.
+Their goal was not only transcription, they wanted to use their notation for spelling reform (especially English), for teaching literacy and foreign languages, and in particular for developing writing systems for unwritten languages.
 
@@
 The first instrumental measurements in the 1870-80s showed unequivocally that spoken language is comprised of sounds waves; it is dynamic and not discrete.
-Given the fluid nature of the speech stream, sounds differ from one utterance to the next and the same sound may differ in different environments.
-However, many of these changes are predictable in the context of those environments and they need not necessarily form distinctive lexical contrasts.
+Given the fluid nature of the speech stream, sounds differ from one utterance to the next and the same sound may differ in different phonetic environments.
+However, many of these alterations are predictable in the context of their environments and they need not necessarily form distinctive lexical contrasts.
 
-These findings left the early phoneticians in a conundrum given the aims of developing precise phonetic notion, but also practical alphabetic-based writing.
+These findings left the early phoneticians in a conundrum given the aims of developing precise phonetic notion, but also practically-minded alphabetic-based writing.
 It is in this vein that the phonetic and phonemic distinction was introduced very early into transcription.
-The etic vs emic principle took hold, in essence, even before terminology for the distinction was introduced and adopted.
+The etic vs emic principle took hold, in essence, even before the dichotomy was widely accepted and before terminology for linguistic distinctiveness was introduced and adopted.
 
 A transcription system that showed both contrastive and non-contrastive elements emerged, taking form notably in the International Phonetic Alphabet (which itself has been updated from time-to-time over the last nearly 120 years).
-The association between phonetic and phonemic would be maintained in the International Phonetic Alphabet and through the various instantiations of phonological theory, which developed from the work of the early phoneticians.
+The association between phonetic and phonemic would be maintained in the International Phonetic Alphabet and through the various instantiations of phonological theory, which developed on the work by early phoneticians.
 
-The model for the principle of contrast was alphabetic writing, in particular shallow orthography, i.e.\ a one-to-one mapping between sounds and symbol.
-Traditional alphabetic orthography, however, did not scale to precise phonetic notation, so the former was sprinkled with diacritics.
-This dichotomy  causes many issues -- linguistic, theoretic, and recently technological -- which we bring to light in this book.
+The model for phonemic contrast was alphabetic writing, in particular shallow orthography, i.e.\ one-to-one mappings between sound and symbol.
+Traditional alphabetic orthography, however, did not scale to precise phonetic notation, so the former was sprinkled with diacritics to denote non-contrastive phenomena.
+This dichotomy between phonemic and phonetic and its implementation in writing and consistent encoding causes many issues -- linguistic, theoretic, and most recently technological -- which we bring to light in this book.

### 2016-01-07 book/chapters/orthography_profiles.tex
adding R implementation info

@@
 
-\section{Examples}
+To illustrate the practical application of orthography profiles, we have
+implemented two different versions of the above specifications, one in Python
+and one in R. These two implementations have rather different implementation
+histories, and we are not 100\% sure that they will in all situations give the
+same results. Also the perfomance with larger datasets differs, and the code is
+not always as clean as we would like it to be. In sum, the two implementations
+should be considered as `proof of concept' and not as the final word on the
+practical application of the specifications above. In our own experience, the
+current implementations are sufficiently fast and stable to be useful for
+academic practice (e.g. checking data consistency, or analyzing and
+transliterating small to medium sized data sets), but they should probably
+better not be used for full-scale industry applications.
+
+\subsection*{Installing the R implementation}
+
+The R implementation is available in the package \texttt{qlcData}, which is 
+directly available from the central R repository CRAN (Comprehensive R Archive 
+Network). The R software environment itself has to be downloaded from its 
+website.\footnote{\url{https://www.r-project.org}} After starting the included 
+R program, the \texttt{qlcData} package for dealing with orthography profiles can be 
+simply installed as follows:
+
+<<eval=FALSE>>==
+# download and install the qlcData software
+install.packages('qlcData') 
+# load the software, so it can be used
+library(qlcData) 
+@
+
+\noindent The version available through CRAN is the most recent stable version.
+To obtain the most recent bug-fixes and experimental additions, please use the
+development versions, which is available on
+GitHub.\footnote{\url{http://github.com/cysouw/qlcData}} This development
+version can be easily installed using the github-install helper software from the
+\texttt{devtools} package.
+
+<<eval=FALSE>>==
+# download and install helper software
+install.packages('devtools') 
+# install the qlcData package from GitHub
+devtools::install_github('cysouw/qlcData')
+# load the software, so it can be used 
+library(qlcData) 
+@
+
+\noindent Inside the \texttt{qlcData} package, there are two functions for
+orthography processing, \texttt{write.profile} and \texttt{tokenize}. R includes
+help files with illustrative examples, and also a so-called `vignette' with
+explanations and examples.
+
+<<eval=FALSE>>==
+# view help files
+help(write.profile)
+help(tokenize)
+# view vignette with explanation and examples
+vignette('orthography_processing')
+@
+
+\noindent Basically, the idea is to use \texttt{write.profile} to produce a
+basic orthography profile from some data and then \texttt{tokenize} to apply the
+(possibly edited) profile on some data, as exemplified in the next section. This
+can of course be performed inside R, but additionally there are two more
+interfaces to the R code supplied in the \texttt{qlcData} package: (i) bash
+executables and (ii) `shiny' webapps.
+
+The bash executables are little files providing an interface to the R code that
+can be used in a shell on a UNIX-alike machine. The location of these
+executables is a bit hidden away by the install procedure of R packages. The
+location can be found by the following command in R. These executables can be 
+used from here, or they can be moved and/or copied to any location as wanted.
+
+<<eval=FALSE>>==
+# show the path to the bash executables
+file.path(find.package('qlcData'), 'exec')
+@
+
+\noindent Another option is to add an alias for these executables to the bash
+profile. The following R function can help adding an alias.
+
+<<eval=FALSE>>==
+# help function to add an alias to the bash profile
+add_alias('tokenize')
+@
+
+After this addition
+it should be possible to access the \texttt{tokenize} function from the shell.
+Try \texttt{tokenize --help} to test the functionality.
+
+<<help, engine='bash'>>=
+tokenize --help
+@
+
+To make the functionality even more accessible, we have prepared webapps with 
+the \texttt{shiny} framework for the R functions. These webapps are available 
+online at \url{TODO}. The webapps are also included inside the \texttt{qlcData} 
+package and can be started with the following helper function:
+
+<<eval=FALSE>>==
+launch_shiny('tokenize')
+@
+
+\subsection*{Installung the Python implementation}
+
+TO DO
+
+\section{Abstract examples}
+
+\subsection*{Error reporting}
 
@@
 
-<<>>==
+<<>>=
 test <- c('AABB', '\u0041\u0410\u0042\u0412')
@@
 @
+
+\subsection*{Different ways to write a profile}
+
+The preparation of an orthography profile from some data might sound like a 
+trivial problem, but actually there are various different ways in which strings 
+can be separated into graphemes by \texttt{write.profile}. Consider the 
+following example string. The default settings of \texttt{write.profile} 
+separates the string into Unicode graphemes according to grapheme clusters 
+(`user-perceived characters', see Section~\ref{the-unicode-approach}).
+
+<<>>=
+example <- '\u00d9\u00da\u00db\u0055\u0300\u0055\u0301\u0055\u0302'
+profile_1 <- write.profile(example)
+@
+
+<<results = 'asis', echo = FALSE>>=
+colnames(profile_1) <- c('Gr.', 'Freq.', 'Codepoint', 'Unicode Name')
+print(xtable(profile_1
+        , caption = 'Profile 1 (default settings, splitting grapheme clusters)'
+        , label = 'tab:profile1'
+        )
+  , include.rownames = FALSE
+  , booktabs = TRUE
+  , size = 'scriptsize'
+  , table.placement = 'H'
+  )
+@
+
+\noindent Some of these graphemes are single codepoints, others are combinations
+of two codepoints. By specifying the separator as empty, it is possible to split
+the string into codepoints.
+
+<<>>=
+profile_2 <- write.profile(example, sep = "")
+@
+
+<<results = 'asis', echo = FALSE>>=
+colnames(profile_2) <- c('Grapheme', 'Frequency', 'Codepoint', 'Unicode Name')
+print(xtable(profile_2
+        , caption = 'Profile 2 (splitting by codepoints)'
+        , label = 'tab:profile2'
+        )
+  , include.rownames = FALSE
+  , booktabs = TRUE
+  , size = 'scriptsize'
+  , table.placement = 'H'
+  )
+@
+
+\noindent Some characters look identical, although they are encoded differently. 
+Unicode offers different ways of normalization, which can be invoked here as 
+well (see Section~\ref{pitfall-canonical-equivalence}).
+
+<<>>=
+# after NFC normalization unicode codepoints have changed
+profile_3 <- write.profile(example, normalize = "NFC", sep = "")
+# NFD normalization gives yet another structure of the codepoints
+profile_4 <- write.profile(example, normalize = "NFD", sep = "")
+@
+
+<<results = 'asis', echo = FALSE>>=
+colnames(profile_3) <- c('Grapheme', 'Frequency', 'Codepoint', 'Unicode Name')
+print(xtable(profile_3
+        , caption = 'Profile 3 (splitting by NFC codepoints)'
+        , label = 'tab:profile3'
+        )
+  , include.rownames = FALSE
+  , booktabs = TRUE
+  , size = 'scriptsize'
+  , table.placement = 'H'
+  )
+@
+
+<<results = 'asis', echo = FALSE>>=
+colnames(profile_4) <- c('Grapheme', 'Frequency', 'Codepoint', 'Unicode Name')
+print(xtable(profile_4
+        , caption = 'Profile 4 (splitting by NFD codepoints)'
+        , label = 'tab:profile4'
+        )
+  , include.rownames = FALSE
+  , booktabs = TRUE
+  , size = 'scriptsize'
+  , table.placement = 'H'
+  )
+@
+
+\noindent It is important to realize that for Unicode grapheme definitions, NFC 
+and NFD normalization are equivalent. This can be shown by normalizing the 
+example in either NFC or NFD, but using default separation in 
+\texttt{write.profile}. To be precise, default separation means setting \texttt{sep = NULL} 
+but that has not be added explicitly.
+
+<<>>=
+# note that NFC and NFD normalization are identical
+# for unicode grapheme definitions
+profile_5 <- write.profile(example, normalize = "NFD")
+profile_6 <- write.profile(example, normalize = "NFC")
+@
+
+<<results = 'asis', echo = FALSE>>=
+colnames(profile_5) <- c('Gr.', 'Freq.', 'Codepoint', 'Unicode Name')
+print(xtable(profile_5
+        , caption = 'Profile 5 (splitting by graphemes after NFD)'
+        , label = 'tab:profile5'
+        )
+  , include.rownames = FALSE
+  , booktabs = TRUE
+  , size = 'scriptsize'
+  , table.placement = 'H'
+  )
+@
+
+<<results = 'asis', echo = FALSE>>=
+colnames(profile_6) <- c('Gr.', 'Freq.', 'Codepoint', 'Unicode Name')
+print(xtable(profile_6
+        , caption = 'Profile 6 (splitting by graphemes after NFC)'
+        , label = 'tab:profile6'
+        )
+  , include.rownames = FALSE
+  , booktabs = TRUE
+  , size = 'scriptsize'
+  , table.placement = 'H'
+  )
+@

### 2016-01-07 book/unicode.tex
adding R implementation info

--- 2015-12-10 book/unicode.tex
@@
 
+\include{chapters/preface}
 \include{chapters/introduction}
@@
 
-\section{Examples}
+To illustrate the practical application of orthography profiles, we have
+implemented two different versions of the above specifications, one in Python
+and one in R. These two implementations have rather different implementation
+histories, and we are not 100\% sure that they will in all situations give the
+same results. Also the perfomance with larger datasets differs, and the code is
+not always as clean as we would like it to be. In sum, the two implementations
+should be considered as `proof of concept' and not as the final word on the
+practical application of the specifications above. In our own experience, the
+current implementations are sufficiently fast and stable to be useful for
+academic practice (e.g. checking data consistency, or analyzing and
+transliterating small to medium sized data sets), but they should probably
+better not be used for full-scale industry applications.
+
+\subsection*{Installing the R implementation}
+
+The R implementation is available in the package \texttt{qlcData}, which is 
+directly available from the central R repository CRAN (Comprehensive R Archive 
+Network). The R software environment itself has to be downloaded from its 
+website.\footnote{\url{https://www.r-project.org}} After starting the included 
+R program, the \texttt{qlcData} package for dealing with orthography profiles can be 
+simply installed as follows:
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# download and install the qlcData software}
+\hlkwd{install.packages}\hlstd{(}\hlstr{"qlcData"}\hlstd{)}
+\hlcom{# load the software, so it can be used}
+\hlkwd{library}\hlstd{(qlcData)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+\noindent The version available through CRAN is the most recent stable version.
+To obtain the most recent bug-fixes and experimental additions, please use the
+development versions, which is available on
+GitHub.\footnote{\url{http://github.com/cysouw/qlcData}} This development
+version can be easily installed using the github-install helper software from the
+\texttt{devtools} package.
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# download and install helper software}
+\hlkwd{install.packages}\hlstd{(}\hlstr{"devtools"}\hlstd{)}
+\hlcom{# install the qlcData package from GitHub}
+\hlstd{devtools}\hlopt{::}\hlkwd{install_github}\hlstd{(}\hlstr{"cysouw/qlcData"}\hlstd{)}
+\hlcom{# load the software, so it can be used}
+\hlkwd{library}\hlstd{(qlcData)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+\noindent Inside the \texttt{qlcData} package, there are two functions for
+orthography processing, \texttt{write.profile} and \texttt{tokenize}. R includes
+help files with illustrative examples, and also a so-called `vignette' with
+explanations and examples.
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# view help files}
+\hlkwd{help}\hlstd{(write.profile)}
+\hlkwd{help}\hlstd{(tokenize)}
+\hlcom{# view vignette with explanation and examples}
+\hlkwd{vignette}\hlstd{(}\hlstr{"orthography_processing"}\hlstd{)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+\noindent Basically, the idea is to use \texttt{write.profile} to produce a
+basic orthography profile from some data and then \texttt{tokenize} to apply the
+(possibly edited) profile on some data, as exemplified in the next section. This
+can of course be performed inside R, but additionally there are two more
+interfaces to the R code supplied in the \texttt{qlcData} package: (i) bash
+executables and (ii) `shiny' webapps.
+
+The bash executables are little files providing an interface to the R code that
+can be used in a shell on a UNIX-alike machine. The location of these
+executables is a bit hidden away by the install procedure of R packages. The
+location can be found by the following command in R. These executables can be 
+used from here, or they can be moved and/or copied to any location as wanted.
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# show the path to the bash executables}
+\hlkwd{file.path}\hlstd{(}\hlkwd{find.package}\hlstd{(}\hlstr{"qlcData"}\hlstd{),} \hlstr{"exec"}\hlstd{)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+\noindent Another option is to add an alias for these executables to the bash
+profile. The following R function can help adding an alias.
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# help function to add an alias to the bash profile}
+\hlkwd{add_alias}\hlstd{(}\hlstr{"tokenize"}\hlstd{)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+After this addition
+it should be possible to access the \texttt{tokenize} function from the shell.
+Try \texttt{tokenize --help} to test the functionality.
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+tokenize --help
+\end{alltt}
+
+\begin{verbatim}
+## Error in running command bash
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+To make the functionality even more accessible, we have prepared webapps with 
+the \texttt{shiny} framework for the R functions. These webapps are available 
+online at \url{TODO}. The webapps are also included inside the \texttt{qlcData} 
+package and can be started with the following helper function:
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlkwd{launch_shiny}\hlstd{(}\hlstr{"tokenize"}\hlstd{)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+\subsection*{Installung the Python implementation}
+
+TO DO
+
+\section{Abstract examples}
+
+\subsection*{Error reporting}
 
@@
 \end{knitrout}
+
+\subsection*{Different ways to write a profile}
+
+The preparation of an orthography profile from some data might sound like a 
+trivial problem, but actually there are various different ways in which strings 
+can be separated into graphemes by \texttt{write.profile}. Consider the 
+following example string. The default settings of \texttt{write.profile} 
+separates the string into Unicode graphemes according to grapheme clusters 
+(`user-perceived characters', see Section~\ref{the-unicode-approach}).
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlstd{example} \hlkwb{<-} \hlstr{"ÙÚÛÙÚÛ"}
+\hlstd{profile_1} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+% latex table generated in R 3.2.2 by xtable 1.7-4 package
+% Thu Jan  7 01:22:44 2016
+\begin{table}[H]
+\centering
+{\scriptsize
+\begin{tabular}{llll}
+  \toprule
+Gr. & Freq. & Codepoint & Unicode Name \\ 
+  \midrule
+Ú & 1 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
+  Ú & 1 & U+0055, U+0301 & LATIN CAPITAL LETTER U, COMBINING ACUTE ACCENT \\ 
+  Ù & 1 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
+  Ù & 1 & U+0055, U+0300 & LATIN CAPITAL LETTER U, COMBINING GRAVE ACCENT \\ 
+  Û & 1 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
+  Û & 1 & U+0055, U+0302 & LATIN CAPITAL LETTER U, COMBINING CIRCUMFLEX ACCENT \\ 
+   \bottomrule
+\end{tabular}
+}
+\caption{Profile 1 (default settings, splitting grapheme clusters)} 
+\label{tab:profile1}
+\end{table}
+
+\noindent Some of these graphemes are single codepoints, others are combinations
+of two codepoints. By specifying the separator as empty, it is possible to split
+the string into codepoints.
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlstd{profile_2} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+% latex table generated in R 3.2.2 by xtable 1.7-4 package
+% Thu Jan  7 01:22:44 2016
+\begin{table}[H]
+\centering
+{\scriptsize
+\begin{tabular}{llll}
+  \toprule
+Grapheme & Frequency & Codepoint & Unicode Name \\ 
+  \midrule
+́ & 1 & U+0301 & COMBINING ACUTE ACCENT \\ 
+  ̀ & 1 & U+0300 & COMBINING GRAVE ACCENT \\ 
+  ̂ & 1 & U+0302 & COMBINING CIRCUMFLEX ACCENT \\ 
+  U & 3 & U+0055 & LATIN CAPITAL LETTER U \\ 
+  Ú & 1 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
+  Ù & 1 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
+  Û & 1 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
+   \bottomrule
+\end{tabular}
+}
+\caption{Profile 2 (splitting by codepoints)} 
+\label{tab:profile2}
+\end{table}
+
+\noindent Some characters look identical, although they are encoded differently. 
+Unicode offers different ways of normalization, which can be invoked here as 
+well (see Section~\ref{pitfall-canonical-equivalence}).
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# after NFC normalization unicode codepoints have changed}
+\hlstd{profile_3} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example,} \hlkwc{normalize} \hlstd{=} \hlstr{"NFC"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{)}
+\hlcom{# NFD normalization gives yet another structure of the codepoints}
+\hlstd{profile_4} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example,} \hlkwc{normalize} \hlstd{=} \hlstr{"NFD"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+% latex table generated in R 3.2.2 by xtable 1.7-4 package
+% Thu Jan  7 01:22:44 2016
+\begin{table}[H]
+\centering
+{\scriptsize
+\begin{tabular}{llll}
+  \toprule
+Grapheme & Frequency & Codepoint & Unicode Name \\ 
+  \midrule
+Ú & 2 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
+  Ù & 2 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
+  Û & 2 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
+   \bottomrule
+\end{tabular}
+}
+\caption{Profile 3 (splitting by NFC codepoints)} 
+\label{tab:profile3}
+\end{table}
+
+% latex table generated in R 3.2.2 by xtable 1.7-4 package
+% Thu Jan  7 01:22:44 2016
+\begin{table}[H]
+\centering
+{\scriptsize
+\begin{tabular}{llll}
+  \toprule
+Grapheme & Frequency & Codepoint & Unicode Name \\ 
+  \midrule
+́ & 2 & U+0301 & COMBINING ACUTE ACCENT \\ 
+  ̀ & 2 & U+0300 & COMBINING GRAVE ACCENT \\ 
+  ̂ & 2 & U+0302 & COMBINING CIRCUMFLEX ACCENT \\ 
+  U & 6 & U+0055 & LATIN CAPITAL LETTER U \\ 
+   \bottomrule
+\end{tabular}
+}
+\caption{Profile 4 (splitting by NFD codepoints)} 
+\label{tab:profile4}
+\end{table}
+
+\noindent It is important to realize that for Unicode grapheme definitions, NFC 
+and NFD normalization are equivalent. This can be shown by normalizing the 
+example in either NFC or NFD, but using default separation in 
+\texttt{write.profile}. To be precise, default separation means setting \texttt{sep = NULL} 
+but that has not be added explicitly.
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# note that NFC and NFD normalization are identical for unicode grapheme}
+\hlcom{# definitions}
+\hlstd{profile_5} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example,} \hlkwc{normalize} \hlstd{=} \hlstr{"NFD"}\hlstd{)}
+\hlstd{profile_6} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example,} \hlkwc{normalize} \hlstd{=} \hlstr{"NFC"}\hlstd{)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+% latex table generated in R 3.2.2 by xtable 1.7-4 package
+% Thu Jan  7 01:22:44 2016
+\begin{table}[H]
+\centering
+{\scriptsize
+\begin{tabular}{llll}
+  \toprule
+Gr. & Freq. & Codepoint & Unicode Name \\ 
+  \midrule
+Ú & 2 & U+0055, U+0301 & LATIN CAPITAL LETTER U, COMBINING ACUTE ACCENT \\ 
+  Ù & 2 & U+0055, U+0300 & LATIN CAPITAL LETTER U, COMBINING GRAVE ACCENT \\ 
+  Û & 2 & U+0055, U+0302 & LATIN CAPITAL LETTER U, COMBINING CIRCUMFLEX ACCENT \\ 
+   \bottomrule
+\end{tabular}
+}
+\caption{Profile 5 (splitting by graphemes after NFD)} 
+\label{tab:profile5}
+\end{table}
+
+% latex table generated in R 3.2.2 by xtable 1.7-4 package
+% Thu Jan  7 01:22:44 2016
+\begin{table}[H]
+\centering
+{\scriptsize
+\begin{tabular}{llll}
+  \toprule
+Gr. & Freq. & Codepoint & Unicode Name \\ 
+  \midrule
+Ú & 2 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
+  Ù & 2 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
+  Û & 2 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
+   \bottomrule
+\end{tabular}
+}
+\caption{Profile 6 (splitting by graphemes after NFC)} 
+\label{tab:profile6}
+\end{table}
 
@@
 % latex table generated in R 3.2.2 by xtable 1.7-4 package
-% Thu Dec 10 18:07:34 2015
+% Wed Jan  6 22:54:44 2016
 \begin{table}[H]

### 2016-06-22 book/chapters/orthography_profiles.tex
minor changes

--- 2015-10-02 book/chapters/orthography_profiles.tex
@@
 \end{enumerate}
-
-\section{Examples}
-
-[Here should a few abstract short simple examples be added]
-
-\
-
-Note that to deal with ambiguous parsing cases, we can use the Unicode approach
-using the zero width joiner. This is actually a non-joiner (the name is
-confusing): the idea is to add this character into the text to identify cases in
-which a sequence of characters is not supposed to be a complex grapheme (even
-though the sequence is in the orthography profile)

### 2016-06-22 book/chapters/profiles_examples.tex
minor changes

--- 2015-12-15 book/unicode.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%                                              %%%
-%%%     Language Science Press Master File       %%%
-%%%         follow the instructions below        %%%
-%%%                                              %%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Available implementations}
+\label{implementations}
 
-% Everything following a % is ignored
-% Some lines start with %. Remove the % to include them
+To illustrate the practical application of orthography profiles, we have
+implemented two different versions of the above specifications, one in Python
+and one in R. These two implementations have rather different implementation
+histories, and we are not 100\% sure that they will in all situations give the
+same results. Also the perfomance with larger datasets differs, and the code is
+not always as clean as we would like it to be. In sum, the two implementations
+should be considered as `proof of concept' and not as the final word on the
+practical application of the specifications above. In our own experience, the
+current implementations are sufficiently fast and stable to be useful for
+academic practice (e.g. checking data consistency, or analyzing and
+transliterating small to medium sized data sets), but they should probably
+better not be used for full-scale industry applications.
 
-\documentclass[output=inprep,
-%  long|short|inprep
-%  ,blackandwhite
-%  ,smallfont
-%  ,draftmode  
-		biblatex
-		]{LSP/langsci}\usepackage[]{graphicx}\usepackage[]{color}
-%% maxwidth is the original width if it is less than linewidth
-%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
-\makeatletter
-\def\maxwidth{ %
-  \ifdim\Gin@nat@width>\linewidth
-    \linewidth
-  \else
-    \Gin@nat@width
-  \fi
-}
-\makeatother
+\subsection*{Installing the R implementation}
 
-\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
-\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
-\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
-\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
-\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
-\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
-\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
-\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
-\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
-\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+The R implementation is available in the package \texttt{qlcData}, which is 
+directly available from the central R repository CRAN (Comprehensive R Archive 
+Network). The R software environment itself has to be downloaded from its 
+website.\footnote{\url{https://www.r-project.org}} After starting the included 
+R program, the \texttt{qlcData} package for dealing with orthography profiles can be 
+simply installed as follows:
 
-\usepackage{framed}
-\makeatletter
-\newenvironment{kframe}{%
- \def\at@end@of@kframe{}%
- \ifinner\ifhmode%
-  \def\at@end@of@kframe{\end{minipage}}%
-  \begin{minipage}{\columnwidth}%
- \fi\fi%
- \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
- \colorbox{shadecolor}{##1}\hskip-\fboxsep
-     % There is no \\@totalrightmargin, so:
-     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
- \MakeFramed {\advance\hsize-\width
-   \@totalleftmargin\z@ \linewidth\hsize
-   \@setminipage}}%
- {\par\unskip\endMakeFramed%
- \at@end@of@kframe}
-\makeatother
+<<eval=FALSE>>==
+# download and install the qlcData software
+install.packages('qlcData') 
+# load the software, so it can be used
+library(qlcData) 
+@
 
-\definecolor{shadecolor}{rgb}{.97, .97, .97}
-\definecolor{messagecolor}{rgb}{0, 0, 0}
-\definecolor{warningcolor}{rgb}{1, 0, 1}
-\definecolor{errorcolor}{rgb}{1, 0, 0}
-\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+\noindent The version available through CRAN is the most recent stable version.
+To obtain the most recent bug-fixes and experimental additions, please use the
+development versions, which is available on
+GitHub.\footnote{\url{http://github.com/cysouw/qlcData}} This development
+version can be easily installed using the github-install helper software from the
+\texttt{devtools} package.
 
-\usepackage{alltt}
+<<eval=FALSE>>==
+# download and install helper software
+install.packages('devtools') 
+# install the qlcData package from GitHub
+devtools::install_github('cysouw/qlcData')
+# load the software, so it can be used 
+library(qlcData) 
+@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%                                              %%%
-%%%          additional packages                 %%%
-%%%                                              %%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\noindent Inside the \texttt{qlcData} package, there are two functions for
+orthography processing, \texttt{write.profile} and \texttt{tokenize}. R includes
+help files with illustrative examples, and also a so-called `vignette' with
+explanations and examples.
 
-% put all additional commands you need in the 
-% following files. If you do not know what this might 
-% mean, you can safely ignore this section
-\usepackage{verbatim}
-\input{localmetadata.tex}
-\input{localpackages.tex}
-\input{localhyphenation.tex}
-\input{localcommands.tex}
-\bibliography{localbibliography}
+<<eval=FALSE>>==
+# view help files
+help(write.profile)
+help(tokenize)
+# view vignette with explanation and examples
+vignette('orthography_processing')
+@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%                                              %%%
-%%%             Frontmatter                      %%%
-%%%                                              %%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\begin{document}
-\maketitle
-\frontmatter
-% %% uncomment if you have preface and/or acknowledgements
-% \include{chapters/preface}
-% \include{chapters/acknowledgments}
-% \include{chapters/abbreviations}
-\tableofcontents
-\mainmatter%
+\noindent Basically, the idea is to use \texttt{write.profile} to produce a
+basic orthography profile from some data and then \texttt{tokenize} to apply the
+(possibly edited) profile on some data, as exemplified in the next section. This
+can of course be performed inside R, but additionally there are two more
+interfaces to the R code supplied in the \texttt{qlcData} package: (i) bash
+executables and (ii) `shiny' webapps.
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%                                              %%%
-%%%             knitr settings                   %%%
-%%%                                              %%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+The bash executables are little files providing an interface to the R code that
+can be used in a shell on a UNIX-alike machine. The location of these
+executables is a bit hidden away by the install procedure of R packages. The
+location can be found by the following command in R. These executables can be 
+used from here, or they can be linked and/or copied to any location as wanted.
 
-% Here are settings for knitr, which will be removed in the .tex file
-% spacing of code-chunks is set with the "knitrout" environment in localcommands.tex
+<<eval=FALSE>>==
+# show the path to the bash executables
+file.path(find.package('qlcData'), 'exec')
+@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%                                              %%%
-%%%             Chapters                         %%%
-%%%                                              %%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\noindent For example, a good way to use the executables in a terminal is to
+make softlinks (using \texttt{ln}) from the executables to a directory in your
+PATH, e.g. to \texttt{/usr/local/bin/}. The two executables are named
+\texttt{tokenize} and \texttt{writeprofile}
 
-\include{chapters/preface}
-\include{chapters/introduction}
-\include{chapters/pitfalls}
-\include{chapters/ipa_background}
-\include{chapters/ipa_meets_unicode}
-\include{chapters/orthography_profiles}
-\include{chapters/use_cases}
+<<link, engine='bash', size='scriptsize', tidy=FALSE, eval=FALSE>>==
+# make two softlinks to the R executables in bash
+ln -is `Rscript -e 'cat(file.path(find.package("qlcData"),
+ "exec", "tokenize"))'` /usr/local/bin
+ln -is `Rscript -e 'cat(file.path(find.package("qlcData"),
+ "exec", "writeprofile"))'` /usr/local/bin
+@
 
-\chapter{Testing code inclusion}
+\noindent After this addition it should be possible to access the \texttt{tokenize}
+function from the shell. Try \texttt{tokenize --help} to test the functionality.
 
-Trying to include code through knitr. 
-Different languages can be easily 
-included!
+<<help, size='tiny', engine='bash'>>=
+tokenize --help
+@
 
-There is a bug for non-R engines: the fontsize settings and line spaceing are not correct. I have 
-filed an issue, hopefully this will be fixed soon.
+To make the functionality even more accessible, we have prepared webapps with 
+the \texttt{shiny} framework for the R functions. These webapps are available 
+online at \url{TODO}. The webapps are also included inside the \texttt{qlcData} 
+package and can be started with the following helper function:
 
-For syntax colouring, install the package `highlight' and for nice tables 
-install the package `xtable'.
+<<eval=FALSE>>==
+launch_shiny('tokenize')
+@
 
-\begin{knitrout}\scriptsize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
-echo 'this is a simple bash example'
-ls | wc
-## this is a simple bash example
-##       18      18     306
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+\subsection*{Installung the Python implementation}
 
-\begin{knitrout}\scriptsize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
-## note that bash chunks are executed as one, with all output at the end.
-## You will have to make different chunks to get output separated.
-## And add manual linebreaks!
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+TO DO
 
-Spacing of lines is fine. I think without background looks better
+\section{Abstract examples}
 
-\begin{knitrout}\scriptsize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
-# adding comments inside code? Probably not a good idea
-(example <- "this is a simple R example")
-## [1] "this is a simple R example"
+\subsection*{Error reporting}
 
-test <- 3 + 4
-test
-## [1] 7
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+In the following example, it looks as if we have two identical strings,
+\texttt{AABB}. However, this is just an surface-impression delivered by the
+current font, which renders Latin and Cyrillic capitals identically. We can
+identify this problem when we produce an orthography profile from the strings.
+Using here the R implementation of orthography profiles, we first assign the two
+strings to a variable \texttt{test}, and then produce an orthography profile
+with the function \texttt{write.profile}. As it turns out, some of the letters 
+are Cyrillic.
 
-\begin{knitrout}\scriptsize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
-library(qlcData)
-\end{verbatim}
+<<>>=
+test <- c('AABB', '\u0041\u0410\u0042\u0412')
+write.profile(test)
+@
 
-{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in library(qlcData): there is no package called 'qlcData'}}\begin{verbatim}
-profile <- write.profile("AΑА")
-\end{verbatim}
+\noindent The working of error-message reporting can also nicely be illustrated with this
+example. Suppose we made an orthography profile with just the Latin <A> and <B> 
+as possible graphemes, then this profile would not be sufficient to tokenize the 
+strings. There are graphemes in the data that are not in the profile, so the 
+tokenization produces an error, which can be used to fix the encoding. In the 
+example below, we can see that the Cyrillic encoding is found in the second 
+string of the \texttt{test} input.
 
-{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(expr, envir, enclos): could not find function "{}write.profile"{}}}\end{kframe}
-\end{knitrout}
+<<>>=
+test <- c('AABB', '\u0041\u0410\u0042\u0412')
+tokenize( test, profile = c('A', 'B') )
+@
 
-\begin{kframe}
+\subsection*{Different ways to write a profile}
 
-{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in print(xtable(profile, caption = "{}test"{}, label = "{}bla"{}), include.rownames = FALSE, : could not find function "{}xtable"{}}}\end{kframe}
+The function \texttt{write.profile} can be used to prepare a skeleton for an
+orthography profile from some data. The preparation of an orthography profile
+from some data might sound like a trivial problem, but actually there are
+various different ways in which strings can be separated into graphemes by
+\texttt{write.profile}. Consider the following example string. The default
+settings of \texttt{write.profile} separates the string into Unicode graphemes
+according to grapheme clusters (`user-perceived characters', see
+Section~\ref{the-unicode-approach}).
 
-You can refer to variables by using \textbackslash Sexpr{}. 
-For example, the length of the example string is 26.
-Crossreferencing also works, see Table~\ref{bla}.
+<<>>=
+example <- '\u00d9\u00da\u00db\u0055\u0300\u0055\u0301\u0055\u0302'
+profile_1 <- write.profile(example)
+@
 
-\begin{knitrout}\scriptsize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
-print('bla' + 'bla')
-print(3+4)
-## blabla
-## 7
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+<<results = 'asis', echo = FALSE>>=
+colnames(profile_1) <- c('Gr.', 'Freq.', 'Codepoint', 'Unicode Name')
+print(xtable(profile_1
+        , caption = 'Profile 1 (default settings, splitting grapheme clusters)'
+        , label = 'tab:profile1'
+        )
+  , include.rownames = FALSE
+  , booktabs = TRUE
+  , size = 'scriptsize'
+  , table.placement = 'H'
+  )
+@
 
-You have to add the option ``engine.path='python3'' to get Python 3.
+\noindent Some of these graphemes are single codepoints, others are combinations
+of two codepoints. By specifying the separator as empty, it is possible to split
+the string into codepoints.
 
-\begin{knitrout}\scriptsize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
-print('python3 needs an extra tag')
-uni = 'aɽɮz'
-print(uni)
-## python3 needs an extra tag
-## aɽɮz
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+<<>>=
+profile_2 <- write.profile(example, sep = "")
+@
 
-Testing: does referencing to variables also work in Python? It does give an error, so no... !
+<<results = 'asis', echo = FALSE>>=
+colnames(profile_2) <- c('Grapheme', 'Frequency', 'Codepoint', 'Unicode Name')
+print(xtable(profile_2
+        , caption = 'Profile 2 (splitting by codepoints)'
+        , label = 'tab:profile2'
+        )
+  , include.rownames = FALSE
+  , booktabs = TRUE
+  , size = 'scriptsize'
+  , table.placement = 'H'
+  )
+@
 
-% TODOS:
+\noindent Some characters look identical, although they are encoded differently. 
+Unicode offers different ways of normalization, which can be invoked here as 
+well (see Section~\ref{pitfall-canonical-equivalence}).
 
-% Lots of characters within <> were lost in translation...
-% some things to note in the translation to LaTeX:
-% tildes in URLs broken -- turned into \textasciitilde{}
-% <a> etc. seem to sometimes be lost
-% section in some cases should be "chapter"
+<<>>=
+# after NFC normalization unicode codepoints have changed
+profile_3 <- write.profile(example, normalize = "NFC", sep = "")
+# NFD normalization gives yet another structure of the codepoints
+profile_4 <- write.profile(example, normalize = "NFD", sep = "")
+@
 
-% other stuff to fix / proof read
-% references
-% sections, e.g. Section 4, Pitfall 4)
-% URLs
-% `', ``''
-% reinsertion of a lot of the glyphs, etc.
-% fix ligatures
-% all glyphs need to be checked
-% all <>s need to be checked and/or readded
+<<results = 'asis', echo = FALSE>>=
+colnames(profile_3) <- c('Grapheme', 'Frequency', 'Codepoint', 'Unicode Name')
+print(xtable(profile_3
+        , caption = 'Profile 3 (splitting by NFC codepoints)'
+        , label = 'tab:profile3'
+        )
+  , include.rownames = FALSE
+  , booktabs = TRUE
+  , size = 'scriptsize'
+  , table.placement = 'H'
+  )
+@
 
-\begin{comment}
-===
-Double quotation marks are generally used for distancing, in particular in the following situations:
+<<results = 'asis', echo = FALSE>>=
+colnames(profile_4) <- c('Grapheme', 'Frequency', 'Codepoint', 'Unicode Name')
+print(xtable(profile_4
+        , caption = 'Profile 4 (splitting by NFD codepoints)'
+        , label = 'tab:profile4'
+        )
+  , include.rownames = FALSE
+  , booktabs = TRUE
+  , size = 'scriptsize'
+  , table.placement = 'H'
+  )
+@
 
-1.
-when a passage from another work is cited in the text (e.g.
-According to Takahashi (2009: 33), “quotatives were never used in subordinate clauses in Old Japanese”); but block quotations do not have quotation marks;
-2.
-when a technical term is mentioned that the author does not want to adopt, but wants to mention, e.g.
-This is sometimes called “pseudo-conservatism”, but I will not use this term here, as it could lead to confusion.
+\noindent It is important to realize that for Unicode grapheme definitions, NFC 
+and NFD normalization are equivalent. This can be shown by normalizing the 
+example in either NFC or NFD, but using default separation in 
+\texttt{write.profile}. To be precise, default separation means setting \texttt{sep = NULL} 
+but that has not be added explicitly.
 
-Single quotation marks are used exclusively for linguistic meanings, as in the following: Latin habere ‘have’ is not cognate with Old English hafian ‘have’.
+<<>>=
+# note that NFC and NFD normalization are identical
+# for unicode grapheme definitions
+profile_5 <- write.profile(example, normalize = "NFD")
+profile_6 <- write.profile(example, normalize = "NFC")
+@
 
-===
-so, we should normally use double quotation in most of our cases :-).
-In general though, I would like us to try and remove double quotations as much as possible.
-Mostly is thus signals uncertainty on our part :-).
+<<results = 'asis', echo = FALSE>>=
+colnames(profile_5) <- c('Gr.', 'Freq.', 'Codepoint', 'Unicode Name')
+print(xtable(profile_5
+        , caption = 'Profile 5 (splitting by graphemes after NFD)'
+        , label = 'tab:profile5'
+        )
+  , include.rownames = FALSE
+  , booktabs = TRUE
+  , size = 'scriptsize'
+  , table.placement = 'H'
+  )
+@
 
-\end{comment}
+<<results = 'asis', echo = FALSE>>=
+colnames(profile_6) <- c('Gr.', 'Freq.', 'Codepoint', 'Unicode Name')
+print(xtable(profile_6
+        , caption = 'Profile 6 (splitting by graphemes after NFC)'
+        , label = 'tab:profile6'
+        )
+  , include.rownames = FALSE
+  , booktabs = TRUE
+  , size = 'scriptsize'
+  , table.placement = 'H'
+  )
+@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%                                              %%%
-%%%             Backmatter                       %%%
-%%%                                              %%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\noindent Note that these different profiles can also be produced using the bash 
+executable \texttt{writeprofile} (see \ref{implementations} for notes on how to 
+install the bash executable). Exactly this example is also included in the help 
+file of the executable.
 
-% There is normally no need to change the backmatter section
-\input{backmatter.tex}
-\end{document}
+<<writehelp, size='tiny', engine='bash'>>=
+writeprofile --help
+@
 
-% you can create your book by running
-% xelatex lsp-skeleton.tex
-%
-% you can also try a simple 
-% make
-% on the commandline
+\subsection*{Using an orthography profile skeleton}

### 2016-06-22 book/unicode.tex
minor changes

--- 2015-12-15 book/unicode.tex
@@
 \include{chapters/ipa_meets_unicode}
+
 \include{chapters/orthography_profiles}
+
+\section{Available implementations}
+\label{implementations}
+
+To illustrate the practical application of orthography profiles, we have
+implemented two different versions of the above specifications, one in Python
+and one in R. These two implementations have rather different implementation
+histories, and we are not 100\% sure that they will in all situations give the
+same results. Also the perfomance with larger datasets differs, and the code is
+not always as clean as we would like it to be. In sum, the two implementations
+should be considered as `proof of concept' and not as the final word on the
+practical application of the specifications above. In our own experience, the
+current implementations are sufficiently fast and stable to be useful for
+academic practice (e.g. checking data consistency, or analyzing and
+transliterating small to medium sized data sets), but they should probably
+better not be used for full-scale industry applications.
+
+\subsection*{Installing the R implementation}
+
+The R implementation is available in the package \texttt{qlcData}, which is 
+directly available from the central R repository CRAN (Comprehensive R Archive 
+Network). The R software environment itself has to be downloaded from its 
+website.\footnote{\url{https://www.r-project.org}} After starting the included 
+R program, the \texttt{qlcData} package for dealing with orthography profiles can be 
+simply installed as follows:
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# download and install the qlcData software}
+\hlkwd{install.packages}\hlstd{(}\hlstr{"qlcData"}\hlstd{)}
+\hlcom{# load the software, so it can be used}
+\hlkwd{library}\hlstd{(qlcData)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+\noindent The version available through CRAN is the most recent stable version.
+To obtain the most recent bug-fixes and experimental additions, please use the
+development versions, which is available on
+GitHub.\footnote{\url{http://github.com/cysouw/qlcData}} This development
+version can be easily installed using the github-install helper software from the
+\texttt{devtools} package.
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# download and install helper software}
+\hlkwd{install.packages}\hlstd{(}\hlstr{"devtools"}\hlstd{)}
+\hlcom{# install the qlcData package from GitHub}
+\hlstd{devtools}\hlopt{::}\hlkwd{install_github}\hlstd{(}\hlstr{"cysouw/qlcData"}\hlstd{)}
+\hlcom{# load the software, so it can be used}
+\hlkwd{library}\hlstd{(qlcData)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+\noindent Inside the \texttt{qlcData} package, there are two functions for
+orthography processing, \texttt{write.profile} and \texttt{tokenize}. R includes
+help files with illustrative examples, and also a so-called `vignette' with
+explanations and examples.
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# view help files}
+\hlkwd{help}\hlstd{(write.profile)}
+\hlkwd{help}\hlstd{(tokenize)}
+\hlcom{# view vignette with explanation and examples}
+\hlkwd{vignette}\hlstd{(}\hlstr{"orthography_processing"}\hlstd{)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+\noindent Basically, the idea is to use \texttt{write.profile} to produce a
+basic orthography profile from some data and then \texttt{tokenize} to apply the
+(possibly edited) profile on some data, as exemplified in the next section. This
+can of course be performed inside R, but additionally there are two more
+interfaces to the R code supplied in the \texttt{qlcData} package: (i) bash
+executables and (ii) `shiny' webapps.
+
+The bash executables are little files providing an interface to the R code that
+can be used in a shell on a UNIX-alike machine. The location of these
+executables is a bit hidden away by the install procedure of R packages. The
+location can be found by the following command in R. These executables can be 
+used from here, or they can be linked and/or copied to any location as wanted.
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# show the path to the bash executables}
+\hlkwd{file.path}\hlstd{(}\hlkwd{find.package}\hlstd{(}\hlstr{"qlcData"}\hlstd{),} \hlstr{"exec"}\hlstd{)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+\noindent For example, a good way to use the executables in a terminal is to
+make softlinks (using \texttt{ln}) from the executables to a directory in your
+PATH, e.g. to \texttt{/usr/local/bin/}. The two executables are named
+\texttt{tokenize} and \texttt{writeprofile}
+
+\begin{knitrout}\scriptsize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\noindent
+\ttfamily
+\hlstd{}\hlslc{\#\ make\ two\ softlinks\ to\ the\ R\ executables\ in\ bash}\hspace*{\fill}\\
+\hlstd{}\hlkwc{ln\ }\hlstd{}\hlopt{{-}}\hlstd{is\ }\hlstr{`Rscript\ {-}e\ 'cat(file.path(find.package("qlcData"),}\hspace*{\fill}\\
+\hlstr{\ "exec",\ "tokenize"))'`}\hlstd{\ }\hlopt{/}\hlstd{usr}\hlopt{/}\hlstd{}\hlkwb{local}\hlstd{}\hlopt{/}\hlstd{bin}\hspace*{\fill}\\
+\hlkwc{ln\ }\hlstd{}\hlopt{{-}}\hlstd{is\ }\hlstr{`Rscript\ {-}e\ 'cat(file.path(find.package("qlcData"),}\hspace*{\fill}\\
+\hlstr{\ "exec",\ "writeprofile"))'`}\hlstd{\ }\hlopt{/}\hlstd{usr}\hlopt{/}\hlstd{}\hlkwb{local}\hlstd{}\hlopt{/}\hlstd{bin}\hspace*{\fill}
+\mbox{}
+\normalfont
+\end{kframe}
+\end{knitrout}
+
+\noindent After this addition it should be possible to access the \texttt{tokenize}
+function from the shell. Try \texttt{tokenize --help} to test the functionality.
+
+\begin{knitrout}\tiny
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\noindent
+\ttfamily
+\hlstd{tokenize\ }\hlopt{{-}{-}}\hlstd{}\hlkwb{help}\hlstd{}\hspace*{\fill}
+\mbox{}
+\normalfont
+
+\begin{verbatim}
+## USAGE: 
+##   tokenize [-hlrd] [-s SEP -p REPL] [-t TRANS -m MISSING] [<STRINGS> <PROFILE>]
+## 
+## DESCRIPTION:
+##   Using the function tokenize() from the R-package qlcData for tokenization and
+##   transliteration of character strings based on an orthography profile. Details
+##   http://www.rdocumentation.org/packages/qlcData/functions/tokenize.html
+## 
+##   STRINGS can be piped through. When no PROFILE is specified, default Unicode
+##   tokenization is performed. Note that the R-code allows for even more options,
+##   not all options are made available here for ease of use.
+##   Errors are likewise (not yet) returned
+## 
+## OPTIONS:
+##   -h, --help      Showing this help text
+##   -l, --linear    Use linear method instead of default global method
+##   -r, --regex     Use regex matching, including contexts in matching graphemes
+##   -d, --NFD       Use NFD normalization instead of default NFC
+##   -s SEP          Separator to be inserted after tokenization
+##                     defaults to space [default: SPACE]
+##   -p REPL         Replacement symbol in case separator occurs in the STRINGS
+##                     defaults to nothing [default: NULL]
+##   -t TRANS        Column in profile to use for transliteration [default: NULL]
+##   -m MISSING      Character to be inserted for characters not in the PROFILE
+##                     defaults to DOUBLE QUESTION MARK at U+2047 [default: ⁇]
+## 
+## EXAMPLES:
+##   Try the difference between the following:
+## 
+##     tokenize áüî
+##     tokenize --NFD -s '__' áüî
+##   
+##   Piping strings should work:
+##   
+##     ls | tokenize
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+To make the functionality even more accessible, we have prepared webapps with 
+the \texttt{shiny} framework for the R functions. These webapps are available 
+online at \url{TODO}. The webapps are also included inside the \texttt{qlcData} 
+package and can be started with the following helper function:
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlkwd{launch_shiny}\hlstd{(}\hlstr{"tokenize"}\hlstd{)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+\subsection*{Installung the Python implementation}
+
+TO DO
+
+\section{Abstract examples}
+
+\subsection*{Error reporting}
+
+In the following example, it looks as if we have two identical strings,
+\texttt{AABB}. However, this is just an surface-impression delivered by the
+current font, which renders Latin and Cyrillic capitals identically. We can
+identify this problem when we produce an orthography profile from the strings.
+Using here the R implementation of orthography profiles, we first assign the two
+strings to a variable \texttt{test}, and then produce an orthography profile
+with the function \texttt{write.profile}. As it turns out, some of the letters 
+are Cyrillic.
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlstd{test} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"AABB"}\hlstd{,} \hlstr{"AАBВ"}\hlstd{)}
+\hlkwd{write.profile}\hlstd{(test)}
+\end{alltt}
+\begin{verbatim}
+##   Grapheme Frequency Codepoint                UnicodeName
+## 1        A         3    U+0041     LATIN CAPITAL LETTER A
+## 2        B         3    U+0042     LATIN CAPITAL LETTER B
+## 3        А         1    U+0410  CYRILLIC CAPITAL LETTER A
+## 4        В         1    U+0412 CYRILLIC CAPITAL LETTER VE
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+\noindent The working of error-message reporting can also nicely be illustrated with this
+example. Suppose we made an orthography profile with just the Latin <A> and <B> 
+as possible graphemes, then this profile would not be sufficient to tokenize the 
+strings. There are graphemes in the data that are not in the profile, so the 
+tokenization produces an error, which can be used to fix the encoding. In the 
+example below, we can see that the Cyrillic encoding is found in the second 
+string of the \texttt{test} input.
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlstd{test} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"AABB"}\hlstd{,} \hlstr{"AАBВ"}\hlstd{)}
+\hlkwd{tokenize}\hlstd{(test,} \hlkwc{profile} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{"A"}\hlstd{,} \hlstr{"B"}\hlstd{))}
+\end{alltt}
+
+{\ttfamily\noindent\color{warningcolor}{\#\# Warning in tokenize(test, profile = c("{}A"{}, "{}B"{})): \\\#\# There were unknown characters found in the input data.\\\#\# Check output\$errors for a table with all problematic strings.}}\begin{verbatim}
+## $strings
+##   originals tokenized
+## 1      AABB   A A B B
+## 2      AАBВ   A ⁇ B ⁇
+## 
+## $profile
+##   matched_rules Grapheme
+## 1             3        B
+## 2             3        A
+## 
+## $errors
+##   originals  errors
+## 2      AАBВ A ⁇ B ⁇
+## 
+## $missing
+##   Grapheme Frequency Codepoint                UnicodeName
+## 1        А         1    U+0410  CYRILLIC CAPITAL LETTER A
+## 2        В         1    U+0412 CYRILLIC CAPITAL LETTER VE
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+\subsection*{Different ways to write a profile}
+
+The preparation of an orthography profile from some data might sound like a 
+trivial problem, but actually there are various different ways in which strings 
+can be separated into graphemes by \texttt{write.profile}. Consider the 
+following example string. The default settings of \texttt{write.profile} 
+separates the string into Unicode graphemes according to grapheme clusters 
+(`user-perceived characters', see Section~\ref{the-unicode-approach}).
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlstd{example} \hlkwb{<-} \hlstr{"ÙÚÛÙÚÛ"}
+\hlstd{profile_1} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+% latex table generated in R 3.3.0 by xtable 1.8-2 package
+% Wed Jun 22 12:03:05 2016
+\begin{table}[H]
+\centering
+\begingroup\scriptsize
+\begin{tabular}{llll}
+  \toprule
+Gr. & Freq. & Codepoint & Unicode Name \\ 
+  \midrule
+Ú & 1 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
+  Ú & 1 & U+0055, U+0301 & LATIN CAPITAL LETTER U, COMBINING ACUTE ACCENT \\ 
+  Ù & 1 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
+  Ù & 1 & U+0055, U+0300 & LATIN CAPITAL LETTER U, COMBINING GRAVE ACCENT \\ 
+  Û & 1 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
+  Û & 1 & U+0055, U+0302 & LATIN CAPITAL LETTER U, COMBINING CIRCUMFLEX ACCENT \\ 
+   \bottomrule
+\end{tabular}
+\endgroup
+\caption{Profile 1 (default settings, splitting grapheme clusters)} 
+\label{tab:profile1}
+\end{table}
+
+\noindent Some of these graphemes are single codepoints, others are combinations
+of two codepoints. By specifying the separator as empty, it is possible to split
+the string into codepoints.
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlstd{profile_2} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+% latex table generated in R 3.3.0 by xtable 1.8-2 package
+% Wed Jun 22 12:03:05 2016
+\begin{table}[H]
+\centering
+\begingroup\scriptsize
+\begin{tabular}{llll}
+  \toprule
+Grapheme & Frequency & Codepoint & Unicode Name \\ 
+  \midrule
+́ & 1 & U+0301 & COMBINING ACUTE ACCENT \\ 
+  ̀ & 1 & U+0300 & COMBINING GRAVE ACCENT \\ 
+  ̂ & 1 & U+0302 & COMBINING CIRCUMFLEX ACCENT \\ 
+  U & 3 & U+0055 & LATIN CAPITAL LETTER U \\ 
+  Ú & 1 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
+  Ù & 1 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
+  Û & 1 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
+   \bottomrule
+\end{tabular}
+\endgroup
+\caption{Profile 2 (splitting by codepoints)} 
+\label{tab:profile2}
+\end{table}
+
+\noindent Some characters look identical, although they are encoded differently. 
+Unicode offers different ways of normalization, which can be invoked here as 
+well (see Section~\ref{pitfall-canonical-equivalence}).
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# after NFC normalization unicode codepoints have changed}
+\hlstd{profile_3} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example,} \hlkwc{normalize} \hlstd{=} \hlstr{"NFC"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{)}
+\hlcom{# NFD normalization gives yet another structure of the codepoints}
+\hlstd{profile_4} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example,} \hlkwc{normalize} \hlstd{=} \hlstr{"NFD"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+% latex table generated in R 3.3.0 by xtable 1.8-2 package
+% Wed Jun 22 12:03:05 2016
+\begin{table}[H]
+\centering
+\begingroup\scriptsize
+\begin{tabular}{llll}
+  \toprule
+Grapheme & Frequency & Codepoint & Unicode Name \\ 
+  \midrule
+Ú & 2 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
+  Ù & 2 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
+  Û & 2 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
+   \bottomrule
+\end{tabular}
+\endgroup
+\caption{Profile 3 (splitting by NFC codepoints)} 
+\label{tab:profile3}
+\end{table}
+
+% latex table generated in R 3.3.0 by xtable 1.8-2 package
+% Wed Jun 22 12:03:05 2016
+\begin{table}[H]
+\centering
+\begingroup\scriptsize
+\begin{tabular}{llll}
+  \toprule
+Grapheme & Frequency & Codepoint & Unicode Name \\ 
+  \midrule
+́ & 2 & U+0301 & COMBINING ACUTE ACCENT \\ 
+  ̀ & 2 & U+0300 & COMBINING GRAVE ACCENT \\ 
+  ̂ & 2 & U+0302 & COMBINING CIRCUMFLEX ACCENT \\ 
+  U & 6 & U+0055 & LATIN CAPITAL LETTER U \\ 
+   \bottomrule
+\end{tabular}
+\endgroup
+\caption{Profile 4 (splitting by NFD codepoints)} 
+\label{tab:profile4}
+\end{table}
+
+\noindent It is important to realize that for Unicode grapheme definitions, NFC 
+and NFD normalization are equivalent. This can be shown by normalizing the 
+example in either NFC or NFD, but using default separation in 
+\texttt{write.profile}. To be precise, default separation means setting \texttt{sep = NULL} 
+but that has not be added explicitly.
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# note that NFC and NFD normalization are identical for unicode grapheme}
+\hlcom{# definitions}
+\hlstd{profile_5} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example,} \hlkwc{normalize} \hlstd{=} \hlstr{"NFD"}\hlstd{)}
+\hlstd{profile_6} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example,} \hlkwc{normalize} \hlstd{=} \hlstr{"NFC"}\hlstd{)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+% latex table generated in R 3.3.0 by xtable 1.8-2 package
+% Wed Jun 22 12:03:05 2016
+\begin{table}[H]
+\centering
+\begingroup\scriptsize
+\begin{tabular}{llll}
+  \toprule
+Gr. & Freq. & Codepoint & Unicode Name \\ 
+  \midrule
+Ú & 2 & U+0055, U+0301 & LATIN CAPITAL LETTER U, COMBINING ACUTE ACCENT \\ 
+  Ù & 2 & U+0055, U+0300 & LATIN CAPITAL LETTER U, COMBINING GRAVE ACCENT \\ 
+  Û & 2 & U+0055, U+0302 & LATIN CAPITAL LETTER U, COMBINING CIRCUMFLEX ACCENT \\ 
+   \bottomrule
+\end{tabular}
+\endgroup
+\caption{Profile 5 (splitting by graphemes after NFD)} 
+\label{tab:profile5}
+\end{table}
+
+% latex table generated in R 3.3.0 by xtable 1.8-2 package
+% Wed Jun 22 12:03:05 2016
+\begin{table}[H]
+\centering
+\begingroup\scriptsize
+\begin{tabular}{llll}
+  \toprule
+Gr. & Freq. & Codepoint & Unicode Name \\ 
+  \midrule
+Ú & 2 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
+  Ù & 2 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
+  Û & 2 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
+   \bottomrule
+\end{tabular}
+\endgroup
+\caption{Profile 6 (splitting by graphemes after NFC)} 
+\label{tab:profile6}
+\end{table}
+
+\noindent Note that these different profiles can also be produced using the bash 
+executable \texttt{writeprofile} (see \ref{implementations} for notes on how to 
+install the bash executable). Exactly this example is also included in the help 
+file of the executable.
+
+\begin{knitrout}\tiny
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\noindent
+\ttfamily
+\hlstd{writeprofile\ }\hlopt{{-}{-}}\hlstd{}\hlkwb{help}\hlstd{}\hspace*{\fill}
+\mbox{}
+\normalfont
+
+\begin{verbatim}
+## USAGE: 
+##   writeprofile [-hie] [-s SEP] [-n NORMALIZE] [<STRINGS>]
+## 
+## DESCRIPTION:
+##   Using the function write.profile() from the R-package qlcData to prepare 
+##   a skeleton for an orthography profile to be used with tokenize. Details:
+##   http://www.rdocumentation.org/packages/qlcData/functions/readwrite_orthography
+## 
+##   STRINGS can be piped through. The results is a tab-delimited file for manual
+##   editing. Note that the R-code allows for even more options, not all options 
+##   are made available here for ease of use.
+## 
+## OPTIONS:
+##   -h, --help      Showing this help text
+##   -i, --info      Add columns with Unicode information to profile
+##   -e, --editing   Add empty columns for easy editing of profile
+##   -s SEP          Separator used to separate the strings
+##                     Often useful is option UNI to split by unicode codepoints
+##                     defaults to unicode character definition [default: NULL]
+##   -n NORMALIZE    Normalization used to make profile. Possible options: NFD, NFC
+##                     defaults to no normalization [default: NULL]
+##                     
+## EXAMPLES:
+##   Note the differences between the following variants:
+## 
+##     example='ÙÚÛÙÚÛ'
+##     echo -e $example
+##     echo -e $example | writeprofile -i
+##     echo -e $example | writeprofile -i -s UNI
+##     echo -e $example | writeprofile -i -s UNI -n NFC
+##     echo -e $example | writeprofile -i -s UNI -n NFD
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
 \include{chapters/use_cases}
@@
 
-\begin{knitrout}\scriptsize
+\begin{knitrout}\footnotesize
 \definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\noindent
+\ttfamily
+\hlstd{}\hlkwb{echo\ }\hlstd{}\hlstr{'this\ is\ a\ simple\ bash\ example'}\hlstd{}\hspace*{\fill}\\
+\hlkwc{ls\ }\hlstd{\textbar \ }\hlkwc{wc}\hlstd{}\hspace*{\fill}
+\mbox{}
+\normalfont
+
 \begin{verbatim}
-echo 'this is a simple bash example'
-ls | wc
 ## this is a simple bash example
-##       18      18     306
+##       11      11     184
 \end{verbatim}
@@
 
-\begin{knitrout}\scriptsize
+\begin{knitrout}\footnotesize
 \definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
@@
 
-\begin{knitrout}\scriptsize
+\begin{knitrout}\footnotesize
 \definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# adding comments inside code? Probably not a good idea}
+\hlstd{(example} \hlkwb{<-} \hlstr{"this is a simple R example"}\hlstd{)}
+\end{alltt}
 \begin{verbatim}
-# adding comments inside code? Probably not a good idea
-(example <- "this is a simple R example")
 ## [1] "this is a simple R example"
+\end{verbatim}
+\begin{alltt}
 
-test <- 3 + 4
-test
+\hlstd{test} \hlkwb{<-} \hlnum{3} \hlopt{+} \hlnum{4}
+\hlstd{test}
+\end{alltt}
+\begin{verbatim}
 ## [1] 7
@@
 
-\begin{knitrout}\scriptsize
+\begin{knitrout}\footnotesize
 \definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
-library(qlcData)
-\end{verbatim}
-
-{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in library(qlcData): there is no package called 'qlcData'}}\begin{verbatim}
-profile <- write.profile("AΑА")
-\end{verbatim}
-
-{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(expr, envir, enclos): could not find function "{}write.profile"{}}}\end{kframe}
+\begin{alltt}
+\hlkwd{library}\hlstd{(qlcData)}
+\hlstd{profile} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(}\hlstr{"AΑА"}\hlstd{)}
+\end{alltt}
+\end{kframe}
 \end{knitrout}
 
-\begin{kframe}
-
-{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in print(xtable(profile, caption = "{}test"{}, label = "{}bla"{}), include.rownames = FALSE, : could not find function "{}xtable"{}}}\end{kframe}
+% latex table generated in R 3.3.0 by xtable 1.8-2 package
+% Wed Jun 22 10:58:23 2016
+\begin{table}[H]
+\centering
+\begingroup\small
+\begin{tabular}{llll}
+  \toprule
+Grapheme & Frequency & Codepoint & UnicodeName \\ 
+  \midrule
+A & 1 & U+0041 & LATIN CAPITAL LETTER A \\ 
+  Α & 1 & U+0391 & GREEK CAPITAL LETTER ALPHA \\ 
+  А & 1 & U+0410 & CYRILLIC CAPITAL LETTER A \\ 
+   \bottomrule
+\end{tabular}
+\endgroup
+\caption{test} 
+\label{bla}
+\end{table}
 
@@
 
-\begin{knitrout}\scriptsize
+\begin{knitrout}\footnotesize
 \definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\noindent
+\ttfamily
+\hlstd{}\hlkwa{print}\hlstd{}\hlopt{(}\hlstd{}\hlstr{'bla'}\hlstd{\ }\hlopt{+\ }\hlstd{}\hlstr{'bla'}\hlstd{}\hlopt{)}\hspace*{\fill}\\
+\hlstd{}\hlkwa{print}\hlstd{}\hlopt{(}\hlstd{}\hlnum{3}\hlstd{}\hlopt{+}\hlstd{}\hlnum{4}\hlstd{}\hlopt{)}\hlstd{}\hspace*{\fill}
+\mbox{}
+\normalfont
+
 \begin{verbatim}
-print('bla' + 'bla')
-print(3+4)
 ## blabla
@@
 
-\begin{knitrout}\scriptsize
+\begin{knitrout}\footnotesize
 \definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\noindent
+\ttfamily
+\hlstd{}\hlkwa{print}\hlstd{}\hlopt{(}\hlstd{}\hlstr{'python3\ needs\ an\ extra\ tag'}\hlstd{}\hlopt{)}\hspace*{\fill}\\
+\hlstd{uni\ }\hlopt{=\ }\hlstd{}\hlstr{'aɽɮz'}\hlstd{}\hspace*{\fill}\\
+\hlkwa{print}\hlstd{}\hlopt{(}\hlstd{uni}\hlopt{)}\hlstd{}\hspace*{\fill}
+\mbox{}
+\normalfont
+
 \begin{verbatim}
-print('python3 needs an extra tag')
-uni = 'aɽɮz'
-print(uni)
 ## python3 needs an extra tag

### 2016-06-30 book/chapters/implementation.tex
adding R examples

\chapter{Implementation}
\label{implementation}

\section{Available implementations}

To illustrate the practical application of orthography profiles as defined in
Chapter \ref{orthography-profiles}, we have implemented two different versions
of the specifications presented there: one in Python and one in R. These two
implementations have rather different implementation histories, and we are not
100\% sure that they will in all situations give the same results. Also the
perfomance with larger datasets differs, and the code is not always as clean as
we would like it to be. In sum, the two implementations should be considered as
`proof of concept' and not as the final word on the practical application of the
specifications above. In our own experience, the current implementations are
sufficiently fast and stable to be useful for academic practice (e.g. checking
data consistency, or analyzing and transliterating small to medium sized data
sets), but they should probably better not be used for full-scale industry
applications.

This chapter will introduce the implementations, and give practical step-by-step 
guidelines for installing them and using them. Various simple, and sometimes 
somewhat abstract, examples will be discussed to show the different options 
available, and to illustrate the intended usage of orthography profiles. In the 
next chapter we will then discuss a few full-scale examples to illustrate the 
application and the usefulness of orthography profiles.

\subsection*{Installing the R implementation}

The R implementation is available in the package \texttt{qlcData}, which is 
directly available from the central R repository CRAN (Comprehensive R Archive 
Network). The R software environment itself has to be downloaded from its 
website.\footnote{\url{https://www.r-project.org}} After starting the included 
R program, the \texttt{qlcData} package for dealing with orthography profiles can be 
simply installed as follows:

<<eval=FALSE>>=
# download and install the qlcData software
install.packages('qlcData') 
# load the software, so it can be used
library(qlcData) 
@

\noindent The version available through CRAN is the most recent stable version.
To obtain the most recent bug-fixes and experimental additions, please use the
development versions, which is available on
GitHub.\footnote{\url{http://github.com/cysouw/qlcData}} This development
version can be easily installed using the github-install helper software from the
\texttt{devtools} package.

<<eval=FALSE>>=
# download and install helper software
install.packages('devtools') 
# install the qlcData package from GitHub
devtools::install_github('cysouw/qlcData')
# load the software, so it can be used 
library(qlcData) 
@

\noindent Inside the \texttt{qlcData} package, there are two functions for
orthography processing, \texttt{write.profile} and \texttt{tokenize}. R includes
help files with illustrative examples, and also a so-called `vignette' with
explanations and examples.

<<eval=FALSE>>=
# view help files
help(write.profile)
help(tokenize)
# view vignette with explanation and examples
vignette('orthography_processing')
@

\noindent Basically, the idea is to use \texttt{write.profile} to produce a
basic orthography profile from some data and then \texttt{tokenize} to apply the
(possibly edited) profile on some data, as exemplified in the next section. This
can of course be performed though R, but additionally there are two more
interfaces to the R code supplied in the \texttt{qlcData} package: (i) bash
executables and (ii) `shiny' webapps.

The bash executables are little files providing an interface to the R code that
can be used in a shell on a UNIX-alike machine. The location of these
executables is a bit hidden away by the install procedure of R packages. The
location can be found by the following command in R. These executables can be 
used from here, or they can be linked and/or copied to any location as wanted.

<<eval=FALSE>>=
# show the path to the bash executables
file.path(find.package('qlcData'), 'exec')
@

\noindent For example, a good way to use the executables in a terminal is to
make softlinks (using \texttt{ln}) from the executables to a directory in your
PATH, e.g. to \texttt{/usr/local/bin/}. The two executables are named
\texttt{tokenize} and \texttt{writeprofile}, and the links can be made directly 
by using Rscript to get the paths to the executables within the terminal.

<<eval=FALSE, tidy=FALSE, size='scriptsize', engine='bash'>>=
# get the paths to the R executables in bash
pathT=`Rscript -e 'cat(file.path(find.package("qlcData"), "exec", "tokenize"))'`
pathW=`Rscript -e 'cat(file.path(find.package("qlcData"), "exec", "writeprofile"))'`

# make softlinks to the R executables in /usr/local/bin
ln -is $pathT $pathW /usr/local/bin
@

\noindent After this softlink it should be possible to access the \texttt{tokenize}
function from the shell. Try \texttt{tokenize --help} to test the functionality.

<<size='scriptsize', engine='bash'>>=
tokenize --help
@

To make the functionality even more accessible, we have prepared webapps with 
the \texttt{shiny} framework for the R functions. These webapps are available 
online at \url{TODO}. The webapps are also included inside the \texttt{qlcData} 
package and can be started with the following helper function:

<<eval=FALSE>>=
launch_shiny('tokenize')
@

TO DO

\subsection*{Installung the Python implementation}

TO DO

\section{Error reporting}
\label{error-reporting}

In the following example, it looks as if we have two identical strings,
\texttt{AABB}. However, this is just an surface-impression delivered by the
current font, which renders Latin and Cyrillic capitals identically. We can
identify this problem when we produce an orthography profile from the strings.
Using here the R implementation of orthography profiles, we first assign the two
strings to a variable \texttt{test}, and then produce an orthography profile
with the function \texttt{write.profile}. As it turns out, some of the letters 
are Cyrillic.

<<>>=
test <- c('AABB', '\u0041\u0410\u0042\u0412')
write.profile(test)
@

\noindent The working of error-message reporting can also nicely be illustrated
with this example. Suppose we made an orthography profile with just the two
Latin letters <A> and <B> as possible graphemes, then this profile would not be
sufficient to tokenize the strings. There are graphemes in the data that are not
in the profile, so the tokenization produces an error, which can be used to fix
the encoding. In the example below, we can see that the Cyrillic encoding is
found in the second string of the \texttt{test} input.

<<>>=
test <- c('AABB', '\u0041\u0410\u0042\u0412')
tokenize( test, profile = c('A', 'B') )
@

\section{Different ways to write a profile}
\label{write-profile}

The function \texttt{write.profile} can be used to prepare a skeleton for an
orthography profile from some data. The preparation of an orthography profile
from some data might sound like a trivial problem, but actually there are
various different ways in which strings can be separated into graphemes by
\texttt{write.profile}. Consider the following example string. The default
settings of \texttt{write.profile} separates the string into Unicode graphemes
according to grapheme clusters (`user-perceived characters', see
Section~\ref{the-unicode-approach}).

<<>>=
example <- '\u00d9\u00da\u00db\u0055\u0300\u0055\u0301\u0055\u0302'
profile_1 <- write.profile(example)
@

<<results = 'asis', echo = FALSE>>=
colnames(profile_1) <- c('Gr.', 'Freq.', 'Codepoint', 'Unicode Name')
print(xtable(profile_1
        , caption = 'Profile 1 (default settings, splitting grapheme clusters)'
        , label = 'tab:profile1'
        )
  , include.rownames = FALSE
  , booktabs = TRUE
  , size = 'scriptsize'
  , table.placement = 'H'
  )
@

\noindent Some of these graphemes are single codepoints, others are combinations
of two codepoints. By specifying the splitting separator as empty, it is
possible to split the string into Unicode codepoints.

<<>>=
profile_2 <- write.profile(example, sep = "")
@

<<results = 'asis', echo = FALSE>>=
colnames(profile_2) <- c('Grapheme', 'Frequency', 'Codepoint', 'Unicode Name')
print(xtable(profile_2
        , caption = 'Profile 2 (splitting by codepoints)'
        , label = 'tab:profile2'
        )
  , include.rownames = FALSE
  , booktabs = TRUE
  , size = 'scriptsize'
  , table.placement = 'H'
  )
@

\noindent Some characters look identical, although they are encoded differently. 
Unicode offers different ways of normalization, which can be invoked here as 
well (see Section~\ref{pitfall-canonical-equivalence}).

<<>>=
# after NFC normalization unicode codepoints have changed
profile_3 <- write.profile(example, normalize = "NFC", sep = "")
# NFD normalization gives yet another structure of the codepoints
profile_4 <- write.profile(example, normalize = "NFD", sep = "")
@

<<results = 'asis', echo = FALSE>>=
colnames(profile_3) <- c('Grapheme', 'Frequency', 'Codepoint', 'Unicode Name')
print(xtable(profile_3
        , caption = 'Profile 3 (splitting by NFC codepoints)'
        , label = 'tab:profile3'
        )
  , include.rownames = FALSE
  , booktabs = TRUE
  , size = 'scriptsize'
  , table.placement = 'H'
  )
@

<<results = 'asis', echo = FALSE>>=
colnames(profile_4) <- c('Grapheme', 'Frequency', 'Codepoint', 'Unicode Name')
print(xtable(profile_4
        , caption = 'Profile 4 (splitting by NFD codepoints)'
        , label = 'tab:profile4'
        )
  , include.rownames = FALSE
  , booktabs = TRUE
  , size = 'scriptsize'
  , table.placement = 'H'
  )
@

\noindent It is important to realize that for Unicode grapheme definitions, NFC 
and NFD normalization are equivalent. This can be shown by normalizing the 
example in either NFC or NFD, but using default separation in 
\texttt{write.profile}. To be precise, default separation means setting \texttt{sep = NULL} 
but that has not be added explicitly.

<<tidy = FALSE>>=
# note that NFC and NFD normalization are identical
# for unicode grapheme definitions
profile_5 <- write.profile(example, normalize = "NFD")
profile_6 <- write.profile(example, normalize = "NFC")
@

<<results = 'asis', echo = FALSE>>=
colnames(profile_5) <- c('Gr.', 'Freq.', 'Codepoint', 'Unicode Name')
print(xtable(profile_5
        , caption = 'Profile 5 (splitting by graphemes after NFD)'
        , label = 'tab:profile5'
        )
  , include.rownames = FALSE
  , booktabs = TRUE
  , size = 'scriptsize'
  , table.placement = 'H'
  )
@

<<results = 'asis', echo = FALSE>>=
colnames(profile_6) <- c('Gr.', 'Freq.', 'Codepoint', 'Unicode Name')
print(xtable(profile_6
        , caption = 'Profile 6 (splitting by graphemes after NFC)'
        , label = 'tab:profile6'
        )
  , include.rownames = FALSE
  , booktabs = TRUE
  , size = 'scriptsize'
  , table.placement = 'H'
  )
@

\noindent Note that these different profiles can also be produced using the bash 
executable \texttt{writeprofile} (see \ref{implementations} for notes on how to 
install the bash executable). Exactly this example is also included in the help 
file of the executable.

<<size='scriptsize', engine='bash'>>=
## show help
writeprofile --help
@

\section{Using an orthography profile skeleton}
\label{profile-skeleton}

A common workflow to use these functions is to first make a skeleton for an
orthography profile and then edit this profile by hand. For example, Table
\ref{tab:profile_skeleton1} shows the profile skeleton after a few graphemes have
been added to the file. Note that in this example, the profile is written to the
desktop, and this file has to be edited manually. We simply add a few
multigraphs to the column \texttt{Grapheme} and leave the other columns empty.
these new graphemes are then included in the graphemic parsing.

<<eval=FALSE>>=
# a few words to be graphemically parsed
example <- c("mishmash", "mishap", "mischief", "scheme")
# write a profile skeleton to a file
write.profile(example, file = "~/Desktop/profile_skeleton.txt")
# edit the profile, and then use the edited profile to tokenize
tokenize(example, profile = "~/Desktop/profile_skeleton.txt")$strings
@

<<echo=FALSE>>=
example <- c("shampoo", "mishap", "mischief", "scheme")
profile_skeleton <- write.profile(example)
profile_skeleton <- rbind( c('sh','','','')
                         , c('ch','','','')
                         , c('sch','','','')
                         , c('ie','','','')
                         , c('oo','','','')
                         , profile_skeleton
                         )
tokenize(example, profile = profile_skeleton)$strings
@

<<results='asis', echo=FALSE>>=
print(xtable(profile_skeleton
        , caption = 'Manually edited profile skeleton'
        , label = 'tab:profile_skeleton1'
        )
  , include.rownames = FALSE
  , booktabs = TRUE
  , size = 'scriptsize'
  , table.placement = 'htb'
  )
@

For quick solutions, it is also possible to leave out the Unicode information in
the profile skeleton by using the option \texttt{info = FALSE}. It is also
possible not to use a separate file at all, but process everything within R. In
simply situations this is often useful (see below), but in general we prefer to
handle everything through a separately saved orthography profile. This profile
often contains highly useful information that is nicely coded and saved inside
this one file, and can thus be easily distributed and shared. Doing the same as
above completely within R might look as follows:

<<>>=
# make a profile, just select the column 'Grapheme'
profile <- write.profile(example)[,"Grapheme"]
# extend the profile with multigraphs
profile <- c("sh", "ch", "sch", "ie", "oo", profile)
# use the profile to tokenize
tokenize(example, profile)$strings
@

To document a specific case of graphemic parsing, it is highly useful to save
all results of the tokenization to file by using the option \texttt{file.out},
for example as follows:

<<eval=FALSE, tidy=FALSE>>= 
# save the results to various files
tokenize( example
         , profile = "~/Desktop/profile_skeleton.txt"
         , file.out = "~/Desktop/result"
        )
@

\noindent This will lead to the following four files being written:

\begin{itemize}

   \item \textsc{result\_strings.tsv}:\\ A tab-separated file with the original
         and the tokenized/transliterated strings

   \item \textsc{result\_profile.tsv}:\\ A tab-separated file with the
         graphemes with added frequencies of occurence in the data. The lines
         in the file is re-ordered according to the order that resulted from the
         ordering specifications.

   \item \textsc{result\_errors.tsv}:\\ A tab-separated file with all original
         strings that contain unmatched parts. Unmatched parts are indicated
         with the character as specified with the option \texttt{missing}. By
         default the character \textsc{double question mark} <⁇> at
         \uni{2047} is used. When there are no errors, then this file is 
         absent.

    \item \textsc{result\_missing.tsv}:\\ A tab-separated file with the graphemes
          that are missing from the original orthography profile, as indicated in
          the errors. When there are no errors, then this file is absent.

\end{itemize}

\section{Rule ordering}
\label{rule-ordering}

Not yet everything is correct in this graphemic parsing. The sequence <sh> in
`mishap' should not be a digraph, and likewise the sequence <sch> in `mischief'
should of course be separated into <s> and <ch>. One of the important issues to
get the graphemic parsing right is the order in which graphemes are parsed. For
example, currently the grapheme <sch> is parsed before the grapheme <ch>,
leading to <m\ i\ sch\ ie\ f> instead of the intended <m\ i\ s\ ch\ ie\ f>. The
reason that <sch> is parsed before <ch> is that by default longer graphemes are
parsed before shorter ones. Our experience is that in most cases this is
expected behaviour. You can change the ordering by specifying the option
\texttt{ordering}. Setting this option to \texttt{NULL} results in no
preferential ordering, i.e. the graphemes are parsed in the order of the
profile, from top to bottom. Now `mischief' is parsed correctly, but `scheme' is
wrong. So ordering is not the solution in this case.

<<eval=FALSE, tidy=FALSE>>= 
# do not reorder the profile
# just apply the graphemes from top to bottom
tokenize( example
         , profile = "~/Desktop/profile_skeleton.txt"
         , ordering = NULL
        )$strings
@

<<echo=FALSE>>=
tokenize( example, profile = profile_skeleton, ordering = NULL)$strings
@

There are various additional options for rule ordering implemented. Please check
the help description in R \texttt{help(tokenize)} for more details on the
possible rule ordering specificiations. In summary, there are four different 
ordering options, that can also be combined:

\begin{itemize}

   \item \textsc{size}\\
         order the lines in the profile by the size of the
         grapheme, largest first. Size is measured by number of Unicode
         characters after normalization as specified in the option
         \texttt{normalize}. For example, <é> has a size of 1 with
         \texttt{normalize = "NFC"}, but a size of 2 with
         \texttt{normalize = "NFD"}.

   \item \textsc{context}\\
         order the lines by whether they have any context
         specified (see next section). Lines with context coming first. Note
         that this only works when the option \texttt{regex = TRUE} is also chosen.

  \item \textsc{reverse}\\
        order the lines from bottom to top.

  \item \textsc{frequency}\\
         order the lines by the frequency with which they
         match in the specified strings before tokenization, least frequent
         coming first. This frequency of course depends crucially on the
         available strings, so it will lead to different orderings when applied
         to different data. Also note that this frequency is (necessarily)
         measured before graphemes are identified, so these ordering frequencies
         are not the same as the final frequencies shown in the output.
         Frequency of course also strongly differs on whether context is used
         for the matching through \texttt{regex = TRUE}.

\end{itemize}

By specifying more than one ordering, these orderings are used to break ties,
e.g. the default setting \texttt{ordering = c("size", "context", "reverse")}
will first order by size, and for those with the same size, it will order by
whether there is any context specified or not. For lines that are still tied
(i.e. the have the same size and both/neither have context) the order will be
reversed compared to the order as attested in the profile. Reversing order can
be useful because hand-written profiles tend to put general rules before
specific rules, which mostly should be applied in reverse order.

It is important to realize that different ordering of the rules does not have
so-called `feeding' and `bleeding' effects as known from finite-state rewrite
rules.\footnote{`Bleeding' is the effect that the application of a rule changes
the string, so as to prevent a next rule to apply. `Feeding' is the opposite: a
specific rule will only be applied becuase a previous rule changed the string
already. The interaction of rules with such `feeding' and `bleeding' effects is
extremely difficult to predict.} The graphemic parsing advocated here works
crucially different from rewrite rules in that there is nothing being rewritten:
each line in an orthography profile specifies a grapheme to be `captured' in the 
string. All lines in the profile are processed in a specifed order (as determined
by the option \texttt{ordering}). At the processing of a specific line, all 
matching graphemes in the data are marked as `captured', but not changed. 
Captured parts cannot be captured again, but they can still be used to match 
contexts of other lines in the profile. Only when all lines are processed the 
captured graphemes are separated (and possibly transliterated). In this way the 
result of the application of the `rules' is rather easy to predict.

To document the process of graphemic tokenization, it is strongly encouraged to

\section{Contextually specified graphemes}
\label{contextual-specification}

To refine a profile, it is also possible to add graphemes with contextual
specifications. An orthography profile can have columns called \texttt{Left} and
\texttt{Right} to specify the context in which the Grapheme is to be
separated.\footnote{The columns names \textttf{Left}, \textttf{Right} and
\textttf{Grapheme} are currently hard coded, so exactly these column names
should be used for these effects to take place. The position of the columns in
the profile is unimportant. So the column \textttf{Left} can occur anywhere.} We
are adding an extra line to the profile specifying that <s> is a grapheme when
it occurs after <mi>. These contextually specified graphemes are based on
regular expressions (so you can actually use regular expressions in the
description of the context). For these to be included in the graphemic parsing 
we have to specify the option \texttt{regex = TRUE}. This contextually specified 
grapheme should actually be handled first, so we could try \texttt{ordering = 
NULL}. However, we can also explicitly specifiy that rules with contextual 
information should be applied first by using \texttt{ordering = "context"}. That 
gives the right results for this toy example.

<<eval=FALSE, tidy=FALSE>>=
# add a contextual grapheme, and then use the edited profile to tokenize
tokenize( example
         , profile = "~/Desktop/profile_skeleton.txt"
         , regex = TRUE
         , ordering = "context"
        )$strings
@

<<echo=FALSE>>=
Left <- rep('', times = nrow(profile_skeleton))
profile_skeleton <- as.matrix(cbind(Left, profile_skeleton))
profile_skeleton <- rbind(c('mi','s','','',''), profile_skeleton)
tokenize(example, profile = profile_skeleton, regex = TRUE, order='context')$strings
@

<<results='asis', echo=FALSE>>=
print(xtable(profile_skeleton
        , caption = 'Orthography profile with contextual specification for <s>'
        , label = 'tab:profile_skeleton2'
        )
  , include.rownames = FALSE
  , booktabs = TRUE
  , size = 'scriptsize'
  , table.placement = 'htb'
  )
@

\section{Profile skeleton with columns for editing}
\label{profile-editing})

When it is expected that context might be important for a profile, then the 
profile skeleton can be written to a file with columns prepared for the 
contextual specifications by using the option \texttt{editing = TRUE}.

### 2016-06-30 book/unicode.tex
adding R examples

@@
 \include{chapters/ipa_meets_unicode}
-
 \include{chapters/orthography_profiles}
 
+\chapter{Implementation}
+\label{implementation}
+
 \section{Available implementations}
-\label{implementations}
 
-To illustrate the practical application of orthography profiles, we have
-implemented two different versions of the above specifications, one in Python
-and one in R. These two implementations have rather different implementation
-histories, and we are not 100\% sure that they will in all situations give the
-same results. Also the perfomance with larger datasets differs, and the code is
-not always as clean as we would like it to be. In sum, the two implementations
-should be considered as `proof of concept' and not as the final word on the
-practical application of the specifications above. In our own experience, the
-current implementations are sufficiently fast and stable to be useful for
-academic practice (e.g. checking data consistency, or analyzing and
-transliterating small to medium sized data sets), but they should probably
-better not be used for full-scale industry applications.
+To illustrate the practical application of orthography profiles as defined in
+Chapter \ref{orthography-profiles}, we have implemented two different versions
+of the specifications presented there: one in Python and one in R. These two
+implementations have rather different implementation histories, and we are not
+100\% sure that they will in all situations give the same results. Also the
+perfomance with larger datasets differs, and the code is not always as clean as
+we would like it to be. In sum, the two implementations should be considered as
+`proof of concept' and not as the final word on the practical application of the
+specifications above. In our own experience, the current implementations are
+sufficiently fast and stable to be useful for academic practice (e.g. checking
+data consistency, or analyzing and transliterating small to medium sized data
+sets), but they should probably better not be used for full-scale industry
+applications.
+
+This chapter will introduce the implementations, and give practical step-by-step 
+guidelines for installing them and using them. Various simple, and sometimes 
+somewhat abstract, examples will be discussed to show the different options 
+available, and to illustrate the intended usage of orthography profiles. In the 
+next chapter we will then discuss a few full-scale examples to illustrate the 
+application and the usefulness of orthography profiles.
 
@@
 (possibly edited) profile on some data, as exemplified in the next section. This
-can of course be performed inside R, but additionally there are two more
+can of course be performed though R, but additionally there are two more
 interfaces to the R code supplied in the \texttt{qlcData} package: (i) bash
@@
 PATH, e.g. to \texttt{/usr/local/bin/}. The two executables are named
-\texttt{tokenize} and \texttt{writeprofile}
+\texttt{tokenize} and \texttt{writeprofile}, and the links can be made directly 
+by using Rscript to get the paths to the executables within the terminal.
 
@@
 \definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\noindent
-\ttfamily
-\hlstd{}\hlslc{\#\ make\ two\ softlinks\ to\ the\ R\ executables\ in\ bash}\hspace*{\fill}\\
-\hlstd{}\hlkwc{ln\ }\hlstd{}\hlopt{{-}}\hlstd{is\ }\hlstr{`Rscript\ {-}e\ 'cat(file.path(find.package("qlcData"),}\hspace*{\fill}\\
-\hlstr{\ "exec",\ "tokenize"))'`}\hlstd{\ }\hlopt{/}\hlstd{usr}\hlopt{/}\hlstd{}\hlkwb{local}\hlstd{}\hlopt{/}\hlstd{bin}\hspace*{\fill}\\
-\hlkwc{ln\ }\hlstd{}\hlopt{{-}}\hlstd{is\ }\hlstr{`Rscript\ {-}e\ 'cat(file.path(find.package("qlcData"),}\hspace*{\fill}\\
-\hlstr{\ "exec",\ "writeprofile"))'`}\hlstd{\ }\hlopt{/}\hlstd{usr}\hlopt{/}\hlstd{}\hlkwb{local}\hlstd{}\hlopt{/}\hlstd{bin}\hspace*{\fill}
-\mbox{}
-\normalfont
+\begin{alltt}
+# get the paths to the R executables in bash
+pathT=`Rscript -e 'cat(file.path(find.package("qlcData"), "exec", "tokenize"))'`
+pathW=`Rscript -e 'cat(file.path(find.package("qlcData"), "exec", "writeprofile"))'`
+
+# make softlinks to the R executables in /usr/local/bin
+ln -is $pathT $pathW /usr/local/bin
+\end{alltt}
 \end{kframe}
@@
 
-\noindent After this addition it should be possible to access the \texttt{tokenize}
+\noindent After this softlink it should be possible to access the \texttt{tokenize}
 function from the shell. Try \texttt{tokenize --help} to test the functionality.
 
-\begin{knitrout}\tiny
+\begin{knitrout}\scriptsize
 \definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\noindent
-\ttfamily
-\hlstd{tokenize\ }\hlopt{{-}{-}}\hlstd{}\hlkwb{help}\hlstd{}\hspace*{\fill}
-\mbox{}
-\normalfont
+\begin{alltt}
+tokenize --help
+\end{alltt}
 
@@
 
+TO DO
+
 \subsection*{Installung the Python implementation}
@@
 
-\section{Abstract examples}
-
-\subsection*{Error reporting}
+\section{Error reporting}
+\label{error-reporting}
 
@@
 
-\noindent The working of error-message reporting can also nicely be illustrated with this
-example. Suppose we made an orthography profile with just the Latin <A> and <B> 
-as possible graphemes, then this profile would not be sufficient to tokenize the 
-strings. There are graphemes in the data that are not in the profile, so the 
-tokenization produces an error, which can be used to fix the encoding. In the 
-example below, we can see that the Cyrillic encoding is found in the second 
-string of the \texttt{test} input.
+\noindent The working of error-message reporting can also nicely be illustrated
+with this example. Suppose we made an orthography profile with just the two
+Latin letters <A> and <B> as possible graphemes, then this profile would not be
+sufficient to tokenize the strings. There are graphemes in the data that are not
+in the profile, so the tokenization produces an error, which can be used to fix
+the encoding. In the example below, we can see that the Cyrillic encoding is
+found in the second string of the \texttt{test} input.
 
@@
 
-\subsection*{Different ways to write a profile}
+\section{Different ways to write a profile}
+\label{write-profile}
 
-The preparation of an orthography profile from some data might sound like a 
-trivial problem, but actually there are various different ways in which strings 
-can be separated into graphemes by \texttt{write.profile}. Consider the 
-following example string. The default settings of \texttt{write.profile} 
-separates the string into Unicode graphemes according to grapheme clusters 
-(`user-perceived characters', see Section~\ref{the-unicode-approach}).
+The function \texttt{write.profile} can be used to prepare a skeleton for an
+orthography profile from some data. The preparation of an orthography profile
+from some data might sound like a trivial problem, but actually there are
+various different ways in which strings can be separated into graphemes by
+\texttt{write.profile}. Consider the following example string. The default
+settings of \texttt{write.profile} separates the string into Unicode graphemes
+according to grapheme clusters (`user-perceived characters', see
+Section~\ref{the-unicode-approach}).
 
@@
 % latex table generated in R 3.3.0 by xtable 1.8-2 package
-% Wed Jun 22 12:03:05 2016
+% Thu Jun 30 18:15:48 2016
 \begin{table}[H]
@@
 \noindent Some of these graphemes are single codepoints, others are combinations
-of two codepoints. By specifying the separator as empty, it is possible to split
-the string into codepoints.
+of two codepoints. By specifying the splitting separator as empty, it is
+possible to split the string into Unicode codepoints.
 
@@
 \begin{alltt}
-\hlcom{# note that NFC and NFD normalization are identical for unicode grapheme}
-\hlcom{# definitions}
+\hlcom{# note that NFC and NFD normalization are identical}
+\hlcom{# for unicode grapheme definitions}
 \hlstd{profile_5} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example,} \hlkwc{normalize} \hlstd{=} \hlstr{"NFD"}\hlstd{)}
@@
 % latex table generated in R 3.3.0 by xtable 1.8-2 package
-% Wed Jun 22 12:03:05 2016
+% Thu Jun 30 18:15:49 2016
 \begin{table}[H]
@@
 
-\begin{knitrout}\tiny
+\begin{knitrout}\scriptsize
 \definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\noindent
-\ttfamily
-\hlstd{writeprofile\ }\hlopt{{-}{-}}\hlstd{}\hlkwb{help}\hlstd{}\hspace*{\fill}
-\mbox{}
-\normalfont
+\begin{alltt}
+## show help
+writeprofile --help
+\end{alltt}
 
@@
 
+\section{Using an orthography profile skeleton}
+\label{profile-skeleton}
+
+A common workflow to use these functions is to first make a skeleton for an
+orthography profile and then edit this profile by hand. For example, Table
+\ref{tab:profile_skeleton1} shows the profile skeleton after a few graphemes have
+been added to the file. Note that in this example, the profile is written to the
+desktop, and this file has to be edited manually. We simply add a few
+multigraphs to the column \texttt{Grapheme} and leave the other columns empty.
+these new graphemes are then included in the graphemic parsing.
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# a few words to be graphemically parsed}
+\hlstd{example} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"mishmash"}\hlstd{,} \hlstr{"mishap"}\hlstd{,} \hlstr{"mischief"}\hlstd{,} \hlstr{"scheme"}\hlstd{)}
+\hlcom{# write a profile skeleton to a file}
+\hlkwd{write.profile}\hlstd{(example,} \hlkwc{file} \hlstd{=} \hlstr{"~/Desktop/profile_skeleton.txt"}\hlstd{)}
+\hlcom{# edit the profile, and then use the edited profile to tokenize}
+\hlkwd{tokenize}\hlstd{(example,} \hlkwc{profile} \hlstd{=} \hlstr{"~/Desktop/profile_skeleton.txt"}\hlstd{)}\hlopt{$}\hlstd{strings}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+##   originals    tokenized
+## 1   shampoo  sh a m p oo
+## 2    mishap   m i sh a p
+## 3  mischief m i sch ie f
+## 4    scheme    sch e m e
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+% latex table generated in R 3.3.0 by xtable 1.8-2 package
+% Thu Jun 30 18:15:49 2016
+\begin{table}[htb]
+\centering
+\begingroup\scriptsize
+\begin{tabular}{llll}
+  \toprule
+Grapheme & Frequency & Codepoint & UnicodeName \\ 
+  \midrule
+sh &  &  &  \\ 
+  ch &  &  &  \\ 
+  sch &  &  &  \\ 
+  ie &  &  &  \\ 
+  oo &  &  &  \\ 
+  a & 2 & U+0061 & LATIN SMALL LETTER A \\ 
+  c & 2 & U+0063 & LATIN SMALL LETTER C \\ 
+  e & 3 & U+0065 & LATIN SMALL LETTER E \\ 
+  f & 1 & U+0066 & LATIN SMALL LETTER F \\ 
+  h & 4 & U+0068 & LATIN SMALL LETTER H \\ 
+  i & 3 & U+0069 & LATIN SMALL LETTER I \\ 
+  m & 4 & U+006D & LATIN SMALL LETTER M \\ 
+  o & 2 & U+006F & LATIN SMALL LETTER O \\ 
+  p & 2 & U+0070 & LATIN SMALL LETTER P \\ 
+  s & 4 & U+0073 & LATIN SMALL LETTER S \\ 
+   \bottomrule
+\end{tabular}
+\endgroup
+\caption{Manually edited profile skeleton} 
+\label{tab:profile_skeleton1}
+\end{table}
+
+For quick solutions, it is also possible to leave out the Unicode information in
+the profile skeleton by using the option \texttt{info = FALSE}. It is also
+possible not to use a separate file at all, but process everything within R. In
+simply situations this is often useful (see below), but in general we prefer to
+handle everything through a separately saved orthography profile. This profile
+often contains highly useful information that is nicely coded and saved inside
+this one file, and can thus be easily distributed and shared. Doing the same as
+above completely within R might look as follows:
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# make a profile, just select the column 'Grapheme'}
+\hlstd{profile} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example)[,} \hlstr{"Grapheme"}\hlstd{]}
+\hlcom{# extend the profile with multigraphs}
+\hlstd{profile} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"sh"}\hlstd{,} \hlstr{"ch"}\hlstd{,} \hlstr{"sch"}\hlstd{,} \hlstr{"ie"}\hlstd{,} \hlstr{"oo"}\hlstd{, profile)}
+\hlcom{# use the profile to tokenize}
+\hlkwd{tokenize}\hlstd{(example, profile)}\hlopt{$}\hlstd{strings}
+\end{alltt}
+\begin{verbatim}
+##   originals    tokenized
+## 1   shampoo  sh a m p oo
+## 2    mishap   m i sh a p
+## 3  mischief m i sch ie f
+## 4    scheme    sch e m e
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+To document a specific case of graphemic parsing, it is highly useful to save
+all results of the tokenization to file by using the option \texttt{file.out},
+for example as follows:
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# save the results to various files}
+\hlkwd{tokenize}\hlstd{( example}
+         \hlstd{,} \hlkwc{profile} \hlstd{=} \hlstr{"~/Desktop/profile_skeleton.txt"}
+         \hlstd{,} \hlkwc{file.out} \hlstd{=} \hlstr{"~/Desktop/result"}
+        \hlstd{)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+This will lead to the following four files being written:
+
+\begin{itemize}
+
+   \item \textsc{result_strings.tsv}:\\ A tab-separated file with the original
+         and the tokenized/transliterated strings
+
+   \item \textsc{result_profile.tsv}:\\ A tab-separated file with the
+         graphemes with added frequencies of occurence in the data. The lines
+         in the file is re-ordered according to the order that resulted from the
+         ordering specifications.
+
+   \item \textsc{result_errors.tsv}:\\ A tab-separated file with all original
+         strings that contain unmatched parts. Unmatched parts are indicated
+         with the character as specified with the option \texttt{missing}. By
+         default the character \textsc{double question mark} < ⁇ > at
+         \uni{2047} is used. When there are no errors, then this file is 
+         absent.
+
+    \item \textsc{result_missing.tsv}:\\ A tab-separated file with the graphemes
+          that are missing from the original orthography profile, as indicated in
+          the errors. When there are no errors, then this file is absent.
+
+\end{itemize}
+
+\section{Rule ordering}
+\label{rule-ordering}
+
+Not yet everything is correct in this graphemic parsing. The sequence <sh> in
+`mishap' should not be a digraph, and likewise the sequence <sch> in `mischief'
+should of course be separated into <s> and <ch>. One of the important issues to
+get the graphemic parsing right is the order in which graphemes are parsed. For
+example, currently the grapheme <sch> is parsed before the grapheme <ch>,
+leading to <m\ i\ sch\ ie\ f> instead of the intended <m\ i\ s\ ch\ ie\ f>. The
+reason that <sch> is parsed before <ch> is that by default longer graphemes are
+parsed before shorter ones. Our experience is that in most cases this is
+expected behaviour. You can change the ordering by specifying the option
+\texttt{ordering}. Setting this option to \texttt{NULL} results in no
+preferential ordering, i.e. the graphemes are parsed in the order of the
+profile, from top to bottom. Now `mischief' is parsed correctly, but `scheme' is
+wrong. So ordering is not the solution in this case.
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# do not reorder the profile}
+\hlcom{# just apply the graphemes from top to bottom}
+\hlkwd{tokenize}\hlstd{( example}
+         \hlstd{,} \hlkwc{profile} \hlstd{=} \hlstr{"~/Desktop/profile_skeleton.txt"}
+         \hlstd{,} \hlkwc{ordering} \hlstd{=} \hlkwa{NULL}
+        \hlstd{)}\hlopt{$}\hlstd{strings}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+##   originals     tokenized
+## 1   shampoo   sh a m p oo
+## 2    mishap    m i sh a p
+## 3  mischief m i s ch ie f
+## 4    scheme    s ch e m e
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+There are various additional options for rule ordering implemented. Please check
+the help description in R \texttt{help(tokenize)} for more details on the
+possible rule ordering specificiations. In summary, there are four different 
+ordering options, that can also be combined:
+
+\begin{itemize}
+
+   \item \textsc{size}:\\
+         order the lines in the profile by the size of the
+         grapheme, largest first. Size is measured by number of Unicode
+         characters after normalization as specified in the option
+         \texttt{normalize}. For example, <é> has a size of 1 with
+         \texttt{normalize = "NFC"}, but a size of 2 with
+         \texttt{normalize = "NFD"}.
+
+   \item \textsc{context}:\\
+         order the lines by whether they have any context
+         specified (see next section). Lines with context coming first. Note
+         that this only works when the option \texttt{regex = TRUE} is also chosen.
+
+  \item \textsc{reverse}:\\
+        order the lines from bottom to top.
+
+  \item \textsc{frequency}:\\
+         order the lines by the frequency with which they
+         match in the specified strings before tokenization, least frequent
+         coming first. This frequency of course depends crucially on the
+         available strings, so it will lead to different orderings when applied
+         to different data. Also note that this frequency is (necessarily)
+         measured before graphemes are identified, so these ordering frequencies
+         are not the same as the final frequencies shown in the output.
+         Frequency of course also strongly differs on whether context is used
+         for the matching through \texttt{regex = TRUE}.
+
+\end{itemize}
+
+By specifying more than one ordering, these orderings are used to break ties,
+e.g. the default setting \texttt{ordering = c("size", "context", "reverse")}
+will first order by size, and for those with the same size, it will order by
+whether there is any context specified or not. For lines that are still tied
+(i.e. the have the same size and both/neither have context) the order will be
+reversed compared to the order as attested in the profile. Reversing order can
+be useful because hand-written profiles tend to put general rules before
+specific rules, which mostly should be applied in reverse order.
+
+It is important to realize that different ordering of the rules does not have
+so-called `feeding' and `bleeding' effects as known from finite-state rewrite
+rules.\footnote{`Bleeding' is the effect that the application of a rule changes
+the string, so as to prevent a next rule to apply. `Feeding' is the opposite: a
+specific rule will only be applied becuase a previous rule changed the string
+already. The interaction of rules with such `feeding' and `bleeding' effects is
+extremely difficult to predict.} The graphemic parsing advocated here works
+crucially different from rewrite rules in that there is nothing being rewritten:
+each line in an orthography profile specifies a grapheme to be `captured' in the 
+string. All lines in the profile are processed in a specifed order (as determined
+by the option \texttt{ordering}). At the processing of a specific line, all 
+matching graphemes in the data are marked as `captured', but not changed. 
+Captured parts cannot be captured again, but they can still be used to match 
+contexts of other lines in the profile. Only when all lines are processed the 
+captured graphemes are separated (and possibly transliterated). In this way the 
+result of the application of the `rules' is rather easy to predict.
+
+To document the process of graphemic tokenization, it is strongly encouraged to
+
+\section{Contextually specified graphemes}
+\label{contextual-specification}
+
+To refine a profile, it is also possible to add graphemes with contextual
+specifications. An orthography profile can have columns called \texttt{Left} and
+\texttt{Right} to specify the context in which the Grapheme is to be
+separated.\footnote{The columns names \textttf{Left}, \textttf{Right} and
+\textttf{Grapheme} are currently hard coded, so exactly these column names
+should be used for these effects to take place. The position of the columns in
+the profile is unimportant. So the column \textttf{Left} can occur anywhere.} We
+are adding an extra line to the profile specifying that <s> is a grapheme when
+it occurs after <mi>. These contextually specified graphemes are based on
+regular expressions (so you can actually use regular expressions in the
+description of the context). For these to be included in the graphemic parsing 
+we have to specify the option \texttt{regex = TRUE}. This contextually specified 
+grapheme should actually be handled first, so we could try \texttt{ordering = 
+NULL}. However, we can also explicitly specifiy that rules with contextual 
+information should be applied first by using \texttt{ordering = "context"}. That 
+gives the right results for this toy example.
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# add a contextual grapheme, and then use the edited profile to tokenize}
+\hlkwd{tokenize}\hlstd{( example}
+         \hlstd{,} \hlkwc{profile} \hlstd{=} \hlstr{"~/Desktop/profile_skeleton.txt"}
+         \hlstd{,} \hlkwc{regex} \hlstd{=} \hlnum{TRUE}
+         \hlstd{,} \hlkwc{ordering} \hlstd{=} \hlstr{"context"}
+        \hlstd{)}\hlopt{$}\hlstd{strings}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+##   originals     tokenized
+## 1   shampoo   sh a m p oo
+## 2    mishap   m i s h a p
+## 3  mischief m i s ch ie f
+## 4    scheme    s ch e m e
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+% latex table generated in R 3.3.0 by xtable 1.8-2 package
+% Thu Jun 30 18:15:49 2016
+\begin{table}[htb]
+\centering
+\begingroup\scriptsize
+\begin{tabular}{lllll}
+  \toprule
+Left & Grapheme & Frequency & Codepoint & UnicodeName \\ 
+  \midrule
+mi & s &  &  &  \\ 
+   & sh &  &  &  \\ 
+   & ch &  &  &  \\ 
+   & sch &  &  &  \\ 
+   & ie &  &  &  \\ 
+   & oo &  &  &  \\ 
+   & a & 2 & U+0061 & LATIN SMALL LETTER A \\ 
+   & c & 2 & U+0063 & LATIN SMALL LETTER C \\ 
+   & e & 3 & U+0065 & LATIN SMALL LETTER E \\ 
+   & f & 1 & U+0066 & LATIN SMALL LETTER F \\ 
+   & h & 4 & U+0068 & LATIN SMALL LETTER H \\ 
+   & i & 3 & U+0069 & LATIN SMALL LETTER I \\ 
+   & m & 4 & U+006D & LATIN SMALL LETTER M \\ 
+   & o & 2 & U+006F & LATIN SMALL LETTER O \\ 
+   & p & 2 & U+0070 & LATIN SMALL LETTER P \\ 
+   & s & 4 & U+0073 & LATIN SMALL LETTER S \\ 
+   \bottomrule
+\end{tabular}
+\endgroup
+\caption{Orthography profile with contextual specification for <s>} 
+\label{tab:profile_skeleton2}
+\end{table}
+
+\section{Profile skeleton with columns for editing}
+\label{profile-editing})
+
+When it is expected that context might be important for a profile, then the 
+profile skeleton can be written to a file with columns prepared for the 
+contextual specifications by using the option \texttt{editing = TRUE}.
+
 \include{chapters/use_cases}
@@
 \definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\noindent
-\ttfamily
-\hlstd{}\hlkwb{echo\ }\hlstd{}\hlstr{'this\ is\ a\ simple\ bash\ example'}\hlstd{}\hspace*{\fill}\\
-\hlkwc{ls\ }\hlstd{\textbar \ }\hlkwc{wc}\hlstd{}\hspace*{\fill}
-\mbox{}
-\normalfont
+\begin{alltt}
+echo 'this is a simple bash example'
+ls | wc
+\end{alltt}
 
@@
 ## this is a simple bash example
-##       11      11     184
+##       12      12     203
 \end{verbatim}
@@
 % latex table generated in R 3.3.0 by xtable 1.8-2 package
-% Wed Jun 22 10:58:23 2016
+% Thu Jun 30 18:15:49 2016
 \begin{table}[H]
@@
 \definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\noindent
-\ttfamily
-\hlstd{}\hlkwa{print}\hlstd{}\hlopt{(}\hlstd{}\hlstr{'bla'}\hlstd{\ }\hlopt{+\ }\hlstd{}\hlstr{'bla'}\hlstd{}\hlopt{)}\hspace*{\fill}\\
-\hlstd{}\hlkwa{print}\hlstd{}\hlopt{(}\hlstd{}\hlnum{3}\hlstd{}\hlopt{+}\hlstd{}\hlnum{4}\hlstd{}\hlopt{)}\hlstd{}\hspace*{\fill}
-\mbox{}
-\normalfont
+\begin{alltt}
+print('bla' + 'bla')
+print(3+4)
+\end{alltt}
 
@@
 \definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\noindent
-\ttfamily
-\hlstd{}\hlkwa{print}\hlstd{}\hlopt{(}\hlstd{}\hlstr{'python3\ needs\ an\ extra\ tag'}\hlstd{}\hlopt{)}\hspace*{\fill}\\
-\hlstd{uni\ }\hlopt{=\ }\hlstd{}\hlstr{'aɽɮz'}\hlstd{}\hspace*{\fill}\\
-\hlkwa{print}\hlstd{}\hlopt{(}\hlstd{uni}\hlopt{)}\hlstd{}\hspace*{\fill}
-\mbox{}
-\normalfont
+\begin{alltt}
+print('python3 needs an extra tag')
+uni = 'aɽɮz'
+print(uni)
+\end{alltt}
 

### 2016-08-18 book/unicode.tex
update general files

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% Everything following a % is ignored
-% Some lines start with %. Remove the % to include them
 
@@
 % mean, you can safely ignore this section
+
 \usepackage{verbatim}
@@
 \frontmatter
-% %% uncomment if you have preface and/or acknowledgements
 % \include{chapters/preface}
@@
 \section{Available implementations}
+\label{available-implementations}
 
 To illustrate the practical application of orthography profiles as defined in
-Chapter \ref{orthography-profiles}, we have implemented two different versions
-of the specifications presented there: one in Python and one in R. These two
-implementations have rather different implementation histories, and we are not
-100\% sure that they will in all situations give the same results. Also the
-perfomance with larger datasets differs, and the code is not always as clean as
-we would like it to be. In sum, the two implementations should be considered as
-`proof of concept' and not as the final word on the practical application of the
-specifications above. In our own experience, the current implementations are
-sufficiently fast and stable to be useful for academic practice (e.g. checking
-data consistency, or analyzing and transliterating small to medium sized data
-sets), but they should probably better not be used for full-scale industry
-applications.
+the previous Chapter \ref{orthography-profiles}, we have implemented two
+different versions of the specifications presented there: one in Python and one
+in R. These two implementations have rather different implementation histories,
+and we are not 100\% sure that they will in all situations give the same
+results. Also the perfomance with larger datasets differs, and the code is not
+always as clean as we would like it to be. In sum, the two implementations
+should be considered as `proof of concept' and not as the final word on the
+practical application of the specifications above. In our own experience, the
+current implementations are sufficiently fast and stable to be useful for
+academic practice (e.g. checking data consistency, or analyzing and
+transliterating small to medium sized data sets), but they should probably
+better not be used for full-scale industry applications.
 
 This chapter will introduce the implementations, and give practical step-by-step 
-guidelines for installing them and using them. Various simple, and sometimes 
-somewhat abstract, examples will be discussed to show the different options 
+guidelines for installing them and using them. Various simple and sometimes 
+somewhat abstract examples will be discussed to show the different options 
 available, and to illustrate the intended usage of orthography profiles. In the 
@@
 
-\noindent The version available through CRAN is the most recent stable version.
+\noindent The version available through CRAN is the latest stable version.
 To obtain the most recent bug-fixes and experimental additions, please use the
-development versions, which is available on
+development version, which is available on
 GitHub.\footnote{\url{http://github.com/cysouw/qlcData}} This development
@@
 \hlcom{# install the qlcData package from GitHub}
-\hlstd{devtools}\hlopt{::}\hlkwd{install_github}\hlstd{(}\hlstr{"cysouw/qlcData"}\hlstd{)}
+\hlstd{devtools}\hlopt{::}\hlkwd{install_github}\hlstd{(}\hlstr{"cysouw/qlcData"}\hlstd{,} \hlkwc{build_vignettes} \hlstd{=} \hlnum{TRUE}\hlstd{)}
 \hlcom{# load the software, so it can be used}
@@
 
-\section{Error reporting}
+\section{Profiles and error reporting}
 \label{error-reporting}
 
-In the following example, it looks as if we have two identical strings,
-\texttt{AABB}. However, this is just an surface-impression delivered by the
-current font, which renders Latin and Cyrillic capitals identically. We can
-identify this problem when we produce an orthography profile from the strings.
-Using here the R implementation of orthography profiles, we first assign the two
-strings to a variable \texttt{test}, and then produce an orthography profile
-with the function \texttt{write.profile}. As it turns out, some of the letters 
-are Cyrillic.
+The first example of how to use these functions considers finding errors in the
+encoding of texts. In the following example, it looks as if we have two
+identical strings, \texttt{AABB}. However, this is just an surface-impression
+delivered by the current font, which renders Latin and Cyrillic capitals
+identically. We can identify this problem when we produce an orthography profile
+from the strings. Using here the R implementation of orthography profiles, we
+first assign the two strings to a variable \texttt{test}, and then produce an
+orthography profile with the function \texttt{write.profile}. As it turns out,
+some of the letters are Cyrillic.
 
@@
 
-\noindent The working of error-message reporting can also nicely be illustrated
+\noindent The function of error-message reporting can also nicely be illustrated
 with this example. Suppose we made an orthography profile with just the two
@@
 in the profile, so the tokenization produces an error, which can be used to fix
-the encoding. In the example below, we can see that the Cyrillic encoding is
-found in the second string of the \texttt{test} input.
+the encoding (or the profile). In the example below, we can see that the
+Cyrillic encoding is found in the second string of the \texttt{test} input.
 
@@
 various different ways in which strings can be separated into graphemes by
-\texttt{write.profile}. Consider the following example string. The default
-settings of \texttt{write.profile} separates the string into Unicode graphemes
-according to grapheme clusters (`user-perceived characters', see
-Section~\ref{the-unicode-approach}).
+\texttt{write.profile}. Consider the following string of characters called
+\texttt{example} below. The default settings of \texttt{write.profile} separates
+the string into Unicode graphemes according to grapheme clusters
+(`user-perceived characters', see Section~\ref{the-unicode-approach}). As it 
+turns out, ome of these graphemes are single codepoints, others are combinations
+of two codepoints (see Section~\ref{pitfall-characters-are-not-glyphs}).
 
@@
 % latex table generated in R 3.3.0 by xtable 1.8-2 package
-% Thu Jun 30 18:15:48 2016
+% Thu Aug 18 14:19:34 2016
 \begin{table}[H]
@@
 
-\noindent Some of these graphemes are single codepoints, others are combinations
-of two codepoints. By specifying the splitting separator as empty, it is
-possible to split the string into Unicode codepoints.
+\noindent By specifying the splitting separator as the empty string
+\texttt{sep~=~""}, it is possible to split the string into Unicode codepoints,
+thus separating the combining diacritics. The idea behind this option
+\texttt{sep} is that separating by a character allows for user-determined
+separation. The most extreme choice here is the empty string \texttt{sep~=~""},
+which is interpreted as separation everywhere. The other extreme is the default
+setting \texttt{sep~=~NULL}, which means that the separation is not
+user-defined, but relegated to the Unicode grapheme definitions.
 
@@
 \noindent Some characters look identical, although they are encoded differently.
-Unicode offers different ways of normalization, which can be invoked here as 
-well (see Section~\ref{pitfall-canonical-equivalence}).
+Unicode offers different ways of normalization (see
+Section~\ref{pitfall-canonical-equivalence}), which can be invoked here as well
+using the option \texttt{normalize}. NFC-normalization turns everything into the
+precomposed characters, while NFD-normalization separates everything into base
+characters with combining diacritics. Splitting by codepoints (i.e. \texttt{sep~=~""}) shows the results of these two normalizations.
 
@@
 example in either NFC or NFD, but using default separation in
-\texttt{write.profile}. To be precise, default separation means setting \texttt{sep = NULL} 
-but that has not be added explicitly.
+\texttt{write.profile}. To be precise, default separation means setting
+\texttt{sep~=~NULL}, but that has not be added explicitly.
 
@@
 % latex table generated in R 3.3.0 by xtable 1.8-2 package
-% Thu Jun 30 18:15:49 2016
+% Thu Aug 18 14:19:34 2016
 \begin{table}[H]
@@
 \noindent Note that these different profiles can also be produced using the bash
-executable \texttt{writeprofile} (see \ref{implementations} for notes on how to 
-install the bash executable). Exactly this example is also included in the help 
-file of the executable.
+executable \texttt{writeprofile} (see \ref{available-implementations} for notes
+on how to install the bash executable). Exactly this example is also included in
+the help file of the executable.
 
@@
 ## USAGE: 
-##   writeprofile [-hie] [-s SEP] [-n NORMALIZE] [<STRINGS>]
+##   writeprofile [-hiec] [-s SEP] [-n NORMALIZE] [<STRINGS>]
 ## 
@@
 ##   -e, --editing     Add empty columns for easy editing of profile
+##   -c, --codepoints  Separate by Unicode codepoints (overrides any -s setting)
 ##   -s SEP            Separator used to separate the strings
-##                     Often useful is option UNI to split by unicode codepoints
 ##                       defaults to unicode character definition [default: NULL]
@@
 ##     echo -e $example | writeprofile -i
-##     echo -e $example | writeprofile -i -s UNI
-##     echo -e $example | writeprofile -i -s UNI -n NFC
-##     echo -e $example | writeprofile -i -s UNI -n NFD
+##     echo -e $example | writeprofile -ic
+##     echo -e $example | writeprofile -ic -n NFC
+##     echo -e $example | writeprofile -ic -n NFD
 \end{verbatim}
@@
 % latex table generated in R 3.3.0 by xtable 1.8-2 package
-% Thu Jun 30 18:15:49 2016
+% Thu Aug 18 14:19:34 2016
 \begin{table}[htb]
@@
 
-For quick solutions, it is also possible to leave out the Unicode information in
+\noindent It is also possible to leave out the Unicode information in
 the profile skeleton by using the option \texttt{info = FALSE}. It is also
@@
 
-To document a specific case of graphemic parsing, it is highly useful to save
-all results of the tokenization to file by using the option \texttt{file.out},
-for example as follows:
-
-\begin{knitrout}\footnotesize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# save the results to various files}
-\hlkwd{tokenize}\hlstd{( example}
-         \hlstd{,} \hlkwc{profile} \hlstd{=} \hlstr{"~/Desktop/profile_skeleton.txt"}
-         \hlstd{,} \hlkwc{file.out} \hlstd{=} \hlstr{"~/Desktop/result"}
-        \hlstd{)}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
-
-This will lead to the following four files being written:
-
-\begin{itemize}
-
-   \item \textsc{result_strings.tsv}:\\ A tab-separated file with the original
-         and the tokenized/transliterated strings
-
-   \item \textsc{result_profile.tsv}:\\ A tab-separated file with the
-         graphemes with added frequencies of occurence in the data. The lines
-         in the file is re-ordered according to the order that resulted from the
-         ordering specifications.
-
-   \item \textsc{result_errors.tsv}:\\ A tab-separated file with all original
-         strings that contain unmatched parts. Unmatched parts are indicated
-         with the character as specified with the option \texttt{missing}. By
-         default the character \textsc{double question mark} < ⁇ > at
-         \uni{2047} is used. When there are no errors, then this file is 
-         absent.
-
-    \item \textsc{result_missing.tsv}:\\ A tab-separated file with the graphemes
-          that are missing from the original orthography profile, as indicated in
-          the errors. When there are no errors, then this file is absent.
-
-\end{itemize}
-
 \section{Rule ordering}
@@
 
-Not yet everything is correct in this graphemic parsing. The sequence <sh> in
-`mishap' should not be a digraph, and likewise the sequence <sch> in `mischief'
-should of course be separated into <s> and <ch>. One of the important issues to
-get the graphemic parsing right is the order in which graphemes are parsed. For
-example, currently the grapheme <sch> is parsed before the grapheme <ch>,
-leading to <m\ i\ sch\ ie\ f> instead of the intended <m\ i\ s\ ch\ ie\ f>. The
-reason that <sch> is parsed before <ch> is that by default longer graphemes are
-parsed before shorter ones. Our experience is that in most cases this is
-expected behaviour. You can change the ordering by specifying the option
-\texttt{ordering}. Setting this option to \texttt{NULL} results in no
-preferential ordering, i.e. the graphemes are parsed in the order of the
-profile, from top to bottom. Now `mischief' is parsed correctly, but `scheme' is
-wrong. So ordering is not the solution in this case.
+Not yet everything is correct in this graphemic parsing of the example discussed
+previously. The sequence <sh> in `mishap' should not be a digraph, and
+conversely the sequence <sch> in `mischief' should of course be separated into
+<s> and <ch>. One of the important issues to get the graphemic parsing right is
+the order in which graphemes are parsed. For example, currently the grapheme
+<sch> is parsed before the grapheme <ch>, leading to <m\ i\ sch\ ie\ f> instead
+of the intended <m\ i\ s\ ch\ ie\ f>. The reason that <sch> is parsed before
+<ch> is that by default longer graphemes are parsed before shorter ones. Our
+experience is that in most cases this is expected behaviour. You can change the
+ordering by specifying the option \texttt{ordering}. Setting this option to
+\texttt{NULL} results in no preferential ordering, i.e. the graphemes are parsed
+in the order of the profile, from top to bottom. Now `mischief' is parsed
+correctly, but `scheme' is wrong. So this ordering is not the solution in this
+case.
 
@@
 
-   \item \textsc{size}:\\
-         order the lines in the profile by the size of the
+   \item \textsc{size}\\
+         This option orders the lines in the profile by the size of the
          grapheme, largest first. Size is measured by number of Unicode
@@
 
-   \item \textsc{context}:\\
-         order the lines by whether they have any context
-         specified (see next section). Lines with context coming first. Note
-         that this only works when the option \texttt{regex = TRUE} is also chosen.
-
-  \item \textsc{reverse}:\\
-        order the lines from bottom to top.
-
-  \item \textsc{frequency}:\\
-         order the lines by the frequency with which they
+   \item \textsc{context}\\ This option orders the lines by whether they have
+           any context specified (see next section). Lines with context will
+           then be used first. Note that this only works when the option
+           \texttt{regex = TRUE} is also chosen (otherwise context
+           specifications are not used).
+
+   \item \textsc{reverse}\\ This option orders the lines from bottom to top.
+         Reversing order can be useful because hand-written profiles tend to put
+         general rules before specific rules, which mostly should be applied in
+         reverse order.
+
+  \item \textsc{frequency}\\
+         This option orders the lines by the frequency with which they
          match in the specified strings before tokenization, least frequent
@@
 (i.e. the have the same size and both/neither have context) the order will be
-reversed compared to the order as attested in the profile. Reversing order can
-be useful because hand-written profiles tend to put general rules before
-specific rules, which mostly should be applied in reverse order.
+reversed compared to the order as attested in the profile, because most
+hand-written specifications of graphemes will first write the general rule,
+followed by more specific regularities. To get the right tokenization, these 
+rules should in most cases be applied in reverse order.
 
@@
 
-To document the process of graphemic tokenization, it is strongly encouraged to
+To document a specific case of graphemic parsing, it is highly useful to save
+all results of the tokenization to file by using the option \texttt{file.out},
+for example as follows:
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# save the results to various files}
+\hlkwd{tokenize}\hlstd{( example}
+         \hlstd{,} \hlkwc{profile} \hlstd{=} \hlstr{"~/Desktop/profile_skeleton.txt"}
+         \hlstd{,} \hlkwc{file.out} \hlstd{=} \hlstr{"~/Desktop/result"}
+        \hlstd{)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+\noindent This will lead to the following four files being written. Crucially, a
+new profile is produced with the re-ordered orthography profile. To reproduce
+the tokenization, this re-ordered profile can be used with the option
+\texttt{ordering~=~NULL}.
+
+\begin{itemize}
+
+   \item \textsc{result\_strings.tsv}:\\ A tab-separated file with the original
+         and the tokenized/transliterated strings
+
+   \item \textsc{result\_profile.tsv}:\\ A tab-separated file with the
+         graphemes with added frequencies of occurence in the data. The lines
+         in the file is re-ordered according to the order that resulted from the
+         ordering specifications (see Section~\ref{rule-ordering}).
+
+   \item \textsc{result\_errors.tsv}:\\ A tab-separated file with all original
+         strings that contain unmatched parts. Unmatched parts are indicated
+         with the character as specified with the option \texttt{missing}. By
+         default the character \textsc{double question mark} <⁇> at
+         \uni{2047} is used. When there are no errors, then this file is 
+         absent.
+
+    \item \textsc{result\_missing.tsv}:\\ A tab-separated file with the graphemes
+          that are missing from the original orthography profile, as indicated in
+          the errors. When there are no errors, then this file is absent.
+
+\end{itemize}
 
@@
 specifications. An orthography profile can have columns called \texttt{Left} and
-\texttt{Right} to specify the context in which the Grapheme is to be
+\texttt{Right} to specify the context in which the grapheme is to be
 separated.\footnote{The columns names \textttf{Left}, \textttf{Right} and
-\textttf{Grapheme} are currently hard coded, so exactly these column names
+\textttf{Grapheme} are currently hard-coded, so exactly these column names
 should be used for these effects to take place. The position of the columns in
-the profile is unimportant. So the column \textttf{Left} can occur anywhere.} We
-are adding an extra line to the profile specifying that <s> is a grapheme when
-it occurs after <mi>. These contextually specified graphemes are based on
-regular expressions (so you can actually use regular expressions in the
-description of the context). For these to be included in the graphemic parsing 
-we have to specify the option \texttt{regex = TRUE}. This contextually specified 
-grapheme should actually be handled first, so we could try \texttt{ordering = 
-NULL}. However, we can also explicitly specifiy that rules with contextual 
-information should be applied first by using \texttt{ordering = "context"}. That 
-gives the right results for this toy example.
+the profile is unimportant. So the column \textttf{Left} can occur anywhere.}
+For example, we are adding an extra line to the profile from above, resulting in
+the profile shown in Table~\ref{tab:profile_skeleton2}. The extra line specifies
+that <s> is a grapheme when it occurs after <mi>. Such contextually-specified
+graphemes are based on regular expressions so you can also use regular
+expressions in the description of the context. For such contextually specified
+graphemes to be included in the graphemic parsing we have to specify the option
+\texttt{regex = TRUE}. This contextually specified grapheme should actually be
+handled first, so we could try \texttt{ordering = NULL}. However, we can also
+explicitly specifiy that rules with contextual information should be applied
+first by using \texttt{ordering = "context"}. That gives the right results for
+this toy example.
 
@@
 
+It is important to realize that with the option \texttt{regex = TRUE} all
+content in the profile is treated as regular expressions, so the characters with
+special meaning in regular expressions should be either omitted or escaped (by
+putting a <\ \backslash\ > \textsc{reverse solidus} at \uni{005C} before the
+character). Specifically, this concerns the following characters:
+
+\begin{itemize}
+
+  \item[] <-> \textsc{hyphen-minus} at \uni{002D}
+  \item[] <!> \textsc{exclamation mark} at \uni{0021}
+  \item[] <?> \textsc{question mark} at \uni{003F}
+  \item[] <.> \textsc{full stop} at \uni{002E}
+  \item[] <(> \textsc{left parenthesis} at \uni{0028}
+  \item[] <)> \textsc{right parenthesis} at \uni{0029}
+  \item[] <[> \textsc{left square bracket} at \uni{005B}
+  \item[] <]> \textsc{right square bracket} at \uni{005D}
+  \item[] <\{> \textsc{left curly bracket} at \uni{007B}
+  \item[] <\}> \textsc{right curly bracket} at \uni{007D}
+  \item[] <|> \textsc{vertical line} at \uni{007C}
+  \item[] <*> \textsc{asterisk} at \uni{002A}
+  \item[] <\backslash> \textsc{reverse solidus} at \uni{005C}
+  \item[] <ˆ> \textsc{circumflex accent} at \uni{005E}
+  \item[] <+> \textsc{plus sign} at \uni{002B}
+  \item[] <\$> \textsc{dollar sign} at \uni{0024}
+
+\end{itemize}
+
 \section{Profile skeleton with columns for editing}
-\label{profile-editing})
+\label{profile-editing}
 
 When it is expected that context might be important for a profile, then the
-profile skeleton can be written to a file with columns prepared for the 
-contextual specifications by using the option \texttt{editing = TRUE}.
-
-\include{chapters/use_cases}
-
-\chapter{Testing code inclusion}
-
-Trying to include code through knitr. 
-Different languages can be easily 
-included!
-
-There is a bug for non-R engines: the fontsize settings and line spaceing are not correct. I have 
-filed an issue, hopefully this will be fixed soon.
-
-For syntax colouring, install the package `highlight' and for nice tables 
-install the package `xtable'.
-
-\begin{knitrout}\footnotesize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-echo 'this is a simple bash example'
-ls | wc
-\end{alltt}
-
-\begin{verbatim}
-## this is a simple bash example
-##       12      12     203
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
-
-\begin{knitrout}\footnotesize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
-## note that bash chunks are executed as one, with all output at the end.
-## You will have to make different chunks to get output separated.
-## And add manual linebreaks!
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
-
-Spacing of lines is fine. I think without background looks better
-
-\begin{knitrout}\footnotesize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# adding comments inside code? Probably not a good idea}
-\hlstd{(example} \hlkwb{<-} \hlstr{"this is a simple R example"}\hlstd{)}
-\end{alltt}
-\begin{verbatim}
-## [1] "this is a simple R example"
-\end{verbatim}
-\begin{alltt}
-
-\hlstd{test} \hlkwb{<-} \hlnum{3} \hlopt{+} \hlnum{4}
-\hlstd{test}
-\end{alltt}
-\begin{verbatim}
-## [1] 7
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
-
-\begin{knitrout}\footnotesize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlkwd{library}\hlstd{(qlcData)}
-\hlstd{profile} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(}\hlstr{"AΑА"}\hlstd{)}
+profile skeleton can be created with columns prepared for the contextual
+specifications. This is done by using the option \texttt{editing = TRUE} (cf.
+Table~\ref{tab:profile_editing_1} for a toy profile of some Italian words).
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlstd{example} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{'cane'}\hlstd{,} \hlstr{'cena'}\hlstd{,} \hlstr{'cine'}\hlstd{)}
+\hlkwd{write.profile}\hlstd{(example}
+              \hlstd{,} \hlkwc{file} \hlstd{=} \hlstr{"~/Desktop/profile_skeleton.txt"}
+              \hlstd{,} \hlkwc{editing} \hlstd{=} \hlnum{TRUE}
+              \hlstd{,} \hlkwc{info} \hlstd{=} \hlnum{FALSE}
+              \hlstd{)}
 \end{alltt}
@@
 % latex table generated in R 3.3.0 by xtable 1.8-2 package
-% Thu Jun 30 18:15:49 2016
-\begin{table}[H]
+% Thu Aug 18 14:19:34 2016
+\begin{table}[htb]
 \centering
-\begingroup\small
-\begin{tabular}{llll}
+\begingroup\scriptsize
+\begin{tabular}{lllll}
   \toprule
-Grapheme & Frequency & Codepoint & UnicodeName \\ 
+Left & Grapheme & Right & Class & Replacement \\ 
   \midrule
-A & 1 & U+0041 & LATIN CAPITAL LETTER A \\ 
-  Α & 1 & U+0391 & GREEK CAPITAL LETTER ALPHA \\ 
-  А & 1 & U+0410 & CYRILLIC CAPITAL LETTER A \\ 
+ & a &  &  & a \\ 
+   & c &  &  & c \\ 
+   & e &  &  & e \\ 
+   & i &  &  & i \\ 
+   & n &  &  & n \\ 
    \bottomrule
@@
 \endgroup
-\caption{test} 
-\label{bla}
+\caption{Orthography profile with empty columns for editing contexts} 
+\label{tab:profile_editing_1}
 \end{table}
 
-You can refer to variables by using \textbackslash Sexpr{}. 
-For example, the length of the example string is 26.
-Crossreferencing also works, see Table~\ref{bla}.
-
-\begin{knitrout}\footnotesize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-print('bla' + 'bla')
-print(3+4)
-\end{alltt}
-
+Besides the columns \texttt{Left}, \texttt{Grapheme}, and \texttt{Right} as
+discussed in the previous sections, there are also columns \texttt{Class} and
+\texttt{Replacement}. The column \texttt{Class} can be used to specify classes
+of graphemes that can then be used in the contextual specification. The column
+\texttt{Replacement} is just a copy of the column \textt{Grapheme}, providing a
+skeleton to specify transliteration. The name of the columne
+\texttt{Replacement} is not fixed -- there can actually be multiple columns with 
+different kinds of transliterations in a single profile.
+
+To achieve contextually determined replacements it is possible to use a regular
+expression in the contextual column. For example, consider the edited toy
+profile for Italian in Table~\ref{tab:profile_editing_2} (where <c> becomes /k/
+except before <i,e>, then it becomes /tʃ/).
+
+% latex table generated in R 3.3.0 by xtable 1.8-2 package
+% Thu Aug 18 14:19:34 2016
+\begin{table}[htb]
+\centering
+\begingroup\scriptsize
+\begin{tabular}{lllll}
+  \toprule
+Left & Grapheme & Right & Class & IPA \\ 
+  \midrule
+ & c & [ie] &  & tʃ \\ 
+   & a &  &  & a \\ 
+   & n &  &  & n \\ 
+   & c &  &  & k \\ 
+   & e &  &  & e \\ 
+   & i &  &  & i \\ 
+   \bottomrule
+\end{tabular}
+\endgroup
+\caption{Orthography profile with regex as context} 
+\label{tab:profile_editing_2}
+\end{table}
+
+To use this profile, you have to add the option \texttt{regex = TRUE}. Also note
+that we have changed the name of the transliteration-column, so we have to tell
+the tokenization process to use this column to transliterate. This is done by
+adding the option \texttt{transliterate = "IPA"}.
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# add a contextual grapheme, and then use the edited profile to tokenize}
+\hlkwd{tokenize}\hlstd{( example}
+         \hlstd{,} \hlkwc{profile} \hlstd{=} \hlstr{"~/Desktop/profile_skeleton.txt"}
+         \hlstd{,} \hlkwc{regex} \hlstd{=} \hlnum{TRUE}
+         \hlstd{,} \hlkwc{transliterate} \hlstd{=} \hlstr{"IPA"}
+        \hlstd{)}\hlopt{$}\hlstd{strings}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
 \begin{verbatim}
-## blabla
-## 7
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
-
-You have to add the option ``engine.path='python3'' to get Python 3.
-
-\begin{knitrout}\footnotesize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-print('python3 needs an extra tag')
-uni = 'aɽɮz'
-print(uni)
-\end{alltt}
-
+##   originals tokenized transliterated
+## 1      cane   c a n e        k a n e
+## 2      cena   c e n a       tʃ e n a
+## 3      cine   c i n e       tʃ i n e
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+Another equivalent possibility is to use a column \texttt{Class} to specify a
+class of graphemes, and then use this class in the specification of context.
+This is useful to keep track of recurrent classes in larger profiles. You are
+free to use any class-name you like, as long as it doesn't clash with the rest
+of the profile. The example shown in Table~\ref{tab:profile_editing_3} should 
+give the same result as obtained previously by using a regular expression.
+
+% latex table generated in R 3.3.0 by xtable 1.8-2 package
+% Thu Aug 18 14:19:34 2016
+\begin{table}[htb]
+\centering
+\begingroup\scriptsize
+\begin{tabular}{lllll}
+  \toprule
+Left & Grapheme & Right & Class & IPA \\ 
+  \midrule
+ & c & Vfront &  & tʃ \\ 
+   & a &  &  & a \\ 
+   & n &  &  & n \\ 
+   & c &  &  & k \\ 
+   & e &  & Vfront & e \\ 
+   & i &  & Vfront & i \\ 
+   \bottomrule
+\end{tabular}
+\endgroup
+\caption{Orthography profile with Class as context} 
+\label{tab:profile_editing_3}
+\end{table}
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# add a class, and then use the edited profile to tokenize}
+\hlkwd{tokenize}\hlstd{( example}
+         \hlstd{,} \hlkwc{profile} \hlstd{=} \hlstr{"~/Desktop/profile_skeleton.txt"}
+         \hlstd{,} \hlkwc{regex} \hlstd{=} \hlnum{TRUE}
+         \hlstd{,} \hlkwc{transliterate} \hlstd{=} \hlstr{"IPA"}
+        \hlstd{)}\hlopt{$}\hlstd{strings}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
 \begin{verbatim}
-## python3 needs an extra tag
-## aɽɮz
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
-
-Testing: does referencing to variables also work in Python? It does give an error, so no... !
+##   originals tokenized transliterated
+## 1      cane   c a n e        k a n e
+## 2      cena   c e n a       tʃ e n a
+## 3      cine   c i n e       tʃ i n e
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+\section{Formatting grapheme separation}
+\label{formattingseparation}
+
+In all examples above we have used the default formatting for grapheme
+separation using space as a separator, which is obtained by the default setting
+\texttt{sep = " "}. It is possible to specify any other separator here,
+including the empty string, i.e. \texttt{sep = ""}. This will not show the
+graphemic tokenization anymore (although it has of course been used in the
+background).
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# Use the empty string as separator}
+\hlkwd{tokenize}\hlstd{( example}
+         \hlstd{,} \hlkwc{profile} \hlstd{=} \hlstr{"~/Desktop/profile_skeleton.txt"}
+         \hlstd{,} \hlkwc{regex} \hlstd{=} \hlnum{TRUE}
+         \hlstd{,} \hlkwc{transliterate} \hlstd{=} \hlstr{"IPA"}
+         \hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}
+        \hlstd{)}\hlopt{$}\hlstd{strings}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+##   originals tokenized transliterated
+## 1      cane      cane           kane
+## 2      cena      cena          tʃena
+## 3      cine      cine          tʃine
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+Normally, the separator specified should not occur in the data. If it does,
+unexpected things might happen, so consider removing the chosen seperator from
+your strings first. However, there is also an option \texttt{sep.replace} to
+replace the separator with something else. When \code{sep.replace} is specified,
+this mark is inserted in the string at those places where the separator occurs.
+Typical usage in linguistics would be \texttt{sep = " ", sep.replace = "\#"} adding
+spaces between graphemes and replacing spaces in the input string by hashes in
+the output string.
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# Replace separator in string to be tokenized}
+\hlkwd{tokenize}\hlstd{(} \hlstr{"test test test"}
+         \hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{" "}
+         \hlstd{,} \hlkwc{sep.replace} \hlstd{=} \hlstr{"#"}
+        \hlstd{)}\hlopt{$}\hlstd{strings}\hlopt{$}\hlstd{tokenized}
+\end{alltt}
+\begin{verbatim}
+## [1] "t e s t # t e s t # t e s t"
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+\section{Remaining issues}
+
+Given a set of graphemes, there are at least two different methods to tokenize
+strings. The first is called \texttt{method = "global"} here: this approach
+takes the first grapheme in the profile, then matches this grapheme globally at
+all places in the string, and then turns to the next string in the profile. To
+us, this seems to be the most intuitive way for linguists to deal with graphemic
+parsing. The other approach is called \texttt{method = "linear"} here: this
+approach walks through the string from left to right. At the first character it
+looks through all graphemes whether there is any match, and then walks further
+to the end of the match and starts again. This approach is more alike to
+finite-state rewrite rules (though note that it still works differently from
+such rewrite rules, see Section~\ref{rule-ordering}).
+
+In some special cases these two methods can lead to different results, but these
+special situations are very unlikely to happen in natural language. The example
+below shows that a string \texttt{'abc'} can be parsed differently in case of a
+very special profile with a very special ordering of the graphemes.
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# different parsing methods can lead to different results}
+\hlcom{# the global method first catches 'bc'}
+\hlkwd{tokenize}\hlstd{(} \hlstr{"abc"}
+         \hlstd{,} \hlkwc{profile} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{"bc"}\hlstd{,}\hlstr{"ab"}\hlstd{,}\hlstr{"a"}\hlstd{,}\hlstr{"c"}\hlstd{)}
+         \hlstd{,} \hlkwc{order} \hlstd{=} \hlkwa{NULL}
+         \hlstd{,} \hlkwc{method} \hlstd{=} \hlstr{"global"}
+         \hlstd{)}\hlopt{$}\hlstd{strings}
+\end{alltt}
+\begin{verbatim}
+##   originals tokenized
+## 1       abc      a bc
+\end{verbatim}
+\begin{alltt}
+\hlcom{# the linear method catches the first grapheme, which is 'ab'}
+\hlkwd{tokenize}\hlstd{(} \hlstr{"abc"}
+         \hlstd{,} \hlkwc{profile} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{"bc"}\hlstd{,}\hlstr{"ab"}\hlstd{,}\hlstr{"a"}\hlstd{,}\hlstr{"c"}\hlstd{)}
+         \hlstd{,} \hlkwc{order} \hlstd{=} \hlkwa{NULL}
+         \hlstd{,} \hlkwc{method} \hlstd{=} \hlstr{"linear"}
+         \hlstd{)}\hlopt{$}\hlstd{strings}
+\end{alltt}
+\begin{verbatim}
+##   originals tokenized
+## 1       abc      ab c
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+The current R-implementation has a limitation when regular expressions are used. 
+The problem is that overlapping matches are not captured. Everything works as 
+expected without regular expressions, but there might be warnings/errors in case of 
+\texttt{regex = TRUE}. However, just as in the previous issue, this should very 
+rarely happen in natural language.
+
+The problem can be exemplified by a sequence <bbbb> in which a 
+grapheme <bb> should be matched. With the default \texttt{regex = FALSE} 
+there are three possible matches, but with \texttt{regex = TRUE} only the first 
+two or the last two are matched. The middle two are not matched because they 
+overlap with the other matches. In the example below this leads to an error, 
+because the second <bb> is not matched. We have not been able to produce a real 
+example in any natural language in which this limitation might lead to an error.
+
+\begin{knitrout}\footnotesize
+\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# Everything perfekt without regular expressions}
+\hlkwd{tokenize}\hlstd{(} \hlstr{"abbb"}
+        \hlstd{,} \hlkwc{profile} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{"ab"}\hlstd{,}\hlstr{"bb"}\hlstd{)}
+        \hlstd{,} \hlkwc{order} \hlstd{=} \hlkwa{NULL}
+        \hlstd{,} \hlkwc{regex} \hlstd{=} \hlnum{FALSE}
+        \hlstd{)}\hlopt{$}\hlstd{strings}
+\end{alltt}
+\begin{verbatim}
+##   originals tokenized
+## 1      abbb     ab bb
+\end{verbatim}
+\begin{alltt}
+\hlcom{# Matching with regular expressions does not catch overlap}
+\hlkwd{tokenize}\hlstd{(} \hlstr{"abbb"}
+        \hlstd{,} \hlkwc{profile} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{"ab"}\hlstd{,}\hlstr{"bb"}\hlstd{)}
+        \hlstd{,} \hlkwc{order} \hlstd{=} \hlkwa{NULL}
+        \hlstd{,} \hlkwc{regex} \hlstd{=} \hlnum{TRUE}
+        \hlstd{)}\hlopt{$}\hlstd{strings}
+\end{alltt}
+
+{\ttfamily\noindent\color{warningcolor}{\#\# Warning in tokenize("{}abbb"{}, profile = c("{}ab"{}, "{}bb"{}), order = NULL, regex = TRUE): \\\#\# There were unknown characters found in the input data.\\\#\# Check output\$errors for a table with all problematic strings.}}\begin{verbatim}
+##   originals tokenized
+## 1      abbb    ab ⁇ ⁇
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+\include{chapters/use_cases}
+% knit_child("chapters/codetesting.tex")
 

### 2016-08-18 book/chapters/orthography_profiles.tex
minor changes to chapter

@@
 central concept in our proposal is the \textsc{orthography profile}, a simple
-tab-separated CSV text file, that characterizes and documents a writing system.
+tab-separated text file, that characterizes and documents a writing system.
 We also offer basic implementations in Python and R to assist with the
@@
 
-Given these goals, Unicode locale descriptions (see Section~\ref{terminology})
-might seem like the ideal orthography profiles. However, there are various
-practical obstacles preventing the use of such locale descriptions in the daily
-linguistic practice, namely: (i) the XML-structure is too verbose to easily and
-quickly produce or correct manually, (ii) locale descriptions are designed for a
-wide scope on information (like date formats or names of weekdays) most of which
-is not applicable for documenting writing systems, and (iii) most crucially,
-even if someone made the effort to produce a technically correct locale
-description for a specific source at hand, then it is nigh impossible to deploy
-the description. This is because a locale description has to be submitted to and
-accepted by the Unicode Common Locale Data Repository. The repository is
-(rightly so) not interested in descriptions that only apply to a limited set of
-sources (e.g.~descriptions for only a single dictionary).
+Given these goals, Unicode locale descriptions (see
+Section~\ref{the-unicode-approach}) might seem like the ideal orthography
+profiles. However, there are various practical obstacles preventing the use of
+such locale descriptions in the daily linguistic practice, namely: (i) the
+XML-structure is too verbose to easily and quickly produce or correct manually,
+(ii) locale descriptions are designed for a wide scope on information (like date
+formats or names of weekdays) most of which is not applicable for documenting
+writing systems, and (iii) most crucially, even if someone made the effort to
+produce a technically correct locale description for a specific source at hand,
+then it is nigh impossible to deploy the description. This is because a locale
+description has to be submitted to and accepted by the Unicode Common Locale
+Data Repository. The repository is (rightly so) not interested in descriptions
+that only apply to a limited set of sources (e.g.~descriptions for only a single
+dictionary).
 
@@
 
-Finally, to arrive at the graphemic tokenization in
-(3), tailored grapheme clusters are needed (as for example specified in an
-orthography profile). For example, this orthography profile would specify that
-the sequence of characters, and form a single grapheme, and that and form a
-grapheme. The orthography profile could also specify orthographic rules,
-e.g.~when tokenization graphemes, in say English, the in the forms and should be
-treated as distinct sequences depending on their contexts.
+Finally, to arrive at the graphemic tokenization in (3), tailored grapheme
+clusters are needed (possibly specified in an orthography profile). For example,
+an orthography profile might specify that the sequence of characters <tsʰ> form
+a single grapheme. The orthography profile could also specify orthographic
+rules, e.g.~when tokenizing graphemes, in say English, the sequences <sh> in the
+forms <mishap> and <mishmash> should be treated as distinct sequences depending
+on their contexts.
 
@@
 and rules that are needed to adequately model a writing system. An important
-assumption of our work is that we assume a resource is encoded in Unicode (or
+assumption is that we assume a resource is encoded in Unicode (or
 has been converted to Unicode). Any data source that the Unicode Standard is
@@
 
-The logic of orthographic parsing ...
+The crucial difference between our current proposal and traditional
+computational approaches to transliteration is the strict separation between
+tokenization and transliteration. Most computational approaches to
+transliteration are based on finite-state transducers (including the
+transliteration as described in the Unicode Locale Data Markup
+Language.\footnote{See \url{http://www.unicode.org/reports/tr35/}} Finite-state 
+transducers attempt to describe the mapping from input to output string directly 
+as a set of rewrite rules. Although such systems are computationally well 
+understood, we feel that they are not well-suited for day-to-day linguistic 
+practice. First, by forcing a first step of grapheme tokenization, our system 
+tries to keep close to the logic of the writing system. Second, by separating 
+tokenization from transliteration there is no problem with `feeding' and 
+`bleeding' of rules, common with transducers (cf.~Section~\ref{rule-ordering}
+).
 
-Note that to deal with ambiguous parsing cases, we can use the Unicode approach
-using the zero width joiner. This is actually a non-joiner (the name is
-confusing): the idea is to add this character into the text to identify cases in
-which a sequence of characters is not supposed to be a complex grapheme (even
-though the sequence is in the orthography profile)
+Note that to deal with ambiguous parsing cases, it is still possible to use the
+Unicode approach including the \textsc{zero-width non-joiner} character at
+\uni{200C} into the text. The idea is to add this character into the text to
+identify cases in which a sequence of characters is \textit{not} supposed to be
+a complex grapheme cluster (even though the sequence is in the orthography
+profile).
 

### 2016-08-18 book/chapters/implementation.tex
adding new chapter with R examples

@@
 # install the qlcData package from GitHub
-devtools::install_github('cysouw/qlcData')
+devtools::install_github('cysouw/qlcData', build_vignettes = TRUE)
 # load the software, so it can be used 
@@
 
-To document a specific case of graphemic parsing, it is highly useful to save
-all results of the tokenization to file by using the option \texttt{file.out},
-for example as follows:
-
-<<eval=FALSE, tidy=FALSE>>= 
-# save the results to various files
-tokenize( example
-         , profile = "~/Desktop/profile_skeleton.txt"
-         , file.out = "~/Desktop/result"
-        )
-@
-
-\noindent This will lead to the following four files being written:
-
-\begin{itemize}
-
-   \item \textsc{result\_strings.tsv}:\\ A tab-separated file with the original
-         and the tokenized/transliterated strings
-
-   \item \textsc{result\_profile.tsv}:\\ A tab-separated file with the
-         graphemes with added frequencies of occurence in the data. The lines
-         in the file is re-ordered according to the order that resulted from the
-         ordering specifications.
-
-   \item \textsc{result\_errors.tsv}:\\ A tab-separated file with all original
-         strings that contain unmatched parts. Unmatched parts are indicated
-         with the character as specified with the option \texttt{missing}. By
-         default the character \textsc{double question mark} <⁇> at
-         \uni{2047} is used. When there are no errors, then this file is 
-         absent.
-
-    \item \textsc{result\_missing.tsv}:\\ A tab-separated file with the graphemes
-          that are missing from the original orthography profile, as indicated in
-          the errors. When there are no errors, then this file is absent.
-
-\end{itemize}
-
 \section{Rule ordering}
@@
    \item \textsc{size}\\
-         order the lines in the profile by the size of the
+         This option orders the lines in the profile by the size of the
          grapheme, largest first. Size is measured by number of Unicode
@@
 
-   \item \textsc{context}\\
-         order the lines by whether they have any context
-         specified (see next section). Lines with context coming first. Note
-         that this only works when the option \texttt{regex = TRUE} is also chosen.
+   \item \textsc{context}\\ This option orders the lines by whether they have
+           any context specified (see next section). Lines with context will
+           then be used first. Note that this only works when the option
+           \texttt{regex = TRUE} is also chosen (otherwise context
+           specifications are not used).
 
-  \item \textsc{reverse}\\
-        order the lines from bottom to top.
+   \item \textsc{reverse}\\ This option orders the lines from bottom to top.
+         Reversing order can be useful because hand-written profiles tend to put
+         general rules before specific rules, which mostly should be applied in
+         reverse order.
 
   \item \textsc{frequency}\\
-         order the lines by the frequency with which they
+         This option orders the lines by the frequency with which they
          match in the specified strings before tokenization, least frequent
@@
 
-To document the process of graphemic tokenization, it is strongly encouraged to
+To document a specific case of graphemic parsing, it is highly useful to save
+all results of the tokenization to file by using the option \texttt{file.out},
+for example as follows:
+
+<<eval=FALSE, tidy=FALSE>>= 
+# save the results to various files
+tokenize( example
+         , profile = "~/Desktop/profile_skeleton.txt"
+         , file.out = "~/Desktop/result"
+        )
+@
+
+\noindent This will lead to the following four files being written. Crucially, a
+new profile is produced with the re-ordered orthography profile. To reproduce
+the tokenization, this re-ordered profile can be used with the option
+\texttt{ordering~=~NULL}.
+
+\begin{itemize}
+
+   \item \textsc{result\_strings.tsv}:\\ A tab-separated file with the original
+         and the tokenized/transliterated strings
+
+   \item \textsc{result\_profile.tsv}:\\ A tab-separated file with the
+         graphemes with added frequencies of occurence in the data. The lines
+         in the file is re-ordered according to the order that resulted from the
+         ordering specifications (see Section~\ref{rule-ordering}).
+
+   \item \textsc{result\_errors.tsv}:\\ A tab-separated file with all original
+         strings that contain unmatched parts. Unmatched parts are indicated
+         with the character as specified with the option \texttt{missing}. By
+         default the character \textsc{double question mark} <⁇> at
+         \uni{2047} is used. When there are no errors, then this file is 
+         absent.
+
+    \item \textsc{result\_missing.tsv}:\\ A tab-separated file with the graphemes
+          that are missing from the original orthography profile, as indicated in
+          the errors. When there are no errors, then this file is absent.
+
+\end{itemize}
 
@@
 
+It is important to realize that with the option \texttt{regex = TRUE} all
+content in the profile is treated as regular expressions, so the characters with
+special meaning in regular expressions should be either omitted or escaped (by
+putting a <\ \backslash\ > \textsc{reverse solidus} at \uni{005C} before the
+character). Specifically, this concerns the following characters:
+
+\begin{itemize}
+
+  \item[] <-> \textsc{hyphen-minus} at \uni{002D}
+  \item[] <!> \textsc{exclamation mark} at \uni{0021}
+  \item[] <?> \textsc{question mark} at \uni{003F}
+  \item[] <.> \textsc{full stop} at \uni{002E}
+  \item[] <(> \textsc{left parenthesis} at \uni{0028}
+  \item[] <)> \textsc{right parenthesis} at \uni{0029}
+  \item[] <[> \textsc{left square bracket} at \uni{005B}
+  \item[] <]> \textsc{right square bracket} at \uni{005D}
+  \item[] <\{> \textsc{left curly bracket} at \uni{007B}
+  \item[] <\}> \textsc{right curly bracket} at \uni{007D}
+  \item[] <|> \textsc{vertical line} at \uni{007C}
+  \item[] <*> \textsc{asterisk} at \uni{002A}
+  \item[] <\backslash> \textsc{reverse solidus} at \uni{005C}
+  \item[] <ˆ> \textsc{circumflex accent} at \uni{005E}
+  \item[] <+> \textsc{plus sign} at \uni{002B}
+  \item[] <\$> \textsc{dollar sign} at \uni{0024}
+
+\end{itemize}
+
 \section{Profile skeleton with columns for editing}
-\label{profile-editing})
+\label{profile-editing}
 
 When it is expected that context might be important for a profile, then the
-profile skeleton can be written to a file with columns prepared for the 
-contextual specifications by using the option \texttt{editing = TRUE}.
+profile skeleton can be created with columns prepared for the contextual
+specifications. This is done by using the option \texttt{editing = TRUE} (cf.
+Table~\ref{tab:profile_editing_1} for a toy profile of some Italian words).
+
+<<eval=FALSE, tidy=FALSE>>=
+example <- c('cane', 'cena', 'cine')
+write.profile(example
+              , file = "~/Desktop/profile_skeleton.txt"
+              , editing = TRUE
+              , info = FALSE
+              )
+@
+
+<<results='asis', echo=FALSE>>=
+example <- c('cane', 'cena', 'cine')
+profile_editing <- write.profile(example, editing=T, info=F)
+print(xtable(profile_editing
+        , caption = 'Orthography profile with empty columns for editing contexts'
+        , label = 'tab:profile_editing_1'
+        )
+  , include.rownames = FALSE
+  , booktabs = TRUE
+  , size = 'scriptsize'
+  , table.placement = 'htb'
+  )
+@
+
+Besides the columns \texttt{Left}, \texttt{Grapheme}, and \texttt{Right} as
+discussed in the previous sections, there are also columns \texttt{Class} and
+\texttt{Replacement}. The column \texttt{Class} can be used to specify classes
+of graphemes that can then be used in the contextual specification. The column
+\texttt{Replacement} is just a copy of the column \textt{Grapheme}, providing a
+skeleton to specify transliteration. The name of the columne
+\texttt{Replacement} is not fixed -- there can actually be multiple columns with 
+different kinds of transliterations in a single profile.
+
+To achieve contextually determined replacements it is possible to use a regular
+expression in the contextual column. For example, consider the edited toy
+profile for Italian in Table~\ref{tab:profile_editing_2} (where <c> becomes /k/
+except before <i,e>, then it becomes /tʃ/).
+
+<<results='asis', echo=FALSE>>=
+colnames(profile_editing)[5] <- 'IPA'
+profile_editing[2,5] <- 'k'
+profile_editing <- rbind(c('', 'c', '[ie]', '', 'tʃ'), profile_editing)
+profile_editing <- profile_editing[c(1,2,6,3,4,5),]
+print(xtable(profile_editing
+        , caption = 'Orthography profile with regex as context'
+        , label = 'tab:profile_editing_2'
+        )
+  , include.rownames = FALSE
+  , booktabs = TRUE
+  , size = 'scriptsize'
+  , table.placement = 'htb'
+  )
+@
+
+To use this profile, you have to add the option \texttt{regex = TRUE}. Also note
+that we have changed the name of the transliteration-column, so we have to tell
+the tokenization process to use this column to transliterate. This is done by
+adding the option \texttt{transliterate = "IPA"}.
+
+<<eval=FALSE, tidy=FALSE>>=
+# add a contextual grapheme, and then use the edited profile to tokenize
+tokenize( example
+         , profile = "~/Desktop/profile_skeleton.txt"
+         , regex = TRUE
+         , transliterate = "IPA"
+        )$strings
+@
+
+<<echo=FALSE>>=
+tokenize(example
+  , profile = profile_editing
+  , regex = TRUE
+  , transliterate = 'IPA'
+  )$strings
+@
+
+Another equivalent possibility is to use a column \texttt{Class} to specify a
+class of graphemes, and then use this class in the specification of context.
+This is useful to keep track of recurrent classes in larger profiles. You are
+free to use any class-name you like, as long as it doesn't clash with the rest
+of the profile. The example shown in Table~\ref{tab:profile_editing_3} should 
+give the same result as obtained previously by using a regular expression.
+
+<<results='asis', echo=FALSE>>=
+profile_editing[1,3] <- 'Vfront'
+profile_editing[5,4] <- 'Vfront'
+profile_editing[6,4] <- 'Vfront'
+print(xtable(profile_editing
+        , caption = 'Orthography profile with Class as context'
+        , label = 'tab:profile_editing_3'
+        )
+  , include.rownames = FALSE
+  , booktabs = TRUE
+  , size = 'scriptsize'
+  , table.placement = 'htb'
+  )
+@
+
+<<eval=FALSE, tidy=FALSE>>=
+# add a class, and then use the edited profile to tokenize
+tokenize( example
+         , profile = "~/Desktop/profile_skeleton.txt"
+         , regex = TRUE
+         , transliterate = "IPA"
+        )$strings
+@
+
+<<echo=FALSE>>=
+tokenize(example
+  , profile = profile_editing
+  , regex = TRUE
+  , transliterate = 'IPA'
+  )$strings
+@
+
+\section{Formatting grapheme separation}
+\label{formattingseparation}
+
+In all examples above we have used the default formatting for grapheme
+separation using space as a separator, which is obtained by the default setting
+\texttt{sep = " "}. It is possible to specify any other separator here,
+including the empty string, i.e. \texttt{sep = ""}. This will not show the
+graphemic tokenization anymore (although it has of course been used in the
+background).
+
+<<eval=FALSE, tidy=FALSE>>=
+# Use the empty string as separator
+tokenize( example
+         , profile = "~/Desktop/profile_skeleton.txt"
+         , regex = TRUE
+         , transliterate = "IPA"
+         , sep = ""
+        )$strings
+@
+
+<<echo=FALSE>>=
+tokenize(example
+  , profile = profile_editing
+  , regex = TRUE
+  , transliterate = 'IPA'
+  , sep = ''
+  )$strings
+@
+
+Normally, the separator specified should not occur in the data. If it does,
+unexpected things might happen, so consider removing the chosen seperator from
+your strings first. However, there is also an option \texttt{sep.replace} to
+replace the separator with something else. When \code{sep.replace} is specified,
+this mark is inserted in the string at those places where the separator occurs.
+Typical usage in linguistics would be \texttt{sep = " ", sep.replace = "\#"} adding
+spaces between graphemes and replacing spaces in the input string by hashes in
+the output string.
+
+<<tidy=FALSE>>=
+# Replace separator in string to be tokenized
+tokenize( "test test test"
+         , sep = " "
+         , sep.replace = "#"
+        )$strings$tokenized
+@
+
+\section{Remaining issues}
+
+Given a set of graphemes, there are at least two different methods to tokenize
+strings. The first is called \texttt{method = "global"} here: this approach
+takes the first grapheme in the profile, then matches this grapheme globally at
+all places in the string, and then turns to the next string in the profile. To
+us, this seems to be the most intuitive way for linguists to deal with graphemic
+parsing. The other approach is called \texttt{method = "linear"} here: this
+approach walks through the string from left to right. At the first character it
+looks through all graphemes whether there is any match, and then walks further
+to the end of the match and starts again. This approach is more alike to
+finite-state rewrite rules (though note that it still works differently from
+such rewrite rules, see Section~\ref{rule-ordering}).
+
+In some special cases these two methods can lead to different results, but these
+special situations are very unlikely to happen in natural language. The example
+below shows that a string \texttt{'abc'} can be parsed differently in case of a
+very special profile with a very special ordering of the graphemes.
+
+<<tidy=FALSE>>=
+# different parsing methods can lead to different results
+# the global method first catches 'bc'
+tokenize( "abc"
+         , profile = c("bc","ab","a","c")
+         , order = NULL
+         , method = "global"
+         )$strings
+# the linear method catches the first grapheme, which is 'ab'
+tokenize( "abc"
+         , profile = c("bc","ab","a","c")
+         , order = NULL
+         , method = "linear"
+         )$strings
+@
+
+The current R-implementation has a limitation when regular expressions are used. 
+The problem is that overlapping matches are not captured. Everything works as 
+expected without regular expressions, but there might be warnings/errors in case of 
+\texttt{regex = TRUE}. However, just as in the previous issue, this should very 
+rarely happen in natural language.
+
+The problem can be exemplified by a sequence <bbbb> in which a 
+grapheme <bb> should be matched. With the default \texttt{regex = FALSE} 
+there are three possible matches, but with \texttt{regex = TRUE} only the first 
+two or the last two are matched. The middle two are not matched because they 
+overlap with the other matches. In the example below this leads to an error, 
+because the second <bb> is not matched. We have not been able to produce a real 
+example in any natural language in which this limitation might lead to an error.
+
+<<tidy=FALSE>>=
+# Everything perfekt without regular expressions
+tokenize( "abbb"
+        , profile = c("ab","bb")
+        , order = NULL
+        , regex = FALSE
+        )$strings
+# Matching with regular expressions does not catch overlap
+tokenize( "abbb"
+        , profile = c("ab","bb")
+        , order = NULL
+        , regex = TRUE
+        )$strings
+@

### 2016-08-18 book/chapters/implementation.tex
typos

@@
 
-\noindent The version available through CRAN is the latest stable version.
+The version available through CRAN is the latest stable version.
 To obtain the most recent bug-fixes and experimental additions, please use the
@@
 
-\noindent Inside the \texttt{qlcData} package, there are two functions for
+Inside the \texttt{qlcData} package, there are two functions for
 orthography processing, \texttt{write.profile} and \texttt{tokenize}. R includes
@@
 
-\noindent Basically, the idea is to use \texttt{write.profile} to produce a
+Basically, the idea is to use \texttt{write.profile} to produce a
 basic orthography profile from some data and then \texttt{tokenize} to apply the
@@
 
-\noindent For example, a good way to use the executables in a terminal is to
+For example, a good way to use the executables in a terminal is to
 make softlinks (using \texttt{ln}) from the executables to a directory in your
@@
 
-\noindent After this softlink it should be possible to access the \texttt{tokenize}
-function from the shell. Try \texttt{tokenize --help} to test the functionality.
+After inserting this softlink it should be possible to access the
+\texttt{tokenize} function from the shell. Try \texttt{tokenize --help} to test
+the functionality.
 
-<<size='scriptsize', engine='bash'>>=
+<<size='scriptsize', engine='bash', tidy=FALSE>>=
 tokenize --help
@@
 
-The first example of how to use these functions considers finding errors in the
+The first example of how to use these functions concerns finding errors in the
 encoding of texts. In the following example, it looks as if we have two
@@
 
-\noindent The function of error-message reporting can also nicely be illustrated
+The function of error-message reporting can also nicely be illustrated
 with this example. Suppose we made an orthography profile with just the two
@@
 (`user-perceived characters', see Section~\ref{the-unicode-approach}). As it 
-turns out, ome of these graphemes are single codepoints, others are combinations
+turns out, some of these graphemes are single codepoints, others are combinations
 of two codepoints (see Section~\ref{pitfall-characters-are-not-glyphs}).
@@
 
-\noindent By specifying the splitting separator as the empty string
+By specifying the splitting separator as the empty string
 \texttt{sep~=~""}, it is possible to split the string into Unicode codepoints,
@@
 
-\noindent Some characters look identical, although they are encoded differently.
+Some characters look identical, although they are encoded differently.
 Unicode offers different ways of normalization (see
@@
 
-\noindent It is important to realize that for Unicode grapheme definitions, NFC
+It is important to realize that for Unicode grapheme definitions, NFC
 and NFD normalization are equivalent. This can be shown by normalizing the
@@
 
-\noindent Note that these different profiles can also be produced using the bash
+Note that these different profiles can also be produced using the bash
 executable \texttt{writeprofile} (see \ref{available-implementations} for notes
@@
 <<size='scriptsize', engine='bash'>>=
-## show help
 writeprofile --help
@@
 
-\noindent It is also possible to leave out the Unicode information in
+It is also possible to leave out the Unicode information in
 the profile skeleton by using the option \texttt{info = FALSE}. It is also
 possible not to use a separate file at all, but process everything within R. In
-simply situations this is often useful (see below), but in general we prefer to
+simple situations this is often useful (see below), but in general we prefer to
 handle everything through a separately saved orthography profile. This profile
@@
 
-\noindent This will lead to the following four files being written. Crucially, a
+This will lead to the following four files being written. Crucially, a
 new profile is produced with the re-ordered orthography profile. To reproduce
@@
 \texttt{Replacement} is just a copy of the column \textt{Grapheme}, providing a
-skeleton to specify transliteration. The name of the columne
+skeleton to specify transliteration. The name of the column
 \texttt{Replacement} is not fixed -- there can actually be multiple columns with 
@@
 separation using space as a separator, which is obtained by the default setting
-\texttt{sep = " "}. It is possible to specify any other separator here,
+\texttt{sep~=~"~"}. It is possible to specify any other separator here,
 including the empty string, i.e. \texttt{sep = ""}. This will not show the
@@
 Given a set of graphemes, there are at least two different methods to tokenize
-strings. The first is called \texttt{method = "global"} here: this approach
+strings. The first is called \texttt{method = "global"}. This approach
 takes the first grapheme in the profile, then matches this grapheme globally at
@@
 us, this seems to be the most intuitive way for linguists to deal with graphemic
-parsing. The other approach is called \texttt{method = "linear"} here: this
+parsing. The other approach is called \texttt{method = "linear"}. This
 approach walks through the string from left to right. At the first character it
@@
 finite-state rewrite rules (though note that it still works differently from
-such rewrite rules, see Section~\ref{rule-ordering}).
+such rewrite rules, see Section~\ref{rule-ordering}). The global method is used 
+by default.
 
-In some special cases these two methods can lead to different results, but these
-special situations are very unlikely to happen in natural language. The example
-below shows that a string \texttt{'abc'} can be parsed differently in case of a
-very special profile with a very special ordering of the graphemes.
+In some special cases these two tokenization methods can lead to different
+results, but these special situations are very unlikely to happen in natural
+language. The example below shows that a string \texttt{'abc'} can be parsed
+differently in case of a very special profile with a very special ordering of
+the graphemes.
 
@@
          )$strings
+
 # the linear method catches the first grapheme, which is 'ab'
@@
 
-The current R-implementation has a limitation when regular expressions are used. 
-The problem is that overlapping matches are not captured. Everything works as 
-expected without regular expressions, but there might be warnings/errors in case of 
-\texttt{regex = TRUE}. However, just as in the previous issue, this should very 
-rarely happen in natural language.
+Further, the current R-implementation has a limitation when regular expressions
+are used. The problem is that overlapping matches are not captured when using
+regular expressions.\footnote{This restriction is an effect of the underlyingly
+used ICU implementation of the Unicode standard as implemented in R through the
+package \texttt{stringi}.} Everything works as expected without regular
+expressions, but there might be warnings/errors in case of \texttt{regex =
+TRUE}. However, just as in the previous issue, this problem should only very
+rarely (when at all) happen in natural language data.
 
@@
         )$strings
+
 # Matching with regular expressions does not catch overlap

### 2016-08-18 book/unicode.tex
typos

@@
 
-\noindent After this softlink it should be possible to access the \texttt{tokenize}
-function from the shell. Try \texttt{tokenize --help} to test the functionality.
+After inserting this softlink it should be possible to access the
+\texttt{tokenize} function from the shell. Try \texttt{tokenize --help} to test
+the functionality.
 
@@
 % latex table generated in R 3.3.0 by xtable 1.8-2 package
-% Thu Aug 18 14:19:34 2016
+% Thu Aug 18 15:23:10 2016
 \begin{table}[H]
@@
 \begin{alltt}
-## show help
 writeprofile --help
@@
 % latex table generated in R 3.3.0 by xtable 1.8-2 package
-% Thu Aug 18 14:19:34 2016
+% Thu Aug 18 15:23:11 2016
 \begin{table}[htb]
@@
 finite-state rewrite rules (though note that it still works differently from
-such rewrite rules, see Section~\ref{rule-ordering}).
-
-In some special cases these two methods can lead to different results, but these
-special situations are very unlikely to happen in natural language. The example
-below shows that a string \texttt{'abc'} can be parsed differently in case of a
-very special profile with a very special ordering of the graphemes.
+such rewrite rules, see Section~\ref{rule-ordering}). The global method is used 
+by default.
+
+In some special cases these two tokenization methods can lead to different
+results, but these special situations are very unlikely to happen in natural
+language. The example below shows that a string \texttt{'abc'} can be parsed
+differently in case of a very special profile with a very special ordering of
+the graphemes.
 
@@
 \begin{alltt}
+
 \hlcom{# the linear method catches the first grapheme, which is 'ab'}
@@
 \begin{alltt}
+
 \hlcom{# Matching with regular expressions does not catch overlap}

### 2016-08-18 book/chapters/introduction.tex
fixes typo
typos

@@
 
-\subsubsection*{Structure of this bok}
+\subsubsection*{Structure of this book}
 
@@
 
-In the nineteenth century, electric telegraphy lead to a new approach
-in which written language characters were encoded by signals sent through a
-copper wire. Originally, \textsc{bisignal codes} were used, consisting of two
-different signals. For example, Carl Friedrich Gauss in 1833 used positive and
-negative current \citep[282]{Mania2008}. More famous and influential, Samuel
-Morse in 1836 used long and short pulses. In those bisignal codes each character
-from the written language was encoded with a different number of signals
-(between one and five), so two different separators are needed: one between
-signals and one between characters. For example, in Morse-code there is a short
-pause between signals and a long pause between characters.\footnote{Actually,
-Morse-code also includes an extra long pause between words. Interestingly, it
-took a long time to consider the written word boundary---using white-space---as
-a bona-fide character that should simply be encoded with its own code point.
-This happened only with the revision of the Baudot-code (see below) by Donald
-Murray in 1901, in which he introduced a specific white-space code. This
-principle has been followed ever since.}
+In the nineteenth century, electric telegraphy lead to a next innovation in
+which written language characters were encoded by signals sent through a copper
+wire. Originally, \textsc{bisignal codes} were used, consisting of two different
+signals. For example, Carl Friedrich Gauss in 1833 used positive and negative
+current \citep[282]{Mania2008}. More famous and influential, Samuel Morse in
+1836 used long and short pulses. In those bisignal codes each character from the
+written language was encoded with a different number of signals (between one and
+five), so two different separators are needed: one between signals and one
+between characters. For example, in Morse-code there is a short pause between
+signals and a long pause between characters.\footnote{Actually, Morse-code also
+includes an extra long pause between words. Interestingly, it took a long time
+to consider the written word boundary---using white-space---as a bona-fide
+character that should simply be encoded with its own code point. This happened
+only with the revision of the Baudot-code (see below) by Donald Murray in 1901,
+in which he introduced a specific white-space code. This principle has been
+followed ever since.}
 

### 2017-01-20 book/chapters/implementation.tex
cleanup

@@
 # make softlinks to the R executables in /usr/local/bin
-ln -is $pathT $pathW /usr/local/bin
+# you will have to enter your user's password!
+sudo ln -is $pathT $pathW /usr/local/bin
 @
@@
 <<>>=
-test <- c('AABB', '\u0041\u0410\u0042\u0412')
+( test <- c('AABB', '\u0041\u0410\u0042\u0412') )
 write.profile(test)
@@
 used ICU implementation of the Unicode standard as implemented in R through the
-package \texttt{stringi}.} Everything works as expected without regular
+package \textttf{stringi}.} Everything works as expected without regular
 expressions, but there might be warnings/errors in case of \texttt{regex =
@@
 
-The problem can be exemplified by a sequence <bbbb> in which a 
-grapheme <bb> should be matched. With the default \texttt{regex = FALSE} 
-there are three possible matches, but with \texttt{regex = TRUE} only the first 
-two or the last two are matched. The middle two are not matched because they 
+The problem can be exemplified by a sequence <bbbb> in which a grapheme <bb>
+should be matched. With the default \texttt{regex = FALSE} there are three
+possible matches, but with \texttt{regex = TRUE} only the first two <b>s or the
+last two <b>s are matched. The middle two <b>s are not matched because they
 overlap with the other matches. In the example below this leads to an error,
-because the second <bb> is not matched. We have not been able to produce a real 
-example in any natural language in which this limitation might lead to an error.
+because the second <bb> is not matched. However, we have not been able to
+produce a real example in any natural language in which this limitation might
+lead to an error.
 

### 2017-01-20 book/chapters/introduction.tex
cleanup

@@
 model the historical and areal processes that underlie linguistic
-diversity.\footnote{
-cf.~\citet{Steiner_etal2011,List2012,List2012a,ListMoran2013,MoranProkic2013}}
-In this work, it is crucial to be able to flexibly process across numerous
+diversity. In this work, it is crucial to be able to flexibly process across numerous
 resources with different orthographies. In many cases even different resources
@@
 report errors to be corrected.
+
+% \footnote{cf.~\citet{Steiner_etal2011,List2012,List2012a,ListMoran2013,MoranProkic2013}}
 
@@
 
-Before we deal with these proposals we will extensively discuss the theoretical
+Before we deal with these proposals we will first discuss the theoretical
 background on text encoding, on the Unicode standard, and on the International
 Phonetic Alphabet. In the remainder of this chapter, we give an extended
-introduction to the notion of encoding (Section~\ref{encoding}) and writing
-systems, both from a linguistic perspective
+introduction to the notion of encoding (Section~\ref{encoding}) and the
+principles of writing systems, both from a linguistic perspective
 (Section~\ref{linguistic-terminology}) and from the perspective of the Unicode
@@
 slightly more in-depth survey of the many techniques that are available in the
-standard. A good appreciation for the solutions that the Unicode Standard also
-allows for a thorough understanding of possible pitfalls that one might
-encounter when using it (Chapter~\ref{unicode-pitfalls}). As an example of the
-current state-of-the-art, we discuss the rather problematic marriage of the
-International Phonetic Alphabet (IPA) with the Unicode Standard (Chapter
+standard. A good appreciation for the solutions that the Unicode Standard
+provides, also allows for a thorough understanding of possible pitfalls that one
+might encounter when using it (Chapter~\ref{unicode-pitfalls}). As an example of
+the current state-of-the-art, we discuss the International Phonetic Alphabet
+(IPA, Chapter~\ref{the-international-phonetic-alphabet}) and its rather
+problematic marriage with the Unicode Standard (Chapter
 \ref{ipa-meets-unicode}).
 
-The second part of the book (Sections~\ref{orthography-profiles} and
-\ref{use-cases}) describes our proposals for how to deal with the Unicode
-Standard in the daily practice of (comparative) linguistics. First, we discuss the
+The second part of the book describes our proposals for how to deal with the
+Unicode Standard in the daily practice of linguistics. First, we discuss the
 challenges of characterizing a writing system. To solve these problems, we
 propose the notions of orthography profiles, closely related to Unicode locale
-descriptions (Chapter~\ref{orthography-profiles}). Finally, we discuss practical
-issues with actual examples (Chapter~\ref{use-cases}). We provide reference
+descriptions (Chapter~\ref{orthography-profiles}). We provide reference
 implementation of our proposals in R and in Python, available as open-source
-libraries.
+libraries. An introduction to these implementations of our proposals is
+described in Chapter~\ref{implementation}. Finally, we discuss practical issues
+with actual examples in Chapter~\ref{use-cases}.
 
@@
 history as summarized in this section, we have used mostly rather ephemeral
-internet sources. When not references by traditional literature in the
+internet sources. When not referenced by traditional literature in the
 bibliography, we have used \url{http://www.unicode.org/history/} and various

### 2017-01-20 book/chapters/ipa_background.tex
cleanup

@@
 phenomenon, as if there were only one approved style.'' Clearly not all
-phoneticians agree (or will likely ever agree of course) on all aspects of the
+phoneticians agree (or will likely ever agree) on all aspects of the
 IPA or on transcription practices. As noted above, there have been several
@@
 Rather, the IPA aims to provide a resource that allows users to express any
-analysis so that it is widely understood. Basically, the IPA allows for both a `narrow' phonetic transcription and a
-`broad' phonological transcription. A `narrow' phonetic transcription may freely
-use the all symbols in the IPA chart with direct reference to their phonetic
-value, i.e. the transcriber can indicate with the symbols <ŋ͡m> that the
-phonetic value of the attested sound is a simultaneous labial and velar closure
-which is voiced and contains nasal airflow, independently of the phonemic status
-of this sound in the language in question.
+analysis so that it is widely understood. Specifically, the IPA allows for both
+a `narrow' phonetic transcription and a `broad' phonological transcription. A
+`narrow' phonetic transcription may freely use the all symbols in the IPA chart
+with direct reference to their phonetic value, i.e. the transcriber can indicate
+with the symbols <ŋ͡m> that the phonetic value of the attested sound is a
+simultaneous labial and velar closure which is voiced and contains nasal
+airflow, independently of the phonemic status of this sound in the language in
+question.
 
@@
 and [t̪], respectively. Similarly, the five vowels of Greek can be represented
-phonemically in IPA as /ieaou/, though phonetically they are closer to [iεauɔ].
+phonemically in IPA as /ieaou/, though phonetically they are closer to [iεaɔu].
 Similarly, the Japanese five vowel system can also be transcribed in IPA as
@@
 
-\citep[164]{IPA1999}.}
+\citep[164]{IPA1999}.
 
@@
 
-\begin(comment)
+\begin{comment}
 
@@
 EslingGaylord1993
+
+\end{comment}
 

### 2017-01-20 book/chapters/preface.tex
adds notes
cleanup

@@
 
-% Early phoneticians: Richard Lepsius, Isaac Pitman, Alexander J. Ellis, and Alexander Melville Bell
-% Later ones: Henry Sweet, L'abbé Jean-Pierre Rousselot, Eduard Sievers, Paul Passy, and Jens Otto Harry Jesperson
+% Very early attmept at notion for linguistic prosody: Joshua Steele (Prosodia Rationalis, 1779)
 
-In the 1840s, several researchers and teachers began developing notation to accurately transcribe articulatory gestures and their acoustic correlates.
-Their goal was not only transcription, they wanted to use their notation for spelling reform (especially English), for teaching literacy and foreign languages, and in particular for developing writing systems for unwritten languages.
+% Early phoneticians: Richard Lepsius (standard alphabet 1854, 1855), Isaac Pitman (Stenographic Soundhand 1837, Phonotypy 1844, English Phonotypic Alphabet 1845 with Ellis), Alexander J.
+Ellis (English Phonotypic Alphabet 1845), and Alexander Melville Bell (visibile speech 1864, 1867)
 
-At first, the challenge of devising a universal phonetic notation seemed straightforward: normalize the relationship between graphemes and sounds (<c> in `cake' -> /k/; <c> in `cell' -> /s/) to remove inconsistencies introduced by the historical divergence of written and spoken forms (e.g.\ in English and French orthographies), and keep adding new symbol-sound mappings until all sounds are accounted for.
+% Later phoneticians: Henry Sweet (Romic alphabet, aka Romic reform, in A Handbook of Phonetics 1877 - later becomes basis for IPA), L'abbé Jean-Pierre Rousselot ("founder of experimental phonetics"), Eduard Sievers, Paul Passy (founded the Phonetic Teachers' Association in 1886, later became IPAssociation), and Jens Otto Harry Jesperson
+
+% Instrumental measurements: Ludimar Hermann (used the phonograph to test theories of vowel production; coined formant)
+
+In the 1840s, several researchers and teachers began developing notation to
+accurately transcribe articulatory gestures and their acoustic correlates. Their
+goal was not only transcription, they wanted to use their notation for spelling
+reform (especially English), for teaching literacy and foreign languages, and in
+particular for developing writing systems for unwritten languages.
+
+At first, the challenge of devising a universal phonetic notation seemed
+straightforward: normalize the relationship between graphemes and sounds (<c> in
+`cake' -> /k/; <c> in `cell' -> /s/) to remove inconsistencies introduced by the
+historical divergence of written and spoken forms (e.g.\ in English and French
+orthographies), and keep adding new symbol-sound mappings until all sounds are
+accounted for.
 
@@
 
-This approach was successful for the first few languages investigated.
-But as the early phoneticians were confronted with more and more articulations, they encountered a wealth of data that they had not expected.
-The in-depth exploration of articulatory phonetics led to the discovery that the number of possible spoken sounds is infinite.
-\cite{Hockett1995} notes:
+This approach was successful for the first few languages investigated. But as
+the early phoneticians were confronted with more and more articulations, they
+encountered a wealth of data that they had not expected. The in-depth
+exploration of articulatory phonetics led to the discovery that the number of
+possible spoken sounds is infinite. \cite{Hockett1995} notes:
 
 \begin{quote}
-``But as they observed articulation more and more closely, and as they took more and more languages into consideration, the number of sounds they could discriminate and the number of symbols needed to denote them both increased seemingly without limit.
-In fact, as early as 1848, Ellis asserted that in principle the number of different sounds is endless.
-With the development of instrumental phonetics, largely by L'abbe Rousselt, it soon turned out that, given sufficiently accurate measuring devices, that is in fact the case.''
+
+But as they observed articulation more and more closely, and as
+they took more and more languages into consideration, the number of sounds they
+could discriminate and the number of symbols needed to denote them both
+increased seemingly without limit. In fact, as early as 1848, Ellis asserted
+that in principle the number of different sounds is endless. With the
+development of instrumental phonetics, largely by L'Abbé Rousselot, it soon
+turned out that, given sufficiently accurate measuring devices, that is in fact
+the case.
+
 \end{quote}
@@
 
-The first instrumental measurements in the 1870-80s showed unequivocally that spoken language is comprised of sounds waves; it is dynamic and not discrete.
-Given the fluid nature of the speech stream, sounds differ from one utterance to the next and the same sound may differ in different phonetic environments.
-However, many of these alterations are predictable in the context of their environments and they need not necessarily form distinctive lexical contrasts.
+The first instrumental measurements in the 1870-80s showed unequivocally that
+spoken language is comprised of sounds waves; it is dynamic and not discrete.
+Given the fluid nature of the speech stream, sounds differ from one utterance to
+the next and the same sound may differ in different phonetic environments.
+However, many of these alterations are predictable in the context of their
+environments and they need not necessarily form distinctive lexical contrasts.
 
-These findings left the early phoneticians in a conundrum given the aims of developing precise phonetic notion, but also practically-minded alphabetic-based writing.
-It is in this vein that the phonetic and phonemic distinction was introduced very early into transcription.
-The etic vs emic principle took hold, in essence, even before the dichotomy was widely accepted and before terminology for linguistic distinctiveness was introduced and adopted.
+These findings left the early phoneticians in a conundrum given the aims of
+developing precise phonetic notion, but also practically-minded alphabetic-based
+writing. It is in this vein that the phonetic and phonemic distinction was
+introduced very early into transcription. The etic vs emic principle took hold,
+in essence, even before the dichotomy was widely accepted and before terminology
+for linguistic distinctiveness was introduced and adopted.
 
-A transcription system that showed both contrastive and non-contrastive elements emerged, taking form notably in the International Phonetic Alphabet (which itself has been updated from time-to-time over the last nearly 120 years).
-The association between phonetic and phonemic would be maintained in the International Phonetic Alphabet and through the various instantiations of phonological theory, which developed on the work by early phoneticians.
+A transcription system that showed both contrastive and non-contrastive elements
+emerged, taking form notably in the International Phonetic Alphabet (which
+itself has been updated from time-to-time over the last nearly 120 years). The
+association between phonetic and phonemic would be maintained in the
+International Phonetic Alphabet and through the various instantiations of
+phonological theory, which developed on the work by early phoneticians.
 
-The model for phonemic contrast was alphabetic writing, in particular shallow orthography, i.e.\ one-to-one mappings between sound and symbol.
-Traditional alphabetic orthography, however, did not scale to precise phonetic notation, so the former was sprinkled with diacritics to denote non-contrastive phenomena.
-This dichotomy between phonemic and phonetic and its implementation in writing and consistent encoding causes many issues -- linguistic, theoretic, and most recently technological -- which we bring to light in this book.
+The model for phonemic contrast was alphabetic writing, in particular shallow
+orthography, i.e.\ one-to-one mappings between sound and symbol. Traditional
+alphabetic orthography, however, did not scale to precise phonetic notation, so
+the former was sprinkled with diacritics to denote non-contrastive phenomena.
+This dichotomy between phonemic and phonetic and its implementation in writing
+and consistent encoding causes many issues -- linguistic, theoretic, and most
+recently technological -- which we bring to light in this book.

### 2017-01-20 book/unicode.tex
typos
cleanup

@@
 \newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+\let\hlipl\hlkwb
 
@@
 \frontmatter
-% \include{chapters/preface}
+\include{chapters/preface}
 % \include{chapters/acknowledgments}
@@
 
-\include{chapters/preface}
+% \include{chapters/preface}
 \include{chapters/introduction}
@@
 \hlcom{# download and install the qlcData software}
-\hlkwd{install.packages}\hlstd{(}\hlstr{"qlcData"}\hlstd{)}
+\hlkwd{install.packages}\hlstd{(}\hlstr{'qlcData'}\hlstd{)}
 \hlcom{# load the software, so it can be used}
@@
 \hlcom{# download and install helper software}
-\hlkwd{install.packages}\hlstd{(}\hlstr{"devtools"}\hlstd{)}
+\hlkwd{install.packages}\hlstd{(}\hlstr{'devtools'}\hlstd{)}
 \hlcom{# install the qlcData package from GitHub}
-\hlstd{devtools}\hlopt{::}\hlkwd{install_github}\hlstd{(}\hlstr{"cysouw/qlcData"}\hlstd{,} \hlkwc{build_vignettes} \hlstd{=} \hlnum{TRUE}\hlstd{)}
-\hlcom{# load the software, so it can be used}
+\hlstd{devtools}\hlopt{::}\hlkwd{install_github}\hlstd{(}\hlstr{'cysouw/qlcData'}\hlstd{,} \hlkwc{build_vignettes} \hlstd{=} \hlnum{TRUE}\hlstd{)}
+\hlcom{# load the software, so it can be used }
 \hlkwd{library}\hlstd{(qlcData)}
@@
 \hlcom{# view vignette with explanation and examples}
-\hlkwd{vignette}\hlstd{(}\hlstr{"orthography_processing"}\hlstd{)}
+\hlkwd{vignette}\hlstd{(}\hlstr{'orthography_processing'}\hlstd{)}
 \end{alltt}
@@
 \hlcom{# show the path to the bash executables}
-\hlkwd{file.path}\hlstd{(}\hlkwd{find.package}\hlstd{(}\hlstr{"qlcData"}\hlstd{),} \hlstr{"exec"}\hlstd{)}
+\hlkwd{file.path}\hlstd{(}\hlkwd{find.package}\hlstd{(}\hlstr{'qlcData'}\hlstd{),} \hlstr{'exec'}\hlstd{)}
 \end{alltt}
@@
 # make softlinks to the R executables in /usr/local/bin
-ln -is $pathT $pathW /usr/local/bin
+# you will have to enter your user's password!
+sudo ln -is $pathT $pathW /usr/local/bin
 \end{alltt}
@@
 \begin{alltt}
-\hlkwd{launch_shiny}\hlstd{(}\hlstr{"tokenize"}\hlstd{)}
+\hlkwd{launch_shiny}\hlstd{(}\hlstr{'tokenize'}\hlstd{)}
 \end{alltt}
@@
 \begin{alltt}
-\hlstd{test} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"AABB"}\hlstd{,} \hlstr{"AАBВ"}\hlstd{)}
+\hlstd{( test} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{'AABB'}\hlstd{,} \hlstr{'\textbackslash{}u0041\textbackslash{}u0410\textbackslash{}u0042\textbackslash{}u0412'}\hlstd{) )}
+\end{alltt}
+\begin{verbatim}
+## [1] "AABB" "AАBВ"
+\end{verbatim}
+\begin{alltt}
 \hlkwd{write.profile}\hlstd{(test)}
@@
 \begin{alltt}
-\hlstd{test} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"AABB"}\hlstd{,} \hlstr{"AАBВ"}\hlstd{)}
-\hlkwd{tokenize}\hlstd{(test,} \hlkwc{profile} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{"A"}\hlstd{,} \hlstr{"B"}\hlstd{))}
+\hlstd{test} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{'AABB'}\hlstd{,} \hlstr{'\textbackslash{}u0041\textbackslash{}u0410\textbackslash{}u0042\textbackslash{}u0412'}\hlstd{)}
+\hlkwd{tokenize}\hlstd{( test,} \hlkwc{profile} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{'A'}\hlstd{,} \hlstr{'B'}\hlstd{) )}
 \end{alltt}
@@
 \begin{alltt}
-\hlstd{example} \hlkwb{<-} \hlstr{"ÙÚÛÙÚÛ"}
+\hlstd{example} \hlkwb{<-} \hlstr{'\textbackslash{}u00d9\textbackslash{}u00da\textbackslash{}u00db\textbackslash{}u0055\textbackslash{}u0300\textbackslash{}u0055\textbackslash{}u0301\textbackslash{}u0055\textbackslash{}u0302'}
 \hlstd{profile_1} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example)}
@@
 
-% latex table generated in R 3.3.0 by xtable 1.8-2 package
-% Thu Aug 18 15:23:10 2016
+% latex table generated in R 3.3.2 by xtable 1.8-2 package
+% Fri Jan 20 14:31:00 2017
 \begin{table}[H]
@@
 
-% latex table generated in R 3.3.0 by xtable 1.8-2 package
-% Thu Aug 18 15:23:11 2016
+% latex table generated in R 3.3.2 by xtable 1.8-2 package
+% Fri Jan 20 14:31:01 2017
 \begin{table}[htb]
@@
 \hlcom{# make a profile, just select the column 'Grapheme'}
-\hlstd{profile} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example)[,} \hlstr{"Grapheme"}\hlstd{]}
+\hlstd{profile} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example)[,}\hlstr{"Grapheme"}\hlstd{]}
 \hlcom{# extend the profile with multigraphs}

### 2017-02-02 book/chapters/implementation.tex
updates book for initial public release

--- 2016-08-18 book/chapters/implementation.tex
@@
 
-\section{Available implementations}
-\label{available-implementations}
-
 To illustrate the practical application of orthography profiles as defined in
 the previous Chapter \ref{orthography-profiles}, we have implemented two 
-different versions of the specifications presented there: one in Python and one
-in R. These two implementations have rather different implementation histories,
+versions of the specifications presented there: one in Python and one
+in R.
+
+These two libraries have rather different implementation histories,
 and we are not 100\% sure that they will in all situations give the same
-results. Also the perfomance with larger datasets differs, and the code is not
+results. We do however provide a test suite for each implementation. 
+Users should refer to the tests cases and to the documentation in each release 
+for specifics about each implementation. Note also that function names differ 
+in the two implementations.
+
+Note also that the performance with larger datasets differs, and the code is not
 always as clean as we would like it to be. In sum, the two implementations
@@
 somewhat abstract examples will be discussed to show the different options 
-available, and to illustrate the intended usage of orthography profiles. In the 
-next chapter we will then discuss a few full-scale examples to illustrate the 
-application and the usefulness of orthography profiles.
+available, and to illustrate the intended usage of orthography profiles.
 
-\subsection*{Installing the R implementation}
+This chapter is split into two sections, one for the Python implementation, the 
+other for R. Each section has a number of subsections that ...
+
+In addition to the material presented here to get users started, we maintain 
+several case studies that use each OP implementation in action. These use cases 
+are an the accompanying Github repository available with this book at: 
+\url{https://github.com/unicode-cookbook/recipes/}.
+
+% TODO: point to the case studies
+% In the next chapter we will then discuss a few full-scale examples to illustrate the application and the usefulness of orthography profiles.
+
+\section{Python}
+\label{python-implementations}
+
+The Python package is available both as a command line interface (CLI) 
+and as an appplication programming interface (API).
+
+\subsection*{Installation}
+
+To install the Python package from the Python Package Index (PyPI) run:
+
+\begin{lstlisting}[language=bash]
+  $ pip install segments
+\end{lstlisting}
+
+\noindent on the command line. This will give you access to both 
+the CLI and programmatic functionality in Python scripts, when 
+you import the \texttt{segments} library.
+
+You can also install the \texttt{segments} 
+package with from the Github repository:
+
+\begin{lstlisting}[language=bash]
+  $ git clone https://github.com/bambooforest/segments.git
+  $ cd segments
+  $ python setup.py develop
+\end{lstlisting}
+
+\subsection*{Command line interface}
+
+From the command line, access \texttt{segments} and its 
+various arguments. For help, run:
+
+\begin{lstlisting}[language=bash]
+  $ segments -h
+\end{lstlisting}
+
+\noindent Here is an example of how to create and use an orthography 
+profile for segmentation. Create a text file:
+
+\begin{lstlisting}[language=bash]
+  $ more text.txt
+  aäaaöaaüaa
+\end{lstlisting}
+
+\noindent Now look at the profile:
+
+\begin{lstlisting}[language=bash,texcl=true]
+  $ cat text.txt | segments profile
+  Grapheme        frequency       mapping
+  a       7       a
+  ä       1       ä
+  ü       1       ü
+  ö       1       ö
+\end{lstlisting}
+
+\noindent Write the profile to a file:
+
+\begin{lstlisting}[language=bash]
+  $ cat text.txt | segments profile > profile.prf
+\end{lstlisting}
+
+\noindent Edit the profile:
+
+\begin{lstlisting}[language=bash]
+  $ more profile.prf
+  Grapheme        frequency       mapping
+  aa      0       x
+  a       7       a
+  ä       1       ä
+  ü       1       ü
+  ö       1       ö
+\end{lstlisting}
+
+\noindent Now tokenize the text without profile:
+
+\begin{lstlisting}[language=bash]
+  $ cat text.txt | segments tokenize
+  a ä a a ö a a ü a a	
+\end{lstlisting}
+
+\noindent And with profile:
+\begin{lstlisting}[language=bash]
+  $ cat text.txt | segments --profile=profile.prf tokenize
+  a ä aa ö aa ü aa
+
+  $ cat text.txt | segments --mapping=mapping --profile=profile.prf tokenize
+  a ä x ö x ü x
+\end{lstlisting}
+
+\subsection*{Application programming interface}
+The \texttt{segments} API can be accesses by importing 
+the package into Python a script. Here is an example of 
+how to import the libraries, create a tokenizer object, 
+tokenize a string and create a profile.
+
+\begin{lstlisting}[language=python]
+>>> from __future__ import unicode_literals, print_function
+>>> from segments import Profile, Tokenizer
+>>> t = Tokenizer()
+>>> t('abcd')
+'a b c d'
+>>> prf = Profile({'Grapheme': 'ab', 'mapping': 'x'}, \
+{'Grapheme': 'cd', 'mapping': 'y'})
+>>> print(prf)
+mapping Grapheme
+ab  x
+cd  y
+>>> t = Tokenizer(profile=prf)
+>>> t('abcd')
+'ab cd'
+>>> t('abcd', column='mapping')
+'x y'
+\end{lstlisting}
+
+\section{R}
+\label{r-implementation}
+
+% \subsection*{Installing the R implementation}
+\subsection*{Installation}
 
@@
 
-TO DO
+% TO DO
 
-\subsection*{Installung the Python implementation}
-
-TO DO
-
-\section{Profiles and error reporting}
+\subsection*{Profiles and error reporting}
 \label{error-reporting}
@@
 
-\section{Different ways to write a profile}
+\subsection*{Different ways to write a profile}
 \label{write-profile}
@@
 Note that these different profiles can also be produced using the bash
-executable \texttt{writeprofile} (see \ref{available-implementations} for notes
+executable \texttt{writeprofile} (see \ref{r-implemenation} for notes
 on how to install the bash executable). Exactly this example is also included in
@@
 
-\section{Using an orthography profile skeleton}
+\subsection*{Using an orthography profile skeleton}
 \label{profile-skeleton}
@@
 
-\section{Rule ordering}
+\subsection*{Rule ordering}
 \label{rule-ordering}
@@
 
-\section{Contextually specified graphemes}
+\subsection*{Contextually specified graphemes}
 \label{contextual-specification}
@@
 
-\section{Profile skeleton with columns for editing}
+\subsection*{Profile skeleton with columns for editing}
 \label{profile-editing}
@@
 
-\section{Formatting grapheme separation}
+\subsection*{Formatting grapheme separation}
 \label{formattingseparation}
@@
 
-\section{Remaining issues}
+\subsection*{Remaining issues}
 

### 2017-02-02 book/chapters/introduction.tex
updates book for initial public release

--- 2016-08-18 book/chapters/introduction.tex
@@
 the enormous variability of existing writing systems for the world's languages
-is a fact of life scholars have to accept and should try to adapt to as good as
-possible. Our plea in this book is a proposal how to do exactly that: formalize 
+is a fact of life scholars have to accept and should try to adapt to as well as
+we can. Our plea in this book is a proposal how to do exactly that: formalize 
 knowledge about individual writing systems in a form that is both easy to use 
@@
 diversity.\footnote{
-cf.~\citet{Steiner_etal2011,List2012,List2012a,ListMoran2013,MoranProkic2013}}
+Cf.~\citet{Steiner_etal2011,List2012,List2012a,ListMoran2013,MoranProkic2013}}
 In this work, it is crucial to be able to flexibly process across numerous
@@
 allows for a thorough understanding of possible pitfalls that one might
-encounter when using it (Chapter~\ref{unicode-pitfalls}). As an example of the
-current state-of-the-art, we discuss the rather problematic marriage of the
-International Phonetic Alphabet (IPA) with the Unicode Standard (Chapter
-\ref{ipa-meets-unicode}).
+encounter when using it in general (Chapter~\ref{unicode-pitfalls}).
+Linguists are more often interested in using the Unicode Standard with the International Phonetic Alphabet (IPA).
+We first provide a history of the development of the IPA and early attempts to encode it electronically (Chapter~\ref{the-international-phonetic-alphabet}) before we discuss the rather problematic marriage of the International Phonetic Alphabet (IPA) with the Unicode Standard (Chapter\ref{ipa-meets-unicode}).
 
-The second part of the book (Sections~\ref{orthography-profiles} and
-\ref{use-cases}) describes our proposals for how to deal with the Unicode
+In the second part of the book (Chapters~\ref{orthography-profiles} and
+\ref{implementation}) we describe our proposals of how to deal with the Unicode
 Standard in the daily practice of (comparative) linguistics. First, we discuss the
@@
 descriptions (Chapter~\ref{orthography-profiles}). Finally, we discuss practical
-issues with actual examples (Chapter~\ref{use-cases}). We provide reference
+issues with actual examples (Chapter~\ref{implementation}). We provide reference
 implementation of our proposals in R and in Python, available as open-source
@@
 
-The following conventions are followed in this paper. All phonemic and phonetic
+\subsubsection*{Conventions}
+
+The following conventions are followed in this book. All phonemic and phonetic
 representations are given in the International Phonetic Alphabet (IPA), unless

### 2017-02-02 book/chapters/ipa_background.tex
updates book for initial public release

--- 2015-12-10 book/chapters/ipa_background.tex
@@
 % ==========================
+
+In this chapter we present a brief history of the IPA 
+(Section~\ref{IPAhistory}), which itself dates 
+back to the late 19th century, not long after the creation of the first 
+typewriter with a QWERTY 
+keyboard. An understanding of the IPA and its premises and principles 
+(Section~\ref{IPApremises-principles}) leads to a better 
+appreciation of the challenges that the IPA association 
+faced when digitally encoding the IPA's set of symbols and 
+diacritics (Section~\ref{EncodingIPA}). Occurring a bit over one hundred 
+years after the IPA's inception, the encoding challenge was no small feat 
+(Section~\ref{need-for-multilingual-environment}); many 
+linguists have encountered the pitfalls when the twain shall meet 
+(Chapter~\ref{ipa-meets-unicode}).
 
@@
 \textit{Association}) has long maintained a standard alphabet, the
-\textsc{international phonetic alphabet} (henceforth \textit{IPA}), which is a
-common standard in linguistics to transcribe sounds of spoken languages. It was
+\textsc{international phonetic alphabet} or IPA, which is a
+standard in linguistics to transcribe sounds of spoken languages. It was
 first published in 1888 as an international system of phonetic transcription for
@@
 over time. It is natural then that the IPA is modified occasionally to
-accommodate scientific innovations and discoveries. In fact, this is part of the
-Association's mandate. These changes are captured in the recurring revisions to
-the IPA.
+accommodate scientific innovations and discoveries. In fact, updates are part of the
+Association's mandate. These changes are captured in revisions to the IPA chart.
 %\footnote{For a detailed history see:
@@
 
-Over the years there have been several revisions, many minor. For example, the
-articulation labels -- what often are call `features' even though the IPA
+Over the years there have been several revisions, but mostly minor. Articulation 
+labels -- what are often called `features' even though the IPA
 deliberately avoids this term -- have changed (e.g.\ terms like `lips', `throat'
@@
 have been added to reflect contrastive sounds found in the world's very diverse
-phonological systems. The current construction and use of the IPA are guided by
-principles outlined in the \textsc{handbook of the international phonetic
-association} \citep{IPA1999}, henceforth simply called \textit{Handbook}.
+phonological systems. The use of the IPA is guided by principles outlined in 
+the \textsc{handbook of the international phonetic association} \citep{IPA1999}, 
+henceforth simply called \textit{Handbook}.
 
@@
 Several styles of transcription with the IPA are possible, as illustrated in the
-\textit{Handbook}, and they are all valid.\footnote{For illustration of
-different possibilities, see the 29 languages and their transcriptions in the
-original and initial \textit{Illustrations of the IPA}
-\citep[41--154]{IPA2007}} Therefore, there are different but equivalent
-transcriptions, or as \citet[64]{Ladefoged1990a} captures it, ``perhaps now that
-the Association has been explicit in its eclectic approach, outsiders to the
+\textit{Handbook}, and they are all valid.\footnote{For an illustration of
+the differences, see the 29 languages and their transcriptions in the
+\textit{Illustrations of the IPA}\citep[41--154]{IPA2007}.} Therefore, there are 
+different but equivalent transcriptions, or as noted by \citet[64]{Ladefoged1990a}, 
+``perhaps now that the Association has been explicit in its eclectic approach, outsiders to the
 Association will no longer speak of \textit{the} IPA transcription of a given
 phenomenon, as if there were only one approved style.'' Clearly not all
-phoneticians agree (or will likely ever agree of course) on all aspects of the
-IPA or on transcription practices. As noted above, there have been several
-revisions in the IPA's long history, but the current version (2005) is
-strikingly similar to the 1926 version, showing the viability of the IPA as a
+phoneticians agree, nor are they likely to ever completely agree, on all aspects of the
+IPA or on transcription approaches and practices in general. As noted above, 
+there have been several revisions in the IPA's long history, but the current version (2005) is
+strikingly similar to the 1926 version, which shows the viability of the IPA as a
 common standard for linguistic transcription.
@@
 % ==========================
-\section{Premises}
+
+\section{Premises and principles}
+\label{IPApremises-principles}
+% ==========================
+\subsection*{Premises}
 \label{IPApremises}
-% ==========================
 
@@
 % ==========================
-\section{Principles}
+\subsection*{Principles}
 \label{IPAprinciples}
@@
 
-Transcription has essentially two parts. The first is a text containing IPA 
-symbols and the second is a set of conventions for interpreting those 
-symbols (and their combinations).
-
-First, the symbols of IPA crucially distinguish between letter-like symbols and
-diacritics. The use of the letter-like symbols to represent a language's sounds
-is guided by the principle of contrast; where two words are distinguishable by
-phonemic contrast, those contrasts should be transcribed with different letter
-symbols (not diacritics). Allophonic distinction falls under the rubric of
-diacritically-distinguished symbols, e.g. the difference in English between and
-unrealized /p/ in [stop̚] vs. an aspirated /p/ in [spʰot] should be marked with
-a diacritic. In sum:
+Essentially transcription has two parts: a text containing symbols and a set 
+of conventions for interpreting those symbols (and their combinations). 
+The symbols of the IPA distinguish between letter-like symbols and
+diacritics (symbol modifiers). The use of the letter-like symbols to represent 
+a language's sounds is guided by the principle of contrast; where two words 
+are distinguishable by phonemic contrast, those contrasts should be transcribed 
+with different letter symbols (and not just diacritics). Allophonic distinction 
+falls under the rubric of diacritically-distinguished symbols, e.g.\ the 
+difference in English between an aspirated /p/ in [spʰot] and 
+an unrealeased /p/ in [stop̚].
 
@@
 	\item Different letter-like symbols should be used whenever
-          a language employs two sound contrastively.
+          a language employs two sounds contrastively.
 	\item When two sounds in a language are not known to be contrastive, the same
@@
 
-\noindent Yet, in some situations diacritics can be used to mark phonemic
+\noindent Yet, in some situations diacritics are used to mark phonemic
 contrasts. The \textit{Handbook} recommends to limit the use of phonemic
@@
 Rather, the IPA aims to provide a resource that allows users to express any
-analysis so that it is widely understood. Basically, the IPA allows for both a `narrow' phonetic transcription and a
-`broad' phonological transcription. A `narrow' phonetic transcription may freely
-use the all symbols in the IPA chart with direct reference to their phonetic
-value, i.e. the transcriber can indicate with the symbols <ŋ͡m> that the
-phonetic value of the attested sound is a simultaneous labial and velar closure
-which is voiced and contains nasal airflow, independently of the phonemic status
-of this sound in the language in question.
+analysis so that it is widely understood. Basically, the IPA allows for both a 
+`narrow' phonetic transcription and a `broad' phonological transcription. 
+A `narrow' phonetic transcription may freely use the all symbols in the IPA 
+chart with direct reference to their phonetic value, i.e. the transcriber can 
+indicate with the symbols <ŋ͡m> that the phonetic value of the attested sound 
+is a simultaneous labial and velar closure which is voiced and contains nasal 
+airflow, independently of the phonemic status of this sound in the language in 
+question.
 
@@
 % ==========================
-\section{Computational representation}
+
+\section{IPA encodings}
+\label{EncodingIPA}
 % ==========================
 
-In 1989 an IPA revision convention was held in Kiel, Germany. As per other
-revisions, there was expansion and revisions to phonetic symbols in the IPA
-chart. Notably the marking of tone was extended with the addition of a second
-system for marking linguistic tones (the so-called `Chao' tone letters).
-Importantly, however, this was also the first revision to address issues of
-computational representations for the IPA.
+In 1989, an IPA revision convention was held in Kiel, Germany. As in previous meetings, 
+there were changes made to the repertoire of phonetic symbols in the IPA chart, which 
+reflected what had been discovered, described and cataloged by linguists about the 
+phonological systems in the world's languages in the interim. Personal computers 
+were also becoming more commonplace, and linguists were using them to create databases. 
+A cogent example is the UCLA Phonological 
+Segment Inventory Database (UPSID; \citealt{Maddieson1984}), which was expanded 
+\citep{MaddiesonPrecoda1990} and then encoded and distributed in a computer program 
+\citep{MaddiesonPrecoda1992}.\footnote{It could be installed via 
+floppy disk on an IBM PC, or compatible, running 
+DOS with 1MB free disk space and 360K available RAM.} The programmers used 
+only ASCII characters to maximize compatibility (e.g.\ <kpW> for $[$kpʷ$]$), but 
+were faced with unavoidable arbitrary mappings between ASCII letters and 
+punctuation, and the more than 900 segment types documented 
+in their sample of world's languages' phonological systems. The developers 
+devised a system of base characters with secondary diacritic marks 
+(e.g.\ in the previous example <kp>, the base character, is modified with <W>). 
+This encoding approach is 
+also used in SAMPA and X-SAMPA (Section~\ref{sampa-xsampa}) and in the 
+ASJP.\footnote{See the ASJP use case in the online supplementary 
+materials to this book: \url{https://github.com/unicode-cookbook/use_cases}.} 
+But before UPSID, SAMPA and ASJP, IPA was encoded with numbers.
 
+% \footnote{The marking of tone was extended (from characteristically Africanist practice) to include a second system for marking linguistic tones (the `Chao tone letters'), a much used convention based on musical staff to describe pitch in by Yuen Ren Chao (1892--1982).}
+\subsection*{IPA Numbers}
 Prior to the Kiel Convention for the modern revision of the IPA in 1989,
@@
 
-The working group insightfully decided that the computing-coding convention 
+The working group decided insightfully that the computing-coding convention 
 for the IPA should be independent of computer environments or formats (e.g.\ ASCII), 
@@
 
-\noindent Thus the coding convention was not meant to address the engineering 
+\noindent The coding convention was not meant to address the engineering 
 aspects of the actual encoding in computers (cf.\ \cite{Anderson1984}). However, 
@@
 creating mapping tables from various computer encodings, fonts, phonetic-character-set 
-software, etc., to common IPA Numbers, and thus symbols.\footnote{Remember, at 
+software, etc., to common IPA Numbers, and therefore symbols.\footnote{Remember, at 
 this time in the late 1980s there was no stable multilingual computing environment. 
@@
 computers for their research and many were quickly adopting electronic mail or 
-discussion boards like Usenet as a medium for international exchanges. Most of these systems ran on 8-bit 
-hardware systems using a 7-bit ASCII character encoding.}
+discussion boards like Usenet as a medium for international exchanges. 
+Most of these systems ran on 8-bit hardware systems using a 7-bit ASCII character encoding.}
 
@@
 the various series by appending them, e.g. [\charis{ⱱ}] 184. Former non-IPA
-symbols (or often-used but non-offical IPA symbols) for consonants, vowels and
+symbols (or often-used but non-official IPA symbols) for consonants, vowels and
 diacritics are numbered from x99 backwards. For example, the voiceless and
@@
 % ==========================
-\section{SAMPA and X-SAMPA}
+\subsection*{SAMPA and X-SAMPA}
+\label{sampa-xsampa}
 % ==========================
@@
 practical recommendations for language resources, including notes on encoding,
-can be found in \citet{BirdSimons2003}} For example, some chose to represent
+can be found in \citet{BirdSimons2003}.} For example, some chose to represent
 unavailable graphemes with substitutes, e.g.~the combination of <ng> to
@@
 % ==========================
+
 \section{The need for a single multilingual environment}
+\label{need-for-multilingual-environment}
 % ==========================
@@
 
-% ===============================================================
+% ==========================
 
-\begin(comment)
-
-\subsection{IPA (practical) extensions}
-
-The IPA extensions cover speech sounds beyond the sound systems of the world's
-languages, including symbols for pathological speech and paralinguistic
-functions (e.g.\ Braille?).
-
-- pathological speech
-- paralinguistic functions (Braille?)
-
-As noted in Section \ref{}, extensions to the IPA -- in particular
-``large-scale'' extensions --
-
-% Englebretson2009.pdf
-http://www.ruf.rice.edu/~reng/englebretson2009.pdf
-
-X-SAMPA was an ASCII transliteration and is still purposeful for several
-programs
-
-There is also the extensions of the IPA designed for disordered speech, although
-the symbols are occasionally used for transcribing phonetic detail in ``normal''
-speech or also cross-linguistic
-rarities.\footnote{\url{http://phoible.github.io/conventions/}}
-
-% Unicode has its own history; for a summary:
-\url{http://www.unicode.org/history/summary.html}
-
-https://www.phon.ucl.ac.uk/home/wells/ipa-unicode.htm
-
-Computer codes for phonetic symbols
-EslingGaylord1993
-
-% ==========================
+\begin{comment}
 \section{Unicode and ISO 10646}
@@
 table of commercial importance as measured by GDP of countries using certain 
-writing systems \citep[2]{unicode88}.
+writing systems \citep[2]{Becker1988}.
 
-The original Unicode manifesto is \cite{unicode88.pdf}.\footnote{http://www.unicode.org/history/unicode88.pdf} 
+The original Unicode manifesto is \cite{Becker1988}.\footnote{http://www.unicode.org/history/unicode88.pdf} 
 Its aim was for a reliable international multilingual text encoding standard 

### 2017-02-02 book/chapters/pitfalls.tex
minor changes
updates book for initial public release

--- 2015-11-13 book/chapters/pitfalls.tex
@@
 \textsc{latin small letter g} at \uni{0067}). In contrast, the symbols
-<{\fontspec{EversonMono}AАΑᎪᗅᴀꓮ𝖠𝙰}> are all different code points,
+<{\fontspec{EversonMono}AАΑᎪᗅᴀꓮ𐊠𝖠𝙰}> are all different code points,
 although they look highly similar---in some cases even sharing exactly the same
@@
 %	\item[] <{\fontspec{EversonMono}Ａ}> \textsc{fullwidth latin capital letter a}, at \uni{FF21} 
-	\item[] <{\fontspec{EversonMono}ʠ}> \textsc{carian letter a}, at \uni{102A0} 
+	\item[] <{\fontspec{EversonMono}𐊠}> \textsc{carian letter a}, at \uni{102A0} 
 %	\item[] <{\fontspec{EversonMono}̀}> \textsc{old italic letter a}, at \uni{10300} 
@@
 sufficient for the large majority of Unicode characters, but not for all of
-them. A special system of \textsc{surrogates} is defined within the Unicode Standard to
-deal with these additional characters. The effect is a more disk-space efficient
-encoding (approximately half the size), while adding a limited computational
-overhead to manage the surrogates.
+them. A special system of \textsc{surrogates} is defined within the Unicode
+Standard to deal with these additional characters. The effect is a more
+disk-space efficient encoding (approximately half the size), while adding a
+limited computational overhead to manage the surrogates.
 

### 2017-02-02 book/unicode.tex
updates book for initial public release

--- 2016-08-18 book/unicode.tex
@@
 
-\include{chapters/preface}
 \include{chapters/introduction}
@@
 somewhat abstract examples will be discussed to show the different options 
-available, and to illustrate the intended usage of orthography profiles. In the 
-next chapter we will then discuss a few full-scale examples to illustrate the 
-application and the usefulness of orthography profiles.
-
-\subsection*{Installing the R implementation}
+available, and to illustrate the intended usage of orthography profiles.
+
+This chapter is split into two sections, one for the Python implementation, the 
+other for R. Each section has a number of subsections that ...
+
+In addition to the material presented here to get users started, we maintain 
+several case studies that use each OP implementation in action. These use cases 
+are an the accompanying Github repository available with this book at: 
+\url{https://github.com/unicode-cookbook/recipes/}.
+
+% TODO: point to the case studies
+% In the next chapter we will then discuss a few full-scale examples to illustrate the application and the usefulness of orthography profiles.
+
+\section{Python}
+\label{python-implementations}
+
+The Python package is available both as a command line interface (CLI) 
+and as an appplication programming interface (API).
+
+\subsection*{Installation}
+
+To install the Python package from the Python Package Index (PyPI) run:
+
+\begin{lstlisting}[language=bash]
+  $ pip install segments
+\end{lstlisting}
+
+\noindent on the command line. This will give you access to both 
+the CLI and programmatic functionality in Python scripts, when 
+you import the \texttt{segments} library.
+
+You can also install the \texttt{segments} 
+package with from the Github repository:
+
+\begin{lstlisting}[language=bash]
+  $ git clone https://github.com/bambooforest/segments.git
+  $ cd segments
+  $ python setup.py develop
+\end{lstlisting}
+
+\subsection*{Command line interface}
+
+From the command line, access \texttt{segments} and its 
+various arguments. For help, run:
+
+\begin{lstlisting}[language=bash]
+  $ segments -h
+\end{lstlisting}
+
+\noindent Here is an example of how to create and use an orthography 
+profile for segmentation. Create a text file:
+
+\begin{lstlisting}[language=bash]
+  $ more text.txt
+  aäaaöaaüaa
+\end{lstlisting}
+
+\noindent Now look at the profile:
+
+\begin{lstlisting}[language=bash,texcl=true]
+  $ cat text.txt | segments profile
+  Grapheme        frequency       mapping
+  a       7       a
+  ä       1       ä
+  ü       1       ü
+  ö       1       ö
+\end{lstlisting}
+
+\noindent Write the profile to a file:
+
+\begin{lstlisting}[language=bash]
+  $ cat text.txt | segments profile > profile.prf
+\end{lstlisting}
+
+\noindent Edit the profile:
+
+\begin{lstlisting}[language=bash]
+  $ more profile.prf
+  Grapheme        frequency       mapping
+  aa      0       x
+  a       7       a
+  ä       1       ä
+  ü       1       ü
+  ö       1       ö
+\end{lstlisting}
+
+\noindent Now tokenize the text without profile:
+
+\begin{lstlisting}[language=bash]
+  $ cat text.txt | segments tokenize
+  a ä a a ö a a ü a a	
+\end{lstlisting}
+
+\noindent And with profile:
+\begin{lstlisting}[language=bash]
+  $ cat text.txt | segments --profile=profile.prf tokenize
+  a ä aa ö aa ü aa
+
+  $ cat text.txt | segments --mapping=mapping --profile=profile.prf tokenize
+  a ä x ö x ü x
+\end{lstlisting}
+
+\subsection*{Application programming interface}
+The \texttt{segments} API can be accesses by importing 
+the package into Python a script. Here is an example of 
+how to import the libraries, create a tokenizer object, 
+tokenize a string and create a profile.
+
+\begin{lstlisting}[language=python]
+>>> from __future__ import unicode_literals, print_function
+>>> from segments import Profile, Tokenizer
+>>> t = Tokenizer()
+>>> t('abcd')
+'a b c d'
+>>> prf = Profile({'Grapheme': 'ab', 'mapping': 'x'}, \
+{'Grapheme': 'cd', 'mapping': 'y'})
+>>> print(prf)
+mapping Grapheme
+ab  x
+cd  y
+>>> t = Tokenizer(profile=prf)
+>>> t('abcd')
+'ab cd'
+>>> t('abcd', column='mapping')
+'x y'
+\end{lstlisting}
+
+\section{R}
+\label{r-implementation}
+
+% \subsection*{Installing the R implementation}
+\subsection*{Installation}
 
@@
 \definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-# get the paths to the R executables in bash
-pathT=`Rscript -e 'cat(file.path(find.package("qlcData"), "exec", "tokenize"))'`
-pathW=`Rscript -e 'cat(file.path(find.package("qlcData"), "exec", "writeprofile"))'`
-
-# make softlinks to the R executables in /usr/local/bin
-ln -is $pathT $pathW /usr/local/bin
-\end{alltt}
+\noindent
+\ttfamily
+\hlstd{}\hlslc{\#\ get\ the\ paths\ to\ the\ R\ executables\ in\ bash}\hspace*{\fill}\\
+\hlstd{pathT}\hlopt{=}\hlstd{}\hlstr{`Rscript\ {-}e\ 'cat(file.path(find.package("qlcData"),\ "exec",\ "tokenize"))'`}\hlstd{}\hspace*{\fill}\\
+\hlstd{pathW}\hlopt{=}\hlstd{}\hlstr{`Rscript\ {-}e\ 'cat(file.path(find.package("qlcData"),\ "exec",\ "writeprofile"))'`}\hlstd{}\hspace*{\fill}\\
+\hlstd{}\hspace*{\fill}\\
+\hlstd{}\hlslc{\#\ make\ softlinks\ to\ the\ R\ executables\ in\ /usr/local/bin}\hspace*{\fill}\\
+\hlstd{}\hlkwc{ln\ }\hlstd{}\hlkwb{{-}is\ }\hlstd{}\hlkwd{\$pathT\ \$pathW\ }\hlstd{}\hlopt{/}\hlstd{usr}\hlopt{/}\hlstd{local}\hlopt{/}\hlstd{bin}\hspace*{\fill}
+\mbox{}
+\normalfont
 \end{kframe}
@@
 \definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-tokenize --help
-\end{alltt}
+\noindent
+\ttfamily
+\hlstd{tokenize\ }\hlkwb{{-}{-}help}\hlstd{}\hspace*{\fill}
+\mbox{}
+\normalfont
 
 \begin{verbatim}
-## USAGE: 
-##   tokenize [-hlrd] [-s SEP -p REPL] [-t TRANS -m MISSING] [<STRINGS> <PROFILE>]
-## 
-## DESCRIPTION:
-##   Using the function tokenize() from the R-package qlcData for tokenization and
-##   transliteration of character strings based on an orthography profile. Details
-##   http://www.rdocumentation.org/packages/qlcData/functions/tokenize.html
-## 
-##   STRINGS can be piped through. When no PROFILE is specified, default Unicode
-##   tokenization is performed. Note that the R-code allows for even more options,
-##   not all options are made available here for ease of use.
-##   Errors are likewise (not yet) returned
-## 
-## OPTIONS:
-##   -h, --help      Showing this help text
-##   -l, --linear    Use linear method instead of default global method
-##   -r, --regex     Use regex matching, including contexts in matching graphemes
-##   -d, --NFD       Use NFD normalization instead of default NFC
-##   -s SEP          Separator to be inserted after tokenization
-##                     defaults to space [default: SPACE]
-##   -p REPL         Replacement symbol in case separator occurs in the STRINGS
-##                     defaults to nothing [default: NULL]
-##   -t TRANS        Column in profile to use for transliteration [default: NULL]
-##   -m MISSING      Character to be inserted for characters not in the PROFILE
-##                     defaults to DOUBLE QUESTION MARK at U+2047 [default: ⁇]
-## 
-## EXAMPLES:
-##   Try the difference between the following:
-## 
-##     tokenize áüî
-##     tokenize --NFD -s '__' áüî
-##   
-##   Piping strings should work:
-##   
-##     ls | tokenize
+## Error in running command bash
 \end{verbatim}
@@
 
-TO DO
-
-\subsection*{Installung the Python implementation}
-
-TO DO
-
-\section{Profiles and error reporting}
+% TO DO
+
+\subsection*{Profiles and error reporting}
 \label{error-reporting}
@@
 
-% latex table generated in R 3.3.0 by xtable 1.8-2 package
-% Thu Aug 18 15:23:10 2016
+% latex table generated in R 3.3.2 by xtable 1.8-2 package
+% Thu Feb  2 11:38:04 2017
 \begin{table}[H]
@@
 \definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-writeprofile --help
-\end{alltt}
+\noindent
+\ttfamily
+\hlstd{writeprofile\ }\hlkwb{{-}{-}help}\hlstd{}\hspace*{\fill}
+\mbox{}
+\normalfont
 
 \begin{verbatim}
-## USAGE: 
-##   writeprofile [-hiec] [-s SEP] [-n NORMALIZE] [<STRINGS>]
-## 
-## DESCRIPTION:
-##   Using the function write.profile() from the R-package qlcData to prepare 
-##   a skeleton for an orthography profile to be used with tokenize. Details:
-##   http://www.rdocumentation.org/packages/qlcData/functions/readwrite_orthography
-## 
-##   STRINGS can be piped through. The results is a tab-delimited file for manual
-##   editing. Note that the R-code allows for even more options, not all options 
-##   are made available here for ease of use.
-## 
-## OPTIONS:
-##   -h, --help        Showing this help text
-##   -i, --info        Add columns with Unicode information to profile
-##   -e, --editing     Add empty columns for easy editing of profile
-##   -c, --codepoints  Separate by Unicode codepoints (overrides any -s setting)
-##   -s SEP            Separator used to separate the strings
-##                       defaults to unicode character definition [default: NULL]
-##   -n NORMALIZE      Normalization used to make profile. Possible options: NFD, NFC
-##                       defaults to no normalization [default: NULL]
-##                     
-## EXAMPLES:
-##   Note the differences between the following variants:
-## 
-##     example='ÙÚÛÙÚÛ'
-##     echo -e $example
-##     echo -e $example | writeprofile -i
-##     echo -e $example | writeprofile -ic
-##     echo -e $example | writeprofile -ic -n NFC
-##     echo -e $example | writeprofile -ic -n NFD
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
-
-\section{Using an orthography profile skeleton}
+## Error in running command bash
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+\subsection*{Using an orthography profile skeleton}
 \label{profile-skeleton}
@@
 
-% latex table generated in R 3.3.0 by xtable 1.8-2 package
-% Thu Aug 18 15:23:11 2016
+% latex table generated in R 3.3.2 by xtable 1.8-2 package
+% Thu Feb  2 11:38:04 2017
 \begin{table}[htb]
@@
 
-% latex table generated in R 3.3.0 by xtable 1.8-2 package
-% Thu Aug 18 15:23:11 2016
+% latex table generated in R 3.3.2 by xtable 1.8-2 package
+% Thu Feb  2 11:38:05 2017
 \begin{table}[htb]
@@
 
-\include{chapters/use_cases}
+% \include{chapters/use_cases}
 % knit_child("chapters/codetesting.tex")
-
-% TODOS:
-
-% Lots of characters within <> were lost in translation...
-% some things to note in the translation to LaTeX:
-% tildes in URLs broken -- turned into \textasciitilde{}
-% <a> etc. seem to sometimes be lost
-% section in some cases should be "chapter"
-
-% other stuff to fix / proof read
-% references
-% sections, e.g. Section 4, Pitfall 4)
-% URLs
-% `', ``''
-% reinsertion of a lot of the glyphs, etc.
-% fix ligatures
-% all glyphs need to be checked
-% all <>s need to be checked and/or readded
-
-\begin{comment}
-===
-Double quotation marks are generally used for distancing, in particular in the following situations:
-
-1.
-when a passage from another work is cited in the text (e.g.
-According to Takahashi (2009: 33), “quotatives were never used in subordinate clauses in Old Japanese”); but block quotations do not have quotation marks;
-2.
-when a technical term is mentioned that the author does not want to adopt, but wants to mention, e.g.
-This is sometimes called “pseudo-conservatism”, but I will not use this term here, as it could lead to confusion.
-
-Single quotation marks are used exclusively for linguistic meanings, as in the following: Latin habere ‘have’ is not cognate with Old English hafian ‘have’.
-
-===
-so, we should normally use double quotation in most of our cases :-).
-In general though, I would like us to try and remove double quotations as much as possible.
-Mostly is thus signals uncertainty on our part :-).
-
-\end{comment}
 

### 2017-02-02 use_cases/JIPA/manual.R
rms use cases (now in recipes repo)

@@ -1,80 +1,0 @@
-#' ---
-#' title: "Tokenization of JIPA transcriptions"
-#' author: "Michael Cysouw"
-#' date: "`r Sys.Date()`"
-#' ---
-
-library(qlcData)
-
-#' ### Check automatic tokenization
-# check it against hand-made tokenization
-# adhoc function to compare tokenization with hand-prepared results
-
-compare <- function(jipa, profile) {
-	text_in <- paste0("data/", jipa, "_input.txt")
-	text_out <- paste0("data/", jipa, "_output.txt")
-	
-	input <- scan(text_in, what = "character", sep = "\n")
-	output <- scan(text_out, what = "character", sep = "\n")
-	
-	tokenized <- tokenize(input
-						, profile = profile
-						, sep = " "
-						, sep.replace = "#"
-						, regex = TRUE
-						, normalize = "NFD"
-						)						
-
-	all.equal(output, tokenized$strings$tokenized)
-
-}
-
-# all true as valid IPA!
-profile <- "profiles/validIPA2005_graphemes.tsv"
-compare("Brazilian_Portuguese", profile)
-compare("Kabiye", profile)
-compare("Zurich_German", profile)
-compare("Vietnamese", profile)
-
-# not all conform to the strict IPA interpretation
-profile <- "profiles/strictIPA2005_graphemes.tsv"
-compare("Brazilian_Portuguese", profile)
-compare("Kabiye", profile)
-compare("Zurich_German", profile)
-compare("Vietnamese", profile)
-
-#' ### Tailored grapheme clusters
-# list all different tailored graphemes with frequencies
-# write the profiles with the tailored graphemes to the sandbox
-# disaply requires the knitr library
-require("knitr")	
-
-tailored <- function(jipa) {
-	input <- scan(paste0("data/", jipa, "_input.txt")
-				, what = "character"
-				, sep = "\n"
-				)
-	
-	tokenized <- tokenize(input
-						, profile = "profiles/validIPA2005_graphemes.tsv"
-						, sep = " "
-						, sep.replace = ""
-						, regex = TRUE
-						, normalize = "NFD"
-						)
-
-	profile <- write.profile(tokenized$strings$tokenized
-				, sep = " "
-				, file.out = paste0("sandbox/", jipa, "_profile.tsv")
-				)
-				
-		
-	invisible(knitr::kable(profile))
-}
-
-# write all to sandbox
-tailored("Brazilian_Portuguese")
[…]

### 2017-02-02 use_cases/dutch/code/correctwordlist.R

@@ -1,41 +1,0 @@
-#' ---
-#' title: "Preparation of dutch wordlist"
-#' author: "Michael Cysouw"
-#' date: "1 October 2015"
-#' ---
-
-library(qlcTokenize)
-
-# get examples from handmade profile
-examples <- read.delim("../data/1_dutch_IPA_profile.tsv", quote = "")$Examples
-examples <- unlist(strsplit(as.character(examples), ", "))
-
-# get examples from corpus
-nld10K <- read.delim("../sources/nld_mixed_2012_10K-words.txt", header = FALSE, quote = "")
-corpus_words <- as.character(nld10K[,2])
-
-# write draft profile of corpus
-write.profile(corpus_words
-			, file = "../sandbox/nld10K_draft_profile.tsv"
-			, editing =  TRUE
-			)
-
-# edit the profile by hand, removing strange characters
-# and adding some conditional constraints of character occurrence
-# this leads to the profile in "data/nld10K_corrected_profile.tsv"
-
-# remove all words with obviously non-dutch characters
-# just remove all words that include some strange characters
-tokens <- tokenize(corpus_words
-					, profile = "nld10K_corrected_profile.tsv"
-					, transliterate = "Replacement"
-					, sep = ""
-					, regex = TRUE
-					)
-errors <- as.numeric(rownames(tokens$errors))
-corpus_words <- tokens$strings[-errors,]$transliterated
-
-# combine the corpus with the hand-selected examples
-# write these to a file
-words <- sort(unique(c(examples, corpus_words)))
-write(words, file = "../sandbox/2_dutch_words.txt")

### 2017-02-02 use_cases/dutch/manual.R

@@ -1,32 +1,0 @@
-#' ---
-#' title: "IPA-transliteration of dutch words"
-#' author: "Michael Cysouw"
-#' date: "`r Sys.Date()`"
-#' ---
-
-# load libraries
-# knitr is needed for a nice display of the tables
-library(qlcData)
-require(knitr)
-
-# read the wordlist with dutch words
-words <- scan(file = "data/2_dutch_words.txt", what = "character")
-
-# use the hand-made orthography profile in `data/dutch_IPA_profile.tsv`
-dutch <- tokenize(words
-				, profile = "data/1_dutch_IPA_profile.tsv"
-				, transliterate = "IPA"
-				, file = "sandbox/3_dutch"
-				, regex = TRUE
-				, sep = ""
-				)
-
-# there are only a few errors coming from the tokenization
-knitr::kable(dutch$errors)
-
-# the IPA transliterations seems to be fine
-# there are some errors, but only a few
-# let's just check a random selection 
-# (note that this selection will change every time this code is executed)
-selection <- sample(1:nrow(dutch$strings), 50)
-knitr::kable(dutch$strings[selection, c("originals", "transliterated")])

### 2017-02-02 book/chapters/introduction.tex

--- 2017-01-20 book/chapters/introduction.tex
@@
 slightly more in-depth survey of the many techniques that are available in the
-standard. A good appreciation for the solutions that the Unicode Standard
-provides, also allows for a thorough understanding of possible pitfalls that one
-might encounter when using it (Chapter~\ref{unicode-pitfalls}). As an example of
-the current state-of-the-art, we discuss the International Phonetic Alphabet
-(IPA, Chapter~\ref{the-international-phonetic-alphabet}) and its rather
-problematic marriage with the Unicode Standard (Chapter
-\ref{ipa-meets-unicode}).
+standard. A good appreciation for the solutions that the Unicode Standard also
+allows for a thorough understanding of possible pitfalls that one might
+encounter when using it in general (Chapter~\ref{unicode-pitfalls}).
+Linguists are more often interested in using the Unicode Standard with the International Phonetic Alphabet (IPA).
+We first provide a history of the development of the IPA and early attempts to encode it electronically (Chapter~\ref{the-international-phonetic-alphabet}) before we discuss the rather problematic marriage of the International Phonetic Alphabet (IPA) with the Unicode Standard (Chapter\ref{ipa-meets-unicode}).
 
-The second part of the book describes our proposals for how to deal with the
-Unicode Standard in the daily practice of linguistics. First, we discuss the
+In the second part of the book (Chapters~\ref{orthography-profiles} and
+\ref{implementation}) we describe our proposals of how to deal with the Unicode
+Standard in the daily practice of (comparative) linguistics. First, we discuss the
 challenges of characterizing a writing system. To solve these problems, we
 propose the notions of orthography profiles, closely related to Unicode locale
-descriptions (Chapter~\ref{orthography-profiles}). We provide reference
+descriptions (Chapter~\ref{orthography-profiles}). Finally, we discuss practical
+issues with actual examples (Chapter~\ref{implementation}). We provide reference
 implementation of our proposals in R and in Python, available as open-source

### 2017-02-02 book/chapters/ipa_meets_unicode.tex
minor changes

@@
 linguist uses <ɑ> for a low back unrounded vowel.\footnote{One example is
-\citet[75]{Vidal2001a}, in which the author states: ``The definition of Pilagá
+\citet[75]{Vidal2001}, in which the author states: ``The definition of Pilagá
 /a/ as [+back] results from its behavior in certain phonological contexts. For
@@
 % ==========================
-\section{Pitfall: Future revisions to the IPA}
+\section{Pitfall: Revisions to the IPA}
 \label{ipa-revisions}

### 2017-02-02 book/chapters/preface.tex

--- 2015-12-15 book/chapters/preface.tex
@@
 
-% ==========================
-% \section{Preface}
-% \label{preface}
-% ==========================
+This text is meant as a practical guide for linguists, and programmers, who work with data in multilingual computational environments.
+We introduce the basic concepts needed to understand how writing systems and character encodings function, and how they work together.
 
-% Early phoneticians: Richard Lepsius, Isaac Pitman, Alexander J. Ellis, and Alexander Melville Bell
-% Later ones: Henry Sweet, L'abbé Jean-Pierre Rousselot, Eduard Sievers, Paul Passy, and Jens Otto Harry Jesperson
+The intersection of the Unicode Standard and the International Phonetic Alphabet is often not met without frustration by users.
+Nevertheless, the two standards have provided language researchers with a consistent computational architecture needed to process, publish and analyze data from many different languages.
+We bring to light common, but not always transparent, pitfalls that researchers face when working with Unicode and IPA.
 
-In the 1840s, several researchers and teachers began developing a notation to accurately transcribe articulatory gestures and their acoustic correlates.
-Their goal was not only transcription, they also wanted to use their notation for spelling reform (especially English), for teaching literacy and foreign languages, and in particular for developing writing systems for unwritten languages.
+Our research uses quantitative methods to compare languages and uncover and clarify their phylogenetic relations.
+However, the majority of lexical data available from the world's languages is in author- or document-specific orthographies.
+Having identified and overcome the pitfalls involved in making writing systems and character encodings syntactically and semantically interoperable (to the extent that they can be), we created a suite of open-source Python and R tools to work with languages using profiles that adequately describe their orthographic conventions.
+Using orthography profiles and these tools allows users to segment text, analyze it, identify errors, and to transform it into different written forms.
 
-At first, the challenge of devising a universal phonetic notation seemed straightforward: normalize the relationship between graphemes and sounds (<c> in `cake' -> /k/; <c> in `cell' -> /s/) to remove inconsistencies introduced by the historical divergence of written and spoken forms (e.g.\ in English and French orthographies), and keep adding new symbol-sound mappings until all sounds are accounted for.
+We welcome comments and corrections of this text, our source code, and the use case studies that are supplement to this book. Please use the issue tracker or email us directly.
 
-% These early transcription system pioneers found success with the first few languages that they encountered.
-Grapheme replacement and normalization of symbols across different orthographies seemed to work.
-But as they were confronted with more articulations and different notational devices, early phoneticians encountered a wealth of potentially important data that they had not expected.
+\bigbreak
+\noindent Steven Moran {\textless}bambooforest@gmail.com{\textgreater}
 
-This approach was successful for the first few languages investigated.
-But as the early phoneticians were confronted with more and more articulations, they encountered a wealth of data that they had not expected.
-The in-depth exploration of articulatory phonetics led to the discovery that the number of possible spoken sounds is infinite.
-\cite{Hockett1995} notes:
-
-\begin{quote}
-``But as they observed articulation more and more closely, and as they took more and more languages into consideration, the number of sounds they could discriminate and the number of symbols needed to denote them both increased seemingly without limit.
-In fact, as early as 1848, Ellis asserted that in principle the number of different sounds is endless.
-With the development of instrumental phonetics, largely by L'abbe Rousselt, it soon turned out that, given sufficiently accurate measuring devices, that is in fact the case.''
-\end{quote}
-
-% Hockett 1995: "And so the phoneticians began to develop notations that would show only the constant and distinctive features and let the transient and predictable ones be provided for by cover statements"
-
-% The phonemic principle emerged and the model for it was alphabetic writing.
-
-The first instrumental measurements in the 1870-80s showed unequivocally that spoken language is comprised of sounds waves; it is dynamic and not discrete.
-Given the fluid nature of the speech stream, sounds differ from one utterance to the next and the same sound may differ in different environments.
-However, many of these changes are predictable in the context of those environments and they need not necessarily form distinctive lexical contrasts.
-
-These findings left the early phoneticians in a conundrum given the aims of developing precise phonetic notion, but also practical alphabetic-based writing.
-It is in this vein that the phonetic and phonemic distinction was introduced very early into transcription.
-The etic vs emic principle took hold, in essence, even before terminology for the distinction was introduced and adopted.
-
-A transcription system that showed both contrastive and non-contrastive elements emerged, taking form notably in the International Phonetic Alphabet (which itself has been updated from time-to-time over the last nearly 120 years).
-The association between phonetic and phonemic would be maintained in the International Phonetic Alphabet and through the various instantiations of phonological theory, which developed from the work of the early phoneticians.
-
-The model for the principle of contrast was alphabetic writing, in particular shallow orthography, i.e.\ a one-to-one mapping between sounds and symbol.
-Traditional alphabetic orthography, however, did not scale to precise phonetic notation, so the former was sprinkled with diacritics.
-This dichotomy  causes many issues -- linguistic, theoretic, and recently technological -- which we bring to light in this book.
+\noindent Michael Cysouw {\textless}cysouw@mac.com{\textgreater}

### 2017-02-02 book/unicode.tex
nicely compiled version with updates merged in

--- 2017-02-02 book/unicode.tex
@@
 \hlstd{}\hlslc{\#\ make\ softlinks\ to\ the\ R\ executables\ in\ /usr/local/bin}\hspace*{\fill}\\
-\hlstd{}\hlkwc{ln\ }\hlstd{}\hlkwb{{-}is\ }\hlstd{}\hlkwd{\$pathT\ \$pathW\ }\hlstd{}\hlopt{/}\hlstd{usr}\hlopt{/}\hlstd{local}\hlopt{/}\hlstd{bin}\hspace*{\fill}
+\hlstd{}\hlslc{\#\ you\ will\ have\ to\ enter\ your\ user's\ password!}\hspace*{\fill}\\
+\hlstd{sudo\ }\hlkwc{ln\ }\hlstd{}\hlkwb{{-}is\ }\hlstd{}\hlkwd{\$pathT\ \$pathW\ }\hlstd{}\hlopt{/}\hlstd{usr}\hlopt{/}\hlstd{local}\hlopt{/}\hlstd{bin}\hspace*{\fill}
 \mbox{}
@@
 \begin{alltt}
-\hlstd{test} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"AABB"}\hlstd{,} \hlstr{"AАBВ"}\hlstd{)}
+\hlstd{(test} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"AABB"}\hlstd{,} \hlstr{"AАBВ"}\hlstd{))}
+\end{alltt}
+\begin{verbatim}
+## [1] "AABB" "AАBВ"
+\end{verbatim}
+\begin{alltt}
 \hlkwd{write.profile}\hlstd{(test)}
@@
 % latex table generated in R 3.3.2 by xtable 1.8-2 package
-% Thu Feb  2 11:38:04 2017
+% Thu Feb  2 15:27:40 2017
 \begin{table}[H]
@@
 % latex table generated in R 3.3.2 by xtable 1.8-2 package
-% Thu Feb  2 11:38:04 2017
+% Thu Feb  2 15:27:40 2017
 \begin{table}[htb]
@@
 % latex table generated in R 3.3.2 by xtable 1.8-2 package
-% Thu Feb  2 11:38:05 2017
+% Thu Feb  2 15:27:40 2017
 \begin{table}[htb]

### 2017-02-09 book/chapters/ipa_meets_unicode.tex
removes some todos

@@
 some ad-hoc decisions had to be taken given the vast variability of phonological
-segments described by linguists.
+segments described by linguists. The most recent version of the conventions 
+is published online by~\citet{MoranMcCloy2014}.\footnote{\url{http://phoible.github.io/conventions/}}
 
@@
 In the course of collecting a large sample of phoneme systems across the world's
-languages \citep{Moran2012}, we found that in order to preserve distinctions
-both within and across language descriptions, additions to the approved IPA
-glyph set were needed. Wherever possible these additions were drawn from the
-extIPA symbols for disordered
-speech.\footnote{\textbf{Add better reference here: }\url{https://www.internationalphoneticassociation.org/sites/default/files/extIPAChart2008.pdf}}
-This section describes our proposed additions to the IPA glyph set. These
+languages \citep{Moran2012}, \citet{Moran_etal2014} found that in order 
+to preserve distinctions both within and across language descriptions, 
+additions to the approved IPA glyph set were needed. Wherever possible 
+these additions were drawn from the extIPA symbols for disordered
+speech \citep{Duckworth_etal1990}.\footnote{\url{https://www.internationalphoneticassociation.org/sites/default/files/extIPAChart2008.pdf}}
+This section describes these proposed additions to the IPA glyph set. The
 additions are not part of the official IPA recommendations, so they should be 
-used with care.
+used with caution.
 
@@
       <\dia{032E}> \textsc{combining breve below} at \uni{032E}. 
-      \textbf{WHAT EXACTLY IS THIS?}
 \item \textsc{Coronal non-sibilant} \newline

### 2017-12-15 book/chapters/ipa_background.tex
fixes issue #5

@@
 the differences, see the 29 languages and their transcriptions in the
-\textit{Illustrations of the IPA}\citep[41--154]{IPA2007}.} Therefore, there are 
+\textit{Illustrations of the IPA} \citep[41--154]{IPA2007}.} Therefore, there are 
 different but equivalent transcriptions, or as noted by \citet[64]{Ladefoged1990a}, 

### 2017-12-15 book/chapters/preface.tex
deals with issue #8

@@
 
-This text is meant as a practical guide for linguists, and programmers, who work with data in multilingual computational environments.
-We introduce the basic concepts needed to understand how writing systems and character encodings function, and how they work together.
+This text is meant as a practical guide for linguists, and programmers, who work with data in multilingual computational environments.\footnote{\url{https://github.com/unicode-cookbook/cookbook}} We introduce the basic concepts needed to understand how writing systems and character encodings function, and how they work together.
 
@@
 
-Our research uses quantitative methods to compare languages and uncover and clarify their phylogenetic relations.
+In our daily working lives, we use quantitative methods to compare languages and uncover and clarify their phylogenetic relations.
 However, the majority of lexical data available from the world's languages is in author- or document-specific orthographies.
@@
 
-We welcome comments and corrections of this text, our source code, and the use case studies that are supplement to this book. Please use the issue tracker or email us directly.
+We welcome comments and corrections of this text, our source code, and the use case studies that are supplement to this book.\footnote{\url{https://github.com/unicode-cookbook/recipes}} Please use the issue tracker, email us directly, or find the book on PaperHive, where readers can leave questions and comments.\footnote{\url{https://paperhive.org/}}
 

### 2017-12-26 book/chapters/implementation.tex
closes issue #23
work towards issue 16 - wip

@@
 always as clean as we would like it to be. In sum, the two implementations
-should be considered as `proof of concept' and not as the final word on the
+should be considered as proof of concept and not as the final word on the
 practical application of the specifications above. In our own experience, the
@@
 \begin{lstlisting}[language=bash]
-  $ git clone https://github.com/bambooforest/segments.git
+  $ git clone https://github.com/cldf/segments
   $ cd segments
@@
 orthography processing, \texttt{write.profile} and \texttt{tokenize}. R includes
-help files with illustrative examples, and also a so-called `vignette' with
+help files with illustrative examples, and also a so-called vignette with
 explanations and examples.
@@
 interfaces to the R code supplied in the \texttt{qlcData} package: (i) bash
-executables and (ii) `shiny' webapps.
+executables and (ii) shiny webapps.
 
@@
 the string into Unicode graphemes according to grapheme clusters
-(`user-perceived characters', see Section~\ref{the-unicode-approach}). As it 
+(called user-perceived characters, see Section~\ref{the-unicode-approach}). As it 
 turns out, some of these graphemes are single codepoints, others are combinations
@@
 It is important to realize that different ordering of the rules does not have
-so-called `feeding' and `bleeding' effects as known from finite-state rewrite
-rules.\footnote{`Bleeding' is the effect that the application of a rule changes
-the string, so as to prevent a next rule to apply. `Feeding' is the opposite: a
+so-called feeding and bleeding effects as known from finite-state rewrite
+rules.\footnote{Bleeding is the effect that the application of a rule changes
+the string, so as to prevent a next rule to apply. Feeding is the opposite: a
 specific rule will only be applied becuase a previous rule changed the string
-already. The interaction of rules with such `feeding' and `bleeding' effects is
+already. The interaction of rules with such feeding and bleeding effects is
 extremely difficult to predict.} The graphemic parsing advocated here works
 crucially different from rewrite rules in that there is nothing being rewritten:
-each line in an orthography profile specifies a grapheme to be `captured' in the 
+each line in an orthography profile specifies a grapheme to be captured in the 
 string. All lines in the profile are processed in a specifed order (as determined
 by the option \texttt{ordering}). At the processing of a specific line, all 
-matching graphemes in the data are marked as `captured', but not changed. 
+matching graphemes in the data are marked as captured, but not changed. 
 Captured parts cannot be captured again, but they can still be used to match 
@@
 captured graphemes are separated (and possibly transliterated). In this way the 
-result of the application of the `rules' is rather easy to predict.
+result of the application of the rules' is rather easy to predict.
 

### 2017-12-26 book/chapters/introduction.tex
work towards issue 16 - wip

@@
 development of encoding of telegraphic communication focus of cryptography,
-e.g.~\citet{Singh1999}, and forego the rather interesting story of ``open'',
+e.g.~\citet{Singh1999}, and forego the rather interesting story of open,
 i.e.~non-cryptographic, encoding that is related here.} This history turns out
@@
 still in limited use, for example utilizing flags or flashing lights. The
-``innovation'' of those semaphoric systems was that all characters of the
+innovation of those semaphoric systems was that all characters of the
 written language were replaced one-to-one by visual signals. Since then, all
@@
 for all written symbols, including punctuation and digits. As a solution, the
-Baudot code uses a so-called ``shift'' character, which signifies that from
+Baudot code uses a so-called shift character, which signifies that from
 that point onwards---until shifted back---a different encoding is used, allowing
@@
 refer to the way a particular language is written. In this sense the term refers
-to the writing system of a particular language, as, for example, in \emph{`the
-Serbian writing system uses two scripts: Latin and Cyrillic.'} Second, the term
+to the writing system of a particular language, as, for example, in \emph{the
+Serbian writing system uses two scripts: Latin and Cyrillic}. Second, the term
 writing system may also refer to a type of symbolic system as used among the
 world's languages to represent the language, as, for example, in
-\emph{`alphabetic writing system.'} In this latter sense the term refers to how
+\emph{alphabetic writing system}. In this latter sense the term refers to how
 scripts have been classified according to the way that they encode language, as
-in, for example, \emph{`the Latin and Cyrillic scripts are both alphabetic
-writing systems.'} To avoid confusion, this second notion of writing system
+in, for example, \emph{the Latin and Cyrillic scripts are both alphabetic
+writing systems}. To avoid confusion, this second notion of writing system
 would more aptly have been called \textsc{script system}.

### 2017-12-26 book/chapters/ipa_meets_unicode.tex
work towards issue 16 - wip

@@
 strict encoding guidelines. Our main goal in this chapter is to present the
-encoding issues and propose recommendations for a ``strict'' IPA encoding for
+encoding issues and propose recommendations for a strict IPA encoding for
 situations in which cross-resource consistency is crucial.
@@
 are also included, see~\cite[Appendix~C]{Moran2012}. The 159 characters from 12 
-code blocks shown here are the characters proposed for ``strict'' IPA encoding, 
+code blocks shown here are the characters proposed for strict IPA encoding, 
 as discussed in Section~\ref{ipa-recommendations}.}
@@
         postalveolar click (not retroflex). This naming is probably best seen as
-        an error in the Unicode Standard. For the ``real'' retroflex click, see 
+        an error in the Unicode Standard. For the real retroflex click, see 
         Section~\ref{ipa-additions}.}
@@
            \textsc{modifier letter apostrophe} at \uni{02BC}, respectively. 
-   \item There are two different ``dashed''-l characters in IPA, namely the <ɫ> 	      \textsc{latin small letter l with middle tilde} at \uni{026B} and the <ɬ>	      \textsc{latin small letter l with belt} at \uni{026C}.
+   \item There are two different dashed-l characters in IPA, namely the <ɫ> 	      \textsc{latin small letter l with middle tilde} at \uni{026B} and the <ɬ>	      \textsc{latin small letter l with belt} at \uni{026C}.
 These of course
@@
 not acceptable when the goal is interoperability between resources written in
-IPA.\@ We consider it crucial to distinguish between ``valid'' IPA, for which it
+IPA.\@ We consider it crucial to distinguish between valid IPA, for which it
 is sufficient that any phonetically-trained reader is able to understand the
-transcription, and ``strict'' IPA, which should be standardized on a single
+transcription, and strict IPA, which should be standardized on a single
 unique encoding for each sound, so search will work across resources. We are
@@
 principle (see Section~\ref{pitfall-characters-are-not-graphemes}) is that fixed
-sequences in a writing system have to be specified as ``tailored'' grapheme
+sequences in a writing system have to be specified as tailored grapheme
 clusters. Only in case the sequence is not a cluster, then this has to be
@@
 \uni{0334}. The main reason some phoneticians like to use this in IPA is to mark
-the ``dark'' <l> in English codas, using the character <ɫ> \textsc{latin small
+the dark <l> in English codas, using the character <ɫ> \textsc{latin small
 letter l with middle tilde} at \uni{026B}. This character is not canonically
@@
       types of frictionalized sounds: First, click consonants where the release
-      of the anterior closure involves an ingressive ``sucking'' sound similar
-      to a fricative, for example <kǃ͓ʰ>; second, ``frictionalized'' vowels
+      of the anterior closure involves an ingressive sucking sound similar
+      to a fricative, for example <kǃ͓ʰ>; second, frictionalized vowels
       (sounds that are phonologically vocalic, but with sufficiently close
@@
       <\dia{02C0}> \textsc{modifier letter glottal stop} at \uni{02C0}, unless
-      it is clear that either ``ejective'' or ``creaky voicing'' are the
+      it is clear that either ejective or creaky voicing are the
       intended sounds (in which cases the standard IPA diacritics should be

### 2017-12-26 book/chapters/pitfalls.tex
work towards issue 16 - wip

@@
 Unicode Standard behaves differently as expected, we think it is important not
-to dismiss Unicode as ``wrong'' or ``deficient'', because our
+to dismiss Unicode as wrong or deficient, because our
 experience is that in almost all cases the behavior of the Unicode Standard has
@@
 Section~\ref{pitfall-different-notions-of-diacritics} for a discussion of the
-notion of diacritic). The Unicode Standard calls such combination ``grapheme
-clusters''. Complex graphemes consisting of multiple base characters,
-like <sch>, are called ``tailored grapheme clusters'' in Unicode parlance (see
+notion of diacritic). The Unicode Standard calls such combination \textit{grapheme
+clusters||. Complex graphemes consisting of multiple base characters,
+like <sch>, are called \textit{tailored grapheme clusters} in Unicode parlance (see
 Section~\ref{the-unicode-approach}).
@@
 Inspecting the Unicode Standard, there appear to be special Unicode characters
-to ``glue'' together characters into larger tailored grapheme clusters,
+to glue together characters into larger tailored grapheme clusters,
 specifically the \textsc{zero width joiner} at \uni{200D} and the
@@
 not join characters, but explicitly separate them. The zero-width joiner (ZWJ)
-can be used to solve special problems related to ordering (called ``collation''
+can be used to solve special problems related to ordering (called \textit{collation}
 in Unicode parlance). The combining grapheme joiner (CGJ) can be used to
@@
 Section~\ref{pitfall-ipa-homoglyphs}). In a sense, these names can be seen as
-``errors.'' However, it is probably better to realize that such names are just
+errors. However, it is probably better to realize that such names are just
 convenience labels that are not going to be changed. Just like the block names
@@
 
-The actual intended ``meaning'' of a Unicode codepoint is a combination of the
+The actual intended meaning of a Unicode codepoint is a combination of the
 name, the block and the character properties (see
@@
 characters with attached extras are not decomposed. However, in the following
-cases the ``extras'' actually exist as combining diacritics, so there is also 
+cases the extras actually exist as combining diacritics, so there is also 
 the possibility to construct a character by using a base character with those 
@@
 
-  \item First, there are the combining characters designated as ``combining
-        overlay'' in the Unicode Standard, like <\dia{0334}>
+  \item First, there are the combining characters designated as \textit{combining
+        overlay} in the Unicode Standard, like <\dia{0334}>
         \textsc{combining tilde overlay} at \uni{0334} or <\dia{0335}>
@@
   \item Second, the same situation also occurs with combining characters
-        designated as ``combining hook'', like 
+        designated as \textit{combining hook}, like 
         <{\fontspec{CharisSIL}{\large ◌}}\symbol{"0321}> \textsc{combining
@@
          use orthography profiles (see Chapter~\ref{orthography-profiles}).
-   \item To deal with ``tailored'' grapheme clusters
+   \item To deal with tailored grapheme clusters
          (Section~\ref{pitfall-characters-are-not-graphemes}), use Unicode Locale 

### 2017-12-26 book/unicode.tex
removes some todos
work towards issue 16 - wip

--- 2017-02-02 book/unicode.tex
@@
 		biblatex
-		]{LSP/langsci}\usepackage[]{graphicx}\usepackage[]{color}
-%% maxwidth is the original width if it is less than linewidth
-%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
-\makeatletter
-\def\maxwidth{ %
-  \ifdim\Gin@nat@width>\linewidth
-    \linewidth
-  \else
-    \Gin@nat@width
-  \fi
-}
-\makeatother
-
-\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
-\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
-\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
-\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
-\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
-\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
-\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
-\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
-\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
-\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
-\let\hlipl\hlkwb
-
-\usepackage{framed}
-\makeatletter
-\newenvironment{kframe}{%
- \def\at@end@of@kframe{}%
- \ifinner\ifhmode%
-  \def\at@end@of@kframe{\end{minipage}}%
-  \begin{minipage}{\columnwidth}%
- \fi\fi%
- \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
- \colorbox{shadecolor}{##1}\hskip-\fboxsep
-     % There is no \\@totalrightmargin, so:
-     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
- \MakeFramed {\advance\hsize-\width
-   \@totalleftmargin\z@ \linewidth\hsize
-   \@setminipage}}%
- {\par\unskip\endMakeFramed%
- \at@end@of@kframe}
-\makeatother
-
-\definecolor{shadecolor}{rgb}{.97, .97, .97}
-\definecolor{messagecolor}{rgb}{0, 0, 0}
-\definecolor{warningcolor}{rgb}{1, 0, 1}
-\definecolor{errorcolor}{rgb}{1, 0, 0}
-\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
-
-\usepackage{alltt}
+		]{LSP/langsci}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+
 \begin{document}
@@
 % spacing of code-chunks is set with the "knitrout" environment in localcommands.tex
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%                                              %%%
-%%%             Chapters                         %%%
-%%%                                              %%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\include{chapters/introduction}
-\include{chapters/pitfalls}
-\include{chapters/ipa_background}
-\include{chapters/ipa_meets_unicode}
-\include{chapters/orthography_profiles}
-
-\chapter{Implementation}
-\label{implementation}
-
-To illustrate the practical application of orthography profiles as defined in
-the previous Chapter \ref{orthography-profiles}, we have implemented two 
-versions of the specifications presented there: one in Python and one
-in R.
-
-These two libraries have rather different implementation histories,
-and we are not 100\% sure that they will in all situations give the same
-results. We do however provide a test suite for each implementation. 
-Users should refer to the tests cases and to the documentation in each release 
-for specifics about each implementation. Note also that function names differ 
-in the two implementations.
-
-Note also that the performance with larger datasets differs, and the code is not
-always as clean as we would like it to be. In sum, the two implementations
-should be considered as `proof of concept' and not as the final word on the
-practical application of the specifications above. In our own experience, the
-current implementations are sufficiently fast and stable to be useful for
-academic practice (e.g. checking data consistency, or analyzing and
-transliterating small to medium sized data sets), but they should probably
-better not be used for full-scale industry applications.
-
-This chapter will introduce the implementations, and give practical step-by-step 
-guidelines for installing them and using them. Various simple and sometimes 
-somewhat abstract examples will be discussed to show the different options 
-available, and to illustrate the intended usage of orthography profiles.
-
-This chapter is split into two sections, one for the Python implementation, the 
-other for R. Each section has a number of subsections that ...
-
-In addition to the material presented here to get users started, we maintain 
-several case studies that use each OP implementation in action. These use cases 
-are an the accompanying Github repository available with this book at: 
-\url{https://github.com/unicode-cookbook/recipes/}.
-
-% TODO: point to the case studies
-% In the next chapter we will then discuss a few full-scale examples to illustrate the application and the usefulness of orthography profiles.
-
-\section{Python}
-\label{python-implementations}
-
-The Python package is available both as a command line interface (CLI) 
-and as an appplication programming interface (API).
-
-\subsection*{Installation}
-
-To install the Python package from the Python Package Index (PyPI) run:
-
-\begin{lstlisting}[language=bash]
-  $ pip install segments
-\end{lstlisting}
-
-\noindent on the command line. This will give you access to both 
-the CLI and programmatic functionality in Python scripts, when 
-you import the \texttt{segments} library.
-
-You can also install the \texttt{segments} 
-package with from the Github repository:
-
-\begin{lstlisting}[language=bash]
-  $ git clone https://github.com/bambooforest/segments.git
-  $ cd segments
-  $ python setup.py develop
-\end{lstlisting}
-
-\subsection*{Command line interface}
-
-From the command line, access \texttt{segments} and its 
-various arguments. For help, run:
-
-\begin{lstlisting}[language=bash]
-  $ segments -h
-\end{lstlisting}
-
-\noindent Here is an example of how to create and use an orthography 
-profile for segmentation. Create a text file:
-
-\begin{lstlisting}[language=bash]
-  $ more text.txt
-  aäaaöaaüaa
-\end{lstlisting}
-
-\noindent Now look at the profile:
-
-\begin{lstlisting}[language=bash,texcl=true]
-  $ cat text.txt | segments profile
-  Grapheme        frequency       mapping
-  a       7       a
-  ä       1       ä
-  ü       1       ü
-  ö       1       ö
-\end{lstlisting}
-
-\noindent Write the profile to a file:
-
-\begin{lstlisting}[language=bash]
-  $ cat text.txt | segments profile > profile.prf
-\end{lstlisting}
-
-\noindent Edit the profile:
-
-\begin{lstlisting}[language=bash]
-  $ more profile.prf
-  Grapheme        frequency       mapping
-  aa      0       x
-  a       7       a
-  ä       1       ä
-  ü       1       ü
-  ö       1       ö
-\end{lstlisting}
-
-\noindent Now tokenize the text without profile:
-
-\begin{lstlisting}[language=bash]
-  $ cat text.txt | segments tokenize
-  a ä a a ö a a ü a a	
-\end{lstlisting}
-
-\noindent And with profile:
-\begin{lstlisting}[language=bash]
-  $ cat text.txt | segments --profile=profile.prf tokenize
-  a ä aa ö aa ü aa
-
-  $ cat text.txt | segments --mapping=mapping --profile=profile.prf tokenize
-  a ä x ö x ü x
-\end{lstlisting}
-
-\subsection*{Application programming interface}
-The \texttt{segments} API can be accesses by importing 
-the package into Python a script. Here is an example of 
-how to import the libraries, create a tokenizer object, 
-tokenize a string and create a profile.
-
-\begin{lstlisting}[language=python]
->>> from __future__ import unicode_literals, print_function
->>> from segments import Profile, Tokenizer
->>> t = Tokenizer()
->>> t('abcd')
-'a b c d'
->>> prf = Profile({'Grapheme': 'ab', 'mapping': 'x'}, \
-{'Grapheme': 'cd', 'mapping': 'y'})
->>> print(prf)
-mapping Grapheme
-ab  x
-cd  y
->>> t = Tokenizer(profile=prf)
->>> t('abcd')
-'ab cd'
->>> t('abcd', column='mapping')
-'x y'
-\end{lstlisting}
-
-\section{R}
-\label{r-implementation}
-
-% \subsection*{Installing the R implementation}
-\subsection*{Installation}
-
-The R implementation is available in the package \texttt{qlcData}, which is 
-directly available from the central R repository CRAN (Comprehensive R Archive 
-Network). The R software environment itself has to be downloaded from its 
-website.\footnote{\url{https://www.r-project.org}} After starting the included 
-R program, the \texttt{qlcData} package for dealing with orthography profiles can be 
-simply installed as follows:
-
-\begin{knitrout}\footnotesize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# download and install the qlcData software}
-\hlkwd{install.packages}\hlstd{(}\hlstr{"qlcData"}\hlstd{)}
-\hlcom{# load the software, so it can be used}
-\hlkwd{library}\hlstd{(qlcData)}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
-
-The version available through CRAN is the latest stable version.
-To obtain the most recent bug-fixes and experimental additions, please use the
-development version, which is available on
-GitHub.\footnote{\url{http://github.com/cysouw/qlcData}} This development
-version can be easily installed using the github-install helper software from the
-\texttt{devtools} package.
-
-\begin{knitrout}\footnotesize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# download and install helper software}
-\hlkwd{install.packages}\hlstd{(}\hlstr{"devtools"}\hlstd{)}
-\hlcom{# install the qlcData package from GitHub}
-\hlstd{devtools}\hlopt{::}\hlkwd{install_github}\hlstd{(}\hlstr{"cysouw/qlcData"}\hlstd{,} \hlkwc{build_vignettes} \hlstd{=} \hlnum{TRUE}\hlstd{)}
-\hlcom{# load the software, so it can be used}
-\hlkwd{library}\hlstd{(qlcData)}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
-
-Inside the \texttt{qlcData} package, there are two functions for
-orthography processing, \texttt{write.profile} and \texttt{tokenize}. R includes
-help files with illustrative examples, and also a so-called `vignette' with
-explanations and examples.
-
-\begin{knitrout}\footnotesize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# view help files}
-\hlkwd{help}\hlstd{(write.profile)}
-\hlkwd{help}\hlstd{(tokenize)}
-\hlcom{# view vignette with explanation and examples}
-\hlkwd{vignette}\hlstd{(}\hlstr{"orthography_processing"}\hlstd{)}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
-
-Basically, the idea is to use \texttt{write.profile} to produce a
-basic orthography profile from some data and then \texttt{tokenize} to apply the
-(possibly edited) profile on some data, as exemplified in the next section. This
-can of course be performed though R, but additionally there are two more
-interfaces to the R code supplied in the \texttt{qlcData} package: (i) bash
-executables and (ii) `shiny' webapps.
-
-The bash executables are little files providing an interface to the R code that
-can be used in a shell on a UNIX-alike machine. The location of these
-executables is a bit hidden away by the install procedure of R packages. The
-location can be found by the following command in R. These executables can be 
-used from here, or they can be linked and/or copied to any location as wanted.
-
-\begin{knitrout}\footnotesize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# show the path to the bash executables}
-\hlkwd{file.path}\hlstd{(}\hlkwd{find.package}\hlstd{(}\hlstr{"qlcData"}\hlstd{),} \hlstr{"exec"}\hlstd{)}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
-
-For example, a good way to use the executables in a terminal is to
-make softlinks (using \texttt{ln}) from the executables to a directory in your
-PATH, e.g. to \texttt{/usr/local/bin/}. The two executables are named
-\texttt{tokenize} and \texttt{writeprofile}, and the links can be made directly 
-by using Rscript to get the paths to the executables within the terminal.
-
-\begin{knitrout}\scriptsize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\noindent
-\ttfamily
-\hlstd{}\hlslc{\#\ get\ the\ paths\ to\ the\ R\ executables\ in\ bash}\hspace*{\fill}\\
-\hlstd{pathT}\hlopt{=}\hlstd{}\hlstr{`Rscript\ {-}e\ 'cat(file.path(find.package("qlcData"),\ "exec",\ "tokenize"))'`}\hlstd{}\hspace*{\fill}\\
-\hlstd{pathW}\hlopt{=}\hlstd{}\hlstr{`Rscript\ {-}e\ 'cat(file.path(find.package("qlcData"),\ "exec",\ "writeprofile"))'`}\hlstd{}\hspace*{\fill}\\
-\hlstd{}\hspace*{\fill}\\
-\hlstd{}\hlslc{\#\ make\ softlinks\ to\ the\ R\ executables\ in\ /usr/local/bin}\hspace*{\fill}\\
-\hlstd{}\hlkwc{ln\ }\hlstd{}\hlkwb{{-}is\ }\hlstd{}\hlkwd{\$pathT\ \$pathW\ }\hlstd{}\hlopt{/}\hlstd{usr}\hlopt{/}\hlstd{local}\hlopt{/}\hlstd{bin}\hspace*{\fill}
-\mbox{}
-\normalfont
-\end{kframe}
-\end{knitrout}
-
-After inserting this softlink it should be possible to access the
-\texttt{tokenize} function from the shell. Try \texttt{tokenize --help} to test
-the functionality.
-
-\begin{knitrout}\scriptsize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\noindent
-\ttfamily
-\hlstd{tokenize\ }\hlkwb{{-}{-}help}\hlstd{}\hspace*{\fill}
-\mbox{}
-\normalfont
-
-\begin{verbatim}
-## Error in running command bash
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
-
-To make the functionality even more accessible, we have prepared webapps with 
-the \texttt{shiny} framework for the R functions. These webapps are available 
-online at \url{TODO}. The webapps are also included inside the \texttt{qlcData} 
-package and can be started with the following helper function:
-
-\begin{knitrout}\footnotesize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlkwd{launch_shiny}\hlstd{(}\hlstr{"tokenize"}\hlstd{)}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
-
-% TO DO
-
-\subsection*{Profiles and error reporting}
-\label{error-reporting}
-
-The first example of how to use these functions concerns finding errors in the
-encoding of texts. In the following example, it looks as if we have two
-identical strings, \texttt{AABB}. However, this is just an surface-impression
-delivered by the current font, which renders Latin and Cyrillic capitals
-identically. We can identify this problem when we produce an orthography profile
-from the strings. Using here the R implementation of orthography profiles, we
-first assign the two strings to a variable \texttt{test}, and then produce an
-orthography profile with the function \texttt{write.profile}. As it turns out,
-some of the letters are Cyrillic.
-
-\begin{knitrout}\footnotesize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlstd{test} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"AABB"}\hlstd{,} \hlstr{"AАBВ"}\hlstd{)}
-\hlkwd{write.profile}\hlstd{(test)}
-\end{alltt}
-\begin{verbatim}
-##   Grapheme Frequency Codepoint                UnicodeName
-## 1        A         3    U+0041     LATIN CAPITAL LETTER A
-## 2        B         3    U+0042     LATIN CAPITAL LETTER B
-## 3        А         1    U+0410  CYRILLIC CAPITAL LETTER A
-## 4        В         1    U+0412 CYRILLIC CAPITAL LETTER VE
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
-
-The function of error-message reporting can also nicely be illustrated
-with this example. Suppose we made an orthography profile with just the two
-Latin letters <A> and <B> as possible graphemes, then this profile would not be
-sufficient to tokenize the strings. There are graphemes in the data that are not
-in the profile, so the tokenization produces an error, which can be used to fix
-the encoding (or the profile). In the example below, we can see that the
-Cyrillic encoding is found in the second string of the \texttt{test} input.
-
-\begin{knitrout}\footnotesize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlstd{test} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"AABB"}\hlstd{,} \hlstr{"AАBВ"}\hlstd{)}
-\hlkwd{tokenize}\hlstd{(test,} \hlkwc{profile} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{"A"}\hlstd{,} \hlstr{"B"}\hlstd{))}
-\end{alltt}
-
-{\ttfamily\noindent\color{warningcolor}{\#\# Warning in tokenize(test, profile = c("{}A"{}, "{}B"{})): \\\#\# There were unknown characters found in the input data.\\\#\# Check output\$errors for a table with all problematic strings.}}\begin{verbatim}
-## $strings
-##   originals tokenized
-## 1      AABB   A A B B
-## 2      AАBВ   A ⁇ B ⁇
-## 
-## $profile
-##   matched_rules Grapheme
-## 1             3        B
-## 2             3        A
-## 
-## $errors
-##   originals  errors
-## 2      AАBВ A ⁇ B ⁇
-## 
-## $missing
-##   Grapheme Frequency Codepoint                UnicodeName
-## 1        А         1    U+0410  CYRILLIC CAPITAL LETTER A
-## 2        В         1    U+0412 CYRILLIC CAPITAL LETTER VE
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
-
-\subsection*{Different ways to write a profile}
-\label{write-profile}
-
-The function \texttt{write.profile} can be used to prepare a skeleton for an
-orthography profile from some data. The preparation of an orthography profile
-from some data might sound like a trivial problem, but actually there are
-various different ways in which strings can be separated into graphemes by
-\texttt{write.profile}. Consider the following string of characters called
-\texttt{example} below. The default settings of \texttt{write.profile} separates
-the string into Unicode graphemes according to grapheme clusters
-(`user-perceived characters', see Section~\ref{the-unicode-approach}). As it 
-turns out, some of these graphemes are single codepoints, others are combinations
-of two codepoints (see Section~\ref{pitfall-characters-are-not-glyphs}).
-
-\begin{knitrout}\footnotesize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlstd{example} \hlkwb{<-} \hlstr{"ÙÚÛÙÚÛ"}
-\hlstd{profile_1} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example)}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
-
-% latex table generated in R 3.3.2 by xtable 1.8-2 package
-% Thu Feb  2 11:38:04 2017
-\begin{table}[H]
-\centering
-\begingroup\scriptsize
-\begin{tabular}{llll}
-  \toprule
-Gr. & Freq. & Codepoint & Unicode Name \\ 
-  \midrule
-Ú & 1 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
-  Ú & 1 & U+0055, U+0301 & LATIN CAPITAL LETTER U, COMBINING ACUTE ACCENT \\ 
-  Ù & 1 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
-  Ù & 1 & U+0055, U+0300 & LATIN CAPITAL LETTER U, COMBINING GRAVE ACCENT \\ 
-  Û & 1 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
-  Û & 1 & U+0055, U+0302 & LATIN CAPITAL LETTER U, COMBINING CIRCUMFLEX ACCENT \\ 
-   \bottomrule
-\end{tabular}
-\endgroup
-\caption{Profile 1 (default settings, splitting grapheme clusters)} 
-\label{tab:profile1}
-\end{table}
-
-By specifying the splitting separator as the empty string
-\texttt{sep~=~""}, it is possible to split the string into Unicode codepoints,
-thus separating the combining diacritics. The idea behind this option
-\texttt{sep} is that separating by a character allows for user-determined
-separation. The most extreme choice here is the empty string \texttt{sep~=~""},
-which is interpreted as separation everywhere. The other extreme is the default
-setting \texttt{sep~=~NULL}, which means that the separation is not
-user-defined, but relegated to the Unicode grapheme definitions.
-
-\begin{knitrout}\footnotesize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlstd{profile_2} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{)}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
-
-% latex table generated in R 3.3.2 by xtable 1.8-2 package
-% Thu Feb  2 11:38:04 2017
-\begin{table}[H]
-\centering
-\begingroup\scriptsize
-\begin{tabular}{llll}
-  \toprule
-Grapheme & Frequency & Codepoint & Unicode Name \\ 
-  \midrule
-́ & 1 & U+0301 & COMBINING ACUTE ACCENT \\ 
-  ̀ & 1 & U+0300 & COMBINING GRAVE ACCENT \\ 
-  ̂ & 1 & U+0302 & COMBINING CIRCUMFLEX ACCENT \\ 
-  U & 3 & U+0055 & LATIN CAPITAL LETTER U \\ 
-  Ú & 1 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
-  Ù & 1 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
-  Û & 1 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
-   \bottomrule
-\end{tabular}
-\endgroup
-\caption{Profile 2 (splitting by codepoints)} 
-\label{tab:profile2}
-\end{table}
-
-Some characters look identical, although they are encoded differently.
-Unicode offers different ways of normalization (see
-Section~\ref{pitfall-canonical-equivalence}), which can be invoked here as well
-using the option \texttt{normalize}. NFC-normalization turns everything into the
-precomposed characters, while NFD-normalization separates everything into base
-characters with combining diacritics. Splitting by codepoints (i.e. \texttt{sep~=~""}) shows the results of these two normalizations.
-
-\begin{knitrout}\footnotesize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# after NFC normalization unicode codepoints have changed}
-\hlstd{profile_3} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example,} \hlkwc{normalize} \hlstd{=} \hlstr{"NFC"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{)}
-\hlcom{# NFD normalization gives yet another structure of the codepoints}
-\hlstd{profile_4} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example,} \hlkwc{normalize} \hlstd{=} \hlstr{"NFD"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{)}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
-
-% latex table generated in R 3.3.2 by xtable 1.8-2 package
-% Thu Feb  2 11:38:04 2017
-\begin{table}[H]
-\centering
-\begingroup\scriptsize
-\begin{tabular}{llll}
-  \toprule
-Grapheme & Frequency & Codepoint & Unicode Name \\ 
-  \midrule
-Ú & 2 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
-  Ù & 2 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
-  Û & 2 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
-   \bottomrule
-\end{tabular}
-\endgroup
-\caption{Profile 3 (splitting by NFC codepoints)} 
-\label{tab:profile3}
-\end{table}
-
-% latex table generated in R 3.3.2 by xtable 1.8-2 package
-% Thu Feb  2 11:38:04 2017
-\begin{table}[H]
-\centering
-\begingroup\scriptsize
-\begin{tabular}{llll}
-  \toprule
-Grapheme & Frequency & Codepoint & Unicode Name \\ 
-  \midrule
-́ & 2 & U+0301 & COMBINING ACUTE ACCENT \\ 
-  ̀ & 2 & U+0300 & COMBINING GRAVE ACCENT \\ 
-  ̂ & 2 & U+0302 & COMBINING CIRCUMFLEX ACCENT \\ 
-  U & 6 & U+0055 & LATIN CAPITAL LETTER U \\ 
-   \bottomrule
-\end{tabular}
-\endgroup
-\caption{Profile 4 (splitting by NFD codepoints)} 
-\label{tab:profile4}
-\end{table}
-
-It is important to realize that for Unicode grapheme definitions, NFC
-and NFD normalization are equivalent. This can be shown by normalizing the
-example in either NFC or NFD, but using default separation in
-\texttt{write.profile}. To be precise, default separation means setting
-\texttt{sep~=~NULL}, but that has not be added explicitly.
-
-\begin{knitrout}\footnotesize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# note that NFC and NFD normalization are identical}
-\hlcom{# for unicode grapheme definitions}
-\hlstd{profile_5} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example,} \hlkwc{normalize} \hlstd{=} \hlstr{"NFD"}\hlstd{)}
-\hlstd{profile_6} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example,} \hlkwc{normalize} \hlstd{=} \hlstr{"NFC"}\hlstd{)}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
-
-% latex table generated in R 3.3.2 by xtable 1.8-2 package
-% Thu Feb  2 11:38:04 2017
-\begin{table}[H]
-\centering
-\begingroup\scriptsize
-\begin{tabular}{llll}
-  \toprule
-Gr. & Freq. & Codepoint & Unicode Name \\ 
-  \midrule
-Ú & 2 & U+0055, U+0301 & LATIN CAPITAL LETTER U, COMBINING ACUTE ACCENT \\ 
-  Ù & 2 & U+0055, U+0300 & LATIN CAPITAL LETTER U, COMBINING GRAVE ACCENT \\ 
-  Û & 2 & U+0055, U+0302 & LATIN CAPITAL LETTER U, COMBINING CIRCUMFLEX ACCENT \\ 
-   \bottomrule
-\end{tabular}
-\endgroup
-\caption{Profile 5 (splitting by graphemes after NFD)} 
-\label{tab:profile5}
-\end{table}
-
-% latex table generated in R 3.3.2 by xtable 1.8-2 package
-% Thu Feb  2 11:38:04 2017
-\begin{table}[H]
-\centering
-\begingroup\scriptsize
-\begin{tabular}{llll}
-  \toprule
-Gr. & Freq. & Codepoint & Unicode Name \\ 
-  \midrule
-Ú & 2 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
-  Ù & 2 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
-  Û & 2 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
-   \bottomrule
-\end{tabular}
-\endgroup
-\caption{Profile 6 (splitting by graphemes after NFC)} 
-\label{tab:profile6}
-\end{table}
-
-Note that these different profiles can also be produced using the bash
-executable \texttt{writeprofile} (see \ref{r-implemenation} for notes
-on how to install the bash executable). Exactly this example is also included in
-the help file of the executable.
-
-\begin{knitrout}\scriptsize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\noindent
-\ttfamily
-\hlstd{writeprofile\ }\hlkwb{{-}{-}help}\hlstd{}\hspace*{\fill}
-\mbox{}
-\normalfont
-
-\begin{verbatim}
-## Error in running command bash
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
-
-\subsection*{Using an orthography profile skeleton}
-\label{profile-skeleton}
-
-A common workflow to use these functions is to first make a skeleton for an
-orthography profile and then edit this profile by hand. For example, Table
-\ref{tab:profile_skeleton1} shows the profile skeleton after a few graphemes have
-been added to the file. Note that in this example, the profile is written to the
-desktop, and this file has to be edited manually. We simply add a few
-multigraphs to the column \texttt{Grapheme} and leave the other columns empty.
-these new graphemes are then included in the graphemic parsing.
-
-\begin{knitrout}\footnotesize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# a few words to be graphemically parsed}
-\hlstd{example} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"mishmash"}\hlstd{,} \hlstr{"mishap"}\hlstd{,} \hlstr{"mischief"}\hlstd{,} \hlstr{"scheme"}\hlstd{)}
-\hlcom{# write a profile skeleton to a file}
-\hlkwd{write.profile}\hlstd{(example,} \hlkwc{file} \hlstd{=} \hlstr{"~/Desktop/profile_skeleton.txt"}\hlstd{)}
-\hlcom{# edit the profile, and then use the edited profile to tokenize}
-\hlkwd{tokenize}\hlstd{(example,} \hlkwc{profile} \hlstd{=} \hlstr{"~/Desktop/profile_skeleton.txt"}\hlstd{)}\hlopt{$}\hlstd{strings}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
-
-\begin{knitrout}\footnotesize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
-##   originals    tokenized
-## 1   shampoo  sh a m p oo
-## 2    mishap   m i sh a p
-## 3  mischief m i sch ie f
-## 4    scheme    sch e m e
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
-
-% latex table generated in R 3.3.2 by xtable 1.8-2 package
-% Thu Feb  2 11:38:04 2017
-\begin{table}[htb]
-\centering
-\begingroup\scriptsize
-\begin{tabular}{llll}
-  \toprule
-Grapheme & Frequency & Codepoint & UnicodeName \\ 
-  \midrule
-sh &  &  &  \\ 
-  ch &  &  &  \\ 
-  sch &  &  &  \\ 
-  ie &  &  &  \\ 
-  oo &  &  &  \\ 
-  a & 2 & U+0061 & LATIN SMALL LETTER A \\ 
-  c & 2 & U+0063 & LATIN SMALL LETTER C \\ 
-  e & 3 & U+0065 & LATIN SMALL LETTER E \\ 
-  f & 1 & U+0066 & LATIN SMALL LETTER F \\ 
-  h & 4 & U+0068 & LATIN SMALL LETTER H \\ 
-  i & 3 & U+0069 & LATIN SMALL LETTER I \\ 
-  m & 4 & U+006D & LATIN SMALL LETTER M \\ 
-  o & 2 & U+006F & LATIN SMALL LETTER O \\ 
-  p & 2 & U+0070 & LATIN SMALL LETTER P \\ 
-  s & 4 & U+0073 & LATIN SMALL LETTER S \\ 
-   \bottomrule
-\end{tabular}
-\endgroup
-\caption{Manually edited profile skeleton} 
-\label{tab:profile_skeleton1}
-\end{table}
-
-It is also possible to leave out the Unicode information in
-the profile skeleton by using the option \texttt{info = FALSE}. It is also
-possible not to use a separate file at all, but process everything within R. In
-simple situations this is often useful (see below), but in general we prefer to
-handle everything through a separately saved orthography profile. This profile
-often contains highly useful information that is nicely coded and saved inside
-this one file, and can thus be easily distributed and shared. Doing the same as
-above completely within R might look as follows:
-
-\begin{knitrout}\footnotesize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# make a profile, just select the column 'Grapheme'}
-\hlstd{profile} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example)[,} \hlstr{"Grapheme"}\hlstd{]}
-\hlcom{# extend the profile with multigraphs}
-\hlstd{profile} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"sh"}\hlstd{,} \hlstr{"ch"}\hlstd{,} \hlstr{"sch"}\hlstd{,} \hlstr{"ie"}\hlstd{,} \hlstr{"oo"}\hlstd{, profile)}
-\hlcom{# use the profile to tokenize}
-\hlkwd{tokenize}\hlstd{(example, profile)}\hlopt{$}\hlstd{strings}
-\end{alltt}
-\begin{verbatim}
-##   originals    tokenized
-## 1   shampoo  sh a m p oo
-## 2    mishap   m i sh a p
-## 3  mischief m i sch ie f
-## 4    scheme    sch e m e
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
-
-\subsection*{Rule ordering}
-\label{rule-ordering}
-
-Not yet everything is correct in this graphemic parsing of the example discussed
-previously. The sequence <sh> in `mishap' should not be a digraph, and
-conversely the sequence <sch> in `mischief' should of course be separated into
-<s> and <ch>. One of the important issues to get the graphemic parsing right is
-the order in which graphemes are parsed. For example, currently the grapheme
-<sch> is parsed before the grapheme <ch>, leading to <m\ i\ sch\ ie\ f> instead
-of the intended <m\ i\ s\ ch\ ie\ f>. The reason that <sch> is parsed before
-<ch> is that by default longer graphemes are parsed before shorter ones. Our
-experience is that in most cases this is expected behaviour. You can change the
-ordering by specifying the option \texttt{ordering}. Setting this option to
-\texttt{NULL} results in no preferential ordering, i.e. the graphemes are parsed
-in the order of the profile, from top to bottom. Now `mischief' is parsed
-correctly, but `scheme' is wrong. So this ordering is not the solution in this
-case.
-
-\begin{knitrout}\footnotesize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# do not reorder the profile}
-\hlcom{# just apply the graphemes from top to bottom}
-\hlkwd{tokenize}\hlstd{( example}
-         \hlstd{,} \hlkwc{profile} \hlstd{=} \hlstr{"~/Desktop/profile_skeleton.txt"}
-         \hlstd{,} \hlkwc{ordering} \hlstd{=} \hlkwa{NULL}
-        \hlstd{)}\hlopt{$}\hlstd{strings}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
-
-\begin{knitrout}\footnotesize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
-##   originals     tokenized
-## 1   shampoo   sh a m p oo
-## 2    mishap    m i sh a p
-## 3  mischief m i s ch ie f
-## 4    scheme    s ch e m e
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
-
-There are various additional options for rule ordering implemented. Please check
-the help description in R \texttt{help(tokenize)} for more details on the
-possible rule ordering specificiations. In summary, there are four different 
-ordering options, that can also be combined:
-
-\begin{itemize}
-
-   \item \textsc{size}\\
-         This option orders the lines in the profile by the size of the
-         grapheme, largest first. Size is measured by number of Unicode
-         characters after normalization as specified in the option
-         \texttt{normalize}. For example, <é> has a size of 1 with
-         \texttt{normalize = "NFC"}, but a size of 2 with
-         \texttt{normalize = "NFD"}.
-
-   \item \textsc{context}\\ This option orders the lines by whether they have
-           any context specified (see next section). Lines with context will
-           then be used first. Note that this only works when the option
-           \texttt{regex = TRUE} is also chosen (otherwise context
-           specifications are not used).
-
-   \item \textsc{reverse}\\ This option orders the lines from bottom to top.
-         Reversing order can be useful because hand-written profiles tend to put
-         general rules before specific rules, which mostly should be applied in
-         reverse order.
-
-  \item \textsc{frequency}\\
-         This option orders the lines by the frequency with which they
-         match in the specified strings before tokenization, least frequent
-         coming first. This frequency of course depends crucially on the
-         available strings, so it will lead to different orderings when applied
-         to different data. Also note that this frequency is (necessarily)
-         measured before graphemes are identified, so these ordering frequencies
-         are not the same as the final frequencies shown in the output.
-         Frequency of course also strongly differs on whether context is used
-         for the matching through \texttt{regex = TRUE}.
-
-\end{itemize}
-
-By specifying more than one ordering, these orderings are used to break ties,
-e.g. the default setting \texttt{ordering = c("size", "context", "reverse")}
-will first order by size, and for those with the same size, it will order by
-whether there is any context specified or not. For lines that are still tied
-(i.e. the have the same size and both/neither have context) the order will be
-reversed compared to the order as attested in the profile, because most
-hand-written specifications of graphemes will first write the general rule,
-followed by more specific regularities. To get the right tokenization, these 
-rules should in most cases be applied in reverse order.
-
-It is important to realize that different ordering of the rules does not have
-so-called `feeding' and `bleeding' effects as known from finite-state rewrite
-rules.\footnote{`Bleeding' is the effect that the application of a rule changes
-the string, so as to prevent a next rule to apply. `Feeding' is the opposite: a
-specific rule will only be applied becuase a previous rule changed the string
-already. The interaction of rules with such `feeding' and `bleeding' effects is
-extremely difficult to predict.} The graphemic parsing advocated here works
-crucially different from rewrite rules in that there is nothing being rewritten:
-each line in an orthography profile specifies a grapheme to be `captured' in the 
-string. All lines in the profile are processed in a specifed order (as determined
-by the option \texttt{ordering}). At the processing of a specific line, all 
-matching graphemes in the data are marked as `captured', but not changed. 
-Captured parts cannot be captured again, but they can still be used to match 
-contexts of other lines in the profile. Only when all lines are processed the 
-captured graphemes are separated (and possibly transliterated). In this way the 
-result of the application of the `rules' is rather easy to predict.
-
-To document a specific case of graphemic parsing, it is highly useful to save
-all results of the tokenization to file by using the option \texttt{file.out},
-for example as follows:
-
-\begin{knitrout}\footnotesize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# save the results to various files}
-\hlkwd{tokenize}\hlstd{( example}
-         \hlstd{,} \hlkwc{profile} \hlstd{=} \hlstr{"~/Desktop/profile_skeleton.txt"}
-         \hlstd{,} \hlkwc{file.out} \hlstd{=} \hlstr{"~/Desktop/result"}
-        \hlstd{)}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
-
-This will lead to the following four files being written. Crucially, a
-new profile is produced with the re-ordered orthography profile. To reproduce
-the tokenization, this re-ordered profile can be used with the option
-\texttt{ordering~=~NULL}.
-
-\begin{itemize}
-
-   \item \textsc{result\_strings.tsv}:\\ A tab-separated file with the original
-         and the tokenized/transliterated strings
-
-   \item \textsc{result\_profile.tsv}:\\ A tab-separated file with the
-         graphemes with added frequencies of occurence in the data. The lines
-         in the file is re-ordered according to the order that resulted from the
-         ordering specifications (see Section~\ref{rule-ordering}).
-
-   \item \textsc{result\_errors.tsv}:\\ A tab-separated file with all original
-         strings that contain unmatched parts. Unmatched parts are indicated
-         with the character as specified with the option \texttt{missing}. By
-         default the character \textsc{double question mark} <⁇> at
-         \uni{2047} is used. When there are no errors, then this file is 
-         absent.
-
-    \item \textsc{result\_missing.tsv}:\\ A tab-separated file with the graphemes
-          that are missing from the original orthography profile, as indicated in
-          the errors. When there are no errors, then this file is absent.
-
-\end{itemize}
-
-\subsection*{Contextually specified graphemes}
-\label{contextual-specification}
-
-To refine a profile, it is also possible to add graphemes with contextual
-specifications. An orthography profile can have columns called \texttt{Left} and
-\texttt{Right} to specify the context in which the grapheme is to be
-separated.\footnote{The columns names \textttf{Left}, \textttf{Right} and
-\textttf{Grapheme} are currently hard-coded, so exactly these column names
-should be used for these effects to take place. The position of the columns in
-the profile is unimportant. So the column \textttf{Left} can occur anywhere.}
-For example, we are adding an extra line to the profile from above, resulting in
-the profile shown in Table~\ref{tab:profile_skeleton2}. The extra line specifies
-that <s> is a grapheme when it occurs after <mi>. Such contextually-specified
-graphemes are based on regular expressions so you can also use regular
-expressions in the description of the context. For such contextually specified
-graphemes to be included in the graphemic parsing we have to specify the option
-\texttt{regex = TRUE}. This contextually specified grapheme should actually be
-handled first, so we could try \texttt{ordering = NULL}. However, we can also
-explicitly specifiy that rules with contextual information should be applied
-first by using \texttt{ordering = "context"}. That gives the right results for
-this toy example.
-
-\begin{knitrout}\footnotesize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# add a contextual grapheme, and then use the edited profile to tokenize}
-\hlkwd{tokenize}\hlstd{( example}
-         \hlstd{,} \hlkwc{profile} \hlstd{=} \hlstr{"~/Desktop/profile_skeleton.txt"}
-         \hlstd{,} \hlkwc{regex} \hlstd{=} \hlnum{TRUE}
-         \hlstd{,} \hlkwc{ordering} \hlstd{=} \hlstr{"context"}
-        \hlstd{)}\hlopt{$}\hlstd{strings}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
-
-\begin{knitrout}\footnotesize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
-##   originals     tokenized
-## 1   shampoo   sh a m p oo
-## 2    mishap   m i s h a p
-## 3  mischief m i s ch ie f
-## 4    scheme    s ch e m e
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
-
-% latex table generated in R 3.3.2 by xtable 1.8-2 package
-% Thu Feb  2 11:38:05 2017
-\begin{table}[htb]
-\centering
-\begingroup\scriptsize
-\begin{tabular}{lllll}
-  \toprule
-Left & Grapheme & Frequency & Codepoint & UnicodeName \\ 
-  \midrule
-mi & s &  &  &  \\ 
-   & sh &  &  &  \\ 
-   & ch &  &  &  \\ 
-   & sch &  &  &  \\ 
-   & ie &  &  &  \\ 
-   & oo &  &  &  \\ 
-   & a & 2 & U+0061 & LATIN SMALL LETTER A \\ 
-   & c & 2 & U+0063 & LATIN SMALL LETTER C \\ 
-   & e & 3 & U+0065 & LATIN SMALL LETTER E \\ 
-   & f & 1 & U+0066 & LATIN SMALL LETTER F \\ 
-   & h & 4 & U+0068 & LATIN SMALL LETTER H \\ 
-   & i & 3 & U+0069 & LATIN SMALL LETTER I \\ 
-   & m & 4 & U+006D & LATIN SMALL LETTER M \\ 
-   & o & 2 & U+006F & LATIN SMALL LETTER O \\ 
-   & p & 2 & U+0070 & LATIN SMALL LETTER P \\ 
-   & s & 4 & U+0073 & LATIN SMALL LETTER S \\ 
-   \bottomrule
-\end{tabular}
-\endgroup
-\caption{Orthography profile with contextual specification for <s>} 
-\label{tab:profile_skeleton2}
-\end{table}
-
-It is important to realize that with the option \texttt{regex = TRUE} all
-content in the profile is treated as regular expressions, so the characters with
-special meaning in regular expressions should be either omitted or escaped (by
-putting a <\ \backslash\ > \textsc{reverse solidus} at \uni{005C} before the
-character). Specifically, this concerns the following characters:
-
-\begin{itemize}
-
-  \item[] <-> \textsc{hyphen-minus} at \uni{002D}
-  \item[] <!> \textsc{exclamation mark} at \uni{0021}
-  \item[] <?> \textsc{question mark} at \uni{003F}
-  \item[] <.> \textsc{full stop} at \uni{002E}
-  \item[] <(> \textsc{left parenthesis} at \uni{0028}
-  \item[] <)> \textsc{right parenthesis} at \uni{0029}
-  \item[] <[> \textsc{left square bracket} at \uni{005B}
-  \item[] <]> \textsc{right square bracket} at \uni{005D}
-  \item[] <\{> \textsc{left curly bracket} at \uni{007B}
-  \item[] <\}> \textsc{right curly bracket} at \uni{007D}
-  \item[] <|> \textsc{vertical line} at \uni{007C}
-  \item[] <*> \textsc{asterisk} at \uni{002A}
-  \item[] <\backslash> \textsc{reverse solidus} at \uni{005C}
-  \item[] <ˆ> \textsc{circumflex accent} at \uni{005E}
-  \item[] <+> \textsc{plus sign} at \uni{002B}
-  \item[] <\$> \textsc{dollar sign} at \uni{0024}
-
-\end{itemize}
-
-\subsection*{Profile skeleton with columns for editing}
-\label{profile-editing}
-
-When it is expected that context might be important for a profile, then the
-profile skeleton can be created with columns prepared for the contextual
-specifications. This is done by using the option \texttt{editing = TRUE} (cf.
-Table~\ref{tab:profile_editing_1} for a toy profile of some Italian words).
-
-\begin{knitrout}\footnotesize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlstd{example} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{'cane'}\hlstd{,} \hlstr{'cena'}\hlstd{,} \hlstr{'cine'}\hlstd{)}
-\hlkwd{write.profile}\hlstd{(example}
-              \hlstd{,} \hlkwc{file} \hlstd{=} \hlstr{"~/Desktop/profile_skeleton.txt"}
-              \hlstd{,} \hlkwc{editing} \hlstd{=} \hlnum{TRUE}
-              \hlstd{,} \hlkwc{info} \hlstd{=} \hlnum{FALSE}
-              \hlstd{)}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
-
-% latex table generated in R 3.3.2 by xtable 1.8-2 package
-% Thu Feb  2 11:38:05 2017
-\begin{table}[htb]
-\centering
-\begingroup\scriptsize
-\begin{tabular}{lllll}
-  \toprule
-Left & Grapheme & Right & Class & Replacement \\ 
-  \midrule
- & a &  &  & a \\ 
-   & c &  &  & c \\ 
-   & e &  &  & e \\ 
-   & i &  &  & i \\ 
-   & n &  &  & n \\ 
-   \bottomrule
-\end{tabular}
-\endgroup
-\caption{Orthography profile with empty columns for editing contexts} 
-\label{tab:profile_editing_1}
-\end{table}
-
-Besides the columns \texttt{Left}, \texttt{Grapheme}, and \texttt{Right} as
-discussed in the previous sections, there are also columns \texttt{Class} and
-\texttt{Replacement}. The column \texttt{Class} can be used to specify classes
-of graphemes that can then be used in the contextual specification. The column
-\texttt{Replacement} is just a copy of the column \textt{Grapheme}, providing a
-skeleton to specify transliteration. The name of the column
-\texttt{Replacement} is not fixed -- there can actually be multiple columns with 
-different kinds of transliterations in a single profile.
-
-To achieve contextually determined replacements it is possible to use a regular
-expression in the contextual column. For example, consider the edited toy
-profile for Italian in Table~\ref{tab:profile_editing_2} (where <c> becomes /k/
-except before <i,e>, then it becomes /tʃ/).
-
-% latex table generated in R 3.3.2 by xtable 1.8-2 package
-% Thu Feb  2 11:38:05 2017
-\begin{table}[htb]
-\centering
-\begingroup\scriptsize
-\begin{tabular}{lllll}
-  \toprule
-Left & Grapheme & Right & Class & IPA \\ 
-  \midrule
- & c & [ie] &  & tʃ \\ 
-   & a &  &  & a \\ 
-   & n &  &  & n \\ 
-   & c &  &  & k \\ 
-   & e &  &  & e \\ 
-   & i &  &  & i \\ 
-   \bottomrule
-\end{tabular}
-\endgroup
-\caption{Orthography profile with regex as context} 
-\label{tab:profile_editing_2}
-\end{table}
-
-To use this profile, you have to add the option \texttt{regex = TRUE}. Also note
-that we have changed the name of the transliteration-column, so we have to tell
-the tokenization process to use this column to transliterate. This is done by
-adding the option \texttt{transliterate = "IPA"}.
-
-\begin{knitrout}\footnotesize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# add a contextual grapheme, and then use the edited profile to tokenize}
-\hlkwd{tokenize}\hlstd{( example}
-         \hlstd{,} \hlkwc{profile} \hlstd{=} \hlstr{"~/Desktop/profile_skeleton.txt"}
-         \hlstd{,} \hlkwc{regex} \hlstd{=} \hlnum{TRUE}
-         \hlstd{,} \hlkwc{transliterate} \hlstd{=} \hlstr{"IPA"}
-        \hlstd{)}\hlopt{$}\hlstd{strings}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
-
-\begin{knitrout}\footnotesize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
-##   originals tokenized transliterated
-## 1      cane   c a n e        k a n e
-## 2      cena   c e n a       tʃ e n a
-## 3      cine   c i n e       tʃ i n e
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
-
-Another equivalent possibility is to use a column \texttt{Class} to specify a
-class of graphemes, and then use this class in the specification of context.
-This is useful to keep track of recurrent classes in larger profiles. You are
-free to use any class-name you like, as long as it doesn't clash with the rest
-of the profile. The example shown in Table~\ref{tab:profile_editing_3} should 
-give the same result as obtained previously by using a regular expression.
-
-% latex table generated in R 3.3.2 by xtable 1.8-2 package
-% Thu Feb  2 11:38:05 2017
-\begin{table}[htb]
-\centering
-\begingroup\scriptsize
-\begin{tabular}{lllll}
-  \toprule
-Left & Grapheme & Right & Class & IPA \\ 
-  \midrule
- & c & Vfront &  & tʃ \\ 
-   & a &  &  & a \\ 
-   & n &  &  & n \\ 
-   & c &  &  & k \\ 
-   & e &  & Vfront & e \\ 
-   & i &  & Vfront & i \\ 
-   \bottomrule
-\end{tabular}
-\endgroup
-\caption{Orthography profile with Class as context} 
-\label{tab:profile_editing_3}
-\end{table}
-
-\begin{knitrout}\footnotesize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# add a class, and then use the edited profile to tokenize}
-\hlkwd{tokenize}\hlstd{( example}
-         \hlstd{,} \hlkwc{profile} \hlstd{=} \hlstr{"~/Desktop/profile_skeleton.txt"}
-         \hlstd{,} \hlkwc{regex} \hlstd{=} \hlnum{TRUE}
-         \hlstd{,} \hlkwc{transliterate} \hlstd{=} \hlstr{"IPA"}
-        \hlstd{)}\hlopt{$}\hlstd{strings}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
-
-\begin{knitrout}\footnotesize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
-##   originals tokenized transliterated
-## 1      cane   c a n e        k a n e
-## 2      cena   c e n a       tʃ e n a
-## 3      cine   c i n e       tʃ i n e
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
-
-\subsection*{Formatting grapheme separation}
-\label{formattingseparation}
-
-In all examples above we have used the default formatting for grapheme
-separation using space as a separator, which is obtained by the default setting
-\texttt{sep~=~"~"}. It is possible to specify any other separator here,
-including the empty string, i.e. \texttt{sep = ""}. This will not show the
-graphemic tokenization anymore (although it has of course been used in the
-background).
-
-\begin{knitrout}\footnotesize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# Use the empty string as separator}
-\hlkwd{tokenize}\hlstd{( example}
-         \hlstd{,} \hlkwc{profile} \hlstd{=} \hlstr{"~/Desktop/profile_skeleton.txt"}
-         \hlstd{,} \hlkwc{regex} \hlstd{=} \hlnum{TRUE}
-         \hlstd{,} \hlkwc{transliterate} \hlstd{=} \hlstr{"IPA"}
-         \hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}
-        \hlstd{)}\hlopt{$}\hlstd{strings}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
-
-\begin{knitrout}\footnotesize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
-##   originals tokenized transliterated
-## 1      cane      cane           kane
-## 2      cena      cena          tʃena
-## 3      cine      cine          tʃine
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
-
-Normally, the separator specified should not occur in the data. If it does,
-unexpected things might happen, so consider removing the chosen seperator from
-your strings first. However, there is also an option \texttt{sep.replace} to
-replace the separator with something else. When \code{sep.replace} is specified,
-this mark is inserted in the string at those places where the separator occurs.
-Typical usage in linguistics would be \texttt{sep = " ", sep.replace = "\#"} adding
-spaces between graphemes and replacing spaces in the input string by hashes in
-the output string.
-
-\begin{knitrout}\footnotesize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# Replace separator in string to be tokenized}
-\hlkwd{tokenize}\hlstd{(} \hlstr{"test test test"}
-         \hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{" "}
-         \hlstd{,} \hlkwc{sep.replace} \hlstd{=} \hlstr{"#"}
-        \hlstd{)}\hlopt{$}\hlstd{strings}\hlopt{$}\hlstd{tokenized}
-\end{alltt}
-\begin{verbatim}
-## [1] "t e s t # t e s t # t e s t"
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
-
-\subsection*{Remaining issues}
-
-Given a set of graphemes, there are at least two different methods to tokenize
-strings. The first is called \texttt{method = "global"}. This approach
-takes the first grapheme in the profile, then matches this grapheme globally at
-all places in the string, and then turns to the next string in the profile. To
-us, this seems to be the most intuitive way for linguists to deal with graphemic
-parsing. The other approach is called \texttt{method = "linear"}. This
-approach walks through the string from left to right. At the first character it
-looks through all graphemes whether there is any match, and then walks further
-to the end of the match and starts again. This approach is more alike to
-finite-state rewrite rules (though note that it still works differently from
-such rewrite rules, see Section~\ref{rule-ordering}). The global method is used 
-by default.
-
-In some special cases these two tokenization methods can lead to different
-results, but these special situations are very unlikely to happen in natural
-language. The example below shows that a string \texttt{'abc'} can be parsed
-differently in case of a very special profile with a very special ordering of
-the graphemes.
-
-\begin{knitrout}\footnotesize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# different parsing methods can lead to different results}
-\hlcom{# the global method first catches 'bc'}
-\hlkwd{tokenize}\hlstd{(} \hlstr{"abc"}
-         \hlstd{,} \hlkwc{profile} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{"bc"}\hlstd{,}\hlstr{"ab"}\hlstd{,}\hlstr{"a"}\hlstd{,}\hlstr{"c"}\hlstd{)}
-         \hlstd{,} \hlkwc{order} \hlstd{=} \hlkwa{NULL}
-         \hlstd{,} \hlkwc{method} \hlstd{=} \hlstr{"global"}
-         \hlstd{)}\hlopt{$}\hlstd{strings}
-\end{alltt}
-\begin{verbatim}
-##   originals tokenized
-## 1       abc      a bc
-\end{verbatim}
-\begin{alltt}
-
-\hlcom{# the linear method catches the first grapheme, which is 'ab'}
-\hlkwd{tokenize}\hlstd{(} \hlstr{"abc"}
-         \hlstd{,} \hlkwc{profile} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{"bc"}\hlstd{,}\hlstr{"ab"}\hlstd{,}\hlstr{"a"}\hlstd{,}\hlstr{"c"}\hlstd{)}
-         \hlstd{,} \hlkwc{order} \hlstd{=} \hlkwa{NULL}
-         \hlstd{,} \hlkwc{method} \hlstd{=} \hlstr{"linear"}
-         \hlstd{)}\hlopt{$}\hlstd{strings}
-\end{alltt}
-\begin{verbatim}
-##   originals tokenized
-## 1       abc      ab c
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
-
-Further, the current R-implementation has a limitation when regular expressions
-are used. The problem is that overlapping matches are not captured when using
-regular expressions.\footnote{This restriction is an effect of the underlyingly
-used ICU implementation of the Unicode standard as implemented in R through the
-package \texttt{stringi}.} Everything works as expected without regular
-expressions, but there might be warnings/errors in case of \texttt{regex =
-TRUE}. However, just as in the previous issue, this problem should only very
-rarely (when at all) happen in natural language data.
-
-The problem can be exemplified by a sequence <bbbb> in which a 
-grapheme <bb> should be matched. With the default \texttt{regex = FALSE} 
-there are three possible matches, but with \texttt{regex = TRUE} only the first 
-two or the last two are matched. The middle two are not matched because they 
-overlap with the other matches. In the example below this leads to an error, 
-because the second <bb> is not matched. We have not been able to produce a real 
-example in any natural language in which this limitation might lead to an error.
-
-\begin{knitrout}\footnotesize
-\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# Everything perfekt without regular expressions}
-\hlkwd{tokenize}\hlstd{(} \hlstr{"abbb"}
-        \hlstd{,} \hlkwc{profile} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{"ab"}\hlstd{,}\hlstr{"bb"}\hlstd{)}
-        \hlstd{,} \hlkwc{order} \hlstd{=} \hlkwa{NULL}
-        \hlstd{,} \hlkwc{regex} \hlstd{=} \hlnum{FALSE}
-        \hlstd{)}\hlopt{$}\hlstd{strings}
-\end{alltt}
-\begin{verbatim}
-##   originals tokenized
-## 1      abbb     ab bb
-\end{verbatim}
-\begin{alltt}
-
-\hlcom{# Matching with regular expressions does not catch overlap}
-\hlkwd{tokenize}\hlstd{(} \hlstr{"abbb"}
-        \hlstd{,} \hlkwc{profile} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{"ab"}\hlstd{,}\hlstr{"bb"}\hlstd{)}
-        \hlstd{,} \hlkwc{order} \hlstd{=} \hlkwa{NULL}
-        \hlstd{,} \hlkwc{regex} \hlstd{=} \hlnum{TRUE}
-        \hlstd{)}\hlopt{$}\hlstd{strings}
-\end{alltt}
-
-{\ttfamily\noindent\color{warningcolor}{\#\# Warning in tokenize("{}abbb"{}, profile = c("{}ab"{}, "{}bb"{}), order = NULL, regex = TRUE): \\\#\# There were unknown characters found in the input data.\\\#\# Check output\$errors for a table with all problematic strings.}}\begin{verbatim}
-##   originals tokenized
-## 1      abbb    ab ⁇ ⁇
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
-
-% \include{chapters/use_cases}
-% knit_child("chapters/codetesting.tex")
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%                                              %%%
-%%%             Backmatter                       %%%
-%%%                                              %%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% There is normally no need to change the backmatter section
-\input{backmatter.tex}
-\end{document}
-
-% you can create your book by running
-% xelatex lsp-skeleton.tex
-%
-% you can also try a simple 
-% make
-% on the commandline

### 2017-12-27 book/chapters/ipa_meets_unicode.tex
deals with chapters 2 and 3 -- wip

@@
       is confusingly referred to as \textsc{latin letter retroflex click} in the
-      Unicode standard, which is probably best seen as an error.
+      Unicode Standard, which is probably best seen as an error.
 \item \textsc{Voiced retroflex implosive} \newline 

### 2017-12-27 book/chapters/pitfalls.tex
deals with chapters 2 and 3 -- wip

@@
 computational environment for written language. In those cases in which the
-Unicode Standard behaves differently as expected, we think it is important not
+Unicode Standard behaves differently than expected, we think it is important not
 to dismiss Unicode as wrong or deficient, because our
@@
 contextually-sensitive variants (isolated, word initial, medial and final).
+
 Second, a single character may be rendered as a sequence of multiple glyphs. For
@@
 consonant and vowel, which are rendered as two adjacent glyphs by fonts that
-supports Tamil. Third, a single glyph may be a combination of multiple
-characters. For example, the ligature <ﬁ>, a single glyph, is the result of two
+supports Tamil, e.g. \textsc{tamil letter au} at \uni{0B94} represents a single 
+character <ஔ>, composed of two glyphs <ஓ> and <ன>. Perhaps confusingly, 
+in the Unicode Standard there are also two individual characters, 
+i.e.\ \textsc{tamil letter oo} at \uni{0B93} and 
+\uni{0BA9} \textsc{tamil letter nnna}, each of which is a glyph. Another example is 
+Sinhala \textsc{sinhala vowel sign kombu deka} at \uni{0DDB} <ෛ>, which is 
+visually two glyphs, each represented by \textsc{sinhala vowel sign kombuva} 
+at \uni{0DD9} < ෙ>.
+
+Third, a single glyph may be a combination of multiple characters. For example, 
+the ligature <ﬁ>, a single glyph, is the result of two
 characters, <f> and <i>, that have undergone glyph substitution by font
@@
 experience with rendering on the side of the software industry is developing 
-quickly, so we can expect the situation only to get better. In the meantime one 
-can try to correct faulty layout by tweaking baseline and/or kerning (when such 
-option are available).
+quickly, so we can expect the situation to only get better. % In the meantime one 
+% can try to correct faulty layout by tweaking baseline and/or kerning (when such 
+% option are available).
 

### 2017-12-27 book/chapters/ipa_meets_unicode.tex
deals with chapter 4 revisions -- wip

@@
 between an open back unrounded vowel, which is encoded in IPA with a
-`single-story' <ɑ> \textsc{latin small letter alpha} at \uni{0251}, and the open
-front unrounded vowel, which is encoded in IPA with a `double-story' <a>
+``single-story'' <ɑ> \textsc{latin small letter alpha} at \uni{0251}, and the open
+front unrounded vowel, which is encoded in IPA with a ``double-story'' <a>
 \textsc{latin small letter a} at \uni{0061}, basically treating them as
@@
 sufficiently urgent to introduce even more <a> homographs in Unicode (which
-already lead to much confusion, see Section~\ref{pitfall-homoglyphs}).}
+already lead to much confusion, see Section~\ref{pitfall-homoglyphs}).} This 
+approach was taken by the Language Science Press to distinguish 
+between the two different orthographic <a>'s in Chakali in 
+\cite{Brindle2017}.\footnote{\url{http://langsci-press.org/catalog/book/74}}
 
@@
 IPA encoding we propose to standardize on the ring below. As a principle, for
-strict IPA encoding only one options should be allowed for all diacritics.
+strict IPA encoding only one option should be allowed for all diacritics.
 
@@
 
-   \item Tone letters are normally written at the end of the syllable,
+   \item Tone letters <˥, ˦, ˧, ˨, ˩> from \uni{02E5}--\uni{02E9} 
+   		 are normally written at the end of the syllable,
          possibly occurring immediately adjacent to a consonant in the coda of
@@
 like strokes (e.g.~in the <ɨ>), retroflex hooks (e.g.~in <ʐ>) or rhotic hooks
-(e.g.~in <ɝ>) will not be decomposed, but strangely enough a cedilla (like in
-<ç>) will be decomposed (see Section~\ref{pitfall-missing-decomposition}). In
+(e.g.~in <ɝ>) will not be decomposed. In
 general, Unicode decomposition does not behave like a feature decomposition as
@@
 
-By Unicode Canonical Combining Classes, the diacritics on top of a character,
+By Unicode Canonical Combining Classes, overlay diacritics 
 like <\dia{0334}> (Combining Class number 1), always come before diacritics
@@
 changes to specific symbols (in previous charts) were debated and 
-the Association's members certain decisions. The prevailing mood at 
+the Association's members made certain decisions. The prevailing mood at 
 the convention was not to change specific symbols unless a strong 
@@
 \begin{itemize}
-	\item symbols for clicks were changed from <ʇ~ʖ~ʗ> to <ǀ~ǁ~ǃ>
+	\item Symbols for clicks were changed from <ʇ~ʖ~ʗ> to <ǀ~ǁ~ǃ>
        because the latter were the symbols used by nearly all Khoisanists and
        Bantuists.
-	\item the Americanist tradition of using using <\dia{030C}>, a
+	\item The Americanist tradition of using using <\dia{030C}>, a
        \textsc{combining caron} at \uni{030C} for all post-alveolar sounds, like
@@
 \noindent These decisions have practical consequences for transcribers 
-of IPA, particularly those who wish to follow recommended practices of 
+of IPA, particularly those who wish to follow current recommended practices of 
 encoding electronic text in the Unicode Standard. For example, the Unicode 
-Standard contains removed symbols as labels them as clicks, e.g. <ʇ> 
-\textsc{latin small letter turned t} at U+0287 in the \textsc{IPA Extensions} 
-block has the comment `dental click (sound of ``tsk tsk'')'. In this case, 
-the IPA transcriber must know the status of click symbols in the current 
-version of the IPA and then identify those characters within the Unicode Standard.
+Standard contains \textsc{latin small letter turned t} <ʇ> at U+0287, 
+which is no longer part of the IPA. It still exists, however, in the Unicode 
+\textsc{IPA Extensions} block, with the comment ``dental click (sound of `tsk tsk')''. 
+In such cases, the IPA transcriber must know the status of legacy symbols in the current 
+version of the IPA and the correct characters in the Unicode Standard.
 
@@
 In the course of collecting a large sample of phoneme systems across the world's
-languages \citep{Moran2012}, \citet{Moran_etal2014} found that in order 
+languages, \citet{Moran_etal2014} found that in order 
 to preserve distinctions both within and across language descriptions, 

### 2018-01-13 book/chapters/_remove/originaltext.tex
issues in chapter 2

@@
 Within the Unicode structure many of these specifications can be captured by so-called `Locales' or `Locale Descriptions', so we are moving to a new distinction of `universal' (Unicode Standard) versus `language-particular' (Locale Description).
-The major gain is a much larger compatibility on the universal level (because Unicode standardizes a much larger portion of writing system diversity), and much better possibilities for automatic processing on the language-particular level (because Locale Descriptions are computer readable specifications).
+The major gain is a much larger compatibility on the universal level (because Unicode standardizes a much larger portion of writing system diversity), and much better possibilities for automated processing on the language-particular level (because Locale Descriptions are computer readable specifications).
 
@@
 
-For the automatic processing of the profiles, the following technical standards will be expected:
+For the automated processing of the profiles, the following technical standards will be expected:
 \begin{enumerate}

### 2018-01-13 book/chapters/_remove/terminology.tex
issues in chapter 2

--- 2015-06-05 chapters/_remove/terminology.tex
@@
 much larger portion of writing system diversity), and much better possibilities
-for automatic processing on the language-particular level (because Locale
+for automated processing on the language-particular level (because Locale
 Descriptions are computer readable specifications).

### 2018-01-13 book/chapters/ipa_meets_unicode.tex

--- 2017-12-26 book/chapters/ipa_meets_unicode.tex
@@
 strict encoding guidelines. Our main goal in this chapter is to present the
-encoding issues and propose recommendations for a strict IPA encoding for
+encoding issues and propose recommendations for a ``strict'' IPA encoding for
 situations in which cross-resource consistency is crucial.
@@
 are also included, see~\cite[Appendix~C]{Moran2012}. The 159 characters from 12 
-code blocks shown here are the characters proposed for strict IPA encoding, 
+code blocks shown here are the characters proposed for ``strict'' IPA encoding, 
 as discussed in Section~\ref{ipa-recommendations}.}
@@
         postalveolar click (not retroflex). This naming is probably best seen as
-        an error in the Unicode Standard. For the real retroflex click, see 
+        an error in the Unicode Standard. For the ``real'' retroflex click, see 
         Section~\ref{ipa-additions}.}
@@
            \textsc{modifier letter apostrophe} at \uni{02BC}, respectively. 
-   \item There are two different dashed-l characters in IPA, namely the <ɫ> 	      \textsc{latin small letter l with middle tilde} at \uni{026B} and the <ɬ>	      \textsc{latin small letter l with belt} at \uni{026C}.
+   \item There are two different ``dashed''-l characters in IPA, namely the <ɫ> 	      \textsc{latin small letter l with middle tilde} at \uni{026B} and the <ɬ>	      \textsc{latin small letter l with belt} at \uni{026C}.
 These of course
@@
 not acceptable when the goal is interoperability between resources written in
-IPA.\@ We consider it crucial to distinguish between valid IPA, for which it
+IPA.\@ We consider it crucial to distinguish between ``valid'' IPA, for which it
 is sufficient that any phonetically-trained reader is able to understand the
-transcription, and strict IPA, which should be standardized on a single
+transcription, and ``strict'' IPA, which should be standardized on a single
 unique encoding for each sound, so search will work across resources. We are
@@
 principle (see Section~\ref{pitfall-characters-are-not-graphemes}) is that fixed
-sequences in a writing system have to be specified as tailored grapheme
+sequences in a writing system have to be specified as ``tailored'' grapheme
 clusters. Only in case the sequence is not a cluster, then this has to be
@@
 \uni{0334}. The main reason some phoneticians like to use this in IPA is to mark
-the dark <l> in English codas, using the character <ɫ> \textsc{latin small
+the ``dark'' <l> in English codas, using the character <ɫ> \textsc{latin small
 letter l with middle tilde} at \uni{026B}. This character is not canonically
@@
       types of frictionalized sounds: First, click consonants where the release
-      of the anterior closure involves an ingressive sucking sound similar
-      to a fricative, for example <kǃ͓ʰ>; second, frictionalized vowels
+      of the anterior closure involves an ingressive ``sucking'' sound similar
+      to a fricative, for example <kǃ͓ʰ>; second, ``frictionalized'' vowels
       (sounds that are phonologically vocalic, but with sufficiently close
@@
       <\dia{02C0}> \textsc{modifier letter glottal stop} at \uni{02C0}, unless
-      it is clear that either ejective or creaky voicing are the
+      it is clear that either ``ejective'' or ``creaky voicing'' are the
       intended sounds (in which cases the standard IPA diacritics should be
@@
 159 different unicode codepoints that form the basis of strict-IPA encoding
-(107 letters, 36 diacritics and 16 remaining symbols). Each of
-these tables shows a typical glyph, and then lists the Unicode Codepoint,
+(107 letters, 36 diacritics and 16 remaining symbols). We also make these tables available 
+online in CSV format.\footnote{\url{https://github.com/unicode-cookbook/cookbook/tree/master/book/tables}} 
+Each of these tables shows a typical glyph, and then lists the Unicode Codepoint,
 Unicode Name and IPA description for each symbol. Further, there is a table with 

### 2018-01-13 book/chapters/ipa_background.tex

@@
 allowed. Because the IPA does not provide phonological analyses for specific
-languages, the IPA does not define a single ``correct'' transcription system.
+languages, the IPA does not define a single correct transcription system.
 Rather, the IPA aims to provide a resource that allows users to express any
@@
 
-In contrast, the basic goal of a phonemic ``broad'' transcription is to distinguish all
+In contrast, the basic goal of a phonemic broad transcription is to distinguish all
 words in a language with a minimal number of transcription symbols
@@
 representations of the IPA, which dealt mainly with the assignment of characters
-on the keyboard to IPA symbols. The process of assigning standardized ``computer
-codes'' to phonetic symbols was assigned to the \textsc{workgroup on computer
+on the keyboard to IPA symbols. The process of assigning standardized computer
+codes to phonetic symbols was assigned to the \textsc{workgroup on computer
 coding} (henceforth \textit{working group}) at the Kiel Convention. This working
@@
 revision they were removed. Ligatures like <~ʧ,~ʤ~> are included as formerly 
-recognized IPA symbols; they are assigned to the ``200 series'' of IPA numbers 
+recognized IPA symbols; they are assigned to the 200 series of IPA numbers 
 as members of the group of symbols formerly recognized by the IPA. To ensure 
@@
 
-The IPA Number is simply expressed as a ``three-digit number numerical directory
-of digit triples'' \cite{}. The numbering scheme specifies three-digit codes,
+The IPA Number is simply expressed as a three-digit number numerical directory
+of digit triples. The numbering scheme specifies three-digit codes,
 the first digit of which indicates the symbol's category
@@
 own submissions as part of a review process by ISO for approval based on 
-both ``informatic'' and phonetic criteria \citep[86]{EslingGaylord1993}.
+both informatic and phonetic criteria \citep[86]{EslingGaylord1993}.
 

### 2018-01-13 book/chapters/ipa_meets_unicode.tex

--- 2017-12-27 book/chapters/ipa_meets_unicode.tex
@@
       is confusingly referred to as \textsc{latin letter retroflex click} in the
-      Unicode Standard, which is probably best seen as an error.
+      Unicode standard, which is probably best seen as an error.
 \item \textsc{Voiced retroflex implosive} \newline 

### 2018-01-13 book/chapters/implementation.tex

@@
 To illustrate the practical application of orthography profiles as defined in
-the previous Chapter \ref{orthography-profiles}, we have implemented two 
+the previous Chapter~\ref{orthography-profiles}, we have implemented two 
 versions of the specifications presented there: one in Python and one
@@
 regular expressions.\footnote{This restriction is an effect of the underlyingly
-used ICU implementation of the Unicode standard as implemented in R through the
+used ICU implementation of the Unicode Standard as implemented in R through the
 package \textttf{stringi}.} Everything works as expected without regular

### 2018-01-13 book/chapters/introduction.tex

@@
 \citep[44]{Olumuyiw2013}. However, writing systems (and the languages they
-represent) are normally full of homographs (and homophones), which is normally
+represent) are normally full of homographs (and homophones), which is 
 not a problem at all for speakers of the language. More importantly, writing is
@@
 for linguists in daily practice, and at the same time computer-readable for 
-automatic processing.
+automated processing.
 
@@
 Before we deal with these proposals we will first discuss the theoretical
-background on text encoding, on the Unicode standard, and on the International
+background on text encoding, on the Unicode Standard, and on the International
 Phonetic Alphabet. In the remainder of this chapter, we give an extended
@@
 Linguists are more often interested in using the Unicode Standard with the International Phonetic Alphabet (IPA).
-We first provide a history of the development of the IPA and early attempts to encode it electronically (Chapter~\ref{the-international-phonetic-alphabet}) before we discuss the rather problematic marriage of the International Phonetic Alphabet (IPA) with the Unicode Standard (Chapter\ref{ipa-meets-unicode}).
+We first provide a history of the development of the IPA and early attempts to encode it electronically (Chapter~\ref{the-international-phonetic-alphabet}) before we discuss the rather problematic marriage of the International Phonetic Alphabet (IPA) with the Unicode Standard (Chapter~\ref{ipa-meets-unicode}).
 
@@
 represented in small capital letters (e.g.~\textsc{latin small letter schwa})
-and code points are expressed as U\emph{+n} where \emph{n} is a four to six
+and code points are expressed as U\emph{+n}, where \emph{n} is a four to six
 digit hexadecimal number, e.g.~\uni{0256}, which can be rendered as the glyph <ə>.
@@
 Serbian writing system uses two scripts: Latin and Cyrillic}. Second, the term
-writing system may also refer to a type of symbolic system as used among the
-world's languages to represent the language, as, for example, in
+writing system may also refer to a type of symbolic system as, for example, in
 \emph{alphabetic writing system}. In this latter sense the term refers to how
@@
 synchronic phonemic rules change the pronunciation, as for example with final
-devoicing in many Germanic languages). All orthographies are language-specific
+devoicing in many Germanic languages). Orthographies are language-specific
 (and often even resource-specific), although individual symbols or rules might
@@
 Unicode standardizes a much larger portion of writing system diversity), and
-much better possibilities for automatic processing on the language-particular
+much better possibilities for automated processing on the language-particular
 level (because locale descriptions are computer readable specifications).
@@
 perspective of Unicode they are \textit{the same thing}. In this approach, a
-\textsc{font} is then simply a collection of glyphs linked to code points.
+\textsc{font} is then simply a collection of glyphs connected to code points.
 Allography is not specified in Unicode (expect for a few exceptional cases, due
@@
 traditions. For example, the corresponding lowercase character to \uni{0047} is
-\uni{0067} (though note that the relation between uppercase and lowercase is in
+\uni{0067}.\footnote{Note that the relation between uppercase and lowercase is in
 many situations much more complex than this, and Unicode has further
-specifications for those cases). Another use of properties is to define the
+specifications for those cases.} Another use of properties is to define the
 script of a character.\footnote{The Unicode Glossary defines the term \textsc{script} as
@@
 definitional statement. For example, many IPA symbols are not located in the
-aptly-names block \textsc{IPA extensions}, but in other blocks
+aptly-named block \textsc{IPA extensions}, but in other blocks
 (see Section~\ref{pitfall-no-complete-ipa-block}).
@@
 sequence. Note that this specific sequence can also be represented with a single
-so-called \textsc{precomposed} code point, \textsc{latin small letter n with
+so-called \textsc{precomposed code point}, \textsc{latin small letter n with
 tilde} at \uni{00F1}, but this is not the case for all multi-code point
@@
 clusters are locale-dependent graphemes, i.e.~such combination of characters do
-not function as text elements in all situations. For example, the sequence
-<c>~+~<h> for the Slovak digraph <ch> or the sequence <ky> in the Sisaala
+not function as text elements in all situations. Examples include the sequence
+<c>~+~<h> for the Slovak digraph <ch> and the sequence <ky> in the Sisaala
 practical orthography (pronounced as IPA /tʃ/,~\cite{Moran2006}). These grapheme

### 2018-01-13 book/chapters/ipa_background.tex

@@
 (Section~\ref{IPApremises-principles}) leads to a better 
-appreciation of the challenges that the IPA association 
+appreciation of the challenges that the International Phonetic Association 
 faced when digitally encoding the IPA's set of symbols and 
@@
 a chart and were given phonetic feature labels (e.g.\ for manner of
-articulation among others `plosives', `nasales', `fricatives', for place of
-articulation among others `bronchiales', `laryngales', `labiales' and for vowels
-e.g.\ `fermées', `mi-fermées', `mi-ouvertes', `ouvertes'). Throughout the last
+articulation among others ``plosives'', ``nasales'', ``fricatives'', for place of
+articulation among others ``bronchiales'', ``laryngales'', ``labiales'' and for vowels
+e.g.\ ``fermées'', ``mi-fermées'', ``mi-ouvertes'', ``ouvertes''). Throughout the last
 century, the structure of the chart has changed with increases in phonetic
@@
 labels -- what are often called `features' even though the IPA
-deliberately avoids this term -- have changed (e.g.\ terms like `lips', `throat'
-or `rolled' are not used anymore). Phonetic symbol values have changed (e.g.\
+deliberately avoids this term -- have changed (e.g.\ terms like ``lips'', ``throat''
+or ``rolled'' are not used anymore). Phonetic symbol values have changed (e.g.\
 voiceless is no longer marked by <h>). Symbols have been dropped (e.g.\ the
@@
 unambiguously represent phonetic details. However, since phonetic detail could
-potentially include anything, e.g. something like `deep voice', the IPA
+potentially include anything, e.g. something like ``deep voice'', the IPA
 restricts phonetic detail to linguistically relevant aspects of speech.
@@
 description, and a way of representing the contrasting sounds that occur in a
-language. Thus [m] is equivalent to `voiced bilabial nasal', and is also a way
+language. Thus [m] is equivalent to ``voiced bilabial nasal'', and is also a way
 of representing one of the contrasting nasal sounds that occur in English and
@@
 falls under the rubric of diacritically-distinguished symbols, e.g.\ the 
-difference in English between an aspirated /p/ in [spʰot] and 
-an unrealeased /p/ in [stop̚].
+difference in English between an aspirated /p/ and /t/ in [spʰot̚] and 
+an unrealeased /p/ in [stʰop̚].
 
@@
 analysis so that it is widely understood. Basically, the IPA allows for both a 
-`narrow' phonetic transcription and a `broad' phonological transcription. 
-A `narrow' phonetic transcription may freely use the all symbols in the IPA 
+\textit{narrow} phonetic transcription and a \textit{broad} phonological transcription. 
+A narrow phonetic transcription may freely use the all symbols in the IPA 
 chart with direct reference to their phonetic value, i.e. the transcriber can 
@@
 phonemically in IPA as /ieaou/, though phonetically they are closer to [iεaɔu].
-Similarly, the Japanese five vowel system can also be transcribed in IPA as
+The Japanese five vowel system can also be transcribed in IPA as
 /ieaou/, while the phonetic targets in this case are closer to [ieaoɯ].
@@
 Note also that there can be different systems of phonemic transcription for the
-same variety of a language, so two different resources on the `same' language
+same variety of a language, so two different resources on the ``same'' language
 might use different symbols that represent the same sound. The differences may
@@
 profile (see Chapter~\ref{orthography-profiles}), linking the selected phonemic 
-transcription symbols to `narrow' phonetic transcriptions, possibly also including 
+transcription symbols to narrow phonetic transcriptions, possibly also including 
 specifications of contextual interpretation.
@@
 ASJP.\footnote{See the ASJP use case in the online supplementary 
-materials to this book: \url{https://github.com/unicode-cookbook/use_cases}.} 
+materials to this book: \url{https://github.com/unicode-cookbook/recipes}.} 
 But before UPSID, SAMPA and ASJP, IPA was encoded with numbers.
@@
 on the keyboard to IPA symbols. The process of assigning standardized computer
-codes to phonetic symbols was assigned to the \textsc{workgroup on computer
+codes to phonetic symbols was given to the \textsc{workgroup on computer
 coding} (henceforth \textit{working group}) at the Kiel Convention. This working
@@
 
-The IPA Number is simply expressed as a three-digit number numerical directory
-of digit triples. The numbering scheme specifies three-digit codes,
-the first digit of which indicates the symbol's category
-\citep{Esling1990,EslingGaylord1993}:
+The IPA Number is expressed as a three-digit number. The first digit 
+indicates the symbol's category \citep{Esling1990,EslingGaylord1993}:
 
@@
 character encoding by mapping custom made fonts to specific code points.\footnote{For 
-example, SIL's popular font SIL IPA 1990}. However,
+example, SIL's popular font SIL IPA 1990.}. However,
 one linguist's electronic text would not render properly on another linguist's
@@
 One solution was the ASCII-ification of the IPA, which simply involved 
-defining keyboard-able sequences consisting of ASCII combinations as an IPA symbol
-codings. For example, \cite{Wells1987} provides an in-depth description of IPA
+defining keyboard-able sequences consisting of ASCII combinations as IPA symbols. 
+For example, \cite{Wells1987} provides an in-depth description of IPA
 codings from country-to-country. Later ASCII-IPAs include Kirshenbaum (created
@@
 \url{http://www.let.rug.nl/kleiweg/L04/} and \url{http://www.splitstree.org/},
-respectively}
+respectively.}
 
@@
 formal proposal to be submitted to various international standards
-organizations, several of which were trying to tackle (and in a sense `win') the
+organizations, several of which were trying to tackle (and in a sense win) the
 multilingual computing environment problem (cf. Section~\ref{encoding}).
@@
 assigned UCS codes in 16 bit representation (in hexadecimal) and were published
-Tables 2 and 3 in \cite{EslingGaylord1993}, which include a graphical
+as Tables 2 and 3 in \cite{EslingGaylord1993}, which include a graphical
 representation of the IPA Symbol, its IPA Name, phonetic description, IPA
@@
 With subsequent revisions to the IPA, one might have expected that the Unicode
-Consortium would update the Unicode Standard in a way that is inline with the
-development linguistic insights. However, updates that go against the principle
+Consortium would update the Unicode Standard in a way that is in line with the
+development of linguistic insights. However, updates that go against the principle
 of maintaining backwards compatibility lose out, i.e.\ it is more important to
-deal with the pitfalls created along the way then it is to change the standard.
+deal with the pitfalls created along the way than it is to change the standard.
 Therefore, many of the pitfalls we encounter today when using Unicode IPA are
@@
 editors and the Unicode Working Group (the predecessor to the Unicode Technical 
-Committee) worked together to lay out the the proposed standard's structure and 
+Committee) worked together to lay out the proposed standard's structure and 
 content. At this time, the proposal contained no code charts nor block descriptions.
@@
 1990.\footnote{\url{http://www.iso.org/iso/catalogue_detail.htm?csnumber=56921}} 
-The `Universal Multiple-Octet Coded Character Set' or simply UCS was the first 
+The \textit{Universal Multiple-Octet Coded Character Set} or simply UCS was the first 
 officially standardized character encoding with the aim of including all 
@@
 
-ISO/IEC 10646 is partly based on ISO/IEC 8859, a series of of ASCII-based 
+ISO/IEC 10646 is partly based on ISO/IEC 8859, a series of ASCII-based 
 standard character encodings published in 1987 that use a single bit 8-byte 

### 2018-01-13 book/chapters/orthography_profiles.tex
deals with chapters 2 and 3 -- wip

@@
 <\dia{0303}> + <\dia{0330}> or <o> + <\dia{0330}> + <\dia{0303}>. Although these two variants are visually homoglyphs,
-computationally they are different. Unicode normalization should be applied to
+computationally they are different (see Sections \ref{pitfall-ipa-homoglyphs} and \ref{pitfall-homoglyphs-in-IPA} ). 
+Unicode normalization should be applied to
 this string to reorder the code points into a canonical order, allowing the data
@@
 characters are underspecified for the direction in which they modify a host
-character. For example, can indicate either pre- or post-aspiration (whereas the
+character. For example, it can indicate either pre- or post-aspiration (whereas the
 nasalization or creaky diacritic is defined in the Unicode Standard to apply to
@@
 identified throughout the whole data (a step which we call
-\textsc{tokenization}), and possibly simply returned as such, possibly including
+\textsc{tokenization}), and simply returned as such, possibly including
 error messages about any parts of the data that are not specified by the
@@
 \subsection*{Implementation}
-For the automatic processing of the profiles, the following technical standards
+For the automated processing of the profiles, the following technical standards
 will be expected:
@@
         should be performed. 
-		\item \textsc{to treat the profile literal}, i.e.~to not interpret regular
+		\item \textsc{to treat the profile literally}, i.e.~to not interpret regular
         expression metacharacters. Matching graphemes literally often leads to

### 2018-02-04 book/chapters/practical_recommendations.tex
addresses issue #17

\chapter{Practical recommendations}
\label{practical-recommendations}

% A section which is missing in something called a "Cookbook" would be
% $ practical recommendations on how to input Unicode characters. There are
% various character selection tools, shortcuts on the keyboard, the
% shapecatcher website references at several places, or the Wikipedia
% lists of glyphs and fileformat.info. Having all this in one section
% would be handy for the user. It is of course unrelated to the
% orthography profiles, but I imagine that many people will use this book
% as a primer on IPA+Unicode and actually disregard the last two chapters.
% For this group, such a summary would be useful.

In this chapter, we provide some practical recommendations for quickly finding information about the Unicode Standard, the International Phonetic Alphabet (IPA), and tools and resources for working with each. This chapter is meant to be a short guide for novice users who are not interested in the programmatic aspects presented in the next two chapters. Instead, we aim here at ordinary working linguists who want to read about the basics and who want to easily insert special characters into their digital documents and applications.

\section{Unicode}
We discussed the Unicode Consortium's approach to computationally encoding writing systems in detail in Section \ref{the-unicode-approach}. The common pitfalls that we have encountered when using the Unicode Standard are discussed in detail in Chapter \ref{unicode-pitfalls}. Together these chapters provide users with an in-depth background about the hurdles they may encounter when using the Unicode Standard for encoding linguistic data or for developing multilingual applications. The provided inforation is by no means exhaustive and it is aimed at both linguists and programmers. For general background information about Unicode and character encodings, try these resources:

\begin{itemize}
	\item \url{http://www.unicode.org/standard/WhatIsUnicode.html}
	\item \url{https://en.wikipedia.org/wiki/Unicode}
	\item \url{https://www.w3.org/International/articles/definitions-characters/}
\end{itemize}

% \section{Unicode character pickers}
For practical purposes, users need a way to insert special characters (i.e.\ characters that are not easily entered via their keyboards) into documents and software applications. There are a few basic approaches for inserting special characters. One way is to use software-specific functionality, when it is available. For example, Microsoft Word has an insert-special-symbol-or-character function that allows users to scroll through a table of special characters across different scripts. Special characters can be then inserted into the document by clicking on them. Another way is to install a system-wide application for special character insertion. We have long been fans of the PopChar application from Ergonis Software, which is a small program that can insert Unicode characters.\footnote{\url{http://www.ergonis.com/products/popcharx/}}

There are also web-based so-called Unicode character pickers available through the browser that allow for the creation and insert of special characters, which can then be copied \& pasted into documents or software applications. For example:

\begin{itemize}
	\item \url{https://unicode-table.com/en/}
	\item \url{https://r12a.github.io/pickers/}
\end{itemize}

Yet another option for special character insertion includes operating system-specific shortcuts. For example on the Mac, holding down a key on the keyboard for a second, say <u>, triggers a pop up with the options <û, ü, ù, ú, ū> which can then be chosen by number (1--5). This method is convenient for occasionally inserting special characters, but it is burdensome for inserting many characters.

Therefore, there are many applications for installing language-specific or script-specific keyboards. These programs typically override keys or keystrokes on the user's keyboard and therefore allow them to quickly keyboard the insertion of specal characters (once the layout of the new keyboard is mastered). Many operating systems provide such keyboards as part of the base operating system, e.g.\ Microsoft Windows, Linux, and OS X, have special character Unicode eyboards pre-installed. Seach for how to access these pre-installed options on your operating system.

Lastly, there are third party applications that provide virtual keyboards. They can be language-specific or devoted specifically to IPA. Two popular programs are:

\begin{itemize}
	\item \url{https://keyman.com/}
	\item \url{http://scripts.sil.org/ipa-sil_keyboard}
\end{itemize}

\section{IPA}
In Chapter \ref{the-international-phonetic-alphabet} we described in detail the history of the International Phonetic Alphabet and how it became encoded in the Unicode Standard. In Chapter \ref{ipa-meets-unicode} we describe the resulting pitfalls from their marriage. These two chapters provide a very detailed overview of the challenges that users face when working with the two standards.

For general information about the International Phonetic Alphabet, the standard text is the \textit{Handbook of the International Phonetic Association: A Guide to the Use of the International Phonetic Alphabet} \cite{IPA2007}. It described in detail the principles and premises of the IPA that we summarize in Section \ref{IPApremises-principles}. The Association makes available information about itself  online\footnote{\url{https://www.internationalphoneticassociation.org/}} and it also makes available the International Phonetic Alphabet in chart form.\footnote{\url{https://www.internationalphoneticassociation.org/content/ipa-chart}} Wikipedia also has a comprehensive article about the IPA.\footnote{\url{https://en.wikipedia.org/wiki/International_Phonetic_Alphabet}}

There are several good Unicode IPA character pickers available on the Web, including:

\begin{itemize}
	\item \url{https://r12a.github.io/pickers/ipa/}
	\item \url{https://westonruter.github.io/ipa-chart/keyboard/}
	\item \url{http://ipa.typeit.org/}
\end{itemize}

\noindent Various linguistics departments also provide information about IPA fonts, software, and inserting Unicode IPA characters. Some good ones are:

\begin{itemize}
	\item \url{http://www.phon.ucl.ac.uk/resource/phonetics/}
	\item \url{https://www.york.ac.uk/language/current/resources/freeware/ipa-fonts-and-software/}
\end{itemize}

Regarding fonts that display Unicode IPA correctly, many linguists turn to the IPA Unicode fonts developed by SIL Internnational. The complete SIL font list is available online.\footnote{\url{http://scripts.sil.org/SILFontList}} There is also a page that describes IPA transcription using the SIL fonts and provides an informative discussion on deciding which font to use.\footnote{\url{http://scripts.sil.org/ipahome}} Traditionally, IPA fonts popular with linguists were created and maintained by SIL International, so it is often the case in our experience that we encounter linguistics data in legacy IPA fonts, i.e.\ pre-Unicode fonts such as SIL IPA93.\footnote{\url{http://scripts.sil.org/FontFAQ_IPA93}} SIL International does a good job of describing how to convert from legacy IPA fonts to Unicode IPA. By far the most popular fonts by SIL International are Doulos SIL and Charis SIL:

\begin{itemize}
	\item \url{https://software.sil.org/doulos/}}
	\item \url{https://software.sil.org/charis/}
\end{itemize}

\section{Advanced topics}
If you have made it this far, and you are eager to know more about the technological aspects of the Unicode Standard and how they relate to software programming, we recommend two light-hearted blog posts on the topic. The classic blog post about what programmers should know about the Unicode Standard is Joel Spolsky's ``The Absolute Minimum Every Software Developer Absolutely, Positively Must Know About Unicode and Character Sets (No Excuses!)''.\footnote{\url{https://www.joelonsoftware.com/2003/10/08/the-absolute-minimum-every-software-developer-absolutely-positively-must-know-about-unicode-and-character-sets-no-excuses/}} A more recent blogpost, with a bit more of the technical details, is by David C. Zentgraf and is titled, ``What Every Programmer Absolutely, Positively Needs To Know About Encodings And Character Sets To Work With Text''.\footnote{\url{http://kunststube.net/encoding/}} This post is aimed at software developers and uses the PHP language for examples.

Here are some online resources that we find particularly useful for finding more information about individual Unicode characters and also for converting between encodings:

\begin{itemize}
	\item \url{http://www.fileformat.info/}
	\item \url{https://unicodelookup.com/}
	\item \url{https://r12a.github.io/scripts/featurelist/}
	\item \url{https://r12a.github.io/app-conversion/}
\end{itemize}

For users of Python, see the standard documentation on how to use Unicode in your programming applications.\footnote{\url{https://docs.python.org/3/howto/unicode.html}} For \latex users, the TIPA package is useful for inserting IPA characters into your typeset documents. See these resources:

\begin{itemize}
	\item \url{http://www.tug.org/tugboat/tb17-2/tb51rei.pdf}
	\item \url{https://ctan.org/pkg/tipa}
	\item \url{http://ptmartins.info/tex/tipacheatsheet.pdf}
\end{itemize}

Lastly, we leave you with some Unicode humor for making it this far:

\begin{itemize}
	\item \url{https://xkcd.com/380/}
	\item \url{https://xkcd.com/1137/}
	\item \url{http://www.commitstrip.com/en/2014/06/17/unicode-7-et-ses-nouveaux-emoji/}
	\item \url{http://www.i18nguy.com/humor/unicode-haiku.html}
\end{itemize}

### 2018-02-13 book/chapters/practical_recommendations.tex
closes #17

@@
 
-In this chapter, we provide some practical recommendations for quickly finding information about the Unicode Standard, the International Phonetic Alphabet (IPA), and tools and resources for working with each.
-This chapter is meant to be a short guide for novice users who are not interested in the programmatic aspects presented in the next two chapters.
-Instead, we aim here at ordinary working linguists who want to read about the basics and who want to easily insert special characters into their digital documents and applications.
+This chapter is meant to be a short guide for novice users who are not interested in the programmatic aspects presented in Chapters \ref{orthography-profiles} and \ref{implementation}.
+Instead, we provide links to quickly find general information about the Unicode Standard and the International Phonetic Alphabet (IPA).
+And we target ordinary working linguists who want to know how to easily insert special characters into their digital documents and applications.
 
 \section{Unicode}
-We discussed the Unicode Consortium's approach to computationally encoding writing systems in detail in Section \ref{the-unicode-approach}.
+We discussed the Unicode Consortium's approach to computationally encoding writing systems in Chapter \ref{the-unicode-approach}.
 The common pitfalls that we have encountered when using the Unicode Standard are discussed in detail in Chapter \ref{unicode-pitfalls}.
-Together these chapters provide users with an in-depth background about the hurdles they may encounter when using the Unicode Standard for encoding linguistic data or for developing multilingual applications.
-The provided inforation is by no means exhaustive and it is aimed at both linguists and programmers.
-For general background information about Unicode and character encodings, try these resources:
+Together these chapters provide users with an in-depth background about the hurdles they may encounter when using the Unicode Standard for encoding their data or for developing multilingual applications.
+For general background information about Unicode and character encodings, see these resources:
 
@@
 Another way is to install a system-wide application for special character insertion.
-We have long been fans of the PopChar application from Ergonis Software, which is a small program that can insert Unicode characters.\footnote{\url{http://www.ergonis.com/products/popcharx/}}
+We have long been fans of the PopChar application from Ergonis Software, which is a small program that can insert most Unicode characters (note however that the full version requires a paid subscription).\footnote{\url{http://www.ergonis.com/products/popcharx/}}
 
-There are also web-based so-called Unicode character pickers available through the browser that allow for the creation and insert of special characters, which can then be copied \& pasted into documents or software applications.
-For example:
+There are also Web-based Unicode character pickers available through the browser that allow for the creation and insert of special characters, which can then be copied \& pasted into documents or software applications.
+For example, try:
 
@@
 Yet another option for special character insertion includes operating system-specific shortcuts.
-For example on the Mac, holding down a key on the keyboard for a second, say <u>, triggers a pop up with the options <û, ü, ù, ú, ū> which can then be chosen by number (1--5).
-This method is convenient for occasionally inserting special characters, but it is burdensome for inserting many characters.
+For example on the Mac, holding down a key on the keyboard for a second, say <u>, triggers a pop up with the options <û, ü, ù, ú, ū> which can then be inserted by keying the associated number (1--5).
+This method is convenient for occasionally inserting type accented characters, but the full range of special characters is limited and this method is burdensome for rapidly inserting many different characters.
+For complete access to special characters, Mac provides a Keyboard Viewer application available in the Keyboard pane of the System Preferences.
 
-Therefore, there are many applications for installing language-specific or script-specific keyboards.
-These programs typically override keys or keystrokes on the user's keyboard and therefore allow them to quickly keyboard the insertion of specal characters (once the layout of the new keyboard is mastered).
-Many operating systems provide such keyboards as part of the base operating system, e.g.\ Microsoft Windows, Linux, and OS X, have special character Unicode eyboards pre-installed.
-Seach for how to access these pre-installed options on your operating system.
+On Windows, accented characters can be inserted by using alt-key shortcuts, i.e.\ holding down the alt-key and keying in a sequence of numbers (which typically reflect the Unicode character's decimal representation).
+For example, \textsc{latin small letter c with cedilla} at \uni{00E7} with the decimal code 231 can be inserted by holding the alt-key and keying the sequence 0231.
+Again, this method is burdensome for rapidly inserting characters.
+For access to the full range of Unicode characters, the Character Map program comes preinstalled on all Microsoft Operating systems.
 
-Lastly, there are third party applications that provide virtual keyboards. They can be language-specific or devoted specifically to IPA. Two popular programs are:
+There are also many third party applications that provide virtual keyboards.
+These programs typically override keys or keystrokes on the user's keyboard  allowing them to quickly keyboard special characters (once the layout of the new keyboard is mastered).
+They can be language-specific or devoted specifically to IPA.
+Two popular programs are:
 
@@
 \section{IPA}
-In Chapter \ref{the-international-phonetic-alphabet} we described in detail the history of the International Phonetic Alphabet and how it became encoded in the Unicode Standard.
+In Chapter \ref{the-international-phonetic-alphabet} we described in detail the history and principles of the International Phonetic Alphabet (IPA) and how it became encoded in the Unicode Standard.
 In Chapter \ref{ipa-meets-unicode} we describe the resulting pitfalls from their marriage.
-These two chapters provide a very detailed overview of the challenges that users face when working with the two standards.
+These two chapters provide a detailed overview of the challenges that users face when working with the two standards.
 
-For general information about the International Phonetic Alphabet, the standard text is the \textit{Handbook of the International Phonetic Association: A Guide to the Use of the International Phonetic Alphabet} \cite{IPA2007}.
-It described in detail the principles and premises of the IPA that we summarize in Section \ref{IPApremises-principles}.
-The Association makes available information about itself  online\footnote{\url{https://www.internationalphoneticassociation.org/}} and it also makes available the International Phonetic Alphabet in chart form.\footnote{\url{https://www.internationalphoneticassociation.org/content/ipa-chart}} Wikipedia also has a comprehensive article about the IPA.\footnote{\url{https://en.wikipedia.org/wiki/International_Phonetic_Alphabet}}
+For general information about the IPA, the standard text is the \textit{Handbook of the International Phonetic Association: A Guide to the Use of the International Phonetic Alphabet} \citep{IPA2007}.
+The handbook describes in detail the principles and premises of the IPA, which we have summarized in Section \ref{IPApremises-principles}.
+The handbook also provides many examples of how to use the IPA.
+The Association also makes available information about itself online\footnote{\url{https://www.internationalphoneticassociation.org/}} and it provides the most current IPA charts.\footnote{\url{https://www.internationalphoneticassociation.org/content/ipa-chart}} Wikipedia also has a comprehensive article about the IPA.\footnote{\url{https://en.wikipedia.org/wiki/International_Phonetic_Alphabet}}
 
-There are several good Unicode IPA character pickers available on the Web, including:
+There are several good Unicode IPA character pickers available on the Web and through the browser, including:
 
@@
 
-\noindent Various linguistics departments also provide information about IPA fonts, software, and inserting Unicode IPA characters. Some good ones are:
+\noindent Various linguistics departments also provide information about IPA fonts, software, and inserting Unicode IPA characters. Two useful resources are:
 
@@
 The complete SIL font list is available online.\footnote{\url{http://scripts.sil.org/SILFontList}} There is also a page that describes IPA transcription using the SIL fonts and provides an informative discussion on deciding which font to use.\footnote{\url{http://scripts.sil.org/ipahome}} Traditionally, IPA fonts popular with linguists were created and maintained by SIL International, so it is often the case in our experience that we encounter linguistics data in legacy IPA fonts, i.e.\ pre-Unicode fonts such as SIL IPA93.\footnote{\url{http://scripts.sil.org/FontFAQ_IPA93}} SIL International does a good job of describing how to convert from legacy IPA fonts to Unicode IPA.
-By far the most popular fonts by SIL International are Doulos SIL and Charis SIL:
+The most popular Unicode IPA fonts are Doulos SIL and Charis SIL:
 
@@
 
-\section{Advanced topics}
-If you have made it this far, and you are eager to know more about the technological aspects of the Unicode Standard and how they relate to software programming, we recommend two light-hearted blog posts on the topic.
-The classic blog post about what programmers should know about the Unicode Standard is Joel Spolsky's ``The Absolute Minimum Every Software Developer Absolutely, Positively Must Know About Unicode and Character Sets (No Excuses!)''.\footnote{\url{https://www.joelonsoftware.com/2003/10/08/the-absolute-minimum-every-software-developer-absolutely-positively-must-know-about-unicode-and-character-sets-no-excuses/}} A more recent blogpost, with a bit more of the technical details, is by David C.
-Zentgraf and is titled, ``What Every Programmer Absolutely, Positively Needs To Know About Encodings And Character Sets To Work With Text''.\footnote{\url{http://kunststube.net/encoding/}} This post is aimed at software developers and uses the PHP language for examples.
-
-Here are some online resources that we find particularly useful for finding more information about individual Unicode characters and also for converting between encodings:
+Lastly, here are some online resources that we find particularly useful for finding more information about individual Unicode characters and also for converting between encodings:
 
@@
 
-For users of Python, see the standard documentation on how to use Unicode in your programming applications.\footnote{\url{https://docs.python.org/3/howto/unicode.html}} For \latex users, the TIPA package is useful for inserting IPA characters into your typeset documents.
+\section{For programmers and potential programmers}
+If you have made it this far, and you are eager to know more about the technological aspects of the Unicode Standard and how they relate to software programming, we recommend two light-hearted blog posts on the topic.
+The classic blog post about what programmers should know about the Unicode Standard is Joel Spolsky's \textit{The Absolute Minimum Every Software Developer Absolutely, Positively Must Know About Unicode and Character Sets (No Excuses!)}.\footnote{\url{https://www.joelonsoftware.com/2003/10/08/the-absolute-minimum-every-software-developer-absolutely-positively-must-know-about-unicode-and-character-sets-no-excuses/}} A more recent blogpost, with a bit more of the technical details, is by David C.
+Zentgraf and is titled, \textit{What Every Programmer Absolutely, Positively Needs To Know About Encodings And Character Sets To Work With Text}.\footnote{\url{http://kunststube.net/encoding/}} This post is aimed at software developers and uses the PHP language for examples.
+
+For users of Python, see the standard documentation on how to use Unicode in your programming applications.\footnote{\url{https://docs.python.org/3/howto/unicode.html}} For R users we recommend the \textsc{stringi} library.\footnote{\url{https://cran.r-project.org/web/packages/stringi/index.html}} For \LaTeX users, the TIPA package is useful for inserting IPA characters into your typeset documents.
 See these resources:
@@
 \end{itemize}
+
+\noindent But we find it much easier to use the Unicode-aware XeTeX/XeLaTeX typesetting system.\footnote{\url{http://xetex.sourceforge.net/}} Unicode characters can be directly inserted into your Tex documents and compiled into typeset PDF with XeLaTeX.
 

### 2018-02-13 book/unicode.tex
Practical implementations (#27)
* addresses issue #17

* closes #17

--- 2017-02-02 book/unicode.tex
@@
 \include{chapters/ipa_meets_unicode}
+\include{chapters/practical_recommendations}
 \include{chapters/orthography_profiles}
@@
 ## $profile
-##   matched_rules Grapheme
-## 1             3        B
-## 2             3        A
+##   Grapheme Frequency
+## 1        B         3
+## 2        A         3
 ## 
@@
 
-% latex table generated in R 3.3.2 by xtable 1.8-2 package
-% Thu Feb  2 11:38:04 2017
+% latex table generated in R 3.4.3 by xtable 1.8-2 package
+% Sat Feb  3 16:28:50 2018
 \begin{table}[H]
@@
 
-% latex table generated in R 3.3.2 by xtable 1.8-2 package
-% Thu Feb  2 11:38:04 2017
+% latex table generated in R 3.4.3 by xtable 1.8-2 package
+% Sat Feb  3 16:28:51 2018
 \begin{table}[H]
@@
 
-% latex table generated in R 3.3.2 by xtable 1.8-2 package
-% Thu Feb  2 11:38:04 2017
+% latex table generated in R 3.4.3 by xtable 1.8-2 package
+% Sat Feb  3 16:28:51 2018
 \begin{table}[htb]
@@
 
-% latex table generated in R 3.3.2 by xtable 1.8-2 package
-% Thu Feb  2 11:38:05 2017
+% latex table generated in R 3.4.3 by xtable 1.8-2 package
+% Sat Feb  3 16:28:51 2018
 \begin{table}[htb]

### 2018-02-13 book/chapters/introduction.tex
keeps capitalization in line
this change deals with reviewer 1 s suggestion to split out The Unicode approach into its own chapter

@@
-\chapter{Writing Systems}
+\chapter{Writing systems}
 \label{writing_systems}
@@
 
-\
+\subsubsection*{Summary}
 
@@
 definitions developed for the computational processing of the world's writing
-systems, which we describe in the next section.
-
-% ==========================
-\section{The Unicode approach}
-\label{the-unicode-approach}
-% ==========================
-
-The conceptualization and terminology of writing systems was rejuvenated by
-the development of the Unicode Standard, with major input from Mark Davis,
-co-founder and long-term president of the Unicode Consortium. For many years,
-``exotic'' writing systems and phonetic transcription systems on personal
-computers were constrained by the American Standard Code for Information
-Interchange (ASCII) character encoding scheme, based on the Latin script, which
-only allowed for a strongly limited number of different symbols to be encoded.
-This implied that users could either use and adopt the (extended) Latin alphabet
-or they could assign new symbols to the small number of code points in the ASCII
-encoding scheme to be rendered by a specifically designed font
-\citep{BirdSimons2003}. In this situation, it was necessary to specify the font
-together with each document to ensure the rightful display of its content. To
-alleviate this problem of assigning different symbols to the same code points,
-in the late 80's and early 90's the Unicode Consortium set itself the ambitious
-goal of developing a single universal character encoding to provide a unique
-number, a code point, for every character in the world's writing systems.
-Nowadays, the Unicode Standard is the default encoding of the technologies that
-support the World Wide Web and for all modern operating systems, software and
-programming languages.
-
-\subsubsection*{The Unicode Standard}
-
-The Unicode Standard represents a massive step forward because it aims to
-eradicate the distinction between universal (ASCII) versus language-particular
-(font) by adding as much as possible language-specific information into the
-universal standard. However, there are still language/resource-specific
-specifications necessary for the proper usage of Unicode, as will be discussed
-below. Within the Unicode structure many of these specifications can be captured
-by so-called \textsc{locale descriptions}, so we are moving to a new distinction
-of universal (Unicode Standard) versus language-particular (locale description).
-The major gain is a much larger compatibility on the universal level (because
-Unicode standardizes a much larger portion of writing system diversity), and
-much better possibilities for automated processing on the language-particular
-level (because locale descriptions are computer readable specifications).
-
-Each version of the Unicode Standard (\citealp{Unicode2014}, as of writing at
-version 7) consists of a set of specifications and guidelines that include (i) a
-core specification, (ii) code charts, (iii) standard annexes and (iv) a
-character database.\footnote{All documents of the Unicode Standard are available
-at \url{http://www.unicode.org/versions/latest/}. For a quick survey of the use
-of terminology inside the Unicode Standard, their glossary is particularly
-useful, available at \url{http://www.unicode.org/glossary/}. For a general
-introduction to the principles of Unicode, Chapter 2 of the core specification,
-called \textsc{general structure}, is particularly insightful. Different from
-many other documents of the Unicode Standard, this general introduction is
-relatively easy to read and illustrated with many interesting examples from
-various orthography traditions all over the world.} The \textsc{core
-specification} is a book directed toward human readers that describes the formal
-standard for encoding multilingual text. The \textsc{code charts} provide a
-humanly readable online reference to the character contents of the Unicode
-Standard in the form of PDF files. The \textsc{Unicode Standard Annexes (UAX)}
-are a set of technical standards that describe the implementation of the Unicode
-Standard for software development, Web standards, and programming languages.
-The \textsc{Unicode Character Database (UCD)} is a set of computer-readable text
-files that describe the character properties, including a set of rich character
-and writing system semantics, for each character in the Unicode Standard. In
-this section, we introduce the basic Unicode concepts, but we will leave out
-many details. Please consult the above mentioned full documentation for a more
-detailed discussion. Further note that the Unicode Standard is exactly that,
-namely a standard. It normatively describes notions and rules to be followed. In
-the actual practice of applying this standard in a computational setting, a
-specific implementation is necessary. The most widely used implementation of the
-Unicode Standard is the \textsc{International Components for Unicode (ICU)},
-which offers C/C++ and Java libraries implementing the Unicode
-Standard.\footnote{More information about the ICU is available here:
-\url{http://icu-project.org}.}
-
-\subsubsection*{Character encoding system}
-
-The Unicode Standard is a \textsc{character encoding system} which
-goal it is to support the interchange and processing of written characters and
-text in a computational setting. Underlyingly, the character encoding is
-represented by a range of numerical values called a \textsc{code space}, which
-is used to encode a set of characters. A \textsc{code point} is a unique
-non-negative integer within a code space (i.e.~within a certain numerical
-range). In the Unicode Standard character encoding system, an \textsc{abstract
-character}, for example the \textsc{latin small letter p}, is mapped to a
-particular code point, in this case the decimal value 112, normally represented in
-hexadecimal, which then looks in Unicode parlance as
-\uni{0070}.\footnote{Hexadecimal (base-16) 0070 is equivalent to decimal
-(base-10) 112, which can be calculated by considering that $(0\cdot16^3) +
-(0\cdot16^2) + (7\cdot16^1) + (0\cdot16^0) = 7\cdot16 = 112$. Underlyingly,
-computers will of course treat this code point binary (base-2) as 11100000, as
-can be seen by calculating that $(1\cdot2^7) + (1\cdot2^6) + (1\cdot2^5) +
-(0\cdot2^4) + (0\cdot2^3) + (0\cdot2^2) + (0\cdot2^1) + (0\cdot2^0) = 64 + 32 +
-16 = 112$.} That encoded abstract character is rendered on a computer screen (or
-printed page) as a \textsc{glyph}, e.g. <p>, depending on the \textsc{font} and
-the context in which that character appears.
-
-In Unicode Standard terminology, an (abstract) \textsc{character} is the basic
-encoding unit. The term \textsc{character} can be quite confusing due to its
-alternative definitions across different scientific disciplines and because in
-general the word \textsc{character} means many different things to different
-people. It is therefore often preferable to refer to Unicode characters simply
-as \textsc{code points}, because there is a one-to-one mapping between Unicode
-characters and their numeric representation. In the Unicode approach, a
-character refers to the abstract meaning and/or general shape, rather than a
-specific shape, though in code tables some form of visual representation is
-essential for the reader's understanding. Unicode defines characters as
-abstractions of orthographic symbols, and it does not define visualizations for
-these characters (although it does presents examples). In contrast, a
-\textsc{glyph} is a concrete graphical representation of a character as it
-appears when rendered (or rasterized) and displayed on an electronic device or
-on printed paper. For example, <g {\large \textit{g}} \textbf{g}
-{\fontspec{ArialMT} {\small g} \textit{g} \textbf{g}}> are different glyphs of the
-same character, i.e.~they may be rendered differently depending on the
-typography being used, but they all share the same code point. From the
-perspective of Unicode they are \textit{the same thing}. In this approach, a
-\textsc{font} is then simply a collection of glyphs connected to code points.
-Allography is not specified in Unicode (expect for a few exceptional cases, due
-to legacy encoding issues), but can be specified in a font as a
-\textsc{contextual variant} (a.k.a.~presentation form).
-
-Each code point in the Unicode Standard is associated with a set of
-\textsc{character properties} as defined by the Unicode character property
-model.\footnote{The character property model is described in
-\url{http://www.unicode.org/reports/tr23/}, but the actual properties are
-described in \url{http://www.unicode.org/reports/tr44/}. A simplified overview
-of the properties is available at
-\url{http://userguide.icu-project.org/strings/properties}. The actual code
-tables listing all properties for all Unicode code points are available at
-\url{http://www.unicode.org/Public/UCD/latest/ucd/}.} Basically, those
-properties are just a long list of values for each character. For example, code
-point \uni{0047} has the following properties (among many others): 
-\begin{itemize}
-	\item Name: LATIN CAPITAL LETTER G 
-	\item Alphabetic: YES 
-	\item Uppercase: YES 
-	\item Script: LATIN 
-	\item Extender: NO 
-	\item Simple\_Lowercase\_Mapping: 0067 
-\end{itemize}
-
-These properties contain the basic information of the Unicode Standard and they
-are necessary to define the correct behavior and conformance required for
-interoperability in and across different software implementations (as defined in
-the Unicode Standard Annexes). The character properties assigned to each code
-point is based on each character's behavior in the real-world writing
-traditions. For example, the corresponding lowercase character to \uni{0047} is
-\uni{0067}.\footnote{Note that the relation between uppercase and lowercase is in
-many situations much more complex than this, and Unicode has further
-specifications for those cases.} Another use of properties is to define the
-script of a character.\footnote{The Unicode Glossary defines the term \textsc{script} as
-a ``collection of letters and other written signs used to represent textual
-information in one or more writing systems. For example, Russian is written with
-a subset of the Cyrillic script; Ukrainian is written with a different subset.
-The Japanese writing system uses several scripts.''} In practice, script is
-simply defined for each character as the explicit \textsc{script} property in
-the Unicode Character Database.
-
-One frequently referenced property is the \textsc{block} property, which is
-often used in software applications to impose some structure to the large number
-of Unicode characters. Each character in Unicode belongs to a specific block.
-These blocks are basically an organizational structure to alleviate the
-administrative burden of keeping Unicode up-to-date. Blocks consist of
-characters that in some way belong together, so that characters are easier to
-find. Some blocks are connected with a specific script, like the Hebrew block or
-the Gujarati block. However, blocks are predefined ranges of code points, and
-often there will come a point after which the range is completely filled. Any
-extra characters will have to be assigned somewhere else. There is, for example,
-a block \textsc{Arabic}, which contains most Arabic symbols. However, there is
-also a block \textsc{Arabic Supplement}, \textsc{Arabic Presentation Forms-A}
-and \textsc{Arabic Presentation Form-B}. The situation with Latin symbols is
-even more extreme. In general, the names for block should not be taken as a
-definitional statement. For example, many IPA symbols are not located in the
-aptly-named block \textsc{IPA extensions}, but in other blocks
-(see Section~\ref{pitfall-no-complete-ipa-block}).
-
-\subsubsection*{Grapheme clusters}
-
-There are many cases in which a sequence of characters (i.e.~a sequence of more
-than one code point) represents what a user perceives as an individual unit in a
-particular orthographic writing system. For this reason the Unicode Standard
-differentiates between \textsc{abstract character} and \textsc{user-perceived
-character}. Sequences of multiple code points that correspond to a single
-user-perceived characters are called grapheme clusters in Unicode parlance.
-Grapheme clusters come in two flavors: (default) grapheme clusters and tailored
-grapheme clusters.
-
-The (default) \textsc{grapheme clusters} are locale-independent graphemes,
-i.e.~they always apply when a particular combination of characters occurs
-independent of the writing system in which they are used. These character
-combinations are defined in the Unicode Standard as functioning as one
-\textsc{text element}.\footnote{The Unicode Glossary defines text element as:
-``A minimum unit of text in relation to a particular text process, in the
-context of a given writing system. In general, the mapping between text elements
-and code points is many-to-many.''} The simplest example of a grapheme cluster
-is a base character followed by a letter modifier character. For example, the
-sequence <n> + <\dia{0303}> (i.e.~\textsc{latin small letter n} at \uni{006E}, followed
-by \textsc{combining tilde} at \uni{0303}) combines visually into <ñ>, a
-user-perceived character in writing systems like that of Spanish. So, what the
-user perceives as a single character actually involves a multi-code-point
-sequence. Note that this specific sequence can also be represented with a single
-so-called \textsc{precomposed code point}, \textsc{latin small letter n with
-tilde} at \uni{00F1}, but this is not the case for all multi-code point
-character sequences. The problem that there multiple encodings possible for the
-same text element has been acknowledged early on in the Unicode Standard
-(e.g.~for <ñ>, the sequence \uni{006E} \uni{0303} should in all situations be
-treated identically to the precomposed \uni{00F1}), and a system of
-\textsc{canonical equivalence} is available for such situations. Basically, the
-Unicode Standard offers different kind of normalizations to either decompose all
-precomposed characters (called \textsc{NFD}, \textsc{normalization form
-canonical decomposition}), or compose as much as possible combinations (called
-\textsc{NFC}, \textsc{normalization form canonical composition}). In current
-practice of software development, NFC seems to be preferred in most situations
-and is widely proposed as the preferred canonical form.
-
-More difficult for text processing, because less standardized, is what the
-Unicode Standard terms \textsc{tailored grapheme clusters}. Tailored grapheme
-clusters are locale-dependent graphemes, i.e.~such combination of characters do
-not function as text elements in all situations. Examples include the sequence
-<c>~+~<h> for the Slovak digraph <ch> and the sequence <ky> in the Sisaala
-practical orthography (pronounced as IPA /tʃ/,~\cite{Moran2006}). These grapheme
-clusters are \textsc{tailored} in the sense that they must be specified on a
-language-by-language or writing-system-by-writing-system basis. The Unicode
-Standard provides technological specifications for creating locale specific data
-in so-called \textsc{Unicode Locale Descriptions}, i.e.~a set of specification
-that defines a set of language-specific elements (e.g.~tailored grapheme
-clusters, collation order, capitalization-equivalence), as well as other special
-information, like how to format numbers, dates, or currencies. Locale
-descriptions are saved in the \textsc{Common Locale Data Repository
-(CLDR)},\footnote{More information about the CLDR can be found here:
-\url{http://cldr.unicode.org/}.} a repository of
-language-specific definitions of writing system properties, each of which
-describes specific usages of characters. Each locale can be encoded in a
-document using the \textsc{Locale Data Markup Language (LDML)}. LDML is an XML
-format and vocabulary for the exchange of structured locale data. Unicode Locale
-Descriptions allow users to define language- or even resource-specific writing
-systems or orthographies.\footnote{The Unicode Glossary defines \textsc{writing
-system} only very loosely, as it is not a central concept in the Unicode
-Standard. A writing system is, ``A set of rules for using one or more scripts to
-write a particular language. Examples include the American English writing
-system, the British English writing system, the French writing system, and the
-Japanese writing system.''} However, there are various drawbacks of locale
-descriptions for the daily practice of linguistic work in a multilingual setting.
+systems, which we describe in the next Chapter.

### 2018-02-13 book/chapters/unicode_approach.tex
this change deals with reviewer 1 s suggestion to split out The Unicode approach into its own chapter

--- 2015-06-20 chapters/introduction.tex
@@
-\chapter{Writing Systems}
-\label{writing_systems}
-
 % ==========================
-\section{Introduction}
-\label{introduction}
+\chapter{The Unicode approach}
+\label{the-unicode-approach}
 % ==========================
 
-Writing systems arise and develop in a complex mixture of cultural,
-technological and practical pressures. They tend to be highly conservative, in
-that people who have learned to read and write in a specific way---however
-impractical or tedious---are mostly unwilling to change their habits. Writers
-tend to resist spelling reforms. In all literate societies there exists a strong
-socio-political mainstream that tries to force unification of writing (for
-example by strongly enforcing ``right'' from ``wrong'' writing in schools).
-However, there is also a large community of users who take as many liberties in
-their writing as they can get away with.
-
-For example, the writing of tone diacritics in Yoruba is often proclaimed to be
-the right way to write, although many users of Yoruba orthography seem to be
-perfectly fine with leaving them out. As pointed out by the proponents of the
-official rules, there are some homographs when leaving out the tone diacritics
-\citep[44]{Olumuyiw2013}. However, writing systems (and the languages they
-represent) are normally full of homophones, which is normally not a problem at
-all for speakers of the language. More importantly, writing is not just a purely
-functional tool, but just as importantly it is a mechanism to signal social
-affiliation. By showing that you \textit{know the rules} of expressing yourself
-in writing, others will more easily accept you as a worthy participant in their
-group. And that just as well holds for obeying to the official rules when
-writing a job application, as for obeying to the informal rules when writing an
-SMS to classmates in school. The case of Yoruba writing is an exemplary case, as
-even after more than a century of efforts to standardize the writing systems,
-there is still a wide range of variation in daily use \citep{Olumuyiw2013}.
-
-The sometimes cumbersome and sometimes illogical structure, and the enormous
-variability of existing writing systems is a fact of life scholars have to
-accept and should try to adapt to as good as possible. Our plea here is a
-proposal for a formalization to do exactly that.
-
-When considering the worldwide linguistic diversity, including all
-lesser-studied and endangered languages, there exist numerous different
-orthographies using symbols from the same scripts. For example, there are
-hundreds of orthographies using Latin-based alphabetic scripts. All of these
-orthographies use the same symbols, but these symbols differ in meaning and
-usage throughout the various orthographies. To be able to computationally use
-and compare different orthographies, we need a way to specify all orthographic
-idiosyncrasies in a computer-readable format (a process called
-\textsc{tailoring} in Unicode parlance). We call such specifications
-\textsc{orthography profiles}. Ideally, these specifications have to be
-integrated into so-called Unicode locale descriptions, though we will argue that
-in practice this is often not the most useful solution for the kind of problems
-arising in the daily practice of linguistics. Consequently, a central goal of
-this paper is to flesh out the linguistic challenges for locale descriptions,
-and work out suggestions to improve their structure for usage in a linguistic
-context. Conversely, we also aim to improve linguists' understanding and
-appreciation for the accomplishments of the Unicode Consortium in the
-development of the Unicode Standard.
-
-The necessity to computationally use and compare different orthographies most
-forcefully arises in the context of language comparison. Concretely, in our
-current research our goal is to develop quantitative methods for language
-comparison and historical analysis in order to investigate worldwide linguistic
-variation and to model the historical and areal processes that underlie
-linguistic diversity,
-cf.~\citet{Steiner_etal2011,List2012,List2012a,ListMoran2013,MoranProkic2013}.
-In this work, it is crucial to be able to flexibly process across numerous
-resources with different orthographies. In many cases even different resources
-on the \textit{same} language use different orthographic conventions. Another
-orthographic challenge that we encounter regularly in our linguistic practice is
-electronic resources on a particular language that claim to follow a specific
-orthographic convention (often a resource-specific convention), but on closer
-inspection such resources are almost always not consistently encoded. Thus a
-second goal of our orthography profiles is to allow for an easy specification of
-orthographic conventions, and use such profiles to check consistency and to
-report errors to be corrected.
-
-A central step in our proposed solution to this problem is the tailored grapheme
-separation of strings of symbols, a process we call \textsc{grapheme
-tokenization}. Basically, given some strings of symbols (e.g.~morphemes, words,
-sentences) in a specific source, our first processing step is to specify how
-these strings have to be separated into graphemes, considering the specific
-orthographic conventions used in a particular source document. Our experience is
-that such a graphemic tokenization can be performed without extensive in-depth
-knowledge about the phonetic and phonological details of the language in
-question. For example, the specification that $<$ou$>$ is a grapheme of English
-is a much easier task than to specify what exactly the phonetic values of this
-grapheme are in any specific occurrence in English words. Grapheme separation is
-a task that can be performed relatively reliably and with limited availability
-of time and resources (compare, for example, the task of creating a complete
-phonetic or phonological normalization).
-
-Although grapheme tokenization is only one part of the solution, it is an
-important and highly fruitful processing step. Given a grapheme tokenization,
-various subsequent tasks become easier, like (a) temporarily reducing the
-orthography in a processing pipeline, e.g.~only distinguishing high versus low
-vowels; (b) normalizing orthographies across sources (often including temporary
-reduction of oppositions), e.g.~specifying an (approximate) mapping to the
-International Phonetic Alphabet; (c) using co-occurrence statistics across
-different languages (or different sources in the same language) to estimate the
-probability of grapheme matches, e.g.~with the goal to find regular sound
-changes between related languages or transliterations between different sources
-in the same language.
-
-Before we deal with these proposals, in the first part of this paper (Sections
-~\ref{encoding} through~\ref{ipa-meets-unicode}) we give an extended introduction
-to the notion of encoding (Section~\ref{encoding}) and writing systems, both
-from a linguistic perspective and from the perspective of the Unicode Consortium
-(Section~\ref{terminology}). We consider the Unicode Standard to be a
-breakthrough (and ongoing) development that fundamentally changed the way we
-look at writing systems, and we aim to provide here a slightly more in-depth
-survey of the many techniques that are available in the standard. A good
-appreciation for the solutions that the Unicode Standard also allows for a
-thorough understanding of possible pitfalls that one might encounter when using
-it (Section~\ref{unicode-pitfalls}). As an example of the current
-state-of-the-art, we discuss the rather problematic marriage of the
-International Phonetic Alphabet (IPA) with the Unicode Standard (Section
-\ref{ipa-meets-unicode}).
-
-The second part of the paper (Sections~\ref{orthography-profiles} and
-\ref{use-cases}) describes our proposals for how to deal with the Unicode
-Standard in the daily practice of (comparative) linguists. First, we discuss the
-challenges of characterizing a writing system. To solve these problems, we
-propose the notions of orthography profiles, closely related to Unicode locale
-descriptions (Section~\ref{orthography-profiles}). Finally, we discuss practical
-issues with actual examples (Section~\ref{use-cases}). We provide reference
-implementation of our proposals in R and in Python, available as open-source
-libraries.
-
-The following conventions are followed in this paper. All phonemic and phonetic
-representations are given in the International Phonetic Alphabet (IPA), unless
-noted otherwise \citep{IPA2005}. Standard conventions are used for
-distinguishing between graphemic < >, phonemic / / and phonetic [ ]
-representations. For character descriptions, we follow the notational
-conventions of the Unicode Standard \citep{Unicode2014}. Character names are
-represented in small capital letters (e.g.~\textsc{latin small letter schwa})
-and code points are expressed as U\emph{+n} where \emph{n} is a four to six
-digit hexadecimal number, e.g.~\uni{0256}, which can be rendered as the glyph <ə>.
-
-% ==========================
-\section{Encoding}
-\label{encoding}
-% ==========================
-
-There are many in-depth histories of the origin and development of writing
-systems \citep[e.g.~][]{Robinson1995,Powell2012}, a story that we therefore will
-not repeat here. However, the history of turning writing into computer readable
-code is not so often told, so we decided to offer a short survey of the major
-developments of such encoding here.\footnote{Because of the recent
-history as summarized in this section, we have used mostly rather ephemeral
-internet sources. When not references by traditional literature in the
-bibliography, we have used \url{http://www.unicode.org/history/} and various
-Wikipedia pages for the information presented here. A useful survey of the
-historical development of the physical hardware of telegraphy and
-telecommunication is \citet{Huurdeman2003}. Most books that discuss the
-development of encoding of telegraphic communication focus of cryptography,
-e.g.~\citet{Singh1999}, and forego the rather interesting story of ``open'',
-i.e.~non-cryptographic, encoding that is related here.} This history turns out
-to be intimately related to the history of telegraphic communication.
-
-\subsubsection*{Telegraphy}
-
-Writing systems have existed for roughly 6000 years, allowing people to exchange
-messages through time and space. Additionally, to quickly bridge large geographic
-distances, telegraphic systems of communication (from Greek \emph{τῆλε
-γράφειν} `distant writing') have a long and widespread history since ancient
-times. The most common telegraphic systems worldwide are so-called whistled
-languages \citep{Meyer2015}, but also drumming languages \citep{Meyer2012} and
-signalling by smoke, fire, flags or even change in water levels through
-hydraulic pressure have been used as forms of telegraphy.
-
-Telegraphy was reinvigorated in the end of the eighteenth century through the
-introduction of so-called semaphoric systems by Claude Chapelle to convey
-messages over large distances. Originally, various specially designed
-contraptions were used to send messages. Today, descendants of these systems are
-still in limited use, for example utilizing flags or flashing lights. The
-``innovation'' of those semaphoric systems was that all characters of the
-written language were replaced one-to-one by visual signals. Since then, all
-telegraphic systems have taken this principle, namely that any language to be
-transmitted first has to be turned into some orthographic system, which
-subsequently is encoded for transmission by the sender, and then turned back
-into orthographic representation at the receiver side.\footnote{Sound and
-video-based telecommunication of course takes a different approach by ignoring
-the written version of language and directly encode sound waves or light
-patterns.} This of course implies that the usefulness of any such telegraphic
-encoding completely depends on the sometimes rather haphazard structure of
-orthographic systems.
-
-In the nineteenth century, electric telegraphy lead to a new approach
-in which written language characters were encoded by signals sent through a
-copper wire. Originally, \textsc{bisignal codes} were used, consisting of two
-different signals. For example, Carl Friedrich Gauss in 1833 used positive and
-negative current \citep[282]{Mania2008}. More famous and influential, Samuel
-Morse in 1836 used long and short pulses. In those bisignal codes each character
-from the written language was encoded with a different number of signals
-(between one and five), so two different separators are needed: one between
-signals and one between characters. For example, in Morse-code there is a short
-pause between signals and a long pause between characters.\footnote{Actually,
-Morse-code also includes an extra long pause between words. Interestingly, it
-took a long time to consider the written word boundary---using white-space---as
-a bona-fide character that should simply be encoded with its own code point.
-This happened only with the revision of the Baudot-code (see below) by Donald
-Murray in 1901, in which he introduced a specific white-space code. This
-principle has been followed ever since.}
-
-\subsubsection*{Binary encoding}
-
-From those bisignal encodings, true \textsc{binary codes} developed with a fixed
-length of signals per character. In such systems only a single separator between
-signals is needed, because the separation between characters can be established
-by counting until a fixed number of signals has passed.\footnote{Of course, no
-explicit separator is needed at all when the timing of the signals is known, which is
-the principle used in all modern telecommunication systems. An important modern
-consideration is also how to know where to start counting when you did not catch
-the start of a message, something that is known in Unicode as \textsc{self
-synchronization}.} In the context of electric telegraphy, such a binary code
-system was first established by Émile Baudot in 1870, using a fixed combination
-of five signals for each written character.\footnote{True binary codes have a
-longer history, going at least back to the Baconian cipher devised by Francis
-Bacon in 1605. However, the proposal by Baudot was the quintessential proposal
-leading to all modern systems.} There are $2^5 = 32$ possible combination when
-using five binary signals; an encoding today designated as 5-bit. These
-codes are sufficient for all Latin letters, but of course they do not suffice
-for all written symbols, including punctuation and digits. As a solution, the
-Baudot code uses a so-called ``shift'' character, which signifies that from
-that point onwards---until shifted back---a different encoding is used, allowing
-for yet another set of 32 codes. In effect, this means that the Baudot code, and
-the \textsc{International Telegraph Alphabet} (ITA) derived from it, had an
-extra bit of information, so the encoding is actually 6-bit (with $2^6
-= 64$ different possible characters). For decades, this encoding was the
-standard for all telegraphy and it is still in limited use today.
-
-To also allow for different uppercase and lowercase letters and for a large
-variety of control characters to be used in the newly developing technology of
-computers, the American Standards Association decided to propose a new 7-bit
-encoding in 1963 (with $2^7 = 128$ different possible characters), known as the
-\textsc{American Standard Code for Information Interchange} (ASCII), geared
-towards the encoding of English orthography. With the ascent of other
-orthographies in computer usage, the wish to encode further variation of Latin
-letters (like German <ß> or various letters with diacritics, like <è>) led the
-Digital Equipment Corporation to introduce an 8-bit \textsc{Multinational
-Character Set} (MCS, with $2^8 = 256$ different possible characters), first used
-with the introduction of the VT{\large 220} Terminal in 1983.
-
-Because 256 characters were clearly not enough for the many different characters
-needed in the world's writing systems, the ISO/IEC~8859 standard in 1987
-extended the MCS to include 16 different 8-bit code pages. For example, part 5
-was used for Cyrillic characters, part 6 for Arabic, and part 7 for
-Greek.\footnote{In effect, because $16 = 2^4$, this means that ISO/IEC~8859 was
-actually an $8+4=12$-bit encoding, though with very many duplicates by design,
-namely all ASCII codes were repeated in each 8-bit code page. To be precise,
-ISO/IEC~8859 used the 7-bit ASCII as the basis for each code page, and defined
-16 different 7-bit extensions, leading to $(1+16)\cdot{2^7} = 2,176$ possible
-characters. However, because of overlap and not-assigned codes points the actual
-number of symbols was much smaller.} This system almost immediately was
-understood to be insufficient and impractical, so various initiatives to extend
-and reorganize the encoding started in the 1980s. This led, for example, to
-various proprietary encodings from Microsoft (e.g.~Windows Latin 1) and Apple
-(e.g.~Mac OS Roman), which one still sometimes encounters today.
-
-More wide-ranging, various people in the 1980s started to develop true
-international code sets. In the United States, a group of computer scientists
-formed the \textsc{Unicode Consortium}, proposing a 16-bit encoding in 1991
-(with $2^{16} = 65,536$ different possible characters). At the same time in
-Europe, the \textsc{International Organization for Standardization} (ISO) was
-working on ISO~10646 to supplant the ISO/IEC~8859 standard. Their first draft of
-the \textsc{Universal Character Set} (UCS) in 1990 was 31-bit (with
-theoretically $2^{31} = 2,147,483,648$ possible characters, but because of some
-technical restrictions only 679,477,248 were allowed). Since 1991, the Unicode
-Consortium and the ISO jointly develop the \textsc{unicode standard}, or
-ISO/IEC~10646, leading to the current system including the original 16-bit
-Unicode proposal as the \textsc{basic multilingual plane}, and 16 additional
-planes of 16-bit for further extensions (with in total $(1+16) \cdot 2^{16} =
-1,114,112$ possible characters). The most recent version of the Unicode Standard
-(currently at version number 7.0) was published in June 2014 and it defines
-112,956 different characters \citep{Unicode2014}.
-
-\
-
-\noindent In the next section we provide a very brief overview of the linguistic terminology
-concerning writing systems before turning to the slightly different
-computational terminology in the next section on the Unicode Standard.
-
-% ==========================
-\section{Linguistic terminology}
-\label{linguistic-terminology}
-% ==========================
-
-Linguistically speaking, a \textsc{writing system} is a symbolic system that
-uses visible or tactile signs to represent language in a systematic way. The
-term writing system has two mutually exclusive meanings. First, it may
-refer to the way a particular language is written. In this sense the term refers
-to the writing system of a particular language, as, for example, in \emph{`the
-Serbian writing system uses two scripts: Latin and Cyrillic.'} Second, the term
-writing system may also refer to a type of symbolic system as used among the
-world's languages to represent the language, as, for example, in
-\emph{`alphabetic writing system.'} In this latter sense the term refers to how
-scripts have been classified according to the way that they encode language, as
-in, for example, \emph{`the Latin and Cyrillic scripts are both alphabetic
-writing systems.'} To avoid confusion, this second notion of writing system
-would more aptly have been called \textsc{script system}.
-
-\subsubsection*{Writing systems}
-
-Focussing on the first sense, we distinguish two different kinds of 
-writing systems used for a particular language, namely transcriptions and
-orthographies.
-
-First, \textsc{transcription} is a scientific procedure (and also
-the result of that procedure) for graphically representing the sounds of human
-speech at the phonetic level. It incorporates a set of unambiguous symbols to
-represent speech sounds, including conventions that specify how these symbols
-should be combined. A transcription system is a specific system of symbols and
-rules used for transcription of the sounds of a spoken language variety. In
-principle, a transcription system should be language-independent, in that it
-should be applicable to all spoken human languages. The \textsc{International
-Phonetic Alphabet} (IPA) is a commonly used transcription system that provides a
-medium for transcribing languages at the phonetic level. However, there is a
-long history of alternative kinds of transcription systems
-\citep[see][]{Kemp2006} and today various alternatives are in widespread use
-(e.g.~X-SAMPA and Cyrillic-based phonetic transcription systems). Many
-users of IPA do not follow the standard to the letter, and many dialects based
-on the IPA have emerged, e.g.~the Africanist and Americanist transcription
-systems. Note that IPA symbols are also often used to represent language on a
-phonemic level. It is important to realize that in this usage the IPA
-symbols are not a transcription system, but rather an orthography (though with
-strong links to the pronunciation). Further, a transcription system does not
-need to be as highly detailed as the IPA.\ It can also be a system of broad sound
-classes. Although such an approximative transcription is not normally used in
-linguistics, it is widespread in technological approaches
-\citetext{\citealp[Soundex and variants, e.g.~][391--392]{Knuth1973};
-\citealp{postel1969,Beider2008}}, and it is sometimes fruitfully used in
-automatic approaches to historical linguistics
-\citep{Dolgopolsky1986,List2012esslli,Brown2013}.
-
-Second, an \textsc{orthography} specifies the symbols, punctuations, and the
-rules in which a specific language is written in a standardized way.
-Orthographies are often based on a phonemic analysis, but they almost always
-include idiosyncrasies because of historical developments (like sound changes or
-loans) and because of the widely-followed principle of lexical integrity
-(i.e.~the attempt to write the same lexical root in a consistent way, also when
-synchronic phonemic rules change the pronunciation, as for example with final
-devoicing in many Germanic languages). All orthographies are language-specific
-(and often even resource-specific), although individual symbols or rules might
-be shared between languages. A \textsc{practical orthography} is a strongly
-phoneme-based writing system designed for practical use by speakers. The mapping
-relation between phonemes and graphemes in practical orthographies is purposely
-shallow, i.e.~there is mostly a systematic and faithful mapping from a phoneme
-to a grapheme. Practical orthographies are intended to jumpstart written
-materials development by correlating a writing system with the sound units of a
-language \citep[cf.~][]{MeinhofJones1928}. Symbols from the IPA are often used
-by linguists in the development of such practical orthographies for languages
-without writing systems, though this usage of IPA symbols should not be confused
-with transcription (as defined above).
-
-Further, a \textsc{transliteration} is a mapping between two different
-orthographies. It is the process of ``recording the graphic symbols of one
-writing system in terms of the corresponding graphic symbols of a second writing
-system'' \citep[396]{Kemp2006}. In straightforward cases, such a transliteration
-is simply a matter of replacing one symbol with another. However, there are
-widespread complications, like one-to-many or many-to-many mappings, which are
-not always easy, or even possible, to solve without listing all cases
-individually \citep[cf.~][Ch.~2]{Moran2012}.
-
-\subsubsection*{Script systems}
-
-Different kinds of writing systems are classified into script
-systems. A \textsc{script} is a collection of distinct symbols as
-employed by one or more orthographies. For example, both Serbian and Russian are
-written with subsets of the Cyrillic script. A single language, like Serbian or
-Japanese, can also be written using orthographies based on different scripts.
-Over the years linguists have classified script systems in a variety of ways,
-with the tripartite classification of logographic, syllabic, and alphabetic
-remaining the most popular, even though there are at least half a dozen
-different types of script systems that can be distinguished
-\citep{Daniels1990,Daniels1996}.
-
-Breaking it down further, a script consists of \textsc{graphemes}, and graphemes
-consist of \textsc{characters}. In the linguistic terminology of writing
-systems, a \textsc{character} is a general term for any self-contained element
-in a writing system.\footnote{There is a second interpretation of the term
-\textsc{character}, i.e.~a conventional term for a unit in the Chinese writing
-system \citep{Daniels1996}. This interpretation will not be further explored in
-this paper.} Although in literate societies most people have a strong intuition
-about what the characters are in their particular orthography or orthographies,
-it turns out that the separation of an orthography into separate characters is
-far from trivial. The widespread intuitive notion of a character is strongly
-biased towards educational traditions, like the alphabet taught at schools, and
-technological possibilities, like the available type pieces in a printer's job
-case, the keys on a typewriter, or the symbols displayed in Microsoft Word's
-symbol browser. In practice, characters often consist of multiple building
-blocks, each of which could be considered a character in its own right. For
-example, although a Chinese character may be considered to be a single basic
-unanalyzable unit, at a more fine-grained level of analysis the internal
-structure of Chinese characters is often comprised of smaller semantic and
-phonetic units that should be considered characters \citep{Sproat2000}. In
-alphabetic scripts, this problem is most forcefully exemplified by diacritics.
-
-A \textsc{diacritic} is a mark, or series of marks, that may be above, below, or
-through other characters \citep{Gaultney2002}. Diacritics are sometimes used to
-distinguish homophonous words, but they are more often used to indicate a
-modified pronunciation \citep[xli]{DanielsBright1996}. The central question is
-whether, for example, <e>, <è>, <a> and <à> should be considered four
-characters, or different combinations of three characters. In general, multiple
-characters together can form another character, and it is not always possible to
-decide on principled grounds what should be the basic building blocks of an
-orthography.
-
-For that reason, it is better to analyze an orthography as a collection of
-graphemes. A \textsc{grapheme} is the basic, minimally distinctive symbol of a
-particular writing system, alike to the phoneme is an abstract representation of
-a distinct sound in a specific language. The term \textsc{grapheme} was modeled
-after the term \textsc{phoneme} and represents a contrastive graphical unit in a
-writing system (see \citealp{Kohrt1986} for a historical overview of the term
-grapheme). Most importantly, a single grapheme regularly consists of multiple
-characters, like <th>, <ou> and <gh> in English (note that each character in
-these graphemes is also a separate grapheme in English). Such complex graphemes
-are often used to represent single phonemes. So, a combination of characters is
-used to represent a single phoneme. Note that the opposite is also found in
-writing systems, in cases in which a single character represents a combination
-of two or more phonemes. For example, <x> in English orthography represents a
-combination of the phonemes /k/ and /s/.
-
-Further, conditioned or free variants of a grapheme are called
-\textsc{allographs}. For example, the distinctive forms of Greek sigma are
-conditioned, with <σ> being used word-internally and <ς> being used at the end
-of a word. In sum, there are many-to-many relationships between phonemes and
-graphemes as they are expressed in the myriad of language- and resource-specific
-orthographies.
-
-\
-
-\noindent This exposition of the linguistic terminology involved in describing writing
-systems has been purposely brief. We have highlighted some of the linguistic
-notions that are pertinent, yet sometimes confused with, the technological
-definitions developed for the computational processing of the world's writing
-systems, which we describe in the next section.
-
-% ==========================
-\section{The Unicode approach}
-\label{the-unicode-approach}
-% ==========================
+\section{Background}
 
@@
 
-\subsubsection*{The Unicode Standard}
+\section{The Unicode Standard}
 
@@
 
-\subsubsection*{Character encoding system}
+\section{Character encoding system}
 
@@
 definitional statement. For example, many IPA symbols are not located in the
-aptly-names block \textsc{IPA extensions}, but in other blocks
+aptly-named block \textsc{IPA extensions}, but in other blocks
 (see Section~\ref{pitfall-no-complete-ipa-block}).
 
-\subsubsection*{Grapheme clusters}
+\section{Grapheme clusters}
 

### 2018-02-26 book/chapters/implementation.tex
includes the broken out chapters

@@
 
-To illustrate the practical application of orthography profiles as defined in
-the previous Chapter~\ref{orthography-profiles}, we have implemented two 
-versions of the specifications presented there: one in Python and one
-in R.
+\section{Overview}
+To illustrate the practical application of orthography profiles, as defined in
+Chapter~\ref{orthography-profiles}, we have implemented two 
+versions of the specifications presented there: one in Python\footnote{\url{https://www.python.org/}} and one
+in R.\footnote{\url{https://www.r-project.org/}}
 
@@
 Users should refer to the tests cases and to the documentation in each release 
-for specifics about each implementation. Note also that function names differ 
-in the two implementations.
+for specifics about each implementation. Due to different naming standards, 
+function names unfortunately differ in the two implementations.
 
-Note also that the performance with larger datasets differs, and the code is not
+Note that the performance with larger datasets differs, and the code is not
 always as clean as we would like it to be. In sum, the two implementations
@@
 current implementations are sufficiently fast and stable to be useful for
-academic practice (e.g. checking data consistency, or analyzing and
+academic practice (e.g.\ checking data consistency, or analyzing and
 transliterating small to medium sized data sets), but they should probably 
-better not be used for full-scale industry applications.
+not be used for full-scale industry applications without some adaptations.
 
-This chapter will introduce the implementations, and give practical step-by-step 
-guidelines for installing them and using them. Various simple and sometimes 
+This chapter introduces the implementations, and gives practical step-by-step 
+guidelines for installing and using them. Various simple and sometimes 
 somewhat abstract examples will be discussed to show the different options 
@@
 
-This chapter is split into two sections, one for the Python implementation, the 
-other for R. Each section has a number of subsections that ...
+First we explain how to install Python and  R.
+There are many ways of installing these programming languages and differences in approach may depend on your operating system.
+Search engines are your friend if you run into installation problems.
+Our solutions below are tailored to our personal experiences.
 
-In addition to the material presented here to get users started, we maintain 
-several case studies that use each OP implementation in action. These use cases 
-are an the accompanying Github repository available with this book at: 
-\url{https://github.com/unicode-cookbook/recipes/}.
+Sections \ref{python-implementations} and \ref{r-implementation} discuss the Python and R implementations, 
+respectively. In addition to the material presented here to get users started, we maintain 
+several case studies that use orthography profiles in action. For convenience, 
+we make these use cases available as Jupyter Notebooks\footnote{\url{http://jupyter.org/}} 
+in the Github repository of code recipes which accompanies this book online.\footnote{\url{https://github.com/unicode-cookbook/recipes/}}
 
-% TODO: point to the case studies
-% In the next chapter we will then discuss a few full-scale examples to illustrate the application and the usefulness of orthography profiles.
+\section{How to install Python and R}
+When one encounters problems installing software, or bugs in programming code, search engines are your friend.
+Installation problems and incomprehensible error messages have often already been encountered and solved by other users.
+Such issues are typically overcome by searching the Web, e.g.\ try simply copying and pasting the output of an error message into a search engine.
+In our experience, answers are often already answered somewhere online.
+We are fans of Stack Exchange\footnote{\url{https://stackexchange.com/}} -- a network of question-and-answer websites -- which are extremely helpful in solving issues regarding software installation or bugs in code.
+
+Searching the Web for ``install r and python'' returns numerous tutorials on how to set up your machine for scientific data analysis.
+Note there is no single correct setup for a particular computer or operating system.
+Both Python and R are available for Windows, Mac, and Unix operating systems from the Python and R project websites.
+Another option is to use a so-called package manager, i.e.\ a software program that allows the user to manage software packages and their dependencies.
+On Mac, we use Homebrew\footnote{\url{https://brew.sh/}}, a simple to install (via the Terminal App) free and open source package management system.
+Follow the instructions on the their website and then use Homebrew to install R and Python.
+
+Alternatively for R, RStudio\footnote{\url{https://www.rstudio.com/}} provides a free and open source integrated development environment (IDE).
+This application can be downloaded and installed (for Mac, Windows or Unix) and it includes its own R installation and R package manager.
+For developing in Python, we recommend the free community version of PyCharm\footnote{\url{https://www.jetbrains.com/pycharm/}}, an IDE which is also available for Mac, Windows, and Unix.
+
+Once you have R or Python (or both) installed on your computer, you are ready to use the orthography profiles software libraries presented in the next two sections.
+We make the setup instructions, and the examples of how to work with and create orthography profiles, available online in Jupyter Notebooks.
+These so-called notebooks are an interface where you can run and rerun code in the browser.
+These notebooks are easily viewed via our Github respository of recipes.\footnote{\url{https://github.com/unicode-cookbook/recipes}}
+
+Depending on your computer's operating system, it should be simple enough to install the Jupyter (or other) notebook service and use the notebooks that we (and many others) provide via Github.
+Our orthography profile receipes and how to install the R or Python libraries for working with orthography profiles.\footnote{\url{https://github.com/unicode-cookbook/recipes}}
+
+We use Github,\footnote{\url{https://github.com/}} a web-based version control system source code management.
+This cookbook\footnote{\url{}} and the recipes\footnote{\url{https://github.com/unicode-cookbook/recipes.git}} for working with orthography profiles are available for cloning\footnote{\url{https://help.github.com/articles/cloning-a-repository/}} for working with on your local machine.
+Use your favorite search engine to figure out how to install Git in your computer or learn more at Git.\footnote{\url{https://git-scm.com/}}
 
@@
 
-To make the functionality even more accessible, we have prepared webapps with 
-the \texttt{shiny} framework for the R functions. These webapps are available 
-online at \url{TODO}. The webapps are also included inside the \texttt{qlcData} 
-package and can be started with the following helper function:
+% TODO
+% To make the functionality even more accessible, we have prepared webapps with 
+% the \texttt{shiny} framework for the R functions. These webapps are available 
+% online at \url{TODO}. The webapps are also included inside the \texttt{qlcData} 
+% package and can be started with the following helper function:
 
-<<eval=FALSE>>=
-launch_shiny('tokenize')
-@
-
-% TO DO
+% <<eval=FALSE>>=
+% launch_shiny('tokenize')
+% @
 

### 2018-02-26 book/chapters/introduction.tex
closes issue #6
closes issue #7

@@
 representations are given in the International Phonetic Alphabet (IPA), unless
-noted otherwise \citep{IPA2005}. Standard conventions are used for
+noted otherwise \citep{IPA2015}. Standard conventions are used for
 distinguishing between graphemic < >, phonemic / / and phonetic [ ]
 representations. For character descriptions, we follow the notational
-conventions of the Unicode Standard \citep{Unicode2014}. Character names are
+conventions of the Unicode Standard \citep{Unicode2017}. Character names are
 represented in small capital letters (e.g.~\textsc{latin small letter schwa})
@@
 1,114,112$ possible characters). The most recent version of the Unicode Standard
-(currently at version number 7.0) was published in June 2014 and it defines
-112,956 different characters \citep{Unicode2014}.
+(currently at version number 10.0.0) was published in June 2017 and it defines
+136,690 different characters \citep{Unicode2017}.
 

### 2018-02-26 book/chapters/pitfalls.tex
fixes compiling
closes issue #7

@@
 notion of diacritic). The Unicode Standard calls such combination \textit{grapheme
-clusters||. Complex graphemes consisting of multiple base characters,
+clusters}. Complex graphemes consisting of multiple base characters,
 like <sch>, are called \textit{tailored grapheme clusters} in Unicode parlance (see
@@
 Format.''} The details of this issue is extensively explained in section 2.5 of
-the Unicode Core Specification~\citet{Unicode2014}.
+the Unicode Core Specification~\citet{Unicode2017}.
 

### 2018-02-26 book/chapters/unicode_approach.tex
closes issue #7

@@
 
-Each version of the Unicode Standard (\citealp{Unicode2014}, as of writing at
-version 7) consists of a set of specifications and guidelines that include (i) a
+Each version of the Unicode Standard (\citealp{Unicode2017}, as of writing at
+version 10) consists of a set of specifications and guidelines that include (i) a
 core specification, (ii) code charts, (iii) standard annexes and (iv) a

### 2018-02-27 book/chapters/ipa_background.tex
closes issue #6
fixes bibliography; updates IPA reference

@@
 accommodate scientific innovations and discoveries. In fact, updates are part of the
-Association's mandate. These changes are captured in revisions to the IPA chart.
-%\footnote{For a detailed history see:
-%\url{https://en.wikipedia.org/wiki/History\_of\_the\_International\_Phonetic_Alphabet}}
+Association's mandate. These changes are captured in revisions to the IPA chart.\footnote{For a detailed history see:
+\url{https://en.wikipedia.org/wiki/History\_of\_the\_International\_Phonetic_Alphabet}}
 
@@
 the differences, see the 29 languages and their transcriptions in the
-\textit{Illustrations of the IPA} \citep[41--154]{IPA2007}.} Therefore, there are 
+\textit{Illustrations of the IPA} \citep[41--154]{IPA1999}.} Therefore, there are 
 different but equivalent transcriptions, or as noted by \citet[64]{Ladefoged1990a}, 
@@
 transcribe speech in written form.
+
+% TODO: cite Moran 2018
 
@@
 IPA symbols and they were also illustrated in IPA chart form (see
-\cite[84]{EslingGaylord1993} or \cite[App. 2]{IPA2007}). The numbers were
+\cite[84]{EslingGaylord1993} or \cite[App. 2]{IPA1999}). The numbers were
 assigned in linear order (e.g.\ [p] 101, [b] 102, [t] 103...) following the IPA

### 2018-02-27 book/chapters/practical_recommendations.tex
fixes bibliography; updates IPA reference

@@
 
-For general information about the IPA, the standard text is the \textit{Handbook of the International Phonetic Association: A Guide to the Use of the International Phonetic Alphabet} \citep{IPA2007}.
+For general information about the IPA, the standard text is the \textit{Handbook of the International Phonetic Association: A Guide to the Use of the International Phonetic Alphabet} \citep{IPA1999}.
 The handbook describes in detail the principles and premises of the IPA, which we have summarized in Section \ref{IPApremises-principles}.

### 2018-02-27 book/chapters/introduction.tex
updates structue of this book section to include new chapter arrangement

@@
 introduction to the notion of encoding (Section~\ref{encoding}) and the
-principles of writing systems, both from a linguistic perspective
-(Section~\ref{linguistic-terminology}) and from the perspective of the Unicode
-Consortium (Section~\ref{the-unicode-approach}). We consider the Unicode
+principles of writing systems from a linguistic perspective
+(Section~\ref{linguistic-terminology}). In Chapter~\ref{the-unicode-approach}, we 
+discuss the notions of encoding and writing systems from the perspective of the Unicode
+Consortium. We consider the Unicode
 Standard to be a breakthrough (and ongoing) development that fundamentally
@@
 Linguists are more often interested in using the Unicode Standard with the International Phonetic Alphabet (IPA).
-We first provide a history of the development of the IPA and early attempts to encode it electronically (Chapter~\ref{the-international-phonetic-alphabet}) before we discuss the rather problematic marriage of the International Phonetic Alphabet (IPA) with the Unicode Standard (Chapter~\ref{ipa-meets-unicode}).
+We first provide a history of the development of the IPA and early attempts to encode it electronically (Chapter~\ref{the-international-phonetic-alphabet}) before we discuss the rather problematic marriage of the IPA with the Unicode Standard (Chapter~\ref{ipa-meets-unicode}).
 
-In the second part of the book (Chapters~\ref{orthography-profiles} and
+In the second part of the book (Chapters~\ref{practical-recommendations}, \ref{orthography-profiles} \&
 \ref{implementation}) we describe our proposals of how to deal with the Unicode
-Standard in the daily practice of (comparative) linguistics. First, we discuss the
-challenges of characterizing a writing system. To solve these problems, we
+Standard in the daily practice of (comparative) linguistics. First, we provide some 
+practical recommendations for using the Unicode Standard and IPA for ordinary 
+working linguists and for computer programmers (Chapter~\ref{practical-recommendations}). 
+Second, we discuss the
+challenges of characterizing a writing system; to solve these problems, we
 propose the notions of orthography profiles, closely related to Unicode locale
-descriptions (Chapter~\ref{orthography-profiles}). Finally, we discuss practical
-issues with actual examples (Chapter~\ref{implementation}). We provide reference
-implementation of our proposals in R and in Python, available as open-source
-libraries. An introduction to these implementations of our proposals is
-described in Chapter~\ref{implementation}. Finally, we discuss practical issues
-with actual examples in Chapter~\ref{use-cases}.
+descriptions (Chapter~\ref{orthography-profiles}).
+Lastly, we provide an introduction to two open source libraries that we have developed, in Python and R, for working with linguistic data and orthography profiles (Chapter~\ref{implementation}).
 

### 2018-03-06 book/reviews/responses_reviewer1.txt
these changes almost complete the remaining reviews

# Responses to reviewer 1

*** Our responses prefixed with "***" (which is sometimes also used to delimit the comments from the reviewer for easier answering).

*** We kindly thank the reviewer -- we know reviewing (especially a book) is a time consuming endeavor and we feel that their comments have helped us substantially improve the cookbook. (Note that chapter numbers mentioned by the reviewer may have changed, e..g we split chapter 1 into chapters 1 and 2, as suggested.)

===========
As stated in the preface, the book not only targets linguists, but also
programmers. This review is written by a programmer from the perspective
of a programmer.

I want to start out saying that this is a much needed book. I personally 
often felt a need to educate linguists about unicode and the representation 
of text in computers and was hoping for a reference with enough detail *and* 
the proper language fit for linguists. The "Unicode Cookbook for Linguists"
feels like it is this reference, so I highly recommend publication of it by
Language Science Press.

I also welcome and second the call to linguists to take IPA and Unicode for
what they are: Bona fide, best effort standards with well-defined use cases.
The invitation to treat these standards as useful tools in a pragmatic way
rather than complaining about their perceived shortcomings is repeated several
times throughout the book - but bears repeating, because reluctance to
accept and adhere to common standards in language documentation, not a rare 
trait among linguists, has prevented the field from more meaningfully employ
quantitative methods of language comparison.

The first chapter of the "Unicode Cookbook for Linguists" provides a good
introduction to the problem space of written language and how to compare it.
Thus, it describes the motivation and problems for standardization efforts.
The last section of chapter 1, "1.4 The Unicode approach" may fit better as
first section in the second chapter, at least in my mental model of the 
outline of the book as

1. The problem space
2. Proposed solution from an IT perspective
3. Proposed solution from a linguistic perspective
4. Identification and description of the "impedance mismatch" between the two
5. Recommendation how to bridge the gap between the two standards

*** We have separated out the Unicode Approach subsection of chapter 1 into its own chapter (2) with these subsections that appear in the TOC:

- Background
- The Unicode Standard
- Character encoding system
- Grapheme clusters

Anyway, chapters 2 and 3 introduce the two most important attempts at
standardizing written language for linguists working on computers: The
Unicode standard and the International Phonetic Alphabet. These two standards
are described with enough history and detail to appreciate the results, but
also highlighting - with the benefit of hindsight - the shortcomings.

*** Thanks!

Chapter 2, "Unicode Pitfalls", is arranged as a list of common problems when
working with Unicode. Even just looking at the some section names, 
"Characters are not glyphs", "Characters are not graphs", "Blocks", "Names",
hints at the main problem of using Unicode being one of terminology.

In particular for programmers or researchers using a computing environment like
python or R, this is certainly the case, because many of these environments -
being mainly used and developed in the English-speaking world - get away with
not distinguishing two things, namely "sequences of code points" and "text
encoded as bit patterns". Now looking up "Character encoding" on Wikipedia [1]
we learn that the term "encoding" is used for all levels - code points or
bit patterns. But in programming, the standard terminology seems to have
settled on reserving "encoding" for the bit-pattern level, while using terms
like "string" or "unicode text" for sequences of Unicode code points [2].

So while it may be technically correct to say

    The Unicode Standard is a CHARACTER ENCODING SYSTEM (page 13)

this sounds a bit odd to the Unicode-aware programmer, who would rather say
"UTF-8 is a Unicode Encoding system". So the solution of the Unicode Standard
for the problem of many mutually conflicting encoding systems was to introduce
a new layer of indirection between character and bit pattern - the code point.
Unfortunately, the Unicode Consortium itself seems to shy away from clearer
*new* terminology, when it compares Unicode to encoding systems (not saying
Unicode is one, though), but does not coin a new term for what Unicode actually
is, but rather defining it purely functional as "Unicode provides a unique number 
for every character" [3].
It's hard to tell whether attempting to coin such a term could have succeeded
with this book, but it is a bit sad, that the blurry terminology used everywhere
is perpetuated here. Still, the authors should try not to distract or confuse
the reader as with footnote 10 on page 13, where the short excursion into
numeral systems seems to suggest, that Unicode code point U+0070 is always
"encoded" as bit pattern 1110 0000.

*** We appreciate this keen comment from the programmer's perspective. In attempt to not confuse the targeted audiences (ordinary working linguists, potential and amateur programmers), we've added a footnote to this sentence "The Unicode Standard is a CHARACTER ENCODING SYSTEM (page 13)", which we think clarifies the terminological issue for advanced readers:

The Unicode Standard is a \textsc{character encoding system} which
goal it is to support the interchange and processing of written characters and
text in a computational setting.\footnote{An insightful reviewer notes that the term \textsc{encoding} is used for both sequences of code points and text encoded as bit patterns. Hence a Unicode-aware programmer might prefer to say that UTF-8, UTF-16, etc., are Unicode encoding systems. The issue is that the Unicode Standard introduces a layer of indirection between characters and bit patterns, i.e. the code point, which can be encoded differently by different encoding systems}

We have also removed the footnote: "Hexadecimal (base-16) 0070 is equivalent to decimal
(base-10) 112..."

Furthermore, we thought it pertinent to add an additional Pitfall: Encodings, which we think clarifies the reviewer's points about the terminological mismatch of "encoding". We put this section before the file formats pitfall.

***

Section "Pitfall: File formats" may profit most from a more transparent
terminology, distinguishing levels of encoding, rather than talking about how
texts "appear inside some kind of computer file". This section should also
mention that line breaks are actually part of the text, and as such also
covered by the Unicode standard. Again regarding line breaks, the 
recommendation "that everybody use this encoding [only LF] whenever possible"
(page 31) seems less well motivated (and specific) than most other 
recommendations in the book. Some common formats, e.g. CSV (as specified by 
RFC 4180 [4]), specify "CRLF" as line break. Thus, the perceived "strong 
tendency" towards "LF" may be just that.

*** See above.

One Unicode pitfall which may be worth adding is the problem with incomplete
implementations of the Unicode Standard. This is a problem most big-enough
standards suffer from, e.g. SQL, XML, XSLT. So although the Unicode Standard
mandates that comparison of Unicode text should always be done using normalized
text, this is not how the equality operator "==" is implemented for Unicode
text in the Python programming language.

*** We added a very brief Pitfall that addresses this point (and shamelessly took the examples from the reviewer). Thanks!

Pitfall: Incomplete implementations

Another pitfall that we encounter when using the Unicode Standard is its incomplete implementation in different standards and programming languages, e.g.\ SQL, XML, XLST, Python. For example, although the Unicode Standard mandates that the comparison of Unicode text be done using normalized text, this is not the csae with the equality operator ``=='' in Python. Furthermore, it is not always transparent what the operating system or specific software applications do when text is being copied and pasted. For example, copy and pasting the character sequence \uni{0061} \textsc{latin small letter a} <a> and \uni{0301} \textsc{combining acute accent} <\dia{0301}>, visually <á>, into the text editor TextWrangler leaves the sequence decomposed as two characters. But when pasting the decomposed sequence into RStudio, and other software programs, the sequence becomes precomposed as \uni{00E1} \textsc{latin small letter a with acute}, i.e.\ <á>.

***

Chapter 3 introduces "The International Phonetic Alphabet". As with Unicode,
the authors stress the characterization of IPA as a standard, which makes some
premises and is guided by some principles, but overall tries to solve an
interoperability problem in a pragmatic way. I.e. when you want to use text
on a computer, you should use Unicode; if you want to compare written language
across languages, you should use IPA.

It may be worth highlighting the parallel between "IPA Numbers" and "Unicode 
Code Points", i.e. the fact, that IPA and Unicode came up with the solution of
an additional layer of indirection between symbols/characters and encoding
(on bit-pattern level); or the parallel between SAMPA/X-SAMPA and the multitude
of pre-Unicode encodings for text. The fact that the authors added enough
historical background about both IPA and Unicode makes it possible to relate
these parallels in time, and prevent dismissal of attempts like SAMPA as
misguided - because now we have Unicode.

*** We added a note about this to the last paragraph in Chapter 3:

It was a long journey, but the goal of achieving a single multilingual computing
environment has largely been accomplished. As such, we should not dismiss the IPA Numbers 
or pre-Unicode encoding attempts, such as SAMPA/X-SAMPA, as misguided. The parallels 
between the IPA Numbers and Unicode Code points, for example, is striking because both the IPA and the Unicode Consortium came up with the solution of an additional layer of indirection between symbols/characters and encoding on the bit pattern level. SAMPA/X-SAMPA is also still useful as an input method for IPA in ASCII and required by some software.

***

Again, the authors should take care to keep terminology as consistent as 
possible, e.g. in a passage like

    […] the computer-coding convention for the IPA should be independent of 
    computer environments and formats (e.g. ASCII), i.e. the IPA Number was
    not meant to be implemented directly in a computer encoding (page 41)

"computer-coding convention", "format" and "computer encoding" seem to refer
to the same thing, i.e. encoding on bit-pattern level. In particular listing
ASCII as an example of a format makes a programmer wonder whether this new
term is significant, because up to this point, ASCII was referred to as a
character encoding.

*** We've updated this passage:

The working group decided insightfully that the computing-coding convention 
for the IPA should be independent of computer environments or formats, 
i.e.\ the IPA Number was not meant to be encoded at the bit pattern level.

Chapter 4, "IPA meets Unicode", describes what it means for IPA that Unicode
has become *the* standard for handling text in a computing environment. We
have learned so far that 
- we have to use Unicode to use text with a computer, and
- we have to use IPA to compare written language.
Thus, chapter 4 gives recommendations how to compare written language with a 
computer. As with chapter 2, this chapter is organized as list of pitfalls,
augmented with a section on recommendations. While it may seem overly verbose
or boring that most of the pitfalls contain somewhat extensive lists of
symbols/characters/Unicode code blocks, I think this is overall the right
approach: As the autors show, making IPA work with Unicode is largely a
matter of spelling out special cases, and how to deal with them. It seems
important, though, to have the tables starting at page 70 available in a
computer readable format, and have the book point to the location of these
files.

*** Nice suggestion! We have put these tables online here in machine-readable format: https://github.com/unicode-cookbook/cookbook/tree/master/book/tables

Chapter 5 introduces "Orthography profiles" as a tool to specify writing
systems in such a way that transliteration (e.g. to a suitable subset of IPA)
becomes simple and transparent; thus, a tool to get from the Unicode domain
of handling text with a computer to the IPA domain of comparing written
language with a computer. Having used orthography profiles myself in projects,
I can already testify to their usefulness.

Like a "real" programmer would do, whenever a chapter contains a section called
"Formal specification", I'll skip right ahead to it.

File Format:

A1: "ideally using […]" is not exactly language you would like to see in a
specification; instead the specification should follow RFC 2119 [5], 
"Key words for use in RFCs to Indicate Requirement Levels". So in this case,
I guess an orthography profile must be normalized following NFC (or NFD, if 
specified in the metadata). Specifying a convention for line endings and BOM
seems overly strict, since most computing environments will transparently
handle both alternatives: E.g. using the Python programming language, decoding
a file using the encoding "utf-8-sig" will strip away a BOM, if present, and
reading a file in text mode will interpret both "LF" and "CRLF" as line
endings.

*** Thank you for the link to RFC 2219 -- we have changed our wording in accordance with their keywords. For example, we've updated this paragraph to:

A1. A profile is a unicode UTF-8 encoded text file normalized following NFC (or NFD if speci ed in the metadata) that includes the information pertinent to the orthography.^2

[2] See Section 3.12 in which we suggest to use NFC, no-BOM and LF line breaks because of our experience in working with linguistic data in this format. A keen reviewer notes that specifying a convention for line endings and BOM is overly strict because most computing environments (now) transparently handle both alternatives. For example, using Python a  file can be decoded using the encoding “utf-8-sig”, which strips away the BOM (if present) and reads an input full in text mode, so that both line feed variants “LF” and “CRLF” will be stripped.

A2: "tabs and newline […] will lead to problems", thus they should be
explicitely disallowed!

*** Agreed! We have updated this to:

A2. A profile is a CSV file with an obligatory header line. A minimal pro le must have a single column with the header “Grapheme”. For any ad- ditional columns, the name in the header of the CSV  le must be speci ed.  e actual ordering of the columns is unimportant. Cells should not con- tain tabs or newlines.  e header list must be delimited in the same way as the rest of the  le’s contents. Each record must be kept on a separate line. Separate lines with comments are not allowed. Comments that belong to specific lines must be put in a separate column of the CSV file, e.g.~add a column called \textsc{comments}.

Note we also merged in A4 into A2, previously:

A4. Separate lines with comments are not allowed. Comments that be- long to speci c lines will have to be put in a separate column of the CSV  le, e.g. add a column called comments.

***

A3: The metadata file seems to be highly under-specified. At the very least,
there should be a list of recognized tags providing semantics for the data.
But since machine-readability is a requirement for the file anyway, a format
like JSON-LD [6], or even more appropriate, the JSON-LD dialect specified in
"Metadata Vocabulary for Tabular Data" [7]  may be a better option, because it
allows easy inclusion of Dublin Core [8] metadata - as well as a specification
of the exact CSV dialect used for the actual orthography profile.
It should be noted that JSON-LD metadata is the choice for datasets conforming
to the Cross-Linguistic Data Formats [9] standard.

*** Agreed. In practice, we haven't actually used these metadata files. We've updated this paragraph to "SHOULD" via RFC 2119 and we RECOMMEND they follow the practices suggested by the reviewer and implemented in CLDF:

A3. Metadata should be added in a separate UTF-8 text file using the JSON-DL dialect speci ed in Metadata Vocabulary for Tabular Data.^ 3  This
metadata format allows easy inclusion of Dublin Core metadata,^4 which should be used to specify information about the orthographic description in the orthography pro le.  e orthography pro le metadata should min- imally include: (i) author, (ii) date, (iii) title of the pro le, (iv) a stable lan- guage identi er encoded in BCP 47/ISO 639-3 of the target language of the pro le, and (v) bibliographic data for resource(s) that illustrate the or- thography described in the pro le. Further, (vi) the tokenization method and (vii) the Unicode normalization form used should be documented here (see below). In accordance with the Metadata Vocabulary for Tabular Data, the metadata’s  lename must consist of the orthography pro le’s  lename appended with “-metadata.json”.^5

[3] https://www.w3.org/TR/tabular-metadata/
[4] http://dublincore.org/
[5] JSON-DL metadata is also the choice for datasets conforming to the Cross-Linguistic Data Formats standard, see: http://cldf.clld.org/.

A7: "If regular expressions are used, then […]": The specification does not
state how a processing application can determine if that is the case - other
than through additional user input. So specifiying whether regular expressions
are used should be required in the metadata file. In general, allowing regular
expressions as input is somewhat problematic, because regular expression
syntax as well as functionality is somewhat under-specified across regex
implementations in different computing environments. The closest to a standard
in this area might be "POSIX Basic Regular Expressions" [10], but these do not
include lookahead and lookbehind functionality as requred by B3. So allowing
regular expressions as part of orthography profiles may lead to different
results of applying the same profile to the same data on different platforms -
which clearly goes against the idea of this specification.

*** TODO

Overall, it seems somewhat strange to sacrifice
CSV standards conformance (no quoting of cell values, "LF" line endings, 
tab as separator) in order to achieve better read-/writeability by humans. This
puts a bigger burden on the authors in terms of specification (they can no 
longer just point to FC 4180) and it makes it harder to use tools like 
LibreOffice Calc or MS Excel to edit Orthography profiles.

*** Agreed (SM). We have made the formal specification lacker (which of course put the onus the software developers):

A1. A profile is a unicode UTF-8 encoded text file normalized following NFC (or NFD if speci ed in the metadata) that includes the information pertinent to the orthography.2

A2. A profile is a CSV file with an obligatory header line. A minimal pro le must have a single column with the header “Grapheme”. For any ad- ditional columns, the name in the header of the CSV  le must be speci ed.  e actual ordering of the columns is unimportant. Cells should not con- tain tabs or newlines.  e header list must be delimited in the same way as the rest of the  le’s contents. Each record must be kept on a separate line.

[2] See Section 3.12 in which we suggest to use NFC, no-BOM and LF line breaks because of the pitfalls they avoid. A keen reviewer notes, however, that specifying a convention for line endings and BOM is overly strict because most computing environments (now) transparently handle both alternatives. For example, using Python a  le can be decoded using the encoding 	“utf-8-sig”, which strips away the BOM (if present) and reads an input full in text mode, so that both line feed variants “LF” and “CRLF” will be stripped.

Implementation:

B1: As noted for A7, to make the basic handling possible in an automated way,
there must be a way to figure out that the profile is "basic".

*** TODO

B2: If - as stated - "ordering […] might be crucial", it must be specified, i.e.
members a class specified as context must be tried in order of occurrence in the
profile.

*** TODO

B4: I would rather spin the functionality to suggest an optimal ordering of rules
off to a separate tool, and limit the specification to apply rules in order of
occurrence. This will lead to simpler implementation and more transparent 
relation of input and output.

*** TODO

Language like

    The actual implementation of the profile on some text-string will function 
    as follows: (page 83)

sounds a bit weird in a technical specification. In general, the presentation
may benefit by adopting the conventions as used for describing algorithms, i.e.
specifying
- input
- output
- a series of steps to get from input to output

*** TODO

I did not review the Python and R implementations, and suggest to spin these off
to a more suitable format, e.g. Jupyter Notebooks [11].

[1] https://en.wikipedia.org/wiki/Character_encoding
[2] https://docs.python.org/3/howto/unicode.html#encodings
[3] http://www.unicode.org/standard/WhatIsUnicode.html
[4] https://www.ietf.org/rfc/rfc4180.txt
[5] https://www.ietf.org/rfc/rfc2119.txt
[6] https://json-ld.org/
[7] https://www.w3.org/TR/tabular-metadata/
[8] http://dublincore.org/
[9] http://cldf.clld.org/
[10] http://www.regular-expressions.info/posix.html
[11] http://jupyter.org/

*** In addition to having this material in the PDF version of the cookbook, we have spun off the implementations (and installation instructions) into Jupyter notebooks (https://github.com/unicode-cookbook/cookbook/issues/22):

For installation of the R and Python libraries (in the README):
https://github.com/unicode-cookbook/recipes https://github.com/unicode-cookbook/recipes

R basics:
https://github.com/unicode-cookbook/recipes/blob/master/Basics/R.ipynb https://github.com/unicode-cookbook/recipes/blob/master/Basics/R.ipynb

Python basics:
https://github.com/unicode-cookbook/recipes/blob/master/Basics/python.ipynb https://github.com/unicode-cookbook/recipes/blob/master/Basics/python.ipynb

The four other recipes are also now in Jupyter notebooks.

### 2018-03-06 book/reviews/responses_reviewer2.txt
these changes almost complete the remaining reviews

Review "The Unicode Cookbook for Linguistics"

This books gives an overview of the intersection of the IPA alphabet and
Unicode, preceded by a discussion of foundational concepts in writing
systems, and followed by the suggestion to establish "Orthography
profiles" as a theoretical concept, which are are in turn implemented in
Python and R.

*** Our responses prefixed with "***" (which is sometimes also used to delimit the comments from the reviewer for easier answering).

*** We thank the reviewer for a very thorough reading of our manuscript and for many helpful (and time consuming to fix/update) suggestions, which we know have made this cookbook ever the more useful for its intended audiences.

Global evaluation:
The book is concise and can serve as a useful manual for Linguists Who
Have Always Wondered About Unicode But Never Dared To Ask. The
exposition of the background is in general well organised and easy to
follow (for a technical paper). The writing is good in general, but
sometimes repetitive. Minor suggestions for reorganisation of passages
are given below. A higher amount of illustrations and charts would make
that part more accessible.

The orthography profiles section is less well written and at times a bit
confused. It does not seem completely finished either, since there are
dangling references and literal TODO in the text. For this section, it
will be necessary to have a very clear picture about the intended
audience. For instance, do the authors assume the readers know bash,
python and R? Most passages do, but some passages explain things that a
tech-savvy audience will well be aware of and find tedious. The language
of the implementation part is of less high quality than the introductory
parts.

*** We have thoroughly updated the orthography profiles and implementation chapters (in line with comments from reviewer 1, in which we make the orthography profile specification more precise, and we additionally spin off the implementation parts with code into Jupyter notebooks online).

The authors should check the use of single and double quotes across the
paper, which is incoherent. As far as I can tell, all instances of
material currently enquoted should be in double quotes. Also, the
authors should check Special Capitalization, which is used in some
places. Finally, the bibliography has recurrent issues relating to
naming of organizations, capitalization, and missing places of
publication and needs to be thoroughly revised.

*** We have gone through the manuscript and updated the quotes following these principles:

- Double quotation marks are generally used for distancing, in particular in the following situations:

	- when a passage from another work is cited in the text (e.g. According to Takahashi (2009: 33), “quotatives were never used in subordinate clauses in Old Japanese”); but block quotations do not have quotation marks;

	- when a technical term is mentioned that the author does not want to adopt, but wants to mention, e.g. This is sometimes called “pseudo-conservatism”, but I will not use this term here, as it could lead to confusion.

- Single quotation marks are used exclusively for linguistic meanings, as in the following: Latin habere ‘have’ is not cognate with Old English hafian ‘have’.

***

A section which is missing in something called a "Cookbook" would be
practical recommendations on how to input Unicode characters. There are
various character selection tools, shortcuts on the keyboard, the
shapecatcher website references at several places, or the Wikipedia
lists of glyphs and fileformat.info. Having all this in one section
would be handy for the user. It is of course unrelated to the
orthography profiles, but I imagine that many people will use this book
as a primer on IPA+Unicode and actually disregard the last two chapters.
For this group, such a summary would be useful.

*** We have included a new chapter titled "Practical recommendations" aimed at ordinary working linguists, which appears before the last two chapters (aimed at programmers).

Overall recommendation: publish with minor revisions.
=============================================================================

Detailed comments

Preface

- our research": who is "we"? I first I took this to mean the Unicode
research, but apparently the day-to-day research of the authors in other
fields is intended here.

*** Fixed: https://github.com/unicode-cookbook/cookbook/commit/e29788678c2647300339fdb1bab3e19eed58df6b

- "We welcome comments ...": this might be a good place to add
information about the book being available on the Paperhive platform,
where readers can leave questions and comments.

*** Fixed: https://github.com/unicode-cookbook/cookbook/commit/e29788678c2647300339fdb1bab3e19eed58df6b

Chapter 1
- 1.1.: two instances of "normally" in close succession

*** Fixed.

- automatic processing --> automated processing

*** Fixed throughout the book.

- Chapter4 --> Chapter 4

*** Fixed.

- are expressed as U + n where --> are expressed as U + n, where

*** Fixed.

- Chapter?? --> resolve reference

*** Fixed.

Section "Telegraphy"
- it is slightly odd that whistling and drumming are treated as
instances of teleGRAPHY here. Neither whistling nor drumming use
writing, so telePHONY or teleSONY might be more appropriate (but have
obvious other issues).

*** Agreed.

- Footnote 6 "In effect ..." must be rephrased (or deleted, not sure if
this information is actually relevant)

*** Deleted.

1.3 Linguistic terminology
- the definition of writing system is odd. It says "system that uses
visible or tactile signs". That would be true for sign languages as well
(and also for communication systems used by blind signers), but we would
not want to call these signers' systems a "writing system". A better
definition could be "shapes on a support", where the shapes could be
perceived with the eyes or, as with Braille, with the fingers. The
support could be paper, clay, or a screen. You might find yet another
definition which suits your purposes, but the current definition should
be revised in any case.

*** This definiton is one of many for the term writing system, e.g. Coulmas (1993, pg 1):

"At least six meanings of 'writing' can be distinguished: (1) a system of recording language by means of visible or tactile marks..."

who also notes:

"Every attempt at a universal definition of writing runs the risk of being either ad hoc or anachronistic, or informed by cultural bias."

Braille is the classic example of a tactile writing system. We will keep the definition as is.

- in the same section: do not italicize quoted strings.

*** Fixed.

- "as used among the world's languages to represent the language": delete

*** Deleted.

- "All orthographies are language-specific": suggest dropping "All"
unless this is an empirical claim and not a general truth.

*** Dropped.

Section script systems
- "Graphemes consist of characters": rewrite this whole passage. Some
problems are given below.

We've updated this passage to make things clearer by introducing what a grapheme is earlier.

- It is unclear to me what a character is. Maybe "a" and "b" are
characters used for the graphemes <a> and <b>, and "c" and "h" are
characters used in the grapheme <ch>?

We more clearly define character.

- "In practice, characters often consist of multiple building blocks,
each of which could be considered a character in its own right". So
characters consist of characters? Are there some "terminal characters",
which do not consist of other characters? Or do "bdpq" all consist of a
stroke and an eye, which are then characters???
This has to become clearer. In this overview section, it would be
helpful to set off the definitions, e.g.
(1) A GRAPHEME is ....
....
(2) A CHARACTER is ...
...
(3) A GLYPH is ...

- Footnote 7 refers to characters in Chinese (maybe radicals are meant
here???), and then states that this will not be further explored. But
about 10 lines down the page, the authors say "although a Chinese
character ...", which does look like a further exploration indeed.
Suggest dropping footnote 7, or alternatively make the argument with an
example other than Chinese.

*** We kept the footnote, but drop the not further explored part, since as you mention, we do highlight the fact (mainly for Chinese linguists) that "character" does have a different meaning in regard to Chinese writing systems:

In the linguistic terminology of writ- ing systems, a character is a general term for any self-contained element in a writing system.6 In technological terminology, a character refers to the elec- tronic encoding of a component in a writing system that has semantic value (see Section 2.3).

[6] Note that there is a second interpretation of the term character, i.e. a conventional term for a unit in the Chinese writing system (Daniels 1996).

- "A diacritic mark ... may be above, below, or through": add left and
right diacritics, e.g. Hindi long and short <i> attach to either side.

*** Good point, we updated this sentence:

A \textsc{diacritic} is a mark, or series of marks, that may be above, below, 
before, after, through, around, or between other characters \citep{Gaultney2002}.

- "A GRAPHEME is the basic, minimally distinctive symbol ..., alike to
the phoneme, which is an abstract representation ...": Does this entail,
then, that the grapheme is an abstract representation as well? Possibly
instantiated by an (allo)graph? The abstract vs. concrete nature of the
grapheme might merit some discussion here.

*** Indeed. We make this clear by stating:

A grapheme is the basic, minimally distinctive symbol of a particular writ- ing system, alike to the phoneme is an abstract representation of a distinct sound in a speci c language.  The term grapheme was modeled a er the term phoneme and represents a contrastive graphical unit in a writing system (see Kohrt 1986 for a historical overview of the term grapheme).

- "For example, <x> in English orthography represents a combination of
the phonemes /k/ and /s/.": give an example, e.g. /IndEks/. Note that
"Xerox" or "example" use <x> for other phonemes or phoneme combinations,
and that /sOks/ uses the phonemes /ks/ which are not represented by <x>.

- It might be interesting to briefly mention the non-trivial mapping of
phonemes on graphemes, e.g. give/gin/jingle, where the graphemes <g>/<j>
and the phonemes /g/dZ/ have a complex mapping.

*** Thanks for these thorough comments. We've updated this sentence to:

For example, <x> in English orthography sometimes represents a
combination of the phonemes /k/ and /s/, as in the word `index' [ˈɪnˌdɛks]; other times 
it is pronounced as /z/, such as in the words `Xerox' [ˈzɪrˌɑks] and `example' [ɪɡˈzæmpəl]. As one can see, there can be non-trivial mappings between graphemes and phonemes in orthographies like English, e.g.\ `give', `gin', `jingle', where the graphemes 
<g> and <j> and the phonemes /g/ and /dʒ/ have a complex mapping.

***

1.4.
The Unicode [Ss]tandard
- "distinction between universal (ASCII) versus language-particular
(font).": The meaning of this escapes me.

*** Fixed throughout.

- "However, ... ": the discussion of locales should not be found this
early on in the introduction of Unicode. Suggest moving it down.

*** As suggested by reviewer 1, we have spun the Unicode Approach section (previously in Chapter 1) into its own chapter, now Chapter 2.

Character encoding system
- "the character encoding is represented ... , which is used to encode a
set of characters": rephrase or delete

*** What's wrong with this?

- "non-negative": delete. You are talking to a human audience. No human
will try negative integers as a code point.

*** That's the core definition -- we keep it.

- Footnote 10: the equivalent of hex 0070 would be binary
0000000011100000. Since you use leading zeros in hex, you should do the
same for binary. But this footnote can probably be deleted. People who
know about different base encodings will find this trivial, and people
who don't will find this frightening and will not profit from the
description.

*** We have removed this footnote.

- "A font is a set of glyphs linked to code points": --> "A font is a
set of connections between glyphs and code points"

*** Updated: "In this approach, a \textsc{font} is then simply a collection of glyphs connected to code points."

- "(though note ...)" : put in footnote

*** Fixed.

- aptly-names --> aptly-named

*** Fixed.

- "PRECOMPOSED code point" --> "PRECOMPOSED CODE POINT" (or "PRECOMPOSED
WHATEVER", but not bare "PRECOMPOSED")

*** Fixed.

- "the Unicode Standard offers different kinds of normalization ...":
normalization is a tough concept to grasp. Find exactly one place where
to discuss it and keep discussions of it elsewhere to a minimum. Either
don't discuss it there at all, or mention it and cross-refer to the
in-depth discussion.

*** Ok, we've added a pointer to where we talk about Unicode Normalization in detail:

We discuss Unicode Normalization in detail in Section \ref{pitfall-canonical-equivalence}.

- "For example, the sequence <c> ...": this sentence no verb.

*** Now: "Examples include the sequence..."

- instead of Slovak and Sisaala, readers might find English <ch> or
French <oe> more useful. In the context of <oe>, one could also discuss
whether <oe> and <œ> are the same grapheme, and what the status of its
composing characters is, if any. A little chart or illustration
exemplifying graphemes, characters, glyphs, tailored graphemes and the
relation between then (meronymic and mapping) might be useful here.

*** Sure readers might, but we like these examples. The Slovak example also illustrates a use case that the Unicode Standard describes in detail (http://unicode.org/reports/tr29/), namely that: " “ch” may be considered a grapheme cluster in Slovak, for processes such as collation."

We've updated the text and we point to the Unicode Standard's annexes for text segmentation and normalization, which we feel, provide excellent illustrations of their 
decisions.

*** We've removed this footnote.

Footnote 10: the equivalent of hex 0070 would be binary
0000000011100000. Since you use leading zeros in hex, you should do the
same for binary. But this footnote can probably be deleted. People who
know about different base encodings will find this trivial, and people
who don't will find this frightening and will not profit from the
description.

Chapter 2
- Next to the notion of "pitfall", it may or may not be useful to
introduce the notion of "gotcha". This is synonymous, as far as I can
tell, but it emphasizes that there is a deliberate design decision which
users are often not aware of and which leads to unsuspected (but
completely legit) behaviour.

*** This is a nice suggestion, but I don't want to divide the notion of pitfall further because it might confused the reader.

- "differently as expected" --> than
- "practical use cases-that" --> than

*** Fixed.

Section 2.2
- It appears that before "First", there would be "Zeroth: a 1:1 mapping,
e.g. <t>:'t'"

*** We say this in the preceding paragraph, which should suffice: "A single Unicode character may of course be rendered as a single glyph."

- Subheadings should be useful here.

*** This seems like overkill for such a small paragraph. Instead I've split the large paragraph into smaller paragraphs, e.g. "first", "second".

- I do not get the Tamil part. Are you talking about U+0B94ஔ, which
visually looks like a combination of U+0B93ஓ and U+0BA9ன? If so, give
these glyphs. But for the case in point, Sinhala might actually be a
more pertinant example. Sinhala U+0DD9 ෙ is one code point and is made
up of one glyph, which looks like a spiral. U+0DDB ෛ is one code point
and is made of twice that spirally glyph. From what I understand from
the passage, this would be a suitable example. If not, I have probably
misunderstood the passage, and the meaning should be clarified.

*** I added the glyphs and the additional example: Second, a single character may be rendered as a sequence of multiple glyphs. For example, in Tamil one Unicode character may result in a combination of a consonant and vowel, which are rendered as two adjacent glyphs by fonts that supports Tamil, e.g. \textsc{tamil letter au} at \uni{0B94} represents a single character <ஔ>, composed of two glyphs <ஓ> and <ன>. Perhaps confusingly, in the Unicode Standard there are also two individual characters, 
i.e.\ \textsc{tamil letter oo} at \uni{0B93} and \uni{0BA9} \textsc{tamil letter nnna}, each of which is a glyph. Another example is Sinhala \textsc{sinhala vowel sign kombu deka} at \uni{0DDB} <ෛ>, which is visually two glyphs, each represented by \textsc{sinhala vowel sign kombuva} at \uni{0DD9} < ෙ>.

- in the discussion of the storeys of <a> and <g>, the elements
currently put between <> are not graphemes but glyphs. Do not use angle
brackets here, but find a way to design glyphs, e.g \fbox{}ing them

*** I think this is overkill -- graphical elements, whether Unicode characters of glyphs, are provided in <>'s, as is standard throughout the writing systems literature.

- "tweaking baseline and/or kerning". These concepts cannot be assumed
familiar. Either delete this passage, or add explanations in a footnote,
or put the complete baseline/kerning passage in a footnote and expand.

*** Removed.

Chapter 3
- general comment: replace single quotes by double quotes everywhere.
Occasionally, italics might be an option, too.

*** We understood that quotes and double quotes should follow these principles:

- Double quotation marks are generally used for distancing, in particular in the following situations:

	- when a passage from another work is cited in the text (e.g. According to Takahashi (2009: 33), “quotatives were never used in subordinate clauses in Old Japanese”); but block quotations do not have quotation marks;

	- when a technical term is mentioned that the author does not want to adopt, but wants to mention, e.g. This is sometimes called “pseudo-conservatism”, but I will not use this term here, as it could lead to confusion.

- Single quotation marks are used exclusively for linguistic meanings, as in the following: Latin habere ‘have’ is not cognate with Old English hafian ‘have’.

*** So I updated the quotation style accordingly (I hope). The old terms in this chapter are in double quotes, standard terms are now italicized.

- the IPA Association --> the International Phonetic Association

*** Fixed.

- when the twain shall meet --> when the twain met OR when the twain
should meet

*** This is a fix phrase ("never the twain shall meet" meaning "used to suggest that two things are too different to coexist") and I would like to keep it as is.

- Footnote 2 "IPA(" --> "IPA ("

*** Fixed.

Section Principles
- "the difference in English between an aspirated /p/ in [] or an
unreleased /p/ in []": I believe that if aspiration of /p/ is marked,
then aspiration for the t's should be marked as well (and unreleased
final /t/, if applicable). Maybe find another pair of examples.

*** For the case of completeness, we added /t/: "difference in English between an aspirated /p/ and /t/ in [spʰot̚] and an unrealeased /p/ in [stʰop̚]."

- German, Dutch, English and French /t/: The mapping of dental/aspirated
to the languages seems odd. Standard German /t/ is not dental, but Dutch
/t/ is. European French /t/ is not dental (but Canadian French /t/ is).
If this is taken from some other resource, a reference should be given.
In any case, that list cannot be taken to be self-evident.

*** We've updated this passage to reflect the dental/alveolar areas:

A phonemic transcription includes the conventions of a particular language's phonological rules. These rules determine the realization of that language's phonemes. For example, in the transcription of German, Dutch, English and French a symbol <t> might be used for the voiceless plosive in the alveolar and dental areas.

- "Similarly... Similarly" in close succession

*** Fixed.

Section IPA numbers
"assigning ... assigned ... assigned". More lexical variation, please.

*** Fixed: "The process of assigning standardized computer
codes to phonetic symbols was given to the \textsc{workgroup on computer..."

"three-digit number numerical directory of digit triples". This seems
like a pleonastic tautology to me.

*** Simplified: "The IPA Number is expressed as a three-digit number. The first digit 
indicates the symbol's category \citep{Esling1990,EslingGaylord1993}:"

- Footnotemark ".⁷.": delete one period

*** Fixed.

- "as an IPA symbol codings": not sure what is meant here, but agreement
is odd

*** Fixed: "One solution was the ASCII-ification of the IPA, which simply involved 
defining keyboard-able sequences consisting of ASCII combinations as IPA symbols."

- "(Computer 1985,1986,1988)": please fix this in the bib file

*** Fixed.

- "16 bit" --> "16-bit"

*** Fixed.

- "were published tables" --> "were published as tables"

*** Fixed.

- "inline" --> "in line"

*** Fixed.

- "development linguistic insights" --> "development of linguistic
insights."

*** Fixed.

- "along the line then" --> "along the line than"

*** Fixed.

Chapter 4
Footnote 1: Moran 2012 --> Moran (2012)

** We use parentheses in reference to the author vs in reference to the work, e.g. as noted by Moran (2012) vs as stated in Moran 2012. We can change this for LSP if this is not consistent with your practices.

4.3.
the first bullet point discusses the use of the apostrophe for the
glottal stop. This is indeed related to phonetics and Unicode, but the
glottal stop is not represented by an apostrophe-like thing in IPA, so
this problem is unrelated to the chapter entitled "IPA meets Unicode".
Restrict this to ejective marker.

*** I think this depends on one's typography know-how. I am reluctant to change the description because although in the strict sense it is not an apostrophe, in Umgangssprache I often hear it referred to as one. Better to keep it in -- also because the apostrophe is used to represent ejectives due to easy keyboarding. @cysouw ?

- discussion of slanted glyphs: it might be worthwhile to note that we
had to use the slanted approach for "A dictionary and grammatical
outline of Chakali" (http://langsci-press.org/catalog/book/74), where
the two different a's are used in the orthography.

*** I added a sentence noting this: "This approach was taken by the Language Science Press to distinguish between the two different orthographic <a>'s in Chakali in 
\cite{Brindle2017}.\footnote{\url{http://langsci-press.org/catalog/book/74}}"

4.5. multiple encoding options
- another issue which might be relevant here is the "i dot-suppression".
When putting an acute on an <i>, the dot will vanish (already in the
precomposed form, so this does not seem to be a font choice.). This will
make the combination of <i>+<´> look like the combination of <ı>+<´>.
Related to that is the unavailability of dotless barred i, so <ɨ>+<´>
will give you <́ɨ>, with both dot and acute. This is a case where the
combining overlay bar is useful, I will return to this below. I am not
sure whether the dot-suppression should be discussed here, or whether it
might be some other kind of pitfall, but it should be mentioned somewhere.

*** It seems most appropriate to mention this problem at the end of Section 5.3 Pitfall: IPA homoglyphs in Unicode and before and Section 5.4 Pitfall: Homoglyphs in IPA since it highlights both issues. Thus we write:

Lastly, before we move on from the pitfall of IPA homoglyphs in Unicode to the pitfall of 
homoglyphs in IPA, we are aware of one example that illustrates both pitfalls. 
Consider for example what one reviewer coined \textsc{i dot-suppression}. 
When combining, say the \textsc{latin small letter i} <i> at \uni{0069} with 
the \textsc{comining acute accent} <\dia{0301}> at \uni{0301}, the result is 
 the combination of these two characters into <í> or the associated 
precomposed form \textsc{latin small letter i with acute} <í> at \uni{00ED}. 
Typographically, the accent mark replaces the dot. In IPA, the <í> denotes 
a high front unrounded vowel with high tone. However, the result of losing 
the dot makes this IPA symbol look very similar to the near-high near-front 
unrounded vowel <ɪ>, when it also has the high tone marker: <ɪ́>.

- Footnote 7: when discussing superscript numbers for tone, it could be
worthwhile to say a word about Unicode superscript numbers
(https://en.wikipedia.org/wiki/Superscripts_and_Subscripts) vs. using
the text processor superscript formatting.

*** We've updated this footnote:

[8] Not sanctioned by the IPA, but nevertheless widely attested, is the use of superscript numbers for marking tones, also proposed by Chao (1930). One issue to note here is that superscript numbers can be either regular numbers that are forma ed as superscript with a text processor, or they can be separate superscript characters, as defined in the Unicode Standard (see: https://en.wikipedia.org/wiki/Superscripts_and_Subscripts). This divide means that searching text is dependent on how the author formatted or encoded the superscript numbers.

- "only one options" --> "only one option"

*** Fixed.

- missing decomposition: the case of accented barred i could be
discussed here.

*** Fixed.

4.9.
- "Tone letters are normally written ...": give the Unicode code points
here as well, in analogy to bullet points 1 and 3.

*** Added.

4.10.
- "but strangely enough ...": delete. The cedilla issue has been
discussed before.

*** Removed.

- "the diacritics on top of a character": to me, "on top" is a synonym
of "above", so an acute would be on top of an <e> for instance. My
interpretation might be unwarranted, but maybe another preposition like
"across" can be found, or something like "overlay".

*** Good call! Updated with the term overlay: "overlay diacritics like <\dia{0334}>"

- "removed symbols as labels them as clicks": rephrase

*** This paragraph has been updated:

These decisions have practical consequences for transcribers 
of IPA, particularly those who wish to follow current recommended practices of 
encoding electronic text in the Unicode Standard. For example, the Unicode 
Standard contains \textsc{latin small letter turned t} <ʇ> at U+0287, 
which is no longer part of the IPA. It still exists, however, in the Unicode 
\textsc{IPA Extensions} block, with the comment ``dental click (sound of `tsk tsk')''. 
In such cases, the IPA transcriber must know the status of legacy symbols in the current 
version of the IPA and the correct characters in the Unicode Standard.

- "Mc Laughlin": delete space?

*** Fixed.

4.12.
- Drop Moran (2012) in first sentence.

*** Fixed.

4.13.
- is widened-IPA a superset of valid-IPA or only a superset of
strict-IPA? A diagram showing the relation of the different sets might
be handy.

*** We'll pass on a pie chart and instead give the number of characters and charts for the user.

- I feel that superscript nasals for presnalized consonants should be
discussed somewhere. I am not sure as to their status in IPA, but they
are certainly used a lot. The current list only gives the superscript n,
but at least m and engma should be discussed.

*** We added a footnote regarding these phonetic symbols that come from non-IPA traditions:

The Unicode Standard defines the Phonetic Extensions block that defines symbols used in phonetic notation systems, including the Uralic Phonetic Alphabet, Americanist and Russianist phonetic notations, Oxford English and American dictionaries, etc. Among other symbols, the Phonetic Extensions block includes the superscript letters <m, ŋ, b>, which are not valid IPA characters, although we have seem them used in linguistic practice.

5.1
- "Although these variants are visually homoglyphs": this has been
discussed under normalization. Shorten or delete this and add a cross-ref

*** We make reference to where the reader can get more inforamtion

- "For example can": rephrase

*** Fixed.

5.2.
- "possibly ... possibly": rephrase

*** Fixed.

5.3. Formal specification
- give a figure with an open file where you indicate the relevant
portions at the very beginning of this section. The prose description is
very dense, and an illustration will make it easier to relate.

*** We don't understand this comment. The formal description is meant to be formal.

Section Implementation
- can orthography formats handle negative lookahead/lookbehind? Note
that this is different from a positive look for the set complement if
the string looked for is longer than 1.

*** They can handle anything that can be encoded with regular expressions (and if the regular expression has that functionality).

B6.
I feel that a language which has both <ch> and <sch> will get you
different results depending on global or linear algorithm.
"Ascha" should give you "A sch a" with the linear approach, but "a s ch
a" with the global approach iff "ch" is matched before "sch".

[ ] todo

C8.
- "To treat the profile literal" --> literally

*** Fixed.

Chapter 6
- "Each section has a number of subsections that ... "

*** Fixed.

- the example "aäaaöaüaa" will confuse a lot of ordinary working
linguists. There are too many vowels, and the representation of "aa" by
<x> is weird. I suggest using the German word "Schächtelchen", where you
have both <sch> and <ch> to show the importance of ordering, and a
dieresis to show decomposition (which seem to be the main didactic
aspects of this passage to me).

*** Schächtelchen is a great word and we appreciate the suggestion that "aäaaöaüaa" is confusing. So we have added Schächtelchen as an example to both the book and the online recipes.

We therefore built a different example:

"ĉháɾã̌ctʼɛ↗ʐː| k͡p"

which captures a hypothetical form from which the child can generalize.

- the code examples should be numbered

*** We don't agree. These examples stand alone and are not necessarily sequential.

- the tabs in the code should have the same length, so that the headers
and the columns align

*** Fixed.

- "edit the profile: more": --> "edit the profile: vim" (or any other
editor, but not more)

***

- "The location can be found in the following command in R.": provide
the command (or take care of the page break)

***

- "Try tokenize to test the functionality": why is there an error
message????

***

- "These webapps are available online at TODO": provide URL

***

- "we first assign the two strings to a variable test": use a more
descriptive name than "test"

***

Section Using an orthography profile skeleton
- " It is also possible ...It is also possible": rephrase

*** Fixed.

- maybe this passage should be written with Italian or Spanish
orthography as an example since English orthography is such a mess. Why
are "ie" and "oo" digraphs? If so, would "ti" in "nation" count as a
digraph as well? And why is the second "e" in "scheme" listed at all? It
does not correspond to any phoneme (but arguably it influences the first
"e"). Also, for me, "sch" in "scheme" consist of two phonemes (/sk/),
just like "sch" in "mischief (/stS/)". But the passage "Now mischief is
parsed correctly, but scheme is wrong" suggests that the author sees
this differently???

- Footnote 3: delete. Feeding and bleeding are needlessly complicating
the exposition

- RESULT_ERRORS.TSV: I stronly suggest to either create an empty file
when no errors are reported, or to include "no errors" in the file. The
meaningful absence of computer files can cause all kinds of problems.

- "Such contextually-specified graphemes are based on regular
expressions so you can also use regular expressions in the descriptions
of the context": who would've thought that!

- "It is important to realize ... ": delete. The need for escape
sequences has been discussed earlier.

- Table 6.11: I was wondering whether graphemes can be in several
classes. "Y" could be in the consonant class and in the vowel class for
instance. Can orthography profiles handle this, and if they can, how
would this be entered?

*** Note, we have also added a section that highlights the four additional use cases that we present as Jupyter Notebooks in our Github repository of orthography profile recipes:

https://github.com/unicode-cookbook/recipes/

### 2018-03-06 book/chapters/introduction.tex
these changes almost complete the remaining reviews

@@
 was used for Cyrillic characters, part 6 for Arabic, and part 7 for
-Greek.\footnote{In effect, because $16 = 2^4$, this means that ISO/IEC~8859 was
-actually an $8+4=12$-bit encoding, though with very many duplicates by design,
-namely all ASCII codes were repeated in each 8-bit code page. To be precise,
-ISO/IEC~8859 used the 7-bit ASCII as the basis for each code page, and defined
-16 different 7-bit extensions, leading to $(1+16) \cdot {2^7} = 2,176$ possible
-characters. However, because of overlap and not-assigned codes points the actual
-number of symbols was much smaller.} This system almost immediately was
+Greek. % \footnote{In effect, because $16 = 2^4$, this means that ISO/IEC~8859 was
+%actually an $8+4=12$-bit encoding, though with very many duplicates by design,
+%namely all ASCII codes were repeated in each 8-bit code page. To be precise,
+%ISO/IEC~8859 used the 7-bit ASCII as the basis for each code page, and defined
+%16 different 7-bit extensions, leading to $(1+16) \cdot {2^7} = 2,176$ possible
+%characters. However, because of overlap and not-assigned codes points the actual
+%number of symbols was much smaller.} 
+This system almost immediately was
 understood to be insufficient and impractical, so various initiatives to extend
@@
 
-Breaking it down further, a script consists of \textsc{graphemes}, and graphemes
-consist of \textsc{characters}. In the linguistic terminology of writing
+Breaking it down further, a script consists of \textsc{graphemes}, which are writing 
+system-specific minimally distinctive symbols (see below). Graphemes may consist of one or more 
+\textsc{characters}. The term \textsc{character} is overladen. In the linguistic terminology of writing
 systems, a \textsc{character} is a general term for any self-contained element
-in a writing system.\footnote{There is a second interpretation of the term
-\textsc{character}, i.e.~a conventional term for a unit in the Chinese writing
-system \citep{Daniels1996}. This interpretation will not be further explored in
-this paper.} Although in literate societies most people have a strong intuition
+in a writing system. A second interpretation is used as a conventional term for a unit in the Chinese writing
+system \citep{Daniels1996}. In technological terminology, a \textsc{character} 
+refers to the electronic encoding of a component in a writing system that has semantic 
+value (see Section \ref{character-encoding-system}). Thus in this work we must navigate 
+between the general linguistic and technological terms for \textsc{character} 
+and \textsc{grapheme} because of how these notions are defined and how they relate at the intersection 
+between the International Phonetic Alphabet and the Unicode Standard (Chapter \ref{ipa-meets-unicode}).
+
+Although in literate societies most people have a strong intuition
 about what the characters are in their particular orthography or orthographies,
@@
 
-A \textsc{diacritic} is a mark, or series of marks, that may be above, below, or
-through other characters \citep{Gaultney2002}. Diacritics are sometimes used to
+A \textsc{diacritic} is a mark, or series of marks, that may be above, below, 
+before, after, through, around, or between other characters \citep{Gaultney2002}. Diacritics are sometimes used to
 distinguish homophonous words, but they are more often used to indicate a
@@
 whether, for example, <e>, <è>, <a> and <à> should be considered four
-characters, or different combinations of three characters. In general, multiple
-characters together can form another character, and it is not always possible to
+characters, or different combinations of three characters, i.e.\ <a>, <e>, and <\dia{0300}>. 
+In general, multiple characters together can form another character, and it is not always possible to
 decide on principled grounds what should be the basic building blocks of an
@@
 writing systems, in cases in which a single character represents a combination
-of two or more phonemes. For example, <x> in English orthography represents a
-combination of the phonemes /k/ and /s/.
+of two or more phonemes. For example, <x> in English orthography sometimes represents a
+combination of the phonemes /k/ and /s/, as in the word `index' [ˈɪnˌdɛks]; other times 
+it is pronounced as /z/, such as in the words `Xerox' [ˈzɪrˌɑks] and `example' [ɪɡˈzæmpəl]. 
+As one can see, there can be non-trivial mappings between graphemes and phonemes in 
+orthographies like English, e.g.\ `give', `gin', `jingle', where the graphemes 
+<g> and <j> and the phonemes /g/ and /dʒ/ have a complex mapping.
 

### 2018-03-06 book/chapters/ipa_background.tex
these changes almost complete the remaining reviews

@@
 difference in English between an aspirated /p/ and /t/ in [spʰot̚] and 
-an unrealeased /p/ in [stʰop̚].
+an unreleased /p/ in [stʰop̚].
 
@@
 the articulatory properties of the IPA symbols themselves are rather succinctly
-summarised by the normative description in the \textit{Handbook}, it is common
+summarized by the normative description in the \textit{Handbook}, it is common
 in practical applications that the transcribed symbols do not precisely
@@
 German, Dutch, English and French a symbol <t> might be used for the voiceless
-plosive in the alveolar area. This symbol is sufficient for a succinct
+plosive in the alveolar and dental areas. This symbol is sufficient for a succinct
 transcription of these languages because there is no further phonemic
@@
 % \footnote{The marking of tone was extended (from characteristically Africanist practice) to include a second system for marking linguistic tones (the `Chao tone letters'), a much used convention based on musical staff to describe pitch in by Yuen Ren Chao (1892--1982).}
+
 \subsection*{IPA Numbers}
@@
 The working group decided insightfully that the computing-coding convention 
-for the IPA should be independent of computer environments or formats (e.g.\ ASCII), 
-i.e.\ the IPA Number was not meant to be implemented directly in a computer 
-encoding. The working group report's declaration includes the following explanatory 
+for the IPA should be independent of computer environments or formats, 
+i.e.\ the IPA Number was not meant to be encoded at the bit pattern level.
+The working group report's declaration includes the following explanatory 
 remarks \citep[82]{International1989report}:
@@
 character sets of ISO/IEC 10646 and the Unicode Standard later converged (see
-Section~\ref{the-unicode-approach}), the IPA proposal was
+Chapter~\ref{the-unicode-approach}), the IPA proposal was
 included in the Unicode Standard Version 1.0 -- largely as we know it
@@
 It was a long journey, but the goal of achieving a single multilingual computing
-environment has largely been accomplished. Current users, however, must cope
+environment has largely been accomplished. As such, we should not dismiss the IPA Numbers 
+or pre-Unicode encoding attempts, such as SAMPA/X-SAMPA, as misguided. The parallels 
+between the IPA Numbers and Unicode Code points, for example, is striking because both the IPA and 
+the Unicode Consortium came up with the solution of an additional layer of indirection (an abstraction layer) 
+between symbols/characters and encoding on the bit pattern level. SAMPA/X-SAMPA is also still useful 
+as an input method for IPA in ASCII and required by some software.
+
+Current users of the Unicode Standard must cope
 with the pitfalls that were dug along the way, as will be discussed in the next

### 2018-03-06 book/chapters/ipa_meets_unicode.tex
updates section name
closes issue #6
closes issue #7
updates IPA reference
these changes almost complete the remaining reviews

@@
 various pitfalls with the encoding of IPA in the Unicode Standard. We will
-specifically refer to the 2005 version of the IPA \citep{IPA2005} and the 7.0
-version of Unicode \citep{Unicode2014}.
+specifically refer to the 2015 version of the IPA \citep{IPA2015} and the 10.0.0 version of Unicode \citep{Unicode2017}.
 
@@
         Section~\ref{pitfall-different-notions-of-diacritics}). To harmonize the 
-        bevavior of <ʰ> with other superscript letters, we propose to 
+        behavior of <ʰ> with other superscript letters, we propose to 
         standardize on the modifier letter at \uni{02B0} (see 
@@
 instance, uvular and pharyngeal consonants only occur around /a/ and /o/. Hence,
-the characterization of /a/ and /o/ as a natural class of (i.e., [+back]
+the characterization of /a/ and /o/ as a natural class of (i.e.\ [+back]
 vowels), as opposed to /i/ and /e/.''} They simply use <a> as long as there is
@@
 
+Lastly, before we move on from the pitfall of IPA homoglyphs in Unicode to the pitfall of 
+homoglyphs in IPA, we are aware of one example that illustrates both pitfalls. 
+Consider for example what one reviewer coined \textsc{i dot-suppression}. 
+When combining, say the \textsc{latin small letter i} <i> at \uni{0069} with 
+the \textsc{combining acute accent} <\dia{0301}> at \uni{0301}, the result is 
+ the combination of these two characters into <í> or the associated 
+precomposed form \textsc{latin small letter i with acute} <í> at \uni{00ED}. 
+Typographically, the accent mark replaces the dot. In IPA, the <í> denotes 
+a high front unrounded vowel with high tone. However, the result of losing 
+the dot makes this IPA symbol look very similar to the near-high near-front 
+unrounded vowel <ɪ>, when it also has the high tone marker: <ɪ́>. To boot, 
+when an accent mark is added to \textsc{latin small letter i with stroke} <ɨ> at 
+\uni{0268}, the dot is not suppressed but retained, i.e. <ɨ́>.\footnote{According 
+to the Unicode Standard, \textsc{latin small letter i with stroke} <ɨ> at 
+U+0268 cannot be decomposed into, say, \textsc{latin small letter i} <i> at 
+U+0069 and \textsc{combining short stroke overlay} <\dia{0335}> at U+0335. 
+We discuss the pitfall of missing decomposition forms 
+in Section \ref{pitfall-missing-decomposition}.}
+
 % ==========================
@@
 The first case in which the IPA allows for different encodings is the question
-of how to transcribe tone. There is an old tradition to use diacritics on vowels
-to mark different tone levels, e.g. <ȅèée̋>.\footnote{To make things even more
-complex, there are at least two different Unicode homoglyphs for the low and
+of how to transcribe tone (cf.\ \cite{Maddieson1990}). There is an old tradition 
+to use diacritics on vowels to mark different tone levels \citep{IPA1949}. Prior to the 1989 Kiel 
+convention, IPA-approved tone symbols included diacritics above or below the vowel or syllable, 
+e.g.\ high and low tones marked with macrons (<\diaf{0304}>, <\diaf{0331}>), and acute and grave 
+accents for high rising tone <\diaf{0301}>, low rising tone <\diaf{0317}>, high falling tone 
+<\diaf{0300}> and low falling tone <\diaf{0316}>. These tone symbols, however,  
+had failed to catch on (probably) due to aesthetic objections and matters of adequacy for 
+transcription \citep[29]{Maddieson1990}.
+
+After the 1989 Kiel convention, the accent tone symbols were updated to the tradition 
+that we are familiar with today and which was already in wide use by Africanists and others, 
+namely level tones <\diaf{030B}, \diaf{0301}, \diaf{0304}, \diaf{0300}, \diaf{030F}> 
+and contour tones <\diaf{030C}, \diaf{0302}, \diaf{1DC4}, \diaf{1DC5}, \diaf{1DC8}>.\footnote{To make things even more
+complicated, there are at least two different Unicode homoglyphs for the low and
 high level tones, namely <\diaf{0340}> \textsc{combining grave tone mark} at
@@
 vs.<\diaf{0301}> \textsc{combining acute accent} at \unif{0301} for high tone.} 
-The IPA also proposes the option of tone letters, e.g. <˥˦˧˨˩>, which are much
-less often used, but are more consistent for contours.\footnote{Not sanctioned
-by the IPA, but nevertheless widely attested, is the usage of superscript
-numbers for marking tones.} Tone letters in the IPA have five different levels,
+In addition, the IPA also adopted tone letters developed by \citet{Chao1930}, e.g. <˥˦˧˨˩>, 
+which were in wide use by Asianists.\footnote{Not sanctioned
+by the IPA, but nevertheless widely attested, is the use of superscript
+numbers for marking tones, also proposed by \citet{Chao1930}. One issue to note here is 
+that superscript numbers can be either regular numbers that are formatted as 
+superscript with a text processor, or they can be separate superscript characters, 
+as defined in the Unicode Standard (see: \url{https://en.wikipedia.org/wiki/Superscripts_and_Subscripts}). 
+This divide means that searching text is dependent 
+on how the author formatted or encoded the superscript numbers.} 
+Tone letters in the IPA have five different levels,
 and sequences of these letters can be used to indicate contours. Well-designed
@@
 both the keyboard \textsc{latin small letter g} and the \textsc{latin small
-letter script g} are valid input characters for the voiced velar plosive.
+letter script g} are valid input characters for the voiced velar plosive \citep[19]{IPA1999}.
 Unfortunately, this decision further introduces ambiguity for linguists trying
@@
 definition combined with the character before them (to form so-called default
-grapheme clusters, see Section~\ref{the-unicode-approach}). In contrast, Spacing
+grapheme clusters, see Chapter~\ref{the-unicode-approach}). In contrast, Spacing
 Modifier Letters are by definition \emph{not} combined into grapheme clusters
@@
 
-	\item[] Tone letters, like: 
+	\item[] Tone letters, including but nut limited to: 
 	\begin{itemize} 
@@
 	  \item[] <˨> \textsc{modifier letter low tone bar} at \uni{02E8}
-	  \item[] and others like this
-	\end{itemize}
-
-	\item[] Superscript letters, like:
+	\end{itemize}
+
+	\item[] Superscript letters, including but not limited to:\footnote{The Unicode Standard defines the 
+	  \textit{Phonetic Extensions} block that defines symbols used in phonetic notation 
+	  systems, including the Uralic Phonetic Alphabet, Americanist and Russianist phonetic notations, 
+	  Oxford English and American dictionaries, etc. Among other symbols, the 
+	  \textit{Phonetic Extensions} block includes 
+	  the superscript letters <\textsuperscript{m, ŋ, b}>, which are not valid 
+	  IPA characters, although we have seen them used in linguistic practice.}
 	\begin{itemize}
@@
 	  \item[] <\dia{207F}> \textsc{superscript latin small letter n} at \uni{207F}
-	  \item[] and many more like this
 	\end{itemize}
@@
          character), so there is no consistent direction in which this
-         diactritic should bind. Note that there is yet another homoglyph,
+         diacritic should bind. Note that there is yet another homoglyph,
          namely the \textsc{combining latin small letter h} at \uni{036A}.
@@
 accordance with the principles of the IPA, as outlined in Section
-\ref{the-international-phonetic-alphabet}, distinct symbols are favoured for
+\ref{the-international-phonetic-alphabet}, distinct symbols are favored for
 cases of phonological contrast. Further, also convenience of display in the chart
@@
       There are some rare articulations that make
-      use of an epilaryngeal phonation mechanism (e.g.,~the “sphincteric vowels”
+      use of an epilaryngeal phonation mechanism (e.g.\ the “sphincteric vowels”
       of~!Xóõ). To represent these vowels, we propose to use the modifier <\dia{1D31}>
@@
 % ==========================
-\section{Recommendations}
+\section{Unicode IPA Recommendations}
 \label{ipa-recommendations}
@@
 Informally speaking, valid-IPA represents the current state of the IPA
-\citep{IPA2005}. Strict-IPA represents a more constrained version of IPA, while
+\citep{IPA2015}. Strict-IPA represents a more constrained version of IPA, while
 widened-IPA is a slightly extended version of IPA, allowing a few more symbols.

### 2018-03-06 book/chapters/orthography_profiles.tex
these changes almost complete the remaining reviews

@@
 Given these goals, Unicode locale descriptions (see
-Section~\ref{the-unicode-approach}) might seem like the ideal orthography
+Chapter~\ref{the-unicode-approach}) might seem like the ideal orthography
 profiles. However, there are various practical obstacles preventing the use of
@@
 transliteration as described in the Unicode Locale Data Markup
-Language.\footnote{See \url{http://www.unicode.org/reports/tr35/}} Finite-state 
+Language).\footnote{See \url{http://www.unicode.org/reports/tr35/}} Finite-state 
 transducers attempt to describe the mapping from input to output string directly 
@@
 tokenization from transliteration there is no problem with `feeding' and 
-`bleeding' of rules, common with transducers (cf.~Section~\ref{rule-ordering}
+`bleeding' of rules, common with transducers (cf.~Section~\ref{r-implementation}
 ).
@@
 need for multiple different orthography profiles to be applied in sequence (see
-Section~\ref{use-cases} on various exemplary use-cases). The result of any such
+Section~\ref{r-implementation} on various exemplary use-cases). The result of any such
 workflow will normally be a cleaned dataset and an explicit description of the
@@
 	\def\labelenumi{A\arabic{enumi}.} 
-	\item \textsc{A profile is a unicode UTF-8 encoded text file}, ideally
-       using NFC, no-BOM, and LF (see Section~\ref{pitfall-file-formats}), that
-       includes the information pertinent to the orthography. 
-	\item \textsc{A profile is a tab-separated CSV file with an obligatory header
-       line}. A minimal profile can have just a single column, in which case
-       there will of course be no tabs, but the first line will still be the
-       header. For all columns we assume the name in the header of the CSV file
-       to be crucial. The actual ordering of the columns is unimportant. Because 
-       tabs and newlines are field separators of the profile, these characters 
-       will lead to problems when they occur in the original data.\footnote{The main
-       reason to choose for tab-delimited, and not for quoted comma-separated files
-       is that we want the profiles to be easily manipulated by hand. Tab-separation 
-       is strongly preferred for this reason.}
-	\item \textsc{Metadata are added in a separate UTF-8 text file} using a basic
+	\item \textsc{A profile is a unicode UTF-8 encoded text file} 
+	   normalized following NFC (or NFD if specified in the metadata)
+       that includes the information pertinent to the orthography.\footnote{See 
+	   Section~\ref{pitfall-file-formats} in which we suggest to use NFC, 
+	   no-BOM and LF line breaks because of the pitfalls they avoid. 
+	   A keen reviewer notes, however, that specifying 
+	   a convention for line endings and BOM is overly strict because most 
+	   computing environments (now) transparently handle both alternatives. 
+	   For example, using Python a file can be decoded using the encoding 
+	   ``utf-8-sig'', which strips away the BOM (if present) and reads 
+	   an input full in text mode, so that both line feed variants ``LF'' and 
+	   ``CRLF'' will be stripped.}
+
+%	\item \textsc{A profile is a unicode UTF-8 encoded text file}, ideally
+%       using NFC, no-BOM, and LF (see Section~\ref{pitfall-file-formats}), that
+%       includes the information pertinent to the orthography. 
+	\item \textsc{A profile is a CSV file with an obligatory header
+       line}. A minimal profile must have a single column with the header ``Grapheme''. 
+	   For any additional columns, the name in the header of the CSV file
+       must be specified. The actual ordering of the columns is unimportant. Cells should 
+	   not contain tabs or newlines. The header list must be delimited 
+	   in the same way as the rest of the file's contents. Each record must be kept 
+	   on a separate line. Separate lines with comments are not allowed. Comments that
+       belong to specific lines must be put in a separate column of
+       the CSV file, e.g.~add a column called \textsc{comments}.
+
+\begin{comment}	   
+	\item \textsc{Metadata should be added in a separate UTF-8 text file} using a basic
        \textsc{tag: value} format. Metadata about the orthographic description
@@
        described in the profile. Further, (vi) the tokenization method and (vii)
-       the unicode normalisation used should be documented here (see below).
-    	\item \textsc{Separate lines with comments are not allowed}. Comments that
-       belong to specific lines will have to be put in a separate column of
-       the CSV file, e.g.~add a column called \textsc{comments}.            
+       the unicode normalization used should be documented here (see below).
+\end{comment}
+
+	\item \textsc{Metadata should be added in a separate UTF-8 text file} using 
+	   the JSON-DL dialect specified in \textit{Metadata Vocabulary for Tabular 
+	   Data}.\footnote{\url{https://www.w3.org/TR/tabular-metadata/}} This metadata 
+	   format allows easy inclusion of Dublin Core metadata,\footnote{\url{http://dublincore.org/}} 
+	   which should be used to specify information about the orthographic description 
+	   in the orthography profile. The orthography profile metadata should minimally include: 
+	   (i) author, (ii) date, (iii) title of the profile, (iv) a stable language identifier
+       encoded in BCP 47/ISO 639-3 of the target language of the profile, and (v)
+       bibliographic data for resource(s) that illustrate the orthography
+       described in the profile. Further, (vi) the tokenization method and (vii)
+       the Unicode normalization form used should be documented here (see below).
+	   In accordance with the \textit{Metadata Vocabulary for Tabular 
+	   Data}, the metadata's filename must consist of the orthography 
+	   profile's filename appended with ``-metadata.json''.\footnote{JSON-DL metadata 
+	   is also the choice for datasets conforming to the Cross-Linguistic Data Formats standard, 
+	   see: \url{http://cldf.clld.org/}.}
+
 \end{enumerate}

### 2018-03-06 book/chapters/pitfalls.tex
these changes almost complete the remaining reviews

@@
 like <sch>, are called \textit{tailored grapheme clusters} in Unicode parlance (see
-Section~\ref{the-unicode-approach}).
+Chapter~\ref{the-unicode-approach}).
 
@@
 name, the block and the character properties (see
-Section~\ref{the-unicode-approach}). Further details about the underlying intentions 
+Chapter~\ref{the-unicode-approach}). Further details about the underlying intentions 
 with which a character should be used
@@
 hidden.\footnote{All proposals and other documents that are the basis of Unicode
-decisions are avaialbe at \url{http://www.unicode.org/L2/all-docs.html}. The
+decisions are available at \url{http://www.unicode.org/L2/all-docs.html}. The
 actual decisions that make up the Unicode Standard are documented in the minutes
@@
 % ==========================
+\section{Pitfall: Encodings}
+\label{encoding}
+% ==========================
+
+% Section "Pitfall: File formats" may profit most from a more transparent terminology, distinguishing levels of encoding, rather than talking about how texts "appear inside some kind of computer file".
+
+Before we discuss the pitfall of different file formats in Section \ref{pitfall-file-formats}, it is pertinent to point out that the common usage of the term \textsc{encoding}, unfortunately, does not distinguish between \textit{encoded} sequences of code points and text \textit{encoded} as bit patterns.
+Recall, a code point is simply a numerical representation of some defined entity; or in other words, a code point is a character encoding-specific unique identifier or ID.
+In the Unicode Standard encoding, code points are numbers that serve as unique identifiers, each of which is associated with a set of character properties defined by the Unicode Consortium in the Unicode Character Database.\footnote{\url{https://www.unicode.org/ucd/}} The number of each code point can be \textit{encoded} in various formats, including as a decimal integer (e.g.\ 112), as an 8-bit binary sequence (01110000) or hexadecimal (0070).
+This example Unicode code point, \uni{0070}, represents \textsc{latin small letter p} and its associated Unicode properties, such as it belongs to the category Letter, Lowercase [Ll], in the Basic Latin block, and that its title case and upper case is associated with code point \uni{0050}.\footnote{See also Chapter \ref{the-unicode-approach}.}
+
+% 0070;LATIN SMALL LETTER P;Ll;0;L;;;;;N;;;0050;;0050
+
+The other meaning of encoding has to do with the fact that computers represent data and instructions in patterns of bits.
+A bit pattern is a combination of binary digits arranged in a sequence.
+And how these sequences are carved up into bit patterns is determined by how they are \textit{encoded}.
+Thus the term \textsc{encoding} is used for both sequences of code points and text encoded as bit patterns.
+Hence a Unicode-aware programmer might prefer to say that UTF-8, UTF-16, etc., are Unicode encoding systems because they determine how sequences of bit patterns are determined, which are then mapped to characters.
+The terminological issue here is that the Unicode Standard introduces a layer of indirection between characters and bit patterns, i.e.
+the code point, which can be encoded differently by different encoding systems.
+
+% Characters encoded, but not seen.
+Note also that all computer character encodings include so-called \textsc{control characters}, which are non-printable sometimes action-inducing characters, such as the null character, bell code, backspace, escape, delete, and line feed.
+Control-characters can interact with encoding schemes.
+For example, some programming languages make use of the null character to mark the end of a string.
+Line breaks are part of the text, and as such as covered by the Unicode Standard.
+But they can be problematic because line breaks differs from operating system-to-operating system in how they are encoded.
+These variants are discussed in Section \ref{pitfall-file-formats}.
+
+% This section should also mention that line breaks are actually part of the text, and as such also covered by the Unicode standard.
+Again regarding line breaks, the recommendation "that everybody use this encoding [only LF] whenever possible" (page 31) seems less well motivated (and specific) than most other recommendations in the book.
+Some common formats, e.g.
+CSV (as specified by  RFC 4180 [4]), specify "CRLF" as line break.
+Thus, the perceived "strong tendency" towards "LF" may be just that.
+
+% ==========================
 \section{Pitfall: File formats}
@@
 % ==========================
+\section{Pitfall: Incomplete implementations}
+\label{incomplete-implementations}
+% ==========================
+Another pitfall that we encounter when using the Unicode Standard is its incomplete implementation in different standards and programming languages, e.g.\ SQL, XML, XLST, Python.
+For example, although the Unicode Standard mandates that the comparison of Unicode text be done using normalized text, this is not the case with the equality operator ``=='' in Python.
+Furthermore, it is not always transparent what the operating system or specific software applications do when text is being copied and pasted.
+For example, copy and pasting the character sequence \uni{0061} \textsc{latin small letter a} <a> and \uni{0301} \textsc{combining acute accent} <\dia{0301}>, visually <á>, into the text editor TextWrangler leaves the sequence decomposed as two characters.
+But when pasting the decomposed sequence into RStudio, and other software programs, the sequence becomes precomposed as \uni{00E1} \textsc{latin small letter a with acute}, i.e.\ <á>.
+
+% One Unicode pitfall which may be worth adding is the problem with incomplete implementations of the Unicode Standard.
+This is a problem most big-enough standards suffer from, e.g.
+SQL, XML, XSLT.
+So although the Unicode Standard mandates that comparison of Unicode text should always be done using normalized text, this is not how the equality operator "==" is implemented for Unicode text in the Python programming language.
+
+% ==========================
 \section{Recommendations}

### 2018-03-06 book/chapters/unicode_approach.tex
these changes almost complete the remaining reviews

@@
 \section{Character encoding system}
+\label{character-encoding-system}
 
@@
 goal it is to support the interchange and processing of written characters and
-text in a computational setting. Underlyingly, the character encoding is
+text in a computational setting.\footnote{An insightful reviewer notes that the term \textsc{encoding} is used for both sequences of code points and text encoded as bit patterns.
+Hence a Unicode-aware programmer might prefer to say that UTF-8, UTF-16, etc., are Unicode encoding systems.
+The issue is that the Unicode Standard introduces a layer of indirection between characters and bit patterns, i.e.
+the code point, which can be encoded differently by different encoding systems.} Underlyingly, the character encoding is
 represented by a range of numerical values called a \textsc{code space}, which
@@
 hexadecimal, which then looks in Unicode parlance as
-\uni{0070}.\footnote{Hexadecimal (base-16) 0070 is equivalent to decimal
-(base-10) 112, which can be calculated by considering that $(0\cdot16^3) +
-(0\cdot16^2) + (7\cdot16^1) + (0\cdot16^0) = 7\cdot16 = 112$. Underlyingly,
-computers will of course treat this code point binary (base-2) as 11100000, as
-can be seen by calculating that $(1\cdot2^7) + (1\cdot2^6) + (1\cdot2^5) +
-(0\cdot2^4) + (0\cdot2^3) + (0\cdot2^2) + (0\cdot2^1) + (0\cdot2^0) = 64 + 32 +
-16 = 112$.} That encoded abstract character is rendered on a computer screen (or
+\uni{0070}.
+%\footnote{Hexadecimal (base-16) 0070 is equivalent to decimal
+%(base-10) 112, which can be calculated by considering that $(0\cdot16^3) +
+%(0\cdot16^2) + (7\cdot16^1) + (0\cdot16^0) = 7\cdot16 = 112$. Underlyingly,
+%computers may treat this code point as an 8-bit binary (base-2) sequence (11100000), as
+%can be seen by calculating that $(1\cdot2^7) + (1\cdot2^6) + (1\cdot2^5) +
+%(0\cdot2^4) + (0\cdot2^3) + (0\cdot2^2) + (0\cdot2^1) + (0\cdot2^0) = 64 + 32 +
+%16 = 112$.} 
+That encoded abstract character is rendered on a computer screen (or
 printed page) as a \textsc{glyph}, e.g. <p>, depending on the \textsc{font} and
@@
 canonical decomposition}), or compose as much as possible combinations (called
-\textsc{NFC}, \textsc{normalization form canonical composition}). In current
-practice of software development, NFC seems to be preferred in most situations
-and is widely proposed as the preferred canonical form.
+\textsc{NFC}, \textsc{normalization form canonical composition}).\footnote{See the 
+Unicode Standard Annex \#15, Unicode Normalization Forms (\url{http://unicode.org/reports/tr15/}), 
+which provides a detailed description of normalization algorithms and illustrated examples.} 
+In current practice of software development, NFC seems to be preferred in most situations
+and is widely proposed as the preferred canonical form. We discuss Unicode Normalization 
+in detail in Section \ref{pitfall-canonical-equivalence}.
 
 More difficult for text processing, because less standardized, is what the
-Unicode Standard terms \textsc{tailored grapheme clusters}. Tailored grapheme
-clusters are locale-dependent graphemes, i.e.~such combination of characters do
+Unicode Standard terms \textsc{tailored grapheme clusters}.\footnote{\url{http://unicode.org/reports/tr29/}} 
+Tailored grapheme clusters are locale-dependent graphemes, i.e.~such combination of characters do
 not function as text elements in all situations. Examples include the sequence
@@
 clusters are \textsc{tailored} in the sense that they must be specified on a
-language-by-language or writing-system-by-writing-system basis. The Unicode
-Standard provides technological specifications for creating locale specific data
+language-by-language or writing-system-by-writing-system basis. They are also 
+grapheme clusters in these orthographies for processes such as collation.\footnote{\url{https://www.unicode.org/glossary/#collation}}
+
+The Unicode Standard provides technological specifications for creating locale specific data
 in so-called \textsc{Unicode Locale Descriptions}, i.e.~a set of specification

### 2018-03-09 book/reviews/responses_reviewer1.txt
ready for review two

@@
 A2.
-A profile is a CSV file with an obligatory header line.
+A profile is a delimited text file with an obligatory header line.
 A minimal pro le must have a single column with the header “Grapheme”.
-For any ad- ditional columns, the name in the header of the CSV  le must be speci ed.
+For any additional columns, the name in the header must be speci ed.
 e actual ordering of the columns is unimportant.
-Cells should not con- tain tabs or newlines.
 e header list must be delimited in the same way as the rest of the  le’s contents.
@@
 Separate lines with comments are not allowed.
-Comments that belong to specific lines must be put in a separate column of the CSV file, e.g.~add a column called \textsc{comments}.
+Comments that belong to speci c lines must be put in a separate column of the  le, e.g.
+add a column called comments.
 
@@
 
-A3. Metadata should be added in a separate UTF-8 text file using the JSON-DL dialect speci ed in Metadata Vocabulary for Tabular Data.^ 3  This
-metadata format allows easy inclusion of Dublin Core metadata,^4 which should be used to specify information about the orthographic description in the orthography pro le.
-e orthography pro le metadata should min- imally include: (i) author, (ii) date, (iii) title of the pro le, (iv) a stable lan- guage identi er encoded in BCP 47/ISO 639-3 of the target language of the pro le, and (v) bibliographic data for resource(s) that illustrate the or- thography described in the pro le.
-Further, (vi) the tokenization method and (vii) the Unicode normalization form used should be documented here (see below).
-In accordance with the Metadata Vocabulary for Tabular Data, the metadata’s  lename must consist of the orthography pro le’s  lename appended with “-metadata.json”.^5
+A3.
+Metadata should be added in a separate UTF-8 text file using the JSON-LD dialect speci ed in Metadata Vocabulary for Tabular Data.3  is metadata format allows easy inclusion of Dublin Core metadata,4 which should be used to specify information about the orthographic description in the orthography pro le.5  e orthography pro le metadata should min- imally include provenance information including: (i) author, (ii) date, (iii) title of the pro le, and (iv) bibliographic data for resource(s) that illustrate the orthography described in the pro le.
+Crucially, the metadata should also specify (iv) a stable language identi er of the target language of the pro le using BCP 47/ISO 639-3 or Glo ocode as per the CLDF ontology.6 Further, the metadata  le should provide information about the orthogra- phy pro le’s structure and contents, including: (vi) its dialect description,7 (vii) the tokenization method (see below), and (ix) proper column descrip- tions,8 which describe how a column should be interpreted and processed (e.g.
+whether they should be processed as regular expressions; see below).
+Finally, in accordance with the Metadata Vocabulary for Tabular Data, the metadata’s  lename should consist of the orthography pro le’s  lename appended with “-metadata.json”.9
 
-[3] https://www.w3.org/TR/tabular-metadata/
-[4] http://dublincore.org/
-[5] JSON-DL metadata is also the choice for datasets conforming to the Cross-Linguistic Data Formats standard, see: http://cldf.clld.org/.
+3 https://www.w3.org/TR/tabular-metadata/
+4 http://dublincore.org/
+5 http://w3c.github.io/csvw/metadata/#dfn-common-property
+6 http://cldf.clld.org/v1.0/terms.rdf
+7 http://w3c.github.io/csvw/metadata/#dfn-dialect-descriptions
+8 http://w3c.github.io/csvw/metadata/#dfn-datatype-description
+9 JSON-LD metadata is also the choice for datasets conforming to the Cross-Linguistic Data
+Formats standard, see: http://cldf.clld.org/.
 
@@
 
-*** TODO
+*** W
 
@@
 
-*** Agreed (SM). We have made the formal specification lacker (which of course put the onus the software developers):
+*** Agreed (SM). The formal specification now requires a delimited text file, in which proper column descriptions are defined in the metadata JSON-LD file following:
 
-A1. A profile is a unicode UTF-8 encoded text file normalized following NFC (or NFD if speci ed in the metadata) that includes the information pertinent to the orthography.2
-
-A2.
-A profile is a CSV file with an obligatory header line.
-A minimal pro le must have a single column with the header “Grapheme”.
-For any ad- ditional columns, the name in the header of the CSV  le must be speci ed.
-e actual ordering of the columns is unimportant.
-Cells should not con- tain tabs or newlines.
-e header list must be delimited in the same way as the rest of the  le’s contents.
-Each record must be kept on a separate line.
-
-[2] See Section 3.12 in which we suggest to use NFC, no-BOM and LF line breaks because of the pitfalls they avoid.
-A keen reviewer notes, however, that specifying a convention for line endings and BOM is overly strict because most computing environments (now) transparently handle both alternatives.
-For example, using Python a  le can be decoded using the encoding 	“utf-8-sig”, which strips away the BOM (if present) and reads an input full in text mode, so that both line feed variants “LF” and “CRLF” will be stripped.
+http://w3c.github.io/csvw/metadata/#dfn-dialect-descriptions
+http://w3c.github.io/csvw/metadata/#dfn-datatype-description
 
@@
 
-*** TODO
+*** See directly above.
 
@@
 
-*** TODO
+*** Updated to:
+
+Note that in some cases the ordering inside this regular expression might be crucial, therefore it must be specified as so in the metadata file.
 
@@
 
-*** TODO
+*** We make this clearer:
+
+The regular expressions will be applied in the order as specified
+in the profile, from top to bottom.} A software implementation can offer
+help in figuring out the optimal ordering of the regular expressions, but
+then it should be made explicit in the orthograpy profile because regular 
+expressions are executed in order from top to bottom.
 
@@
 
-*** TODO
+*** We've updated the text to be more specification-like, as suggested above.
 

### 2018-03-09 book/reviews/responses_reviewer2.txt
all but the final reviewer comments on the specification
ready for review two

@@
 
-[ ] todo
+*** This is true, thus we write:
+
+Theoretically, these approaches will lead to different results, though in practice of actual natural language orthographies they almost always lead to the same result.
 
@@
 
-*** Schächtelchen is a great word and we appreciate the suggestion that "aäaaöaüaa" is confusing. So we have added Schächtelchen as an example to both the book and the online recipes.
+*** Schächtelchen is a great word and we appreciate the suggestion that "aäaaöaüaa" is confusing. So we have updated the example with "Schächtelchen" (also online).
 
-We therefore built a different example:
+We also built a different example:
 
@@
 
-which captures a hypothetical form from which the child can generalize.
+which captures a hypothetical form with all the intricacies of how IPA is encoded in Unicode.
 
@@
 
-***
+*** Fixed.
 
@@
 
-***
+*** No.
 
@@
 
+*** Recall we are tokenizing orthographic conventions, not their phonetic or phonemic realizations.
+
 - Footnote 3: delete. Feeding and bleeding are needlessly complicating
 the exposition
+
+*** We don't agree. Feeding and bleeding is an essential concept and this is the minimum we should do to make the reader aware of it.
 
@@
 
+*** An empty file suggests that some error was triggered.
+
 - "Such contextually-specified graphemes are based on regular
@@
 
+*** Thanks!
+
 - "It is important to realize ... ": delete. The need for escape
 sequences has been discussed earlier.
+
+*** Fixed.
 
@@
 
+*** This would be a orthography-specific case, for which we should create an example online.
+
 *** Note, we have also added a section that highlights the four additional use cases that we present as Jupyter Notebooks in our Github repository of orthography profile recipes:

### 2018-03-09 book/chapters/orthography_profiles.tex
ready for review two

@@
 central concept in our proposal is the \textsc{orthography profile}, a simple
-tab-separated text file, that characterizes and documents a writing system.
+delimited text file, that characterizes and documents a writing system.
 We also offer basic implementations in Python and R to assist with the
@@
 <\dia{0303}> + <\dia{0330}> or <o> + <\dia{0330}> + <\dia{0303}>. Although these two variants are visually homoglyphs,
-computationally they are different (see Sections \ref{pitfall-ipa-homoglyphs} and \ref{pitfall-homoglyphs-in-IPA} ). 
+computationally they are different (see Sections \ref{pitfall-ipa-homoglyphs} and \ref{pitfall-homoglyphs-in-IPA}). 
 Unicode normalization should be applied to
@@
 tokenization from transliteration there is no problem with `feeding' and 
-`bleeding' of rules, common with transducers (cf.~Section~\ref{r-implementation}
-).
+`bleeding' of rules, common with transducers (cf.~Section~\ref{r-implementation}).
 
@@
 need for multiple different orthography profiles to be applied in sequence (see
-Section~\ref{r-implementation} on various exemplary use-cases). The result of any such
+Sections~\ref{python-implementations} \& \ref{r-implementation} on various exemplary use-cases). The result of any such
 workflow will normally be a cleaned dataset and an explicit description of the
@@
 
-%	\item \textsc{A profile is a unicode UTF-8 encoded text file}, ideally
-%       using NFC, no-BOM, and LF (see Section~\ref{pitfall-file-formats}), that
-%       includes the information pertinent to the orthography. 
-	\item \textsc{A profile is a CSV file with an obligatory header
+	\item \textsc{A profile is a delimited text file with an obligatory header
        line}. A minimal profile must have a single column with the header ``Grapheme''. 
-	   For any additional columns, the name in the header of the CSV file
-       must be specified. The actual ordering of the columns is unimportant. Cells should 
-	   not contain tabs or newlines. The header list must be delimited 
+	   For any additional columns, the name in the header 
+       must be specified. The actual ordering of the columns is unimportant. 
+	   The header list must be delimited 
 	   in the same way as the rest of the file's contents. Each record must be kept 
@@
        belong to specific lines must be put in a separate column of
-       the CSV file, e.g.~add a column called \textsc{comments}.
+       the file, e.g.~add a column called \textsc{comments}.
 
@@
 	\item \textsc{Metadata should be added in a separate UTF-8 text file} using 
-	   the JSON-DL dialect specified in \textit{Metadata Vocabulary for Tabular 
+	   the JSON-LD dialect specified in \textit{Metadata Vocabulary for Tabular 
 	   Data}.\footnote{\url{https://www.w3.org/TR/tabular-metadata/}} This metadata 
@@
 	   which should be used to specify information about the orthographic description 
-	   in the orthography profile. The orthography profile metadata should minimally include: 
-	   (i) author, (ii) date, (iii) title of the profile, (iv) a stable language identifier
-       encoded in BCP 47/ISO 639-3 of the target language of the profile, and (v)
+	   in the orthography profile.\footnote{\url{http://w3c.github.io/csvw/metadata/#dfn-common-property}} 
+	   The orthography profile metadata should minimally include provenance information including: 
+	   (i) author, (ii) date, (iii) title of the profile, and (iv)
        bibliographic data for resource(s) that illustrate the orthography
-       described in the profile. Further, (vi) the tokenization method and (vii)
-       the Unicode normalization form used should be documented here (see below).
-	   In accordance with the \textit{Metadata Vocabulary for Tabular 
-	   Data}, the metadata's filename must consist of the orthography 
-	   profile's filename appended with ``-metadata.json''.\footnote{JSON-DL metadata 
+       described in the profile. Crucially, the metadata should also specify 
+	   (iv) a stable language identifier of the target language of the profile
+       using BCP 47/ISO 639-3 or Glottocode as per the CLDF ontology.\footnote{\url{http://cldf.clld.org/v1.0/terms.rdf}}	   	   
+	   Further, the metadata file should provide information about the orthography profile's structure and contents, 
+	   including: (vi) its dialect description,\footnote{\url{http://w3c.github.io/csvw/metadata/#dfn-dialect-descriptions}} 
+	   (vii) the tokenization method (see below), and (ix) proper column           
+	   descriptions,\footnote{\url{http://w3c.github.io/csvw/metadata/#dfn-datatype-description}} 
+	   which describe how a column should be interpreted and processed (e.g.\ whether they 
+	   should be processed as regular expressions; see below).
+	   Finally, in accordance with the \textit{Metadata Vocabulary for Tabular 
+	   Data}, the metadata's filename should consist of the orthography 
+	   profile's filename appended with ``-metadata.json''.\footnote{JSON-LD metadata 
 	   is also the choice for datasets conforming to the Cross-Linguistic Data Formats standard, 
@@
        can use regular expression metacharacters.} If regular expressions are
-       used, then all literal usage of the special symbols, like full stops <.>
+       used, then the must be specified in the metadata file as such, 
+	   and all literal usage of the special symbols, like full stops <.>
        or dollar signs <\$> (so-called \textsc{metacharacters}) have to be
@@
        specified for the elements of the class. Note that in some cases the 
-       ordering inside this regular expression might be crucial.
+       ordering inside this regular expression might be crucial, therefore it must be 
+	   specified as so in the metadata file.
 	\item \textsc{The \textsc{left} and \textsc{right} contexts will be included
@@
        help in figuring out the optimal ordering of the regular expressions, but
-       should then explicitly report on the order used.      
+       then it should be made explicit in the orthograpy profile because regular 
+	   expressions are executed in order from top to bottom.
 \end{enumerate}
@@
        or transliterated}. In this way, there is no necessity for the user to
-       consider `feeding' and `bleeding' situations, in which the application of
+       consider \textit{feeding} and \textit{bleeding} situations, in which the application of
        a rule either changes the text so another rule suddenly applies (feeding)
-       or prevents another rule to apply (`bleeding'). 
+       or prevents another rule to apply (bleeding). 
 	\item \textsc{The matching of the graphemes can occur either globally or
-       linearly. }From a computer science perspective, the most natural way to
+       linearly.} From a computer science perspective, the most natural way to
        match graphemes from a profile in some text is by walking linearly

### 2018-03-11 book/chapters/pitfalls.tex
resolves issue #28

@@
 supports Tamil, e.g. \textsc{tamil letter au} at \uni{0B94} represents a single 
-character <ஔ>, composed of two glyphs <ஓ> and <ன>. Perhaps confusingly, 
+character <\tamil{ஔ}>, composed of two glyphs <\tamil{ஓ}> and <\tamil{ன}>. Perhaps confusingly, 
 in the Unicode Standard there are also two individual characters, 
@@
 \uni{0BA9} \textsc{tamil letter nnna}, each of which is a glyph. Another example is 
-Sinhala \textsc{sinhala vowel sign kombu deka} at \uni{0DDB} <ෛ>, which is 
+Sinhala \textsc{sinhala vowel sign kombu deka} at \uni{0DDB} <\sinhala{ෛ}>, which is 
 visually two glyphs, each represented by \textsc{sinhala vowel sign kombuva} 
-at \uni{0DD9} < ෙ>.
+at \uni{0DD9} <\sinhala{ ෙ}>.
 

### 2018-03-12 book/chapters/introduction.tex
fixes section reference

@@
 terminology concerning writing systems before turning to the slightly different
-computational terminology in the subsequent section on the Unicode Standard.
+computational terminology in the subsequent chapter on the Unicode Standard.
 

### 2018-03-12 book/chapters/introduction.tex
Resolves issue #29

@@
 languages \citep{Meyer2015}, but also drumming languages \citep{Meyer2012} and
-signalling by smoke, fire, flags or even change in water levels through
+signaling by smoke, fire, flags or even change in water levels through
 hydraulic pressure have been used as forms of telegraphy.

### 2018-03-12 book/chapters/ipa_background.tex
Resolves issue #29

@@
 the International Phonetic Association in representing the IPA to the
-\textsc{international standards organisation} (ISO) and to the \textsc{text
+\textsc{international organization for standardization} (ISO) and to the \textsc{text
 encoding initiative} (TEI) \citep{EslingGaylord1993}. The working group's
@@
 own submissions as part of a review process by ISO for approval based on 
-both informatic and phonetic criteria \citep[86]{EslingGaylord1993}.
+both informatics and phonetic criteria \citep[86]{EslingGaylord1993}.
 

### 2018-03-12 book/chapters/ipa_meets_unicode.tex
Resolves issue #29

@@
 	\item The Americanist tradition of using using <\dia{030C}>, a
-       \textsc{combining caron} at \uni{030C} for all post-alveolar sounds, like
+       \textsc{combining caron} at \uni{030C} for all postalveolar sounds, like
        in <š~ž~č~ǰ>, was not adopted because the Association members at the
@@
 \noindent At the end of this chapter we have added a few longish tables summarizing all
-159 different unicode codepoints that form the basis of strict-IPA encoding
+159 different unicode code points that form the basis of strict-IPA encoding
 (107 letters, 36 diacritics and 16 remaining symbols). We also make these tables available 
 online in CSV format.\footnote{\url{https://github.com/unicode-cookbook/cookbook/tree/master/book/tables}} 
-Each of these tables shows a typical glyph, and then lists the Unicode Codepoint,
+Each of these tables shows a typical glyph, and then lists the Unicode Code point,
 Unicode Name and IPA description for each symbol. Further, there is a table with 

### 2018-03-12 book/chapters/orthography_profiles.tex
Resolves issue #29

@@
        highly useful to add extra information on the graphemes in further
-       columns, like for example Unicode codepoints, Unicode names, frequency of
+       columns, like for example Unicode code points, Unicode names, frequency of
        occurrence, examples of occurrence, explanation of the contextual
@@
        help in figuring out the optimal ordering of the regular expressions, but
-       then it should be made explicit in the orthograpy profile because regular 
+       then it should be made explicit in the orthography profile because regular 
 	   expressions are executed in order from top to bottom.

### 2018-03-12 book/chapters/pitfalls.tex
fixes section reference
Resolves issue #29

@@
 rendering (see also Section~\ref{pitfall-faulty-rendering}). Like
-contextually-determined glyphs, ligatures are (intended) artefacts of text
+contextually-determined glyphs, ligatures are (intended) artifacts of text
 processing instructions. Finally, a single glyph may be a part of a
@@
 \textsc{combining grapheme joiner} at \uni{034F}. However, these characters are
-confusingly named (cf.~Section~\ref{pitfall-names}). Both codepoints actually do
+confusingly named (cf.~Section~\ref{pitfall-names}). Both code points actually do
 not join characters, but explicitly separate them. The zero-width joiner (ZWJ)
@@
 attempt to show a real glyph, but only shows the hexadecimal code inside a box
-for each character, so a user can at least see the Unicode codepoint of the
+for each character, so a user can at least see the Unicode code point of the
 character to be displayed.\footnote{The SIL Unicode BMP Fallback Font is
@@
 Consortium, but on a complex interplay of different mechanisms happening in a
-computer to turn Unicode codepoints into visual symbols. We will only sketch a
+computer to turn Unicode code points into visual symbols. We will only sketch a
 few aspects of this complex interplay here.
@@
 
-The actual intended meaning of a Unicode codepoint is a combination of the
+The actual intended meaning of a Unicode code point is a combination of the
 name, the block and the character properties (see
@@
 that is ever proposed, discussed and decided in relation to a specific
-codepoint of interest, so many of the details are often somewhat
+code point of interest, so many of the details are often somewhat
 hidden.\footnote{All proposals and other documents that are the basis of Unicode
@@
 \section{Pitfall: Encodings}
-\label{encoding}
+\label{encodings}
 % ==========================
@@
          rendering (see Section~\ref{pitfall-faulty-rendering}).
-   \item Do not trust the names of codepoints as a definition of the character
+   \item Do not trust the names of code points as a definition of the character
          (see Section~\ref{pitfall-names}). Also do not trust Unicode blocks as

### 2018-06-04 book/chapters/orthography_profiles.tex
deals with most of Cozens' comments

@@
 fruitfully be published alongside sources (e.g.~in digital archives). Also, in
-high-level linguistic analyzes in which the graphemic detail is of central
+high-level linguistic analyses in which the graphemic detail is of central
 importance (e.g.~phonotactic or comparative-historical studies), orthography
@@
 such locale descriptions in the daily linguistic practice, namely: (i) the
-XML-structure is too verbose to easily and quickly produce or correct manually,
-(ii) locale descriptions are designed for a wide scope on information (like date
+XML structure is too verbose to easily and quickly produce or correct manually,
+(ii) locale descriptions are designed for a wide scope of information (like date
 formats or names of weekdays) most of which is not applicable for documenting
@@
 produce a technically correct locale description for a specific source at hand,
-then it is nigh impossible to deploy the description. This is because a locale
+then it is well-nigh impossible to deploy the description. This is because a locale
 description has to be submitted to and accepted by the Unicode Common Locale
@@
 
-The major challenge then is developing an infrastructure to identify the
+The major challenge, then, is developing an infrastructure to identify the
 elements that are individual graphemes in a source, specifically for the
@@
 and strings must be tokenized, which may include taking orthographic rules into
-account (e.g.~between vowels is /n/ and after a vowel but before a consonant is
+account (e.g.~a nasal sound may be transcribed as /n/ when it appears between 
+two vowels, but when it appears between a vowel and a consonant it becomes 
 a nasalized vowel /Ṽ/).
@@
 source-by-source approach is required before an orthography profile can be
-created. For example, almost each available lexicon on the world's languages is
+created. For example, almost every available lexicon on the world's languages is
 idiosyncratic in its orthography and thus requires lexicon-specific approaches
@@
 In (1), the string <tsʰṍ̰shi> has been tokenized into ten Unicode code points
-(using NFD normalization), delimited here by space. Unicode tokenization is
+(using NFD normalization), delimited here by space. Unicode normalization is
 required because sequences of code points can differ in their visual and logical
@@
 this string to reorder the code points into a canonical order, allowing the data
-to be treated canonically equivalently for search and comparison.
+to be treated canonically equivalent for search and comparison.
 
@@
 grapheme clusters, as specified by the Unicode Standard. The combining character
-sequence <õ̰> is normalized and visually grouped together. Note that, the
-\textsc{modifier letter small h} at \uni{02B0}, is not grouped with any 
+sequence <õ̰> is normalized and visually grouped together. Note that the
+\textsc{modifier letter small h} at \uni{02B0} is not grouped with any 
 other character. This is because it
-belongs to Spacing Modifier Letters category in the Unicode Standard. These
-characters are underspecified for the direction in which they modify a host
+belongs to the Spacing Modifier Letters category. The Unicode Standard 
+does not specify the direction that these characters modify a host
 character. For example, it can indicate either pre- or post-aspiration (whereas the
@@
 Finally, to arrive at the graphemic tokenization in (3), tailored grapheme
-clusters are needed (possibly specified in an orthography profile). For example,
+clusters are needed, possibly as specified in an orthography profile. For example,
 an orthography profile might specify that the sequence of characters <tsʰ> form
 a single grapheme. The orthography profile could also specify orthographic
-rules, e.g.~when tokenizing graphemes, in say English, the sequences <sh> in the
+rules, e.g.~when tokenizing graphemes in English, the sequences <sh> in the
 forms <mishap> and <mishmash> should be treated as distinct sequences depending
@@
 and rules that are needed to adequately model a writing system. An important
-assumption is that we assume a resource is encoded in Unicode (or
-has been converted to Unicode). Any data source that the Unicode Standard is
-unable to capture, will also not be captured by an orthography profile.
+assumption is that we assume a resource is encoded in Unicode or
+has been converted to Unicode. Any data source that the Unicode Standard is
+unable to capture will also not be captured by an orthography profile.
 
-Informally, an orthography profile specifies the graphemes (or, in Unicode
-parlance, \textsc{tailored grapheme clusters}) that are expected to occur in any
+Informally, an orthography profile specifies the graphemes -- in Unicode
+parlance \textsc{tailored grapheme clusters} -- that are expected to occur in any
 data to be analyzed or checked for consistency. These graphemes are first
-identified throughout the whole data (a step which we call
-\textsc{tokenization}), and simply returned as such, possibly including
+identified throughout the whole data, a step which we call
+\textsc{tokenization}, and simply returned as such, possibly including
 error messages about any parts of the data that are not specified by the
 orthography profile. Once the graphemes are identified, they might also be
-changed into other graphemes (a step which we call \textsc{transliteration}).
+changed into other graphemes -- a step which we call \textsc{transliteration}.
 When a grapheme has different possible transliterations, then these differences
@@
 Note that to deal with ambiguous parsing cases, it is still possible to use the
-Unicode approach including the \textsc{zero-width non-joiner} character at
+Unicode approach of including the \textsc{zero-width non-joiner} character at
 \uni{200C} into the text. The idea is to add this character into the text to
 identify cases in which a sequence of characters is \textit{not} supposed to be
-a complex grapheme cluster (even though the sequence is in the orthography
-profile).
+a complex grapheme cluster -- even though the sequence is in the orthography
+profile.
 
@@
 	\def\labelenumi{A\arabic{enumi}.} 
-	\item \textsc{A profile is a unicode UTF-8 encoded text file} 
+	\item \textsc{A profile is a unicode utf-\scalebox{0.8}{8} encoded text file} 
 	   normalized following NFC (or NFD if specified in the metadata)
@@
 	   Section~\ref{pitfall-file-formats} in which we suggest to use NFC, 
-	   no-BOM and LF line breaks because of the pitfalls they avoid. 
-	   A keen reviewer notes, however, that specifying 
+	   no-BOM and LF line breaks because of the pitfalls they avoid. A keen reviewer notes, however, that specifying 
 	   a convention for line endings and BOM is overly strict because most 
@@
 	\item \textsc{A profile is a delimited text file with an obligatory header
-       line}. A minimal profile must have a single column with the header ``Grapheme''. 
+       line}. A minimal profile must have a single column with the header \texttt{Grapheme}. 
 	   For any additional columns, the name in the header 
@@
 \begin{comment}	   
-	\item \textsc{Metadata should be added in a separate UTF-8 text file} using a basic
+	\item \textsc{Metadata should be added in a separate utf-\scalebox{0.8}{8} text file} using a basic
        \textsc{tag: value} format. Metadata about the orthographic description
@@
        described in the profile. Crucially, the metadata should also specify 
-	   (iv) a stable language identifier of the target language of the profile
+	   (v) a stable language identifier of the target language of the profile
        using BCP 47/ISO 639-3 or Glottocode as per the CLDF ontology.\footnote{\url{http://cldf.clld.org/v1.0/terms.rdf}}	   	   
@@
        context of the grapheme}, to be designated with the headers \texttt{Left}
-       and \texttt{Right}, respectively. The same grapheme can occur multiple
+       and \texttt{Right} respectively. The same grapheme can occur multiple
        times with different contextual specifications, for example to
@@
        can use regular expression metacharacters.} If regular expressions are
-       used, then the must be specified in the metadata file as such, 
+       used, then they must be specified in the metadata file as such, 
 	   and all literal usage of the special symbols, like full stops <.>
@@
        <\textbackslash.> or <\textbackslash\$>). Note that any specification of
-       context automatically expects regular expressions, so it is probably
+       context automatically expects regular expressions, so it is 
        better to always escape all regular expression metacharacters when used
@@
        highly useful to add extra information on the graphemes in further
-       columns, like for example Unicode code points, Unicode names, frequency of
-       occurrence, examples of occurrence, explanation of the contextual
+       columns, such as Unicode code points, Unicode names, frequency of
+       occurrence, examples of occurrence, explanation of contextual
        restrictions, or comments. 
@@
        a rule either changes the text so another rule suddenly applies (feeding)
-       or prevents another rule to apply (bleeding). 
+       or prevents another rule from applying (bleeding). 
 	\item \textsc{The matching of the graphemes can occur either globally or
@@
        match graphemes from a profile in some text is by walking linearly
-       through the text-string from left to right, and at each position go
+       through the text-string from left to right, and at each position going
        through all graphemes in the profile to see which one matches, then go to
@@
        happens when very many different combinations of diacritics occur in the
-       data. An NFD-profile can then be used to first check which individual
+       data. An NFD profile can then be used to first check which individual
        diacritics are used, before turning to the more cumbersome inspection of
@@
        matched by the profile, should be reported to the user as errors.}
-       Typically, the unmatched character are replaced in the tokenization by a
+       Typically, the unmatched characters are replaced in the tokenization by a
        user-specified symbol-string.       
@@
         expression metacharacters. Matching graphemes literally often leads to
-        strong speed increase, and would allow users to not needing to worry
+        significant speed increase, and would allow users to not have to worry
         about escaping metacharacters. However, in our experience all actually
@@
 		\item \textsc{the original text-strings} to be processed in the specified
-        unicode normalization, i.e.~in either NFC or NFD as specified by the
+        Unicode normalization, i.e.~in either NFC or NFD as specified by the
         user. 
@@
         text-strings lead to errors. 
-		\item \textsc{a reordered profile}, when any automatic reordering is offered 
+		\item \textsc{a reordered profile}, when any automatic reordering is offered. 
 	\end{enumerate}

### 2018-06-04 book/chapters/introduction.tex
fixes en dashes throughout

@@
 technological and practical pressures. They tend to be highly conservative, in
-that people who have learned to read and write in a specific way---however
-impractical or tedious---are mostly unwilling to change their habits. Writers
+that people who have learned to read and write in a specific way -- however
+impractical or tedious -- are mostly unwilling to change their habits. Writers
 tend to resist spelling reforms. In all literate societies there exists a strong
@@
 includes an extra long pause between words. Interestingly, it took a long time
-to consider the written word boundary---using white-space---as a bona-fide
+to consider the written word boundary -- using white-space -- as a bona-fide
 character that should simply be encoded with its own code point. This happened

### 2018-06-04 book/chapters/pitfalls.tex
fixes en dashes throughout

@@
 executes without error but does something different than the user
-expected---often unknowingly---a \textsc{pitfall}.
+expected -- often unknowingly -- a \textsc{pitfall}.
 
@@
 <{\fontspec{EversonMono}AАΑᎪᗅᴀꓮ𐊠𝖠𝙰}> are all different code points,
-although they look highly similar---in some cases even sharing exactly the same
+although they look highly similar -- in some cases even sharing exactly the same
 glyph in some fonts. All these different A-like characters include the following

### 2018-06-05 book/chapters/introduction.tex
fix en dashes

@@
 Baudot code uses a so-called shift character, which signifies that from
-that point onwards---until shifted back---a different encoding is used, allowing
+that point onwards -- until shifted back -- a different encoding is used, allowing
 for yet another set of 32 codes. In effect, this means that the Baudot code, and

### 2018-06-05 book/chapters/orthography_profiles.tex
deals with paperhivers comments for chapter 7 Orthography profiles

@@
 and strings must be tokenized, which may include taking orthographic rules into
-account (e.g.~a nasal sound may be transcribed as /n/ when it appears between 
+account (e.g.~a nasal sound may be transcribed as <n> when it appears between 
 two vowels, but when it appears between a vowel and a consonant it becomes 
-a nasalized vowel /Ṽ/).
+a nasalized vowel <Ṽ>).
 
@@
        then be used in the description of the graphemes or the context. The
-       identifiers should of course be chosen such that they do not conflate
-       with any symbols used in the orthography themselves. Note that such
+       identifiers should of course be chosen so that they do not conflate
+       with any symbols used in the orthography. Note that such
        classes only refer to the graphemes, not to the context. 
@@
 	\item \textsc{Each line of a profile will be interpreted as a regular
-       expression. }Software applications using profiles can also offer to
+       expression.} Software applications using profiles can also offer to
        interpret a profile in the literal sense to avoid the necessity for the
-       user to escape regular expressions metacharacters in the profile.
-       However, this only is possible when no contexts or classes are described,
+       user to escape regular expression metacharacters in the profile.
+       However, this is only possible when no contexts or classes are described,
        so this seems only useful in the most basic orthographies. 
@@
        ordering inside this regular expression might be crucial, therefore it must be 
-	   specified as so in the metadata file.
+	   specified as such in the metadata file.
 	\item \textsc{The \textsc{left} and \textsc{right} contexts will be included
@@
 		\item \textsc{the tokenization method}, i.e.~whether the tokenization should
-        proceed \textsc{global} or \textsc{linear} (see B6 above). 
+        proceed as \textsc{global} or \textsc{linear} (see B6 above). 
 		\item \textsc{unicode normalization}, i.e.~whether the text-string and profile
@@
 		\def\labelenumii{C\arabic{enumii}.} \setcounter{enumii}{5} 
-		\item \textsc{to assist in the ordering of the graphemes.} In our experience, it
-        makes sense to apply larger graphemes before shorter graphemes, and to
-        apply graphemes with context before graphemes without context. Further,
+		\item \textsc{to assist in the ordering of the graphemes.} In our experience 
+		working with idiosyncratic transcriptions and orthographies from low-resource languages, it
+        is helpful to identify multi-sequence graphemes before single graphemes, and to
+        identify graphemes with context before graphemes without context. Further,
         frequently relevant rules might be applied after rarely relevant rules
@@
         <a> and <A> are treated equally. This will be useful in many concrete
-        cases, although the user should be warned that case matching does not
-        function universally in the same way across orthographies. Ideally,
+        cases (such as search or collation), although the user should be warned that case matching does not
+        function universally in the same way across orthographies.\footnote{For example 
+		compare the different first-letter capitalization practices of the digraphs 
+		<ǋ> and <Ĳ> (single-character ligatures in the Unicode Standard) 
+		in the Latin-based scripts of Southern-Slavic languages and 
+		Dutch, respectively.} Ideally,
         users should prepare orthography profiles with all lowercase and
@@
         expression metacharacters. Matching graphemes literally often leads to
-        significant speed increase, and would allow users to not have to worry
+        significant speed increase, and ensures that users do not have to worry
         about escaping metacharacters. However, in our experience all actually
@@
         strings, if transliteration is requested. 
-		\item \textsc{a survey of all errors encountered}, ideally both in which
-        text-strings any errors occurred and which characters in the
+		\item \textsc{a survey of all errors encountered}, ideally both (i) in which
+        text-strings any errors occurred and (ii) which characters in the
         text-strings lead to errors. 

### 2018-06-05 book/chapters/ipa_meets_unicode.tex
work towards paperhivers chapter 8

@@
 \noindent At the end of this chapter we have added a few longish tables summarizing all
-159 different unicode code points that form the basis of strict-IPA encoding
+159 different Unicode code points that form the basis of strict-IPA encoding
 (107 letters, 36 diacritics and 16 remaining symbols). We also make these tables available 

### 2018-06-05 book/chapters/introduction.tex
fixes abbreviations, address paperhivers through chapter 1

@@
 socio-political mainstream that tries to force unification of writing (for
-example by strongly enforcing ``right'' from ``wrong'' writing in schools).
-However, there is also a large community of users who take as many liberties in
+example by strongly enforcing ``right'' from ``wrong'' spelling in schools).
+However, there are also communities of users who take as many liberties in
 their writing as they can get away with.
@@
 expressing yourself in writing, others will more easily accept you as a worthy
-participant in their group. And that just as well holds for obeying to the
-official rules when writing a job application, as for obeying to the informal
-rules when writing an SMS to classmates in school. The case of Yoruba writing is
+participant in their group -- whether it means following the official rules 
+when writing a job application or conforming to the informal rules when writing text 
+messages. The case of Yoruba writing is
 an exemplary case, as even after more than a century of efforts to standardize
@@
 The resulting cumbersome and often illogical structure of writing systems, and
-the enormous variability of existing writing systems for the world's languages
-is a fact of life scholars have to accept and should try to adapt to as well as
-we can. Our plea in this book is a proposal how to do exactly that: formalize 
-knowledge about individual writing systems in a form that is both easy to use 
+the enormous variability of existing writing systems for the world's languages, 
+is a fact of life that scholars have to accept and they should try to adapt to as well as
+they can. Our goal in this book is a proposal for how to do exactly that: formalize 
+knowledge about individual writing systems in a form that is easy to use 
 for linguists in daily practice, and at the same time computer-readable for 
@@
 
-When considering the worldwide linguistic diversity, including the many
+When considering worldwide linguistic diversity, including the many
 lesser-studied and endangered languages, there exist numerous different
@@
 and compare different orthographies, we need a way to specify all orthographic
-idiosyncrasies in a computer-readable format (a process called
-\textsc{tailoring} in Unicode parlance). We call such specifications
+idiosyncrasies in a computer-readable format. We call such specifications
 \textsc{orthography profiles}. Ideally, these specifications have to be
-integrated into so-called Unicode locale descriptions, though we will argue that
-in practice this is often not the most useful solution for the kind of problems
-arising in the daily practice of linguistics. Consequently, a central goal of
+integrated into so-called Unicode Locales,\footnote{\url{http://cldr.unicode.org/locale_faq-html}} 
+though we will argue that in practice this is often not the most useful solution for the kind of problems
+arising in the daily practice of many linguists. Consequently, a central goal of
 this book is to flesh out the linguistic challenges for locale descriptions,
@@
 
-The necessity to computationally use and compare different orthographies most
-forcefully arises in the context of language comparison. Concretely, the proper
-processing of orthographies and transcriptions becomes crucial for the
+The need to use computational methods to compare different orthographies arises most
+forcefully in the context of language comparison. Concretely, the proper
+processing of orthographies and transcription systems becomes crucial for the
 development of quantitative methods for language comparison and historical
-reconstruction in order to investigate worldwide linguistic variation and to
+reconstruction, in order to investigate worldwide linguistic variation and to
 model the historical and areal processes that underlie linguistic
-diversity. In this work, it is crucial to be able to flexibly process across numerous
-resources with different orthographies. In many cases even different resources
+diversity. In this work, it is crucial to be able to flexibly process numerous
+resources in different orthographies. In many cases even different resources
 on the same language use different orthographic conventions. Another
@@
 orthographic convention (often a resource-specific convention), but on closer
-inspection such resources are almost always not consistently encoded. Thus a
+inspection such resources are almost always not consistently encoded. Thus, a
 second goal of our orthography profiles is to allow for an easy specification of
@@
 sentences) in a specific source, our first processing step is to specify how
-these strings have to be separated into graphemes, considering the specific
+these strings should be separated into graphemes, considering the specific
 orthographic conventions used in a particular source document. Our experience is
@@
 words. Grapheme separation is a task that can be performed relatively reliably
-and with limited availability of time and resources (compare, for example, the
+and with limited time and resources (compare, for example, the
 daunting task of creating a complete phonetic or phonological normalization).
@@
 important and highly fruitful processing step. Given a grapheme tokenization,
-various subsequent tasks become easier, like (a) temporarily reducing the
+various subsequent tasks become easier, for instance (a) temporarily reducing the
 orthography in a processing pipeline, e.g.~only distinguishing high versus low
@@
 slightly more in-depth survey of the many techniques that are available in the
-standard. A good appreciation for the solutions that the Unicode Standard also
-allows for a thorough understanding of possible pitfalls that one might
-encounter when using it in general (Chapter~\ref{unicode-pitfalls}).
+standard. A good appreciation for the solutions that the Unicode Consortium has created 
+allows for a thorough understanding of the possible pitfalls that one might
+encounter when using the Unicode Standard in general (Chapter~\ref{unicode-pitfalls}).
 Linguists are more often interested in using the Unicode Standard with the International Phonetic Alphabet (IPA).
@@
 
-The following conventions are followed in this book. All phonemic and phonetic
+The following conventions are adhered to in this book. All phonemic and phonetic
 representations are given in the International Phonetic Alphabet (IPA), unless
@@
 representations. For character descriptions, we follow the notational
-conventions of the Unicode Standard \citep{Unicode2017}. Character names are
+conventions of the Unicode Standard \citep{Unicode2018}. Character names are
 represented in small capital letters (e.g.~\textsc{latin small letter schwa})
@@
 languages \citep{Meyer2015}, but also drumming languages \citep{Meyer2012} and
-signaling by smoke, fire, flags or even change in water levels through
+signaling by smoke, fire, flags, or even changes in water levels through
 hydraulic pressure have been used as forms of telegraphy.
 
-Telegraphy was reinvigorated in the end of the eighteenth century through the
+Telegraphy was reinvigorated at the end of the eighteenth century through the
 introduction of so-called semaphoric systems by Claude Chapelle to convey
@@
 written language were replaced one-to-one by visual signals. Since then, all
-telegraphic systems have taken this principle, namely that any language to be
+telegraphic systems have adopted this principle,\footnote{Sound- and
+video-based telecommunication take a different approach by ignoring
+the written version of language and they directly encode sound waves or light
+patterns.} namely that any language to be
 transmitted first has to be turned into some orthographic system, which
 subsequently is encoded for transmission by the sender, and then turned back
-into orthographic representation at the receiver side.\footnote{Sound and
-video-based telecommunication of course takes a different approach by ignoring
-the written version of language and directly encode sound waves or light
-patterns.} This of course implies that the usefulness of any such telegraphic
+into orthographic representation at the receiver side. This of course implies 
+that the usefulness of any such telegraphic
 encoding completely depends on the sometimes rather haphazard structure of
@@
 
-In the nineteenth century, electric telegraphy lead to a next innovation in
+In the nineteenth century, electric telegraphy led to a further innovation in
 which written language characters were encoded by signals sent through a copper
@@
 of five signals for each written character.\footnote{True binary codes have a
-longer history, going at least back to the Baconian cipher devised by Francis
+longer history, going back at least to the Baconian cipher devised by Francis
 Bacon in 1605. However, the proposal by Baudot was the quintessential proposal
-leading to all modern systems.} There are $2^5 = 32$ possible combination when
+leading to all modern systems.} There are $2^5 = 32$ possible combinations when
 using five binary signals; an encoding today designated as 5-bit. These
@@
 towards the encoding of English orthography. With the ascent of other
-orthographies in computer usage, the wish to encode further variation of Latin
-letters (like German <ß> or various letters with diacritics, like <è>) led the
+orthographies in computer usage, the wish to encode further variations of Latin
+letters (including German <ß> and various letters with diacritics, e.g.\ <è>) led the
 Digital Equipment Corporation to introduce an 8-bit \textsc{Multinational
@@
 
-Because 256 characters were clearly not enough for the many different characters
+Because 256 characters were clearly not enough for the unique representation of 
+many different characters
 needed in the world's writing systems, the ISO/IEC~8859 standard in 1987
@@
 %number of symbols was much smaller.} 
-This system almost immediately was
+This system was almost immediately 
 understood to be insufficient and impractical, so various initiatives to extend
@@
 
-More wide-ranging, various people in the 1980s started to develop true
+In the 1980s various people started to develop true
 international code sets. In the United States, a group of computer scientists
@@
 Europe, the \textsc{international organization for standardization} (ISO) was
-working on ISO~10646 to supplant the ISO/IEC~8859 standard. Their first draft of
+working on ISO~10646 to replace the ISO/IEC~8859 standard. Their first draft of
 the \textsc{universal character set} (UCS) in 1990 was 31-bit (with
@@
 1,114,112$ possible characters). The most recent version of the Unicode Standard
-(currently at version number 10.0.0) was published in June 2017 and it defines
-136,690 different characters \citep{Unicode2017}.
+(currently at version number 11.0.0) was published in June 2018 and it defines
+137,374 different characters \citep{Unicode2018}.
 
@@
 
-Focussing on the first sense, we distinguish two different kinds of 
-writing systems used for a particular language, namely transcriptions and
-orthographies.
-
-First, \textsc{transcription} is a scientific procedure (and also
+Focusing on the first sense of \textsc{writing system} described above, we distinguish 
+between two different kinds of writing systems used for a particular language, namely transcriptions and
+orthographies. First, \textsc{transcription} is a scientific procedure (and also
 the result of that procedure) for graphically representing the sounds of human
@@
 materials development by correlating a writing system with the sound units of a
-language \citep[cf.~][]{MeinhofJones1928}. Symbols from the IPA are often used
+language \citep{MeinhofJones1928}. Symbols from the IPA are often used
 by linguists in the development of such practical orthographies for languages
@@
 graphemes. A \textsc{grapheme} is the basic, minimally distinctive symbol of a
-particular writing system, alike to the phoneme is an abstract representation of
-a distinct sound in a specific language. The term \textsc{grapheme} was modeled
-after the term \textsc{phoneme} and represents a contrastive graphical unit in a
+particular writing system. It was modeled
+after the term \textsc{phoneme} (an abstract representation of
+a distinct sound in a specific language) and as such it represents a contrastive graphical unit in a
 writing system (see \citealp{Kohrt1986} for a historical overview of the term
-grapheme). Most importantly, a single grapheme regularly consists of multiple
+\textsc{grapheme}). Most importantly, a single grapheme regularly consists of multiple
 characters, like <th>, <ou> and <gh> in English (note that each character in
@@
 systems has been purposely brief. We have highlighted some of the linguistic
-notions that are pertinent, yet sometimes confused with, the technological
+notions that are pertinent to, yet sometimes confused with, the technological
 definitions developed for the computational processing of the world's writing

### 2018-06-05 book/chapters/ipa_background.tex
fixes abbreviations, address paperhivers through chapter 1

@@
 unambiguously represent phonetic details. However, since phonetic detail could
-potentially include anything, e.g. something like ``deep voice'', the IPA
+potentially include anything, e.g.\ something like ``deep voice'', the IPA
 restricts phonetic detail to linguistically relevant aspects of speech.
@@
 	\item obviating the design of a (large) number of new symbols when a 
-		  single diacritic suffices (e.g. nasalized vowels, aspirated stops).
+		  single diacritic suffices (e.g.\ nasalized vowels, aspirated stops).
 
@@
 A narrow phonetic transcription may freely use the all symbols in the IPA 
-chart with direct reference to their phonetic value, i.e. the transcriber can 
+chart with direct reference to their phonetic value, i.e.\ the transcriber can 
 indicate with the symbols <ŋ͡m> that the phonetic value of the attested sound 
@@
 301, [e] 302, [ɛ] 303. The system allows for the addition of new symbols within
-the various series by appending them, e.g. [\charis{ⱱ}] 184. Former non-IPA
+the various series by appending them, e.g.\ [\charis{ⱱ}] 184. Former non-IPA
 symbols (or often-used but non-official IPA symbols) for consonants, vowels and
@@
 represent IPA symbols with ASCII
-character sequences \citep{Wells1987,Wells_etal1992}, using e.g. <p\textbackslash> 
+character sequences \citep{Wells1987,Wells_etal1992}, using e.g.\ <p\textbackslash> 
 for [ɸ]. SAMPA was developed by a group of speech scientists from nine countries 
@@
 (reportedly) uniquely parsable \citep{Wells1995}. However, like the IPA, X-SAMPA
-has different notations for encoding the same phonetic phenomena (cf. Section~
+has different notations for encoding the same phonetic phenomena (cf.\ Section~
 \ref{pitfall-multiple-options-ipa}).
@@
 organizations, several of which were trying to tackle (and in a sense win) the
-multilingual computing environment problem (cf. Section~\ref{encoding}).
+multilingual computing environment problem (cf.\ Section~\ref{encoding}).
 Basically, everyone -- from corporations to governments to language scientists

### 2018-06-05 book/chapters/ipa_meets_unicode.tex
fixes abbreviations, address paperhivers through chapter 1

@@
 various pitfalls with the encoding of IPA in the Unicode Standard. We will
-specifically refer to the 2015 version of the IPA \citep{IPA2015} and the 10.0.0 version of Unicode \citep{Unicode2017}.
+specifically refer to the 2015 version of the IPA \citep{IPA2015} and the 11.0.0 version of Unicode \citep{Unicode2018}.
 
@@
 when an accent mark is added to \textsc{latin small letter i with stroke} <ɨ> at 
-\uni{0268}, the dot is not suppressed but retained, i.e. <ɨ́>.\footnote{According 
+\uni{0268}, the dot is not suppressed but retained, i.e.\ <ɨ́>.\footnote{According 
 to the Unicode Standard, \textsc{latin small letter i with stroke} <ɨ> at 
@@
 vs.<\diaf{0301}> \textsc{combining acute accent} at \unif{0301} for high tone.} 
-In addition, the IPA also adopted tone letters developed by \citet{Chao1930}, e.g. <˥˦˧˨˩>, 
+In addition, the IPA also adopted tone letters developed by \citet{Chao1930}, e.g.\ <˥˦˧˨˩>, 
 which were in wide use by Asianists.\footnote{Not sanctioned
@@
 and ligatures. Multigraphs are groups of characters (in the context of IPA e.g.
-<tʃ> or <ou>) while ligatures are single characters (e.g. <ʧ> \textsc{latin
+<tʃ> or <ou>) while ligatures are single characters (e.g.\ <ʧ> \textsc{latin
 small letter tesh digraph} at \uni{02A7}). Ligatures arose in the context of
@@
 instances of <tʰʷ>, and vice versa. Likewise, a creaky voiced syllabic dental
-nasal can be encoded in various orders, e.g. <n̪̰̩>, <n̩̰̪> or <n̩̪̰>.
+nasal can be encoded in various orders, e.g.\ <n̪̰̩>, <n̩̰̪> or <n̩̪̰>.
 

### 2018-06-05 book/chapters/orthography_profiles.tex
work towards paperhivers chapter 8
fixes abbreviations, address paperhivers through chapter 1

@@
 
-Given these goals, Unicode locale descriptions (see
+Given these goals, Unicode Locales (see
 Chapter~\ref{the-unicode-approach}) might seem like the ideal orthography
@@
        described in the profile. Further, (vi) the tokenization method and (vii)
-       the unicode normalization used should be documented here (see below).
+       the Unicode normalization used should be documented here (see below).
 \end{comment}
@@
        result. Still, we suggest that any software application using orthography
-       profiles should offer both approaches (i.e. \textsc{global} or
+       profiles should offer both approaches (i.e.\ \textsc{global} or
        \textsc{linear}) to the user. The approach used should be documented in
@@
        all combinations. We suggest that any software application using
-       orthography profiles should offer both approaches (i.e. \textsc{NFC} or
+       orthography profiles should offer both approaches (i.e.\ \textsc{NFC} or
        \textsc{NFD}) to the user. The approach used should be documented in the

### 2018-06-05 book/chapters/pitfalls.tex
fixes abbreviations, address paperhivers through chapter 1

@@
 consonant and vowel, which are rendered as two adjacent glyphs by fonts that
-supports Tamil, e.g. \textsc{tamil letter au} at \uni{0B94} represents a single 
+supports Tamil, e.g.\ \textsc{tamil letter au} at \uni{0B94} represents a single 
 character <\tamil{ஔ}>, composed of two glyphs <\tamil{ஓ}> and <\tamil{ன}>. Perhaps confusingly, 
@@
 To solve the issue of tailored grapheme clusters, Unicode offers some assistance
-in the form of the Unicode Locale Descriptions. However, in the practice of
+in the form of the Unicode Locales.\footnote{\url{http://cldr.unicode.org/locale_faq-html}} 
+However, in the practice of
 linguistic research, this is not a real solution. For that reason we propose to
@@
 Hence a Unicode-aware programmer might prefer to say that UTF-8, UTF-16, etc., are Unicode encoding systems because they determine how sequences of bit patterns are determined, which are then mapped to characters.
-The terminological issue here is that the Unicode Standard introduces a layer of indirection between characters and bit patterns, i.e.
-the code point, which can be encoded differently by different encoding systems.
+The terminological issue here is that the Unicode Standard introduces a layer of indirection between characters and bit patterns, i.e.\ the code point, which can be encoded differently by different encoding systems.
 
@@
 Again regarding line breaks, the recommendation "that everybody use this encoding [only LF] whenever possible" (page 31) seems less well motivated (and specific) than most other recommendations in the book.
-Some common formats, e.g.
-CSV (as specified by  RFC 4180 [4]), specify "CRLF" as line break.
+Some common formats, e.g.\ CSV (as specified by  RFC 4180 [4]), specify "CRLF" as line break.
 Thus, the perceived "strong tendency" towards "LF" may be just that.
@@
 Format.''} The details of this issue is extensively explained in section 2.5 of
-the Unicode Core Specification~\citet{Unicode2017}.
+the Unicode Core Specification~\citet{Unicode2018}.
 
@@
 % One Unicode pitfall which may be worth adding is the problem with incomplete implementations of the Unicode Standard.
-This is a problem most big-enough standards suffer from, e.g.
-SQL, XML, XSLT.
+This is a problem most big-enough standards suffer from, e.g.\ SQL, XML, XSLT.
 So although the Unicode Standard mandates that comparison of Unicode text should always be done using normalized text, this is not how the equality operator "==" is implemented for Unicode text in the Python programming language.

### 2018-06-05 book/chapters/preface.tex
fixes abbreviations, address paperhivers through chapter 1

@@
 
-This text is meant as a practical guide for linguists, and programmers, who work with data in multilingual computational environments.\footnote{\url{https://github.com/unicode-cookbook/cookbook}} We introduce the basic concepts needed to understand how writing systems and character encodings function, and how they work together.
+This text is meant as a practical guide for linguists and programmers, who work with data in multilingual computational environments.\footnote{\url{https://github.com/unicode-cookbook/cookbook}} We introduce the basic concepts needed to understand how writing systems and character encodings function, and how they work together.
 
-The intersection of the Unicode Standard and the International Phonetic Alphabet is often not met without frustration by users.
+The intersection of the Unicode Standard and the International Phonetic Alphabet is often met with frustration by users.
 Nevertheless, the two standards have provided language researchers with a consistent computational architecture needed to process, publish and analyze data from many different languages.

### 2018-06-05 book/chapters/unicode_approach.tex
fixes abbreviations, address paperhivers through chapter 1

@@
 
-Each version of the Unicode Standard (\citealp{Unicode2017}, as of writing at
-version 10) consists of a set of specifications and guidelines that include (i) a
+Each version of the Unicode Standard (\citealp{Unicode2018}, as of writing at
+version 11) consists of a set of specifications and guidelines that include (i) a
 core specification, (ii) code charts, (iii) standard annexes and (iv) a
@@
 Hence a Unicode-aware programmer might prefer to say that UTF-8, UTF-16, etc., are Unicode encoding systems.
-The issue is that the Unicode Standard introduces a layer of indirection between characters and bit patterns, i.e.
-the code point, which can be encoded differently by different encoding systems.} Underlyingly, the character encoding is
+The issue is that the Unicode Standard introduces a layer of indirection between characters and bit patterns, i.e.\ the code point, which can be encoded differently by different encoding systems.} Underlyingly, the character encoding is
 represented by a range of numerical values called a \textsc{code space}, which
@@
 That encoded abstract character is rendered on a computer screen (or
-printed page) as a \textsc{glyph}, e.g. <p>, depending on the \textsc{font} and
+printed page) as a \textsc{glyph}, e.g.\ <p>, depending on the \textsc{font} and
 the context in which that character appears.
@@
 The Unicode Standard provides technological specifications for creating locale specific data
-in so-called \textsc{Unicode Locale Descriptions}, i.e.~a set of specification
+in so-called \textsc{Unicode Locales}, i.e.~a set of specification
 that defines a set of language-specific elements (e.g.~tailored grapheme

### 2018-06-05 book/chapters/use_cases.tex
comments out some stuff we probably won't use in the use cases
fixes abbreviations, address paperhivers through chapter 1

--- 2015-06-18 chapters/use_cases.tex
@@
 
-\section{Tokenization and error checking} The most basic use case is
-tokenization of language data. Tokenization comes in three types:
+% ==========================
+\section{Tokenization and error checking}
+\label{usecase-tokenization}
+% ==========================
+
+The most basic use case is tokenization of language data. Tokenization comes in three types:
 
@@
 
+\begin{comment}
 \section{Cross-orthographic analysis of writing systems}
@@
 % any measure of writing systems complexity?
-
-\begin{comment}
 
@@
 character property also has a character property value. So each General Category
-is denoted by a property value by a single letter abbreviation, e.g. ``L'' for
+is denoted by a property value by a single letter abbreviation, e.g.\ ``L'' for
 Letter. Each General Category property also has subcategories that are also
@@
 expression engine includes functionality to match characters at the level of
-script, then it provides the ability to identify characters, e.g. \match{Greek},
+script, then it provides the ability to identify characters, e.g.\ \match{Greek},
 which tests whether letters in a text belong to the Greek script or not. This
@@
 
+\begin{comment}
 \section{PHOIBLE}
@@
 \citet{Moran2012;MoranMcCloy2014} mark tones as singletons with Space Modifier
-Letters, e.g. <˦> for a phonemic high tone, instead of accent diacritics,
+Letters, e.g.\ <˦> for a phonemic high tone, instead of accent diacritics,
 alleviating potential conflicts.
@@
 Building on these works...
+
+\end{comment}

### 2018-06-06 book/chapters/introduction.tex
paperhivers up to page 18

@@
 systems \citep[e.g.~][]{Robinson1995,Powell2012}, a story that we therefore will
-not repeat here. However, the history of turning writing into computer readable
+not repeat here. However, the history of turning writing into machine-readable
 code is not so often told, so we decided to offer a short survey of the major

### 2018-06-06 book/chapters/ipa_meets_unicode.tex
paperhivers up to page 18

@@
 core specification. Specifically related to IPA, various phonetic combinations
-of characters (typically affricates) are available as single code-points in the
+of characters (typically affricates) are available as single code points in the
 Unicode Standard, but are designated \textsc{digraphs}. Such glyphs might be used by

### 2018-06-06 book/chapters/unicode_approach.tex
paperhivers up to page 18

@@
 eradicate the distinction between universal (ASCII) versus language-particular
-(font) by adding as much as possible language-specific information into the
+(font) by adding as much language-specific information as possible into the
 universal standard. However, there are still language/resource-specific
@@
 of universal (Unicode Standard) versus language-particular (locale description).
-The major gain is a much larger compatibility on the universal level (because
-Unicode standardizes a much larger portion of writing system diversity), and
+The major gain is much larger compatibility on the universal level (because
+Unicode standardizes a much greater portion of writing system diversity), and
 much better possibilities for automated processing on the language-particular
-level (because locale descriptions are computer readable specifications).
+level (because locale descriptions are machine-readable specifications).
 
@@
 introduction to the principles of Unicode, Chapter 2 of the core specification,
-called \textsc{general structure}, is particularly insightful. Different from
-many other documents of the Unicode Standard, this general introduction is
+called \textsc{general structure}, is particularly insightful. Unlike 
+many other documents in the Unicode Standard, this general introduction is
 relatively easy to read and illustrated with many interesting examples from
-various orthography traditions all over the world.} The \textsc{core
-specification} is a book directed toward human readers that describes the formal
+various orthographic traditions from all over the world.} The \textsc{core
+specification} is a book aimed at human readers that describes the formal
 standard for encoding multilingual text. The \textsc{code charts} provide a
-humanly readable online reference to the character contents of the Unicode
+human-readable online reference to the character contents of the Unicode
 Standard in the form of PDF files. The \textsc{Unicode Standard Annexes (UAX)}
@@
 this section, we introduce the basic Unicode concepts, but we will leave out
-many details. Please consult the above mentioned full documentation for a more
+many details. Please consult the above-mentioned full documentation for a more
 detailed discussion. Further note that the Unicode Standard is exactly that,
@@
 
-The Unicode Standard is a \textsc{character encoding system} which
-goal it is to support the interchange and processing of written characters and
+The Unicode Standard is a \textsc{character encoding system} whose
+goal is to support the interchange and processing of written characters and
 text in a computational setting.\footnote{An insightful reviewer notes that the term \textsc{encoding} is used for both sequences of code points and text encoded as bit patterns.
@@
 abstractions of orthographic symbols, and it does not define visualizations for
-these characters (although it does presents examples). In contrast, a
+these characters (although it does present examples). In contrast, a
 \textsc{glyph} is a concrete graphical representation of a character as it
@@
 \textsc{font} is then simply a collection of glyphs connected to code points.
-Allography is not specified in Unicode (expect for a few exceptional cases, due
+Allography is not specified in Unicode (barring a few exceptional cases, due
 to legacy encoding issues), but can be specified in a font as a
-\textsc{contextual variant} (a.k.a.~presentation form).
+\textsc{contextual variant} (aka presentation form).
 
@@
 the Unicode Standard Annexes). The character properties assigned to each code
-point is based on each character's behavior in the real-world writing
+point are based on each character's behavior in real-world writing
 traditions. For example, the corresponding lowercase character to \uni{0047} is
@@
 One frequently referenced property is the \textsc{block} property, which is
-often used in software applications to impose some structure to the large number
+often used in software applications to impose some structure on the large number
 of Unicode characters. Each character in Unicode belongs to a specific block.
@@
 also a block \textsc{Arabic Supplement}, \textsc{Arabic Presentation Forms-A}
-and \textsc{Arabic Presentation Form-B}. The situation with Latin symbols is
-even more extreme. In general, the names for block should not be taken as a
+and \textsc{Arabic Presentation Forms-B}. The situation with Latin symbols is
+even more extreme. In general, the names of blocks should not be taken as a
 definitional statement. For example, many IPA symbols are not located in the
@@
 character}. Sequences of multiple code points that correspond to a single
-user-perceived characters are called grapheme clusters in Unicode parlance.
+user-perceived characters are called \textsc{grapheme clusters} in Unicode parlance.
 Grapheme clusters come in two flavors: (default) grapheme clusters and tailored
@@
 by \textsc{combining tilde} at \uni{0303}) combines visually into <ñ>, a
-user-perceived character in writing systems like that of Spanish. So, what the
+user-perceived character in writing systems like that of Spanish. In effect, what the
 user perceives as a single character actually involves a multi-code-point
 sequence. Note that this specific sequence can also be represented with a single
-so-called \textsc{precomposed code point}, \textsc{latin small letter n with
-tilde} at \uni{00F1}, but this is not the case for all multi-code point
-character sequences. The problem that there multiple encodings possible for the
-same text element has been acknowledged early on in the Unicode Standard
-(e.g.~for <ñ>, the sequence \uni{006E} \uni{0303} should in all situations be
-treated identically to the precomposed \uni{00F1}), and a system of
-\textsc{canonical equivalence} is available for such situations. Basically, the
-Unicode Standard offers different kind of normalizations to either decompose all
-precomposed characters (called \textsc{NFD}, \textsc{normalization form
-canonical decomposition}), or compose as much as possible combinations (called
-\textsc{NFC}, \textsc{normalization form canonical composition}).\footnote{See the 
+so-called \textsc{precomposed code point}, the \textsc{latin small letter n with
+tilde} at \uni{00F1}, but this is not the case for all multi-code-point
+character sequences. A solution to the problem of multiple encodings for the
+same text element was developed early on in the Unicode Standard. It is called 
+\textsc{canonical equivalence}, e.g.~for <ñ>, the sequence \uni{006E} \uni{0303} 
+should in all situations be treated identically to the precomposed \uni{00F1}. By doing so, 
+Unicode can also support special or precomposed characters in legacy character sets. 
+To determine canonical equivalence, the Unicode Standard offers different kinds of normalization to either 
+decompose precomposed characters (called \textsc{NFD} for \textsc{normalization form
+canonical decomposition}) or to combine sequences of code points into precomposed 
+characters (called \textsc{NFC} for \textsc{normalization form canonical composition}).\footnote{See the 
 Unicode Standard Annex \#15, Unicode Normalization Forms (\url{http://unicode.org/reports/tr15/}), 
 which provides a detailed description of normalization algorithms and illustrated examples.} 
-In current practice of software development, NFC seems to be preferred in most situations
-and is widely proposed as the preferred canonical form. We discuss Unicode Normalization 
+In current software development practice, NFC seems to be preferred in most situations
+and is widely proposed as the preferred canonical form. We discuss Unicode normalization 
 in detail in Section \ref{pitfall-canonical-equivalence}.
@@
 language-by-language or writing-system-by-writing-system basis. They are also 
-grapheme clusters in these orthographies for processes such as collation.\footnote{\url{https://www.unicode.org/glossary/#collation}}
+grapheme clusters in these orthographies for processes such as collation (i.e.\ sorting).\footnote{\url{https://www.unicode.org/glossary/#collation}}
 
-The Unicode Standard provides technological specifications for creating locale specific data
-in so-called \textsc{Unicode Locales}, i.e.~a set of specification
-that defines a set of language-specific elements (e.g.~tailored grapheme
+The Unicode Standard provides technical specifications for creating locale specific data
+in so-called \textsc{Unicode Locales}, i.e.~specifications 
+that define a set of language-specific elements (e.g.~tailored grapheme
 clusters, collation order, capitalization-equivalence), as well as other special

### 2018-06-06 book/chapters/orthography_profiles.tex
deals with paperhivers through chapter 2

@@
 profiles. However, there are various practical obstacles preventing the use of
-such locale descriptions in the daily linguistic practice, namely: (i) the
+Unicode Locales in the daily linguistic practice, namely: (i) the
 XML structure is too verbose to easily and quickly produce or correct manually,
-(ii) locale descriptions are designed for a wide scope of information (like date
+(ii) Unicode Locales are designed for a wide scope of information (like date
 formats or names of weekdays) most of which is not applicable for documenting
 writing systems, and (iii) most crucially, even if someone made the effort to
-produce a technically correct locale description for a specific source at hand,
+produce a technically correct Unicode Locale for a specific source at hand,
 then it is well-nigh impossible to deploy the description. This is because a locale

### 2018-06-06 book/chapters/unicode_approach.tex
deals with paperhivers through chapter 2

@@
 below. Within the Unicode structure many of these specifications can be captured
-by so-called \textsc{locale descriptions}, so we are moving to a new distinction
-of universal (Unicode Standard) versus language-particular (locale description).
+by so-called \textsc{Unicode Locales}, so we are moving to a new distinction
+of universal (Unicode Standard) versus language-particular (Unicode Locale).
 The major gain is much larger compatibility on the universal level (because
@@
 much better possibilities for automated processing on the language-particular
-level (because locale descriptions are machine-readable specifications).
+level (because Unicode Locales are machine-readable specifications).
 
@@
 specifications for those cases.} Another use of properties is to define the
-script of a character.\footnote{The Unicode Glossary defines the term \textsc{script} as
+script of a character.\footnote{The Glossary of Unicode Terms defines the term \textsc{script} as
 a ``collection of letters and other written signs used to represent textual
@@
 combinations are defined in the Unicode Standard as functioning as one
-\textsc{text element}.\footnote{The Unicode Glossary defines text element as:
+\textsc{text element}.\footnote{The Glossary of Unicode Terms defines \textsc{text element} as:
 ``A minimum unit of text in relation to a particular text process, in the
@@
 language-by-language or writing-system-by-writing-system basis. They are also 
-grapheme clusters in these orthographies for processes such as collation (i.e.\ sorting).\footnote{\url{https://www.unicode.org/glossary/#collation}}
+grapheme clusters in these orthographies for processes such as collation (i.e.\ sorting).\footnote{\url{https://www.unicode.org/glossary/\#collation}}
 
@@
 Descriptions allow users to define language- or even resource-specific writing
-systems or orthographies.\footnote{The Unicode Glossary defines \textsc{writing
+systems or orthographies.\footnote{The Glossary of Unicode Terms defines \textsc{writing
 system} only very loosely, as it is not a central concept in the Unicode
@@
 system, the British English writing system, the French writing system, and the
-Japanese writing system.''} However, there are various drawbacks of locale
-descriptions for the daily practice of linguistic work in a multilingual setting.
+Japanese writing system.''} However, Unicode Locales have various drawbacks 
+for the daily practice of scientific linguistic research in a multilingual setting.

### 2018-06-06 book/chapters/pitfalls.tex
deals with paperhivers through chapter 2
through page 21

@@
 with other linguists. This section is not meant as a criticism of the decisions
-made by the Unicode Consortium; on the contrary, we aim to highlight where the
+made by the Unicode Consortium; rather we aim to highlight where the
 technological aspects of the Unicode Standard diverge from many users'
@@
 also some cases in which the Unicode Standard has made decisions that
-theoretically could have been taken differently, but for some reason or another
+theoretically could have been made differently, but for some reason or another
 (mostly very good reasons) were accepted as they are now. We call such behavior that
@@
 experience is that in almost all cases the behavior of the Unicode Standard has
-been particularly well thought through. The Unicode Consortium has a more
+been particularly well thought through. The Unicode Consortium has a 
 wide-ranging view of matters and often examines important practical use-cases
-that from a linguistic point of view are normally not considered. Our general
+that are not normally considered from a linguistic point of view. Our general
 guideline for dealing with the Unicode Standard is to accept it as it is, and
-not to battle windmills. Alternatively, of course, it is possible to actively
+not to tilt at windmills. Alternatively, of course, it is possible to actively
 engage in the development of the standard itself, an effort that is highly
@@
 character is the abstract notion of a symbol in a writing system, while a glyph
-is the concrete drawing of such a symbol. In practice, there is a complex
+is the visual representation of such a symbol. In practice, there is a complex
 interaction between characters and glyphs. A single Unicode character may of
@@
 consonant and vowel, which are rendered as two adjacent glyphs by fonts that
-supports Tamil, e.g.\ \textsc{tamil letter au} at \uni{0B94} represents a single 
+support Tamil, e.g.\ \textsc{tamil letter au} at \uni{0B94} represents a single 
 character <\tamil{ஔ}>, composed of two glyphs <\tamil{ஓ}> and <\tamil{ன}>. Perhaps confusingly, 
@@
 <{\fontspec{Courier}g}> in the Linux Libertine and Courier fonts, respectively,
-because their typefaces are designed differently. Furthermore, font face may
+because their typefaces are designed differently. Furthermore, the font face may
 change the visual appearance of a character, for example Times New Roman
@@
 
-In sum, character-to-glyph mappings are complex technological issues that the
+In sum, character-to-glyph mappings are complex technical issues that the
 Unicode Consortium has had to address in the development of the Unicode
-Standard, but for the lay user they can be utterly confusing because visual
+Standard. However, they can can be utterly confusing for the lay user because visual
 rendering does not (necessarily) indicate logical encoding.
@@
 
-The Unicode Standard is a character encoding system, and not a writing system
-encoding system. This most forcefully becomes clear with the notion of grapheme.
+The Unicode Standard encodes characters. This becomes most clear with the notion of grapheme.
 From a linguistic point of view, graphemes are the basic building blocks of a
 writing system (see Section~\ref{linguistic-terminology}). It is extremely
-common, up to the point of being universally attested, that writing systems use
-combinations of multiple symbols as a single grapheme, like <sch>, <th> or <ei>.
-There is no possibility to encode such complex graphemes using the Unicode Standard.
+common for writing systems to use 
+combinations of multiple symbols (or letters) as a single grapheme, such as <sch>, <th> or <ei>.
+There is no way to encode such complex graphemes using the Unicode Standard.
 
-The Unicode Standard deals with complex graphemes only inasmuch they consist of
+The Unicode Standard deals with complex graphemes only inasmuch as they consist of
 base characters with diacritics (see
 Section~\ref{pitfall-different-notions-of-diacritics} for a discussion of the
-notion of diacritic). The Unicode Standard calls such combination \textit{grapheme
+notion of diacritic). The Unicode Standard calls such combinations \textit{grapheme
 clusters}. Complex graphemes consisting of multiple base characters,
-like <sch>, are called \textit{tailored grapheme clusters} in Unicode parlance (see
+like <sch>, are called \textit{tailored grapheme clusters} (see
 Chapter~\ref{the-unicode-approach}).
 
-Inspecting the Unicode Standard, there appear to be special Unicode characters
-to glue together characters into larger tailored grapheme clusters,
+There are special Unicode characters that 
+glue together characters into larger tailored grapheme clusters,
 specifically the \textsc{zero width joiner} at \uni{200D} and the
@@
 To solve the issue of tailored grapheme clusters, Unicode offers some assistance
-in the form of the Unicode Locales.\footnote{\url{http://cldr.unicode.org/locale_faq-html}} 
+in the form of Unicode Locales.\footnote{\url{http://cldr.unicode.org/locale_faq-html}} 
 However, in the practice of
-linguistic research, this is not a real solution. For that reason we propose to
+linguistic research, this is not a real solution. To address this issue, we propose to
 use orthography profiles (see Chapter~\ref{orthography-profiles}). Basically,
-both orthography profiles and locale descriptions offer a way to specify
+both orthography profiles and Unicode Locales offer a way to specify
 tailored grapheme clusters. For example, for English one could specify that <sh>
@@
 The central problem behind the missing glyphs is that designing actual glyphs
-includes a lot of different considerations and it is a time-consuming process.
-Many traditional expectations of how specific characters should look like have
+includes many different considerations and it is a time-consuming process. 
+Many traditional expectations of what specific characters should look like have
 to be taken into account when designing glyphs. Those expectations are often not
 well documented, and it is mostly up to the knowledge and experience of the font
-designer to try and conform to them as good as possible. Therefore, most
-designers produce fonts only including glyphs for certain parts of the Unicode
-Standard, namely for those characters they feel comfortable with. At the same
+designer to try and conform to them as well as possible. Therefore, most
+designers produce fonts that only include glyphs for certain parts of the Unicode
+Standard. At the same
 time, the number of characters defined by the Unicode Standard is growing with
-each new version, so it is neigh impossible for any designer to produce glyphs
-for all characters. The result of this is that, almost necessarily, each font
+each new version, so it is nigh impossible for any designer to produce glyphs
+for every character. The result of this is that, out of necessity, each font
 only includes glyphs for a subset of the characters in the Unicode Standard.
 
-The simple solution to missing glyphs is thus to install additional fonts
+A simple solution to missing glyphs is to install additional fonts
 providing additional glyphs. For the more exotic characters there is often not

### 2018-06-06 book/chapters/ipa_meets_unicode.tex
paperhivers up to page 25 or so...

@@
 IPA font Doulos SIL
-(\url{http://scripts.sil.org/cms/scripts/page.php?item\_id=DoulosSIL}) does not
+(\url{https://software.sil.org/doulos/}) does not
 have real italics. This leads some word-processing software, like Microsoft

### 2018-06-08 book/chapters/ipa_background.tex
paperhivers comments through chapter 3

@@
 In this chapter we present a brief history of the IPA 
-(Section~\ref{IPAhistory}), which itself dates 
+(Section~\ref{IPAhistory}), which dates 
 back to the late 19th century, not long after the creation of the first 
@@
 faced when digitally encoding the IPA's set of symbols and 
-diacritics (Section~\ref{EncodingIPA}). Occurring a bit over one hundred 
-years after the IPA's inception, the encoding challenge was no small feat 
+diacritics (Section~\ref{EncodingIPA}). Occurring a little over a hundred years after 
+the inception of the IPA, its encoding was a major challenge 
 (Section~\ref{need-for-multilingual-environment}); many 
-linguists have encountered the pitfalls when the twain shall meet 
+linguists have encountered the pitfalls when the two are used together 
 (Chapter~\ref{ipa-meets-unicode}).
@@
 using words in different languages. In 1900 the symbols were first organized into
-a chart and were given phonetic feature labels (e.g.\ for manner of
-articulation among others ``plosives'', ``nasales'', ``fricatives'', for place of
-articulation among others ``bronchiales'', ``laryngales'', ``labiales'' and for vowels
-e.g.\ ``fermées'', ``mi-fermées'', ``mi-ouvertes'', ``ouvertes''). Throughout the last
+a chart and were given phonetic feature labels, e.g.\ for manner of
+articulation among others \textit{plosives}, \textit{nasales}, \textit{fricatives}, for place of
+articulation among others \textit{bronchiales}, \textit{laryngales}, \textit{labiales} and for vowels
+e.g.\ \textit{fermées}, \textit{mi-fermées}, \textit{mi-ouvertes}, \textit{ouvertes}. Throughout the last
 century, the structure of the chart has changed with increases in phonetic
@@
 accommodate scientific innovations and discoveries. In fact, updates are part of the
-Association's mandate. These changes are captured in revisions to the IPA chart.\footnote{For a detailed history see:
-\url{https://en.wikipedia.org/wiki/History\_of\_the\_International\_Phonetic_Alphabet}}
+Association's mandate. These changes are captured in revisions to the IPA chart.\footnote{For a detailed history, 
+we refer the reader to:
+\url{https://en.wikipedia.org/wiki/History\_of\_the\_International\_Phonetic_Alphabet}.}
 
-Over the years there have been several revisions, but mostly minor. Articulation 
-labels -- what are often called `features' even though the IPA
-deliberately avoids this term -- have changed (e.g.\ terms like ``lips'', ``throat''
-or ``rolled'' are not used anymore). Phonetic symbol values have changed (e.g.\
-voiceless is no longer marked by <h>). Symbols have been dropped (e.g.\ the
-caret diacritic denoting `long and narrow' is no longer used). And many symbols
+Over the years there have been several revisions, but mostly minor ones. Articulation 
+labels -- what are often called \textit{features} even though the IPA
+deliberately avoids this term -- have changed, e.g.\ terms like \textit{lips}, \textit{throat}
+or \textit{rolled} are no longer used. Phonetic symbol values have changed, e.g.\
+voiceless is no longer marked by <h>. Symbols have been dropped, e.g.\ the
+caret diacritic denoting `long and narrow' is no longer used. And many symbols
 have been added to reflect contrastive sounds found in the world's very diverse
 phonological systems. The use of the IPA is guided by principles outlined in 
-the \textsc{handbook of the international phonetic association} \citep{IPA1999}, 
+the \textit{Handbook of the International Phonetic Association} \citep{IPA1999}, 
 henceforth simply called \textit{Handbook}.
@@
        sounds in the world's (spoken) languages;
-	\item its chart has a linguistic basis (specifically a phonological bias)
+	\item its chart has a linguistic basis (and specifically a phonological bias)
        rather than just being a general phonetic notation scheme;
@@
 restricts phonetic detail to linguistically relevant aspects of speech.
-Phonological consideration thus inextricably play a roll in transcription. In
+Phonological considerations thus inextricably play a roll in transcription. In
 other words, phonetic observations beyond quantitative acoustic analysis are
@@
 \textit{Report on the 1989 Kiel convention} published in the \textit{Journal of
-the International Phonetic Association} \citep[67-68]{International1989report}:
+the International Phonetic Association} \citep[67--68]{International1989report}:
 
@@
 \citep[27]{IPA1999}. Distinctive sounds became later known as \textsc{phonemes}
-and the IPA has developed historically into a notational devise with a strictly
+and the IPA has developed historically into a notational device with a strictly
 segmental phonemic view. A phoneme is an abstract theoretical notion derived
@@
 
-Essentially transcription has two parts: a text containing symbols and a set 
+Essentially, transcription has two parts: a text containing symbols and a set 
 of conventions for interpreting those symbols (and their combinations). 

### 2018-06-08 book/chapters/ipa_background.tex
through paperhivers chapter 4

@@
 falls under the rubric of diacritically-distinguished symbols, e.g.\ the 
-difference in English between an aspirated /p/ and /t/ in [spʰot̚] and 
-an unreleased /p/ in [stʰop̚].
+difference in English between an aspirated /p/ in [pʰæt] and 
+an unreleased /p/ in [stop̚].
 
@@
 \textit{narrow} phonetic transcription and a \textit{broad} phonological transcription. 
-A narrow phonetic transcription may freely use the all symbols in the IPA 
+A narrow phonetic transcription may freely use all symbols in the IPA 
 chart with direct reference to their phonetic value, i.e.\ the transcriber can 
@@
 
-In contrast, the basic goal of a phonemic broad transcription is to distinguish all
+In contrast, the basic goal of a broad phonemic transcription is to distinguish all
 words in a language with a minimal number of transcription symbols
@@
 phonemically in IPA as /ieaou/, though phonetically they are closer to [iεaɔu].
-The Japanese five vowel system can also be transcribed in IPA as
+The Japanese five-vowel system can also be transcribed in IPA as
 /ieaou/, while the phonetic targets in this case are closer to [ieaoɯ].
@@
 on the keyboard to IPA symbols. The process of assigning standardized computer
-codes to phonetic symbols was given to the \textsc{workgroup on computer
+codes to phonetic symbols was given to the \textit{workgroup on computer
 coding} (henceforth \textit{working group}) at the Kiel Convention. This working
-group was assigned to consider the following tasks
+group was given the following tasks
 \citep{Esling1990,EslingGaylord1993}:
@@
 a first step in formalizing the IPA computationally because it would give 
-each symbol an unambiguous numerical identifier called an \textsc{IPA number}. 
+each symbol an unambiguous numerical identifier called an \textsc{ipa number}. 
 The numbering system was to be comprehensive enough to support future revisions 
@@
 Although the Association had never officially approved a set of names 
-for the IPA symbols, each IPA symbol received a unique \textsc{IPA name}. 
+for the IPA symbols, each IPA symbol received a unique \textsc{ipa name}. 
 Many symbols already had an informal name (or two) used by linguists, but 
@@
 for the IPA should be independent of computer environments or formats, 
-i.e.\ the IPA Number was not meant to be encoded at the bit pattern level.
+i.e.\ the \textsc{ipa number} was not meant to be encoded at the bit pattern level.
 The working group report's declaration includes the following explanatory 
@@
 aspects of the actual encoding in computers (cf.\ \cite{Anderson1984}). However, 
-it was meant to serve as a basis for a communication-interchange standard for 
+it was meant to serve as a basis for a interchange standard for 
 creating mapping tables from various computer encodings, fonts, phonetic-character-set 
-software, etc., to common IPA Numbers, and therefore symbols.\footnote{Remember, at 
+software, etc., to common IPA numbers, and therefore symbols.\footnote{Remember, at 
 this time in the late 1980s there was no stable multilingual computing environment. 
@@
 
-The working group assigned each IPA symbol to a unique three-digit number, i.e.\
-an IPA Number. Encoded in this number scheme implicitly includes information
+The working group assigned each IPA symbol to a unique three-digit IPA number. 
+Encoded in this number scheme implicitly is information
 about the status of each symbol (see below). The IPA numbers were listed with the
@@
 revision of 1989 and its update in 1996. Although the numbering scheme still 
-exists, in practice is it superseded by the Unicode codification of symbols.
+exists, in practice it is superseded by the Unicode codification of symbols.
 
@@
 as members of the group of symbols formerly recognized by the IPA. To ensure 
-backwards compatibility, legacy IPA symbols would retain an IPA Number and 
-an IPA Name for reference purposes. As we discuss below, this decision is 
+backwards compatibility, legacy IPA symbols would retain an IPA number and 
+an IPA name for reference purposes. As we discuss below, this decision is 
 later reflected in the Unicode Standard as many legacy IPA symbols still reside in 
-the \textsc{IPA extensions} block.
+the \textsc{IPA Extensions} block.
 
-The IPA Number is expressed as a three-digit number. The first digit 
+The IPA number is expressed as a three-digit number. The first digit 
 indicates the symbol's category \citep{Esling1990,EslingGaylord1993}:
@@
 symbols (or often-used but non-official IPA symbols) for consonants, vowels and
-diacritics are numbered from x99 backwards. For example, the voiceless and
+diacritics are numbered from 299 backwards. For example, the voiceless and
 voiced postalveolar affricates and fricatives <~č,~ǰ,~š,~ž~> are assigned the
@@
 tables that reference ASCII representations of the IPA, and this early
-computational representation of the IPA became the basis X-SAMPA and for the
+computational representation of the IPA became the basis for X-SAMPA and for the
 inclusion of the IPA into the Unicode Standard version 1.0.
@@
 True to the working group's aim, the IPA numbers provided a mechanism for 
-a communication interchange standard for creating mapping tables to various 
+an interchange standard for creating mapping tables to various 
 computer encodings. For example, the IPA coding system was used as a mapping 
@@
 represent <ŋ>. Tech-savvy linguists redefined selected characters from a
-character encoding by mapping custom made fonts to specific code points.\footnote{For 
-example, SIL's popular font SIL IPA 1990.}. However,
+character encoding by mapping custom-made fonts to specific code points.\footnote{For 
+example, SIL's popular font SIL IPA 1990.} However,
 one linguist's electronic text would not render properly on another linguist's
@@
 defining keyboard-able sequences consisting of ASCII combinations as IPA symbols. 
-For example, \cite{Wells1987} provides an in-depth description of IPA
+For example, \citet{Wells1987} provides an in-depth description of IPA
 codings from country-to-country. Later ASCII-IPAs include Kirshenbaum (created
 in 1992 in a Usenet group and named after its lead developer who was at
-Hewlett-Packard Laboratories) and Worldbet (published in 1993 by
-\cite{Hieronymus1993}, who was at AT\&T Laboratories). 
+Hewlett-Packard Laboratories) and Worldbet (published by
+\citet{Hieronymus1993}, who was at AT\&T Laboratories). 
 The most successful effort was SAMPA (Speech Assessment
@@
 
-During the 1980's, it became increasingly clear that an adequate solution 
+During the 1980s, it became increasingly clear that an adequate solution 
 to the problem of multilingual computing environments was needed. Linguists 
@@
 encoding initiative} (TEI) \citep{EslingGaylord1993}. The working group's
-formalization of the IPA, i.e.\ a full listing of agreed upon computer codings
+formalization of the IPA, i.e.\ a full listing of agreed-upon computer codings
 for phonetic symbols, was used in developing writing system descriptions, which
@@
 characters submitted to them in May 1993. The set of IPA characters were
-assigned UCS codes in 16 bit representation (in hexadecimal) and were published
+assigned UCS codes in 16-bit representation (in hexadecimal) and were published
 as Tables 2 and 3 in \cite{EslingGaylord1993}, which include a graphical
-representation of the IPA Symbol, its IPA Name, phonetic description, IPA
-Number, UCS Code and AFII Code.\footnote{The \textsc{association for font
-information interchange} (AFII) was an international database of glyphs created
+representation of the IPA symbol, its IPA name, phonetic description, IPA
+number, UCS code and AFII code.\footnote{The Association for Font
+Information Interchange (AFII) was an international database of glyphs created
 to promote the standardization of font data required to produce ISO/IEC 10646.} When the
@@
 It was a long journey, but the goal of achieving a single multilingual computing
-environment has largely been accomplished. As such, we should not dismiss the IPA Numbers 
+environment has largely been accomplished. As such, we should not dismiss the IPA numbers 
 or pre-Unicode encoding attempts, such as SAMPA/X-SAMPA, as misguided. The parallels 
-between the IPA Numbers and Unicode Code points, for example, is striking because both the IPA and 
+between the IPA numbers and Unicode Code points, for example, are striking because both the IPA and 
 the Unicode Consortium came up with the solution of an additional layer of indirection (an abstraction layer) 
@@
 Unicode 88 provided the basic principles for the Unicode Standard's design -- 
-pushing for 16 bit representations of characters with a clear distinction 
+pushing for 16-bit representations of characters with a clear distinction 
 between characters and glyphs. Some of the contents of this status proposal 

### 2018-06-08 book/chapters/unicode_approach.tex
through paperhivers chapter 4

@@
 alleviate this problem of assigning different symbols to the same code points,
-in the late 80's and early 90's the Unicode Consortium set itself the ambitious
+in the late 80s and early 90s the Unicode Consortium set itself the ambitious
 goal of developing a single universal character encoding to provide a unique

### 2018-06-08 book/chapters/ipa_meets_unicode.tex
deals with paperhivers comments through chapter 5

@@
 keyboard. Some further problematic homoglyphs related to encoding IPA in the
-Unicode Standard are the following.
-
-\begin{itemize}
-
-   \item The uses of the apostrophe has led to long discussions on the Unicode
+Unicode Standard are the following:
+
+\begin{itemize}
+
+   \item The uses of the apostrophe have led to long discussions on the Unicode
         Standard email list. An English keyboard inputs the symbol
@@
         character, namely <\dia{02BC}>, i.e.~the \textsc{modifier letter apostrophe} 
-        at \uni{02BC}, which unfortunately looks mostly extremely similar to
+        at \uni{02BC}, which unfortunately looks extremely similar to
         \uni{2019}. 
@@
         line} at \uni{007C}.
-  \item For the marking of length there is a special Unicode characters, namely
+  \item For the marking of length there is a special Unicode character, namely
         <\dia{02D0}> \textsc{modifier letter triangular colon} at \uni{02D0}. However,
@@
 <ɑ> use the same glyph. For example, in Linux Libertine (the font of this book)
-the italics of these characters are highly similar <\textit{a}> and
+the italics of these characters are highly similar, <\textit{a}> and
 <\textit{ɑ}>, while in Charis SIL they are identical: <\textit{\charis{a}}> and
@@
 
-The first case in which the IPA allows for different encodings is the question
+\noindent The first case in which the IPA allows for different encodings is the question
 of how to transcribe tone (cf.\ \cite{Maddieson1990}). There is an old tradition 
@@
 % "Either of the variant letter shapes [g] and [g] may be used to represent the voiced velar plosive."
+% https://www.cambridge.org/core/journals/journal-of-the-international-phonetic-association/information/instructions-contributors
 
@@
 both the keyboard \textsc{latin small letter g} and the \textsc{latin small
-letter script g} are valid input characters for the voiced velar plosive \citep[19]{IPA1999}.
+letter script g} are valid input characters for the voiced velar plosive 
+\citep[19]{IPA1999}.\footnote{Note however that the current instructions for contributors to 
+the Journal of the International Phonetic Association requires the use of 
+opentail <ɡ> and not looptail <g>.}
 Unfortunately, this decision further introduces ambiguity for linguists trying
@@
 Finally, the IPA states that ``diacritics may be placed above a symbol with a
-descender''. For example, for marking marking of voiceless pronunciation of
+descender''. For example, for marking of voiceless pronunciation of
 voiced segments the IPA uses the ring diacritic. Originally, the ring should be
@@
 In the major revision of the IPA in 1932, affricates were represented by two
-consonants <tʃ>, ligatures <ʧ>, or with the tie-bar <t͡ʃ>. In the
-1938 revision the tie-bar's semantics were broadened to indicate simultaneous
-articulation, as for example in labial velars such as <k͡p>. Thus, the tie-bar is a
+consonants <tʃ>, ligatures <ʧ>, or with the tie bar <t͡ʃ>. In the
+1938 revision the tie bar's semantics were broadened to indicate simultaneous
+articulation, as for example in labial velars such as <k͡p>. Thus, the tie bar is a
 convenient diacritic for visually tokenizing input strings into chunks of
@@
 combine more than two base characters into one tie bar, but this is not possible
-with Unicode (i.e.~there is no possibility to draw a tie bar over three of four
+with Unicode (i.e.~there is no possibility to draw a tie bar over three or four
 characters). It is possible to indicate such larger groups by repeating the tie
@@
 
-Most importantly though, in comparison to normal Unicode processing, the tie-bar
+Most importantly though, in comparison to normal Unicode processing, the tie bar
 actually takes a reversed approach to complex graphemes. Basically, the Unicode
@@
 Although many combinations of base character with diacritic are treated as
-canonical equivalent with precomposed characters, there are a few combinations
+ with precomposed characters, there are a few combinations
 in IPA that allow for multiple, apparently identical, encodings that are not
-canonical equivalent (see Section~\ref{pitfall-canonical-equivalence} on the
+ (see Section~\ref{pitfall-canonical-equivalence} on the
 principle of canonical equivalence). For that reason, the following elements
@@
 characters exist as well as precomposed combination in Unicode, and these
-precomposed characters should preferably be used. When instead combinations of a base character with
-diacritic are used, then these combinations are not canonical equivalent to the
+precomposed characters should preferably be used. When combinations of a base character with
+diacritic are used instead, then these combinations are not  to the
 precomposed combinations. This means that any search will not find both at the
@@
 A similar problem arises with the rhotic hook. There are two precomposed
-characters in Unicode with a rhotic hook, which are not canonical equivalent 
+characters in Unicode with a rhotic hook, which are not  
 with a combination of the vowel with a separately encoded hook:
@@
 Reversely, note that the <ç> \textsc{latin small letter c with cedilla} at
-\uni{00E7} is canonically equivalent with <c> with <\dia{0327}>
+\uni{00E7} is  with <c> with <\dia{0327}>
 \textsc{combining cedilla} at \uni{0327}, so it will be separated into two
@@
 
-Another pitfall relates to the question of what are diacritics. The problem is that
-the meaning of the term diacritics as used by the IPA is not the same as is used
+Another pitfall relates to the question, what is a diacritic? The problem is that
+the meaning of the term \textsc{diacritic} as used by the IPA is not the same as it is used
 in the Unicode Standard. Specifically, diacritics in the IPA-sense are either
@@
 
-	\item[] Tone letters, including but nut limited to: 
+	\item[] Tone letters, including but not limited to: 
 	\begin{itemize} 
@@
 
-If intended, then any default combination of Spacing Modifiers
+\noindent If intended, then any default combination of Spacing Modifiers
 with the preceding character can be specified in orthography profiles (see
@@
 back from using <tʰʷ> instead (with exactly the same intended meaning). However,
-from a technical standpoint, these two sequences are different, e.g.~if both
+from a technical standpoint, these two sequences are different; if both
 sequences are used in a document, searching for <tʷʰ> will not find any
@@
 
-By Unicode Canonical Combining Classes, overlay diacritics 
-like <\dia{0334}> (Combining Class number 1), always come before diacritics
+According to Unicode Canonical Combining Classes, overlay diacritics 
+like <\dia{0334}> (Combining Class number 1) always come before diacritics
 below (Combining Class number 220), which in turn always come before diacritics
@@
 \noindent Then, when a character sequence contains more than one character of the Spacing
-Modifier Letters, these will be placed after all combining diacritic marks in the
+Modifier Letters, these will be placed after all combining diacritical marks in the
 following order:
@@
 \ref{the-international-phonetic-alphabet}, distinct symbols are favored for
-cases of phonological contrast. Further, also convenience of display in the chart
-must be taken into account when arguing for or against the inclusion or deletion
+cases of phonological contrast. Further, convenience of display in the chart 
+must also be taken into account when arguing for or against the inclusion or deletion
 of IPA symbols in the IPA chart. Finally, the inclusion or deletion of symbols
@@
 voiceless implosives in Africa; and (iii) that the sounds are sufficiently rare
-as to not need a whole new row of symbols in the chart. Ladefoged favored
+so as not to need a whole new row of symbols in the chart. Ladefoged favored
 symbolizing the sounds using a voiceless diacritic ring below voiced implosives,
@@
 implosives containing voiceless and voiced
-pairs.\footnote{https://en.wikipedia.org/wiki/File:IPA\_as\_of\_1989.png} But
-already in the next revision, in 1993 (with an update in 1996), the voiceless
+pairs.\footnote{https://en.wikipedia.org/wiki/File:IPA\_as\_of\_1989.png} 
+But in the next revision, in 1993 (with an update in 1996), the voiceless
 implosives were dropped. The implosives row from the IPA consonantal chart
@@
 The point of this pitfall is to highlight that revisions to the IPA will 
-continue into the future, albeit they are infrequent. Nevertheless, 
+continue into the future, albeit infrequently. Nevertheless, 
 given the Unicode Standard's principle of maintaining backwards compatibility 
@@
 There is the possibility that at a later revision of the IPA, symbols that 
-are not currently encoded in the Unicode Standard are added to the IPA -- 
+are not currently encoded in the Unicode Standard will be added to the IPA -- 
 although we think this is unlikely.
@@
       is confusingly referred to as \textsc{latin letter retroflex click} in the
-      Unicode Standard, which is probably best seen as an error.
+      Unicode Standard, which is probably best considered an error.
 \item \textsc{Voiced retroflex implosive} \newline 

### 2018-06-08 book/chapters/orthography_profiles.tex
deals with all paperhiver comments except a few time consuming ones

@@
 this string to reorder the code points into a canonical order, allowing the data
-to be treated canonically equivalent for search and comparison.
+to be treated  for search and comparison.
 

### 2018-06-08 book/chapters/practical_recommendations.tex
Resolves issue #29
deals with all paperhiver comments except a few time consuming ones

@@
 Instead, we provide links to quickly find general information about the Unicode Standard and the International Phonetic Alphabet (IPA).
-And we target ordinary working linguists who want to know how to easily insert special characters into their digital documents and applications.
+We target ordinary working linguists who want to know how to easily insert special characters into their digital documents and applications.
 
@@
 
-There are also Web-based Unicode character pickers available through the browser that allow for the creation and insert of special characters, which can then be copied \& pasted into documents or software applications.
+There are also web-based Unicode character pickers available through the browser that allow for the creation and insertion of special characters, which can then be copied \& pasted into documents or software applications.
 For example, try:
@@
 Again, this method is burdensome for rapidly inserting characters.
-For access to the full range of Unicode characters, the Character Map program comes preinstalled on all Microsoft Operating systems.
+For access to the full range of Unicode characters, the Character Map program comes preinstalled on all Microsoft operating systems.
 
-There are also many third party applications that provide virtual keyboards.
-These programs typically override keys or keystrokes on the user's keyboard  allowing them to quickly keyboard special characters (once the layout of the new keyboard is mastered).
+There are also many third-party applications that provide custom keyboard layouts.
+These programs typically override keys or keystrokes on the user's keyboard allowing them to quickly enter special characters (once the layout of the new keyboard is mastered).
 They can be language-specific or devoted specifically to IPA.
@@
 
-There are several good Unicode IPA character pickers available on the Web and through the browser, including:
+There are several good Unicode IPA character pickers available through the browser, including:
 
@@
 
-Regarding fonts that display Unicode IPA correctly, many linguists turn to the IPA Unicode fonts developed by SIL Internnational.
+Regarding fonts that display Unicode IPA correctly, many linguists turn to the IPA Unicode fonts developed by SIL International.
 The complete SIL font list is available online.\footnote{\url{http://scripts.sil.org/SILFontList}} There is also a page that describes IPA transcription using the SIL fonts and provides an informative discussion on deciding which font to use.\footnote{\url{http://scripts.sil.org/ipahome}} Traditionally, IPA fonts popular with linguists were created and maintained by SIL International, so it is often the case in our experience that we encounter linguistics data in legacy IPA fonts, i.e.\ pre-Unicode fonts such as SIL IPA93.\footnote{\url{http://scripts.sil.org/FontFAQ_IPA93}} SIL International does a good job of describing how to convert from legacy IPA fonts to Unicode IPA.
@@
 
-For users of Python, see the standard documentation on how to use Unicode in your programming applications.\footnote{\url{https://docs.python.org/3/howto/unicode.html}} For R users we recommend the \textsc{stringi} library.\footnote{\url{https://cran.r-project.org/web/packages/stringi/index.html}} For \LaTeX users, the TIPA package is useful for inserting IPA characters into your typeset documents.
+For users of Python, see the standard documentation on how to use Unicode in your programming applications.\footnote{\url{https://docs.python.org/3/howto/unicode.html}} For R users we recommend the \textsc{stringi} library.\footnote{\url{https://cran.r-project.org/web/packages/stringi/index.html}} For \LaTeX~users the \textsc{TIPA} package is useful for inserting IPA characters into your typeset documents.
 See these resources:
@@
 
-\noindent But we find it much easier to use the Unicode-aware XeTeX/XeLaTeX typesetting system.\footnote{\url{http://xetex.sourceforge.net/}} Unicode characters can be directly inserted into your Tex documents and compiled into typeset PDF with XeLaTeX.
+\noindent But we find it much easier to use the Unicode-aware \hologo{XeTeX} typesetting system.\footnote{\url{http://xetex.sourceforge.net/}} Unicode characters can be directly inserted into your \hologo{TeX} documents and compiled into typeset PDF with \hologo{XeLaTeX}.
 

### 2018-06-08 book/chapters/unicode_approach.tex
deals with all paperhiver comments except a few time consuming ones

@@
 are a set of technical standards that describe the implementation of the Unicode
-Standard for software development, Web standards, and programming languages.
+Standard for software development, web standards, and programming languages.
 The \textsc{Unicode Character Database (UCD)} is a set of computer-readable text

### 2018-06-08 book/chapters/introduction.tex
deals with paperhivers through chapter 2
deals with all paperhiver comments

@@
 though we will argue that in practice this is often not the most useful solution for the kind of problems
-arising in the daily practice of many linguists. Consequently, a central goal of
-this book is to flesh out the linguistic challenges for locale descriptions,
-and work out suggestions to improve their structure for usage in a linguistic
-context. Conversely, we also aim to improve linguists' understanding and
-appreciation for the accomplishments of the Unicode Consortium in the
-development of the Unicode Standard.
+arising in the daily practice of many linguists.
+Consequently, a central goal of this book is to flesh out the linguistic-specific challenges regarding Unicode Locales and to work out suggestions to simplify their structure for usage in a linguistic context.
+Conversely, we also aim to improve linguists' understanding and appreciation for the accomplishments of the Unicode Consortium in the development of the Unicode Standard.
 
@@
 forcefully in the context of language comparison. Concretely, the proper
-processing of orthographies and transcription systems becomes crucial for the
+processing of orthographies and transcription systems becomes critical for the
 development of quantitative methods for language comparison and historical
-reconstruction, in order to investigate worldwide linguistic variation and to
+reconstruction. In order to investigate worldwide linguistic variation and to
 model the historical and areal processes that underlie linguistic
-diversity. In this work, it is crucial to be able to flexibly process numerous
+diversity, it is crucial that we are able to flexibly process numerous
 resources in different orthographies. In many cases even different resources
@@
 in a writing system. A second interpretation is used as a conventional term for a unit in the Chinese writing
-system \citep{Daniels1996}. In technological terminology, a \textsc{character} 
+system \citep{Daniels1996}. In technical terminology, a \textsc{character} 
 refers to the electronic encoding of a component in a writing system that has semantic 
 value (see Section \ref{character-encoding-system}). Thus in this work we must navigate 
-between the general linguistic and technological terms for \textsc{character} 
+between the general linguistic and technical terms for \textsc{character} 
 and \textsc{grapheme} because of how these notions are defined and how they relate at the intersection 
@@
 combination of the phonemes /k/ and /s/, as in the word `index' [ˈɪnˌdɛks]; other times 
-it is pronounced as /z/, such as in the words `Xerox' [ˈzɪrˌɑks] and `example' [ɪɡˈzæmpəl]. 
-As one can see, there can be non-trivial mappings between graphemes and phonemes in 
+it is pronounced as /z/, such as in the words `Xerox' [ˈzɪrˌɑks]; and in `example' [ɪɡˈzæmpəl] 
+it is a combination of /g/ and /s/. As one can see, there can be non-trivial mappings between graphemes and phonemes in 
 orthographies like English, e.g.\ `give', `gin', `jingle', where the graphemes 
@@
 systems has been purposely brief. We have highlighted some of the linguistic
-notions that are pertinent to, yet sometimes confused with, the technological
+notions that are pertinent to, yet sometimes confused with, the technical
 definitions developed for the computational processing of the world's writing

### 2018-06-08 book/chapters/orthography_profiles.tex
deals with all paperhiver comments

@@
 As an example of graphemic tokenization, note the three different levels of
-technological and linguistic elements that interact in the hypothetical lexical
+technical and linguistic elements that interact in the hypothetical lexical
 form <tsʰṍ̰shi>:

### 2018-06-08 book/chapters/practical_recommendations.tex
deals with all paperhiver comments

@@
 \section{For programmers and potential programmers}
-If you have made it this far, and you are eager to know more about the technological aspects of the Unicode Standard and how they relate to software programming, we recommend two light-hearted blog posts on the topic.
+If you have made it this far, and you are eager to know more about the technical aspects of the Unicode Standard and how they relate to software programming, we recommend two light-hearted blog posts on the topic.
 The classic blog post about what programmers should know about the Unicode Standard is Joel Spolsky's \textit{The Absolute Minimum Every Software Developer Absolutely, Positively Must Know About Unicode and Character Sets (No Excuses!)}.\footnote{\url{https://www.joelonsoftware.com/2003/10/08/the-absolute-minimum-every-software-developer-absolutely-positively-must-know-about-unicode-and-character-sets-no-excuses/}} A more recent blogpost, with a bit more of the technical details, is by David C.

### 2018-06-11 book/langsci/graphics/langsci-advertisement.tex
migrate to newer class

\thispagestyle{empty}
\pgfdeclarelayer{lspadvert_bg} % Please make sure to never use lspadvert_... PGF layers in any document
\pgfsetlayers{lspadvert_bg,main}

\begin{tikzpicture}[remember picture, overlay,bg/.style={outer sep=0}]
	\begin{pgfonlayer}{lspadvert_bg} % background layer
	\node at (current page.center) [bg, fill=lsGuidelinesGray, minimum height=22cm, minimum width=15cm] (lspadvert_bg1) {}; % Die können wir noch dynamisch bestimmen
	\end{pgfonlayer}	
	\node [execute at begin node={}, font=\fontsize{52pt}{16.75mm}\selectfont, below right = 10mm and 7.5mm of lspadvert_bg1.north west, text width=13.7cm, align=left] (lspadvert_covertitle) {\color{black}\lsCoverTitleFont{Did you like this book?\par}}; % x = 15mm - 7.5mm ; y = 17.5mm - 7.5mm

	\node [font=\fontsize{25pt}{10mm}\selectfont,below = 8mm of lspadvert_covertitle.south, text width=137mm] (lspadvert_coversubtitle) {\color{black} \lsCoverSubTitleFont \nohyphens{This book was brought to you for free}\par};

	\node (lspadvert_logo) [above left = 10mm and 7.5mm of lspadvert_bg1.south east] {\color{black}\includegraphics{\logopath langsci_logo_nocolor.pdf}};
	\node [above right = 48.8mm and -.1mm of lspadvert_bg1.south west, rectangle, fill=white, minimum size=17pt] (lspadvert_square) {};
	\path let \p1 = (lspadvert_square.north east), \p2 = (lspadvert_covertitle.west) in node at (\x2,\y1) (lspadvert_seriesinfo) [font=\fontsize{17pt}{7.5mm}\selectfont, right, align=flush left, text width=90mm, anchor=north west] {\color{black}\lsCoverSeriesFont{\nohyphens{Please help us in providing free access to linguistic research worldwide. Visit \url{http://www.langsci-press.org/donate} to provide financial support or register as a community proofreader or typesetter \mbox{at \url{http://www.langsci-press.org/register}.}}\par}};
\end{tikzpicture}

### 2018-06-11 book/langsci/seriesinfo/calseries-info.tex
migrate to newer class

%% -*- coding:utf-8 -*-
{\large Contemporary African Linguistics}

\bigskip

Editors: Akinbiyi Akinlabi,  Laura J. Downing

\bigskip

In this series:

\begin{enumerate}
\item Payne, Doris L., Sara Pacchiarotti \& Mokaya Bosire (eds.). Diversity in African languages: Selected papers from the 46th Annual Conference on African Linguistics.
\item Persohn, Bastian. The verb in Nyakyusa: A focus on tense, aspect and modality.
\item Kandybowicz, Jason, Travis Major \& Harold Torrence (eds.). African linguistics on the prairie. Selected papers from the 45th Annual Conference on African Linguistics.
\end{enumerate}

\vfill

% \gdef\lsISSN{XXXX-YYYY}%
% \hfill ISSN: \lsISSN

### 2018-06-11 book/langsci/seriesinfo/cfls-info.tex
migrate to newer class

--- 2015-09-29 book/LSP/lsp-seriesinfo/cfls-info.tex
@@
 \textbf{Series editors}\\
-Mark Dingemanse \textit{Max Planck Institute for Psycholinguistics}   \\
-Nick J. Enfield \textit{University of Sydney}
+Mark Dingemanse, \textit{Max Planck Institute for Psycholinguistics}   \\
+N. J. Enfield, \textit{University of Sydney}
 
@@
 
-\textbf{Editorial board} \\
-Balthasar Bickel \textit{University of Zürich}   \\
-Claire Bowern \textit{Yale University}          \\
-Elizabeth Couper-Kuhlen \textit{University of Helsinki}    \\
-William Croft \textit{University of New Mexico}            \\
-Rose-Marie Déchaine \textit{University of British Columbia}      \\
-William A. Foley \textit{University of Sydney}                   \\
-William F. Hanks \textit{University of California at Berkeley}   \\
-Paul Kockelman \textit{Yale University}           \\
-Keren Rice \textit{University of Toronto}         \\
-Sharon Rose \textit{University of California at San Diego }       \\
-Frederick J. Newmeyer \textit{University of Washington}        \\
-Wendy Sandler \textit{University of Haifa }            \\
-Dan Sperber \textit{Jean Nicod Institute}     \\
+\textbf{Editorial board}\\
+Balthasar Bickel, \textit{University of Zürich},
+Claire Bowern, \textit{Yale University},
+Elizabeth Couper-Kuhlen, \textit{University of Helsinki},
+William Croft, \textit{University of New Mexico},
+Rose-Marie Déchaine, \textit{University of British Columbia},
+William A. Foley, \textit{University of Sydney} ,
+William F. Hanks, \textit{University of California at Berkeley},
+Paul Kockelman, \textit{Yale University},
+Keren Rice, \textit{University of Toronto},
+Sharon Rose, \textit{University of California at San Diego},
+Frederick J. Newmeyer, \textit{University of Washington},
+Wendy Sandler, \textit{University of Haifa},
+Dan Sperber \textit{Central European University}
 
+\bigskip
 \bigskip
@@
 Only with clarity about conceptual foundations can we pose coherent research questions, design critical experiments, and collect crucial data.
-
-\medskip
-
+%
 This series publishes short and accessible books that explore well-defined topics in the conceptual foundations of language science.
@@
 \bigskip
+\bigskip
 
@@
 \begin{enumerate}
-\item Nick J. Enfield. \textit{Natural Causes of Language}.
+\item N. J. Enfield. \textit{Natural causes of language}.
 
 \end{enumerate}
-
-% 
-% \vfill
-% 
-% \gdef\lsISSN{2363-5568}%
-% \hfill ISSN: \lsISSN

### 2018-06-11 book/langsci/seriesinfo/classics-info.tex
migrate to newer class

%% -*- coding:utf-8 -*-
{\large Classics in Linguistics}

\bigskip

Chief Editors: Martin Haspelmath, Stefan Müller

\bigskip

In this series:

\begin{enumerate}
\item Lehmann, Christian. Thoughts on grammaticalization.
\item Schütze, Carson T. The empirical base of linguistics: Grammaticality judgments and linguistic methodology.
\item Bickerton, Derek. Roots of language.
\item von der Gabelentz, Georg. Die Sprachwissenschaft:  Ihre Aufgaben, Methoden und bisherigen Ergebnisse.
\item Stefan Müller, Marga Reis \& Frank Richter (Hrsg): Beiträge zur deutschen Grammatik: Gesammelte Schriften von Tilman N.\ Höhle.
\end{enumerate}

\vfill

% \gdef\lsISSN{XXXX-YYYY}%
% \hfill ISSN: \lsISSN

### 2018-06-11 book/langsci/seriesinfo/eotms-info.tex
migrate to newer class

%% -*- coding:utf-8 -*-
{\large Em­pir­i­cal­ly Ori­ent­ed The­o­ret­i­cal Mor­phol­o­gy and Syn­tax}

\bigskip

Chief Editor: Stefan Müller \\
Consulting Editors: Berthold Crysmann, Laura Kallmeyer

\bigskip

In this series:

\begin{enumerate}
\item Lichte, Timm. Syntax und Valenz: Zur Modellierung kohärenter und elliptischer Strukturen mit Baumadjunktionsgrammatiken 
\item Bîlbîie, Gabriela. Grammaire des constructions elliptiques: Une étude comparative des phrases sans verbe en roumain et en français 
\item Bowern, Claire, Laurence Horn \& Raffaella Zanuttini (eds.). On looking into words (and beyond): Structures, Relations, Analyses

\end{enumerate}

### 2018-06-11 book/langsci/seriesinfo/ogs-info.tex
migrate to newer class

%% -*- coding:utf-8 -*-
{\large Open Generative Syntax}

\bigskip

Editors:   Elena Anagnostopoulou,
    Mark Baker,
    Roberta D’Alessandro,
    David Pesetsky,
    Susi Wurmbrand

\bigskip

In this series:

\begin{enumerate}
\item Bailey, Laura R. \& Michelle Sheehan (eds.). Order and structure in syntax I: Word order and syntactic structure.
\item Sheehan, Michelle \& Laura R. Bailey (eds.).  Order and structure in syntax II: Subjecthood and argument structure.
\item Bacskai-Atkari, Julia. Deletion phenomena in comparative constructions: English comparatives in a cross-linguistic perspective.
\end{enumerate}

\vfill

% \gdef\lsISSN{XXXX-YYYY}%
% \hfill ISSN: \lsISSN

### 2018-06-11 book/langsci/seriesinfo/sidl-info.tex
migrate to newer class

%% -*- coding:utf-8 -*-
{\large Studies in Diversity Linguistics}

\bigskip

Chief Editor: Martin Haspelmath \\
Consulting Editors: Fernando Zúñiga, Peter Arkadiev, Ruth Singer, Pilar Valen­zuela

\bigskip

In this series:

\begin{enumerate}
\item Handschuh, Corinna. A typology of marked-S languages.
\item Rießler, Michael. Adjective attribution.
\item Klamer, Marian (ed.). The Alor-Pantar languages: History and typology.
\item Berghäll, Liisa. A grammar of Mauwake (Papua New Guinea).
\item Wilbur, Joshua. A grammar of Pite Saami.
\item Dahl, Östen. Grammaticalization in the North: Noun phrase morphosyntax in Scandinavian vernaculars.
\item Schackow, Diana.    A grammar of Yakkha.
\item Liljegren, Henrik. A grammar of Palula.
\item Shimelman, Aviva. A grammar of Yauyos Quechua.
\item Rudin, Catherine \& Bryan James Gordon (eds.). Advances in the study of Siouan languages and linguistics.
\item Kluge, Angela. A grammar of Papuan Malay.
\item Kieviet, Paulus. A grammar of Rapa Nui.
\item Michaud, Alexis. Tone in Yongning Na: Lexical tones and morphotonology.
\item Enfield, N.\,J (ed.).  Dependencies in language: On the causal ontology of linguistic systems .
\item Gutman, Ariel. Attributive constructions in North-Eastern Neo-Aramaic.
\item Bisang, Walter \& Andrej Malchukov (eds.). Unity and diversity in grammaticalization scenarios.
\item Stenzel, Kristine \& Patrizia Paggio (eds.). On this and other worlds: Voices from Amazonia.
\item Paggio, Patrizia and Albert Gatt (eds.). The languages of Malta.
\item Ilja A. Seržant \& Alena Witzlack-Makarevich (eds.). The diachronic typology of differential argument marking.

% \item Yakpo 
% \item Aguilar
% \item Riesberg
% \item di Garbo
% \item Floyd
% \item Haspelmath
% \item Veselinova
\end{enumerate}

### 2018-06-11 book/langsci/seriesinfo/silp-info.tex
migrate to newer class

--- 2015-05-29 LSP/lsp-seriesinfo/silp-info.tex
@@
 
-{\raggedleft\IfFileExists{./LSP/lsp-seriesinfo/labphon-logo.pdf}{\includegraphics[width=5cm]{./LSP/lsp-seriesinfo/labphon-logo.pdf}}{LabPhon-Logo}}
+{\raggedleft\IfFileExists{./langsci/seriesinfo/labphon-logo.pdf}{\includegraphics[width=5cm]{./langsci/seriesinfo/labphon-logo.pdf}}{LabPhon-Logo}}
 
@@
 \begin{enumerate}
-\item Cangemi, Francesco. Prosodic detail in Neapolitan Italian
+\item Cangemi, Francesco. Prosodic detail in Neapolitan Italian.
+\item Drager, Katie. linguistic variation, identity construction, and cognition.
 \end{enumerate}
-
-\vfill
-
-\gdef\lsISSN{2363-5576}%
-\hfill ISSN: \lsISSN

### 2018-06-11 book/langsci/seriesinfo/tbls-info.tex
migrate to newer class

%% -*- coding:utf-8 -*-
{\large Textbooks in Language Sciences}

\bigskip

Editors: Stefan Müller, Martin Haspelmath  \\
Editorial Board: Claude Hagège, Marianne Mithun, Anatol Stefanowitsch, Foong Ha Yap

\bigskip

In this series:

\begin{enumerate}
\item Müller, Stefan. Grammatical theory: From transformational grammar to constraint-based approaches.
\item Schäfer, Roland. Einführung in die grammatische Beschreibung des Deutschen.
\item Freitas, Maria João \& Ana Lúcia Santos (eds.). Aquisição de língua materna e não materna: Questões gerais e dados do português.
\item Roussarie, Laurent . Sémantique formelle : Introduction à la grammaire de Montague.
\item Kroeger, Paul. Analyzing meaning: An introduction to semantics and pragmatics
\end{enumerate}

### 2018-06-11 book/langsci/seriesinfo/tmnlp-info.tex
migrate to newer class

%% -*- coding:utf-8 -*-
{\large Translation and Multilingual Natural Language Processing}

\bigskip

Editors: Oliver Czulo (Universität Leipzig),
Silvia Hansen-Schirra (Johannes Gutenberg-Universität Mainz),
Stella Neumann (RWTH Aachen),
Reinhard Rapp (Johannes Gutenberg-Universität Mainz)

\bigskip

In this series:

\begin{enumerate}
\item Fantinuoli, Claudio \& Federico Zanettin (eds.). New directions in corpus-based translation studies.
\item Hansen-Schirra, Silvia \& Sambor Grucza (eds.). Eyetracking and Applied Linguistics.
\item Neumann, Stella, Oliver Čulo \& Silvia Hansen-Schirra (eds.). Annotation, exploitation and evaluation of parallel corpora: TC3\,I.
\item Czulo, Oliver  \& Silvia Hansen-Schirra (eds.). Crossroads between Contrastive Linguistics, Translation Studies and Machine Translation: TC3\,II.
\item Rehm, Georg, Felix Sasaki, Daniel Stein \& Andreas Witt (eds.). Language technologies for a multilingual Europe: TC3\,III.
\item Menzel, Katrin, Ekaterina Lapshinova-Koltunski \& Kerstin Anna Kunz  (eds.). New perspectives on cohesion and coherence: Implications for translation.
\item Hansen-Schirra, Silvia, Oliver Czulo \& Sascha Hofmann (eds). Empirical modelling of translation and interpreting.
\item Svoboda, Tomáš, Łucja Biel \& Krzysztof Łoboda (eds.). Quality aspects in institutional translation.
% \item SMMC
% \item LVDV
% \item WF
\end{enumerate}

### 2018-06-11 book/chapters/unicode_approach.tex
adds backslashes after eg

@@
 that define a set of language-specific elements (e.g.~tailored grapheme
-clusters, collation order, capitalization-equivalence), as well as other special
+clusters, collation order, capitalization equivalence), as well as other special
 information, like how to format numbers, dates, or currencies. Locale

### 2018-06-11 book/chapters/orthography_profiles.tex
updates
closes #33

@@
 	   Data}.\footnote{\url{https://www.w3.org/TR/tabular-metadata/}} This metadata 
-	   format allows easy inclusion of Dublin Core metadata,\footnote{\url{http://dublincore.org/}} 
+	   format allows for easy inclusion of Dublin Core metadata,\footnote{\url{http://dublincore.org/}} 
 	   which should be used to specify information about the orthographic description 
@@
 
-\subsection*{Implementation}
+\subsection*{Processing}
 For the automated processing of the profiles, the following technical standards
@@
 	\def\labelenumi{B\arabic{enumi}.} 
-	\item \textsc{Each line of a profile will be interpreted as a regular
-       expression.} Software applications using profiles can also offer to
-       interpret a profile in the literal sense to avoid the necessity for the
-       user to escape regular expression metacharacters in the profile.
-       However, this is only possible when no contexts or classes are described,
-       so this seems only useful in the most basic orthographies. 
+	\item \textsc{Each line of a profile will be interpreted according to the content type of the column as specified in the profile metadata}. Content types include literal and regular expression.
 	\item \textsc{The \textsc{class} column will be used to produce explicit

### 2018-06-11 book/chapters/orthography_profiles.tex
closes #31

@@
 
-	\item \textsc{Metadata should be added in a separate UTF-8 text file} using 
+	\item \textsc{Metadata should be added in a separate utf-\scalebox{0.8}{8} text file} using 
 	   the JSON-LD dialect specified in \textit{Metadata Vocabulary for Tabular 
@@
 	   including: (vi) its dialect description,\footnote{\url{http://w3c.github.io/csvw/metadata/#dfn-dialect-descriptions}} 
-	   (vii) the tokenization method (see below), and (ix) proper column           
-	   descriptions,\footnote{\url{http://w3c.github.io/csvw/metadata/#dfn-datatype-description}} 
+	   and (vii) proper column descriptions,\footnote{\url{http://w3c.github.io/csvw/metadata/#dfn-datatype-description}} 
 	   which describe how a column should be interpreted and processed (e.g.\ whether they 

### 2018-06-11 book/localmetadata.tex
ready for review two
migrate to newer class
add some of the proofreaders

@@
 \BackBody{Change backbody in  localmetadata.tex}
-\dedication{Change dedication in localmetadata.tex}
+% \dedication{Change dedication in localmetadata.tex}
 \typesetter{Change typesetter in localmetadata.tex}
-\proofreader{Change proofreaders in localmetadata.tex}
-\author{Steven Moran \& Michael Cysouw}
-\renewcommand{\lsISBN}{978--3--000000--00--0}                     
-\renewcommand{\lsSeries}{dummyseries} 
-% use lowercase acronym, e.g. sidl, eotms, tgdi
-\renewcommand{\lsSeriesNumber}{99} 
-%will be assigned when the book enters the proofreading stage
-\renewcommand{\lsURL}{http://langsci-press.org/catalog/book/0} 
-% contact the coordinator for the right number
+\proofreader{Sandra Auderset, Aleksandrs Berdičevskis, Rosey Billington, Simon Cozens, Aniefon Daniel, Bev Erasmus, Amir Ghorbanpour, David Lukeš, Hugh Paterson III, Stephen Pepper, Katja Politt, Felix Rau, James Tauber, Luigi Talamo, Jeroen van de Weijer, Viola Wiegand, Alena Witzlack-Makarevich, and Esther Yap}
+\author{Steven Moran \and Michael Cysouw}
+\renewcommand{\lsISBNdigital}{978-3-96110-090-3 }                     
+\renewcommand{\lsISBNhardcover}{978-3-96110-091-0}                     
+\renewcommand{\lsSeries}{tmnlp}  
+\renewcommand{\lsSeriesNumber}{10}  
+\renewcommand{\lsID}{176}

### 2018-06-12 book/localmetadata.tex
adds blurb on back and proofreaders

@@
 \subtitle{Managing writing systems using orthography profiles}
-\BackTitle{Change backtitle in  localmetadata.tex}
-\BackBody{Change backbody in  localmetadata.tex}
+% \BackTitle{Change backtitle in  localmetadata.tex}
+\BackBody{This text is a practical guide for linguists, and programmers, who work with data in multilingual computational environments.
+We introduce the basic concepts needed to understand how writing systems and character encodings function, and how they work together at the intersection between the Unicode Standard and the International Phonetic Alphabet.
+Although these standards are often met with frustration by users, they nevertheless provide language researchers and programmers with a consistent computational architecture needed to process, publish and analyze lexical data from the world's languages.
+Thus we bring to light common, but not always transparent, pitfalls which researchers face when working with Unicode and IPA.
+Having identified and overcome these pitfalls involved in making writing systems and character encodings syntactically and semantically interoperable (to the extent that they can be), we created a suite of open-source Python and R tools to work with languages using orthography profiles that describe author- or document-specific orthographic conventions.
+In this cookbook we describe a formal specification of orthography profiles and provide recipes using open source tools to show how users can segment text, analyze it, identify errors, and to transform it into different written forms for comparative linguistics research.}
 % \dedication{Change dedication in localmetadata.tex}
-\typesetter{Change typesetter in localmetadata.tex}
-\proofreader{Sandra Auderset, Aleksandrs Berdičevskis, Rosey Billington, Simon Cozens, Aniefon Daniel, Bev Erasmus, Amir Ghorbanpour, David Lukeš, Hugh Paterson III, Stephen Pepper, Katja Politt, Felix Rau, James Tauber, Luigi Talamo, Jeroen van de Weijer, Viola Wiegand, Alena Witzlack-Makarevich, and Esther Yap}
+\typesetter{Steven Moran, Michael Cysouw}
+\proofreader{Sandra Auderset, Aleksandrs Berdičevskis, Rosey Billington, Varun deCastro-Arrazola, Simon Cozens, Aniefon Daniel, Bev Erasmus, Amir Ghorbanpour, Linda Leembruggen, David Lukeš, Antonio Machicao y Priemer, Hugh Paterson III, Stephen Pepper, Katja Politt, Felix Rau, Lea Schäfer, Benedikt Singpiel, James Tauber, Luigi Talamo, Jeroen van de Weijer, Viola Wiegand, Alena Witzlack-Makarevich, and Esther Yap}
 \author{Steven Moran \and Michael Cysouw}

### 2018-06-17 book/chapters/orthography_profiles.tex
closes #30

@@
 
-At this point in the course of rapid ongoing developments, we are left with a
-situation in which the Unicode Standard offers a highly detailed and flexible
-approach to deal computationally with writing systems, but it has unfortunately
-not influenced the linguistic practice very much. In many practical situations,
-the Unicode Standard is far too complex for the day-to-day practice in
-linguistics because it does not offer practical solutions for the down-to-earth
-problems of many linguists. In this section, we propose some simple practical
-guidelines and methods to improve on this situation.
+The Unicode Standard offers a very detailed technical approach 
+for characterizing writing systems computationally. As such, it is 
+sometimes too complex for the day-to-day practice of many linguists, 
+as exemplified by the need to understand the common pitfalls that we discussed in Chapters 
+\ref{unicode-pitfalls} \& \ref{ipa-meets-unicode}. Therefore, in this section we propose some simple 
+guidelines for linguists working in multilingual environments.
 
-Our central aims for linguistics, to be approached with a Unicode-based
-solution, are: (i) to improve the consistency of the encoding of sources, (ii)
+Our aims for adopting a Unicode-based
+solution are: (i) to improve the consistency of the encoding of sources, (ii)
 to transparently document knowledge about the writing system (including
@@
 Unicode Locales in the daily linguistic practice, namely: (i) the
-XML structure is too verbose to easily and quickly produce or correct manually,
+XML structure\footnote{\url{http://unicode.org/reports/tr35/}} is too verbose to easily and quickly produce or correct manually,
 (ii) Unicode Locales are designed for a wide scope of information (like date
@@
 transliteration as described in the Unicode Locale Data Markup
-Language).\footnote{See \url{http://www.unicode.org/reports/tr35/}} Finite-state 
+Language).\footnote{\url{http://www.unicode.org/reports/tr35/}} Finite-state 
 transducers attempt to describe the mapping from input to output string directly 
@@
 refined, while at the same time inconsistencies and errors in the data to be
-tokenized are corrected. In some more complex use-cases there might even be a
+tokenized are corrected. In some more complex use cases there might even be a
 need for multiple different orthography profiles to be applied in sequence (see
-Sections~\ref{python-implementations} \& \ref{r-implementation} on various exemplary use-cases). The result of any such
+Sections~\ref{python-implementations} \& \ref{r-implementation} on various exemplary use cases). The result of any such
 workflow will normally be a cleaned dataset and an explicit description of the
@@
 	\item \textsc{A profile is a unicode utf-\scalebox{0.8}{8} encoded text file} 
-	   normalized following NFC (or NFD if specified in the metadata)
-       that includes the information pertinent to the orthography.\footnote{See 
+	   % normalized following NFC (or NFD if specified in the metadata)
+       that includes information pertinent to the orthography.\footnote{See 
 	   Section~\ref{pitfall-file-formats} in which we suggest to use NFC, 
@@
 	\item \textsc{The columns \texttt{Grapheme}, \texttt{Left} and \texttt{Right}
-       can use regular expression metacharacters.} If regular expressions are
+       can use regular expression \\ metacharacters.} If regular expressions are
        used, then they must be specified in the metadata file as such, 
@@
        specific transliteration. 
-	\item \textsc{Any other columns can be added freely, but will mostly be ignored
+	\item \textsc{Any other columns can be added freely, but will be typically ignored
        by any software application using the profiles}. As orthography profiles
        are also intended to be read and interpreted by humans, it is often
-       highly useful to add extra information on the graphemes in further
+       very useful to add extra information about the graphemes in further
        columns, such as Unicode code points, Unicode names, frequency of
@@
        expression like:
-       \texttt{(a\textbar{}e\textbar{}i\textbar{}o\textbar{}u\textbar{}ei\textbar{}au)}.
+       \texttt{(au\textbar{}ei\textbar{}a\textbar{}e\textbar{}i\textbar{}o\textbar{}u}).
        Only the graphemes themselves are included here, not any contexts
-       specified for the elements of the class. Note that in some cases the 
-       ordering inside this regular expression might be crucial, therefore it must be 
-	   specified as such in the metadata file.
+       specified for the elements of the class. Note that the 
+       ordering inside this regular expression is crucial (e.g.\ regular expressions are greedy, so
+	   longest matches should be placed before matching substrings).
 	\item \textsc{The \textsc{left} and \textsc{right} contexts will be included
@@
        \texttt{Left} and \texttt{Right} are combined with the column
-       \texttt{Grapheme} to form a complex regular expression like:\ 
+       \texttt{Grapheme} to form a complex regular expression like:\\ 
        \texttt{(?\textless{}=Left)Grapheme(?=Right)}. 
@@
        the metadata as \textsc{tokenization method}. 
-	\item \textsc{The matching of the graphemes can occur either in NFC or NFD.} By
-       default, both the profile and the text-string to be tokenized should be
-       treated as NFC (see Section~\ref{pitfall-canonical-equivalence}). However, in some use-cases it turns out to
+	\item \textsc{The matching of the graphemes can occur either in nfc or nfd.} 
+	   The Unicode Standard states that software is free to compose or decompose the character stream from one 
+	   representation to another. However, Unicode conformant software must treat canonically equivalent sequences in 
+	   NFC and NFD as the same. It is up to the orthography profile creator how they choose to encode their profile. 
+	   Several sources suggest to use NFC when possible for text encoding,\footnote{\url{http://www.win.tue.nl/~aeb/linux/uc/nfc_vs_nfd.html}} 
+	   including SIL International with regard to data archiving.\footnote{\url{http://scripts.sil.org/cms/scripts/page.php?item_id=NFC_vs_NFD}} 
+	   In our experience, in some use cases it turns out to
        be practical to treat both text and profile as NFD. This typically
-       happens when very many different combinations of diacritics occur in the
+       happens when many different combinations of diacritics occur in the
        data. An NFD profile can then be used to first check which individual
@@
        orthography profiles should offer both approaches (i.e.\ \textsc{NFC} or
-       \textsc{NFD}) to the user. The approach used should be documented in the
+       \textsc{NFD}) to the user. The approach used can be documented in the
        metadata as \textsc{unicode normalization}. 
@@
         about escaping metacharacters. However, in our experience all actually
-        interesting use-cases of orthography profiles include some contexts,
+        interesting use cases of orthography profiles include some contexts,
         which automatically prevents any literal interpretation.

### 2018-06-18 book/chapters/ipa_meets_unicode.tex
numerous edits

@@
 the diacritics are distinguished by their position relative to the base
-character.\footnote{See
-\url{http://unicode.org/reports/tr44/\#Canonical\_Combining\_Class\_Values} for
-a detailed description.} When applying a Unicode normalization (NFC or NFD, see
+character.\footnote{For a detailed description, see: 
+\url{http://unicode.org/reports/tr44/\#Canonical\_Combining\_Class\_Values}.} 
+When applying a Unicode normalization (NFC or NFD, see
 Section~\ref{pitfall-canonical-equivalence}), the diacritics in different
@@
 	\item The Americanist tradition of using using <\dia{030C}>, a
-       \textsc{combining caron} at \uni{030C} for all postalveolar sounds, like
+       \textsc{combining caron}\\ at \uni{030C} for all postalveolar sounds, like
        in <š~ž~č~ǰ>, was not adopted because the Association members at the
@@
       using the voiceless glyph for the plain phoneme, and then 
-      <\dia{0348}> \textsc{combining double vertical line below} at
+      <\dia{0348}> \textsc{combining double vertical line below} at \\
       \uni{0348} to mark the fortis articulation, and/or <\dia{0349}>
@@
       <\dia{0347}> \textsc{combining equals sign below} at \uni{0347} to mark
-      the non-sibilant phoneme.
+      the non-sibilant \\ phoneme.
 \item \textsc{Glottalization} \newline 

### 2018-06-18 book/chapters/orthography_profiles.tex
numerous edits

@@
 <\dia{0303}> + <\dia{0330}> or <o> + <\dia{0330}> + <\dia{0303}>. Although these two variants are visually homoglyphs,
-computationally they are different (see Sections \ref{pitfall-ipa-homoglyphs} and \ref{pitfall-homoglyphs-in-IPA}). 
+computationally they are different (see Sections \ref{pitfall-ipa-homoglyphs} \& \ref{pitfall-homoglyphs-in-IPA}). 
 Unicode normalization should be applied to

### 2018-06-18 book/chapters/practical_recommendations.tex
remove extra curly
numerous edits

@@
 
-This chapter is meant to be a short guide for novice users who are not interested in the programmatic aspects presented in Chapters \ref{orthography-profiles} and \ref{implementation}.
+This chapter is meant to be a short guide for novice users who are not interested in the programmatic aspects presented in Chapters \ref{orthography-profiles} \& \ref{implementation}.
 Instead, we provide links to quickly find general information about the Unicode Standard and the International Phonetic Alphabet (IPA).
@@
 \begin{itemize}
-	\item \url{https://software.sil.org/doulos/}}
+	\item \url{https://software.sil.org/doulos/}
 	\item \url{https://software.sil.org/charis/}

### 2018-06-18 book/chapters/preface.tex
numerous edits

--- 2017-02-02 book/chapters/preface.tex
@@
 
-This text is meant as a practical guide for linguists, and programmers, who work with data in multilingual computational environments.
+This text is meant as a practical guide for linguists and programmers, who work with data in multilingual computational environments.
 We introduce the basic concepts needed to understand how writing systems and character encodings function, and how they work together.
 
-The intersection of the Unicode Standard and the International Phonetic Alphabet is often not met without frustration by users.
+The intersection of the Unicode Standard and the International Phonetic Alphabet is often met with frustration by users.
 Nevertheless, the two standards have provided language researchers with a consistent computational architecture needed to process, publish and analyze data from many different languages.
@@
 
-Our research uses quantitative methods to compare languages and uncover and clarify their phylogenetic relations.
+We use quantitative methods in our research to compare languages to uncover and clarify their phylogenetic relationships.
 However, the majority of lexical data available from the world's languages is in author- or document-specific orthographies.
-Having identified and overcome the pitfalls involved in making writing systems and character encodings syntactically and semantically interoperable (to the extent that they can be), we created a suite of open-source Python and R tools to work with languages using profiles that adequately describe their orthographic conventions.
-Using orthography profiles and these tools allows users to segment text, analyze it, identify errors, and to transform it into different written forms.
+Having identified and overcome the pitfalls involved in making writing systems and character encodings syntactically and semantically interoperable (to the extent that they can be), we created a suite of open-source Python and R software packages to work with languages using profiles that adequately describe their orthographic conventions.
+Using orthography profiles and these tools allows users to tokenize and transliterate text from diverse sources, so that they can be meaningfully compared and analyzed.
 
-We welcome comments and corrections of this text, our source code, and the use case studies that are supplement to this book. Please use the issue tracker or email us directly.
+We welcome comments and corrections regarding this book, our source code, and the supplemental case studies that we provide online.\footnote{\url{https://github.com/unicode-cookbook/}} Please use the issue tracker, email us directly, or make suggestions on PaperHive.\footnote{\url{https://paperhive.org/}}
 

### 2018-06-18 book/chapters/unicode_approach.tex
numerous edits

@@
 character database.\footnote{All documents of the Unicode Standard are available
-at \url{http://www.unicode.org/versions/latest/}. For a quick survey of the use
+at: \url{http://www.unicode.org/versions/latest/}. For a quick survey of the use
 of terminology inside the Unicode Standard, their glossary is particularly
-useful, available at \url{http://www.unicode.org/glossary/}. For a general
+useful, available at: \url{http://www.unicode.org/glossary/}. For a general
 introduction to the principles of Unicode, Chapter 2 of the core specification,
@@
 described in \url{http://www.unicode.org/reports/tr44/}. A simplified overview
-of the properties is available at
+of the properties is available at: 
 \url{http://userguide.icu-project.org/strings/properties}. The actual code
-tables listing all properties for all Unicode code points are available at
+tables listing all properties for all Unicode code points are available at: 
 \url{http://www.unicode.org/Public/UCD/latest/ucd/}.} Basically, those

### 2018-06-21 book/chapters/orthography_profiles.tex
fixes urls

@@
 	   which should be used to specify information about the orthographic description 
-	   in the orthography profile.\footnote{\url{http://w3c.github.io/csvw/metadata/#dfn-common-property}} 
+	   in the orthography profile.\footnote{\url{http://w3c.github.io/csvw/metadata/\#dfn-common-property}} 
 	   The orthography profile metadata should minimally include provenance information including: 
@@
 	   Further, the metadata file should provide information about the orthography profile's structure and contents, 
-	   including: (vi) its dialect description,\footnote{\url{http://w3c.github.io/csvw/metadata/#dfn-dialect-descriptions}} 
-	   and (vii) proper column descriptions,\footnote{\url{http://w3c.github.io/csvw/metadata/#dfn-datatype-description}} 
+	   including: (vi) its dialect description,\footnote{\url{http://w3c.github.io/csvw/metadata/\#dfn-dialect-descriptions}} 
+	   and (vii) proper column descriptions,\footnote{\url{http://w3c.github.io/csvw/metadata/\#dfn-datatype-description}} 
 	   which describe how a column should be interpreted and processed (e.g.\ whether they 

### 2018-06-23 book/chapters/preface.tex
ready for release, except doi

@@
 
-This text is meant as a practical guide for linguists and programmers, who work with data in multilingual computational environments.
+This text is meant as a practical guide for linguists and programmers who work with data in multilingual computational environments.
 We introduce the basic concepts needed to understand how writing systems and character encodings function, and how they work together.
@@
 The intersection of the Unicode Standard and the International Phonetic Alphabet is often met with frustration by users.
-Nevertheless, the two standards have provided language researchers with a consistent computational architecture needed to process, publish and analyze data from many different languages.
+Nevertheless, the two standards have provided language researchers with the computational architecture needed to process, publish and analyze data from many different languages.
 We bring to light common, but not always transparent, pitfalls that researchers face when working with Unicode and IPA.
 
-We use quantitative methods in our research to compare languages to uncover and clarify their phylogenetic relationships.
+In our research, we use quantitative methods to compare languages to uncover and clarify their phylogenetic relationships.
 However, the majority of lexical data available from the world's languages is in author- or document-specific orthographies.
-Having identified and overcome the pitfalls involved in making writing systems and character encodings syntactically and semantically interoperable (to the extent that they can be), we created a suite of open-source Python and R software packages to work with languages using profiles that adequately describe their orthographic conventions.
-Using orthography profiles and these tools allows users to tokenize and transliterate text from diverse sources, so that they can be meaningfully compared and analyzed.
+Having identified and overcome the pitfalls involved in making writing systems and character encodings syntactically and semantically interoperable (to the extent that they can be), we have created a suite of open-source Python and R software packages to work with languages using profiles that adequately describe their orthographic conventions.
+Using these tools in combination with orthography profiles allows users to tokenize and transliterate text from diverse sources, so that they can be meaningfully compared and analyzed.
 

### 2018-06-23 book/chapters/unicode_approach.tex
ready for release, except doi

@@
 Each version of the Unicode Standard (\citealp{Unicode2018}, as of writing at
-version 11) consists of a set of specifications and guidelines that include (i) a
+version 11.0.0) consists of a set of specifications and guidelines that include (i) a
 core specification, (ii) code charts, (iii) standard annexes and (iv) a
@@
 <c>~+~<h> for the Slovak digraph <ch> and the sequence <ky> in the Sisaala
-practical orthography (pronounced as IPA /tʃ/,~\cite{Moran2006}). These grapheme
+practical orthography, which is pronounced as IPA /tʃ/ \citep{Moran2006}. These grapheme
 clusters are \textsc{tailored} in the sense that they must be specified on a

### 2018-06-29 book/langsci/seriesinfo/tmnlp-info.tex
adds latest tmnlp series
nordhoff requested changes

@@
 Silvia Hansen-Schirra (Johannes Gutenberg-Universität Mainz),
-Stella Neumann (RWTH Aachen),
 Reinhard Rapp (Johannes Gutenberg-Universität Mainz)
@@
 \item Svoboda, Tomáš, Łucja Biel \& Krzysztof Łoboda (eds.). Quality aspects in institutional translation.
+\item Fox, Wendy. Can integrated titles improve the viewing experience? Investigating the impact of subtitling on the reception and enjoyment of film using eye tracking and questionnaire data.
+\item Moran, Steven \& Michael Cysouw. The Unicode cookbook for linguists: Managing writing systems using orthography profiles
 % \item SMMC

### 2018-07-30 book/chapters/introduction.tex
minor edits in introduction

@@
 computers, the American Standards Association decided to propose a new 7-bit
-encoding in 1963 (with $2^7 = 128$ different possible characters), known as the
+encoding in 1963 (with 2\textsuperscript{7} = 128 different possible characters), 
+known as the
 \textsc{American Standard Code for Information Interchange} (ASCII), geared
@@
 Digital Equipment Corporation to introduce an 8-bit \textsc{Multinational
-Character Set} (MCS, with $2^8 = 256$ different possible characters), first used
-with the introduction of the VT{\large 220} Terminal in 1983.
+Character Set} (MCS, with 2\textsuperscript{8} = 256 different possible characters), first used with the introduction of the VT{\large 220} Terminal in 1983.
 
@@
 formed the \textsc{unicode consortium}, proposing a 16-bit encoding in 1991
-(with $2^{16} = 65,536$ different possible characters). At the same time in
+(with 2\textsuperscript{16} = 65,536 different possible characters). At the same time in
 Europe, the \textsc{international organization for standardization} (ISO) was
@@
 the \textsc{universal character set} (UCS) in 1990 was 31-bit (with
-theoretically $2^{31} = 2,147,483,648$ possible characters, but because of some
+theoretically 2\textsuperscript{31} = 2,147,483,648 possible characters, but because of some
 technical restrictions only 679,477,248 were allowed). Since 1991, the Unicode
@@
 Unicode proposal as the \textsc{basic multilingual plane}, and 16 additional
-planes of 16-bit for further extensions (with in total $(1+16) \cdot 2^{16} =
-1,114,112$ possible characters). The most recent version of the Unicode Standard
+planes of 16-bit for further extensions (with in total (1+16)\times 2\textsuperscript{16} =
+1,114,112 possible characters). The most recent version of the Unicode Standard
 (currently at version number 11.0.0) was published in June 2018 and it defines

### 2018-07-30 book/chapters/introduction.tex
small caps in introduction

@@
 system-specific minimally distinctive symbols (see below). Graphemes may consist of one or more 
-\textsc{characters}. The term \textsc{character} is overladen. In the linguistic terminology of writing
-systems, a \textsc{character} is a general term for any self-contained element
+\textsc{characters}. The term `character' is overladen. In the linguistic terminology of writing
+systems, a character is a general term for any self-contained element
 in a writing system. A second interpretation is used as a conventional term for a unit in the Chinese writing
-system \citep{Daniels1996}. In technical terminology, a \textsc{character} 
+system \citep{Daniels1996}. In technical terminology, a character}
 refers to the electronic encoding of a component in a writing system that has semantic 

### 2018-07-30 book/chapters/unicode_approach.tex
minor edits to chapter 2 unicode approach

@@
 same character, i.e.~they may be rendered differently depending on the
-typography being used, but they all share the same code point. From the
+typography being used, but they all share the same code point.\footnote{The first three g's 
+are regular, italic and bold versions of Linux Libertine, the main font used in this book, 
+while the last three are regular, italic and bold version of Arial.} From the
 perspective of Unicode they are \textit{the same thing}. In this approach, a
@@
 Japanese writing system.''} However, Unicode Locales have various drawbacks 
-for the daily practice of scientific linguistic research in a multilingual setting.
+for the daily practice of scientific linguistic research in a multilingual setting (see Section~\ref{characterizing-writing-systems}).

### 2018-07-31 book/chapters/ipa_background.tex
adds backslashes after eg
corrections chapter 4 IPA

@@
 (Section~\ref{need-for-multilingual-environment}); many 
-linguists have encountered the pitfalls when the two are used together 
+linguists have encountered pitfalls when the two are used together 
 (Chapter~\ref{ipa-meets-unicode}).
@@
 Over the years there have been several revisions, but mostly minor ones. Articulation 
-labels -- what are often called \textit{features} even though the IPA
+labels -- what are often called \textit{features}, even though the IPA
 deliberately avoids this term -- have changed, e.g.\ terms like \textit{lips}, \textit{throat}
@@
 This encoding approach is 
-also used in SAMPA and X-SAMPA (Section~\ref{sampa-xsampa}) and in the 
+also used in SAMPA and X-SAMPA (see below) and in the 
 ASJP.\footnote{See the ASJP use case in the online supplementary 
@@
 % "The set of IPA symbols and their numbers were used to draw up an entity set within SGML by the Text Encoding Initiative (TEl).
-The name of each entity is formed by 'IPA' preceding the number, e.g.
-IPA304 is the rEIentity name of lower-case A.
+The name of each entity is formed by 'IPA' preceding the number, e.g.\ IPA304 is the rEIentity name of lower-case A.
 These symbols can be processed as IPA symbols and represented on paper and screen with the appropriate local font by modifying the :entity replacement text.

### 2018-07-31 book/chapters/ipa_meets_unicode.tex
typo in Chapter 5

@@
 to try to be as consistent as possible across a wide range of use cases, and it
-does place linguistic traditions above other orthographic choices. Furthermore,
+does not place linguistic traditions above other orthographic choices. Furthermore,
 when we look at the history of how the IPA met Unicode, we see that many of the

### 2018-08-09 book/chapters/introduction.tex
Post release updates (#44)
* update gitignore

* minor edits in introduction

* small caps in introduction

* minor edits to chapter 2 unicode approach

* corrections chapter 3 unicode pitfalls

* corrections chapter 4 IPA

* layout of tables in chapter 5

* typo in Chapter 5

* update bibliography

* more latex gitignore

* addition to Chapter 7 orthography profiles

* corrections to chapter 8 implementation

* Updated tex and pdf files

* requested changes

* fix alignment on p.107

* change bash explanation

* edits in last paragraph; copy up the cookbook pdf

@@
 in a writing system. A second interpretation is used as a conventional term for a unit in the Chinese writing
-system \citep{Daniels1996}. In technical terminology, a character}
+system \citep{Daniels1996}. In technical terminology, a character
 refers to the electronic encoding of a component in a writing system that has semantic 

### 2018-08-09 book/chapters/orthography_profiles.tex
Post release updates (#44)
* update gitignore

* minor edits in introduction

* small caps in introduction

* minor edits to chapter 2 unicode approach

* corrections chapter 3 unicode pitfalls

* corrections chapter 4 IPA

* layout of tables in chapter 5

* typo in Chapter 5

* update bibliography

* more latex gitignore

* addition to Chapter 7 orthography profiles

* corrections to chapter 8 implementation

* Updated tex and pdf files

* requested changes

* fix alignment on p.107

* change bash explanation

* edits in last paragraph; copy up the cookbook pdf

@@
 	   Section~\ref{pitfall-file-formats} in which we suggest to use NFC, 
-	   no-BOM and LF line breaks because of the pitfalls they avoid. A keen reviewer notes, however, that specifying 
-	   a convention for line endings and BOM is overly strict because most 
+	   no-BOM and LF line breaks because of the pitfalls they avoid. Specifying 
+	   a convention for line endings and BOM is often overly strict because most 
 	   computing environments (now) transparently handle both alternatives. 
@@
 	   an input full in text mode, so that both line feed variants ``LF'' and 
-	   ``CRLF'' will be stripped.}
+	   ``CRLF'' will be stripped. However, note that most shells (e.g. bash) will not 
+	   behave properly with CRLF line endings.}
 
